### 2020-01-22 Inversion_results.Rmd
Many major changes. 1. Best-fit param changed so that not just top 10% of best-fits for each par.sam

---
title: "BCI Water Uptake Depths"
author: "Rutuja Chitra-Tarak"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Rooting profiles used for inversion

```{r Chunk 0.01, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "100%", fig.pos = 'H', fig.cap = ""}
knitr::include_graphics(file.path("figures/rooting_profiles_for_inversion.jpeg"))
```
## All size-classes

```{r Chunk 2.00, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "100%", fig.pos = 'H', fig.cap = ""}
knitr::include_graphics(file.path("figures/UDI/mean/sp_uptake.depth_all_sizes_cor0.3_mean_skIqAp.jpeg"))
knitr::include_graphics(file.path("figures/UDI/mean/violinplot_all_sizes_cor0.3_mean_skIqAp.jpeg"))
```

### Species Uptake Depth Index versus delta Deuterium in Xylem sap water

#### *New Model*
```{r Chunk 0.15, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "100%", fig.pos = 'H', fig.cap = ""}
knitr::include_graphics(file.path("figures/UDI_confidence/mean/Comparison_with_Meinzer1999_deltaD_vs.modelled_uptake.depth_cor0.3_comm_with outliers.jpeg"))
knitr::include_graphics(file.path("figures/UDI_confidence/mean/Comparison_with_Meinzer1999_deltaD_vs.modelled_uptake.depth_cor0.3_comm_without outliers.jpeg"))
knitr::include_graphics(file.path("figures/UDI_confidence/mean/Comparison_with_Meinzer1999_deltaD_vs.modelled_uptake.depth_cor0.3_sp_with outliers.jpeg"))
knitr::include_graphics(file.path("figures/UDI_confidence/mean/Comparison_with_Meinzer1999_deltaD_vs.modelled_uptake.depth_cor0.3_sp_without outliers.jpeg"))
```
#### *Old Model*

```{r Chunk 0.16, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "100%", fig.pos = 'H', fig.cap = ""}
knitr::include_graphics(file.path("figures/UDI_confidence/mean/Old_model_Comparison_with_Meizer1999_isotope_vs.modelled_uptake.depth.tiff"))
```

#### Meinzer et al. 1999: Relation between delta Deuterium in soil water versus Depth

```{r Chunk 0.17, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "100%", fig.pos = 'H', fig.cap = ""}
knitr::include_graphics(file.path("figures/UDI_confidence/Meinzer_etal_1999_Depth_vs_soil_deltaD.jpeg"))
```

#### New Model: Water-Uptake Depth versus Meinzer et al 1999 based water-uptake depth

```{r Chunk 0.18, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "100%", fig.pos = 'H', fig.cap = ""}
knitr::include_graphics(file.path("figures/UDI_confidence/mean/Comparison_with_Meinzer1999_isotopic_vs.modelled_uptake.depth_cor0.3_sp_without outliers.jpeg"))
```

### Species-size growth by mean interval Growth Factor

```{r Chunk 0.10, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "100%", fig.pos = 'H', fig.cap = ""}
knitr::include_graphics(file.path("figures/mean_growth_by_gfac_by_size.jpeg"))
```

### Species-size growth by mean interval Solar Radiation

```{r Chunk 0.101, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "100%", fig.pos = 'H', fig.cap = ""}

knitr::include_graphics(file.path("figures/mean_growth_by_solar_radiation_by_size.jpeg"))
```

### Mean Rsq between
```{r Chunk 0.11, eval = FALSE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "100%", fig.pos = 'H', fig.cap = ""}

knitr::include_graphics(file.path("figures/UDI/mean/UDI_mean.rsq_size_class_cor0.3_2019-10-14_5000_mean_5_rsq_vs_UDI.jpeg"))
```

### Does a species' water-Uptake Depth Index explain its mean mortality rate?

#### Mortality continues after the 1982-83 El-Nino drought

```{r Chunk 3.00, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "100%", fig.pos = 'H', fig.cap = ""}

knitr::include_graphics(file.path("figures/mortality/sp_Mortality_rate_by_period_above1cmDBH_aboveN50.jpeg"))
```

##### Species mean abundance cutoff n = 1
```{r Chunk 3.021, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "100%", fig.pos = 'H', fig.cap = ""}
knitr::include_graphics(file.path("figures/mortality/mean/sp_Mortality_rate_by_period_udi_by_size_aboveN1_tlpcommn.jpeg"))
knitr::include_graphics(file.path("figures/mortality/mean/sp_Mortality_rate_by_period_udi_by_size_aboveN1_tlpsp.jpeg"))
```

##### Species mean abundance cutoff n = 10
```{r Chunk 3.022, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "100%", fig.pos = 'H', fig.cap = ""}
knitr::include_graphics(file.path("figures/mortality/mean/sp_Mortality_rate_by_period_udi_by_size_aboveN10_tlpcommn.jpeg"))
knitr::include_graphics(file.path("figures/mortality/mean/sp_Mortality_rate_by_period_udi_by_size_aboveN10_tlpsp.jpeg"))
```

##### Species mean abundance cutoff n = 50
```{r Chunk 3.023, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "100%", fig.pos = 'H', fig.cap = ""}
knitr::include_graphics(file.path("figures/mortality/mean/sp_Mortality_rate_by_period_udi_by_size_aboveN50_tlpcommn.jpeg"))
knitr::include_graphics(file.path("figures/mortality/mean/sp_Mortality_rate_by_period_udi_by_size_aboveN50_tlpsp.jpeg"))
```

### Species Water-Uptake Depth Index is not correlated with its mean growth rate.

```{r Chunk 3.031, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "100%", fig.pos = 'H', fig.cap = ""}
knitr::include_graphics(file.path("figures/mortality/mean/sp_mean_Growth_vs_udi_size_tlpcomm.jpeg"))
```

### Species growth vs mortality rate

```{r Chunk 3.04, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "100%", fig.pos = 'H', fig.cap = ""}
knitr::include_graphics(file.path("figures/mortality/mean/sp_mean_Growth_vs_mrate.jpeg"))
```

### Is Water-Uptake Depth Index related to other hydraulic traits?
#### TLP at Common/Community level
#### TLPlevel = comm, *Tiny*
```{r Chunk 4.01, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "100%", fig.pos = 'H', fig.cap = ""}
knitr::include_graphics(file.path("figures/traits/mean/pca/sp_UDI_traits_democomm_tiny.tiff"))
```
#### TLPlevel = comm, *Small*
```{r Chunk 4.02, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "100%", fig.pos = 'H', fig.cap = ""}
knitr::include_graphics(file.path("figures/traits/mean/pca/sp_UDI_traits_democomm_small.tiff"))
```
#### TLPlevel = comm, *Medium*
```{r Chunk 4.03, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "100%", fig.pos = 'H', fig.cap = ""}
knitr::include_graphics(file.path("figures/traits/mean/pca/sp_UDI_traits_democomm_medium.tiff"))
```
#### TLPlevel = comm, *Large*
```{r Chunk 4.04, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "100%", fig.pos = 'H', fig.cap = ""}
knitr::include_graphics(file.path("figures/traits/mean/pca/sp_UDI_traits_democomm_large.tiff"))
```

#### TLP at Species level
#### TLPlevel = sp, *Tiny*
```{r Chunk 5.01, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "100%", fig.pos = 'H', fig.cap = ""}
knitr::include_graphics(file.path("figures/traits/mean/pca/sp_UDI_traits_demosp_tiny.tiff"))
```
#### TLPlevel = sp, *Small*
```{r Chunk 5.02, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "100%", fig.pos = 'H', fig.cap = ""}
knitr::include_graphics(file.path("figures/traits/mean/pca/sp_UDI_traits_demosp_small.tiff"))
```
#### TLPlevel = sp, *Medium*
```{r Chunk 5.03, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "100%", fig.pos = 'H', fig.cap = ""}
knitr::include_graphics(file.path("figures/traits/mean/pca/sp_UDI_traits_demosp_medium.tiff"))
```
#### TLPlevel = sp, *Large*
```{r Chunk 5.04, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "100%", fig.pos = 'H', fig.cap = ""}
knitr::include_graphics(file.path("figures/traits/mean/pca/sp_UDI_traits_demosp_large.tiff"))
```

#### TLP at Common/Community level
```{r Chunk 4.11, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "100%", fig.pos = 'H', fig.cap = ""}
knitr::include_graphics(file.path("figures/traits/mean/sp_UDI_vs_traits_corrplot_tlpcomm_tiny_demo.jpeg"))
knitr::include_graphics(file.path("figures/traits/mean/sp_UDI_vs_traits_corrplot_tlpcomm_small_demo.jpeg"))
knitr::include_graphics(file.path("figures/traits/mean/sp_UDI_vs_traits_corrplot_tlpcomm_medium_demo.jpeg"))
knitr::include_graphics(file.path("figures/traits/mean/sp_UDI_vs_traits_corrplot_tlpcomm_large_demo.jpeg"))
```

#### TLP at Species level
```{r Chunk 5.11, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "100%", fig.pos = 'H', fig.cap = ""}
knitr::include_graphics(file.path("figures/traits/mean/sp_UDI_vs_traits_corrplot_tlpsp_tiny_demo.jpeg"))
knitr::include_graphics(file.path("figures/traits/mean/sp_UDI_vs_traits_corrplot_tlpsp_small_demo.jpeg"))
knitr::include_graphics(file.path("figures/traits/mean/sp_UDI_vs_traits_corrplot_tlpsp_medium_demo.jpeg"))
knitr::include_graphics(file.path("figures/traits/mean/sp_UDI_vs_traits_corrplot_tlpsp_large_demo.jpeg"))
```

### 2020-01-22 Workflow.Rmd
Many major changes. 1. Best-fit param changed so that not just top 10% of best-fits for each par.sam

---
title: "Workflow of Inverse Model for Rooting Profiles"
author: "Rutuja"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, include = TRUE)
```

## How is this project set-up

##### Load swp or gfac data from 
## Then using a combination of a range of hypothetical root profiles obtained from 2.rooting_profiles.R and SWP ensemble simulations generate a btran time-series using equation : Btran= i=1 to i=z RootFrac_z * f(swp_z),
where f(swp_z) is a growth.factor_z based on a linear step relationship with swp_z defined based on maximum value in BCI tlp data from Kunert's.
Code/6.0_Growth_correlate_table_from_swp.R

```{r Chunk 1.0, eval = FALSE}
source("R/1.GLUEsetup_part1_BCI.R")
GLUEsetup_part1(current.folder = "2019-10-14_5000", intervals = 5) # this can take 30 min to get btran by nsam
```

```{r Chunk 2.2.0, eval = FALSE}
source("R/4.1.GLUEsetup_part2_BCI.R")
GLUEsetup_part2(growth.type = "med", dbh.residuals = "on", solar.residuals = "off", 
                growth.selection = "size_class_predefined_cc_scaled") #"size_class_varying_non_cc") #n.ensembles = 100000
```

```{r Chunk 2.3.0, eval = TRUE}
# This sources "R/4.2GLUErun_part2_func_BCI.R"
source("R/4.3.GLUErun_part2_BCI.R")
GLUE_run(splevel = "off", goodness.fit = 0.3)
GLUE_run(splevel = "on", goodness.fit = 0.3)
```

```{r Chunk 2.4.0, eval = TRUE}
source("R/4.4.Best-fit_param_BCI.R")
bestfit.param(splevel = "on")
bestfit.param(splevel = "off")
```

```{r Chunk 2.4.1, eval = TRUE}
source("R/4.5.prep.udi.calculator.R")
prep.udi.calculator(splevel = "on")
prep.udi.calculator(splevel = "off")
```

```{r Chunk 2.4.2, eval = TRUE}
source("R/4.6.udi.calculator.R")
udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.5) ## udi is not calculated below this rsq.thresh
udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.5)
```

```{r Chunk 2.4.4, eval = TRUE}
source("R/4.7.summarising_udi_by_sp_size.R")
summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.7)
summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.7)
```

```{r Chunk 2.5.0, eval = TRUE}
source("R/10.Comparison_with_Meinzer1999.R")
```

```{r Chunk 2.5.1, eval = FALSE}
source("R/5.species_graphs_BCI.R")
source("R/9.5mortality_BCI.R")
source("R/13.hydraulic_traits_BCI.R")
```

### 2020-01-22 R/1.GLUEsetup_part1_BCI.R

@@ -1,0 +1,245 @@
+# This file imports the data and sets up the 'Extra' variable
+
+rm(list=ls())
+gc()
+graphics.off()
+if (!require("pacman")) install.packages("pacman"); library(pacman)
+pacman::p_load(tidyverse, doParallel, foreach, data.table)
+
+GLUEsetup_part1 <- function(current.folder = "2019-10-14_5000", intervals = 5) {
+  if(!dir.exists(file.path("results", current.folder))) {dir.create(file.path("results", current.folder))}
+
+  ####-------------------------------------
+  #### 1. To enable hydro time series into intervals
+  ####-------------------------------------
+  ##
+  ## Storing census med and interval start and end dates--------------
+  # load("data-raw/CTFScensuses/bci.tree1.Rdata")
+  # load("data-raw/CTFScensuses/bci.tree2.Rdata")
+  # load("data-raw/CTFScensuses/bci.tree3.Rdata")
+  # load("data-raw/CTFScensuses/bci.tree4.Rdata")
+  # load("data-raw/CTFScensuses/bci.tree5.Rdata")
+  # load("data-raw/CTFScensuses/bci.tree6.Rdata")
+  # load("data-raw/CTFScensuses/bci.tree7.Rdata")
+  # load("data-raw/CTFScensuses/bci.tree8.Rdata")
+  # head(bci.tree1)
+  #
+  # census.years <- c(1982, 1985, 1990, 1995, 2000, 2005, 2010, 2015)
+  # bci.tree1$census <- census.years[1]
+  # bci.tree2$census <- census.years[2]
+  # bci.tree3$census <- census.years[3]
+  # bci.tree4$census <- census.years[4]
+  # bci.tree5$census <- census.years[5]
+  # bci.tree6$census <- census.years[6]
+  # bci.tree7$census <- census.years[7]
+  # bci.tree8$census <- census.years[8]
+  #
+  # bci.tree <- rbind.data.frame(bci.tree1, bci.tree2, bci.tree3, bci.tree4, bci.tree5, bci.tree6, bci.tree7, bci.tree8)
+  # bci.tree$ExactDate <- as.Date(bci.tree$ExactDate)
+  # ggplot(bci.tree, aes(x = ExactDate)) +
+  #   geom_histogram() +
+  #   facet_wrap(~ census)
+  # library(dplyr)
+  # mediandates <- summarise(bci.tree %>% group_by(census), mediandate = median(ExactDate, na.rm = T))
[…]

### 2020-01-22 R/10.Comparison_with_Meinzer1999.R

@@ -1,0 +1,357 @@
+## Comparison with Meinzer et al. 1999 results.
+
+#
+rm(list=ls())
+gc()
+library(tidyverse)
+# graphics info
+theme_set(theme_bw())
+theme_update(text = element_text(size = 14),
+             panel.grid.major = element_blank(),
+             panel.grid.minor = element_blank(),
+             strip.background = element_blank()
+)
+require(scales)
+rev_sqrt_trans <- function() {
+  scales::trans_new(
+    name = "rev_sqrt",
+    transform = function(x) -sqrt(abs(x)),
+    inverse = function(x) x^2);
+}
+
+# load interval and working.iter
+load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
+load(file.path("results/4.1GLUEsetup_part2_BCI.RData")) # has working.iter and growth and si matrix
+
+intervals <- info$intervals
+n.ensembles <- growth_by_si.info$n.ensembles
+growth.type <- growth_by_si.info$growth.type
+growth.selection <- growth_by_si.info$growth.selection
+dbh.residuals <- growth_by_si.info$dbh.residuals
+si.type <- growth_by_si.info$si.type
+goodness.fit <- 0.3
+soil.depths <- unique(info$root.param.long$depth)
+dryseason <- "on"
+##
+load(file = paste("results/splevel/ds.bestfit_cor", goodness.fit, "_", si.type, "_", n.ensembles, "_", growth.type, "_", growth.selection, "_", dbh.residuals, "_", intervals, "_id_dryseason_", dryseason, ".Rdata", sep = ""))
+ds <- ds.bestfit
+load(file = paste("results/commlevel/ds.bestfit_cor", goodness.fit, "_", si.type, "_", n.ensembles, "_", growth.type, "_", growth.selection, "_", dbh.residuals, "_", intervals, "_id_dryseason_", dryseason, ".Rdata", sep = ""))
+ds <- rbind(ds, ds.bestfit) %>% mutate(tlplevel = as.factor(tlplevel)) %>% subset(!is.na(udi)) %>% droplevels()
+
+iso <- read.csv("data-raw/traits/isotopes/Meinzer1999_Table1.csv", na.strings = c("NA",""), header = T, row.names = NULL, check.names = F)
+
+head(iso)
+colnames(iso)[1] <- "species"
[…]

### 2020-01-22 R/13.hydraulic_traits_BCI.R

@@ -1,0 +1,180 @@
+##-------------------------
+## Date : 22 Jul 2018
+## Author : Rutuja
+## Title : Uptake depth and hydrualic traits
+##-------------------------
+## Do species with shallow uptake depth have greater drought tolerance? (more -ve TLP)
+
+rm(list = ls())
+gc()
+if (!require("pacman")) install.packages("pacman"); library(pacman)
+pacman::p_load(tidyverse, ggcorrplot, Hmisc, PerformanceAnalytics, corrplot, devtools)
+install_github("vqv/ggbiplot")
+  # graphics info
+theme_set(theme_bw())
+theme_update(text = element_text(size=14),
+             panel.grid.major = element_blank(),
+             panel.grid.minor = element_blank(),
+             strip.background = element_blank()
+)
+## adding uptake depth index
+load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
+load("results/4.1GLUEsetup_part2_BCI.RData") # has n.ensembles and growth and si matrix
+load("results/demo.sp_size.RData")
+intervals <- info$intervals
+n.ensembles <- growth_by_si.info$n.ensembles
+growth.type <- growth_by_si.info$growth.type
+si.type <- growth_by_si.info$si.type
+goodness.fit <- 0.3
+soil.depths <- unique(info$root.param.long$depth)
+##
+load(file = paste("results/splevel/ds.bestfit_cor", goodness.fit, "_", si.type, "_", n.ensembles, "_", growth.type, "_", intervals, "_id.Rdata", sep = ""))
+ds <- ds.bestfit
+load(file = paste("results/commlevel/ds.bestfit_cor", goodness.fit, "_", si.type, "_", n.ensembles, "_", growth.type, "_", intervals, "_id.Rdata", sep = ""))
+ds <- rbind(ds, ds.bestfit) %>% mutate(tlplevel = as.factor(tlplevel)) %>%
+  subset(!is.na(udi)) %>%
+  mutate(size = factor(size, levels = c("tiny", "small", "medium", "large"))) %>%
+  droplevels()
+hyd <- read.csv("data-raw/traits/ht1_20200103.csv") #  # Brett's data
+hyd <- hyd %>% select(-genus, -species, -deciduousness, -site) %>% mutate(sp = tolower(sp))
[…]

### 2020-01-22 R/2.rooting_profiles.R

@@ -1,0 +1,182 @@
+## Estimating parameter b (fates_rootb_par) range for rooting profiles
+
+
+rm(list=ls())
+
+if (!require("pacman")) install.packages("pacman"); library(pacman)
+pacman::p_load(ncdf4, easyNCDF, tidyverse)
+
+require(scales);
+rev_sqrt_trans <- function() {
+  scales::trans_new(
+    name = "rev_sqrt",
+    transform = function(x) -sqrt(abs(x)),
+    inverse = function(x) x^2);
+}
+theme_set(theme_bw())
+
+####--------------------------------------------
+#### Rooting profiles for inversion
+####--------------------------------------------
+
+load(file = file.path("data-raw/psi.rda"))
+
+soil.depths <- unique(psi$depth)
+# max.rds.df <- data.frame(max.rds = soil.depths[-length(soil.depths)]) %>%
+#   mutate(lag.max.rds = lag(max.rds, default = 0),
+#          diffby2 = (max.rds - lag.max.rds)/2,
+#          half.depths = max.rds - diffby2)
+# max.rds <- c(max.rds.df$max.rds, max.rds.df$half.depths)
+by.param = 0.1
+exponents1plus <- data.frame(x = seq(-6, 6, by = by.param)) %>% mutate(y = exp(x))
+plot(data = exponents1plus, y~x)
+# exponents.full <- rbind(exponents01, exponents1plus)
+# plot(data = exponents.full, y~x)
+max.rds <- soil.depths
+
+profiles.list <- vector("list", length(max.rds))
+profiles.dfs <- vector("list", length(max.rds))
+
+exponents <- exponents1plus$y #seq(0.01, 30, by = 0.1) #c(seq(0.1, 1, length.out = 10), seq(1.2, 2.6, by = 0.2), seq(3, 15, length.out = 20))
+nsam <- length(max.rds)*length(exponents)
+## each rf.sam is associated with one maxD & power combination
+for (i in 1: length(max.rds)) {
+  maxD <- max.rds[i]
+  profiles.list[[i]] <- vector("list", length(exponents))
+  for (j in 1:length(exponents)) {
+    rf.sam <- (i-1)*length(exponents) + j
+    power <- exponents[j]
+    profiles.list[[i]][[j]] <- data.frame(rf.sam = rf.sam, maxD = maxD, power = power,
+                                     base = seq(0, 1, length.out = 100)) %>%
+      mutate(abs.depth = base*maxD,
+             cum.root.frac = base^power,
[…]

### 2020-01-22 R/4.0.Species_growth_BCI_Condit_data.R

@@ -1,0 +1,864 @@
+# for 50 ha obs species groups
+rm(list = ls())
+source(file = "R/utilities.R")
+if (!require("pacman")) install.packages("pacman"); library(pacman)
+pacman::p_load(tidyverse, purrrlyr, scales, grid, gridExtra)
+# graphics info
+tex <- element_text(size = 16, face = "plain") # , family = "gara"
+tex <- element_text(size = 5, face = "plain")
+my.theme <-  theme(axis.text = tex, axis.title = tex,
+                   title = tex, legend.title = tex, legend.text = tex, strip.text.y = tex, strip.text.x = tex)
+my.bg <- theme_bw() + theme(panel.grid.major.y = element_blank(), panel.grid.major.x = element_line(colour = "black", size = 0.25),
+                            panel.grid.minor = element_blank())
+my.adjust <- theme(axis.title.y = element_text(vjust = 1), axis.title.x = element_text(vjust = -0.6), title = element_text(vjust = 2))
+
+# time series info
+years <- c(1982, seq(1985, 2015, by = 5))
+ncensus  <- length(years)
+nint <- length(years) - 1
+# sp4 <- c("ANOL", "TECG", "LAGL", "TERT")
+##------------------------------------------------
+## Collating tree.full data for all censuses
+##------------------------------------------------
+## All data frames have same number of rows. Each row represent a tree,
+## and a given row number across censuses represent the same tree
+load("data-raw/CTFScensuses/BCI.tree1.Rdata")
+load("data-raw/CTFScensuses/BCI.tree2.Rdata")
+load("data-raw/CTFScensuses/BCI.tree3.Rdata")
+load("data-raw/CTFScensuses/BCI.tree4.Rdata")
+load("data-raw/CTFScensuses/BCI.tree5.Rdata")
+load("data-raw/CTFScensuses/BCI.tree6.Rdata")
+load("data-raw/CTFScensuses/BCI.tree7.Rdata")
+load("data-raw/CTFScensuses/BCI.tree8.Rdata")
+
+length(unique(bci.tree1$sp))
+# all censuses in one list
+tree.full <- bind_rows(bci.tree2, bci.tree3, bci.tree4, bci.tree5, bci.tree6, bci.tree7, bci.tree8)
+# > nrow(bci.tree.full1)
+# [1] 423617
+# > nrow(bci.tree.full1)*8
+# [1] 3388936
+###------------selection----------
[…]

### 2020-01-22 R/4.1.GLUEsetup_part2_BCI.R

@@ -1,0 +1,85 @@
+## Best-fit_param_BCI_fun
+## load btran
+rm(list = ls())
+gc()
+graphics.off()
+if (!require("pacman")) install.packages("pacman"); library(pacman)
+pacman::p_load(tidyverse)
+
+GLUEsetup_part2 <- function(growth.type, dbh.residuals, solar.residuals, growth.selection) { #n.ensembles = 5000
+  # n.ensembles = 100000,  #which ensemble set to use
+    #growth.selection  <- "size_class_predefined_cc_scaled"
+    load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
+    intervals <- info$intervals
+    ## load Result of 3.GLUErun_part1_BCI.R:
+    btran <- info$btran
+    n.ensembles <- length(btran)
+    ## Observed growth
+    ## individual, species or sp by size level
+    growth.name <- load(file =  paste0("results/sp_size.", growth.type, "_growth_dbh.residuals_", dbh.residuals, "_", intervals, "_", growth.selection, ".Rdata"))
+    growth <- get(growth.name); rm(growth.name)
+    sp_size.name <- load(file = paste0("results/sp_size.", growth.type, ".names_", intervals, "_", growth.selection, ".Rdata"))
+    growth.meta <- data.frame(sp_size = get(sp_size.name)) %>%
+      separate(sp_size, into = c("sp", "size"), sep = "_", remove = FALSE)
+    rm(sp_size.name)
+
+    # growth.type <- "med_growth_sp_by_size"
+    # load(paste0("results/", growth.type, "_intervals_", intervals ,".Rdata"))
+    # growth.long <- med_growth_sp_by_size %>% unite("sp_size", sp, size, remove = TRUE) %>% select(-interval.2) %>%
+    #   pivot_longer(cols = starts_with("interval."), names_to = "interval", names_prefix = "interval.", values_to = "growth") %>% mutate(interval = as.numeric(interval))
+    # growth.prep <- split(growth.long, growth.long$sp_size)
+    # growth <- lapply(growth.prep, function(x) {x %>% select(-sp_size)})
+    #
+    # sp_size.median.names <- names(growth)
+    # growth.meta <- data.frame(sp_size = sp_size.median.names) %>%
+    #   separate(sp_size, into = c("sp", "size"), sep = "_", remove = FALSE)
+
+
+    # growth.type <- "med_growth_sp_by_size"
+    ## species or size level
[…]

### 2020-01-22 R/4.1_Preparing_stem_growth_time_series.R

@@ -1,0 +1,685 @@
+
+# for 50 ha obs species groups
+rm(list = ls())
+gc()
+if (!require("pacman")) install.packages("pacman"); library(pacman)
+pacman::p_load(tidyverse, purrrlyr, scales, grid, gridExtra, lme4)
+
+# graphics info
+theme_set(theme_bw())
+theme_update(text = element_text(size = 14),
+             panel.grid.major = element_blank(),
+             panel.grid.minor = element_blank(),
+             strip.background = element_blank()
+)
+# time series info
+years <- seq(1990, 2015, by = 5)
+ncensus  <- length(years)
+nint <- length(years) - 1
+intervals <- nint
+# sp4 <- c("ANOL", "TECG", "LAGL", "TERT")
+##------------------------------------------------
+## Collating tree.full data for all censuses
+##------------------------------------------------
+# https://repository.si.edu/handle/10088/20925
+# load("data-raw/CTFScensuses/BCI.stem1.Rdata") # 1982-1985
+# load("data-raw/CTFScensuses/BCI.stem2.Rdata") # 1985-1990
+load("data-raw/CTFScensuses/BCI.stem3.Rdata")
+load("data-raw/CTFScensuses/BCI.stem4.Rdata")
+load("data-raw/CTFScensuses/BCI.stem5.Rdata")
+load("data-raw/CTFScensuses/BCI.stem6.Rdata")
+load("data-raw/CTFScensuses/BCI.stem7.Rdata")
+load("data-raw/CTFScensuses/BCI.stem8.Rdata")
+
+length(unique(bci.stem3$sp))
+# all censuses in one list
+stem.full <- bind_rows(bci.stem3, bci.stem4, bci.stem5, bci.stem6, bci.stem7, bci.stem8)
+# > nrow(stem.full)
+# [1] 4740360
+# > nrow(stem.full)*6
+# [1] 28442160
+###------------selection----------
+census <- 3:8
+###-------------------------------
+stem.full$census <- rep(census, each = nrow(bci.stem3))
+## remove duplicated tags if any
+nrow(stem.full) # 4740360
+stem.full <- stem.full %>% mutate(ExactDate = as.Date(ExactDate))
+stem.full <- stem.full %>% group_by(census) %>% distinct(stemID, .keep_all= TRUE)
+nrow(stem.full)
+# 4740360
+# so no duplicate stemIDs;
[…]

### 2020-01-22 R/4.2.GLUErun_part2_func_BCI.R

@@ -1,0 +1,75 @@
+## function to calculate rsq between observed growth vs. simulated SI matrix
+growth_by_si.func <- function(growth_by_si.info, goodness.fit, statistic, splevel){
+
+  if (splevel == "on"){
+    sp.with.tlp <- names(info.2$sp.si)
+    growth.meta <- growth_by_si.info$growth.meta %>%
+      subset(sp %in% sp.with.tlp)
+    sp_size.on  <- growth.meta$sp_size #.full %>% subset(size == "large" || size == "small") %>% select(-size, -sp, -interval.2) %>% as.matrix()
+    growth <- growth_by_si.info$growth[sp_size.on]
+    sp_size <- names(growth)
+    ### to create si for each sp_size, sp.si elements should be repeated as sps in growth
+    sp.in.sp_size <- stringr::str_split_fixed(names(growth), "_", 2)[,1]
+    sp.si.match <- info.2$sp.si[sp.in.sp_size] ## this repeats sp.si elements (with sp names) in the order of sp.in.sp_size
+    ## need to repeat elements of si according to growth[[j]]$interval
+    ## since intervals are from 3:7 (8-intervals), and so are corresponding si vec, this would mean removing 3-1 = 2
+    location.adj <- as.numeric(strsplit(colnames(sp.si.match[[1]]),".", fixed = TRUE)[[1]][2]) - 1
+    sp.si.match <- lapply(sp.si.match, unname)
+  } else {
+    growth <- growth_by_si.info$growth
+    sp_size <- names(growth)
+    location.adj <- as.numeric(strsplit(colnames(growth_by_si.info$si),".", fixed = TRUE)[[1]][2]) - 1
+    si <- unname(growth_by_si.info$si)
+  }
+  if (growth_by_si.info$dbh.residuals == "on") {
+    growth <- lapply(growth, function (x) {
+      x %>% rename(growth = dbh.residuals)
+    })
+  }
+  n.rf.sam <- length(unique(growth_by_si.info$si.param.rel$rf.sam))
+  n.par.sam <- length(unique(growth_by_si.info$si.param.rel$par.sam))
+  intervals <- info$intervals
+
+  ncor <- detectCores() - 1
+  cl <- parallel::makeForkCluster(ncor)
+  doParallel::registerDoParallel(cl)
+
+  GLUE <- vector("list", length(sp_size))
+  Sys.time()
[…]

### 2020-01-22 R/4.4.Best-fit_param_BCI.R

@@ -1,0 +1,81 @@
+## Not anymore: Finding best fit water-stress/growth model parameters for species growth time series
+## Just vectorises Rsquares with decreasing order and notes down the order to relate to par.sam and rf.sam from growth_by_si.info$si.param.rel
+
+# for 50 ha obs species groups
+rm(list = ls())
+gc()
+if (!require("pacman")) install.packages("pacman"); library(pacman)
+pacman::p_load(tidyverse, doParallel, foreach, data.table)
+
+bestfit.param <- function(splevel) {
+  # load interval and n.ensembles
+  load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
+  load(file.path("results/4.1GLUEsetup_part2_BCI.RData")) # has n.ensembles and growth and si matrix
+
+  intervals <- info$intervals
+  n.ensembles <- growth_by_si.info$n.ensembles
+  growth.type <- growth_by_si.info$growth.type ## this is not data but type such as "individual"
+  growth.selection <- growth_by_si.info$growth.selection
+  si.type <- growth_by_si.info$si.type
+  dbh.residuals <- growth_by_si.info$dbh.residuals
+  goodness.fit <- 0.3 # rsq0.3
+  if (splevel == "on") {
+    level.folder <- "splevel"
+  } else {
+    level.folder <- "commlevel"
+  }
+  ncor <- detectCores() - 2
+
+  if (splevel == "on") {
+    load(paste0("results/sp.GLUE.rsq_", si.type, "_", n.ensembles, "_",
+              growth.type, "_", growth.selection, "_", dbh.residuals, "_", intervals, "_cor", goodness.fit, ".Rdata"))
+    GLUE.rsq <- sp.GLUE.rsq
+    rm(sp.GLUE.rsq)
+  } else {
+    load(paste0("results/GLUE.rsq_", si.type, "_", n.ensembles, "_",
+                growth.type, "_", growth.selection, "_", dbh.residuals, "_", intervals, "_cor", goodness.fit, ".Rdata"))
+  }
+  sp_size <- names(GLUE.rsq)
+  sp <- stringr::str_split(sp_size, "_", simplify = TRUE)[, 1]
+  size <- stringr::str_split(sp_size, "_", simplify = TRUE)[, 2]
[…]

### 2020-01-22 R/4.6.udi.calculator.R

@@ -1,0 +1,235 @@
+## Finding best fit water-stress/growth model parameters for species growth time series
+
+# .rs.restartR()
+# for 50 ha obs species groups
+rm(list = ls())
+for (i in 1:5) {gc()}
+
+if (!require("pacman")) install.packages("pacman"); library(pacman)
+pacman::p_load(tidyverse, doParallel, foreach, data.table)
+
+udi.calculator <- function(splevel, dryseason,  rsq.thresh) {
+  # load interval and n.ensembles
+  load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
+  load(file.path("results/4.1GLUEsetup_part2_BCI.RData")) # has n.ensembles and growth and si matrix
+
+  intervals <- info$intervals
+  n.ensembles <- growth_by_si.info$n.ensembles
+  growth.type <- growth_by_si.info$growth.type ## this is not data but type such as "individual"
+  growth.selection <- growth_by_si.info$growth.selection
+  dbh.residuals <- growth_by_si.info$dbh.residuals
+  si.type <- growth_by_si.info$si.type
+  si.param.rel <- growth_by_si.info$si.param.rel
+  goodness.fit <- 0.3 # rsq0.3
+  ncor <- detectCores() - 2
+
+  if (splevel == "on") {
+    level.folder <- "splevel"
+    } else {
+    level.folder <- "commlevel"
+    }
+
+  rf.sam <- unique(si.param.rel$rf.sam)
+  n.par.sam <- length(unique(si.param.rel$par.sam)) ## this is the sequence repeated in growth_by_si.info$si.param.rel
+  par.sam <- si.param.rel$par.sam[1:n.par.sam] ## this is the sequence used when GLUE.rsq was constructed
+  n.rf.sam <- length(rf.sam)
+  rm(info); rm(growth_by_si.info)
+  ## sdt is chosen based on dryseason on/off
+  load(file = paste0("results/", level.folder, "/GLUE.rsq.vector_cor", goodness.fit, "_", si.type, "_", n.ensembles, "_", growth.type, "_", growth.selection, "_", dbh.residuals, "_", intervals, ".Rdata"), envir = parent.frame(), verbose = FALSE)
+  sp_size <- row.names(GLUE.rsq.vector)
+  sp <- stringr::str_split(sp_size, "_", simplify = TRUE)[, 1]
[…]

### 2020-01-22 R/4.6.udi.calculator_best10_per_par.sam.R

@@ -1,0 +1,230 @@
+## Finding best fit water-stress/growth model parameters for species growth time series
+
+# .rs.restartR()
+# for 50 ha obs species groups
+rm(list = ls())
+for (i in 1:5) {gc()}
+
+if (!require("pacman")) install.packages("pacman"); library(pacman)
+pacman::p_load(tidyverse, doParallel, foreach, data.table)
+
+udi.calculator <- function(splevel, dryseason) {
+  # load interval and n.ensembles
+  load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
+  load(file.path("results/4.1GLUEsetup_part2_BCI.RData")) # has n.ensembles and growth and si matrix
+
+  intervals <- info$intervals
+  n.ensembles <- growth_by_si.info$n.ensembles
+  growth.type <- growth_by_si.info$growth.type ## this is not data but type such as "individual"
+  growth.selection <- growth_by_si.info$growth.selection
+  dbh.residuals <- growth_by_si.info$dbh.residuals
+  si.type <- growth_by_si.info$si.type
+  goodness.fit <- 0.3 # rsq0.3
+  n.best <- 10
+  ncor <- detectCores() - 2
+
+  # growth_by_si.info$growth only has growth matrix for given info$intervals
+  # to get corresponding species-size, use growth meta
+
+  n.par.sam <- length(unique(growth_by_si.info$si.param.rel$par.sam))
+  par.sam <- growth_by_si.info$si.param.rel$par.sam[1:n.par.sam] ## this is the sequence used when GLUE.rsq was constructed
+  rm(growth_by_si.info)
+  ## sdt is chosen based on dryseason on/off
+  if (splevel == "on") {
+    load(file = paste("results/splevel/best10.type_cor", goodness.fit, "_", si.type, "_", n.ensembles, "_", growth.type, "_", growth.selection, "_", dbh.residuals, "_", intervals, ".Rdata", sep = ""), envir = parent.frame(), verbose = FALSE)
+    load(file = paste("results/splevel/sdt.list_cor", goodness.fit, "_", si.type, "_", n.ensembles, "_", growth.type, "_", growth.selection, "_", dbh.residuals, "_", intervals, "_dryseason_", dryseason, ".Rdata", sep = ""), envir = parent.frame(), verbose = FALSE)
+  } else {
[…]

### 2020-01-22 R/4.7.summarising_udi_by_sp_size.R

@@ -1,0 +1,74 @@
+rm(list = ls())
+gc()
+if (!require("pacman")) install.packages("pacman"); library(pacman)
+pacman::p_load(tidyverse, data.table)
+
+summarise.udi <- function(splevel, dryseason, rsq.thresh) {
+  load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
+  load(file.path("results/4.1GLUEsetup_part2_BCI.RData")) # has n.ensembles and growth and si matrix
+
+  intervals <- info$intervals
+  n.ensembles <- growth_by_si.info$n.ensembles
+  growth.type <- growth_by_si.info$growth.type ## this is not data but type such as "individual"
+  si.type <- growth_by_si.info$si.type
+  growth.selection <- growth_by_si.info$growth.selection
+  dbh.residuals <- growth_by_si.info$dbh.residuals
+  goodness.fit <- 0.3 # rsq0.3
+  rsq.thresh <-  rsq.thresh
+  n.best.fit <- 100
+  if (splevel == "on") {
+    level.folder <- "splevel"
+  } else {
+    level.folder <- "commlevel"
+  }
+
+  load(file = paste("results/", level.folder, "/ds.bestfit.all_cor", goodness.fit, "_", si.type, "_", n.ensembles, "_", growth.type, "_", growth.selection, "_", dbh.residuals, "_", intervals, "_id_dryseason_", dryseason, ".Rdata", sep = ""))
+  ds.bestfit <- ds.bestfit.all %>% group_by(sp_size) %>%
+    arrange(sp_size, rsq) %>%
+    # limiting udi to the cases that are top 100
+    mutate(sdi.test = ifelse(rsq < rsq.thresh, NA, sdi),
+           udi.test = ifelse(rsq < rsq.thresh, NA, udi),
+           rsq.test = ifelse(rsq < rsq.thresh, NA, rsq),
+           udi.best = ifelse(rsq == max(rsq, na.rm = TRUE), udi, NA),
+           sdi.best = ifelse(rsq == max(rsq, na.rm = TRUE), sdi, NA)) %>%
+    summarise(udi.best = mean(udi.best, na.rm = TRUE),
+              udi = median(udi, na.rm = TRUE), #sum(udi.test*rsq.test, na.rm = TRUE)/sum(rsq.test, na.rm = TRUE),
+              udi.sd = sd(udi.test, na.rm = TRUE),
+              n = n(),
+              udi.se = udi.sd/(n^0.5),
[…]

### 2020-01-22 R/5.species_graphs_BCI.R

@@ -1,0 +1,259 @@
+##
+rm(list = ls())
+gc()
+if (!require("pacman")) install.packages("pacman"); library(pacman)
+pacman::p_load(tidyverse, scales)
+
+# graphics info
+theme_set(theme_bw())
+theme_update(text = element_text(size = 14),
+             panel.grid.major = element_blank(),
+             panel.grid.minor = element_blank(),
+             strip.background = element_blank()
+)
+
+# load interval and n.ensembles
+load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
+load(file.path("results/4.1GLUEsetup_part2_BCI.RData")) # has n.ensembles and growth and si matrix
+
+intervals <- info$intervals
+n.ensembles <- growth_by_si.info$n.ensembles
+growth.type <- growth_by_si.info$growth.type
+growth.selection <- growth_by_si.info$growth.selection
+dbh.residuals <- growth_by_si.info$dbh.residuals
+si.type <- growth_by_si.info$si.type
+runs <- info$root.param
+goodness.fit <- 0.3
+dryseason = "on"
+soil.depths <- unique(info$root.param.long$depth)
+if(!dir.exists(file.path("figures/UDI", growth.type))) {dir.create(file.path("figures/UDI", growth.type))}
+
+##
+load(file = paste("results/splevel/ds.bestfit_cor", goodness.fit, "_", si.type, "_", n.ensembles, "_", growth.type, "_", growth.selection, "_", dbh.residuals, "_", intervals, "_id_dryseason_", dryseason, ".Rdata", sep = ""))
+load(file = paste("results/splevel/ds.bestfit.longer_cor", goodness.fit, "_", si.type, "_", n.ensembles, "_", growth.type, "_", growth.selection, "_", dbh.residuals, "_", intervals, "_id_dryseason_", dryseason, ".Rdata", sep = ""))
+ds <- ds.bestfit
+dss <- ds.bestfit.longer
+load(file = paste("results/commlevel/ds.bestfit_cor", goodness.fit, "_", si.type, "_", n.ensembles, "_", growth.type, "_", growth.selection, "_", dbh.residuals, "_", intervals, "_id_dryseason_", dryseason, ".Rdata", sep = ""))
[…]

### 2020-01-22 R/9.5mortality_BCI.R

@@ -1,0 +1,294 @@
+rm(list=ls())
+gc()
+# load("/Library/Frameworks/R.framework/Versions/3.4/Resources/library/CTFSRPackage/CTFSRPackage.Rdata")
+if (!require("pacman")) install.packages("pacman"); library(pacman)
+pacman::p_load(tidyverse, hms)
+
+# graphics info
+tex <- element_text(size = 16, face = "plain") # , family = "gara"
+tex <- element_text(size = 5, face = "plain")
+my.theme <-  theme(axis.text = tex, axis.title = tex,
+                   title = tex, legend.title = tex, legend.text = tex, strip.text.y = tex, strip.text.x = tex)
+my.bg <- theme_bw() + theme(panel.grid.major.y = element_blank(), panel.grid.major.x = element_line(colour = "black", size = 0.25),
+                            panel.grid.minor = element_blank())
+my.adjust <- theme(axis.title.y = element_text(vjust = 1), axis.title.x = element_text(vjust = -0.6), title = element_text(vjust = 2))
+theme_set(theme_bw())
+theme_update(text = element_text(size=14),
+             panel.grid.major = element_blank(),
+             panel.grid.minor = element_blank(),
+             strip.background = element_blank()
+)
+## load full tree data
+## only single stem present for a tree, usually the largest in dbh, status refers to the entire tree
+load("data-raw/CTFScensuses/bci.tree1.Rdata")
+load("data-raw/CTFScensuses/bci.tree2.Rdata")
+load("data-raw/CTFScensuses/bci.tree3.Rdata")
+load("data-raw/CTFScensuses/bci.tree4.Rdata")
+load("data-raw/CTFScensuses/bci.tree5.Rdata")
+load("data-raw/CTFScensuses/bci.tree6.Rdata")
+load("data-raw/CTFScensuses/bci.tree7.Rdata")
+load("data-raw/CTFScensuses/bci.tree8.Rdata")
+
+head(bci.tree1)
+
+census.years <- c(1982, 1985, 1990, 1995, 2000, 2005, 2010, 2015)
+bci.tree1$census <- census.years[1]
+bci.tree2$census <- census.years[2]
+bci.tree3$census <- census.years[3]
+bci.tree4$census <- census.years[4]
+bci.tree5$census <- census.years[5]
+bci.tree6$census <- census.years[6]
+bci.tree7$census <- census.years[7]
+bci.tree8$census <- census.years[8]
+
[…]

### 2020-03-20 Workflow.Rmd
-mortality rates now in % per year -includes Panama traits

@@
 source("R/4.1.GLUEsetup_part2_BCI.R")
-GLUEsetup_part2(growth.type = "med", dbh.residuals = "on", solar.residuals = "off", 
+GLUEsetup_part2(growth.type = "med", dbh.residuals = "off", solar.residuals = "off", 
                 growth.selection = "size_class_predefined_cc_scaled") #"size_class_varying_non_cc") #n.ensembles = 100000
+# growth.type = "med"; dbh.residuals = "off"; solar.residuals = "off"; growth.selection = "size_class_predefined_cc_scaled"
 ```
@@
 GLUE_run(splevel = "off", goodness.fit = 0.3)
-GLUE_run(splevel = "on", goodness.fit = 0.3)
+# GLUE_run(splevel = "on", goodness.fit = 0.3)
 ```
 
-```{r Chunk 2.4.0, eval = TRUE}
-source("R/4.4.Best-fit_param_BCI.R")
-bestfit.param(splevel = "on")
-bestfit.param(splevel = "off")
-```
+<!-- ```{r Chunk 2.4.0, eval = TRUE} -->
+<!-- source("R/4.4.Best-fit_param_BCI.R") -->
+<!-- # bestfit.param(splevel = "on") -->
+<!-- bestfit.param(splevel = "off") -->
+<!-- ``` -->
 
@@
 source("R/4.6.udi.calculator.R")
-udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.5) ## udi is not calculated below this rsq.thresh
-udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.5)
+##---needs to be run manually on macbook for splevel = "off" and iso.subset = "off"---
+# udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.5, iso.subset = "on") ## udi is not calculated below this rsq.thresh
+# # splevel = "on"; dryseason = "off"; rsq.thresh = 0.5; iso.subset = "on"
+# udi.calculator(splevel = "on", dryseason = "off", rsq.thresh = 0.5, iso.subset = "on") 
+# udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.5, iso.subset = "on")
+udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.5, iso.subset = "on")
+udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.5, iso.subset = "on")
+# udi.calculator(splevel = "off", dryseason = "1992", rsq.thresh = 0.5, iso.subset = "off") 
+# udi.calculator(splevel = "on", dryseason = "1992", rsq.thresh = 0.5, iso.subset = "off")
 ```
@@
 source("R/4.7.summarising_udi_by_sp_size.R")
-summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.7)
-summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.7)
+# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.5, root.selection = "on", iso.subset = "on")
+# # splevel = "on"; dryseason = "off"; rsq.thresh = 0.5; root.selection = "on"; iso.subset = "off"
+# summarise.udi(splevel = "on", dryseason = "off", rsq.thresh = 0.5, root.selection = "on", iso.subset = "on")
+# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.5, root.selection = "on", iso.subset = "on")
+# summarise.udi(splevel = "off", dryseason = "off", rsq.thresh = 0.5, root.selection = "on", iso.subset = "on")
+summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.5, root.selection = "on", iso.subset = "on")
+summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.5, root.selection = "on", iso.subset = "on")
 ```

### 2020-03-20 R/10.Comparison_with_Meinzer1999.R

@@ -3,7 +3,7 @@
 #
 rm(list=ls())
 gc()
-library(tidyverse)
+pacman::p_load(tidyverse, scales, ggpmisc)
 # graphics info
 theme_set(theme_bw())
 theme_update(text = element_text(size = 14),
@@ -11,7 +11,7 @@ theme_update(text = element_text(size = 14),
              panel.grid.minor = element_blank(),
              strip.background = element_blank()
 )
-require(scales)
+
 rev_sqrt_trans <- function() {
   scales::trans_new(
     name = "rev_sqrt",
@@ -32,21 +32,28 @@ si.type <- growth_by_si.info$si.type
 goodness.fit <- 0.3
 soil.depths <- unique(info$root.param.long$depth)
 dryseason <- "on"
+root.selection <- "on"
+iso.subset <- "on"
 ##
-load(file = paste("results/splevel/ds.bestfit_cor", goodness.fit, "_", si.type, "_", n.ensembles, "_", growth.type, "_", growth.selection, "_", dbh.residuals, "_", intervals, "_id_dryseason_", dryseason, ".Rdata", sep = ""))
-ds <- ds.bestfit
-load(file = paste("results/commlevel/ds.bestfit_cor", goodness.fit, "_", si.type, "_", n.ensembles, "_", growth.type, "_", growth.selection, "_", dbh.residuals, "_", intervals, "_id_dryseason_", dryseason, ".Rdata", sep = ""))
-ds <- rbind(ds, ds.bestfit) %>% mutate(tlplevel = as.factor(tlplevel)) %>% subset(!is.na(udi)) %>% droplevels()
 
-iso <- read.csv("data-raw/traits/isotopes/Meinzer1999_Table1.csv", na.strings = c("NA",""), header = T, row.names = NULL, check.names = F)
+file.extension.base4 <- paste0(goodness.fit, "_", si.type, "_", n.ensembles, "_", growth.type, "_", growth.selection, "_", dbh.residuals, "_", intervals, "_dryseason_", dryseason, "_iso.subset_", iso.subset, "_root.selection_", root.selection)
+
+load(file = paste0("results/splevel/ds.bestfit_cor", file.extension.base4, ".Rdata"))
+ds <- ds.bestfit
[…]

### 2020-03-20 R/11.spatial_hydrological_niche.R

@@ -1,0 +1,222 @@
+##-------------------------
+## Date : 22 Jul 2018
+## Edited: Jan 23 2020
+## Author : Rutuja
+## Title : Spatial hydrological niche
+##-------------------------
+rm(list = ls())
+
+if (!require("pacman")) install.packages("pacman"); library(pacman)
+pacman::p_load(tidyverse, devtools, data.table, reshape2, fgeo, remotes)
+# remotes::install_github("forestgeo/bciex")
+# had to load XQuartz for Mac first
+# install.packages("installr")
+# available.packages("utils")
+# packageUrl<- "https://cran.r-project.org/src/contrib/Archive/R.utils/R.utils_2.6.0.tar.gz"
+# install.packages(packageUrl, repos=NULL, type='source')
+# graphics info
+theme_set(theme_bw())
+theme_update(text = element_text(size = 14),
+             panel.grid.major = element_blank(),
+             panel.grid.minor = element_blank(),
+             strip.background = element_blank()
+)
+require(scales)
+rev_sqrt_trans <- function() {
+  scales::trans_new(
+    name = "rev_sqrt",
+    transform = function(x) -sqrt(abs(x)),
+    inverse = function(x) x^2);
+}
+file.path.spatial <- file.path("figures", "spatial")
+if(!dir.exists(file.path.spatial)) {dir.create(file.path.spatial)}
+#------------------
+## map trees at BCI color coded by UDImean
+# load tree locations data
+load("data-raw/CTFScensuses/BCI.tree8.Rdata")
+# load interval and working.iter
+load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
+load(file.path("results/4.1GLUEsetup_part2_BCI.RData")) # has working.iter and growth and si matrix
+
+intervals <- info$intervals
+n.ensembles <- growth_by_si.info$n.ensembles
+growth.type <- growth_by_si.info$growth.type
+growth.selection <- growth_by_si.info$growth.selection
+dbh.residuals <- growth_by_si.info$dbh.residuals
+si.type <- growth_by_si.info$si.type
+goodness.fit <- 0.3
+soil.depths <- unique(info$root.param.long$depth)
+dryseason <- "on"
+root.selection <- "on"
+rm(info); rm(growth_by_si.info)
+##
[…]

### 2020-03-20 R/13.hydraulic_traits_BCI.R

@@ -8,8 +8,9 @@
 rm(list = ls())
 gc()
 if (!require("pacman")) install.packages("pacman"); library(pacman)
-pacman::p_load(tidyverse, ggcorrplot, Hmisc, PerformanceAnalytics, corrplot, devtools)
-install_github("vqv/ggbiplot")
+pacman::p_load(tidyverse, ggcorrplot, Hmisc, PerformanceAnalytics,
+               corrplot, devtools, multcompView, agricolae, readxl, ggpmisc)
+# install_github("vqv/ggbiplot")
   # graphics info
 theme_set(theme_bw())
 theme_update(text = element_text(size=14),
@@ -17,164 +18,529 @@ theme_update(text = element_text(size=14),
              panel.grid.minor = element_blank(),
              strip.background = element_blank()
 )
+require(scales)
+rev_sqrt_trans <- function() {
+  scales::trans_new(
+    name = "rev_sqrt",
+    transform = function(x) -sqrt(abs(x)),
+    inverse = function(x) x^2);
+}
 ## adding uptake depth index
 load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
 load("results/4.1GLUEsetup_part2_BCI.RData") # has n.ensembles and growth and si matrix
 load("results/demo.sp_size.RData")
+load("results/demo.sp.RData") ## Mortality rate only for those cases for which avg.abundance greater than 10
 intervals <- info$intervals
 n.ensembles <- growth_by_si.info$n.ensembles
 growth.type <- growth_by_si.info$growth.type
+growth.selection <- growth_by_si.info$growth.selection
 si.type <- growth_by_si.info$si.type
 goodness.fit <- 0.3
 soil.depths <- unique(info$root.param.long$depth)
 ##
-load(file = paste("results/splevel/ds.bestfit_cor", goodness.fit, "_", si.type, "_", n.ensembles, "_", growth.type, "_", intervals, "_id.Rdata", sep = ""))
+dbh.residuals <- "on"#growth_by_si.info$dbh.residuals
+dryseason <- "on"
+root.selection <- "on"
+iso.subset <- "off"
+##
[…]

### 2020-03-20 R/2.rooting_profiles.R

@@ -67,6 +67,9 @@ for (i in 1: length(max.rds)) {
 }
 
 profiles <- do.call(rbind, profiles.dfs)
+write.csv(profiles, "results/root_profiles.csv", row.names = FALSE)
+profiles <- read.csv(file = file.path("results/root_profiles.csv"), header = TRUE)
+
 n.profiles <- unique(profiles$rf.sam)
 length(n.profiles)
 ggplot(profiles %>% subset(maxD == max.rds[length(max.rds)]), aes(y = abs.depth, x = cum.root.frac, color = maxD)) +
@@ -89,7 +92,17 @@ ggplot(profiles %>% subset(!is.na(depth)),
   scale_y_continuous(trans="rev_sqrt", breaks = signif(round(soil.depths, 2), 2))
 ggsave(file.path(paste0("figures/rooting_profiles_for_inversion_ELM_FATES_depths.jpeg")), height = 8, width = 8, units ='in')
 
-write.csv(profiles, "results/root_profiles.csv", row.names = FALSE)
+### only those rf.sam in which root fraction does not increase with depth
+
+ggplot(profiles %>% subset(power < 0.5),
+       aes(y = abs.depth, x = cum.root.frac, color = as.factor(rf.sam))) +
+  ylab("Depths (m)") + xlab("Cumulative Root Fraction (0-1)") +
+  geom_line(aes(group = rf.sam), alpha = 0.7, show.legend = FALSE) +
+  scale_y_continuous(trans="rev_sqrt", breaks = signif(round(soil.depths, 2), 2))
+ggsave(file.path(paste0("figures/rooting_profiles_for_inversion_power.threshold_0.5.jpeg")), height = 8, width = 8, units ='in')
+
+select.rf.sam <- unique(profiles$rf.sam[profiles$power < 0.5])
+write.csv(select.rf.sam, "results/rf.sam_power.threshold_0.5.csv", row.names = FALSE)
 
 profiles.sub <- profiles %>% select(rf.sam, depth, cum.root.frac) %>%
   subset(!is.na(depth))

### 2020-03-20 R/4.1.1.sizeclass.mode.R

@@ -1,0 +1,57 @@
+rm(list = ls())
+gc()
+if (!require("pacman")) install.packages("pacman"); library(pacman)
+pacman::p_load(tidyverse, data.table)
+## Tree canopy classification based on maximum diameter
+## only single stem present for a tree, usually the largest in dbh, status refers to the entire tree
+load("data-raw/CTFScensuses/bci.stem1.Rdata")
+load("data-raw/CTFScensuses/bci.tree1.Rdata")
+load("data-raw/CTFScensuses/bci.tree2.Rdata")
+load("data-raw/CTFScensuses/bci.tree3.Rdata")
+load("data-raw/CTFScensuses/bci.tree4.Rdata")
+load("data-raw/CTFScensuses/bci.tree5.Rdata")
+load("data-raw/CTFScensuses/bci.tree6.Rdata")
+load("data-raw/CTFScensuses/bci.tree7.Rdata")
+load("data-raw/CTFScensuses/bci.tree8.Rdata")
+
+head(bci.tree1)
+
+census.years <- c(1982, 1985, 1990, 1995, 2000, 2005, 2010, 2015)
+bci.tree1$census <- census.years[1]
+bci.tree2$census <- census.years[2]
+bci.tree3$census <- census.years[3]
+bci.tree4$census <- census.years[4]
+bci.tree5$census <- census.years[5]
+bci.tree6$census <- census.years[6]
+bci.tree7$census <- census.years[7]
+bci.tree8$census <- census.years[8]
+
+bci.tree <- rbind.data.frame(bci.tree1, bci.tree2, bci.tree3, bci.tree4, bci.tree5, bci.tree6, bci.tree7, bci.tree8)
+
+sp.bci.tree <- split(bci.tree, bci.tree$sp)
+sp.max.dbh <- lapply(sp.bci.tree, function (x) {
+  as.numeric(quantile(x$dbh, 0.99, na.rm = TRUE))
+})
+
+size.lims <- data.frame(mode = c("Understorey", "Transient", "Canopy", "Large Canopy"),
+                        mean.99 = c(9.8, 14.3, 27.8, 68.4),
+                        se.99 = c(2.4, 9.4, 7.0, 18.5)) %>%
+  mutate(upr = mean.99 + se.99,
+         lwr = mean.99 - se.99,
[…]

### 2020-03-20 R/4.1_Preparing_stem_growth_time_series.R

@@ -506,6 +506,7 @@ theme_update(text = element_text(size = 14),
 )
 
 load(file.path("data-raw/swp.gfac.rda"))
+load(file.path("data-raw/psi.rda"))
 census.meds <- readr::read_rds("results/census.mediandates.rds")
 census.beg <- census.meds[3: length(census.meds)]
 cut.breaks <- census.beg
@@ -525,6 +526,22 @@ plot.swp.gfac.stat.3 <- ggplot(swp.gfac.stat.2, aes(x = interval.yrs, y = mean))
   scale_fill_viridis_c("Growth\nFactor\n[unitless, 0-1]", trans = "reverse", option = "plasma") +
   ggtitle("Average Growth Factor by interval and depth across best-fit ensembles")
 ggsave("swp.gfac_mean_across_params.top.few_full_interval_no-depth.jpeg", plot = plot.swp.gfac.stat.3, path =
+         file.path("figures"), height = 5, width = 5, units='in')
+
+psi <- psi %>%
+  mutate(interval.yrs = cut(date, include.lowest = TRUE, breaks = cut.breaks,
+                            labels = cut.labels.2, right = TRUE))
+
+psi.stat.2 <- psi %>%
+  subset(!is.na(interval.yrs)) %>%
+  group_by(interval.yrs) %>%
+  summarise(mean = mean(psi, na.rm = TRUE)) %>%
+  droplevels()
+plot.psi.stat.3 <- ggplot(psi.stat.2, aes(x = interval.yrs, y = mean)) +
+  geom_point() + ylab("PSI (MPa)") + xlab("Census Interval") +
+  scale_fill_viridis_c("Soil\nWater\nPotential\n[MPa]", trans = "reverse", option = "plasma") +
+  ggtitle("Mean Soil Water Potential by interval and depth across best-fit ensembles")
+ggsave("psi_mean_across_params.top.few_full_interval_no-depth.jpeg", plot = plot.psi.stat.3, path =
          file.path("figures"), height = 5, width = 5, units='in')
 
 ####------------------------------------------------------

### 2020-03-20 R/4.2.GLUErun_part2_func2_BCI.R

@@ -1,5 +1,5 @@
 ## function to calculate rsq between observed growth vs. simulated SI matrix
-growth_by_si.func <- function(growth_by_si.info, goodness.fit, statistic, splevel){
+growth_by_si.func2 <- function(growth_by_si.info, goodness.fit, statistic, splevel){
 
   if (splevel == "on"){
     sp.with.tlp <- names(info.2$sp.si)
@@ -46,21 +46,11 @@ growth_by_si.func <- function(growth_by_si.info, goodness.fit, statistic, spleve
     }
     GLUE.sub <- apply(si.matrix, 1, function (x) {
       sim <- x[obs$interval - location.adj]
-      cor.val <- cor(obs$median, sim, method = "pearson", use = "complete.obs")
-      if (!is.na(cor.val) & (abs(cor.val) > (goodness.fit - 0.01))) {
-        if (statistic == "rsq") {
-          value <- cor.val^2 } else {
-            value <- cor.val
-            }
-      } else {
-        value <- NA
-      }
-      # list(cor = cor, rsq = rsq)
-      # model <- lm(y ~ obs)
-      # summary(model)$r.squared
-      # list(model = model, rsq = rsq)
-    }
-    )
+      model <- lm.fit(matrix(sim), matrix(obs$median))
+      within.ci <- data.table::between(model$fitted.values, obs$lwr, obs$upr)
+      # return(all(data.table::between(model$fitted.values, obs$lwr, obs$upr), na.rm = FALSE))
+      return(sum(within.ci, na.rm = TRUE))
+    })
     GLUE.list[[jj]] <- unlist(GLUE.sub)
   }
   Sys.time()
@@ -69,6 +59,8 @@ growth_by_si.func <- function(growth_by_si.info, goodness.fit, statistic, spleve
   parallel::stopCluster(cl)
   GLUE <- data.frame(do.call(rbind, GLUE.list))
   row.names(GLUE) <- sp_size
+  # apply(GLUE, 1, function(x){sum(x, na.rm = TRUE)}) ## if all were TRUE
+  # matches.cutoff <- 5; cond <- apply(GLUE, 1, function(x){length(x[x >= matches.cutoff])}); length(cond[cond != 0])/length(cond)
   # 12 min for splevel = "on" sp-size & ~1500 n.rf.sam, 33 min for splevel = "off"
   return(GLUE)
 }

### 2020-03-20 R/4.3.GLUErun_part2_BCI.R

@@ -1,64 +1,67 @@
+rm(list = ls())
+gc()
+if (!require("pacman"))
+  install.packages("pacman")
+library(pacman)
+pacman::p_load(tidyverse, doParallel, foreach)
 
-  rm(list = ls())
-  gc()
-  if (!require("pacman")) install.packages("pacman"); library(pacman)
-  pacman::p_load(tidyverse, doParallel, foreach)
-
-  load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
-  intervals <- info$intervals
-  load("results/GLUEsetup_part1.5_BCI.RData")
-  load(file.path("results/4.1GLUEsetup_part2_BCI.RData"))
-  n.ensembles <- growth_by_si.info$n.ensembles
-  source(file.path("R/4.2.GLUErun_part2_func_BCI.R"))
-  GLUE_run <- function(splevel, goodness.fit){
-    if(splevel == "on") {
-      sp.GLUE.rsq <- growth_by_si.func(growth_by_si.info = growth_by_si.info,
-                                       statistic = "rsq", splevel = "on", goodness.fit)
-
-      save(sp.GLUE.rsq, file = paste0("results/sp.GLUE.rsq_",
-                                      growth_by_si.info$si.type, "_", n.ensembles, "_",
-                                      growth_by_si.info$growth.type, "_",
-                                      growth_by_si.info$growth.selection, "_",
-                                      growth_by_si.info$dbh.residuals, "_",
-                                      info$intervals, "_cor", goodness.fit, ".Rdata"))
-      rm(sp.GLUE.rsq)
-
-      # sp.GLUE.cor <- growth_by_si.func(growth_by_si.info = growth_by_si.info,
-      #                                  statistic = "cor", splevel = "on", goodness.fit)
-      # save(sp.GLUE.cor, file = paste0("results/sp.GLUE.cor_", growth_by_si.info$si.type, "_", n.ensembles, "_",
-      #                                 growth_by_si.info$growth.type, "_",
-      #                                 info$intervals, "_cor", goodness.fit, ".Rdata"))
-
-    } else {
-    # Rprof(file = "growth_by_si.out", interval = 0.02,
-    #       line.profiling = TRUE)
-      beg <- Sys.time()
-      beg
-      GLUE.rsq <- growth_by_si.func(growth_by_si.info = growth_by_si.info,
[…]

### 2020-03-20 R/4.4.Best-fit_param_BCI.R

@@ -1,81 +1,56 @@
-## Not anymore: Finding best fit water-stress/growth model parameters for species growth time series
-## Just vectorises Rsquares with decreasing order and notes down the order to relate to par.sam and rf.sam from growth_by_si.info$si.param.rel
-
-# for 50 ha obs species groups
-rm(list = ls())
-gc()
-if (!require("pacman")) install.packages("pacman"); library(pacman)
-pacman::p_load(tidyverse, doParallel, foreach, data.table)
-
-bestfit.param <- function(splevel) {
-  # load interval and n.ensembles
-  load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
-  load(file.path("results/4.1GLUEsetup_part2_BCI.RData")) # has n.ensembles and growth and si matrix
-
-  intervals <- info$intervals
-  n.ensembles <- growth_by_si.info$n.ensembles
-  growth.type <- growth_by_si.info$growth.type ## this is not data but type such as "individual"
-  growth.selection <- growth_by_si.info$growth.selection
-  si.type <- growth_by_si.info$si.type
-  dbh.residuals <- growth_by_si.info$dbh.residuals
-  goodness.fit <- 0.3 # rsq0.3
-  if (splevel == "on") {
-    level.folder <- "splevel"
-  } else {
-    level.folder <- "commlevel"
-  }
-  ncor <- detectCores() - 2
-
-  if (splevel == "on") {
-    load(paste0("results/sp.GLUE.rsq_", si.type, "_", n.ensembles, "_",
-              growth.type, "_", growth.selection, "_", dbh.residuals, "_", intervals, "_cor", goodness.fit, ".Rdata"))
-    GLUE.rsq <- sp.GLUE.rsq
-    rm(sp.GLUE.rsq)
-  } else {
-    load(paste0("results/GLUE.rsq_", si.type, "_", n.ensembles, "_",
-                growth.type, "_", growth.selection, "_", dbh.residuals, "_", intervals, "_cor", goodness.fit, ".Rdata"))
-  }
-  sp_size <- names(GLUE.rsq)
-  sp <- stringr::str_split(sp_size, "_", simplify = TRUE)[, 1]
-  size <- stringr::str_split(sp_size, "_", simplify = TRUE)[, 2]
[…]

### 2020-03-20 R/4.5.prep.udi.calculator.R

@@ -37,6 +37,8 @@ prep.udi.calculator <- function(splevel) {
   depth.thick <- data.frame(depth = as.numeric(unique(rdt$depth))) %>%
     mutate(thickness = depth - lag(depth, default = 0)) %>% data.table()
 
+  file.extension.base1 <- paste0(goodness.fit, "_", si.type, "_", n.ensembles, "_", growth.type, "_", growth.selection, "_", dbh.residuals, "_", intervals)
+
   if (splevel == "on") {
     load(file.path("results/sp.hydro.output.chosen.RData"))
     sp.hydro.output.chosen <- lapply(sp.hydro.output.chosen, function(x){
@@ -47,25 +49,58 @@ prep.udi.calculator <- function(splevel) {
         paw.df, on = c("date","depth","par.sam"), `:=`(paw = i.paw)][,
                                                                      tw := paw*thickness*1000][, !c("paw", "thickness")], depth)
     })
-    save(sdt.list, file = paste("results/splevel/sdt.list_cor", goodness.fit, "_", si.type, "_", n.ensembles, "_", growth.type, "_", growth.selection, "_", dbh.residuals, "_", intervals, "_dryseason_off.Rdata", sep = ""))
+    save(sdt.list, file = paste0("results/splevel/sdt.list_cor", file.extension.base1, "_dryseason_off.Rdata"))
     rm(sp.hydro.output.chosen)
-    ### Subsetting to dryseason----
-    sdt.list <- lapply(sdt.list, function(x) {
-      x[, ':='(month = format(as.Date(date), "%m"), year = format(as.Date(date), "%Y"))][month %chin% c("03", "04") & year == "1997"] ## Choosing Feb, March and April
+    sdt.list.full <- sdt.list
+    # load(file = paste0("results/splevel/sdt.list_cor", file.extension.base1, "_dryseason_off.Rdata"))
+    ### Subsetting to late dryseason 1997----
+    sdt.list <- lapply(sdt.list.full, function(x) {
+      x[, ':='(month = format(as.Date(date), "%m"), year = format(as.Date(date), "%Y"))][month %chin% c("03", "04") & year == "1997"] ## Choosing March and April
     })
     sdt.list <- lapply(sdt.list,  droplevels)
[…]

### 2020-03-20 R/4.6.udi.calculator.R

@@ -8,7 +8,7 @@ for (i in 1:5) {gc()}
 if (!require("pacman")) install.packages("pacman"); library(pacman)
 pacman::p_load(tidyverse, doParallel, foreach, data.table)
 
-udi.calculator <- function(splevel, dryseason,  rsq.thresh) {
+udi.calculator <- function(splevel, dryseason, rsq.thresh, iso.subset) {
   # load interval and n.ensembles
   load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
   load(file.path("results/4.1GLUEsetup_part2_BCI.RData")) # has n.ensembles and growth and si matrix
@@ -29,24 +29,35 @@ udi.calculator <- function(splevel, dryseason,  rsq.thresh) {
     level.folder <- "commlevel"
     }
 
-  rf.sam <- unique(si.param.rel$rf.sam)
+  rf.sam.vec <- unique(si.param.rel$rf.sam)
   n.par.sam <- length(unique(si.param.rel$par.sam)) ## this is the sequence repeated in growth_by_si.info$si.param.rel
-  par.sam <- si.param.rel$par.sam[1:n.par.sam] ## this is the sequence used when GLUE.rsq was constructed
-  n.rf.sam <- length(rf.sam)
+  par.sam.vec <- si.param.rel$par.sam[1:n.par.sam] ## this is the sequence used when GLUE.rsq was constructed
+  n.rf.sam <- length(rf.sam.vec)
   rm(info); rm(growth_by_si.info)
   ## sdt is chosen based on dryseason on/off
-  load(file = paste0("results/", level.folder, "/GLUE.rsq.vector_cor", goodness.fit, "_", si.type, "_", n.ensembles, "_", growth.type, "_", growth.selection, "_", dbh.residuals, "_", intervals, ".Rdata"), envir = parent.frame(), verbose = FALSE)
+  file.extension.base1 <- paste0(goodness.fit, "_", si.type, "_", n.ensembles, "_", growth.type, "_", growth.selection, "_", dbh.residuals, "_", intervals)
+  file.extension.base2 <- paste0(file.extension.base1, "_dryseason_", dryseason)
+  file.extension.base3 <- paste0(file.extension.base2, "_iso.subset_", iso.subset)
+
+  load(file = paste0("results/", level.folder, "/GLUE.rsq_cor", file.extension.base1 , ".Rdata"), envir = parent.frame(), verbose = FALSE)
[…]

### 2020-03-20 R/4.7.summarising_udi_by_sp_size.R

@@ -3,11 +3,12 @@ gc()
 if (!require("pacman")) install.packages("pacman"); library(pacman)
 pacman::p_load(tidyverse, data.table)
 
-summarise.udi <- function(splevel, dryseason, rsq.thresh) {
+summarise.udi <- function(splevel, dryseason, rsq.thresh, root.selection, iso.subset) {
   load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
   load(file.path("results/4.1GLUEsetup_part2_BCI.RData")) # has n.ensembles and growth and si matrix
 
   intervals <- info$intervals
+  rm(info)
   n.ensembles <- growth_by_si.info$n.ensembles
   growth.type <- growth_by_si.info$growth.type ## this is not data but type such as "individual"
   si.type <- growth_by_si.info$si.type
@@ -22,53 +23,77 @@ summarise.udi <- function(splevel, dryseason, rsq.thresh) {
     level.folder <- "commlevel"
   }
 
-  load(file = paste("results/", level.folder, "/ds.bestfit.all_cor", goodness.fit, "_", si.type, "_", n.ensembles, "_", growth.type, "_", growth.selection, "_", dbh.residuals, "_", intervals, "_id_dryseason_", dryseason, ".Rdata", sep = ""))
-  ds.bestfit <- ds.bestfit.all %>% group_by(sp_size) %>%
+  file.extension.base3 <- paste0(goodness.fit, "_", si.type, "_", n.ensembles, "_", growth.type, "_", growth.selection, "_", dbh.residuals, "_", intervals, "_dryseason_", dryseason, "_iso.subset_", iso.subset)
+  file.extension.base4 <- paste0(file.extension.base3, "_root.selection_", root.selection)
+
+  load(file = paste0("results/", level.folder, "/ds.bestfit.all_cor", file.extension.base3, ".Rdata"))
+
+  if (root.selection == "on") {
+    select.rf.sam <- read.csv(file = file.path("results/rf.sam_power.threshold_0.5.csv"), header = TRUE)
+    ds.bestfit.all <- ds.bestfit.all %>% subset(rf.sam %in% select.rf.sam$x)
+  } # 702 root profiles selected
+    ds.bestfit.select <- ds.bestfit.all %>%
+      #subset(!is.na(udi)) %>%
+      group_by(sp_size) %>%
[…]

### 2020-03-20 R/9.5mortality_BCI.R

@@ -2,7 +2,7 @@ rm(list=ls())
 gc()
 # load("/Library/Frameworks/R.framework/Versions/3.4/Resources/library/CTFSRPackage/CTFSRPackage.Rdata")
 if (!require("pacman")) install.packages("pacman"); library(pacman)
-pacman::p_load(tidyverse, hms)
+pacman::p_load(tidyverse, hms, ggpmisc)
 
 # graphics info
 tex <- element_text(size = 16, face = "plain") # , family = "gara"
@@ -18,8 +18,23 @@ theme_update(text = element_text(size=14),
              panel.grid.minor = element_blank(),
              strip.background = element_blank()
 )
+
+census.years <- c(1982, 1985, 1990, 1995, 2000, 2005, 2010, 2015)
+
+require(scales)
+rev_sqrt_trans <- function() {
+  scales::trans_new(
+    name = "rev_sqrt",
+    transform = function(x) -sqrt(abs(x)),
+    inverse = function(x) x^2);
+}
+####**********************************************************************************************
+##### ---- If you dont need to recalculate demographic rates, jump to "Load demographic data" ----
+####**********************************************************************************************
+
 ## load full tree data
 ## only single stem present for a tree, usually the largest in dbh, status refers to the entire tree
+load("data-raw/CTFScensuses/bci.stem1.Rdata")
 load("data-raw/CTFScensuses/bci.tree1.Rdata")
 load("data-raw/CTFScensuses/bci.tree2.Rdata")
 load("data-raw/CTFScensuses/bci.tree3.Rdata")
@@ -31,7 +46,6 @@ load("data-raw/CTFScensuses/bci.tree8.Rdata")
 
 head(bci.tree1)
 
-census.years <- c(1982, 1985, 1990, 1995, 2000, 2005, 2010, 2015)
 bci.tree1$census <- census.years[1]
 bci.tree2$census <- census.years[2]
 bci.tree3$census <- census.years[3]
@@ -43,6 +57,7 @@ bci.tree8$census <- census.years[8]
 
 bci.tree <- rbind.data.frame(bci.tree1, bci.tree2, bci.tree3, bci.tree4, bci.tree5, bci.tree6, bci.tree7, bci.tree8)
 range(bci.tree$dbh, na.rm = T)
+
 ## trees that are in status dead have no size.
[…]

### 2020-04-02 Workflow.Rmd
Moved R files to a new folder code

--- 2020-01-22 Workflow.Rmd
@@
 
-##### Load swp or gfac data from 
+### 1.GLUEsetup_part1_BCI.R
+
+ * Loads best.fit ensembles of Soil Water Potential (SWP) from bci.elm.fates.hydro::psi of a specific version (given by the "current folder")
+ * Loads hypothetical rooting profiles generated by R/3.090_root_parameters_b.R
+ * Loads Turgor Loss Point (TLP) data
+ * Loads tree growth census dates for specific no. of intervals
+ * Calculates btran time-series by census-interval for each root profile for all the best.fit SWP ensemble
+  + by individual species with species-specific TLP (N of btran time series == No. of species x No. of SWP ensembles x No. of root profiles)
+  + at community-mean TLP (No. of btran time series == No. of SWP ensembles x No. of root profiles)
+  * Saves all the above info in a list called info, except species level btran is saved in info.2
+
 ## Then using a combination of a range of hypothetical root profiles obtained from 2.rooting_profiles.R and SWP ensemble simulations generate a btran time-series using equation : Btran= i=1 to i=z RootFrac_z * f(swp_z),
-where f(swp_z) is a growth.factor_z based on a linear step relationship with swp_z defined based on maximum value in BCI tlp data from Kunert's.
+where f(swp_z) is a growth.factor_z based on a linear step relationship with swp_z ap species specific or community mean TLP level.
 Code/6.0_Growth_correlate_table_from_swp.R
@@
 ```{r Chunk 1.0, eval = FALSE}
-source("R/1.GLUEsetup_part1_BCI.R")
+source("code/1.GLUEsetup_part1_BCI.R")
 GLUEsetup_part1(current.folder = "2019-10-14_5000", intervals = 5) # this can take 30 min to get btran by nsam
@@
 
-```{r Chunk 2.2.0, eval = FALSE}
-source("R/4.1.GLUEsetup_part2_BCI.R")
-GLUEsetup_part2(growth.type = "med", dbh.residuals = "on", solar.residuals = "off", 
+```{r Chunk 2, eval = FALSE}
+source("code/4.1.GLUEsetup_part2_BCI.R")
+GLUEsetup_part2(growth.type = "med", dbh.residuals = "off", solar.residuals = "off", 
                 growth.selection = "size_class_predefined_cc_scaled") #"size_class_varying_non_cc") #n.ensembles = 100000
+# growth.type = "med"; dbh.residuals = "off"; solar.residuals = "off"; growth.selection = "size_class_predefined_cc_scaled"
 ```
 
-```{r Chunk 2.3.0, eval = TRUE}
-# This sources "R/4.2GLUErun_part2_func_BCI.R"
-source("R/4.3.GLUErun_part2_BCI.R")
+```{r Chunk 3, eval = TRUE}
+# This sources "code/4.2GLUErun_part2_func_BCI.R"
+source("code/4.3.GLUErun_part2_BCI.R")
 GLUE_run(splevel = "off", goodness.fit = 0.3)
-GLUE_run(splevel = "on", goodness.fit = 0.3)
+# GLUE_run(splevel = "on", goodness.fit = 0.3)
 ```
 
-```{r Chunk 2.4.0, eval = TRUE}
-source("R/4.4.Best-fit_param_BCI.R")
-bestfit.param(splevel = "on")
-bestfit.param(splevel = "off")
-```
+<!-- ```{r Chunk 2.4.0, eval = TRUE} -->
+<!-- source("code/4.4.Best-fit_param_BCI.R") -->
+<!-- # bestfit.param(splevel = "on") -->
+<!-- bestfit.param(splevel = "off") -->
+<!-- ``` -->
 
-```{r Chunk 2.4.1, eval = TRUE}
-source("R/4.5.prep.udi.calculator.R")
+```{r Chunk 4, eval = TRUE}
+source("code/4.5.prep.udi.calculator.R")
 prep.udi.calculator(splevel = "on")
@@
 
-```{r Chunk 2.4.2, eval = TRUE}
-source("R/4.6.udi.calculator.R")
-udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.5) ## udi is not calculated below this rsq.thresh
-udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.5)
+```{r Chunk 5, eval = TRUE}
+source("code/4.6.udi.calculator.R")
+##---needs to be run manually on macbook for splevel = "off" and iso.subset = "off"---
+# udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.5, iso.subset = "on") ## udi is not calculated below this rsq.thresh
+# # splevel = "on"; dryseason = "off"; rsq.thresh = 0.5; iso.subset = "on"
+# udi.calculator(splevel = "on", dryseason = "off", rsq.thresh = 0.5, iso.subset = "on") 
+# udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.5, iso.subset = "on")
+udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.5, iso.subset = "on")
+udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.5, iso.subset = "on")
+# udi.calculator(splevel = "off", dryseason = "1992", rsq.thresh = 0.5, iso.subset = "off") 
+# udi.calculator(splevel = "on", dryseason = "1992", rsq.thresh = 0.5, iso.subset = "off")
 ```
 
-```{r Chunk 2.4.4, eval = TRUE}
-source("R/4.7.summarising_udi_by_sp_size.R")
-summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.7)
-summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.7)
+```{r Chunk 6, eval = TRUE}
+source("code/4.7.summarising_udi_by_sp_size.R")
+# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.5, root.selection = "on", iso.subset = "on")
+# # splevel = "on"; dryseason = "off"; rsq.thresh = 0.5; root.selection = "on"; iso.subset = "off"
+# summarise.udi(splevel = "on", dryseason = "off", rsq.thresh = 0.5, root.selection = "on", iso.subset = "on")
+# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.5, root.selection = "on", iso.subset = "on")
+# summarise.udi(splevel = "off", dryseason = "off", rsq.thresh = 0.5, root.selection = "on", iso.subset = "on")
+summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.5, root.selection = "on", iso.subset = "on")
+summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.5, root.selection = "on", iso.subset = "on")
 ```
 
-```{r Chunk 2.5.0, eval = TRUE}
-source("R/10.Comparison_with_Meinzer1999.R")
+```{r Chunk 7, eval = TRUE}
+source("code/10.Comparison_with_Meinzer1999.R")
 ```
 
-```{r Chunk 2.5.1, eval = FALSE}
-source("R/5.species_graphs_BCI.R")
-source("R/9.5mortality_BCI.R")
-source("R/13.hydraulic_traits_BCI.R")
+```{r Chunk 8, eval = FALSE}
+source("code/9.6_mortality_BCI.R")
+demo.graphs(level.folder = "commlevel", n.threshold = 30, dbh.residuals = "on", 
+           dryseason = "on", root.selection = "on", iso.subset = "off")
+# level.folder = "commlevel"; n.threshold = 30; dbh.residuals = "on";
+#            dryseason = "on"; root.selection = "on"; iso.subset = "off"
 ```
+
+```{r Chunk 9, eval = FALSE}
+source("code/5.species_graphs_BCI.R")
+source("code/13.hydraulic_traits_BCI.R")
+```

### 2020-04-02 R/9.5mortality_BCI.R

@@ -1,549 +1,0 @@
-rm(list=ls())
-gc()
-# load("/Library/Frameworks/R.framework/Versions/3.4/Resources/library/CTFSRPackage/CTFSRPackage.Rdata")
-if (!require("pacman")) install.packages("pacman"); library(pacman)
-pacman::p_load(tidyverse, hms, ggpmisc)
-
-# graphics info
-tex <- element_text(size = 16, face = "plain") # , family = "gara"
-tex <- element_text(size = 5, face = "plain")
-my.theme <-  theme(axis.text = tex, axis.title = tex,
-                   title = tex, legend.title = tex, legend.text = tex, strip.text.y = tex, strip.text.x = tex)
-my.bg <- theme_bw() + theme(panel.grid.major.y = element_blank(), panel.grid.major.x = element_line(colour = "black", size = 0.25),
-                            panel.grid.minor = element_blank())
-my.adjust <- theme(axis.title.y = element_text(vjust = 1), axis.title.x = element_text(vjust = -0.6), title = element_text(vjust = 2))
-theme_set(theme_bw())
-theme_update(text = element_text(size=14),
-             panel.grid.major = element_blank(),
-             panel.grid.minor = element_blank(),
-             strip.background = element_blank()
-)
-
-census.years <- c(1982, 1985, 1990, 1995, 2000, 2005, 2010, 2015)
-
-require(scales)
-rev_sqrt_trans <- function() {
-  scales::trans_new(
-    name = "rev_sqrt",
-    transform = function(x) -sqrt(abs(x)),
-    inverse = function(x) x^2);
-}
-####**********************************************************************************************
-##### ---- If you dont need to recalculate demographic rates, jump to "Load demographic data" ----
-####**********************************************************************************************
-
-## load full tree data
-## only single stem present for a tree, usually the largest in dbh, status refers to the entire tree
-load("data-raw/CTFScensuses/bci.stem1.Rdata")
-load("data-raw/CTFScensuses/bci.tree1.Rdata")
-load("data-raw/CTFScensuses/bci.tree2.Rdata")
-load("data-raw/CTFScensuses/bci.tree3.Rdata")
[…]

### 2020-04-02 code/9.5_demographic_data_BCI.R

@@ -1,0 +1,267 @@
+rm(list=ls())
+gc()
+# load("/Library/Frameworks/R.framework/Versions/3.4/Resources/library/CTFSRPackage/CTFSRPackage.Rdata")
+if (!require("pacman")) install.packages("pacman"); library(pacman)
+pacman::p_load(tidyverse, hms, ggpmisc)
+
+# graphics info
+theme_set(theme_bw())
+theme_update(text = element_text(size=14),
+             panel.grid.major = element_blank(),
+             panel.grid.minor = element_blank(),
+             strip.background = element_blank()
+)
+
+census.years <- c(1982, 1985, 1990, 1995, 2000, 2005, 2010, 2015)
+
+require(scales)
+rev_sqrt_trans <- function() {
+  scales::trans_new(
+    name = "rev_sqrt",
+    transform = function(x) -sqrt(abs(x)),
+    inverse = function(x) x^2);
+}
+####**********************************************************************************************
+##### ---- If you dont need to recalculate demographic rates, jump to "Load demographic data" ----
+####**********************************************************************************************
+
+## load full tree data
+## only single stem present for a tree, usually the largest in dbh, status refers to the entire tree
+load("data-raw/CTFScensuses/bci.stem1.Rdata")
+load("data-raw/CTFScensuses/bci.tree1.Rdata")
+load("data-raw/CTFScensuses/bci.tree2.Rdata")
+load("data-raw/CTFScensuses/bci.tree3.Rdata")
+load("data-raw/CTFScensuses/bci.tree4.Rdata")
+load("data-raw/CTFScensuses/bci.tree5.Rdata")
+load("data-raw/CTFScensuses/bci.tree6.Rdata")
+load("data-raw/CTFScensuses/bci.tree7.Rdata")
+load("data-raw/CTFScensuses/bci.tree8.Rdata")
+
+head(bci.tree1)
+
+bci.tree1$census <- census.years[1]
+bci.tree2$census <- census.years[2]
+bci.tree3$census <- census.years[3]
+bci.tree4$census <- census.years[4]
+bci.tree5$census <- census.years[5]
+bci.tree6$census <- census.years[6]
+bci.tree7$census <- census.years[7]
+bci.tree8$census <- census.years[8]
+
[…]

### 2020-04-05 Workflow.Rmd
Code rearrangement after rooting profiles changed; and together wit the last commit edits towards running a sensitivity analyses by removing several dry season months in Btran calculations for deciduous species

@@
 ```{r Chunk 1.0, eval = FALSE}
-source("code/1.GLUEsetup_part1_BCI.R")
+source("code/4.1.1.GLUEsetup_part1_BCI.R")
 GLUEsetup_part1(current.folder = "2019-10-14_5000", intervals = 5) # this can take 30 min to get btran by nsam
@@
 ```{r Chunk 2, eval = FALSE}
-source("code/4.1.GLUEsetup_part2_BCI.R")
-GLUEsetup_part2(growth.type = "med", dbh.residuals = "off", solar.residuals = "off", 
+source("code/4.1.2.GLUEsetup_part2_BCI.R")
+GLUEsetup_part2(growth.type = "med", dbh.residuals = "on", solar.residuals = "off", 
                 growth.selection = "size_class_predefined_cc_scaled") #"size_class_varying_non_cc") #n.ensembles = 100000
-# growth.type = "med"; dbh.residuals = "off"; solar.residuals = "off"; growth.selection = "size_class_predefined_cc_scaled"
+# growth.type = "med"; dbh.residuals = "on"; solar.residuals = "off"; growth.selection = "size_class_predefined_cc_scaled"
 ```
@@
 source("code/4.3.GLUErun_part2_BCI.R")
-GLUE_run(splevel = "off", goodness.fit = 0.3)
-# GLUE_run(splevel = "on", goodness.fit = 0.3)
+# splevel = "on"; goodness.fit = 0.3; drop.months = "Mar"
+GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "Feb")
+GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "Mar")
+GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "FebMar")
+GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "Feb")
+GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "Mar")
+GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "FebMar")
 ```
@@
 source("code/4.5.prep.udi.calculator.R")
-prep.udi.calculator(splevel = "on")
-prep.udi.calculator(splevel = "off")
+# splevel = "on"; drop.months = "JanFebMar"
+prep.udi.calculator(splevel = "on", drop.months = "Feb")
+prep.udi.calculator(splevel = "on", drop.months = "FebMar")
+prep.udi.calculator(splevel = "on", drop.months = "Mar")
+
+prep.udi.calculator(splevel = "off", drop.months = "Feb")
+prep.udi.calculator(splevel = "off", drop.months = "FebMar")
+prep.udi.calculator(splevel = "off", drop.months = "Mar")
 ```
@@
 # udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.5, iso.subset = "on") ## udi is not calculated below this rsq.thresh
-# # splevel = "on"; dryseason = "off"; rsq.thresh = 0.5; iso.subset = "on"
-# udi.calculator(splevel = "on", dryseason = "off", rsq.thresh = 0.5, iso.subset = "on") 
-# udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.5, iso.subset = "on")
-udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.5, iso.subset = "on")
-udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.5, iso.subset = "on")
-# udi.calculator(splevel = "off", dryseason = "1992", rsq.thresh = 0.5, iso.subset = "off") 
-# udi.calculator(splevel = "on", dryseason = "1992", rsq.thresh = 0.5, iso.subset = "off")
+# splevel = "on"; dryseason = "on"; rsq.thresh = 0.5;  n.ci.thresh = 3; iso.subset = "on"; drop.months = "JanFebMar"
+udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.3, 
+               n.ci.thresh = 3, iso.subset = "on", drop.months = "Feb")
+udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.3, 
+               n.ci.thresh = 3, iso.subset = "on", drop.months = "FebMar")
+udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.3, 
+               n.ci.thresh = 3, iso.subset = "on", drop.months = "Mar")
+
+udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.3, 
+               n.ci.thresh = 3, iso.subset = "on", drop.months = "Feb")
+udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.3, 
+               n.ci.thresh = 3, iso.subset = "on", drop.months = "FebMar")
+udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.3, 
+               n.ci.thresh = 3, iso.subset = "on", drop.months = "Mar")
+
+udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.3, 
+               n.ci.thresh = 3, iso.subset = "off", drop.months = "Feb")
+udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.3, 
+               n.ci.thresh = 3, iso.subset = "off", drop.months = "FebMar")
+udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.3, 
+               n.ci.thresh = 3, iso.subset = "off", drop.months = "Mar")
+
+udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.3, 
+               n.ci.thresh = 3, iso.subset = "off", drop.months = "Feb")
+udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.3, 
+               n.ci.thresh = 3, iso.subset = "off", drop.months = "FebMar")
+udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.3, 
+               n.ci.thresh = 3, iso.subset = "off", drop.months = "Mar")
 ```
@@
 source("code/4.7.summarising_udi_by_sp_size.R")
-# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.5, root.selection = "on", iso.subset = "on")
-# # splevel = "on"; dryseason = "off"; rsq.thresh = 0.5; root.selection = "on"; iso.subset = "off"
-# summarise.udi(splevel = "on", dryseason = "off", rsq.thresh = 0.5, root.selection = "on", iso.subset = "on")
-# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.5, root.selection = "on", iso.subset = "on")
-# summarise.udi(splevel = "off", dryseason = "off", rsq.thresh = 0.5, root.selection = "on", iso.subset = "on")
-summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.5, root.selection = "on", iso.subset = "on")
-summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.5, root.selection = "on", iso.subset = "on")
+summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.5, root.selection = "exp", iso.subset = "on", drop.months = "Feb")
+summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.5, root.selection = "exp", iso.subset = "on", drop.months = "FebMar")
+# # splevel = "on"; dryseason = "off"; rsq.thresh = 0.5; root.selection = "exp"; iso.subset = "off"
+summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.5, root.selection = "exp", iso.subset = "on", drop.months = "Jan")
+
+summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.5, root.selection = "exp", iso.subset = "on", drop.months = "FebMar")
+summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.5, root.selection = "exp", iso.subset = "on", drop.months = "Feb")
+summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.5, root.selection = "exp", iso.subset = "on", drop.months = "Mar")
+
+summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.5, root.selection = "exp", iso.subset = "off", drop.months = "Feb")
+summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.5, root.selection = "exp", iso.subset = "off", drop.months = "FebMar")
+summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.5, root.selection = "exp", iso.subset = "off", drop.months = "Mar")
+
+summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.5, root.selection = "exp", iso.subset = "off", drop.months = "FebMar")
+summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.5, root.selection = "exp", iso.subset = "off", drop.months = "Feb")
+summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.5, root.selection = "exp", iso.subset = "off", drop.months = "Mar")
 ```
@@
 source("code/10.Comparison_with_Meinzer1999.R")
+iso.compare(goodness.fit = 0.3, dryseason = "on", 
+            root.selection = "exp", iso.subset = "on", drop.months = "None")
+iso.compare(goodness.fit = 0.3, dryseason = "on", 
+            root.selection = "exp", iso.subset = "on", drop.months = "JanFebMar")
+iso.compare(goodness.fit = 0.3, dryseason = "on", 
+            root.selection = "exp", iso.subset = "on", drop.months = "JanFeb")
+iso.compare(goodness.fit = 0.3, dryseason = "on", 
+            root.selection = "exp", iso.subset = "on", drop.months = "Jan")
+# goodness.fit = 0.3; dryseason = "on"; root.selection = "exp"; iso.subset = "on"; drop.months = "None"
 ```

### 2020-04-05 code/10.Comparison_with_Meinzer1999.R

@@ -1,528 +1,543 @@
 ## Comparison with Meinzer et al. 1999 results.
 
-#
 rm(list=ls())
 gc()
-pacman::p_load(tidyverse, scales, ggpmisc)
-# graphics info
-theme_set(theme_bw())
-theme_update(text = element_text(size = 14),
-             panel.grid.major = element_blank(),
-             panel.grid.minor = element_blank(),
-             strip.background = element_blank()
-)
 
-rev_sqrt_trans <- function() {
-  scales::trans_new(
-    name = "rev_sqrt",
-    transform = function(x) -sqrt(abs(x)),
-    inverse = function(x) x^2);
-}
+iso.compare <- function(goodness.fit = goodness.fit,
+                         dryseason = dryseason,
+                         root.selection =  root.selection,
+                         iso.subset = iso.subset, drop.months = drop.months){
+  pacman::p_load(tidyverse, scales, ggpmisc)
+  # graphics info
+  theme_set(theme_bw())
+  theme_update(text = element_text(size = 14),
+               panel.grid.major = element_blank(),
+               panel.grid.minor = element_blank(),
+               strip.background = element_blank()
+  )
 
-# load interval and working.iter
-load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
-load(file.path("results/4.1GLUEsetup_part2_BCI.RData")) # has working.iter and growth and si matrix
+  rev_sqrt_trans <- function() {
+    scales::trans_new(
+      name = "rev_sqrt",
+      transform = function(x) -sqrt(abs(x)),
+      inverse = function(x) x^2);
+  }
 
-intervals <- info$intervals
-n.ensembles <- growth_by_si.info$n.ensembles
-growth.type <- growth_by_si.info$growth.type
-growth.selection <- growth_by_si.info$growth.selection
-dbh.residuals <- growth_by_si.info$dbh.residuals
-si.type <- growth_by_si.info$si.type
-goodness.fit <- 0.3
-soil.depths <- unique(info$root.param.long$depth)
-dryseason <- "on"
-root.selection <- "on"
-iso.subset <- "on"
-##
+  # load interval and working.iter
[…]

### 2020-04-05 code/4.1.1.GLUEsetup_part1_BCI.R

@@ -1,19 +1,21 @@
 # This file imports the data and sets up the 'Extra' variable
 
 rm(list=ls())
-gc()
+for (i in 1:5) {gc()}
 graphics.off()
 if (!require("pacman")) install.packages("pacman"); library(pacman)
 pacman::p_load(tidyverse, doParallel, foreach, data.table)
 
-GLUEsetup_part1 <- function(current.folder = "2019-10-14_5000", intervals = 5) {
+GLUEsetup_part1 <- function(current.folder = current.folder, intervals = intervals) {
+  # current.folder = "2019-10-14_5000"; intervals = 5
+  if(!dir.exists(file.path("results"))) {dir.create(file.path("results"))}
   if(!dir.exists(file.path("results", current.folder))) {dir.create(file.path("results", current.folder))}
 
   ####-------------------------------------
   #### 1. To enable hydro time series into intervals
   ####-------------------------------------
   ##
-  ## Storing census med and interval start and end dates--------------
+  # # Storing census med and interval start and end dates--------------
   # load("data-raw/CTFScensuses/bci.tree1.Rdata")
   # load("data-raw/CTFScensuses/bci.tree2.Rdata")
   # load("data-raw/CTFScensuses/bci.tree3.Rdata")
@@ -39,7 +41,6 @@ GLUEsetup_part1 <- function(current.folder = "2019-10-14_5000", intervals = 5) {
   # ggplot(bci.tree, aes(x = ExactDate)) +
   #   geom_histogram() +
   #   facet_wrap(~ census)
-  # library(dplyr)
   # mediandates <- summarise(bci.tree %>% group_by(census), mediandate = median(ExactDate, na.rm = T))
   # census.meds <- mediandates$mediandate
   # saveRDS(census.meds, "results/census.mediandates.rds") #-------
@@ -68,7 +69,7 @@ GLUEsetup_part1 <- function(current.folder = "2019-10-14_5000", intervals = 5) {
   ##-------------------------------------
   #root.nsam.long <- read.csv(file = file.path("data", current.folder, "params.root_fraction_by_depth_long.csv"), header = TRUE)
   # Generated from R/3.090_root_parameters_b.R
-  pro.df <- read.csv(file = file.path("results/root.profiles.long.csv"), header = TRUE)
[…]

### 2020-04-05 code/4.2.GLUErun_part2_func2_BCI.R

@@ -1,66 +1,0 @@
-## function to calculate rsq between observed growth vs. simulated SI matrix
-growth_by_si.func2 <- function(growth_by_si.info, goodness.fit, statistic, splevel){
-
-  if (splevel == "on"){
-    sp.with.tlp <- names(info.2$sp.si)
-    growth.meta <- growth_by_si.info$growth.meta %>%
-      subset(sp %in% sp.with.tlp)
-    sp_size.on  <- growth.meta$sp_size #.full %>% subset(size == "large" || size == "small") %>% select(-size, -sp, -interval.2) %>% as.matrix()
-    growth <- growth_by_si.info$growth[sp_size.on]
-    sp_size <- names(growth)
-    ### to create si for each sp_size, sp.si elements should be repeated as sps in growth
-    sp.in.sp_size <- stringr::str_split_fixed(names(growth), "_", 2)[,1]
-    sp.si.match <- info.2$sp.si[sp.in.sp_size] ## this repeats sp.si elements (with sp names) in the order of sp.in.sp_size
-    ## need to repeat elements of si according to growth[[j]]$interval
-    ## since intervals are from 3:7 (8-intervals), and so are corresponding si vec, this would mean removing 3-1 = 2
-    location.adj <- as.numeric(strsplit(colnames(sp.si.match[[1]]),".", fixed = TRUE)[[1]][2]) - 1
-    sp.si.match <- lapply(sp.si.match, unname)
-  } else {
-    growth <- growth_by_si.info$growth
-    sp_size <- names(growth)
-    location.adj <- as.numeric(strsplit(colnames(growth_by_si.info$si),".", fixed = TRUE)[[1]][2]) - 1
-    si <- unname(growth_by_si.info$si)
-  }
-  # if (growth_by_si.info$dbh.residuals == "on") {
-  #   growth <- lapply(growth, function (x) {
-  #     x %>% rename(growth = dbh.residuals)
-  #   })
-  # }
-  n.rf.sam <- length(unique(growth_by_si.info$si.param.rel$rf.sam))
-  n.par.sam <- length(unique(growth_by_si.info$si.param.rel$par.sam))
-  intervals <- info$intervals
-
-  ncor <- detectCores() - 1
-  cl <- parallel::makeForkCluster(ncor)
-  doParallel::registerDoParallel(cl)
-
[…]

### 2020-04-05 code/4.3.GLUErun_part2_BCI.R

@@ -7,54 +7,69 @@ pacman::p_load(tidyverse, doParallel, foreach)
 
 load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
 intervals <- info$intervals
-load("results/GLUEsetup_part1.5_BCI.RData")
+load("results/GLUEsetup_part1.2_BCI.RData")
+load("results/GLUEsetup_part1.3_BCI.RData")
 load(file.path("results/4.1GLUEsetup_part2_BCI.RData"))
 n.ensembles <- growth_by_si.info$n.ensembles
-source(file.path("R/4.2.GLUErun_part2_func_BCI.R"))
-source(file.path("R/4.2.GLUErun_part2_func2_BCI.R"))
-GLUE_run <- function(splevel, goodness.fit) {
+source(file.path("code/4.2.GLUErun_part2_func_BCI.R"))
+
+GLUE_run <- function(splevel, goodness.fit, drop.months) {
+
   if (splevel == "on") {
     level.folder <- "splevel"
   } else {
     level.folder <- "commlevel"
   }
+  if(!dir.exists(file.path("results", level.folder))) {dir.create(file.path("results", level.folder))}
+
   file.suffix <-
-    paste0(growth_by_si.info$si.type, "_", n.ensembles, "_", growth_by_si.info$growth.type, "_",
+    paste0("drop.months", drop.months, "_cor", goodness.fit, "_", growth_by_si.info$si.type, "_", n.ensembles, "_", growth_by_si.info$growth.type, "_",
            growth_by_si.info$growth.selection, "_",
            growth_by_si.info$dbh.residuals, "_",
-           info$intervals, "_cor", goodness.fit, ".Rdata")
-  # ###-----------------
-  # ###---------GLUE.rsq----
-  # ###-----------------
-  # beg <- Sys.time()
-  # beg
-  # GLUE.rsq <-
-  #   growth_by_si.func(
-  #     growth_by_si.info = growth_by_si.info,
-  #     statistic = "rsq",
-  #     splevel = splevel,
-  #     goodness.fit
-  #   )
-  # end = Sys.time()
-  # (end - beg) * nrow(growth_by_si.info$growth) / 60 / 60 / 100
-  # end - beg
-  # # 24 min on macbook
-  # # 26 secs at both levels ; 47.21 sec for parallel at lower levels # 10 n.ensembles & 12 growth
[…]

### 2020-04-05 code/4.5.prep.udi.calculator.R

@@ -7,7 +7,7 @@ gc()
 if (!require("pacman")) install.packages("pacman"); library(pacman)
 pacman::p_load(tidyverse, doParallel, foreach, data.table)
 
-prep.udi.calculator <- function(splevel) {
+prep.udi.calculator <- function(splevel, drop.months) {
   # load interval and n.ensembles
   load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
   load(file.path("results/4.1GLUEsetup_part2_BCI.RData")) # has n.ensembles and growth and si matrix
@@ -37,11 +37,14 @@ prep.udi.calculator <- function(splevel) {
   depth.thick <- data.frame(depth = as.numeric(unique(rdt$depth))) %>%
     mutate(thickness = depth - lag(depth, default = 0)) %>% data.table()
 
-  file.extension.base1 <- paste0(goodness.fit, "_", si.type, "_", n.ensembles, "_", growth.type, "_", growth.selection, "_", dbh.residuals, "_", intervals)
+  file.extension.base1 <- paste0("drop.months", drop.months, "_cor", goodness.fit, "_",
+                                 si.type, "_", n.ensembles, "_", growth.type, "_",
+                                 growth.selection, "_", dbh.residuals, "_", intervals)
 
   if (splevel == "on") {
-    load(file.path("results/sp.hydro.output.chosen.RData"))
-    sp.hydro.output.chosen <- lapply(sp.hydro.output.chosen, function(x){
+    load("results/sp.hydro.output.chosen.RData")
+    ## Taking hydro with the prescribed drop.months
+    sp.hydro.output.chosen <- lapply(sp.hydro.output.chosen[[drop.months]], function(x){
       data.table(x, key = "depth")[, date := as.factor(date)]
     })
     sdt.list <- lapply(sp.hydro.output.chosen, function(x){
@@ -49,49 +52,50 @@ prep.udi.calculator <- function(splevel) {
         paw.df, on = c("date","depth","par.sam"), `:=`(paw = i.paw)][,
                                                                      tw := paw*thickness*1000][, !c("paw", "thickness")], depth)
     })
-    save(sdt.list, file = paste0("results/splevel/sdt.list_cor", file.extension.base1, "_dryseason_off.Rdata"))
[…]

### 2020-04-05 code/4.6.udi.calculator.R

@@ -8,7 +8,8 @@ for (i in 1:5) {gc()}
 if (!require("pacman")) install.packages("pacman"); library(pacman)
 pacman::p_load(tidyverse, doParallel, foreach, data.table)
 
-udi.calculator <- function(splevel, dryseason, rsq.thresh, iso.subset) {
+udi.calculator <- function(splevel = splevel, dryseason = dryseason, rsq.thresh = rsq.thresh,
+                           n.ci.thresh = n.ci.thresh, iso.subset = iso.subset, drop.months = drop.months) {
   # load interval and n.ensembles
   load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
   load(file.path("results/4.1GLUEsetup_part2_BCI.RData")) # has n.ensembles and growth and si matrix
@@ -19,9 +20,9 @@ udi.calculator <- function(splevel, dryseason, rsq.thresh, iso.subset) {
   growth.selection <- growth_by_si.info$growth.selection
   dbh.residuals <- growth_by_si.info$dbh.residuals
   si.type <- growth_by_si.info$si.type
-  si.param.rel <- growth_by_si.info$si.param.rel
+  si.param.rel <- growth_by_si.info$si.param.rel[[drop.months]]
   goodness.fit <- 0.3 # rsq0.3
-  ncor <- detectCores() - 2
+  ncor <- detectCores() - 1
 
   if (splevel == "on") {
     level.folder <- "splevel"
@@ -35,35 +36,39 @@ udi.calculator <- function(splevel, dryseason, rsq.thresh, iso.subset) {
   n.rf.sam <- length(rf.sam.vec)
   rm(info); rm(growth_by_si.info)
   ## sdt is chosen based on dryseason on/off
-  file.extension.base1 <- paste0(goodness.fit, "_", si.type, "_", n.ensembles, "_", growth.type, "_", growth.selection, "_", dbh.residuals, "_", intervals)
+  file.extension.base1 <- paste0("drop.months", drop.months, "_cor", goodness.fit, "_", si.type, "_", n.ensembles, "_",
+                                 growth.type, "_", growth.selection, "_", dbh.residuals, "_",
+                                 intervals)
   file.extension.base2 <- paste0(file.extension.base1, "_dryseason_", dryseason)
   file.extension.base3 <- paste0(file.extension.base2, "_iso.subset_", iso.subset)
 
[…]

### 2020-04-11 Workflow.Rmd
added likelihood calculcation

@@
 source("code/4.3.GLUErun_part2_BCI.R")
-# splevel = "on"; goodness.fit = 0.3; drop.months = "Mar"
+# splevel = "off"; goodness.fit = 0.3; drop.months = "Mar"
+GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "None")
+GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "Jan")
 GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "Feb")
 GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "Mar")
+GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "JanFeb")
 GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "FebMar")
+GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "JanFebMar")
+GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "None")
+GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "Jan")
 GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "Feb")
 GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "Mar")
+GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "JanFeb")
 GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "FebMar")
+GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "JanFebMar")
 ```
@@
 # # splevel = "on"; dryseason = "off"; rsq.thresh = 0.5; root.selection = "exp"; iso.subset = "off"
-summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.5, root.selection = "exp", iso.subset = "on", drop.months = "Jan")
+summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.5, root.selection = "exp", iso.subset = "on", drop.months = "Mar")
 

### 2020-04-11 code/10.Comparison_with_Meinzer1999.R

@@ -47,7 +47,7 @@ iso.compare <- function(goodness.fit = goodness.fit,
   ds <- rbind(ds, ds.bestfit)
 
   ds <- ds %>% mutate(tlplevel = as.factor(tlplevel)) %>%
-    subset(!is.na(udi)) %>% droplevels()
+    subset(!is.na(udi.med.rsq)) %>% droplevels()
 
   iso <- read.csv("data-raw/traits/isotopes/Meinzer1999_Table1_Xylem_Sap_deltaD_Fig4.csv", na.strings = c("NA", ""), header = T, row.names = NULL, check.names = F)
 
@@ -91,8 +91,8 @@ iso.compare <- function(goodness.fit = goodness.fit,
   iso.sp <- iso$sp #[!is.na(iso$Xylem_sap_deltaD_permil)]
   save(iso.sp, file = "results/sp_with_isotopic_record.Rdata")
   write.csv(iso, file.path(paste0("data-raw/traits/isotopes/Meinzer1999_Table1_Xylem_Sap_deltaD_Fig4_sp_code.csv")), row.names = FALSE)
-  ### More data from March1997 from Fig 5B (includes data in Fig 5A)
 
+  ### More data from March1997 from Fig 5B (includes data in Fig 5A)
   iso.2.raw <- read.csv("data-raw/traits/isotopes/Meinzer1999_Xylem_Sap_deltaD_March97_DBH_Fig5B.csv", na.strings = c("NA", ""), header = T, row.names = NULL, check.names = F)
   iso.change <- read.csv("data-raw/traits/isotopes/Meinzer1999_Xylem_Sap_deltaD_by_delta_SapFlow_&_monthly_LeafFall_Fig7AB.csv", na.strings = c("NA", ""), header = T, row.names = NULL, check.names = F)
 
@@ -106,49 +106,51 @@ iso.compare <- function(goodness.fit = goodness.fit,
   iso.2 <- iso.2 %>% left_join(iso %>% select(sp, Phenology), by = "sp") %>%
     left_join(iso.change, by = "sp")
 
-  ####-----------
+  #### Plotting soil isotopic variation from Meinzer et al and Jackson et al -----------
   iso.soil.1 <- read.csv("data-raw/traits/isotopes/Meinzer1999_Fig2A_soil_deltaD_BCI_data_Mar_Apr_1997.csv", na.strings = c("NA",""), header = T, row.names = NULL, check.names = F)
[…]

### 2020-04-11 code/4.1.1.GLUEsetup_part1_BCI.R

@@ -155,6 +155,7 @@ GLUEsetup_part1 <- function(current.folder = current.folder, intervals = interva
   beg <- Sys.time()
   drop.months.vec <- list("Feb", "Mar", c("Feb", "Mar"))
   btran.nsam.int <- vector(mode = "list", length = length(drop.months.vec))
+  names(btran.nsam.int) <- lapply(drop.months.vec, paste, collapse = "")
   for (k in 1: length(drop.months.vec)){
     sdt.k <- sdt[, ':='(month = format(as.Date(date), "%b"))][!month %chin% drop.months.vec[[k]]]
     ## saving btran by nsam would be too big
@@ -172,7 +173,6 @@ GLUEsetup_part1 <- function(current.folder = current.folder, intervals = interva
     (end - beg)
     print(k)
   }
-  names(btran.nsam.int) <- lapply(drop.months.vec, paste, collapse = "")
   parallel::stopCluster(cl)
   # 4 min for root.nsam = 599
   # 1.15 hrs for nsam 10K n.best 100 1990-2018
@@ -186,6 +186,8 @@ GLUEsetup_part1 <- function(current.folder = current.folder, intervals = interva
   beg <- Sys.time()
   n.ensembles <- length(btran.nsam.int[[1]])
   btran.matrix <- si.param.rel <- vector(mode = "list", length = length(drop.months.vec))
+  names(btran.matrix) <- names(si.param.rel) <- lapply(drop.months.vec, paste, collapse = "")
+
   for (k in 1: length(drop.months.vec)){
     btran.mat.list <- lapply(btran.nsam.int[[k]], function(x) {
       btran.wide <- x %>% pivot_wider(names_from = interval, values_from = btran) %>% as.matrix()
@@ -201,13 +203,13 @@ GLUEsetup_part1 <- function(current.folder = current.folder, intervals = interva
                                     par.sam = rep(par.sam.vec, times = n.ensembles))
     print(k)
   }
-  names(btran.matrix) <- names(si.param.rel) <- lapply(drop.months.vec, paste, collapse = "")
 
   info.3 <- list(btran = btran.nsam.int,
                  btran.matrix = btran.matrix,
                  si.param.rel = si.param.rel,
                  drop.months.vec = drop.months.vec)
   save(file = "results/GLUEsetup_part1.3_BCI.RData", info.3)
+  # load("results/GLUEsetup_part1.3_BCI.RData")
 
[…]

### 2020-04-11 code/4.3.GLUErun_part2_BCI.R

@@ -30,36 +30,55 @@ GLUE_run <- function(splevel, goodness.fit, drop.months) {
   ###-----------------
   ###---------GLUE.rsq----
   ###-----------------
-  beg <- Sys.time()
-  beg
-  # Function from source(file.path("code/4.2.GLUErun_part2_func_BCI.R"))
-  GLUE.rsq <-
-    growth_by_si.func(
-      fun.version = "rsq",
-      growth_by_si.info = growth_by_si.info,
-      statistic = "rsq",
-      splevel = splevel,
-      goodness.fit = goodness.fit,
-      drop.months = drop.months
-    )
-  # fun.version = "rsq"; growth_by_si.info = growth_by_si.info; statistic = "rsq"; splevel = splevel;
-  # goodness.fit = goodness.fit; drop.months = drop.months
-  end = Sys.time()
-  (end - beg) * nrow(growth_by_si.info$growth) / 60 / 60 / 100
-  end - beg
-  # 24 min on macbook
-  # 26 secs at both levels ; 47.21 sec for parallel at lower levels # 10 n.ensembles & 12 growth
-  ## took 15 hours on the server for 1k n.ensembles and nrow(growth) = 158107
-  # takes 11 min on server for ~580 growth ts n.ensembles = 10000 n.best = 100
-  save(GLUE.rsq,
-       file = paste0("results/", level.folder, "/GLUE.rsq_", file.suffix))
-  rm(GLUE.rsq)
+  # beg <- Sys.time()
+  # beg
+  # # Function from source(file.path("code/4.2.GLUErun_part2_func_BCI.R"))
+  # GLUE.rsq <-
+  #   growth_by_si.func(
+  #     fun.version = "rsq",
+  #     growth_by_si.info = growth_by_si.info,
+  #     statistic = "rsq",
+  #     splevel = splevel,
+  #     goodness.fit = goodness.fit,
+  #     drop.months = drop.months
+  #   )
+  # # fun.version = "rsq"; growth_by_si.info = growth_by_si.info; statistic = "rsq"; splevel = splevel;
+  # # goodness.fit = goodness.fit; drop.months = drop.months
+  # end = Sys.time()
+  # (end - beg) * nrow(growth_by_si.info$growth) / 60 / 60 / 100
+  # end - beg
+  # # 24 min on macbook
+  # # 26 secs at both levels ; 47.21 sec for parallel at lower levels # 10 n.ensembles & 12 growth
+  # ## took 15 hours on the server for 1k n.ensembles and nrow(growth) = 158107
[…]

### 2020-04-18 Workflow.Rmd
Renamed files to sort in the sequence they are used in the workflow

--- 2020-04-02 Workflow.Rmd
@@
 date: "`r format(Sys.time(), '%d %B, %Y')`"
-output: html_document
+output: github_document
 ---
@@
 ```{r setup, include = FALSE}
-knitr::opts_chunk$set(echo = TRUE, include = TRUE)
+knitr::opts_chunk$set(echo = TRUE, include = TRUE, eval = FALSE)
 ```
@@
 
-### 1.GLUEsetup_part1_BCI.R
+## Prerequisite data
+
+_*Growth data*_
+
+_*Water Availability*_
+
+ Soil Water Potential data -- best-fit 100 ensembles from ELM-FATES during the study period.
+ This is sourced from data-package `bci.elm.fates.hydro`
+
+### Gather all pre-requisite data in a place or two
+
+`GLUEsetup_part1() in 1.GLUEsetup_part1_BCI.R`
 
@@
  * Loads tree growth census dates for specific no. of intervals
- * Calculates btran time-series by census-interval for each root profile for all the best.fit SWP ensemble
+ * Converts swp into a growth factor (range 0-1) using a step-fuction, such that growth factor is 0 at sp-level or community mean level Turgor Loss Point.
+Species wise growth factor is stores separately (sp.hydro.output.chosen)
+
+Then using a combination of a range of hypothetical root profiles obtained from 2.rooting_profiles.R and SWP based growth factor ensemble simulations generate a btran time-series using equation : Btran= i=1 to i=z RootFrac_z * f(swp_z),
+where f(swp_z) is a growth.factor_z based on a linear step relationship with swp_z ap species specific or community mean TLP level.
+
+ * Calculates btran time-series by census-interval for each rooting profile for all the best.fit SWP ensemble
   + by individual species with species-specific TLP (N of btran time series == No. of species x No. of SWP ensembles x No. of root profiles)
   + at community-mean TLP (No. of btran time series == No. of SWP ensembles x No. of root profiles)
-  * Saves all the above info in a list called info, except species level btran is saved in info.2
-
-## Then using a combination of a range of hypothetical root profiles obtained from 2.rooting_profiles.R and SWP ensemble simulations generate a btran time-series using equation : Btran= i=1 to i=z RootFrac_z * f(swp_z),
-where f(swp_z) is a growth.factor_z based on a linear step relationship with swp_z ap species specific or community mean TLP level.
-Code/6.0_Growth_correlate_table_from_swp.R
+  + both community and species level btran is calculated by dropping none or one or more number of dry season months as sublists
+  * Saves ELM-FATES hydro output, rooting profiles, census dates, number of intervals in a list called info, community level btran in a list named info.3, and species level btran is saved in info.2, along with sub-list names of dropped months
 
 ```{r Chunk 1.0, eval = FALSE}
-source("code/1.GLUEsetup_part1_BCI.R")
+source("code/4.1.1.GLUEsetup_part1_BCI.R")
 GLUEsetup_part1(current.folder = "2019-10-14_5000", intervals = 5) # this can take 30 min to get btran by nsam
@@
 
+### GLUEsetup_part2()
+
+This function stores together growth data, growth meta-data, and btran data (at community mean TLP level by dropped months) in a list named growth_by_si.info.
+
 ```{r Chunk 2, eval = FALSE}
-source("code/4.1.GLUEsetup_part2_BCI.R")
-GLUEsetup_part2(growth.type = "med", dbh.residuals = "off", solar.residuals = "off", 
+source("code/4.1.2.GLUEsetup_part2_BCI.R")
+GLUEsetup_part2(growth.type = "med", dbh.residuals = "on", solar.residuals = "off", 
                 growth.selection = "size_class_predefined_cc_scaled") #"size_class_varying_non_cc") #n.ensembles = 100000
-# growth.type = "med"; dbh.residuals = "off"; solar.residuals = "off"; growth.selection = "size_class_predefined_cc_scaled"
+# growth.type = "med"; dbh.residuals = "on"; solar.residuals = "off"; growth.selection = "size_class_predefined_cc_scaled"
 ```
 
-```{r Chunk 3, eval = TRUE}
+###GLUE_run()
+
+This function generates a few statistics for the fit between growth and btran:
+
+Rsq, negative likelihood and the number of data points of growth factor that pass through 95% confidence interval of growth.
+TLP level and any months to drop while calculating growth factor need be specified.
+Threshold (goodness.fit) as to when to calculate Rs value also needs to be specified.
+
+```{r Chunk 3, eval = FALSE}
 # This sources "code/4.2GLUErun_part2_func_BCI.R"
 source("code/4.3.GLUErun_part2_BCI.R")
-GLUE_run(splevel = "off", goodness.fit = 0.3)
-# GLUE_run(splevel = "on", goodness.fit = 0.3)
+# splevel = "on"; goodness.fit = 0.3; drop.months = "None"
+GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "None")
+GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "Jan")
+GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "Feb")
+GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "Mar")
+GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "JanFeb")
+GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "FebMar")
+GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "JanFebMar")
+GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "None")
+GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "Jan")
+GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "Feb")
+GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "Mar")
+GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "JanFeb")
+GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "FebMar")
+GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "JanFebMar")
 ```
 
-<!-- ```{r Chunk 2.4.0, eval = TRUE} -->
+<!-- ```{r Chunk 2.4.0, eval = FALSE} -->
 <!-- source("code/4.4.Best-fit_param_BCI.R") -->
@@
 <!-- ``` -->
+### prep.udi.calculator()
 
-```{r Chunk 4, eval = TRUE}
+This subsets SWP-based growth factor data (species-specific or community level) to various specified time period or months: late dryseason (Mar Apr) 1997, Jan 1997, dryseason 1992 etc.
+For example, restricted to early and late 1997 dry season to later calculate UDI only for this period to compare with isotopic tree Sylem sap data that is collected in early and later 1997.
+
+```{r Chunk 4, eval = FALSE}
 source("code/4.5.prep.udi.calculator.R")
-prep.udi.calculator(splevel = "on")
-prep.udi.calculator(splevel = "off")
+# splevel = "on"; drop.months = "JanFebMar"
+prep.udi.calculator(splevel = "on", drop.months = "Feb")
+prep.udi.calculator(splevel = "on", drop.months = "FebMar")
+prep.udi.calculator(splevel = "on", drop.months = "Mar")
+
+prep.udi.calculator(splevel = "off", drop.months = "Feb")
+prep.udi.calculator(splevel = "off", drop.months = "FebMar")
+prep.udi.calculator(splevel = "off", drop.months = "Mar")
+prep.udi.calculator(splevel = "off", drop.months = "None")
+prep.udi.calculator(splevel = "off", drop.months = "Jan")
+prep.udi.calculator(splevel = "off", drop.months = "JanFeb")
+prep.udi.calculator(splevel = "off", drop.months = "JanFebMar")
 ```
 
-```{r Chunk 5, eval = TRUE}
+### udi.calculator() calculates Uptake Depth Index
+
+This calculates Uptake Depth Index and SDI (water Stress Depth Index, without taking into account total water availability) for each simulation and stores corresponding values of model Rsq, negative loLikelihood, number of data points that pass through confidence interval of growth, species and size in one data-frame named "ds.bestfit.all.Rdata"
+
+UDI or SDI is only calculated for a simulation for which R-sq is greater than the given therehold (e.g.
+rsq.thresh = 0.3) and if at least 3 data-points have passed through 95% CI of growth data calculated based on (e.g.
+n.ci.thresh = 3)
+
+```{r Chunk 5, eval = FALSE}
 source("code/4.6.udi.calculator.R")
@@
 # udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.5, iso.subset = "on") ## udi is not calculated below this rsq.thresh
-# # splevel = "on"; dryseason = "off"; rsq.thresh = 0.5; iso.subset = "on"
-# udi.calculator(splevel = "on", dryseason = "off", rsq.thresh = 0.5, iso.subset = "on") 
-# udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.5, iso.subset = "on")
-udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.5, iso.subset = "on")
-udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.5, iso.subset = "on")
-# udi.calculator(splevel = "off", dryseason = "1992", rsq.thresh = 0.5, iso.subset = "off") 
-# udi.calculator(splevel = "on", dryseason = "1992", rsq.thresh = 0.5, iso.subset = "off")
+# splevel = "on"; dryseason = "on"; rsq.thresh = 0.5;  n.ci.thresh = 3; iso.subset = "on"; drop.months = "JanFebMar"
+udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.3,
+               n.ci.thresh = 3, iso.subset = "on", drop.months = "None")
+udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.3,
+               n.ci.thresh = 3, iso.subset = "on", drop.months = "Jan")
+udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.3,
+               n.ci.thresh = 3, iso.subset = "on", drop.months = "Feb")
+udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.3,
+               n.ci.thresh = 3, iso.subset = "on", drop.months = "Mar")
+udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.3,
+               n.ci.thresh = 3, iso.subset = "on", drop.months = "JanFeb")
+udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.3,
+               n.ci.thresh = 3, iso.subset = "on", drop.months = "FebMar")
+udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.3,
+               n.ci.thresh = 3, iso.subset = "on", drop.months = "JanFebMar")
+
+udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, 
+               n.ci.thresh = 3, iso.subset = "on", drop.months = "None")
+udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, 
+               n.ci.thresh = 3, iso.subset = "on", drop.months = "Jan")
+udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.3,
+               n.ci.thresh = 3, iso.subset = "on", drop.months = "Feb")
+udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.3,
+               n.ci.thresh = 3, iso.subset = "on", drop.months = "Mar")
+udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, 
+               n.ci.thresh = 3, iso.subset = "on", drop.months = "JanFeb")
+udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.3,
+               n.ci.thresh = 3, iso.subset = "on", drop.months = "FebMar")
+udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, 
+               n.ci.thresh = 3, iso.subset = "on", drop.months = "JanFebMar")
 ```
+
+## Summarise Uptake Depth Index by species
+
+UDI corresponding to model with maximum Rsq, minimum negative likelihood 
+median, sd, se of UDI for top 10 rsq or bottom 10 negative likeliood, that are also greater than the Rsq threshold.
 
@@
 source("code/4.7.summarising_udi_by_sp_size.R")
-# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.5, root.selection = "on", iso.subset = "on")
-# # splevel = "on"; dryseason = "off"; rsq.thresh = 0.5; root.selection = "on"; iso.subset = "off"
-# summarise.udi(splevel = "on", dryseason = "off", rsq.thresh = 0.5, root.selection = "on", iso.subset = "on")
-# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.5, root.selection = "on", iso.subset = "on")
-# summarise.udi(splevel = "off", dryseason = "off", rsq.thresh = 0.5, root.selection = "on", iso.subset = "on")
-summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.5, root.selection = "on", iso.subset = "on")
-summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.5, root.selection = "on", iso.subset = "on")
+# splevel = "on"; dryseason = "on"; rsq.thresh = 0.3; root.selection = "exp"; iso.subset = "on"; drop.months = "None"
+# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "None")
+# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "Jan")
+# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "Feb")
+# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "Mar")
+# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "JanFeb")
+# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "FebMar")
+# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "JanFebMar")
+# 
+# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "None")
+# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "Jan")
+# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "Feb")
+# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "Mar")
+# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "JanFeb")
+# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "FebMar")
+# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "JanFebMar")
+
+summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "None")
+summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "Jan")
+summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "Feb")
+summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "Mar")
+summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "JanFeb")
+summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "FebMar")
+summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "JanFebMar")
+
+summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "None")
+summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "Jan")
+summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "Feb")
+summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "Mar")
+summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "JanFeb")
+summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "FebMar")
+summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "JanFebMar")
 ```
 
-```{r Chunk 7, eval = TRUE}
-source("code/10.Comparison_with_Meinzer1999.R")
+## Sensitivity of UDI of deciduous species to inclusion of deciduous months
+
+```{r Chunk 7, eval = FALSE}
+source("code/4.8.Sensitivity analysis for UDI of deciduous species.R")
 ```
@@
 ```{r Chunk 8, eval = FALSE}
+source("code/10.Comparison_with_Meinzer1999.R")
+iso.compare(goodness.fit = 0.3, dryseason = "on", 
+            root.selection = "exp", iso.subset = "on", drop.months = "None")
+iso.compare(goodness.fit = 0.3, dryseason = "on", 
+            root.selection = "exp", iso.subset = "on", drop.months = "Jan")
+iso.compare(goodness.fit = 0.3, dryseason = "on", 
+            root.selection = "exp", iso.subset = "on", drop.months = "Feb")
+iso.compare(goodness.fit = 0.3, dryseason = "on", 
+            root.selection = "exp", iso.subset = "on", drop.months = "Mar")
+iso.compare(goodness.fit = 0.3, dryseason = "on", 
+            root.selection = "exp", iso.subset = "on", drop.months = "JanFeb")
+iso.compare(goodness.fit = 0.3, dryseason = "on", 
+            root.selection = "exp", iso.subset = "on", drop.months = "FebMar")
+iso.compare(goodness.fit = 0.3, dryseason = "on", 
+            root.selection = "exp", iso.subset = "on", drop.months = "JanFebMar")
+# goodness.fit = 0.3; dryseason = "on"; root.selection = "exp"; iso.subset = "on"; drop.months = "None"
+```
+
+```{r Chunk 9, eval = FALSE}
 source("code/9.6_mortality_BCI.R")
@@
 
-```{r Chunk 9, eval = FALSE}
+```{r Chunk 10, eval = FALSE}
 source("code/5.species_graphs_BCI.R")

### 2020-04-19 Workflow.Rmd
Updated file number prefix to scripts. Aslo for GLUEsetup_part2_BCI.R, but removed corresponding numbered prefix to GLUEsetup_part2_BCI.Rdata

@@
 
-```{r Chunk 1.0, eval = FALSE}
-source("code/4.1.1.GLUEsetup_part1_BCI.R")
+```{r Chunk 1.0, eval = TRUE}
+source("code/03.0_GLUEsetup_part1_BCI.R")
 GLUEsetup_part1(current.folder = "2019-10-14_5000", intervals = 5) # this can take 30 min to get btran by nsam
@@
 
-```{r Chunk 2, eval = FALSE}
-source("code/4.1.2.GLUEsetup_part2_BCI.R")
+```{r Chunk 2, eval = TRUE}
+source("code/04.0_GLUEsetup_part2_BCI.R")
 GLUEsetup_part2(growth.type = "med", dbh.residuals = "on", solar.residuals = "off", 
@@
 
-```{r Chunk 3, eval = FALSE}
+```{r Chunk 3, eval = TRUE}
 # This sources "code/4.2GLUErun_part2_func_BCI.R"
-source("code/4.3.GLUErun_part2_BCI.R")
+source("code/05.1_GLUErun_part2_BCI.R")
 # splevel = "on"; goodness.fit = 0.3; drop.months = "None"
@@
 ```{r Chunk 4, eval = FALSE}
-source("code/4.5.prep.udi.calculator.R")
+source("code/06.0_prep.udi.calculator.R")
 # splevel = "on"; drop.months = "JanFebMar"
@@
 ```{r Chunk 5, eval = FALSE}
-source("code/4.6.udi.calculator.R")
+source("code/06.1_udi.calculator.R")
 ##---needs to be run manually on macbook for splevel = "off" and iso.subset = "off"---
@@
 ```{r Chunk 6, eval = TRUE}
-source("code/4.7.summarising_udi_by_sp_size.R")
-# splevel = "on"; dryseason = "on"; rsq.thresh = 0.3; root.selection = "exp"; iso.subset = "on"; drop.months = "None"
-# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "None")
-# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "Jan")
-# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "Feb")
-# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "Mar")
-# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "JanFeb")
-# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "FebMar")
-# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "JanFebMar")
+source("code/07.0_summarising_udi_by_sp_size.R")
+# splevel = "off"; dryseason = "on"; rsq.thresh = 0.3; root.selection = "exp"; iso.subset = "off"; drop.months = "None"; n.rank = 1000
+# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 100, drop.months = "None")
+# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 100, drop.months = "Jan")
+# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 100, drop.months = "Feb")
+# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 100, drop.months = "Mar")
+# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 100, drop.months = "JanFeb")
+# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 100, drop.months = "FebMar")
+# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 100, drop.months = "JanFebMar")
 # 
-# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "None")
-# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "Jan")
-# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "Feb")
-# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "Mar")
-# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "JanFeb")
-# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "FebMar")
-# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "JanFebMar")
+# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 100, drop.months = "None")
+# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 100, drop.months = "Jan")
+# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 100, drop.months = "Feb")
+# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 100, drop.months = "Mar")
+# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 100, drop.months = "JanFeb")
+# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 100, drop.months = "FebMar")
+# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 100, drop.months = "JanFebMar")
 
-summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "None")
-summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "Jan")
-summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "Feb")
-summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "Mar")
-summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "JanFeb")
-summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "FebMar")
-summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "JanFebMar")
+summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "None")
+summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "Jan")
+summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "Feb")
+summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "Mar")
+summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "JanFeb")
+summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "FebMar")
+summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "JanFebMar")
 
-summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "None")
-summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "Jan")
-summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "Feb")
-summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "Mar")
-summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "JanFeb")
-summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "FebMar")
-summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "JanFebMar")
+summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "None")
+summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "Jan")
+summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "Feb")
+summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "Mar")
+summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "JanFeb")
+summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "FebMar")
+summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "JanFebMar")
+
+summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "None")
+summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "Jan")
+summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "Feb")
+summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "Mar")
+summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "JanFeb")
+summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "FebMar")
+summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "JanFebMar")
+
+summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "None")
+summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "Jan")
+summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "Feb")
+summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "Mar")
+summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "JanFeb")
+summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "FebMar")
+summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "JanFebMar")
 ```
@@
 ```{r Chunk 7, eval = FALSE}
-source("code/4.8.Sensitivity analysis for UDI of deciduous species.R")
+source("code/08.0_Sensitivity analysis for UDI of deciduous species.R")
 ```
@@
 ```{r Chunk 8, eval = FALSE}
-source("code/10.Comparison_with_Meinzer1999.R")
+source("code/09.0_Comparison_with_Meinzer1999.R")
 iso.compare(goodness.fit = 0.3, dryseason = "on", 
@@
 ```{r Chunk 9, eval = FALSE}
-source("code/9.6_mortality_BCI.R")
+source("code/11.1_mortality_BCI.R")
 demo.graphs(level.folder = "commlevel", n.threshold = 30, dbh.residuals = "on", 

### 2020-04-19 code/05.1_GLUErun_part2_BCI.R

@@ -17,9 +17,9 @@ load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
 intervals <- info$intervals
 load("results/GLUEsetup_part1.2_BCI.RData")
 load("results/GLUEsetup_part1.3_BCI.RData")
-load(file.path("results/4.1GLUEsetup_part2_BCI.RData"))
+load(file.path("results/GLUEsetup_part2_BCI.RData"))
 n.ensembles <- growth_by_si.info$n.ensembles
-source(file.path("code/4.2.GLUErun_part2_func_BCI.R"))
+source(file.path("code/05.0_GLUErun_part2_func_BCI.R"))
 
 GLUE_run <- function(splevel, goodness.fit, drop.months) {
 

### 2020-04-19 code/06.0_prep.udi.calculator.R

@@ -15,7 +15,7 @@ pacman::p_load(tidyverse, doParallel, foreach, data.table)
 prep.udi.calculator <- function(splevel, drop.months) {
   # load interval and n.ensembles
   load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
-  load(file.path("results/4.1GLUEsetup_part2_BCI.RData")) # has n.ensembles and growth and si matrix
+  load(file.path("results/GLUEsetup_part2_BCI.RData")) # has n.ensembles and growth and si matrix
 
   intervals <- info$intervals
   n.ensembles <- growth_by_si.info$n.ensembles

### 2020-04-19 code/06.1_udi.calculator.R

@@ -17,7 +17,7 @@ udi.calculator <- function(splevel = splevel, dryseason = dryseason, rsq.thresh 
                            n.ci.thresh = n.ci.thresh, iso.subset = iso.subset, drop.months = drop.months) {
   # load interval and n.ensembles
   load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
-  load(file.path("results/4.1GLUEsetup_part2_BCI.RData")) # has n.ensembles and growth and si matrix
+  load(file.path("results/GLUEsetup_part2_BCI.RData")) # has n.ensembles and growth and si matrix
 
   intervals <- info$intervals
   n.ensembles <- growth_by_si.info$n.ensembles

### 2020-04-19 code/07.0_summarising_udi_by_sp_size.R

@@ -14,7 +14,7 @@ summarise.udi <- function(splevel = splevel, dryseason = dryseason, rsq.thresh =
                           root.selection = root.selection, iso.subset = iso.subset,
                           n.rank = n.rank, drop.months = drop.months) {
   load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
-  load(file.path("results/4.1GLUEsetup_part2_BCI.RData")) # has n.ensembles and growth and si matrix
+  load(file.path("results/GLUEsetup_part2_BCI.RData")) # has n.ensembles and growth and si matrix
 
   intervals <- info$intervals
   rm(info)
@@ -69,7 +69,6 @@ summarise.udi <- function(splevel = splevel, dryseason = dryseason, rsq.thresh =
            root.95.tops.ll = ifelse(ll.rank > n.rank, NA, root.95),
            root.75.tops.ll = ifelse(ll.rank > n.rank, NA, root.75),
            max.root.tops.ll = ifelse(ll.rank > n.rank, NA, max.root)) %>%
-      subset(!is.na(rsq.tops.rsq)) %>%
       arrange(sp_size, ll.rank, rsq.rank)
     ds.bestfit <- ds.bestfit.select %>%
     summarise(udi.best.rsq = mean(udi.best.rsq, na.rm = TRUE),

### 2020-04-19 code/09.0_Comparison_with_Meinzer1999.R

@@ -25,7 +25,7 @@ iso.compare <- function(goodness.fit = goodness.fit,
 
   # load interval and working.iter
   load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
-  load(file.path("results/4.1GLUEsetup_part2_BCI.RData")) # has working.iter and growth and si matrix
+  load(file.path("results/GLUEsetup_part2_BCI.RData")) # has working.iter and growth and si matrix
 
   soil.depths <- unique(info$root.param.long$depth)
   intervals <- info$intervals

### 2020-04-19 code/11.1_mortality_BCI.R

@@ -55,7 +55,7 @@ demo.graphs <- function(level.folder = level.folder, n.threshold = n.threshold,
 
   ## adding uptake depth index
   load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
-  load(file.path("results/4.1GLUEsetup_part2_BCI.RData")) # has n.ensembles and growth and si matrix
+  load(file.path("results/GLUEsetup_part2_BCI.RData")) # has n.ensembles and growth and si matrix
 
   intervals <- info$intervals
   n.ensembles <- growth_by_si.info$n.ensembles

### 2020-04-19 code/12.0_hydraulic_traits_BCI.R

@@ -27,7 +27,7 @@ rev_sqrt_trans <- function() {
 }
 ## adding uptake depth index
 load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
-load("results/4.1GLUEsetup_part2_BCI.RData") # has n.ensembles and growth and si matrix
+load("results/GLUEsetup_part2_BCI.RData") # has n.ensembles and growth and si matrix
 load("results/demo.sp_size.RData")
 load("results/demo.sp.RData") ## Mortality rate only for those cases for which avg.abundance greater than 10
 intervals <- info$intervals

### 2020-04-19 code/13.0_tree_habitat_vs_UDI.R

@@ -34,7 +34,7 @@ if(!dir.exists(file.path.spatial)) {dir.create(file.path.spatial)}
 
 # load interval and working.iter
 load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
-load(file.path("results/4.1GLUEsetup_part2_BCI.RData")) # has working.iter and growth and si matrix
+load(file.path("results/GLUEsetup_part2_BCI.RData")) # has working.iter and growth and si matrix
 
 intervals <- info$intervals
 n.ensembles <- growth_by_si.info$n.ensembles

### 2020-04-19 code/14.0_spatial_hydrological_niche.R

@@ -36,7 +36,7 @@ if(!dir.exists(file.path.spatial)) {dir.create(file.path.spatial)}
 load("data-raw/CTFScensuses/BCI.tree8.Rdata")
 # load interval and working.iter
 load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
-load(file.path("results/4.1GLUEsetup_part2_BCI.RData")) # has working.iter and growth and si matrix
+load(file.path("results/GLUEsetup_part2_BCI.RData")) # has working.iter and growth and si matrix
 
 intervals <- info$intervals
 n.ensembles <- growth_by_si.info$n.ensembles

### 2020-04-20 Workflow.Rmd
Formatted

--- 2020-04-18 Workflow.Rmd
@@
 date: "`r format(Sys.time(), '%d %B, %Y')`"
-output: github_document
+output: 
+  github_document
 ---
@@
 
-## How is this project set-up
+# Model Setup
 
-## Prerequisite data
+#### Pre-requisite data
 
-_*Growth data*_
+*Growth data*
 
-_*Water Availability*_
+```{r Chunk 01.0, eval = FALSE}
+source("code/01.0_Preparing_stem_growth_time_series.R")
+```
+
+*Water Availability*
 
 Soil Water Potential data -- best-fit 100 ensembles from ELM-FATES during the study period.
- This is sourced from data-package `bci.elm.fates.hydro`
+This is sourced from data-package `bci.elm.fates.hydro` with library(bci.elm.fates.hydro)
 
-### Gather all pre-requisite data in a place or two
+#### Potential rooting profile trait-space
 
-`GLUEsetup_part1() in 1.GLUEsetup_part1_BCI.R`
+```{r Chunk 02.0, eval = FALSE}
+source("code/02.0_Rooting_profiles.R")
+```
+
+#### Calculate btran and organise
+
+Gather all pre-requisites in a place or two
+
+1. `GLUEsetup_part1() in 03.0_GLUEsetup_part1_BCI.R`
 
@@
 
-```{r Chunk 1.0, eval = FALSE}
-source("code/4.1.1.GLUEsetup_part1_BCI.R")
-GLUEsetup_part1(current.folder = "2019-10-14_5000", intervals = 5) # this can take 30 min to get btran by nsam
+```{r Chunk 03.0, eval = FALSE}
+source("code/03.0_GLUEsetup_part1_BCI.R")
+GLUEsetup_part1(current.folder = "2019-10-14_5000", intervals = 5)
 ```
 
-### GLUEsetup_part2()
+2. `GLUEsetup_part2() in 04.0_GLUEsetup_part2_BCI.R`
 
@@
 
-```{r Chunk 2, eval = FALSE}
-source("code/4.1.2.GLUEsetup_part2_BCI.R")
+```{r Chunk 04.0, eval = FALSE}
+source("code/04.0_GLUEsetup_part2_BCI.R")
 GLUEsetup_part2(growth.type = "med", dbh.residuals = "on", solar.residuals = "off", 
@@
 
-###GLUE_run()
+# Model Parameterisation
+
+#### Run model for all ensembles
+
+`GLUE_run() in 05.1_GLUErun_part2_BCI.R`
 
@@
 
-Rsq, negative likelihood and the number of data points of growth factor that pass through 95% confidence interval of growth.
+Rsq, residuals, sum of negative log likelihood and the number of data points of growth factor that pass through 95% confidence interval of growth.
 TLP level and any months to drop while calculating growth factor need be specified.
@@
 
-```{r Chunk 3, eval = FALSE}
+```{r Chunk 05.1, eval = FALSE}
 # This sources "code/4.2GLUErun_part2_func_BCI.R"
-source("code/4.3.GLUErun_part2_BCI.R")
+source("code/05.1_GLUErun_part2_BCI.R")
 # splevel = "on"; goodness.fit = 0.3; drop.months = "None"
@@
 
-<!-- ```{r Chunk 2.4.0, eval = FALSE} -->
-<!-- source("code/4.4.Best-fit_param_BCI.R") -->
-<!-- # bestfit.param(splevel = "on") -->
-<!-- bestfit.param(splevel = "off") -->
-<!-- ``` -->
-### prep.udi.calculator()
+### Get water availability and btran during a specific period
+
+`prep.udi.calculator() in 06.0_prep.udi.calculator.R`
 
 This subsets SWP-based growth factor data (species-specific or community level) to various specified time period or months: late dryseason (Mar Apr) 1997, Jan 1997, dryseason 1992 etc.
-For example, restricted to early and late 1997 dry season to later calculate UDI only for this period to compare with isotopic tree Sylem sap data that is collected in early and later 1997.
+For example, restricted to early and late 1997 dry season to later calculate UDI only for this period to compare with isotopic tree Xylem sap data that is collected in early and later 1997.
 
-```{r Chunk 4, eval = FALSE}
-source("code/4.5.prep.udi.calculator.R")
+```{r Chunk 06.0, eval = FALSE}
+source("code/06.0_prep.udi.calculator.R")
 # splevel = "on"; drop.months = "JanFebMar"
@@
 
-### udi.calculator() calculates Uptake Depth Index
+#### Calculate water Uptake Depth Index for a specific period
+
+`udi.calculator() in 06.1_udi.calculator.R`
 
@@
 
-```{r Chunk 5, eval = FALSE}
-source("code/4.6.udi.calculator.R")
+```{r Chunk 06.1, eval = FALSE}
+source("code/06.1_udi.calculator.R")
 ##---needs to be run manually on macbook for splevel = "off" and iso.subset = "off"---
@@
 # splevel = "on"; dryseason = "on"; rsq.thresh = 0.5;  n.ci.thresh = 3; iso.subset = "on"; drop.months = "JanFebMar"
-udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.3,
+udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.3,
                n.ci.thresh = 3, iso.subset = "on", drop.months = "None")
-udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.3,
+udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.3,
                n.ci.thresh = 3, iso.subset = "on", drop.months = "Jan")
-udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.3,
+udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.3,
                n.ci.thresh = 3, iso.subset = "on", drop.months = "Feb")
-udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.3,
+udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.3,
                n.ci.thresh = 3, iso.subset = "on", drop.months = "Mar")
-udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.3,
+udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.3,
                n.ci.thresh = 3, iso.subset = "on", drop.months = "JanFeb")
-udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.3,
+udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.3,
                n.ci.thresh = 3, iso.subset = "on", drop.months = "FebMar")
-udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.3,
+udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.3,
                n.ci.thresh = 3, iso.subset = "on", drop.months = "JanFebMar")
 
-udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, 
+udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.3, 
                n.ci.thresh = 3, iso.subset = "on", drop.months = "None")
-udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, 
+udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.3, 
                n.ci.thresh = 3, iso.subset = "on", drop.months = "Jan")
-udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.3,
+udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.3,
                n.ci.thresh = 3, iso.subset = "on", drop.months = "Feb")
-udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.3,
+udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.3,
                n.ci.thresh = 3, iso.subset = "on", drop.months = "Mar")
-udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, 
+udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.3, 
                n.ci.thresh = 3, iso.subset = "on", drop.months = "JanFeb")
-udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.3,
+udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.3,
                n.ci.thresh = 3, iso.subset = "on", drop.months = "FebMar")
-udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, 
+udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.3, 
                n.ci.thresh = 3, iso.subset = "on", drop.months = "JanFebMar")
@@
 
-## Summarise Uptake Depth Index by species
+#### Summarise Uptake Depth Index by species for best-fit models
 
-UDI corresponding to model with maximum Rsq, minimum negative likelihood 
-median, sd, se of UDI for top 10 rsq or bottom 10 negative likeliood, that are also greater than the Rsq threshold.
+UDI corresponding to model with maximum Rsq, likelihood 
+median, upr & lwr confidence interval for UDI for top ranking models
 
-```{r Chunk 6, eval = TRUE}
-source("code/4.7.summarising_udi_by_sp_size.R")
-# splevel = "on"; dryseason = "on"; rsq.thresh = 0.3; root.selection = "exp"; iso.subset = "on"; drop.months = "None"
-# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "None")
-# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "Jan")
-# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "Feb")
-# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "Mar")
-# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "JanFeb")
-# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "FebMar")
-# summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "JanFebMar")
+```{r Chunk 07.0, eval = FALSE}
+source("code/07.0_summarising_udi_by_sp_size.R")
+# splevel = "off"; dryseason = "jan"; rsq.thresh = 0.3; root.selection = "exp"; iso.subset = "on"; drop.months = "None"; n.rank = 1000
+
+summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 100, drop.months = "None")
+summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 100, drop.months = "Jan")
+summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 100, drop.months = "Feb")
+summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 100, drop.months = "Mar")
+summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 100, drop.months = "JanFeb")
+summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 100, drop.months = "FebMar")
+summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 100, drop.months = "JanFebMar")
+
+summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 100, drop.months = "None")
+summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 100, drop.months = "Jan")
+summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 100, drop.months = "Feb")
+summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 100, drop.months = "Mar")
+summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 100, drop.months = "JanFeb")
+summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 100, drop.months = "FebMar")
+summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 100, drop.months = "JanFebMar")
+
+# summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "None")
+# summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "Jan")
+# summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "Feb")
+# summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "Mar")
+# summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "JanFeb")
+# summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "FebMar")
+# summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "JanFebMar")
 # 
-# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "None")
-# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "Jan")
-# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "Feb")
-# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "Mar")
-# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "JanFeb")
-# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "FebMar")
-# summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "on", n.rank = 50, drop.months = "JanFebMar")
-
-summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "None")
-summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "Jan")
-summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "Feb")
-summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "Mar")
-summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "JanFeb")
-summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "FebMar")
-summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "JanFebMar")
-
-summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "None")
-summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "Jan")
-summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "Feb")
-summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "Mar")
-summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "JanFeb")
-summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "FebMar")
-summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 1000, drop.months = "JanFebMar")
+# summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "None")
+# summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "Jan")
+# summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "Feb")
+# summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "Mar")
+# summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "JanFeb")
+# summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "FebMar")
+# summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, root.selection = "exp", iso.subset = "off", n.rank = 100, drop.months = "JanFebMar")
 ```
 
-## Sensitivity of UDI of deciduous species to inclusion of deciduous months
+# Evaluation
 
-```{r Chunk 7, eval = FALSE}
-source("code/4.8.Sensitivity analysis for UDI of deciduous species.R")
+#### Sensitivity of UDI of deciduous species to inclusion of deciduous months
+
+```{r Chunk 08.0, eval = FALSE}
+source("code/08.0_Sensitivity analysis for UDI of deciduous species.R")
+# goodness.fit = 0.3; dryseason = "on"; root.selection = "exp"; iso.subset = "off"
+sensitivity.analysis(goodness.fit = 0.3, dryseason = "on", 
+            root.selection = "exp", iso.subset = "off")
+sensitivity.analysis(goodness.fit = 0.3, dryseason = "jan", 
+            root.selection = "exp", iso.subset = "off")
 ```
 
-```{r Chunk 8, eval = FALSE}
-source("code/10.Comparison_with_Meinzer1999.R")
+#### Comparison with deuterium isotopic signatures of water sourcing depths
+
+<!-- For example: -->
+<!-- `source("code/09.0_Comparison_with_Meinzer1999.R") -->
+<!-- iso.compare(goodness.fit = 0.3, dryseason = "on",  -->
+<!--             root.selection = "exp", iso.subset = "on", drop.months = "None")` -->
+
+```{r Chunk 09.0, eval = FALSE}
+source("code/09.0_Comparison_with_Meinzer1999.R")
 iso.compare(goodness.fit = 0.3, dryseason = "on", 
@@
 
-```{r Chunk 9, eval = FALSE}
-source("code/9.6_mortality_BCI.R")
+#### Plot species and size-wise Uptake Depth Index
+```{r Chunk 10.0, eval = FALSE}
+source("code/10.0_species_graphs_BCI.R")
+```
+
+# Evaluating whole plant drought strategies
+
+#### Calculate mortality rates from forest inventory data
+
+```{r Chunk 11.0, eval = FALSE}
+source("code/11.0_demographic_data_BCI.R")
+```
+
+#### Analyse mortality against UDI
+```{r Chunk 11.1, eval = FALSE}
+source("code/11.1_mortality_BCI.R")
 demo.graphs(level.folder = "commlevel", n.threshold = 30, dbh.residuals = "on", 
@@
 
-```{r Chunk 10, eval = FALSE}
-source("code/5.species_graphs_BCI.R")
-source("code/13.hydraulic_traits_BCI.R")
+#### Analyse functional traits against UDI
+```{r Chunk 12.0, eval = FALSE}
+source("code/12.0_hydraulic_traits_BCI.R")
 ```

### 2020-07-06 manuscript.rmd
New file

---
title: "Drought-sensitive trees survive on deep water access in a tropical forest community"
author: ''
date: ''
output:
  bookdown::html_document2:
    number_sections: no
    theme: flatly
  bookdown::pdf_document2: default
  bookdown::word_document2:
        reference_docx: manuscriptstyle.docx
csl: new-phytologist.csl
bibliography: references.bib
---

```{r, include=FALSE}
if(!require("pacman", quietly=TRUE))install.packages("pacman")
suppressPackageStartupMessages(
  pacman::p_load(bookdown, knitr, pander, citr))
options(tinytex.verbose = TRUE)
```

\
**Corresponding author**: 
Rutuja Chitra-Tarak

\
E: arutuj@gmail.com

\
T: +1(505)

# Summary

- Deep water access (WA~Deep~) is believed to be an important drought-tolerance mechanism, albeit little studied. Vulnerability to embolism ($\Psi_{88, Stem}$) and safety margins ($\Psi_{min} - \Psi_{88, Stem}$) are shown to be predictors of drought-induced mortality risk, but deep water access may delay plants from reaching these critical thresholds. Here we test its role in drought-tolerance strategies of a tropical forest community (Panama) with a novel data-model approach.
- We inversely modelled Effective Rooting Depth (ERD) for 32 large, canopy tree species by linking species growth dynamics (xx trees, 6 censuses, 1990-2015) to VPD, $\Psi_{soil}$ by depth in the whole-soil column, and a novel community-scale model of species vulnerability curves ($K_{Leaf}$ vs. $\Psi_{x}$) linking data on the same for 21 species with community-scale softer traits. We evaluated ERD by leveraging existing isotopic tracer data ($\delta^2H$). 
- We find that deeper ERD is associated with higher hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margin, but lower mean mortality rate (*R*^2^ = 0.2, *P* = 0.0xx). Deeper ERD species spent less time below critical hydraulic thredholds indicating that ERD buffers drought-stress.  
- The role of deep water access in alleviating mortality of drought-sensitive trees have important implications for modeling the future of tropical species and forests under climate change.

Keywords: rooting depths, drought tolerance, mortality, tropical forest, climate change

# Introduction
\
Intensifying droughts [ref] threaten tropical forests, a biome that plays a disproportionately large role in the global carbon and energy cycle, and supports half of the global biodiversity [@Bonan:2008; @Wright:20050th]. Tropical forests are considered an especially drought-vulnerable biome given their narrower range of safety margins [@Choat:2012; @Oliveira:2019], and widespread drought-induced mortality has already been observed across the tropics, especially for large and light-wooded trees [@Bennett:2015; @Chitra-Tarak:2018; @Hilker:2014; @Phillips:2010]. However, state-of-the-art dynamic global vegetation models (DGVMs) fail to capture drought-induced vegetation dynamics in the tropics [@Galbraith:2010; @Powell:2013], as underlying mechanisms of drought-tolerance are not yet fully understood, while data are limited.

\
Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism. Lack of data on trees rooting or water-sourcing depths has been a key bottlneck, followed by data on water-availability at those depths. A recent meta-analysis shows that maximum rooting depths are known for less than 1% of the World’s 45-53 thousand tree species [@Fan:2017; @Silk:2000]. Further, only a small fraction of those are tropical, even though ~95% of the World’s tree diversity resides in the tropics [@Silk:2000]. Tracer techniques like stable isotopes provide an indirect way of trees' water sourcing depths in which isotopic composition of water in the non-transpiring plant tissue is matched with corresponding gradient in the vertical soil depth, but such data is also available only for ~0.05% of World's tree species [@Evaristo:2016]. Rooting or water-sourcing depths are especially challenging to obtain at community-scale for species-rich tropical forests.

\
Traditionally, species are thought to employ three general mechanisms to deal with drought via a variety of associated traits: (1) avoidance--dry season dormancy via deciduous leaf-habit, (2) delay--greater water access via deep rootedness or valley/riparian habitat preference; or reduced water loss via water storage and stomatal control, and (3) physiological tolerance--continued physiological function at low leaf water potential (Poorter 2008). Recent understanding of

- traits in global change
- trait-demography are traits really functional (link with plant performance and fitness)
Understanding and explaining the belowground mechanisms and trade‐offs that drive variation in root traits, resource acquisition and plant performance across species

- Safety-efficiency trade-off in hydraulic architecture--why deep rooted trees may be both acquisitive with unsafe xylem, yet an overall safe strategy--so long as their high resource environment is stable. 
- consequence on performance
- the problem of large trees--large tree mortality in the tropics

\
In this study, ….growth PLC and water ...This study x times increased the mean number of Percent Loss of Hydraulic Conductivity data reported for or tropical forests on by xx %

\
We test the hypotheses that: in this hyperdiverse forest, interspecific variation in effective rooting depths of canopy trees aligns with a conservative-acquisitive trade-off axis such that trees with deeper effective rooting depths rely on a relatively stable, larger water resource and are thus associated with acquisitive hydraulic architecture: with higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrow hydraulic safety margins. Despite having a drought-vulnerable xylem, in this humid forest availability of a reliable water at depth allows trees with deeper effective rooting depths to escape the meteorological drought, resulting in lower mortality rates compared to trees with shallow effective rooting depths.

# Materials and Methods
\
In the following sections we describe the study site, the models and data used to estimate and evaluate (1) the soil matric potential dynamics by depth over the forests’ rooting zone, (2) species-specific percent loss of leaf conductivity curves (PLC), and (3) species-specific effective rooting depths. These are followed by analyses to address the proposed hypotheses. Note that only new data on leaf physiological traits collected during this study are described below, details of all other datasets employed are described in the Supplementary Information.

## The Study Site
\
We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterisation in terms of long-term forest demography, plant functional ecology, hydrology, including availability of an independent isotopic tracer study on trees’ water-sourcing depths (Meinzer et al.). The 1,560 hectare Barro Colorado Island was formed after the river Chagres was dammed in 1914 that created Gatun Lake, the main channel of the Panama Canal. Rainfall at BCI is seasonal with a mean annual rainfall of 2687 mm (± 515 SD over 1983-2017) and a four-month long dry season from mid-December through mid-April  (≤ 100 mm rain/month). The entire Island is forested and classified as a tropical moist forest. Although percentage of canopy species that are at least partly deciduous during the dry season is 28%, the maximum percentage of deciduous individuals in the canopy is only 6.3%, and the percentage of crown area that is deciduous is estimated at 9.7% (Condit et al., 2000). Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 (Condit, 1998; Hubbell et al., 1999) (Anderson-Texeira xx).

\
In the 50 ha plot all stems above 10 mm are mapped, tagged with a unique number, identified to species level and measured every 5 years through 1985-2015 for diameter growth, mortality as well as recruitment of new stems into the 10 mm size class (Hubbell, S.P., Condit, R., and Foster, R.B., 2010; Condit et al., 2012). This inventory represents 321 woody species across the eight censuses. Aside from a 2 ha patch of ~ 100 year old secondary forest at the north-eastern corner of the plot the remaining 48 ha represent old growth forest that was never clear cut for agriculture (Piperno 1992). The plot (100 m x 50 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) (Harms 2001). The plot covers most of the upper catchment of Conrad Creek. The shallow valley of this creek is the main interruption to the otherwise mostly flat or gently sloping plot. The valley gives rise to the 1.5-ha swamp at the end of the wet season. Soils of the plot are homogenous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot (BCI soil survey report). The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil. Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding. Detailed descriptions of the climate, geology, flora and fauna of BCI can be found in Croat (1978), Leigh et al. (1982) and Gentry (1990).

## Model ELM-FATES for Soil Matric Potentials

### Model description

\
To estimate dynamics of water availability by depth over the forest’s root zone, we implemented a 1-D hydrological model, the Energy Exascale Earth System Land Model (ELM), coupled with a Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES). FATES is a state-of-the-art, community-based, open-source model widely used for studying climate-vegetation interactions. FATES is considered a next generation vegetation dynamics model for ELM, with a size-structured group of plants (cohorts) and successional trajectory-based patches using the ecosystem demography approach...... (How ELM talks to FATES)...Detailed descriptions can be found elsewhere (Koven et al., 2020) Fisher et al ).

\
Below we describe the equation of water balance in ELM-FATES that is most relevant for our study site, even though ELM-FATES is capable of incorporating numerous other hydrological processes relevant for wetlands and lakes and snow (CLM4.5 Tech Note). ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water ΔWcan, surface water ΔWsfc, soil water ΔWliq, i and water in the unconfined aquifer ΔWa (omitting processes relevant to snow or wetlands and lakes), and can be described as

 $\Delta W_{can} + \Delta W_{sfc} + \sum_{z=1}^{15} \Delta W_{liq,z} + \Delta W_{a} = \left(q_{rain} - E_{v} - E_{g} - q_{over} - q_{h2osfc} - q_{drai}\right).\Delta t$,        Eq. (1)

where, $q_{rain}$ is rainfall, $E_{v}$ is evapotranspiration (ET) of the forest, $E_{g}$ is ground evaporation, $q_{over}$ is surface runoff, $q_{h2osfc}$ is runoff from surface water storage, $q_{drai}$ is sub-surface drainage (all in mm H2O) and $\Delta t$ is the time-step. z are the 15 soil layers exponentially distributed in depth, here, (r soil.depths) (m).

### Model parameterization
\
To parameterise the model, we identified eleven parameters relevant for in eq (1), determined their ranges based on literature for the study site or for the tropics (Table 1), and sampled 10, 000 ensembles from this parameter space with Latin Hypercube Sampling.

\
We ran ELM-FATES for these 10, 000 ensembles with hourly climate drivers from 1985-2018 initialised with the stand structure of the 50 ha plot in a single site mode (0.1 degree x 0.1 degree centered at 9.1530 $^\circ$N, 280.2539 $^\circ$E). As our key interest here was in soil water availability rather than forest dynamics, we ran ELM-FATES in simplified modes: static stand structure and without plant hydrodynamics, and with a single Plant Functional Type (PFT) of evergreen trees. The latter choice was because only 9.7% of BCI crown area is dry-season deciduous and addition of dry-deciduous PFT did not significantly alter results (not shown).

### Model evaluation
\
We calibrated the model against three key fluxes in Eq. (1), namely, (i) evapotranspiration ($E_{v} + E_{g}$) from the eddy flux tower, (ii) stream discharge for the Conrad Catchment ($q_{over} + q_{drai}$) and (iii) Volumetric Water Content (VWC) obtained with Time Domain Reflectometry (TDR) at three depths (10, 40 and 100 cm; relating to $W_{liq, z}$) at point location as well as opportunistic dry season VWC measured over those depths for the entire 50-ha plot (Figure \@ref(fig:hydro-calib). Among the 10,000 parameter ensembles we chose 100 ensembles that best-fit all three fluxes simultaneously (add details...that is those that maximised the sum of regression coefficients between observed and simulated variables.) See [Table 1](#param) for best-fit parameter ranges.

```{r hydro-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap ="\\label{lab-hydro-calib}Model evaluation: Observations (red line) of Volumetric Water Content, Evapotranspiration and stream discharge aginst simulations (blue lines) from 100 best-fit parameter ensembles runs of ELM-FATES. This figure will be reformatted to compose individual figures into one."}
p1 <- cowplot::ggdraw() + cowplot::draw_image("figures/2019-10-14_5000/best-fits/swc_Obs_vs_model_daily_horizontal_with_steph_mean.jpeg", scale = 1)
p2 <- cowplot::ggdraw() + cowplot::draw_image("figures/2019-10-14_5000/best-fits/QRUNOFF_Obs_vs_model_monthly.jpeg", scale = 1)
p3 <- cowplot::ggdraw() +
  cowplot::draw_image("figures/2019-10-14_5000/best-fits/ET_Obs_vs_model_monthly.jpeg", scale = 1)
bottom_row <- cowplot::plot_grid(p2, p3, labels = c('B', 'C'), label_size = 12, ncol = 2)
cowplot::plot_grid(p1, bottom_row, labels = c('A', ''), label_size = 12, ncol = 1, rel_widths = c(2, 1))
```

## Model for Percent Loss of Leaf Conductivity

```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', fig.cap = "\\label{lab-comm-plc-curves}Modelled 'vulnerability curves' for standardized *K*~Leaf~ vs. $\\Psi$~Leaf~ for 189 tree species from Panama, color coded by Wood Specific Gravity (units)."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/std.k.spkmax_by_psi_color_by_Specific Gravity_predicted_AB.jpeg"))
```

See Figure \@ref(fig:comm-plc-curves)

## Model for Effective Rooting Depths (ERD)

### Model parameterization

### Model evaluation

Figure \@ref(fig:erd-iso)

```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-iso}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE; m) versus dry season stable isotopic concentration $\\delta$^2^H (mean $\\pm$ (add error); ‰) for seven canopy species from Barro Colorado Island (data from [@Meinzer:1999]"}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_phenology_Meinzer_gr.Psi.VPD.jpeg"))
```

## ERD Coordination with Above-ground Hydraulic Traits

To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD with maximum stem hydraulic conductance  ($K_{max, Stem}$), leaf Turgor Loss Point ($\Psi_{TLP}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88, Stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88, Stem}$).

## Mortality Analyses and Extreme Drought Experience

To test whether ERD plays a role in alleviating mortality risk, we regressed species ERD against mean mortality rates over the 33 year record (1985-2015). To assess whether this may be due to deeper ERD species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potentials in the soil zone matching species ERD (𝚿ERD) were below species P80, leaf.

All statistical analyses were done in R statistical environment (version 3.5.3, @R:2020).

# Results

## Effective Rooting Depths, Vulnerability to Embolism and Hydraulic Safety Margins

\
In this forest modelled Effective Rooting Depths (ERD) for the 32 large, canopy tree species varied from 0.5 m to 13 m (Figure \@ref(fig:sp-erd)). For the subset of species for which hydraulic properties were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max, Stem}$; *R*^2^ = 0.39, *P* = 0.098, *n* = 8; Figure \@ref(fig:erd-hyd)A), less negative leaf Turgor Loss Point ($\Psi_{TLP}$); *R*^2^ = 0.6, p = 0.024, *n* = 8; Figure \@ref(fig:erd-hyd)B), greater vulnerability to embolism from cavitation with less negative ($\Psi_{88, Stem}$) (*R*^2^ = 0.72, p = 0.007, *n* = 8; Figure \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88, Stem}$; *R*^2^ = 0.74, p = 0.013, *n* = 8; Figure \@ref(fig:erd-hyd)D).

```{r sp-erd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap="\\label{lab-sp-erd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for 32 large, canopy species of Barro Colorado Island. See Table S2 for scientific names of species codes"}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/ERD_by_sp_large_canopy.jpeg"))
```

```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap ="\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus hydraulic propertiese for seven large, canopy species found on Barro Colorado Island. *R*^2^ and *P*-values are given in panel insets."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/Select_traits_depth.jpeg"))
```

## Effective Rooting Depths, Experienced Drought and Mortality

\
Species with deeper ERD showed lower mean mortality rates across the 33 year record (1982-2015; *R*^2^ = 0.13, *P* = 0.029, *n* = xx; Figure \@ref(fig:erd-mort)). For five out of seven census intervals, species with shallower ERD had significantly higher mortality anomalies (MortalityCensus Interval, sp - Mean(MortalityCensus Interval, sp) than deeper ERD species; Figure (\@ref(fig:erd-mort-anom)). Analysis of time spent under catastrophic soil drought showed that the proportion of days species $\Psi_{z = ERD}$ was more negative than species $\Psi_{80, Leaf}$ decreased with species ERD (Figure \@ref(fig:erd-time)). Species with ERD deeper than 1 m did not cross species $\Psi_{80, Leaf}$ at all for the entire 25 year period for which soil water potential was modelled (1990-2015).

```{r erd-mort, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort}Mortality rate (mean $\\pm$ (add error)) versus Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for xx canopy species found on Barro Colorado Island. *R*^2^ and *P*-values are given in panel insets."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_rate_by rdi.gr.jpeg"))
```

```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-time}Species modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{x, z}$ was below $\\Psi_{80,Leaf}$ over 1990-2015, where z is the soil depth matching species ERD--for xx canopy species found on Barro Colorado Island. (Are *R*^2^ and *P*-values needed here?)"}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr.jpeg"))
```

# Discussion

### ERD not correlated with growth
# Limitations & Future directions:

- not hydrodynamics, 1D to 3D

- Temperate rooting depths shallow?  Shenk and Jackson

-transient depth of water uptake, which is known to vary seasonally (Ellsworth & Sternberg, 2015), 
In fact, associations between soil water access by roots and aboveground drought survival traits such as resistance of xylem to drought‐induced cavitation (Pockman & Sperry, 2000; Pratt et al ., 2007) or phenology (Jackson et al ., 1995; Hasselquist et al ., 2010), are fairly reliable trade‐off functions in plant eco‐physiology.

- Deep ERD species may possibly be associated with locations of greater water availability rather than having deep rootedness. Future studies should employ a distributed hydrological model to separate these factors.

Future studies: high resolution growth data to better constrain the model. 
  -water storage (capacitance) should be considered
  -shifts in root water-uptake by depth
  -deci

## Acknowledgements
## Data Availability
## Supplementary Information

```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts}Extreme droughts by census interval and some representative depths"}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q2.5.jpeg"))
```

```{r iso-data, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-iso-data}Stable isotopic concetration ($\\delta^2H$) of soil water by depth measured in March 1997. Data from @Meinzer:1999"}
knitr::include_graphics(file.path("figures/UDI_confidence/Meinzer_etal_1999_Depth_vs_soil_deltaD.jpeg"))
```

```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-other-erd-mods}All ERD models tested. See Eqns xxxx"}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_sp_color_Meinzer.jpeg"))
```

```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-obs-plc-curves} 'vulnerability curves' fitted to observed data shown as standardized *K*~Leaf~ vs. $\\Psi$~Leaf~ for 21 tree species from Panama, color coded by Wood Specific Gravity (units)."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/std.k.commkmax_by_psi_color_by_Specific Gravity_predicted_AB.jpeg"))
```

```{r ab-fit, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', fresults = "asis", fig.cap = "\\label{lab-ab-fit}Parameters of 'vulnerability curve' (Eq.) fitted to data versus predicted from soft traits (Eq.)"}
p1 <- cowplot::ggdraw() + cowplot::draw_image("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/A_data_vs_model.jpeg", scale = 1)
p2 <- cowplot::ggdraw() + cowplot::draw_image("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/B_data_vs_model.jpeg", scale = 1)
cowplot::plot_grid(p1, p2, ncol = 2, labels = "AUTO")
```

```{r erd-growth, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-growth}Scatterplot between Effective Rooting Deoth (ERD) and mean growth rate $\\pm$ (add error)for xx large, canopy species"}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/adult_Growth_vs_udi_with_outliers_sp.jpeg"))
```

```{r erd-mort-anom, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-anom}Mortality anomaly in each census interval during the study period plotted against ERD"}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/diff.mortality_by rdi.gr.jpeg"))
```

#### Table 1: ELM-FATES parameters chosen for enseble generation, their description, chosen ranges and references and/or rationale.  {#param}

| S.N. | Parameter | Description | Range (Units) | Method of determination with reference
| -- | ------ | ------------- | ------------- |  ------------------------------------------------------------------------------------------------- |
| 1 | fates_leaf_ BB_slope | stomatal slope parameter, as per Ball-Berry | 4 - 16 Unitless  | The range of Ball Berry parameter fitted across several studies as compiled in Table 2 of (Medlyn et al. 2011) | 
| 2 | fates_leaf_ slatop | Specific Leaf Area (SLA) at the top of canopy, projected area basis | 0.0042 - 0.0400 m^2^/gC | Based on the observed range of Leaf Mass Area (LMA)--24.97 to 235.8 g per m^2^-- for individual tree variation across 51 tree species in Barro Colorado Island (this study), and assumption of 50% carbon content in biomass. |  
| 3 | fates_leaf_ vcmax25top | Maximum carboxylation rate of Rubisco at 25 $^\circ C$, canopy top | 22.7 - 92.5 umol CO^2^/m^2^/s | Observed range of Vcmax at 25 $^\circ C$ for tropical tree species under relative radiation of 50% or greater, thereby excluding highly shaded leaves. Values derived with a conversion specific for CLM model are used. (Ali et al. 2015) | 
| 4 | fates_roota_ par | CLM rooting distribution parameter | 5.9 - 7.4 per m | Parameter a in Eq. 2 in (Zeng 2001) that regulates the shape of the rooting profile. Range corresponds to this parameter specified for BATS (or IGBP) land cover classified Deciduous Broadleaf Trees (5.9 m-1) and Evergreen Broadleaf Trees  (7.4 m-1) as given in Table 1 (or 2) of (Zeng 2001). | 
| 5 | fates_rootb_ par | CLM rooting distribution parameter | 0.217 - 1.956  per m | Parameter b in Eq. 2 in (Zeng 2001) that regulates the depth of the rooting profile. Chosen range of *b* is derived using this equation so as to fit the observed range of rooting depth (d~r~) of 2 - 18 m for Tropical Deciduous Forest (mean ± Standard Error (SE); 3.7 ± 0.5, *n* = 5 trees; min = 2, max = 4.7)  and Tropical Evergreen Forest (mean $\pm$ S.E.; 7.3 $\pm$ 0.5, n = 3 trees and 3 communities; min = 2, max = 18 m) combined (Canadell et al. 1996). Besides the direct observation of roots at 18 m included by (Nepstad et al. 1994) in Paragominas, eastern Amazonia that is included in the above study; in Tapajos, eastern Amazonia water extraction by roots was also inferred up to 18 m. (Davidson et al. Forest Science 2011) |
| 6 | fates_smpsc | Soil water potential at full stomatal closure | 113000 - 242000 mm | Based on observed range of -1.13 to -2.42 MPa for leaf turgor loss point for individual tree variation across 49 BCI tree species. This study| 
| 7 | aveDTB | Distance to bedrock | 3 - 18 m | Ben Turner; pers. comm. | 
| 8 | fmax | The maximum fractional saturated area |  unitless | Empirical | 
| 9 | HKSAT | Soil hydraulic conductivity profile | 0.007 - 0.014 mm/s for $\le$ 12.5 cm; 5.56e-05, 0.0003 for $\ge$ 60 cm | To account for high macroporosity and direct flow paths in tropical soils(Broedel et al. 2017; Tomasella and Hodnett 1996)(Kinner and Stallard 2004; Broedel et al. 2017; Tomasella and Hodnett 1996) For Conrad catchment observed median Ksat (95%CI) for 12.5 cm depth is 38.3 mm/hr (25.4-51.2, *n* = 75), while for 60 cm depth 0.7 (0.2, 1.2, *n* = 40) mm/hr. For reference a storm with 12.5 mm/hr rainfall intensity has a 0.2 probability of occuring in any given rainfall event (Kinner and Stallard 2004). | 
| 10 | HKSAT_ADJ | Adjusting factor for soil hydraulic conductivity | 1 - 8 unitless | To account for high macroporosity and direct flow paths in tropical soils that is not accounted for by small soil core samples (Broedel et al. 2017; Tomasella and Hodnett 1996)(Kinner and Stallard 2004; Broedel et al. 2017; Tomasella and Hodnett 1996) | 
| 11 | fpi_max | Maximum interception fraction of precipitation | 0.05-0.44 | Based on throughfall data from (Zimmermann et al. 2010) and precipitation data for BCI from STRI Physical Monitoring program, defined for precipitation events greater than 10 mm. |

# References

### 2020-07-09 manuscript.rmd
changed to large trees demo rates instead of adult

@@
     theme: flatly
+    fig_caption: yes
   bookdown::pdf_document2: default
@@
         reference_docx: manuscriptstyle.docx
+        fig_caption: yes
 csl: new-phytologist.csl
@@
 suppressPackageStartupMessages(
-  pacman::p_load(bookdown, knitr, pander, citr))
+  pacman::p_load(bookdown, knitr, pander, citr, broom))
 options(tinytex.verbose = TRUE)
+source("code/load.R")
+```
+
+```{r setup}
+library(knitr)
+opts_knit$set(eval.after = "fig.cap")
 ```
@@
 Rutuja Chitra-Tarak
-
 \
 E: arutuj@gmail.com
-
 \
@@
 
-- Deep water access (WA~Deep~) is believed to be an important drought-tolerance mechanism, albeit little studied.
+- Deep water access (WA~Deep~) is believed to be an important drought-tolerance mechanism, but under-studied.
 Vulnerability to embolism ($\Psi_{88, Stem}$) and safety margins ($\Psi_{min} - \Psi_{88, Stem}$) are shown to be predictors of drought-induced mortality risk, but deep water access may delay plants from reaching these critical thresholds.
@@
 \
-Intensifying droughts [ref] threaten tropical forests, a biome that plays a disproportionately large role in the global carbon and energy cycle, and supports half of the global biodiversity [@Bonan:2008; @Wright:20050th].
-Tropical forests are considered an especially drought-vulnerable biome given their narrower range of safety margins [@Choat:2012; @Oliveira:2019], and widespread drought-induced mortality has already been observed across the tropics, especially for large and light-wooded trees [@Bennett:2015; @Chitra-Tarak:2018; @Hilker:2014; @Phillips:2010].
-However, state-of-the-art dynamic global vegetation models (DGVMs) fail to capture drought-induced vegetation dynamics in the tropics [@Galbraith:2010; @Powell:2013], as underlying mechanisms of drought-tolerance are not yet fully understood, while data are limited.
+Intensifying droughts threaten tropical forests, a biome that plays a disproportionately large role in the global carbon and energy cycle, and supports half of the global biodiversity [@Bonan:2008; @Wright:20050th].
+Tropical forests are considered an especially drought-vulnerable biome based on current estimates of trees' hydraulic safety margins [@Choat:2012; @Oliveira:2019].
+Widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak:2018; @Hilker:2014; @Phillips:2010].
+However, state-of-the-art dynamic global vegetation models (DGVMs) fail to capture drought-induced vegetation dynamics in the tropics [@Galbraith:2010; @Powell:2013], as underlying mechanisms of drought-tolerance are not fully understood.
 
@@
 Lack of data on trees rooting or water-sourcing depths has been a key bottlneck, followed by data on water-availability at those depths.
-A recent meta-analysis shows that maximum rooting depths are known for less than 1% of the World’s 45-53 thousand tree species [@Fan:2017; @Silk:2000].
+A recent meta-analysis documented maximum rooting depths for 318 tree species known for less than 1% of the World’s 45-53 thousand tree species [@Fan:2017; @Silk:2000].
 Further, only a small fraction of those are tropical, even though ~95% of the World’s tree diversity resides in the tropics [@Silk:2000].
 Tracer techniques like stable isotopes provide an indirect way of trees' water sourcing depths in which isotopic composition of water in the non-transpiring plant tissue is matched with corresponding gradient in the vertical soil depth, but such data is also available only for ~0.05% of World's tree species [@Evaristo:2016].
-Rooting or water-sourcing depths are especially challenging to obtain at community-scale for species-rich tropical forests.
+Community-scale data collection for rooting or water-sourcing depths is especially challenging for species-rich tropical forests.
 
@@
 
-- traits in global change
+Trees' form and function are outcome of the environment in which trees have evolved. And as cliamte changes  and Traits link to plant environment - trait - performance 
+- traits in global change as environment changes
 - trait-demography are traits really functional (link with plant performance and fitness)
 Understanding and explaining the belowground mechanisms and trade‐offs that drive variation in root traits, resource acquisition and plant performance across species
+
+Differential dynamics of water environment by depth
 
@@
 \
-In this study, ….growth PLC and water ...This study x times increased the mean number of Percent Loss of Hydraulic Conductivity data reported for or tropical forests on by xx %
+In this study, what we do differently: which soil depth dynamics link dynamics of water dynamics in whole-soil column, atmospheric drought via
+...This study increased the mean number of tropical forest species with data on Percent Loss of Hydraulic Conductivity data by xx %
 
@@
 This inventory represents 321 woody species across the eight censuses.
-Aside from a 2 ha patch of ~ 100 year old secondary forest at the north-eastern corner of the plot the remaining 48 ha represent old growth forest that was never clear cut for agriculture (Piperno 1992).
+Aside from a 2 ha patch of ~100 year old secondary forest at the north-eastern corner of the plot the remaining 48 ha represent old growth forest that was never clear cut for agriculture (Piperno 1992).
 The plot (100 m x 50 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) (Harms 2001).
@@
 Below we describe the equation of water balance in ELM-FATES that is most relevant for our study site, even though ELM-FATES is capable of incorporating numerous other hydrological processes relevant for wetlands and lakes and snow (CLM4.5 Tech Note).
-ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water ΔWcan, surface water ΔWsfc, soil water ΔWliq, i and water in the unconfined aquifer ΔWa (omitting processes relevant to snow or wetlands and lakes), and can be described as
+ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\Delta W_{can}$, surface water $\Delta W_{sfc}$, soil water $\Delta W_{liq, z}$, and water in the unconfined aquifer $\Delta W_{a}$ (omitting processes relevant to snow or wetlands and lakes), and can be described as
 
- $\Delta W_{can} + \Delta W_{sfc} + \sum_{z=1}^{15} \Delta W_{liq,z} + \Delta W_{a} = \left(q_{rain} - E_{v} - E_{g} - q_{over} - q_{h2osfc} - q_{drai}\right).\Delta t$,        Eq. (1)
+\begin{equation}
+\Delta W_{can} + \Delta W_{sfc} + \sum_{z=1}^{15} \Delta W_{liq,z} + \Delta W_{a} = \left(q_{rain} - E_{v} - E_{g} - q_{over} - q_{h2osfc} - q_{drai}\right)\Delta t
+(\#eq:water-balance)
+\end{equation}
 
@@
 \
-To parameterise the model, we identified eleven parameters relevant for in eq (1), determined their ranges based on literature for the study site or for the tropics (Table 1), and sampled 10, 000 ensembles from this parameter space with Latin Hypercube Sampling.
+To parameterise the model, we identified eleven parameters relevant for in Eq.
+\@ref(eq:water-balance), determined their ranges based on literature for the study site or for the tropics ([Table 1](#param)), and sampled 10, 000 ensembles from this parameter space with Latin Hypercube Sampling.
 
 \
-We ran ELM-FATES for these 10, 000 ensembles with hourly climate drivers from 1985-2018 initialised with the stand structure of the 50 ha plot in a single site mode (0.1 degree x 0.1 degree centered at 9.1530 $^\circ$N, 280.2539 $^\circ$E).
+We ran ELM-FATES for these 10,000 ensembles with hourly climate drivers from 1985-2018 initialised with the stand structure of the 50 ha plot in a single site mode (0.1$^\circ$ x 0.1$^\circ$ centered at 9.1530 $^\circ$N, 280.2539 $^\circ$E).
 As our key interest here was in soil water availability rather than forest dynamics, we ran ELM-FATES in simplified modes: static stand structure and without plant hydrodynamics, and with a single Plant Functional Type (PFT) of evergreen trees.
-The latter choice was because only 9.7% of BCI crown area is dry-season deciduous and addition of dry-deciduous PFT did not significantly alter results (not shown).
+The latter waws chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of dry-deciduous PFT did not significantly alter results (not shown).
 
@@
 We calibrated the model against three key fluxes in Eq.
-(1), namely, (i) evapotranspiration ($E_{v} + E_{g}$) from the eddy flux tower, (ii) stream discharge for the Conrad Catchment ($q_{over} + q_{drai}$) and (iii) Volumetric Water Content (VWC) obtained with Time Domain Reflectometry (TDR) at three depths (10, 40 and 100 cm; relating to $W_{liq, z}$) at point location as well as opportunistic dry season VWC measured over those depths for the entire 50-ha plot (Figure \@ref(fig:hydro-calib).
+\@ref(eq:water-balance), namely, (i) evapotranspiration ($E_{v} + E_{g}$) from the eddy flux tower, (ii) stream discharge for the Conrad Catchment ($q_{over} + q_{drai}$) and (iii) Volumetric Water Content (VWC) obtained with Time Domain Reflectometry (TDR) at three depths (10, 40 and 100 cm; relating to $W_{liq, z}$) at point location as well as opportunistic dry season VWC measured over those depths for the entire 50-ha plot (Figure \@ref(fig:hydro-calib).
 Among the 10,000 parameter ensembles we chose 100 ensembles that best-fit all three fluxes simultaneously (add details...that is those that maximised the sum of regression coefficients between observed and simulated variables.) See [Table 1](#param) for best-fit parameter ranges.
@@
 
-## Model for Percent Loss of Leaf Conductivity
+## Model for Leaf Vulnerability Curves at community-scale
 
-```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', fig.cap = "\\label{lab-comm-plc-curves}Modelled 'vulnerability curves' for standardized *K*~Leaf~ vs.
-$\\Psi$~Leaf~ for 189 tree species from Panama, color coded by Wood Specific Gravity (units)."}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/std.k.spkmax_by_psi_color_by_Specific Gravity_predicted_AB.jpeg"))
+We measured ($K_{x, Leaf}$ vs.
+$\Psi_{x, Leaf}$) on adult trees of `r length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))` common species (see SI x).
+We fitted exponential, sigmoidal, logistic and exponential form of equations to species-level data as prescribed by [@GUYOT:2011] and based on AIC criteria chose exponential form as the best-fitting one (Fig.
+\@ref(fig:obs-plc-curves)), described as
+
+\begin{equation}
+K_{Leaf} = y_0 + Ae^{-B\Psi_{Leaf}} + error
+(\#eq:vc-exp)
+\end{equation}
+
+where $A$ and $B$ are fitted species-specific parameters.
+
+We fitted parameters $A$ and $B$ in Eq.
+\@ref(eq:vc-exp) with polynomial equations a functions of two traits that were available at community-scale, namely, $WSG$ and $LMA$, defined as mean wood specific gravity after drying at 100$^\circ$C (g cm^{-3}^) and mean leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules (g m^{-2}^) for leaves receiving direct sunlight, respectively.
+
+\begin{equation}
+A = a_3WSG^2 + a_2B + a_1B^2 + a_0 + error
+(\#eq:a-mod)
+\end{equation}
+
+\begin{equation}
+B = b_3WSG^2 + b_2WSG.LMA + b_1LMA^2 + b_0 + error
+(\#eq:b-mod)
+\end{equation}
+
+where $(a_0, a_1, a_2, a_3, b_0, b_1, b_2, b_3)$ are coefficients.
+$WSG$ & parameter $B$ explained large variation in parameter $A$ (Eq.
+\@ref(eq:a-mod); Adj.
+R^2^ = `r signif(glance(k_by_psi.models[[2]])$adj.r.squared, 2)`; *P* = `r  signif(glance(k_by_psi.models[[2]])$p.value, 2)`; also see Fig.
+\@ref(fig:ab-fit)).
+$WSG$ & $LMA$ explained large variation in parameter $B$ (Eq.
+\@ref(eq:b-mod); Adj.
+*R*^2^ = `r signif(glance(k_by_psi.models[[1]])$adj.r.squared, 2)`; *P* = `r  signif(glance(k_by_psi.models[[1]])$p.value, 2)`).
+We thus sequentially used Eq.
+(\#eq:b-mod) and (\#eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r length(unique(data.model.AB$sp))` species for which WSG and LMA data were available and fitted vulnerability curves (see Fig.
+\@ref(fig:comm-plc-curves)).
+
+```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', fig.cap = paste("\\label{lab-comm-plc-curves} Modelled standardized 'Vulnerability curves' for  $K_{x, Leaf}$ vs.
+$\\Psi_{x, Leaf}$ at community-scale based on communit-scale data on $WSG$ & $LMA$ and their relationships with vulnerability curve parameters measured for a few species.
+Each line represents a species, color coded by $WSG$.")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_Wood Specific Gravity_predicted_AB.jpeg"))
 ```
-
-See Figure \@ref(fig:comm-plc-curves)
 
@@
 
-To test whether ERD plays a role in alleviating mortality risk, we regressed species ERD against mean mortality rates over the 33 year record (1985-2015).
-To assess whether this may be due to deeper ERD species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potentials in the soil zone matching species ERD (𝚿ERD) were below species P80, leaf.
+We calculated mortality rates for large trees (>=10 cm DBH) of xx canopy species for each of the seven census interval ($int$) in the 33 year record (1985-2015; $Mort_{int}$) and a species mean mortality rate across those intervals ($Mort_{mean} = \frac{1}{int}\sum_{int=1}^{6} Mort_{int}$).
+To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$).
+To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{x, ERD}$) were below species $\Psi_{80,Leaf}$.
 
@@
 
-```{r sp-erd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap="\\label{lab-sp-erd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for 32 large, canopy species of Barro Colorado Island.
-See Table S2 for scientific names of species codes"}
+```{r sp-erd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap= paste0("\\label{lab-sp-erd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for ", length(unique(erd.data$R2)) ," large, canopy species of Barro Colorado Island.
+See Table S2 for scientific names of species codes")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/ERD_by_sp_large_canopy.jpeg"))
@@
 
-```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap ="\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus hydraulic propertiese for seven large, canopy species found on Barro Colorado Island.
+```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus hydraulic propertiese for seven large, canopy species found on Barro Colorado Island.
 *R*^2^ and *P*-values are given in panel insets."}
@@
 \
-Species with deeper ERD showed lower mean mortality rates across the 33 year record (1982-2015; *R*^2^ = 0.13, *P* = 0.029, *n* = xx; Figure \@ref(fig:erd-mort)).
-For five out of seven census intervals, species with shallower ERD had significantly higher mortality anomalies (MortalityCensus Interval, sp - Mean(MortalityCensus Interval, sp) than deeper ERD species; Figure (\@ref(fig:erd-mort-anom)).
-Analysis of time spent under catastrophic soil drought showed that the proportion of days species $\Psi_{z = ERD}$ was more negative than species $\Psi_{80, Leaf}$ decreased with species ERD (Figure \@ref(fig:erd-time)).
-Species with ERD deeper than 1 m did not cross species $\Psi_{80, Leaf}$ at all for the entire 25 year period for which soil water potential was modelled (1990-2015).
+Species with deeper $ERD$ showed lower mean mortality rates across the 33 year record (1982-2015; *R*^2^ = 0.13, *P* = 0.029, *n* = xx; Figure \@ref(fig:erd-mort)).
+For five out of seven census intervals, species with shallower $ERD$ had significantly higher mortality anomalies ($Mort_{int}$ - $Mort{mean}$ than deeper ERD species; Figure (\@ref(fig:erd-mort-anom)).
+Analysis of time spent under catastrophic soil drought showed that the proportion of days species $\Psi_{x,~ERD}$ was more negative than species $\Psi_{80, Leaf}$ decreased with species ERD (Figure \@ref(fig:erd-time)).
+Species with ERD deeper than 1 m did not cross species $\Psi_{80, Leaf}$ at all for the entire 25 year period for which soil water potential $\Psi_{x,~z}$ was modelled (1990-2015).
 
@@
 ```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-obs-plc-curves} 'vulnerability curves' fitted to observed data shown as standardized *K*~Leaf~ vs.
-$\\Psi$~Leaf~ for 21 tree species from Panama, color coded by Wood Specific Gravity (units)."}
+$\\Psi$~Leaf~ for `r length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))` tree species from Panama, color coded by Wood Specific Gravity (units)."}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/std.k.commkmax_by_psi_color_by_Specific Gravity_predicted_AB.jpeg"))
@@
 
-```{r ab-fit, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', fresults = "asis", fig.cap = "\\label{lab-ab-fit}Parameters of 'vulnerability curve' (Eq.) fitted to data versus predicted from soft traits (Eq.)"}
+```{r ab-fit, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', fresults = "asis", fig.cap = "\\label{lab-ab-fit}Parameters $A$ & $B$ of 'vulnerability curve' fitted to data versus predicted from WSG and LMA"}
 p1 <- cowplot::ggdraw() + cowplot::draw_image("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/A_data_vs_model.jpeg", scale = 1)
@@
 2011) |
-| 2 | fates_leaf_ slatop | Specific Leaf Area (SLA) at the top of canopy, projected area basis | 0.0042 - 0.0400 m^2^/gC | Based on the observed range of Leaf Mass Area (LMA)--24.97 to 235.8 g per m^2^-- for individual tree variation across 51 tree species in Barro Colorado Island (this study), and assumption of 50% carbon content in biomass.
+| 2 | fates_leaf_ slatop | Specific Leaf Area (SLA) at the top of canopy, projected area basis | 0.0042 - 0.0400 $m^2/gC$ | Based on the observed range of Leaf Mass Area (LMA)--24.97 to 235.8 $g/m^2$-- for individual tree variation across 51 tree species in Barro Colorado Island (this study), and assumption of 50% carbon content in biomass.
 |
-| 3 | fates_leaf_ vcmax25top | Maximum carboxylation rate of Rubisco at 25 $^\circ C$, canopy top | 22.7 - 92.5 umol CO^2^/m^2^/s | Observed range of Vcmax at 25 $^\circ C$ for tropical tree species under relative radiation of 50% or greater, thereby excluding highly shaded leaves.
+| 3 | fates_leaf_ vcmax25top | Maximum carboxylation rate of Rubisco at 25 $^\circ$C, canopy top | 22.7 - 92.5 umol $CO_2/m^2/s$ | Observed range of Vcmax at 25 $^\circ$C for tropical tree species under relative radiation of 50% or greater, thereby excluding highly shaded leaves.
 Values derived with a conversion specific for CLM model are used.
@@
 2 in (Zeng 2001) that regulates the shape of the rooting profile.
-Range corresponds to this parameter specified for BATS (or IGBP) land cover classified Deciduous Broadleaf Trees (5.9 m-1) and Evergreen Broadleaf Trees  (7.4 m-1) as given in Table 1 (or 2) of (Zeng 2001).
+Range corresponds to this parameter specified for BATS (or IGBP) land cover classified Deciduous Broadleaf Trees (5.9 $m^{-1}$ and Evergreen Broadleaf Trees  (7.4 $m^{-1}$) as given in Table 1 (or 2) of (Zeng 2001).
 |
-| 5 | fates_rootb_ par | CLM rooting distribution parameter | 0.217 - 1.956  per m | Parameter b in Eq.
+| 5 | fates_rootb_ par | CLM rooting distribution parameter | 0.217 - 1.956 $m^{-1}$ | Parameter b in Eq.
 2 in (Zeng 2001) that regulates the depth of the rooting profile.
@@
 2017; Tomasella and Hodnett 1996) |
-| 11 | fpi_max | Maximum interception fraction of precipitation | 0.05-0.44 | Based on throughfall data from (Zimmermann et al.
+| 11 | fpi_max | Maximum interception fraction of precipitation | 0.05-0.44 unitless | Based on throughfall data from (Zimmermann et al.
 2010) and precipitation data for BCI from STRI Physical Monitoring program, defined for precipitation events greater than 10 mm.
@@
 
+### Appendix A
+
+#### 1. 
 # References

### 2020-07-10 manuscript.rmd
edited figure captions to add inline code

@@
 
-```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', fig.cap = paste("\\label{lab-comm-plc-curves} Modelled standardized 'Vulnerability curves' for  $K_{x, Leaf}$ vs.
-$\\Psi_{x, Leaf}$ at community-scale based on communit-scale data on $WSG$ & $LMA$ and their relationships with vulnerability curve parameters measured for a few species.
+```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-plc-curves} Modelled standardized 'Vulnerability curves' ($K_{x, Leaf}$ vs.
+$\\Psi_{x, Leaf}$) at community-scale (", nrow(data.model.AB$sp)," species) based on species data on $WSG$ & $LMA$ and their relationships with vulnerability curve parameters measured on ", sp.soft.filled$AB.data.model," species.
 Each line represents a species, color coded by $WSG$.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_Wood Specific Gravity_predicted_AB.jpeg"))
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB.jpeg"))
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_LMALAM_AVD_predicted_AB.jpeg"))
 ```
@@
 
-```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-iso}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE; m) versus dry season stable isotopic concentration $\\delta$^2^H (mean $\\pm$ (add error); ‰) for seven canopy species from Barro Colorado Island (data from [@Meinzer:1999]"}
+```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-iso}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE; m) versus dry season stable isotopic concentration $\\delta$^2^H (mean $\\pm$ (add error); ‰) for ", length(subset(ml.rsq.combine.sub, source == "Meinzer et al.1999 Fig.
+4" & corr.func == "gr.Psi.VPD")$sp)," canopy species from Barro Colorado Island (data from [@Meinzer:1999]")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_phenology_Meinzer_gr.Psi.VPD.jpeg"))
@@
 
-```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus hydraulic propertiese for seven large, canopy species found on Barro Colorado Island.
+```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus hydraulic propertiese for seven large, canopy species found on Barro Colorado Island (Panama).
 *R*^2^ and *P*-values are given in panel insets."}
@@
 
-```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-time}Species modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{x, z}$ was below $\\Psi_{80,Leaf}$ over 1990-2015, where z is the soil depth matching species ERD--for xx canopy species found on Barro Colorado Island.
-(Are *R*^2^ and *P*-values needed here?)"}
+```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{x, z}$ was below $\\Psi_{80,Leaf}$ over 1990-2015, where z is the soil depth matching species ERD--for ", ," canopy species found on Barro Colorado Island.
+(Are *R*^2^ and *P*-values needed here?)")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr.jpeg"))
@@
 knitr::include_graphics(file.path("figures/UDI_confidence/Meinzer_etal_1999_Depth_vs_soil_deltaD.jpeg"))
+```
+
+```{r leaf-cover, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-cover} Seasonal variation in Leaf Cover Fraction for ", length(unique(sp.leaf_cover.for.model$sp[sp.leaf_cover.for.model$deciduousness != "Evergreen"]))," non-evergreen species at Barro Colorado Island.
+Leaf Cover Fraction for evergreen species (n = ", length(unique(sp.leaf_cover.for.model$sp[sp.leaf_cover.for.model$deciduousness == "Evergreen"])),") is assumed at 1 through the year.")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/leaf.cover_BCI_single_panel_evergreensAt100.jpeg"))
 ```

### 2020-07-12 manuscript.rmd
1. Added a what we did para in intro 2. Re-wrote hypotheses

@@
 output:
+  bookdown::word_document2:
+        reference_docx: manuscriptstyle.docx
+        fig_caption: yes
+  bookdown::pdf_document2: default
   bookdown::html_document2:
@@
     fig_caption: yes
-  bookdown::pdf_document2: default
-  bookdown::word_document2:
-        reference_docx: manuscriptstyle.docx
-        fig_caption: yes
+bibliography: references.bib
 csl: new-phytologist.csl
-bibliography: references.bib
 ---
@@
 
-- Deep water access (WA~Deep~) is believed to be an important drought-tolerance mechanism, but under-studied.
+- Deep water access (WA~Deep~) is believed to be an important, drought-tolerance mechanism, but under-studied due to lack of data.
 Vulnerability to embolism ($\Psi_{88, Stem}$) and safety margins ($\Psi_{min} - \Psi_{88, Stem}$) are shown to be predictors of drought-induced mortality risk, but deep water access may delay plants from reaching these critical thresholds.
@@
 We evaluated ERD by leveraging existing isotopic tracer data ($\delta^2H$).
-- We find that deeper ERD is associated with higher hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margin, but lower mean mortality rate (*R*^2^ = 0.2, *P* = 0.0xx).
+- We find that deeper ERD is associated with higher hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margin, but lower mean mortality rate.
 Deeper ERD species spent less time below critical hydraulic thredholds indicating that ERD buffers drought-stress.
@@
 Intensifying droughts threaten tropical forests, a biome that plays a disproportionately large role in the global carbon and energy cycle, and supports half of the global biodiversity [@Bonan:2008; @Wright:20050th].
-Tropical forests are considered an especially drought-vulnerable biome based on current estimates of trees' hydraulic safety margins [@Choat:2012; @Oliveira:2019].
+Tropical forests are considered an especially drought-vulnerable biome based on current estimates of trees' hydraulic safety margins [@Choat:2012; but see @Oliveira:2019].
 Widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak:2018; @Hilker:2014; @Phillips:2010].
@@
 Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism.
-Lack of data on trees rooting or water-sourcing depths has been a key bottlneck, followed by data on water-availability at those depths.
-A recent meta-analysis documented maximum rooting depths for 318 tree species known for less than 1% of the World’s 45-53 thousand tree species [@Fan:2017; @Silk:2000].
+This mechanism assumes that water availability varies belowground by vertical depth and so do rooting depths of co-occuring trees, by species and size.
+Consequently trees' in the same forest differ in their drought experience, and thus in their responses in term of growth and mortality.
+A key bottlneck in testing this mechanism community-wide has been lack of data on trees' rooting or water-sourcing depths, followed by data on water-availability at those depths.
+A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 1% of the World’s 45-53 thousand tree species [@Fan:2017; @Silk:2000].
 Further, only a small fraction of those are tropical, even though ~95% of the World’s tree diversity resides in the tropics [@Silk:2000].
-Tracer techniques like stable isotopes provide an indirect way of trees' water sourcing depths in which isotopic composition of water in the non-transpiring plant tissue is matched with corresponding gradient in the vertical soil depth, but such data is also available only for ~0.05% of World's tree species [@Evaristo:2016].
-Community-scale data collection for rooting or water-sourcing depths is especially challenging for species-rich tropical forests.
+Use of stable isotopes of water as a tracer provides an indirect way of trees' water sourcing depths by matching isotopic concentration in plant tissue to it's gradient vertically belowground.
+Such data is available only for ~0.05% of World's tree species [@Evaristo:2016].
+In general, community-scale data collection for rooting or water-sourcing depths is especially challenging for species-rich tropical forests.
+Further, to estimate plant available water in the vertical soil column requires hydrological data and modeling, but these are rarely co-located with forest inventory sites (but see Johnson, Porporato).
 
 \
-Traditionally, species are thought to employ three general mechanisms to deal with drought via a variety of associated traits: (1) avoidance--dry season dormancy via deciduous leaf-habit, (2) delay--greater water access via deep rootedness or valley/riparian habitat preference; or reduced water loss via water storage and stomatal control, and (3) physiological tolerance--continued physiological function at low leaf water potential (Poorter 2008).
-Recent understanding of
+Here, we obtained daily dynamics of plant water potential ($\Psi_{soil, z}$) in the whole-soil column of a neo-tropical forest (BCI, Panama) with a locally parametarised and calibrated 1-D hydrological water balance model.
+We parametarised this model using local soil properties such as hydraulic conductivity by depth and water retention curves, and calibrated it against new data on soil moisture dynamics by depth, as well as catchment evapotranspiration and stream discharge.
+We then estimated the effective rooting depths for xx large, canopy tree species as the depth ($z$) whose soil water dynamics ($\Psi_{soil, z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015) with a novel inverse model.
+This model incorporated impact of atmospheric drought (VPD) on growth as well as new data on species-specific leaf hydraulic conductivity responses to water stress ($K_{Leaf}$ vs.
+$\Psi_{x}$).
+As vulnerability curves were collected for a small fraction of species (hard trait), we developed a new model that predicted the parameters of these curves from Leaf Mass Area and Wood Specific Gravity i.e.
+leveraging soft traits available community-wide.
+We evaluated the inversely estimated ERD against exisiting data on plant and soil water isotopes ($\delta^2H$).
 
-Trees' form and function are outcome of the environment in which trees have evolved. And as cliamte changes  and Traits link to plant environment - trait - performance 
-- traits in global change as environment changes
-- trait-demography are traits really functional (link with plant performance and fitness)
-Understanding and explaining the belowground mechanisms and trade‐offs that drive variation in root traits, resource acquisition and plant performance across species
-
-Differential dynamics of water environment by depth
-
-- Safety-efficiency trade-off in hydraulic architecture--why deep rooted trees may be both acquisitive with unsafe xylem, yet an overall safe strategy--so long as their high resource environment is stable.
-- consequence on performance
-- the problem of large trees--large tree mortality in the tropics
-
-\
-In this study, what we do differently: which soil depth dynamics link dynamics of water dynamics in whole-soil column, atmospheric drought via
-...This study increased the mean number of tropical forest species with data on Percent Loss of Hydraulic Conductivity data by xx %
-
-\
-We test the hypotheses that: in this hyperdiverse forest, interspecific variation in effective rooting depths of canopy trees aligns with a conservative-acquisitive trade-off axis such that trees with deeper effective rooting depths rely on a relatively stable, larger water resource and are thus associated with acquisitive hydraulic architecture: with higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrow hydraulic safety margins.
-Despite having a drought-vulnerable xylem, in this humid forest availability of a reliable water at depth allows trees with deeper effective rooting depths to escape the meteorological drought, resulting in lower mortality rates compared to trees with shallow effective rooting depths.
+We test three hypotheses under an overarching one that interspecific variation in effective rooting depths of trees aligns with a conservative-acquisitive trade-off axis.
+In this forest (i) dynamics of soil water potential at depth does not cross catastrophic levels even through multi-annual droughts because of high total annual precipitation to evapotranspiration ratio; (ii) canpoy trees with deeper effective rooting depths (ERD) show acquisitive hydraulic architecture: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins; finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure.
+To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and mortality outcomes for a tropical forest community.
 
@@
 \
-In the following sections we describe the study site, the models and data used to estimate and evaluate (1) the soil matric potential dynamics by depth over the forests’ rooting zone, (2) species-specific percent loss of leaf conductivity curves (PLC), and (3) species-specific effective rooting depths.
+In the following sections we describe the study site, the models and data used to estimate and evaluate (1) the soil water potential dynamics by depth over the forests’ rooting zone, (2) species-specific percent loss of leaf conductivity curves (PLC), and (3) species-specific effective rooting depths.
 These are followed by analyses to address the proposed hypotheses.
@@
 
-```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-iso}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE; m) versus dry season stable isotopic concentration $\\delta$^2^H (mean $\\pm$ (add error); ‰) for ", length(subset(ml.rsq.combine.sub, source == "Meinzer et al.1999 Fig.
-4" & corr.func == "gr.Psi.VPD")$sp)," canopy species from Barro Colorado Island (data from [@Meinzer:1999]")}
+```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-iso}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE; m) versus dry season stable isotopic concentration $\\delta$^2^H (mean $\\pm$ (add error); ‰) for ", length(unique(erd.data$sp))," canopy species from Barro Colorado Island (data from [@Meinzer:1999]")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_phenology_Meinzer_gr.Psi.VPD.jpeg"))
@@
 
-We calculated mortality rates for large trees (>=10 cm DBH) of xx canopy species for each of the seven census interval ($int$) in the 33 year record (1985-2015; $Mort_{int}$) and a species mean mortality rate across those intervals ($Mort_{mean} = \frac{1}{int}\sum_{int=1}^{6} Mort_{int}$).
+We calculated mortality rates for large trees (>=10 cm DBH) of xx canopy species for each of the seven census interval ($int$) in the 33 year record (1985-2015; $Mort_{int}$) and a species mean mortality rate across those intervals ($Mort_{mean} = \frac{1}{int}\sum_{int=1}^{6} {Mort_{int}}$).
 To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$).
@@
 ```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus hydraulic propertiese for seven large, canopy species found on Barro Colorado Island (Panama).
-*R*^2^ and *P*-values are given in panel insets."}
+*R*^2^ and *P*-values are given in panel insets.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/Select_traits_depth.jpeg"))
@@
 
-```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{x, z}$ was below $\\Psi_{80,Leaf}$ over 1990-2015, where z is the soil depth matching species ERD--for ", ," canopy species found on Barro Colorado Island.
+```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{x, z}$ was below $\\Psi_{80,Leaf}$ over 1990-2015, where z is the soil depth matching species ERD--for xx canopy species found on Barro Colorado Island.
 (Are *R*^2^ and *P*-values needed here?)")}
@@
 - Temperate rooting depths shallow?  Shenk and Jackson
+
+- lightening explains 1/3rd of mortality
+- forest is realtively stable over 40 years.
 
@@
 $\\Psi$~Leaf~ for `r length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))` tree species from Panama, color coded by Wood Specific Gravity (units)."}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/std.k.commkmax_by_psi_color_by_Specific Gravity_predicted_AB.jpeg"))
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB_for_data_sp.jpeg"))
 ```

### 2020-07-12 manuscript.rmd
Added intro para on realised water-environments

@@
 
-```{r, include=FALSE}
+```{r, include=FALSE, echo = FALSE}
 if(!require("pacman", quietly=TRUE))install.packages("pacman")
@@
 
-```{r setup}
+```{r setup, echo = FALSE}
 library(knitr)
@@
 \
-T: +1(505)
+T: +1(505) 665 1647
 
@@
 
-Keywords: rooting depths, drought tolerance, mortality, tropical forest, climate change
+Keywords: rooting depths, drought tolerance, mortality, tropical forest, climate change; life-history strategies
 
@@
 Intensifying droughts threaten tropical forests, a biome that plays a disproportionately large role in the global carbon and energy cycle, and supports half of the global biodiversity [@Bonan:2008; @Wright:20050th].
-Tropical forests are considered an especially drought-vulnerable biome based on current estimates of trees' hydraulic safety margins [@Choat:2012; but see @Oliveira:2019].
+Tropical forests are considered an especially drought-vulnerable biome based on observed ranges of trees' hydraulic safety margins [@Choat:2012; but see @Oliveira:2019].
 Widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak:2018; @Hilker:2014; @Phillips:2010].
@@
 
+Trait-based plant ecology aims to predict impacts of environmental change on community composition and function using relevant plant traits as currency.
+To reach this goal it is imperative to establish relationships between plant's environment, traits and demographic outcomes, but this remains to be established for trees' realised water environments and water-access through droughts.
+Huge efforts have been devoted to linking same-for-all-trees drought indices, above-ground traits and demographic rates.
+These have resulted in above-ground hydraulic traits such as proxies of hydraulic failure as suggested indicators of mortality risk.
+A recent meta-analysis found greater drought-induced mortality rates in species with high vulnerability to drought-induced embolism ($\Psi_{88, Stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88, Stem}$), but counter examples also exist (Hoffmann 2011; Paddock 2013), indicating whether plants actually reach critical thresholds of hydraulic failure depends on their realised drought intensity.
+A rare study that accounted for realised water availability for four tree species in a rare, centennial drought (Texas, USA) found that deep rooted species died less despite being more vulnerable to xylem embolism (high $\Psi_{88, Stem}$) and having narrower safety-margins ($\Psi_{min} - \Psi_{88, Stem}$).
+A species sourcing deep water showed lower hydraulic failure and lower mortality in an extreme drought in an Italian Karst system even though it had narrower hydrulic safety margin and high $\Psi_{88, Stem}$ than other two species limited to shallow-water.
+Chitra-Tarak et al.
+(2018) proposed a general model to link realised water availability to demographic outcomes through droughts at tree community-scale.
+This study found that species accessing deep-water in a tropical dry forest study system died more in a rare, centennial drought compared to shallow-water species.
+Authors concluded that this was consistent with expectation of deep water usage as an aquisitive strategy relying on stable and reliable water resource at depth, which exhausted in the multi-annual drought resulting in reduced survival, compared to persistent species associated with shallow-water that was relatively more available via infiltration during this drought.
+However, species estimates of water-access modelled by this study were not corroborated with rooting/water-sourcing depths, or related to hydraulic traits.
+
 \
 Here, we obtained daily dynamics of plant water potential ($\Psi_{soil, z}$) in the whole-soil column of a neo-tropical forest (BCI, Panama) with a locally parametarised and calibrated 1-D hydrological water balance model.
-We parametarised this model using local soil properties such as hydraulic conductivity by depth and water retention curves, and calibrated it against new data on soil moisture dynamics by depth, as well as catchment evapotranspiration and stream discharge.
 We then estimated the effective rooting depths for xx large, canopy tree species as the depth ($z$) whose soil water dynamics ($\Psi_{soil, z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015) with a novel inverse model.
-This model incorporated impact of atmospheric drought (VPD) on growth as well as new data on species-specific leaf hydraulic conductivity responses to water stress ($K_{Leaf}$ vs.
-$\Psi_{x}$).
-As vulnerability curves were collected for a small fraction of species (hard trait), we developed a new model that predicted the parameters of these curves from Leaf Mass Area and Wood Specific Gravity i.e.
-leveraging soft traits available community-wide.
+This model incorporated impact of atmospheric drought (VPD) on growth and new data on species leaf vulnerability curves ($K_{Leaf}$ vs.
+$\Psi_{x}$), which we scaled-up to community level via scaling relationships with commonly available traits.
 We evaluated the inversely estimated ERD against exisiting data on plant and soil water isotopes ($\delta^2H$).
 
-We test three hypotheses under an overarching one that interspecific variation in effective rooting depths of trees aligns with a conservative-acquisitive trade-off axis.
-In this forest (i) dynamics of soil water potential at depth does not cross catastrophic levels even through multi-annual droughts because of high total annual precipitation to evapotranspiration ratio; (ii) canpoy trees with deeper effective rooting depths (ERD) show acquisitive hydraulic architecture: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins; finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure.
+We test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns them on a conservative-acquisitive trade-off axis.
+Specifically, in this forest (i) dynamics of soil water potential at depth does not cross catastrophic levels even during multi-annual droughts because of high total annual precipitation to evapotranspiration ratio; (ii) trees with deeper effective rooting depths (ERD) show acquisitive hydraulic architecture: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins; finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure.
 To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and mortality outcomes for a tropical forest community.
@@
 \
-We ran ELM-FATES for these 10,000 ensembles with hourly climate drivers from 1985-2018 initialised with the stand structure of the 50 ha plot in a single site mode (0.1$^\circ$ x 0.1$^\circ$ centered at 9.1530 $^\circ$N, 280.2539 $^\circ$E).
+We ran ELM-FATES for these 10,000 ensembles with hourly climate drivers from 1985-2018 initialised with the stand structure of the 50 ha plot in a single site mode ($0.1^\circ$ x $0.1^\circ$ centered at $9.1530^\circ$N, $280.2539^\circ$E).
 As our key interest here was in soil water availability rather than forest dynamics, we ran ELM-FATES in simplified modes: static stand structure and without plant hydrodynamics, and with a single Plant Functional Type (PFT) of evergreen trees.
@@
 2011) |
-| 2 | fates_leaf_ slatop | Specific Leaf Area (SLA) at the top of canopy, projected area basis | 0.0042 - 0.0400 $m^2/gC$ | Based on the observed range of Leaf Mass Area (LMA)--24.97 to 235.8 $g/m^2$-- for individual tree variation across 51 tree species in Barro Colorado Island (this study), and assumption of 50% carbon content in biomass.
+| 2 | fates_leaf_ slatop | Specific Leaf Area (SLA) at the top of canopy, projected area basis | 0.0042 - 0.0400 $m^2gC^{-1}$ | Based on the observed range of Leaf Mass Area (LMA)--24.97 to 235.8 $gm^{-2}$-- for individual tree variation across 51 tree species in Barro Colorado Island (this study), and assumption of 50% carbon content in biomass.
 |
-| 3 | fates_leaf_ vcmax25top | Maximum carboxylation rate of Rubisco at 25 $^\circ$C, canopy top | 22.7 - 92.5 umol $CO_2/m^2/s$ | Observed range of Vcmax at 25 $^\circ$C for tropical tree species under relative radiation of 50% or greater, thereby excluding highly shaded leaves.
+| 3 | fates_leaf_ vcmax25top | Maximum carboxylation rate of Rubisco at $25^\circ$C, canopy top | 22.7 - 92.5 umol $CO^2m^2s^{-1}$ | Observed range of Vcmax at  $25^\circ$C for tropical tree species under relative radiation of 50% or greater, thereby excluding highly shaded leaves.
 Values derived with a conversion specific for CLM model are used.
@@
 | 8 | fmax | The maximum fractional saturated area |  unitless | Empirical | 
-| 9 | HKSAT | Soil hydraulic conductivity profile | 0.007 - 0.014 mm/s for $\le$ 12.5 cm; 5.56e-05, 0.0003 for $\ge$ 60 cm | To account for high macroporosity and direct flow paths in tropical soils(Broedel et al.
-2017; Tomasella and Hodnett 1996)(Kinner and Stallard 2004; Broedel et al.
-2017; Tomasella and Hodnett 1996) For Conrad catchment observed median Ksat (95%CI) for 12.5 cm depth is 38.3 mm/hr (25.4-51.2, *n* = 75), while for 60 cm depth 0.7 (0.2, 1.2, *n* = 40) mm/hr.
-For reference a storm with 12.5 mm/hr rainfall intensity has a 0.2 probability of occuring in any given rainfall event (Kinner and Stallard 2004).
+| 9 | HKSAT | Soil hydraulic conductivity profile | 0.007 - 0.014 mms{-1} for $\le$ 12.5 cm; 5.56e-05, 0.0003 for $\ge$ 60 cm | To account for high macroporosity and direct flow paths in tropical soils(Broedel et al.
+2017; Tomasella and Hodnett 1996) (Kinner and Stallard 2004; Broedel et al.
+2017; Tomasella and Hodnett 1996) For Conrad catchment observed median Ksat (95% CI) for 12.5 cm depth is 38.3 mm/hr (25.4-51.2, *n* = 75), while for 60 cm depth 0.7 (0.2, 1.2, *n* = 40) mmhr^{-1}^.
+For reference a storm with 12.5 mmhr^{-1}^ rainfall intensity has a 0.2 probability of occuring in any given rainfall event (Kinner and Stallard 2004).
 |
@@
 2017; Tomasella and Hodnett 1996) |
-| 11 | fpi_max | Maximum interception fraction of precipitation | 0.05-0.44 unitless | Based on throughfall data from (Zimmermann et al.
+| 11 | fpi_max | Maximum interception fraction of precipitation | 0.05 - 0.44 unitless | Based on throughfall data from (Zimmermann et al.
 2010) and precipitation data for BCI from STRI Physical Monitoring program, defined for precipitation events greater than 10 mm.
@@
 
-### Appendix A
-
-#### 1. 
 # References

### 2020-07-13 manuscript.rmd
Added Inverse model methods section Other smaller edits

@@
 output:
-  bookdown::word_document2:
-        reference_docx: manuscriptstyle.docx
-        fig_caption: yes
-  bookdown::pdf_document2: default
   bookdown::html_document2:
@@
     theme: flatly
+    fig_caption: yes
+  bookdown::pdf_document2: default
+  bookdown::word_document2:
+        reference_docx: manuscriptstyle.docx
         fig_caption: yes
@@
 - Deep water access (WA~Deep~) is believed to be an important, drought-tolerance mechanism, but under-studied due to lack of data.
-Vulnerability to embolism ($\Psi_{88, Stem}$) and safety margins ($\Psi_{min} - \Psi_{88, Stem}$) are shown to be predictors of drought-induced mortality risk, but deep water access may delay plants from reaching these critical thresholds.
+Vulnerability to embolism ($\Psi_{88,~Stem}$) and safety margins ($\Psi_{min} - \Psi_{88,~Stem}$) are shown to be predictors of drought-induced mortality risk, but deep water access may delay plants from reaching these critical thresholds.
 Here we test its role in drought-tolerance strategies of a tropical forest community (Panama) with a novel data-model approach.
-- We inversely modelled Effective Rooting Depth (ERD) for 32 large, canopy tree species by linking species growth dynamics (xx trees, 6 censuses, 1990-2015) to VPD, $\Psi_{soil}$ by depth in the whole-soil column, and a novel community-scale model of species vulnerability curves ($K_{Leaf}$ vs.
-$\Psi_{x}$) linking data on the same for 21 species with community-scale softer traits.
+- We inversely modelled Effective Rooting Depth (ERD) for 32 large, canopy tree species by linking species growth dynamics (xx trees, 6 censuses, 1990-2015) to VPD, $\Psi_{x,~z}$ by depth $z$in the whole-soil column, and a novel community-scale model of species vulnerability curves ($K_{Leaf}$ vs.
+$\Psi_{x,~Leaf}$) linking data on the same for 21 species with community-scale softer traits.
 We evaluated ERD by leveraging existing isotopic tracer data ($\delta^2H$).
@@
 A key bottlneck in testing this mechanism community-wide has been lack of data on trees' rooting or water-sourcing depths, followed by data on water-availability at those depths.
-A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 1% of the World’s 45-53 thousand tree species [@Fan:2017; @Silk:2000].
-Further, only a small fraction of those are tropical, even though ~95% of the World’s tree diversity resides in the tropics [@Silk:2000].
+A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the World’s 65 thousand tree species [@Fan:2017; @Beech:2017].
+Further, only a small fraction of those are tropical, even though ~60-80 % (40 - 53 throusand species) of the World’s tree diversity resides in the tropics [@Slik:2015].
 Use of stable isotopes of water as a tracer provides an indirect way of trees' water sourcing depths by matching isotopic concentration in plant tissue to it's gradient vertically belowground.
@@
 In general, community-scale data collection for rooting or water-sourcing depths is especially challenging for species-rich tropical forests.
-Further, to estimate plant available water in the vertical soil column requires hydrological data and modeling, but these are rarely co-located with forest inventory sites (but see Johnson, Porporato).
+Further, to estimate plant available water in the vertical soil column requires hydrological data and modeling, but these are rarely co-located with forest inventory sites (but see [@Johnson:2018fad;@Kumagai:2011;@Ivanov:2012].
 
@@
 These have resulted in above-ground hydraulic traits such as proxies of hydraulic failure as suggested indicators of mortality risk.
-A recent meta-analysis found greater drought-induced mortality rates in species with high vulnerability to drought-induced embolism ($\Psi_{88, Stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88, Stem}$), but counter examples also exist (Hoffmann 2011; Paddock 2013), indicating whether plants actually reach critical thresholds of hydraulic failure depends on their realised drought intensity.
-A rare study that accounted for realised water availability for four tree species in a rare, centennial drought (Texas, USA) found that deep rooted species died less despite being more vulnerable to xylem embolism (high $\Psi_{88, Stem}$) and having narrower safety-margins ($\Psi_{min} - \Psi_{88, Stem}$).
-A species sourcing deep water showed lower hydraulic failure and lower mortality in an extreme drought in an Italian Karst system even though it had narrower hydrulic safety margin and high $\Psi_{88, Stem}$ than other two species limited to shallow-water.
+A recent meta-analysis found greater drought-induced mortality rates in species with high vulnerability to drought-induced embolism ($\Psi_{88,~Stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~Stem}$), but counter examples also exist (Hoffmann 2011; Paddock 2013), indicating whether plants actually reach critical thresholds of hydraulic failure depends on their realised drought intensity.
+A rare study that accounted for realised water availability for four tree species in a rare, centennial drought (Texas, USA) found that deep rooted species died less despite being more vulnerable to xylem embolism (high $\Psi_{88,~Stem}$) and having narrower safety-margins ($\Psi_{min} - \Psi_{88,~Stem}$).
+A species sourcing deep water showed lower hydraulic failure and lower mortality in an extreme drought in an Italian Karst system even though it had narrower hydrulic safety margin and high $\Psi_{88,~Stem}$ than other two species limited to shallow-water.
 Chitra-Tarak et al.
@@
 \
-Here, we obtained daily dynamics of plant water potential ($\Psi_{soil, z}$) in the whole-soil column of a neo-tropical forest (BCI, Panama) with a locally parametarised and calibrated 1-D hydrological water balance model.
-We then estimated the effective rooting depths for xx large, canopy tree species as the depth ($z$) whose soil water dynamics ($\Psi_{soil, z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015) with a novel inverse model.
+Here, we obtained daily dynamics of plant water potential ($\Psi_{x,~z}$) by depth $z$ in the whole-soil column of a neo-tropical forest (BCI, Panama) with a locally parametarised and calibrated 1-D hydrological water balance model.
+We then estimated the effective rooting depths for xx large, canopy tree species as the depth ($z$) whose soil water dynamics ($\Psi_{x,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015) with a novel inverse model.
 This model incorporated impact of atmospheric drought (VPD) on growth and new data on species leaf vulnerability curves ($K_{Leaf}$ vs.
-$\Psi_{x}$), which we scaled-up to community level via scaling relationships with commonly available traits.
+$\Psi_{x,~Leaf}$), which we scaled-up to community level via scaling relationships with commonly available traits.
 We evaluated the inversely estimated ERD against exisiting data on plant and soil water isotopes ($\delta^2H$).
@@
 \
-We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterisation in terms of long-term forest demography, plant functional ecology, hydrology, including availability of an independent isotopic tracer study on trees’ water-sourcing depths (Meinzer et al.).
+We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterisation in terms of long-term forest demography, plant functional ecology, hydrology, including availability of an independent isotopic tracer study on trees’ water-sourcing depths [@Meinzer:1999].
 The 1,560 hectare Barro Colorado Island was formed after the river Chagres was dammed in 1914 that created Gatun Lake, the main channel of the Panama Canal.
@@
 The entire Island is forested and classified as a tropical moist forest.
-Although percentage of canopy species that are at least partly deciduous during the dry season is 28%, the maximum percentage of deciduous individuals in the canopy is only 6.3%, and the percentage of crown area that is deciduous is estimated at 9.7% (Condit et al., 2000).
-Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 (Condit, 1998; Hubbell et al., 1999) (Anderson-Texeira xx).
+Although percentage of canopy species that are at least partly deciduous during the dry season is 28%, the maximum percentage of deciduous individuals in the canopy is only 6.3%, and the percentage of crown area that is deciduous is estimated at 9.7% [@Condit:2000].
+Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit:1998395; @Hubbell:1999; @Anderson-Teixeira:2014].
 
 \
-In the 50 ha plot all stems above 10 mm are mapped, tagged with a unique number, identified to species level and measured every 5 years through 1985-2015 for diameter growth, mortality as well as recruitment of new stems into the 10 mm size class (Hubbell, S.P., Condit, R., and Foster, R.B., 2010; Condit et al., 2012).
+In the 50 ha plot all stems above 10 mm are mapped, tagged with a unique number, identified to species level and measured every 5 years through 1985-2015 for diameter growth, mortality as well as recruitment of new stems into the 10 mm size class [@Condit:2017].
 This inventory represents 321 woody species across the eight censuses.
-Aside from a 2 ha patch of ~100 year old secondary forest at the north-eastern corner of the plot the remaining 48 ha represent old growth forest that was never clear cut for agriculture (Piperno 1992).
-The plot (100 m x 50 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) (Harms 2001).
+Aside from a 2 ha patch of ~100 year old secondary forest at the north-eastern corner of the plot the remaining 48 ha represent old growth forest that was never clear cut for agriculture.
+The plot (100 m x 50 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001].
 The plot covers most of the upper catchment of Conrad Creek.
@@
 Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding.
-Detailed descriptions of the climate, geology, flora and fauna of BCI can be found in Croat (1978), Leigh et al.
-(1982) and Gentry (1990).
+Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere.
 
@@
 To estimate dynamics of water availability by depth over the forest’s root zone, we implemented a 1-D hydrological model, the Energy Exascale Earth System Land Model (ELM), coupled with a Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES).
-FATES is a state-of-the-art, community-based, open-source model widely used for studying climate-vegetation interactions.
+FATES is a state-of-the-art, community-based, open-source model increasingly used for studying climate-vegetation interactions.
 FATES is considered a next generation vegetation dynamics model for ELM, with a size-structured group of plants (cohorts) and successional trajectory-based patches using the ecosystem demography approach......
-(How ELM talks to FATES)...Detailed descriptions can be found elsewhere (Koven et al., 2020) Fisher et al ).
+(How ELM talks to FATES)...Detailed descriptions can be found elsewhere [@Koven:2020; @Fisher: ).
 
@@
 
-We measured ($K_{x, Leaf}$ vs.
-$\Psi_{x, Leaf}$) on adult trees of `r length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))` common species (see SI x).
-We fitted exponential, sigmoidal, logistic and exponential form of equations to species-level data as prescribed by [@GUYOT:2011] and based on AIC criteria chose exponential form as the best-fitting one (Fig.
-\@ref(fig:obs-plc-curves)), described as
+We measured ($K_{x,~Leaf}$ vs.
+$\Psi_{x,~Leaf}$) on adult trees of `r length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))` common species (see SI x).
+We fitted exponential, sigmoidal, logistic forms of equations to species-level data as prescribed by [@Guyot:2011] and based on AIC criteria chose exponential form (Fig.
+S\@ref(fig:obs-plc-curves)), described as
 
 \begin{equation}
-K_{Leaf} = y_0 + Ae^{-B\Psi_{Leaf}} + error
+K_{Leaf} = Ae^{-B\Psi_{Leaf}} + error
 (\#eq:vc-exp)
@@
 R^2^ = `r signif(glance(k_by_psi.models[[2]])$adj.r.squared, 2)`; *P* = `r  signif(glance(k_by_psi.models[[2]])$p.value, 2)`; also see Fig.
-\@ref(fig:ab-fit)).
+S\@ref(fig:ab-fit)).
 $WSG$ & $LMA$ explained large variation in parameter $B$ (Eq.
@@
 
-```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-plc-curves} Modelled standardized 'Vulnerability curves' ($K_{x, Leaf}$ vs.
-$\\Psi_{x, Leaf}$) at community-scale (", nrow(data.model.AB$sp)," species) based on species data on $WSG$ & $LMA$ and their relationships with vulnerability curve parameters measured on ", sp.soft.filled$AB.data.model," species.
+```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-plc-curves} Modelled standardized 'Vulnerability curves' ($K_{x,~Leaf}$ vs.
+$\\Psi_{x,~Leaf}$) at community-scale (", nrow(data.model.AB$sp)," species) based on species data on $WSG$ & $LMA$ and their relationships with vulnerability curve parameters measured on ", sp.soft.filled$AB.data.model," species.
 Each line represents a species, color coded by $WSG$.")}
@@
 
+With 
+\
+\begin{equation}
+K_{Leaf,z,n} = Ae^{-B\Psi_{z,n}}
+(\#eq:k-soil)
+\end{equation}
+
+\begin{equation}
+Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{Leaf,z,n} + error
+(\#eq:M1)
+\end{equation}
+
+\begin{equation}
+Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{Leaf,n} StdVPD_{n} + error
+(\#eq:M2)
+\end{equation}
+
+\begin{equation}
+Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{Leaf,z,n} LeafFrac_{doy} + error
+(\#eq:M3)
+\end{equation}
+
+\begin{equation}
+Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{Leaf,z,n} LeafFrac_{doy}StdVPD_{n} + error
+(\#eq:M4)
+\end{equation}
+
 ### Model parameterization
+
+Growth Data: To minimise effect of light variation among trees and census intervals we selected only large trees (>= 30 cm DBH) and species whose maximum height was at least 30 m (hereafter, canopy species).
+We chose only those trees for which diameter (DBH) record was present across all size censuses (1990, 19905, 2000, 2005, 2010, 2015).
+We calculated individual tree ($i$) growth rates ($Gr_{i,t}$ cmyr^{-1}) for each of the five census intervals ($t$).
+As tree growth rates increase with size, to keep growth rates comparable across census intervals, we obtained residuals from a DBH model of tree growth based on a trees' DBH at the beginning of a census onterval ($GrResid_{i,t} = Gr_{i,t} - f(DBH_{i, t})$).
+The diameter model of growth was fitted based on all trees' data pooled together as there was insufficient species-specific data for large trees (SI ).
+We then standardised (centered and scaled) an individual tree time series ($GrResid_{i,t}$) and obtained species-level (*n* = 32) times-series of medians ($Growth_{t}$) that had data on at least three trees (median = xx).
+
+As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between GPP (gCm^2^day^-1^) and VPD (kPa; $GPP = f(VPD) + error$;  SI) and used standardised $f(VPD)$ (range 0-1) as $StdVPD$ in Eqs.
+\@ref(eq:M2)- \@ref(eq:M4).
+
+Leaf fraction...
+
+For each model (\@ref(eq:M2)- \@ref(eq:M4)) and species combination ($sp$), we obtained goodness-of-fit (*R*^2^) and Pearson's $\rho$ for all soil depths $z$, and for each sp, model we chose the $z$ with highest *R*^2^ as the Effective Rooting Depth (ERD); after retaining species with positive $\rho$ and *R*^2^ >= 0.1.
 
@@
 
-Figure \@ref(fig:erd-iso)
+For each model, species ERD was regressed against species Xylem sap $\delta^2H$ (%~$\circ$~; *n* = 7).
+We used Xylem sap $\delta^2H$ values given in [@Meinzer:2001] for dry-season (March 1997) as these showed largest seasonal divergence among species, with a corresponding gradient of $\delta^2H$ in the soil.
+We removed species that are deciduous in March as xylem sap $\delta^2H$ of these species may not be linked with water sourced in March and thus may not comparable to other species with leaves present.
+Fig.
+S\@ref(fig:iso-data).
+We also removed *Guapira* *spp*, whose xylem sap value fell in the uncertain ranges of soil depth.
+
+Models with VPD and without Leaf Fraction (Eqs.
+\@ref(eq:M2)) had highest goodness-of fit (*R*^2^ = 0.51); followed by model with both VPD and Leaf Fraction (Eq.
+\@ref(eq:M4); Fig.
+S\@ref(fig:other-erd-mods)).
+Models with leaf fraction had only three species with sufficient, indvidual species-level goodness-of-fit (Eqs.
+\@ref(eq:M3 & eq:M4) versus seven for models without (Eqs.
+\@ref(eq:M1) & \@ref(eq:M2)).
+For rest of the analyses we therefore used species ERD estimates from model (\@ref(eq:M2)) as it was also a parsimonious model compared to \@ref(eq:M4) (Fig.
+\@ref(fig:erd-iso)).
 
@@
 
-To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD with maximum stem hydraulic conductance  ($K_{max, Stem}$), leaf Turgor Loss Point ($\Psi_{TLP}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88, Stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88, Stem}$).
+To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD with maximum stem hydraulic conductance  ($K_{max, Stem}$), leaf Turgor Loss Point ($\Psi_{TLP}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~Stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~Stem}$).
 
@@
 
-We calculated mortality rates for large trees (>=10 cm DBH) of xx canopy species for each of the seven census interval ($int$) in the 33 year record (1985-2015; $Mort_{int}$) and a species mean mortality rate across those intervals ($Mort_{mean} = \frac{1}{int}\sum_{int=1}^{6} {Mort_{int}}$).
+We calculated mortality rates for large trees (>=10 cm DBH) of xx canopy species for each of the seven census interval ($t$) in the 33 year record (1985-2015; $Mort_{t}$) and a species mean mortality rate across those intervals ($Mort_{mean} = \frac{1}{t}\sum_{t=1}^{6} {Mort_{t}}$).
 To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$).
@@
 In this forest modelled Effective Rooting Depths (ERD) for the 32 large, canopy tree species varied from 0.5 m to 13 m (Figure \@ref(fig:sp-erd)).
-For the subset of species for which hydraulic properties were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max, Stem}$; *R*^2^ = 0.39, *P* = 0.098, *n* = 8; Figure \@ref(fig:erd-hyd)A), less negative leaf Turgor Loss Point ($\Psi_{TLP}$); *R*^2^ = 0.6, p = 0.024, *n* = 8; Figure \@ref(fig:erd-hyd)B), greater vulnerability to embolism from cavitation with less negative ($\Psi_{88, Stem}$) (*R*^2^ = 0.72, p = 0.007, *n* = 8; Figure \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88, Stem}$; *R*^2^ = 0.74, p = 0.013, *n* = 8; Figure \@ref(fig:erd-hyd)D).
+For the subset of species for which hydraulic properties were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max, Stem}$; *R*^2^ = 0.39, *P* = 0.098, *n* = 8; Figure \@ref(fig:erd-hyd)A), less negative leaf Turgor Loss Point ($\Psi_{TLP}$); *R*^2^ = 0.6, p = 0.024, *n* = 8; Figure \@ref(fig:erd-hyd)B), greater vulnerability to embolism from cavitation with less negative ($\Psi_{88,~Stem}$) (*R*^2^ = 0.72, p = 0.007, *n* = 8; Figure \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~Stem}$; *R*^2^ = 0.74, p = 0.013, *n* = 8; Figure \@ref(fig:erd-hyd)D).
 
-```{r sp-erd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap= paste0("\\label{lab-sp-erd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for ", length(unique(erd.data$R2)) ," large, canopy species of Barro Colorado Island.
+```{r sp-erd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap= paste0("\\label{lab-sp-erd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for 32 large, canopy species of Barro Colorado Island.
 See Table S2 for scientific names of species codes")}
@@
 ```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{x, z}$ was below $\\Psi_{80,Leaf}$ over 1990-2015, where z is the soil depth matching species ERD--for xx canopy species found on Barro Colorado Island.
+EAch circle resepresents a species.
+Circles are jittered vertically around the named depths (ticks) to be able to distinguish visually.
 (Are *R*^2^ and *P*-values needed here?)")}
@@
 | -- | ------ | ------------- | ------------- |  ------------------------------------------------------------------------------------------------- |
-| 1 | fates_leaf_ BB_slope | stomatal slope parameter, as per Ball-Berry | 4 - 16 Unitless  | The range of Ball Berry parameter fitted across several studies as compiled in Table 2 of (Medlyn et al.
-2011) |
+| 1 | fates_leaf_ BB_slope | stomatal slope parameter, as per Ball-Berry | 4 - 16 Unitless  | The range of Ball Berry parameter fitted across several studies as compiled in Table 2 of [@Medlyn:2012] |
 | 2 | fates_leaf_ slatop | Specific Leaf Area (SLA) at the top of canopy, projected area basis | 0.0042 - 0.0400 $m^2gC^{-1}$ | Based on the observed range of Leaf Mass Area (LMA)--24.97 to 235.8 $gm^{-2}$-- for individual tree variation across 51 tree species in Barro Colorado Island (this study), and assumption of 50% carbon content in biomass.
 |
-| 3 | fates_leaf_ vcmax25top | Maximum carboxylation rate of Rubisco at $25^\circ$C, canopy top | 22.7 - 92.5 umol $CO^2m^2s^{-1}$ | Observed range of Vcmax at  $25^\circ$C for tropical tree species under relative radiation of 50% or greater, thereby excluding highly shaded leaves.
+| 3 | fates_leaf_ vcmax25top | Maximum carboxylation rate of Rubisco at $25^\circ$C, canopy top | 22.7 - 92.5 umol $CO^2m^2s^{-1}$ | Observed range of $V_{cmax}$ at  $25^\circ$C for tropical tree species under relative radiation of 50% or greater, thereby excluding highly shaded leaves.
 Values derived with a conversion specific for CLM model are used.
-(Ali et al.
-2015) |
+[@Ali:2015] |
 | 4 | fates_roota_ par | CLM rooting distribution parameter | 5.9 - 7.4 per m | Parameter a in Eq.
-2 in (Zeng 2001) that regulates the shape of the rooting profile.
-Range corresponds to this parameter specified for BATS (or IGBP) land cover classified Deciduous Broadleaf Trees (5.9 $m^{-1}$ and Evergreen Broadleaf Trees  (7.4 $m^{-1}$) as given in Table 1 (or 2) of (Zeng 2001).
+2 in [@Zeng:2001] that regulates the shape of the rooting profile.
+Range corresponds to this parameter specified for BATS (or IGBP) land cover classified Deciduous Broadleaf Trees (5.9 $m^{-1}$ and Evergreen Broadleaf Trees  (7.4 $m^{-1}$) as given in Table 1 of [@Zeng:2001].
 |
 | 5 | fates_rootb_ par | CLM rooting distribution parameter | 0.217 - 1.956 $m^{-1}$ | Parameter b in Eq.
-2 in (Zeng 2001) that regulates the depth of the rooting profile.
-Chosen range of *b* is derived using this equation so as to fit the observed range of rooting depth (d~r~) of 2 - 18 m for Tropical Deciduous Forest (mean ± Standard Error (SE); 3.7 ± 0.5, *n* = 5 trees; min = 2, max = 4.7)  and Tropical Evergreen Forest (mean $\pm$ S.E.; 7.3 $\pm$ 0.5, n = 3 trees and 3 communities; min = 2, max = 18 m) combined (Canadell et al.
-1996).
-Besides the direct observation of roots at 18 m included by (Nepstad et al.
-1994) in Paragominas, eastern Amazonia that is included in the above study; in Tapajos, eastern Amazonia water extraction by roots was also inferred up to 18 m.
-(Davidson et al.
-Forest Science 2011) |
+2 in [@Zeng:2001] that regulates the depth of the rooting profile.
+Chosen range of *b* is derived using this equation so as to fit the observed range of rooting depth (d~r~) of 2 - 18 m for Tropical Deciduous Forest (mean ± Standard Error (SE); 3.7 ± 0.5, *n* = 5 trees; min = 2, max = 4.7)  and Tropical Evergreen Forest (mean $\pm$ S.E.; 7.3 $\pm$ 0.5, n = 3 trees and 3 communities; min = 2, max = 18 m) combined [@Canadell:1996].
+Besides the direct observation of roots at 18 m included by [Nepstad:1994] in Paragominas, eastern Amazonia that is included in the above study; in Tapajos, eastern Amazonia water extraction by roots was also inferred up to 18 m.
+[Davidson:2011] |
 | 6 | fates_smpsc | Soil water potential at full stomatal closure | 113000 - 242000 mm | Based on observed range of -1.13 to -2.42 MPa for leaf turgor loss point for individual tree variation across 49 BCI tree species.
@@
 | 8 | fmax | The maximum fractional saturated area |  unitless | Empirical | 
-| 9 | HKSAT | Soil hydraulic conductivity profile | 0.007 - 0.014 mms{-1} for $\le$ 12.5 cm; 5.56e-05, 0.0003 for $\ge$ 60 cm | To account for high macroporosity and direct flow paths in tropical soils(Broedel et al.
-2017; Tomasella and Hodnett 1996) (Kinner and Stallard 2004; Broedel et al.
-2017; Tomasella and Hodnett 1996) For Conrad catchment observed median Ksat (95% CI) for 12.5 cm depth is 38.3 mm/hr (25.4-51.2, *n* = 75), while for 60 cm depth 0.7 (0.2, 1.2, *n* = 40) mmhr^{-1}^.
-For reference a storm with 12.5 mmhr^{-1}^ rainfall intensity has a 0.2 probability of occuring in any given rainfall event (Kinner and Stallard 2004).
+| 9 | HKSAT | Soil hydraulic conductivity profile | 0.007 - 0.014 mms{-1} for $\le$ 12.5 cm; 5.56e-05, 0.0003 for $\ge$ 60 cm |  For Conrad catchment observed median $K_{sat}$ (95% CI) for 12.5 cm depth is 38.3 mmhr^{-1}^ (25.4-51.2, *n* = 75), while for 60 cm depth 0.7 (0.2, 1.2, *n* = 40) mmhr^{-1}^.
+For reference, a storm with 12.5 mmhr^{-1}^ rainfall intensity has a 0.2 probability of occuring in any given rainfall event [@Kinner:2004].
 |
-| 10 | HKSAT_ADJ | Adjusting factor for soil hydraulic conductivity | 1 - 8 unitless | To account for high macroporosity and direct flow paths in tropical soils that is not accounted for by small soil core samples (Broedel et al.
-2017; Tomasella and Hodnett 1996)(Kinner and Stallard 2004; Broedel et al.
-2017; Tomasella and Hodnett 1996) |
-| 11 | fpi_max | Maximum interception fraction of precipitation | 0.05 - 0.44 unitless | Based on throughfall data from (Zimmermann et al.
-2010) and precipitation data for BCI from STRI Physical Monitoring program, defined for precipitation events greater than 10 mm.
+| 10 | HKSAT_ADJ | Adjusting factor for soil hydraulic conductivity | 1 - 8 unitless | To account for high macroporosity and direct flow paths in tropical soils that is not accounted for by small soil core samples [@Broedel:2017; @Kinner:2004; @Tomasella:1998] |
+| 11 | fpi_max | Maximum interception fraction of precipitation | 0.05 - 0.44 unitless | Based on throughfall data from [@Zimmermann:2010kzl] and precipitation data for BCI from STRI Physical Monitoring program, defined for precipitation events greater than 10 mm.
 |

### 2020-07-13 code/load.R

@@ -95,12 +95,12 @@ load(file = file.path(results.folder, "coh.sp.summ.Rdata"))
 # #******************************************************
 # ## output
 # #******************************************************
-# load(file = file.path(results.folder, "ml.rsq.combine.best.Rdata"))
-# ###
-# ml.rsq.combine.best <- ml.rsq.combine.best %>% left_join(bci.traits %>% dplyr::select(sp, form1), by = "sp") %>%
-#   mutate(depth = as.numeric(depth))
-# erd.data <- ml.rsq.combine.best %>%
-#   subset(corr >= 0 & form1 == "T" &
-#            corr.func == "gr.Psi.VPD" &
-#            R2 >= 0.1)
+load(file = file.path(results.folder, "ml.rsq.combine.best.Rdata"))
+###
+ml.rsq.combine.best <- ml.rsq.combine.best %>% left_join(bci.traits %>% dplyr::select(sp, form1), by = "sp") %>%
+  mutate(depth = as.numeric(depth))
+erd.data <- ml.rsq.combine.best %>%
+  subset(corr >= 0 & form1 == "T" & sp != "guapst" &
+           corr.func == "gr.Psi.VPD" & source == "Meinzer et al.1999 Fig. 4" &
+           R2 >= 0.1)
 

### 2020-07-15 manuscript.rmd
1. mortality analyses 2. discussion points 3. other smaller edits

@@
 
+\
 Trait-based plant ecology aims to predict impacts of environmental change on community composition and function using relevant plant traits as currency.
@@
 
-We test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns them on a conservative-acquisitive trade-off axis.
-Specifically, in this forest (i) dynamics of soil water potential at depth does not cross catastrophic levels even during multi-annual droughts because of high total annual precipitation to evapotranspiration ratio; (ii) trees with deeper effective rooting depths (ERD) show acquisitive hydraulic architecture: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins; finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure.
+\
+We test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns species on a conservative-acquisitive trade-off axis.
+Specifically, in this forest (i) dynamics of soil water potential at depth does not cross critical plant thresholds even during multi-annual droughts because of high total annual precipitation to evapotranspiration ratio (not usre if this should be stated as a hypothesis or just an explanation of the results); (ii) trees with deeper effective rooting depths (ERD) show acquisitive hydraulic architecture: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins; finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure.
 To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and mortality outcomes for a tropical forest community.
@@
 
-```{r hydro-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap ="\\label{lab-hydro-calib}Model evaluation: Observations (red line) of Volumetric Water Content, Evapotranspiration and stream discharge aginst simulations (blue lines) from 100 best-fit parameter ensembles runs of ELM-FATES.
+```{r hydro-calib, dev='png', fig.show='hide', eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap ="\\label{lab-hydro-calib}Model evaluation: Observations (red line) of Volumetric Water Content, Evapotranspiration and stream discharge aginst simulations (blue lines) from 100 best-fit parameter ensembles runs of ELM-FATES.
 This figure will be reformatted to compose individual figures into one."}
@@
 
-## Model for Leaf Vulnerability Curves at community-scale
-
+## Model for Leaf Vulnerability Curves at Community-scale
+\
 We measured ($K_{x,~Leaf}$ vs.
@@
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB.jpeg"))
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_LMALAM_AVD_predicted_AB.jpeg"))
+# knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_LMALAM_AVD_predicted_AB.jpeg"))
 ```
@@
 
-With 
+-- Write about why models were designed this way.
+
 \
@@
 
-Growth Data: To minimise effect of light variation among trees and census intervals we selected only large trees (>= 30 cm DBH) and species whose maximum height was at least 30 m (hereafter, canopy species).
-We chose only those trees for which diameter (DBH) record was present across all size censuses (1990, 19905, 2000, 2005, 2010, 2015).
-We calculated individual tree ($i$) growth rates ($Gr_{i,t}$ cmyr^{-1}) for each of the five census intervals ($t$).
-As tree growth rates increase with size, to keep growth rates comparable across census intervals, we obtained residuals from a DBH model of tree growth based on a trees' DBH at the beginning of a census onterval ($GrResid_{i,t} = Gr_{i,t} - f(DBH_{i, t})$).
+Growth Data: To minimise effect of light variation among trees and census intervals we selected only large trees (>= 30 cm dbh, diamater at breast height) and species whose maximum height was at least 30 m (hereafter, canopy species).
+We chose only those trees for which diameter (dbh) record was present across all size censuses (1990, 1995, 2000, 2005, 2010, 2015).
+We calculated individual tree ($i$) growth rates ($Gr_{i,t}$ (cm.yr^{-1})) for each of the five census intervals ($t$) as the   difference between tree dbh measured in successive censuses, divided by the inter-census interval based on the exact date of measurements.
+As tree growth rates increase with size, to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census onterval ($Resid_{i,t} = Gr_{i,t} - f(DBH_{i, t})$).
 The diameter model of growth was fitted based on all trees' data pooled together as there was insufficient species-specific data for large trees (SI ).
-We then standardised (centered and scaled) an individual tree time series ($GrResid_{i,t}$) and obtained species-level (*n* = 32) times-series of medians ($Growth_{t}$) that had data on at least three trees (median = xx).
+We then standardised (centered and scaled) an individual tree time series ($Resid_{i,t}$) and obtained species-level (*n* = 32) times-series of medians ($Growth_{t}$) that had data on at least three trees (median = xx).
 
@@
 
-For each model (\@ref(eq:M2)- \@ref(eq:M4)) and species combination ($sp$), we obtained goodness-of-fit (*R*^2^) and Pearson's $\rho$ for all soil depths $z$, and for each sp, model we chose the $z$ with highest *R*^2^ as the Effective Rooting Depth (ERD); after retaining species with positive $\rho$ and *R*^2^ >= 0.1.
+For each model \@ref(eq:M1)- \@ref(eq:M4) and species combination ($sp$), we obtained goodness-of-fit (*R*^2^) and Pearson's $\rho$ for all soil depths $z$, and for each sp, model combination we chose the $z$ with highest *R*^2^ as the Effective Rooting Depth (ERD); after retaining species with positive $\rho$ and *R*^2^ >= 0.1.
 
@@
 
-Models with VPD and without Leaf Fraction (Eqs.
-\@ref(eq:M2)) had highest goodness-of fit (*R*^2^ = 0.51); followed by model with both VPD and Leaf Fraction (Eq.
-\@ref(eq:M4); Fig.
+\
+Model with VPD and without Leaf Fraction (Eq.
+\@ref(eq:M2)) had highest goodness-of fit (*R*^2^ = 0.51; *p* = 0.072), followed by followed by model without both VPD and and Leaf Fraction (\@ref(eq:M1); Fig.
 S\@ref(fig:other-erd-mods)).
-Models with leaf fraction had only three species with sufficient, indvidual species-level goodness-of-fit (Eqs.
-\@ref(eq:M3 & eq:M4) versus seven for models without (Eqs.
-\@ref(eq:M1) & \@ref(eq:M2)).
-For rest of the analyses we therefore used species ERD estimates from model (\@ref(eq:M2)) as it was also a parsimonious model compared to \@ref(eq:M4) (Fig.
+Models with leaf fraction had only three species with sufficient species-level goodness-of-fit (Eqs.
+\@ref(eq:M3)-\@ref(eq:M4)) versus seven for models without (Eqs.
+\@ref(eq:M1)-\@ref(eq:M2)).
+For rest of the analyses we therefore used species ERD estimates from model with VPD and without Leaf Fraction (Eq.
+\@ref(eq:M2); Fig.
 \@ref(fig:erd-iso)).
 
-```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-iso}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE; m) versus dry season stable isotopic concentration $\\delta$^2^H (mean $\\pm$ (add error); ‰) for ", length(unique(erd.data$sp))," canopy species from Barro Colorado Island (data from [@Meinzer:1999]")}
+```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-iso}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE; m) versus dry season stable isotopic concentration $\\delta$^2^H (mean $\\pm$ (add error); ‰) for ", length(unique(erd.iso$sp))," canopy species from Barro Colorado Island (data from [@Meinzer:1999]")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_phenology_Meinzer_gr.Psi.VPD.jpeg"))
@@
 
-We calculated mortality rates for large trees (>=10 cm DBH) of xx canopy species for each of the seven census interval ($t$) in the 33 year record (1985-2015; $Mort_{t}$) and a species mean mortality rate across those intervals ($Mort_{mean} = \frac{1}{t}\sum_{t=1}^{6} {Mort_{t}}$).
+We calculated mortality rates for large trees (>= 10 cm dbh) of species with ERD estimates (all canopy species with maximum height >= 30 m) and average adbunance of at least 20 trees (dropped one species with average abundance of 9 trees).
+For each of the seven census intervals ($t$) in the 33 year record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successives censuses (cen~{1}~ and cen~{2}~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{t}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50 ha plot of a species present in cen~{1}~ and dead in cen~{2}~, respectively, and t is the duration based on mean dates of cen~{1}~ and cen~{2}~.
+We calculated species mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{i}\sum_{i=1}^{6} {Mort_{t}}$ and, finally, species mortality anoamalies by intervals as $Mort_{anomaly}$ = $Mort_{t}$ - $Mort_{mean}$.
 To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$).
-To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{x, ERD}$) were below species $\Psi_{80,Leaf}$.
+To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{x, ERD}$) were below species $\Psi_{crit}$, defined as $\Psi_{80,~Leaf}$.
 
@@
 Species with deeper $ERD$ showed lower mean mortality rates across the 33 year record (1982-2015; *R*^2^ = 0.13, *P* = 0.029, *n* = xx; Figure \@ref(fig:erd-mort)).
-For five out of seven census intervals, species with shallower $ERD$ had significantly higher mortality anomalies ($Mort_{int}$ - $Mort{mean}$ than deeper ERD species; Figure (\@ref(fig:erd-mort-anom)).
-Analysis of time spent under catastrophic soil drought showed that the proportion of days species $\Psi_{x,~ERD}$ was more negative than species $\Psi_{80, Leaf}$ decreased with species ERD (Figure \@ref(fig:erd-time)).
-Species with ERD deeper than 1 m did not cross species $\Psi_{80, Leaf}$ at all for the entire 25 year period for which soil water potential $\Psi_{x,~z}$ was modelled (1990-2015).
+For five out of seven census intervals, species with shallower $ERD$ had significantly higher mortality anomalies than deeper ERD species; Figure (\@ref(fig:erd-mort-anom)).
+Analysis of time spent under catastrophic soil drought showed that the proportion of days species $\Psi_{x,~ERD}$ was more negative than species $\Psi_{crit}$ decreased with species ERD (Figure \@ref(fig:erd-time)).
+Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25 year period for which soil water potential $\Psi_{x,~z}$ was modelled (1990-2015), as soil depths at and below 1.7 m remained above `r round(psi.1.7.min$median, 2)` [`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`] (median [95% CI]) through all the dryseason and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).
 
-```{r erd-mort, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort}Mortality rate (mean $\\pm$ (add error)) versus Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for xx canopy species found on Barro Colorado Island.
+```{r erd-mort, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort}Mortality rate (mean $\\pm$ 1SE) versus Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for xx canopy species found on Barro Colorado Island.
 *R*^2^ and *P*-values are given in panel insets."}
@@
 
-```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{x, z}$ was below $\\Psi_{80,Leaf}$ over 1990-2015, where z is the soil depth matching species ERD--for xx canopy species found on Barro Colorado Island.
-EAch circle resepresents a species.
+```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{x, z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~Leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for xx canopy species found on Barro Colorado Island.
+Each circle resepresents a species.
 Circles are jittered vertically around the named depths (ticks) to be able to distinguish visually.
@@
 
-### ERD not correlated with growth
-# Limitations & Future directions:
+Our analyses revealed that trees with deeper effective rooting depths (ERD) showed acquisitive hydraulic architecture, with higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species.
+Yet, deeper ERD species showed lower mortality rates over the 33 year record (1982-2015).
+Analysis of time spent below ($\Psi_{crit}$) in the soil zone of ERD showed that this was likely due to low realised risk of hydraulic failure for deeper ERD species.
+In this moist forest ($\Psi_{z}$) for depths at and below 1.7 m remained above ($\Psi_{crit}$) of the most sensitive tree species through all the dryseason and drought periods over 1990-2015, period over which ($\Psi_{z}$) was modelled.
+We thus show, for the first time, that interspecific variation in effective rooting depths of trees aligns species on a conservative-acquisitive trade-off axism, by demonstrating a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a tropical forest community.
 
-- not hydrodynamics, 1D to 3D
+## Limitations & Future directions
 
-- Temperate rooting depths shallow?  Shenk and Jackson
+Our modeling framework had a few limitations. 
+- The inverse model for ERD was parametarised on 5-yr growth rates, which vastly increased the degree of freedom to fit the model.
+Future studies should use high-frequency growth time series such as those obtained from dendrometer bands.
+To infer growth from such diameter time series, however, would especially require accounting for diameter changes due to hydrostatic flexing [@Chitra-Tarak:2015].
 
-- lightening explains 1/3rd of mortality
-- forest is realtively stable over 40 years.
+- Future studies should investigate seasonality in water-sourcing depths investigated.
 
--transient depth of water uptake, which is known to vary seasonally (Ellsworth & Sternberg, 2015), 
-In fact, associations between soil water access by roots and aboveground drought survival traits such as resistance of xylem to drought‐induced cavitation (Pockman & Sperry, 2000; Pratt et al ., 2007) or phenology (Jackson et al ., 1995; Hasselquist et al ., 2010), are fairly reliable trade‐off functions in plant eco‐physiology.
+- Although we attempted a coarse estimate of species leaf cover seasonality by combining  leaf-fall data from litter traps and forest-level LAI data for leaf-on periods, more accuarate estimations of leaf-cover seasonality are required, such as those obtained from drone images [@Park: 2019]
+
+-To enable estimation of inverse model for the larger community across size classes would need accounting for changing light environments on growth; we minimised this issue by limiting our ERD inversion to large trees of species with max height >= 30m and thus whose canopy's are most likely to be fully exposed to light.
+
+- Future studies would benefit from modeling the complete hydrodynamics of water potential gradients from soil through plant tissues (roots-stems-leaves) to atmosphere continuum (SPAC) [@Johnson:2017].
+However, to parameterise such a model at community-scale would need community-scale data on hard traits such as stem and root vulnerability curves; if not, model development to up-scale data from a few species, for example, via softer traits as we did for leaf vulnerability curves.
+(Note that In this forest LMA and WSG did not scale with stem vulnerability curves).
+Water storage (capacitance) should also be considered.
 
 - Deep ERD species may possibly be associated with locations of greater water availability rather than having deep rootedness.
-Future studies should employ a distributed hydrological model to separate these factors.
+A 3-D hydrology model would provide spatially distributed estimates of water dynamics in the whole-soil/rock column and depth to water table to link with individual trees water experience [@Schwantes:2018].
+Forest inventory sites would benefit from investing data collection efforts on water availability dynamics by depth and water tables covering spatial heterogeneity so as to better constrain such models.
 
-Future studies: high resolution growth data to better constrain the model. 
-  -water storage (capacitance) should be considered
-  -shifts in root water-uptake by depth
-  -deci
+- Variation in mortality explained by ERD is low. Other causes.... Based on a year-long study, lightening explains ~4.5 % of mortality for trees >=10 cm dbh and ~ 40% for trees >=60 cm dbh.
+
+-Forest is realtively stable over 35 years.
+Future studies could investigate if ongoing change in the duration, frequency and intensity of droughts entails sufficient exhaustion of deep water availablity so as to cause mortality of deep ERD species.
+
+## Conclusions:
 
@@
 ```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts}Extreme droughts by census interval and some representative depths"}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q2.5.jpeg"))
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q2.5_depths_in_inverse_model.jpeg"))
 ```
@@
 ```{r leaf-cover, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-cover} Seasonal variation in Leaf Cover Fraction for ", length(unique(sp.leaf_cover.for.model$sp[sp.leaf_cover.for.model$deciduousness != "Evergreen"]))," non-evergreen species at Barro Colorado Island.
-Leaf Cover Fraction for evergreen species (n = ", length(unique(sp.leaf_cover.for.model$sp[sp.leaf_cover.for.model$deciduousness == "Evergreen"])),") is assumed at 1 through the year.")}
+Leaf Cover Fraction for evergreen species (*n* = ", length(unique(sp.leaf_cover.for.model$sp[sp.leaf_cover.for.model$deciduousness == "Evergreen"])),") is assumed at 1 through the year.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/leaf.cover_BCI_single_panel_evergreensAt100.jpeg"))

### 2020-07-22 manuscript.rmd
added Brent's edits. other edits

@@
 We evaluated ERD by leveraging existing isotopic tracer data ($\delta^2H$).
-- We find that deeper ERD is associated with higher hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margin, but lower mean mortality rate.
-Deeper ERD species spent less time below critical hydraulic thredholds indicating that ERD buffers drought-stress.
+- We find that deeper ERD is associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margin, but lower mean mortality rate.
+Deeper ERD species spent less time below critical hydraulic thresholds indicating that ERD buffers drought-stress.
 - The role of deep water access in alleviating mortality of drought-sensitive trees have important implications for modeling the future of tropical species and forests under climate change.
@@
 Intensifying droughts threaten tropical forests, a biome that plays a disproportionately large role in the global carbon and energy cycle, and supports half of the global biodiversity [@Bonan:2008; @Wright:20050th].
-Tropical forests are considered an especially drought-vulnerable biome based on observed ranges of trees' hydraulic safety margins [@Choat:2012; but see @Oliveira:2019].
+Tropical forests are considered an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw]: here, historic climatic variation is relatively small[@Mora:2013], and the observed ranges of trees' hydraulic safety margins are relatively narrow [@Choat:2012; @Oliveira:2019].
 Widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak:2018; @Hilker:2014; @Phillips:2010].
@@
 Consequently trees' in the same forest differ in their drought experience, and thus in their responses in term of growth and mortality.
-A key bottlneck in testing this mechanism community-wide has been lack of data on trees' rooting or water-sourcing depths, followed by data on water-availability at those depths.
+A key bottlneck in community-wide testing of this mechanism has been lack of data on trees' rooting or water-sourcing depths, followed by data on water-availability at those depths.
 A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the World’s 65 thousand tree species [@Fan:2017; @Beech:2017].
 Further, only a small fraction of those are tropical, even though ~60-80 % (40 - 53 throusand species) of the World’s tree diversity resides in the tropics [@Slik:2015].
-Use of stable isotopes of water as a tracer provides an indirect way of trees' water sourcing depths by matching isotopic concentration in plant tissue to it's gradient vertically belowground.
-Such data is available only for ~0.05% of World's tree species [@Evaristo:2016].
+Use of stable isotopes of water as a tracer provides an indirect way of trees' water sourcing depths by matching isotopic value in plant tissue to variations in soil pore water vertically belowground.
+Such data are available only for ~0.05% of World's tree species [@Evaristo:2016].
 In general, community-scale data collection for rooting or water-sourcing depths is especially challenging for species-rich tropical forests.
-Further, to estimate plant available water in the vertical soil column requires hydrological data and modeling, but these are rarely co-located with forest inventory sites (but see [@Johnson:2018fad;@Kumagai:2011;@Ivanov:2012].
+Further, to estimate plant available water with soil depth requires hydrological data and modeling, but these are rarely co-located with forest inventory sites (but see [@Johnson:2018fad;@Kumagai:2011;@Ivanov:2012].
 
@@
 Trait-based plant ecology aims to predict impacts of environmental change on community composition and function using relevant plant traits as currency.
-To reach this goal it is imperative to establish relationships between plant's environment, traits and demographic outcomes, but this remains to be established for trees' realised water environments and water-access through droughts.
+To reach this goal it is imperative to establish relationships between plant's environment, traits and demographic outcomes, but this remains to be a challenge.
 Huge efforts have been devoted to linking same-for-all-trees drought indices, above-ground traits and demographic rates.
-These have resulted in above-ground hydraulic traits such as proxies of hydraulic failure as suggested indicators of mortality risk.
+These efforts have resulted in above-ground hydraulic traits such as proxies of hydraulic failure as suggested indicators of mortality risk.
 A recent meta-analysis found greater drought-induced mortality rates in species with high vulnerability to drought-induced embolism ($\Psi_{88,~Stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~Stem}$), but counter examples also exist (Hoffmann 2011; Paddock 2013), indicating whether plants actually reach critical thresholds of hydraulic failure depends on their realised drought intensity.
-A rare study that accounted for realised water availability for four tree species in a rare, centennial drought (Texas, USA) found that deep rooted species died less despite being more vulnerable to xylem embolism (high $\Psi_{88,~Stem}$) and having narrower safety-margins ($\Psi_{min} - \Psi_{88,~Stem}$).
-A species sourcing deep water showed lower hydraulic failure and lower mortality in an extreme drought in an Italian Karst system even though it had narrower hydrulic safety margin and high $\Psi_{88,~Stem}$ than other two species limited to shallow-water.
+A rare study that accounted for realised water availability for four tree species in a rare, centennial drought (Texas, USA) found that deep rooted species died less despite being more vulnerable to xylem embolism (high $\Psi_{88,~Stem}$) and having narrower safety-margins ($\Psi_{min} - \Psi_{88,~Stem}$) [@Johnson:2017].
+A species sourcing deep water showed lower hydraulic failure and lower mortality in an extreme drought in an Italian Karst system even though it had narrower hydrulic safety margin and high $\Psi_{88,~Stem}$ than other two species limited to shallow-water[@Nardini:2015].
 Chitra-Tarak et al.
@@
 This study found that species accessing deep-water in a tropical dry forest study system died more in a rare, centennial drought compared to shallow-water species.
-Authors concluded that this was consistent with expectation of deep water usage as an aquisitive strategy relying on stable and reliable water resource at depth, which exhausted in the multi-annual drought resulting in reduced survival, compared to persistent species associated with shallow-water that was relatively more available via infiltration during this drought.
-However, species estimates of water-access modelled by this study were not corroborated with rooting/water-sourcing depths, or related to hydraulic traits.
+They concluded that this was consistent with expectation of deep water usage as an aquisitive strategy relying on stable and reliable water resource at depth, which exhausted in the multi-annual drought resulting in reduced survival, compared to persistent species associated with shallow-water that was relatively more available via infiltration during this drought.
+However, species estimates of water-access modelled by this study were not corroborated becuase of lack of data on rooting/water-sourcing depths, or related hydraulic traits.
 
@@
 We test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns species on a conservative-acquisitive trade-off axis.
-Specifically, in this forest (i) dynamics of soil water potential at depth does not cross critical plant thresholds even during multi-annual droughts because of high total annual precipitation to evapotranspiration ratio (not usre if this should be stated as a hypothesis or just an explanation of the results); (ii) trees with deeper effective rooting depths (ERD) show acquisitive hydraulic architecture: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins; finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure.
+Specific hypotehses applicable to this forest included: (i) dynamics of soil water potential at depth does not cross critical plant thresholds even during multi-annual droughts because of high total annual precipitation to evapotranspiration ratio (not sure if this should be stated as a hypothesis or just an explanation of the results); (ii) trees with deeper effective rooting depths (ERD) show acquisitive hydraulic architecture: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins; and finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure.
 To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and mortality outcomes for a tropical forest community.
@@
 We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterisation in terms of long-term forest demography, plant functional ecology, hydrology, including availability of an independent isotopic tracer study on trees’ water-sourcing depths [@Meinzer:1999].
-The 1,560 hectare Barro Colorado Island was formed after the river Chagres was dammed in 1914 that created Gatun Lake, the main channel of the Panama Canal.
+The 1,560 hectare island was formed after the river Chagres was dammed in 1914 that created Gatun Lake, the main channel of the Panama Canal.
 Rainfall at BCI is seasonal with a mean annual rainfall of 2687 mm (± 515 SD over 1983-2017) and a four-month long dry season from mid-December through mid-April (≤ 100 mm rain/month).
-The entire Island is forested and classified as a tropical moist forest.
+The entire island is forested and classified as a tropical moist forest.
 Although percentage of canopy species that are at least partly deciduous during the dry season is 28%, the maximum percentage of deciduous individuals in the canopy is only 6.3%, and the percentage of crown area that is deciduous is estimated at 9.7% [@Condit:2000].
@@
 Below we describe the equation of water balance in ELM-FATES that is most relevant for our study site, even though ELM-FATES is capable of incorporating numerous other hydrological processes relevant for wetlands and lakes and snow (CLM4.5 Tech Note).
-ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\Delta W_{can}$, surface water $\Delta W_{sfc}$, soil water $\Delta W_{liq, z}$, and water in the unconfined aquifer $\Delta W_{a}$ (omitting processes relevant to snow or wetlands and lakes), and can be described as
+ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\Delta W_{can}$, surface water $\Delta W_{sfc}$, soil water $\Delta W_{liq, z}$, and water in an unconfined aquifer $\Delta W_{a}$ (omitting processes relevant to snow or wetlands and lakes), and can be described as
 
@@
 As our key interest here was in soil water availability rather than forest dynamics, we ran ELM-FATES in simplified modes: static stand structure and without plant hydrodynamics, and with a single Plant Functional Type (PFT) of evergreen trees.
-The latter waws chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of dry-deciduous PFT did not significantly alter results (not shown).
+The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of dry-deciduous PFT did not significantly alter results (not shown).
 
@@
 
-```{r hydro-calib, dev='png', fig.show='hide', eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap ="\\label{lab-hydro-calib}Model evaluation: Observations (red line) of Volumetric Water Content, Evapotranspiration and stream discharge aginst simulations (blue lines) from 100 best-fit parameter ensembles runs of ELM-FATES.
+```{r hydro-calib, dev='png', fig.show='hide', eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap ="\\label{lab-hydro-calib}Model evaluation: Observations (red line) of Volumetric Water Content, Evapotranspiration and stream discharge against simulations (blue lines) from 100 best-fit parameter ensembles runs of ELM-FATES.
 This figure will be reformatted to compose individual figures into one."}
@@
 For each model, species ERD was regressed against species Xylem sap $\delta^2H$ (%~$\circ$~; *n* = 7).
-We used Xylem sap $\delta^2H$ values given in [@Meinzer:2001] for dry-season (March 1997) as these showed largest seasonal divergence among species, with a corresponding gradient of $\delta^2H$ in the soil.
+We used Xylem sap $\delta^2H$ values given in [@Meinzer:2001] for the dry-season (March 1997) as these showed largest seasonal divergence among species, with a corresponding gradient of $\delta^2H$ in the soil.
 We removed species that are deciduous in March as xylem sap $\delta^2H$ of these species may not be linked with water sourced in March and thus may not comparable to other species with leaves present.
@@
 S\@ref(fig:iso-data).
-We also removed *Guapira* *spp*, whose xylem sap value fell in the uncertain ranges of soil depth.
+We also removed *Guapira* *spp*, whose xylem sap isotope value was not constrained by the soil water isotope depth profile.
 
 \
-Model with VPD and without Leaf Fraction (Eq.
+The model with VPD and without Leaf Fraction (Eq.
 \@ref(eq:M2)) had highest goodness-of fit (*R*^2^ = 0.51; *p* = 0.072), followed by followed by model without both VPD and and Leaf Fraction (\@ref(eq:M1); Fig.
@@
 \@ref(eq:M1)-\@ref(eq:M2)).
-For rest of the analyses we therefore used species ERD estimates from model with VPD and without Leaf Fraction (Eq.
+For rest of the analyses we therefore used species ERD estimates from the model with VPD and without Leaf Fraction (Eq.
 \@ref(eq:M2); Fig.
@@
 
-```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-iso}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE; m) versus dry season stable isotopic concentration $\\delta$^2^H (mean $\\pm$ (add error); ‰) for ", length(unique(erd.iso$sp))," canopy species from Barro Colorado Island (data from [@Meinzer:1999]")}
+```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-iso}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE; m) versus dry season stable isotopic concentration $\\delta$^2^H (mean $\\pm$ (add error); ‰) for ", length(unique(erd.iso$sp))," canopy species from Barro Colorado Island (data from [@Meinzer:1999]").
+See Table S2 for scientific names corresponding to species codes.}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_phenology_Meinzer_gr.Psi.VPD.jpeg"))
@@
 
-We calculated mortality rates for large trees (>= 10 cm dbh) of species with ERD estimates (all canopy species with maximum height >= 30 m) and average adbunance of at least 20 trees (dropped one species with average abundance of 9 trees).
-For each of the seven census intervals ($t$) in the 33 year record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successives censuses (cen~{1}~ and cen~{2}~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{t}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50 ha plot of a species present in cen~{1}~ and dead in cen~{2}~, respectively, and t is the duration based on mean dates of cen~{1}~ and cen~{2}~.
-We calculated species mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{i}\sum_{i=1}^{6} {Mort_{t}}$ and, finally, species mortality anoamalies by intervals as $Mort_{anomaly}$ = $Mort_{t}$ - $Mort_{mean}$.
+We calculated mortality rates for large trees (>= 10 cm dbh) of species with ERD estimates (all canopy species with maximum height >= 30 m) and average adbunance of at least 20 trees (one species was eleminated with an average abundance of 9 trees).
+For each of the seven census intervals ($t$) in the 33 year record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successives censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{t}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50 ha plot of a species present in cen~1~ and dead in cen~2~, respectively, and t is the duration based on mean dates of cen~1~ and cen~2~.
+We calculated species mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{i=1}^{6} {Mort_{t}}$ and, finally, species mortality anoamalies by intervals as $Mort_{anomaly}$ = $Mort_{t}$ - $Mort_{mean}$.
 To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$).
@@
 ```{r sp-erd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap= paste0("\\label{lab-sp-erd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for 32 large, canopy species of Barro Colorado Island.
-See Table S2 for scientific names of species codes")}
+See Table S2 for scientific names corresponding to species codes.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/ERD_by_sp_large_canopy.jpeg"))
@@
 
-```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus hydraulic propertiese for seven large, canopy species found on Barro Colorado Island (Panama).
+```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus hydraulic propertiese for eight large, canopy species found on Barro Colorado Island (Panama).
 *R*^2^ and *P*-values are given in panel insets.")}
@@
 ```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{x, z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~Leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for xx canopy species found on Barro Colorado Island.
-Each circle resepresents a species.
+Each circle represents a species.
 Circles are jittered vertically around the named depths (ticks) to be able to distinguish visually.
@@
 Analysis of time spent below ($\Psi_{crit}$) in the soil zone of ERD showed that this was likely due to low realised risk of hydraulic failure for deeper ERD species.
-In this moist forest ($\Psi_{z}$) for depths at and below 1.7 m remained above ($\Psi_{crit}$) of the most sensitive tree species through all the dryseason and drought periods over 1990-2015, period over which ($\Psi_{z}$) was modelled.
-We thus show, for the first time, that interspecific variation in effective rooting depths of trees aligns species on a conservative-acquisitive trade-off axism, by demonstrating a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a tropical forest community.
+We thus show, for the first time, that interspecific variation in trees' effective rooting depths aligns species on an acquisitive-conservative strategy axis by demonstrating a link between trees' realised drought experience, above-ground hydraulic architecture and long-term mortality outcomes.
+On the aquisitive end were species with hydraulically efficient, but risky suit of traits optimised for an environment of reliable, abundant source of water at depth, which actualised to a lower associated mortality-risk, whilst at the conservative end were species with inefficient, but safer xylem, associated with a water environment of high variability.
+
+Acquisitive strategy is thought to be optimised for high resource environments with high performance outcome and a large decline in performance under stresses or low resource environment; whilst conservative strategy have a moderate performance in high resource environment that declines with decreasing resources, while retaining a competitive advantage over acquisitives under lower resources.
+
+In other contexts aquisitive-conservative trait trade-off is associated with a growth-survival trade-off, with aquisitive species achieving higher potential growth at the cost of lower survival, but in case of ERD axis, aquisitiveness did not relate Normally Our results highlight the need..
+to higher growth, but
+
+## Trait-Trait Coordination
+
+-- trade-offs among differential investment in hydraulic efficiency and safety
+here links with trade-off in water-stress vertically belowground
+-- unexplored (hydraulic capacitance, stomatal control, photosynthetic capacity, high temperature tolerance)
+
+## Trait & Demography
+
+## Environment, Trait & Demography
+
+Mortality could be immediate or may result after decades of drought occurrence.
+Mortality could occur over
 
@@
 
-- Future studies would benefit from modeling the complete hydrodynamics of water potential gradients from soil through plant tissues (roots-stems-leaves) to atmosphere continuum (SPAC) [@Johnson:2017].
-However, to parameterise such a model at community-scale would need community-scale data on hard traits such as stem and root vulnerability curves; if not, model development to up-scale data from a few species, for example, via softer traits as we did for leaf vulnerability curves.
+- Future studies would benefit from modeling the complete hydrodynamics of water potential gradients over the soil - plant (roots, stems, leaves) - atmosphere continuum (SPAC) [@Johnson:2017; @Christoffersen 2017].
+However, to parameterise such a model at community-scale would need community-scale data on hard traits such as stem and root vulnerability curves; if not, model development would be needed to up-scale data from a few species, for example, via softer traits as we did for leaf vulnerability curves.
 (Note that In this forest LMA and WSG did not scale with stem vulnerability curves).
@@
 
-```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts}Extreme droughts by census interval and some representative depths"}
+```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts} Extreme soil droughts by census interval (panels) at three representative depths (0.5, 1 and 1.7 m depicted in different colors).
+Mean soil water potential ($Psi_z$}) for a given DOY and depth z are shown as lines.
+Filled areas show extreme drought periods, when $Psi_z$ declined below 2.5 percentile of $Psi_z$ for the given DOY and depth z in an individual year within the cenus intervals.
+Eextreme drought periods from multiple years may be overlaid within a panel."}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q2.5_depths_in_inverse_model.jpeg"))
@@
 | 8 | fmax | The maximum fractional saturated area |  unitless | Empirical | 
-| 9 | HKSAT | Soil hydraulic conductivity profile | 0.007 - 0.014 mms{-1} for $\le$ 12.5 cm; 5.56e-05, 0.0003 for $\ge$ 60 cm |  For Conrad catchment observed median $K_{sat}$ (95% CI) for 12.5 cm depth is 38.3 mmhr^{-1}^ (25.4-51.2, *n* = 75), while for 60 cm depth 0.7 (0.2, 1.2, *n* = 40) mmhr^{-1}^.
-For reference, a storm with 12.5 mmhr^{-1}^ rainfall intensity has a 0.2 probability of occuring in any given rainfall event [@Kinner:2004].
-|
 | 10 | HKSAT_ADJ | Adjusting factor for soil hydraulic conductivity | 1 - 8 unitless | To account for high macroporosity and direct flow paths in tropical soils that is not accounted for by small soil core samples [@Broedel:2017; @Kinner:2004; @Tomasella:1998] |

### 2020-07-26 manuscript.rmd
1. Revised text 2. Added table for sp code names 3. Filled up info on ELM-FATEs parameterisation table: soil vulnerability curves, soil hydro conductivity calculation and corresponding figures 4. added ksat values in parameter table

@@
     fig_caption: yes
-  bookdown::pdf_document2: default
   bookdown::word_document2:
@@
         fig_caption: yes
+  bookdown::pdf_document2: default
 bibliography: references.bib
@@
 suppressPackageStartupMessages(
-  pacman::p_load(bookdown, knitr, pander, citr, broom))
+  pacman::p_load(bookdown, knitr, pander, citr, broom, kableExtra))
 options(tinytex.verbose = TRUE)
 source("code/load.R")
+source("code/manuscript_data_figs.R")
+load("results/manuWorkSpace.RData")
+# install.packages(c("vctrs", "rlang"))
 ```
@@
 
-- Deep water access (WA~Deep~) is believed to be an important, drought-tolerance mechanism, but under-studied due to lack of data.
+- Deep water access is believed to be an important, drought-tolerance mechanism, but under-studied given the lack of data.
 Vulnerability to embolism ($\Psi_{88,~Stem}$) and safety margins ($\Psi_{min} - \Psi_{88,~Stem}$) are shown to be predictors of drought-induced mortality risk, but deep water access may delay plants from reaching these critical thresholds.
-Here we test its role in drought-tolerance strategies of a tropical forest community (Panama) with a novel data-model approach.
-- We inversely modelled Effective Rooting Depth (ERD) for 32 large, canopy tree species by linking species growth dynamics (xx trees, 6 censuses, 1990-2015) to VPD, $\Psi_{x,~z}$ by depth $z$in the whole-soil column, and a novel community-scale model of species vulnerability curves ($K_{Leaf}$ vs.
-$\Psi_{x,~Leaf}$) linking data on the same for 21 species with community-scale softer traits.
-We evaluated ERD by leveraging existing isotopic tracer data ($\delta^2H$).
-- We find that deeper ERD is associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margin, but lower mean mortality rate.
+Here we test the role of deep-water access in drought-tolerance strategies of a hyper-diverse tropical forest community (Barro Colorado Island, Panama) using a novel data-model approach.
+- We inversely modelled Effective Rooting Depth (ERD) for large trees of `r length(erd.sp)` canopy tree species by linking species growth dynamics (6 censuses, 1990-2015) to VPD, Soil Water Potential ($\Psi_{soil}$) by depth in the whole-soil column, and a new community-scale model of species vulnerability curves ($K_{Leaf}$ vs.
+$\Psi_{Leaf}$) that scaled new data on 21 species with Leaf Mass Area and Wood Specific Gravity.
+We evaluated ERD by leveraging existing isotopic tracer data ($\delta^2H$) on tree xylem sap and soil column.
+- We find that deeper ERD is associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, but lower mortality rates over 35 years.
 Deeper ERD species spent less time below critical hydraulic thresholds indicating that ERD buffers drought-stress.
-- The role of deep water access in alleviating mortality of drought-sensitive trees have important implications for modeling the future of tropical species and forests under climate change.
+- The role of deep water access in alleviating mortality of drought-sensitive trees have important implications for predicting the future of tropical forests under climate change.
 
@@
 Intensifying droughts threaten tropical forests, a biome that plays a disproportionately large role in the global carbon and energy cycle, and supports half of the global biodiversity [@Bonan:2008; @Wright:20050th].
-Tropical forests are considered an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw]: here, historic climatic variation is relatively small[@Mora:2013], and the observed ranges of trees' hydraulic safety margins are relatively narrow [@Choat:2012; @Oliveira:2019].
+Tropical forests are considered an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw]: here, historic climatic variation is relatively small[@Mora.2013], and the observed ranges of trees' hydraulic safety margins are relatively narrow [@Choat:2012; @Oliveira:2019].
 Widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak:2018; @Hilker:2014; @Phillips:2010].
-However, state-of-the-art dynamic global vegetation models (DGVMs) fail to capture drought-induced vegetation dynamics in the tropics [@Galbraith:2010; @Powell:2013], as underlying mechanisms of drought-tolerance are not fully understood.
+However, state-of-the-art dynamic vegetation models (DVMs) fail to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013], as underlying mechanisms of drought-tolerance are not fully understood.
 
@@
 In general, community-scale data collection for rooting or water-sourcing depths is especially challenging for species-rich tropical forests.
-Further, to estimate plant available water with soil depth requires hydrological data and modeling, but these are rarely co-located with forest inventory sites (but see [@Johnson:2018fad;@Kumagai:2011;@Ivanov:2012].
+Further, to estimate plant available water with soil depth requires hydrological data and modeling, but these are rarely co-located with forest inventory sites (but see [@Johnson:2018fad;@Kumagai:2011;@Ivanov:2012]
 
 \
-Trait-based plant ecology aims to predict impacts of environmental change on community composition and function using relevant plant traits as currency.
-To reach this goal it is imperative to establish relationships between plant's environment, traits and demographic outcomes, but this remains to be a challenge.
-Huge efforts have been devoted to linking same-for-all-trees drought indices, above-ground traits and demographic rates.
-These efforts have resulted in above-ground hydraulic traits such as proxies of hydraulic failure as suggested indicators of mortality risk.
-A recent meta-analysis found greater drought-induced mortality rates in species with high vulnerability to drought-induced embolism ($\Psi_{88,~Stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~Stem}$), but counter examples also exist (Hoffmann 2011; Paddock 2013), indicating whether plants actually reach critical thresholds of hydraulic failure depends on their realised drought intensity.
-A rare study that accounted for realised water availability for four tree species in a rare, centennial drought (Texas, USA) found that deep rooted species died less despite being more vulnerable to xylem embolism (high $\Psi_{88,~Stem}$) and having narrower safety-margins ($\Psi_{min} - \Psi_{88,~Stem}$) [@Johnson:2017].
-A species sourcing deep water showed lower hydraulic failure and lower mortality in an extreme drought in an Italian Karst system even though it had narrower hydrulic safety margin and high $\Psi_{88,~Stem}$ than other two species limited to shallow-water[@Nardini:2015].
-Chitra-Tarak et al.
-(2018) proposed a general model to link realised water availability to demographic outcomes through droughts at tree community-scale.
-This study found that species accessing deep-water in a tropical dry forest study system died more in a rare, centennial drought compared to shallow-water species.
-They concluded that this was consistent with expectation of deep water usage as an aquisitive strategy relying on stable and reliable water resource at depth, which exhausted in the multi-annual drought resulting in reduced survival, compared to persistent species associated with shallow-water that was relatively more available via infiltration during this drought.
-However, species estimates of water-access modelled by this study were not corroborated becuase of lack of data on rooting/water-sourcing depths, or related hydraulic traits.
+Establishing relationships between plant's environment, traits and demographic outcomes is imperative for a predictive plant ecology [@Mcgill.2006], but plant's realised water environments are rarely studied.
+Huge efforts are devoted to linking the same-for-all-trees drought indices, above-ground traits and demographic rates.
+These efforts have resulted in identification of above-ground traits such as proxies of hydraulic failure as indicators of mortality risk.
+A recent meta-analysis found greater drought-induced mortality rates in species with cavitation-vulnerable xylem ($\Psi_{88,~Stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~Stem}$) [@Anderegg:2016], but counter examples also exist [@Hoffmann:2011; @Paddock:2013]; indicating whether plants reach critical thresholds of hydraulic failure depends on their realised drought intensity.
+A rare study that accounted for realised water availability for four tree species in a centennial drought (Texas, USA) found that mortality decreased with deep rootedness even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi_{88,~Stem}$) and had narrower safety-margins ($\Psi_{min} - \Psi_{88,~Stem}$) [@Johnson:2018fad].
+A species sourcing deep water in a Karst system (Italy) showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi_{88,~Stem}$ and narrower safety margin compared to two species sourcing shallow-water [@Nardini:2015].
+@Chitra-Tarak:2018 proposed the first community-scale model that linked trees' realised water availability to demographic outcomes through droughts.
+They found that species accessing deep-water in a tropical dry forest (India) died more in a rare, centennial drought compared to shallow-water species.
+They concluded that this was consistent with an 'aquisitive-conservative' species strategy spectrum.
+Aquisitives relied on stable water resource at depth, which exhausted in the multi-annual drought, resulting in reduced survival, whereas conservative species used fluctuating shallow-water and survived the drought as some shallow-water was available via infiltration.
+However, lack of data on rooting/water-sourcing depths and hydraulic traits precluded this study from corroborating modelled estimates of water-uptake depths, or investigating trait coordination.
 
 \
-Here, we obtained daily dynamics of plant water potential ($\Psi_{x,~z}$) by depth $z$ in the whole-soil column of a neo-tropical forest (BCI, Panama) with a locally parametarised and calibrated 1-D hydrological water balance model.
-We then estimated the effective rooting depths for xx large, canopy tree species as the depth ($z$) whose soil water dynamics ($\Psi_{x,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015) with a novel inverse model.
-This model incorporated impact of atmospheric drought (VPD) on growth and new data on species leaf vulnerability curves ($K_{Leaf}$ vs.
-$\Psi_{x,~Leaf}$), which we scaled-up to community level via scaling relationships with commonly available traits.
-We evaluated the inversely estimated ERD against exisiting data on plant and soil water isotopes ($\delta^2H$).
+Trade-offs that shape how trees' make a living in their environment (ecological strategies) provide Dynamic Vegetation Models (DVMs) the essential constraints to represent species-rich communities under changing environment [@Koven:2020].
+This entails indentifying trade-offs among environment, traits and demography.
+A universal 'fast-slow' (aka productive-persistent, or acquisitive-conservative) plant economics spectrum conceptualises plants coordination for all resources and organs, both above- and below-ground [@Reich:2014].
+For water resource observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified a fast-slow spectrum, although examples of all trade-offs observed simultaneously are rare, and belowground trait or environment linkages are weak.
+'Fast' species at the productive end of the spectrum have traits with high capacity to acquire, move, store and use water (high Kmax, low wood density) and achieve high growth potential, but build flimsy, disposable tissues (high SLA) and are less tolerant of low resources, whereas species with slow traits are better protected from C losses and are more drought tolerant (low $\Psi_{50}$, low $\Psi_{TLP}$, larger safety margin).
+Fast strategy is associated with spatially wet sites such as valleys or high rainfall [@Kursar:2009; @Oliveira:2019], but whether and how the fast-slow spectrum holds true for the moisture gradient in the soil column is not clear.
+While deep rootedness may be expected to support a fast strategy with a steady water supply unlike shallow-rooted trees.
+However, this may depend on the exact dynamics of water-availability by depth at a site, depending on precipitation regime and hydrology.
+In pulse-driven arid systems *shallow*-rooted plants are associated with fast traits (high Kmax) that maximise the use of the water-pulse, while *deep*-rooted plants have slow traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013].
 
 \
-We test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns species on a conservative-acquisitive trade-off axis.
-Specific hypotehses applicable to this forest included: (i) dynamics of soil water potential at depth does not cross critical plant thresholds even during multi-annual droughts because of high total annual precipitation to evapotranspiration ratio (not sure if this should be stated as a hypothesis or just an explanation of the results); (ii) trees with deeper effective rooting depths (ERD) show acquisitive hydraulic architecture: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins; and finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure.
-To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and mortality outcomes for a tropical forest community.
+Here, for large trees of xx canopy species of a hyperdiverse neo-tropical moist forest (BCI, Panama) we make a thorough assessment of trade-offs between belowground water environment, above-ground hydraulic traits, estimates of effective rooting depths and demographic outcomes over 35 years that withnessed droughts of various intensity and duration.
+We obtained daily dynamics of plant water potential ($\Psi_{x,~z}$) by depth $z$ in the whole-soil column with a locally parametarised and calibrated 1-D hydrological water balance model.
+Developing a novel inverse model allowed us to estimate species effective rooting depths (ERD), defined as the depth ($z$) whose soil water dynamics ($\Psi_{x,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015).
+This model incorporated impact of atmospheric drought (VPD) on growth and was constrined by above-ground hydraulic traits, namely, species leaf vulnerability curves ($K_{Leaf}$ vs.
+$\Psi_{x,~Leaf}$).
+We collected new data on vulnerability curves for xx species.
+Identifying scaling relationships between these curves and commonly available traits (SLA, specific wood density) allowed us to scale-up these curves to `r length(data.model.AB$sp)` tree species in the community.
+We evaluated ERD against existing data on plant and soil water isotopes ($\delta^2H$) and coordination of ERD with hydraulic traits and mortality rates.
+
+\
+We test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns species on a fast-slow trade-off axis.
+Specific hypotheses applicable to this forest included: (i) dynamics of soil water potential at depth does not cross critical plant thresholds even during multi-annual droughts because of high total annual precipitation to evapotranspiration ratio (not sure if this should be stated as a hypothesis or just an explanation of the results); (ii) trees with deeper ERD show 'fast' hydraulic traits: higher maximum stem hydraulic conductance (Kmax), but greater xylem vulnerability to embolism ($\Psi_{88,~Stem}$), higher $\Psi_{TLP}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~Stem}$); and finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure.
+To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and mortality outcomes for a species-rich forest.
 
@@
 \
-In the following sections we describe the study site, the models and data used to estimate and evaluate (1) the soil water potential dynamics by depth over the forests’ rooting zone, (2) species-specific percent loss of leaf conductivity curves (PLC), and (3) species-specific effective rooting depths.
+In the following sections we describe the study site, the models and data used to estimate and evaluate (1) soil water potential dynamics by depth over the forests’ rooting zone, (2) species-specific leaf vulnerability curves, and (3) species-specific effective rooting depths.
 These are followed by analyses to address the proposed hypotheses.
-Note that only new data on leaf physiological traits collected during this study are described below, details of all other datasets employed are described in the Supplementary Information.
+Note that only new data on leaf vulnerability curves collected during this study are described below, details of all other datasets used are described in the Supplementary Information.
 
-## The Study Site
+## Study Site Description
 \
-We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterisation in terms of long-term forest demography, plant functional ecology, hydrology, including availability of an independent isotopic tracer study on trees’ water-sourcing depths [@Meinzer:1999].
+We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterisation in terms of long-term forest demography, plant functional ecology, hydrology, and water sourcing depth water isotopes on tree species and soil from an independent isotopic tracer study on trees’ water-sourcing depths [@Meinzer:1999].
 The 1,560 hectare island was formed after the river Chagres was dammed in 1914 that created Gatun Lake, the main channel of the Panama Canal.
@@
 Although percentage of canopy species that are at least partly deciduous during the dry season is 28%, the maximum percentage of deciduous individuals in the canopy is only 6.3%, and the percentage of crown area that is deciduous is estimated at 9.7% [@Condit:2000].
-Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit:1998395; @Hubbell:1999; @Anderson-Teixeira:2014].
+Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014].
 
 \
-In the 50 ha plot all stems above 10 mm are mapped, tagged with a unique number, identified to species level and measured every 5 years through 1985-2015 for diameter growth, mortality as well as recruitment of new stems into the 10 mm size class [@Condit:2017].
+In the 50 ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species level and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm size class [@Condit:2017].
 This inventory represents 321 woody species across the eight censuses.
@@
 To estimate dynamics of water availability by depth over the forest’s root zone, we implemented a 1-D hydrological model, the Energy Exascale Earth System Land Model (ELM), coupled with a Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES).
-FATES is a state-of-the-art, community-based, open-source model increasingly used for studying climate-vegetation interactions.
+FATES is a state-of-the-art, community-based, open-source model used for studying climate-vegetation interactions.
 FATES is considered a next generation vegetation dynamics model for ELM, with a size-structured group of plants (cohorts) and successional trajectory-based patches using the ecosystem demography approach......
-(How ELM talks to FATES)...Detailed descriptions can be found elsewhere [@Koven:2020; @Fisher: ).
+(How ELM talks to FATES)...Detailed descriptions can be found elsewhere [@Koven:2020; @Fisher:2014; @Bonan:2013].
 
 \
-Below we describe the equation of water balance in ELM-FATES that is most relevant for our study site, even though ELM-FATES is capable of incorporating numerous other hydrological processes relevant for wetlands and lakes and snow (CLM4.5 Tech Note).
-ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\Delta W_{can}$, surface water $\Delta W_{sfc}$, soil water $\Delta W_{liq, z}$, and water in an unconfined aquifer $\Delta W_{a}$ (omitting processes relevant to snow or wetlands and lakes), and can be described as
+Below we describe the equation of water balance in ELM-FATES that is most relevant for our study site, even though ELM-FATES is capable of incorporating numerous other hydrological processes relevant for wetlands and lakes and snow [@Bonan:2013].
+ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\Delta W_{can}$, surface water $\Delta W_{sfc}$, soil water $\Delta W_{liq, z}$, and water in an unconfined aquifer $\Delta W_{a}$ (omitting processes relevant to snow, wetlands or lakes), and can be described as
 
@@
 
-where, $q_{rain}$ is rainfall, $E_{v}$ is evapotranspiration (ET) of the forest, $E_{g}$ is ground evaporation, $q_{over}$ is surface runoff, $q_{h2osfc}$ is runoff from surface water storage, $q_{drai}$ is sub-surface drainage (all in mm H2O) and $\Delta t$ is the time-step.
-z are the 15 soil layers exponentially distributed in depth, here, (r soil.depths) (m).
+where, $q_{rain}$ is rainfall, $E_{v}$ is evapotranspiration (ET) of the forest, $E_{g}$ is ground evaporation, $q_{over}$ is surface runoff, $q_{h2osfc}$ is runoff from surface water storage, $q_{drai}$ is sub-surface drainage (all in mm H~2~O) and $\Delta t$ is the time-step.
+$z$ represents the 15 soil layers exponentially distributed in depth, here, (r soil.depths) (m).
 
@@
 \
-To parameterise the model, we identified eleven parameters relevant for in Eq.
-\@ref(eq:water-balance), determined their ranges based on literature for the study site or for the tropics ([Table 1](#param)), and sampled 10, 000 ensembles from this parameter space with Latin Hypercube Sampling.
+To parameterise the model, we identified eleven parameters relevant for Eq.
+\@ref(eq:water-balance), determined their ranges based on literature for the study site or for the tropics ([Table 1](#param)), and sampled 5000 ensembles from this parameter space with Latin Hypercube Sampling (LHS).
+In the Conrad catchment soil hydraulic conductivity declined by depth based on existing data for the depths of 12.5 and 60 cm (@Kinner:2004; [Table 1](#param)].
+We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths < 12.5 cm and > 60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig.
+S\@ref(fig:ksat-profiles)).
+For all runs, instead of ELM's default soil texture-based soil water retention curve scheme, we used parameters of a soil water retention curve that we derived from existing data on volumetric water content \theta vs.
+$\Psi_{soil}$ for depths 15, 40 and 100 cm [@Kupers:2019].
+A Campbell (1974) empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences among depths were nominal:
 
-\
-We ran ELM-FATES for these 10,000 ensembles with hourly climate drivers from 1985-2018 initialised with the stand structure of the 50 ha plot in a single site mode ($0.1^\circ$ x $0.1^\circ$ centered at $9.1530^\circ$N, $280.2539^\circ$E).
+\begin{equation}
+\Psi_{soil} = \Psi_e\left(\frac{\theta}{\theta_{s}}\right)^{-b}
+(\#eq:soil-ret-curve)
+\end{equation}
+
+where $\Psi_e$ is the air-entry matric potential (200, mmH~2~O), \theta_{s} is the saturated volumetric water content (0.51, $cm^{3}cm^{-3}$), and $b$, an index for soil pore-size distribution (10, unitless), respectively.
+
+We ran ELM-FATES for the 5000 parameter ensembles with hourly climate drivers from 1985-2018 initialised with the stand structure of the 50 ha plot in a single site mode ($0.1^\circ$ x $0.1^\circ$ centered at $9.1530^\circ$N, $280.2539^\circ$E).
 As our key interest here was in soil water availability rather than forest dynamics, we ran ELM-FATES in simplified modes: static stand structure and without plant hydrodynamics, and with a single Plant Functional Type (PFT) of evergreen trees.
@@
 ### Model evaluation
+
 \
-We calibrated the model against three key fluxes in Eq.
-\@ref(eq:water-balance), namely, (i) evapotranspiration ($E_{v} + E_{g}$) from the eddy flux tower, (ii) stream discharge for the Conrad Catchment ($q_{over} + q_{drai}$) and (iii) Volumetric Water Content (VWC) obtained with Time Domain Reflectometry (TDR) at three depths (10, 40 and 100 cm; relating to $W_{liq, z}$) at point location as well as opportunistic dry season VWC measured over those depths for the entire 50-ha plot (Figure \@ref(fig:hydro-calib).
-Among the 10,000 parameter ensembles we chose 100 ensembles that best-fit all three fluxes simultaneously (add details...that is those that maximised the sum of regression coefficients between observed and simulated variables.) See [Table 1](#param) for best-fit parameter ranges.
+We calibrated the model over 2012-2018 against three key fluxes in Eq.
+\@ref(eq:water-balance), namely, (i) evapotranspiration (ET, $E_{v} + E_{g}$) from the eddy flux tower (monthly, 2012-2017), (ii) stream discharge for the Conrad Catchment ($q_{over} + q_{drai}$; daily, 2012-2018) and (iii) volumetric water content (VWC; $W_{liq, z}$; Fig.
+\@ref(fig:hydro-calib; daily, 2015-2018).
+We collected VWC data (2016-2018) from a point location at the eddy flux tower with Time Domain Reflectometry (TDR) using three horizontal probes at depths 10, 40 and 100 cm and one vertical probe (0-15 cm).
+TDR data for the vertical probe was calibrated based on measurements of gravimetric soil moisture on several occasions (Fig.
+\@ref(fig:TDR-calib)) and mean bulk dry density of 0.8 gcm{-3} (n = 5, cylinder volume = 1250 cm{3}).
+The same calibration equation was used for other probes.
+We also leveraged existing opportunistic dry season VWC measurements for the entire 50-ha plot (15, 40 and 100 cm in 2015 and 15 & 40 cm in 2016; @Kupers:2019).
+Among the 5000 parameter ensembles we chose 100 ensembles that best-fit all three fluxes simultaneously (add details...that is those that maximised the sum of regression coefficients between observed and simulated variables.) See [Table 1](#param) for best-fit parameter ranges.
+We ran ELM-FATES with the chosen 100 ensembles from 1985-2018, with first 5 years used for model spin-up.
 
-```{r hydro-calib, dev='png', fig.show='hide', eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap ="\\label{lab-hydro-calib}Model evaluation: Observations (red line) of Volumetric Water Content, Evapotranspiration and stream discharge against simulations (blue lines) from 100 best-fit parameter ensembles runs of ELM-FATES.
-This figure will be reformatted to compose individual figures into one."}
+```{r hydro-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-hydro-calib} Model evaluation: Observations (red line) of Volumetric Water Content, Evapotranspiration and stream discharge against simulations (blue lines) from 100 best-fit parameter ensemble runs of ELM-FATES.
+This figure will be reformatted to compose individual figures into one.")}
 p1 <- cowplot::ggdraw() + cowplot::draw_image("figures/2019-10-14_5000/best-fits/swc_Obs_vs_model_daily_horizontal_with_steph_mean.jpeg", scale = 1)
@@
 $\Psi_{x,~Leaf}$) on adult trees of `r length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))` common species (see SI x).
-We fitted exponential, sigmoidal, logistic forms of equations to species-level data as prescribed by [@Guyot:2011] and based on AIC criteria chose exponential form (Fig.
+We fitted exponential, sigmoidal, logistic equations to species-level data as prescribed by @Guyot:2011.
+Exponential form was chosen as it best-fitted all but one species based on AIC criteria (Fig.
 S\@ref(fig:obs-plc-curves)), described as
@@
 
-We fitted parameters $A$ and $B$ in Eq.
-\@ref(eq:vc-exp) with polynomial equations a functions of two traits that were available at community-scale, namely, $WSG$ and $LMA$, defined as mean wood specific gravity after drying at 100$^\circ$C (g cm^{-3}^) and mean leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules (g m^{-2}^) for leaves receiving direct sunlight, respectively.
+We identified scaling relationships between fitted parameters $A$ and $B$ in Eq.
+\@ref(eq:vc-exp) and two traits that were available at community-scale, namely, $WSG$ and $LMA$, defined as mean wood specific gravity after drying at 100$^\circ$C (g cm^{-3}^) and mean leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules (g m^{-2}^) for leaves receiving direct sunlight, respectively.
+Fitted polynomial equations are described as,
 
@@
 We thus sequentially used Eq.
-(\#eq:b-mod) and (\#eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r length(unique(data.model.AB$sp))` species for which WSG and LMA data were available and fitted vulnerability curves (see Fig.
+\@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r length(unique(data.model.AB$sp))` species for which WSG and LMA data were available and fitted vulnerability curves (see Fig.
 \@ref(fig:comm-plc-curves)).
@@
 ```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-plc-curves} Modelled standardized 'Vulnerability curves' ($K_{x,~Leaf}$ vs.
-$\\Psi_{x,~Leaf}$) at community-scale (", nrow(data.model.AB$sp)," species) based on species data on $WSG$ & $LMA$ and their relationships with vulnerability curve parameters measured on ", sp.soft.filled$AB.data.model," species.
+$\\Psi_{x,~Leaf}$) at community-scale (", length(data.model.AB$sp)," species) based on data on $WSG$ & $LMA$ and their relationships with vulnerability curve parameters measured on ", sp.soft.filled$AB.data.model," species.
 Each line represents a species, color coded by $WSG$.")}
@@
 The diameter model of growth was fitted based on all trees' data pooled together as there was insufficient species-specific data for large trees (SI ).
-We then standardised (centered and scaled) an individual tree time series ($Resid_{i,t}$) and obtained species-level (*n* = 32) times-series of medians ($Growth_{t}$) that had data on at least three trees (median = xx).
-
+We then standardised (centered and scaled) an individual tree time series ($Resid_{i,t}$) and obtained species-level (*n* = `r length(erd.sp)`) times-series of medians ($Growth_{t}$) that had data on at least three trees (median = xx).
 As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between GPP (gCm^2^day^-1^) and VPD (kPa; $GPP = f(VPD) + error$;  SI) and used standardised $f(VPD)$ (range 0-1) as $StdVPD$ in Eqs.
@@
 
-```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-iso}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE; m) versus dry season stable isotopic concentration $\\delta$^2^H (mean $\\pm$ (add error); ‰) for ", length(unique(erd.iso$sp))," canopy species from Barro Colorado Island (data from [@Meinzer:1999]").
-See Table S2 for scientific names corresponding to species codes.}
+```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-iso}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE; m) versus dry season stable isotopic concentration $\\delta$^2^H (mean $\\pm$ (add error); ‰) for ", length(unique(erd.iso$sp))," canopy species from Barro Colorado Island.
+Data from [@Meinzer:1999].
+See [Table S1]{erd-sp} for scientific names corresponding to species codes.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_phenology_Meinzer_gr.Psi.VPD.jpeg"))
@@
 
-We calculated mortality rates for large trees (>= 10 cm dbh) of species with ERD estimates (all canopy species with maximum height >= 30 m) and average adbunance of at least 20 trees (one species was eleminated with an average abundance of 9 trees).
+We calculated mortality rates for large trees (>= 10 cm dbh) of species with ERD estimates (all canopy species with maximum height >= 30 m) and average adbunance of at least 20 trees (one species was eleminated with an average abundance of 9 trees; *n* = `r length(erd.mrate.sp)`).
 For each of the seven census intervals ($t$) in the 33 year record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successives censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{t}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50 ha plot of a species present in cen~1~ and dead in cen~2~, respectively, and t is the duration based on mean dates of cen~1~ and cen~2~.
@@
 
-All statistical analyses were done in R statistical environment (version 3.5.3, @R:2020).
+All statistical analyses were done in R statistical environment (version 3.5.3, @R:2019).
 
@@
 \
-In this forest modelled Effective Rooting Depths (ERD) for the 32 large, canopy tree species varied from 0.5 m to 13 m (Figure \@ref(fig:sp-erd)).
+In this forest modelled Effective Rooting Depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from 0.5 m to 13 m (Figure \@ref(fig:sp-erd)).
 For the subset of species for which hydraulic properties were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max, Stem}$; *R*^2^ = 0.39, *P* = 0.098, *n* = 8; Figure \@ref(fig:erd-hyd)A), less negative leaf Turgor Loss Point ($\Psi_{TLP}$); *R*^2^ = 0.6, p = 0.024, *n* = 8; Figure \@ref(fig:erd-hyd)B), greater vulnerability to embolism from cavitation with less negative ($\Psi_{88,~Stem}$) (*R*^2^ = 0.72, p = 0.007, *n* = 8; Figure \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~Stem}$; *R*^2^ = 0.74, p = 0.013, *n* = 8; Figure \@ref(fig:erd-hyd)D).
 
-```{r sp-erd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap= paste0("\\label{lab-sp-erd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for 32 large, canopy species of Barro Colorado Island.
-See Table S2 for scientific names corresponding to species codes.")}
+```{r sp-erd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap= paste0("\\label{lab-sp-erd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for ", length(erd.sp)," large, canopy species of Barro Colorado Island.
+See [Table S1]{erd-sp} for scientific names corresponding to species codes.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/ERD_by_sp_large_canopy.jpeg"))
@@
 \
-Species with deeper $ERD$ showed lower mean mortality rates across the 33 year record (1982-2015; *R*^2^ = 0.13, *P* = 0.029, *n* = xx; Figure \@ref(fig:erd-mort)).
+Species with deeper $ERD$ showed lower mean mortality rates across the 33 year record (1982-2015; *R*^2^ = 0.13, *P* = 0.029, *n* = `r length(erd.mrate.sp)`; Figure \@ref(fig:erd-mort)).
 For five out of seven census intervals, species with shallower $ERD$ had significantly higher mortality anomalies than deeper ERD species; Figure (\@ref(fig:erd-mort-anom)).
@@
 
-```{r erd-mort, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort}Mortality rate (mean $\\pm$ 1SE) versus Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for xx canopy species found on Barro Colorado Island.
-*R*^2^ and *P*-values are given in panel insets."}
+```{r erd-mort, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-mort}Mortality rate (mean $\\pm$ 1SE) versus Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for ", length(erd.mrate.sp)," canopy species found on Barro Colorado Island.
+*R*^2^ and *P*-values are given in panel insets.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_rate_by rdi.gr.jpeg"))
@@
 
-```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{x, z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~Leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for xx canopy species found on Barro Colorado Island.
+```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{x, z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~Leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for ", length(erd.mrate.sp)," canopy species found on Barro Colorado Island.
 Each circle represents a species.
@@
 Analysis of time spent below ($\Psi_{crit}$) in the soil zone of ERD showed that this was likely due to low realised risk of hydraulic failure for deeper ERD species.
-We thus show, for the first time, that interspecific variation in trees' effective rooting depths aligns species on an acquisitive-conservative strategy axis by demonstrating a link between trees' realised drought experience, above-ground hydraulic architecture and long-term mortality outcomes.
-On the aquisitive end were species with hydraulically efficient, but risky suit of traits optimised for an environment of reliable, abundant source of water at depth, which actualised to a lower associated mortality-risk, whilst at the conservative end were species with inefficient, but safer xylem, associated with a water environment of high variability.
+We thus show, for the first time, that interspecific variation in trees' effective rooting depths aligns species on a 'fast-slow' strategy axis by demonstrating a link between trees' realised drought experience, above-ground hydraulic architecture and long-term mortality outcomes.
+On the aquisitive end were species with hydraulically efficient, but risky suit of traits optimised for an environment of reliable, abundant source of water at depth, which actualised to a lower associated mortality-risk, whilst at the slow end were species with inefficient, but safer xylem, associated with a water environment of high variability.
 
-Acquisitive strategy is thought to be optimised for high resource environments with high performance outcome and a large decline in performance under stresses or low resource environment; whilst conservative strategy have a moderate performance in high resource environment that declines with decreasing resources, while retaining a competitive advantage over acquisitives under lower resources.
-
-In other contexts aquisitive-conservative trait trade-off is associated with a growth-survival trade-off, with aquisitive species achieving higher potential growth at the cost of lower survival, but in case of ERD axis, aquisitiveness did not relate Normally Our results highlight the need..
-to higher growth, but
-
-## Trait-Trait Coordination
+## Higher xylem embolism vulnerability of species with Deeper Effective Rooting Depths
 
@@
 
-## Trait & Demography
+fast-slow axis is matched by presence of an efficiency-safety trade-off between high water conductance (high Kmax) and cavitation-resistant xylem (low $\Psi_{50}$), although community-wide examples are rare [@Hacke:2006;@Gleason:2016;].
+Low $\Psi_{50}$ is in turn associated with low $ and both are broadly associated with low water availability[@Choat:2012; @Bartlett:2012].
+Slow strategy is associated with larger safety-margin ($\Psi_{min} - \Psi_{88,~Stem}$) in two angiosperm systems.
+High Kmax is associated with photosynthetic rates with faster growth potential.
 
-## Environment, Trait & Demography
+## Lower mortality of species with deeper Effective Rooting Depths
 
-Mortality could be immediate or may result after decades of drought occurrence.
-Mortality could occur over
+Mortality could be immediate or may result after decades of drought occurrence[@Cailleret:2016].
 
@@
 
-- Although we attempted a coarse estimate of species leaf cover seasonality by combining leaf-fall data from litter traps and forest-level LAI data for leaf-on periods, more accuarate estimations of leaf-cover seasonality are required, such as those obtained from drone images [@Park: 2019]
+- Although we attempted a coarse estimate of species leaf cover seasonality by combining leaf-fall data from litter traps and forest-level LAI data for leaf-on periods, more accuarate estimations of leaf-cover seasonality are required, such as those obtained from drone images [@Park.2019]
 
@@
 
-- Future studies would benefit from modeling the complete hydrodynamics of water potential gradients over the soil - plant (roots, stems, leaves) - atmosphere continuum (SPAC) [@Johnson:2017; @Christoffersen 2017].
+- Future studies would benefit from modeling the complete hydrodynamics of water potential gradients over the soil - plant (roots, stems, leaves) - atmosphere continuum (SPAC) [@Johnson:2018fad; @Christoffersen:2016kd2d].
 However, to parameterise such a model at community-scale would need community-scale data on hard traits such as stem and root vulnerability curves; if not, model development would be needed to up-scale data from a few species, for example, via softer traits as we did for leaf vulnerability curves.
@@
 ## Data Availability
+
+#### Table 1: ELM-FATES parameters used to generate ensembles, with description, chosen ranges, references and rationale. No. 1-6 are parameters specific to FATES.  {#param}
+
+| No. | Parameter | Description | Range & Units | Method of determination with reference |
+| -- | ------ | ------------- | ------------- |  ------------------------------------------------------------------------------------------------- | 
+| 1 | fates_leaf_ BB_slope | stomatal slope parameter, as per Ball-Berry | 4 - 16 Unitless  | The range of Ball Berry parameter fitted across several studies as compiled in Table 2 of [@Medlyn:2012] |
+| 2 | fates_leaf_ slatop | Specific Leaf Area (SLA) at the top of canopy, projected area basis | 0.0042 - 0.0400 $m^2gC^{-1}$ | Based on the observed range of Leaf Mass Area (LMA)--24.97 to 235.8 $gm^{-2}$-- for individual tree variation across 51 tree species in Barro Colorado Island (this study), and assumption of 50% carbon content in biomass.
+|
+| 3 | fates_leaf_ vcmax25top | Maximum carboxylation rate of Rubisco at $25^\circ$C, canopy top | 22.7 - 92.5 umol $CO^2m^2s^{-1}$ | Observed range of $V_{cmax}$ at  $25^\circ$C for tropical tree species under relative radiation of 50% or greater, thereby excluding highly shaded leaves.
+Values derived with a conversion specific for CLM model are used.
+[@Ali:2015] |
+| 4 | fates_roota_ par | CLM rooting distribution parameter | 5.9 - 7.4 per m | Parameter a in Eq.
+2 in [@Zeng:2001] that regulates the shape of the rooting profile.
+Range corresponds to this parameter specified for BATS (or IGBP) land cover classified Deciduous Broadleaf Trees (5.9 $m^{-1}$ and Evergreen Broadleaf Trees  (7.4 $m^{-1}$) as given in Table 1 of [@Zeng:2001].
+|
+| 5 | fates_rootb_ par | CLM rooting distribution parameter | 0.217 - 1.956 $m^{-1}$ | Parameter b in Eq.
+2 in [@Zeng:2001] that regulates the depth of the rooting profile.
+Chosen range of *b* is derived using this equation so as to fit the observed range of rooting depth (d~r~) of 2 - 18 m for Tropical Deciduous Forest (mean ± Standard Error (SE); 3.7 ± 0.5, *n* = 5 trees; min = 2, max = 4.7)  and Tropical Evergreen Forest (mean $\pm$ S.E.; 7.3 $\pm$ 0.5, n = 3 trees and 3 communities; min = 2, max = 18 m) combined [@Canadell:1996].
+Besides the direct observation of roots at 18 m included by [Nepstad:1994] in Paragominas, eastern Amazonia that is included in the above study; in Tapajos, eastern Amazonia water extraction by roots was also inferred up to 18 m.
+[@Davidson:2011] |
+| 6 | fates_smpsc | Soil water potential at full stomatal closure | 113000 - 242000 mm | Based on observed range of -1.13 to -2.42 MPa for leaf turgor loss point for individual tree variation across 49 BCI tree species.
+This study.|
+| 7 | aveDTB | Distance to bedrock | 3 - 18 m | Ben Turner; pers. comm. | 
+| 8 | fmax | The maximum fractional saturated area |  unitless | Empirical | 
+| 9 | HKSAT_ADJ | Adjusting factor for soil hydraulic conductivity | 1 - 8 unitless | To account for high macroporosity and direct flow paths in tropical soils that is not accounted for by small soil core samples [@Broedel:2017; @Kinner:2004; @Tomasella:1998] |
+| 10 | HKSAT | Soil hydraulic conductivity profile | 0.007 - 0.014 mm/s for =< 12.5 cm; 5.56e-05, 0.0003 for >= 60 cm | For Conrad catchment observed median Ksat (95% CI) for 12.5 cm depth is 38.3 mm/hr (25.4 - 51.2, n = 75), while for 60 cm depth 0.7 mm/hr (0.2 - 1.2, n = 40) [@Kinner:2004].
+For reference, a storm with 12.5 mm/hr rainfall intensity has a 0.2 probability of occuring in any given rainfall event.
+|
+| 11 | fpi_max | Maximum interception fraction of precipitation | 0.05 - 0.44 unitless | Based on throughfall data from [@Zimmermann:2010kzl] and precipitation data for BCI from STRI Physical Monitoring program, defined for precipitation events greater than 10 mm.
+|
+
 ## Supplementary Information
 
-```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts} Extreme soil droughts by census interval (panels) at three representative depths (0.5, 1 and 1.7 m depicted in different colors).
-Mean soil water potential ($Psi_z$}) for a given DOY and depth z are shown as lines.
-Filled areas show extreme drought periods, when $Psi_z$ declined below 2.5 percentile of $Psi_z$ for the given DOY and depth z in an individual year within the cenus intervals.
-Eextreme drought periods from multiple years may be overlaid within a panel."}
+```{r ksat-profiles, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-ksat-profiles} Generated 5000 curves of soil hydraulic conductivity variation by depth based on Latin Hypercube Sampling within 95% CI of observed data at depths of 12.5 cm and 60 cm [@Kinner:2004], assuming a linear decline between those depths, and values before 12.5 and after 60 cm as that of 12.5 cm and 60 cm, respectively.
+Only 100 curves are shown for clarity.
+Note the log scale on horizontal axis."}
+knitr::include_graphics(file.path("figures/Bootstrapped Ksat profiles within observed 95CI_random_two_hundred.pdf"))
+```
+
+```{r TDR-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-TDR-calib} Volumetric Water Content vs.
+TDR data for a vertical TDR probe (0-15 cm depth) at the base of Eddy Flux Tower."}
+knitr::include_graphics(file.path("figures/Matteos_TDR_SWC_data_linear_model.jpeg"))
+```
+```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts} Extreme soil droughts by census interval (panels) at three representative depths: 0.5, 1 and 1.7 m, depicted in different colors.
+Mean soil water potential ($Psi_z$) for a given DOY and depth z are shown as lines.
+Filled areas show extreme drought periods, when $Psi_z$ declined below 2.5 percentile of $Psi_z$ for the given DOY and depth z in any given year within the cenus intervals.
+Extreme drought periods from multiple years may be overlaid within a panel.
+Wet season is shown in gray as beginning with DOY "}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q2.5_depths_in_inverse_model.jpeg"))
@@
 
-```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-obs-plc-curves} 'vulnerability curves' fitted to observed data shown as standardized *K*~Leaf~ vs.
-$\\Psi$~Leaf~ for `r length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))` tree species from Panama, color coded by Wood Specific Gravity (units)."}
+```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves} Leaf 'vulnerability curves', that is standardized *K*~Leaf~ vs.
+$\\Psi$~Leaf~, fitted to observed data on ",  length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))," tree species from Panama, color coded by Wood Specific Gravity ($gcm^{-3}$).")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB_for_data_sp.jpeg"))
@@
 
-```{r ab-fit, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', fresults = "asis", fig.cap = "\\label{lab-ab-fit}Parameters $A$ & $B$ of 'vulnerability curve' fitted to data versus predicted from WSG and LMA"}
+```{r ab-fit, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ab-fit}Parameters $A$ (panel A) & $B$ (panel B) of leaf vulnerability curves that were fitted to observed data on *K*~Leaf~ vs.
+$\\Psi$~Leaf~ for ", sp.soft.filled$AB.data.model," species vs.
+those that were predicted from models based on species soft-traits, WSG and LMA.")}
 p1 <- cowplot::ggdraw() + cowplot::draw_image("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/A_data_vs_model.jpeg", scale = 1)
@@
 
-#### Table 1: ELM-FATES parameters chosen for enseble generation, their description, chosen ranges and references and/or rationale.  {#param}
+#### Table S1: Scientific names and codes for tree species for which Effective Rooting Depths were estimated. {#erd-sp}
 
-| S.N. | Parameter | Description | Range (Units) | Method of determination with reference
-| -- | ------ | ------------- | ------------- |  ------------------------------------------------------------------------------------------------- |
-| 1 | fates_leaf_ BB_slope | stomatal slope parameter, as per Ball-Berry | 4 - 16 Unitless  | The range of Ball Berry parameter fitted across several studies as compiled in Table 2 of [@Medlyn:2012] |
-| 2 | fates_leaf_ slatop | Specific Leaf Area (SLA) at the top of canopy, projected area basis | 0.0042 - 0.0400 $m^2gC^{-1}$ | Based on the observed range of Leaf Mass Area (LMA)--24.97 to 235.8 $gm^{-2}$-- for individual tree variation across 51 tree species in Barro Colorado Island (this study), and assumption of 50% carbon content in biomass.
-|
-| 3 | fates_leaf_ vcmax25top | Maximum carboxylation rate of Rubisco at $25^\circ$C, canopy top | 22.7 - 92.5 umol $CO^2m^2s^{-1}$ | Observed range of $V_{cmax}$ at  $25^\circ$C for tropical tree species under relative radiation of 50% or greater, thereby excluding highly shaded leaves.
-Values derived with a conversion specific for CLM model are used.
-[@Ali:2015] |
-| 4 | fates_roota_ par | CLM rooting distribution parameter | 5.9 - 7.4 per m | Parameter a in Eq.
-2 in [@Zeng:2001] that regulates the shape of the rooting profile.
-Range corresponds to this parameter specified for BATS (or IGBP) land cover classified Deciduous Broadleaf Trees (5.9 $m^{-1}$ and Evergreen Broadleaf Trees  (7.4 $m^{-1}$) as given in Table 1 of [@Zeng:2001].
-|
-| 5 | fates_rootb_ par | CLM rooting distribution parameter | 0.217 - 1.956 $m^{-1}$ | Parameter b in Eq.
-2 in [@Zeng:2001] that regulates the depth of the rooting profile.
-Chosen range of *b* is derived using this equation so as to fit the observed range of rooting depth (d~r~) of 2 - 18 m for Tropical Deciduous Forest (mean ± Standard Error (SE); 3.7 ± 0.5, *n* = 5 trees; min = 2, max = 4.7)  and Tropical Evergreen Forest (mean $\pm$ S.E.; 7.3 $\pm$ 0.5, n = 3 trees and 3 communities; min = 2, max = 18 m) combined [@Canadell:1996].
-Besides the direct observation of roots at 18 m included by [Nepstad:1994] in Paragominas, eastern Amazonia that is included in the above study; in Tapajos, eastern Amazonia water extraction by roots was also inferred up to 18 m.
-[Davidson:2011] |
-| 6 | fates_smpsc | Soil water potential at full stomatal closure | 113000 - 242000 mm | Based on observed range of -1.13 to -2.42 MPa for leaf turgor loss point for individual tree variation across 49 BCI tree species.
-This study|
-| 7 | aveDTB | Distance to bedrock | 3 - 18 m | Ben Turner; pers. comm. | 
-| 8 | fmax | The maximum fractional saturated area |  unitless | Empirical | 
-| 10 | HKSAT_ADJ | Adjusting factor for soil hydraulic conductivity | 1 - 8 unitless | To account for high macroporosity and direct flow paths in tropical soils that is not accounted for by small soil core samples [@Broedel:2017; @Kinner:2004; @Tomasella:1998] |
-| 11 | fpi_max | Maximum interception fraction of precipitation | 0.05 - 0.44 unitless | Based on throughfall data from [@Zimmermann:2010kzl] and precipitation data for BCI from STRI Physical Monitoring program, defined for precipitation events greater than 10 mm.
-|
+```{r, results = 'asis', echo = FALSE}
+writeLines("td, th { padding : 10px } th { background-color : brown ; color : white; border : 1px solid white; } td { color : brown ; border : 1px solid brown }", con = "mystyle.css")
+knitr::kable(erd.sp.names, format = "html")
+```
 

### 2020-07-29 manuscript.rmd
Added tables in chunks

@@
 suppressPackageStartupMessages(
-  pacman::p_load(bookdown, knitr, pander, citr, broom, kableExtra))
+  pacman::p_load(bookdown, knitr, pander, citr, broom))
 options(tinytex.verbose = TRUE)
@@
 source("code/manuscript_data_figs.R")
-load("results/manuWorkSpace.RData")
+# load("results/manuWorkSpace.RData")
 # install.packages(c("vctrs", "rlang"))
@@
 Intensifying droughts threaten tropical forests, a biome that plays a disproportionately large role in the global carbon and energy cycle, and supports half of the global biodiversity [@Bonan:2008; @Wright:20050th].
-Tropical forests are considered an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw]: here, historic climatic variation is relatively small[@Mora.2013], and the observed ranges of trees' hydraulic safety margins are relatively narrow [@Choat:2012; @Oliveira:2019].
+Tropical forests are considered an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw]: here, historic climatic variation is relatively small [@Mora.2013], and the observed ranges of trees' hydraulic safety margins are relatively narrow [@Choat:2012; @Oliveira:2019].
 Widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak:2018; @Hilker:2014; @Phillips:2010].
@@
 In general, community-scale data collection for rooting or water-sourcing depths is especially challenging for species-rich tropical forests.
-Further, to estimate plant available water with soil depth requires hydrological data and modeling, but these are rarely co-located with forest inventory sites (but see [@Johnson:2018fad;@Kumagai:2011;@Ivanov:2012]
+Further, to estimate plant available water with soil depth requires hydrological data and modeling, but these are rarely co-located with forest inventory sites (but see @Johnson:2018fad;@Kumagai:2011;@Ivanov:2012)
 
@@
 These efforts have resulted in identification of above-ground traits such as proxies of hydraulic failure as indicators of mortality risk.
-A recent meta-analysis found greater drought-induced mortality rates in species with cavitation-vulnerable xylem ($\Psi_{88,~Stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~Stem}$) [@Anderegg:2016], but counter examples also exist [@Hoffmann:2011; @Paddock:2013]; indicating whether plants reach critical thresholds of hydraulic failure depends on their realised drought intensity.
+A recent meta-analysis found greater drought-induced mortality rates in species with cavitation-vulnerable xylem ($\Psi_{88,~Stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~Stem}$) [@Anderegg:2016], but counter examples also exist [@Hoffmann:2011; @Paddock:2013]; indicating whether plants reach critical thresholds of hydraulic failure depends on their realised drought intensity (see Table \@ref(tab:symbols) for trait symbols and their definitions).
 A rare study that accounted for realised water availability for four tree species in a centennial drought (Texas, USA) found that mortality decreased with deep rootedness even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi_{88,~Stem}$) and had narrower safety-margins ($\Psi_{min} - \Psi_{88,~Stem}$) [@Johnson:2018fad].
@@
 For water resource observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified a fast-slow spectrum, although examples of all trade-offs observed simultaneously are rare, and belowground trait or environment linkages are weak.
-'Fast' species at the productive end of the spectrum have traits with high capacity to acquire, move, store and use water (high Kmax, low wood density) and achieve high growth potential, but build flimsy, disposable tissues (high SLA) and are less tolerant of low resources, whereas species with slow traits are better protected from C losses and are more drought tolerant (low $\Psi_{50}$, low $\Psi_{TLP}$, larger safety margin).
+'Fast' species at the productive end of the spectrum have traits with high capacity to acquire, move, store and use water (high Kmax, low wood density) and achieve high growth potential, but build flimsy, disposable tissues (high SLA) and are less tolerant of low resources, whereas species with slow traits are better protected from C losses and are more drought tolerant (low $\Psi_{50}$, low $\Psi_{tlp}$, larger safety margin).
 Fast strategy is associated with spatially wet sites such as valleys or high rainfall [@Kursar:2009; @Oliveira:2019], but whether and how the fast-slow spectrum holds true for the moisture gradient in the soil column is not clear.
@@
 We test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns species on a fast-slow trade-off axis.
-Specific hypotheses applicable to this forest included: (i) dynamics of soil water potential at depth does not cross critical plant thresholds even during multi-annual droughts because of high total annual precipitation to evapotranspiration ratio (not sure if this should be stated as a hypothesis or just an explanation of the results); (ii) trees with deeper ERD show 'fast' hydraulic traits: higher maximum stem hydraulic conductance (Kmax), but greater xylem vulnerability to embolism ($\Psi_{88,~Stem}$), higher $\Psi_{TLP}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~Stem}$); and finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure.
+Specific hypotheses applicable to this forest included: (i) dynamics of soil water potential at depth does not cross critical plant thresholds even during multi-annual droughts because of high total annual precipitation to evapotranspiration ratio (not sure if this should be stated as a hypothesis or just an explanation of the results); (ii) trees with deeper ERD show 'fast' hydraulic traits: higher maximum stem hydraulic conductance (Kmax), but greater xylem vulnerability to embolism ($\Psi_{88,~Stem}$), higher $\Psi_{tlp}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~Stem}$); and finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure.
 To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and mortality outcomes for a species-rich forest.
@@
 To parameterise the model, we identified eleven parameters relevant for Eq.
-\@ref(eq:water-balance), determined their ranges based on literature for the study site or for the tropics ([Table 1](#param)), and sampled 5000 ensembles from this parameter space with Latin Hypercube Sampling (LHS).
-In the Conrad catchment soil hydraulic conductivity declined by depth based on existing data for the depths of 12.5 and 60 cm (@Kinner:2004; [Table 1](#param)].
+\@ref(eq:water-balance), determined their ranges based on literature for the study site or for the tropics (Table \@ref(tab:param)), and sampled 5000 ensembles from this parameter space with Latin Hypercube Sampling (LHS).
+In the Conrad catchment soil hydraulic conductivity declined by depth based on existing data for the depths of 12.5 and 60 cm (@Kinner:2004; (Table \@ref(tab:param)).
 We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths < 12.5 cm and > 60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig.
 S\@ref(fig:ksat-profiles)).
-For all runs, instead of ELM's default soil texture-based soil water retention curve scheme, we used parameters of a soil water retention curve that we derived from existing data on volumetric water content \theta vs.
+For all runs, instead of ELM's default soil texture-based soil water retention curve, we used a curve that we derived from existing data on volumetric water content ($\theta$) vs.
 $\Psi_{soil}$ for depths 15, 40 and 100 cm [@Kupers:2019].
@@
 
-where $\Psi_e$ is the air-entry matric potential (200, mmH~2~O), \theta_{s} is the saturated volumetric water content (0.51, $cm^{3}cm^{-3}$), and $b$, an index for soil pore-size distribution (10, unitless), respectively.
+where $\Psi_e$ is the air-entry matric potential (200, mmH~2~O), $\theta_{s}$ is the saturated volumetric water content (0.51, $cm^{3}cm^{-3}$), and $b$, an index for soil pore-size distribution (10, unitless), respectively.
 
@@
 TDR data for the vertical probe was calibrated based on measurements of gravimetric soil moisture on several occasions (Fig.
-\@ref(fig:TDR-calib)) and mean bulk dry density of 0.8 gcm{-3} (n = 5, cylinder volume = 1250 cm{3}).
+\@ref(fig:TDR-calib)) and mean bulk dry density of 0.8 gcm{-3} (*n* = 5, cylinder volume = 1250 cm{3}).
 The same calibration equation was used for other probes.
 We also leveraged existing opportunistic dry season VWC measurements for the entire 50-ha plot (15, 40 and 100 cm in 2015 and 15 & 40 cm in 2016; @Kupers:2019).
-Among the 5000 parameter ensembles we chose 100 ensembles that best-fit all three fluxes simultaneously (add details...that is those that maximised the sum of regression coefficients between observed and simulated variables.) See [Table 1](#param) for best-fit parameter ranges.
+Among the 5000 parameter ensembles we chose 100 ensembles that best-fit all three fluxes simultaneously (add details...that is those that maximised the sum of regression coefficients between observed and simulated variables.) See Table \ref(tab:param) for best-fit parameter ranges.
 We ran ELM-FATES with the chosen 100 ensembles from 1985-2018, with first 5 years used for model spin-up.
 
-```{r hydro-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-hydro-calib} Model evaluation: Observations (red line) of Volumetric Water Content, Evapotranspiration and stream discharge against simulations (blue lines) from 100 best-fit parameter ensemble runs of ELM-FATES.
-This figure will be reformatted to compose individual figures into one.")}
-p1 <- cowplot::ggdraw() + cowplot::draw_image("figures/2019-10-14_5000/best-fits/swc_Obs_vs_model_daily_horizontal_with_steph_mean.jpeg", scale = 1)
-p2 <- cowplot::ggdraw() + cowplot::draw_image("figures/2019-10-14_5000/best-fits/QRUNOFF_Obs_vs_model_monthly.jpeg", scale = 1)
-p3 <- cowplot::ggdraw() +
-  cowplot::draw_image("figures/2019-10-14_5000/best-fits/ET_Obs_vs_model_monthly.jpeg", scale = 1)
-bottom_row <- cowplot::plot_grid(p2, p3, labels = c('B', 'C'), label_size = 12, ncol = 2)
-cowplot::plot_grid(p1, bottom_row, labels = c('A', ''), label_size = 12, ncol = 1, rel_widths = c(2, 1))
+```{r hydro-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-hydro-calib} Model evaluation: Observations (red lines) versus simulations (blue lines) from 100 best-fit parameter ensemble runs of ELM-FATES for daily Volumetric Water Content (VWC) by depth for horizontal probes at three depths (10, 40 and 100 cm; A) and a vertical probe (0-15 cm; B); monthly stream discharge (C) and monthly Evapotranspiration from eddy flux tower (D).
+Points (red) in (A) show average plot-wide observations of VWC with 95% CI (error bars).[Goodness-of-fit values to be added on the panels.]")}
+knitr::include_graphics(file.path("figures/hydro-calib2.jpeg"))
 ```
@@
 
-To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD with maximum stem hydraulic conductance ($K_{max, Stem}$), leaf Turgor Loss Point ($\Psi_{TLP}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~Stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~Stem}$).
+To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD with maximum stem hydraulic conductance ($K_{max, Stem}$), leaf Turgor Loss Point ($\Psi_{tlp}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~Stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~Stem}$).
 
@@
 In this forest modelled Effective Rooting Depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from 0.5 m to 13 m (Figure \@ref(fig:sp-erd)).
-For the subset of species for which hydraulic properties were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max, Stem}$; *R*^2^ = 0.39, *P* = 0.098, *n* = 8; Figure \@ref(fig:erd-hyd)A), less negative leaf Turgor Loss Point ($\Psi_{TLP}$); *R*^2^ = 0.6, p = 0.024, *n* = 8; Figure \@ref(fig:erd-hyd)B), greater vulnerability to embolism from cavitation with less negative ($\Psi_{88,~Stem}$) (*R*^2^ = 0.72, p = 0.007, *n* = 8; Figure \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~Stem}$; *R*^2^ = 0.74, p = 0.013, *n* = 8; Figure \@ref(fig:erd-hyd)D).
+For the subset of species for which hydraulic properties were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max, Stem}$; *R*^2^ = 0.39, *P* = 0.098, *n* = 8; Figure \@ref(fig:erd-hyd)A), less negative leaf Turgor Loss Point ($\Psi_{tlp}$); *R*^2^ = 0.6, p = 0.024, *n* = 8; Figure \@ref(fig:erd-hyd)B), greater vulnerability to embolism from cavitation with less negative ($\Psi_{88,~Stem}$) (*R*^2^ = 0.72, p = 0.007, *n* = 8; Figure \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~Stem}$; *R*^2^ = 0.74, p = 0.013, *n* = 8; Figure \@ref(fig:erd-hyd)D).
 
@@
 ```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus hydraulic propertiese for eight large, canopy species found on Barro Colorado Island (Panama).
-*R*^2^ and *P*-values are given in panel insets.")}
+*R*^2^ and *P*-values are given in panel insets.[X-Axis labels to be added: Kmax units as in Fig.
+2.
+Water Potentials are in Mpa.]")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/Select_traits_depth.jpeg"))
@@
 
-```{r erd-mort, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-mort}Mortality rate (mean $\\pm$ 1SE) versus Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for ", length(erd.mrate.sp)," canopy species found on Barro Colorado Island.
+```{r erd-mort-changed, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-mort}Mortality rate (mean $\\pm$ 1SE) versus Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for ", length(erd.mrate.sp)," canopy species found on Barro Colorado Island.
 *R*^2^ and *P*-values are given in panel insets.")}
@@
 
-```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{x, z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~Leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for ", length(erd.mrate.sp)," canopy species found on Barro Colorado Island.
+```{r erd-time-new, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{x, z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~Leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for ", length(erd.mrate.sp), " canopy species found on Barro Colorado Island.
 Each circle represents a species.
@@
 fast-slow axis is matched by presence of an efficiency-safety trade-off between high water conductance (high Kmax) and cavitation-resistant xylem (low $\Psi_{50}$), although community-wide examples are rare [@Hacke:2006;@Gleason:2016;].
-Low $\Psi_{50}$ is in turn associated with low $ and both are broadly associated with low water availability[@Choat:2012; @Bartlett:2012].
+Low $\Psi_{50}$ is in turn associated with low $\Psi_{tlp}$ and both are broadly associated with low water availability[@Choat:2012; @Bartlett:2012].
 Slow strategy is associated with larger safety-margin ($\Psi_{min} - \Psi_{88,~Stem}$) in two angiosperm systems.
@@
 ## Data Availability
-
-#### Table 1: ELM-FATES parameters used to generate ensembles, with description, chosen ranges, references and rationale. No. 1-6 are parameters specific to FATES.  {#param}
-
-| No. | Parameter | Description | Range & Units | Method of determination with reference |
-| -- | ------ | ------------- | ------------- |  ------------------------------------------------------------------------------------------------- | 
-| 1 | fates_leaf_ BB_slope | stomatal slope parameter, as per Ball-Berry | 4 - 16 Unitless  | The range of Ball Berry parameter fitted across several studies as compiled in Table 2 of [@Medlyn:2012] |
-| 2 | fates_leaf_ slatop | Specific Leaf Area (SLA) at the top of canopy, projected area basis | 0.0042 - 0.0400 $m^2gC^{-1}$ | Based on the observed range of Leaf Mass Area (LMA)--24.97 to 235.8 $gm^{-2}$-- for individual tree variation across 51 tree species in Barro Colorado Island (this study), and assumption of 50% carbon content in biomass.
-|
-| 3 | fates_leaf_ vcmax25top | Maximum carboxylation rate of Rubisco at $25^\circ$C, canopy top | 22.7 - 92.5 umol $CO^2m^2s^{-1}$ | Observed range of $V_{cmax}$ at  $25^\circ$C for tropical tree species under relative radiation of 50% or greater, thereby excluding highly shaded leaves.
-Values derived with a conversion specific for CLM model are used.
-[@Ali:2015] |
-| 4 | fates_roota_ par | CLM rooting distribution parameter | 5.9 - 7.4 per m | Parameter a in Eq.
-2 in [@Zeng:2001] that regulates the shape of the rooting profile.
-Range corresponds to this parameter specified for BATS (or IGBP) land cover classified Deciduous Broadleaf Trees (5.9 $m^{-1}$ and Evergreen Broadleaf Trees  (7.4 $m^{-1}$) as given in Table 1 of [@Zeng:2001].
-|
-| 5 | fates_rootb_ par | CLM rooting distribution parameter | 0.217 - 1.956 $m^{-1}$ | Parameter b in Eq.
-2 in [@Zeng:2001] that regulates the depth of the rooting profile.
-Chosen range of *b* is derived using this equation so as to fit the observed range of rooting depth (d~r~) of 2 - 18 m for Tropical Deciduous Forest (mean ± Standard Error (SE); 3.7 ± 0.5, *n* = 5 trees; min = 2, max = 4.7)  and Tropical Evergreen Forest (mean $\pm$ S.E.; 7.3 $\pm$ 0.5, n = 3 trees and 3 communities; min = 2, max = 18 m) combined [@Canadell:1996].
-Besides the direct observation of roots at 18 m included by [Nepstad:1994] in Paragominas, eastern Amazonia that is included in the above study; in Tapajos, eastern Amazonia water extraction by roots was also inferred up to 18 m.
-[@Davidson:2011] |
-| 6 | fates_smpsc | Soil water potential at full stomatal closure | 113000 - 242000 mm | Based on observed range of -1.13 to -2.42 MPa for leaf turgor loss point for individual tree variation across 49 BCI tree species.
-This study.|
-| 7 | aveDTB | Distance to bedrock | 3 - 18 m | Ben Turner; pers. comm. | 
-| 8 | fmax | The maximum fractional saturated area |  unitless | Empirical | 
-| 9 | HKSAT_ADJ | Adjusting factor for soil hydraulic conductivity | 1 - 8 unitless | To account for high macroporosity and direct flow paths in tropical soils that is not accounted for by small soil core samples [@Broedel:2017; @Kinner:2004; @Tomasella:1998] |
-| 10 | HKSAT | Soil hydraulic conductivity profile | 0.007 - 0.014 mm/s for =< 12.5 cm; 5.56e-05, 0.0003 for >= 60 cm | For Conrad catchment observed median Ksat (95% CI) for 12.5 cm depth is 38.3 mm/hr (25.4 - 51.2, n = 75), while for 60 cm depth 0.7 mm/hr (0.2 - 1.2, n = 40) [@Kinner:2004].
-For reference, a storm with 12.5 mm/hr rainfall intensity has a 0.2 probability of occuring in any given rainfall event.
-|
-| 11 | fpi_max | Maximum interception fraction of precipitation | 0.05 - 0.44 unitless | Based on throughfall data from [@Zimmermann:2010kzl] and precipitation data for BCI from STRI Physical Monitoring program, defined for precipitation events greater than 10 mm.
-|
-
 ## Supplementary Information
@@
 ```
+
 ```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts} Extreme soil droughts by census interval (panels) at three representative depths: 0.5, 1 and 1.7 m, depicted in different colors.
@@
 
-#### Table S1: Scientific names and codes for tree species for which Effective Rooting Depths were estimated. {#erd-sp}
-
-```{r, results = 'asis', echo = FALSE}
-writeLines("td, th { padding : 10px } th { background-color : brown ; color : white; border : 1px solid white; } td { color : brown ; border : 1px solid brown }", con = "mystyle.css")
-knitr::kable(erd.sp.names, format = "html")
+```{r symbols, results = 'asis', echo = FALSE}
+knitr::kable(symbols.table, caption = 'The symbol, definition of key hydraulic traits or variables used', booktabs = TRUE, escape = FALSE)
 ```
 
+```{r param, results = 'asis', echo = FALSE}
+
+knitr::kable(param.table, caption = 'ELM-FATES parameters used to generate ensembles, with description, chosen ranges, references and rationale.
+No.
+1-6 are parameters specific to FATES.', booktabs = TRUE, escape = FALSE)
+```
+
+```{r spcodes, results = 'asis', echo = FALSE}
+knitr::kable(erd.sp.names, caption = 'Scientific names and codes for tree species for which Effective Rooting Depths were estimated.', booktabs = TRUE
+)
+```
 # References

### 2020-07-29 manuscript.rmd
Changed hydraulic traits organ label from upper to lower case

@@
 - Deep water access is believed to be an important, drought-tolerance mechanism, but under-studied given the lack of data.
-Vulnerability to embolism ($\Psi_{88,~Stem}$) and safety margins ($\Psi_{min} - \Psi_{88,~Stem}$) are shown to be predictors of drought-induced mortality risk, but deep water access may delay plants from reaching these critical thresholds.
+Vulnerability to embolism ($\Psi_{88,~stem}$) and safety margins ($\Psi_{min} - \Psi_{88,~stem}$) are shown to be predictors of drought-induced mortality risk, but deep water access may delay plants from reaching these critical thresholds.
 Here we test the role of deep-water access in drought-tolerance strategies of a hyper-diverse tropical forest community (Barro Colorado Island, Panama) using a novel data-model approach.
-- We inversely modelled Effective Rooting Depth (ERD) for large trees of `r length(erd.sp)` canopy tree species by linking species growth dynamics (6 censuses, 1990-2015) to VPD, Soil Water Potential ($\Psi_{soil}$) by depth in the whole-soil column, and a new community-scale model of species vulnerability curves ($K_{Leaf}$ vs.
-$\Psi_{Leaf}$) that scaled new data on 21 species with Leaf Mass Area and Wood Specific Gravity.
+- We inversely modelled Effective Rooting Depth (ERD) for large trees of `r length(erd.sp)` canopy tree species by linking species growth dynamics (6 censuses, 1990-2015) to VPD, Soil Water Potential ($\Psi_{soil}$) by depth in the whole-soil column, and a new community-scale model of species vulnerability curves ($K_{leaf}$ vs.
+$\Psi_{leaf}$) that scaled new data on 21 species with Leaf Mass Area and Wood Specific Gravity.
 We evaluated ERD by leveraging existing isotopic tracer data ($\delta^2H$) on tree xylem sap and soil column.
@@
 These efforts have resulted in identification of above-ground traits such as proxies of hydraulic failure as indicators of mortality risk.
-A recent meta-analysis found greater drought-induced mortality rates in species with cavitation-vulnerable xylem ($\Psi_{88,~Stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~Stem}$) [@Anderegg:2016], but counter examples also exist [@Hoffmann:2011; @Paddock:2013]; indicating whether plants reach critical thresholds of hydraulic failure depends on their realised drought intensity (see Table \@ref(tab:symbols) for trait symbols and their definitions).
-A rare study that accounted for realised water availability for four tree species in a centennial drought (Texas, USA) found that mortality decreased with deep rootedness even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi_{88,~Stem}$) and had narrower safety-margins ($\Psi_{min} - \Psi_{88,~Stem}$) [@Johnson:2018fad].
-A species sourcing deep water in a Karst system (Italy) showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi_{88,~Stem}$ and narrower safety margin compared to two species sourcing shallow-water [@Nardini:2015].
+A recent meta-analysis found greater drought-induced mortality rates in species with cavitation-vulnerable xylem ($\Psi_{88,~stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~stem}$) [@Anderegg:2016], but counter examples also exist [@Hoffmann:2011; @Paddock:2013]; indicating whether plants reach critical thresholds of hydraulic failure depends on their realised drought intensity (see Table \@ref(tab:symbols) for trait symbols and their definitions).
+A rare study that accounted for realised water availability for four tree species in a centennial drought (Texas, USA) found that mortality decreased with deep rootedness even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi_{88,~stem}$) and had narrower safety-margins ($\Psi_{min} - \Psi_{88,~stem}$) [@Johnson:2018fad].
+A species sourcing deep water in a Karst system (Italy) showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi_{88,~stem}$ and narrower safety margin compared to two species sourcing shallow-water [@Nardini:2015].
 @Chitra-Tarak:2018 proposed the first community-scale model that linked trees' realised water availability to demographic outcomes through droughts.
@@
 Developing a novel inverse model allowed us to estimate species effective rooting depths (ERD), defined as the depth ($z$) whose soil water dynamics ($\Psi_{x,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015).
-This model incorporated impact of atmospheric drought (VPD) on growth and was constrined by above-ground hydraulic traits, namely, species leaf vulnerability curves ($K_{Leaf}$ vs.
-$\Psi_{x,~Leaf}$).
+This model incorporated impact of atmospheric drought (VPD) on growth and was constrined by above-ground hydraulic traits, namely, species leaf vulnerability curves ($K_{leaf}$ vs.
+$\Psi_{leaf}$).
 We collected new data on vulnerability curves for xx species.
@@
 We test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns species on a fast-slow trade-off axis.
-Specific hypotheses applicable to this forest included: (i) dynamics of soil water potential at depth does not cross critical plant thresholds even during multi-annual droughts because of high total annual precipitation to evapotranspiration ratio (not sure if this should be stated as a hypothesis or just an explanation of the results); (ii) trees with deeper ERD show 'fast' hydraulic traits: higher maximum stem hydraulic conductance (Kmax), but greater xylem vulnerability to embolism ($\Psi_{88,~Stem}$), higher $\Psi_{tlp}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~Stem}$); and finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure.
+Specific hypotheses applicable to this forest included: (i) dynamics of soil water potential at depth does not cross critical plant thresholds even during multi-annual droughts because of high total annual precipitation to evapotranspiration ratio (not sure if this should be stated as a hypothesis or just an explanation of the results); (ii) trees with deeper ERD show 'fast' hydraulic traits: higher maximum stem hydraulic conductance (Kmax), but greater xylem vulnerability to embolism ($\Psi_{88,~stem}$), higher $\Psi_{tlp}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$); and finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure.
 To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and mortality outcomes for a species-rich forest.
@@
 \
-We measured ($K_{x,~Leaf}$ vs.
-$\Psi_{x,~Leaf}$) on adult trees of `r length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))` common species (see SI x).
+We measured ($K_{leaf}$ vs.
+$\Psi_{leaf}$) on adult trees of `r length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))` common species (see SI x).
 We fitted exponential, sigmoidal, logistic equations to species-level data as prescribed by @Guyot:2011.
@@
 \begin{equation}
-K_{Leaf} = Ae^{-B\Psi_{Leaf}} + error
+K_{leaf} = Ae^{-B\Psi_{leaf}} + error
 (\#eq:vc-exp)
@@
 
-```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-plc-curves} Modelled standardized 'Vulnerability curves' ($K_{x,~Leaf}$ vs.
-$\\Psi_{x,~Leaf}$) at community-scale (", length(data.model.AB$sp)," species) based on data on $WSG$ & $LMA$ and their relationships with vulnerability curve parameters measured on ", sp.soft.filled$AB.data.model," species.
+```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-plc-curves} Modelled standardized 'Vulnerability curves' ($K_{leaf}$ vs.
+$\\Psi_{leaf}$) at community-scale (", length(data.model.AB$sp)," species) based on data on $WSG$ & $LMA$ and their relationships with vulnerability curve parameters measured on ", sp.soft.filled$AB.data.model," species.
 Each line represents a species, color coded by $WSG$.")}
@@
 \begin{equation}
-K_{Leaf,z,n} = Ae^{-B\Psi_{z,n}}
+K_{leaf,z,n} = Ae^{-B\Psi_{z,n}}
 (\#eq:k-soil)
@@
 \begin{equation}
-Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{Leaf,z,n} + error
+Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,z,n} + error
 (\#eq:M1)
@@
 \begin{equation}
-Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{Leaf,n} StdVPD_{n} + error
+Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,n} StdVPD_{n} + error
 (\#eq:M2)
@@
 \begin{equation}
-Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{Leaf,z,n} LeafFrac_{doy} + error
+Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,z,n} LeafFrac_{doy} + error
 (\#eq:M3)
@@
 \begin{equation}
-Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{Leaf,z,n} LeafFrac_{doy}StdVPD_{n} + error
+Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,z,n} LeafFrac_{doy}StdVPD_{n} + error
 (\#eq:M4)
@@
 
-To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD with maximum stem hydraulic conductance ($K_{max, Stem}$), leaf Turgor Loss Point ($\Psi_{tlp}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~Stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~Stem}$).
+To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD with maximum stem hydraulic conductance ($K_{max, Stem}$), leaf Turgor Loss Point ($\Psi_{tlp}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$).
 
@@
 To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$).
-To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{x, ERD}$) were below species $\Psi_{crit}$, defined as $\Psi_{80,~Leaf}$.
+To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{x, ERD}$) were below species $\Psi_{crit}$, defined as $\Psi_{80,~leaf}$.
 
@@
 In this forest modelled Effective Rooting Depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from 0.5 m to 13 m (Figure \@ref(fig:sp-erd)).
-For the subset of species for which hydraulic properties were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max, Stem}$; *R*^2^ = 0.39, *P* = 0.098, *n* = 8; Figure \@ref(fig:erd-hyd)A), less negative leaf Turgor Loss Point ($\Psi_{tlp}$); *R*^2^ = 0.6, p = 0.024, *n* = 8; Figure \@ref(fig:erd-hyd)B), greater vulnerability to embolism from cavitation with less negative ($\Psi_{88,~Stem}$) (*R*^2^ = 0.72, p = 0.007, *n* = 8; Figure \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~Stem}$; *R*^2^ = 0.74, p = 0.013, *n* = 8; Figure \@ref(fig:erd-hyd)D).
+For the subset of species for which hydraulic properties were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max, Stem}$; *R*^2^ = 0.39, *P* = 0.098, *n* = 8; Figure \@ref(fig:erd-hyd)A), less negative leaf Turgor Loss Point ($\Psi_{tlp}$); *R*^2^ = 0.6, p = 0.024, *n* = 8; Figure \@ref(fig:erd-hyd)B), greater vulnerability to embolism from cavitation with less negative ($\Psi_{88,~stem}$) (*R*^2^ = 0.72, p = 0.007, *n* = 8; Figure \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$; *R*^2^ = 0.74, p = 0.013, *n* = 8; Figure \@ref(fig:erd-hyd)D).
 
@@
 
-```{r erd-time-new, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{x, z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~Leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for ", length(erd.mrate.sp), " canopy species found on Barro Colorado Island.
+```{r erd-time-new, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{x, z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for ", length(erd.mrate.sp), " canopy species found on Barro Colorado Island.
 Each circle represents a species.
@@
 Low $\Psi_{50}$ is in turn associated with low $\Psi_{tlp}$ and both are broadly associated with low water availability[@Choat:2012; @Bartlett:2012].
-Slow strategy is associated with larger safety-margin ($\Psi_{min} - \Psi_{88,~Stem}$) in two angiosperm systems.
+Slow strategy is associated with larger safety-margin ($\Psi_{min} - \Psi_{88,~stem}$) in two angiosperm systems.
 High Kmax is associated with photosynthetic rates with faster growth potential.
@@
 
-```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves} Leaf 'vulnerability curves', that is standardized *K*~Leaf~ vs.
-$\\Psi$~Leaf~, fitted to observed data on ",  length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))," tree species from Panama, color coded by Wood Specific Gravity ($gcm^{-3}$).")}
+```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves} Leaf 'vulnerability curves', that is standardized $K_{leaf}$ vs.
+$\\Psi_{leaf}$, fitted to observed data on ",  length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))," tree species from Panama, color coded by Wood Specific Gravity ($gcm^{-3}$).")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB_for_data_sp.jpeg"))
@@
 
-```{r ab-fit, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ab-fit}Parameters $A$ (panel A) & $B$ (panel B) of leaf vulnerability curves that were fitted to observed data on *K*~Leaf~ vs.
-$\\Psi$~Leaf~ for ", sp.soft.filled$AB.data.model," species vs.
+```{r ab-fit, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ab-fit}Parameters $A$ (panel A) & $B$ (panel B) of leaf vulnerability curves that were fitted to observed data on $K_{leaf}$ vs.
+$\\Psi_{leaf}$ for ", sp.soft.filled$AB.data.model," species vs.
 those that were predicted from models based on species soft-traits, WSG and LMA.")}

### 2020-07-31 manuscript.rmd
Edited symbols

@@
 For water resource observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified a fast-slow spectrum, although examples of all trade-offs observed simultaneously are rare, and belowground trait or environment linkages are weak.
-'Fast' species at the productive end of the spectrum have traits with high capacity to acquire, move, store and use water (high Kmax, low wood density) and achieve high growth potential, but build flimsy, disposable tissues (high SLA) and are less tolerant of low resources, whereas species with slow traits are better protected from C losses and are more drought tolerant (low $\Psi_{50}$, low $\Psi_{tlp}$, larger safety margin).
+'Fast' species at the productive end of the spectrum have traits with high capacity to acquire, move, store and use water (high $K_{max}$, low wood density) and achieve high growth potential, but build flimsy, disposable tissues (high SLA) and are less tolerant of low resources, whereas species with slow traits are better protected from C losses and are more drought tolerant (low $\Psi_{50}$, low $\Psi_{tlp}$, larger safety margin).
 Fast strategy is associated with spatially wet sites such as valleys or high rainfall [@Kursar:2009; @Oliveira:2019], but whether and how the fast-slow spectrum holds true for the moisture gradient in the soil column is not clear.
@@
 However, this may depend on the exact dynamics of water-availability by depth at a site, depending on precipitation regime and hydrology.
-In pulse-driven arid systems *shallow*-rooted plants are associated with fast traits (high Kmax) that maximise the use of the water-pulse, while *deep*-rooted plants have slow traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013].
+In pulse-driven arid systems *shallow*-rooted plants are associated with fast traits (high $K_{max}$) that maximise the use of the water-pulse, while *deep*-rooted plants have slow traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013].
 
@@
 Here, for large trees of xx canopy species of a hyperdiverse neo-tropical moist forest (BCI, Panama) we make a thorough assessment of trade-offs between belowground water environment, above-ground hydraulic traits, estimates of effective rooting depths and demographic outcomes over 35 years that withnessed droughts of various intensity and duration.
-We obtained daily dynamics of plant water potential ($\Psi_{x,~z}$) by depth $z$ in the whole-soil column with a locally parametarised and calibrated 1-D hydrological water balance model.
-Developing a novel inverse model allowed us to estimate species effective rooting depths (ERD), defined as the depth ($z$) whose soil water dynamics ($\Psi_{x,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015).
-This model incorporated impact of atmospheric drought (VPD) on growth and was constrined by above-ground hydraulic traits, namely, species leaf vulnerability curves ($K_{leaf}$ vs.
+We obtained daily dynamics of plant water potential ($\Psi_{soil,~z}$) by depth $z$ in the whole-soil column with a locally parametarised and calibrated 1-D hydrological water balance model.
+Developing a novel inverse model allowed us to estimate species effective rooting depths (ERD), defined as the depth ($z$) whose soil water dynamics ($\Psi_{~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015).
+This model incorporated impact of atmospheric drought (VPD) on growth and was constrained by above-ground hydraulic traits, namely, species leaf vulnerability curves ($K_{leaf}$ vs.
 $\Psi_{leaf}$).
@@
 We test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns species on a fast-slow trade-off axis.
-Specific hypotheses applicable to this forest included: (i) dynamics of soil water potential at depth does not cross critical plant thresholds even during multi-annual droughts because of high total annual precipitation to evapotranspiration ratio (not sure if this should be stated as a hypothesis or just an explanation of the results); (ii) trees with deeper ERD show 'fast' hydraulic traits: higher maximum stem hydraulic conductance (Kmax), but greater xylem vulnerability to embolism ($\Psi_{88,~stem}$), higher $\Psi_{tlp}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$); and finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure.
+Specifically, (i) dynamics of soil water potential at depth does not cross critical plant thresholds even during multi-annual droughts because of high total annual precipitation to evapotranspiration ratio (not sure if this should be stated as a hypothesis or just an explanation of the results); (ii) trees with deeper ERD show 'fast' hydraulic traits: higher maximum stem hydraulic conductance ($K_{max}$), but greater xylem vulnerability to embolism ($\Psi_{88,~stem}$), higher $\Psi_{tlp}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$); and finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure.
 To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and mortality outcomes for a species-rich forest.
@@
 
-where $\Psi_e$ is the air-entry matric potential (200, mmH~2~O), $\theta_{s}$ is the saturated volumetric water content (0.51, $cm^{3}cm^{-3}$), and $b$, an index for soil pore-size distribution (10, unitless), respectively.
+where $\Psi_e$ is the air-entry matric potential (200, mmH~2~O), $\theta_{s}$ is the saturated volumetric water content (0.51, cm^3^cm^-3^), and $b$, an index for soil pore-size distribution (10, unitless), respectively.
 
@@
 \@ref(eq:water-balance), namely, (i) evapotranspiration (ET, $E_{v} + E_{g}$) from the eddy flux tower (monthly, 2012-2017), (ii) stream discharge for the Conrad Catchment ($q_{over} + q_{drai}$; daily, 2012-2018) and (iii) volumetric water content (VWC; $W_{liq, z}$; Fig.
-\@ref(fig:hydro-calib; daily, 2015-2018).
+\@ref(fig:hydro-calib); daily, 2015-2018).
 We collected VWC data (2016-2018) from a point location at the eddy flux tower with Time Domain Reflectometry (TDR) using three horizontal probes at depths 10, 40 and 100 cm and one vertical probe (0-15 cm).
 TDR data for the vertical probe was calibrated based on measurements of gravimetric soil moisture on several occasions (Fig.
-\@ref(fig:TDR-calib)) and mean bulk dry density of 0.8 gcm{-3} (*n* = 5, cylinder volume = 1250 cm{3}).
+\@ref(fig:TDR-calib)) and mean bulk dry density of 0.8 gcm^-3^ (*n* = 5, cylinder volume = 1250 cm^3^).
 The same calibration equation was used for other probes.
 We also leveraged existing opportunistic dry season VWC measurements for the entire 50-ha plot (15, 40 and 100 cm in 2015 and 15 & 40 cm in 2016; @Kupers:2019).
-Among the 5000 parameter ensembles we chose 100 ensembles that best-fit all three fluxes simultaneously (add details...that is those that maximised the sum of regression coefficients between observed and simulated variables.) See Table \ref(tab:param) for best-fit parameter ranges.
+Among the 5000 parameter ensembles we chose 100 ensembles that best-fit all three fluxes simultaneously (add details...that is those that maximised the sum of regression coefficients between observed and simulated variables.) See Table \@ref(tab:param) for best-fit parameter ranges.
 We ran ELM-FATES with the chosen 100 ensembles from 1985-2018, with first 5 years used for model spin-up.
@@
 We identified scaling relationships between fitted parameters $A$ and $B$ in Eq.
-\@ref(eq:vc-exp) and two traits that were available at community-scale, namely, $WSG$ and $LMA$, defined as mean wood specific gravity after drying at 100$^\circ$C (g cm^{-3}^) and mean leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules (g m^{-2}^) for leaves receiving direct sunlight, respectively.
+\@ref(eq:vc-exp) and two traits that were available at community-scale, namely, $WSG$ and $LMA$, defined as mean wood specific gravity after drying at 100$^\circ$C (gcm^-3^) and mean leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules (gm^-2^) for leaves receiving direct sunlight, respectively.
 Fitted polynomial equations are described as,
@@
 \begin{equation}
-K_{leaf,z,n} = Ae^{-B\Psi_{z,n}}
+K_{leaf,z,n} = Ae^{-B\Psi_{soil,~z,n}}
 (\#eq:k-soil)
@@
 We chose only those trees for which diameter (dbh) record was present across all size censuses (1990, 1995, 2000, 2005, 2010, 2015).
-We calculated individual tree ($i$) growth rates ($Gr_{i,t}$ (cm.yr^{-1})) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses, divided by the inter-census interval based on the exact date of measurements.
+We calculated individual tree ($i$) growth rates ($Gr_{i,t}$ (cm.yr^-1^)) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses, divided by the inter-census interval based on the exact date of measurements.
 As tree growth rates increase with size, to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census onterval ($Resid_{i,t} = Gr_{i,t} - f(DBH_{i, t})$).
@@
 
-For each model \@ref(eq:M1)- \@ref(eq:M4) and species combination ($sp$), we obtained goodness-of-fit (*R*^2^) and Pearson's $\rho$ for all soil depths $z$, and for each sp, model combination we chose the $z$ with highest *R*^2^ as the Effective Rooting Depth (ERD); after retaining species with positive $\rho$ and *R*^2^ >= 0.1.
+For each model \@ref(eq:M1)- \@ref(eq:M4) and species combination ($sp$), we obtained goodness-of-fit (*R*^2^) and Pearson's $\rho$ for all soil depths $z$, and for each sp, model combination we chose the $z$ with highest *R*^2^ as the Effective Rooting Depth (ERD); after retaining species with positive Pearson's $\rho$ and *R*^2^ >= 0.1.
 
@@
 The model with VPD and without Leaf Fraction (Eq.
-\@ref(eq:M2)) had highest goodness-of fit (*R*^2^ = 0.51; *p* = 0.072), followed by followed by model without both VPD and and Leaf Fraction (\@ref(eq:M1); Fig.
+\@ref(eq:M2)) had highest goodness-of fit (*R*^2^ = 0.51; *p* = 0.072), followed by followed by model without both VPD and and Leaf Fraction (Eq.
+\@ref(eq:M1); Fig.
 S\@ref(fig:other-erd-mods)).
@@
 Data from [@Meinzer:1999].
-See [Table S1]{erd-sp} for scientific names corresponding to species codes.")}
+See Table  \\@ref(tab:spcodes) for scientific names corresponding to species codes.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_phenology_Meinzer_gr.Psi.VPD.jpeg"))
@@
 
-To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD with maximum stem hydraulic conductance ($K_{max, Stem}$), leaf Turgor Loss Point ($\Psi_{tlp}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$).
+To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD with maximum stem hydraulic conductance ($K_{max, stem}$), leaf turgor loss point ($\Psi_{tlp}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$).
 
@@
 To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$).
-To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{x, ERD}$) were below species $\Psi_{crit}$, defined as $\Psi_{80,~leaf}$.
+To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{ERD}$) were below species $\Psi_{crit}$, defined as $\Psi_{80,~leaf}$.
 
@@
 In this forest modelled Effective Rooting Depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from 0.5 m to 13 m (Figure \@ref(fig:sp-erd)).
-For the subset of species for which hydraulic properties were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max, Stem}$; *R*^2^ = 0.39, *P* = 0.098, *n* = 8; Figure \@ref(fig:erd-hyd)A), less negative leaf Turgor Loss Point ($\Psi_{tlp}$); *R*^2^ = 0.6, p = 0.024, *n* = 8; Figure \@ref(fig:erd-hyd)B), greater vulnerability to embolism from cavitation with less negative ($\Psi_{88,~stem}$) (*R*^2^ = 0.72, p = 0.007, *n* = 8; Figure \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$; *R*^2^ = 0.74, p = 0.013, *n* = 8; Figure \@ref(fig:erd-hyd)D).
+For the subset of species for which hydraulic properties were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max,~stem}$; *R*^2^ = 0.39, *P* = 0.098, *n* = 8; Figure \@ref(fig:erd-hyd)A), less negative leaf  ($\Psi_{tlp}$); *R*^2^ = 0.6, p = 0.024, *n* = 8; Figure \@ref(fig:erd-hyd)B), greater vulnerability to embolism from cavitation with less negative ($\Psi_{88,~stem}$) (*R*^2^ = 0.72, p = 0.007, *n* = 8; Figure \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$; *R*^2^ = 0.74, p = 0.013, *n* = 8; Figure \@ref(fig:erd-hyd)D).
 
 ```{r sp-erd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap= paste0("\\label{lab-sp-erd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for ", length(erd.sp)," large, canopy species of Barro Colorado Island.
-See [Table S1]{erd-sp} for scientific names corresponding to species codes.")}
+See Table  \\@ref(tab:spcodes) for scientific names corresponding to species codes.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/ERD_by_sp_large_canopy.jpeg"))
@@
 ```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus hydraulic propertiese for eight large, canopy species found on Barro Colorado Island (Panama).
-*R*^2^ and *P*-values are given in panel insets.[X-Axis labels to be added: Kmax units as in Fig.
+*R*^2^ and *P*-values are given in panel insets.[X-Axis labels to be added: $K_{max}$ units as in Fig.
 2.
-Water Potentials are in Mpa.]")}
+Water Potentials are in MPa.]")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/Select_traits_depth.jpeg"))
@@
 For five out of seven census intervals, species with shallower $ERD$ had significantly higher mortality anomalies than deeper ERD species; Figure (\@ref(fig:erd-mort-anom)).
-Analysis of time spent under catastrophic soil drought showed that the proportion of days species $\Psi_{x,~ERD}$ was more negative than species $\Psi_{crit}$ decreased with species ERD (Figure \@ref(fig:erd-time)).
-Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25 year period for which soil water potential $\Psi_{x,~z}$ was modelled (1990-2015), as soil depths at and below 1.7 m remained above `r round(psi.1.7.min$median, 2)` [`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`] (median [95% CI]) through all the dryseason and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).
+Analysis of time spent under catastrophic soil drought showed that the proportion of days species $\Psi_{ERD}$ was more negative than species $\Psi_{crit}$ decreased with species ERD (Figure \@ref(fig:erd-time)).
+Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25 year period for which soil water potential $\Psi_{soil,~z}$ was modelled (1990-2015), as soil depths at and below 1.7 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dryseason and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).
 
-```{r erd-mort-changed, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-mort}Mortality rate (mean $\\pm$ 1SE) versus Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for ", length(erd.mrate.sp)," canopy species found on Barro Colorado Island.
+```{r erd-mort, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-mort}Mortality rate (mean $\\pm$ 1SE) versus Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for ", length(erd.mrate.sp)," canopy species found on Barro Colorado Island.
 *R*^2^ and *P*-values are given in panel insets.")}
@@
 
-```{r erd-time-new, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{x, z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for ", length(erd.mrate.sp), " canopy species found on Barro Colorado Island.
+```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{soil,~z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for ", length(erd.mrate.sp), " canopy species found on Barro Colorado Island.
 Each circle represents a species.
@@
 Analysis of time spent below ($\Psi_{crit}$) in the soil zone of ERD showed that this was likely due to low realised risk of hydraulic failure for deeper ERD species.
-We thus show, for the first time, that interspecific variation in trees' effective rooting depths aligns species on a 'fast-slow' strategy axis by demonstrating a link between trees' realised drought experience, above-ground hydraulic architecture and long-term mortality outcomes.
-On the aquisitive end were species with hydraulically efficient, but risky suit of traits optimised for an environment of reliable, abundant source of water at depth, which actualised to a lower associated mortality-risk, whilst at the slow end were species with inefficient, but safer xylem, associated with a water environment of high variability.
+We thus show, for the first time, that interspecific variation in trees' effective rooting depths aligns species on a 'fast-slow' water economic spectrum by demonstrating a link between trees' realised drought experience, hydraulic architecture and long-term mortality outcomes.
+On the 'fast' end of the spectrum were species with hydraulically efficient, but risky suit of hydraulic traits.
+These were associated with, and likely optimised for, an environment of reliable and abundant source of water at depth, which actualised to a lower associated mortality-risk.
+At the slow end were species with inefficient, but safer xylem, associated with a shallow soil- water environment of high variability.
 
@@
 
-fast-slow axis is matched by presence of an efficiency-safety trade-off between high water conductance (high Kmax) and cavitation-resistant xylem (low $\Psi_{50}$), although community-wide examples are rare [@Hacke:2006;@Gleason:2016;].
+fast-slow axis is matched by presence of an efficiency-safety trade-off between high water conductance (high $K_{max}$) and cavitation-resistant xylem (low $\Psi_{50}$), although community-wide examples are rare [@Hacke:2006;@Gleason:2016;].
 Low $\Psi_{50}$ is in turn associated with low $\Psi_{tlp}$ and both are broadly associated with low water availability[@Choat:2012; @Bartlett:2012].
 Slow strategy is associated with larger safety-margin ($\Psi_{min} - \Psi_{88,~stem}$) in two angiosperm systems.
-High Kmax is associated with photosynthetic rates with faster growth potential.
+High $K_{max}$ is associated with photosynthetic rates with faster growth potential.
 
@@
 
-```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-other-erd-mods}All ERD models tested.
-See Eqns xxxx"}
+```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-other-erd-mods}Effective Rooting Depth versus $\\delta^2^H$ for all the models tested.
+See Eqs.
+\\@ref(eq:M1)-\\@ref(eq:M4).")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_sp_color_Meinzer.jpeg"))
@@
 ```
+
 # References

### 2020-08-01 manuscript.rmd
Edited first Discussion para and added two

@@
 \
-We test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns species on a fast-slow trade-off axis.
-Specifically, (i) dynamics of soil water potential at depth does not cross critical plant thresholds even during multi-annual droughts because of high total annual precipitation to evapotranspiration ratio (not sure if this should be stated as a hypothesis or just an explanation of the results); (ii) trees with deeper ERD show 'fast' hydraulic traits: higher maximum stem hydraulic conductance ($K_{max}$), but greater xylem vulnerability to embolism ($\Psi_{88,~stem}$), higher $\Psi_{tlp}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$); and finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure.
-To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and mortality outcomes for a species-rich forest.
+We test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns species on a fast-slow economic spectrum optimised for increasingly reliable water availability by depth such that (i) trees with deeper ERD have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem: higher maximum stem hydraulic conductance ($K_{max}$), but greater xylem vulnerability to embolism ($\Psi_{88,~stem}$), higher $\Psi_{tlp}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$), whereas shallower ERD species are associated with less efficient but safer hydraulic traits; and that, (ii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure, whereas shallower ERD species may have higher mortality rates as a result of higher variability in shallow soil water availability.
+To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.
 
@@
 $\Psi_{soil}$ for depths 15, 40 and 100 cm [@Kupers:2019].
-A Campbell (1974) empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences among depths were nominal:
+A Campbell @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences among depths were nominal:
 
@@
 # Discussion
+\
+Our analyses revealed that trees with deeper effective rooting depths (ERD) were associated higher hydraulic efficiency, but not safety: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species.
+Yet, deeper ERD species showed lower mortality rates than shallower ERD species over the 33 year record (1982-2015).
+Analysis of time spent below ($\Psi_{crit}$) in the soil zone matching ERD $\Psi_{soil,~z = ERD}$ showed that this was likely due to decreasing realised risk of hydraulic failure with depth.
+We thus show, for the first time, that interspecific variation in trees' effective rooting depths aligns species on a 'fast-slow' water economic spectrum by demonstrating a link between trees' realised drought experience, hydraulic architecture and long-term mortality outcomes.
+On the 'fast' end of the spectrum were deep ERD species with hydraulically efficient, but risky suit of traits optimised for an environment of reliable and abundant source of water at depth that actualised to a lower mortality-risk.
+At the slow end of the spectrum were shallow ERD species with inefficient, but safer xylem, likely optimised for high variability in $\Psi_{soil}$ in the shallow soil layers, resulting in higher variance in mortality rates across censuses as $\Psi_{soil,~z = ERD}$ crossed ERD $\Psi_{crit}$.
+Mortality rates of shallow ERD species may have been exacerbated by increased drought events during the study period.
+Previous work shows that half of the extreme dry seasons, estimated as < 1 SD annual water deficit, during 1931-2015 (84 years) occurred during the first 23 years (1982-2005) of the study period.
+[@Condit.2017]
 
-Our analyses revealed that trees with deeper effective rooting depths (ERD) showed acquisitive hydraulic architecture, with higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species.
-Yet, deeper ERD species showed lower mortality rates over the 33 year record (1982-2015).
-Analysis of time spent below ($\Psi_{crit}$) in the soil zone of ERD showed that this was likely due to low realised risk of hydraulic failure for deeper ERD species.
-We thus show, for the first time, that interspecific variation in trees' effective rooting depths aligns species on a 'fast-slow' water economic spectrum by demonstrating a link between trees' realised drought experience, hydraulic architecture and long-term mortality outcomes.
-On the 'fast' end of the spectrum were species with hydraulically efficient, but risky suit of hydraulic traits.
-These were associated with, and likely optimised for, an environment of reliable and abundant source of water at depth, which actualised to a lower associated mortality-risk.
-At the slow end were species with inefficient, but safer xylem, associated with a shallow soil- water environment of high variability.
+### Effective rooting depths at tree community level
+\
+By estimating ERD for large trees of 33 canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species species-rich tropical forests.
+We draw attention to the availability of data on $\delta^2^H$ on trees and water sources belowground--soil or substrate matrix water in the unsaturated vadoze zone or free water below the water table--as important for model calibration/validation.
+The first inverse model of ERD (water uptake depth index) developed by Chitra-Tarak et al @Chitra-Tarak:2018 lacked validation as existing methods were inadequate for extracting water bound to the soil and saprolite found in their dry forest study site (Mudumalai National Park, Southern India; Jean Riotte, pers.
+comm.).
+Our ERD model was distinctly constrained on species leaf vulnerability (PLC) curves.
+We benefitted from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (SLA~lamina~ and Wood Specific Gravity) and parameters of PLC curves for which we collected data on a few species (*n* = 14), a 'hard' trait.
+This allowed us to scale-up PLC curves at the community scale (*n* = 284 species).
+Whether such scaling relationships between hard and soft traits hold true for other forests and their wider applicability remains to be seen as data on PLC curves in species-rich, tropical forests are limited, but counter examples exist [see xx, yy, zz (Lawren, could you suggest references?].
 
-## Higher xylem embolism vulnerability of species with Deeper Effective Rooting Depths
+\
+##### Future Directions for Model Development
+\
+Our model estimated ERD as a single value, unlike the dynamic ERD modelled by Chitra-Tarak et al @Chitra-Tarak:2018, but future model developments should investigate seasonal dynamics of ERD, calibrated on corresponding calibration data on isotopic water sourcing depths.
+Our inverse ERD model was parametarised on 5-yearly growth data, with five data points over a 25 year period, which provided large degree of freedom to fit the model.
+Future studies could better constrain the model with higher resolution growth data such as those from dendrometers.
+This is especially important in modeling seasonally dynamic ERD.
+However, trees' stem diameter change is a sum total of not just growth, the addition of wood, but also $\pm$ changes due to hydrostatic stem flexing, whose importance increases with temporal resolution [@Chitra-Tarak.2016; @Chitra-Tarak.2014].
+Use of dendrometer data would thus entail accounting for stem hydrostatic flexing [@Zweifel:2018].
+In light-limited forests such as BCI, expanding ERD estimates to smaller trees, especially via parametairising over a few growth data points, would likely need model development to account for variation in trees' light environment over time and among trees' of the same size class, an issue we minimised by limiting ERD estimates to canopy trees.
+To estimate topographic variation in ERD within or across species a distributed hydrological model could be used.
+We note that data availability on soil water retention curve ($\Psi_{soil}$vs.
+VWC), hydraulic conductivity ($K_{soil}$) by depth and soil moisture by depth were important for effective calibration of our 1-D hydrological model, and recommend wide-spread data collection of these variables in forested sites as well as water-table levels (not collected in our site) to estimate trees' water environments.
+Finally, developing an inverse ERD model that accounts for movement of water within the soil-plant-atmosphere-continuum (SPAC) would be a step change from the current statistical models [this study; @Chitra-Tarak:2018].
+Although data needs for parametarising such models [@Christoffersen.2016;@Sperry.1998] could be greater (e.g.
+retention curves for roots), limiting community-wide application, they would improve our understanding of plant physiology [@Johnson:2017].
 
--- trade-offs among differential investment in hydraulic efficiency and safety
-here links with trade-off in water-stress vertically belowground
--- unexplored (hydraulic capacitance, stomatal control, photosynthetic capacity, high temperature tolerance)
+### Need for drought risk
 
-fast-slow axis is matched by presence of an efficiency-safety trade-off between high water conductance (high $K_{max}$) and cavitation-resistant xylem (low $\Psi_{50}$), although community-wide examples are rare [@Hacke:2006;@Gleason:2016;].
-Low $\Psi_{50}$ is in turn associated with low $\Psi_{tlp}$ and both are broadly associated with low water availability[@Choat:2012; @Bartlett:2012].
-Slow strategy is associated with larger safety-margin ($\Psi_{min} - \Psi_{88,~stem}$) in two angiosperm systems.
-High $K_{max}$ is associated with photosynthetic rates with faster growth potential.
+Plant traits hold clues for the historical environments that shaped their ecological strategy.
 
-## Lower mortality of species with deeper Effective Rooting Depths
-
-Mortality could be immediate or may result after decades of drought occurrence[@Cailleret:2016].
-
-## Limitations & Future directions
-
-Our modeling framework had a few limitations. 
-- The inverse model for ERD was parametarised on 5-yr growth rates, which vastly increased the degree of freedom to fit the model.
-Future studies should use high-frequency growth time series such as those obtained from dendrometer bands.
-To infer growth from such diameter time series, however, would especially require accounting for diameter changes due to hydrostatic flexing [@Chitra-Tarak:2015].
-
-- Future studies should investigate seasonality in water-sourcing depths investigated.
-
-- Although we attempted a coarse estimate of species leaf cover seasonality by combining leaf-fall data from litter traps and forest-level LAI data for leaf-on periods, more accuarate estimations of leaf-cover seasonality are required, such as those obtained from drone images [@Park.2019]
-
--To enable estimation of inverse model for the larger community across size classes would need accounting for changing light environments on growth; we minimised this issue by limiting our ERD inversion to large trees of species with max height >= 30m and thus whose canopy's are most likely to be fully exposed to light.
-
-- Future studies would benefit from modeling the complete hydrodynamics of water potential gradients over the soil - plant (roots, stems, leaves) - atmosphere continuum (SPAC) [@Johnson:2018fad; @Christoffersen:2016kd2d].
-However, to parameterise such a model at community-scale would need community-scale data on hard traits such as stem and root vulnerability curves; if not, model development would be needed to up-scale data from a few species, for example, via softer traits as we did for leaf vulnerability curves.
-(Note that In this forest LMA and WSG did not scale with stem vulnerability curves).
-Water storage (capacitance) should also be considered.
-
-- Deep ERD species may possibly be associated with locations of greater water availability rather than having deep rootedness.
-A 3-D hydrology model would provide spatially distributed estimates of water dynamics in the whole-soil/rock column and depth to water table to link with individual trees water experience [@Schwantes:2018].
-Forest inventory sites would benefit from investing data collection efforts on water availability dynamics by depth and water tables covering spatial heterogeneity so as to better constrain such models.
-
-- Variation in mortality explained by ERD is low. Other causes.... Based on a year-long study, lightening explains ~4.5 % of mortality for trees >=10 cm dbh and ~ 40% for trees >=60 cm dbh.
-
--Forest is realtively stable over 35 years.
-Future studies could investigate if ongoing change in the duration, frequency and intensity of droughts entails sufficient exhaustion of deep water availablity so as to cause mortality of deep ERD species.
-
-## Conclusions:
+## Conclusions
 

### 2020-08-02 manuscript.rmd
edited a discussion paragraph

@@
 
-```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-iso}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE; m) versus dry season stable isotopic concentration $\\delta$^2^H (mean $\\pm$ (add error); ‰) for ", length(unique(erd.iso$sp))," canopy species from Barro Colorado Island.
+```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-iso}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE; m) versus dry season stable isotopic concentration $\\delta^2H$ (mean $\\pm$ (add error); ‰) for ", length(unique(erd.iso$sp))," canopy species from Barro Colorado Island.
 Data from [@Meinzer:1999].
@@
 By estimating ERD for large trees of 33 canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species species-rich tropical forests.
-We draw attention to the availability of data on $\delta^2^H$ on trees and water sources belowground--soil or substrate matrix water in the unsaturated vadoze zone or free water below the water table--as important for model calibration/validation.
-The first inverse model of ERD (water uptake depth index) developed by Chitra-Tarak et al @Chitra-Tarak:2018 lacked validation as existing methods were inadequate for extracting water bound to the soil and saprolite found in their dry forest study site (Mudumalai National Park, Southern India; Jean Riotte, pers.
+We note the availability of data on $\delta^2^H$ on trees and water sources belowground--soil or substrate matrix water in the unsaturated vadoze zone or free water below the water table--as important for model calibration/validation.
+The first inverse model of ERD (water uptake depth index, *n* = 12 species) developed by Chitra-Tarak et al @Chitra-Tarak:2018 lacked validation as existing methods were inadequate to extract water bound to the soil and saprolite found in their dry forest study site (Mudumalai National Park, Southern India; Jean Riotte, pers.
 comm.).
@@
 Use of dendrometer data would thus entail accounting for stem hydrostatic flexing [@Zweifel:2018].
-In light-limited forests such as BCI, expanding ERD estimates to smaller trees, especially via parametairising over a few growth data points, would likely need model development to account for variation in trees' light environment over time and among trees' of the same size class, an issue we minimised by limiting ERD estimates to canopy trees.
-To estimate topographic variation in ERD within or across species a distributed hydrological model could be used.
+Two of our exploratory ERD models included seasonality in leaf area fraction (LeafFrac), but we did not select as the worsened the fit with growth data and ERD calibration data (Fig.
+S\@ref(fig:other-erd-mods)).
+We interpret this result as that VPD and PLC curves may have adequately explained growth, but also acknowledge that our estimate of seasonality of leaf fraction is a rough estimate based on leaf fall data and assumption of leaf presence during the wet season and we did not take into account inter-annual variation.
+Future studies could make use of data from new technologies such as drone based monitoring of species-specific leaf area fraction in species-rich forests [@Park.2019].
+In forests such as BCI where understorey tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes, especially from a few growth data points, may need accounting for variation in trees' light environment over time and among trees' of the same size class, an issue we minimised by limiting ERD estimates to canopy trees.
+To estimate topographic variation in ERD within or across species future studies may use a distributed hydrological model.
 We note that data availability on soil water retention curve ($\Psi_{soil}$vs.
-VWC), hydraulic conductivity ($K_{soil}$) by depth and soil moisture by depth were important for effective calibration of our 1-D hydrological model, and recommend wide-spread data collection of these variables in forested sites as well as water-table levels (not collected in our site) to estimate trees' water environments.
-Finally, developing an inverse ERD model that accounts for movement of water within the soil-plant-atmosphere-continuum (SPAC) would be a step change from the current statistical models [this study; @Chitra-Tarak:2018].
+VWC), hydraulic conductivity ($K_{soil}$) by depth and soil moisture by depth were important for effective calibration of our 1-D hydrological model, and recommend wide-spread data collection of these variables as well as water-table levels (not collected in our site) in forest-inventory sites to allow for estimation of trees' water environments.
+Finally, developing an inverse ERD model that accounts for movement of water within the soil-plant-atmosphere-continuum (SPAC) would be a step change from the current statistical models of ERD [this study; @Chitra-Tarak:2018].
 Although data needs for parametarising such models [@Christoffersen.2016;@Sperry.1998] could be greater (e.g.
-retention curves for roots), limiting community-wide application, they would improve our understanding of plant physiology [@Johnson:2017].
+retention curves for roots), limiting immediate community-wide application, they promise to improve our understanding of plant physiology [e.g.
+see @Johnson:2017], but which may be potentially harnessed for a community level scale-up.
 

### 2020-08-03 manuscript.rmd
edited a discussion para

@@
 output:
+  bookdown::word_document2:
+        reference_docx: manuscriptstyle.docx
+        fig_caption: yes
   bookdown::html_document2:
@@
     theme: flatly
-    fig_caption: yes
-  bookdown::word_document2:
-        reference_docx: manuscriptstyle.docx
     fig_caption: yes
@@
 **Corresponding author**: 
-Rutuja Chitra-Tarak
+Rutuja Chitra-Tarak, A mispelt word.
 \
@@
 In general, community-scale data collection for rooting or water-sourcing depths is especially challenging for species-rich tropical forests.
-Further, to estimate plant available water with soil depth requires hydrological data and modeling, but these are rarely co-located with forest inventory sites (but see @Johnson:2018fad;@Kumagai:2011;@Ivanov:2012)
+Further, to estimate plant available water with soil depth requires hydrological data and modeling, but these are rarely co-located with forest inventory sites (but see, @Johnson:2018fad; @Kumagai:2011; @Ivanov:2012)
 
@@
 These efforts have resulted in identification of above-ground traits such as proxies of hydraulic failure as indicators of mortality risk.
-A recent meta-analysis found greater drought-induced mortality rates in species with cavitation-vulnerable xylem ($\Psi_{88,~stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~stem}$) [@Anderegg:2016], but counter examples also exist [@Hoffmann:2011; @Paddock:2013]; indicating whether plants reach critical thresholds of hydraulic failure depends on their realised drought intensity (see Table \@ref(tab:symbols) for trait symbols and their definitions).
+A recent meta-analysis found greater drought-induced mortality rates in species with cavitation-vulnerable xylem ($\Psi_{88,~stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~stem}$) [@Anderegg:2016], but counter examples also exist [@Hoffmann:2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013]; indicating whether plants reach critical thresholds of hydraulic failure depends on their realised drought intensity (see Table \@ref(tab:symbols) for trait symbols and their definitions).
 A rare study that accounted for realised water availability for four tree species in a centennial drought (Texas, USA) found that mortality decreased with deep rootedness even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi_{88,~stem}$) and had narrower safety-margins ($\Psi_{min} - \Psi_{88,~stem}$) [@Johnson:2018fad].
@@
 Aquisitives relied on stable water resource at depth, which exhausted in the multi-annual drought, resulting in reduced survival, whereas conservative species used fluctuating shallow-water and survived the drought as some shallow-water was available via infiltration.
-However, lack of data on rooting/water-sourcing depths and hydraulic traits precluded this study from corroborating modelled estimates of water-uptake depths, or investigating trait coordination.
+However, lack of data on rooting or water-sourcing depths and hydraulic traits precluded this study from corroborating modelled estimates of water-uptake depths, or investigating trait coordination.
 
 \
-Trade-offs that shape how trees' make a living in their environment (ecological strategies) provide Dynamic Vegetation Models (DVMs) the essential constraints to represent species-rich communities under changing environment [@Koven:2020].
+Trade-offs shape how trees' make a living in their environment (ecological strategies) and also provide Dynamic Vegetation Models (DVMs) the essential constraints to represent species-rich communities under changing environment [@Koven:2020].
 This entails indentifying trade-offs among environment, traits and demography.
@@
 'Fast' species at the productive end of the spectrum have traits with high capacity to acquire, move, store and use water (high $K_{max}$, low wood density) and achieve high growth potential, but build flimsy, disposable tissues (high SLA) and are less tolerant of low resources, whereas species with slow traits are better protected from C losses and are more drought tolerant (low $\Psi_{50}$, low $\Psi_{tlp}$, larger safety margin).
-Fast strategy is associated with spatially wet sites such as valleys or high rainfall [@Kursar:2009; @Oliveira:2019], but whether and how the fast-slow spectrum holds true for the moisture gradient in the soil column is not clear.
-While deep rootedness may be expected to support a fast strategy with a steady water supply unlike shallow-rooted trees.
+Fast strategy is associated with spatially wet sites such as valleys or high rainfall [@Kursar:2009; @Oliveira:2019], but whether and how the fast-slow spectrum holds true for the moisture dynamics vertically belowground is not clear.
+Deep rootedness may be expected to support a fast strategy with a steady water supply unlike shallow-rooted trees.
 However, this may depend on the exact dynamics of water-availability by depth at a site, depending on precipitation regime and hydrology.
-In pulse-driven arid systems *shallow*-rooted plants are associated with fast traits (high $K_{max}$) that maximise the use of the water-pulse, while *deep*-rooted plants have slow traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013].
+In contrast, in pulse-driven arid systems *shallow*-rooted plants are associated with fast traits (high $K_{max}$) that maximise the use of the water-pulse, while *deep*-rooted plants show slow traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013].
 
 \
-Here, for large trees of xx canopy species of a hyperdiverse neo-tropical moist forest (BCI, Panama) we make a thorough assessment of trade-offs between belowground water environment, above-ground hydraulic traits, estimates of effective rooting depths and demographic outcomes over 35 years that withnessed droughts of various intensity and duration.
+Here, for large trees of length(erd.sp) canopy species of a hyperdiverse neo-tropical moist forest (BCI, Panama) we make a thorough assessment of trade-offs between belowground water environment, above-ground hydraulic traits, estimates of effective rooting depths and demographic outcomes over 34 years that withnessed droughts of various intensity and duration.
 We obtained daily dynamics of plant water potential ($\Psi_{soil,~z}$) by depth $z$ in the whole-soil column with a locally parametarised and calibrated 1-D hydrological water balance model.
@@
 \
-We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterisation in terms of long-term forest demography, plant functional ecology, hydrology, and water sourcing depth water isotopes on tree species and soil from an independent isotopic tracer study on trees’ water-sourcing depths [@Meinzer:1999].
-The 1,560 hectare island was formed after the river Chagres was dammed in 1914 that created Gatun Lake, the main channel of the Panama Canal.
+We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and water sourcing depth water isotopes on tree species and soil from an independent isotopic tracer study on trees’ water-sourcing depths [@Meinzer:1999].
+The 1,560ha island was formed after the river Chagres was dammed in 1914 that created Gatun Lake, the main channel of the Panama Canal.
 Rainfall at BCI is seasonal with a mean annual rainfall of 2687 mm (± 515 SD over 1983-2017) and a four-month long dry season from mid-December through mid-April (≤ 100 mm rain/month).
@@
 Although percentage of canopy species that are at least partly deciduous during the dry season is 28%, the maximum percentage of deciduous individuals in the canopy is only 6.3%, and the percentage of crown area that is deciduous is estimated at 9.7% [@Condit:2000].
-Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014].
+Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014].
 
 \
-In the 50 ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species level and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm size class [@Condit:2017].
+In the 50ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species level and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm size class [@Condit:2017].
 This inventory represents 321 woody species across the eight censuses.
-Aside from a 2 ha patch of ~100 year old secondary forest at the north-eastern corner of the plot the remaining 48 ha represent old growth forest that was never clear cut for agriculture.
+Aside from a 2ha patch of ~100 year old secondary forest at the north-eastern corner of the plot the remaining 48 ha represent old growth forest that was never clear cut for agriculture.
 The plot (100 m x 50 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001].
@@
 where, $q_{rain}$ is rainfall, $E_{v}$ is evapotranspiration (ET) of the forest, $E_{g}$ is ground evaporation, $q_{over}$ is surface runoff, $q_{h2osfc}$ is runoff from surface water storage, $q_{drai}$ is sub-surface drainage (all in mm H~2~O) and $\Delta t$ is the time-step.
-$z$ represents the 15 soil layers exponentially distributed in depth, here, (r soil.depths) (m).
+$z$ represents the depths of exponentially distributed 15 soil layers, here, `r paste(soil.depths, collapse=", ")` (m).
 
@@
 
-We ran ELM-FATES for the 5000 parameter ensembles with hourly climate drivers from 1985-2018 initialised with the stand structure of the 50 ha plot in a single site mode ($0.1^\circ$ x $0.1^\circ$ centered at $9.1530^\circ$N, $280.2539^\circ$E).
+We ran ELM-FATES for the 5000 parameter ensembles with hourly climate drivers from 1985-2018 initialised with the stand structure of the 50ha plot in a single site mode ($0.1^\circ$ x $0.1^\circ$ centered at $9.1530^\circ$N, $280.2539^\circ$E).
 As our key interest here was in soil water availability rather than forest dynamics, we ran ELM-FATES in simplified modes: static stand structure and without plant hydrodynamics, and with a single Plant Functional Type (PFT) of evergreen trees.
@@
 
-```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-iso}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE; m) versus dry season stable isotopic concentration $\\delta^2H$ (mean $\\pm$ (add error); ‰) for ", length(unique(erd.iso$sp))," canopy species from Barro Colorado Island.
+```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-iso}Modelled Effective Rooting Depth (ERD; mean $\\pm$ add error; m) versus dry season stable isotopic concentration $\\delta^2H$ (mean $\\pm$ 1SE; ‰) for ", length(unique(erd.iso$sp))," canopy species from Barro Colorado Island.
 Data from [@Meinzer:1999].
@@
 
-We calculated mortality rates for large trees (>= 10 cm dbh) of species with ERD estimates (all canopy species with maximum height >= 30 m) and average adbunance of at least 20 trees (one species was eleminated with an average abundance of 9 trees; *n* = `r length(erd.mrate.sp)`).
-For each of the seven census intervals ($t$) in the 33 year record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successives censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{t}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50 ha plot of a species present in cen~1~ and dead in cen~2~, respectively, and t is the duration based on mean dates of cen~1~ and cen~2~.
+We calculated mortality rates for large trees (>= 10 cm dbh) of species with ERD estimates (all canopy species with maximum height >= 30 m) and average adbunance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; *n* = `r length(erd.mrate.sp)`).
+For each of the seven census intervals ($t$) in the 34 year record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successives censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{t}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50ha plot of a species present in cen~1~ and dead in cen~2~, respectively, and t is the duration based on mean dates of cen~1~ and cen~2~.
 We calculated species mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{i=1}^{6} {Mort_{t}}$ and, finally, species mortality anoamalies by intervals as $Mort_{anomaly}$ = $Mort_{t}$ - $Mort_{mean}$.
@@
 \
-Species with deeper $ERD$ showed lower mean mortality rates across the 33 year record (1982-2015; *R*^2^ = 0.13, *P* = 0.029, *n* = `r length(erd.mrate.sp)`; Figure \@ref(fig:erd-mort)).
+Species with deeper $ERD$ showed lower mean mortality rates across the 34 year record (1982-2015; *R*^2^ = 0.11, *P* = 0.075, *n* = `r length(erd.mrate.sp)`; Figure \@ref(fig:erd-mort)).
 For five out of seven census intervals, species with shallower $ERD$ had significantly higher mortality anomalies than deeper ERD species; Figure (\@ref(fig:erd-mort-anom)).
 Analysis of time spent under catastrophic soil drought showed that the proportion of days species $\Psi_{ERD}$ was more negative than species $\Psi_{crit}$ decreased with species ERD (Figure \@ref(fig:erd-time)).
-Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25 year period for which soil water potential $\Psi_{soil,~z}$ was modelled (1990-2015), as soil depths at and below 1.7 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dryseason and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).
+Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25 year period for which soil water potential $\Psi_{soil,~z}$ was modelled (1990-2015), as soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dryseasons and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).
 
@@
 Our analyses revealed that trees with deeper effective rooting depths (ERD) were associated higher hydraulic efficiency, but not safety: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species.
-Yet, deeper ERD species showed lower mortality rates than shallower ERD species over the 33 year record (1982-2015).
+Yet, deeper ERD species showed lower mortality rates than shallower ERD species over the 34 year record (1982-2015).
 Analysis of time spent below ($\Psi_{crit}$) in the soil zone matching ERD $\Psi_{soil,~z = ERD}$ showed that this was likely due to decreasing realised risk of hydraulic failure with depth.
@@
 Mortality rates of shallow ERD species may have been exacerbated by increased drought events during the study period.
-Previous work shows that half of the extreme dry seasons, estimated as < 1 SD annual water deficit, during 1931-2015 (84 years) occurred during the first 23 years (1982-2005) of the study period.
+Previous work shows that half of the extreme dry seasons, estimated as < 1 SD annual water deficit, during 1931-2015 (85 yrs) occurred during 1982-2005 (24 yrs) of the study period (1982-2015).
 [@Condit.2017]
 
-### Effective rooting depths at tree community level
+### Realised drought risk integral to assessing drought-vulnerability
+
+Our study brings attention to the need for assessing drought-sensitivity in terms of species realised drought risk.
+and the usefuless of looking at species ecological strategies for that.
+As those are developed in response to plants historical environemnt.
+Plant traits hold clue for the historical environment that shaped their ecological strategy [@Reich:2014].
+
+Our result of lower mortality in deep rooted trees, through a study period of greater recurrence of drought, is consistent with other studies, where among the three or four species studied through a centennial (Texas) or extreme (Italy) drought, deep rooted species showed lower mortality than shallow rooted species.
+Our result contrasts with the finding of @Chitra-Tarak:2018 who found that species identified as effectively deep rooted by their inverse model died *more* in a prolonged, centennial drought.
+Their hydrological model found that in the multi-annual drought deep water was exhausted for over two years, whereas our hydrological modeling at BCI finds that despite higher frequncy of droughts during our study period $\Psi_{soil}$ below the depth of 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dryseasons and drought periods, that is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).
+The difference in our results are likely attributable to site differences in the precipitation regime.
+In their dry forest site (Mudumalai National Park, India) precipitation is low (1095 $\pm$ 296 mm; mean $\pm$ SD) compared to BCI (2687 $\pm$ 515 mm), and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (PET = 9xx $\pm$ xx mm), and thus at BCI deep soil layers are effectively recharged annually [@Chitra-Tarak:2018 and this study].
+We note that the variation in mortality explained by in our study is low (*R*^2^ = 0.11, *P* = 0.075), and other causes may explain observed mortality.
+For example.
+based on a year-long study, lightening explained ~4.5% of mortality for trees >=10 cm dbh and ~40% for trees >=60 cm dbh [@Yanoviak:2019].
+
+### Effective rooting depths at the tree community level
 \
-By estimating ERD for large trees of 33 canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species species-rich tropical forests.
-We note the availability of data on $\delta^2^H$ on trees and water sources belowground--soil or substrate matrix water in the unsaturated vadoze zone or free water below the water table--as important for model calibration/validation.
-The first inverse model of ERD (water uptake depth index, *n* = 12 species) developed by Chitra-Tarak et al @Chitra-Tarak:2018 lacked validation as existing methods were inadequate to extract water bound to the soil and saprolite found in their dry forest study site (Mudumalai National Park, Southern India; Jean Riotte, pers.
-comm.).
-Our ERD model was distinctly constrained on species leaf vulnerability (PLC) curves.
+By estimating ERD for large trees of length(erd.sp) canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species species-rich tropical forests.
+Our chosen ERD model is an advancement over the only other inverse model of ERD (or water uptake depth index, @Chitra-Tarak:2018, *n* = 12 species) in terms of physiologically meaningful representation, fewer parameters, corroboration with tree and soil data on $\delta^2H$.
+Model by @Chitra-Tarak:2018 estimated six species-specific phenomenological parameters: three as root fraction for corresponding lumped vadose zone layers, and three as shape parameters for water uptake vs availability curves.
+For each species, they calculated daily water uptake successively from three layers--the supply, to satisfy the demand of the daily maximum evapo-transpiration, a function of PET and LAI.
+They regressed total supply to total demand ratio over census intervals (termed as growth coefficient) against growth (5 data points), and chose values for six parameters that best-fit (highest R^2^, that is, 2 additional parameters).
+Our inverse ERD model fits only two parameteres by way of regressing depth-specific model against 5 growth data points.
+Our model was constrained on VPD and species leaf vulnerability (PLC) curves, variables that are shown to be physiologically meaningful [add references].
 We benefitted from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (SLA~lamina~ and Wood Specific Gravity) and parameters of PLC curves for which we collected data on a few species (*n* = 14), a 'hard' trait.
 This allowed us to scale-up PLC curves at the community scale (*n* = 284 species).
-Whether such scaling relationships between hard and soft traits hold true for other forests and their wider applicability remains to be seen as data on PLC curves in species-rich, tropical forests are limited, but counter examples exist [see xx, yy, zz (Lawren, could you suggest references?].
+Whether such scaling relationships between hard and soft traits hold true for other forests and their wider applicability remains to be seen as data on PLC curves in tropical forests are limited, but counter examples exist [see xx, yy, zz (Lawren, could you suggest references?].
+We also note the availability of data on $\delta^2^H$ on trees and water sources belowground--soil/substrate matrix water in the vadoze zone or free water below the water table--as important for model calibration/validation.
+Model by Chitra-Tarak et al @Chitra-Tarak:2018 were not corroborated with such data as existing methods were inadequate to extract water bound to the soil and saprolite found in their dry forest study site (Jean Riotte, pers.
+comm.).
+Our model of belowground water-availability (ELM-FATES) was also advanced compared to that used by Chitra-Tarak et al [2018] and was calibrated with data on soil moisture dynamics by depth (15, 40, 100 cm), as well as, evapotranspiration and stream discharge, unlike theirs, which was calibrated on stream discharge and water table levels.
+We lacked data on water table levels which would have been useful to constrain water availability dynamics in the unobserved soil layers.
 
@@
 \
-Our model estimated ERD as a single value, unlike the dynamic ERD modelled by Chitra-Tarak et al @Chitra-Tarak:2018, but future model developments should investigate seasonal dynamics of ERD, calibrated on corresponding calibration data on isotopic water sourcing depths.
+Our model estimated ERD as a single value, unlike the dynamic ERD modelled by @Chitra-Tarak:2018, but future model developments should investigate seasonal dynamics of ERD, calibrated on corresponding calibration data on isotopic water sourcing depths.
 Our inverse ERD model was parametarised on 5-yearly growth data, with five data points over a 25 year period, which provided large degree of freedom to fit the model.
 Future studies could better constrain the model with higher resolution growth data such as those from dendrometers.
-This is especially important in modeling seasonally dynamic ERD.
-However, trees' stem diameter change is a sum total of not just growth, the addition of wood, but also $\pm$ changes due to hydrostatic stem flexing, whose importance increases with temporal resolution [@Chitra-Tarak.2016; @Chitra-Tarak.2014].
+This is especially important for modeling seasonally dynamic ERD.
+However, trees' stem diameter change is a sum total of not just growth, the addition of wood, but also $\pm$ changes due to hydrostatic stem flexing, whose importance increases with temporal resolution [@Chitra-Tarak.2016; @Chitra-Tarak.2015].
 Use of dendrometer data would thus entail accounting for stem hydrostatic flexing [@Zweifel:2018].
-Two of our exploratory ERD models included seasonality in leaf area fraction (LeafFrac), but we did not select as the worsened the fit with growth data and ERD calibration data (Fig.
+Two of our exploratory ERD models included seasonality in leaf area fraction, but we did not select them as they worsened the fit with growth data and evaluation data ($\delta^2H$, Fig.
 S\@ref(fig:other-erd-mods)).
-We interpret this result as that VPD and PLC curves may have adequately explained growth, but also acknowledge that our estimate of seasonality of leaf fraction is a rough estimate based on leaf fall data and assumption of leaf presence during the wet season and we did not take into account inter-annual variation.
+We interpret this result as that VPD and PLC curves may be adequate to explaine growth, but also acknowledge that our estimate of seasonality of leaf fraction is a tentative estimate based on leaf fall data and assumption of leaf presence during the wet season and omits inter-annual variation.
 Future studies could make use of data from new technologies such as drone based monitoring of species-specific leaf area fraction in species-rich forests [@Park.2019].
 In forests such as BCI where understorey tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes, especially from a few growth data points, may need accounting for variation in trees' light environment over time and among trees' of the same size class, an issue we minimised by limiting ERD estimates to canopy trees.
-To estimate topographic variation in ERD within or across species future studies may use a distributed hydrological model.
-We note that data availability on soil water retention curve ($\Psi_{soil}$vs.
-VWC), hydraulic conductivity ($K_{soil}$) by depth and soil moisture by depth were important for effective calibration of our 1-D hydrological model, and recommend wide-spread data collection of these variables as well as water-table levels (not collected in our site) in forest-inventory sites to allow for estimation of trees' water environments.
+To estimate topographic variation in ERD within or across species future studies may use soil moisture dynamics from a distributed hydrological model [e.g.
+@Schwantes:2018].
+We note that data availability on soil water retention curve ($\Psi_{soil}$ vs.
+VWC), hydraulic conductivity ($K_{soil}$) by depth, soil moisture by depth, stream discharge and evapotranspiration were important for effective calibration of our 1-D hydrological model, and recommend wide-spread data collection of these variables as well as water-table levels in forest-inventory sites to allow for estimation of trees' water environments.
 Finally, developing an inverse ERD model that accounts for movement of water within the soil-plant-atmosphere-continuum (SPAC) would be a step change from the current statistical models of ERD [this study; @Chitra-Tarak:2018].
-Although data needs for parametarising such models [@Christoffersen.2016;@Sperry.1998] could be greater (e.g.
+Although data needs for parametarising such models [Christoffersen.2016tbh;@Sperry.1998] could be greater (e.g.
 retention curves for roots), limiting immediate community-wide application, they promise to improve our understanding of plant physiology [e.g.
-see @Johnson:2017], but which may be potentially harnessed for a community level scale-up.
-
-### Need for drought risk
-
-Plant traits hold clues for the historical environments that shaped their ecological strategy.
+see @Johnson:2018fad], but which may be potentially harnessed for a community level scale-up.
 

### 2020-08-03 manuscript.rmd
minor edits

@@
 **Corresponding author**: 
-Rutuja Chitra-Tarak, A mispelt word.
+Rutuja Chitra-Tarak
 \
@@
 This entails indentifying trade-offs among environment, traits and demography.
-A universal 'fast-slow' (aka productive-persistent, or acquisitive-conservative) plant economics spectrum conceptualises plants coordination for all resources and organs, both above- and below-ground [@Reich:2014].
-For water resource observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified a fast-slow spectrum, although examples of all trade-offs observed simultaneously are rare, and belowground trait or environment linkages are weak.
+@Reich:2014 concenptualised a universal 'fast-slow' (aka productive-persistent, or acquisitive-conservative) plant economics spectrum coordinated for all resources and organs, both above- and below-ground.
+specifically for water observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified a fast-slow spectrum, although examples of all trade-offs observed simultaneously are rare, and belowground trait or environment linkages are weak.
 'Fast' species at the productive end of the spectrum have traits with high capacity to acquire, move, store and use water (high $K_{max}$, low wood density) and achieve high growth potential, but build flimsy, disposable tissues (high SLA) and are less tolerant of low resources, whereas species with slow traits are better protected from C losses and are more drought tolerant (low $\Psi_{50}$, low $\Psi_{tlp}$, larger safety margin).
@@
 
-Growth Data: To minimise effect of light variation among trees and census intervals we selected only large trees (>= 30 cm dbh, diamater at breast height) and species whose maximum height was at least 30 m (hereafter, canopy species).
+Growth Data: To minimise effect of light variation among trees and census intervals we selected only large trees (>= 30 cm dbh, diameter at breast height) and species whose maximum height was at least 30 m (hereafter, canopy species).
 We chose only those trees for which diameter (dbh) record was present across all size censuses (1990, 1995, 2000, 2005, 2010, 2015).
 We calculated individual tree ($i$) growth rates ($Gr_{i,t}$ (cm.yr^-1^)) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses, divided by the inter-census interval based on the exact date of measurements.
-As tree growth rates increase with size, to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census onterval ($Resid_{i,t} = Gr_{i,t} - f(DBH_{i, t})$).
+As tree growth rates increase with size, to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($Resid_{i,t} = Gr_{i,t} - f(DBH_{i, t})$).
 The diameter model of growth was fitted based on all trees' data pooled together as there was insufficient species-specific data for large trees (SI ).
-We then standardised (centered and scaled) an individual tree time series ($Resid_{i,t}$) and obtained species-level (*n* = `r length(erd.sp)`) times-series of medians ($Growth_{t}$) that had data on at least three trees (median = xx).
+We then standardized (centered and scaled) an individual tree time series ($Resid_{i,t}$) and obtained species-level (*n* = `r length(erd.sp)`) times-series of medians ($Growth_{t}$) that had data on at least three trees (median = xx).
 As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between GPP (gCm^2^day^-1^) and VPD (kPa; $GPP = f(VPD) + error$;  SI) and used standardised $f(VPD)$ (range 0-1) as $StdVPD$ in Eqs.
@@
 The model with VPD and without Leaf Fraction (Eq.
-\@ref(eq:M2)) had highest goodness-of fit (*R*^2^ = 0.51; *p* = 0.072), followed by followed by model without both VPD and and Leaf Fraction (Eq.
+\@ref(eq:M2)) had highest goodness-of fit (*R*^2^ = 0.51; *p* = 0.072), followed by followed by model without both VPD and abundance Leaf Fraction (Eq.
 \@ref(eq:M1); Fig.
@@
 We calculated mortality rates for large trees (>= 10 cm dbh) of species with ERD estimates (all canopy species with maximum height >= 30 m) and average adbunance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; *n* = `r length(erd.mrate.sp)`).
-For each of the seven census intervals ($t$) in the 34 year record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successives censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{t}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50ha plot of a species present in cen~1~ and dead in cen~2~, respectively, and t is the duration based on mean dates of cen~1~ and cen~2~.
+For each of the seven census intervals ($t$) in the 34-yr record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successive censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{t}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50ha plot of a species present in cen~1~ and dead in cen~2~, respectively, and t is the duration based on mean dates of cen~1~ and cen~2~.
 We calculated species mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{i=1}^{6} {Mort_{t}}$ and, finally, species mortality anoamalies by intervals as $Mort_{anomaly}$ = $Mort_{t}$ - $Mort_{mean}$.
@@
 \
-In this forest modelled Effective Rooting Depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from 0.5 m to 13 m (Figure \@ref(fig:sp-erd)).
-For the subset of species for which hydraulic properties were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max,~stem}$; *R*^2^ = 0.39, *P* = 0.098, *n* = 8; Figure \@ref(fig:erd-hyd)A), less negative leaf  ($\Psi_{tlp}$); *R*^2^ = 0.6, p = 0.024, *n* = 8; Figure \@ref(fig:erd-hyd)B), greater vulnerability to embolism from cavitation with less negative ($\Psi_{88,~stem}$) (*R*^2^ = 0.72, p = 0.007, *n* = 8; Figure \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$; *R*^2^ = 0.74, p = 0.013, *n* = 8; Figure \@ref(fig:erd-hyd)D).
+In this forest modelled Effective Rooting Depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from 0.5 m to 13 m (Fig.
+\@ref(fig:sp-erd)).
+For the subset of species for which hydraulic properties were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max,~stem}$; *R*^2^ = 0.39, *P* = 0.098, *n* = 8; Fig.
+\@ref(fig:erd-hyd)A), less negative leaf  ($\Psi_{tlp}$); *R*^2^ = 0.6, p = 0.024, *n* = 8; Fig.
+\@ref(fig:erd-hyd)B), greater vulnerability to embolism from cavitation with less negative ($\Psi_{88,~stem}$) (*R*^2^ = 0.72, p = 0.007, *n* = 8; Fig.
+\@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$; *R*^2^ = 0.74, p = 0.013, *n* = 8; Fig.
+\@ref(fig:erd-hyd)D).
 
@@
 
-```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus hydraulic propertiese for eight large, canopy species found on Barro Colorado Island (Panama).
+```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus hydraulic properties for eight large, canopy species found on Barro Colorado Island (Panama).
 *R*^2^ and *P*-values are given in panel insets.[X-Axis labels to be added: $K_{max}$ units as in Fig.
@@
 \
-Species with deeper $ERD$ showed lower mean mortality rates across the 34 year record (1982-2015; *R*^2^ = 0.11, *P* = 0.075, *n* = `r length(erd.mrate.sp)`; Figure \@ref(fig:erd-mort)).
-For five out of seven census intervals, species with shallower $ERD$ had significantly higher mortality anomalies than deeper ERD species; Figure (\@ref(fig:erd-mort-anom)).
-Analysis of time spent under catastrophic soil drought showed that the proportion of days species $\Psi_{ERD}$ was more negative than species $\Psi_{crit}$ decreased with species ERD (Figure \@ref(fig:erd-time)).
-Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25 year period for which soil water potential $\Psi_{soil,~z}$ was modelled (1990-2015), as soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dryseasons and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).
+Species with deeper $ERD$ showed lower mean mortality rates across the 34-yr record (1982-2015; *R*^2^ = 0.11, *P* = 0.075, *n* = `r length(erd.mrate.sp)`; Fig.
+\@ref(fig:erd-mort)).
+For five out of seven census intervals, species with shallower $ERD$ had significantly higher mortality anomalies than deeper ERD species; Fig.
+(\@ref(fig:erd-mort-anom)).
+Analysis of time spent under catastrophic soil drought showed that the proportion of days species $\Psi_{ERD}$ was more negative than species $\Psi_{crit}$ decreased with species ERD (Fig.
+\@ref(fig:erd-time)).
+Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25-yrperiod for which soil water potential $\Psi_{soil,~z}$ was modelled (1990-2015), as soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dry seasons and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).
 
@@
 \
-Our analyses revealed that trees with deeper effective rooting depths (ERD) were associated higher hydraulic efficiency, but not safety: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species.
-Yet, deeper ERD species showed lower mortality rates than shallower ERD species over the 34 year record (1982-2015).
+Our analyses revealed that co-occurring large canopy tree species (*n* = ` r length(erd.sp)`) with deeper effective rooting depths (ERD) were associated with higher hydraulic efficiency, but not safety: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species.
+Yet, deeper ERD species showed lower mortality rates than shallower ERD species over the 34-yr record (1982-2015).
 Analysis of time spent below ($\Psi_{crit}$) in the soil zone matching ERD $\Psi_{soil,~z = ERD}$ showed that this was likely due to decreasing realised risk of hydraulic failure with depth.
@@
 Mortality rates of shallow ERD species may have been exacerbated by increased drought events during the study period.
-Previous work shows that half of the extreme dry seasons, estimated as < 1 SD annual water deficit, during 1931-2015 (85 yrs) occurred during 1982-2005 (24 yrs) of the study period (1982-2015).
-[@Condit.2017]
+Previous work shows that half of the extreme dry seasons, estimated as < 1 SD annual water deficit, during 1931-2015 (85-yr) occurred during 1982-2005 (24-yr) of the 34-yr long study period (1982-2015; @Condit.2017)
 
@@
 Our study brings attention to the need for assessing drought-sensitivity in terms of species realised drought risk.
-and the usefuless of looking at species ecological strategies for that.
-As those are developed in response to plants historical environemnt.
-Plant traits hold clue for the historical environment that shaped their ecological strategy [@Reich:2014].
-
-Our result of lower mortality in deep rooted trees, through a study period of greater recurrence of drought, is consistent with other studies, where among the three or four species studied through a centennial (Texas) or extreme (Italy) drought, deep rooted species showed lower mortality than shallow rooted species.
-Our result contrasts with the finding of @Chitra-Tarak:2018 who found that species identified as effectively deep rooted by their inverse model died *more* in a prolonged, centennial drought.
-Their hydrological model found that in the multi-annual drought deep water was exhausted for over two years, whereas our hydrological modeling at BCI finds that despite higher frequncy of droughts during our study period $\Psi_{soil}$ below the depth of 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dryseasons and drought periods, that is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).
+We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins traits, that are identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through a study period of greater recurrence of drought.
+These results are consistent with the rare studies that studied three or four species each with known rooting or water sourcing depths and assesed their hydraulic traits and mortality rates during a centennial (Texas) or extreme (Italy) drought [@Johnson:2018fad; @Nardini:2015].
+Our result of lower mortality in deep rooted trees contrasts with the finding of @Chitra-Tarak:2018, who found that species identified by their inverse ERD model as effectively deep rooted died *more* in a prolonged, centennial drought.
+Their hydrological model found that in the multi-annual drought deep water was exhausted for over two years, whereas our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period $\Psi_{soil}$ below the depth of 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dryseasons and drought periods, that is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).
 The difference in our results are likely attributable to site differences in the precipitation regime.
@@
 
+and the usefulness of looking at species ecological strategies for that.
+As those are developed in response to plants historical environment.
+Plant traits hold clue for the historical environment that shaped their ecological strategy [@Reich:2014].
+
 ### Effective rooting depths at the tree community level
 \
-By estimating ERD for large trees of length(erd.sp) canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species species-rich tropical forests.
+By estimating ERD for large trees of `r length(erd.sp)` canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species species-rich tropical forests.
 Our chosen ERD model is an advancement over the only other inverse model of ERD (or water uptake depth index, @Chitra-Tarak:2018, *n* = 12 species) in terms of physiologically meaningful representation, fewer parameters, corroboration with tree and soil data on $\delta^2H$.
@@
 They regressed total supply to total demand ratio over census intervals (termed as growth coefficient) against growth (5 data points), and chose values for six parameters that best-fit (highest R^2^, that is, 2 additional parameters).
-Our inverse ERD model fits only two parameteres by way of regressing depth-specific model against 5 growth data points.
+Our inverse ERD model fits only two parameters by way of regressing depth-specific model against 5 growth data points.
 Our model was constrained on VPD and species leaf vulnerability (PLC) curves, variables that are shown to be physiologically meaningful [add references].
@@
 Our model estimated ERD as a single value, unlike the dynamic ERD modelled by @Chitra-Tarak:2018, but future model developments should investigate seasonal dynamics of ERD, calibrated on corresponding calibration data on isotopic water sourcing depths.
-Our inverse ERD model was parametarised on 5-yearly growth data, with five data points over a 25 year period, which provided large degree of freedom to fit the model.
+Our inverse ERD model was parametarised on 5-yearly growth data, with five data points over a 25-yrperiod, which provided large degree of freedom to fit the model.
 Future studies could better constrain the model with higher resolution growth data such as those from dendrometers.
@@
 Future studies could make use of data from new technologies such as drone based monitoring of species-specific leaf area fraction in species-rich forests [@Park.2019].
-In forests such as BCI where understorey tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes, especially from a few growth data points, may need accounting for variation in trees' light environment over time and among trees' of the same size class, an issue we minimised by limiting ERD estimates to canopy trees.
+In forests such as BCI where understory tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes, especially from a few growth data points, may need accounting for variation in trees' light environment over time and among trees' of the same size class, an issue we minimised by limiting ERD estimates to canopy trees.
 To estimate topographic variation in ERD within or across species future studies may use soil moisture dynamics from a distributed hydrological model [e.g.
@@
 Finally, developing an inverse ERD model that accounts for movement of water within the soil-plant-atmosphere-continuum (SPAC) would be a step change from the current statistical models of ERD [this study; @Chitra-Tarak:2018].
-Although data needs for parametarising such models [Christoffersen.2016tbh;@Sperry.1998] could be greater (e.g.
+Although data needs for parametarizing such models [Christoffersen.2016tbh;@Sperry.1998] could be greater (e.g.
 retention curves for roots), limiting immediate community-wide application, they promise to improve our understanding of plant physiology [e.g.

### 2020-08-03 manuscript.rmd
symbol edits

@@
 output:
-  bookdown::word_document2:
-        reference_docx: manuscriptstyle.docx
-        fig_caption: yes
   bookdown::html_document2:
@@
     theme: flatly
+    fig_caption: yes
+  bookdown::word_document2:
+        reference_docx: manuscriptstyle.docx
         fig_caption: yes
@@
 - We inversely modelled Effective Rooting Depth (ERD) for large trees of `r length(erd.sp)` canopy tree species by linking species growth dynamics (6 censuses, 1990-2015) to VPD, Soil Water Potential ($\Psi_{soil}$) by depth in the whole-soil column, and a new community-scale model of species vulnerability curves ($K_{leaf}$ vs.
-$\Psi_{leaf}$) that scaled new data on 21 species with Leaf Mass Area and Wood Specific Gravity.
+$\Psi_{leaf}$) that scaled new data on 21 species with Leaf Mass Area~lamina~ and Wood Specific Gravity~100^$\circ$^C~.
 We evaluated ERD by leveraging existing isotopic tracer data ($\delta^2H$) on tree xylem sap and soil column.
@@
 Tropical forests are considered an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw]: here, historic climatic variation is relatively small [@Mora.2013], and the observed ranges of trees' hydraulic safety margins are relatively narrow [@Choat:2012; @Oliveira:2019].
-Widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak:2018; @Hilker:2014; @Phillips:2010].
+Widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010].
 However, state-of-the-art dynamic vegetation models (DVMs) fail to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013], as underlying mechanisms of drought-tolerance are not fully understood.
@@
 A species sourcing deep water in a Karst system (Italy) showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi_{88,~stem}$ and narrower safety margin compared to two species sourcing shallow-water [@Nardini:2015].
-@Chitra-Tarak:2018 proposed the first community-scale model that linked trees' realised water availability to demographic outcomes through droughts.
+@Chitra-Tarak.2018 proposed the first community-scale model that linked trees' realised water availability to demographic outcomes through droughts.
 They found that species accessing deep-water in a tropical dry forest (India) died more in a rare, centennial drought compared to shallow-water species.
@@
 We obtained daily dynamics of plant water potential ($\Psi_{soil,~z}$) by depth $z$ in the whole-soil column with a locally parametarised and calibrated 1-D hydrological water balance model.
-Developing a novel inverse model allowed us to estimate species effective rooting depths (ERD), defined as the depth ($z$) whose soil water dynamics ($\Psi_{~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015).
+Developing a novel inverse model allowed us to estimate species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015).
 This model incorporated impact of atmospheric drought (VPD) on growth and was constrained by above-ground hydraulic traits, namely, species leaf vulnerability curves ($K_{leaf}$ vs.
@@
 \
-In the 50ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species level and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm size class [@Condit:2017].
+In the 50ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species level and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm size class [@Condit.2017].
 This inventory represents 321 woody species across the eight censuses.
@@
 We identified scaling relationships between fitted parameters $A$ and $B$ in Eq.
-\@ref(eq:vc-exp) and two traits that were available at community-scale, namely, $WSG$ and $LMA$, defined as mean wood specific gravity after drying at 100$^\circ$C (gcm^-3^) and mean leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules (gm^-2^) for leaves receiving direct sunlight, respectively.
+\@ref(eq:vc-exp) and two traits that were available at community-scale, namely, $WSG$ (or, WSG~100^$\circ$^C~) and $LMA$ (or, LMA~lamina~), defined, respectively, as mean wood specific gravity after drying at 100$^\circ$C (gcm^-3^) and mean leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules (gm^-2^) for leaves receiving direct sunlight.
 Fitted polynomial equations are described as,
@@
 We chose only those trees for which diameter (dbh) record was present across all size censuses (1990, 1995, 2000, 2005, 2010, 2015).
-We calculated individual tree ($i$) growth rates ($Gr_{i,t}$ (cm.yr^-1^)) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses, divided by the inter-census interval based on the exact date of measurements.
+We calculated individual tree ($i$) growth rates ($Gr_{i,t}$; cm.yr^-1^) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses, divided by the inter-census interval based on the exact date of measurements.
 As tree growth rates increase with size, to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($Resid_{i,t} = Gr_{i,t} - f(DBH_{i, t})$).
@@
 S\@ref(fig:iso-data).
-We also removed *Guapira* *spp*, whose xylem sap isotope value was not constrained by the soil water isotope depth profile.
+We also removed *Guapira* *standleyana*, whose xylem sap isotope value was not constrained by the soil water isotope depth profile.
 
@@
 \@ref(fig:erd-time)).
-Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25-yrperiod for which soil water potential $\Psi_{soil,~z}$ was modelled (1990-2015), as soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dry seasons and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).
+Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25-yr period for which soil water potential $\Psi_{soil,~z}$ was modelled (1990-2015), as soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dry seasons and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).
 
@@
 \
-Our analyses revealed that co-occurring large canopy tree species (*n* = ` r length(erd.sp)`) with deeper effective rooting depths (ERD) were associated with higher hydraulic efficiency, but not safety: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species.
+Our analyses revealed that co-occurring large canopy tree species (*n* = `r length(erd.sp)`) with deeper effective rooting depths (ERD) were associated with higher hydraulic efficiency, but not safety: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species.
 Yet, deeper ERD species showed lower mortality rates than shallower ERD species over the 34-yr record (1982-2015).
@@
 These results are consistent with the rare studies that studied three or four species each with known rooting or water sourcing depths and assesed their hydraulic traits and mortality rates during a centennial (Texas) or extreme (Italy) drought [@Johnson:2018fad; @Nardini:2015].
-Our result of lower mortality in deep rooted trees contrasts with the finding of @Chitra-Tarak:2018, who found that species identified by their inverse ERD model as effectively deep rooted died *more* in a prolonged, centennial drought.
+Our result of lower mortality in deep rooted trees contrasts with the finding of @Chitra-Tarak.2018, who found that species identified by their inverse ERD model as effectively deep rooted died *more* in a prolonged, centennial drought.
 Their hydrological model found that in the multi-annual drought deep water was exhausted for over two years, whereas our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period $\Psi_{soil}$ below the depth of 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dryseasons and drought periods, that is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).
 The difference in our results are likely attributable to site differences in the precipitation regime.
-In their dry forest site (Mudumalai National Park, India) precipitation is low (1095 $\pm$ 296 mm; mean $\pm$ SD) compared to BCI (2687 $\pm$ 515 mm), and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (PET = 9xx $\pm$ xx mm), and thus at BCI deep soil layers are effectively recharged annually [@Chitra-Tarak:2018 and this study].
+In their dry forest site (Mudumalai National Park, India) precipitation is low (1095 $\pm$ 296 mm; mean $\pm$ SD) compared to BCI (2687 $\pm$ 515 mm), and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (PET = 9xx $\pm$ xx mm), and thus at BCI deep soil layers are effectively recharged annually [@Chitra-Tarak.2018 and this study].
 We note that the variation in mortality explained by in our study is low (*R*^2^ = 0.11, *P* = 0.075), and other causes may explain observed mortality.
-For example.
-based on a year-long study, lightening explained ~4.5% of mortality for trees >=10 cm dbh and ~40% for trees >=60 cm dbh [@Yanoviak:2019].
+For example, based on a year-long study, lightening explained ~4.5% of mortality for trees >=10 cm dbh and ~40% for trees >=60 cm dbh [@Yanoviak:2019].
 
-and the usefulness of looking at species ecological strategies for that.
-As those are developed in response to plants historical environment.
-Plant traits hold clue for the historical environment that shaped their ecological strategy [@Reich:2014].
+The trade-offs that we identify between above- and below-ground traits, water environment vertically belowground and mortality are important for representing diversity in Dynamic Global Vegetation Models especially in species-rich forests.
+The hydraulic efficiency-safety trade-off between high values of maximum leaf or stem conductivity (high $K_{max}$) and cavitation-resistance (low $\Psi_{50}$ or $\Psi_{88}$) is hypothesised although community-wide examples are rare [@Hacke:2006; @Gleason:2016].
+We report such a trade-off at both leaf (*n* = 21 species) and branch level (*n* = 27 species, but data measured on trees in SLZ and PNM, only include those species that are found in BCI?; Or shall I add those species numbers and a SI figure for only those species for which these data are available for both leaf and branch?).
+We also found that LMA~lamina~ and WSG~100^$\circ$^C~ were strong predictors of PLC curves for leaf, but not branch (add SI figures).
+High $K_{max}$ is elsewhere associated with photosynthetic rates with faster growth potential, but we did not find a relationship between $K_{max}$ and growth rates (*n* = 21 species).
 
@@
 By estimating ERD for large trees of `r length(erd.sp)` canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species species-rich tropical forests.
-Our chosen ERD model is an advancement over the only other inverse model of ERD (or water uptake depth index, @Chitra-Tarak:2018, *n* = 12 species) in terms of physiologically meaningful representation, fewer parameters, corroboration with tree and soil data on $\delta^2H$.
-Model by @Chitra-Tarak:2018 estimated six species-specific phenomenological parameters: three as root fraction for corresponding lumped vadose zone layers, and three as shape parameters for water uptake vs availability curves.
+Our chosen ERD model is an advancement over the only other inverse model of ERD (or water uptake depth index, @Chitra-Tarak.2018, *n* = 12 species) in terms of physiologically meaningful representation, fewer parameters, corroboration with tree and soil data on $\delta^2H$.
+Model by @Chitra-Tarak.2018 estimated six species-specific phenomenological parameters: three as root fraction for corresponding lumped vadose zone layers, and three as shape parameters for water uptake vs availability curves.
 For each species, they calculated daily water uptake successively from three layers--the supply, to satisfy the demand of the daily maximum evapo-transpiration, a function of PET and LAI.
@@
 Our model was constrained on VPD and species leaf vulnerability (PLC) curves, variables that are shown to be physiologically meaningful [add references].
-We benefitted from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (SLA~lamina~ and Wood Specific Gravity) and parameters of PLC curves for which we collected data on a few species (*n* = 14), a 'hard' trait.
+We benefitted from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA~lamina~ and WSG~100^$\circ$^C~) and parameters of PLC curves for which we collected data on a few species (*n* = 14), a 'hard' trait.
 This allowed us to scale-up PLC curves at the community scale (*n* = 284 species).
 Whether such scaling relationships between hard and soft traits hold true for other forests and their wider applicability remains to be seen as data on PLC curves in tropical forests are limited, but counter examples exist [see xx, yy, zz (Lawren, could you suggest references?].
-We also note the availability of data on $\delta^2^H$ on trees and water sources belowground--soil/substrate matrix water in the vadoze zone or free water below the water table--as important for model calibration/validation.
-Model by Chitra-Tarak et al @Chitra-Tarak:2018 were not corroborated with such data as existing methods were inadequate to extract water bound to the soil and saprolite found in their dry forest study site (Jean Riotte, pers.
+We also note the availability of data on $\delta^2H$ on trees and water sources belowground--soil/substrate matrix water in the vadoze zone or free water below the water table--as important for model calibration/validation.
+Model by @Chitra-Tarak.2018 were not corroborated with such data as existing methods were inadequate to extract water bound to the soil and saprolite found in their dry forest study site (Jean Riotte, pers.
 comm.).
-Our model of belowground water-availability (ELM-FATES) was also advanced compared to that used by Chitra-Tarak et al [2018] and was calibrated with data on soil moisture dynamics by depth (15, 40, 100 cm), as well as, evapotranspiration and stream discharge, unlike theirs, which was calibrated on stream discharge and water table levels.
+Our model of belowground water-availability (ELM-FATES) was also advanced compared to that used by @Chitra-Tarak.2018 and was calibrated with data on soil moisture dynamics by depth (15, 40, 100 cm), as well as, evapotranspiration and stream discharge, unlike theirs, which was calibrated on stream discharge and water table levels.
 We lacked data on water table levels which would have been useful to constrain water availability dynamics in the unobserved soil layers.
 
+#### Future Directions for Model Development
 \
-##### Future Directions for Model Development
-\
-Our model estimated ERD as a single value, unlike the dynamic ERD modelled by @Chitra-Tarak:2018, but future model developments should investigate seasonal dynamics of ERD, calibrated on corresponding calibration data on isotopic water sourcing depths.
-Our inverse ERD model was parametarised on 5-yearly growth data, with five data points over a 25-yrperiod, which provided large degree of freedom to fit the model.
+Our model estimated ERD as a single value, unlike the dynamic ERD modelled by @Chitra-Tarak.2018, but future model developments should investigate seasonal dynamics of ERD, calibrated on corresponding calibration data on isotopic water sourcing depths.
+Our inverse ERD model was parametarised on 5-yearly growth data, with five data points over a 25-yr period, which provided large degree of freedom to fit the model.
 Future studies could better constrain the model with higher resolution growth data such as those from dendrometers.
 This is especially important for modeling seasonally dynamic ERD.
-However, trees' stem diameter change is a sum total of not just growth, the addition of wood, but also $\pm$ changes due to hydrostatic stem flexing, whose importance increases with temporal resolution [@Chitra-Tarak.2016; @Chitra-Tarak.2015].
+However, trees' stem diameter change is a sum total of not just growth, the addition of wood, but also changes due to hydrostatic stem flexing, importance of which increases with temporal resolution [@Chitra-Tarak.2016; @Chitra-Tarak.2015].
 Use of dendrometer data would thus entail accounting for stem hydrostatic flexing [@Zweifel:2018].
@@
 VWC), hydraulic conductivity ($K_{soil}$) by depth, soil moisture by depth, stream discharge and evapotranspiration were important for effective calibration of our 1-D hydrological model, and recommend wide-spread data collection of these variables as well as water-table levels in forest-inventory sites to allow for estimation of trees' water environments.
-Finally, developing an inverse ERD model that accounts for movement of water within the soil-plant-atmosphere-continuum (SPAC) would be a step change from the current statistical models of ERD [this study; @Chitra-Tarak:2018].
-Although data needs for parametarizing such models [Christoffersen.2016tbh;@Sperry.1998] could be greater (e.g.
+Finally, developing an inverse ERD model that accounts for movement of water within the soil-plant-atmosphere-continuum (SPAC) would be a step change from the current statistical models of ERD (this study; @Chitra-Tarak.2018).
+Although data needs for parametarizing such models [@Christoffersen.2016tbh; @Sperry.1998] could be greater (e.g.
 retention curves for roots), limiting immediate community-wide application, they promise to improve our understanding of plant physiology [e.g.
@@
 
-```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-other-erd-mods}Effective Rooting Depth versus $\\delta^2^H$ for all the models tested.
+```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-other-erd-mods}Effective Rooting Depth versus $\\delta^2H$ for all the models tested.
 See Eqs.
@@
 ```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves} Leaf 'vulnerability curves', that is standardized $K_{leaf}$ vs.
-$\\Psi_{leaf}$, fitted to observed data on ",  length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))," tree species from Panama, color coded by Wood Specific Gravity ($gcm^{-3}$).")}
+$\\Psi_{leaf}$, fitted to observed data on ",  length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))," tree species from Panama, color coded by Wood Specific Gravity~100^$\circ$^C~ ($gcm^{-3}$).")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB_for_data_sp.jpeg"))
@@
 
-```{r erd-growth, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-growth}Scatterplot between Effective Rooting Deoth (ERD) and mean growth rate $\\pm$ (add error)for xx large, canopy species"}
+```{r erd-growth, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-growth}Scatterplot between Effective Rooting Deoth (ERD) and mean growth rate $\\pm$ (add error) for xx large, canopy species"}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/adult_Growth_vs_udi_with_outliers_sp.jpeg"))

### 2020-08-03 manuscript.rmd
Minor edits

@@
 \@ref(eq:water-balance), determined their ranges based on literature for the study site or for the tropics (Table \@ref(tab:param)), and sampled 5000 ensembles from this parameter space with Latin Hypercube Sampling (LHS).
-In the Conrad catchment soil hydraulic conductivity declined by depth based on existing data for the depths of 12.5 and 60 cm (@Kinner:2004; (Table \@ref(tab:param)).
+In the Conrad catchment soil hydraulic conductivity declined by depth based on existing data for the depths of 12.5 and 60 cm (@Kinner:2004; Table \@ref(tab:param)).
 We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths < 12.5 cm and > 60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig.
@@
 $\Psi_{soil}$ for depths 15, 40 and 100 cm [@Kupers:2019].
-A Campbell @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences among depths were nominal:
+A @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences among depths were nominal:
 
@@
 
--- Write about why models were designed this way.
-
-\
-\begin{equation}
-K_{leaf,z,n} = Ae^{-B\Psi_{soil,~z,n}}
-(\#eq:k-soil)
-\end{equation}
+Water is a key limiting resource for tree growth.
+Vapour pressure deficit (VPD) and species-specific maximum hydraulic conductivity and its decline with water-stress ($K_{leaf}$ vs.
+$\Psi_{soil}$), are factors known to influence trees water uptake [add references, Chonggang?].
+For dry-deciduous species seasonality in leaf cover may also influence plant water uptake.
+We therefore tested alternate models of growth, aka inverse models for ERD (Eqs.
+\@ref(eq:M2)- \@ref(eq:M4)), in which a species growth dynamics across census intervals was regressed against census interval means of multiplied effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the fraction of leaf cover (all standardised to 0-1, see below).
+Let n~t~ be the total number of days in census interval $t$, then for individual tree species Growth~t~ and depth $z$ these models are described as:
 
 \begin{equation}
-Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,z,n} + error
+Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,~z,~n} + error
 (\#eq:M1)
@@
 \begin{equation}
-Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,n} StdVPD_{n} + error
+Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,~n} StdVPD_{n} + error
 (\#eq:M2)
@@
 \begin{equation}
-Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,z,n} LeafFrac_{doy} + error
+Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,~z,~n} LeafCover_{doy} + error
 (\#eq:M3)
@@
 \begin{equation}
-Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,z,n} LeafFrac_{doy}StdVPD_{n} + error
+Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,~z,~n} LeafCover_{doy}StdVPD_{n} + error
 (\#eq:M4)
@@
 
+We estimated daily value of $K_{leaf,z,n}$ by plugging in $\Psi_{soil,~z,~n}$ in the leaf PLC curves Eq. \@ref(eq:vc-exp) as:
+
+\
+\begin{equation}
+K_{leaf,~z,~n} = Ae^{-B\Psi_{soil,~z,~n}}
+(\#eq:k-soil)
+\end{equation}
+
+We estimated species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015; see below).
+
 ### Model parameterization
 
-Growth Data: To minimise effect of light variation among trees and census intervals we selected only large trees (>= 30 cm dbh, diameter at breast height) and species whose maximum height was at least 30 m (hereafter, canopy species).
+To minimise effect of light variation among trees and census intervals we selected only large trees (>= 30 cm dbh, diameter at breast height) and species whose maximum height was at least 30 m (hereafter, canopy species) that are likely to be fully exposed to the sun.
 We chose only those trees for which diameter (dbh) record was present across all size censuses (1990, 1995, 2000, 2005, 2010, 2015).
@@
 As tree growth rates increase with size, to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($Resid_{i,t} = Gr_{i,t} - f(DBH_{i, t})$).
-The diameter model of growth was fitted based on all trees' data pooled together as there was insufficient species-specific data for large trees (SI ).
+The diameter model of growth was fitted based on all trees' data pooled together as there was insufficient species-specific data for large trees (Do I need to add SI figure?).
 We then standardized (centered and scaled) an individual tree time series ($Resid_{i,t}$) and obtained species-level (*n* = `r length(erd.sp)`) times-series of medians ($Growth_{t}$) that had data on at least three trees (median = xx).
-As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between GPP (gCm^2^day^-1^) and VPD (kPa; $GPP = f(VPD) + error$;  SI) and used standardised $f(VPD)$ (range 0-1) as $StdVPD$ in Eqs.
+
+As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between GPP (gCm^2^day^-1^) data we collected from the Eddy Flux Tower and VPD (kPa; $\bar{GPP} = m_2VPD^2 + m_1VPD + m_0 + error = StdVPD$; Do I need to add SI Fig?) and used standardised $StdVPD$ (range 0-1) in Eqs.
 \@ref(eq:M2)- \@ref(eq:M4).
 
-Leaf fraction...
+To explore the effect of species seasonality in leaf cover on growth, we made an informed assumption of  species-specific seasonal curves for Leaf Cover Fraction ($LeafCover$, range 0-1, unitless).
+For evergreen species we assumed $LeafCover$ = 1 throughtout the year, whereas for species with all other decisuous classes, we assumed $LeafCover$ = 1 during the wet season (DOY = 150 to 300), given high forest leaf area index for this period (data not shown).
+We interpolated species-specific weekly leaf-fall data collected at two similar sites in BCI (1985-2020) to a daily estimate, scaled to express daily leaf fall as a fraction of the total leaf fall in a given year and obtained an average leaf fall fraction by DOY estimate across years ($LeafFall$, range 0-1, unitless).
+For non-evergreen species, we cumulatively deducted $LeafFall$ from $LeafCover$ = 1 beginning DOY = 300 and obtained a $LeafCover$ by DOY curve (\@ref(fig:leaf-cover)).
 
-For each model \@ref(eq:M1)- \@ref(eq:M4) and species combination ($sp$), we obtained goodness-of-fit (*R*^2^) and Pearson's $\rho$ for all soil depths $z$, and for each sp, model combination we chose the $z$ with highest *R*^2^ as the Effective Rooting Depth (ERD); after retaining species with positive Pearson's $\rho$ and *R*^2^ >= 0.1.
+We obtained species-specific parameters $A$ & $B$ for Eq. \@ref(eq:k-soil) either from data when available, or a model based on $LMA$ or $WSG$ as described in the earlier section.
+
+For each model \@ref(eq:M1)- \@ref(eq:M4) and species combination ($sp$), we obtained goodness-of-fit (*R*^2^) and Pearson's $\rho$ for all soil depths $z$, and for each sp and model combination we chose the $z$ with highest *R*^2^ as the Effective Rooting Depth (ERD), only for species with positive Pearson's $\rho$ and *R*^2^ >= 0.1.
 
@@
 \
-The model with VPD and without Leaf Fraction (Eq.
-\@ref(eq:M2)) had highest goodness-of fit (*R*^2^ = 0.51; *p* = 0.072), followed by followed by model without both VPD and abundance Leaf Fraction (Eq.
+The model with VPD and without Leaf Cover (Eq.
+\@ref(eq:M2)) had highest goodness-of fit (*R*^2^ = 0.51; *p* = 0.072), followed by followed by model without both VPD and Leaf Cover (Eq.
 \@ref(eq:M1); Fig.
 S\@ref(fig:other-erd-mods)).
-Models with leaf fraction had only three species with sufficient species-level goodness-of-fit (Eqs.
+Models with Leaf Cover had only three species with sufficient species-level goodness-of-fit (Eqs.
 \@ref(eq:M3)-\@ref(eq:M4)) versus seven for models without (Eqs.
 \@ref(eq:M1)-\@ref(eq:M2)).
-For rest of the analyses we therefore used species ERD estimates from the model with VPD and without Leaf Fraction (Eq.
+For rest of the analyses we therefore used species ERD estimates from the model with VPD and without Leaf Cover (Eq.
 \@ref(eq:M2); Fig.
@@
 To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD with maximum stem hydraulic conductance ($K_{max, stem}$), leaf turgor loss point ($\Psi_{tlp}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$).
+See SI for data collection and estimation of these variables.
 
@@
 
-We calculated mortality rates for large trees (>= 10 cm dbh) of species with ERD estimates (all canopy species with maximum height >= 30 m) and average adbunance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; *n* = `r length(erd.mrate.sp)`).
-For each of the seven census intervals ($t$) in the 34-yr record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successive censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{t}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50ha plot of a species present in cen~1~ and dead in cen~2~, respectively, and t is the duration based on mean dates of cen~1~ and cen~2~.
-We calculated species mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{i=1}^{6} {Mort_{t}}$ and, finally, species mortality anoamalies by intervals as $Mort_{anomaly}$ = $Mort_{t}$ - $Mort_{mean}$.
+We calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$ 30 m) and average adbunance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; *n* = `r length(erd.mrate.sp)`).
+For each of the seven census intervals ($t$) in the 34-yr record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successive censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{time}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50ha plot of a species present in cen~1~ and dead in cen~2~, respectively, and $time$ is the duration based on mean dates of cen~1~ and cen~2~.
+We calculated species mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{t=1}^{6} {Mort_{t}}$ and, finally, species mortality anomalies by intervals as $Mort_{anomaly}$ = $Mort_{t}$ - $Mort_{mean}$.
 To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$).
@@
 Each circle represents a species.
-Circles are jittered vertically around the named depths (ticks) to be able to distinguish visually.
+Circles are colored and jittered vertically around the named depths (ticks) for clarity.
 (Are *R*^2^ and *P*-values needed here?)")}
@@
 Mortality rates of shallow ERD species may have been exacerbated by increased drought events during the study period.
-Previous work shows that half of the extreme dry seasons, estimated as < 1 SD annual water deficit, during 1931-2015 (85-yr) occurred during 1982-2005 (24-yr) of the 34-yr long study period (1982-2015; @Condit.2017)
+Previous work shows that half of the extreme dry seasons, estimated as < 1 SD annual water deficit, during 1931-2015 (85-yr) occurred during 1982-2005 (24-yr) of the 34-yr long study period (1982-2015; @Condit.2017).
 
 ### Realised drought risk integral to assessing drought-vulnerability
-
+\
 Our study brings attention to the need for assessing drought-sensitivity in terms of species realised drought risk.
@@
 Our result of lower mortality in deep rooted trees contrasts with the finding of @Chitra-Tarak.2018, who found that species identified by their inverse ERD model as effectively deep rooted died *more* in a prolonged, centennial drought.
-Their hydrological model found that in the multi-annual drought deep water was exhausted for over two years, whereas our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period $\Psi_{soil}$ below the depth of 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dryseasons and drought periods, that is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).
+Their hydrological model found that in the multi-annual drought deep water was exhausted for over two years, whereas our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period $\Psi_{soil}$ below the depth of 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, (median; `r round(as.numeric(psi.1.7.min$q2.5), 2)`, 95% CI) through all the dryseasons and drought periods, that is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).
 The difference in our results are likely attributable to site differences in the precipitation regime.
@@
 We note that the variation in mortality explained by in our study is low (*R*^2^ = 0.11, *P* = 0.075), and other causes may explain observed mortality.
-For example, based on a year-long study, lightening explained ~4.5% of mortality for trees >=10 cm dbh and ~40% for trees >=60 cm dbh [@Yanoviak:2019].
+For example, based on a year-long study at BCI, lightening explained ~4.5% of mortality for trees >=10 cm dbh and ~40% for trees >=60 cm dbh [@Yanoviak:2019].
 
@@
 
-#### Future Directions for Model Development
+### Future Directions for Model Development
 \
@@
 S\@ref(fig:other-erd-mods)).
-We interpret this result as that VPD and PLC curves may be adequate to explaine growth, but also acknowledge that our estimate of seasonality of leaf fraction is a tentative estimate based on leaf fall data and assumption of leaf presence during the wet season and omits inter-annual variation.
+We interpret this result as that VPD and PLC curves may be adequate to explaine growth, but also acknowledge that our estimate of seasonality of leaf cover is a tentative estimate based on leaf fall data and assumption of leaf presence during the wet season and omits inter-annual variation.
 Future studies could make use of data from new technologies such as drone based monitoring of species-specific leaf area fraction in species-rich forests [@Park.2019].
@@
 VWC), hydraulic conductivity ($K_{soil}$) by depth, soil moisture by depth, stream discharge and evapotranspiration were important for effective calibration of our 1-D hydrological model, and recommend wide-spread data collection of these variables as well as water-table levels in forest-inventory sites to allow for estimation of trees' water environments.
-Finally, developing an inverse ERD model that accounts for movement of water within the soil-plant-atmosphere-continuum (SPAC) would be a step change from the current statistical models of ERD (this study; @Chitra-Tarak.2018).
-Although data needs for parametarizing such models [@Christoffersen.2016tbh; @Sperry.1998] could be greater (e.g.
+Finally, developing an inverse ERD model that accounts for movement of water within the soil-plant-atmosphere-continuum (SPAC) would be a step change from the current statistical models of ERD (this study and @Chitra-Tarak.2018).
+Although data needs for parametarizing such models [@Christoffersen:2016kd2d; @Sperry.1998] could be greater (e.g.
 retention curves for roots), limiting immediate community-wide application, they promise to improve our understanding of plant physiology [e.g.
@@
 
+Yet To be written.
+
 ## Acknowledgements
@@
 
-```{r ksat-profiles, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-ksat-profiles} Generated 5000 curves of soil hydraulic conductivity variation by depth based on Latin Hypercube Sampling within 95% CI of observed data at depths of 12.5 cm and 60 cm [@Kinner:2004], assuming a linear decline between those depths, and values before 12.5 and after 60 cm as that of 12.5 cm and 60 cm, respectively.
+```{r ksat-profiles, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ksat-profiles} Generated 5000 curves of soil hydraulic conductivity variation by depth based on Latin Hypercube Sampling within 95% CI of observed data at depths of 12.5 cm and 60 cm [@Kinner:2004], assuming a linear decline between those depths, and values before 12.5 and after 60 cm as that of 12.5 cm and 60 cm, respectively.
 Only 100 curves are shown for clarity.
-Note the log scale on horizontal axis."}
+Note the log scale on horizontal axis.")}
 knitr::include_graphics(file.path("figures/Bootstrapped Ksat profiles within observed 95CI_random_two_hundred.pdf"))
@@
 ```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts} Extreme soil droughts by census interval (panels) at three representative depths: 0.5, 1 and 1.7 m, depicted in different colors.
-Mean soil water potential ($Psi_z$) for a given DOY and depth z are shown as lines.
-Filled areas show extreme drought periods, when $Psi_z$ declined below 2.5 percentile of $Psi_z$ for the given DOY and depth z in any given year within the cenus intervals.
+Mean soil water potential ($Psi_{soil,~z}$) for a given DOY and depth z are shown as lines.
+Filled areas show extreme drought periods, when $Psi_{soil,~z}$ declined below 2.5 percentile of $Psi_{soil,~z}$ for the given DOY and depth z in any given year within the cenus intervals.
 Extreme drought periods from multiple years may be overlaid within a panel.
-Wet season is shown in gray as beginning with DOY "}
+Wet season is shown in gray as beginning with DOY."}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q2.5_depths_in_inverse_model.jpeg"))
@@
 ```{r iso-data, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-iso-data}Stable isotopic concetration ($\\delta^2H$) of soil water by depth measured in March 1997.
-Data from @Meinzer:1999"}
+Data from @Meinzer:1999."}
 knitr::include_graphics(file.path("figures/UDI_confidence/Meinzer_etal_1999_Depth_vs_soil_deltaD.jpeg"))
@@
 ```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves} Leaf 'vulnerability curves', that is standardized $K_{leaf}$ vs.
-$\\Psi_{leaf}$, fitted to observed data on ",  length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))," tree species from Panama, color coded by Wood Specific Gravity~100^$\circ$^C~ ($gcm^{-3}$).")}
+$\\Psi_{leaf}$, fitted to observed data on ",  length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))," tree species from Panama, color coded by Wood Specific Gravity~100^$\\circ$^C~ ($gcm^{-3}$).")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB_for_data_sp.jpeg"))
@@
 
-```{r erd-growth, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-growth}Scatterplot between Effective Rooting Deoth (ERD) and mean growth rate $\\pm$ (add error) for xx large, canopy species"}
+```{r erd-growth, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-growth}Scatterplot between Effective Rooting Deoth (ERD) and mean growth rate $\\pm$ (add error) for xx large, canopy species."}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/adult_Growth_vs_udi_with_outliers_sp.jpeg"))
@@
 
-```{r erd-mort-anom, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-anom}Mortality anomaly in each census interval during the study period plotted against ERD"}
+```{r erd-mort-anom, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-anom}Mortality anomaly in each census interval during the study period plotted against ERD."}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/diff.mortality_by rdi.gr.jpeg"))
@@
 ```{r symbols, results = 'asis', echo = FALSE}
-knitr::kable(symbols.table, caption = 'The symbol, definition of key hydraulic traits or variables used', booktabs = TRUE, escape = FALSE)
+knitr::kable(symbols.table, caption = 'The symbol, definition of key hydraulic traits or variables used.', booktabs = TRUE, escape = FALSE)
 ```

### 2020-08-05 manuscript.rmd
added error for ERD as 1SE

@@
 \begin{equation}
-Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,~n} StdVPD_{n} + error
+Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,~z,~n} StdVPD_{n} + error
 (\#eq:M2)
@@
 
-```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-iso}Modelled Effective Rooting Depth (ERD; mean $\\pm$ add error; m) versus dry season stable isotopic concentration $\\delta^2H$ (mean $\\pm$ 1SE; ‰) for ", length(unique(erd.iso$sp))," canopy species from Barro Colorado Island.
+```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-iso}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE; m) versus dry season stable isotopic concentration $\\delta^2H$ (mean $\\pm$ 1SE; ‰) for ", length(unique(erd.iso$sp))," canopy species from Barro Colorado Island.
 Data from [@Meinzer:1999].
@@
 
-```{r sp-erd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap= paste0("\\label{lab-sp-erd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for ", length(erd.sp)," large, canopy species of Barro Colorado Island.
+```{r sp-erd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap= paste0("\\label{lab-sp-erd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) for ", length(erd.sp)," large, canopy species of Barro Colorado Island.
 See Table  \\@ref(tab:spcodes) for scientific names corresponding to species codes.")}
@@
 
-```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus hydraulic properties for eight large, canopy species found on Barro Colorado Island (Panama).
+```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus hydraulic properties for eight large, canopy species found on Barro Colorado Island (Panama).
 *R*^2^ and *P*-values are given in panel insets.[X-Axis labels to be added: $K_{max}$ units as in Fig.
@@
 
-```{r erd-mort, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-mort}Mortality rate (mean $\\pm$ 1SE) versus Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for ", length(erd.mrate.sp)," canopy species found on Barro Colorado Island.
+```{r erd-mort, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-mort}Mortality rate (mean $\\pm$ 1SE) versus Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) for ", length(erd.mrate.sp)," canopy species found on Barro Colorado Island.
 *R*^2^ and *P*-values are given in panel insets.")}
@@
 
-```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{soil,~z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for ", length(erd.mrate.sp), " canopy species found on Barro Colorado Island.
+```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{soil,~z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for ", length(erd.mrate.sp), " canopy species found on Barro Colorado Island.
 Each circle represents a species.
@@
 
-```{r erd-growth, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-growth}Scatterplot between Effective Rooting Deoth (ERD) and mean growth rate $\\pm$ (add error) for xx large, canopy species."}
+```{r erd-growth, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-growth}Scatterplot between Effective Rooting Deoth (ERD) and mean growth rate $\\pm$ 1SE for xx large, canopy species."}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/adult_Growth_vs_udi_with_outliers_sp.jpeg"))

### 2020-08-05 manuscript.rmd
minor edit

@@
 The 1,560ha island was formed after the river Chagres was dammed in 1914 that created Gatun Lake, the main channel of the Panama Canal.
-Rainfall at BCI is seasonal with a mean annual rainfall of 2687 mm (± 515 SD over 1983-2017) and a four-month long dry season from mid-December through mid-April (≤ 100 mm rain/month).
+Rainfall at BCI is seasonal with a mean annual rainfall of 2687 mm (± 515 SD; 1983-2017) and a four-month long dry season from mid-December through mid-April (≤ 100 mm rain/month).
 The entire island is forested and classified as a tropical moist forest.

### 2020-08-05 manuscript.rmd
edited source for hydraulic traits figure

@@
 ```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus hydraulic properties for eight large, canopy species found on Barro Colorado Island (Panama).
-*R*^2^ and *P*-values are given in panel insets.[X-Axis labels to be added: $K_{max}$ units as in Fig.
-2.
-Water Potentials are in MPa.]")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/Select_traits_depth.jpeg"))
+*R*^2^ and *P*-values are given in panel insets.")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/erd.stem.traits.jpeg"))
 ```

### 2020-08-07 manuscript.rmd
Incorporated Chonggang's comment; first run

@@
 ```{r, include=FALSE, echo = FALSE}
-if(!require("pacman", quietly=TRUE))install.packages("pacman")
+if(!require("pacman", quietly=TRUE)) install.packages("pacman")
 suppressPackageStartupMessages(
@@
 
-- Deep water access is believed to be an important, drought-tolerance mechanism, but under-studied given the lack of data.
+- Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism.
 Vulnerability to embolism ($\Psi_{88,~stem}$) and safety margins ($\Psi_{min} - \Psi_{88,~stem}$) are shown to be predictors of drought-induced mortality risk, but deep water access may delay plants from reaching these critical thresholds.
-Here we test the role of deep-water access in drought-tolerance strategies of a hyper-diverse tropical forest community (Barro Colorado Island, Panama) using a novel data-model approach.
-- We inversely modelled Effective Rooting Depth (ERD) for large trees of `r length(erd.sp)` canopy tree species by linking species growth dynamics (6 censuses, 1990-2015) to VPD, Soil Water Potential ($\Psi_{soil}$) by depth in the whole-soil column, and a new community-scale model of species vulnerability curves ($K_{leaf}$ vs.
-$\Psi_{leaf}$) that scaled new data on 21 species with Leaf Mass Area~lamina~ and Wood Specific Gravity~100^$\circ$^C~.
-We evaluated ERD by leveraging existing isotopic tracer data ($\delta^2H$) on tree xylem sap and soil column.
-- We find that deeper ERD is associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, but lower mortality rates over 35 years.
-Deeper ERD species spent less time below critical hydraulic thresholds indicating that ERD buffers drought-stress.
-- The role of deep water access in alleviating mortality of drought-sensitive trees have important implications for predicting the future of tropical forests under climate change.
+- We tested the role of deep-water access in survival strategies of a diverse tropical forest community (Barro Colorado Island, Panama) using a novel data-model approach.
+We inversely modelled effective rooting depth (ERD) for large trees of `r length(erd.sp)` canopy tree species by linking species growth dynamics (6 censuses, 1990-2015) to VPD, Soil Water Potential ($\Psi_{soil}$) by depth in the whole-soil column, and a new community-scale model of species vulnerability curves ($K_{leaf}$ vs.
+$\Psi_{leaf}$).
+We evaluated our ERD estimates with existing isotopic tracer data ($\delta^2H$) on tree xylem sap and soil column.
+- We find that deeper ERD is associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates over 35 years.
+Deeper ERD species spent less time below critical hydraulic thresholds indicating that plants mitigate mortality risk from water stress through deep water access.
+- The role of deep water access in alleviating mortality of drought-sensitive trees has important implications for our predictive understanding of tropical forest dynamics under current and future climate.
 
@@
 \
-Intensifying droughts threaten tropical forests, a biome that plays a disproportionately large role in the global carbon and energy cycle, and supports half of the global biodiversity [@Bonan:2008; @Wright:20050th].
-Tropical forests are considered an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw]: here, historic climatic variation is relatively small [@Mora.2013], and the observed ranges of trees' hydraulic safety margins are relatively narrow [@Choat:2012; @Oliveira:2019].
+Tropical forests play a disproportionately large role in the global carbon and energy cycle, and support half of global biodiversity [@Bonan:2008; @Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019].
+Tropical forests are considered an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw].
+Historic climatic variation is relatively small [@Mora.2013], and the observed ranges of trees' hydraulic safety margins are relatively narrow [@Choat:2012; but see @Oliveira:2019].
 Widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010].
-However, state-of-the-art dynamic vegetation models (DVMs) fail to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013], as underlying mechanisms of drought-tolerance are not fully understood.
+However, state-of-the-art dynamic vegetation models (DVMs) fail to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013], as underlying mechanisms of drought-tolerance are not fully understood or quantified.
 
@@
 Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism.
-This mechanism assumes that water availability varies belowground by vertical depth and so do rooting depths of co-occuring trees, by species and size.
-Consequently trees' in the same forest differ in their drought experience, and thus in their responses in term of growth and mortality.
+This mechanism assumes that water availability varies belowground by vertical depth and so do rooting depths of co-occuring trees, by species and size [@Meinzer:1999; @Chitra-Tarak.2018].
+Consequently, trees' in the same forest differ in their drought experience, and thus in their responses in term of growth and mortality.
 A key bottlneck in community-wide testing of this mechanism has been lack of data on trees' rooting or water-sourcing depths, followed by data on water-availability at those depths.
-A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the World’s 65 thousand tree species [@Fan:2017; @Beech:2017].
-Further, only a small fraction of those are tropical, even though ~60-80 % (40 - 53 throusand species) of the World’s tree diversity resides in the tropics [@Slik:2015].
+A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the world’s 65 thousand tree species [@Fan:2017; @Beech:2017].
+Further, only a small fraction of those are tropical, even though ~60-80 % (40 - 53 thousand species) of the World’s tree diversity resides in the tropics [@Slik:2015].
 Use of stable isotopes of water as a tracer provides an indirect way of trees' water sourcing depths by matching isotopic value in plant tissue to variations in soil pore water vertically belowground.
-Such data are available only for ~0.05% of World's tree species [@Evaristo:2016].
+However, such data are extremely rare [@Evaristo:2016].
 In general, community-scale data collection for rooting or water-sourcing depths is especially challenging for species-rich tropical forests.
-Further, to estimate plant available water with soil depth requires hydrological data and modeling, but these are rarely co-located with forest inventory sites (but see, @Johnson:2018fad; @Kumagai:2011; @Ivanov:2012)
 
 \
-Establishing relationships between plant's environment, traits and demographic outcomes is imperative for a predictive plant ecology [@Mcgill.2006], but plant's realised water environments are rarely studied.
-Huge efforts are devoted to linking the same-for-all-trees drought indices, above-ground traits and demographic rates.
-These efforts have resulted in identification of above-ground traits such as proxies of hydraulic failure as indicators of mortality risk.
-A recent meta-analysis found greater drought-induced mortality rates in species with cavitation-vulnerable xylem ($\Psi_{88,~stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~stem}$) [@Anderegg:2016], but counter examples also exist [@Hoffmann:2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013]; indicating whether plants reach critical thresholds of hydraulic failure depends on their realised drought intensity (see Table \@ref(tab:symbols) for trait symbols and their definitions).
-A rare study that accounted for realised water availability for four tree species in a centennial drought (Texas, USA) found that mortality decreased with deep rootedness even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi_{88,~stem}$) and had narrower safety-margins ($\Psi_{min} - \Psi_{88,~stem}$) [@Johnson:2018fad].
-A species sourcing deep water in a Karst system (Italy) showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi_{88,~stem}$ and narrower safety margin compared to two species sourcing shallow-water [@Nardini:2015].
-@Chitra-Tarak.2018 proposed the first community-scale model that linked trees' realised water availability to demographic outcomes through droughts.
-They found that species accessing deep-water in a tropical dry forest (India) died more in a rare, centennial drought compared to shallow-water species.
-They concluded that this was consistent with an 'aquisitive-conservative' species strategy spectrum.
-Aquisitives relied on stable water resource at depth, which exhausted in the multi-annual drought, resulting in reduced survival, whereas conservative species used fluctuating shallow-water and survived the drought as some shallow-water was available via infiltration.
-However, lack of data on rooting or water-sourcing depths and hydraulic traits precluded this study from corroborating modelled estimates of water-uptake depths, or investigating trait coordination.
+Trade-offs among co-occurring plants shape how they make a living in their environment (ecological strategies) and also provide the essential constraints to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d].
+This entails indentifying trade-offs among environment, traits and demography.
+Specifically for the water environment (or resource) observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified an acquisitive-conservative resource economic spectrum similar to the ones for light and nutrients, although, for water, examples of all trade-offs observed simultaneously are rare, and belowground traits or environment linkages are weak (@Reich:2014; @Sterck:2011).
+Acquisitive species have traits with high capacity to acquire, move, store and use water (high $K_{max}$, low wood density) and achieve high growth potential, but build flimsy, disposable tissues (high SLA) and are less tolerant of low resources, whereas species with conservative traits are better protected from carbon losses and are more drought tolerant (low $\Psi_{50}$, low $\Psi_{tlp}$, larger safety margin).
+Acquisitive species are spatially associated with wet sites (high resource) such as valleys or those with high rainfall [@Kursar:2009; @Oliveira:2019], but whether and how the acquisitive-conservative spectrum holds true for the moisture dynamics vertically belowground is not clear.
+Deep rootedness may be expected to support an acquisitive strategy with a steady water supply unlike shallow-rooted trees.
+However, this may depend on the exact dynamics of water-availability by depth at a site, depending on precipitation regime and hydrology.
+In contrast, in pulse-driven arid systems *shallow*-rooted plants are associated with acquisitive traits (high $K_{max}$) that maximise the use of the water-pulse, while *deep*-rooted plants show conservative traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013].
 
 \
-Trade-offs shape how trees' make a living in their environment (ecological strategies) and also provide Dynamic Vegetation Models (DVMs) the essential constraints to represent species-rich communities under changing environment [@Koven:2020].
-This entails indentifying trade-offs among environment, traits and demography.
-@Reich:2014 concenptualised a universal 'fast-slow' (aka productive-persistent, or acquisitive-conservative) plant economics spectrum coordinated for all resources and organs, both above- and below-ground.
-specifically for water observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified a fast-slow spectrum, although examples of all trade-offs observed simultaneously are rare, and belowground trait or environment linkages are weak.
-'Fast' species at the productive end of the spectrum have traits with high capacity to acquire, move, store and use water (high $K_{max}$, low wood density) and achieve high growth potential, but build flimsy, disposable tissues (high SLA) and are less tolerant of low resources, whereas species with slow traits are better protected from C losses and are more drought tolerant (low $\Psi_{50}$, low $\Psi_{tlp}$, larger safety margin).
-Fast strategy is associated with spatially wet sites such as valleys or high rainfall [@Kursar:2009; @Oliveira:2019], but whether and how the fast-slow spectrum holds true for the moisture dynamics vertically belowground is not clear.
-Deep rootedness may be expected to support a fast strategy with a steady water supply unlike shallow-rooted trees.
-However, this may depend on the exact dynamics of water-availability by depth at a site, depending on precipitation regime and hydrology.
-In contrast, in pulse-driven arid systems *shallow*-rooted plants are associated with fast traits (high $K_{max}$) that maximise the use of the water-pulse, while *deep*-rooted plants show slow traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013].
-
-\
-Here, for large trees of length(erd.sp) canopy species of a hyperdiverse neo-tropical moist forest (BCI, Panama) we make a thorough assessment of trade-offs between belowground water environment, above-ground hydraulic traits, estimates of effective rooting depths and demographic outcomes over 34 years that withnessed droughts of various intensity and duration.
-We obtained daily dynamics of plant water potential ($\Psi_{soil,~z}$) by depth $z$ in the whole-soil column with a locally parametarised and calibrated 1-D hydrological water balance model.
-Developing a novel inverse model allowed us to estimate species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015).
-This model incorporated impact of atmospheric drought (VPD) on growth and was constrained by above-ground hydraulic traits, namely, species leaf vulnerability curves ($K_{leaf}$ vs.
-$\Psi_{leaf}$).
-We collected new data on vulnerability curves for xx species.
-Identifying scaling relationships between these curves and commonly available traits (SLA, specific wood density) allowed us to scale-up these curves to `r length(data.model.AB$sp)` tree species in the community.
-We evaluated ERD against existing data on plant and soil water isotopes ($\delta^2H$) and coordination of ERD with hydraulic traits and mortality rates.
-
-\
-We test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns species on a fast-slow economic spectrum optimised for increasingly reliable water availability by depth such that (i) trees with deeper ERD have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem: higher maximum stem hydraulic conductance ($K_{max}$), but greater xylem vulnerability to embolism ($\Psi_{88,~stem}$), higher $\Psi_{tlp}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$), whereas shallower ERD species are associated with less efficient but safer hydraulic traits; and that, (ii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure, whereas shallower ERD species may have higher mortality rates as a result of higher variability in shallow soil water availability.
+Huge efforts are invested in linking the same-for-all-trees drought indices, above-ground traits and demographic rates.
+These efforts have resulted in identification of above-ground traits such as proxies of hydraulic failure as indicators of mortality risk.
+A recent meta-analysis found greater drought-induced mortality rates in species with cavitation-vulnerable xylem ($\Psi_{88,~stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~stem}$) [@Anderegg:2016], but counter examples also exist [@Hoffmann:2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013]; indicating that plants reach critical thresholds of hydraulic failure depending on their realised drought intensity (see Table \@ref(tab:symbols) for trait symbols and their definitions).
+However, plants' realised water environments are rarely studied, especially in the tropical region.
+To better understand how belowground rooting is linked to aboveground hydraulic traits and mortality for co-occuring tropical forest tree species, we make a thorough assessment of trade-offs between the belowground water environment, above-ground hydraulic traits, estimates of effective rooting depths and demographic outcomes over 34 years that witnessed droughts of various intensity and duration for a tropical site.
+Specifically, we test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns species on an acquisitive-conservative economic spectrum optimised for increasingly reliable water availability by depth such that (i) trees with deeper ERD have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem: higher maximum stem hydraulic conductance ($K_{max}$), but greater xylem vulnerability to embolism ($\Psi_{88,~stem}$), higher $\Psi_{tlp}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$), whereas shallower ERD species are associated with less efficient but safer hydraulic traits; and that, (ii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure, whereas shallower ERD species may have higher mortality rates as a result of higher variability in shallow soil water availability.
 To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.
@@
 
+In summary, we obtained daily dynamics of plant water potential ($\Psi_{soil,~z}$) by depth $z$ in the whole-soil column with a locally parametarised and calibrated 1-D hydrological water balance model.
+Developing a novel inverse model allowed us to estimate species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015).
+This model incorporated impact of atmospheric drought (VPD) on growth and was constrained by above-ground hydraulic traits, namely, species leaf vulnerability curves ($K_{leaf}$ vs.
+$\Psi_{leaf}$).
+We collected new data on vulnerability curves for 21 species.
+Identifying scaling relationships between these curves and commonly available traits (Leaf Mass Area~lamina~, Wood Specific Gravity~100^$\circ$^C~; defined below) allowed us to scale-up these curves to `r length(data.model.AB$sp)` tree species in the community (of ~ 320 total).
+We evaluated ERD against existing data on plant and soil water isotopes ($\delta^2H$) and coordination of ERD with hydraulic traits and mortality rates.
+
 ## Study Site Description
@@
 The 1,560ha island was formed after the river Chagres was dammed in 1914 that created Gatun Lake, the main channel of the Panama Canal.
-Rainfall at BCI is seasonal with a mean annual rainfall of 2687 mm (± 515 SD; 1983-2017) and a four-month long dry season from mid-December through mid-April (≤ 100 mm rain/month).
+Rainfall at BCI is seasonal with a mean annual rainfall of 2687 mm (± 515 SD; 1983-2017) and a four-month long dry season from mid-December through mid-April ($\le$ 100 mm rain/month).
 The entire island is forested and classified as a tropical moist forest.
@@
 This inventory represents 321 woody species across the eight censuses.
-Aside from a 2ha patch of ~100 year old secondary forest at the north-eastern corner of the plot the remaining 48 ha represent old growth forest that was never clear cut for agriculture.
+Aside from a 2ha patch of ~100-year-old secondary forest at the north-eastern corner of the plot the remaining 48 ha represent old growth forest that was never clear cut for agriculture.
 The plot (100 m x 50 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001].
@@
 Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding.
-Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere.
+Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere[xx add references].
 
@@
 FATES is a state-of-the-art, community-based, open-source model used for studying climate-vegetation interactions.
-FATES is considered a next generation vegetation dynamics model for ELM, with a size-structured group of plants (cohorts) and successional trajectory-based patches using the ecosystem demography approach......
-(How ELM talks to FATES)...Detailed descriptions can be found elsewhere [@Koven:2020; @Fisher:2014; @Bonan:2013].
+FATES is considered a next generation vegetation dynamics model for ELM, with a size-structured group of plants (cohorts) and successional trajectory-based patches using the ecosystem demography approach.
+FATES couples to ELM by a common interface of water and carbon fluxes.
+Detailed descriptions can be found elsewhere [@Koven:2020; @Fisher:2014; @Bonan:2013].
 
@@
 We also leveraged existing opportunistic dry season VWC measurements for the entire 50-ha plot (15, 40 and 100 cm in 2015 and 15 & 40 cm in 2016; @Kupers:2019).
-Among the 5000 parameter ensembles we chose 100 ensembles that best-fit all three fluxes simultaneously (add details...that is those that maximised the sum of regression coefficients between observed and simulated variables.) See Table \@ref(tab:param) for best-fit parameter ranges.
+Among the 5000 parameter ensembles we chose 100 ensembles that best-fit all three fluxes simultaneously (add details...that is those that maximised the sum of regression coefficients between observed and simulated variables.
+xx add equation) See Table \@ref(tab:param) for best-fit values and ranges for different parameters.
 We ran ELM-FATES with the chosen 100 ensembles from 1985-2018, with first 5 years used for model spin-up.
@@
 We measured ($K_{leaf}$ vs.
-$\Psi_{leaf}$) on adult trees of `r length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))` common species (see SI x).
+$\Psi_{leaf}$) on adult trees of `r length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))` common species (see SI--Joseph please add a section).
 We fitted exponential, sigmoidal, logistic equations to species-level data as prescribed by @Guyot:2011.
-Exponential form was chosen as it best-fitted all but one species based on AIC criteria (Fig.
-S\@ref(fig:obs-plc-curves)), described as
+Exponential form was chosen, because based on AIC criteria it best-fitted all but one species, and for the one species it fitted only slightly worse (Fig.
+S\@ref(fig:obs-plc-curves); we might want to add specie-specific panel of graphs showing data and fitted curves here), described as
 
@@
 where $A$ and $B$ are fitted species-specific parameters.
-
 We identified scaling relationships between fitted parameters $A$ and $B$ in Eq.
-\@ref(eq:vc-exp) and two traits that were available at community-scale, namely, $WSG$ (or, WSG~100^$\circ$^C~) and $LMA$ (or, LMA~lamina~), defined, respectively, as mean wood specific gravity after drying at 100$^\circ$C (gcm^-3^) and mean leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules (gm^-2^) for leaves receiving direct sunlight.
+\@ref(eq:vc-exp) and two traits that were available at community-scale, namely, mean wood specific gravity after drying at 100$^\circ$C ($WSG$ or, WSG~100^$\circ$^C~; gcm^-3^) and mean leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules ($LMA$ or LMA~lamina~; gm^-2^) for leaves receiving direct sunlight.
 Fitted polynomial equations are described as,
@@
 where $(a_0, a_1, a_2, a_3, b_0, b_1, b_2, b_3)$ are coefficients.
-$WSG$ & parameter $B$ explained large variation in parameter $A$ (Eq.
+$WSG$ & parameter $B$ explained a large proportion of variation in parameter $A$ (Eq.
 \@ref(eq:a-mod); Adj.
@@
 S\@ref(fig:ab-fit)).
-$WSG$ & $LMA$ explained large variation in parameter $B$ (Eq.
+$WSG$ & $LMA$ explained a large proportion of variation in parameter $B$ (Eq.
 \@ref(eq:b-mod); Adj.
@@
 $\\Psi_{leaf}$) at community-scale (", length(data.model.AB$sp)," species) based on data on $WSG$ & $LMA$ and their relationships with vulnerability curve parameters measured on ", sp.soft.filled$AB.data.model," species.
-Each line represents a species, color coded by $WSG$.")}
+Each line represents a species, color coded by $WSG$ with blue indicating a smaller value of WGS and red indicating a larger value of WSG.[xx chagne legend to a scale not just two values.]")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB.jpeg"))
@@
 Water is a key limiting resource for tree growth.
-Vapour pressure deficit (VPD) and species-specific maximum hydraulic conductivity and its decline with water-stress ($K_{leaf}$ vs.
-$\Psi_{soil}$), are factors known to influence trees water uptake [add references, Chonggang?].
+Vapour pressure deficit (VPD) and water supply as determined by soil moisture are factors known to influence trees water uptake [@Lawrence.2019].
+In this study, we used leaf conductivity ($K_leaf$) determined by rhizosphere soil water potential ($\Psi_{soil}$) as an approximation of the soil water limiting factor.
 For dry-deciduous species seasonality in leaf cover may also influence plant water uptake.
 We therefore tested alternate models of growth, aka inverse models for ERD (Eqs.
-\@ref(eq:M2)- \@ref(eq:M4)), in which a species growth dynamics across census intervals was regressed against census interval means of multiplied effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the fraction of leaf cover (all standardised to 0-1, see below).
-Let n~t~ be the total number of days in census interval $t$, then for individual tree species Growth~t~ and depth $z$ these models are described as:
+\@ref(eq:M2)- \@ref(eq:M4)), in which a species' growth dynamics across census intervals was regressed against census interval means of multiplied effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the fraction of leaf cover (all standardised to 0-1, see below).
+Let n~t~ be the total number of days in census interval $t$, then for individual tree species Growth~t~ and depth $z$ these models are described as follows (xxadd additive models):
 
@@
 
-To minimise effect of light variation among trees and census intervals we selected only large trees (>= 30 cm dbh, diameter at breast height) and species whose maximum height was at least 30 m (hereafter, canopy species) that are likely to be fully exposed to the sun.
+To minimise the effect of light variation among trees, we selected only large trees (>= 30 cm dbh, diameter at breast height) and species whose maximum height was at least 30 m (hereafter, canopy species) that are likely to be fully exposed to the sun.
 We chose only those trees for which diameter (dbh) record was present across all size censuses (1990, 1995, 2000, 2005, 2010, 2015).
-We calculated individual tree ($i$) growth rates ($Gr_{i,t}$; cm.yr^-1^) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses, divided by the inter-census interval based on the exact date of measurements.
+For each individual tree (i), we calculated growth rates ($Gr_{i,t}$; cm.yr^-1^) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses, divided by the inter-census interval based on the exact date of measurements.
 As tree growth rates increase with size, to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($Resid_{i,t} = Gr_{i,t} - f(DBH_{i, t})$).
-The diameter model of growth was fitted based on all trees' data pooled together as there was insufficient species-specific data for large trees (Do I need to add SI figure?).
+The diameter model of growth was fitted based on all trees' data pooled together as there was insufficient species-specific data for large trees (xx Add SI figure showing the model).
 We then standardized (centered and scaled) an individual tree time series ($Resid_{i,t}$) and obtained species-level (*n* = `r length(erd.sp)`) times-series of medians ($Growth_{t}$) that had data on at least three trees (median = xx).
 
-As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between GPP (gCm^2^day^-1^) data we collected from the Eddy Flux Tower and VPD (kPa; $\bar{GPP} = m_2VPD^2 + m_1VPD + m_0 + error = StdVPD$; Do I need to add SI Fig?) and used standardised $StdVPD$ (range 0-1) in Eqs.
+There are many factors that could affect plant growth, including soil moisture, VPD, leaf area, temperature and radiation.
+Based on our preliminary analyses, incorporation of temperature and radiation did not improve our model fitting, thus we mainly focused on the impact of soil moisture, VPD and leaf area in our models.
+
+As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between GPP (gCm^2^day^-1^) data we collected from the Eddy Flux Tower and VPD (kPa; $\bar{GPP} = m_2VPD^2 + m_1VPD + m_0 + error = StdVPD$; xxAdd SI Fig) and used standardised \bar{GPP} to represent the impact of VPD on growth ($StdVPD$; range 0-1) in Eqs.
 \@ref(eq:M2)- \@ref(eq:M4).
+Due to coincidence of higher temperature, higher radiation, and lower humidity during the dry season at our study site, the VPD impact may as well be a combined impact of temperature, humidity, and radiation.
 
@@
 For non-evergreen species, we cumulatively deducted $LeafFall$ from $LeafCover$ = 1 beginning DOY = 300 and obtained a $LeafCover$ by DOY curve (\@ref(fig:leaf-cover)).
+We obtained species-specific parameters $A$ & $B$ for Eq.
+\@ref(eq:k-soil) either from data when available, or a model based on $LMA$ or $WSG$ as described in the earlier section.
 
-We obtained species-specific parameters $A$ & $B$ for Eq. \@ref(eq:k-soil) either from data when available, or a model based on $LMA$ or $WSG$ as described in the earlier section.
-
-For each model \@ref(eq:M1)- \@ref(eq:M4) and species combination ($sp$), we obtained goodness-of-fit (*R*^2^) and Pearson's $\rho$ for all soil depths $z$, and for each sp and model combination we chose the $z$ with highest *R*^2^ as the Effective Rooting Depth (ERD), only for species with positive Pearson's $\rho$ and *R*^2^ >= 0.1.
+For the combination of each model \@ref(eq:M1)- \@ref(eq:M4), species and hydrological realisation ($sp$), we obtained AIC, goodness-of-fit (*R*^2^) and Pearson's $\rho$ for each soil depth $z$.
+For each model-species-hydrological realisation we retained only one depth each with the least AIC as the best-fit, and retained only the subset of depths that had a positive Pearson's $\rho$ and *R*^2^ >= 0.5.
+For each model-species, we defined Effective Rooting Depth (ERD) as the median of this subset of depths and calculated standard error.
 
@@
 S\@ref(fig:iso-data).
-We also removed *Guapira* *standleyana*, whose xylem sap isotope value was not constrained by the soil water isotope depth profile.
+We also removed *Guapira* *standleyana*, whose xylem sap isotope value was not well constrained by the soil water isotope depth profile.
+
+## ERD Coordination with Above-ground Hydraulic Traits
+
+To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD with maximum stem hydraulic conductance ($K_{max, stem}$), leaf turgor loss point ($\Psi_{tlp}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$).
+See SI for data collection and estimation of these variables (--Brett please add a section).
+
+## Mortality Analyses and Extreme Drought Experience
+
+We calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$ 30 m) and average adbunance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; *n* = `r length(erd.mrate.sp)`).
+For each of the seven census intervals ($t$) in the 34-yr record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successive censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{time}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50ha plot of a species present in cen~1~ and dead in cen~2~, respectively, and $time$ is the duration based on mean dates of cen~1~ and cen~2~.
+We calculated species mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{t=1}^{6} {Mort_{t}}$ and, finally, species mortality anomalies by intervals as $Mort_{anomaly}$ = $Mort_{t}$ - $Mort_{mean}$.
+To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$).
+To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{ERD}$) were below species $\Psi_{crit}$, defined as $\Psi_{80,~leaf}$.
+
+All statistical analyses were done in R statistical environment (version 3.5.3, @R:2019).
+
+# Results
+
+## Effective Rooting Depths, Vulnerability to Embolism and Hydraulic Safety Margins
 
 \
-The model with VPD and without Leaf Cover (Eq.
+In comparison to the measured isotopic signal, our results showed that the model with VPD and without Leaf Cover (Eq.
 \@ref(eq:M2)) had highest goodness-of fit (*R*^2^ = 0.51; *p* = 0.072), followed by followed by model without both VPD and Leaf Cover (Eq.
@@
 
-## ERD Coordination with Above-ground Hydraulic Traits
-
-To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD with maximum stem hydraulic conductance ($K_{max, stem}$), leaf turgor loss point ($\Psi_{tlp}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$).
-See SI for data collection and estimation of these variables.
-
-## Mortality Analyses and Extreme Drought Experience
-
-We calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$ 30 m) and average adbunance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; *n* = `r length(erd.mrate.sp)`).
-For each of the seven census intervals ($t$) in the 34-yr record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successive censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{time}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50ha plot of a species present in cen~1~ and dead in cen~2~, respectively, and $time$ is the duration based on mean dates of cen~1~ and cen~2~.
-We calculated species mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{t=1}^{6} {Mort_{t}}$ and, finally, species mortality anomalies by intervals as $Mort_{anomaly}$ = $Mort_{t}$ - $Mort_{mean}$.
-To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$).
-To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{ERD}$) were below species $\Psi_{crit}$, defined as $\Psi_{80,~leaf}$.
-
-All statistical analyses were done in R statistical environment (version 3.5.3, @R:2019).
-
-# Results
-
-## Effective Rooting Depths, Vulnerability to Embolism and Hydraulic Safety Margins
-
 \
-In this forest modelled Effective Rooting Depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from 0.5 m to 13 m (Fig.
+In this forest modelled effective rooting depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from 0.5 m to 13 m (Fig.
 \@ref(fig:sp-erd)).
@@
 \
-Our analyses revealed that co-occurring large canopy tree species (*n* = `r length(erd.sp)`) with deeper effective rooting depths (ERD) were associated with higher hydraulic efficiency, but not safety: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species.
-Yet, deeper ERD species showed lower mortality rates than shallower ERD species over the 34-yr record (1982-2015).
+Our analyses revealed that co-occurring large canopy tree species (*n* = `r length(erd.sp)`) with deeper effective rooting depths (ERD) were associated with higher hydraulic efficiency, but not safety.
+Deep ERD species showed higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species.
+This trade-off translated into greater drought tolerance, as deeper ERD species showed lower mortality rates than shallower ERD species over the 34-yr record (1982-2015).
 Analysis of time spent below ($\Psi_{crit}$) in the soil zone matching ERD $\Psi_{soil,~z = ERD}$ showed that this was likely due to decreasing realised risk of hydraulic failure with depth.
-We thus show, for the first time, that interspecific variation in trees' effective rooting depths aligns species on a 'fast-slow' water economic spectrum by demonstrating a link between trees' realised drought experience, hydraulic architecture and long-term mortality outcomes.
-On the 'fast' end of the spectrum were deep ERD species with hydraulically efficient, but risky suit of traits optimised for an environment of reliable and abundant source of water at depth that actualised to a lower mortality-risk.
-At the slow end of the spectrum were shallow ERD species with inefficient, but safer xylem, likely optimised for high variability in $\Psi_{soil}$ in the shallow soil layers, resulting in higher variance in mortality rates across censuses as $\Psi_{soil,~z = ERD}$ crossed ERD $\Psi_{crit}$.
+We thus show, for the first time, that interspecific variation in trees' effective rooting depths aligns species on a conservative-acquisitive water economic spectrum by demonstrating a link between trees' realised drought experience, hydraulic architecture and long-term mortality outcomes.
+On the acquisitive end of the spectrum were deep ERD species with hydraulically efficient, but risky suit of traits optimised for an environment of reliable and abundant source of water at depth that actualised to a lower mortality-risk.
+At the conservative end of the spectrum were shallow ERD species with inefficient, but safer xylem, likely optimised for high variability in $\Psi_{soil}$ in the shallow soil layers, resulting in higher variance in mortality rates across censuses as $\Psi_{soil,~z = ERD}$ crossed ERD $\Psi_{crit}$.
 Mortality rates of shallow ERD species may have been exacerbated by increased drought events during the study period.
@@
 Our study brings attention to the need for assessing drought-sensitivity in terms of species realised drought risk.
-We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins traits, that are identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through a study period of greater recurrence of drought.
+We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins traits, which are commonly  are identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through a study period of greater recurrence of drought.
 These results are consistent with the rare studies that studied three or four species each with known rooting or water sourcing depths and assesed their hydraulic traits and mortality rates during a centennial (Texas) or extreme (Italy) drought [@Johnson:2018fad; @Nardini:2015].
+@Johnson:2018fad found that mortality decreased with deep rooting even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi_{88,~stem}$) and had narrower safety-margins ($\Psi_{min} - \Psi_{88,~stem}$).
+@Nardini:2015 found that a species sourcing deep water in a Karst system (Italy) showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi_{88,~stem}$ and narrower safety margin compared to two species sourcing shallow-water.
 Our result of lower mortality in deep rooted trees contrasts with the finding of @Chitra-Tarak.2018, who found that species identified by their inverse ERD model as effectively deep rooted died *more* in a prolonged, centennial drought.
@@
 We note that the variation in mortality explained by in our study is low (*R*^2^ = 0.11, *P* = 0.075), and other causes may explain observed mortality.
-For example, based on a year-long study at BCI, lightening explained ~4.5% of mortality for trees >=10 cm dbh and ~40% for trees >=60 cm dbh [@Yanoviak:2019].
-
-The trade-offs that we identify between above- and below-ground traits, water environment vertically belowground and mortality are important for representing diversity in Dynamic Global Vegetation Models especially in species-rich forests.
-The hydraulic efficiency-safety trade-off between high values of maximum leaf or stem conductivity (high $K_{max}$) and cavitation-resistance (low $\Psi_{50}$ or $\Psi_{88}$) is hypothesised although community-wide examples are rare [@Hacke:2006; @Gleason:2016].
-We report such a trade-off at both leaf (*n* = 21 species) and branch level (*n* = 27 species, but data measured on trees in SLZ and PNM, only include those species that are found in BCI?; Or shall I add those species numbers and a SI figure for only those species for which these data are available for both leaf and branch?).
-We also found that LMA~lamina~ and WSG~100^$\circ$^C~ were strong predictors of PLC curves for leaf, but not branch (add SI figures).
-High $K_{max}$ is elsewhere associated with photosynthetic rates with faster growth potential, but we did not find a relationship between $K_{max}$ and growth rates (*n* = 21 species).
 
@@
 Our chosen ERD model is an advancement over the only other inverse model of ERD (or water uptake depth index, @Chitra-Tarak.2018, *n* = 12 species) in terms of physiologically meaningful representation, fewer parameters, corroboration with tree and soil data on $\delta^2H$.
-Model by @Chitra-Tarak.2018 estimated six species-specific phenomenological parameters: three as root fraction for corresponding lumped vadose zone layers, and three as shape parameters for water uptake vs availability curves.
+The model in @Chitra-Tarak.2018 estimated six species-specific phenomenological parameters: three as root fraction for corresponding lumped vadose zone layers, and three as shape parameters for water uptake vs availability curves.
 For each species, they calculated daily water uptake successively from three layers--the supply, to satisfy the demand of the daily maximum evapo-transpiration, a function of PET and LAI.
@@
 We also note the availability of data on $\delta^2H$ on trees and water sources belowground--soil/substrate matrix water in the vadoze zone or free water below the water table--as important for model calibration/validation.
-Model by @Chitra-Tarak.2018 were not corroborated with such data as existing methods were inadequate to extract water bound to the soil and saprolite found in their dry forest study site (Jean Riotte, pers.
+The model in @Chitra-Tarak.2018 were not corroborated with such data as existing methods were inadequate to extract water bound to the soil and saprolite found in their dry forest study site (Jean Riotte, pers.
 comm.).
-Our model of belowground water-availability (ELM-FATES) was also advanced compared to that used by @Chitra-Tarak.2018 and was calibrated with data on soil moisture dynamics by depth (15, 40, 100 cm), as well as, evapotranspiration and stream discharge, unlike theirs, which was calibrated on stream discharge and water table levels.
+The model of belowground water-availability in this study (ELM-FATES) is far more advanced compared to that used by @Chitra-Tarak.2018 and was calibrated with data on soil moisture dynamics by depth (15, 40, 100 cm), as well as, evapotranspiration and stream discharge.
+@Chitra-Tarak.2018 was calibrated only on stream discharge and water table levels.
 We lacked data on water table levels which would have been useful to constrain water availability dynamics in the unobserved soil layers.
@@
 \
-Our model estimated ERD as a single value, unlike the dynamic ERD modelled by @Chitra-Tarak.2018, but future model developments should investigate seasonal dynamics of ERD, calibrated on corresponding calibration data on isotopic water sourcing depths.
-Our inverse ERD model was parametarised on 5-yearly growth data, with five data points over a 25-yr period, which provided large degree of freedom to fit the model.
-Future studies could better constrain the model with higher resolution growth data such as those from dendrometers.
+The trade-offs that we identify between above- and below-ground traits, water environment vertically belowground and mortality are important for representing diversity in Dynamic Global Vegetation Models especially in species-rich forests.
+The hydraulic efficiency-safety trade-off between high values of maximum leaf or stem conductivity (high $K_{max}$) and cavitation-resistance (low $\Psi_{50}$ or $\Psi_{88}$) is hypothesised although community-wide examples are rare [@Hacke:2006; @Gleason:2016].
+We report such a trade-off at both leaf (*n* = 21 species) and branch level (*n* = 27 species).
+We also found that LMA~lamina~ and WSG~100^$\circ$^C~ were strong predictors of PLC curves for leaf, but not branch (add SI figures).
+High $K_{max}$ is elsewhere associated with photosynthetic rates with faster growth potential, but we did not find a relationship between $K_{max}$ and growth rates (*n* = 21 species).
+These relationships between ERD and aboveground hydraulic traits will provide an important insight on how to model the rooting depth and its coordination or trade-off with other traits to better represent the functional diversity in tropical forests.
+
+\
+Our model estimated ERD as a single value, unlike the dynamic ERD modelled by @Chitra-Tarak.2018, but future model developments should investigate seasonal dynamics of ERD, using calibrated on corresponding calibration data on isotopic water sourcing depths.
+Our inverse ERD model was parametarised on 5-yearly growth data, with five data points over a 25-yr period, which decouples climate events and demographic outcomes.
+Future studies could better constrain the model with higher resolution growth data such as those from dendrometer bands.
 This is especially important for modeling seasonally dynamic ERD.
@@
 Future studies could make use of data from new technologies such as drone based monitoring of species-specific leaf area fraction in species-rich forests [@Park.2019].
-In forests such as BCI where understory tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes, especially from a few growth data points, may need accounting for variation in trees' light environment over time and among trees' of the same size class, an issue we minimised by limiting ERD estimates to canopy trees.
+In forests such as BCI where understory tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes, especially from a few growth data points, may need accounting for variation in trees' light environment over time and among trees of the same size class, an issue we minimised by limiting ERD estimates to canopy trees.
 To estimate topographic variation in ERD within or across species future studies may use soil moisture dynamics from a distributed hydrological model [e.g.
@@
 retention curves for roots), limiting immediate community-wide application, they promise to improve our understanding of plant physiology [e.g.
-see @Johnson:2018fad], but which may be potentially harnessed for a community level scale-up.
+see @Johnson:2018fad].
 
@@
 
-```{r erd-growth, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-growth}Scatterplot between Effective Rooting Deoth (ERD) and mean growth rate $\\pm$ 1SE for xx large, canopy species."}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/adult_Growth_vs_udi_with_outliers_sp.jpeg"))
-```
-
 ```{r erd-mort-anom, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-anom}Mortality anomaly in each census interval during the study period plotted against ERD."}

### 2020-08-07 manuscript.rmd
1. Added PET values 2. PLC curves 3. Join A & B goodness of fit curves

@@
 ---
-title: "Drought-sensitive trees survive on deep water access in a tropical forest community"
+title: "Drought-sensitive trees survive on deep-water access in a tropical forest community"
 author: ''
@@
 - Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism.
-Vulnerability to embolism ($\Psi_{88,~stem}$) and safety margins ($\Psi_{min} - \Psi_{88,~stem}$) are shown to be predictors of drought-induced mortality risk, but deep water access may delay plants from reaching these critical thresholds.
-- We tested the role of deep-water access in survival strategies of a diverse tropical forest community (Barro Colorado Island, Panama) using a novel data-model approach.
-We inversely modelled effective rooting depth (ERD) for large trees of `r length(erd.sp)` canopy tree species by linking species growth dynamics (6 censuses, 1990-2015) to VPD, Soil Water Potential ($\Psi_{soil}$) by depth in the whole-soil column, and a new community-scale model of species vulnerability curves ($K_{leaf}$ vs.
-$\Psi_{leaf}$).
-We evaluated our ERD estimates with existing isotopic tracer data ($\delta^2H$) on tree xylem sap and soil column.
+Vulnerability to embolism and hydraulic safety margins are shown to be predictors of drought-induced mortality risk, but deep-water access may delay plants from reaching these critical thresholds.
+Here, we tested the role of deep-water access in survival strategies of a diverse tropical forest community (Panama) using a novel data-model approach.
+- We inversely modelled effective rooting depth (ERD) for large trees of `r length(erd.sp)` canopy species by linking growth dynamics (1990-2015) to vapour pressure deficit, matric potentials in the whole-soil column, and a new community-scale model of species leaf vulnerability curves.
+We evaluated ERD against existing isotopic tracer data on trees and soil.
 - We find that deeper ERD is associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates over 35 years.
-Deeper ERD species spent less time below critical hydraulic thresholds indicating that plants mitigate mortality risk from water stress through deep water access.
-- The role of deep water access in alleviating mortality of drought-sensitive trees has important implications for our predictive understanding of tropical forest dynamics under current and future climate.
-
-Keywords: rooting depths, drought tolerance, mortality, tropical forest, climate change; life-history strategies
+Deeper ERD species spent less time below critical hydraulic thresholds indicating that trees' mitigate water-stress related mortality risk through deep water access.
+- The role of deep-water access in alleviating mortality of drought-sensitive trees has important implications for our predictive understanding of tropical forest dynamics under current and future climate.
+\
+Keywords: belowground traits, deep-water access, drought-induced mortality, drought tolerance, rooting depths, tropical forest, life-history strategies
 
@@
 Note that only new data on leaf vulnerability curves collected during this study are described below, details of all other datasets used are described in the Supplementary Information.
-
+\
 In summary, we obtained daily dynamics of plant water potential ($\Psi_{soil,~z}$) by depth $z$ in the whole-soil column with a locally parametarised and calibrated 1-D hydrological water balance model.
@@
 The 1,560ha island was formed after the river Chagres was dammed in 1914 that created Gatun Lake, the main channel of the Panama Canal.
-Rainfall at BCI is seasonal with a mean annual rainfall of 2687 mm (± 515 SD; 1983-2017) and a four-month long dry season from mid-December through mid-April ($\le$ 100 mm rain/month).
+Rainfall at BCI is seasonal with a mean annual rainfall of `r round(mean(clim$Precip))` mm ($pm$ `r round(sd(clim$Precip))` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a four-month long dry season from mid-December through mid-April ($\le$ 100 mm rain/month).
+Mean annual Penman-Monteith evapo-transpiration is `r round(mean(clim$pet.PM))` mm ($pm$ `r round(sd(clim$pet.PM))` SD; `r min(clim$Year)`-`r max(clim$Year)`).
 The entire island is forested and classified as a tropical moist forest.
@@
 where $\Psi_e$ is the air-entry matric potential (200, mmH~2~O), $\theta_{s}$ is the saturated volumetric water content (0.51, cm^3^cm^-3^), and $b$, an index for soil pore-size distribution (10, unitless), respectively.
-
+\
 We ran ELM-FATES for the 5000 parameter ensembles with hourly climate drivers from 1985-2018 initialised with the stand structure of the 50ha plot in a single site mode ($0.1^\circ$ x $0.1^\circ$ centered at $9.1530^\circ$N, $280.2539^\circ$E).
@@
 Exponential form was chosen, because based on AIC criteria it best-fitted all but one species, and for the one species it fitted only slightly worse (Fig.
-S\@ref(fig:obs-plc-curves); we might want to add specie-specific panel of graphs showing data and fitted curves here), described as
+S\@ref(fig:obs-vuln-curves); we might want to add specie-specific panel of graphs showing data and fitted curves here), described as
 
@@
 \@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r length(unique(data.model.AB$sp))` species for which WSG and LMA data were available and fitted vulnerability curves (see Fig.
-\@ref(fig:comm-plc-curves)).
+\@ref(fig:comm-vuln-curves)).
 
-```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-plc-curves} Modelled standardized 'Vulnerability curves' ($K_{leaf}$ vs.
+```{r comm-vuln-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-vuln-curves} Modelled 'vulnerability curves' ($K_{leaf}$ vs.
 $\\Psi_{leaf}$) at community-scale (", length(data.model.AB$sp)," species) based on data on $WSG$ & $LMA$ and their relationships with vulnerability curve parameters measured on ", sp.soft.filled$AB.data.model," species.
-Each line represents a species, color coded by $WSG$ with blue indicating a smaller value of WGS and red indicating a larger value of WSG.[xx chagne legend to a scale not just two values.]")}
+Each line represents a species, color coded by $WSG$ with blue tones indicating smaller values of $WSG$ and red tones indicating larger values of $WSG$.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB.jpeg"))
-# knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_LMALAM_AVD_predicted_AB.jpeg"))
 ```
@@
 ## Model for Effective Rooting Depths (ERD)
-
+\
 Water is a key limiting resource for tree growth.
@@
 \@ref(eq:M2)- \@ref(eq:M4)), in which a species' growth dynamics across census intervals was regressed against census interval means of multiplied effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the fraction of leaf cover (all standardised to 0-1, see below).
-Let n~t~ be the total number of days in census interval $t$, then for individual tree species Growth~t~ and depth $z$ these models are described as follows (xxadd additive models):
+Let n~t~ be the total number of days in census interval $t$, then for individual tree species Growth~t~ and depth $z$ these models are described as follows (add additive models xx):
 
@@
 \end{equation}
-
+\
 We estimated daily value of $K_{leaf,z,n}$ by plugging in $\Psi_{soil,~z,~n}$ in the leaf PLC curves Eq. \@ref(eq:vc-exp) as:
 
-\
 \begin{equation}
@@
 \end{equation}
-
+\
 We estimated species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015; see below).
@@
 ### Model parameterization
-
+\
 To minimise the effect of light variation among trees, we selected only large trees (>= 30 cm dbh, diameter at breast height) and species whose maximum height was at least 30 m (hereafter, canopy species) that are likely to be fully exposed to the sun.
@@
 We then standardized (centered and scaled) an individual tree time series ($Resid_{i,t}$) and obtained species-level (*n* = `r length(erd.sp)`) times-series of medians ($Growth_{t}$) that had data on at least three trees (median = xx).
-
+\
 There are many factors that could affect plant growth, including soil moisture, VPD, leaf area, temperature and radiation.
 Based on our preliminary analyses, incorporation of temperature and radiation did not improve our model fitting, thus we mainly focused on the impact of soil moisture, VPD and leaf area in our models.
-
-As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between GPP (gCm^2^day^-1^) data we collected from the Eddy Flux Tower and VPD (kPa; $\bar{GPP} = m_2VPD^2 + m_1VPD + m_0 + error = StdVPD$; xxAdd SI Fig) and used standardised \bar{GPP} to represent the impact of VPD on growth ($StdVPD$; range 0-1) in Eqs.
-\@ref(eq:M2)- \@ref(eq:M4).
+\
+As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between mean daily values by month of gross primary productivity (gCm^2^day^-1^) and VPD we collected from the eddy flux tower (kPa; $\bar{GPP} = -18.8VPD^2 + 14.5VPD + 5.02 + error = StdVPD$; 2012-2017; Fig.
+S\@ref(fig:gpp-vpd)) and used standardised $\bar{GPP}$ to represent the impact of VPD on growth ($StdVPD$; range 0-1) in Eqs.
+\@ref(eq:M2) and \@ref(eq:M4).
 Due to coincidence of higher temperature, higher radiation, and lower humidity during the dry season at our study site, the VPD impact may as well be a combined impact of temperature, humidity, and radiation.
-
+\
 To explore the effect of species seasonality in leaf cover on growth, we made an informed assumption of species-specific seasonal curves for Leaf Cover Fraction ($LeafCover$, range 0-1, unitless).
@@
 We obtained species-specific parameters $A$ & $B$ for Eq.
-\@ref(eq:k-soil) either from data when available, or a model based on $LMA$ or $WSG$ as described in the earlier section.
-
-For the combination of each model \@ref(eq:M1)- \@ref(eq:M4), species and hydrological realisation ($sp$), we obtained AIC, goodness-of-fit (*R*^2^) and Pearson's $\rho$ for each soil depth $z$.
+\@ref(eq:k-soil) either from data when available, or a model based on $LMA$ or $WSG$ as described in the earlier section, and used standardised K within species ($Std.K$; range 0-1) in Eqs.
+\@ref(eq:M1)- \@ref(eq:M4).
+\
+For the combination of each model \@ref(eq:M1)- \@ref(eq:M4), species and hydrological realisation, we obtained AIC, goodness-of-fit (*R*^2^) and Pearson's $\rho$ for each soil depth $z$.
 For each model-species-hydrological realisation we retained only one depth each with the least AIC as the best-fit, and retained only the subset of depths that had a positive Pearson's $\rho$ and *R*^2^ >= 0.5.
-For each model-species, we defined Effective Rooting Depth (ERD) as the median of this subset of depths and calculated standard error.
+For each model-species, we defined effective rooting depth (ERD) as the median of this subset of depths and calculated standard error.
 
@@
 ## ERD Coordination with Above-ground Hydraulic Traits
-
+\
 To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD with maximum stem hydraulic conductance ($K_{max, stem}$), leaf turgor loss point ($\Psi_{tlp}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$).
@@
 ## Mortality Analyses and Extreme Drought Experience
-
+\
 We calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$ 30 m) and average adbunance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; *n* = `r length(erd.mrate.sp)`).
@@
 To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{ERD}$) were below species $\Psi_{crit}$, defined as $\Psi_{80,~leaf}$.
-
+\
 All statistical analyses were done in R statistical environment (version 3.5.3, @R:2019).
@@
 The difference in our results are likely attributable to site differences in the precipitation regime.
-In their dry forest site (Mudumalai National Park, India) precipitation is low (1095 $\pm$ 296 mm; mean $\pm$ SD) compared to BCI (2687 $\pm$ 515 mm), and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (PET = 9xx $\pm$ xx mm), and thus at BCI deep soil layers are effectively recharged annually [@Chitra-Tarak.2018 and this study].
-We note that the variation in mortality explained by in our study is low (*R*^2^ = 0.11, *P* = 0.075), and other causes may explain observed mortality.
+In their dry forest site (Mudumalai National Park, India) precipitation is low (1095 $\pm$ 296 mm; mean $\pm$ SD) compared to BCI (`r round(mean(clim$Precip))` $pm$ `r round(sd(clim$Precip))` mm), and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (PET = `r round(mean(clim$pet.PM))` ($pm$ `r round(sd(clim$pet.PM))` mm), and thus at BCI deep soil layers are effectively recharged annually [@Chitra-Tarak.2018 and this study].
+We note that the variation in mortality explained by ERD in our study is low (*R*^2^ = 0.11, *P* = 0.075), and other causes may explain observed mortality.
 
@@
 This allowed us to scale-up PLC curves at the community scale (*n* = 284 species).
-Whether such scaling relationships between hard and soft traits hold true for other forests and their wider applicability remains to be seen as data on PLC curves in tropical forests are limited, but counter examples exist [see xx, yy, zz (Lawren, could you suggest references?].
+We find that per cent loss of conductivity was more rapid for soft wooded and low LMA species as $\Psi_{leaf}$ declined (Fig.
+S\@ref(fig:comm-plc-curves)A & S\@ref(fig:comm-plc-curves)B).
+Whether such scaling relationships between hard and soft traits hold true for other forests, and their wider applicability, remains to be seen as data on PLC curves are limited in tropical forests, but counter examples exist [see xx, yy, zz; Lawren, could you suggest references?
+Is our $K_{leaf}$ vs.
+$\Psi_{leaf}$ data the first reported for a tropical forest?].
 We also note the availability of data on $\delta^2H$ on trees and water sources belowground--soil/substrate matrix water in the vadoze zone or free water below the water table--as important for model calibration/validation.
@@
 
+```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd} Scatterplot of mean daily gross primary productivity (GPP) and mean daily vapour pressure deficit (VPD) by month measured at the eddy flux tower in Barro Colorado Island over 2012-2017.
+Red line shows a fit from a polynomial model described in inset, and used to standardise VPD employed in equations \\@ref(eq:M2) and \\@ref(eq:M4)."}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/gpp_by_env_variables/monthly_gpp.vpd.jpeg"))
+```
+
 ```{r iso-data, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-iso-data}Stable isotopic concetration ($\\delta^2H$) of soil water by depth measured in March 1997.
@@
 
-```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves} Leaf 'vulnerability curves', that is standardized $K_{leaf}$ vs.
+```{r obs-vuln-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-vuln-curves} Leaf 'vulnerability curves', that is standardized $K_{leaf}$ vs.
 $\\Psi_{leaf}$, fitted to observed data on ",  length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))," tree species from Panama, color coded by Wood Specific Gravity~100^$\\circ$^C~ ($gcm^{-3}$).")}
@@
 those that were predicted from models based on species soft-traits, WSG and LMA.")}
-p1 <- cowplot::ggdraw() + cowplot::draw_image("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/A_data_vs_model.jpeg", scale = 1)
-p2 <- cowplot::ggdraw() + cowplot::draw_image("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/B_data_vs_model.jpeg", scale = 1)
-cowplot::plot_grid(p1, p2, ncol = 2, labels = "AUTO")
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/A_B_data_vs_model.tiff"))
 ```
@@
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/diff.mortality_by rdi.gr.jpeg"))
+```
+
+```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-plc-curves} Modelled 'vulnerability curves' ($K_{leaf}$ vs.
+$\\Psi_{leaf}$) at community-scale (", length(data.model.AB$sp)," species) shown in terms of loss of conductivity (%) (A & B).
+Each line represents a species, color coded by $WSG$ (A), or $LMA$ (B), with blue tones indicating smaller values and red tones indicating larger values of $WSG$ (A), or $LMA$ (B)")}
+p1 <- cowplot::ggdraw() + cowplot::draw_image("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/std.k.spkmax_by_psi_color_by_SG100C_AVG_predicted_AB.jpeg", scale = 1)
+p2 <- cowplot::ggdraw() + cowplot::draw_image("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/std.k.spkmax_by_psi_color_by_LMALAM_AVD_predicted_AB.jpeg", scale = 1)
+cowplot::plot_grid(p1, p2, ncol = 2, labels = "AUTO")
 ```
@@
 ```{r param, results = 'asis', echo = FALSE}
-
 knitr::kable(param.table, caption = 'ELM-FATES parameters used to generate ensembles, with description, chosen ranges, references and rationale.

### 2020-08-08 manuscript.rmd
1. coefficient valeus given for models of A & B 2. tiff instead of jpeg

@@
 - The role of deep-water access in alleviating mortality of drought-sensitive trees has important implications for our predictive understanding of tropical forest dynamics under current and future climate.
+
 \
@@
 Note that only new data on leaf vulnerability curves collected during this study are described below, details of all other datasets used are described in the Supplementary Information.
+
 \
@@
 We collected new data on vulnerability curves for 21 species.
-Identifying scaling relationships between these curves and commonly available traits (Leaf Mass Area~lamina~, Wood Specific Gravity~100^$\circ$^C~; defined below) allowed us to scale-up these curves to `r length(data.model.AB$sp)` tree species in the community (of ~ 320 total).
+Identifying scaling relationships between these curves and commonly available traits (Leaf Mass Area~lamina~, Wood Specific Gravity~100^$\circ$^C~; defined below) allowed us to scale-up these curves to `r length(data.model.AB$sp)` tree species in the community (of 321 total).
 We evaluated ERD against existing data on plant and soil water isotopes ($\delta^2H$) and coordination of ERD with hydraulic traits and mortality rates.
@@
 Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding.
-Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere[xx add references].
+Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere [@Croat.1978; Leigh.1982; @Gentry.1990].
 
@@
 ```{r hydro-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-hydro-calib} Model evaluation: Observations (red lines) versus simulations (blue lines) from 100 best-fit parameter ensemble runs of ELM-FATES for daily Volumetric Water Content (VWC) by depth for horizontal probes at three depths (10, 40 and 100 cm; A) and a vertical probe (0-15 cm; B); monthly stream discharge (C) and monthly Evapotranspiration from eddy flux tower (D).
-Points (red) in (A) show average plot-wide observations of VWC with 95% CI (error bars).[Goodness-of-fit values to be added on the panels.]")}
+Points (red) in (A) show average plot-wide observations of VWC with 95% CI (error bars).[Goodness-of-fit values to be added on the panels.]xx")}
 knitr::include_graphics(file.path("figures/hydro-calib2.jpeg"))
@@
 \begin{equation}
-A = a_3WSG^2 + a_2B + a_1B^2 + a_0 + error
+A = `r acf[1]` `r acf[2]`WSG `r acf[3]`WSG^2 + `r acf[4]`WSG.B + `r acf[5]`B `r acf.6`B^2 + error
 (\#eq:a-mod)
@@
 \begin{equation}
-B = b_3WSG^2 + b_2WSG.LMA + b_1LMA^2 + b_0 + error
+B = `r bcf[1]` `r bcf[2]`B + `r bcf[3]`B^2 + `r bcf[4]`B.LMA + `r bcf[5]`LMA `r bcf[6]`LMA^2 + error
 (\#eq:b-mod)
@@
 
-where $(a_0, a_1, a_2, a_3, b_0, b_1, b_2, b_3)$ are coefficients.
 $WSG$ & parameter $B$ explained a large proportion of variation in parameter $A$ (Eq.
@@
 Each line represents a species, color coded by $WSG$ with blue tones indicating smaller values of $WSG$ and red tones indicating larger values of $WSG$.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB.jpeg"))
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB.tiff"))
 ```
@@
 We calculated species mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{t=1}^{6} {Mort_{t}}$ and, finally, species mortality anomalies by intervals as $Mort_{anomaly}$ = $Mort_{t}$ - $Mort_{mean}$.
-To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$).
-To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{ERD}$) were below species $\Psi_{crit}$, defined as $\Psi_{80,~leaf}$.
+To test whether ERD plays a role in alleviating mortality risk, we regressed species ERD against species ($Mort_{mean}$).
+To assess whether this may be due to deeper ERD species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{ERD}$) were below species $\Psi_{crit}$, defined as $\Psi_{80,~leaf}$.
 \
@@
 See Table  \\@ref(tab:spcodes) for scientific names corresponding to species codes.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/ERD_by_sp_large_canopy.jpeg"))
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/ERD_by_sp_large_canopy.tiff"))
 ```
@@
 *R*^2^ and *P*-values are given in panel insets.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/erd.stem.traits.jpeg"))
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/erd.stem.traits.tiff"))
 ```
@@
 \
-Species with deeper $ERD$ showed lower mean mortality rates across the 34-yr record (1982-2015; *R*^2^ = 0.11, *P* = 0.075, *n* = `r length(erd.mrate.sp)`; Fig.
+Species with deeper ERD showed lower mean mortality rates across the 34-yr record (1982-2015; *R*^2^ = 0.11, *P* = 0.075, *n* = `r length(erd.mrate.sp)`; Fig.
 \@ref(fig:erd-mort)).
-For five out of seven census intervals, species with shallower $ERD$ had significantly higher mortality anomalies than deeper ERD species; Fig.
+For five out of seven census intervals, species with shallower ERD had significantly higher mortality anomalies than deeper ERD species; Fig.
 (\@ref(fig:erd-mort-anom)).
@@
 *R*^2^ and *P*-values are given in panel insets.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_rate_by rdi.gr.jpeg"))
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_rate_by rdi.gr.tiff"))
 ```
@@
 (Are *R*^2^ and *P*-values needed here?)")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr.jpeg"))
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr.tiff"))
 ```
@@
 Wet season is shown in gray as beginning with DOY."}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q2.5_depths_in_inverse_model.jpeg"))
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q2.5_depths_in_inverse_model.tiff"))
 ```
 
-```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd} Scatterplot of mean daily gross primary productivity (GPP) and mean daily vapour pressure deficit (VPD) by month measured at the eddy flux tower in Barro Colorado Island over 2012-2017.
+```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd} Scatterplot of mean daily gross primary productivity (GPP) and mean daily vapour pressure deficit (VPD) by month measured at the eddy flux tower in Barro Colorado Island over 2012-2017.
 Red line shows a fit from a polynomial model described in inset, and used to standardise VPD employed in equations \\@ref(eq:M2) and \\@ref(eq:M4)."}
@@
 $\\Psi_{leaf}$, fitted to observed data on ",  length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))," tree species from Panama, color coded by Wood Specific Gravity~100^$\\circ$^C~ ($gcm^{-3}$).")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB_for_data_sp.jpeg"))
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB_for_data_sp.tiff"))
 ```
@@
 ```{r erd-mort-anom, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-anom}Mortality anomaly in each census interval during the study period plotted against ERD."}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/diff.mortality_by rdi.gr.jpeg"))
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/diff.mortality_by rdi.gr.tiff"))
 ```
 
-```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-plc-curves} Modelled 'vulnerability curves' ($K_{leaf}$ vs.
+```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-plc-curves} Modelled 'vulnerability curves' ($K_{leaf}$ vs.
 $\\Psi_{leaf}$) at community-scale (", length(data.model.AB$sp)," species) shown in terms of loss of conductivity (%) (A & B).
 Each line represents a species, color coded by $WSG$ (A), or $LMA$ (B), with blue tones indicating smaller values and red tones indicating larger values of $WSG$ (A), or $LMA$ (B)")}
-p1 <- cowplot::ggdraw() + cowplot::draw_image("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/std.k.spkmax_by_psi_color_by_SG100C_AVG_predicted_AB.jpeg", scale = 1)
-p2 <- cowplot::ggdraw() + cowplot::draw_image("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/std.k.spkmax_by_psi_color_by_LMALAM_AVD_predicted_AB.jpeg", scale = 1)
+p1 <- cowplot::ggdraw() + cowplot::draw_image("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/std.k.spkmax_by_psi_color_by_SG100C_AVG_predicted_AB.tiff", scale = 1)
+p2 <- cowplot::ggdraw() + cowplot::draw_image("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/std.k.spkmax_by_psi_color_by_LMALAM_AVD_predicted_AB.tiff", scale = 1)
 cowplot::plot_grid(p1, p2, ncol = 2, labels = "AUTO")

### 2020-08-09 manuscript.rmd
1. From tiff back to jpeg 2. Added SI figure for dbh model 3. Revised description of growth data processing

--- 2020-08-07 manuscript.rmd
@@
 \begin{equation}
-A = a_3WSG^2 + a_2B + a_1B^2 + a_0 + error
+A = `r acf[1]` `r acf[2]`WSG `r acf[3]`WSG^2 + `r acf[4]`WSGB + `r acf[5]`B `r acf.6`B^2 + error
 (\#eq:a-mod)
@@
 \begin{equation}
-B = b_3WSG^2 + b_2WSG.LMA + b_1LMA^2 + b_0 + error
+B = `r bcf[1]` `r bcf[2]`B + `r bcf[3]`B^2 + `r bcf[4]`BLMA + `r bcf[5]`LMA `r bcf[6]`LMA^2 + error
 (\#eq:b-mod)
@@
 \@ref(eq:M2)- \@ref(eq:M4)), in which a species' growth dynamics across census intervals was regressed against census interval means of multiplied effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the fraction of leaf cover (all standardised to 0-1, see below).
-Let n~t~ be the total number of days in census interval $t$, then for individual tree species Growth~t~ and depth $z$ these models are described as follows (add additive models xx):
+Let n~t~ be the total number of days in census interval $t$, then for individual tree species Growth~t~ and depth $z$ these models are described as follows:
 
@@
 \
-To minimise the effect of light variation among trees, we selected only large trees (>= 30 cm dbh, diameter at breast height) and species whose maximum height was at least 30 m (hereafter, canopy species) that are likely to be fully exposed to the sun.
-We chose only those trees for which diameter (dbh) record was present across all size censuses (1990, 1995, 2000, 2005, 2010, 2015).
-For each individual tree (i), we calculated growth rates ($Gr_{i,t}$; cm.yr^-1^) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses, divided by the inter-census interval based on the exact date of measurements.
-As tree growth rates increase with size, to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($Resid_{i,t} = Gr_{i,t} - f(DBH_{i, t})$).
-The diameter model of growth was fitted based on all trees' data pooled together as there was insufficient species-specific data for large trees (xx Add SI figure showing the model).
-We then standardized (centered and scaled) an individual tree time series ($Resid_{i,t}$) and obtained species-level (*n* = `r length(erd.sp)`) times-series of medians ($Growth_{t}$) that had data on at least three trees (median = xx).
+For growth estiamtes, to minimise the effect of light variation among trees, we selected only large trees ($\ge$ 30 cm dbh, diameter at breast height) and species whose maximum height was at least 30 m (hereafter, canopy species) and thus are likely to be fully exposed to the sun.
+We selected trees for which height of diameter measurement was unchanged across all censuses (1990, 1995, 2000, 2005, 2010, 2015).
+For each individual tree (*i*), we calculated growth rates ($Gr_{i,t}$; cm.yr^-1^) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses, divided by the inter-census interval based on the exact date of measurements.
+We eliminated positive outliers as any growth rates amounting to > 75 mm.yr^-1^, as this was the highest rate observed with confidence for fastest growing species [@Condit.2017].
+Negative outliers were removed based on known measurement error, $\sigma_{e} = 0.006214d + 0.903$, where *d* is dbh.
+Any stem whose later dbh measurement fell < 4$\sigma_{e}$ the earlier measurement was removed [@Condit.2017].
+As tree growth rates vary with size, to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($Resid_{i,t} = Gr_{i,t} - f(d_{i, t})$).
+The diameter model of growth ($f(d_{i, t})$), a B-spline based polynomial equation with 5 degrees of freedom, was fitted to growth rates from all trees pooled together as there was insufficient species-specific data for large trees (Fig.
+S\@ref(fig:dbh-model))).
+We retained only those trees that had $Resid_{i,t}$ estimates across all censuses (complete records, total trees = `r sum(g.n$n)`) and standardized (centered and scaled) individual tree time series of $Resid_{i,t}$.
+We then obtained species-level (*n* = `r length(erd.sp)`) time-series of medians of $Resid_{i,t}$ ($Growth_{t}$).
+These were obtained for only those species with complete records on at least three trees (median *n* = `r median(g.n$n)`, max *n* = `r max(g.n$n)` trees).
 \
@@
 We also found that LMA~lamina~ and WSG~100^$\circ$^C~ were strong predictors of PLC curves for leaf, but not branch (add SI figures).
-High $K_{max}$ is elsewhere associated with photosynthetic rates with faster growth potential, but we did not find a relationship between $K_{max}$ and growth rates (*n* = 21 species).
+High $K_{max}$ is elsewhere associated with photosynthetic rates with faster growth potential, but we did not find a relationship between $K_{max}$ and growth rates (*n* = 21 species; xx add figure).
 These relationships between ERD and aboveground hydraulic traits will provide an important insight on how to model the rooting depth and its coordination or trade-off with other traits to better represent the functional diversity in tropical forests.
@@
 
-```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd} Scatterplot of mean daily gross primary productivity (GPP) and mean daily vapour pressure deficit (VPD) by month measured at the eddy flux tower in Barro Colorado Island over 2012-2017.
+```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd} Scatterplot of mean daily gross primary productivity (GPP) and mean daily vapour pressure deficit (VPD) by month measured at the eddy flux tower in Barro Colorado Island over 2012-2017.
 Red line shows a fit from a polynomial model described in inset, and used to standardise VPD employed in equations \\@ref(eq:M2) and \\@ref(eq:M4)."}
@@
 
-```{r obs-vuln-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-vuln-curves} Leaf 'vulnerability curves', that is standardized $K_{leaf}$ vs.
-$\\Psi_{leaf}$, fitted to observed data on ",  length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))," tree species from Panama, color coded by Wood Specific Gravity~100^$\\circ$^C~ ($gcm^{-3}$).")}
+```{r obs-vuln-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-vuln-curves} Leaf 'vulnerability curves', that is, $K_{leaf}$ vs.
+$\\Psi_{leaf}$, fitted to observed data on ",  length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))," tree species from Panama, color coded by Wood Specific Gravity~100^$\\circ$^C~ (gcm^-3^).")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB_for_data_sp.jpeg"))
@@
 
-```{r ab-fit, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ab-fit}Parameters $A$ (panel A) & $B$ (panel B) of leaf vulnerability curves that were fitted to observed data on $K_{leaf}$ vs.
+```{r ab-fit, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ab-fit}Parameters $A$ (panel A) & $B$ (panel B) of leaf vulnerability curves that were fitted to observed data on $K_{leaf}$ vs.
 $\\Psi_{leaf}$ for ", sp.soft.filled$AB.data.model," species vs.
 those that were predicted from models based on species soft-traits, WSG and LMA.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/A_B_data_vs_model.tiff"))
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/A_B_data_vs_model.jpeg"))
+```
+
+```{r dbh-model, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-dbh-model} A scatterplot of tree growth rates (circles) from all census intervals versus dbh at the beginning of the cenus interval is overlaid with a dbh model of growth (spline).
+Growth rates are color coded by census interval.
+The model equation is shown in inset along with goodness-of-fit(*R*^2^)")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/dbh/growthrate_dbh_predicted_growth_with_spline_degree_5_dbh_cutoff_100_model.jpeg"))
 ```
@@
 
-```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-plc-curves} Modelled 'vulnerability curves' ($K_{leaf}$ vs.
+```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-plc-curves} Modelled 'vulnerability curves' ($K_{leaf}$ vs.
 $\\Psi_{leaf}$) at community-scale (", length(data.model.AB$sp)," species) shown in terms of loss of conductivity (%) (A & B).

### 2020-08-09 manuscript.rmd
model comparison moved back to methods

@@
 
+The model with VPD and without Leaf Cover had highest goodness-of fit (*R*^2^ = 0.51; *p* = 0.072), followed by model without both VPD and Leaf Cover (Eq.
+\@ref(eq:M1); Fig.
+S\@ref(fig:other-erd-mods)).
+Models with Leaf Cover had only three species with sufficient species-level goodness-of-fit (Eqs.
+\@ref(eq:M3)-\@ref(eq:M4)) versus seven for models without (Eqs.
+\@ref(eq:M1)-\@ref(eq:M2)).
+For rest of the analyses we therefore used species ERD estimates from the model with VPD and without Leaf Cover (Eq.
+\@ref(eq:M2)).
+
 ## ERD Coordination with Above-ground Hydraulic Traits
@@
 \
-In comparison to the measured isotopic signal, our results showed that the model with VPD and without Leaf Cover (Eq.
-\@ref(eq:M2)) had highest goodness-of fit (*R*^2^ = 0.51; *p* = 0.072), followed by followed by model without both VPD and Leaf Cover (Eq.
-\@ref(eq:M1); Fig.
-S\@ref(fig:other-erd-mods)).
-Models with Leaf Cover had only three species with sufficient species-level goodness-of-fit (Eqs.
-\@ref(eq:M3)-\@ref(eq:M4)) versus seven for models without (Eqs.
-\@ref(eq:M1)-\@ref(eq:M2)).
-For rest of the analyses we therefore used species ERD estimates from the model with VPD and without Leaf Cover (Eq.
-\@ref(eq:M2); Fig.
+In comparison to the measured isotopic signal, our results showed that the ERD model with VPD and without Leaf Cover (Eq.
+\@ref(eq:M2)) had highest goodness-of fit (*R*^2^ = 0.51; *p* = 0.072) and was thus the chosen model of ERD for rest of the analyses.
+Notably, two species, *Luehea* *seemannii* and *Trichilia tuberculata*, that @Meinzer:1999 found to have xylem sap $\delta^2H$ values between soil water and groundwater, suggesting that these species sourced most of the water from depths > 1 m, were also identified by our model with ERD > 1 m (2.9 m for both species).
+(Fig.
 \@ref(fig:erd-iso)).
@@
 \
-In this forest modelled effective rooting depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from 0.5 m to 13 m (Fig.
+In this forest modelled effective rooting depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from `r min(df.erd.to.plot$depth)` m to `r max(df.erd.to.plot$depth)` m (Fig.
 \@ref(fig:sp-erd)).
@@
 ```{r sp-erd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap= paste0("\\label{lab-sp-erd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) for ", length(erd.sp)," large, canopy species of Barro Colorado Island.
-See Table  \\@ref(tab:spcodes) for scientific names corresponding to species codes.")}
+See Table  \\@ref(tab:spcodes) for complete scientific names.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/ERD_by_sp_large_canopy.jpeg"))
@@
 
-```{r iso-data, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-iso-data}Stable isotopic concetration ($\\delta^2H$) of soil water by depth measured in March 1997.
+```{r iso-data, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-iso-data}Stable isotopic ratio ($\\delta^2H$) of soil water by depth measured at Barro Colorado Island in March 1997.
 Data from @Meinzer:1999."}
@@
 The model equation is shown in inset along with goodness-of-fit(*R*^2^)")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/dbh/growthrate_dbh_predicted_growth_with_spline_degree_5_dbh_cutoff_100_model.jpeg"))
+knitr::include_graphics(file.path("figures/dbh/growthrate_dbh_predicted_growth_with_spline_degree_5_dbh_cutoff_100_model.jpeg"))
 ```

### 2020-08-09 code/15.0_PhenoDemoTraitsPsi.R

@@ -907,60 +907,6 @@ ml.rsq.combine.sub <- ml.rsq.combine.sub %>% transform(models.plot1 = factor(cor
                                                                                                              expression(italic(K[italic('leaf')]*'LeafCover')),
                                                                                                                         expression(italic(K[italic('leaf')]*'VPD'*'LeafCover')))))
 
-p3.2 <- ggplot(ml.rsq.combine.sub %>% subset(source == "Meinzer et al.1999 Fig. 4"),
-             aes(x = Xylem_sap_deltaD_permil, y = depth)) + #HSMTLP.80L)) +
-  coord_cartesian(ylim = c(13, 0.3)) +
-  geom_smooth(method = "lm", se = TRUE, color = "black", size = 0.5, formula = formula) +
-  geom_errorbarh(aes(xmax = Xylem_sap_deltaD_permil + se,
-                     xmin = Xylem_sap_deltaD_permil - se, color = sp),
-                 size = 0.5, height = 0.05) +
-  geom_errorbar(aes(ymax = depth + depth.se, ymin = depth - depth.se, color = deciduousness), size = 0.5, height = 0.05) +
-  facet_wrap( ~ models.plot1, nrow = 1) +
-  geom_text(aes(x =  Xylem_sap_deltaD_permil, y = depth, label = sp, color = sp), nudge_y = 0.07, nudge_x = 0,
-            size = 3) +
-  ylab(expression("Water Uptake Depth (m)")) + xlab(xylem.label) +
-  scale_y_continuous(trans=reverselog_trans(10), breaks = unique(ml.rsq.combine$depth)) +
-  stat_poly_eq(aes(label = paste(..rr.label..)),
-               npcx = 0.98, npcy = 0.13, rr.digits = 2,
-               formula = formula, parse = TRUE, size = 4) +
-  stat_fit_glance(method = 'lm',
-                  method.args = list(formula = formula),
-                  geom = 'text_npc',
-                  aes(label = paste("P = ", round(..p.value.., digits = 3), sep = "")),
-                  npcx = 0.98, npcy = 0.05, size = 4) +
-  geom_point(shape = 21, color = "white", aes(fill = sp), alpha = 1, size = 3) +
-  theme(legend.position = "top",
-        legend.direction = "horizontal") +
-  guides(fill = "none", color = "none")
[…]

### 2020-08-10 manuscript.rmd
1. Added objective function for calibration 2. Edited discussion text

@@
 We calibrated the model over 2012-2018 against three key fluxes in Eq.
-\@ref(eq:water-balance), namely, (i) evapotranspiration (ET, $E_{v} + E_{g}$) from the eddy flux tower (monthly, 2012-2017), (ii) stream discharge for the Conrad Catchment ($q_{over} + q_{drai}$; daily, 2012-2018) and (iii) volumetric water content (VWC; $W_{liq, z}$; Fig.
-\@ref(fig:hydro-calib); daily, 2015-2018).
-We collected VWC data (2016-2018) from a point location at the eddy flux tower with Time Domain Reflectometry (TDR) using three horizontal probes at depths 10, 40 and 100 cm and one vertical probe (0-15 cm).
-TDR data for the vertical probe was calibrated based on measurements of gravimetric soil moisture on several occasions (Fig.
+\@ref(eq:water-balance), namely, (i) evapotranspiration ($ET \simeq E_{v} + E_{g}$) from the eddy flux tower (2012-2017), (ii) stream discharge for the Conrad Catchment ($q_{run} \simeq q_{over} + q_{drai}$; 2012-2018) and (iii) volumetric water content (VWC, $\theta_{z} \simeq W_{liq, z}$; 2012-2018).
+We collected VWC data (2012-2018) from a point location at the eddy flux tower with a vertical Time Domain Reflectometry (TDR) probe ($\theta_{0-0.15m}$).
+This data was calibrated based on measurements of gravimetric soil moisture at the vertical probe (0-15 cm) on several occasions (Fig.
 \@ref(fig:TDR-calib)) and mean bulk dry density of 0.8 gcm^-3^ (*n* = 5, cylinder volume = 1250 cm^3^).
-The same calibration equation was used for other probes.
-We also leveraged existing opportunistic dry season VWC measurements for the entire 50-ha plot (15, 40 and 100 cm in 2015 and 15 & 40 cm in 2016; @Kupers:2019).
-Among the 5000 parameter ensembles we chose 100 ensembles that best-fit all three fluxes simultaneously (add details...that is those that maximised the sum of regression coefficients between observed and simulated variables.
-xx add equation) See Table \@ref(tab:param) for best-fit values and ranges for different parameters.
+We also leveraged existing opportunistic dry season VWC measurements by depth for the entire 50-ha plot ($\theta_{0.15m}$, $\theta_{0.4m}$ and $\theta_{1m}$ in 2015 and $\theta_{0.15m}$, $\theta_{0.4m}$ in 2016; @Kupers:2019).
+Beginning in 2016 we additionally installed three horizontal probes at depths of 0.15, 0.4 and 1m, but depth specific calibration equation was not available, so we applied the same calibration equation to these data as that of the vertical probe.
+However, for VWC calibration we only used data from the vertical probe to capture the long-term dynamics and plot-wide measurements to capture the dynamics by depth.
+To find parameter ensembles that best-fit all three fluxes simultaneously, we calculated an objective function for each of the 5000 ensembles based on root mean square error (RMSE) between observations and simulations.
+Let $f(obs,~sim)$ be normalised RMSE (0-1) between observations and simulations, based on monthly sums for $ET$ and $q_{run}$ and daily avarages for $\theta$s, then the objective function ($Z$) is described as:
+
+\begin{equation}
+Z = \frac{1}{6}\sum \left(f(ET,~E_{v} + E_{g}) + f(q_{run},~q_{over} + q_{drai}) + f(\theta_{0-0.15m}, {W_{liq,~\bar{z}}})  + f(\theta_{0.15m}, W_{liq,~`r soil.depths[4]`m}) + f(\theta_{0.4m}, W_{liq,~`r soil.depths[6]`m}) + f(\theta_{1m}, W_{liq,~`r soil.depths[8]`m})\right)
+(\#eq:obj-func)
+\end{equation}
+
+where $W_{liq,~\bar{z}}$ is the daily average of $W_{liq,~z}$ for $z$ = `r paste(soil.depths[1:4], collapse=", ")` m.
+To ensure calibration with $\theta_{0.1m}$, we first selected ensembles with  $f(\theta_{1m}, W_{liq,~`r soil.depths[8]`m})$ < 0.2 (~300 ensembles) and among them selected 100 ensembles with lowest values for the objective function $Z$.
+Model simulations corresponding to these 100 best-fit ensembles are shown against observed fluxes in Fig.
+\@ref(fig:hydro-calib), including VWC from horizontal probes that were not used for calibration.
+See Table \@ref(tab:param) for the range of best-fit values for different parameters.
 We ran ELM-FATES with the chosen 100 ensembles from 1985-2018, with first 5 years used for model spin-up.
@@
 S\@ref(fig:comm-plc-curves)A & S\@ref(fig:comm-plc-curves)B).
-Whether such scaling relationships between hard and soft traits hold true for other forests, and their wider applicability, remains to be seen as data on PLC curves are limited in tropical forests, but counter examples exist [see xx, yy, zz; Lawren, could you suggest references?
-Is our $K_{leaf}$ vs.
-$\Psi_{leaf}$ data the first reported for a tropical forest?].
+Whether such scaling relationships between hard and soft traits hold true for other forests, and their wider applicability, remains to be seen as data on PLC curves are limited in tropical forests, but counter examples exist [Lawren, could you suggest references?
+Also, folks, is our $K_{leaf}$ vs.
+$\Psi_{leaf}$ data the first reported for a tropical forest?
+or could you add how rare it is?].
 We also note the availability of data on $\delta^2H$ on trees and water sources belowground--soil/substrate matrix water in the vadoze zone or free water below the water table--as important for model calibration/validation.
@@
 We report such a trade-off at both leaf (*n* = 21 species) and branch level (*n* = 27 species).
-We also found that LMA~lamina~ and WSG~100^$\circ$^C~ were strong predictors of PLC curves for leaf, but not branch (add SI figures).
-High $K_{max}$ is elsewhere associated with photosynthetic rates with faster growth potential, but we did not find a relationship between $K_{max}$ and growth rates (*n* = 21 species; xx add figure).
-These relationships between ERD and aboveground hydraulic traits will provide an important insight on how to model the rooting depth and its coordination or trade-off with other traits to better represent the functional diversity in tropical forests.
+We also found that LMA~lamina~ and WSG~100^$\circ$^C~ were strong predictors of PLC curves for leaves, but not for xylem (not shown).
+Acquisitive traits such as high $K_{max}$ are expected to be associated with faster growth rates.
+We found a positive but not  significant relationship of adult growth rates ($\ge$ 10cm dbh) with $K_{max, ~leaf}$ (*R*^2^ = 0.13; *p* = 0.126, *n* = `leaf.traits.sp`) and no relationship with $K_{max, ~stem}$ (*n* = `stem.traits.sp` species).
+The relationships between ERD and aboveground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits in order to better represent the functional diversity of tropical forests.
 

### 2020-08-10 manuscript.rmd
Added RMSE on hydro-calib

@@
 \@ref(fig:TDR-calib)) and mean bulk dry density of 0.8 gcm^-3^ (*n* = 5, cylinder volume = 1250 cm^3^).
-We also leveraged existing opportunistic dry season VWC measurements by depth for the entire 50-ha plot ($\theta_{0.15m}$, $\theta_{0.4m}$ and $\theta_{1m}$ in 2015 and $\theta_{0.15m}$, $\theta_{0.4m}$ in 2016; @Kupers:2019).
+We also leveraged existing opportunistic dry season gravimetric measurements by depth for the entire 50-ha plot and converted them to VWC units assuming bulk density of 0.8 gcm^-3^  ($\theta_{0.15m}$, $\theta_{0.4m}$ and $\theta_{1m}$ in 2015 and $\theta_{0.15m}$, $\theta_{0.4m}$ in 2016; @Kupers:2019).
 Beginning in 2016 we additionally installed three horizontal probes at depths of 0.15, 0.4 and 1m, but depth specific calibration equation was not available, so we applied the same calibration equation to these data as that of the vertical probe.
@@
 To find parameter ensembles that best-fit all three fluxes simultaneously, we calculated an objective function for each of the 5000 ensembles based on root mean square error (RMSE) between observations and simulations.
-Let $f(obs,~sim)$ be normalised RMSE (0-1) between observations and simulations, based on monthly sums for $ET$ and $q_{run}$ and daily avarages for $\theta$s, then the objective function ($Z$) is described as:
+Let $f(obs,~sim)$ be normalised RMSE (0-1) between observations and simulations, based on monthly sums for $ET$ and $q_{run}$ and daily avarages for $\theta$s, then the objective function can be described as:
 
 \begin{equation}
-Z = \frac{1}{6}\sum \left(f(ET,~E_{v} + E_{g}) + f(q_{run},~q_{over} + q_{drai}) + f(\theta_{0-0.15m}, {W_{liq,~\bar{z}}})  + f(\theta_{0.15m}, W_{liq,~`r soil.depths[4]`m}) + f(\theta_{0.4m}, W_{liq,~`r soil.depths[6]`m}) + f(\theta_{1m}, W_{liq,~`r soil.depths[8]`m})\right)
+Objective Function = \frac{1}{6}\sum \left(f(ET,~E_{v} + E_{g}) + f(q_{run},~q_{over} + q_{drai}) + f(\theta_{0-0.15m}, {W_{liq,~\bar{z}}})  + f(\theta_{0.15m}, W_{liq,~`r soil.depths[4]`m}) + f(\theta_{0.4m}, W_{liq,~`r soil.depths[6]`m}) + f(\theta_{1m}, W_{liq,~`r soil.depths[8]`m})\right)
 (\#eq:obj-func)
@@
 
-```{r hydro-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-hydro-calib} Model evaluation: Observations (red lines) versus simulations (blue lines) from 100 best-fit parameter ensemble runs of ELM-FATES for daily Volumetric Water Content (VWC) by depth for horizontal probes at three depths (10, 40 and 100 cm; A) and a vertical probe (0-15 cm; B); monthly stream discharge (C) and monthly Evapotranspiration from eddy flux tower (D).
-Points (red) in (A) show average plot-wide observations of VWC with 95% CI (error bars).[Goodness-of-fit values to be added on the panels.]xx")}
+```{r hydro-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-hydro-calib} Model evaluation: Observations (red lines) versus simulations (blue lines) from 100 best-fit parameter ensemble runs of ELM-FATES for daily Volumetric Water Content (VWC) by depth for horizontal Time Domain Reflectometry (TDR) probes at three depths (10, 40 and 100 cm; A) and a vertical probe (0-15 cm; B); monthly stream discharge (C) and monthly Evapotranspiration from eddy flux tower (D).
+Points (red) in (A) show average plot-wide observations of VWC with 95% CI (error bars).
+VWC data from the horizontal probes (lines in A) were not used for ELM-FATES calibration.
+Values in inset are average standardised RMSE across the 100 simulations.
+For panel A, values at the top are for plot-wide VWC and those at the bottom are for TDR probes.")}
 knitr::include_graphics(file.path("figures/hydro-calib2.jpeg"))
@@
 
-```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus hydraulic properties for eight large, canopy species found on Barro Colorado Island (Panama).
+```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus hydraulic properties for eight large, canopy species found on Barro Colorado Island (Panama).
 *R*^2^ and *P*-values are given in panel insets.")}
@@
 Note the log scale on horizontal axis.")}
-knitr::include_graphics(file.path("figures/Bootstrapped Ksat profiles within observed 95CI_random_two_hundred.pdf"))
+knitr::include_graphics(file.path("figures/Bootstrapped Ksat profiles within observed 95CI_random_two_hundred.jpeg"))
 ```
@@
 
-```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts} Extreme soil droughts by census interval (panels) at three representative depths: 0.5, 1 and 1.7 m, depicted in different colors.
+```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts} Extreme soil droughts by census interval (panels) at three representative depths: 0.5, 1 and 1.7 m, depicted in different colors.
 Mean soil water potential ($Psi_{soil,~z}$) for a given DOY and depth z are shown as lines.
@@
 
-```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd} Scatterplot of mean daily gross primary productivity (GPP) and mean daily vapour pressure deficit (VPD) by month measured at the eddy flux tower in Barro Colorado Island over 2012-2017.
+```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "20%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd} Scatterplot of mean daily gross primary productivity (GPP) and mean daily vapour pressure deficit (VPD) by month measured at the eddy flux tower in Barro Colorado Island over 2012-2017.
 Red line shows a fit from a polynomial model described in inset, and used to standardise VPD employed in equations \\@ref(eq:M2) and \\@ref(eq:M4)."}
@@
 Each line represents a species, color coded by $WSG$ (A), or $LMA$ (B), with blue tones indicating smaller values and red tones indicating larger values of $WSG$ (A), or $LMA$ (B)")}
-p1 <- cowplot::ggdraw() + cowplot::draw_image("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/std.k.spkmax_by_psi_color_by_SG100C_AVG_predicted_AB.jpeg", scale = 1)
-p2 <- cowplot::ggdraw() + cowplot::draw_image("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/std.k.spkmax_by_psi_color_by_LMALAM_AVD_predicted_AB.jpeg", scale = 1)
-cowplot::plot_grid(p1, p2, ncol = 2, labels = "AUTO")
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/plot.comm.plc.jpeg"))
 ```
 
+\
 ```{r symbols, results = 'asis', echo = FALSE}
@@
 ```
-
+\
 ```{r param, results = 'asis', echo = FALSE}
@@
 ```
-
+\
 ```{r spcodes, results = 'asis', echo = FALSE}

### 2020-08-29 manuscript.rmd
Numerous edits

@@
 Here, we tested the role of deep-water access in survival strategies of a diverse tropical forest community (Panama) using a novel data-model approach.
-- We inversely modelled effective rooting depth (ERD) for large trees of `r length(erd.sp)` canopy species by linking growth dynamics (1990-2015) to vapour pressure deficit, matric potentials in the whole-soil column, and a new community-scale model of species leaf vulnerability curves.
+- We inversely modeled effective rooting depth (ERD) for large trees of `r length(erd.sp)` canopy species by linking growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and a new community-scale model of species leaf vulnerability curves.
 We evaluated ERD against existing isotopic tracer data on trees and soil.
 - We find that deeper ERD is associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates over 35 years.
-Deeper ERD species spent less time below critical hydraulic thresholds indicating that trees' mitigate water-stress related mortality risk through deep water access.
+Deeper ERD species spent less time below critical hydraulic thresholds indicating that trees mitigate water-stress related mortality risk through deep water access.
 - The role of deep-water access in alleviating mortality of drought-sensitive trees has important implications for our predictive understanding of tropical forest dynamics under current and future climate.
@@
 \
-Keywords: belowground traits, deep-water access, drought-induced mortality, drought tolerance, rooting depths, tropical forest, life-history strategies
+Keywords: below-ground traits, deep-water access, drought-induced mortality, drought tolerance, rooting depths, tropical forest, life-history strategies
 
@@
 Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism.
-This mechanism assumes that water availability varies belowground by vertical depth and so do rooting depths of co-occuring trees, by species and size [@Meinzer:1999; @Chitra-Tarak.2018].
-Consequently, trees' in the same forest differ in their drought experience, and thus in their responses in term of growth and mortality.
-A key bottlneck in community-wide testing of this mechanism has been lack of data on trees' rooting or water-sourcing depths, followed by data on water-availability at those depths.
+This mechanism assumes that water availability varies below-ground by vertical depth and so do rooting depths of co-occurring trees, by species and size [@Meinzer:1999; @Chitra-Tarak.2018].
+Consequently, trees in the same forest differ in their drought experience, and thus in their responses in term of growth and mortality.
+A key bottleneck in community-wide testing of this mechanism has been lack of data on trees' rooting or water-sourcing depths, followed by data on water-availability at those depths.
 A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the world’s 65 thousand tree species [@Fan:2017; @Beech:2017].
-Further, only a small fraction of those are tropical, even though ~60-80 % (40 - 53 thousand species) of the World’s tree diversity resides in the tropics [@Slik:2015].
-Use of stable isotopes of water as a tracer provides an indirect way of trees' water sourcing depths by matching isotopic value in plant tissue to variations in soil pore water vertically belowground.
+Further, only a small fraction of those are tropical, even though ~60-80 % (40 - 53 thousand species) of the world’s tree diversity resides in the tropics [@Slik:2015].
+Use of stable isotopes of water as a tracer provides an indirect way of trees' water sourcing depths by matching isotopic value in plant tissue to variations in soil pore water vertically below-ground.
 However, such data are extremely rare [@Evaristo:2016].
@@
 \
-Trade-offs among co-occurring plants shape how they make a living in their environment (ecological strategies) and also provide the essential constraints to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d].
-This entails indentifying trade-offs among environment, traits and demography.
-Specifically for the water environment (or resource) observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified an acquisitive-conservative resource economic spectrum similar to the ones for light and nutrients, although, for water, examples of all trade-offs observed simultaneously are rare, and belowground traits or environment linkages are weak (@Reich:2014; @Sterck:2011).
-Acquisitive species have traits with high capacity to acquire, move, store and use water (high $K_{max}$, low wood density) and achieve high growth potential, but build flimsy, disposable tissues (high SLA) and are less tolerant of low resources, whereas species with conservative traits are better protected from carbon losses and are more drought tolerant (low $\Psi_{50}$, low $\Psi_{tlp}$, larger safety margin).
-Acquisitive species are spatially associated with wet sites (high resource) such as valleys or those with high rainfall [@Kursar:2009; @Oliveira:2019], but whether and how the acquisitive-conservative spectrum holds true for the moisture dynamics vertically belowground is not clear.
+Trade-offs shape how plants make a living in their environment (ecological strategies) and also provide the essential constraints to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010].
+This entails identifying trade-offs among environment, traits and demography.
+Specifically for the water environment (or, resource), observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified an acquisitive-conservative resource economic spectrum similar to those for light and nutrients, although, for water, examples of all trade-offs observed simultaneously are rare, and below-ground traits or environment linkages are weak [@Reich:2014; @Sterck:2011].
+Acquisitive species have traits with high capacity to acquire, move, store and use water (high $K_{max}$, low wood density) and achieve high growth potential, but build flimsy, disposable tissues (high SLA) and are less tolerant of low resources, whereas species with conservative traits are better protected from carbon losses and are more drought tolerant (low $\Psi_{50}$, low $\Psi_{tlp}$, larger safety margin; see Table 1 for trait symbols and their definitions).
+Acquisitive species are spatially associated with wet sites (high resource) such as valleys or those with high rainfall [@Kursar:2009; @Oliveira:2019], but whether and how the acquisitive-conservative spectrum holds true for the water dynamics vertically below-ground is not clear.
 Deep rootedness may be expected to support an acquisitive strategy with a steady water supply unlike shallow-rooted trees.
 However, this may depend on the exact dynamics of water-availability by depth at a site, depending on precipitation regime and hydrology.
-In contrast, in pulse-driven arid systems *shallow*-rooted plants are associated with acquisitive traits (high $K_{max}$) that maximise the use of the water-pulse, while *deep*-rooted plants show conservative traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013].
+In pulse-driven arid systems *shallow*-rooted plants are associated with acquisitive traits (high $K_{max}$) that maximise the use of the water-pulse, while *deep*-rooted plants show conservative traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013 and references therein].
 
@@
 These efforts have resulted in identification of above-ground traits such as proxies of hydraulic failure as indicators of mortality risk.
-A recent meta-analysis found greater drought-induced mortality rates in species with cavitation-vulnerable xylem ($\Psi_{88,~stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~stem}$) [@Anderegg:2016], but counter examples also exist [@Hoffmann:2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013]; indicating that plants reach critical thresholds of hydraulic failure depending on their realised drought intensity (see Table \@ref(tab:symbols) for trait symbols and their definitions).
-However, plants' realised water environments are rarely studied, especially in the tropical region.
-To better understand how belowground rooting is linked to aboveground hydraulic traits and mortality for co-occuring tropical forest tree species, we make a thorough assessment of trade-offs between the belowground water environment, above-ground hydraulic traits, estimates of effective rooting depths and demographic outcomes over 34 years that witnessed droughts of various intensity and duration for a tropical site.
-Specifically, we test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns species on an acquisitive-conservative economic spectrum optimised for increasingly reliable water availability by depth such that (i) trees with deeper ERD have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem: higher maximum stem hydraulic conductance ($K_{max}$), but greater xylem vulnerability to embolism ($\Psi_{88,~stem}$), higher $\Psi_{tlp}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$), whereas shallower ERD species are associated with less efficient but safer hydraulic traits; and that, (ii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure, whereas shallower ERD species may have higher mortality rates as a result of higher variability in shallow soil water availability.
-To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.
+A recent meta-analysis found greater drought-induced mortality rates in species with cavitation-vulnerable xylem ($\Psi_{88,~stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~stem}$) [@Anderegg:2016], but counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013]; indicating that plants reach critical thresholds of hydraulic failure depending on their realized drought intensity.
+However, plants' realized water environments are rarely studied, especially in the tropical region.
+To better understand how below-ground rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tropical forest tree species, we make a thorough assessment for a tropical forest of trade-offs between the below-ground water environment, tress' above-ground hydraulic traits, estimates of effective rooting depths and demographic outcomes over 35 years that witnessed droughts of various intensity and duration.
+Specifically, we test an overarching hypothesis that inter-specific variation in trees' effective rooting depths aligns species on an acquisitive-conservative economic spectrum optimized for increasingly reliable water availability by depth such that (i) trees with deeper ERD have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem: higher maximum stem hydraulic conductivity ($K_{max}$), but greater xylem vulnerability to embolism ($\Psi_{88,~stem}$), higher $\Psi_{tlp}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$), whereas shallower ERD species are associated with less efficient but safer hydraulic traits; and that, (ii) deeper ERD species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallower ERD species may have higher mortality rates as a result of higher variability in shallow soil water availability.
+To our knowledge this is the first study to demonstrate a mechanistic link between trees' realized drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.
+
+\
+```{r symbols, results = 'asis', echo = FALSE}
+knitr::kable(symbols.table, caption = 'The symbol, definition and units of key traits used.', booktabs = TRUE, escape = FALSE)
+```
 
@@
 \
-In the following sections we describe the study site, the models and data used to estimate and evaluate (1) soil water potential dynamics by depth over the forests’ rooting zone, (2) species-specific leaf vulnerability curves, and (3) species-specific effective rooting depths.
+In the following sections we describe the study site, the models and data used to estimate and evaluate (1) soil water potential dynamics by depth in the whole-soil column, (2) species-specific leaf vulnerability curves, and (3) species-specific effective rooting depths.
 These are followed by analyses to address the proposed hypotheses.
@@
 \
-In summary, we obtained daily dynamics of plant water potential ($\Psi_{soil,~z}$) by depth $z$ in the whole-soil column with a locally parametarised and calibrated 1-D hydrological water balance model.
-Developing a novel inverse model allowed us to estimate species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015).
+In summary, we obtained daily dynamics of plant water potential ($\Psi_{soil,~z}$) by depth $z$ in the whole-soil column with a locally parameterized and calibrated 1-D hydrological water balance model.
+We developed a novel inverse model to estimate species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015).
 This model incorporated impact of atmospheric drought (VPD) on growth and was constrained by above-ground hydraulic traits, namely, species leaf vulnerability curves ($K_{leaf}$ vs.
@@
 We collected new data on vulnerability curves for 21 species.
-Identifying scaling relationships between these curves and commonly available traits (Leaf Mass Area~lamina~, Wood Specific Gravity~100^$\circ$^C~; defined below) allowed us to scale-up these curves to `r length(data.model.AB$sp)` tree species in the community (of 321 total).
-We evaluated ERD against existing data on plant and soil water isotopes ($\delta^2H$) and coordination of ERD with hydraulic traits and mortality rates.
+Identifying scaling relationships between these curves and commonly available traits (Leaf Mass Area~lamina~, Wood Specific Gravity~100^$\circ$^C~; see Table \@ref(tab:symbols) for definitions) allowed us to scale-up these curves to `r length(data.model.AB$sp)` tree species in the community (of 321 total).
+We evaluated ERD against existing data on plant and soil water isotopes ($\delta^2H$) and tested coordination of ERD with hydraulic traits and mortality rates.
 
@@
 \
-We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and water sourcing depth water isotopes on tree species and soil from an independent isotopic tracer study on trees’ water-sourcing depths [@Meinzer:1999].
+We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and data on stable water isotopes on tree species and soil from an independent study on trees’ water-sourcing depths [@Meinzer:1999].
 The 1,560ha island was formed after the river Chagres was dammed in 1914 that created Gatun Lake, the main channel of the Panama Canal.
-Rainfall at BCI is seasonal with a mean annual rainfall of `r round(mean(clim$Precip))` mm ($pm$ `r round(sd(clim$Precip))` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a four-month long dry season from mid-December through mid-April ($\le$ 100 mm rain/month).
-Mean annual Penman-Monteith evapo-transpiration is `r round(mean(clim$pet.PM))` mm ($pm$ `r round(sd(clim$pet.PM))` SD; `r min(clim$Year)`-`r max(clim$Year)`).
+Rainfall at BCI is seasonal with a mean annual rainfall of `r round(mean(clim$Precip))` mm ($\pm$ `r round(sd(clim$Precip))` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a four-month long dry season from mid-December through mid-April ($\le$ 100 mm rain month^-1^).
+Mean annual Penman-Monteith evapotranspiration is `r round(mean(clim$pet.PM))` mm ($\pm$ `r round(sd(clim$pet.PM))` SD; `r min(clim$Year)`-`r max(clim$Year)`).
 The entire island is forested and classified as a tropical moist forest.
@@
 This inventory represents 321 woody species across the eight censuses.
-Aside from a 2ha patch of ~100-year-old secondary forest at the north-eastern corner of the plot the remaining 48 ha represent old growth forest that was never clear cut for agriculture.
+Aside from a 2ha patch of ~100-year-old secondary forest at the north-eastern corner of the plot the remaining 48ha represent old growth forest that was never clear cut for agriculture.
 The plot (100 m x 50 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001].
@@
 The valley gives rise to the 1.5-ha swamp at the end of the wet season.
-Soils of the plot are homogenous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot (BCI soil survey report).
+Soils of the plot are homogeneous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot [@Baillie.2017].
 The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil.
 Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding.
-Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere [@Croat.1978; Leigh.1982; @Gentry.1990].
+Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere [@Croat.1978; @Leigh.1982; @Gentry.1990].
 
@@
 ### Model description
-
 \
-To estimate dynamics of water availability by depth over the forest’s root zone, we implemented a 1-D hydrological model, the Energy Exascale Earth System Land Model (ELM), coupled with a Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES).
+To estimate dynamics of water availability by depth over the forest’s rooting zone, we implemented a 1-D hydrological model, the Energy Exascale Earth System Land Model (ELM), coupled with a Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES).
 FATES is a state-of-the-art, community-based, open-source model used for studying climate-vegetation interactions.
-FATES is considered a next generation vegetation dynamics model for ELM, with a size-structured group of plants (cohorts) and successional trajectory-based patches using the ecosystem demography approach.
+FATES is a next generation vegetation dynamics model for ELM, with a size-structured group of plants (cohorts) and successional trajectory-based patches based on the ecosystem demography approach.
 FATES couples to ELM by a common interface of water and carbon fluxes.
@@
 \
-Below we describe the equation of water balance in ELM-FATES that is most relevant for our study site, even though ELM-FATES is capable of incorporating numerous other hydrological processes relevant for wetlands and lakes and snow [@Bonan:2013].
-ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\Delta W_{can}$, surface water $\Delta W_{sfc}$, soil water $\Delta W_{liq, z}$, and water in an unconfined aquifer $\Delta W_{a}$ (omitting processes relevant to snow, wetlands or lakes), and can be described as
+Below we describe an equation of water balance in ELM-FATES that is most relevant for our study site, even though ELM-FATES is capable of incorporating numerous other hydrological processes relevant for wetlands and lakes and snow [@Bonan:2013].
+ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\Delta W_{can}$, surface water $\Delta W_{sfc}$, soil water $\Delta W_{liq, z}$, and water in an unconfined aquifer $\Delta W_{a}$ (omitting processes relevant to snow, wetlands or lakes), and can be described as:
 
@@
 \end{equation}
-
 where, $q_{rain}$ is rainfall, $E_{v}$ is evapotranspiration (ET) of the forest, $E_{g}$ is ground evaporation, $q_{over}$ is surface runoff, $q_{h2osfc}$ is runoff from surface water storage, $q_{drai}$ is sub-surface drainage (all in mm H~2~O) and $\Delta t$ is the time-step.
-$z$ represents the depths of exponentially distributed 15 soil layers, here, `r paste(soil.depths, collapse=", ")` (m).
+$z$ represents the depths of exponentially distributed 15 soil layers: $\simeq$ 0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13, 21 and 35 (m).
 
@@
 \
-To parameterise the model, we identified eleven parameters relevant for Eq.
-\@ref(eq:water-balance), determined their ranges based on literature for the study site or for the tropics (Table \@ref(tab:param)), and sampled 5000 ensembles from this parameter space with Latin Hypercube Sampling (LHS).
-In the Conrad catchment soil hydraulic conductivity declined by depth based on existing data for the depths of 12.5 and 60 cm (@Kinner:2004; Table \@ref(tab:param)).
+To parameterize the model, we identified eleven parameters relevant for Eq.
+\@ref(eq:water-balance), determined their ranges based on literature for the study site or for the tropics (Table \@ref(tab:param)), and sampled 5000 ensembles from this global parameter space with Latin Hypercube Sampling (LHS).
+In the Conrad catchment existing data on soil hydraulic conductivity for the depths of 12.5 and 60 cm showed a decline by depth (@Kinner:2004; Table \@ref(tab:param)).
 We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths < 12.5 cm and > 60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig.
-S\@ref(fig:ksat-profiles)).
-For all runs, instead of ELM's default soil texture-based soil water retention curve, we used a curve that we derived from existing data on volumetric water content ($\theta$) vs.
-$\Psi_{soil}$ for depths 15, 40 and 100 cm [@Kupers:2019].
-A @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences among depths were nominal:
+S\@ref(fig:ksat-profiles) [Need to fix SI Figs.
+numbering to begin at 1]).
+For all runs, instead of ELM's default soil texture-based soil water retention curve, we used a curve that we derived using existing data on gravimetric water content vs.
+$\Psi_{soil}$ for depths of 0.15, 0.4 and 1 m [@Kupers:2019].
+We converted gravimetric water content to volumetric (VWC, $\theta$) using a bulk density value of of 0.8 g cm^-3^ measured at a point location (see below).
+A @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences among depths were nominal,
 
@@
 \end{equation}
+where $\Psi_e$ is the air-entry matric potential (200, mm H~2~O), $\theta_{s}$ is the saturated volumetric water content (0.51, cm^3^ cm^-3^), and $b$, an index for soil pore-size distribution (10, unitless), respectively.
 
-where $\Psi_e$ is the air-entry matric potential (200, mmH~2~O), $\theta_{s}$ is the saturated volumetric water content (0.51, cm^3^cm^-3^), and $b$, an index for soil pore-size distribution (10, unitless), respectively.
 \
-We ran ELM-FATES for the 5000 parameter ensembles with hourly climate drivers from 1985-2018 initialised with the stand structure of the 50ha plot in a single site mode ($0.1^\circ$ x $0.1^\circ$ centered at $9.1530^\circ$N, $280.2539^\circ$E).
-As our key interest here was in soil water availability rather than forest dynamics, we ran ELM-FATES in simplified modes: static stand structure and without plant hydrodynamics, and with a single Plant Functional Type (PFT) of evergreen trees.
+We ran ELM-FATES for the 5000 parameter ensembles with hourly climate drivers from 1985-2018 initialized with the stand structure of the 50ha plot in a single site mode ($0.1^\circ$ x $0.1^\circ$ centered at $9.1530^\circ$N, $280.2539^\circ$E).
+As our key interest here was in soil water availability rather than forest dynamics, we ran ELM-FATES in simplified modes: static stand structure and without plant hydrodynamics, and with a single plant functional type (PFT) of evergreen trees.
 The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of dry-deciduous PFT did not significantly alter results (not shown).
@@
 ### Model evaluation
-
 \
 We calibrated the model over 2012-2018 against three key fluxes in Eq.
-\@ref(eq:water-balance), namely, (i) evapotranspiration ($ET \simeq E_{v} + E_{g}$) from the eddy flux tower (2012-2017), (ii) stream discharge for the Conrad Catchment ($q_{run} \simeq q_{over} + q_{drai}$; 2012-2018) and (iii) volumetric water content (VWC, $\theta_{z} \simeq W_{liq, z}$; 2012-2018).
-We collected VWC data (2012-2018) from a point location at the eddy flux tower with a vertical Time Domain Reflectometry (TDR) probe ($\theta_{0-0.15m}$).
+\@ref(eq:water-balance), namely, (i) evapotranspiration ($ET \simeq E_{v} + E_{g}$) from the eddy flux tower (2012-2017; see [SI](#SIData)--RCT:"Matteo, please add a section"), (ii) stream discharge for the Conrad Catchment ($q_{run} \simeq q_{over} + q_{drai}$; 2012-2018) and (iii) volumetric water content (VWC, $\theta_{z} \simeq W_{liq, z}$; 2012-2018).
+We collected VWC data (2012-2018) from a point location at the eddy flux tower with a vertical time domain reflectometry (TDR) probe ($\theta_{0-0.15 m}$).
 This data was calibrated based on measurements of gravimetric soil moisture at the vertical probe (0-15 cm) on several occasions (Fig.
-\@ref(fig:TDR-calib)) and mean bulk dry density of 0.8 gcm^-3^ (*n* = 5, cylinder volume = 1250 cm^3^).
-We also leveraged existing opportunistic dry season gravimetric measurements by depth for the entire 50-ha plot and converted them to VWC units assuming bulk density of 0.8 gcm^-3^  ($\theta_{0.15m}$, $\theta_{0.4m}$ and $\theta_{1m}$ in 2015 and $\theta_{0.15m}$, $\theta_{0.4m}$ in 2016; @Kupers:2019).
-Beginning in 2016 we additionally installed three horizontal probes at depths of 0.15, 0.4 and 1m, but depth specific calibration equation was not available, so we applied the same calibration equation to these data as that of the vertical probe.
-However, for VWC calibration we only used data from the vertical probe to capture the long-term dynamics and plot-wide measurements to capture the dynamics by depth.
-To find parameter ensembles that best-fit all three fluxes simultaneously, we calculated an objective function for each of the 5000 ensembles based on root mean square error (RMSE) between observations and simulations.
-Let $f(obs,~sim)$ be normalised RMSE (0-1) between observations and simulations, based on monthly sums for $ET$ and $q_{run}$ and daily avarages for $\theta$s, then the objective function can be described as:
+S\@ref(fig:TDR-calib)) and mean bulk dry density of 0.8 g cm^-3^ (*n* = 5, cylinder volume = 1250 cm^3^).
+We also leveraged existing opportunistic dry season gravimetric measurements by depth for the entire 50-ha plot, converted to VWC units assuming a bulk density of 0.8 g cm^-3^ ($\theta_{0.15 m}$, $\theta_{0.4m}$ and $\theta_{1 m}$ in 2015 and $\theta_{0.15m}$, $\theta_{0.4m}$ in 2016; @Kupers:2019).
+Beginning in 2016 we additionally installed three horizontal probes at depths of 0.15, 0.4 and 1 m, but as depth specific calibration equation was not available, we applied the same calibration equation to these data as that for the vertical probe.
+However, for VWC calibration we used data from only the vertical TDR probe ($\theta_{0-0.15 m}$) to capture the long-term dynamics and the plot-wide measurements to capture the dynamics by depth.
+Here, in general, our interest was to capture the soil moisture dynamics by depth, so we performed model calibration in a two step process: first, we prioritized model fit with $\theta_{0.1m}$.
+We chose ensembles that minimized root mean square error (RMSE) between $\theta_{0.1m}$ and W~liq,`r soil.depths[8]`m~ (~300 ensembles with standaridised RMSE (range 0-1) across ensembles < 0.2).
+Second, we calculated an objective function for each of the 5000 ensembles by equally weighing standardized RMSE between observations and simulations across all fluxes, and chose 100 ensembles, within the 300 that best-fit $\theta_{0-0.15 m}$, that had the lowest values of the objective function.
+Let $f(obs,~sim)$ be RMSE between observations and simulations standardized across ensembles (range 0-1), based on monthly sums for $ET$ and $q_{run}$, respectively, and daily averages for $\theta$s, then the objective function can be described as
 
 \begin{equation}
-Objective Function = \frac{1}{6}\sum \left(f(ET,~E_{v} + E_{g}) + f(q_{run},~q_{over} + q_{drai}) + f(\theta_{0-0.15m}, {W_{liq,~\bar{z}}})  + f(\theta_{0.15m}, W_{liq,~`r soil.depths[4]`m}) + f(\theta_{0.4m}, W_{liq,~`r soil.depths[6]`m}) + f(\theta_{1m}, W_{liq,~`r soil.depths[8]`m})\right)
+Objective Function = \frac{1}{6}\sum \left(f(ET,~E_{v} + E_{g}) + f(q_{run},~q_{over} + q_{drai}) + f(\theta_{0-0.15m}, {W_{liq,~\bar{z}}}) + f(\theta_{0.15m}, W_{liq,~`r soil.depths[4]`m}) + f(\theta_{0.4m}, W_{liq,~`r soil.depths[6]`m}) + f(\theta_{1 m}, W_{liq,~`r soil.depths[8]`m})\right)
 (\#eq:obj-func)
 \end{equation}
-
 where $W_{liq,~\bar{z}}$ is the daily average of $W_{liq,~z}$ for $z$ = `r paste(soil.depths[1:4], collapse=", ")` m.
-To ensure calibration with $\theta_{0.1m}$, we first selected ensembles with  $f(\theta_{1m}, W_{liq,~`r soil.depths[8]`m})$ < 0.2 (~300 ensembles) and among them selected 100 ensembles with lowest values for the objective function $Z$.
-Model simulations corresponding to these 100 best-fit ensembles are shown against observed fluxes in Fig.
-\@ref(fig:hydro-calib), including VWC from horizontal probes that were not used for calibration.
 See Table \@ref(tab:param) for the range of best-fit values for different parameters.
 We ran ELM-FATES with the chosen 100 ensembles from 1985-2018, with first 5 years used for model spin-up.
-
-```{r hydro-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-hydro-calib} Model evaluation: Observations (red lines) versus simulations (blue lines) from 100 best-fit parameter ensemble runs of ELM-FATES for daily Volumetric Water Content (VWC) by depth for horizontal Time Domain Reflectometry (TDR) probes at three depths (10, 40 and 100 cm; A) and a vertical probe (0-15 cm; B); monthly stream discharge (C) and monthly Evapotranspiration from eddy flux tower (D).
-Points (red) in (A) show average plot-wide observations of VWC with 95% CI (error bars).
-VWC data from the horizontal probes (lines in A) were not used for ELM-FATES calibration.
-Values in inset are average standardised RMSE across the 100 simulations.
-For panel A, values at the top are for plot-wide VWC and those at the bottom are for TDR probes.")}
-knitr::include_graphics(file.path("figures/hydro-calib2.jpeg"))
-```
 
@@
 We measured ($K_{leaf}$ vs.
-$\Psi_{leaf}$) on adult trees of `r length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))` common species (see SI--Joseph please add a section).
-We fitted exponential, sigmoidal, logistic equations to species-level data as prescribed by @Guyot:2011.
-Exponential form was chosen, because based on AIC criteria it best-fitted all but one species, and for the one species it fitted only slightly worse (Fig.
-S\@ref(fig:obs-vuln-curves); we might want to add specie-specific panel of graphs showing data and fitted curves here), described as
+$\Psi_{leaf}$) on adult trees of `r length(sp.exp.param$sp)` common species (see [SI](#SIData)--RCT:"Joseph, please add a section").
+We fitted four different functions to species-spefic data using the *anneal* optimization function with the package likelihood in R: exponential without an intercept ($K_{leaf} = Ae^{-B\Psi_{leaf}}$), exponential with an intercept ($K_{leaf} = C + Ae^{-B\Psi_{leaf}}$), logistic ($K_{leaf} = \frac{A}{1 + \left(\frac{\Psi_{leaf}}{x_{0}}\right)^B}$) and sigmoidal ($K_{leaf} = \frac{A}{1 + e^{-\left(\frac{\Psi_{leaf} - x_{0}}{B}\right)}}$) [@Guyot:2011].
+The differences among AICc values for these functions were not large within species (Fig.
+S\@ref(fig:leaf-vcurves-data1) & S\@ref(fig:leaf-vcurves-data2)).
+We chose the exponential function without the intercept to fit observed data across species, and as a general function to scale-up to the community, because it had the lowest AICc for most species, captured the form of the observed data across all species well, and had the lowest number of parameters, that is, two:
 
@@
 \end{equation}
-
 where $A$ and $B$ are fitted species-specific parameters.
+From these fits we estimated species $K_{max,~leaf}$, $\Psi_{50,~leaf}$ and  $\Psi_{80,~leaf}$ (see Table 1).
 We identified scaling relationships between fitted parameters $A$ and $B$ in Eq.
-\@ref(eq:vc-exp) and two traits that were available at community-scale, namely, mean wood specific gravity after drying at 100$^\circ$C ($WSG$ or, WSG~100^$\circ$^C~; gcm^-3^) and mean leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules ($LMA$ or LMA~lamina~; gm^-2^) for leaves receiving direct sunlight.
-Fitted polynomial equations are described as,
+\@ref(eq:vc-exp) and two traits that were available at the community-scale, namely,  WSG~100^$\circ$^C~ (hereafter, $WSG$, g cm^-3^) and LMA~lamina~ (hereafter, $LMA$; g m^-2^) (see SI for description of how data gaps were filled.
+RCT:"Joe, please add a reference if these data are published, else add an SI section.").
+We fitted polynomial equations described as,
 
@@
 
-$WSG$ & parameter $B$ explained a large proportion of variation in parameter $A$ (Eq.
-\@ref(eq:a-mod); Adj.
-R^2^ = `r signif(glance(k_by_psi.models[[2]])$adj.r.squared, 2)`; *P* = `r signif(glance(k_by_psi.models[[2]])$p.value, 2)`; also see Fig.
-S\@ref(fig:ab-fit)).
-$WSG$ & $LMA$ explained a large proportion of variation in parameter $B$ (Eq.
-\@ref(eq:b-mod); Adj.
-*R*^2^ = `r signif(glance(k_by_psi.models[[1]])$adj.r.squared, 2)`; *P* = `r  signif(glance(k_by_psi.models[[1]])$p.value, 2)`).
-We thus sequentially used Eq.
-\@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r length(unique(data.model.AB$sp))` species for which WSG and LMA data were available and fitted vulnerability curves (see Fig.
-\@ref(fig:comm-vuln-curves)).
-
-```{r comm-vuln-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-vuln-curves} Modelled 'vulnerability curves' ($K_{leaf}$ vs.
-$\\Psi_{leaf}$) at community-scale (", length(data.model.AB$sp)," species) based on data on $WSG$ & $LMA$ and their relationships with vulnerability curve parameters measured on ", sp.soft.filled$AB.data.model," species.
-Each line represents a species, color coded by $WSG$ with blue tones indicating smaller values of $WSG$ and red tones indicating larger values of $WSG$.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB.jpeg"))
-```
+As these fits explained large proportion of variation in parameters $A$ and $B$ (see results), we sequentially used Eq.
+\@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r length(unique(bci.AB$sp[!is.na(bci.AB$A)]))` species for which WSG and LMA were both available, retained `r length(unique(data.model.AB$sp))` for which parameters $A$ and $B$ were both positive, and fitted vulnerability curves to those.
 
@@
 Water is a key limiting resource for tree growth.
-Vapour pressure deficit (VPD) and water supply as determined by soil moisture are factors known to influence trees water uptake [@Lawrence.2019].
-In this study, we used leaf conductivity ($K_leaf$) determined by rhizosphere soil water potential ($\Psi_{soil}$) as an approximation of the soil water limiting factor.
+Vapor pressure deficit (VPD) and water supply as determined by soil moisture are factors known to influence trees water uptake [@Lawrence.2019].
+In this study, we used leaf conductivity ($K_{leaf}$) determined by rhizosphere soil water potential ($\Psi_{soil}$) as an approximation of the soil water limiting factor.
 For dry-deciduous species seasonality in leaf cover may also influence plant water uptake.
 We therefore tested alternate models of growth, aka inverse models for ERD (Eqs.
-\@ref(eq:M2)- \@ref(eq:M4)), in which a species' growth dynamics across census intervals was regressed against census interval means of multiplied effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the fraction of leaf cover (all standardised to 0-1, see below).
+\@ref(eq:M2)- \@ref(eq:M6)), in which a species' growth dynamics across census intervals was regressed against census interval means of multiplied or additive effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the fraction of leaf cover (all standardized to range from 0-1; see below).
 Let n~t~ be the total number of days in census interval $t$, then for individual tree species Growth~t~ and depth $z$ these models are described as follows:
@@
 \begin{equation}
-Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,~z,~n} LeafCover_{doy} + error
+Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,~z,~n} + StdVPD_{n} + error
 (\#eq:M3)
@@
 \begin{equation}
-Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,~z,~n} LeafCover_{doy}StdVPD_{n} + error
+Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,~z,~n} LeafCover_{doy} + error
 (\#eq:M4)
 \end{equation}
-\
-We estimated daily value of $K_{leaf,z,n}$ by plugging in $\Psi_{soil,~z,~n}$ in the leaf PLC curves Eq. \@ref(eq:vc-exp) as:
+
+\begin{equation}
+Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,~z,~n}StdVPD_{n}LeafCover_{doy} + error
+(\#eq:M5)
+\end{equation}
+
+\begin{equation}
+Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} (StdK_{leaf,~z,~n} + StdVPD_{n}) LeafCover_{doy} + error
+(\#eq:M6)
+\end{equation}
+We estimated daily value of $K_{leaf,z,n}$ by substituting $\Psi_{soil,~z,~n}$ in the leaf vulnerability curves Eq. \@ref(eq:vc-exp) as:
 
@@
 \end{equation}
-\
+
 We estimated species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015; see below).
@@
 \
-For growth estiamtes, to minimise the effect of light variation among trees, we selected only large trees ($\ge$ 30 cm dbh, diameter at breast height) and species whose maximum height was at least 30 m (hereafter, canopy species) and thus are likely to be fully exposed to the sun.
+For growth estimates, to minimize the effect of light variation among trees, we selected only large trees ($\ge$ 30 cm dbh, the diameter at breast height) and species whose maximum height was $\ge$ 30 m (hereafter, canopy species) and thus are likely to be fully exposed to the sun.
 We selected trees for which height of diameter measurement was unchanged across all censuses (1990, 1995, 2000, 2005, 2010, 2015).
-For each individual tree (*i*), we calculated growth rates ($Gr_{i,t}$; cm.yr^-1^) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses, divided by the inter-census interval based on the exact date of measurements.
-We eliminated positive outliers as any growth rates amounting to > 75 mm.yr^-1^, as this was the highest rate observed with confidence for fastest growing species [@Condit.2017].
+For each individual tree (*i*), we calculated growth rates ($Gr_{i,t}$; cm yr^-1^) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses, divided by the inter-census interval based on the exact date of measurements.
+We eliminated positive outliers as any growth rates amounting to > 75 mm yr^-1^, as this was the highest rate observed with confidence for fastest growing species [@Condit.2017].
 Negative outliers were removed based on known measurement error, $\sigma_{e} = 0.006214d + 0.903$, where *d* is dbh.
 Any stem whose later dbh measurement fell < 4$\sigma_{e}$ the earlier measurement was removed [@Condit.2017].
-As tree growth rates vary with size, to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($Resid_{i,t} = Gr_{i,t} - f(d_{i, t})$).
+As tree growth rates vary with size [@Muller-Landau:2006], to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($Resid_{i,t} = Gr_{i,t} - f(d_{i, t})$).
 The diameter model of growth ($f(d_{i, t})$), a B-spline based polynomial equation with 5 degrees of freedom, was fitted to growth rates from all trees pooled together as there was insufficient species-specific data for large trees (Fig.
-S\@ref(fig:dbh-model))).
-We retained only those trees that had $Resid_{i,t}$ estimates across all censuses (complete records, total trees = `r sum(g.n$n)`) and standardized (centered and scaled) individual tree time series of $Resid_{i,t}$.
+S\@ref(fig:dbh-model)).
+We retained only those trees that had $Resid_{i,t}$ estimates for all census intervals (total trees = `r sum(g.n$n)`) and standardized (centered and scaled) individual tree time series of $Resid_{i,t}$.
 We then obtained species-level (*n* = `r length(erd.sp)`) time-series of medians of $Resid_{i,t}$ ($Growth_{t}$).
 These were obtained for only those species with complete records on at least three trees (median *n* = `r median(g.n$n)`, max *n* = `r max(g.n$n)` trees).
+
 \
@@
 Based on our preliminary analyses, incorporation of temperature and radiation did not improve our model fitting, thus we mainly focused on the impact of soil moisture, VPD and leaf area in our models.
+
 \
-As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between mean daily values by month of gross primary productivity (gCm^2^day^-1^) and VPD we collected from the eddy flux tower (kPa; $\bar{GPP} = -18.8VPD^2 + 14.5VPD + 5.02 + error = StdVPD$; 2012-2017; Fig.
-S\@ref(fig:gpp-vpd)) and used standardised $\bar{GPP}$ to represent the impact of VPD on growth ($StdVPD$; range 0-1) in Eqs.
-\@ref(eq:M2) and \@ref(eq:M4).
-Due to coincidence of higher temperature, higher radiation, and lower humidity during the dry season at our study site, the VPD impact may as well be a combined impact of temperature, humidity, and radiation.
+As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between mean daily values by month of gross primary productivity (gC m^2^ day^-1^) and VPD we collected from the eddy flux tower (kPa; $\bar{GPP} = -18.8VPD^2 + 14.5VPD + 5.02 + error = StdVPD$; 2012-2017; Fig.
+S\@ref(fig:gpp-vpd)) and used standardized $\bar{GPP}$ to represent the impact of VPD on growth (hereafter, $StdVPD$; range 0-1) in Eqs.
+\@ref(eq:M2), \@ref(eq:M3) \@ref(eq:M5) and \@ref(eq:M6).
+Due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season at our study site, the impact of VPD may as well be a combined impact of temperature, humidity, and radiation.
+
 \
 To explore the effect of species seasonality in leaf cover on growth, we made an informed assumption of species-specific seasonal curves for Leaf Cover Fraction ($LeafCover$, range 0-1, unitless).
-For evergreen species we assumed $LeafCover$ = 1 throughtout the year, whereas for species with all other decisuous classes, we assumed $LeafCover$ = 1 during the wet season (DOY = 150 to 300), given high forest leaf area index for this period (data not shown).
+For evergreen species we assumed $LeafCover$ = 1 throughout the year, whereas for non-evergreen species (various deciduous classes), we assumed $LeafCover$ = 1 during the wet season (DOY = 150 to 300) given the high forest-level leaf area index for this period (not shown), and obtained an estimate of $LeafCover$ for the rest of the year as follows.
 We interpolated species-specific weekly leaf-fall data collected at two similar sites in BCI (1985-2020) to a daily estimate, scaled to express daily leaf fall as a fraction of the total leaf fall in a given year and obtained an average leaf fall fraction by DOY estimate across years ($LeafFall$, range 0-1, unitless).
-For non-evergreen species, we cumulatively deducted $LeafFall$ from $LeafCover$ = 1 beginning DOY = 300 and obtained a $LeafCover$ by DOY curve (\@ref(fig:leaf-cover)).
+Thus for non-evergreen species, we cumulatively deducted $LeafFall$ from $LeafCover$ = 1 beginning DOY = 300 and obtained a $LeafCover$ by DOY curve (Fig.
+S\@ref(fig:leaf-cover)).
 We obtained species-specific parameters $A$ & $B$ for Eq.
-\@ref(eq:k-soil) either from data when available, or a model based on $LMA$ or $WSG$ as described in the earlier section, and used standardised K within species ($Std.K$; range 0-1) in Eqs.
-\@ref(eq:M1)- \@ref(eq:M4).
+\@ref(eq:k-soil) either from data when available, or a model based on $LMA$ or $WSG$ as described in the earlier section, and used standardized K within species ($Std.K$; range 0-1) in Eqs.
+\@ref(eq:M1)- \@ref(eq:M6).
+
 \
-For the combination of each model \@ref(eq:M1)- \@ref(eq:M4), species and hydrological realisation, we obtained AIC, goodness-of-fit (*R*^2^) and Pearson's $\rho$ for each soil depth $z$.
-For each model-species-hydrological realisation we retained only one depth each with the least AIC as the best-fit, and retained only the subset of depths that had a positive Pearson's $\rho$ and *R*^2^ >= 0.5.
-For each model-species, we defined effective rooting depth (ERD) as the median of this subset of depths and calculated standard error.
+For the combination of each model \@ref(eq:M1)- \@ref(eq:M6), species and 100 hydrological realizations, we obtained AIC, goodness-of-fit (*R*^2^) and Pearson's $\rho$ for each soil depth $z$.
+For each combination of model-species-hydrological realization we retained only one depth with the least AIC score (the best-fit), and within those depths retained the ones that had a positive corresponding Pearson's $\rho$ and *R*^2^ >= 0.5 as a valid set.
+For each model-species combination, we defined effective rooting depth (ERD) as the median of these subset of depths and calculated standard error.
 
 ### Model evaluation
-
-For each model, species ERD was regressed against species Xylem sap $\delta^2H$ (%~$\circ$~; *n* = 7).
-We used Xylem sap $\delta^2H$ values given in [@Meinzer:2001] for the dry-season (March 1997) as these showed largest seasonal divergence among species, with a corresponding gradient of $\delta^2H$ in the soil.
-We removed species that are deciduous in March as xylem sap $\delta^2H$ of these species may not be linked with water sourced in March and thus may not comparable to other species with leaves present.
-Fig.
-S\@ref(fig:iso-data).
-We also removed *Guapira* *standleyana*, whose xylem sap isotope value was not well constrained by the soil water isotope depth profile.
-
-The model with VPD and without Leaf Cover had highest goodness-of fit (*R*^2^ = 0.51; *p* = 0.072), followed by model without both VPD and Leaf Cover (Eq.
-\@ref(eq:M1); Fig.
-S\@ref(fig:other-erd-mods)).
-Models with Leaf Cover had only three species with sufficient species-level goodness-of-fit (Eqs.
-\@ref(eq:M3)-\@ref(eq:M4)) versus seven for models without (Eqs.
-\@ref(eq:M1)-\@ref(eq:M2)).
-For rest of the analyses we therefore used species ERD estimates from the model with VPD and without Leaf Cover (Eq.
-\@ref(eq:M2)).
+\
+For each model, species ERD was regressed against species Xylem sap $\delta^2H$ (permil; *n* = 7).
+We used Xylem sap $\delta^2H$ values given in @Meinzer:2001 for the dry-season of March 1997 as these showed largest seasonal divergence among species, with a corresponding gradient of $\delta^2H$ in the soil (Fig.
+S\@ref(fig:iso-data)).
+We removed species that are likely to be deciduous in March as xylem sap $\delta^2H$ of these species may not be linked with water sourced at the time of measurement and thus may not be comparable with other species that had leaves.
+We also removed *G.
+standleyana*, whose xylem sap isotope value was not well constrained by the soil water isotope depth profile.
 
@@
 To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD with maximum stem hydraulic conductance ($K_{max, stem}$), leaf turgor loss point ($\Psi_{tlp}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$).
-See SI for data collection and estimation of these variables (--Brett please add a section).
+See [SI](#SIData) for data collection and estimation of these variables (RCT:"Brett, please add a section").
 
@@
 \
-We calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$ 30 m) and average adbunance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; *n* = `r length(erd.mrate.sp)`).
-For each of the seven census intervals ($t$) in the 34-yr record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successive censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{time}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50ha plot of a species present in cen~1~ and dead in cen~2~, respectively, and $time$ is the duration based on mean dates of cen~1~ and cen~2~.
-We calculated species mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{t=1}^{6} {Mort_{t}}$ and, finally, species mortality anomalies by intervals as $Mort_{anomaly}$ = $Mort_{t}$ - $Mort_{mean}$.
+We calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$ 30 m) and average abundance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; *n* = `r length(erd.mrate.sp)`).
+For each of the seven census intervals ($t$) in the 35-yr record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (% yr^-1^) for two successive censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{time}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees of a species present in cen~1~ and dead in cen~2~, respectively, in the 50ha plot and $time$ is the duration based on mean dates of cen~1~ and cen~2~.
+For each species we calculated mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{t=1}^{6} {Mort_{t}}$.
 To test whether ERD plays a role in alleviating mortality risk, we regressed species ERD against species ($Mort_{mean}$).
-To assess whether this may be due to deeper ERD species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{ERD}$) were below species $\Psi_{crit}$, defined as $\Psi_{80,~leaf}$.
+To assess whether this may be due to deeper ERD species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the $\Psi_{soil,~z}$ simulation period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{soil,~z = ERD}$) were below species $\Psi_{crit}$, which was defined as $\Psi_{50,~leaf}$.
+
 \
@@
 # Results
+## Soil Water Dynamics By Depth
+\
+The ELM-FATES simulations with the chosen best-fit parameter ensembles (Table \@ref(tab:param)) successfully captured the dynamics of soil moisture by depth (Fig.
+\@ref(fig:hydro-calib)A & Fig.
+\@ref(fig:hydro-calib)B), including the observations from the horizontal TDR probes that were not included in the calibration.
+These simulations also simultaneously captured the dynamics in stream discharge (Fig.
+\@ref(fig:hydro-calib)C) and evapotranspiration (Fig.
+\@ref(fig:hydro-calib)D).
+The reduction in parameter range in the best-fit ensembles compared to the global ranges tested showed that the above calibration was sensitive to the following parameters: the Ball-Berry stomatal slope parameter, the ELM rooting distribution parameter that regulates the depth of the rooting profile, soil hydraulic conductivity profile especially at depth, and our adjustment factor to soil hydraulic conductivity to account for macroporosity and direct flow paths (Table \@ref(tab:param)).
+Soil Water Potential dynamics ($\Psi_{soil,~z}$) by depth (z = 0.1 through 13) obtained from the best-fit simulations revealed that census intervals varied in soil droughts in terms of intensity at depth, seasonality and duration (Fig.
+S\@ref(fig:swp-droughts)).
+Nevertheless, across the simulation period (1990-2015) $\Psi_{soil,~z}$ remained above -0.5 MPa from depths 1.7 to 13 m (Fig.
+S\@ref(fig:swp-all-depths)).
+
+```{r hydro-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-hydro-calib}Observations (red lines) versus simulations (blue lines) from 100 best-fit parameter ensemble runs of ELM-FATES are shown for daily Volumetric Water Content (VWC) by depth for horizontal time domain reflectometry (TDR) probes at three depths (0.1, 0.4 and 1 m; A) and a vertical probe (0-0.15 m; B); monthly stream discharge (C) and monthly evapotranspiration from eddy flux tower (D).
+Points (red) in (A) show average plot-wide observations of VWC with 95% CI (error bars).
+VWC data from the horizontal probes (lines in A) were not used for ELM-FATES calibration.
+Values in inset are average standardised RMSE across the 100 simulations.
+RMSE was standardized to range 0-1 across the 5000 simulations.
+For panel A, values at the top are for plot-wide VWC and those at the bottom are for TDR probes.")}
+knitr::include_graphics(file.path("figures/hydro-calib2.jpeg"))
+```
+
+## Leaf Vulnerability Curves
+\
+For the leaf vulnerability curve parameters $A$ & $B$ fitted to the observed data on $K_{leaf}$ vs.
+$\Psi_{leaf}$, $WSG$ & parameter $B$ explained a large proportion of variation in parameter $A$ (Eq.
+\@ref(eq:a-mod); *Adj.
+R*^2^ = `r afit.r2`; *P* `r  afit.p`; also see Fig.
+S\@ref(fig:ab-fit)).
+$WSG$ & $LMA$ explained a large proportion of variation in parameter $B$ (Eq.
+\@ref(eq:b-mod); Adj.
+*R*^2^ = `r bfit.r2`; *P* `r  bfit.p`).
+WSG alone explained `r round(a.wsg.r2*100,0)` % of variation in $A$ ($A = `r a.wsg.cf[1]` `r a.wsg.cf[2]`B + `r a.wsg.cf[3]`B^2 + error$; *P* `r  a.wsg.p`) and `r round(b.wsg.r2*100,0)` % of variation in $B$ ($B = `r b.wsg.cf[1]` `r b.wsg.cf[2]`B + `r b.wsg.cf[3]`B^2 + error$; *P* `r  b.wsg.p`).
+Parameter $B$ alone explained `r round(a.b.r2*100,0)` % of variation in $A$ (*P* `r  a.b.p`), while $LMA$ alone was weakly related to $A$ and was not related to $B$ (not shown).
+The leaf vulnerability curves fitted to observed data on `r length(sp.exp.param$sp)` species are shown in Figs.
+S\@ref(fig:leaf-vcurves-data1), S\@ref(fig:leaf-vcurves-data2) and S\@ref(fig:obs-plc-curves) and those predicted for the `r length(unique(data.model.AB$sp))` species in the community are shown in Fig.
+\@ref(fig:comm-vuln-curves).
+Corresponding parameters for the community are given in Table S\@ref(tab:param) (also see Fig.
+\@ref(fig:comm-plc-curves)).
+
+```{r comm-vuln-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-vuln-curves}Modelled 'vulnerability curves' ($K_{leaf}$ vs.
+$\\Psi_{leaf}$) at community-scale (", length(data.model.AB$sp)," species) based on data on $WSG$ & $LMA$ and their relationships with vulnerability curve parameters measured on ", length(sp.exp.param$sp)," species.
+Each line represents a species, color coded by $WSG$ with blue tones indicating smaller values of $WSG$ and red tones indicating larger values of $WSG$.")}
+
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB.jpeg"))
+```
 
 ## Effective Rooting Depths, Vulnerability to Embolism and Hydraulic Safety Margins
-
 \
-In comparison to the measured isotopic signal, our results showed that the ERD model with VPD and without Leaf Cover (Eq.
-\@ref(eq:M2)) had highest goodness-of fit (*R*^2^ = 0.51; *p* = 0.072) and was thus the chosen model of ERD for rest of the analyses.
-Notably, two species, *Luehea* *seemannii* and *Trichilia tuberculata*, that @Meinzer:1999 found to have xylem sap $\delta^2H$ values between soil water and groundwater, suggesting that these species sourced most of the water from depths > 1 m, were also identified by our model with ERD > 1 m (2.9 m for both species).
-(Fig.
+In comparison to the measured isotopic signal ($\delta^2H$), our results showed that the ERD model with a multiplicative effect of VPD and without Leaf Cover (Eq.
+\@ref(eq:M2)) had highest goodness-of-fit (*R*^2^ = 0.85; *p* = 0.008; Fig.
+\@ref(fig:erd-iso)).
+All the other ERD models showed a positive but non-significant relationship with $\delta^2H$ (Fig.
+S\@ref(fig:other-erd-mods)).
+The model with an additive effect of VPD and those with Leaf Cover had only three-to-four species with sufficient species-level goodness-of-fit (Eqs.
+\@ref(eq:M3)-\@ref(eq:M6)) versus six for models without (Eqs.
+\@ref(eq:M1)-\@ref(eq:M2)).
+For the rest of the analyses we therefore used species ERD estimates from the model with a multiplicative effect of VPD, without Leaf Cover (Eq.
+\@ref(eq:M2)).
+Notably, two species, *L.
+seemannii* and *T.
+tuberculata*, that @Meinzer:1999 found to have xylem sap $\delta^2H$ values between soil water and groundwater, suggesting that these species sourced most of the water from depths > 1 m, were also identified by our model with ERD > 1 m (2.9 m for both species; Fig.
 \@ref(fig:erd-iso)).
@@
 Data from [@Meinzer:1999].
-See Table  \\@ref(tab:spcodes) for scientific names corresponding to species codes.")}
+See Table  \\@ref(tab:ab-param) for the full form of scientific names.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_phenology_Meinzer_gr.Psi.VPD.jpeg"))
@@
 \
-In this forest modelled effective rooting depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from `r min(df.erd.to.plot$depth)` m to `r max(df.erd.to.plot$depth)` m (Fig.
+In this forest modeled effective rooting depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from `r min(df.erd.to.plot$depth)` m to `r max(df.erd.to.plot$depth)` m (Fig.
 \@ref(fig:sp-erd)).
-For the subset of species for which hydraulic properties were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max,~stem}$; *R*^2^ = 0.39, *P* = 0.098, *n* = 8; Fig.
-\@ref(fig:erd-hyd)A), less negative leaf ($\Psi_{tlp}$); *R*^2^ = 0.6, p = 0.024, *n* = 8; Fig.
-\@ref(fig:erd-hyd)B), greater vulnerability to embolism from cavitation with less negative ($\Psi_{88,~stem}$) (*R*^2^ = 0.72, p = 0.007, *n* = 8; Fig.
+For the subset of species for which hydraulic properties were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max,~stem}$, *R*^2^ = 0.39, *P* = 0.098, *n* = 8; Fig.
+\@ref(fig:erd-hyd)A), less negative leaf $\Psi_{tlp}$ (*R*^2^ = 0.6, p = 0.024, *n* = 8; Fig.
+\@ref(fig:erd-hyd)B), less negative $\Psi_{88,~stem}$ and thus greater vulnerability to embolism from cavitation (*R*^2^ = 0.72, p = 0.007, *n* = 8; Fig.
 \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$; *R*^2^ = 0.74, p = 0.013, *n* = 8; Fig.
@@
 ```{r sp-erd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap= paste0("\\label{lab-sp-erd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) for ", length(erd.sp)," large, canopy species of Barro Colorado Island.
-See Table  \\@ref(tab:spcodes) for complete scientific names.")}
+See Table  \\@ref(tab:ab-param) for the full form of scientific names.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/ERD_by_sp_large_canopy.jpeg"))
@@
 
-```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus hydraulic properties for eight large, canopy species found on Barro Colorado Island (Panama).
+```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus hydraulic properties for seven large, canopy species found on Barro Colorado Island (Panama).
 *R*^2^ and *P*-values are given in panel insets.")}
@@
 ## Effective Rooting Depths, Experienced Drought and Mortality
-
 \
@@
 \@ref(fig:erd-mort)).
-For five out of seven census intervals, species with shallower ERD had significantly higher mortality anomalies than deeper ERD species; Fig.
-(\@ref(fig:erd-mort-anom)).
+For five out of seven census intervals, species with shallower ERD had significantly higher mortality anomalies than deeper ERD species (Fig.
+\@ref(fig:erd-mort-census)).
 Analysis of time spent under catastrophic soil drought showed that the proportion of days species $\Psi_{ERD}$ was more negative than species $\Psi_{crit}$ decreased with species ERD (Fig.
 \@ref(fig:erd-time)).
-Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25-yr period for which soil water potential $\Psi_{soil,~z}$ was modelled (1990-2015), as soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dry seasons and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).
+Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25-yr period for which soil water potential $\Psi_{soil,~z}$ was modeled (1990-2015), as soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dry seasons and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).
 
@@
 
-```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{soil,~z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for ", length(erd.mrate.sp), " canopy species found on Barro Colorado Island.
+```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\\Psi_{soil,~z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for ", length(erd.mrate.sp), " canopy species found on Barro Colorado Island.
 Each circle represents a species.
-Circles are colored and jittered vertically around the named depths (ticks) for clarity.
-(Are *R*^2^ and *P*-values needed here?)")}
+Circles are colored and jittered vertically around the named depths (ticks) for clarity.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr.jpeg"))
@@
 Deep ERD species showed higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species.
-This trade-off translated into greater drought tolerance, as deeper ERD species showed lower mortality rates than shallower ERD species over the 34-yr record (1982-2015).
-Analysis of time spent below ($\Psi_{crit}$) in the soil zone matching ERD $\Psi_{soil,~z = ERD}$ showed that this was likely due to decreasing realised risk of hydraulic failure with depth.
-We thus show, for the first time, that interspecific variation in trees' effective rooting depths aligns species on a conservative-acquisitive water economic spectrum by demonstrating a link between trees' realised drought experience, hydraulic architecture and long-term mortality outcomes.
-On the acquisitive end of the spectrum were deep ERD species with hydraulically efficient, but risky suit of traits optimised for an environment of reliable and abundant source of water at depth that actualised to a lower mortality-risk.
-At the conservative end of the spectrum were shallow ERD species with inefficient, but safer xylem, likely optimised for high variability in $\Psi_{soil}$ in the shallow soil layers, resulting in higher variance in mortality rates across censuses as $\Psi_{soil,~z = ERD}$ crossed ERD $\Psi_{crit}$.
+This trade-off translated into greater drought tolerance, as deeper ERD species showed lower mortality rates than shallower ERD species over the 35-yr record (1982-2015).
+Analysis of time spent below ($\Psi_{crit}$) in the soil zone matching ERD $\Psi_{soil,~z = ERD}$ showed that this was likely due to decreasing realized risk of hydraulic failure with depth.
+We thus show, for the first time, that inter-specific variation in trees' effective rooting depths aligns species on a conservative-acquisitive water economic spectrum by demonstrating a link between trees' realized drought experience, hydraulic architecture and long-term mortality outcomes.
+On the acquisitive end of the spectrum were deep ERD species with hydraulically efficient, but risky suit of traits optimized for an environment of reliable and abundant source of water at depth that actualized to a lower mortality-risk.
+At the conservative end of the spectrum were shallow ERD species with inefficient, but safer xylem, likely optimized for high variability in $\Psi_{soil}$ in the shallow soil layers, resulting in higher variance in mortality rates across censuses as $\Psi_{soil,~z = ERD}$ crossed ERD $\Psi_{crit}$.
 Mortality rates of shallow ERD species may have been exacerbated by increased drought events during the study period.
-Previous work shows that half of the extreme dry seasons, estimated as < 1 SD annual water deficit, during 1931-2015 (85-yr) occurred during 1982-2005 (24-yr) of the 34-yr long study period (1982-2015; @Condit.2017).
+@Condit.2017 showed that half of the extreme dry seasons (< 1 SD annual water deficit) during 1931-2015 (85-yr) occurred during 1982-2005 (24-yr) of the 34-yr long study period (1982-2015).
 
@@
 \
-Our study brings attention to the need for assessing drought-sensitivity in terms of species realised drought risk.
-We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins traits, which are commonly  are identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through a study period of greater recurrence of drought.
-These results are consistent with the rare studies that studied three or four species each with known rooting or water sourcing depths and assesed their hydraulic traits and mortality rates during a centennial (Texas) or extreme (Italy) drought [@Johnson:2018fad; @Nardini:2015].
+Our study brings attention to the need for assessing drought-sensitivity in terms of species realized drought risk.
+We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins, that is, traits that are commonly identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through a study period of greater recurrence of drought.
+These results are consistent with the rare studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during a centennial (Texas) or extreme (Italy) drought [@Johnson:2018fad; @Nardini:2015].
 @Johnson:2018fad found that mortality decreased with deep rooting even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi_{88,~stem}$) and had narrower safety-margins ($\Psi_{min} - \Psi_{88,~stem}$).
 @Nardini:2015 found that a species sourcing deep water in a Karst system (Italy) showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi_{88,~stem}$ and narrower safety margin compared to two species sourcing shallow-water.
-Our result of lower mortality in deep rooted trees contrasts with the finding of @Chitra-Tarak.2018, who found that species identified by their inverse ERD model as effectively deep rooted died *more* in a prolonged, centennial drought.
-Their hydrological model found that in the multi-annual drought deep water was exhausted for over two years, whereas our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period $\Psi_{soil}$ below the depth of 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, (median; `r round(as.numeric(psi.1.7.min$q2.5), 2)`, 95% CI) through all the dryseasons and drought periods, that is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).
+Our result of lower mortality in deep rooted trees contrasts with the finding of @Chitra-Tarak.2018 in which species identified by their inverse ERD model as effectively deep rooted died *more* in a prolonged, centennial drought.
+Their hydrological model revealed that the multi-annual drought exhausted the deep water reservoir for over two years, whereas our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period, $\Psi_{soil}$ remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) below the depth of 1 m through all the dryseasons and drought periods, that is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).
 The difference in our results are likely attributable to site differences in the precipitation regime.
-In their dry forest site (Mudumalai National Park, India) precipitation is low (1095 $\pm$ 296 mm; mean $\pm$ SD) compared to BCI (`r round(mean(clim$Precip))` $pm$ `r round(sd(clim$Precip))` mm), and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (PET = `r round(mean(clim$pet.PM))` ($pm$ `r round(sd(clim$pet.PM))` mm), and thus at BCI deep soil layers are effectively recharged annually [@Chitra-Tarak.2018 and this study].
+In their dry forest site (Mudumalai National Park, India) precipitation is low: 1095 $\pm$ 296 mm (mean $\pm$ SD) compared to `r round(mean(clim$Precip))` $\pm$ `r round(sd(clim$Precip))` mm at BCI, and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (PET = `r round(mean(clim$pet.PM))` $\pm$ `r round(sd(clim$pet.PM))` mm), and thus at BCI deep soil layers are effectively recharged annually (Fig.
+S\@ref(fig:swp-all-depths); @Chitra-Tarak.2018).
 We note that the variation in mortality explained by ERD in our study is low (*R*^2^ = 0.11, *P* = 0.075), and other causes may explain observed mortality.
@@
 By estimating ERD for large trees of `r length(erd.sp)` canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species species-rich tropical forests.
-Our chosen ERD model is an advancement over the only other inverse model of ERD (or water uptake depth index, @Chitra-Tarak.2018, *n* = 12 species) in terms of physiologically meaningful representation, fewer parameters, corroboration with tree and soil data on $\delta^2H$.
+Our chosen ERD model is an advancement over the only other inverse model of ERD (@Chitra-Tarak.2018; *n* = 12 species) in terms of physiologically meaningful representation, fewer parameters and corroboration with tree and soil data on $\delta^2H$.
 The model in @Chitra-Tarak.2018 estimated six species-specific phenomenological parameters: three as root fraction for corresponding lumped vadose zone layers, and three as shape parameters for water uptake vs availability curves.
-For each species, they calculated daily water uptake successively from three layers--the supply, to satisfy the demand of the daily maximum evapo-transpiration, a function of PET and LAI.
-They regressed total supply to total demand ratio over census intervals (termed as growth coefficient) against growth (5 data points), and chose values for six parameters that best-fit (highest R^2^, that is, 2 additional parameters).
+For each species, they calculated daily water uptake successively from three layers, that is, the supply, to satisfy the demand of the daily maximum evapotranspiration, a function of PET and LAI.
+They regressed total supply to total demand ratio over census intervals as a growth coefficient against growth (5 data points), and chose values for six parameters that best-fit (highest *R*^2^, that is, 2 additional parameters).
 Our inverse ERD model fits only two parameters by way of regressing depth-specific model against 5 growth data points.
-Our model was constrained on VPD and species leaf vulnerability (PLC) curves, variables that are shown to be physiologically meaningful [add references].
-We benefitted from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA~lamina~ and WSG~100^$\circ$^C~) and parameters of PLC curves for which we collected data on a few species (*n* = 14), a 'hard' trait.
-This allowed us to scale-up PLC curves at the community scale (*n* = 284 species).
-We find that per cent loss of conductivity was more rapid for soft wooded and low LMA species as $\Psi_{leaf}$ declined (Fig.
-S\@ref(fig:comm-plc-curves)A & S\@ref(fig:comm-plc-curves)B).
-Whether such scaling relationships between hard and soft traits hold true for other forests, and their wider applicability, remains to be seen as data on PLC curves are limited in tropical forests, but counter examples exist [Lawren, could you suggest references?
-Also, folks, is our $K_{leaf}$ vs.
-$\Psi_{leaf}$ data the first reported for a tropical forest?
-or could you add how rare it is?].
-We also note the availability of data on $\delta^2H$ on trees and water sources belowground--soil/substrate matrix water in the vadoze zone or free water below the water table--as important for model calibration/validation.
-The model in @Chitra-Tarak.2018 were not corroborated with such data as existing methods were inadequate to extract water bound to the soil and saprolite found in their dry forest study site (Jean Riotte, pers.
+Our model was constrained on VPD and species leaf vulnerability curves, variables that are shown to be physiologically meaningful.
+We benefited from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA~lamina~ and WSG~100^$\circ$^C~) and parameters of leaf vulnerability (or percent loss of conductance, (PLC)) curves for which we collected data on a few species (*n* = `r length(sp.exp.param$sp)`), a 'hard' trait.
+This allowed us to scale-up the leaf vulnerability curves at the community scale (*n* = `r length(ab.table$Species)` species, see Table S\@ref(tab:ab-param) for model predicted parameters).
+We find that leaf conductance declined more rapidly with declining $\Psi_{leaf}$ for species with higher values of WSG and LMA (Fig.
+S\@ref(fig:obs-plc-curves)).
+Whether such scaling relationships between hard and soft traits hold true for other forests, and their wider applicability, remains to be seen as data on PLC curves are limited in tropical forests, but supportive and counter examples both exist [@McCulloh.2011; @Hoffmann.2011] (RCT: "K~leaf~ folks, Suggest if there are more references from the tropics, or relevant ones elsewhere.
+Also, could you comment on how rare this $K_{leaf}$ vs.
+$\Psi_{leaf}$ dataset is?").
+We also note that the availability of data on $\delta^2H$ on trees and water sources below-ground--from soil or substrate pore water in the vadose zone and free water below the water table--as important for model calibration or validation.
+The model in @Chitra-Tarak.2018 were not corroborated with such data as existing methods were inadequate to extract water bound to the saprolite found in their dry forest study site (Jean Riotte, pers.
 comm.).
-The model of belowground water-availability in this study (ELM-FATES) is far more advanced compared to that used by @Chitra-Tarak.2018 and was calibrated with data on soil moisture dynamics by depth (15, 40, 100 cm), as well as, evapotranspiration and stream discharge.
+The model of below-ground water-availability in this study (ELM-FATES) is far more advanced compared to that used by @Chitra-Tarak.2018 and was calibrated with data on long-term soil moisture dynamics at a point location (0-0.15m) and plot-wide measurements by depth (0.15, 0.4 and 1 m), as well as, evapotranspiration and stream discharge.
 @Chitra-Tarak.2018 was calibrated only on stream discharge and water table levels.
 We lacked data on water table levels which would have been useful to constrain water availability dynamics in the unobserved soil layers.
+As in @Chitra-Tarak.2018, we also interpret our ERD estimates as revealing relative differences among species' effective rooting depths rather than in absolute terms.
 
@@
 \
-The trade-offs that we identify between above- and below-ground traits, water environment vertically belowground and mortality are important for representing diversity in Dynamic Global Vegetation Models especially in species-rich forests.
-The hydraulic efficiency-safety trade-off between high values of maximum leaf or stem conductivity (high $K_{max}$) and cavitation-resistance (low $\Psi_{50}$ or $\Psi_{88}$) is hypothesised although community-wide examples are rare [@Hacke:2006; @Gleason:2016].
-We report such a trade-off at both leaf (*n* = 21 species) and branch level (*n* = 27 species).
-We also found that LMA~lamina~ and WSG~100^$\circ$^C~ were strong predictors of PLC curves for leaves, but not for xylem (not shown).
+The trade-offs that we identify between above- and below-ground traits, the water environment vertically below-ground and mortality rates are important for representing diversity in Dynamic Global Vegetation Models especially in species-rich forests.
+The hydraulic efficiency-safety trade-off between high values of maximum leaf or stem conductivity (high $K_{max}$) and cavitation-resistance (low $\Psi_{50}$ or $\Psi_{88}$) is hypothesized although community-wide examples are rare [@Hacke:2006; @Gleason.2016].
+We report such a trade-off at both leaf (*n* = `r length(sp.exp.param$sp)` species) and stem level (*n* = `r length(hyd$KmaxS[!is.na(hyd$KmaxS)])` species; see Fig.S\@ref(fig:se-trade-offs) [RCT:"Nobby et al and Brett, are such trade-offs part of your other manuscripts?
+If not, we could consider moving these Figs.
+to the results section."]).
+We also found that LMA~lamina~ and WSG~100^$\circ$^C~ were strong predictors of PLC curves for leaves, but not for stems (not shown).
 Acquisitive traits such as high $K_{max}$ are expected to be associated with faster growth rates.
-We found a positive but not  significant relationship of adult growth rates ($\ge$ 10cm dbh) with $K_{max, ~leaf}$ (*R*^2^ = 0.13; *p* = 0.126, *n* = `leaf.traits.sp`) and no relationship with $K_{max, ~stem}$ (*n* = `stem.traits.sp` species).
-The relationships between ERD and aboveground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits in order to better represent the functional diversity of tropical forests.
+We found a positive but non- significant relationship of adult growth rates ($\ge$ 10cm dbh) with $K_{max, ~leaf}$ (*R*^2^ = 0.13; *p* = 0.126, *n* = `r length(sp.exp.param$sp)`) and no relationship with $K_{max, ~stem}$ (*n* = `r stem.traits.sp` species).
+The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits in order to better represent the functional diversity of tropical forests.
 
 \
-Our model estimated ERD as a single value, unlike the dynamic ERD modelled by @Chitra-Tarak.2018, but future model developments should investigate seasonal dynamics of ERD, using calibrated on corresponding calibration data on isotopic water sourcing depths.
-Our inverse ERD model was parametarised on 5-yearly growth data, with five data points over a 25-yr period, which decouples climate events and demographic outcomes.
+Our model estimated ERD as a single value, unlike the dynamic ERD modeled by @Chitra-Tarak.2018, but future model developments should investigate seasonal dynamics of ERD, using corresponding calibration data on isotopically determined water-sourcing depths.
+Our inverse ERD model was parameterized on 5-yearly growth data, with five data points over a 25-yr period, which decouples climate events and demographic outcomes.
 Future studies could better constrain the model with higher resolution growth data such as those from dendrometer bands.
@@
 Use of dendrometer data would thus entail accounting for stem hydrostatic flexing [@Zweifel:2018].
-Two of our exploratory ERD models included seasonality in leaf area fraction, but we did not select them as they worsened the fit with growth data and evaluation data ($\delta^2H$, Fig.
+Three of our exploratory ERD models included seasonality in leaf area fraction, but we did not select them as they worsened the fit with growth data and evaluation data ($\delta^2H$, Fig.
 S\@ref(fig:other-erd-mods)).
-We interpret this result as that VPD and PLC curves may be adequate to explaine growth, but also acknowledge that our estimate of seasonality of leaf cover is a tentative estimate based on leaf fall data and assumption of leaf presence during the wet season and omits inter-annual variation.
+We interpret this result as that VPD and PLC curves may be adequate to explain growth, but also acknowledge that our estimates of seasonality of leaf cover is a tentative estimate combining leaf fall data and assumption of leaf presence during the wet season, and omits inter-annual variation.
 Future studies could make use of data from new technologies such as drone based monitoring of species-specific leaf area fraction in species-rich forests [@Park.2019].
-In forests such as BCI where understory tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes, especially from a few growth data points, may need accounting for variation in trees' light environment over time and among trees of the same size class, an issue we minimised by limiting ERD estimates to canopy trees.
+In forests such as BCI where understory tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes, especially from a few growth data points, may need accounting for variation in trees' light environment over time and among trees of the same size class, an issue we minimized by limiting ERD estimates to canopy trees.
 To estimate topographic variation in ERD within or across species future studies may use soil moisture dynamics from a distributed hydrological model [e.g.
@@
 We note that data availability on soil water retention curve ($\Psi_{soil}$ vs.
-VWC), hydraulic conductivity ($K_{soil}$) by depth, soil moisture by depth, stream discharge and evapotranspiration were important for effective calibration of our 1-D hydrological model, and recommend wide-spread data collection of these variables as well as water-table levels in forest-inventory sites to allow for estimation of trees' water environments.
+VWC), hydraulic conductivity ($K_{soil}$) by depth, soil moisture by depth, stream discharge and evapotranspiration were important for effective calibration of our 1-D hydrological model.
+We recommend wide-spread data collection of these variables as well as water-table levels in forest-inventory sites to allow for estimation of trees' water environments.
 Finally, developing an inverse ERD model that accounts for movement of water within the soil-plant-atmosphere-continuum (SPAC) would be a step change from the current statistical models of ERD (this study and @Chitra-Tarak.2018).
-Although data needs for parametarizing such models [@Christoffersen:2016kd2d; @Sperry.1998] could be greater (e.g.
-retention curves for roots), limiting immediate community-wide application, they promise to improve our understanding of plant physiology [e.g.
+Although data needs for parameterizing such models [@Christoffersen:2016kd2d; @Sperry.1998] could be greater (e.g.
+retention curves for roots), which limits immediate community-wide application, such models hold greater promise to improve our understanding of plant physiology [e.g.
 see @Johnson:2018fad].
 
-## Conclusions
+# Conclusions
 
-Yet To be written.
+Establishing relationships between plant's environment, traits and demographic outcomes is imperative for developing a predictive plant ecology, but trees realized water environments and water-access are rarely studied.
+To our knowledge this is the first study to demonstrate a mechanistic link between trees' realized drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.
+The role of deep-water access that we find in alleviating mortality of drought-sensitive trees has important implications for our predictive understanding of tropical forest dynamics under current and future climate.
+Our community-scale framework for modeling effective rooting depths or leaf vulnerability curves indicate the possibilities in expanding the use of critical, hard-earned observations in species-rich forests towards community-scale generalizations.
 
-## Acknowledgements
-## Data Availability
-## Supplementary Information
+# Acknowledgements
 
-```{r ksat-profiles, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ksat-profiles} Generated 5000 curves of soil hydraulic conductivity variation by depth based on Latin Hypercube Sampling within 95% CI of observed data at depths of 12.5 cm and 60 cm [@Kinner:2004], assuming a linear decline between those depths, and values before 12.5 and after 60 cm as that of 12.5 cm and 60 cm, respectively.
+This research was supported as part of the Next Generation Ecosystem Experiments-Tropics, funded by the U.S.
+Department of Energy, Office of Science, Office of Biological and Environmental Research (LANL DOE Contract #7537545).
+Funding was also provided to RCT and SMM through an NSF grant (1137366).
+LR was deputed to Indo-French Cell for Water Sciences on an IRD Fellowship.
+An NSF grant to Smithsonian Institution’s CTFS-ForestGEO (Dimensions 1046113) also partially supported this work via RCT and SMM.
+**Add acknowledgements for other co-authors.
+We acknowledge the following for contributions to the 50-ha plot project: R.
+Foster as plot founder and the first botanist able to identify so many trees in a diverse forest; R.
+Pérez and S.
+Aguilar for species identification; S.
+Lao for data management; S.
+Dolins for database design; plus hundreds of field workers for the census work, now over 2 million tree measurements; the National Science Foundation, Smithsonian Tropical Research Institute, and MacArthur Foundation for the bulk of the financial support.
+
+# Data Availability
+
+**RCT: Upload on NGET-Archive and site DOIs:
+(1) Boris to upload climate forcing data
+(2) Rutuja to upload R scripts for ELM-FATES parameter generations, ELM-FATES modified code, and all R scripts to reproduce manuscript.
+\
+Barro Colorado Forest Census Plot Data, 2012 Version [@Condit.2012], can be obtained from DOI http://dx.doi.org/10.5479/data.bci.20130603.
+More recent versions of the data can be requested directly from R.
+Condit (conditr@gmail.com).
+
+# Supplementary Information
+
+## Data Descriptions {#SIData}
+
+### Data at the Eddy Flux Tower
+Matteo to add. ET, TDR and may be vary briefly, GPP and VPD.
+
+### $K_{leaf}$ vs. $\Psi_{leaf}$
+Joseph et al. to add.
+
+### $K_{stem}$ vs. $\Psi_{stem}$
+Brett to add.
+
+### WSG~100^$\circ$^C~ and LMA~lamina~
+Joe please add, if any.
+
+### Data gaps in WSG~100^$\circ$^C~ and LMA~lamina~
+
+Among the `r length(sp.exp.param$sp)` species with observed data on ($K_{leaf}$ vs.
+$\Psi_{leaf}$) used to build the community-level models of leaf vulnerability curves (Eqs.
+\@ref(eq:b-mod) and \@ref(eq:a-mod)), two lacked data on WSG~100^$\circ$^C~ and five on LMA~lamina~.
+We filled three gaps in LMA~lamina~ using predictions from a strong linear relationship between LMA~lamina~ and LMA~disc~, the latter defined as the mean leaf mass per unit area measured for a 1.483 cm^2^ leaf disc taken to avoid veins (g m^-2^) for leaves receiving direct sunlight ($\bar{LMA_{lamina}} = `r disc.cf[1]` + `r disc.cf[2]`LMA_{disc}$; Adj.
+*R*^2^ = `r disc.r2`; *P* `r disc.p`, *n* = `r disc.n`).
+We filled the two remaining gaps in LMA~lamina~ using predictions from a linear relationship between LMA~lamina~ and LMA~lamina2~, as measured during the smaller, $K_{leaf}$ campaign ($\bar{LMA_{lamina}} = `r lma.cf[1]` + `r lma.cf[2]`LMA_{lamina2}$; Adj.
+*R*^2^ = `r lma.r2`; *P* `r lma.p`, *n* = `r lma.n`).
+We chose not to fill the two data gaps in WSG~100^$\circ$^C~ as available data were poor predictors of WSG~100^$\circ$^C~.
+
+\
+At the community level, we filled data gaps in $LMA_{lamina}$ for `r sum(c.lma.gaps[-which(names(c.lma.gaps) == "original")])` species (of total `r sum(c.lma.gaps)`)  using linear relationships with other variables: `r c.lma.gaps[[which(names(c.lma.gaps) == "filled-LMA-DISC")]]` species with $LMA_{disc}$, `r c.lma.gaps[[which(names(c.lma.gaps) == "filled-LMA-Nobby")]]` with $LMA_{lamina2}$, and `r c.lma.gaps[[which(names(c.lma.gaps) == "filled-LEAF")]]` with $LMA_{leaf}$, defined as the mean leaf mass per unit area measured for the entire leaf including the petiole (g m^-2^) for leaves receiving direct sunlight ($\bar{LMA_{lamina}} = `r leaf.cf[1]` + `r leaf.cf[2]`LMA_{leaf}$; Adj.
+*R*^2^ = `r leaf.r2`; *P* `r leaf.p`, *n* = `r leaf.n`).
+For the `r sum(c.lma.gaps)` species thus with $LMA_{lamina}$, we gap-filled data on  WSG~100^$\circ$^C~ for `r sum(c.wsg.gaps[-which(names(c.wsg.gaps) == "original")])` species using a linear relationship with wood specific gravity reported by @Chave:2009 ($\bar{WSG_{100^{\circ}C}} = `r chave.cf[1]` + `r chave.cf[2]`WSG_{chave}$; Adj.
+*R*^2^ = `r chave.r2`; *P* `r chave.p`, *n* = `r chave.n`).
+
+# Supplementary Figures
+
+```{r ksat-profiles, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ksat-profiles}Generated 5000 curves of soil hydraulic conductivity variation by depth based on Latin Hypercube Sampling within 95% CI of observed data at depths of 12.5 cm and 60 cm [@Kinner:2004], assuming a linear decline between those depths, and values before 12.5 and after 60 cm as that of 12.5 cm and 60 cm, respectively.
 Only 100 curves are shown for clarity.
@@
 
-```{r TDR-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-TDR-calib} Volumetric Water Content vs.
-TDR data for a vertical TDR probe (0-15 cm depth) at the base of Eddy Flux Tower."}
+```{r TDR-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-TDR-calib}Volumetric Water Content vs.
+TDR data for a vertical TDR probe (0-15 cm depth) at the base of the eddy flux tower near the 50ha plot at BCI."}
 knitr::include_graphics(file.path("figures/Matteos_TDR_SWC_data_linear_model.jpeg"))
@@
 
-```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts} Extreme soil droughts by census interval (panels) at three representative depths: 0.5, 1 and 1.7 m, depicted in different colors.
-Mean soil water potential ($Psi_{soil,~z}$) for a given DOY and depth z are shown as lines.
-Filled areas show extreme drought periods, when $Psi_{soil,~z}$ declined below 2.5 percentile of $Psi_{soil,~z}$ for the given DOY and depth z in any given year within the cenus intervals.
-Extreme drought periods from multiple years may be overlaid within a panel.
-Wet season is shown in gray as beginning with DOY."}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q2.5_depths_in_inverse_model.jpeg"))
+```{r leaf-vcurves-data1, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-vcurves-data}Fitted curves vs.
+observed data on $K_{leaf}$ vs.
+$\\Psi_{leaf}$ for ", length(sp.exp.param$sp)," tree species from BCI.
+Curves are color coded by the model used.
+AICc for each model are given in inset (RCT:Add models and their reference).
+Only nine species are shown here.
+The remaining twelve species are shown in Fig.
+\\@ref(fig:leaf-vcurves-data2).")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/vcurves_fits_to_data_by_sp.1.jpeg"))
 ```
 
-```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "20%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd} Scatterplot of mean daily gross primary productivity (GPP) and mean daily vapour pressure deficit (VPD) by month measured at the eddy flux tower in Barro Colorado Island over 2012-2017.
-Red line shows a fit from a polynomial model described in inset, and used to standardise VPD employed in equations \\@ref(eq:M2) and \\@ref(eq:M4)."}
+```{r leaf-vcurves-data2, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-vcurves-data}See Fig.
+\\@ref(fig:leaf-vcurves-data1) caption.")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/vcurves_fits_to_data_by_sp.2.jpeg"))
+```
+
+```{r dbh-model, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-dbh-model}A scatterplot of tree growth rates (circles) from all census intervals versus dbh at the beginning of the census interval is overlaid with a dbh model of growth (spline).
+Growth rates are color coded by census interval.
+The model equation is shown in inset along with goodness-of-fit (Adj.
+*R*^2^).")}
+knitr::include_graphics(file.path("figures/dbh/growthrate_dbh_predicted_growth_with_spline_degree_5_dbh_cutoff_100_model.jpeg"))
+```
+
+```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "35%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd}Scatterplot of mean daily gross primary productivity (GPP) and mean daily vapor pressure deficit (VPD) by month measured at the eddy flux tower in Barro Colorado Island over 2012-2017.
+Red line shows a fit from a polynomial model described in inset, and used to standardize VPD employed in equations \\@ref(eq:M2) and \\@ref(eq:M4)."}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/gpp_by_env_variables/monthly_gpp.vpd.jpeg"))
+```
+
+```{r leaf-cover, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-cover}Seasonal variation in Leaf Cover Fraction for ", length(unique(sp.leaf_cover.for.model$sp[sp.leaf_cover.for.model$deciduousness != "Evergreen"]))," non-evergreen species at Barro Colorado Island.
+Leaf Cover Fraction for evergreen species (*n* = ", length(unique(sp.leaf_cover.for.model$sp[sp.leaf_cover.for.model$deciduousness == "Evergreen"])),") is assumed at 1 through the year.")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/leaf.cover_BCI_single_panel_evergreensAt100.jpeg"))
 ```
@@
 
-```{r leaf-cover, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-cover} Seasonal variation in Leaf Cover Fraction for ", length(unique(sp.leaf_cover.for.model$sp[sp.leaf_cover.for.model$deciduousness != "Evergreen"]))," non-evergreen species at Barro Colorado Island.
-Leaf Cover Fraction for evergreen species (*n* = ", length(unique(sp.leaf_cover.for.model$sp[sp.leaf_cover.for.model$deciduousness == "Evergreen"])),") is assumed at 1 through the year.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/leaf.cover_BCI_single_panel_evergreensAt100.jpeg"))
+```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts}Extreme soil droughts by census interval (panels) at three representative depths: 0.5, 1 and 1.7 m, depicted in different colors.
+Mean soil water potential ($Psi_{soil,~z}$) for a given DOY and depth z are shown as lines.
+Filled areas show extreme drought periods, when $Psi_{soil,~z}$ declined below 2.5 percentile of $Psi_{soil,~z}$ for the given DOY and depth z in any given year within the census intervals.
+Extreme drought periods from multiple years may be overlaid within a panel.
+Wet season is shown in gray beginning with DOY 120."}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q2.5_depths_in_inverse_model.jpeg"))
 ```
 
-```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-other-erd-mods}Effective Rooting Depth versus $\\delta^2H$ for all the models tested.
+```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($Psi_{soil,~z}$) depth z over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines).
+Vertical gray lines show median dates of the demographic censuses at the BCI 50ha plot."}
+knitr::include_graphics(file.path("figures/2019-10-14_5000/best-fits/psi_model_daily_all_depths_params.top.few_full.jpeg"))
+```
+
+```{r ab-fit, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ab-fit}Parameters $A$ (panel A) & $B$ (panel B) of leaf vulnerability curves that were fitted to observed data on $K_{leaf}$ vs.
+$\\Psi_{leaf}$ for ", length(sp.exp.param$sp)," species (Eq.
+\\@ref(eq:vc-exp)) vs.
+those that were predicted from models based on species 'soft-traits', $WSG$ and $LMA$.
+(Eqs.
+\\@ref(eq:a-mod) & \\@ref(eq:b-mod)).
+Goodness-of-fit (*R*^2^) and p-values are given in inset.")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/A_B_data_vs_model.jpeg"))
+```
+
+```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}'Vulnerability curves' fitted to observed $K_{leaf}$ vs.
+$\\Psi_{leaf}$ data on ", length(sp.exp.param$sp)," tree species from BCI (A & B) with equation \\@ref(eq:vc-exp)--in absolute terms (A, B), as well as, in terms of loss of conductivity (%) (C, D).
+Each line represents a species, color coded by $WSG$ (A, C), or $LMA$ (B, D), with blue tones indicating smaller values of $WSG$ or $LMA$ and red tones indicating larger values.")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/plot.obs.plc.jpeg"))
+```
+
+```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-other-erd-mods}Effective Rooting Depth versus $\\delta^2H$ for each of the ERD models tested (panels).
+For each ERD model only those species have an estimate of ERD that have passed the test of Pearson's $\\rho$ > 0 and *R*^2^ > 0.5 during ERD parameterisation.
+The number of species with an ERD estimate as well as $\\delta^2H$.
+or each model are shown in inset, along with the goodness-of-fit (*R*^2^) and p-values for a linear relationship between ERD and $\\delta^2H$.
 See Eqs.
-\\@ref(eq:M1)-\\@ref(eq:M4).")}
+\\@ref(eq:M1)-\\@ref(eq:M6).
+See Table  \\@ref(tab:ab-param) for the full form of scientific names.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_sp_color_Meinzer.jpeg"))
@@
 
-```{r obs-vuln-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-vuln-curves} Leaf 'vulnerability curves', that is, $K_{leaf}$ vs.
-$\\Psi_{leaf}$, fitted to observed data on ",  length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))," tree species from Panama, color coded by Wood Specific Gravity~100^$\\circ$^C~ (gcm^-3^).")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB_for_data_sp.jpeg"))
-```
-
-```{r ab-fit, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ab-fit}Parameters $A$ (panel A) & $B$ (panel B) of leaf vulnerability curves that were fitted to observed data on $K_{leaf}$ vs.
-$\\Psi_{leaf}$ for ", sp.soft.filled$AB.data.model," species vs.
-those that were predicted from models based on species soft-traits, WSG and LMA.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/A_B_data_vs_model.jpeg"))
-```
-
-```{r dbh-model, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-dbh-model} A scatterplot of tree growth rates (circles) from all census intervals versus dbh at the beginning of the cenus interval is overlaid with a dbh model of growth (spline).
-Growth rates are color coded by census interval.
-The model equation is shown in inset along with goodness-of-fit(*R*^2^)")}
-knitr::include_graphics(file.path("figures/dbh/growthrate_dbh_predicted_growth_with_spline_degree_5_dbh_cutoff_100_model.jpeg"))
-```
-
-```{r erd-mort-anom, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-anom}Mortality anomaly in each census interval during the study period plotted against ERD."}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/diff.mortality_by rdi.gr.jpeg"))
-```
-
-```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-plc-curves} Modelled 'vulnerability curves' ($K_{leaf}$ vs.
-$\\Psi_{leaf}$) at community-scale (", length(data.model.AB$sp)," species) shown in terms of loss of conductivity (%) (A & B).
-Each line represents a species, color coded by $WSG$ (A), or $LMA$ (B), with blue tones indicating smaller values and red tones indicating larger values of $WSG$ (A), or $LMA$ (B)")}
+```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-plc-curves}Modelled 'vulnerability curves' ($K_{leaf}$ vs.
+$\\Psi_{leaf}$) at the community-scale (", length(ab.table$Species)," species) predicted from soft traits, $WSG$ and $LMA$, with equations \\@ref(eq:a-mod) and \\@ref(eq:b-mod) in absolute terms (A, B) as well as in terms of loss of conductivity (%) (C, D).
+Each line represents a species, color coded by $WSG$ (A, C), or $LMA$ (B, D), with blue tones indicating smaller values and red tones indicating larger values of $WSG$, or $LMA$.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/plot.comm.plc.jpeg"))
@@
 
-\
-```{r symbols, results = 'asis', echo = FALSE}
-knitr::kable(symbols.table, caption = 'The symbol, definition of key hydraulic traits or variables used.', booktabs = TRUE, escape = FALSE)
+```{r erd-mort-census, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-census}Species mortality rate in each census interval during the study period plotted against ERD."}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_by rdi.gr.jpeg"))
 ```
-\
+
+```{r se-trade-offs, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-se-trade-offs}Efficiency-safety trade-off between $K_{max,~leaf}$ and $\\Psi_{88, leaf}$ (A) and $K_{max,~stem}$ and $\\Psi_{88,~stem}$ based on observed data on ", length(sp.exp.param$sp)," species and ", length(hyd$KmaxS[!is.na(hyd$KmaxS)]), " species, respectively.
+The fitted equations of the form, $K_{max}$ = a + b.log($\\Psi_{88}$), are given in inset along with goodness-of-fit (*R*^2^) and *p*-values.")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/leaf.stem.trade_off.jpeg"))
+```
+
+#Supplementary Tables
+
 ```{r param, results = 'asis', echo = FALSE}
-knitr::kable(param.table, caption = 'ELM-FATES parameters used to generate ensembles, with description, chosen ranges, references and rationale.
-No.
-1-6 are parameters specific to FATES.', booktabs = TRUE, escape = FALSE)
+knitr::kable(param.table, caption = 'ELM-FATES parameters used to generate ensembles, with description, prescribed global ranges, rationale for the choice of ranges and references.
+Ranges for best-fit chosen ensembles are also given.
+Note that no.
+1-6 are parameters specific to FATES.', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(param.table))))
 ```
-\
-```{r spcodes, results = 'asis', echo = FALSE}
-knitr::kable(erd.sp.names, caption = 'Scientific names and codes for tree species for which Effective Rooting Depths were estimated.', booktabs = TRUE
-)
+
+#####
+```{r ab-param, results = 'asis', echo = FALSE}
+knitr::kable(ab.table, caption = 'Modelled parameters A & B for leaf vulnerability curves ($K_{leaf}$ vs.
+$\\Psi_{leaf}$) for 233 species in Panama predicted from soft traits, $WSG$ and $LMA$, using Eqs.
+\\@ref(eq:a-mod) and \\@ref(eq:b-mod).
+Corresponding curves are of the exponetial form as in Eq.
+\\@ref(eq:vc-exp).', booktabs = TRUE, escape = FALSE)
 ```

### 2020-08-29 manuscript.rmd
1. Removed community level curves for 200+ species and moved figure for curves on obs species there. 2. Removed corresponding param table 3. Removed figures for leaf vulnerability curves fit to data on individual species 4. Removed figures on trade-offs beetween K and Psi80 for both leaf and stem

@@
 Corresponding parameters for the community are given in Table S\@ref(tab:param) (also see Fig.
-\@ref(fig:comm-plc-curves)).
+\@ref(fig:obs-plc-curves)).
 
-```{r comm-vuln-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-vuln-curves}Modelled 'vulnerability curves' ($K_{leaf}$ vs.
-$\\Psi_{leaf}$) at community-scale (", length(data.model.AB$sp)," species) based on data on $WSG$ & $LMA$ and their relationships with vulnerability curve parameters measured on ", length(sp.exp.param$sp)," species.
-Each line represents a species, color coded by $WSG$ with blue tones indicating smaller values of $WSG$ and red tones indicating larger values of $WSG$.")}
-
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB.jpeg"))
+```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}'Vulnerability curves' fitted to observed $K_{leaf}$ vs.
+$\\Psi_{leaf}$ data on ", length(sp.exp.param$sp)," tree species from BCI (A & B) with equation \\@ref(eq:vc-exp)--in absolute terms (A, B), as well as, in terms of loss of conductivity (%) (C, D).
+Each line represents a species, color coded by $WSG$ (A, C), or $LMA$ (B, D), with blue tones indicating smaller values of $WSG$ or $LMA$ and red tones indicating larger values.")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/plot.obs.plc.jpeg"))
 ```
@@
 
-```{r leaf-vcurves-data1, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-vcurves-data}Fitted curves vs.
-observed data on $K_{leaf}$ vs.
-$\\Psi_{leaf}$ for ", length(sp.exp.param$sp)," tree species from BCI.
-Curves are color coded by the model used.
-AICc for each model are given in inset (RCT:Add models and their reference).
-Only nine species are shown here.
-The remaining twelve species are shown in Fig.
-\\@ref(fig:leaf-vcurves-data2).")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/vcurves_fits_to_data_by_sp.1.jpeg"))
-```
-
-```{r leaf-vcurves-data2, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-vcurves-data}See Fig.
-\\@ref(fig:leaf-vcurves-data1) caption.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/vcurves_fits_to_data_by_sp.2.jpeg"))
-```
-
 ```{r dbh-model, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-dbh-model}A scatterplot of tree growth rates (circles) from all census intervals versus dbh at the beginning of the census interval is overlaid with a dbh model of growth (spline).
@@
 
-```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}'Vulnerability curves' fitted to observed $K_{leaf}$ vs.
-$\\Psi_{leaf}$ data on ", length(sp.exp.param$sp)," tree species from BCI (A & B) with equation \\@ref(eq:vc-exp)--in absolute terms (A, B), as well as, in terms of loss of conductivity (%) (C, D).
-Each line represents a species, color coded by $WSG$ (A, C), or $LMA$ (B, D), with blue tones indicating smaller values of $WSG$ or $LMA$ and red tones indicating larger values.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/plot.obs.plc.jpeg"))
-```
-
 ```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-other-erd-mods}Effective Rooting Depth versus $\\delta^2H$ for each of the ERD models tested (panels).
@@
 
-```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-plc-curves}Modelled 'vulnerability curves' ($K_{leaf}$ vs.
-$\\Psi_{leaf}$) at the community-scale (", length(ab.table$Species)," species) predicted from soft traits, $WSG$ and $LMA$, with equations \\@ref(eq:a-mod) and \\@ref(eq:b-mod) in absolute terms (A, B) as well as in terms of loss of conductivity (%) (C, D).
-Each line represents a species, color coded by $WSG$ (A, C), or $LMA$ (B, D), with blue tones indicating smaller values and red tones indicating larger values of $WSG$, or $LMA$.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/plot.comm.plc.jpeg"))
-```
-
 ```{r erd-mort-census, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-census}Species mortality rate in each census interval during the study period plotted against ERD."}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_by rdi.gr.jpeg"))
-```
-
-```{r se-trade-offs, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-se-trade-offs}Efficiency-safety trade-off between $K_{max,~leaf}$ and $\\Psi_{88, leaf}$ (A) and $K_{max,~stem}$ and $\\Psi_{88,~stem}$ based on observed data on ", length(sp.exp.param$sp)," species and ", length(hyd$KmaxS[!is.na(hyd$KmaxS)]), " species, respectively.
-The fitted equations of the form, $K_{max}$ = a + b.log($\\Psi_{88}$), are given in inset along with goodness-of-fit (*R*^2^) and *p*-values.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/leaf.stem.trade_off.jpeg"))
 ```
@@
 
-#####
-```{r ab-param, results = 'asis', echo = FALSE}
-knitr::kable(ab.table, caption = 'Modelled parameters A & B for leaf vulnerability curves ($K_{leaf}$ vs.
-$\\Psi_{leaf}$) for 233 species in Panama predicted from soft traits, $WSG$ and $LMA$, using Eqs.
-\\@ref(eq:a-mod) and \\@ref(eq:b-mod).
-Corresponding curves are of the exponetial form as in Eq.
-\\@ref(eq:vc-exp).', booktabs = TRUE, escape = FALSE)
-```
-
 # References

### 2020-08-29 manuscript.rmd
added vpd model coefficients dynamically

@@
 \
-As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between mean daily values by month of gross primary productivity (gC m^2^ day^-1^) and VPD we collected from the eddy flux tower (kPa; $\bar{GPP} = -18.8VPD^2 + 14.5VPD + 5.02 + error = StdVPD$; 2012-2017; Fig.
+As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between mean daily values by month of gross primary productivity (gC m^2^ day^-1^) we collected from the eddy flux tower and VPD (kPa; $\bar{GPP} = `r vpd.cf[1]` + `r vpd.cf[2]`VPD + `r vpd.cf[3]`VPD^2 + error = StdVPD$; 2012-2017; Fig.
 S\@ref(fig:gpp-vpd)) and used standardized $\bar{GPP}$ to represent the impact of VPD on growth (hereafter, $StdVPD$; range 0-1) in Eqs.

### 2020-08-30 manuscript.rmd
1. added fig for ERD species : obs+modelled plc curves 2. references to comm plc curves 3. removed suffix to WSG and LMA 4. removed references to trade-off figs.

@@
 
-## Model for Leaf Vulnerability Curves at Community-scale
+## Leaf Vulnerability Curves
 \
@@
 We fitted four different functions to species-spefic data using the *anneal* optimization function with the package likelihood in R: exponential without an intercept ($K_{leaf} = Ae^{-B\Psi_{leaf}}$), exponential with an intercept ($K_{leaf} = C + Ae^{-B\Psi_{leaf}}$), logistic ($K_{leaf} = \frac{A}{1 + \left(\frac{\Psi_{leaf}}{x_{0}}\right)^B}$) and sigmoidal ($K_{leaf} = \frac{A}{1 + e^{-\left(\frac{\Psi_{leaf} - x_{0}}{B}\right)}}$) [@Guyot:2011].
-The differences among AICc values for these functions were not large within species (Fig.
-S\@ref(fig:leaf-vcurves-data1) & S\@ref(fig:leaf-vcurves-data2)).
+The differences among AICc values for these functions were not large within species.
 We chose the exponential function without the intercept to fit observed data across species, and as a general function to scale-up to the community, because it had the lowest AICc for most species, captured the form of the observed data across all species well, and had the lowest number of parameters, that is, two:
@@
 We identified scaling relationships between fitted parameters $A$ and $B$ in Eq.
-\@ref(eq:vc-exp) and two traits that were available at the community-scale, namely,  WSG~100^$\circ$^C~ (hereafter, $WSG$, g cm^-3^) and LMA~lamina~ (hereafter, $LMA$; g m^-2^) (see SI for description of how data gaps were filled.
-RCT:"Joe, please add a reference if these data are published, else add an SI section.").
+\@ref(eq:vc-exp) and two traits that were available at the community-scale, namely,  WSG and LMA (@Wright:2010; see SI for description of how data gaps were filled.
 We fitted polynomial equations described as,
@@
 Parameter $B$ alone explained `r round(a.b.r2*100,0)` % of variation in $A$ (*P* `r  a.b.p`), while $LMA$ alone was weakly related to $A$ and was not related to $B$ (not shown).
-The leaf vulnerability curves fitted to observed data on `r length(sp.exp.param$sp)` species are shown in Figs.
-S\@ref(fig:leaf-vcurves-data1), S\@ref(fig:leaf-vcurves-data2) and S\@ref(fig:obs-plc-curves) and those predicted for the `r length(unique(data.model.AB$sp))` species in the community are shown in Fig.
-\@ref(fig:comm-vuln-curves).
-Corresponding parameters for the community are given in Table S\@ref(tab:param) (also see Fig.
-\@ref(fig:obs-plc-curves)).
-
-```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}'Vulnerability curves' fitted to observed $K_{leaf}$ vs.
-$\\Psi_{leaf}$ data on ", length(sp.exp.param$sp)," tree species from BCI (A & B) with equation \\@ref(eq:vc-exp)--in absolute terms (A, B), as well as, in terms of loss of conductivity (%) (C, D).
-Each line represents a species, color coded by $WSG$ (A, C), or $LMA$ (B, D), with blue tones indicating smaller values of $WSG$ or $LMA$ and red tones indicating larger values.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/plot.obs.plc.jpeg"))
-```
+Leaf vulnerability curves fitted to observed data on `r length(sp.exp.param$sp)` species as well as percent loss of conuctivity curves are shown in Fig.
+S\@ref(fig:obs-plc-curves).
+Vulnerability curves were predicted using Eqs.
+\@ref(eq:a-mod)-\@ref(eq:b-mod) for `r sp.n.model` species for which such curves were unavailable (out of `r length(unique(erd.iso$sp))` species for which ERD was determined (Fig.
+S\@ref(fig:obs-comm-plc-curves).
 
@@
 
-```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus hydraulic properties for seven large, canopy species found on Barro Colorado Island (Panama).
-*R*^2^ and *P*-values are given in panel insets.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/erd.stem.traits.jpeg"))
+```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Pairwise correlations between Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) and hydraulic properties for seven large, canopy species found on Barro Colorado Island (Panama).
+Scatterplots with linear fits are shown in the lower half.
+Corresponding Spearmon's $\\rho$ are given in the diagonally opposite location in the upper half, along with significance levels, where **, * and .
+represent *P*-value $\\le$ 0.01, 0.05 and 0.1, respectively.
+Diagonal shows density distribution of each variable.")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/erd.stem.traits_cor.chart.jpeg"))
 ```
@@
 The hydraulic efficiency-safety trade-off between high values of maximum leaf or stem conductivity (high $K_{max}$) and cavitation-resistance (low $\Psi_{50}$ or $\Psi_{88}$) is hypothesized although community-wide examples are rare [@Hacke:2006; @Gleason.2016].
-We report such a trade-off at both leaf (*n* = `r length(sp.exp.param$sp)` species) and stem level (*n* = `r length(hyd$KmaxS[!is.na(hyd$KmaxS)])` species; see Fig.S\@ref(fig:se-trade-offs) [RCT:"Nobby et al and Brett, are such trade-offs part of your other manuscripts?
-If not, we could consider moving these Figs.
-to the results section."]).
-We also found that LMA~lamina~ and WSG~100^$\circ$^C~ were strong predictors of PLC curves for leaves, but not for stems (not shown).
+Such a trade-off is found at our study site at both leaf and stem level (Zailaa et al.
+to be submitted; Wolfe et al.
+in revision).
+We also found that LMAand WSG were strong predictors of PLC curves for leaves, but not for stems (not shown).
 Acquisitive traits such as high $K_{max}$ are expected to be associated with faster growth rates.
@@
 
+```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}'Vulnerability curves' fitted to observed $K_{leaf}$ vs.
+$\\Psi_{leaf}$ data on ", length(sp.exp.param$sp)," tree species from BCI (A & B) with equation \\@ref(eq:vc-exp)--in absolute terms (A, B), as well as, in terms of loss of conductivity (%) (C, D).
+Each line represents a species, color coded by $WSG$ (A, C), or $LMA$ (B, D), with blue tones indicating smaller values of $WSG$ or $LMA$ and red tones indicating larger values.")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/plot.obs.plc.jpeg"))
+```
+
+```{r obs-comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-comm-plc-curves}'Vulnerability curves' for species with ERD estimates.
+Curves in  data on ", length(sp.exp.param$sp)," tree species from BCI (A & B) with equation \\@ref(eq:vc-exp)--in absolute terms (A, B), as well as, in terms of loss of conductivity (%) (C, D).
+Each line represents a species, color coded by source of the curve: fitted to observed data (yellow ochre) on $K_{leaf}$ vs.
+$\\Psi_{leaf}$ or based on predicted model parameters $A$ and $B$ from soft traits.
+(gray).")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/obs.mod_plccurves_fits_ggplot_no_col.jpeg"))
+```
+
 ```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-other-erd-mods}Effective Rooting Depth versus $\\delta^2H$ for each of the ERD models tested (panels).

### 2020-08-30 manuscript.rmd
1. Part of changes as per co-author comments: NAte, Joe, Charlie, Stuart

@@
 
-- Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism.
+- Among the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied.
 Vulnerability to embolism and hydraulic safety margins are shown to be predictors of drought-induced mortality risk, but deep-water access may delay plants from reaching these critical thresholds.
@@
 Tropical forests play a disproportionately large role in the global carbon and energy cycle, and support half of global biodiversity [@Bonan:2008; @Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019].
-Tropical forests are considered an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw].
-Historic climatic variation is relatively small [@Mora.2013], and the observed ranges of trees' hydraulic safety margins are relatively narrow [@Choat:2012; but see @Oliveira:2019].
-Widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010].
+Tropical forests are logically considered to be an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw], yet mortality events in the tropics are rarely as large as those in the temperate and boreal zones (McDowell et al.
+2018), leading to questions regarding the role of large trait-diversity in mitigating mortality events.
+Despite this, widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010] and tropical mortality rates are increasing in some regions [@Brienen.2015; @Hubau.2020].
 However, state-of-the-art dynamic vegetation models (DVMs) fail to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013], as underlying mechanisms of drought-tolerance are not fully understood or quantified.
@@
 This mechanism assumes that water availability varies below-ground by vertical depth and so do rooting depths of co-occurring trees, by species and size [@Meinzer:1999; @Chitra-Tarak.2018].
-Consequently, trees in the same forest differ in their drought experience, and thus in their responses in term of growth and mortality.
+Consequently, trees in the same forest differ in their drought experience, and thus in their growth and mortality responses.
 A key bottleneck in community-wide testing of this mechanism has been lack of data on trees' rooting or water-sourcing depths, followed by data on water-availability at those depths.
-A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the world’s 65 thousand tree species [@Fan:2017; @Beech:2017].
-Further, only a small fraction of those are tropical, even though ~60-80 % (40 - 53 thousand species) of the world’s tree diversity resides in the tropics [@Slik:2015].
-Use of stable isotopes of water as a tracer provides an indirect way of trees' water sourcing depths by matching isotopic value in plant tissue to variations in soil pore water vertically below-ground.
+A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the world’s 60 thousand tree species [@Fan:2017; @Beech:2017].
+Further, only a small fraction of those are tropical, even though ~70-90 % (40 - 53 thousand species) of the world’s tree diversity resides in the tropics [@Slik:2015].
+Use of stable isotopes of water as a tracer provides an indirect measure of water sourcing depths by matching isotopic value in plant tissue to variations in soil pore water with depth.
 However, such data are extremely rare [@Evaristo:2016].
@@
 Specifically for the water environment (or, resource), observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified an acquisitive-conservative resource economic spectrum similar to those for light and nutrients, although, for water, examples of all trade-offs observed simultaneously are rare, and below-ground traits or environment linkages are weak [@Reich:2014; @Sterck:2011].
-Acquisitive species have traits with high capacity to acquire, move, store and use water (high $K_{max}$, low wood density) and achieve high growth potential, but build flimsy, disposable tissues (high SLA) and are less tolerant of low resources, whereas species with conservative traits are better protected from carbon losses and are more drought tolerant (low $\Psi_{50}$, low $\Psi_{tlp}$, larger safety margin; see Table 1 for trait symbols and their definitions).
+Acquisitive species have traits with high capacity to acquire, move, store and use water (high $K_{max}$, low wood density; see Table 1 for trait symbols and their definitions) and achieve high growth potential, but build flimsy, disposable tissues (high SLA) and are less tolerant of low resource availablity, whereas species with conservative traits are better protected from carbon losses and are more drought tolerant (low $\Psi_{50}$, low $\Psi_{tlp}$, larger safety margin).
 Acquisitive species are spatially associated with wet sites (high resource) such as valleys or those with high rainfall [@Kursar:2009; @Oliveira:2019], but whether and how the acquisitive-conservative spectrum holds true for the water dynamics vertically below-ground is not clear.
-Deep rootedness may be expected to support an acquisitive strategy with a steady water supply unlike shallow-rooted trees.
+Deep rooting may be expected to support an acquisitive strategy with a steady water supply unlike shallow-rooted trees.
 However, this may depend on the exact dynamics of water-availability by depth at a site, depending on precipitation regime and hydrology.
-In pulse-driven arid systems *shallow*-rooted plants are associated with acquisitive traits (high $K_{max}$) that maximise the use of the water-pulse, while *deep*-rooted plants show conservative traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013 and references therein].
+In pulse-driven systems *shallow*-rooted plants are associated with acquisitive traits (high $K_{max}$) that maximise the use of the water-pulse, while *deep*-rooted plants show conservative traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013 and references therein].
 
 \
-Huge efforts are invested in linking the same-for-all-trees drought indices, above-ground traits and demographic rates.
+Significant efforts have been invested into identifying and linking universal drought indices, above-ground traits, and demographic rates.
 These efforts have resulted in identification of above-ground traits such as proxies of hydraulic failure as indicators of mortality risk.
-A recent meta-analysis found greater drought-induced mortality rates in species with cavitation-vulnerable xylem ($\Psi_{88,~stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~stem}$) [@Anderegg:2016], but counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013]; indicating that plants reach critical thresholds of hydraulic failure depending on their realized drought intensity.
-However, plants' realized water environments are rarely studied, especially in the tropical region.
-To better understand how below-ground rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tropical forest tree species, we make a thorough assessment for a tropical forest of trade-offs between the below-ground water environment, tress' above-ground hydraulic traits, estimates of effective rooting depths and demographic outcomes over 35 years that witnessed droughts of various intensity and duration.
-Specifically, we test an overarching hypothesis that inter-specific variation in trees' effective rooting depths aligns species on an acquisitive-conservative economic spectrum optimized for increasingly reliable water availability by depth such that (i) trees with deeper ERD have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem: higher maximum stem hydraulic conductivity ($K_{max}$), but greater xylem vulnerability to embolism ($\Psi_{88,~stem}$), higher $\Psi_{tlp}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$), whereas shallower ERD species are associated with less efficient but safer hydraulic traits; and that, (ii) deeper ERD species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallower ERD species may have higher mortality rates as a result of higher variability in shallow soil water availability.
+While correlations between aboveground hydraulic traits and mortality have been identified [@Anderegg:2016], numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013].
+However, plants' realized water environments, specifically their ability to acquire water via differing rooting depths, are rarely studied, especially in the tropical region.
+To better understand how below-ground rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tropical forest tree species, we make a thorough assessment for a tropical forest of trade-offs between the below-ground water environment, above-ground hydraulic traits, estimates of effective rooting depths and demographic outcomes over 35 years that witnessed droughts of various intensity and duration.
+Specifically, we test an overarching hypothesis that inter-specific variation in trees' effective rooting depths aligns species on an acquisitive-conservative economic spectrum optimized for increasingly reliable water availability by depth.
+Our hypotheses are that (i) trees with deeper ERD have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem: higher maximum stem hydraulic conductivity ($K_{max}$), but greater xylem vulnerability to embolism ($\Psi_{88,~stem}$), higher $\Psi_{tlp}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$), whereas shallower ERD species are associated with less efficient but safer hydraulic traits; and that, (ii) deeper ERD species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallower ERD species may have higher mortality rates as a result of higher variability in shallow soil water availability.
 To our knowledge this is the first study to demonstrate a mechanistic link between trees' realized drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.
@@
 We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and data on stable water isotopes on tree species and soil from an independent study on trees’ water-sourcing depths [@Meinzer:1999].
-The 1,560ha island was formed after the river Chagres was dammed in 1914 that created Gatun Lake, the main channel of the Panama Canal.
-Rainfall at BCI is seasonal with a mean annual rainfall of `r round(mean(clim$Precip))` mm ($\pm$ `r round(sd(clim$Precip))` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a four-month long dry season from mid-December through mid-April ($\le$ 100 mm rain month^-1^).
-Mean annual Penman-Monteith evapotranspiration is `r round(mean(clim$pet.PM))` mm ($\pm$ `r round(sd(clim$pet.PM))` SD; `r min(clim$Year)`-`r max(clim$Year)`).
-The entire island is forested and classified as a tropical moist forest.
-Although percentage of canopy species that are at least partly deciduous during the dry season is 28%, the maximum percentage of deciduous individuals in the canopy is only 6.3%, and the percentage of crown area that is deciduous is estimated at 9.7% [@Condit:2000].
-Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014].
+The 1,560ha island was formed after the river Chagres was dammed in 1914 to create Gatun Lake and complete the Panama Canal.
+Rainfall at BCI is seasonal with a mean annual rainfall of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a four-month long dry season from mid-December through mid-April ($\le$ 100 mm rain month^-1^).
+The entire island is forested and classified as a tropical moist forest in the Holdridge Life Zone system.
+Although 28% of canopy tree species are at least partly deciduous during the dry season, the maximum percentage of deciduous individuals in the canopy is only 6.3%, and the percentage of crown area that is deciduous is estimated at 9.7% [@Condit:2000].
+Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014].
 
 \
-In the 50ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species level and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm size class [@Condit.2017].
+In the 50 ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species level and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm size class [@Condit.2017].
 This inventory represents 321 woody species across the eight censuses.
 Aside from a 2ha patch of ~100-year-old secondary forest at the north-eastern corner of the plot the remaining 48ha represent old growth forest that was never clear cut for agriculture.
-The plot (100 m x 50 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001].
-The plot covers most of the upper catchment of Conrad Creek.
-The shallow valley of this creek is the main interruption to the otherwise mostly flat or gently sloping plot.
-The valley gives rise to the 1.5-ha swamp at the end of the wet season.
+The plot (1000 m x 500 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001].
+The plot is situated in the Conrad catchment.
 Soils of the plot are homogeneous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot [@Baillie.2017].
@@
 \
-To estimate dynamics of water availability by depth over the forest’s rooting zone, we implemented a 1-D hydrological model, the Energy Exascale Earth System Land Model (ELM), coupled with a Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES).
+To estimate dynamics of water availability by depth over the forest’s rooting zone, we implemented a 1-D hydrological model, the Energy Exascale Earth System Land Model (ELM), coupled with the Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES).
 FATES is a state-of-the-art, community-based, open-source model used for studying climate-vegetation interactions.
-FATES is a next generation vegetation dynamics model for ELM, with a size-structured group of plants (cohorts) and successional trajectory-based patches based on the ecosystem demography approach.
+FATES is a next generation vegetation dynamics model for ELM, with a size-structured group of plants (cohorts) and successional trajectory-based patches based on the ecosystem demography approach [@Moorcroft.2001].
 FATES couples to ELM by a common interface of water and carbon fluxes.
-Detailed descriptions can be found elsewhere [@Koven:2020; @Fisher:2014; @Bonan:2013].
+Detailed descriptions for ELM and FATES can be found elsewhere [@Koven:2020; @Fisher:2014; @Bonan:2013].
 
@@
 Below we describe an equation of water balance in ELM-FATES that is most relevant for our study site, even though ELM-FATES is capable of incorporating numerous other hydrological processes relevant for wetlands and lakes and snow [@Bonan:2013].
-ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\Delta W_{can}$, surface water $\Delta W_{sfc}$, soil water $\Delta W_{liq, z}$, and water in an unconfined aquifer $\Delta W_{a}$ (omitting processes relevant to snow, wetlands or lakes), and can be described as:
+ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\Delta W_{can}$, surface water $\Delta W_{sfc}$, soil water $\Delta W_{liq, z}$, and water in an unconfined aquifer $\Delta W_{a}$ (omitting processes relevant to snow, wetlands or lakes).
+The equation can be described as:
 
@@
 where, $q_{rain}$ is rainfall, $E_{v}$ is evapotranspiration (ET) of the forest, $E_{g}$ is ground evaporation, $q_{over}$ is surface runoff, $q_{h2osfc}$ is runoff from surface water storage, $q_{drai}$ is sub-surface drainage (all in mm H~2~O) and $\Delta t$ is the time-step.
-$z$ represents the depths of exponentially distributed 15 soil layers: $\simeq$ 0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13, 21 and 35 (m).
+$z$ represents the depths of 15 exponentially distributed soil layers: $\simeq$ 0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13, 21 and 35 (m).
 
@@
 \
-We ran ELM-FATES for the 5000 parameter ensembles with hourly climate drivers from 1985-2018 initialized with the stand structure of the 50ha plot in a single site mode ($0.1^\circ$ x $0.1^\circ$ centered at $9.1530^\circ$N, $280.2539^\circ$E).
-As our key interest here was in soil water availability rather than forest dynamics, we ran ELM-FATES in simplified modes: static stand structure and without plant hydrodynamics, and with a single plant functional type (PFT) of evergreen trees.
+We ran ELM-FATES for the 5000 parameter ensembles with hourly climate drivers from 1985-2018 [@Faybishenko.2020] initialized with the stand structure of the 50 ha plot [@Condit.2019a] in a single site mode ($0.1^\circ$ x $0.1^\circ$ centered at $9.1530^\circ$N, $280.2539^\circ$E).
+As our key interest here was in soil water availability rather than forest dynamics, we ran ELM-FATES in a lower-complexity configuration: static stand structure, and with a single plant functional type (PFT) of evergreen trees.
 The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of dry-deciduous PFT did not significantly alter results (not shown).
+Holding the stand structure static rather than allowing it to emerge from physiologically-mediated competition between trees removes a set of internal model feedbacks and thus enables a cleaner experimental design where the only differences between model simulations are the direct physiological effects of perturbed model parameters.
 
@@
 We calibrated the model over 2012-2018 against three key fluxes in Eq.
-\@ref(eq:water-balance), namely, (i) evapotranspiration ($ET \simeq E_{v} + E_{g}$) from the eddy flux tower (2012-2017; see [SI](#SIData)--RCT:"Matteo, please add a section"), (ii) stream discharge for the Conrad Catchment ($q_{run} \simeq q_{over} + q_{drai}$; 2012-2018) and (iii) volumetric water content (VWC, $\theta_{z} \simeq W_{liq, z}$; 2012-2018).
+\@ref(eq:water-balance), namely, (i) evapotranspiration ($ET \simeq E_{v} + E_{g}$) from the eddy flux tower (2012-2017; see [SI](#SIData)), (ii) stream discharge for the Conrad Catchment ($q_{run} \simeq q_{over} + q_{drai}$; 2012-2018) and (iii) soil volumetric water content (VWC, $\theta_{z} \simeq W_{liq, z}$; 2012-2018).
 We collected VWC data (2012-2018) from a point location at the eddy flux tower with a vertical time domain reflectometry (TDR) probe ($\theta_{0-0.15 m}$).
@@
 We also leveraged existing opportunistic dry season gravimetric measurements by depth for the entire 50-ha plot, converted to VWC units assuming a bulk density of 0.8 g cm^-3^ ($\theta_{0.15 m}$, $\theta_{0.4m}$ and $\theta_{1 m}$ in 2015 and $\theta_{0.15m}$, $\theta_{0.4m}$ in 2016; @Kupers:2019).
-Beginning in 2016 we additionally installed three horizontal probes at depths of 0.15, 0.4 and 1 m, but as depth specific calibration equation was not available, we applied the same calibration equation to these data as that for the vertical probe.
+Beginning in 2016 we additionally installed three horizontal probes at depths of 0.15, 0.4 and 1 m and we applied the same calibration equation to these data as that for the vertical probe.
 However, for VWC calibration we used data from only the vertical TDR probe ($\theta_{0-0.15 m}$) to capture the long-term dynamics and the plot-wide measurements to capture the dynamics by depth.
@@
 \begin{equation}
-Objective Function = \frac{1}{6}\sum \left(f(ET,~E_{v} + E_{g}) + f(q_{run},~q_{over} + q_{drai}) + f(\theta_{0-0.15m}, {W_{liq,~\bar{z}}}) + f(\theta_{0.15m}, W_{liq,~`r soil.depths[4]`m}) + f(\theta_{0.4m}, W_{liq,~`r soil.depths[6]`m}) + f(\theta_{1 m}, W_{liq,~`r soil.depths[8]`m})\right)
+Objective Function = \frac{1}{6}\sum \left(f(ET,~E_{v} + E_{g}) + f(q_{run},~q_{over} + q_{drai}) + f(\theta_{0-0.15m}, {W_{liq,~\overline{z}}}) + f(\theta_{0.15m}, W_{liq,~`r soil.depths[4]`m}) + f(\theta_{0.4m}, W_{liq,~`r soil.depths[6]`m}) + f(\theta_{1 m}, W_{liq,~`r soil.depths[8]`m})\right)
 (\#eq:obj-func)
 \end{equation}
-where $W_{liq,~\bar{z}}$ is the daily average of $W_{liq,~z}$ for $z$ = `r paste(soil.depths[1:4], collapse=", ")` m.
+where $W_{liq,~\overline{z}}$ is the daily average of $W_{liq,~z}$ for $z$ = `r paste(soil.depths[1:4], collapse=", ")` m.
 See Table \@ref(tab:param) for the range of best-fit values for different parameters.
@@
 We measured ($K_{leaf}$ vs.
-$\Psi_{leaf}$) on adult trees of `r length(sp.exp.param$sp)` common species (see [SI](#SIData)--RCT:"Joseph, please add a section").
-We fitted four different functions to species-spefic data using the *anneal* optimization function with the package likelihood in R: exponential without an intercept ($K_{leaf} = Ae^{-B\Psi_{leaf}}$), exponential with an intercept ($K_{leaf} = C + Ae^{-B\Psi_{leaf}}$), logistic ($K_{leaf} = \frac{A}{1 + \left(\frac{\Psi_{leaf}}{x_{0}}\right)^B}$) and sigmoidal ($K_{leaf} = \frac{A}{1 + e^{-\left(\frac{\Psi_{leaf} - x_{0}}{B}\right)}}$) [@Guyot:2011].
-The differences among AICc values for these functions were not large within species.
+$\Psi_{leaf}$) on adult trees of `r length(sp.exp.param$sp)` common species (see [SI](#SIData)).
+We fitted four different functions to species-specific data using the *anneal* optimization function with the package likelihood in R: exponential without an intercept ($K_{leaf} = Ae^{-B\Psi_{leaf}}$), exponential with an intercept ($K_{leaf} = C + Ae^{-B\Psi_{leaf}}$), logistic ($K_{leaf} = \frac{A}{1 + \left(\frac{\Psi_{leaf}}{x_{0}}\right)^B}$) and sigmoidal ($K_{leaf} = \frac{A}{1 + e^{-\left(\frac{\Psi_{leaf} - x_{0}}{B}\right)}}$) [@Guyot:2011].
+The differences among AICc values for these functions were not large within species (Zailaa et al.
+to be submitted).
 We chose the exponential function without the intercept to fit observed data across species, and as a general function to scale-up to the community, because it had the lowest AICc for most species, captured the form of the observed data across all species well, and had the lowest number of parameters, that is, two:
@@
 Water is a key limiting resource for tree growth.
-Vapor pressure deficit (VPD) and water supply as determined by soil moisture are factors known to influence trees water uptake [@Lawrence.2019].
+Vapor pressure deficit (VPD) and water supply as determined by soil moisture are factors known to influence water uptake [@Lawrence.2019].
 In this study, we used leaf conductivity ($K_{leaf}$) determined by rhizosphere soil water potential ($\Psi_{soil}$) as an approximation of the soil water limiting factor.
@@
 For growth estimates, to minimize the effect of light variation among trees, we selected only large trees ($\ge$ 30 cm dbh, the diameter at breast height) and species whose maximum height was $\ge$ 30 m (hereafter, canopy species) and thus are likely to be fully exposed to the sun.
-We selected trees for which height of diameter measurement was unchanged across all censuses (1990, 1995, 2000, 2005, 2010, 2015).
+We selected trees for which the height of the diameter measurements were unchanged across all censuses (1990, 1995, 2000, 2005, 2010, 2015).
 For each individual tree (*i*), we calculated growth rates ($Gr_{i,t}$; cm yr^-1^) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses, divided by the inter-census interval based on the exact date of measurements.
@@
 We retained only those trees that had $Resid_{i,t}$ estimates for all census intervals (total trees = `r sum(g.n$n)`) and standardized (centered and scaled) individual tree time series of $Resid_{i,t}$.
-We then obtained species-level (*n* = `r length(erd.sp)`) time-series of medians of $Resid_{i,t}$ ($Growth_{t}$).
-These were obtained for only those species with complete records on at least three trees (median *n* = `r median(g.n$n)`, max *n* = `r max(g.n$n)` trees).
+We then obtained `r length(erd.sp)` species-level time-series of medians of $Resid_{i,t}$ ($Growth_{t}$).
+These were obtained for only those species with complete records on at least three trees (median *n* = `r median(g.n$n)`, max *n* = `r max(g.n$n)` trees per species).
 
@@
 \
-As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between mean daily values by month of gross primary productivity (gC m^2^ day^-1^) we collected from the eddy flux tower and VPD (kPa; $\bar{GPP} = `r vpd.cf[1]` + `r vpd.cf[2]`VPD + `r vpd.cf[3]`VPD^2 + error = StdVPD$; 2012-2017; Fig.
-S\@ref(fig:gpp-vpd)) and used standardized $\bar{GPP}$ to represent the impact of VPD on growth (hereafter, $StdVPD$; range 0-1) in Eqs.
+As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between mean daily values by month of gross primary productivity (gC m^2^ day^-1^) we collected from the eddy flux tower located 75 m from the 50-ha plot and VPD (kPa; $\overline{GPP} = `r vpd.cf[1]` + `r vpd.cf[2]`VPD + `r vpd.cf[3]`VPD^2 + error = StdVPD$; 2012-2017; Fig.
+S\@ref(fig:gpp-vpd)) and used standardized $\overline{GPP}$ to represent the impact of VPD on growth (hereafter, $StdVPD$; range 0-1) in Eqs.
 \@ref(eq:M2), \@ref(eq:M3) \@ref(eq:M5) and \@ref(eq:M6).
@@
 \
-To explore the effect of species seasonality in leaf cover on growth, we made an informed assumption of species-specific seasonal curves for Leaf Cover Fraction ($LeafCover$, range 0-1, unitless).
+To explore the effect of species seasonality in leaf cover on growth, we assumed of species-specific seasonal curves for Leaf Cover Fraction ($LeafCover$, range 0-1, unitless).
 For evergreen species we assumed $LeafCover$ = 1 throughout the year, whereas for non-evergreen species (various deciduous classes), we assumed $LeafCover$ = 1 during the wet season (DOY = 150 to 300) given the high forest-level leaf area index for this period (not shown), and obtained an estimate of $LeafCover$ for the rest of the year as follows.
@@
 \
-For each model, species ERD was regressed against species Xylem sap $\delta^2H$ (permil; *n* = 7).
-We used Xylem sap $\delta^2H$ values given in @Meinzer:2001 for the dry-season of March 1997 as these showed largest seasonal divergence among species, with a corresponding gradient of $\delta^2H$ in the soil (Fig.
+For each model, species ERD was regressed against species xylem sap $\delta^2H$ (permil; *n* = 7).
+We used xylem sap $\delta^2H$ values given in @Meinzer:2001 for the dry-season of March 1997 as these showed largest seasonal divergence among species, with a corresponding gradient of $\delta^2H$ in the soil (Fig.
 S\@ref(fig:iso-data)).
@@
 We calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$ 30 m) and average abundance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; *n* = `r length(erd.mrate.sp)`).
-For each of the seven census intervals ($t$) in the 35-yr record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (% yr^-1^) for two successive censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{time}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees of a species present in cen~1~ and dead in cen~2~, respectively, in the 50ha plot and $time$ is the duration based on mean dates of cen~1~ and cen~2~.
+For each of the seven census intervals ($t$) in the 35-yr record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (% yr^-1^) for two successive censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{time}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees of a species present in cen~1~ and dead in cen~2~, respectively, in the 50 ha plot and $time$ is the duration based on mean dates of cen~1~ and cen~2~.
 For each species we calculated mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{t=1}^{6} {Mort_{t}}$.
@@
 
-```{r hydro-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-hydro-calib}Observations (red lines) versus simulations (blue lines) from 100 best-fit parameter ensemble runs of ELM-FATES are shown for daily Volumetric Water Content (VWC) by depth for horizontal time domain reflectometry (TDR) probes at three depths (0.1, 0.4 and 1 m; A) and a vertical probe (0-0.15 m; B); monthly stream discharge (C) and monthly evapotranspiration from eddy flux tower (D).
+```{r hydro-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-hydro-calib}Observations (red lines) versus simulations (blue lines) from 100 best-fit parameter ensemble runs of ELM-FATES are shown for daily Volumetric Water Content (VWC) by depth for horizontal time domain reflectometry (TDR) probes at three depths (0.1, 0.4 and 1 m; A) and a vertical probe (0-0.15 m; B); monthly stream discharge (Discharge, C) and monthly evapotranspiration from eddy flux tower (ET, D).
 Points (red) in (A) show average plot-wide observations of VWC with 95% CI (error bars).
 VWC data from the horizontal probes (lines in A) were not used for ELM-FATES calibration.
-Values in inset are average standardised RMSE across the 100 simulations.
+Values in inset are average standardized RMSE across the 100 simulations.
 RMSE was standardized to range 0-1 across the 5000 simulations.
-For panel A, values at the top are for plot-wide VWC and those at the bottom are for TDR probes.")}
+For panel A, values at the top are for plot-wide VWC and those at the bottom are for TDR probes.
+VWC is in units of cm^3^ cm^-3^, while ET and Discharge in mm month^-1^.")}
 knitr::include_graphics(file.path("figures/hydro-calib2.jpeg"))
@@
 In comparison to the measured isotopic signal ($\delta^2H$), our results showed that the ERD model with a multiplicative effect of VPD and without Leaf Cover (Eq.
-\@ref(eq:M2)) had highest goodness-of-fit (*R*^2^ = 0.85; *p* = 0.008; Fig.
+\@ref(eq:M2)) had highest goodness-of-fit (*R*^2^ = 0.85; *P* = 0.008; Fig.
 \@ref(fig:erd-iso)).
@@
 
-```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Pairwise correlations between Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) and hydraulic properties for seven large, canopy species found on Barro Colorado Island (Panama).
-Scatterplots with linear fits are shown in the lower half.
-Corresponding Spearmon's $\\rho$ are given in the diagonally opposite location in the upper half, along with significance levels, where **, * and .
-represent *P*-value $\\le$ 0.01, 0.05 and 0.1, respectively.
+```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Pairwise correlations among Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) and hydraulic properties for seven large, canopy species found on Barro Colorado Island (Panama).
+Scatterplots with linear fits are shown in the lower-left triangle.
+Corresponding Spearmon's $\\rho$ are given in the diagonally opposite location in the upper-right triangle.
+*P*-values $\\le$ 0.01, 0.05 and 0.1 are represented by symbols **, * and .
+, respectively.
 Diagonal shows density distribution of each variable.")}
@@
 \@ref(fig:erd-time)).
-Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25-yr period for which soil water potential $\Psi_{soil,~z}$ was modeled (1990-2015), as soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dry seasons and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).
+Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25-yr period for which soil water potential $\Psi_{soil,~z}$ was modeled (1990-2015), as soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`;  median, 95% CI) through all the dry seasons and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).
 
@@
 
-```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\\Psi_{soil,~z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for ", length(erd.mrate.sp), " canopy species found on Barro Colorado Island.
+```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time}Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\\Psi_{soil,~z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for ", length(erd.mrate.sp), " canopy species found on Barro Colorado Island.
 Each circle represents a species.
@@
 This trade-off translated into greater drought tolerance, as deeper ERD species showed lower mortality rates than shallower ERD species over the 35-yr record (1982-2015).
-Analysis of time spent below ($\Psi_{crit}$) in the soil zone matching ERD $\Psi_{soil,~z = ERD}$ showed that this was likely due to decreasing realized risk of hydraulic failure with depth.
+Analysis of time spent below ($\Psi_{crit}$) in the soil zone matching ERD $\Psi_{soil,~z = ERD}$ showed that this was likely due to decreasing realized risk of hydraulic failure with increasing rooting depth.
 We thus show, for the first time, that inter-specific variation in trees' effective rooting depths aligns species on a conservative-acquisitive water economic spectrum by demonstrating a link between trees' realized drought experience, hydraulic architecture and long-term mortality outcomes.
-On the acquisitive end of the spectrum were deep ERD species with hydraulically efficient, but risky suit of traits optimized for an environment of reliable and abundant source of water at depth that actualized to a lower mortality-risk.
-At the conservative end of the spectrum were shallow ERD species with inefficient, but safer xylem, likely optimized for high variability in $\Psi_{soil}$ in the shallow soil layers, resulting in higher variance in mortality rates across censuses as $\Psi_{soil,~z = ERD}$ crossed ERD $\Psi_{crit}$.
+On the acquisitive end of the spectrum were deep ERD species with a  hydraulically efficient, but risky, suite of traits optimized for an environment of reliable and abundant source of water at depth that resulted in a lower mortality risk.
+At the conservative end of the spectrum were shallow ERD species with inefficient, but safer, xylem likely optimized for high variability in $\Psi_{soil}$ in the shallow soil layers, and resulting in higher variance in mortality rates across censuses as $\Psi_{soil,~z = ERD}$ crossed ERD $\Psi_{crit}$.
 Mortality rates of shallow ERD species may have been exacerbated by increased drought events during the study period.
@@
 We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins, that is, traits that are commonly identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through a study period of greater recurrence of drought.
-These results are consistent with the rare studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during a centennial (Texas) or extreme (Italy) drought [@Johnson:2018fad; @Nardini:2015].
+These results are consistent with the rare studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015].
 @Johnson:2018fad found that mortality decreased with deep rooting even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi_{88,~stem}$) and had narrower safety-margins ($\Psi_{min} - \Psi_{88,~stem}$).
-@Nardini:2015 found that a species sourcing deep water in a Karst system (Italy) showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi_{88,~stem}$ and narrower safety margin compared to two species sourcing shallow-water.
-Our result of lower mortality in deep rooted trees contrasts with the finding of @Chitra-Tarak.2018 in which species identified by their inverse ERD model as effectively deep rooted died *more* in a prolonged, centennial drought.
+@Nardini:2015 found that a species sourcing deep water in a Karst system showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi_{88,~stem}$ and narrower safety margin compared to two species sourcing shallow-water.
+
+\
+Our result of lower mortality in deep rooted trees contrasts with the finding of @Chitra-Tarak.2018 in which species identified by their inverse ERD model as effectively deep rooted died *more* in a prolonged, extreme drought.
 Their hydrological model revealed that the multi-annual drought exhausted the deep water reservoir for over two years, whereas our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period, $\Psi_{soil}$ remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) below the depth of 1 m through all the dryseasons and drought periods, that is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).
 The difference in our results are likely attributable to site differences in the precipitation regime.
-In their dry forest site (Mudumalai National Park, India) precipitation is low: 1095 $\pm$ 296 mm (mean $\pm$ SD) compared to `r round(mean(clim$Precip))` $\pm$ `r round(sd(clim$Precip))` mm at BCI, and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (PET = `r round(mean(clim$pet.PM))` $\pm$ `r round(sd(clim$pet.PM))` mm), and thus at BCI deep soil layers are effectively recharged annually (Fig.
+In their dry forest site (Mudumalai National Park, India) precipitation is low: 1095 $\pm$ 296 mm (mean $\pm$ SD) compared to `r round(rain.man.stats$rain)` $\pm$ `r round(rain.man.stats$rain.sd)` mm (`r min(clim$Year)`-`r max(clim$Year)`) at BCI, and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (Penman-Monteith PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (Penman-Monteith PET = `r round(mean(clim$pet.PM))` $\pm$ `r round(sd(clim$pet.PM))` mm;`r min(clim$Year)`-`r max(clim$Year)`; this study, @Guo.2016), and thus at BCI deep soil layers are effectively recharged annually (Fig.
 S\@ref(fig:swp-all-depths); @Chitra-Tarak.2018).
-We note that the variation in mortality explained by ERD in our study is low (*R*^2^ = 0.11, *P* = 0.075), and other causes may explain observed mortality.
+The variation in mortality explained by ERD in our study is low (*R*^2^ = 0.11, *P* = 0.075), and other causes may explain observed mortality [@McDowell.2018].
 
@@
 By estimating ERD for large trees of `r length(erd.sp)` canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species species-rich tropical forests.
-Our chosen ERD model is an advancement over the only other inverse model of ERD (@Chitra-Tarak.2018; *n* = 12 species) in terms of physiologically meaningful representation, fewer parameters and corroboration with tree and soil data on $\delta^2H$.
+Our chosen ERD model is an advance over the only other inverse model of ERD (@Chitra-Tarak.2018; *n* = 12 species) in terms of physiologically meaningful representation, fewer parameters and corroboration with tree and soil data on $\delta^2H$.
 The model in @Chitra-Tarak.2018 estimated six species-specific phenomenological parameters: three as root fraction for corresponding lumped vadose zone layers, and three as shape parameters for water uptake vs availability curves.
 For each species, they calculated daily water uptake successively from three layers, that is, the supply, to satisfy the demand of the daily maximum evapotranspiration, a function of PET and LAI.
-They regressed total supply to total demand ratio over census intervals as a growth coefficient against growth (5 data points), and chose values for six parameters that best-fit (highest *R*^2^, that is, 2 additional parameters).
-Our inverse ERD model fits only two parameters by way of regressing depth-specific model against 5 growth data points.
+They regressed total supply to total demand ratio over census intervals as a growth coefficient against growth (five data points), and chose values for six parameters that best-fit (highest *R*^2^, that is, two additional parameters).
+Our inverse ERD model fits only two parameters by way of regressing depth-specific model against five growth data points.
 Our model was constrained on VPD and species leaf vulnerability curves, variables that are shown to be physiologically meaningful.
-We benefited from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA~lamina~ and WSG~100^$\circ$^C~) and parameters of leaf vulnerability (or percent loss of conductance, (PLC)) curves for which we collected data on a few species (*n* = `r length(sp.exp.param$sp)`), a 'hard' trait.
+
+\
+We benefited from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA and WSG) and parameters of leaf vulnerability (or percent loss of conductance, (PLC)) curves for which we collected data on a few species (*n* = `r length(sp.exp.param$sp)`), a 'hard' trait.
 This allowed us to scale-up the leaf vulnerability curves at the community scale (*n* = `r length(ab.table$Species)` species, see Table S\@ref(tab:ab-param) for model predicted parameters).
@@
 $\Psi_{leaf}$ dataset is?").
+
+\
 We also note that the availability of data on $\delta^2H$ on trees and water sources below-ground--from soil or substrate pore water in the vadose zone and free water below the water table--as important for model calibration or validation.
@@
 Acquisitive traits such as high $K_{max}$ are expected to be associated with faster growth rates.
-We found a positive but non- significant relationship of adult growth rates ($\ge$ 10cm dbh) with $K_{max, ~leaf}$ (*R*^2^ = 0.13; *p* = 0.126, *n* = `r length(sp.exp.param$sp)`) and no relationship with $K_{max, ~stem}$ (*n* = `r stem.traits.sp` species).
+We found a positive but non- significant relationship of adult growth rates ($\ge$ 10cm dbh) with $K_{max, ~leaf}$ (*R*^2^ = 0.13; *P* = 0.126, *n* = `r length(sp.exp.param$sp)`) and no relationship with $K_{max, ~stem}$ (*n* = `r stem.traits.sp` species).
 The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits in order to better represent the functional diversity of tropical forests.
@@
 LR was deputed to Indo-French Cell for Water Sciences on an IRD Fellowship.
-An NSF grant to Smithsonian Institution’s CTFS-ForestGEO (Dimensions 1046113) also partially supported this work via RCT and SMM.
+An NSF grant to Smithsonian Institution’s ForestGEO (Dimensions 1046113) also partially supported this work via RCT and SMM.
 **Add acknowledgements for other co-authors.
@@
 \
-Barro Colorado Forest Census Plot Data, 2012 Version [@Condit.2012], can be obtained from DOI http://dx.doi.org/10.5479/data.bci.20130603.
-More recent versions of the data can be requested directly from R.
-Condit (conditr@gmail.com).
+Barro Colorado Forest Census Plot Data can be obtained online[@Condit.2019a;@Condit.2019b].
 
@@
 
-### WSG~100^$\circ$^C~ and LMA~lamina~
-Joe please add, if any.
-
-### Data gaps in WSG~100^$\circ$^C~ and LMA~lamina~
+### Data gaps in WSG and LMA
 
@@
 $\Psi_{leaf}$) used to build the community-level models of leaf vulnerability curves (Eqs.
-\@ref(eq:b-mod) and \@ref(eq:a-mod)), two lacked data on WSG~100^$\circ$^C~ and five on LMA~lamina~.
-We filled three gaps in LMA~lamina~ using predictions from a strong linear relationship between LMA~lamina~ and LMA~disc~, the latter defined as the mean leaf mass per unit area measured for a 1.483 cm^2^ leaf disc taken to avoid veins (g m^-2^) for leaves receiving direct sunlight ($\bar{LMA_{lamina}} = `r disc.cf[1]` + `r disc.cf[2]`LMA_{disc}$; Adj.
+\@ref(eq:b-mod) and \@ref(eq:a-mod)), two lacked data on WSG and five on LMA~lamina~.
+We filled three gaps in LMA using predictions from a strong linear relationship between LMA and LMA~disc~, the latter defined as the mean leaf mass per unit area measured for a 1.483 cm^2^ leaf disc taken to avoid veins (g m^-2^) for leaves receiving direct sunlight ($\overline{LMA} = `r disc.cf[1]` + `r disc.cf[2]`LMA_{disc}$; Adj.
 *R*^2^ = `r disc.r2`; *P* `r disc.p`, *n* = `r disc.n`).
-We filled the two remaining gaps in LMA~lamina~ using predictions from a linear relationship between LMA~lamina~ and LMA~lamina2~, as measured during the smaller, $K_{leaf}$ campaign ($\bar{LMA_{lamina}} = `r lma.cf[1]` + `r lma.cf[2]`LMA_{lamina2}$; Adj.
+We filled the two remaining gaps in LMA using predictions from a linear relationship between LMA and LMA~lamina~, as measured during the smaller, $K_{leaf}$ campaign ($\overline{LMA} = `r lma.cf[1]` + `r lma.cf[2]`LMA_{lamina}$; Adj.
 *R*^2^ = `r lma.r2`; *P* `r lma.p`, *n* = `r lma.n`).
-We chose not to fill the two data gaps in WSG~100^$\circ$^C~ as available data were poor predictors of WSG~100^$\circ$^C~.
+We chose not to fill the two data gaps in WSG as available data were poor predictors of WSG.
 
 \
-At the community level, we filled data gaps in $LMA_{lamina}$ for `r sum(c.lma.gaps[-which(names(c.lma.gaps) == "original")])` species (of total `r sum(c.lma.gaps)`)  using linear relationships with other variables: `r c.lma.gaps[[which(names(c.lma.gaps) == "filled-LMA-DISC")]]` species with $LMA_{disc}$, `r c.lma.gaps[[which(names(c.lma.gaps) == "filled-LMA-Nobby")]]` with $LMA_{lamina2}$, and `r c.lma.gaps[[which(names(c.lma.gaps) == "filled-LEAF")]]` with $LMA_{leaf}$, defined as the mean leaf mass per unit area measured for the entire leaf including the petiole (g m^-2^) for leaves receiving direct sunlight ($\bar{LMA_{lamina}} = `r leaf.cf[1]` + `r leaf.cf[2]`LMA_{leaf}$; Adj.
+At the community level, we filled data gaps in $LMA$ for `r sum(c.lma.gaps[-which(names(c.lma.gaps) == "original")])` species (of total `r sum(c.lma.gaps)`)  using linear relationships with other variables: `r c.lma.gaps[[which(names(c.lma.gaps) == "filled-LMA-DISC")]]` species with $LMA_{disc}$, `r c.lma.gaps[[which(names(c.lma.gaps) == "filled-LMA-Nobby")]]` with $LMA_{lamina}$, and `r c.lma.gaps[[which(names(c.lma.gaps) == "filled-LEAF")]]` with $LMA_{leaf}$, defined as the mean leaf mass per unit area measured for the entire leaf including the petiole (g m^-2^) for leaves receiving direct sunlight ($\overline{LMA} = `r leaf.cf[1]` + `r leaf.cf[2]`LMA_{leaf}$; Adj.
 *R*^2^ = `r leaf.r2`; *P* `r leaf.p`, *n* = `r leaf.n`).
-For the `r sum(c.lma.gaps)` species thus with $LMA_{lamina}$, we gap-filled data on  WSG~100^$\circ$^C~ for `r sum(c.wsg.gaps[-which(names(c.wsg.gaps) == "original")])` species using a linear relationship with wood specific gravity reported by @Chave:2009 ($\bar{WSG_{100^{\circ}C}} = `r chave.cf[1]` + `r chave.cf[2]`WSG_{chave}$; Adj.
+For the `r sum(c.lma.gaps)` species thus with $LMA$, we gap-filled data on  WSG for `r sum(c.wsg.gaps[-which(names(c.wsg.gaps) == "original")])` species using a linear relationship with wood specific gravity reported by @Chave:2009 ($\overline{WSG} = `r chave.cf[1]` + `r chave.cf[2]`WSG_{chave}$; Adj.
 *R*^2^ = `r chave.r2`; *P* `r chave.p`, *n* = `r chave.n`).
@@
 ```{r TDR-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-TDR-calib}Volumetric Water Content vs.
-TDR data for a vertical TDR probe (0-15 cm depth) at the base of the eddy flux tower near the 50ha plot at BCI."}
+TDR data for a vertical TDR probe (0-15 cm depth) at the base of the eddy flux tower near the 50 ha plot at BCI."}
 knitr::include_graphics(file.path("figures/Matteos_TDR_SWC_data_linear_model.jpeg"))
@@
 ```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($Psi_{soil,~z}$) depth z over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines).
-Vertical gray lines show median dates of the demographic censuses at the BCI 50ha plot."}
+Vertical gray lines show median dates of the demographic censuses at the BCI 50 ha plot."}
 knitr::include_graphics(file.path("figures/2019-10-14_5000/best-fits/psi_model_daily_all_depths_params.top.few_full.jpeg"))
@@
 \\@ref(eq:a-mod) & \\@ref(eq:b-mod)).
-Goodness-of-fit (*R*^2^) and p-values are given in inset.")}
+Goodness-of-fit (*R*^2^) and *P*-values are given in inset.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/A_B_data_vs_model.jpeg"))
@@
 
-```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}'Vulnerability curves' fitted to observed $K_{leaf}$ vs.
+```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}Vulnerability curves fitted to observed $K_{leaf}$ vs.
 $\\Psi_{leaf}$ data on ", length(sp.exp.param$sp)," tree species from BCI (A & B) with equation \\@ref(eq:vc-exp)--in absolute terms (A, B), as well as, in terms of loss of conductivity (%) (C, D).
@@
 
-```{r obs-comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-comm-plc-curves}'Vulnerability curves' for species with ERD estimates.
+```{r obs-comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-comm-plc-curves}Vulnerability curves for species with ERD estimates.
 Curves in  data on ", length(sp.exp.param$sp)," tree species from BCI (A & B) with equation \\@ref(eq:vc-exp)--in absolute terms (A, B), as well as, in terms of loss of conductivity (%) (C, D).
@@
 The number of species with an ERD estimate as well as $\\delta^2H$.
-or each model are shown in inset, along with the goodness-of-fit (*R*^2^) and p-values for a linear relationship between ERD and $\\delta^2H$.
+or each model are shown in inset, along with the goodness-of-fit (*R*^2^) and *P*-values for a linear relationship between ERD and $\\delta^2H$.
 See Eqs.

### 2020-08-30 manuscript.rmd
Equation alignment changed

@@
     fig_caption: yes
-  bookdown::word_document2:
-        reference_docx: manuscriptstyle.docx
-        fig_caption: yes
-  bookdown::pdf_document2: default
 bibliography: references.bib
@@
 \begin{equation}
+\begin{split}
 \Delta W_{can} + \Delta W_{sfc} + \sum_{z=1}^{15} \Delta W_{liq,z} + \Delta W_{a} = \left(q_{rain} - E_{v} - E_{g} - q_{over} - q_{h2osfc} - q_{drai}\right)\Delta t
+\end{split}
 (\#eq:water-balance)
 \end{equation}
+
 where, $q_{rain}$ is rainfall, $E_{v}$ is evapotranspiration (ET) of the forest, $E_{g}$ is ground evaporation, $q_{over}$ is surface runoff, $q_{h2osfc}$ is runoff from surface water storage, $q_{drai}$ is sub-surface drainage (all in mm H~2~O) and $\Delta t$ is the time-step.
@@
 Second, we calculated an objective function for each of the 5000 ensembles by equally weighing standardized RMSE between observations and simulations across all fluxes, and chose 100 ensembles, within the 300 that best-fit $\theta_{0-0.15 m}$, that had the lowest values of the objective function.
-Let $f(obs,~sim)$ be RMSE between observations and simulations standardized across ensembles (range 0-1), based on monthly sums for $ET$ and $q_{run}$, respectively, and daily averages for $\theta$s, then the objective function can be described as
+Let $f(obs,~sim)$ be RMSE between observations and simulations standardized across ensembles (range 0-1), based on monthly sums for $ET$ and $q_{run}$, respectively, and daily averages for $\theta$s, then the objective function ($\overline{O}$) can be described as
 
 \begin{equation}
-Objective Function = \frac{1}{6}\sum \left(f(ET,~E_{v} + E_{g}) + f(q_{run},~q_{over} + q_{drai}) + f(\theta_{0-0.15m}, {W_{liq,~\overline{z}}}) + f(\theta_{0.15m}, W_{liq,~`r soil.depths[4]`m}) + f(\theta_{0.4m}, W_{liq,~`r soil.depths[6]`m}) + f(\theta_{1 m}, W_{liq,~`r soil.depths[8]`m})\right)
+\overline{O} = \frac{1}{6}\sum \left(f(ET,~E_{v} + E_{g}) + f(q_{run},~q_{over} + q_{drai}) + f(\theta_{0-0.15m}, {W_{liq,~\overline{z}}}) + f(\theta_{0.15m}, W_{liq,~`r soil.depths[4]`m}) + f(\theta_{0.4m}, W_{liq,~`r soil.depths[6]`m}) + f(\theta_{1 m}, W_{liq,~`r soil.depths[8]`m})\right)
 (\#eq:obj-func)
 \end{equation}
+
 where $W_{liq,~\overline{z}}$ is the daily average of $W_{liq,~z}$ for $z$ = `r paste(soil.depths[1:4], collapse=", ")` m.
@@
 We therefore tested alternate models of growth, aka inverse models for ERD (Eqs.
-\@ref(eq:M2)- \@ref(eq:M6)), in which a species' growth dynamics across census intervals was regressed against census interval means of multiplied or additive effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the fraction of leaf cover (all standardized to range from 0-1; see below).
+\@ref(eq:M2)-\@ref(eq:M6)), in which a species' growth dynamics across census intervals was regressed against census interval means of multiplied or additive effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the fraction of leaf cover (all standardized to range from 0-1; see below).
 Let n~t~ be the total number of days in census interval $t$, then for individual tree species Growth~t~ and depth $z$ these models are described as follows:
@@
 S\@ref(fig:gpp-vpd)) and used standardized $\overline{GPP}$ to represent the impact of VPD on growth (hereafter, $StdVPD$; range 0-1) in Eqs.
-\@ref(eq:M2), \@ref(eq:M3) \@ref(eq:M5) and \@ref(eq:M6).
+\@ref(eq:M2), \@ref(eq:M3), \@ref(eq:M5) and \@ref(eq:M6).
 Due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season at our study site, the impact of VPD may as well be a combined impact of temperature, humidity, and radiation.
@@
 \@ref(eq:k-soil) either from data when available, or a model based on $LMA$ or $WSG$ as described in the earlier section, and used standardized K within species ($Std.K$; range 0-1) in Eqs.
-\@ref(eq:M1)- \@ref(eq:M6).
+\@ref(eq:M1)-\@ref(eq:M6).
 
 \
-For the combination of each model \@ref(eq:M1)- \@ref(eq:M6), species and 100 hydrological realizations, we obtained AIC, goodness-of-fit (*R*^2^) and Pearson's $\rho$ for each soil depth $z$.
+For the combination of each model \@ref(eq:M1)-\@ref(eq:M6), species and 100 hydrological realizations, we obtained AIC, goodness-of-fit (*R*^2^) and Pearson's $\rho$ for each soil depth $z$.
 For each combination of model-species-hydrological realization we retained only one depth with the least AIC score (the best-fit), and within those depths retained the ones that had a positive corresponding Pearson's $\rho$ and *R*^2^ >= 0.5 as a valid set.
@@
 
-# Supplementary Information
+\newpage
+
+# Supplementary material {-}
 
@@
 $\Psi_{leaf}$) used to build the community-level models of leaf vulnerability curves (Eqs.
-\@ref(eq:b-mod) and \@ref(eq:a-mod)), two lacked data on WSG and five on LMA~lamina~.
+\@ref(eq:b-mod)-\@ref(eq:a-mod)), two lacked data on WSG and five on LMA~lamina~.
 We filled three gaps in LMA using predictions from a strong linear relationship between LMA and LMA~disc~, the latter defined as the mean leaf mass per unit area measured for a 1.483 cm^2^ leaf disc taken to avoid veins (g m^-2^) for leaves receiving direct sunlight ($\overline{LMA} = `r disc.cf[1]` + `r disc.cf[2]`LMA_{disc}$; Adj.
@@
 (Eqs.
-\\@ref(eq:a-mod) & \\@ref(eq:b-mod)).
+\\@ref(eq:a-mod)-\\@ref(eq:b-mod)).
 Goodness-of-fit (*R*^2^) and *P*-values are given in inset.")}

### 2020-09-19 manuscript.rmd
incorporated some of the co-authors' comments

@@
 output:
+  bookdown::html_document2:
+    number_sections: no
+    theme: flatly
+    fig_caption: yes
   bookdown::word_document2:
@@
   bookdown::pdf_document2: default
-  bookdown::html_document2:
-    number_sections: no
-    theme: flatly
-    fig_caption: yes
 bibliography: references.bib
@@
 
-- Among the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied.
+- Among the mechanisms that plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied.
 Vulnerability to embolism and hydraulic safety margins are shown to be predictors of drought-induced mortality risk, but deep-water access may delay plants from reaching these critical thresholds.
 Here, we tested the role of deep-water access in survival strategies of a diverse tropical forest community (Panama) using a novel data-model approach.
-- We inversely modeled effective rooting depth (ERD) for large trees of `r length(erd.sp)` canopy species by linking growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and a new community-scale model of species leaf vulnerability curves.
+- We inversely modeled effective rooting depth (ERD) for large trees of `r length(erd.sp)` canopy species by linking growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and a species-specific leaf vulnerability curves.
 We evaluated ERD against existing isotopic tracer data on trees and soil.
@@
 Deeper ERD species spent less time below critical hydraulic thresholds indicating that trees mitigate water-stress related mortality risk through deep water access.
-- The role of deep-water access in alleviating mortality of drought-sensitive trees has important implications for our predictive understanding of tropical forest dynamics under current and future climate.
+- The role of deep-water access in alleviating mortality of drought-sensitive trees has important implications for our predictive understanding of forest dynamics under current and future climate.
 
@@
 \
-Tropical forests play a disproportionately large role in the global carbon and energy cycle, and support half of global biodiversity [@Bonan:2008; @Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019].
+Tropical forests play a disproportionately large role in the global carbon and energy cycle [@Bonan:2008], and support half of global biodiversity [@Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019].
 Tropical forests are logically considered to be an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw], yet mortality events in the tropics are rarely as large as those in the temperate and boreal zones (McDowell et al.
@@
 Despite this, widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010] and tropical mortality rates are increasing in some regions [@Brienen.2015; @Hubau.2020].
-However, state-of-the-art dynamic vegetation models (DVMs) fail to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013], as underlying mechanisms of drought-tolerance are not fully understood or quantified.
+However, state-of-the-art dynamic vegetation models (DVMs) struggle to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013; @Powell:2018], as underlying mechanisms of drought-tolerance are not fully understood or quantified.
 
@@
 Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism.
-This mechanism assumes that water availability varies below-ground by vertical depth and so do rooting depths of co-occurring trees, by species and size [@Meinzer:1999; @Chitra-Tarak.2018].
+This mechanism assumes that water availability varies below-ground by vertical depth, and that the rooting depths of colocated trees also vary, by species and size [@Meinzer:1999; @Chitra-Tarak.2018].
 Consequently, trees in the same forest differ in their drought experience, and thus in their growth and mortality responses.
@@
 Further, only a small fraction of those are tropical, even though ~70-90 % (40 - 53 thousand species) of the world’s tree diversity resides in the tropics [@Slik:2015].
-Use of stable isotopes of water as a tracer provides an indirect measure of water sourcing depths by matching isotopic value in plant tissue to variations in soil pore water with depth.
+Use of stable isotopes of water as a tracer provides an indirect measure of water sourcing depths by matching the isotopic value in plant tissue to variations in soil pore water with depth.
 However, such data are extremely rare [@Evaristo:2016].
@@
 \
-Trade-offs shape how plants make a living in their environment (ecological strategies) and also provide the essential constraints to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010].
-This entails identifying trade-offs among environment, traits and demography.
+Trade-offs in plant 'traits' shape how plants make a living in their environment (ecological strategies) and also provide the essential constraints to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010].
+This entails identifying relationships among environment, traits and demography.
 Specifically for the water environment (or, resource), observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified an acquisitive-conservative resource economic spectrum similar to those for light and nutrients, although, for water, examples of all trade-offs observed simultaneously are rare, and below-ground traits or environment linkages are weak [@Reich:2014; @Sterck:2011].
-Acquisitive species have traits with high capacity to acquire, move, store and use water (high $K_{max}$, low wood density; see Table 1 for trait symbols and their definitions) and achieve high growth potential, but build flimsy, disposable tissues (high SLA) and are less tolerant of low resource availablity, whereas species with conservative traits are better protected from carbon losses and are more drought tolerant (low $\Psi_{50}$, low $\Psi_{tlp}$, larger safety margin).
-Acquisitive species are spatially associated with wet sites (high resource) such as valleys or those with high rainfall [@Kursar:2009; @Oliveira:2019], but whether and how the acquisitive-conservative spectrum holds true for the water dynamics vertically below-ground is not clear.
-Deep rooting may be expected to support an acquisitive strategy with a steady water supply unlike shallow-rooted trees.
-However, this may depend on the exact dynamics of water-availability by depth at a site, depending on precipitation regime and hydrology.
-In pulse-driven systems *shallow*-rooted plants are associated with acquisitive traits (high $K_{max}$) that maximise the use of the water-pulse, while *deep*-rooted plants show conservative traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013 and references therein].
+Acquisitive species have traits with high capacity to acquire, move, store and use water (higher hydraulic conductivity, low wood density) and achieve high growth, but build flimsy tissues (high SLA) and are less tolerant of low resource availablity, whereas species with conservative traits are better protected from carbon losses and are more physiologically drought tolerant (greater embolsim resistance, larger safety margin).
+Whether and how the acquisitive-conservative spectrum holds true for the water dynamics vertically below-ground is not clear.
+Acquisitive species are spatially associated with wet sites (high resource) such as valleys or those with high rainfall [@Kursar:2009; @Oliveira:2019].
+Deep rooting may be expected to support an acquisitive strategy by enabling a steady water supply.
+However, the degree to which deep rooting is associated with an acquisitive strategy may depend on the exact dynamics of water availability by depth at a site, depending on precipitation regime and hydrology.
+In pulse-driven systems *shallow*-rooted plants are associated with acquisitive traits (higher hydraulic conductivity) that maximise the use of the water-pulse, while *deep*-rooted plants show conservative traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013 and references therein].
 
@@
 However, plants' realized water environments, specifically their ability to acquire water via differing rooting depths, are rarely studied, especially in the tropical region.
-To better understand how below-ground rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tropical forest tree species, we make a thorough assessment for a tropical forest of trade-offs between the below-ground water environment, above-ground hydraulic traits, estimates of effective rooting depths and demographic outcomes over 35 years that witnessed droughts of various intensity and duration.
-Specifically, we test an overarching hypothesis that inter-specific variation in trees' effective rooting depths aligns species on an acquisitive-conservative economic spectrum optimized for increasingly reliable water availability by depth.
-Our hypotheses are that (i) trees with deeper ERD have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem: higher maximum stem hydraulic conductivity ($K_{max}$), but greater xylem vulnerability to embolism ($\Psi_{88,~stem}$), higher $\Psi_{tlp}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$), whereas shallower ERD species are associated with less efficient but safer hydraulic traits; and that, (ii) deeper ERD species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallower ERD species may have higher mortality rates as a result of higher variability in shallow soil water availability.
+
+\
+To better understand how rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tropical forest tree species, in this paper, we conduct a thorough assessment for a tropical forest of relationships between the water environment in the whole-soil column, above-ground hydraulic traits, estimates of effective rooting depths and demographic outcomes over 35 years that witnessed droughts of various intensity and duration.
+Specifically, we test an overarching hypothesis that inter-specific variation in trees' effective rooting depths aligns species on an acquisitive-conservative economic spectrum, .
+whereby acquisitive trees have rooting depths optimized for increasingly reliable water availability by depth.
+Our hypotheses are that (i) trees with deeper ERD have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallower ERD species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deeper ERD species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallower ERD species may have higher mortality rates as a result of higher variability in shallow soil water availability.
 To our knowledge this is the first study to demonstrate a mechanistic link between trees' realized drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.
@@
 \
-```{r symbols, results = 'asis', echo = FALSE}
-knitr::kable(symbols.table, caption = 'The symbol, definition and units of key traits used.', booktabs = TRUE, escape = FALSE)
+```{r hypo, results = 'asis', echo = FALSE}
+knitr::kable(hypo.table, caption = 'Hypotheses for association between Effective Rooting Depth (ERD) and hydraulic traits.', booktabs = TRUE, escape = FALSE)
 ```
@@
 These are followed by analyses to address the proposed hypotheses.
-Note that only new data on leaf vulnerability curves collected during this study are described below, details of all other datasets used are described in the Supplementary Information.
+Descriptions of datasets used and details of some of the methods are described in the Supplementary Information.
 
 \
-In summary, we obtained daily dynamics of plant water potential ($\Psi_{soil,~z}$) by depth $z$ in the whole-soil column with a locally parameterized and calibrated 1-D hydrological water balance model.
-We developed a novel inverse model to estimate species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015).
-This model incorporated impact of atmospheric drought (VPD) on growth and was constrained by above-ground hydraulic traits, namely, species leaf vulnerability curves ($K_{leaf}$ vs.
-$\Psi_{leaf}$).
-We collected new data on vulnerability curves for 21 species.
-Identifying scaling relationships between these curves and commonly available traits (Leaf Mass Area~lamina~, Wood Specific Gravity~100^$\circ$^C~; see Table \@ref(tab:symbols) for definitions) allowed us to scale-up these curves to `r length(data.model.AB$sp)` tree species in the community (of 321 total).
-We evaluated ERD against existing data on plant and soil water isotopes ($\delta^2H$) and tested coordination of ERD with hydraulic traits and mortality rates.
+In summary, we obtained daily dynamics of plant water potential by depth in the whole-soil column with a locally parameterized and calibrated 1-D hydrological water balance model.
+We developed a novel inverse model to estimate species effective rooting depths (ERD), defined as the depth at which soil water dynamics best explained observed dynamics of species growth.
+This model incorporated the impact of atmospheric drought on growth and was constrained by above-ground hydraulic traits, namely, species leaf vulnerability curves.
+For the latter, we collected new data on vulnerability curves for 21 species and scale-up to the community via scaling relationships with 'softer' traits, namely, Wood Density and Leaf Mass Area.
+We evaluated ERD against independent estimates based on existing data on plant and soil water isotopes.
+We then tested whether deeper ERD was associated with hydraulically efficient but vulnerable traits (Table 1 ), but not greater mortality rates.
 
@@
 The entire island is forested and classified as a tropical moist forest in the Holdridge Life Zone system.
-Although 28% of canopy tree species are at least partly deciduous during the dry season, the maximum percentage of deciduous individuals in the canopy is only 6.3%, and the percentage of crown area that is deciduous is estimated at 9.7% [@Condit:2000].
+Although 28% of canopy tree species are at least partly deciduous during the dry season, this amounts to the maximum percentage of deciduous individuals in the canopy this amounts to is only 6.3%, and the percentage of crown area that is deciduous is estimated at 9.7% [@Condit:2000].
 Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014].
@@
 \
-We measured ($K_{leaf}$ vs.
-$\Psi_{leaf}$) on adult trees of `r length(sp.exp.param$sp)` common species (see [SI](#SIData)).
-We fitted four different functions to species-specific data using the *anneal* optimization function with the package likelihood in R: exponential without an intercept ($K_{leaf} = Ae^{-B\Psi_{leaf}}$), exponential with an intercept ($K_{leaf} = C + Ae^{-B\Psi_{leaf}}$), logistic ($K_{leaf} = \frac{A}{1 + \left(\frac{\Psi_{leaf}}{x_{0}}\right)^B}$) and sigmoidal ($K_{leaf} = \frac{A}{1 + e^{-\left(\frac{\Psi_{leaf} - x_{0}}{B}\right)}}$) [@Guyot:2011].
-The differences among AICc values for these functions were not large within species (Zailaa et al.
-to be submitted).
+We obtained leaf hydraulic vulnerability curves ($K_{leaf}$ vs.
+$\Psi_{leaf}$; see Table \@ref(tab:symbols) for trait symbols and their definitions) for adult trees of 21 common species from Zailaa et al.
+(to be submitted; see SI for brief description of methods).
+
 We chose the exponential function without the intercept to fit observed data across species, and as a general function to scale-up to the community, because it had the lowest AICc for most species, captured the form of the observed data across all species well, and had the lowest number of parameters, that is, two:
@@
 \@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r length(unique(bci.AB$sp[!is.na(bci.AB$A)]))` species for which WSG and LMA were both available, retained `r length(unique(data.model.AB$sp))` for which parameters $A$ and $B$ were both positive, and fitted vulnerability curves to those.
+
+\
+```{r symbols, results = 'asis', echo = FALSE}
+knitr::kable(symbols.table, caption = 'The symbol, definition and units of key traits used.', booktabs = TRUE, escape = FALSE)
+```
 
@@
 
-```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Pairwise correlations among Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) and hydraulic properties for seven large, canopy species found on Barro Colorado Island (Panama).
-Scatterplots with linear fits are shown in the lower-left triangle.
-Corresponding Spearmon's $\\rho$ are given in the diagonally opposite location in the upper-right triangle.
-*P*-values $\\le$ 0.01, 0.05 and 0.1 are represented by symbols **, * and .
-, respectively.
-Diagonal shows density distribution of each variable.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/erd.stem.traits_cor.chart.jpeg"))
+```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Matrix for pairwise correlations (circles) among Effective Rooting Depth (ERD) and hydraulic properties for ", nrow(cor.data)," canopy species found on Barro Colorado Island (Panama).
+Each circle is color coded by the Spearman's $\\rho$ with darker red or blue shades for increasngly positive or negative $\\rho$ values, respectively.
+P-values are given only when they are $\\ge$ 0.05.
+Units are as follows: ERD in m, K~max~ in mmol m^-1^ s^-1^ MPa^-1^), all variables with $\\Psi$ in MPa.")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/erd.stem.traits_cor.matrix_all_p-val.jpeg"))
 ```
@@
 \
-Barro Colorado Forest Census Plot Data can be obtained online[@Condit.2019a;@Condit.2019b].
+Barro Colorado Forest Census Plot Data can be obtained online[@Condit.2019a; @Condit.2019b].
 
@@
 
-### Data at the Eddy Flux Tower
-Matteo to add. ET, TDR and may be vary briefly, GPP and VPD.
+### Microclimatic tower
 
-### $K_{leaf}$ vs. $\Psi_{leaf}$
-Joseph et al. to add.
+The tower used for these measurements is 41 m above ground, on a plateau on BCI [@Pau.2018].
+The eddy covariance system includes a sonic anemometer (CSAT3, Campbell Scientific, Logan, UT) and an open-path infrared CO2/H2O gas analyzer (LI7500, LiCOR.
+Lincoln, NE).
+Hi-frequency (10Hz) measurements were acquired by a datalogger (CR1000, Campbell Scientific) and stored on a local PC.
+Data were processed with a custom program using a standard routine described in [@Detto.2010].
+QA/QC criteria for removing erroneous values are listed in table 1 (see below).
+GPP was derived from daytime values of NEE by adding the corresponding mean daily ecosystem respiration obtained as the intercept of the light response curve [@Lasslop.2010].
+The light curve was fitted on a 15-days moving window using a rectangular hyperbolic function (runs with friction velocity less than 0.4 m s^-1^ were excluded).
+In order to compute daily time integrated budgets, gaps were filled using Artificial Neural Network [@Papale.2003] with hydro-meteorological inputs as predictors (soil moisture, solar radiation, temperature, VPD and air pressure, all measured on the same tower).
+To train the network, the dataset was randomly divided in a training set (70%), a validation set (15%) and a test set (15%).
+A two-layer feed-forward network with 10 sigmoid hidden neurons and linear output neurons, was trained using the Levenberg-Marquardt algorithm until the mean square error (MSE) of the validation set stop improving [@Hagan.1994].
+Performance, in term of MSE, was evaluated using the test set at the end of the training.
+This procedure was repeated 100 times to produce 100 estimates of GPP.
+Training multiple times generates different results due to different initial conditions and random sampling of the three sets.
+Ensemble was obtained as weighted average from the 100 ANN predictions using the MSE of the test set as weights according to:
+f_i=(∑_k▒f_ik^ANN /MSE_ik)/(∑_k▒〖1/MSE_ik 〗)
 
-### $K_{stem}$ vs. $\Psi_{stem}$
-Brett to add.
+The ANN was implemented using the Neural Network Toolbox in Matlab 2014a.
+
+### Leaf maximum hydraulic conductivity and vulnerability to cavitation
+
+The Kleaf was measured under steady state and high irradiance following the Evaporative Flux Method (FEM) [@Sack:2012]  using a flowmeter.
+Briefly, the leaf is excised under filtered DI water and attached to a water source (graduated cylinder) using clear plastic tubing.
+The system includes tubing of a known resistance, separated upstream and downstream by two sensors measuring the transpiration-driven water flow pressure in the EFM system [@Sack.2011].
+The K~leaf~ is calculated as the ratio of the transpiration rate (E~leaf~; mmol s^-1^) to the difference in flow pressure (MPa) upstream and downstream from the resistance tubing, and further standardized to 25°C and by leaf area.
+At stability (CV < 5%), the leaf is left to equilibrate in plastic bags with moist paper towels and measured for \Psi~leaf~ using a Scholander pressure chamber.
+To construct vulnerability curves for each species, four different functions were fitted to species-specific K~leaf~ vs.
+\Psi~leaf~ data using the *anneal* optimization function in the R package likelihood: exponential without an intercept ($K_{leaf} = Ae^{-B\Psi_{leaf}}$), exponential with an intercept ($K_{leaf} = C + Ae^{-B\Psi_{leaf}}$), logistic ($K_{leaf} = \frac{A}{1 + \left(\frac{\Psi_{leaf}}{x_{0}}\right)^B}$) and sigmoidal ($K_{leaf} = \frac{A}{1 + e^{-\left(\frac{\Psi_{leaf} - x_{0}}{B}\right)}}$).
+The differences among AICc values for these functions were not large within species (Zailaa et al.
+to be submitted; (see Zailaa et al.
+for more detailed methods description).
+
+### Stem maximum hydraulic conductivity and vulnerability to cavitation
+
+Brett to edit.
+
+Methods for Stem Maximum hydraulic conductivity and vulnerability to cavitation closely followed [Christman.2012] and [Wolfe.2016].
+Values for some of the species reported here were previously reported by @Dickman.2019 and @Wu.2020.
+The entire dataset and R script for the analyses are archived in the NGEE-Tropics archive (Wolfe 2020, To be done).
+\
+Canopy branches were cut from 4–10 trees of 26 species across three sites (Table SXX).
+The branch segments were enclosed in opaque plastic bags, and brought to the laboratory for measurements.
+Stem segments used for hydraulic conductivity (Kh) measurements were located at least 1 m from the cut end of the branch to minimize the number of open vessels that extended into the segment.
+Maximum vessel length was measured on a subset of samples from each species.
+Among species, maximum vessel length ranged XXX and length of stem between the cut that was made in the air to collect the branch and the section of stem used for Kh measurement was XXX (Table SXX).
+\
+Three leaves that were distal to the Kh segment were sealed in plastic bags and the entire branch was enclosed in black plastic bags for at least 30 minutes to stop transpiration.
+Just prior to measurement of Kh, leaf water potential of the three bagged leaves was measured with a pressure chamber.
+Since transpiration was stopped by bagging the leaves, leaf water potential was assumed to equal stem water potential (REFERENCE).
+\
+For each Kh segment, three measurements of stem water potential were made on stem sections that were located adjacent to the Kh segment.
+Stem sections that were approximately 40 mm in length were placed in nickel-plated aluminum chambers that were connected to Peltier-type psychrometers (Merrill Instruments, Logan, UT) and placed in a water bath at 30°C for 3 h to equilibrate the humidity within the cambers.
+Then water potential was measured using an automated multi-channel micro-voltmeter (CR7, Campbell Scientific, Logan, UT) with a 10 s thermocouple cooling current.
+For samples with water potential < -4 MPa, a 45 s cooling current was used to improve accuracy.
+Each psychrometer probe was calibrated against salt solutions beforehand and the plateau of the psychromatic response curve was used to assess water potential.
+\
+The Kh segments were removed from the branches by repeatedly cutting back the branch on both ends while it was submerged in water to relax the hydraulic tension within the xylem ((REFERENCE)).
+Then the ends were shaved clean with a fresh razor.
+Segments were 11–34 cm in length (mean ± SD = 22 ± 4 cm) and 0.2–1.6 cm in diameter (0.7 ± 0.2 cm).
+The segments were placed in a hydraulic conductivity apparatus similar to that described by Sperry et al.
+(1998) except that flow rate was measured with graduated pipettes instead of a balance.
+The perfusion solution was 10 mM KCl filtered to 0.2 m.
+For each segment, we measured the flow rate at three pressure heads: two that ranged 0.5–2 kPa and at zero pressure.
+Following Torres-Ruiz et al.
+(2012), Kh was calculated as the slope of the flow rate versus pressure gradient (i.e., pressure head divided by segment length) among the three pressure heads.
+Stem area specific hydraulic conductivity was (Ks) was calculated by dividing Kh by the cross-sectional area of stem.
+To calculate cross-sectional stem area, stem diameter was measured twice perpendicularly on both ends of the segment and averaged, then area was calculated assuming a circular cross section.
+\
+To assess vulnerability to cavitation, the branches were allowed to dry in the lab for 0 – 180 hours before they were measured for stem water potential and Ks.
+We used the water potential measured by the psychrometers because they are a more direct measurement.
+It has been noted that bagged-leaf water potential can deviate from stem water potential, especially when branch segments are relatively dry (REFERENCE).
+For some species, we noted this pattern while for others it was not apparent (Fig SXX).
+For each species, Ks was plotted against stem water potential and a Weibull curve was fit through the 90th  percentile of the points, following Wolfe et al.
+(2016)).
+The intercept of the curve was interpreted as maximum Ks and the water potential where the Weibull curve predicted 12, 50, 80, and 88 loss of maximum Ks were interpreted as P12, P50, P80, and P88, respectively.
 
@@
 
-```{r leaf-cover, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-cover}Seasonal variation in Leaf Cover Fraction for ", length(unique(sp.leaf_cover.for.model$sp[sp.leaf_cover.for.model$deciduousness != "Evergreen"]))," non-evergreen species at Barro Colorado Island.
-Leaf Cover Fraction for evergreen species (*n* = ", length(unique(sp.leaf_cover.for.model$sp[sp.leaf_cover.for.model$deciduousness == "Evergreen"])),") is assumed at 1 through the year.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/leaf.cover_BCI_single_panel_evergreensAt100.jpeg"))
+```{r leaf-cover, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-cover}Seasonal variation in Leaf Cover Fraction for ", length(erd.sp)," species at Barro Colorado Island (panels).
+Species are color-coded by deciduousness.")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf.cover_BCI_multi_panel.jpeg"))
 ```
@@
 
-```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($Psi_{soil,~z}$) depth z over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines).
-Vertical gray lines show median dates of the demographic censuses at the BCI 50 ha plot."}
+```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($Psi_{soil,~z}$) by depth z (m; panels) over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines).
+Vertical gray lines demarcate median dates of the demographic censuses at the BCI 50 ha plot."}
 knitr::include_graphics(file.path("figures/2019-10-14_5000/best-fits/psi_model_daily_all_depths_params.top.few_full.jpeg"))
@@
 
+```{r eddy, results = 'asis', echo = FALSE}
+knitr::kable(eddy.table, caption = 'QA/QC procedure applied to eddy covariance fluxes.
+w: vertical wind speed; t: air temperature; q: water vapor density; c: CO~2~ density; u: horizontal wind speed.
+The bar indicates 30-min averaging and the apex indicates fluctuation ($\\overline{x` = x - \\bar{x}}$)', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(eddy.table))))
+```
+
 # References

### 2020-10-08 manuscript.rmd
1. changed objective function equation alignment 2. added ERD equation 3. re-arranged methods with leaf vulnerability curves now a sub-section of ERD parametarization

@@
 ---
-title: "Drought-sensitive trees survive on deep-water access in a tropical forest community"
+title: "Hydraulically-vulnerable trees are buffered by deep-water access in a tropical forest community"
 author: ''
@@
 Despite this, widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010] and tropical mortality rates are increasing in some regions [@Brienen.2015; @Hubau.2020].
-However, state-of-the-art dynamic vegetation models (DVMs) struggle to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013; @Powell:2018], as underlying mechanisms of drought-tolerance are not fully understood or quantified.
+However, state-of-the-art dynamic vegetation models (DVMs) struggle to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013; @Powell.2018], as underlying mechanisms of drought-tolerance are not fully understood or quantified.
 
 \
-Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism.
-This mechanism assumes that water availability varies below-ground by vertical depth, and that the rooting depths of colocated trees also vary, by species and size [@Meinzer:1999; @Chitra-Tarak.2018].
-Consequently, trees in the same forest differ in their drought experience, and thus in their growth and mortality responses.
+Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied.
+This mechanism assumes that water availability varies with soil depth, and that the rooting depths of co-located trees also vary, by species and size [@Meinzer:1999; @Chitra-Tarak.2018; @Brum.2019].
+Consequently, trees in the same forest differ in their drought experience, and thus in their growth and mortality responses [@Chitra-Tarak.2018].
 A key bottleneck in community-wide testing of this mechanism has been lack of data on trees' rooting or water-sourcing depths, followed by data on water-availability at those depths.
-A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the world’s 60 thousand tree species [@Fan:2017; @Beech:2017].
-Further, only a small fraction of those are tropical, even though ~70-90 % (40 - 53 thousand species) of the world’s tree diversity resides in the tropics [@Slik:2015].
-Use of stable isotopes of water as a tracer provides an indirect measure of water sourcing depths by matching the isotopic value in plant tissue to variations in soil pore water with depth.
-However, such data are extremely rare [@Evaristo:2016].
-In general, community-scale data collection for rooting or water-sourcing depths is especially challenging for species-rich tropical forests.
+A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the world’s more than 60 thousand tree species [@Fan:2017; @Beech:2017].
+Further, only a small fraction of those are tropical, even though more than 90% of the world’s tree diversity resides in the tropics [@Slik:2015].
+Use of stable isotopes of water as a tracer provides an indirect measure of water sourcing depths by matching the isotopic value in xylem water to variations in soil pore water with depth.
+However, such data are rare [@Evaristo:2016].
+In general, community-scale data collection for rooting or water-sourcing depths is challenging for species-rich tropical forests.
 
 \
-Trade-offs in plant 'traits' shape how plants make a living in their environment (ecological strategies) and also provide the essential constraints to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010].
-This entails identifying relationships among environment, traits and demography.
+Trade-offs in plant 'traits' shape how plants make a living in their environment (ecological strategies) and also provide the essential constraints to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010; @Bartlett.2019].
+Understanding how plant trade-offs lead to alternative life history strategies entails identifying the topography of trait trade-offs in different environments, and how these relate to demographic rates (growth, recruitment and mortality).
 Specifically for the water environment (or, resource), observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified an acquisitive-conservative resource economic spectrum similar to those for light and nutrients, although, for water, examples of all trade-offs observed simultaneously are rare, and below-ground traits or environment linkages are weak [@Reich:2014; @Sterck:2011].
-Acquisitive species have traits with high capacity to acquire, move, store and use water (higher hydraulic conductivity, low wood density) and achieve high growth, but build flimsy tissues (high SLA) and are less tolerant of low resource availablity, whereas species with conservative traits are better protected from carbon losses and are more physiologically drought tolerant (greater embolsim resistance, larger safety margin).
+Acquisitive species have traits with high capacity to acquire, move, store and use water (higher hydraulic conductivity, low wood density) and achieve high growth, but build flimsy tissues (high specific leaf area) and are less tolerant of low resource availablity, whereas species with conservative traits are better protected from carbon losses and are more physiologically drought tolerant (greater embolism resistance, larger safety margins).
 Whether and how the acquisitive-conservative spectrum holds true for the water dynamics vertically below-ground is not clear.
@@
 Deep rooting may be expected to support an acquisitive strategy by enabling a steady water supply.
-However, the degree to which deep rooting is associated with an acquisitive strategy may depend on the exact dynamics of water availability by depth at a site, depending on precipitation regime and hydrology.
-In pulse-driven systems *shallow*-rooted plants are associated with acquisitive traits (higher hydraulic conductivity) that maximise the use of the water-pulse, while *deep*-rooted plants show conservative traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013 and references therein].
+However, the degree to which deep rooting is associated with an acquisitive strategy may depend on the specific dynamics of water availability by depth at a site, which is affected by precipitation regime and hydrology.
+For example, in pulse-driven systems *shallow*-rooted plants are associated with acquisitive traits (higher hydraulic conductivity) and opportunistic strategies (e.g.
+dry deciduousness) that maximise the use of the water-pulse, while *deep*-rooted plants show conservative traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013 and references therein].
 
@@
 \
-To better understand how rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tropical forest tree species, in this paper, we conduct a thorough assessment for a tropical forest of relationships between the water environment in the whole-soil column, above-ground hydraulic traits, estimates of effective rooting depths and demographic outcomes over 35 years that witnessed droughts of various intensity and duration.
-Specifically, we test an overarching hypothesis that inter-specific variation in trees' effective rooting depths aligns species on an acquisitive-conservative economic spectrum, .
-whereby acquisitive trees have rooting depths optimized for increasingly reliable water availability by depth.
-Our hypotheses are that (i) trees with deeper ERD have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallower ERD species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deeper ERD species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallower ERD species may have higher mortality rates as a result of higher variability in shallow soil water availability.
+To better understand how rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tropical forest tree species, in this paper, we conduct a thorough assessment for a tropical forest of relationships between the water environment in the whole-soil column, above-ground hydraulic traits, estimates of rooting depths and demographic outcomes over 35 years that witnessed droughts of various intensity and duration.
+Specifically, we test an overarching hypothesis that inter-specific variation in trees' rooting depths aligns species on an acquisitive-conservative economic spectrum, whereby acquisitive trees have rooting depths optimized for more reliable water availability by depth.
+Our hypotheses are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallow rooted species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deep rooted species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallow rooted species may have higher mortality rates as a result of higher variability in shallow soil water availability.
 To our knowledge this is the first study to demonstrate a mechanistic link between trees' realized drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.
@@
 ```{r hypo, results = 'asis', echo = FALSE}
-knitr::kable(hypo.table, caption = 'Hypotheses for association between Effective Rooting Depth (ERD) and hydraulic traits.', booktabs = TRUE, escape = FALSE)
+knitr::kable(hypo.table, caption = 'Hypotheses for association between Effective Rooting Depth (ERD) and hydraulic traits.', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(hypo.table))))
 ```
@@
 \
-In the 50 ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species level and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm size class [@Condit.2017].
+In the 50 ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm size class [@Condit.2017].
 This inventory represents 321 woody species across the eight censuses.
@@
 \
-To estimate dynamics of water availability by depth over the forest’s rooting zone, we implemented a 1-D hydrological model, the Energy Exascale Earth System Land Model (ELM), coupled with the Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES).
-FATES is a state-of-the-art, community-based, open-source model used for studying climate-vegetation interactions.
-FATES is a next generation vegetation dynamics model for ELM, with a size-structured group of plants (cohorts) and successional trajectory-based patches based on the ecosystem demography approach [@Moorcroft.2001].
+To estimate dynamics of water availability by depth over the forest’s rooting zone, we used the Energy Exascale Earth System Land Model (ELM), coupled with the Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES).
+ELM is a land-model that, among many features, simulates the physics and conservative dynamics of water, energy and carbon fluxes.
+In particular, soil hydrologic fluxes are vertically resolved among discrete soil layers (1-D) [@Bonan:2013].
+FATES is a community-based, open-source model used for studying climate-vegetation interactions.
+FATES is a vegetation demography model, with a size-structured group of plants (cohorts) and successional trajectory-based patches based on the ecosystem demography [@Moorcroft.2001] approach.
 FATES couples to ELM by a common interface of water and carbon fluxes.
-Detailed descriptions for ELM and FATES can be found elsewhere [@Koven:2020; @Fisher:2014; @Bonan:2013].
+Detailed descriptions for ELM and FATES can be found elsewhere [@Fisher.2010; @Fisher.2015; @Koven:2020].
 
 \
-Below we describe an equation of water balance in ELM-FATES that is most relevant for our study site, even though ELM-FATES is capable of incorporating numerous other hydrological processes relevant for wetlands and lakes and snow [@Bonan:2013].
-ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\Delta W_{can}$, surface water $\Delta W_{sfc}$, soil water $\Delta W_{liq, z}$, and water in an unconfined aquifer $\Delta W_{a}$ (omitting processes relevant to snow, wetlands or lakes).
-The equation can be described as:
+ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\Delta W_{can}$, surface water $\Delta W_{sfc}$, soil water $\Delta W_{liq, i}$ at depth z~i~ and water in an unconfined aquifer $\Delta W_{a}$ (omitting processes relevant to snow, wetlands or lakes), and conservation of these terms is expressed as follows:
 
@@
 \begin{split}
-\Delta W_{can} + \Delta W_{sfc} + \sum_{z=1}^{15} \Delta W_{liq,z} + \Delta W_{a} = \left(q_{rain} - E_{v} - E_{g} - q_{over} - q_{h2osfc} - q_{drai}\right)\Delta t
+\Delta W_{can} + \Delta W_{sfc} + \sum_{i=1}^{15} \Delta W_{liq,i} + \Delta W_{a} = \left(q_{rain} - E_{v} - E_{g} - q_{over} - q_{h2osfc} - q_{drai}\right)\Delta t
 \end{split}
@@
 
-where, $q_{rain}$ is rainfall, $E_{v}$ is evapotranspiration (ET) of the forest, $E_{g}$ is ground evaporation, $q_{over}$ is surface runoff, $q_{h2osfc}$ is runoff from surface water storage, $q_{drai}$ is sub-surface drainage (all in mm H~2~O) and $\Delta t$ is the time-step.
-$z$ represents the depths of 15 exponentially distributed soil layers: $\simeq$ 0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13, 21 and 35 (m).
+where, $q_{rain}$ is rainfall, $E_{v}$ is evapotranspiration (ET) of the forest, $E_{g}$ is ground evaporation, $q_{over}$ is surface runoff, $q_{h2osfc}$ is runoff from surface water storage, $q_{drai}$ is sub-surface drainage (all in mm H~2~O) and $\Delta t$ is the time-step (typically, 1 hour).
+The soil profile is discretized into 15 exponentially distributed soil layers with depth z $\simeq$ 0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13, 21 and 35 (m).
 
@@
 \
-To parameterize the model, we identified eleven parameters relevant for Eq.
-\@ref(eq:water-balance), determined their ranges based on literature for the study site or for the tropics (Table \@ref(tab:param)), and sampled 5000 ensembles from this global parameter space with Latin Hypercube Sampling (LHS).
-In the Conrad catchment existing data on soil hydraulic conductivity for the depths of 12.5 and 60 cm showed a decline by depth (@Kinner:2004; Table \@ref(tab:param)).
+To parameterize ELM-FATES, we identified eleven parameters relevant for Eq.
+\@ref(eq:water-balance), determined their ranges based on literature for the study site or for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling (LHS) of this global parameter space.
+In the Conrad catchment, existing data on soil hydraulic conductivity for the depths of 12.5 and 60 cm showed a decline of conductivity by depth (@Kinner:2004; Table \@ref(tab:param)).
 We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths < 12.5 cm and > 60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig.
@@
 We converted gravimetric water content to volumetric (VWC, $\theta$) using a bulk density value of of 0.8 g cm^-3^ measured at a point location (see below).
-A @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences among depths were nominal,
+A @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences in water retention among depths were nominal,
 
@@
 \
-We ran ELM-FATES for the 5000 parameter ensembles with hourly climate drivers from 1985-2018 [@Faybishenko.2020] initialized with the stand structure of the 50 ha plot [@Condit.2019a] in a single site mode ($0.1^\circ$ x $0.1^\circ$ centered at $9.1530^\circ$N, $280.2539^\circ$E).
+We ran ELM-FATES for the 5000-member parameter ensemble with hourly climate drivers from 1985-2018 [@Faybishenko.2020] initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode.
 As our key interest here was in soil water availability rather than forest dynamics, we ran ELM-FATES in a lower-complexity configuration: static stand structure, and with a single plant functional type (PFT) of evergreen trees.
-The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of dry-deciduous PFT did not significantly alter results (not shown).
+The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of a dry-deciduous PFT did not significantly alter results (not shown).
 Holding the stand structure static rather than allowing it to emerge from physiologically-mediated competition between trees removes a set of internal model feedbacks and thus enables a cleaner experimental design where the only differences between model simulations are the direct physiological effects of perturbed model parameters.
+
+\
+We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq.
+\@ref(eq:water-balance), namely, (i) evapotranspiration ($ET \simeq E_{v} + E_{g}$) from the eddy flux tower (2012-2017; see [SI](#SIData.matteo)), (ii) stream discharge for the Conrad Catchment ($q_{run} \simeq q_{over} + q_{drai}$; 2012-2018) and (iii) soil volumetric water content (VWC, $\theta_{z} \simeq W_{liq, z}$) from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we built with a vertical TDR probe over 0-15 cm depth ($\theta_{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m (1299 samples in total covering all soil types and habitats; see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\theta_{0.15 m}$, $\theta_{0.4m}$ and $\theta_{1 m}$ refer to plot-wide averages).
+
+For ELM-FATES calibration we calculated an objective function for each of the 5000-member ensemble by equally weighting standardized RMSE between observations and simulations across all fluxes and states.
+We describe the objective function $O$ as :
+
+\begin{align*}
+\overline{O} = \frac{1}{6}\sum \left(RMSE^{*}(ET,~E_{v} + E_{g}) + RMSE^{*}(q_{run},~q_{over} + q_{drai}) + RMSE^{*}(\theta_{0-0.15m}, {W_{liq,~\overline{z}}})\\
++ RMSE^{*}(\theta_{0.15m}, W_{liq,~`r soil.depths[4]`m}) + RMSE^{*}(\theta_{0.4m}, W_{liq,~`r soil.depths[6]`m}) + RMSE^{*}(\theta_{1 m}, W_{liq,~`r soil.depths[8]`m})\right)
+(\#eq:obj-func)
+\end{align*}
+
+based on monthly sums for both $ET$ and $q_{run}$ and daily averages for $\theta$s.
+$W_{liq,~\overline{z}}$ is the daily average of $W_{liq,~z}$ for $z$ $\in$ {`r paste(soil.depths[1:4], collapse=", ")`} m.
+Superscript * indicates that RMSE is standardised between 0 and 1 across the 5000 simulations.
+
+Preliminary analyses showed that minimising the objective function $\overline{O}$, that is, simultaneously fitting the model to all states and fluxes did not capture the soil moisture dynamics at 1 m depth well, our key priority for model calibration.
+To ensure model fit with soil moisture dynamics at 1 m depth, we first chose 300 parameter ensembles that minimised RMSE for VWC at the depth of 1m ($RMSE^{*}(\theta_{1 m}, W_{liq,~`r soil.depths[8]`m})$ < 0.2), and among those 300 ensemble members we chose 100 members with the lowest values for the objective function $\overline{O}$.
+See Table \@ref(tab:param) for the ranges of best-fit values for different parameters.
 
@@
 \
-We calibrated the model over 2012-2018 against three key fluxes in Eq.
-\@ref(eq:water-balance), namely, (i) evapotranspiration ($ET \simeq E_{v} + E_{g}$) from the eddy flux tower (2012-2017; see [SI](#SIData)), (ii) stream discharge for the Conrad Catchment ($q_{run} \simeq q_{over} + q_{drai}$; 2012-2018) and (iii) soil volumetric water content (VWC, $\theta_{z} \simeq W_{liq, z}$; 2012-2018).
-We collected VWC data (2012-2018) from a point location at the eddy flux tower with a vertical time domain reflectometry (TDR) probe ($\theta_{0-0.15 m}$).
-This data was calibrated based on measurements of gravimetric soil moisture at the vertical probe (0-15 cm) on several occasions (Fig.
-S\@ref(fig:TDR-calib)) and mean bulk dry density of 0.8 g cm^-3^ (*n* = 5, cylinder volume = 1250 cm^3^).
-We also leveraged existing opportunistic dry season gravimetric measurements by depth for the entire 50-ha plot, converted to VWC units assuming a bulk density of 0.8 g cm^-3^ ($\theta_{0.15 m}$, $\theta_{0.4m}$ and $\theta_{1 m}$ in 2015 and $\theta_{0.15m}$, $\theta_{0.4m}$ in 2016; @Kupers:2019).
-Beginning in 2016 we additionally installed three horizontal probes at depths of 0.15, 0.4 and 1 m and we applied the same calibration equation to these data as that for the vertical probe.
-However, for VWC calibration we used data from only the vertical TDR probe ($\theta_{0-0.15 m}$) to capture the long-term dynamics and the plot-wide measurements to capture the dynamics by depth.
-Here, in general, our interest was to capture the soil moisture dynamics by depth, so we performed model calibration in a two step process: first, we prioritized model fit with $\theta_{0.1m}$.
-We chose ensembles that minimized root mean square error (RMSE) between $\theta_{0.1m}$ and W~liq,`r soil.depths[8]`m~ (~300 ensembles with standaridised RMSE (range 0-1) across ensembles < 0.2).
-Second, we calculated an objective function for each of the 5000 ensembles by equally weighing standardized RMSE between observations and simulations across all fluxes, and chose 100 ensembles, within the 300 that best-fit $\theta_{0-0.15 m}$, that had the lowest values of the objective function.
-Let $f(obs,~sim)$ be RMSE between observations and simulations standardized across ensembles (range 0-1), based on monthly sums for $ET$ and $q_{run}$, respectively, and daily averages for $\theta$s, then the objective function ($\overline{O}$) can be described as
+We evaluated ELM-FATES by calculating RMSE between simulated and observed long-term daily VWC for the depths of 0.1, 0.4 and 1 m (2016-2018), baesd on a dataset we had left-out during calibration.
+We obtained these observations from three horizontal time domain reflectometry probes at the depths of 0.1, 0.4 and 1 m at a point location near the eddy flux tower (see [SI](#SIData.VWC)).
+
+\
+We ran ELM-FATES with the chosen 100 ensemble members from 1985-2018, with the first 5 years used for model spin-up.
+
+## Model for Effective Rooting Depth (ERD)
+\
+Water is a key limiting resource for tree growth.
+Vapor pressure deficit (VPD) and water supply as determined by soil moisture are factors known to influence water uptake [@Lawrence.2019].
+In this study, we used leaf conductivity ($K_{leaf}$) determined by rhizosphere soil water potential ($\Psi_{soil}$) as an approximation of the soil water limiting factor.
+For dry-deciduous species seasonality in leaf cover may also influence plant water uptake.
+We therefore tested alternate models of growth, aka inverse models for ERD, in which a species' growth dynamics across census intervals was regressed against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1 (see below).
+For the chosen model (see Model Evaluation below), daily average growth for species $s$, $G_{s,~t~|~z}$, in the census interval $t$ for depth $z$ and for a given hydrauligoical realisation $h$ is described as follows (see SI for equations for all the tested models):
 
 \begin{equation}
-\overline{O} = \frac{1}{6}\sum \left(f(ET,~E_{v} + E_{g}) + f(q_{run},~q_{over} + q_{drai}) + f(\theta_{0-0.15m}, {W_{liq,~\overline{z}}}) + f(\theta_{0.15m}, W_{liq,~`r soil.depths[4]`m}) + f(\theta_{0.4m}, W_{liq,~`r soil.depths[6]`m}) + f(\theta_{1 m}, W_{liq,~`r soil.depths[8]`m})\right)
-(\#eq:obj-func)
+G_{s,~t~|~z,~h} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} (K^{*}_{leaf,~s,~i~|~z,~h} + VPD^{*}_{i}) LAI^{*}_{s,~doy~|~i}) + \epsilon_{t~|~z,~h}
+(\#eq:M-chosen)
 \end{equation}
 
-where $W_{liq,~\overline{z}}$ is the daily average of $W_{liq,~z}$ for $z$ = `r paste(soil.depths[1:4], collapse=", ")` m.
-See Table \@ref(tab:param) for the range of best-fit values for different parameters.
-We ran ELM-FATES with the chosen 100 ensembles from 1985-2018, with first 5 years used for model spin-up.
+where n~t~ be the total number of days in census interval *t*, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is a normally distributed model error term.
+We estimated daily value of $K_{leaf,~s,~i~|~z,~h}$ for a species $s$ by substituting $\Psi_{soil,~i~|~z,~h}$ in the leaf vulnerability curves defined below in Eq.
+\@ref(eq:vc-exp) using species-specific parameters $A_{s}$ and $B_{s}$ as:
 
-## Leaf Vulnerability Curves
+\begin{equation}
+K_{leaf,~s,~i~|~z,~h} = A_{s}e^{-B_{s}\Psi_{soil,~i~|~z,~h}}
+(\#eq:k-soil)
+\end{equation}
+
+We estimated species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015).
+Using Eq.
+\@ref(eq:M-chosen), for a species $s$, we obtained Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $\rho$ for each soil depth $z$ for each of the 100 best-fit hydrological realizations.
+For each hydrological realization we retained only one depth that corresponded to the least AIC score (the best-fit), and within those depths retained the ones that had Pearson's $r$ >= 0.71 (or, $r^2$ >= 0.5) as a valid set.
+For a species $s$, we defined effective rooting depth (ERD) as the median of these set of depths, described as :
+
+\begin{equation}
+ERD_{s} = \underset{\forall \, h \, \in \, H}{median} \Big\{z,h\,\Big|\, \Big(AIC_{s~|~z,~h} = \underset{\forall \, z \,\in\,Z}{min}(AIC_{s~|~z,~h})\Big) \cap \Big(r_{s~|~z,~h} \ge 0.71\Big)\Big\}
+(\#eq:erd)
+\end{equation} 
+where $H$ $\in$ {1, .., 100}, refer to the 100 hydrological realisations and $Z$ $\in$ {`r paste(soil.depths, collapse=", ")`} m, covering the range of soil depths in ELM-FATES after above parametarision.
+
+### Model parameterization
+
+#### Growth data
+\
+For growth estimates, to minimize the effect of light variation among trees, we selected only large trees ($\ge$ 30 cm dbh, the diameter at breast height) and species whose maximum height was $\ge$ 30 m (hereafter, canopy species) and thus are likely to be fully exposed to the sun.
+We selected trees for which the height of the diameter measurements were unchanged across all censuses (1990, 1995, 2000, 2005, 2010, 2015).
+For each individual tree (*i*), we calculated absolute growth rates ($G_{i,t}$; cm yr^-1^) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses, divided by the inter-census interval based on the exact date of measurements.
+We eliminated positive outliers as any growth rates amounting to > 75 mm yr^-1^, as this was the highest rate observed with confidence for fastest growing species [@Condit.2017].
+Negative outliers were removed based on known measurement error, $\sigma_{e} = 0.006214d + 0.903$, where *d* is dbh.
+Any stem whose later dbh measurement fell < 4$\sigma_{e}$ the earlier measurement was removed [@Condit.2017].
+As tree growth rates vary with size [@Muller-Landau:2006], to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($Resid_{i,t} = G_{i,t} - f(d_{i, t})$).
+The diameter model of growth ($f(d_{i, t})$), a B-spline based polynomial equation with 5 degrees of freedom, was fitted to growth rates from all trees pooled together as there was insufficient species-specific data for large trees (Fig.
+S\@ref(fig:dbh-model)).
+We retained only those trees that had $Resid_{i,t}$ estimates for all census intervals (total trees = `r sum(g.n$n)`) and standardized (centered and scaled) individual tree time series of $Resid_{i,t}$.
+We then obtained `r length(erd.sp)` species-level time-series of medians of $Resid_{i,t}$ ($G_{t}$).
+These were obtained for only those species with complete records on at least three trees (median *n* = `r median(g.n$n)`, max *n* = `r max(g.n$n)` trees per species).
+
+\
+There are many factors that could affect plant growth, including soil moisture, VPD, leaf area, temperature and radiation.
+Based on our preliminary analyses, incorporation of temperature and radiation did not improve our model fitting, thus we mainly focused on the impact of soil moisture, VPD and leaf area in our models.
+
+#### Leaf Vulnerability Curves
 \
@@
 \begin{equation}
-K_{leaf} = Ae^{-B\Psi_{leaf}} + error
+K_{leaf,~s} = A_{s}e^{-B_{s}\Psi_{leaf}} + error
 (\#eq:vc-exp)
@@
 \begin{equation}
-A = `r acf[1]` `r acf[2]`WSG `r acf[3]`WSG^2 + `r acf[4]`WSGB + `r acf[5]`B `r acf.6`B^2 + error
+A_{s} = `r acf[1]` `r acf[2]`B_{s} `r acf[3]`B_{s}^2 + `r acf[4]`B_{s}LMA_{s} + `r acf[5]`LMA_{s} `r acf.6`LMA_{s}^2 + error
 (\#eq:a-mod)
@@
 \begin{equation}
-B = `r bcf[1]` `r bcf[2]`B + `r bcf[3]`B^2 + `r bcf[4]`BLMA + `r bcf[5]`LMA `r bcf[6]`LMA^2 + error
+B_{s} = `r bcf[1]` `r bcf[2]`WSG_{s} + `r bcf[3]`WSG_{s}^2 + `r bcf[4]`WSG_{s}LMA_{s} + `r bcf[5]`LMA_{s} `r bcf[6]`LMA_{s}^2 + error
 (\#eq:b-mod)
@@
 As these fits explained large proportion of variation in parameters $A$ and $B$ (see results), we sequentially used Eq.
-\@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r length(unique(bci.AB$sp[!is.na(bci.AB$A)]))` species for which WSG and LMA were both available, retained `r length(unique(data.model.AB$sp))` for which parameters $A$ and $B$ were both positive, and fitted vulnerability curves to those.
+\@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r length(unique(bci.AB$sp[!is.na(bci.AB$A)]))` species for which WSG and LMA were both available, retained `r length(unique(data.model.AB$sp))` for which parameters $A$ and $B$ were both positive, and fitted vulnerability curves to those (Eq.
+\#ref(eq:vc-exp)).
 
@@
 
-## Model for Effective Rooting Depths (ERD)
+#### VPD
 \
-Water is a key limiting resource for tree growth.
-Vapor pressure deficit (VPD) and water supply as determined by soil moisture are factors known to influence water uptake [@Lawrence.2019].
-In this study, we used leaf conductivity ($K_{leaf}$) determined by rhizosphere soil water potential ($\Psi_{soil}$) as an approximation of the soil water limiting factor.
-For dry-deciduous species seasonality in leaf cover may also influence plant water uptake.
-We therefore tested alternate models of growth, aka inverse models for ERD (Eqs.
-\@ref(eq:M2)-\@ref(eq:M6)), in which a species' growth dynamics across census intervals was regressed against census interval means of multiplied or additive effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the fraction of leaf cover (all standardized to range from 0-1; see below).
-Let n~t~ be the total number of days in census interval $t$, then for individual tree species Growth~t~ and depth $z$ these models are described as follows:
+As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between mean daily values by month of gross primary productivity (gC m^2^ day^-1^) and VPD (kPa) we collected from the eddy flux tower located 75 m from the 50-ha plot (for any day *i*, $GPP_i = `r vpd.cf[1]` + `r vpd.cf[2]`VPD_i + `r vpd.cf[3]`VPD_i^2 + error$; 2012-2017; Fig.
+S\@ref(fig:gpp-vpd)) and used standardized $GPP$ to represent the impact of VPD on growth (hereafter, standardized $GPP$ = $VPD^*$ in Eq.
+\@ref(eq:M-chosen); range 0-1).
+At our study site, due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season, the impact of VPD may as well be a combined impact of not only temperature and humidity, but also radiation.
 
-\begin{equation}
-Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,~z,~n} + error
-(\#eq:M1)
-\end{equation}
-
-\begin{equation}
-Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,~z,~n} StdVPD_{n} + error
-(\#eq:M2)
-\end{equation}
-
-\begin{equation}
-Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,~z,~n} + StdVPD_{n} + error
-(\#eq:M3)
-\end{equation}
-
-\begin{equation}
-Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,~z,~n} LeafCover_{doy} + error
-(\#eq:M4)
-\end{equation}
-
-\begin{equation}
-Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,~z,~n}StdVPD_{n}LeafCover_{doy} + error
-(\#eq:M5)
-\end{equation}
-
-\begin{equation}
-Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} (StdK_{leaf,~z,~n} + StdVPD_{n}) LeafCover_{doy} + error
-(\#eq:M6)
-\end{equation}
-We estimated daily value of $K_{leaf,z,n}$ by substituting $\Psi_{soil,~z,~n}$ in the leaf vulnerability curves Eq. \@ref(eq:vc-exp) as:
-
-\begin{equation}
-K_{leaf,~z,~n} = Ae^{-B\Psi_{soil,~z,~n}}
-(\#eq:k-soil)
-\end{equation}
-
-We estimated species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015; see below).
-
-### Model parameterization
+#### Leaf Area Index (LAI)
 \
-For growth estimates, to minimize the effect of light variation among trees, we selected only large trees ($\ge$ 30 cm dbh, the diameter at breast height) and species whose maximum height was $\ge$ 30 m (hereafter, canopy species) and thus are likely to be fully exposed to the sun.
-We selected trees for which the height of the diameter measurements were unchanged across all censuses (1990, 1995, 2000, 2005, 2010, 2015).
-For each individual tree (*i*), we calculated growth rates ($Gr_{i,t}$; cm yr^-1^) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses, divided by the inter-census interval based on the exact date of measurements.
-We eliminated positive outliers as any growth rates amounting to > 75 mm yr^-1^, as this was the highest rate observed with confidence for fastest growing species [@Condit.2017].
-Negative outliers were removed based on known measurement error, $\sigma_{e} = 0.006214d + 0.903$, where *d* is dbh.
-Any stem whose later dbh measurement fell < 4$\sigma_{e}$ the earlier measurement was removed [@Condit.2017].
-As tree growth rates vary with size [@Muller-Landau:2006], to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($Resid_{i,t} = Gr_{i,t} - f(d_{i, t})$).
-The diameter model of growth ($f(d_{i, t})$), a B-spline based polynomial equation with 5 degrees of freedom, was fitted to growth rates from all trees pooled together as there was insufficient species-specific data for large trees (Fig.
-S\@ref(fig:dbh-model)).
-We retained only those trees that had $Resid_{i,t}$ estimates for all census intervals (total trees = `r sum(g.n$n)`) and standardized (centered and scaled) individual tree time series of $Resid_{i,t}$.
-We then obtained `r length(erd.sp)` species-level time-series of medians of $Resid_{i,t}$ ($Growth_{t}$).
-These were obtained for only those species with complete records on at least three trees (median *n* = `r median(g.n$n)`, max *n* = `r max(g.n$n)` trees per species).
-
-\
-There are many factors that could affect plant growth, including soil moisture, VPD, leaf area, temperature and radiation.
-Based on our preliminary analyses, incorporation of temperature and radiation did not improve our model fitting, thus we mainly focused on the impact of soil moisture, VPD and leaf area in our models.
-
-\
-As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between mean daily values by month of gross primary productivity (gC m^2^ day^-1^) we collected from the eddy flux tower located 75 m from the 50-ha plot and VPD (kPa; $\overline{GPP} = `r vpd.cf[1]` + `r vpd.cf[2]`VPD + `r vpd.cf[3]`VPD^2 + error = StdVPD$; 2012-2017; Fig.
-S\@ref(fig:gpp-vpd)) and used standardized $\overline{GPP}$ to represent the impact of VPD on growth (hereafter, $StdVPD$; range 0-1) in Eqs.
-\@ref(eq:M2), \@ref(eq:M3), \@ref(eq:M5) and \@ref(eq:M6).
-Due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season at our study site, the impact of VPD may as well be a combined impact of temperature, humidity, and radiation.
-
-\
-To explore the effect of species seasonality in leaf cover on growth, we assumed of species-specific seasonal curves for Leaf Cover Fraction ($LeafCover$, range 0-1, unitless).
-For evergreen species we assumed $LeafCover$ = 1 throughout the year, whereas for non-evergreen species (various deciduous classes), we assumed $LeafCover$ = 1 during the wet season (DOY = 150 to 300) given the high forest-level leaf area index for this period (not shown), and obtained an estimate of $LeafCover$ for the rest of the year as follows.
+To explore the effect of species seasonality in leaf phenology on growth, we assumed species-specific seasonal curves for leaf area index ($LAI$, range 0-1, unitless).
+For evergreen species we assumed $LAI$ = 1 throughout the year, whereas for non-evergreen species (various deciduous classes), we assumed $LAI$ = 1 during the wet season (DOY = 150 to 300) given the high forest-level leaf area index for this period (not shown), and obtained an estimate of $LAI$ for the rest of the year as follows.
 We interpolated species-specific weekly leaf-fall data collected at two similar sites in BCI (1985-2020) to a daily estimate, scaled to express daily leaf fall as a fraction of the total leaf fall in a given year and obtained an average leaf fall fraction by DOY estimate across years ($LeafFall$, range 0-1, unitless).
-Thus for non-evergreen species, we cumulatively deducted $LeafFall$ from $LeafCover$ = 1 beginning DOY = 300 and obtained a $LeafCover$ by DOY curve (Fig.
+Thus for non-evergreen species, we cumulatively deducted $LeafFall$ from $LAI$ = 1 beginning DOY = 300 and obtained a $LAI$ by DOY curve (Fig.
 S\@ref(fig:leaf-cover)).
 We obtained species-specific parameters $A$ & $B$ for Eq.
-\@ref(eq:k-soil) either from data when available, or a model based on $LMA$ or $WSG$ as described in the earlier section, and used standardized K within species ($Std.K$; range 0-1) in Eqs.
-\@ref(eq:M1)-\@ref(eq:M6).
-
-\
-For the combination of each model \@ref(eq:M1)-\@ref(eq:M6), species and 100 hydrological realizations, we obtained AIC, goodness-of-fit (*R*^2^) and Pearson's $\rho$ for each soil depth $z$.
-For each combination of model-species-hydrological realization we retained only one depth with the least AIC score (the best-fit), and within those depths retained the ones that had a positive corresponding Pearson's $\rho$ and *R*^2^ >= 0.5 as a valid set.
-For each model-species combination, we defined effective rooting depth (ERD) as the median of these subset of depths and calculated standard error.
+\@ref(eq:k-soil) either from data when available, or a model based on $LMA$ or $WSG$ as described in the earlier section, and used standardized K within species ($K^*$; range 0-1) in Eq.
+\@ref(eq:M-chosen).
 
@@
 \
-For each model, species ERD was regressed against species xylem sap $\delta^2H$ (permil; *n* = 7).
-We used xylem sap $\delta^2H$ values given in @Meinzer:2001 for the dry-season of March 1997 as these showed largest seasonal divergence among species, with a corresponding gradient of $\delta^2H$ in the soil (Fig.
+For each model of ERD (Eq.
+\@ref(eq:M-chosen) and Eqs.
+in SI), species ERD was regressed against species xylem sap $\delta^2H$ (permil).
+We used xylem sap $\delta^2H$ values given in @Meinzer:2001 for the dry-season of March 1997 as this period showed largest seasonal divergence among species, with a corresponding gradient of $\delta^2H$ in the soil (Fig.
 S\@ref(fig:iso-data)).
@@
 standleyana*, whose xylem sap isotope value was not well constrained by the soil water isotope depth profile.
+We were thus left with 7 species with data on $\delta^2H$ as well as modelled estimates of ERD.
 
@@
 To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD with maximum stem hydraulic conductance ($K_{max, stem}$), leaf turgor loss point ($\Psi_{tlp}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$).
-See [SI](#SIData) for data collection and estimation of these variables (RCT:"Brett, please add a section").
+See [SI](#SIData.brett) for data collection and estimation of these variables.
 
@@
 
-```{r hydro-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-hydro-calib}Observations (red lines) versus simulations (blue lines) from 100 best-fit parameter ensemble runs of ELM-FATES are shown for daily Volumetric Water Content (VWC) by depth for horizontal time domain reflectometry (TDR) probes at three depths (0.1, 0.4 and 1 m; A) and a vertical probe (0-0.15 m; B); monthly stream discharge (Discharge, C) and monthly evapotranspiration from eddy flux tower (ET, D).
+```{r hydro-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-hydro-calib}ELM-FATES calibration and evaluations.
+Observations (red lines) versus simulations (blue lines) from 100 best-fit parameter ensemble runs of ELM-FATES are shown for daily Volumetric Water Content (VWC) by depth for horizontal time domain reflectometry (TDR) probes at three depths (0.1, 0.4 and 1 m; A) and a vertical probe (0-0.15 m; B); monthly stream discharge (Discharge, C) and monthly evapotranspiration from eddy flux tower (ET, D).
 Points (red) in (A) show average plot-wide observations of VWC with 95% CI (error bars).
-VWC data from the horizontal probes (lines in A) were not used for ELM-FATES calibration.
-Values in inset are average standardized RMSE across the 100 simulations.
-RMSE was standardized to range 0-1 across the 5000 simulations.
+VWC data from the horizontal probes (lines in A) were not used for ELM-FATES calibration but evaluation.
+Values in inset are average RMSE across the 100 best-fit simulations.
 For panel A, values at the top are for plot-wide VWC and those at the bottom are for TDR probes.
@@
 \
-For the leaf vulnerability curve parameters $A$ & $B$ fitted to the observed data on $K_{leaf}$ vs.
-$\Psi_{leaf}$, $WSG$ & parameter $B$ explained a large proportion of variation in parameter $A$ (Eq.
+For the leaf vulnerability curve parameters $A$ and $B$ fitted to the observed data on $K_{leaf}$ vs.
+$\Psi_{leaf}$, $WSG$ and parameter $B$ explained a large proportion of variation in parameter $A$ (Eq.
 \@ref(eq:a-mod); *Adj.
@@
 S\@ref(fig:ab-fit)).
-$WSG$ & $LMA$ explained a large proportion of variation in parameter $B$ (Eq.
+$WSG$ and $LMA$ explained a large proportion of variation in parameter $B$ (Eq.
 \@ref(eq:b-mod); Adj.
@@
 \
-In comparison to the measured isotopic signal ($\delta^2H$), our results showed that the ERD model with a multiplicative effect of VPD and without Leaf Cover (Eq.
-\@ref(eq:M2)) had highest goodness-of-fit (*R*^2^ = 0.85; *P* = 0.008; Fig.
+In comparison to the measured isotopic signal ($\delta^2H$), our results showed that the ERD model with the effect of LAI and an additive effect of VPD (Eq.
+\@ref(eq:M2)) had highest goodness-of-fit (*R*^2^ = 0.65; *P* = 0.027; Fig.
 \@ref(fig:erd-iso)).
-All the other ERD models showed a positive but non-significant relationship with $\delta^2H$ (Fig.
+The model without LAI and an additive effect of VPD showed no species with sufficient species-level goodness-of-fit.
+All the other ERD models showed a positive but non-significant relationship with $\delta^2H$ at $\alpha$ = 0.05 (Fig.
 S\@ref(fig:other-erd-mods)).
-The model with an additive effect of VPD and those with Leaf Cover had only three-to-four species with sufficient species-level goodness-of-fit (Eqs.
-\@ref(eq:M3)-\@ref(eq:M6)) versus six for models without (Eqs.
-\@ref(eq:M1)-\@ref(eq:M2)).
-For the rest of the analyses we therefore used species ERD estimates from the model with a multiplicative effect of VPD, without Leaf Cover (Eq.
-\@ref(eq:M2)).
+For the rest of the analyses we therefore used species ERD estimates from the model with with the effect of LAI and an additive effect of VPD (Eq.
+\@ref(eq:M-chosen)).
 Notably, two species, *L.
@@
 ```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Matrix for pairwise correlations (circles) among Effective Rooting Depth (ERD) and hydraulic properties for ", nrow(cor.data)," canopy species found on Barro Colorado Island (Panama).
-Each circle is color coded by the Spearman's $\\rho$ with darker red or blue shades for increasngly positive or negative $\\rho$ values, respectively.
+Each circle is color coded by the Spearman's $\\rho$ with red shades for positive $\\rho$ values, while blue shades for negative values, respectively.
 P-values are given only when they are $\\ge$ 0.05.
@@
 We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins, that is, traits that are commonly identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through a study period of greater recurrence of drought.
-These results are consistent with the rare studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015].
+These results are consistent with the two studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015].
 @Johnson:2018fad found that mortality decreased with deep rooting even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi_{88,~stem}$) and had narrower safety-margins ($\Psi_{min} - \Psi_{88,~stem}$).
@@
 S\@ref(fig:obs-plc-curves)).
-Whether such scaling relationships between hard and soft traits hold true for other forests, and their wider applicability, remains to be seen as data on PLC curves are limited in tropical forests, but supportive and counter examples both exist [@McCulloh.2011; @Hoffmann.2011] (RCT: "K~leaf~ folks, Suggest if there are more references from the tropics, or relevant ones elsewhere.
-Also, could you comment on how rare this $K_{leaf}$ vs.
-$\Psi_{leaf}$ dataset is?").
+Whether such scaling relationships between hard and soft traits hold true for other forests, and their wider applicability, remains to be seen as data on PLC curves are limited in tropical forests, but supportive and counter examples both exist [@McCulloh.2011; @Hoffmann.2011].
 
-\
-We also note that the availability of data on $\delta^2H$ on trees and water sources below-ground--from soil or substrate pore water in the vadose zone and free water below the water table--as important for model calibration or validation.
+\ Deurwaerder.2020
+Another important piece of information is stable water isotopes of xylem water and below-ground sources, in the vadose zone and water table.
+Although the interpretation of these data is still under research (Deurwaerder et al., 2020), they are critical for model calibration or validation.
 The model in @Chitra-Tarak.2018 were not corroborated with such data as existing methods were inadequate to extract water bound to the saprolite found in their dry forest study site (Jean Riotte, pers.
@@
 @Chitra-Tarak.2018 was calibrated only on stream discharge and water table levels.
-We lacked data on water table levels which would have been useful to constrain water availability dynamics in the unobserved soil layers.
+We lacked data on water table levels which would have been useful to constrain water availability dynamics in the unobserved soil layers and the spatial variability within the plot.
 As in @Chitra-Tarak.2018, we also interpret our ERD estimates as revealing relative differences among species' effective rooting depths rather than in absolute terms.
@@
 S\@ref(fig:other-erd-mods)).
-We interpret this result as that VPD and PLC curves may be adequate to explain growth, but also acknowledge that our estimates of seasonality of leaf cover is a tentative estimate combining leaf fall data and assumption of leaf presence during the wet season, and omits inter-annual variation.
+We interpret this result as that VPD and PLC curves may be adequate to explain growth, but also acknowledge that our estimates of seasonality of LAI is a tentative estimate combining leaf fall data and assumption of leaf presence during the wet season, and omits inter-annual variation.
 Future studies could make use of data from new technologies such as drone based monitoring of species-specific leaf area fraction in species-rich forests [@Park.2019].
@@
 An NSF grant to Smithsonian Institution’s ForestGEO (Dimensions 1046113) also partially supported this work via RCT and SMM.
+RF was supported by the National Center for Atmospheric Research, which is funded by the National Science Foundation.
 **Add acknowledgements for other co-authors.
-We acknowledge the following for contributions to the 50-ha plot project: R.
-Foster as plot founder and the first botanist able to identify so many trees in a diverse forest; R.
-Pérez and S.
-Aguilar for species identification; S.
-Lao for data management; S.
-Dolins for database design; plus hundreds of field workers for the census work, now over 2 million tree measurements; the National Science Foundation, Smithsonian Tropical Research Institute, and MacArthur Foundation for the bulk of the financial support.
+The BCI forest dynamics research project, founded by Stephen P.
+Hubbell and Robin B.
+Foster and sustained for many years by Richard Condit, is now managed by Stuart Davies, Suzanne Lao, and Rolando Perez under the ForestGEO program of the Smithsonian Tropical Research in Panama.
+Numerous organizations have provided funding, principally the US National Science Foundation, and hundreds of field workers have contributed.
 
@@
 
-## Data Descriptions {#SIData}
+## Data Descriptions
 
-### Microclimatic tower
+### Microclimatic tower {#SIData.matteo}
 
@@
 Ensemble was obtained as weighted average from the 100 ANN predictions using the MSE of the test set as weights according to:
-f_i=(∑_k▒f_ik^ANN /MSE_ik)/(∑_k▒〖1/MSE_ik 〗)
+
+$f_{i} = \frac{\sum_{k}f_{ik}^{ANN}/{MSE_{ik}}}{\sum_{k}1/{MSE_{ik}}}$
 
@@
 
-### Leaf maximum hydraulic conductivity and vulnerability to cavitation
+### Volumetric Water Content
 
-The Kleaf was measured under steady state and high irradiance following the Evaporative Flux Method (FEM) [@Sack:2012]  using a flowmeter.
+We collected volumetric water content (VWC) data (2012-2018) from a point location at the eddy flux tower with a vertical time domain reflectometry (TDR) probe.
+The relationship between the output from the TDR probe and VWC was calibrated based on measurements of gravimetric water content (GWC) on several occasions from representative soils near the vertical probe (0-0.15 m) (Fig.
+S\@ref(fig:TDR-calib)).
+GWC was converted to VWC based on the observed mean bulk dry density of 0.8 g cm^-3^ near the TDR probe (*n* = 5, cylinder volume = 1250 cm^3^).
+Beginning in 2016 we additionally installed three horizontal probes at depths of 0.15, 0.4 and 1 m and we applied the same calibration equation to these data (2016-2018) as that for the vertical probe.
+
+We also leveraged exisiting GWC measurements by @Kupers:2019 at the depths of 0.15, 0.4 and 1 m in summer 2015 (February, March and April) and 0.15, 0.4 in summer 2016 (March)--1299 samples that covered all soil types and habitats in the plot.
+The sampling periods were 6, 5, 10 and 8 days long, respectively.
+The 2016 dry season was associated with the 2015–2016 El Niño (see @Kupers:2019 for further details).
+We converted these GWC measurements to VWC assuming a bulk density of 0.8 g cm^-3^.
+$\theta_{0.15 m}$, $\theta_{0.4 m}$ and $\theta_{1 m}$ refer to the plot-wide VWC averages by depth over each sampling period.
+Middle date of a sampling period is used for comparison with model simulations.
+
+### Leaf maximum hydraulic conductivity and vulnerability to cavitation {#SIData.joseph}
+
+The Kleaf was measured under steady state and high irradiance following the Evaporative Flux Method (FEM) [@Sack.2011]  using a flowmeter.
 Briefly, the leaf is excised under filtered DI water and attached to a water source (graduated cylinder) using clear plastic tubing.
@@
 
-### Stem maximum hydraulic conductivity and vulnerability to cavitation
+### Stem maximum hydraulic conductivity and vulnerability to cavitation {#SIData.brett}
 
@@
 
-### Data gaps in WSG and LMA
+### Data gaps in WSG and LMA {#SIData.rutuja}
 
@@
 *R*^2^ = `r chave.r2`; *P* `r chave.p`, *n* = `r chave.n`).
+
+## Alternative models for Effective Rooting Depth (ERD)
+
+Below we present all the altervative models for Effective Rooting Depth (ERD).
+Daily average species-specific growth, G~t | z~, in the census interval *t* given an ERD of depth *z*, is described as follows:
+
+\begin{equation}
+G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,~i~|~z}^{*}) + \epsilon_{t~|~z}
+(\#eq:M1)
+\end{equation}
+
+\begin{equation}
+G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,~i~|~z}^{*} VPD_{i}^{*}) + \epsilon_{t~|~z}
+(\#eq:M2)
+\end{equation}
+
+\begin{equation}
+G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,~i~|~z}^{*} + VPD_{i}^{*}) + \epsilon_{t~|~z}
+(\#eq:M3)
+\end{equation}
+
+\begin{equation}
+G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,~i~|~z}^{*} LAI_{doy~|~i}^{*}) + \epsilon_{t~|~z}
+(\#eq:M4)
+\end{equation}
+
+\begin{equation}
+G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} (K^{*}_{leaf,~i~|~z} + VPD^{*}_{i}) LAI^{*}_{doy~|~i}) + \epsilon_{t~|~z}
+(\#eq:M5)
+\end{equation}
+
+\begin{equation}
+G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,~i~|~z}^{*}VPD_{i}^{*}LAI_{doy~|~i}^{*}) + \epsilon_{t~|~z}
+(\#eq:M6)
+\end{equation}
+where n~t~ be the total number of days in census interval *t*, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is a normally distributed model error term.
 
@@
 
-```{r leaf-cover, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-cover}Seasonal variation in Leaf Cover Fraction for ", length(erd.sp)," species at Barro Colorado Island (panels).
+```{r leaf-cover, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-cover}Seasonal variation in standardised LAI (range 0-1) for ", length(erd.sp)," species at Barro Colorado Island (panels).
 Species are color-coded by deciduousness.")}

### 2020-10-09 manuscript.rmd
indexed an equation
stream discharge details

@@
 We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq.
-\@ref(eq:water-balance), namely, (i) evapotranspiration ($ET \simeq E_{v} + E_{g}$) from the eddy flux tower (2012-2017; see [SI](#SIData.matteo)), (ii) stream discharge for the Conrad Catchment ($q_{run} \simeq q_{over} + q_{drai}$; 2012-2018) and (iii) soil volumetric water content (VWC, $\theta_{z} \simeq W_{liq, z}$) from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we built with a vertical TDR probe over 0-15 cm depth ($\theta_{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m (1299 samples in total covering all soil types and habitats; see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\theta_{0.15 m}$, $\theta_{0.4m}$ and $\theta_{1 m}$ refer to plot-wide averages).
+\@ref(eq:water-balance), namely, (i) evapotranspiration ($ET \simeq E_{v} + E_{g}$) from the eddy flux tower (2012-2017; see [SI](#SIData.matteo)), (ii) stream discharge from the Conrad catchment ($q_{run} \simeq q_{over} + q_{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC, $\theta_{z} \simeq W_{liq, z}$) from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we built with a vertical TDR probe over 0-15 cm depth ($\theta_{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m (1299 samples in total covering all soil types and habitats; see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\theta_{0.15 m}$, $\theta_{0.4m}$ and $\theta_{1 m}$ refer to plot-wide averages).
 
@@
 \begin{equation}
-K_{leaf,~s} = A_{s}e^{-B_{s}\Psi_{leaf}} + error
+K_{leaf,~s} = A_{s}e^{-B_{s}\Psi_{leaf_{s}}} + error
 (\#eq:vc-exp)
@@
 
+### Stream discharge from the Conrad catchment (#SIData.steveP)
+
+The Conrad weir is located on the gently sloping western side of Barro Colorado Island.
+The weir consists of a 90 degree ‘V’ notch with a rectangular upper section.
+The weir was constructed in several stages between 1993 and 1996.
+Stream stage is recorded at 5-minute intervals.
+
+The weir drains a catchment of approximately 40 ha of which at least 90% consists of the central plateau area of the island.
+Peak flow usually follows peak rain by approximately 2-3 hours.
+Large storms are sometimes preceded by small, initial peaks generated by the rain falling on the steeper slopes close to the weir.
+
+Streamflow data are analyzed by software custom-built by author SP.
+The program is used to correct for blockages of the weir ‘V’ by fallen vegetation (a common occurrence), sensor drift, small gaps and other problems.
+Discharge is then calculated using an empirically derived rating curve calculated for each 1mm stream stage.
+
 ### Volumetric Water Content
@@
 Red line shows a fit from a polynomial model described in inset, and used to standardize VPD employed in equations \\@ref(eq:M2) and \\@ref(eq:M4)."}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/gpp_by_env_variables/monthly_gpp.vpd.jpeg"))
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/gpp_by_env_variables/daily_gpp.vpd.jpeg"))
 ```

### 2020-10-10 manuscript.rmd
growth vs. dbh relationship without color coding by census interval
1. Added a schematic fig. for the methods workflow 2. Updated extreme drought year fig. description

@@
 \
-In summary, we obtained daily dynamics of plant water potential by depth in the whole-soil column with a locally parameterized and calibrated 1-D hydrological water balance model.
+In summary (Fig.
+\@ref(fig:schematic)), we obtained daily dynamics of plant water potential by depth in the whole-soil column with a locally parameterized and calibrated 1-D hydrological water balance model.
 We developed a novel inverse model to estimate species effective rooting depths (ERD), defined as the depth at which soil water dynamics best explained observed dynamics of species growth.
@@
 We then tested whether deeper ERD was associated with hydraulically efficient but vulnerable traits (Table 1 ), but not greater mortality rates.
+
+```{r schematic, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-schematic}A schematic for the methods workflow.")}
+knitr::include_graphics(file.path("figures/schematic.png"))
+```
 
@@
 *R*^2^).")}
-knitr::include_graphics(file.path("figures/dbh/growthrate_dbh_predicted_growth_with_spline_degree_5_dbh_cutoff_100_model.jpeg"))
+knitr::include_graphics(file.path("figures/dbh/growthrate_dbh_predicted_growth_with_spline_degree_5_dbh_cutoff_100_model_no_color.jpeg"))
 ```
@@
 
-```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts}Extreme soil droughts by census interval (panels) at three representative depths: 0.5, 1 and 1.7 m, depicted in different colors.
-Mean soil water potential ($Psi_{soil,~z}$) for a given DOY and depth z are shown as lines.
-Filled areas show extreme drought periods, when $Psi_{soil,~z}$ declined below 2.5 percentile of $Psi_{soil,~z}$ for the given DOY and depth z in any given year within the census intervals.
-Extreme drought periods from multiple years may be overlaid within a panel.
+```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts}Occurrence of extreme soil droughts by census intervals (horizontal panels) at three representative depths (z): 0.1, 0.6 and 1 m (vertical panels).
+Mean (black lines) and lower half of 95% distribution (grey areas) of soil water potential ($Psi_{soil,~z}$) for a given DOY and depth z are shown (same across census intervals).
+$Psi_{soil,~z}$ by DOY for a year (colored line) is only shown if for some part of that year $Psi_{soil,~z}$ declined below the 2.5 percentile distribution.
 Wet season is shown in gray beginning with DOY 120."}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q2.5_depths_in_inverse_model.jpeg"))
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q2.5_by_dpeth.jpeg"))
 ```

### 2020-10-16 manuscript.rmd
1. Several instances of including co-authors suugestions 2. Moved main text Fig and table captions to the end of main text 3. Moved significant droughts figure to the main text 4. Moved goodness-of fit figures for obs vs. predict A and B fits to the main text

@@
 
-- Among the mechanisms that plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied.
-Vulnerability to embolism and hydraulic safety margins are shown to be predictors of drought-induced mortality risk, but deep-water access may delay plants from reaching these critical thresholds.
-Here, we tested the role of deep-water access in survival strategies of a diverse tropical forest community (Panama) using a novel data-model approach.
-- We inversely modeled effective rooting depth (ERD) for large trees of `r length(erd.sp)` canopy species by linking growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and a species-specific leaf vulnerability curves.
-We evaluated ERD against existing isotopic tracer data on trees and soil.
-- We find that deeper ERD is associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates over 35 years.
-Deeper ERD species spent less time below critical hydraulic thresholds indicating that trees mitigate water-stress related mortality risk through deep water access.
-- The role of deep-water access in alleviating mortality of drought-sensitive trees has important implications for our predictive understanding of forest dynamics under current and future climate.
+- Among the mechanisms that plants employ to tolerate drought, deep water access is arguably the most commonly invoked, but under-studied.
+Vulnerability to embolism and hydraulic safety margins are shown to be predictors of drought-induced mortality risk, but deep-water access may delay plants from reaching critical safety thresholds.
+Here, we tested the role of deep-water access in enabling survival within a diverse tropical forest community (Panama) using a novel data-model approach.
+- We inversely modeled effective rooting depth (ERD, as the average depth of water extraction), for large trees of `r length(erd.sp)` canopy species by linking growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and to leaf hydraulic vulnerability curves.
+We validated ERD against existing tree and soil stable water isotope data.
+- Across species, deeper ERD was associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates over 35 years.
+Species with deeper ERD spent less time below critical hydraulic thresholds indicating that trees compensate for water-stress related mortality risk through deeper water access.
+- The role of deep-water access in alleviating mortality of hydraulically-vulnerable trees has important implications for our predictive understanding of forest dynamics under current and future climates.
 
@@
 \
-Tropical forests play a disproportionately large role in the global carbon and energy cycle [@Bonan:2008], and support half of global biodiversity [@Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019].
-Tropical forests are logically considered to be an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw], yet mortality events in the tropics are rarely as large as those in the temperate and boreal zones (McDowell et al.
-2018), leading to questions regarding the role of large trait-diversity in mitigating mortality events.
-Despite this, widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010] and tropical mortality rates are increasing in some regions [@Brienen.2015; @Hubau.2020].
-However, state-of-the-art dynamic vegetation models (DVMs) struggle to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013; @Powell.2018], as underlying mechanisms of drought-tolerance are not fully understood or quantified.
+    Tropical forests play a disproportionately large role in the global carbon and energy cycles [@Bonan:2008], and support half of global biodiversity [@Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019].
+Tropical forests are logically considered to be an especially drought-vulnerable among biomes given the combination of climate risk and vegetation sensitivity [@Meir:2015trw], yet mortality events in the tropics are rarely as large as those in the temperate and boreal zones (McDowell et al.
+2018), leading to questions regarding the role of large trait-diversity and hydraulic strategies in mitigating mortality events.
+Despite this, widespread drought-induced tree mortality has occurred across the tropics for specific functional groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010] and tropical mortality rates are increasing in some regions [@Brienen.2015; @Hubau.2020].
+However, state-of-the-art dynamic global vegetation models (DGVMs) struggle to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013; @Powell.2018], as underlying mechanisms of drought-tolerance are not fully understood or quantified.
 
 \
-Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied.
-This mechanism assumes that water availability varies with soil depth, and that the rooting depths of co-located trees also vary, by species and size [@Meinzer:1999; @Chitra-Tarak.2018; @Brum.2019].
-Consequently, trees in the same forest differ in their drought experience, and thus in their growth and mortality responses [@Chitra-Tarak.2018].
-A key bottleneck in community-wide testing of this mechanism has been lack of data on trees' rooting or water-sourcing depths, followed by data on water-availability at those depths.
+    Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation to stem water storage, deep water access, increased root production, hydraulic redistribution and embolism resistance.
+Deep water access is arguably the most long-standing, yet under-studied mechanism.
+This mechanism assumes that plant available water varies with depth, and that the rooting depths of co-located trees also vary, by species and size [@Meinzer:1999; @Chitra-Tarak.2018; @Brum.2019].
+Consequently, trees in the same forest may differ in their experience during a drought, and thus in their growth and mortality responses [@Chitra-Tarak.2018].
+A key bottleneck in community-wide testing of this mechanism has been a lack of data in both trees' rooting or water-sourcing depths and plant available soil water at those depths.
 A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the world’s more than 60 thousand tree species [@Fan:2017; @Beech:2017].
@@
 Use of stable isotopes of water as a tracer provides an indirect measure of water sourcing depths by matching the isotopic value in xylem water to variations in soil pore water with depth.
-However, such data are rare [@Evaristo:2016].
+However, such data are rare [@Evaristo:2016] and their interpretation is an active area of research given unresolved issues pertaining to soil heterogeneity and the correspondence of soil water extract in the laboratory to actual plant available water [@Adams.2020; @Bowers.2020; @Deurwaerder.2020].
+DNA barcoding of roots [@Jones.2011] may be used to to estimate species-specific rooting depths or rooting profiles, although extent of sampling required may be cost-prohibitive and DNA barcode libraries for tropical forests are still under development.
 In general, community-scale data collection for rooting or water-sourcing depths is challenging for species-rich tropical forests.
@@
 \
-Trade-offs in plant 'traits' shape how plants make a living in their environment (ecological strategies) and also provide the essential constraints to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010; @Bartlett.2019].
-Understanding how plant trade-offs lead to alternative life history strategies entails identifying the topography of trait trade-offs in different environments, and how these relate to demographic rates (growth, recruitment and mortality).
-Specifically for the water environment (or, resource), observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified an acquisitive-conservative resource economic spectrum similar to those for light and nutrients, although, for water, examples of all trade-offs observed simultaneously are rare, and below-ground traits or environment linkages are weak [@Reich:2014; @Sterck:2011].
-Acquisitive species have traits with high capacity to acquire, move, store and use water (higher hydraulic conductivity, low wood density) and achieve high growth, but build flimsy tissues (high specific leaf area) and are less tolerant of low resource availablity, whereas species with conservative traits are better protected from carbon losses and are more physiologically drought tolerant (greater embolism resistance, larger safety margins).
-Whether and how the acquisitive-conservative spectrum holds true for the water dynamics vertically below-ground is not clear.
-Acquisitive species are spatially associated with wet sites (high resource) such as valleys or those with high rainfall [@Kursar:2009; @Oliveira:2019].
-Deep rooting may be expected to support an acquisitive strategy by enabling a steady water supply.
-However, the degree to which deep rooting is associated with an acquisitive strategy may depend on the specific dynamics of water availability by depth at a site, which is affected by precipitation regime and hydrology.
-For example, in pulse-driven systems *shallow*-rooted plants are associated with acquisitive traits (higher hydraulic conductivity) and opportunistic strategies (e.g.
-dry deciduousness) that maximise the use of the water-pulse, while *deep*-rooted plants show conservative traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013 and references therein].
+    Trade-offs in plant 'traits' shape how plants make a living in their environment leading to alternative ecological strategies.
+Characterizing the essential constraints necessary to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010; @Bartlett.2019] entails identifying the topography of trait trade-offs in different environments, and how these relate to demographic rates (growth, recruitment and mortality).
+Significant efforts have been invested into identifying and linking universal drought indices, above-ground traits, and demographic rates.
+These efforts have resulted in identification of correlations between aboveground hydraulic traits such as vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013, @Venturas.2016].
+Deep roots may mitigate hydraulic vulnerability and mortality-risk, yet plants' realized water environments, specifically their ability to acquire water via differing rooting depths, are rarely studied, especially in the tropical region.
+We tested how rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tree species of a tropical forest over 35 years that witnessed droughts of various intensity and duration.
 
 \
-Significant efforts have been invested into identifying and linking universal drought indices, above-ground traits, and demographic rates.
-These efforts have resulted in identification of above-ground traits such as proxies of hydraulic failure as indicators of mortality risk.
-While correlations between aboveground hydraulic traits and mortality have been identified [@Anderegg:2016], numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013].
-However, plants' realized water environments, specifically their ability to acquire water via differing rooting depths, are rarely studied, especially in the tropical region.
+    We used a locally parameterized and calibrated 1-D hydrological water balance model to obtain daily dynamics of soil water potential by depth in the whole-soil column and then applied a novel inverse model to estimate species effective rooting depths (ERD), defined as the depth at which soil water potential best explained observed dynamics of species-specific growth for large trees of `r length(erd.sp)` dominant species (Fig.
+\@ref(fig:schematic)).
+This model incorporated the impact of atmospheric drought on growth and was constrained by above-ground hydraulic traits, namely, species-specific leaf vulnerability curves, by leveraging existing data for 8 species and developing trait-based proxies for the rest.
+We evaluated ERD against existing data on plant and soil water isotopes and tested coordination of ERD with hydraulic traits and mortality rates.
+Specifically, we tested an overarching hypothesis that inter-specific variation in trees' effective rooting depths (ERD) enables increasingly reliable water availability with depth and thus mitigation of hydraulic vulnerability.
+Our hypotheses are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallow rooted species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deep rooted species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallow rooted species may have higher mortality rates as a result of higher variability water availability in shallow soil layers.
+To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.
+
+Fig. \@ref(fig:schematic) here.
+Table \@ref(tab:hypo) here.
+
+# Materials and Methods
+## Study site description
+\
+We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and data on stable water isotopes on tree species and soil from an independent study of tree water-sourcing depths [@Meinzer:1999].
+The 1,560ha island was formed after the river Chagres was dammed in 1914 to create Gatun Lake and complete the Panama Canal.
+The entire island is forested and classified as a tropical moist forest in the Holdridge Life Zone system.
+Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014].
+Rainfall at BCI is seasonal with a mean annual rainfall of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a four-month long dry season from mid-December through mid-April ($\le$ 100 mm rain month^-1^).
+Among the tree *species* found in the forest canopy, 28% are identified to be at least partly deciduous during the dry season, but this amounts to a maximum of 6.3% *trees* in the canopy that are deciduous, and the percentage of total forest crown area that is deciduous is estimated at 9.7% [@Condit:2000].
 
 \
-To better understand how rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tropical forest tree species, in this paper, we conduct a thorough assessment for a tropical forest of relationships between the water environment in the whole-soil column, above-ground hydraulic traits, estimates of rooting depths and demographic outcomes over 35 years that witnessed droughts of various intensity and duration.
-Specifically, we test an overarching hypothesis that inter-specific variation in trees' rooting depths aligns species on an acquisitive-conservative economic spectrum, whereby acquisitive trees have rooting depths optimized for more reliable water availability by depth.
-Our hypotheses are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallow rooted species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deep rooted species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallow rooted species may have higher mortality rates as a result of higher variability in shallow soil water availability.
-To our knowledge this is the first study to demonstrate a mechanistic link between trees' realized drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.
-
-\
-```{r hypo, results = 'asis', echo = FALSE}
-knitr::kable(hypo.table, caption = 'Hypotheses for association between Effective Rooting Depth (ERD) and hydraulic traits.', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(hypo.table))))
-```
-
-# Materials and Methods
-\
-In the following sections we describe the study site, the models and data used to estimate and evaluate (1) soil water potential dynamics by depth in the whole-soil column, (2) species-specific leaf vulnerability curves, and (3) species-specific effective rooting depths.
-These are followed by analyses to address the proposed hypotheses.
-Descriptions of datasets used and details of some of the methods are described in the Supplementary Information.
-
-\
-In summary (Fig.
-\@ref(fig:schematic)), we obtained daily dynamics of plant water potential by depth in the whole-soil column with a locally parameterized and calibrated 1-D hydrological water balance model.
-We developed a novel inverse model to estimate species effective rooting depths (ERD), defined as the depth at which soil water dynamics best explained observed dynamics of species growth.
-This model incorporated the impact of atmospheric drought on growth and was constrained by above-ground hydraulic traits, namely, species leaf vulnerability curves.
-For the latter, we collected new data on vulnerability curves for 21 species and scale-up to the community via scaling relationships with 'softer' traits, namely, Wood Density and Leaf Mass Area.
-We evaluated ERD against independent estimates based on existing data on plant and soil water isotopes.
-We then tested whether deeper ERD was associated with hydraulically efficient but vulnerable traits (Table 1 ), but not greater mortality rates.
-
-```{r schematic, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-schematic}A schematic for the methods workflow.")}
-knitr::include_graphics(file.path("figures/schematic.png"))
-```
-
-## Study Site Description
-\
-We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and data on stable water isotopes on tree species and soil from an independent study on trees’ water-sourcing depths [@Meinzer:1999].
-The 1,560ha island was formed after the river Chagres was dammed in 1914 to create Gatun Lake and complete the Panama Canal.
-Rainfall at BCI is seasonal with a mean annual rainfall of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a four-month long dry season from mid-December through mid-April ($\le$ 100 mm rain month^-1^).
-The entire island is forested and classified as a tropical moist forest in the Holdridge Life Zone system.
-Although 28% of canopy tree species are at least partly deciduous during the dry season, this amounts to the maximum percentage of deciduous individuals in the canopy this amounts to is only 6.3%, and the percentage of crown area that is deciduous is estimated at 9.7% [@Condit:2000].
-Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014].
-
-\
-In the 50 ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm size class [@Condit.2017].
+In the 50 ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm dbh size class [@Condit.2017].
 This inventory represents 321 woody species across the eight censuses.
 Aside from a 2ha patch of ~100-year-old secondary forest at the north-eastern corner of the plot the remaining 48ha represent old growth forest that was never clear cut for agriculture.
-The plot (1000 m x 500 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001].
+The plot (1000 m $\times$ 500 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001].
 The plot is situated in the Conrad catchment.
@@
 
-## Model ELM-FATES for Soil Matric Potentials
+## Model ELM-FATES for soil matric potentials
 
@@
 numbering to begin at 1]).
-For all runs, instead of ELM's default soil texture-based soil water retention curve, we used a curve that we derived using existing data on gravimetric water content vs.
+For greater accuracy, instead of the ELM default soil texture-based pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content vs.
 $\Psi_{soil}$ for depths of 0.15, 0.4 and 1 m [@Kupers:2019].
@@
 \end{equation}
-where $\Psi_e$ is the air-entry matric potential (200, mm H~2~O), $\theta_{s}$ is the saturated volumetric water content (0.51, cm^3^ cm^-3^), and $b$, an index for soil pore-size distribution (10, unitless), respectively.
+where $\Psi_e$ is the air-entry matric potential, $\theta_{s}$ is the saturated volumetric water content, and $b$, an index for soil pore-size distribution, respectively.
+$\Psi_e$, $\theta_{s}$ and $b$ taht best-fit the data were 200 (mm H~2~O), 0.51 (cm^3^ cm^-3^) and 10 (unitless), respectively.
 
@@
 We ran ELM-FATES for the 5000-member parameter ensemble with hourly climate drivers from 1985-2018 [@Faybishenko.2020] initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode.
-As our key interest here was in soil water availability rather than forest dynamics, we ran ELM-FATES in a lower-complexity configuration: static stand structure, and with a single plant functional type (PFT) of evergreen trees.
+As our key interest here was in deriving soil water availability, we ran ELM-FATES in a lower-complexity configuration: static stand structure, and with a single plant functional type (PFT) of evergreen trees.
 The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of a dry-deciduous PFT did not significantly alter results (not shown).
@@
 
+### Model calibration
 \
 We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq.
-\@ref(eq:water-balance), namely, (i) evapotranspiration ($ET \simeq E_{v} + E_{g}$) from the eddy flux tower (2012-2017; see [SI](#SIData.matteo)), (ii) stream discharge from the Conrad catchment ($q_{run} \simeq q_{over} + q_{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC, $\theta_{z} \simeq W_{liq, z}$) from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we built with a vertical TDR probe over 0-15 cm depth ($\theta_{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m (1299 samples in total covering all soil types and habitats; see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\theta_{0.15 m}$, $\theta_{0.4m}$ and $\theta_{1 m}$ refer to plot-wide averages).
+\@ref(eq:water-balance), namely, (i) evapotranspiration ($ET \simeq E_{v} + E_{g}$) from the eddy flux tower (2012-2017; see [SI](#SIData.matteo)), (ii) stream discharge from the Conrad catchment ($q_{run} \simeq q_{over} + q_{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC, $\theta_{z} \simeq W_{liq, z}$) from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we obtained with a vertical TDR probe over 0-15 cm depth ($\theta_{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m (1299 samples in total covering all soil types and habitats; see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\theta_{0.15 m}$, $\theta_{0.4m}$ and $\theta_{1 m}$ refer to plot-wide averages).
 
@@
 based on monthly sums for both $ET$ and $q_{run}$ and daily averages for $\theta$s.
-$W_{liq,~\overline{z}}$ is the daily average of $W_{liq,~z}$ for $z$ $\in$ {`r paste(soil.depths[1:4], collapse=", ")`} m.
+$W_{liq,~\overline{z}}$ is the average of daily $W_{liq,~z}$ for $z$ $\in$ {`r paste(soil.depths[1:4], collapse=", ")`} m.
 Superscript * indicates that RMSE is standardised between 0 and 1 across the 5000 simulations.
@@
 Preliminary analyses showed that minimising the objective function $\overline{O}$, that is, simultaneously fitting the model to all states and fluxes did not capture the soil moisture dynamics at 1 m depth well, our key priority for model calibration.
-To ensure model fit with soil moisture dynamics at 1 m depth, we first chose 300 parameter ensembles that minimised RMSE for VWC at the depth of 1m ($RMSE^{*}(\theta_{1 m}, W_{liq,~`r soil.depths[8]`m})$ < 0.2), and among those 300 ensemble members we chose 100 members with the lowest values for the objective function $\overline{O}$.
+To ensure model fit with soil moisture dynamics at 1 m depth, we first chose 300 parameter ensemble members that minimised RMSE between observations and simulations for VWC at the depth of 1m ($RMSE^{*}(\theta_{1 m}, W_{liq,~`r soil.depths[8]`m})$ < 0.2), and among those 300 ensemble members we chose 100 members with the lowest values for the objective function $\overline{O}$.
 See Table \@ref(tab:param) for the ranges of best-fit values for different parameters.
@@
 
-## Model for Effective Rooting Depth (ERD)
+## Model for effective rooting depth (ERD)
 \
-Water is a key limiting resource for tree growth.
-Vapor pressure deficit (VPD) and water supply as determined by soil moisture are factors known to influence water uptake [@Lawrence.2019].
-In this study, we used leaf conductivity ($K_{leaf}$) determined by rhizosphere soil water potential ($\Psi_{soil}$) as an approximation of the soil water limiting factor.
-For dry-deciduous species seasonality in leaf cover may also influence plant water uptake.
-We therefore tested alternate models of growth, aka inverse models for ERD, in which a species' growth dynamics across census intervals was regressed against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1 (see below).
-For the chosen model (see Model Evaluation below), daily average growth for species $s$, $G_{s,~t~|~z}$, in the census interval $t$ for depth $z$ and for a given hydrauligoical realisation $h$ is described as follows (see SI for equations for all the tested models):
+To estimate effective rooting depth, we tested alternate models for growth (Eqs.
+\@ref(eq:M-chosen) and those in SI), which use effective rooting depth as an input.
+There are many factors that could limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019].
+In this study, we used maximum diurnal leaf conductance ($K_{leaf}$) approximated using rhizosphere soil water potential ($\Psi_{soil}$) as a proxy for the soil water limiting factor on growth.
+In our models $K_{leaf}$ therefore incorporates the effects of rooting depth and rhizosphere water potential.
+For each species' growth dynamics across census intervals was regressed against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1.
+Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting.
+At our study site, due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season, the impact of VPD may as well be a combined impact of not only temperature and humidity, but also radiation.
+
+For the chosen model (see Model validation and selection below), daily average growth for species $s$, $G_{s,~t~|~z}$, in the census interval $t$ for depth $z$ and for a given hydrological realisation $h$ is described as follows:
 
@@
 where n~t~ be the total number of days in census interval *t*, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is a normally distributed model error term.
-We estimated daily value of $K_{leaf,~s,~i~|~z,~h}$ for a species $s$ by substituting $\Psi_{soil,~i~|~z,~h}$ in the leaf vulnerability curves defined below in Eq.
+See SI for equations for all the alternate models tested.
+
+As VPD may have a non-linear relationship with growth, with growth increasing for the lower range of VPD, and decreasing for the higher range of VPD due to stomatal closure (reference), we used a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see Model parametarisation) to represent the impact of VPD on growth and used GPP standardised between 0 and 1 (hereafter, VPD*) in Eq.
+\@ref(eq:M-chosen).
+
+For each species, we estimated diurnal maximum value of $K_{leaf,~s,~i~|~z,~h}$ for a species $s$ by substituting $\Psi_{soil,~i~|~z,~h}$ in the leaf vulnerability curves defined below in Eq.
 \@ref(eq:vc-exp) using species-specific parameters $A_{s}$ and $B_{s}$ as:
@@
 
-We estimated species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015).
+By varying ‘z’, we estimated species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015).
 Using Eq.
-\@ref(eq:M-chosen), for a species $s$, we obtained Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $\rho$ for each soil depth $z$ for each of the 100 best-fit hydrological realizations.
-For each hydrological realization we retained only one depth that corresponded to the least AIC score (the best-fit), and within those depths retained the ones that had Pearson's $r$ >= 0.71 (or, $r^2$ >= 0.5) as a valid set.
+\@ref(eq:M-chosen), for a species $s$, we obtained Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $r$ for each soil depth $z$ for each of the 100 best-fit hydrological realizations.
+For each hydrological realization we retained only one depth that corresponded to the least AIC score (the best-fit) [@Chonggang, we could have retained multiple depths with non-differentiable AICs], and within those depths retained the ones that had Pearson's $r$ >= 0.71 (or, $r^2$ >= 0.5) as a valid set.
 For a species $s$, we defined effective rooting depth (ERD) as the median of these set of depths, described as :
@@
 \
-For growth estimates, to minimize the effect of light variation among trees, we selected only large trees ($\ge$ 30 cm dbh, the diameter at breast height) and species whose maximum height was $\ge$ 30 m (hereafter, canopy species) and thus are likely to be fully exposed to the sun.
+For growth estimates, to minimize the effect of light variation among trees, we selected only large trees ($\ge$ 30 cm dbh, the diameter at breast height) and also species whose maximum height was $\ge$ 30 m (hereafter, canopy species) and thus are likely to be fully exposed to the sun.
 We selected trees for which the height of the diameter measurements were unchanged across all censuses (1990, 1995, 2000, 2005, 2010, 2015).
@@
 
-\
-There are many factors that could affect plant growth, including soil moisture, VPD, leaf area, temperature and radiation.
-Based on our preliminary analyses, incorporation of temperature and radiation did not improve our model fitting, thus we mainly focused on the impact of soil moisture, VPD and leaf area in our models.
-
-#### Leaf Vulnerability Curves
+#### Leaf hydraulic vulnerability curves
 \
@@
 $\Psi_{leaf}$; see Table \@ref(tab:symbols) for trait symbols and their definitions) for adult trees of 21 common species from Zailaa et al.
-(to be submitted; see SI for brief description of methods).
-
-We chose the exponential function without the intercept to fit observed data across species, and as a general function to scale-up to the community, because it had the lowest AICc for most species, captured the form of the observed data across all species well, and had the lowest number of parameters, that is, two:
+(to be submitted; see SI for brief description of methods) described as:
 
@@
 where $A$ and $B$ are fitted species-specific parameters.
-From these fits we estimated species $K_{max,~leaf}$, $\Psi_{50,~leaf}$ and  $\Psi_{80,~leaf}$ (see Table 1).
+These 21 species included `r sp.n.data` of the `r length(erd.sp)` species selected for ERD estimation.
+For the rest `r sp.n.model` species, we obtained vulnerability curves using trait-based proxies.
 We identified scaling relationships between fitted parameters $A$ and $B$ in Eq.
-\@ref(eq:vc-exp) and two traits that were available at the community-scale, namely,  WSG and LMA (@Wright:2010; see SI for description of how data gaps were filled.
+\@ref(eq:vc-exp) and two traits, namely, WSG, the wood specific gravity after drying at 100^$\\circ$^C and LMA, the leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules for leaves receiving direct sunlight (@Wright:2010; see SI for details).
 We fitted polynomial equations described as,
@@
 
-As these fits explained large proportion of variation in parameters $A$ and $B$ (see results), we sequentially used Eq.
-\@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r length(unique(bci.AB$sp[!is.na(bci.AB$A)]))` species for which WSG and LMA were both available, retained `r length(unique(data.model.AB$sp))` for which parameters $A$ and $B$ were both positive, and fitted vulnerability curves to those (Eq.
-\#ref(eq:vc-exp)).
+As these fits explained a large proportion of variation in parameters $A$ and $B$ (see results), we sequentially used Eq.
+\@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r sp.n.model` ERD species without direct data and estimated leaf hydraulic vulnerability curves using Eq.
+\#ref(eq:vc-exp).
+For all the 31 ERD species we estimated species $\Psi_{50,~leaf}$ based on their vulnerability curves (see Table \@ref(tab:symbols) for definition).
 
-\
-```{r symbols, results = 'asis', echo = FALSE}
-knitr::kable(symbols.table, caption = 'The symbol, definition and units of key traits used.', booktabs = TRUE, escape = FALSE)
-```
-
-#### VPD
-\
-As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between mean daily values by month of gross primary productivity (gC m^2^ day^-1^) and VPD (kPa) we collected from the eddy flux tower located 75 m from the 50-ha plot (for any day *i*, $GPP_i = `r vpd.cf[1]` + `r vpd.cf[2]`VPD_i + `r vpd.cf[3]`VPD_i^2 + error$; 2012-2017; Fig.
-S\@ref(fig:gpp-vpd)) and used standardized $GPP$ to represent the impact of VPD on growth (hereafter, standardized $GPP$ = $VPD^*$ in Eq.
-\@ref(eq:M-chosen); range 0-1).
-At our study site, due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season, the impact of VPD may as well be a combined impact of not only temperature and humidity, but also radiation.
+Table \@ref(tab:symbols) here.
 
@@
 We obtained species-specific parameters $A$ & $B$ for Eq.
-\@ref(eq:k-soil) either from data when available, or a model based on $LMA$ or $WSG$ as described in the earlier section, and used standardized K within species ($K^*$; range 0-1) in Eq.
+\@ref(eq:k-soil) either from data when available, or a model based on $LMA$ or $WSG$ as described in the earlier section, and used K standardized between 0 and 1 within species ($K^*$) in Eq.
 \@ref(eq:M-chosen).
 
-### Model evaluation
+#### Vapor Pressure Deficit (VPD)
 \
+We fitted a polynomial curve between mean midday values (11:00 to 15:00 hrs) of gross primary productivity (gC m^2^ day^-1^) and VPD (kPa) observations we collected from the eddy flux tower located 75 m from the 50-ha plot (for any day *i*, $GPP_i = `r vpd.cf[1]` + `r vpd.cf[2]`VPD_i + `r vpd.cf[3]`VPD_i^2 + error$; 2012-2017; Fig.
+S\@ref(fig:gpp-vpd)) and used $GPP$ standardized between 0 and 1 as $VPD^*$ in Eq.
+\@ref(eq:M-chosen)).
+
+### Model validation and selection
+\
+To validate the ERD models we used data for xylem sap $\delta^2H$ (permil) as an independent observation of ERD.
 For each model of ERD (Eq.
 \@ref(eq:M-chosen) and Eqs.
-in SI), species ERD was regressed against species xylem sap $\delta^2H$ (permil).
+in SI), estimated species ERD was regressed against species xylem sap $\delta^2H$ (permil).
 We used xylem sap $\delta^2H$ values given in @Meinzer:2001 for the dry-season of March 1997 as this period showed largest seasonal divergence among species, with a corresponding gradient of $\delta^2H$ in the soil (Fig.
 S\@ref(fig:iso-data)).
-We removed species that are likely to be deciduous in March as xylem sap $\delta^2H$ of these species may not be linked with water sourced at the time of measurement and thus may not be comparable with other species that had leaves.
-We also removed *G.
-standleyana*, whose xylem sap isotope value was not well constrained by the soil water isotope depth profile.
+We removed species that are likely to have no leaves in March as xylem sap $\delta^2H$ of these species may not be linked with water sourced at the time of measurement and thus may not be comparable with other species that had leaves.
+We also removed *Guapira standleyana*, whose xylem sap isotope value was not well constrained by the soil water isotope depth profile.
 We were thus left with 7 species with data on $\delta^2H$ as well as modelled estimates of ERD.
 
-## ERD Coordination with Above-ground Hydraulic Traits
+## ERD coordination with above-ground hydraulic traits
 \
-To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD with maximum stem hydraulic conductance ($K_{max, stem}$), leaf turgor loss point ($\Psi_{tlp}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$).
+To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD against maximum stem hydraulic conductance ($K_{max, stem}$), leaf turgor loss point ($\Psi_{tlp}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$).
 See [SI](#SIData.brett) for data collection and estimation of these variables.
 
-## Mortality Analyses and Extreme Drought Experience
+## Mortality analyses and species-specific drought experience
 \
@@
 For each species we calculated mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{t=1}^{6} {Mort_{t}}$.
-To test whether ERD plays a role in alleviating mortality risk, we regressed species ERD against species ($Mort_{mean}$).
-To assess whether this may be due to deeper ERD species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the $\Psi_{soil,~z}$ simulation period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{soil,~z = ERD}$) were below species $\Psi_{crit}$, which was defined as $\Psi_{50,~leaf}$.
+
+    We regressed species ERD against species ($Mort_{mean}$).
+To assess whether deeper ERD species escape the experience of significant soil drought, we calculated for each speceis the proportion of days in the simulation period (1990-2015) during which soil water potential in the soil zone matching species ERD ($\Psi_{soil,~z = ERD}$) were below species $\Psi_{crit}$.
+$\Psi_{crit}$ was defined as $\Psi_{50,~leaf}$.
 
 \
-All statistical analyses were done in R statistical environment (version 3.5.3, @R:2019).
+All statistical analyses were conducted in R statistical environment (version 3.5.3, @R:2019).
 
 # Results
-## Soil Water Dynamics By Depth
+## Soil water dynamics by depth
 \
 The ELM-FATES simulations with the chosen best-fit parameter ensembles (Table \@ref(tab:param)) successfully captured the dynamics of soil moisture by depth (Fig.
-\@ref(fig:hydro-calib)A & Fig.
+\@ref(fig:hydro-calib) A & Fig.
 \@ref(fig:hydro-calib)B), including the observations from the horizontal TDR probes that were not included in the calibration.
-These simulations also simultaneously captured the dynamics in stream discharge (Fig.
-\@ref(fig:hydro-calib)C) and evapotranspiration (Fig.
+These simulations also captured the dynamics in stream discharge (Fig.
+\@ref(fig:hydro-calib)C) and to a moderate extent the dynamics in evapotranspiration (Fig.
 \@ref(fig:hydro-calib)D).
 The reduction in parameter range in the best-fit ensembles compared to the global ranges tested showed that the above calibration was sensitive to the following parameters: the Ball-Berry stomatal slope parameter, the ELM rooting distribution parameter that regulates the depth of the rooting profile, soil hydraulic conductivity profile especially at depth, and our adjustment factor to soil hydraulic conductivity to account for macroporosity and direct flow paths (Table \@ref(tab:param)).
-Soil Water Potential dynamics ($\Psi_{soil,~z}$) by depth (z = 0.1 through 13) obtained from the best-fit simulations revealed that census intervals varied in soil droughts in terms of intensity at depth, seasonality and duration (Fig.
+
+Fig. \@ref(fig:hydro-calib) here.
+
+  Soil Water Potential dynamics ($\Psi_{soil,~z}$) by depth (z = 0.1 through 13) obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of such extreme years, drought intensity at depth, drought seasonality and duration (Fig.
 S\@ref(fig:swp-droughts)).
@@
 
-```{r hydro-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-hydro-calib}ELM-FATES calibration and evaluations.
-Observations (red lines) versus simulations (blue lines) from 100 best-fit parameter ensemble runs of ELM-FATES are shown for daily Volumetric Water Content (VWC) by depth for horizontal time domain reflectometry (TDR) probes at three depths (0.1, 0.4 and 1 m; A) and a vertical probe (0-0.15 m; B); monthly stream discharge (Discharge, C) and monthly evapotranspiration from eddy flux tower (ET, D).
-Points (red) in (A) show average plot-wide observations of VWC with 95% CI (error bars).
-VWC data from the horizontal probes (lines in A) were not used for ELM-FATES calibration but evaluation.
-Values in inset are average RMSE across the 100 best-fit simulations.
-For panel A, values at the top are for plot-wide VWC and those at the bottom are for TDR probes.
-VWC is in units of cm^3^ cm^-3^, while ET and Discharge in mm month^-1^.")}
-knitr::include_graphics(file.path("figures/hydro-calib2.jpeg"))
-```
+Fig. \@ref(fig:swp-droughts) here.
 
-## Leaf Vulnerability Curves
+## Leaf hydraulic vulnerability curves
 \
-For the leaf vulnerability curve parameters $A$ and $B$ fitted to the observed data on $K_{leaf}$ vs.
+For the leaf vulnerability curve parameters $A$ and $B$ fitted to the observed data for $K_{leaf}$ vs.
 $\Psi_{leaf}$, $WSG$ and parameter $B$ explained a large proportion of variation in parameter $A$ (Eq.
 \@ref(eq:a-mod); *Adj.
-R*^2^ = `r afit.r2`; *P* `r  afit.p`; also see Fig.
-S\@ref(fig:ab-fit)).
+R*^2^ = `r afit.r2`; *P* `r  afit.p`; Fig.
+\@ref(fig:ab-fit)).
 $WSG$ and $LMA$ explained a large proportion of variation in parameter $B$ (Eq.
 \@ref(eq:b-mod); Adj.
-*R*^2^ = `r bfit.r2`; *P* `r  bfit.p`).
-WSG alone explained `r round(a.wsg.r2*100,0)` % of variation in $A$ ($A = `r a.wsg.cf[1]` `r a.wsg.cf[2]`B + `r a.wsg.cf[3]`B^2 + error$; *P* `r  a.wsg.p`) and `r round(b.wsg.r2*100,0)` % of variation in $B$ ($B = `r b.wsg.cf[1]` `r b.wsg.cf[2]`B + `r b.wsg.cf[3]`B^2 + error$; *P* `r  b.wsg.p`).
-Parameter $B$ alone explained `r round(a.b.r2*100,0)` % of variation in $A$ (*P* `r  a.b.p`), while $LMA$ alone was weakly related to $A$ and was not related to $B$ (not shown).
-Leaf vulnerability curves fitted to observed data on `r length(sp.exp.param$sp)` species as well as percent loss of conuctivity curves are shown in Fig.
+*R*^2^ = `r bfit.r2`; *P* `r  bfit.p`; also see Fig.
 S\@ref(fig:obs-plc-curves).
-Vulnerability curves were predicted using Eqs.
-\@ref(eq:a-mod)-\@ref(eq:b-mod) for `r sp.n.model` species for which such curves were unavailable (out of `r length(unique(erd.iso$sp))` species for which ERD was determined (Fig.
-S\@ref(fig:obs-comm-plc-curves).
+Vulnerability curve parameters were predicted for `r sp.n.model` out of `r length(unique(erd.iso$sp))` ERD species without direct data using Eqs.
+\@ref(eq:a-mod)-\@ref(eq:b-mod) (Table S\@ref(tab:ab-param), Fig.
+S\@ref(fig:obs-comm-plc-curves)).
+Fig. \@ref(fig:ab-fit) here.
 
-## Effective Rooting Depths, Vulnerability to Embolism and Hydraulic Safety Margins
+## Effective rooting depths (ERD)
 \
-In comparison to the measured isotopic signal ($\delta^2H$), our results showed that the ERD model with the effect of LAI and an additive effect of VPD (Eq.
-\@ref(eq:M2)) had highest goodness-of-fit (*R*^2^ = 0.65; *P* = 0.027; Fig.
+ERD estimated from the model with K~leaf~, LAI and an additive effect of VPD as explanatory variables (Eq.
+\@ref(eq:M2)) was strongly related to the measured isotopic signal ($\delta^2H$), that is, with highest goodness-of-fit (*R*^2^ = 0.65; *P* = 0.027; Fig.
 \@ref(fig:erd-iso)).
-The model without LAI and an additive effect of VPD showed no species with sufficient species-level goodness-of-fit.
-All the other ERD models showed a positive but non-significant relationship with $\delta^2H$ at $\alpha$ = 0.05 (Fig.
+The model with K~leaf~, but without LAI and an additive effect of VPD showed no species with sufficient species-level goodness-of-fit.
+ERD from all the other models showed a positive, but non-significant relationship with $\delta^2H$ at $\alpha$ = 0.05 (Fig.
 S\@ref(fig:other-erd-mods)).
-For the rest of the analyses we therefore used species ERD estimates from the model with with the effect of LAI and an additive effect of VPD (Eq.
+For the rest of the analyses we therefore used ERD estimates from the best-fit model (Eq.
 \@ref(eq:M-chosen)).
@@
 
-```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-iso}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE; m) versus dry season stable isotopic concentration $\\delta^2H$ (mean $\\pm$ 1SE; ‰) for ", length(unique(erd.iso$sp))," canopy species from Barro Colorado Island.
-Data from [@Meinzer:1999].
-See Table  \\@ref(tab:ab-param) for the full form of scientific names.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_phenology_Meinzer_gr.Psi.VPD.jpeg"))
-```
+Fig. \@ref(fig:erd-iso) here.
 
 \
-In this forest modeled effective rooting depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from `r min(df.erd.to.plot$depth)` m to `r max(df.erd.to.plot$depth)` m (Fig.
+  In this forest, modeled effective rooting depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from `r min(df.erd.to.plot$depth)` m to `r max(df.erd.to.plot$depth)` m (Fig.
 \@ref(fig:sp-erd)).
-For the subset of species for which hydraulic properties were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max,~stem}$, *R*^2^ = 0.39, *P* = 0.098, *n* = 8; Fig.
-\@ref(fig:erd-hyd)A), less negative leaf $\Psi_{tlp}$ (*R*^2^ = 0.6, p = 0.024, *n* = 8; Fig.
-\@ref(fig:erd-hyd)B), less negative $\Psi_{88,~stem}$ and thus greater vulnerability to embolism from cavitation (*R*^2^ = 0.72, p = 0.007, *n* = 8; Fig.
-\@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$; *R*^2^ = 0.74, p = 0.013, *n* = 8; Fig.
+
+Fig. \@ref(fig:sp-erd) here.
+
+## Coordination of ERD with other hydraulic traits
+
+  For the subset of 8 species with ERD for which hydraulic traits data were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max,~stem}$, Spearmans' *r* = `r cor.k`, *P* = `r p.k` at $\alpha$ = 0.05; Fig.
+\@ref(fig:erd-hyd)A), less negative leaf $\Psi_{tlp}$ (*r* = `r cor.tlp`, *P* = `r p.tlp`; Fig.
+\@ref(fig:erd-hyd)B), less negative $\Psi_{88,~stem}$ and thus greater vulnerability to embolism from cavitation (*r* = `r cor.stem.88`, *P* = `r p.stem.88`; Fig.
+\@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$; *r* = `r cor.hsm`, *P* = `r p.hsm`; Fig.
 \@ref(fig:erd-hyd)D).
 
-```{r sp-erd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap= paste0("\\label{lab-sp-erd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) for ", length(erd.sp)," large, canopy species of Barro Colorado Island.
-See Table  \\@ref(tab:ab-param) for the full form of scientific names.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/ERD_by_sp_large_canopy.jpeg"))
-```
+Fig. \@ref(fig:erd-hyd) here.
 
-```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Matrix for pairwise correlations (circles) among Effective Rooting Depth (ERD) and hydraulic properties for ", nrow(cor.data)," canopy species found on Barro Colorado Island (Panama).
-Each circle is color coded by the Spearman's $\\rho$ with red shades for positive $\\rho$ values, while blue shades for negative values, respectively.
-P-values are given only when they are $\\ge$ 0.05.
-Units are as follows: ERD in m, K~max~ in mmol m^-1^ s^-1^ MPa^-1^), all variables with $\\Psi$ in MPa.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/erd.stem.traits_cor.matrix_all_p-val.jpeg"))
-```
-
-## Effective Rooting Depths, Experienced Drought and Mortality
+## Effective rooting depths, experienced drought and mortality
 \
@@
 For five out of seven census intervals, species with shallower ERD had significantly higher mortality anomalies than deeper ERD species (Fig.
-\@ref(fig:erd-mort-census)).
-Analysis of time spent under catastrophic soil drought showed that the proportion of days species $\Psi_{ERD}$ was more negative than species $\Psi_{crit}$ decreased with species ERD (Fig.
+S\@ref(fig:erd-mort-census)).
+Analysis of time spent under significant soil drought showed that the proportion of days species $\Psi_{ERD}$ was more negative than species $\Psi_{crit}$ decreased with species ERD, that is the shallower the ERD, the greater proportion of days $\Psi_{ERD}$ < $\Psi_{crit}$ (Fig.
 \@ref(fig:erd-time)).
@@
 
-```{r erd-mort, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-mort}Mortality rate (mean $\\pm$ 1SE) versus Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) for ", length(erd.mrate.sp)," canopy species found on Barro Colorado Island.
-*R*^2^ and *P*-values are given in panel insets.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_rate_by rdi.gr.jpeg"))
-```
-
-```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time}Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\\Psi_{soil,~z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for ", length(erd.mrate.sp), " canopy species found on Barro Colorado Island.
-Each circle represents a species.
-Circles are colored and jittered vertically around the named depths (ticks) for clarity.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr.jpeg"))
-```
+Fig. \@ref(fig:erd-mort) here.
+Fig. \@ref(fig:erd-time) here.
 
@@
 \
-Our analyses revealed that co-occurring large canopy tree species (*n* = `r length(erd.sp)`) with deeper effective rooting depths (ERD) were associated with higher hydraulic efficiency, but not safety.
-Deep ERD species showed higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species.
-This trade-off translated into greater drought tolerance, as deeper ERD species showed lower mortality rates than shallower ERD species over the 35-yr record (1982-2015).
-Analysis of time spent below ($\Psi_{crit}$) in the soil zone matching ERD $\Psi_{soil,~z = ERD}$ showed that this was likely due to decreasing realized risk of hydraulic failure with increasing rooting depth.
-We thus show, for the first time, that inter-specific variation in trees' effective rooting depths aligns species on a conservative-acquisitive water economic spectrum by demonstrating a link between trees' realized drought experience, hydraulic architecture and long-term mortality outcomes.
-On the acquisitive end of the spectrum were deep ERD species with a  hydraulically efficient, but risky, suite of traits optimized for an environment of reliable and abundant source of water at depth that resulted in a lower mortality risk.
-At the conservative end of the spectrum were shallow ERD species with inefficient, but safer, xylem likely optimized for high variability in $\Psi_{soil}$ in the shallow soil layers, and resulting in higher variance in mortality rates across censuses as $\Psi_{soil,~z = ERD}$ crossed ERD $\Psi_{crit}$.
-Mortality rates of shallow ERD species may have been exacerbated by increased drought events during the study period.
-@Condit.2017 showed that half of the extreme dry seasons (< 1 SD annual water deficit) during 1931-2015 (85-yr) occurred during 1982-2005 (24-yr) of the 34-yr long study period (1982-2015).
+Our analyses revealed that co-occurring large canopy tree species with deeper effective rooting depths (ERD) were associated with higher hydraulic efficiency, but not safety.
+Species with deeper ERD showed higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species.
+Nevertheless, deeper ERD species showed lower mortality rates than shallower ERD species over the 35-yr record (1982-2015).
+Analysis of time spent below ($\Psi_{crit}$) in the soil zone matching ERD $\Psi_{soil,~z = ERD}$ showed that this was likely due to decreasing realized risk of hydraulic failure with increasing access to water for species with deeper ERD.
+We thus demonstrate, for the first time, a link between realized drought experience, hydraulic architecture and long-term mortality outcomes for co-occurring adult tree species.
 
@@
 We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins, that is, traits that are commonly identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through a study period of greater recurrence of drought.
-These results are consistent with the two studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015].
+These results are consistent with the two studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016].
 @Johnson:2018fad found that mortality decreased with deep rooting even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi_{88,~stem}$) and had narrower safety-margins ($\Psi_{min} - \Psi_{88,~stem}$).
@@
 Our result of lower mortality in deep rooted trees contrasts with the finding of @Chitra-Tarak.2018 in which species identified by their inverse ERD model as effectively deep rooted died *more* in a prolonged, extreme drought.
-Their hydrological model revealed that the multi-annual drought exhausted the deep water reservoir for over two years, whereas our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period, $\Psi_{soil}$ remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) below the depth of 1 m through all the dryseasons and drought periods, that is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).
+Their hydrological model revealed that the multi-year drought exhausted the deep soil and even bedrock water availablity.
+In contrast, our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period, $\Psi_{soil}$ below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and droughts, that is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).
 The difference in our results are likely attributable to site differences in the precipitation regime.
@@
 \
-By estimating ERD for large trees of `r length(erd.sp)` canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species species-rich tropical forests.
-Our chosen ERD model is an advance over the only other inverse model of ERD (@Chitra-Tarak.2018; *n* = 12 species) in terms of physiologically meaningful representation, fewer parameters and corroboration with tree and soil data on $\delta^2H$.
-The model in @Chitra-Tarak.2018 estimated six species-specific phenomenological parameters: three as root fraction for corresponding lumped vadose zone layers, and three as shape parameters for water uptake vs availability curves.
-For each species, they calculated daily water uptake successively from three layers, that is, the supply, to satisfy the demand of the daily maximum evapotranspiration, a function of PET and LAI.
-They regressed total supply to total demand ratio over census intervals as a growth coefficient against growth (five data points), and chose values for six parameters that best-fit (highest *R*^2^, that is, two additional parameters).
-Our inverse ERD model fits only two parameters by way of regressing depth-specific model against five growth data points.
-Our model was constrained on VPD and species leaf vulnerability curves, variables that are shown to be physiologically meaningful.
+By estimating ERD for large trees of `r length(erd.sp)` canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species-rich tropical forests.
+Our chosen ERD model is an advance over the only other inverse model of ERD (@Chitra-Tarak.2018; *n* = 12 species) as we utilize physiologically meaningful representation, fewer parameters and corroboration with tree and soil data on $\delta^2H$.
 
 \
-We benefited from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA and WSG) and parameters of leaf vulnerability (or percent loss of conductance, (PLC)) curves for which we collected data on a few species (*n* = `r length(sp.exp.param$sp)`), a 'hard' trait.
+We benefited from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA and WSG) and parameters of leaf vulnerability (or percent loss of conductance, (PLC)) curves for a few species (*n* = `r length(sp.exp.param$sp)`), a 'hard' trait.
 This allowed us to scale-up the leaf vulnerability curves at the community scale (*n* = `r length(ab.table$Species)` species, see Table S\@ref(tab:ab-param) for model predicted parameters).
@@
 
-\ Deurwaerder.2020
-Another important piece of information is stable water isotopes of xylem water and below-ground sources, in the vadose zone and water table.
-Although the interpretation of these data is still under research (Deurwaerder et al., 2020), they are critical for model calibration or validation.
+\ 
+Another important piece of information for understanding effective rooting depth is stable water isotopes of xylem water and below-ground sources, in the vadose zone and water table.
+Although the interpretation of these data is still under research [@Deurwaerder.2020], they are critical for model calibration or validation.
 The model in @Chitra-Tarak.2018 were not corroborated with such data as existing methods were inadequate to extract water bound to the saprolite found in their dry forest study site (Jean Riotte, pers.
@@
 
-### Future Directions for Model Development
+### Future directions for model development
 \
 The trade-offs that we identify between above- and below-ground traits, the water environment vertically below-ground and mortality rates are important for representing diversity in Dynamic Global Vegetation Models especially in species-rich forests.
-The hydraulic efficiency-safety trade-off between high values of maximum leaf or stem conductivity (high $K_{max}$) and cavitation-resistance (low $\Psi_{50}$ or $\Psi_{88}$) is hypothesized although community-wide examples are rare [@Hacke:2006; @Gleason.2016].
-Such a trade-off is found at our study site at both leaf and stem level (Zailaa et al.
-to be submitted; Wolfe et al.
-in revision).
-We also found that LMAand WSG were strong predictors of PLC curves for leaves, but not for stems (not shown).
-Acquisitive traits such as high $K_{max}$ are expected to be associated with faster growth rates.
-We found a positive but non- significant relationship of adult growth rates ($\ge$ 10cm dbh) with $K_{max, ~leaf}$ (*R*^2^ = 0.13; *P* = 0.126, *n* = `r length(sp.exp.param$sp)`) and no relationship with $K_{max, ~stem}$ (*n* = `r stem.traits.sp` species).
-The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits in order to better represent the functional diversity of tropical forests.
+The hydraulic efficiency-safety trade-off between high values of maximum stem conductivity (high $K_{max,~stem}$) and cavitation-resistance (low $\Psi_{50,~stem}$ or $\Psi_{88,~stem}$) is hypothesized although community-wide examples are rare [@Hacke:2006; @Gleason.2016].
+We report such a trade-off (n = 26 species) at our study site stem level ().
+We also found that LMA and WSG were strong predictors of PLC curves for leaves, but not for stems (not shown).
+WSG alone explained `r round(a.wsg.r2*100,0)` % of variation in parameter $A$ ($A = `r a.wsg.cf[1]` `r a.wsg.cf[2]`B + `r a.wsg.cf[3]`B^2 + error$; *P* `r  a.wsg.p`) and `r round(b.wsg.r2*100,0)` % of variation in $B$ ($B = `r b.wsg.cf[1]` `r b.wsg.cf[2]`B + `r b.wsg.cf[3]`B^2 + error$; *P* `r  b.wsg.p`).
+Parameter $B$ alone explained `r round(a.b.r2*100,0)` % of variation in $A$ (*P* `r  a.b.p`), while $LMA$ alone was weakly related to $A$ and was not related to $B$ (not shown).
+The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits, including hydraulic vulnerability, in order to better represent the functional diversity of tropical forests, and their trajectories into the future.
 
 \
-Our model estimated ERD as a single value, unlike the dynamic ERD modeled by @Chitra-Tarak.2018, but future model developments should investigate seasonal dynamics of ERD, using corresponding calibration data on isotopically determined water-sourcing depths.
-Our inverse ERD model was parameterized on 5-yearly growth data, with five data points over a 25-yr period, which decouples climate events and demographic outcomes.
+Our model estimated ERD as a single value, unlike the dynamic ERD modeled by @Chitra-Tarak.2018, but future model developments should investigate seasonal dynamics of ERD, using corresponding calibration data on isotopically determined water-sourcing depths, as species are expected to shift to deeper soil water sources as shallow soils dry out [@Meinzer:1999].
+Our inverse ERD model was parameterized on 5-year growth data, with five data points over a 25-yr period, which decouples climate events and demographic outcomes.
 Future studies could better constrain the model with higher resolution growth data such as those from dendrometer bands.
 This is especially important for modeling seasonally dynamic ERD.
-However, trees' stem diameter change is a sum total of not just growth, the addition of wood, but also changes due to hydrostatic stem flexing, importance of which increases with temporal resolution [@Chitra-Tarak.2016; @Chitra-Tarak.2015].
+However, trees' stem diameter change is a sum total of not just growth, the addition of wood, but also changes due to hydrostatic stem flexing (xylem, phloem and bark hydration), the importance of which increases with temporal resolution [@Chitra-Tarak.2016; @Chitra-Tarak.2015].
 Use of dendrometer data would thus entail accounting for stem hydrostatic flexing [@Zweifel:2018].
@@
 We interpret this result as that VPD and PLC curves may be adequate to explain growth, but also acknowledge that our estimates of seasonality of LAI is a tentative estimate combining leaf fall data and assumption of leaf presence during the wet season, and omits inter-annual variation.
-Future studies could make use of data from new technologies such as drone based monitoring of species-specific leaf area fraction in species-rich forests [@Park.2019].
+Future studies could make use of data from new technologies such as drone based monitoring of species-specific leaf area fraction in species-rich tropical forests [@Park.2019].
 In forests such as BCI where understory tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes, especially from a few growth data points, may need accounting for variation in trees' light environment over time and among trees of the same size class, an issue we minimized by limiting ERD estimates to canopy trees.
@@
 
+# Captions for Tables and Figures
+
+Table \@ref(tab:hypo): Hypotheses for association between Effective Rooting Depth (ERD) and hydraulic traits.
+
+Table \@ref(tab:symbols): The symbol, definition and units of key traits used in the simulations and analyses.
+
+Figure \@ref(fig:schematic): A schematic for the methods workflow.
+
+Figure \@ref(fig:hydro-calib): ELM-FATES calibration and evaluations.
+Observations (red lines) versus simulations (blue lines) from 100 best-fit parameter ensemble runs of ELM-FATES are shown for daily Volumetric Water Content (VWC) by depth for horizontal time domain reflectometry (TDR) probes at three depths (0.1, 0.4 and 1 m; A) and a vertical probe (0-0.15 m; B); monthly stream discharge (Discharge, C) and monthly evapotranspiration from eddy flux tower (ET, D).
+Points (red) in (A) show average plot-wide observations of VWC with 95% CI (error bars).
+VWC data from the horizontal probes (lines in A) were not used for ELM-FATES calibration but evaluation.
+Values in inset are average RMSE across the 100 best-fit simulations.
+For panel A, values at the top are for plot-wide VWC and those at the bottom are for TDR probes.
+VWC is in units of cm^3^ cm^-3^, while ET and Discharge in mm month^-1^.
+
+Figure \@ref(fig:swp-droughts): Occurrence of extreme soil droughts by census intervals (horizontal panels) at three representative depths (z): 0.1, 0.6 and 1 m (vertical panels).
+Mean (black lines) and lower half of 95% distribution (grey areas) of soil water potential ($Psi_{soil,~z}$) for a given DOY and depth z are shown (same across census intervals).
+$Psi_{soil,~z}$ by DOY for a year (colored line) is only shown if for some part of that year $Psi_{soil,~z}$ declined below the 2.5 percentile distribution.
+
+Figure \@ref(fig:ab-fit): Parameters $A$ (panel A) & $B$ (panel B) for the leaf vulnerability curves that were fitted to observed data on $K_{leaf}$ vs.
+$\\Psi_{leaf}$ for `r length(sp.exp.param$sp)` species (Eq.
+\@ref(eq:vc-exp)) vs.
+those that were predicted from model based on trait-proxies.
+(Eqs.
+\@ref(eq:a-mod)-\@ref(eq:b-mod)).
+Goodness-of-fit (*R*^2^) and *P*-values are given in inset.
+
+Figure \@ref(fig:erd-iso): Modelled Effective Rooting Depth (ERD; mean $\pm$ 1SE; m) versus dry season stable isotopic concentration $\delta^2H$ (mean $\pm$ 1SE; ‰) for `r length(unique(erd.iso$sp))` canopy species from Barro Colorado Island.
+Data from [@Meinzer:1999].
+See Table  \@ref(tab:ab-param) for the full form of scientific names.
+
+Figure \@ref(fig:sp-erd): Modelled Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) for `r  length(erd.sp)` large, canopy species of Barro Colorado Island.
+
+Figure \@ref(fig:erd-iso): Matrix for pairwise correlations (circles) among Effective Rooting Depth (ERD) and hydraulic properties for `r nrow(cor.data)` canopy species found on Barro Colorado Island (Panama).
+Each circle is color coded by the Spearman's $\\rho$ with red shades for positive $\\rho$ values, while blue shades for negative values, respectively.
+P-values are given only when they are $\\ge$ 0.05.
+Units are as follows: ERD in m, K~max~ in mmol m^-1^ s^-1^ MPa^-1^), all variables with $\\Psi$ in MPa.
+
+Figure \@ref(fig:erd-mort): Mortality rate (mean $\pm$ 1SE) versus Modelled Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) for `r length(erd.mrate.sp)` canopy species found on Barro Colorado Island.
+*R*^2^ and *P*-values are given in panel insets.
+
+Figure \@ref(fig:erd-time): Modelled species Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\Psi_{soil,~z}$ was below $\Psi_{crit}$, defined as $\Psi_{80,~leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for `r length(erd.mrate.sp)` canopy species found on Barro Colorado Island.
+Each circle represents a species.
+Circles are colored and jittered vertically around the named depths (ticks) for clarity.
+
+# Tables
+
+Hypotheses for association between Effective Rooting Depth (ERD) and hydraulic traits.
+The symbol, definition and units of key traits used.
+
+```{r hypo, results = 'asis', echo = FALSE}
+knitr::kable(hypo.table, caption = '', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(hypo.table))))
+```
+
+```{r symbols, results = 'asis', echo = FALSE}
+knitr::kable(symbols.table, caption = '', booktabs = TRUE, escape = FALSE)
+```
+
+# Figures
+
+```{r schematic, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-schematic}")}
+knitr::include_graphics(file.path("figures/schematic.png"))
+```
+
+```{r hydro-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-hydro-calib}")}
+knitr::include_graphics(file.path("figures/hydro-calib2.jpeg"))
+```
+
+```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts}"}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q2.5_by_depth.jpeg"))
+```
+
+```{r ab-fit, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ab-fit}")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/A_B_data_vs_model.jpeg"))
+```
+
+```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-iso}")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_phenology_Meinzer_gr.Psi.VPD.jpeg"))
+```
+
+```{r sp-erd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap= paste0("\\label{lab-sp-erd}")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/ERD_by_sp_large_canopy.jpeg"))
+```
+
+```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/erd.stem.traits_cor.matrix_all_p-val.jpeg"))
+```
+
+```{r erd-mort, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-mort}")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_rate_by rdi.gr.jpeg"))
+```
+
+```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time}")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr.jpeg"))
+```
+
 \newpage
@@
 
-### Stream discharge from the Conrad catchment (#SIData.steveP)
+### Stream discharge from the Conrad catchment {#SIData.steveP}
 
@@
 
-### Volumetric Water Content
+### Volumetric water content
 
@@
 
-### Leaf maximum hydraulic conductivity and vulnerability to cavitation {#SIData.joseph}
+### Leaf hydraulic conductivity and vulnerability to cavitation {#SIData.joseph}
 
-The Kleaf was measured under steady state and high irradiance following the Evaporative Flux Method (FEM) [@Sack.2011]  using a flowmeter.
-Briefly, the leaf is excised under filtered DI water and attached to a water source (graduated cylinder) using clear plastic tubing.
+The K~leaf~ was measured under steady state and high irradiance following the Evaporative Flux Method (FEM) [@Sack.2012] using a flowmeter.
+Briefly, the leaf was excised under filtered deionised water and attached to a water source (graduated cylinder) using clear plastic tubing.
 The system includes tubing of a known resistance, separated upstream and downstream by two sensors measuring the transpiration-driven water flow pressure in the EFM system [@Sack.2011].
 The K~leaf~ is calculated as the ratio of the transpiration rate (E~leaf~; mmol s^-1^) to the difference in flow pressure (MPa) upstream and downstream from the resistance tubing, and further standardized to 25°C and by leaf area.
-At stability (CV < 5%), the leaf is left to equilibrate in plastic bags with moist paper towels and measured for \Psi~leaf~ using a Scholander pressure chamber.
+At stability (CV < 5%), the leaf is left to equilibrate in plastic bags with moist paper towels and measured for $\Psi$~leaf~ using a Scholander pressure chamber.
 To construct vulnerability curves for each species, four different functions were fitted to species-specific K~leaf~ vs.
-\Psi~leaf~ data using the *anneal* optimization function in the R package likelihood: exponential without an intercept ($K_{leaf} = Ae^{-B\Psi_{leaf}}$), exponential with an intercept ($K_{leaf} = C + Ae^{-B\Psi_{leaf}}$), logistic ($K_{leaf} = \frac{A}{1 + \left(\frac{\Psi_{leaf}}{x_{0}}\right)^B}$) and sigmoidal ($K_{leaf} = \frac{A}{1 + e^{-\left(\frac{\Psi_{leaf} - x_{0}}{B}\right)}}$).
-The differences among AICc values for these functions were not large within species (Zailaa et al.
-to be submitted; (see Zailaa et al.
+$\Psi$~leaf~ data using the *anneal* optimization function in the R package likelihood: exponential without an intercept ($K_{leaf} = Ae^{-B\Psi_{leaf}}$), exponential with an intercept ($K_{leaf} = C + Ae^{-B\Psi_{leaf}}$), logistic ($K_{leaf} = \frac{A}{1 + \left(\frac{\Psi_{leaf}}{x_{0}}\right)^B}$) and sigmoidal ($K_{leaf} = \frac{A}{1 + e^{-\left(\frac{\Psi_{leaf} - x_{0}}{B}\right)}}$) [@Sack.2012].
+The differences among AICc values for these functions were not large within species (see Zailaa et al.
 for more detailed methods description).
+We chose the exponential function without the intercept to fit observed data across species, and as a general function to scale-up to the community, because it had the lowest AICc for most species, captured the form of the observed data across all species well, and had the lowest number of parameters, that is, two.
 
@@
 
-\
-At the community level, we filled data gaps in $LMA$ for `r sum(c.lma.gaps[-which(names(c.lma.gaps) == "original")])` species (of total `r sum(c.lma.gaps)`)  using linear relationships with other variables: `r c.lma.gaps[[which(names(c.lma.gaps) == "filled-LMA-DISC")]]` species with $LMA_{disc}$, `r c.lma.gaps[[which(names(c.lma.gaps) == "filled-LMA-Nobby")]]` with $LMA_{lamina}$, and `r c.lma.gaps[[which(names(c.lma.gaps) == "filled-LEAF")]]` with $LMA_{leaf}$, defined as the mean leaf mass per unit area measured for the entire leaf including the petiole (g m^-2^) for leaves receiving direct sunlight ($\overline{LMA} = `r leaf.cf[1]` + `r leaf.cf[2]`LMA_{leaf}$; Adj.
-*R*^2^ = `r leaf.r2`; *P* `r leaf.p`, *n* = `r leaf.n`).
-For the `r sum(c.lma.gaps)` species thus with $LMA$, we gap-filled data on  WSG for `r sum(c.wsg.gaps[-which(names(c.wsg.gaps) == "original")])` species using a linear relationship with wood specific gravity reported by @Chave:2009 ($\overline{WSG} = `r chave.cf[1]` + `r chave.cf[2]`WSG_{chave}$; Adj.
-*R*^2^ = `r chave.r2`; *P* `r chave.p`, *n* = `r chave.n`).
-
 ## Alternative models for Effective Rooting Depth (ERD)
@@
 Below we present all the altervative models for Effective Rooting Depth (ERD).
-Daily average species-specific growth, G~t | z~, in the census interval *t* given an ERD of depth *z*, is described as follows:
+Daily average species-specific growth, $G_{t~|~z}$, in the census interval *t* given an ERD of depth *z*, is described as follows:
 
@@
 
-```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "35%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd}Scatterplot of mean daily gross primary productivity (GPP) and mean daily vapor pressure deficit (VPD) by month measured at the eddy flux tower in Barro Colorado Island over 2012-2017.
+```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "35%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd}Scatterplot of mean midday (11:00 am to 15:00 pm) gross primary productivity (GPP) and mean midday vapor pressure deficit (VPD) measured at the eddy flux tower in Barro Colorado Island over 2012-2017.
 Red line shows a fit from a polynomial model described in inset, and used to standardize VPD employed in equations \\@ref(eq:M2) and \\@ref(eq:M4)."}
@@
 
-```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts}Occurrence of extreme soil droughts by census intervals (horizontal panels) at three representative depths (z): 0.1, 0.6 and 1 m (vertical panels).
-Mean (black lines) and lower half of 95% distribution (grey areas) of soil water potential ($Psi_{soil,~z}$) for a given DOY and depth z are shown (same across census intervals).
-$Psi_{soil,~z}$ by DOY for a year (colored line) is only shown if for some part of that year $Psi_{soil,~z}$ declined below the 2.5 percentile distribution.
-Wet season is shown in gray beginning with DOY 120."}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q2.5_by_dpeth.jpeg"))
-```
-
 ```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($Psi_{soil,~z}$) by depth z (m; panels) over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines).
@@
 knitr::include_graphics(file.path("figures/2019-10-14_5000/best-fits/psi_model_daily_all_depths_params.top.few_full.jpeg"))
-```
-
-```{r ab-fit, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ab-fit}Parameters $A$ (panel A) & $B$ (panel B) of leaf vulnerability curves that were fitted to observed data on $K_{leaf}$ vs.
-$\\Psi_{leaf}$ for ", length(sp.exp.param$sp)," species (Eq.
-\\@ref(eq:vc-exp)) vs.
-those that were predicted from models based on species 'soft-traits', $WSG$ and $LMA$.
-(Eqs.
-\\@ref(eq:a-mod)-\\@ref(eq:b-mod)).
-Goodness-of-fit (*R*^2^) and *P*-values are given in inset.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/A_B_data_vs_model.jpeg"))
 ```
@@
 
-```{r obs-comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-comm-plc-curves}Vulnerability curves for species with ERD estimates.
-Curves in  data on ", length(sp.exp.param$sp)," tree species from BCI (A & B) with equation \\@ref(eq:vc-exp)--in absolute terms (A, B), as well as, in terms of loss of conductivity (%) (C, D).
-Each line represents a species, color coded by source of the curve: fitted to observed data (yellow ochre) on $K_{leaf}$ vs.
-$\\Psi_{leaf}$ or based on predicted model parameters $A$ and $B$ from soft traits.
-(gray).")}
+```{r obs-comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-comm-plc-curves}Percent loss of vulnerability curves for species with ERD estimates.
+Each line represents a species, color coded by source of the curve: fitted to observed data for $K_{leaf}$ vs.
+$\\Psi_{leaf}$ are in yellow and those based on scaling relationship with WSG and LMA are in gray.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/obs.mod_plccurves_fits_ggplot_no_col.jpeg"))
@@
 or each model are shown in inset, along with the goodness-of-fit (*R*^2^) and *P*-values for a linear relationship between ERD and $\\delta^2H$.
-See Eqs.
-\\@ref(eq:M1)-\\@ref(eq:M6).
-See Table \\@ref(tab:ab-param) for the full form of scientific names.")}
+ERD models in panels A-E correspond to Eq.
+\\@ref(eq:M1)-\\@ref(eq:M6), respectively.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_sp_color_Meinzer.jpeg"))
@@
 
+```{r ab-param, results = 'asis', echo = FALSE}
+knitr::kable(ab.table, caption = 'Leaf vulnerability curves parameters A & B, K~max, leaf~ and $\\Psi$ for the " length(erd.sp)" species with ERD estimates.
+Species for which curves were fitted to observed data for $K_{leaf}$ vs.
+$\\Psi_{leaf}$ are identified with Source as Data, while species for which curves were obtained based on trait-proxies are identified with Source as Model', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(ab.table))))
+```
+
 # References

### 2020-10-17 old_ms_sections.Rmd
first commit

---
title: "Hydraulically-vulnerable trees are buffered by deep-water access in a tropical forest community"
author: ''
date: ''
output:
  bookdown::html_document2:
    number_sections: no
    theme: flatly
    fig_caption: yes
  bookdown::word_document2:
        reference_docx: manuscriptstyle.docx
        fig_caption: yes
  bookdown::pdf_document2: default
bibliography: references.bib
csl: new-phytologist.csl
---

```{r, include=FALSE, echo = FALSE}
if(!require("pacman", quietly=TRUE)) install.packages("pacman")
suppressPackageStartupMessages(
  pacman::p_load(bookdown, knitr, pander, citr, broom))
options(tinytex.verbose = TRUE)
source("code/load.R")
source("code/manuscript_data_figs.R")
```

```{r setup, echo = FALSE}
library(knitr)
opts_knit$set(eval.after = "fig.cap")
```

Specifically for the water environment (or, resource), observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified an acquisitive-conservative resource economic spectrum similar to those for light and nutrients, although, for water, examples of all trade-offs observed simultaneously are rare, and below-ground traits or environment linkages are weak [@Reich:2014; @Sterck:2011]. Acquisitive species have traits with high capacity to acquire, move, store and use water (higher hydraulic conductivity, low wood density) and achieve high growth, but build flimsy tissues (high specific leaf area) and are less tolerant of low resource availablity, whereas species with conservative traits are better protected from carbon losses and are more physiologically drought tolerant (greater embolism resistance, larger safety margins). Whether and how the acquisitive-conservative spectrum holds true for the water dynamics vertically below-ground is not clear. Acquisitive species are spatially associated with wet sites (high resource) such as valleys or those with high rainfall [@Kursar:2009; @Oliveira:2019]. Deep rooting may be expected to allow an acquisitive strategy by enabling a steady water supply. However, the degree to which deep rooting is associated with an acquisitive strategy may depend on the site-specific dynamics of water availability by depth, which is driven by precipitation regime and hydrology. For example, in pulse-driven systems *shallow*-rooted plants are associated with acquisitive traits (higher hydraulic conductivity) and opportunistic strategies (e.g. dry deciduousness) that maximise the use of the water-pulse, while *deep*-rooted plants show conservative traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013 and references therein].

\
At the community level, we filled data gaps in $LMA$ for `r sum(c.lma.gaps[-which(names(c.lma.gaps) == "original")])` species (of total `r sum(c.lma.gaps)`)  using linear relationships with other variables: `r c.lma.gaps[[which(names(c.lma.gaps) == "filled-LMA-DISC")]]` species with $LMA_{disc}$, `r c.lma.gaps[[which(names(c.lma.gaps) == "filled-LMA-Nobby")]]` with $LMA_{lamina}$, and `r c.lma.gaps[[which(names(c.lma.gaps) == "filled-LEAF")]]` with $LMA_{leaf}$, defined as the mean leaf mass per unit area measured for the entire leaf including the petiole (g m^-2^) for leaves receiving direct sunlight ($\overline{LMA} = `r leaf.cf[1]` + `r leaf.cf[2]`LMA_{leaf}$; Adj. *R*^2^ = `r leaf.r2`; *P* `r leaf.p`, *n* = `r leaf.n`). For the `r sum(c.lma.gaps)` species thus with $LMA$, we gap-filled data on  WSG for `r sum(c.wsg.gaps[-which(names(c.wsg.gaps) == "original")])` species using a linear relationship with wood specific gravity reported by @Chave:2009 ($\overline{WSG} = `r chave.cf[1]` + `r chave.cf[2]`WSG_{chave}$; Adj. *R*^2^ = `r chave.r2`; *P* `r chave.p`, *n* = `r chave.n`).

From Discussion first para

Deeper ERD species had on average more hydraulically efficient, but risky, suite of traits optimized for an environment of more reliable and greater water source at depth that resulted in a lower mortality risk. By contrast shallow ERD species had inefficient, but safer, xylem likely optimized for high variability in $\Psi_{soil}$ in the shallow soil layers, and resulting in higher variance in mortality rates across censuses as $\Psi_{soil,~z = ERD}$ crossed ERD $\Psi_{crit}$. Mortality rates of shallow ERD species may have been exacerbated by increased drought events during the study period. @Condit.2017 showed that half of the extreme dry seasons (< 1 SD annual water deficit) during 1931-2015 (85-yr) occurred during 1982-2005 (24-yr) of the 34-yr long study period (1982-2015).

The model in @Chitra-Tarak.2018 estimated six species-specific phenomenological parameters: three as root fraction for corresponding lumped vadose zone layers, and three as shape parameters for water uptake vs availability curves. For each species, they calculated daily water uptake successively from three layers, that is, the supply, to satisfy the demand of the daily maximum evapotranspiration, a function of PET and LAI. They regressed total supply to total demand ratio over census intervals as a growth coefficient against growth (five data points), and chose values for six parameters that best-fit (highest *R*^2^, that is, two additional parameters). Our inverse ERD model fits only two parameters by way of regressing depth-specific model against five growth data points. Our model was constrained on VPD and species leaf vulnerability curves, variables that are shown to be physiologically meaningful.

### 2020-10-17 Appendix.Rmd
First commit

--- 2020-10-16 manuscript.rmd
@@
 ---
-title: "Hydraulically-vulnerable trees are buffered by deep-water access in a tropical forest community"
+title: "Drought-sensitive trees survive on deep water access in a tropical forest community"
+subtitle: 'Appendix'
 author: ''
@@
   bookdown::html_document2:
-    number_sections: no
-    theme: flatly
-    fig_caption: yes
     bookdown::word_document2:
@@
         fig_caption: yes
+  number_sections: no
+theme: flatly
 bookdown::pdf_document2: default
+csl: new-phytologist.csl
 bibliography: references.bib
-csl: new-phytologist.csl
+number_sections: true
+# header-includes: 
+#   \usepackage{float} \floatplacement{figure}{H} 
+#   \setcounter{table}{0}  
+#   \renewcommand{\thetable}{S\arabic{table}} 
+#   \setcounter{figure}{0} 
+#   \renewcommand{\thefigure}{S\arabic{figure}}
 ---
@@
 
-\
-**Corresponding author**: 
-Rutuja Chitra-Tarak
-\
-E: arutuj@gmail.com
-\
-T: +1(505) 665 1647
+\documentclass{article}
 
-# Summary
+\usepackage{chngcntr}
+\begin{document}
 
-- Among the mechanisms that plants employ to tolerate drought, deep water access is arguably the most commonly invoked, but under-studied.
-Vulnerability to embolism and hydraulic safety margins are shown to be predictors of drought-induced mortality risk, but deep-water access may delay plants from reaching critical safety thresholds.
-Here, we tested the role of deep-water access in enabling survival within a diverse tropical forest community (Panama) using a novel data-model approach.
-- We inversely modeled effective rooting depth (ERD, as the average depth of water extraction), for large trees of `r length(erd.sp)` canopy species by linking growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and to leaf hydraulic vulnerability curves.
-We validated ERD against existing tree and soil stable water isotope data.
-- Across species, deeper ERD was associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates over 35 years.
-Species with deeper ERD spent less time below critical hydraulic thresholds indicating that trees compensate for water-stress related mortality risk through deeper water access.
-- The role of deep-water access in alleviating mortality of hydraulically-vulnerable trees has important implications for our predictive understanding of forest dynamics under current and future climates.
+\section{Introduction}
 
-\
-Keywords: below-ground traits, deep-water access, drought-induced mortality, drought tolerance, rooting depths, tropical forest, life-history strategies
+\subsection{Problem description}
 
-# Introduction
-\
-    Tropical forests play a disproportionately large role in the global carbon and energy cycles [@Bonan:2008], and support half of global biodiversity [@Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019].
-Tropical forests are logically considered to be an especially drought-vulnerable among biomes given the combination of climate risk and vegetation sensitivity [@Meir:2015trw], yet mortality events in the tropics are rarely as large as those in the temperate and boreal zones (McDowell et al.
-2018), leading to questions regarding the role of large trait-diversity and hydraulic strategies in mitigating mortality events.
-Despite this, widespread drought-induced tree mortality has occurred across the tropics for specific functional groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010] and tropical mortality rates are increasing in some regions [@Brienen.2015; @Hubau.2020].
-However, state-of-the-art dynamic global vegetation models (DGVMs) struggle to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013; @Powell.2018], as underlying mechanisms of drought-tolerance are not fully understood or quantified.
+\begin{figure}[htp]
+  \centering Figure
+  \caption{Call this figure 1.}
+  \label{fig1}
+\end{figure}
 
-\
-    Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation to stem water storage, deep water access, increased root production, hydraulic redistribution and embolism resistance.
-Deep water access is arguably the most long-standing, yet under-studied mechanism.
-This mechanism assumes that plant available water varies with depth, and that the rooting depths of co-located trees also vary, by species and size [@Meinzer:1999; @Chitra-Tarak.2018; @Brum.2019].
-Consequently, trees in the same forest may differ in their experience during a drought, and thus in their growth and mortality responses [@Chitra-Tarak.2018].
-A key bottleneck in community-wide testing of this mechanism has been a lack of data in both trees' rooting or water-sourcing depths and plant available soil water at those depths.
-A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the world’s more than 60 thousand tree species [@Fan:2017; @Beech:2017].
-Further, only a small fraction of those are tropical, even though more than 90% of the world’s tree diversity resides in the tropics [@Slik:2015].
-Use of stable isotopes of water as a tracer provides an indirect measure of water sourcing depths by matching the isotopic value in xylem water to variations in soil pore water with depth.
-However, such data are rare [@Evaristo:2016] and their interpretation is an active area of research given unresolved issues pertaining to soil heterogeneity and the correspondence of soil water extract in the laboratory to actual plant available water [@Adams.2020; @Bowers.2020; @Deurwaerder.2020].
-DNA barcoding of roots [@Jones.2011] may be used to to estimate species-specific rooting depths or rooting profiles, although extent of sampling required may be cost-prohibitive and DNA barcode libraries for tropical forests are still under development.
-In general, community-scale data collection for rooting or water-sourcing depths is challenging for species-rich tropical forests.
+S\@ref(fig:other-erd-mods)). 
+S\@ref(fig:erd-mort-census)). 
+\appendix
+\counterwithin{figure}{section}
 
-\
-    Trade-offs in plant 'traits' shape how plants make a living in their environment leading to alternative ecological strategies.
-Characterizing the essential constraints necessary to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010; @Bartlett.2019] entails identifying the topography of trait trade-offs in different environments, and how these relate to demographic rates (growth, recruitment and mortality).
-Significant efforts have been invested into identifying and linking universal drought indices, above-ground traits, and demographic rates.
-These efforts have resulted in identification of correlations between aboveground hydraulic traits such as vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013, @Venturas.2016].
-Deep roots may mitigate hydraulic vulnerability and mortality-risk, yet plants' realized water environments, specifically their ability to acquire water via differing rooting depths, are rarely studied, especially in the tropical region.
-We tested how rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tree species of a tropical forest over 35 years that witnessed droughts of various intensity and duration.
+\section{Figures}
 
-\
-    We used a locally parameterized and calibrated 1-D hydrological water balance model to obtain daily dynamics of soil water potential by depth in the whole-soil column and then applied a novel inverse model to estimate species effective rooting depths (ERD), defined as the depth at which soil water potential best explained observed dynamics of species-specific growth for large trees of `r length(erd.sp)` dominant species (Fig.
-\@ref(fig:schematic)).
-This model incorporated the impact of atmospheric drought on growth and was constrained by above-ground hydraulic traits, namely, species-specific leaf vulnerability curves, by leveraging existing data for 8 species and developing trait-based proxies for the rest.
-We evaluated ERD against existing data on plant and soil water isotopes and tested coordination of ERD with hydraulic traits and mortality rates.
-Specifically, we tested an overarching hypothesis that inter-specific variation in trees' effective rooting depths (ERD) enables increasingly reliable water availability with depth and thus mitigation of hydraulic vulnerability.
-Our hypotheses are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallow rooted species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deep rooted species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallow rooted species may have higher mortality rates as a result of higher variability water availability in shallow soil layers.
-To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.
-
-Fig. \@ref(fig:schematic) here.
-Table \@ref(tab:hypo) here.
-
-# Materials and Methods
-## Study site description
-\
-We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and data on stable water isotopes on tree species and soil from an independent study of tree water-sourcing depths [@Meinzer:1999].
-The 1,560ha island was formed after the river Chagres was dammed in 1914 to create Gatun Lake and complete the Panama Canal.
-The entire island is forested and classified as a tropical moist forest in the Holdridge Life Zone system.
-Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014].
-Rainfall at BCI is seasonal with a mean annual rainfall of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a four-month long dry season from mid-December through mid-April ($\le$ 100 mm rain month^-1^).
-Among the tree *species* found in the forest canopy, 28% are identified to be at least partly deciduous during the dry season, but this amounts to a maximum of 6.3% *trees* in the canopy that are deciduous, and the percentage of total forest crown area that is deciduous is estimated at 9.7% [@Condit:2000].
-
-\
-In the 50 ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm dbh size class [@Condit.2017].
-This inventory represents 321 woody species across the eight censuses.
-Aside from a 2ha patch of ~100-year-old secondary forest at the north-eastern corner of the plot the remaining 48ha represent old growth forest that was never clear cut for agriculture.
-The plot (1000 m $\times$ 500 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001].
-The plot is situated in the Conrad catchment.
-Soils of the plot are homogeneous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot [@Baillie.2017].
-The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil.
-Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding.
-Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere [@Croat.1978; @Leigh.1982; @Gentry.1990].
-
-## Model ELM-FATES for soil matric potentials
-
-### Model description
-\
-To estimate dynamics of water availability by depth over the forest’s rooting zone, we used the Energy Exascale Earth System Land Model (ELM), coupled with the Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES).
-ELM is a land-model that, among many features, simulates the physics and conservative dynamics of water, energy and carbon fluxes.
-In particular, soil hydrologic fluxes are vertically resolved among discrete soil layers (1-D) [@Bonan:2013].
-FATES is a community-based, open-source model used for studying climate-vegetation interactions.
-FATES is a vegetation demography model, with a size-structured group of plants (cohorts) and successional trajectory-based patches based on the ecosystem demography [@Moorcroft.2001] approach.
-FATES couples to ELM by a common interface of water and carbon fluxes.
-Detailed descriptions for ELM and FATES can be found elsewhere [@Fisher.2010; @Fisher.2015; @Koven:2020].
-
-\
-ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\Delta W_{can}$, surface water $\Delta W_{sfc}$, soil water $\Delta W_{liq, i}$ at depth z~i~ and water in an unconfined aquifer $\Delta W_{a}$ (omitting processes relevant to snow, wetlands or lakes), and conservation of these terms is expressed as follows:
-
-\begin{equation}
-\begin{split}
-\Delta W_{can} + \Delta W_{sfc} + \sum_{i=1}^{15} \Delta W_{liq,i} + \Delta W_{a} = \left(q_{rain} - E_{v} - E_{g} - q_{over} - q_{h2osfc} - q_{drai}\right)\Delta t
-\end{split}
-(\#eq:water-balance)
-\end{equation}
-
-where, $q_{rain}$ is rainfall, $E_{v}$ is evapotranspiration (ET) of the forest, $E_{g}$ is ground evaporation, $q_{over}$ is surface runoff, $q_{h2osfc}$ is runoff from surface water storage, $q_{drai}$ is sub-surface drainage (all in mm H~2~O) and $\Delta t$ is the time-step (typically, 1 hour).
-The soil profile is discretized into 15 exponentially distributed soil layers with depth z $\simeq$ 0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13, 21 and 35 (m).
-
-### Model parameterization
-\
-To parameterize ELM-FATES, we identified eleven parameters relevant for Eq.
-\@ref(eq:water-balance), determined their ranges based on literature for the study site or for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling (LHS) of this global parameter space.
-In the Conrad catchment, existing data on soil hydraulic conductivity for the depths of 12.5 and 60 cm showed a decline of conductivity by depth (@Kinner:2004; Table \@ref(tab:param)).
-We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths < 12.5 cm and > 60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig.
-S\@ref(fig:ksat-profiles) [Need to fix SI Figs.
-numbering to begin at 1]).
-For greater accuracy, instead of the ELM default soil texture-based pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content vs.
-$\Psi_{soil}$ for depths of 0.15, 0.4 and 1 m [@Kupers:2019].
-We converted gravimetric water content to volumetric (VWC, $\theta$) using a bulk density value of of 0.8 g cm^-3^ measured at a point location (see below).
-A @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences in water retention among depths were nominal,
-
-\begin{equation}
-\Psi_{soil} = \Psi_e\left(\frac{\theta}{\theta_{s}}\right)^{-b}
-(\#eq:soil-ret-curve)
-\end{equation}
-where $\Psi_e$ is the air-entry matric potential, $\theta_{s}$ is the saturated volumetric water content, and $b$, an index for soil pore-size distribution, respectively.
-$\Psi_e$, $\theta_{s}$ and $b$ taht best-fit the data were 200 (mm H~2~O), 0.51 (cm^3^ cm^-3^) and 10 (unitless), respectively.
-
-\
-We ran ELM-FATES for the 5000-member parameter ensemble with hourly climate drivers from 1985-2018 [@Faybishenko.2020] initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode.
-As our key interest here was in deriving soil water availability, we ran ELM-FATES in a lower-complexity configuration: static stand structure, and with a single plant functional type (PFT) of evergreen trees.
-The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of a dry-deciduous PFT did not significantly alter results (not shown).
-Holding the stand structure static rather than allowing it to emerge from physiologically-mediated competition between trees removes a set of internal model feedbacks and thus enables a cleaner experimental design where the only differences between model simulations are the direct physiological effects of perturbed model parameters.
-
-### Model calibration
-\
-We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq.
-\@ref(eq:water-balance), namely, (i) evapotranspiration ($ET \simeq E_{v} + E_{g}$) from the eddy flux tower (2012-2017; see [SI](#SIData.matteo)), (ii) stream discharge from the Conrad catchment ($q_{run} \simeq q_{over} + q_{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC, $\theta_{z} \simeq W_{liq, z}$) from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we obtained with a vertical TDR probe over 0-15 cm depth ($\theta_{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m (1299 samples in total covering all soil types and habitats; see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\theta_{0.15 m}$, $\theta_{0.4m}$ and $\theta_{1 m}$ refer to plot-wide averages).
-
-For ELM-FATES calibration we calculated an objective function for each of the 5000-member ensemble by equally weighting standardized RMSE between observations and simulations across all fluxes and states.
-We describe the objective function $O$ as :
-
-\begin{align*}
-\overline{O} = \frac{1}{6}\sum \left(RMSE^{*}(ET,~E_{v} + E_{g}) + RMSE^{*}(q_{run},~q_{over} + q_{drai}) + RMSE^{*}(\theta_{0-0.15m}, {W_{liq,~\overline{z}}})\\
-+ RMSE^{*}(\theta_{0.15m}, W_{liq,~`r soil.depths[4]`m}) + RMSE^{*}(\theta_{0.4m}, W_{liq,~`r soil.depths[6]`m}) + RMSE^{*}(\theta_{1 m}, W_{liq,~`r soil.depths[8]`m})\right)
-(\#eq:obj-func)
-\end{align*}
-
-based on monthly sums for both $ET$ and $q_{run}$ and daily averages for $\theta$s.
-$W_{liq,~\overline{z}}$ is the average of daily $W_{liq,~z}$ for $z$ $\in$ {`r paste(soil.depths[1:4], collapse=", ")`} m.
-Superscript * indicates that RMSE is standardised between 0 and 1 across the 5000 simulations.
-
-Preliminary analyses showed that minimising the objective function $\overline{O}$, that is, simultaneously fitting the model to all states and fluxes did not capture the soil moisture dynamics at 1 m depth well, our key priority for model calibration.
-To ensure model fit with soil moisture dynamics at 1 m depth, we first chose 300 parameter ensemble members that minimised RMSE between observations and simulations for VWC at the depth of 1m ($RMSE^{*}(\theta_{1 m}, W_{liq,~`r soil.depths[8]`m})$ < 0.2), and among those 300 ensemble members we chose 100 members with the lowest values for the objective function $\overline{O}$.
-See Table \@ref(tab:param) for the ranges of best-fit values for different parameters.
-
-### Model evaluation
-\
-We evaluated ELM-FATES by calculating RMSE between simulated and observed long-term daily VWC for the depths of 0.1, 0.4 and 1 m (2016-2018), baesd on a dataset we had left-out during calibration.
-We obtained these observations from three horizontal time domain reflectometry probes at the depths of 0.1, 0.4 and 1 m at a point location near the eddy flux tower (see [SI](#SIData.VWC)).
-
-\
-We ran ELM-FATES with the chosen 100 ensemble members from 1985-2018, with the first 5 years used for model spin-up.
-
-## Model for effective rooting depth (ERD)
-\
-To estimate effective rooting depth, we tested alternate models for growth (Eqs.
-\@ref(eq:M-chosen) and those in SI), which use effective rooting depth as an input.
-There are many factors that could limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019].
-In this study, we used maximum diurnal leaf conductance ($K_{leaf}$) approximated using rhizosphere soil water potential ($\Psi_{soil}$) as a proxy for the soil water limiting factor on growth.
-In our models $K_{leaf}$ therefore incorporates the effects of rooting depth and rhizosphere water potential.
-For each species' growth dynamics across census intervals was regressed against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1.
-Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting.
-At our study site, due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season, the impact of VPD may as well be a combined impact of not only temperature and humidity, but also radiation.
-
-For the chosen model (see Model validation and selection below), daily average growth for species $s$, $G_{s,~t~|~z}$, in the census interval $t$ for depth $z$ and for a given hydrological realisation $h$ is described as follows:
-
-\begin{equation}
-G_{s,~t~|~z,~h} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} (K^{*}_{leaf,~s,~i~|~z,~h} + VPD^{*}_{i}) LAI^{*}_{s,~doy~|~i}) + \epsilon_{t~|~z,~h}
-(\#eq:M-chosen)
-\end{equation}
-
-where n~t~ be the total number of days in census interval *t*, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is a normally distributed model error term.
-See SI for equations for all the alternate models tested.
-
-As VPD may have a non-linear relationship with growth, with growth increasing for the lower range of VPD, and decreasing for the higher range of VPD due to stomatal closure (reference), we used a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see Model parametarisation) to represent the impact of VPD on growth and used GPP standardised between 0 and 1 (hereafter, VPD*) in Eq.
-\@ref(eq:M-chosen).
-
-For each species, we estimated diurnal maximum value of $K_{leaf,~s,~i~|~z,~h}$ for a species $s$ by substituting $\Psi_{soil,~i~|~z,~h}$ in the leaf vulnerability curves defined below in Eq.
-\@ref(eq:vc-exp) using species-specific parameters $A_{s}$ and $B_{s}$ as:
-
-\begin{equation}
-K_{leaf,~s,~i~|~z,~h} = A_{s}e^{-B_{s}\Psi_{soil,~i~|~z,~h}}
-(\#eq:k-soil)
-\end{equation}
-
-By varying ‘z’, we estimated species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015).
-Using Eq.
-\@ref(eq:M-chosen), for a species $s$, we obtained Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $r$ for each soil depth $z$ for each of the 100 best-fit hydrological realizations.
-For each hydrological realization we retained only one depth that corresponded to the least AIC score (the best-fit) [@Chonggang, we could have retained multiple depths with non-differentiable AICs], and within those depths retained the ones that had Pearson's $r$ >= 0.71 (or, $r^2$ >= 0.5) as a valid set.
-For a species $s$, we defined effective rooting depth (ERD) as the median of these set of depths, described as :
-
-\begin{equation}
-ERD_{s} = \underset{\forall \, h \, \in \, H}{median} \Big\{z,h\,\Big|\, \Big(AIC_{s~|~z,~h} = \underset{\forall \, z \,\in\,Z}{min}(AIC_{s~|~z,~h})\Big) \cap \Big(r_{s~|~z,~h} \ge 0.71\Big)\Big\}
-(\#eq:erd)
-\end{equation} 
-where $H$ $\in$ {1, .., 100}, refer to the 100 hydrological realisations and $Z$ $\in$ {`r paste(soil.depths, collapse=", ")`} m, covering the range of soil depths in ELM-FATES after above parametarision.
-
-### Model parameterization
-
-#### Growth data
-\
-For growth estimates, to minimize the effect of light variation among trees, we selected only large trees ($\ge$ 30 cm dbh, the diameter at breast height) and also species whose maximum height was $\ge$ 30 m (hereafter, canopy species) and thus are likely to be fully exposed to the sun.
-We selected trees for which the height of the diameter measurements were unchanged across all censuses (1990, 1995, 2000, 2005, 2010, 2015).
-For each individual tree (*i*), we calculated absolute growth rates ($G_{i,t}$; cm yr^-1^) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses, divided by the inter-census interval based on the exact date of measurements.
-We eliminated positive outliers as any growth rates amounting to > 75 mm yr^-1^, as this was the highest rate observed with confidence for fastest growing species [@Condit.2017].
-Negative outliers were removed based on known measurement error, $\sigma_{e} = 0.006214d + 0.903$, where *d* is dbh.
-Any stem whose later dbh measurement fell < 4$\sigma_{e}$ the earlier measurement was removed [@Condit.2017].
-As tree growth rates vary with size [@Muller-Landau:2006], to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($Resid_{i,t} = G_{i,t} - f(d_{i, t})$).
-The diameter model of growth ($f(d_{i, t})$), a B-spline based polynomial equation with 5 degrees of freedom, was fitted to growth rates from all trees pooled together as there was insufficient species-specific data for large trees (Fig.
-S\@ref(fig:dbh-model)).
-We retained only those trees that had $Resid_{i,t}$ estimates for all census intervals (total trees = `r sum(g.n$n)`) and standardized (centered and scaled) individual tree time series of $Resid_{i,t}$.
-We then obtained `r length(erd.sp)` species-level time-series of medians of $Resid_{i,t}$ ($G_{t}$).
-These were obtained for only those species with complete records on at least three trees (median *n* = `r median(g.n$n)`, max *n* = `r max(g.n$n)` trees per species).
-
-#### Leaf hydraulic vulnerability curves
-\
-We obtained leaf hydraulic vulnerability curves ($K_{leaf}$ vs.
-$\Psi_{leaf}$; see Table \@ref(tab:symbols) for trait symbols and their definitions) for adult trees of 21 common species from Zailaa et al.
-(to be submitted; see SI for brief description of methods) described as:
-
-\begin{equation}
-K_{leaf,~s} = A_{s}e^{-B_{s}\Psi_{leaf_{s}}} + error
-(\#eq:vc-exp)
-\end{equation}
-where $A$ and $B$ are fitted species-specific parameters.
-These 21 species included `r sp.n.data` of the `r length(erd.sp)` species selected for ERD estimation.
-For the rest `r sp.n.model` species, we obtained vulnerability curves using trait-based proxies.
-We identified scaling relationships between fitted parameters $A$ and $B$ in Eq.
-\@ref(eq:vc-exp) and two traits, namely, WSG, the wood specific gravity after drying at 100^$\\circ$^C and LMA, the leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules for leaves receiving direct sunlight (@Wright:2010; see SI for details).
-We fitted polynomial equations described as,
-
-\begin{equation}
-A_{s} = `r acf[1]` `r acf[2]`B_{s} `r acf[3]`B_{s}^2 + `r acf[4]`B_{s}LMA_{s} + `r acf[5]`LMA_{s} `r acf.6`LMA_{s}^2 + error
-(\#eq:a-mod)
-\end{equation}
-
-\begin{equation}
-B_{s} = `r bcf[1]` `r bcf[2]`WSG_{s} + `r bcf[3]`WSG_{s}^2 + `r bcf[4]`WSG_{s}LMA_{s} + `r bcf[5]`LMA_{s} `r bcf[6]`LMA_{s}^2 + error
-(\#eq:b-mod)
-\end{equation}
-
-As these fits explained a large proportion of variation in parameters $A$ and $B$ (see results), we sequentially used Eq.
-\@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r sp.n.model` ERD species without direct data and estimated leaf hydraulic vulnerability curves using Eq.
-\#ref(eq:vc-exp).
-For all the 31 ERD species we estimated species $\Psi_{50,~leaf}$ based on their vulnerability curves (see Table \@ref(tab:symbols) for definition).
-
-Table \@ref(tab:symbols) here.
-
-#### Leaf Area Index (LAI)
-\
-To explore the effect of species seasonality in leaf phenology on growth, we assumed species-specific seasonal curves for leaf area index ($LAI$, range 0-1, unitless).
-For evergreen species we assumed $LAI$ = 1 throughout the year, whereas for non-evergreen species (various deciduous classes), we assumed $LAI$ = 1 during the wet season (DOY = 150 to 300) given the high forest-level leaf area index for this period (not shown), and obtained an estimate of $LAI$ for the rest of the year as follows.
-We interpolated species-specific weekly leaf-fall data collected at two similar sites in BCI (1985-2020) to a daily estimate, scaled to express daily leaf fall as a fraction of the total leaf fall in a given year and obtained an average leaf fall fraction by DOY estimate across years ($LeafFall$, range 0-1, unitless).
-Thus for non-evergreen species, we cumulatively deducted $LeafFall$ from $LAI$ = 1 beginning DOY = 300 and obtained a $LAI$ by DOY curve (Fig.
-S\@ref(fig:leaf-cover)).
-We obtained species-specific parameters $A$ & $B$ for Eq.
-\@ref(eq:k-soil) either from data when available, or a model based on $LMA$ or $WSG$ as described in the earlier section, and used K standardized between 0 and 1 within species ($K^*$) in Eq.
-\@ref(eq:M-chosen).
-
-#### Vapor Pressure Deficit (VPD)
-\
-We fitted a polynomial curve between mean midday values (11:00 to 15:00 hrs) of gross primary productivity (gC m^2^ day^-1^) and VPD (kPa) observations we collected from the eddy flux tower located 75 m from the 50-ha plot (for any day *i*, $GPP_i = `r vpd.cf[1]` + `r vpd.cf[2]`VPD_i + `r vpd.cf[3]`VPD_i^2 + error$; 2012-2017; Fig.
-S\@ref(fig:gpp-vpd)) and used $GPP$ standardized between 0 and 1 as $VPD^*$ in Eq.
-\@ref(eq:M-chosen)).
-
-### Model validation and selection
-\
-To validate the ERD models we used data for xylem sap $\delta^2H$ (permil) as an independent observation of ERD.
-For each model of ERD (Eq.
-\@ref(eq:M-chosen) and Eqs.
-in SI), estimated species ERD was regressed against species xylem sap $\delta^2H$ (permil).
-We used xylem sap $\delta^2H$ values given in @Meinzer:2001 for the dry-season of March 1997 as this period showed largest seasonal divergence among species, with a corresponding gradient of $\delta^2H$ in the soil (Fig.
-S\@ref(fig:iso-data)).
-We removed species that are likely to have no leaves in March as xylem sap $\delta^2H$ of these species may not be linked with water sourced at the time of measurement and thus may not be comparable with other species that had leaves.
-We also removed *Guapira standleyana*, whose xylem sap isotope value was not well constrained by the soil water isotope depth profile.
-We were thus left with 7 species with data on $\delta^2H$ as well as modelled estimates of ERD.
-
-## ERD coordination with above-ground hydraulic traits
-\
-To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD against maximum stem hydraulic conductance ($K_{max, stem}$), leaf turgor loss point ($\Psi_{tlp}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$).
-See [SI](#SIData.brett) for data collection and estimation of these variables.
-
-## Mortality analyses and species-specific drought experience
-\
-We calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$ 30 m) and average abundance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; *n* = `r length(erd.mrate.sp)`).
-For each of the seven census intervals ($t$) in the 35-yr record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (% yr^-1^) for two successive censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{time}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees of a species present in cen~1~ and dead in cen~2~, respectively, in the 50 ha plot and $time$ is the duration based on mean dates of cen~1~ and cen~2~.
-For each species we calculated mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{t=1}^{6} {Mort_{t}}$.
-
-    We regressed species ERD against species ($Mort_{mean}$).
-To assess whether deeper ERD species escape the experience of significant soil drought, we calculated for each speceis the proportion of days in the simulation period (1990-2015) during which soil water potential in the soil zone matching species ERD ($\Psi_{soil,~z = ERD}$) were below species $\Psi_{crit}$.
-$\Psi_{crit}$ was defined as $\Psi_{50,~leaf}$.
-
-\
-All statistical analyses were conducted in R statistical environment (version 3.5.3, @R:2019).
-
-# Results
-## Soil water dynamics by depth
-\
-The ELM-FATES simulations with the chosen best-fit parameter ensembles (Table \@ref(tab:param)) successfully captured the dynamics of soil moisture by depth (Fig.
-\@ref(fig:hydro-calib) A & Fig.
-\@ref(fig:hydro-calib)B), including the observations from the horizontal TDR probes that were not included in the calibration.
-These simulations also captured the dynamics in stream discharge (Fig.
-\@ref(fig:hydro-calib)C) and to a moderate extent the dynamics in evapotranspiration (Fig.
-\@ref(fig:hydro-calib)D).
-The reduction in parameter range in the best-fit ensembles compared to the global ranges tested showed that the above calibration was sensitive to the following parameters: the Ball-Berry stomatal slope parameter, the ELM rooting distribution parameter that regulates the depth of the rooting profile, soil hydraulic conductivity profile especially at depth, and our adjustment factor to soil hydraulic conductivity to account for macroporosity and direct flow paths (Table \@ref(tab:param)).
-
-Fig. \@ref(fig:hydro-calib) here.
-
-  Soil Water Potential dynamics ($\Psi_{soil,~z}$) by depth (z = 0.1 through 13) obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of such extreme years, drought intensity at depth, drought seasonality and duration (Fig.
-S\@ref(fig:swp-droughts)).
-Nevertheless, across the simulation period (1990-2015) $\Psi_{soil,~z}$ remained above -0.5 MPa from depths 1.7 to 13 m (Fig.
-S\@ref(fig:swp-all-depths)).
-
-Fig. \@ref(fig:swp-droughts) here.
-
-## Leaf hydraulic vulnerability curves
-\
-For the leaf vulnerability curve parameters $A$ and $B$ fitted to the observed data for $K_{leaf}$ vs.
-$\Psi_{leaf}$, $WSG$ and parameter $B$ explained a large proportion of variation in parameter $A$ (Eq.
-\@ref(eq:a-mod); *Adj.
-R*^2^ = `r afit.r2`; *P* `r  afit.p`; Fig.
-\@ref(fig:ab-fit)).
-$WSG$ and $LMA$ explained a large proportion of variation in parameter $B$ (Eq.
-\@ref(eq:b-mod); Adj.
-*R*^2^ = `r bfit.r2`; *P* `r  bfit.p`; also see Fig.
-S\@ref(fig:obs-plc-curves).
-Vulnerability curve parameters were predicted for `r sp.n.model` out of `r length(unique(erd.iso$sp))` ERD species without direct data using Eqs.
-\@ref(eq:a-mod)-\@ref(eq:b-mod) (Table S\@ref(tab:ab-param), Fig.
-S\@ref(fig:obs-comm-plc-curves)).
-Fig. \@ref(fig:ab-fit) here.
-
-## Effective rooting depths (ERD)
-\
-ERD estimated from the model with K~leaf~, LAI and an additive effect of VPD as explanatory variables (Eq.
-\@ref(eq:M2)) was strongly related to the measured isotopic signal ($\delta^2H$), that is, with highest goodness-of-fit (*R*^2^ = 0.65; *P* = 0.027; Fig.
-\@ref(fig:erd-iso)).
-The model with K~leaf~, but without LAI and an additive effect of VPD showed no species with sufficient species-level goodness-of-fit.
-ERD from all the other models showed a positive, but non-significant relationship with $\delta^2H$ at $\alpha$ = 0.05 (Fig.
-S\@ref(fig:other-erd-mods)).
-For the rest of the analyses we therefore used ERD estimates from the best-fit model (Eq.
-\@ref(eq:M-chosen)).
-Notably, two species, *L.
-seemannii* and *T.
-tuberculata*, that @Meinzer:1999 found to have xylem sap $\delta^2H$ values between soil water and groundwater, suggesting that these species sourced most of the water from depths > 1 m, were also identified by our model with ERD > 1 m (2.9 m for both species; Fig.
-\@ref(fig:erd-iso)).
-
-Fig. \@ref(fig:erd-iso) here.
-
-\
-  In this forest, modeled effective rooting depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from `r min(df.erd.to.plot$depth)` m to `r max(df.erd.to.plot$depth)` m (Fig.
-\@ref(fig:sp-erd)).
-
-Fig. \@ref(fig:sp-erd) here.
-
-## Coordination of ERD with other hydraulic traits
-
-  For the subset of 8 species with ERD for which hydraulic traits data were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max,~stem}$, Spearmans' *r* = `r cor.k`, *P* = `r p.k` at $\alpha$ = 0.05; Fig.
-\@ref(fig:erd-hyd)A), less negative leaf $\Psi_{tlp}$ (*r* = `r cor.tlp`, *P* = `r p.tlp`; Fig.
-\@ref(fig:erd-hyd)B), less negative $\Psi_{88,~stem}$ and thus greater vulnerability to embolism from cavitation (*r* = `r cor.stem.88`, *P* = `r p.stem.88`; Fig.
-\@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$; *r* = `r cor.hsm`, *P* = `r p.hsm`; Fig.
-\@ref(fig:erd-hyd)D).
-
-Fig. \@ref(fig:erd-hyd) here.
-
-## Effective rooting depths, experienced drought and mortality
-\
-Species with deeper ERD showed lower mean mortality rates across the 34-yr record (1982-2015; *R*^2^ = 0.11, *P* = 0.075, *n* = `r length(erd.mrate.sp)`; Fig.
-\@ref(fig:erd-mort)).
-For five out of seven census intervals, species with shallower ERD had significantly higher mortality anomalies than deeper ERD species (Fig.
-S\@ref(fig:erd-mort-census)).
-Analysis of time spent under significant soil drought showed that the proportion of days species $\Psi_{ERD}$ was more negative than species $\Psi_{crit}$ decreased with species ERD, that is the shallower the ERD, the greater proportion of days $\Psi_{ERD}$ < $\Psi_{crit}$ (Fig.
-\@ref(fig:erd-time)).
-Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25-yr period for which soil water potential $\Psi_{soil,~z}$ was modeled (1990-2015), as soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`;  median, 95% CI) through all the dry seasons and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).
-
-Fig. \@ref(fig:erd-mort) here.
-Fig. \@ref(fig:erd-time) here.
-
-# Discussion
-\
-Our analyses revealed that co-occurring large canopy tree species with deeper effective rooting depths (ERD) were associated with higher hydraulic efficiency, but not safety.
-Species with deeper ERD showed higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species.
-Nevertheless, deeper ERD species showed lower mortality rates than shallower ERD species over the 35-yr record (1982-2015).
-Analysis of time spent below ($\Psi_{crit}$) in the soil zone matching ERD $\Psi_{soil,~z = ERD}$ showed that this was likely due to decreasing realized risk of hydraulic failure with increasing access to water for species with deeper ERD.
-We thus demonstrate, for the first time, a link between realized drought experience, hydraulic architecture and long-term mortality outcomes for co-occurring adult tree species.
-
-### Realised drought risk integral to assessing drought-vulnerability
-\
-Our study brings attention to the need for assessing drought-sensitivity in terms of species realized drought risk.
-We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins, that is, traits that are commonly identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through a study period of greater recurrence of drought.
-These results are consistent with the two studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016].
-@Johnson:2018fad found that mortality decreased with deep rooting even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi_{88,~stem}$) and had narrower safety-margins ($\Psi_{min} - \Psi_{88,~stem}$).
-@Nardini:2015 found that a species sourcing deep water in a Karst system showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi_{88,~stem}$ and narrower safety margin compared to two species sourcing shallow-water.
-
-\
-Our result of lower mortality in deep rooted trees contrasts with the finding of @Chitra-Tarak.2018 in which species identified by their inverse ERD model as effectively deep rooted died *more* in a prolonged, extreme drought.
-Their hydrological model revealed that the multi-year drought exhausted the deep soil and even bedrock water availablity.
-In contrast, our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period, $\Psi_{soil}$ below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and droughts, that is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).
-The difference in our results are likely attributable to site differences in the precipitation regime.
-In their dry forest site (Mudumalai National Park, India) precipitation is low: 1095 $\pm$ 296 mm (mean $\pm$ SD) compared to `r round(rain.man.stats$rain)` $\pm$ `r round(rain.man.stats$rain.sd)` mm (`r min(clim$Year)`-`r max(clim$Year)`) at BCI, and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (Penman-Monteith PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (Penman-Monteith PET = `r round(mean(clim$pet.PM))` $\pm$ `r round(sd(clim$pet.PM))` mm;`r min(clim$Year)`-`r max(clim$Year)`; this study, @Guo.2016), and thus at BCI deep soil layers are effectively recharged annually (Fig.
-S\@ref(fig:swp-all-depths); @Chitra-Tarak.2018).
-The variation in mortality explained by ERD in our study is low (*R*^2^ = 0.11, *P* = 0.075), and other causes may explain observed mortality [@McDowell.2018].
-
-### Effective rooting depths at the tree community level
-\
-By estimating ERD for large trees of `r length(erd.sp)` canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species-rich tropical forests.
-Our chosen ERD model is an advance over the only other inverse model of ERD (@Chitra-Tarak.2018; *n* = 12 species) as we utilize physiologically meaningful representation, fewer parameters and corroboration with tree and soil data on $\delta^2H$.
-
-\
-We benefited from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA and WSG) and parameters of leaf vulnerability (or percent loss of conductance, (PLC)) curves for a few species (*n* = `r length(sp.exp.param$sp)`), a 'hard' trait.
-This allowed us to scale-up the leaf vulnerability curves at the community scale (*n* = `r length(ab.table$Species)` species, see Table S\@ref(tab:ab-param) for model predicted parameters).
-We find that leaf conductance declined more rapidly with declining $\Psi_{leaf}$ for species with higher values of WSG and LMA (Fig.
-S\@ref(fig:obs-plc-curves)).
-Whether such scaling relationships between hard and soft traits hold true for other forests, and their wider applicability, remains to be seen as data on PLC curves are limited in tropical forests, but supportive and counter examples both exist [@McCulloh.2011; @Hoffmann.2011].
-
-\ 
-Another important piece of information for understanding effective rooting depth is stable water isotopes of xylem water and below-ground sources, in the vadose zone and water table.
-Although the interpretation of these data is still under research [@Deurwaerder.2020], they are critical for model calibration or validation.
-The model in @Chitra-Tarak.2018 were not corroborated with such data as existing methods were inadequate to extract water bound to the saprolite found in their dry forest study site (Jean Riotte, pers.
-comm.).
-The model of below-ground water-availability in this study (ELM-FATES) is far more advanced compared to that used by @Chitra-Tarak.2018 and was calibrated with data on long-term soil moisture dynamics at a point location (0-0.15m) and plot-wide measurements by depth (0.15, 0.4 and 1 m), as well as, evapotranspiration and stream discharge.
-@Chitra-Tarak.2018 was calibrated only on stream discharge and water table levels.
-We lacked data on water table levels which would have been useful to constrain water availability dynamics in the unobserved soil layers and the spatial variability within the plot.
-As in @Chitra-Tarak.2018, we also interpret our ERD estimates as revealing relative differences among species' effective rooting depths rather than in absolute terms.
-
-### Future directions for model development
-\
-The trade-offs that we identify between above- and below-ground traits, the water environment vertically below-ground and mortality rates are important for representing diversity in Dynamic Global Vegetation Models especially in species-rich forests.
-The hydraulic efficiency-safety trade-off between high values of maximum stem conductivity (high $K_{max,~stem}$) and cavitation-resistance (low $\Psi_{50,~stem}$ or $\Psi_{88,~stem}$) is hypothesized although community-wide examples are rare [@Hacke:2006; @Gleason.2016].
-We report such a trade-off (n = 26 species) at our study site stem level ().
-We also found that LMA and WSG were strong predictors of PLC curves for leaves, but not for stems (not shown).
-WSG alone explained `r round(a.wsg.r2*100,0)` % of variation in parameter $A$ ($A = `r a.wsg.cf[1]` `r a.wsg.cf[2]`B + `r a.wsg.cf[3]`B^2 + error$; *P* `r  a.wsg.p`) and `r round(b.wsg.r2*100,0)` % of variation in $B$ ($B = `r b.wsg.cf[1]` `r b.wsg.cf[2]`B + `r b.wsg.cf[3]`B^2 + error$; *P* `r  b.wsg.p`).
-Parameter $B$ alone explained `r round(a.b.r2*100,0)` % of variation in $A$ (*P* `r  a.b.p`), while $LMA$ alone was weakly related to $A$ and was not related to $B$ (not shown).
-The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits, including hydraulic vulnerability, in order to better represent the functional diversity of tropical forests, and their trajectories into the future.
-
-\
-Our model estimated ERD as a single value, unlike the dynamic ERD modeled by @Chitra-Tarak.2018, but future model developments should investigate seasonal dynamics of ERD, using corresponding calibration data on isotopically determined water-sourcing depths, as species are expected to shift to deeper soil water sources as shallow soils dry out [@Meinzer:1999].
-Our inverse ERD model was parameterized on 5-year growth data, with five data points over a 25-yr period, which decouples climate events and demographic outcomes.
-Future studies could better constrain the model with higher resolution growth data such as those from dendrometer bands.
-This is especially important for modeling seasonally dynamic ERD.
-However, trees' stem diameter change is a sum total of not just growth, the addition of wood, but also changes due to hydrostatic stem flexing (xylem, phloem and bark hydration), the importance of which increases with temporal resolution [@Chitra-Tarak.2016; @Chitra-Tarak.2015].
-Use of dendrometer data would thus entail accounting for stem hydrostatic flexing [@Zweifel:2018].
-Three of our exploratory ERD models included seasonality in leaf area fraction, but we did not select them as they worsened the fit with growth data and evaluation data ($\delta^2H$, Fig.
-S\@ref(fig:other-erd-mods)).
-We interpret this result as that VPD and PLC curves may be adequate to explain growth, but also acknowledge that our estimates of seasonality of LAI is a tentative estimate combining leaf fall data and assumption of leaf presence during the wet season, and omits inter-annual variation.
-Future studies could make use of data from new technologies such as drone based monitoring of species-specific leaf area fraction in species-rich tropical forests [@Park.2019].
-In forests such as BCI where understory tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes, especially from a few growth data points, may need accounting for variation in trees' light environment over time and among trees of the same size class, an issue we minimized by limiting ERD estimates to canopy trees.
-To estimate topographic variation in ERD within or across species future studies may use soil moisture dynamics from a distributed hydrological model [e.g.
-@Schwantes:2018].
-We note that data availability on soil water retention curve ($\Psi_{soil}$ vs.
-VWC), hydraulic conductivity ($K_{soil}$) by depth, soil moisture by depth, stream discharge and evapotranspiration were important for effective calibration of our 1-D hydrological model.
-We recommend wide-spread data collection of these variables as well as water-table levels in forest-inventory sites to allow for estimation of trees' water environments.
-Finally, developing an inverse ERD model that accounts for movement of water within the soil-plant-atmosphere-continuum (SPAC) would be a step change from the current statistical models of ERD (this study and @Chitra-Tarak.2018).
-Although data needs for parameterizing such models [@Christoffersen:2016kd2d; @Sperry.1998] could be greater (e.g.
-retention curves for roots), which limits immediate community-wide application, such models hold greater promise to improve our understanding of plant physiology [e.g.
-see @Johnson:2018fad].
-
-# Conclusions
-
-Establishing relationships between plant's environment, traits and demographic outcomes is imperative for developing a predictive plant ecology, but trees realized water environments and water-access are rarely studied.
-To our knowledge this is the first study to demonstrate a mechanistic link between trees' realized drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.
-The role of deep-water access that we find in alleviating mortality of drought-sensitive trees has important implications for our predictive understanding of tropical forest dynamics under current and future climate.
-Our community-scale framework for modeling effective rooting depths or leaf vulnerability curves indicate the possibilities in expanding the use of critical, hard-earned observations in species-rich forests towards community-scale generalizations.
-
-# Acknowledgements
-
-This research was supported as part of the Next Generation Ecosystem Experiments-Tropics, funded by the U.S.
-Department of Energy, Office of Science, Office of Biological and Environmental Research (LANL DOE Contract #7537545).
-Funding was also provided to RCT and SMM through an NSF grant (1137366).
-LR was deputed to Indo-French Cell for Water Sciences on an IRD Fellowship.
-An NSF grant to Smithsonian Institution’s ForestGEO (Dimensions 1046113) also partially supported this work via RCT and SMM.
-RF was supported by the National Center for Atmospheric Research, which is funded by the National Science Foundation.
-**Add acknowledgements for other co-authors.
-The BCI forest dynamics research project, founded by Stephen P.
-Hubbell and Robin B.
-Foster and sustained for many years by Richard Condit, is now managed by Stuart Davies, Suzanne Lao, and Rolando Perez under the ForestGEO program of the Smithsonian Tropical Research in Panama.
-Numerous organizations have provided funding, principally the US National Science Foundation, and hundreds of field workers have contributed.
-
-# Data Availability
-
-**RCT: Upload on NGET-Archive and site DOIs:
-(1) Boris to upload climate forcing data
-(2) Rutuja to upload R scripts for ELM-FATES parameter generations, ELM-FATES modified code, and all R scripts to reproduce manuscript.
-\
-Barro Colorado Forest Census Plot Data can be obtained online[@Condit.2019a; @Condit.2019b].
-
-# Captions for Tables and Figures
-
-Table \@ref(tab:hypo): Hypotheses for association between Effective Rooting Depth (ERD) and hydraulic traits.
-
-Table \@ref(tab:symbols): The symbol, definition and units of key traits used in the simulations and analyses.
-
-Figure \@ref(fig:schematic): A schematic for the methods workflow.
-
-Figure \@ref(fig:hydro-calib): ELM-FATES calibration and evaluations.
-Observations (red lines) versus simulations (blue lines) from 100 best-fit parameter ensemble runs of ELM-FATES are shown for daily Volumetric Water Content (VWC) by depth for horizontal time domain reflectometry (TDR) probes at three depths (0.1, 0.4 and 1 m; A) and a vertical probe (0-0.15 m; B); monthly stream discharge (Discharge, C) and monthly evapotranspiration from eddy flux tower (ET, D).
-Points (red) in (A) show average plot-wide observations of VWC with 95% CI (error bars).
-VWC data from the horizontal probes (lines in A) were not used for ELM-FATES calibration but evaluation.
-Values in inset are average RMSE across the 100 best-fit simulations.
-For panel A, values at the top are for plot-wide VWC and those at the bottom are for TDR probes.
-VWC is in units of cm^3^ cm^-3^, while ET and Discharge in mm month^-1^.
-
-Figure \@ref(fig:swp-droughts): Occurrence of extreme soil droughts by census intervals (horizontal panels) at three representative depths (z): 0.1, 0.6 and 1 m (vertical panels).
-Mean (black lines) and lower half of 95% distribution (grey areas) of soil water potential ($Psi_{soil,~z}$) for a given DOY and depth z are shown (same across census intervals).
-$Psi_{soil,~z}$ by DOY for a year (colored line) is only shown if for some part of that year $Psi_{soil,~z}$ declined below the 2.5 percentile distribution.
-
-Figure \@ref(fig:ab-fit): Parameters $A$ (panel A) & $B$ (panel B) for the leaf vulnerability curves that were fitted to observed data on $K_{leaf}$ vs.
-$\\Psi_{leaf}$ for `r length(sp.exp.param$sp)` species (Eq.
-\@ref(eq:vc-exp)) vs.
-those that were predicted from model based on trait-proxies.
-(Eqs.
-\@ref(eq:a-mod)-\@ref(eq:b-mod)).
-Goodness-of-fit (*R*^2^) and *P*-values are given in inset.
-
-Figure \@ref(fig:erd-iso): Modelled Effective Rooting Depth (ERD; mean $\pm$ 1SE; m) versus dry season stable isotopic concentration $\delta^2H$ (mean $\pm$ 1SE; ‰) for `r length(unique(erd.iso$sp))` canopy species from Barro Colorado Island.
-Data from [@Meinzer:1999].
-See Table  \@ref(tab:ab-param) for the full form of scientific names.
-
-Figure \@ref(fig:sp-erd): Modelled Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) for `r  length(erd.sp)` large, canopy species of Barro Colorado Island.
-
-Figure \@ref(fig:erd-iso): Matrix for pairwise correlations (circles) among Effective Rooting Depth (ERD) and hydraulic properties for `r nrow(cor.data)` canopy species found on Barro Colorado Island (Panama).
-Each circle is color coded by the Spearman's $\\rho$ with red shades for positive $\\rho$ values, while blue shades for negative values, respectively.
-P-values are given only when they are $\\ge$ 0.05.
-Units are as follows: ERD in m, K~max~ in mmol m^-1^ s^-1^ MPa^-1^), all variables with $\\Psi$ in MPa.
-
-Figure \@ref(fig:erd-mort): Mortality rate (mean $\pm$ 1SE) versus Modelled Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) for `r length(erd.mrate.sp)` canopy species found on Barro Colorado Island.
-*R*^2^ and *P*-values are given in panel insets.
-
-Figure \@ref(fig:erd-time): Modelled species Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\Psi_{soil,~z}$ was below $\Psi_{crit}$, defined as $\Psi_{80,~leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for `r length(erd.mrate.sp)` canopy species found on Barro Colorado Island.
-Each circle represents a species.
-Circles are colored and jittered vertically around the named depths (ticks) for clarity.
-
-# Tables
-
-Hypotheses for association between Effective Rooting Depth (ERD) and hydraulic traits.
-The symbol, definition and units of key traits used.
-
-```{r hypo, results = 'asis', echo = FALSE}
-knitr::kable(hypo.table, caption = '', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(hypo.table))))
-```
-
-```{r symbols, results = 'asis', echo = FALSE}
-knitr::kable(symbols.table, caption = '', booktabs = TRUE, escape = FALSE)
-```
-
-# Figures
-
-```{r schematic, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-schematic}")}
-knitr::include_graphics(file.path("figures/schematic.png"))
-```
-
-```{r hydro-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-hydro-calib}")}
-knitr::include_graphics(file.path("figures/hydro-calib2.jpeg"))
-```
-
-```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts}"}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q2.5_by_depth.jpeg"))
-```
-
-```{r ab-fit, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ab-fit}")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/A_B_data_vs_model.jpeg"))
-```
-
-```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-iso}")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_phenology_Meinzer_gr.Psi.VPD.jpeg"))
-```
-
-```{r sp-erd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap= paste0("\\label{lab-sp-erd}")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/ERD_by_sp_large_canopy.jpeg"))
-```
-
-```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/erd.stem.traits_cor.matrix_all_p-val.jpeg"))
-```
-
-```{r erd-mort, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-mort}")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_rate_by rdi.gr.jpeg"))
-```
-
-```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time}")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr.jpeg"))
-```
-
-\newpage
-
-# Supplementary material {-}
-
-## Data Descriptions
-
-### Microclimatic tower {#SIData.matteo}
-
-The tower used for these measurements is 41 m above ground, on a plateau on BCI [@Pau.2018].
-The eddy covariance system includes a sonic anemometer (CSAT3, Campbell Scientific, Logan, UT) and an open-path infrared CO2/H2O gas analyzer (LI7500, LiCOR.
-Lincoln, NE).
-Hi-frequency (10Hz) measurements were acquired by a datalogger (CR1000, Campbell Scientific) and stored on a local PC.
-Data were processed with a custom program using a standard routine described in [@Detto.2010].
-QA/QC criteria for removing erroneous values are listed in table 1 (see below).
-GPP was derived from daytime values of NEE by adding the corresponding mean daily ecosystem respiration obtained as the intercept of the light response curve [@Lasslop.2010].
-The light curve was fitted on a 15-days moving window using a rectangular hyperbolic function (runs with friction velocity less than 0.4 m s^-1^ were excluded).
-In order to compute daily time integrated budgets, gaps were filled using Artificial Neural Network [@Papale.2003] with hydro-meteorological inputs as predictors (soil moisture, solar radiation, temperature, VPD and air pressure, all measured on the same tower).
-To train the network, the dataset was randomly divided in a training set (70%), a validation set (15%) and a test set (15%).
-A two-layer feed-forward network with 10 sigmoid hidden neurons and linear output neurons, was trained using the Levenberg-Marquardt algorithm until the mean square error (MSE) of the validation set stop improving [@Hagan.1994].
-Performance, in term of MSE, was evaluated using the test set at the end of the training.
-This procedure was repeated 100 times to produce 100 estimates of GPP.
-Training multiple times generates different results due to different initial conditions and random sampling of the three sets.
-Ensemble was obtained as weighted average from the 100 ANN predictions using the MSE of the test set as weights according to:
-
-$f_{i} = \frac{\sum_{k}f_{ik}^{ANN}/{MSE_{ik}}}{\sum_{k}1/{MSE_{ik}}}$
-
-The ANN was implemented using the Neural Network Toolbox in Matlab 2014a.
-
-### Stream discharge from the Conrad catchment {#SIData.steveP}
-
-The Conrad weir is located on the gently sloping western side of Barro Colorado Island.
-The weir consists of a 90 degree ‘V’ notch with a rectangular upper section.
-The weir was constructed in several stages between 1993 and 1996.
-Stream stage is recorded at 5-minute intervals.
-
-The weir drains a catchment of approximately 40 ha of which at least 90% consists of the central plateau area of the island.
-Peak flow usually follows peak rain by approximately 2-3 hours.
-Large storms are sometimes preceded by small, initial peaks generated by the rain falling on the steeper slopes close to the weir.
-
-Streamflow data are analyzed by software custom-built by author SP.
-The program is used to correct for blockages of the weir ‘V’ by fallen vegetation (a common occurrence), sensor drift, small gaps and other problems.
-Discharge is then calculated using an empirically derived rating curve calculated for each 1mm stream stage.
-
-### Volumetric water content
-
-We collected volumetric water content (VWC) data (2012-2018) from a point location at the eddy flux tower with a vertical time domain reflectometry (TDR) probe.
-The relationship between the output from the TDR probe and VWC was calibrated based on measurements of gravimetric water content (GWC) on several occasions from representative soils near the vertical probe (0-0.15 m) (Fig.
-S\@ref(fig:TDR-calib)).
-GWC was converted to VWC based on the observed mean bulk dry density of 0.8 g cm^-3^ near the TDR probe (*n* = 5, cylinder volume = 1250 cm^3^).
-Beginning in 2016 we additionally installed three horizontal probes at depths of 0.15, 0.4 and 1 m and we applied the same calibration equation to these data (2016-2018) as that for the vertical probe.
-
-We also leveraged exisiting GWC measurements by @Kupers:2019 at the depths of 0.15, 0.4 and 1 m in summer 2015 (February, March and April) and 0.15, 0.4 in summer 2016 (March)--1299 samples that covered all soil types and habitats in the plot.
-The sampling periods were 6, 5, 10 and 8 days long, respectively.
-The 2016 dry season was associated with the 2015–2016 El Niño (see @Kupers:2019 for further details).
-We converted these GWC measurements to VWC assuming a bulk density of 0.8 g cm^-3^.
-$\theta_{0.15 m}$, $\theta_{0.4 m}$ and $\theta_{1 m}$ refer to the plot-wide VWC averages by depth over each sampling period.
-Middle date of a sampling period is used for comparison with model simulations.
-
-### Leaf hydraulic conductivity and vulnerability to cavitation {#SIData.joseph}
-
-The K~leaf~ was measured under steady state and high irradiance following the Evaporative Flux Method (FEM) [@Sack.2012] using a flowmeter.
-Briefly, the leaf was excised under filtered deionised water and attached to a water source (graduated cylinder) using clear plastic tubing.
-The system includes tubing of a known resistance, separated upstream and downstream by two sensors measuring the transpiration-driven water flow pressure in the EFM system [@Sack.2011].
-The K~leaf~ is calculated as the ratio of the transpiration rate (E~leaf~; mmol s^-1^) to the difference in flow pressure (MPa) upstream and downstream from the resistance tubing, and further standardized to 25°C and by leaf area.
-At stability (CV < 5%), the leaf is left to equilibrate in plastic bags with moist paper towels and measured for $\Psi$~leaf~ using a Scholander pressure chamber.
-To construct vulnerability curves for each species, four different functions were fitted to species-specific K~leaf~ vs.
-$\Psi$~leaf~ data using the *anneal* optimization function in the R package likelihood: exponential without an intercept ($K_{leaf} = Ae^{-B\Psi_{leaf}}$), exponential with an intercept ($K_{leaf} = C + Ae^{-B\Psi_{leaf}}$), logistic ($K_{leaf} = \frac{A}{1 + \left(\frac{\Psi_{leaf}}{x_{0}}\right)^B}$) and sigmoidal ($K_{leaf} = \frac{A}{1 + e^{-\left(\frac{\Psi_{leaf} - x_{0}}{B}\right)}}$) [@Sack.2012].
-The differences among AICc values for these functions were not large within species (see Zailaa et al.
-for more detailed methods description).
-We chose the exponential function without the intercept to fit observed data across species, and as a general function to scale-up to the community, because it had the lowest AICc for most species, captured the form of the observed data across all species well, and had the lowest number of parameters, that is, two.
-
-### Stem maximum hydraulic conductivity and vulnerability to cavitation {#SIData.brett}
-
-Brett to edit.
-
-Methods for Stem Maximum hydraulic conductivity and vulnerability to cavitation closely followed [Christman.2012] and [Wolfe.2016].
-Values for some of the species reported here were previously reported by @Dickman.2019 and @Wu.2020.
-The entire dataset and R script for the analyses are archived in the NGEE-Tropics archive (Wolfe 2020, To be done).
-\
-Canopy branches were cut from 4–10 trees of 26 species across three sites (Table SXX).
-The branch segments were enclosed in opaque plastic bags, and brought to the laboratory for measurements.
-Stem segments used for hydraulic conductivity (Kh) measurements were located at least 1 m from the cut end of the branch to minimize the number of open vessels that extended into the segment.
-Maximum vessel length was measured on a subset of samples from each species.
-Among species, maximum vessel length ranged XXX and length of stem between the cut that was made in the air to collect the branch and the section of stem used for Kh measurement was XXX (Table SXX).
-\
-Three leaves that were distal to the Kh segment were sealed in plastic bags and the entire branch was enclosed in black plastic bags for at least 30 minutes to stop transpiration.
-Just prior to measurement of Kh, leaf water potential of the three bagged leaves was measured with a pressure chamber.
-Since transpiration was stopped by bagging the leaves, leaf water potential was assumed to equal stem water potential (REFERENCE).
-\
-For each Kh segment, three measurements of stem water potential were made on stem sections that were located adjacent to the Kh segment.
-Stem sections that were approximately 40 mm in length were placed in nickel-plated aluminum chambers that were connected to Peltier-type psychrometers (Merrill Instruments, Logan, UT) and placed in a water bath at 30°C for 3 h to equilibrate the humidity within the cambers.
-Then water potential was measured using an automated multi-channel micro-voltmeter (CR7, Campbell Scientific, Logan, UT) with a 10 s thermocouple cooling current.
-For samples with water potential < -4 MPa, a 45 s cooling current was used to improve accuracy.
-Each psychrometer probe was calibrated against salt solutions beforehand and the plateau of the psychromatic response curve was used to assess water potential.
-\
-The Kh segments were removed from the branches by repeatedly cutting back the branch on both ends while it was submerged in water to relax the hydraulic tension within the xylem ((REFERENCE)).
-Then the ends were shaved clean with a fresh razor.
-Segments were 11–34 cm in length (mean ± SD = 22 ± 4 cm) and 0.2–1.6 cm in diameter (0.7 ± 0.2 cm).
-The segments were placed in a hydraulic conductivity apparatus similar to that described by Sperry et al.
-(1998) except that flow rate was measured with graduated pipettes instead of a balance.
-The perfusion solution was 10 mM KCl filtered to 0.2 m.
-For each segment, we measured the flow rate at three pressure heads: two that ranged 0.5–2 kPa and at zero pressure.
-Following Torres-Ruiz et al.
-(2012), Kh was calculated as the slope of the flow rate versus pressure gradient (i.e., pressure head divided by segment length) among the three pressure heads.
-Stem area specific hydraulic conductivity was (Ks) was calculated by dividing Kh by the cross-sectional area of stem.
-To calculate cross-sectional stem area, stem diameter was measured twice perpendicularly on both ends of the segment and averaged, then area was calculated assuming a circular cross section.
-\
-To assess vulnerability to cavitation, the branches were allowed to dry in the lab for 0 – 180 hours before they were measured for stem water potential and Ks.
-We used the water potential measured by the psychrometers because they are a more direct measurement.
-It has been noted that bagged-leaf water potential can deviate from stem water potential, especially when branch segments are relatively dry (REFERENCE).
-For some species, we noted this pattern while for others it was not apparent (Fig SXX).
-For each species, Ks was plotted against stem water potential and a Weibull curve was fit through the 90th  percentile of the points, following Wolfe et al.
-(2016)).
-The intercept of the curve was interpreted as maximum Ks and the water potential where the Weibull curve predicted 12, 50, 80, and 88 loss of maximum Ks were interpreted as P12, P50, P80, and P88, respectively.
-
-### Data gaps in WSG and LMA {#SIData.rutuja}
-
-Among the `r length(sp.exp.param$sp)` species with observed data on ($K_{leaf}$ vs.
-$\Psi_{leaf}$) used to build the community-level models of leaf vulnerability curves (Eqs.
-\@ref(eq:b-mod)-\@ref(eq:a-mod)), two lacked data on WSG and five on LMA~lamina~.
-We filled three gaps in LMA using predictions from a strong linear relationship between LMA and LMA~disc~, the latter defined as the mean leaf mass per unit area measured for a 1.483 cm^2^ leaf disc taken to avoid veins (g m^-2^) for leaves receiving direct sunlight ($\overline{LMA} = `r disc.cf[1]` + `r disc.cf[2]`LMA_{disc}$; Adj.
-*R*^2^ = `r disc.r2`; *P* `r disc.p`, *n* = `r disc.n`).
-We filled the two remaining gaps in LMA using predictions from a linear relationship between LMA and LMA~lamina~, as measured during the smaller, $K_{leaf}$ campaign ($\overline{LMA} = `r lma.cf[1]` + `r lma.cf[2]`LMA_{lamina}$; Adj.
-*R*^2^ = `r lma.r2`; *P* `r lma.p`, *n* = `r lma.n`).
-We chose not to fill the two data gaps in WSG as available data were poor predictors of WSG.
-
-## Alternative models for Effective Rooting Depth (ERD)
-
-Below we present all the altervative models for Effective Rooting Depth (ERD).
-Daily average species-specific growth, $G_{t~|~z}$, in the census interval *t* given an ERD of depth *z*, is described as follows:
-
-\begin{equation}
-G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,~i~|~z}^{*}) + \epsilon_{t~|~z}
-(\#eq:M1)
-\end{equation}
-
-\begin{equation}
-G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,~i~|~z}^{*} VPD_{i}^{*}) + \epsilon_{t~|~z}
-(\#eq:M2)
-\end{equation}
-
-\begin{equation}
-G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,~i~|~z}^{*} + VPD_{i}^{*}) + \epsilon_{t~|~z}
-(\#eq:M3)
-\end{equation}
-
-\begin{equation}
-G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,~i~|~z}^{*} LAI_{doy~|~i}^{*}) + \epsilon_{t~|~z}
-(\#eq:M4)
-\end{equation}
-
-\begin{equation}
-G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} (K^{*}_{leaf,~i~|~z} + VPD^{*}_{i}) LAI^{*}_{doy~|~i}) + \epsilon_{t~|~z}
-(\#eq:M5)
-\end{equation}
-
-\begin{equation}
-G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,~i~|~z}^{*}VPD_{i}^{*}LAI_{doy~|~i}^{*}) + \epsilon_{t~|~z}
-(\#eq:M6)
-\end{equation}
-where n~t~ be the total number of days in census interval *t*, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is a normally distributed model error term.
-
-# Supplementary Figures
-
-```{r ksat-profiles, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ksat-profiles}Generated 5000 curves of soil hydraulic conductivity variation by depth based on Latin Hypercube Sampling within 95% CI of observed data at depths of 12.5 cm and 60 cm [@Kinner:2004], assuming a linear decline between those depths, and values before 12.5 and after 60 cm as that of 12.5 cm and 60 cm, respectively.
-Only 100 curves are shown for clarity.
-Note the log scale on horizontal axis.")}
-knitr::include_graphics(file.path("figures/Bootstrapped Ksat profiles within observed 95CI_random_two_hundred.jpeg"))
-```
-
-```{r TDR-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-TDR-calib}Volumetric Water Content vs.
-TDR data for a vertical TDR probe (0-15 cm depth) at the base of the eddy flux tower near the 50 ha plot at BCI."}
-knitr::include_graphics(file.path("figures/Matteos_TDR_SWC_data_linear_model.jpeg"))
-```
-
-```{r dbh-model, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-dbh-model}A scatterplot of tree growth rates (circles) from all census intervals versus dbh at the beginning of the census interval is overlaid with a dbh model of growth (spline).
-Growth rates are color coded by census interval.
-The model equation is shown in inset along with goodness-of-fit (Adj.
-*R*^2^).")}
-knitr::include_graphics(file.path("figures/dbh/growthrate_dbh_predicted_growth_with_spline_degree_5_dbh_cutoff_100_model_no_color.jpeg"))
-```
-
-```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "35%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd}Scatterplot of mean midday (11:00 am to 15:00 pm) gross primary productivity (GPP) and mean midday vapor pressure deficit (VPD) measured at the eddy flux tower in Barro Colorado Island over 2012-2017.
-Red line shows a fit from a polynomial model described in inset, and used to standardize VPD employed in equations \\@ref(eq:M2) and \\@ref(eq:M4)."}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/gpp_by_env_variables/daily_gpp.vpd.jpeg"))
-```
-
-```{r leaf-cover, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-cover}Seasonal variation in standardised LAI (range 0-1) for ", length(erd.sp)," species at Barro Colorado Island (panels).
-Species are color-coded by deciduousness.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf.cover_BCI_multi_panel.jpeg"))
-```
-
-```{r iso-data, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-iso-data}Stable isotopic ratio ($\\delta^2H$) of soil water by depth measured at Barro Colorado Island in March 1997.
-Data from @Meinzer:1999."}
-knitr::include_graphics(file.path("figures/UDI_confidence/Meinzer_etal_1999_Depth_vs_soil_deltaD.jpeg"))
-```
-
-```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($Psi_{soil,~z}$) by depth z (m; panels) over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines).
-Vertical gray lines demarcate median dates of the demographic censuses at the BCI 50 ha plot."}
-knitr::include_graphics(file.path("figures/2019-10-14_5000/best-fits/psi_model_daily_all_depths_params.top.few_full.jpeg"))
-```
-
-```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}Vulnerability curves fitted to observed $K_{leaf}$ vs.
-$\\Psi_{leaf}$ data on ", length(sp.exp.param$sp)," tree species from BCI (A & B) with equation \\@ref(eq:vc-exp)--in absolute terms (A, B), as well as, in terms of loss of conductivity (%) (C, D).
-Each line represents a species, color coded by $WSG$ (A, C), or $LMA$ (B, D), with blue tones indicating smaller values of $WSG$ or $LMA$ and red tones indicating larger values.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/plot.obs.plc.jpeg"))
-```
-
-```{r obs-comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-comm-plc-curves}Percent loss of vulnerability curves for species with ERD estimates.
-Each line represents a species, color coded by source of the curve: fitted to observed data for $K_{leaf}$ vs.
-$\\Psi_{leaf}$ are in yellow and those based on scaling relationship with WSG and LMA are in gray.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/obs.mod_plccurves_fits_ggplot_no_col.jpeg"))
-```
+\beginsupplement
 
@@
 or each model are shown in inset, along with the goodness-of-fit (*R*^2^) and *P*-values for a linear relationship between ERD and $\\delta^2H$.
-ERD models in panels A-E correspond to Eq.
-\\@ref(eq:M1)-\\@ref(eq:M6), respectively.")}
+See Eqs.
+\\@ref(eq:M1)-\\@ref(eq:M6).
+See Table \\@ref(tab:ab-param) for the full form of scientific names.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_sp_color_Meinzer.jpeg"))
@@
 
-#Supplementary Tables
-
-```{r param, results = 'asis', echo = FALSE}
-knitr::kable(param.table, caption = 'ELM-FATES parameters used to generate ensembles, with description, prescribed global ranges, rationale for the choice of ranges and references.
-Ranges for best-fit chosen ensembles are also given.
-Note that no.
-1-6 are parameters specific to FATES.', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(param.table))))
-```
-
-```{r eddy, results = 'asis', echo = FALSE}
-knitr::kable(eddy.table, caption = 'QA/QC procedure applied to eddy covariance fluxes.
-w: vertical wind speed; t: air temperature; q: water vapor density; c: CO~2~ density; u: horizontal wind speed.
-The bar indicates 30-min averaging and the apex indicates fluctuation ($\\overline{x` = x - \\bar{x}}$)', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(eddy.table))))
-```
-
-```{r ab-param, results = 'asis', echo = FALSE}
-knitr::kable(ab.table, caption = 'Leaf vulnerability curves parameters A & B, K~max, leaf~ and $\\Psi$ for the " length(erd.sp)" species with ERD estimates.
-Species for which curves were fitted to observed data for $K_{leaf}$ vs.
-$\\Psi_{leaf}$ are identified with Source as Data, while species for which curves were obtained based on trait-proxies are identified with Source as Model', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(ab.table))))
-```
-
-# References
+\end{document}

### 2020-10-20 manuscript.rmd
1. Graph for ERD vs. mrate shown for evergeeens 2. Graph for GPP vs VPD changed to monthly

@@
 ```{r erd-mort, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-mort}")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_rate_by rdi.gr.jpeg"))
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_rate_by rdi.gr_evergreen.jpeg"))
 ```
@@
 Red line shows a fit from a polynomial model described in inset, and used to standardize VPD employed in equations \\@ref(eq:M2) and \\@ref(eq:M4)."}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/gpp_by_env_variables/daily_gpp.vpd.jpeg"))
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/gpp_by_env_variables/monthly_gpp.vpd.jpeg"))
 ```

### 2020-10-20 manuscript.rmd
1. formatting
1. Composite drought figure added. 2. older drought figure moved to supplementary

@@
 
-Figure \@ref(fig:swp-droughts): Occurrence of extreme soil droughts by census intervals (horizontal panels) at three representative depths (z): 0.1, 0.6 and 1 m (vertical panels).
-Mean (black lines) and lower half of 95% distribution (grey areas) of soil water potential ($Psi_{soil,~z}$) for a given DOY and depth z are shown (same across census intervals).
-$Psi_{soil,~z}$ by DOY for a year (colored line) is only shown if for some part of that year $Psi_{soil,~z}$ declined below the 2.5 percentile distribution.
-
 Figure \@ref(fig:ab-fit): Parameters $A$ (panel A) & $B$ (panel B) for the leaf vulnerability curves that were fitted to observed data on $K_{leaf}$ vs.
@@
 
-```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts}"}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q2.5_by_depth.jpeg"))
-```
-
 ```{r ab-fit, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ab-fit}")}
@@
 
-```{r erd-mort, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-mort}")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_rate_by rdi.gr_evergreen.jpeg"))
+```{r erd-mort, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-mort}")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/erd_mrate_interval_droughts.jpeg"))
 ```
@@
 
-```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($Psi_{soil,~z}$) by depth z (m; panels) over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines).
+```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($Psi_{soil,~z}$) by depth z (m; panels) over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines).
 Vertical gray lines demarcate median dates of the demographic censuses at the BCI 50 ha plot."}
 knitr::include_graphics(file.path("figures/2019-10-14_5000/best-fits/psi_model_daily_all_depths_params.top.few_full.jpeg"))
+```
+
+```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts}Occurrence of extreme soil droughts by census intervals (horizontal panels) at three representative depths (z): 0.1, 0.6 and 1 m (vertical panels).
+Mean (black lines) and lower half of 95% distribution (grey areas) of soil water potential ($\\Psi_{soil,~z}$) for a given DOY and depth z are shown (same across census intervals).
+$\\Psi_{soil,~z}$ by DOY for a year (colored line) is only shown if for some part of that year $\\Psi_{soil,~z}$ declined below the 2.5 percentile distribution."}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q2.5_by_depth.jpeg"))
 ```

### 2020-10-21 manuscript.rmd
1. changed chosen ERD model 2. changed depths used for ERD 3. LMA and WSG to Roman 4. P to small p roman; rho to r 5. space

@@
 \begin{equation}
-G_{s,~t~|~z,~h} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} (K^{*}_{leaf,~s,~i~|~z,~h} + VPD^{*}_{i}) LAI^{*}_{s,~doy~|~i}) + \epsilon_{t~|~z,~h}
+G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,~i~|~z}^{*} VPD_{i}^{*}) + \epsilon_{t~|~z}
 (\#eq:M-chosen)
@@
 Using Eq.
-\@ref(eq:M-chosen), for a species $s$, we obtained Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $r$ for each soil depth $z$ for each of the 100 best-fit hydrological realizations.
-For each hydrological realization we retained only one depth that corresponded to the least AIC score (the best-fit) [@Chonggang, we could have retained multiple depths with non-differentiable AICs], and within those depths retained the ones that had Pearson's $r$ >= 0.71 (or, $r^2$ >= 0.5) as a valid set.
+\@ref(eq:M-chosen), for a species $s$, we obtained likelihood based on Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $r$ for each soil depth $z$ for each of the 100 best-fit hydrological realizations.
+For each hydrological realization we retained only one depth that corresponded to the least AIC score (the best-fit), and within those depths retained the ones that had Pearson's $r$ >= 0.71 (or, $r^2$ >= 0.5) as a valid set.
 For a species $s$, we defined effective rooting depth (ERD) as the median of these set of depths, described as :
@@
 \begin{equation}
-ERD_{s} = \underset{\forall \, h \, \in \, H}{median} \Big\{z,h\,\Big|\, \Big(AIC_{s~|~z,~h} = \underset{\forall \, z \,\in\,Z}{min}(AIC_{s~|~z,~h})\Big) \cap \Big(r_{s~|~z,~h} \ge 0.71\Big)\Big\}
+ERD_{s} = \underset{\forall \, h \, \in \, H}{median} \Big\{z,h\,\Big|\, \Big(Likelihood_{s~|~z,~h} = \underset{\forall \, z \,\in\,Z}{maximum}(Likelihood_{s~|~z,~h})\Big) \cap \Big(r_{s~|~z,~h} \ge 0.71\Big)\Big\}
 (\#eq:erd)
 \end{equation} 
-where $H$ $\in$ {1, .., 100}, refer to the 100 hydrological realisations and $Z$ $\in$ {`r paste(soil.depths, collapse=", ")`} m, covering the range of soil depths in ELM-FATES after above parametarision.
+where $H$ $\in$ {1, .., 100}, refer to the 100 hydrological realisations and $Z$ $\in$ 0.5, {`r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m, covering the range of soil depths in ELM-FATES after above parametarision [@Chonggang, Depths 0.21, 0.37 and 0.62 were all labelled as 0.5 m for model evaluation.
+How do I explain that?
+Do they need to be retained as their original depths?].
 
@@
 We obtained species-specific parameters $A$ & $B$ for Eq.
-\@ref(eq:k-soil) either from data when available, or a model based on $LMA$ or $WSG$ as described in the earlier section, and used K standardized between 0 and 1 within species ($K^*$) in Eq.
+\@ref(eq:k-soil) either from data when available, or a model based on LMA or WSG as described in the earlier section, and used K standardized between 0 and 1 within species ($K^*$) in Eq.
 \@ref(eq:M-chosen).
@@
 For the leaf vulnerability curve parameters $A$ and $B$ fitted to the observed data for $K_{leaf}$ vs.
-$\Psi_{leaf}$, $WSG$ and parameter $B$ explained a large proportion of variation in parameter $A$ (Eq.
+$\Psi_{leaf}$, WSG and parameter $B$ explained a large proportion of variation in parameter $A$ (Eq.
 \@ref(eq:a-mod); *Adj.
-R*^2^ = `r afit.r2`; *P* `r  afit.p`; Fig.
+R*^2^ = `r afit.r2`; *p* `r  afit.p`; Fig.
 \@ref(fig:ab-fit)).
-$WSG$ and $LMA$ explained a large proportion of variation in parameter $B$ (Eq.
+WSG and LMA explained a large proportion of variation in parameter $B$ (Eq.
 \@ref(eq:b-mod); Adj.
-*R*^2^ = `r bfit.r2`; *P* `r  bfit.p`; also see Fig.
+*R*^2^ = `r bfit.r2`; *p* `r  bfit.p`; also see Fig.
 S\@ref(fig:obs-plc-curves).
@@
 ERD estimated from the model with K~leaf~, LAI and an additive effect of VPD as explanatory variables (Eq.
-\@ref(eq:M2)) was strongly related to the measured isotopic signal ($\delta^2H$), that is, with highest goodness-of-fit (*R*^2^ = 0.65; *P* = 0.027; Fig.
+\@ref(eq:M2)) was strongly related to the measured isotopic signal ($\delta^2H$), that is, with highest goodness-of-fit (*R*^2^ = 0.65; *p* = 0.027; Fig.
 \@ref(fig:erd-iso)).
@@
 
-  For the subset of 8 species with ERD for which hydraulic traits data were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max,~stem}$, Spearmans' *r* = `r cor.k`, *P* = `r p.k` at $\alpha$ = 0.05; Fig.
-\@ref(fig:erd-hyd)A), less negative leaf $\Psi_{tlp}$ (*r* = `r cor.tlp`, *P* = `r p.tlp`; Fig.
-\@ref(fig:erd-hyd)B), less negative $\Psi_{88,~stem}$ and thus greater vulnerability to embolism from cavitation (*r* = `r cor.stem.88`, *P* = `r p.stem.88`; Fig.
-\@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$; *r* = `r cor.hsm`, *P* = `r p.hsm`; Fig.
+  For the subset of 8 species with ERD for which hydraulic traits data were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max,~stem}$, Spearmans' *r* = `r cor.k`, *p* = `r p.k` at $\alpha$ = 0.05; Fig.
+\@ref(fig:erd-hyd)A), less negative leaf $\Psi_{tlp}$ (*r* = `r cor.tlp`, *p* = `r p.tlp`; Fig.
+\@ref(fig:erd-hyd)B), less negative $\Psi_{88,~stem}$ and thus greater vulnerability to embolism from cavitation (*r* = `r cor.stem.88`, *p* = `r p.stem.88`; Fig.
+\@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$; *r* = `r cor.hsm`, *p* = `r p.hsm`; Fig.
 \@ref(fig:erd-hyd)D).
@@
 \
-Species with deeper ERD showed lower mean mortality rates across the 34-yr record (1982-2015; *R*^2^ = 0.11, *P* = 0.075, *n* = `r length(erd.mrate.sp)`; Fig.
+Species with deeper ERD showed lower mean mortality rates across the 34-yr record (1982-2015; *R*^2^ = 0.11, *p* = 0.075, *n* = `r length(erd.mrate.sp)`; Fig.
 \@ref(fig:erd-mort)).
@@
 S\@ref(fig:swp-all-depths); @Chitra-Tarak.2018).
-The variation in mortality explained by ERD in our study is low (*R*^2^ = 0.11, *P* = 0.075), and other causes may explain observed mortality [@McDowell.2018].
+The variation in mortality explained by ERD in our study is low (*R*^2^ = 0.11, *p* = 0.075), and other causes may explain observed mortality [@McDowell.2018].
 
@@
 We also found that LMA and WSG were strong predictors of PLC curves for leaves, but not for stems (not shown).
-WSG alone explained `r round(a.wsg.r2*100,0)` % of variation in parameter $A$ ($A = `r a.wsg.cf[1]` `r a.wsg.cf[2]`B + `r a.wsg.cf[3]`B^2 + error$; *P* `r  a.wsg.p`) and `r round(b.wsg.r2*100,0)` % of variation in $B$ ($B = `r b.wsg.cf[1]` `r b.wsg.cf[2]`B + `r b.wsg.cf[3]`B^2 + error$; *P* `r  b.wsg.p`).
-Parameter $B$ alone explained `r round(a.b.r2*100,0)` % of variation in $A$ (*P* `r  a.b.p`), while $LMA$ alone was weakly related to $A$ and was not related to $B$ (not shown).
+WSG alone explained `r round(a.wsg.r2*100,0)` % of variation in parameter $A$ ($A = `r a.wsg.cf[1]` `r a.wsg.cf[2]`B + `r a.wsg.cf[3]`B^2 + error$; *p* `r  a.wsg.p`) and `r round(b.wsg.r2*100,0)` % of variation in $B$ ($B = `r b.wsg.cf[1]` `r b.wsg.cf[2]`B + `r b.wsg.cf[3]`B^2 + error$; *p* `r  b.wsg.p`).
+Parameter $B$ alone explained `r round(a.b.r2*100,0)` % of variation in $A$ (*p* `r  a.b.p`), while LMA alone was weakly related to $A$ and was not related to $B$ (not shown).
 The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits, including hydraulic vulnerability, in order to better represent the functional diversity of tropical forests, and their trajectories into the future.
@@
 \@ref(eq:a-mod)-\@ref(eq:b-mod)).
-Goodness-of-fit (*R*^2^) and *P*-values are given in inset.
+Goodness-of-fit (*R*^2^) and *p*-values are given in inset.
 
@@
 Figure \@ref(fig:erd-iso): Matrix for pairwise correlations (circles) among Effective Rooting Depth (ERD) and hydraulic properties for `r nrow(cor.data)` canopy species found on Barro Colorado Island (Panama).
-Each circle is color coded by the Spearman's $\\rho$ with red shades for positive $\\rho$ values, while blue shades for negative values, respectively.
-P-values are given only when they are $\\ge$ 0.05.
+Each circle is color coded by the Spearman's *r* with red shades for positive *r* values, while blue shades for negative values, respectively.
+*p*-values are given only when they are $\\ge$ 0.05.
 Units are as follows: ERD in m, K~max~ in mmol m^-1^ s^-1^ MPa^-1^), all variables with $\\Psi$ in MPa.
@@
 Figure \@ref(fig:erd-mort): Mortality rate (mean $\pm$ 1SE) versus Modelled Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) for `r length(erd.mrate.sp)` canopy species found on Barro Colorado Island.
-*R*^2^ and *P*-values are given in panel insets.
+*R*^2^ and *p*-values are given in panel insets.
 
@@
 We filled three gaps in LMA using predictions from a strong linear relationship between LMA and LMA~disc~, the latter defined as the mean leaf mass per unit area measured for a 1.483 cm^2^ leaf disc taken to avoid veins (g m^-2^) for leaves receiving direct sunlight ($\overline{LMA} = `r disc.cf[1]` + `r disc.cf[2]`LMA_{disc}$; Adj.
-*R*^2^ = `r disc.r2`; *P* `r disc.p`, *n* = `r disc.n`).
+*R*^2^ = `r disc.r2`; *p* `r disc.p`, *n* = `r disc.n`).
 We filled the two remaining gaps in LMA using predictions from a linear relationship between LMA and LMA~lamina~, as measured during the smaller, $K_{leaf}$ campaign ($\overline{LMA} = `r lma.cf[1]` + `r lma.cf[2]`LMA_{lamina}$; Adj.
-*R*^2^ = `r lma.r2`; *P* `r lma.p`, *n* = `r lma.n`).
+*R*^2^ = `r lma.r2`; *p* `r lma.p`, *n* = `r lma.n`).
 We chose not to fill the two data gaps in WSG as available data were poor predictors of WSG.
@@
 \begin{equation}
-G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,~i~|~z}^{*} VPD_{i}^{*}) + \epsilon_{t~|~z}
+G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,~i~|~z}^{*} + VPD_{i}^{*}) + \epsilon_{t~|~z}
 (\#eq:M2)
@@
 \begin{equation}
-G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,~i~|~z}^{*} + VPD_{i}^{*}) + \epsilon_{t~|~z}
+G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,~i~|~z}^{*} VPD_{i}^{*}) + \epsilon_{t~|~z}
 (\#eq:M3)
@@
 $\\Psi_{soil,~z}$ by DOY for a year (colored line) is only shown if for some part of that year $\\Psi_{soil,~z}$ declined below the 2.5 percentile distribution."}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q2.5_by_depth.jpeg"))
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q5_by_depth.jpeg"))
 ```
@@
 $\\Psi_{leaf}$ data on ", length(sp.exp.param$sp)," tree species from BCI (A & B) with equation \\@ref(eq:vc-exp)--in absolute terms (A, B), as well as, in terms of loss of conductivity (%) (C, D).
-Each line represents a species, color coded by $WSG$ (A, C), or $LMA$ (B, D), with blue tones indicating smaller values of $WSG$ or $LMA$ and red tones indicating larger values.")}
+Each line represents a species, color coded by WSG (A, C), or LMA (B, D), with blue tones indicating smaller values of WSG or LMA and red tones indicating larger values.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/plot.obs.plc.jpeg"))
@@
 ```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-other-erd-mods}Effective Rooting Depth versus $\\delta^2H$ for each of the ERD models tested (panels).
-For each ERD model only those species have an estimate of ERD that have passed the test of Pearson's $\\rho$ > 0 and *R*^2^ > 0.5 during ERD parameterisation.
+For each ERD model only those species have an estimate of ERD that have passed the test of Pearson's *r* > 0 and *R*^2^ > 0.5 during ERD parameterisation.
 The number of species with an ERD estimate as well as $\\delta^2H$.
-or each model are shown in inset, along with the goodness-of-fit (*R*^2^) and *P*-values for a linear relationship between ERD and $\\delta^2H$.
+or each model are shown in inset, along with the goodness-of-fit (*R*^2^) and *p*-values for a linear relationship between ERD and $\\delta^2H$.
 ERD models in panels A-E correspond to Eq.

### 2020-10-21 manuscript.rmd
Corrected variable symbols from italic to roman

@@
 For greater accuracy, instead of the ELM default soil texture-based pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content vs.
-$\Psi_{soil}$ for depths of 0.15, 0.4 and 1 m [@Kupers:2019].
+$\Psi$~soil~ for depths of 0.15, 0.4 and 1 m [@Kupers:2019].
 We converted gravimetric water content to volumetric (VWC, $\theta$) using a bulk density value of of 0.8 g cm^-3^ measured at a point location (see below).
@@
 There are many factors that could limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019].
-In this study, we used maximum diurnal leaf conductance ($K_{leaf}$) approximated using rhizosphere soil water potential ($\Psi_{soil}$) as a proxy for the soil water limiting factor on growth.
-In our models $K_{leaf}$ therefore incorporates the effects of rooting depth and rhizosphere water potential.
-For each species' growth dynamics across census intervals was regressed against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1.
+In this study, we used maximum diurnal leaf conductance (K~leaf~) approximated using rhizosphere soil water potential ($\Psi$~soil~) as a proxy for the soil water limiting factor on growth.
+In our models K~leaf~ therefore incorporates the effects of rooting depth and rhizosphere water potential.
+For each species' growth dynamics across census intervals was regressed against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, K~leaf~ and the leaf area index (LAI), all standardized between 0 and 1.
 Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting.
@@
 
-For each species, we estimated diurnal maximum value of $K_{leaf,~s,~i~|~z,~h}$ for a species $s$ by substituting $\Psi_{soil,~i~|~z,~h}$ in the leaf vulnerability curves defined below in Eq.
+For each species, we estimated diurnal maximum value of $K~{leaf,~s,~i~|~z,~h}$ for a species $s$ by substituting $\Psi$~soil,~i~|~z,~h~ in the leaf vulnerability curves defined below in Eq.
 \@ref(eq:vc-exp) using species-specific parameters $A_{s}$ and $B_{s}$ as:
@@
 
-By varying ‘z’, we estimated species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015).
+By varying ‘z’, we estimated species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi$~soil,z~) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015).
 Using Eq.
@@
 \
-We obtained leaf hydraulic vulnerability curves ($K_{leaf}$ vs.
-$\Psi_{leaf}$; see Table \@ref(tab:symbols) for trait symbols and their definitions) for adult trees of 21 common species from Zailaa et al.
+We obtained leaf hydraulic vulnerability curves (K~leaf~ vs.
+$\Psi$~leaf~; see Table \@ref(tab:symbols) for trait symbols and their definitions) for adult trees of 21 common species from Zailaa et al.
 (to be submitted; see SI for brief description of methods) described as:
@@
 \#ref(eq:vc-exp).
-For all the 31 ERD species we estimated species $\Psi_{50,~leaf}$ based on their vulnerability curves (see Table \@ref(tab:symbols) for definition).
+For all the 31 ERD species we estimated species $\Psi$~50,leaf~ based on their vulnerability curves (see Table \@ref(tab:symbols) for definition).
 
@@
 We obtained species-specific parameters $A$ & $B$ for Eq.
-\@ref(eq:k-soil) either from data when available, or a model based on LMA or WSG as described in the earlier section, and used K standardized between 0 and 1 within species ($K^*$) in Eq.
+\@ref(eq:k-soil) either from data when available, or a model based on LMA or WSG as described in the earlier section, and used K standardized between 0 and 1 within species (K*) in Eq.
 \@ref(eq:M-chosen).
@@
 \
-To validate the ERD models we used data for xylem sap $\delta^2H$ (permil) as an independent observation of ERD.
+To validate the ERD models we used data for xylem sap $\delta$^2H (permil) as an independent observation of ERD.
 For each model of ERD (Eq.
 \@ref(eq:M-chosen) and Eqs.
-in SI), estimated species ERD was regressed against species xylem sap $\delta^2H$ (permil).
-We used xylem sap $\delta^2H$ values given in @Meinzer:2001 for the dry-season of March 1997 as this period showed largest seasonal divergence among species, with a corresponding gradient of $\delta^2H$ in the soil (Fig.
+in SI), estimated species ERD was regressed against species xylem sap $\delta$^2H (permil).
+We used xylem sap $\delta$^2H values given in @Meinzer:2001 for the dry-season of March 1997 as this period showed largest seasonal divergence among species, with a corresponding gradient of $\delta$^2H in the soil (Fig.
 S\@ref(fig:iso-data)).
-We removed species that are likely to have no leaves in March as xylem sap $\delta^2H$ of these species may not be linked with water sourced at the time of measurement and thus may not be comparable with other species that had leaves.
+We removed species that are likely to have no leaves in March as xylem sap $\delta$^2H of these species may not be linked with water sourced at the time of measurement and thus may not be comparable with other species that had leaves.
 We also removed *Guapira standleyana*, whose xylem sap isotope value was not well constrained by the soil water isotope depth profile.
-We were thus left with 7 species with data on $\delta^2H$ as well as modelled estimates of ERD.
+We were thus left with 7 species with data on $\delta$^2H as well as modelled estimates of ERD.
 
@@
 \
-To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD against maximum stem hydraulic conductance ($K_{max, stem}$), leaf turgor loss point ($\Psi_{tlp}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$).
+To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD against maximum stem hydraulic conductance (K~max,stem~), leaf turgor loss point ($\Psi$~tlp~), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi$~88,stem~), and hydraulic safety margins ($\Psi$~min~ - $\Psi$~88,stem~).
 See [SI](#SIData.brett) for data collection and estimation of these variables.
@@
 We regressed species ERD against species ($Mort_{mean}$).
-To assess whether deeper ERD species escape the experience of significant soil drought, we calculated for each speceis the proportion of days in the simulation period (1990-2015) during which soil water potential in the soil zone matching species ERD ($\Psi_{soil,~z = ERD}$) were below species $\Psi_{crit}$.
-$\Psi_{crit}$ was defined as $\Psi_{50,~leaf}$.
+To assess whether deeper ERD species escape the experience of significant soil drought, we calculated for each speceis the proportion of days in the simulation period (1990-2015) during which soil water potential in the soil zone matching species ERD ($\Psi$~soil,z~=~ERD~) were below species $\Psi$~crit~.
+$\Psi$~crit~ was defined as $\Psi$~50,leaf~.
 
@@
 
-  Soil Water Potential dynamics ($\Psi_{soil,~z}$) by depth (z = 0.1 through 13) obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of such extreme years, drought intensity at depth, drought seasonality and duration (Fig.
+  Soil Water Potential dynamics ($\Psi$~soil,z~) by depth (z = 0.1 through 13) obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of such extreme years, drought intensity at depth, drought seasonality and duration (Fig.
 S\@ref(fig:swp-droughts)).
-Nevertheless, across the simulation period (1990-2015) $\Psi_{soil,~z}$ remained above -0.5 MPa from depths 1.7 to 13 m (Fig.
+Nevertheless, across the simulation period (1990-2015) $\Psi$~soil,z~ remained above -0.5 MPa from depths 1.7 to 13 m (Fig.
 S\@ref(fig:swp-all-depths)).
@@
 \
-For the leaf vulnerability curve parameters $A$ and $B$ fitted to the observed data for $K_{leaf}$ vs.
-$\Psi_{leaf}$, WSG and parameter $B$ explained a large proportion of variation in parameter $A$ (Eq.
+For the leaf vulnerability curve parameters $A$ and $B$ fitted to the observed data for K~leaf~ vs.
+$\Psi$~leaf~, WSG and parameter $B$ explained a large proportion of variation in parameter $A$ (Eq.
 \@ref(eq:a-mod); *Adj.
@@
 ERD estimated from the model with K~leaf~, LAI and an additive effect of VPD as explanatory variables (Eq.
-\@ref(eq:M2)) was strongly related to the measured isotopic signal ($\delta^2H$), that is, with highest goodness-of-fit (*R*^2^ = 0.65; *p* = 0.027; Fig.
+\@ref(eq:M2)) was strongly related to the measured isotopic signal ($\delta$^2H), that is, with highest goodness-of-fit (*R*^2^ = 0.65; *p* = 0.027; Fig.
 \@ref(fig:erd-iso)).
 The model with K~leaf~, but without LAI and an additive effect of VPD showed no species with sufficient species-level goodness-of-fit.
-ERD from all the other models showed a positive, but non-significant relationship with $\delta^2H$ at $\alpha$ = 0.05 (Fig.
+ERD from all the other models showed a positive, but non-significant relationship with $\delta$^2H at $\alpha$ = 0.05 (Fig.
 S\@ref(fig:other-erd-mods)).
@@
 seemannii* and *T.
-tuberculata*, that @Meinzer:1999 found to have xylem sap $\delta^2H$ values between soil water and groundwater, suggesting that these species sourced most of the water from depths > 1 m, were also identified by our model with ERD > 1 m (2.9 m for both species; Fig.
+tuberculata*, that @Meinzer:1999 found to have xylem sap $\delta$^2H values between soil water and groundwater, suggesting that these species sourced most of the water from depths > 1 m, were also identified by our model with ERD > 1 m (2.9 m for both species; Fig.
 \@ref(fig:erd-iso)).
@@
 
-  For the subset of 8 species with ERD for which hydraulic traits data were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max,~stem}$, Spearmans' *r* = `r cor.k`, *p* = `r p.k` at $\alpha$ = 0.05; Fig.
-\@ref(fig:erd-hyd)A), less negative leaf $\Psi_{tlp}$ (*r* = `r cor.tlp`, *p* = `r p.tlp`; Fig.
-\@ref(fig:erd-hyd)B), less negative $\Psi_{88,~stem}$ and thus greater vulnerability to embolism from cavitation (*r* = `r cor.stem.88`, *p* = `r p.stem.88`; Fig.
-\@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$; *r* = `r cor.hsm`, *p* = `r p.hsm`; Fig.
+  For the subset of 8 species with ERD for which hydraulic traits data were available, species with deeper ERD showed greater maximum stem hydraulic conductance (K~max,stem~, Spearmans' *r* = `r cor.k`, *p* = `r p.k` at $\alpha$ = 0.05; Fig.
+\@ref(fig:erd-hyd)A), less negative leaf $\Psi$~tlp~ (*r* = `r cor.tlp`, *p* = `r p.tlp`; Fig.
+\@ref(fig:erd-hyd)B), less negative $\Psi$~88,stem~ and thus greater vulnerability to embolism from cavitation (*r* = `r cor.stem.88`, *p* = `r p.stem.88`; Fig.
+\@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi$~min~ - $\Psi$~88,stem~; *r* = `r cor.hsm`, *p* = `r p.hsm`; Fig.
 \@ref(fig:erd-hyd)D).
@@
 S\@ref(fig:erd-mort-census)).
-Analysis of time spent under significant soil drought showed that the proportion of days species $\Psi_{ERD}$ was more negative than species $\Psi_{crit}$ decreased with species ERD, that is the shallower the ERD, the greater proportion of days $\Psi_{ERD}$ < $\Psi_{crit}$ (Fig.
+Analysis of time spent under significant soil drought showed that the proportion of days species $\Psi$~ERD~ was more negative than species $\Psi$~crit~ decreased with species ERD, that is the shallower the ERD, the greater proportion of days $\Psi$~ERD~ < $\Psi$~crit~ (Fig.
 \@ref(fig:erd-time)).
-Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25-yr period for which soil water potential $\Psi_{soil,~z}$ was modeled (1990-2015), as soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`;  median, 95% CI) through all the dry seasons and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).
+Species with ERD deeper than 1 m did not cross species $\Psi$~crit~ for the entire 25-yr period for which soil water potential $\Psi$~soil,z~ was modeled (1990-2015), as soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`;  median, 95% CI) through all the dry seasons and drought periods, which is above ($\Psi$~crit~) of the most sensitive tree species (-0.28 MPa).
 
@@
 Nevertheless, deeper ERD species showed lower mortality rates than shallower ERD species over the 35-yr record (1982-2015).
-Analysis of time spent below ($\Psi_{crit}$) in the soil zone matching ERD $\Psi_{soil,~z = ERD}$ showed that this was likely due to decreasing realized risk of hydraulic failure with increasing access to water for species with deeper ERD.
+Analysis of time spent below ($\Psi$~crit~) in the soil zone matching ERD $\Psi$~soil,z~=~ERD~ showed that this was likely due to decreasing realized risk of hydraulic failure with increasing access to water for species with deeper ERD.
 We thus demonstrate, for the first time, a link between realized drought experience, hydraulic architecture and long-term mortality outcomes for co-occurring adult tree species.
@@
 These results are consistent with the two studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016].
-@Johnson:2018fad found that mortality decreased with deep rooting even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi_{88,~stem}$) and had narrower safety-margins ($\Psi_{min} - \Psi_{88,~stem}$).
-@Nardini:2015 found that a species sourcing deep water in a Karst system showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi_{88,~stem}$ and narrower safety margin compared to two species sourcing shallow-water.
+@Johnson:2018fad found that mortality decreased with deep rooting even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi$~88,stem~) and had narrower safety-margins ($\Psi$~min~ - $\Psi$~88,stem~).
+@Nardini:2015 found that a species sourcing deep water in a Karst system showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi$~88,stem~ and narrower safety margin compared to two species sourcing shallow-water.
 
@@
 Their hydrological model revealed that the multi-year drought exhausted the deep soil and even bedrock water availablity.
-In contrast, our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period, $\Psi_{soil}$ below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and droughts, that is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).
+In contrast, our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period, $\Psi$~soil~ below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and droughts, that is above ($\Psi$~crit~) of the most sensitive tree species (-0.28 MPa).
 The difference in our results are likely attributable to site differences in the precipitation regime.
@@
 By estimating ERD for large trees of `r length(erd.sp)` canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species-rich tropical forests.
-Our chosen ERD model is an advance over the only other inverse model of ERD (@Chitra-Tarak.2018; *n* = 12 species) as we utilize physiologically meaningful representation, fewer parameters and corroboration with tree and soil data on $\delta^2H$.
+Our chosen ERD model is an advance over the only other inverse model of ERD (@Chitra-Tarak.2018; *n* = 12 species) as we utilize physiologically meaningful representation, fewer parameters and corroboration with tree and soil data on $\delta$^2H.
 
@@
 This allowed us to scale-up the leaf vulnerability curves at the community scale (*n* = `r length(ab.table$Species)` species, see Table S\@ref(tab:ab-param) for model predicted parameters).
-We find that leaf conductance declined more rapidly with declining $\Psi_{leaf}$ for species with higher values of WSG and LMA (Fig.
+We find that leaf conductance declined more rapidly with declining $\Psi$~leaf~ for species with higher values of WSG and LMA (Fig.
 S\@ref(fig:obs-plc-curves)).
@@
 The trade-offs that we identify between above- and below-ground traits, the water environment vertically below-ground and mortality rates are important for representing diversity in Dynamic Global Vegetation Models especially in species-rich forests.
-The hydraulic efficiency-safety trade-off between high values of maximum stem conductivity (high $K_{max,~stem}$) and cavitation-resistance (low $\Psi_{50,~stem}$ or $\Psi_{88,~stem}$) is hypothesized although community-wide examples are rare [@Hacke:2006; @Gleason.2016].
+The hydraulic efficiency-safety trade-off between high values of maximum stem conductivity (high K~max,stem~) and cavitation-resistance (low $\Psi$~50,stem~ or $\Psi$~88,stem~) is hypothesized although community-wide examples are rare [@Hacke:2006; @Gleason.2016].
 We report such a trade-off (n = 26 species) at our study site stem level ().
@@
 Use of dendrometer data would thus entail accounting for stem hydrostatic flexing [@Zweifel:2018].
-Three of our exploratory ERD models included seasonality in leaf area fraction, but we did not select them as they worsened the fit with growth data and evaluation data ($\delta^2H$, Fig.
+Three of our exploratory ERD models included seasonality in leaf area fraction, but we did not select them as they worsened the fit with growth data and evaluation data ($\delta$^2H, Fig.
 S\@ref(fig:other-erd-mods)).
@@
 @Schwantes:2018].
-We note that data availability on soil water retention curve ($\Psi_{soil}$ vs.
-VWC), hydraulic conductivity ($K_{soil}$) by depth, soil moisture by depth, stream discharge and evapotranspiration were important for effective calibration of our 1-D hydrological model.
+We note that data availability on soil water retention curve ($\Psi$~soil~ vs.
+VWC), hydraulic conductivity (K~soil~) by depth, soil moisture by depth, stream discharge and evapotranspiration were important for effective calibration of our 1-D hydrological model.
 We recommend wide-spread data collection of these variables as well as water-table levels in forest-inventory sites to allow for estimation of trees' water environments.
@@
 
-Figure \@ref(fig:ab-fit): Parameters $A$ (panel A) & $B$ (panel B) for the leaf vulnerability curves that were fitted to observed data on $K_{leaf}$ vs.
-$\\Psi_{leaf}$ for `r length(sp.exp.param$sp)` species (Eq.
+Figure \@ref(fig:ab-fit): Parameters $A$ (panel A) & $B$ (panel B) for the leaf vulnerability curves that were fitted to observed data on K~leaf~ vs.
+$\\Psi$~leaf~ for `r length(sp.exp.param$sp)` species (Eq.
 \@ref(eq:vc-exp)) vs.
@@
 
-Figure \@ref(fig:erd-iso): Modelled Effective Rooting Depth (ERD; mean $\pm$ 1SE; m) versus dry season stable isotopic concentration $\delta^2H$ (mean $\pm$ 1SE; ‰) for `r length(unique(erd.iso$sp))` canopy species from Barro Colorado Island.
+Figure \@ref(fig:erd-iso): Modelled Effective Rooting Depth (ERD; mean $\pm$ 1SE; m) versus dry season stable isotopic concentration $\delta$^2H (mean $\pm$ 1SE; ‰) for `r length(unique(erd.iso$sp))` canopy species from Barro Colorado Island.
 Data from [@Meinzer:1999].
@@
 
-Figure \@ref(fig:erd-time): Modelled species Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\Psi_{soil,~z}$ was below $\Psi_{crit}$, defined as $\Psi_{80,~leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for `r length(erd.mrate.sp)` canopy species found on Barro Colorado Island.
+Figure \@ref(fig:erd-time): Modelled species Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\Psi$~soil,z~ was below $\Psi$~crit~, defined as $\Psi$~80,leaf~, over 1990-2015, where z is the soil depth matching species ERD--for `r length(erd.mrate.sp)` canopy species found on Barro Colorado Island.
 Each circle represents a species.
@@
 
-Among the `r length(sp.exp.param$sp)` species with observed data on ($K_{leaf}$ vs.
-$\Psi_{leaf}$) used to build the community-level models of leaf vulnerability curves (Eqs.
+Among the `r length(sp.exp.param$sp)` species with observed data on (K~leaf~ vs.
+$\Psi$~leaf~) used to build the community-level models of leaf vulnerability curves (Eqs.
 \@ref(eq:b-mod)-\@ref(eq:a-mod)), two lacked data on WSG and five on LMA~lamina~.
@@
 *R*^2^ = `r disc.r2`; *p* `r disc.p`, *n* = `r disc.n`).
-We filled the two remaining gaps in LMA using predictions from a linear relationship between LMA and LMA~lamina~, as measured during the smaller, $K_{leaf}$ campaign ($\overline{LMA} = `r lma.cf[1]` + `r lma.cf[2]`LMA_{lamina}$; Adj.
+We filled the two remaining gaps in LMA using predictions from a linear relationship between LMA and LMA~lamina~, as measured during the smaller, K~leaf~ campaign ($\overline{LMA} = `r lma.cf[1]` + `r lma.cf[2]`LMA_{lamina}$; Adj.
 *R*^2^ = `r lma.r2`; *p* `r lma.p`, *n* = `r lma.n`).
@@
 
-```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($Psi_{soil,~z}$) by depth z (m; panels) over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines).
+```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($Psi$~soil,z~) by depth z (m; panels) over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines).
 Vertical gray lines demarcate median dates of the demographic censuses at the BCI 50 ha plot."}
@@
 ```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts}Occurrence of extreme soil droughts by census intervals (horizontal panels) at three representative depths (z): 0.1, 0.6 and 1 m (vertical panels).
-Mean (black lines) and lower half of 95% distribution (grey areas) of soil water potential ($\\Psi_{soil,~z}$) for a given DOY and depth z are shown (same across census intervals).
-$\\Psi_{soil,~z}$ by DOY for a year (colored line) is only shown if for some part of that year $\\Psi_{soil,~z}$ declined below the 2.5 percentile distribution."}
+Mean (black lines) and lower half of 95% distribution (grey areas) of soil water potential ($\\Psi$~soil,z~) for a given DOY and depth z are shown (same across census intervals).
+$\\Psi$~soil,z~ by DOY for a year (colored line) is only shown if for some part of that year $\\Psi$~soil,z~ declined below the 2.5 percentile distribution."}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q5_by_depth.jpeg"))
@@
 
-```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}Vulnerability curves fitted to observed $K_{leaf}$ vs.
-$\\Psi_{leaf}$ data on ", length(sp.exp.param$sp)," tree species from BCI (A & B) with equation \\@ref(eq:vc-exp)--in absolute terms (A, B), as well as, in terms of loss of conductivity (%) (C, D).
+```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}Vulnerability curves fitted to observed K~leaf~ vs.
+$\\Psi$~leaf~ data on ", length(sp.exp.param$sp)," tree species from BCI (A & B) with equation \\@ref(eq:vc-exp)--in absolute terms (A, B), as well as, in terms of loss of conductivity (%) (C, D).
 Each line represents a species, color coded by WSG (A, C), or LMA (B, D), with blue tones indicating smaller values of WSG or LMA and red tones indicating larger values.")}
@@
 ```{r obs-comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-comm-plc-curves}Percent loss of vulnerability curves for species with ERD estimates.
-Each line represents a species, color coded by source of the curve: fitted to observed data for $K_{leaf}$ vs.
-$\\Psi_{leaf}$ are in yellow and those based on scaling relationship with WSG and LMA are in gray.")}
+Each line represents a species, color coded by source of the curve: fitted to observed data for K~leaf~ vs.
+$\\Psi$~leaf~ are in yellow and those based on scaling relationship with WSG and LMA are in gray.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/obs.mod_plccurves_fits_ggplot_no_col.jpeg"))
@@
 ```{r ab-param, results = 'asis', echo = FALSE}
-knitr::kable(ab.table, caption = 'Leaf vulnerability curves parameters A & B, K~max, leaf~ and $\\Psi$ for the " length(erd.sp)" species with ERD estimates.
-Species for which curves were fitted to observed data for $K_{leaf}$ vs.
-$\\Psi_{leaf}$ are identified with Source as Data, while species for which curves were obtained based on trait-proxies are identified with Source as Model', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(ab.table))))
+knitr::kable(ab.table, caption = 'Leaf vulnerability curves parameters A & B, K~max,leaf~ and $\\Psi$ for the " length(erd.sp)" species with ERD estimates.
+Species for which curves were fitted to observed data for K~leaf~ vs.
+$\\Psi$~leaf~ are identified with Source as Data, while species for which curves were obtained based on trait-proxies are identified with Source as Model', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(ab.table))))
 ```

### 2020-10-21 code/manuscript_data_figs.R

@@ -255,26 +255,27 @@ erd.iso_sp_N_by_model <- ml.rsq.combine.sub %>%
 
 p3.2 <- ggplot(ml.rsq.combine.sub,
                aes(x = Xylem_sap_deltaD_permil, y = depth)) +
-  coord_cartesian(ylim = c(13, 0.3)) +
+  # coord_cartesian(ylim = c(13, 0.3)) +
   geom_smooth(data = ml.rsq.combine.sub %>% subset(significant), aes(group = models.plot1),
               method = "lm", se = TRUE, color = "black", size = 0.5, formula = formula) +
   geom_errorbarh(aes(xmax = Xylem_sap_deltaD_permil + se,
                      xmin = Xylem_sap_deltaD_permil - se, color = s.names),
                  size = 0.5, height = 0.05) +
-  geom_errorbar(aes(ymax = depth + depth.se, ymin = depth - depth.se, color = s.names), size = 0.5) +
+  geom_errorbar(aes(ymax = depth + depth.se, ymin = depth - depth.se, color = s.names), size = 0.5, width = 0.2) +
   facet_wrap( ~ models.plot1, nrow = 2) +
-  geom_text(data = erd.iso_sp_N_by_model, aes(x = -60, y = 0.3, label = paste0("n = ", N), group = models.plot1),
-            vjust = "inward", hjust = "inward", inherit.aes = FALSE) +
+  geom_text(data = erd.iso_sp_N_by_model, aes(x = -60, y = 0.01,
+                                              label = sprintf('italic(n)~"="~%.2f', N), group = models.plot1),
+            parse = TRUE, vjust = "inward", hjust = "inward", inherit.aes = FALSE) +
   ylab(expression("Effective Rooting Depth (m)")) + xlab(xylem.label) +
-  scale_y_continuous(trans="reverse", breaks = unique(ml.rsq.combine$depth)) +
+  scale_y_continuous(trans="reverse", breaks = unique(ml.rsq.combine.sub$depth)) +
   stat_poly_eq(aes(label = paste(..rr.label..)),
-               npcx = 0.98, npcy = 0.15, rr.digits = 2,
+               npcx = 0.95, npcy = 0.15, rr.digits = 2,
                formula = formula, parse = TRUE, size = 4) +
   stat_fit_glance(method = 'lm',
                   method.args = list(formula = formula),
                   geom = 'text_npc',
[…]

### 2020-10-21 manuscript.rmd
1. Changed title 2. TRied to add para first line indentation 2. multi-line equation made to work with word 3. other minor formatting

@@
 ---
-title: "Hydraulically-vulnerable trees are buffered by deep-water access in a tropical forest community"
+title: "Hydraulically-vulnerable trees survive on deep-water access in a tropical forest community"
 author: ''
@@
 output:
+  bookdown::pdf_document2: default
   bookdown::html_document2:
@@
         fig_caption: yes
-  bookdown::pdf_document2: default
 bibliography: references.bib
@@
 # Introduction
+
 \     
@@
 
-\
-    Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation to stem water storage, deep water access, increased root production, hydraulic redistribution and embolism resistance.
+\     Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation to stem water storage, deep water access, increased root production, hydraulic redistribution and embolism resistance.
 Deep water access is arguably the most long-standing, yet under-studied mechanism.
@@
 
-\
-    Trade-offs in plant 'traits' shape how plants make a living in their environment leading to alternative ecological strategies.
+\     Trade-offs in plant 'traits' shape how plants make a living in their environment leading to alternative ecological strategies.
 Characterizing the essential constraints necessary to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010; @Bartlett.2019] entails identifying the topography of trait trade-offs in different environments, and how these relate to demographic rates (growth, recruitment and mortality).
@@
 
-\
-    We used a locally parameterized and calibrated 1-D hydrological water balance model to obtain daily dynamics of soil water potential by depth in the whole-soil column and then applied a novel inverse model to estimate species effective rooting depths (ERD), defined as the depth at which soil water potential best explained observed dynamics of species-specific growth for large trees of `r length(erd.sp)` dominant species (Fig.
+\     We used a locally parameterized and calibrated 1-D hydrological water balance model to obtain daily dynamics of soil water potential by depth in the whole-soil column and then applied a novel inverse model to estimate species effective rooting depths (ERD), defined as the depth at which soil water potential best explained observed dynamics of species-specific growth for large trees of `r length(erd.sp)` dominant species (Fig.
 \@ref(fig:schematic)).
@@
 ## Study site description
-\
-We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and data on stable water isotopes on tree species and soil from an independent study of tree water-sourcing depths [@Meinzer:1999].
+\     We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and data on stable water isotopes on tree species and soil from an independent study of tree water-sourcing depths [@Meinzer:1999].
 The 1,560ha island was formed after the river Chagres was dammed in 1914 to create Gatun Lake and complete the Panama Canal.
@@
 
-\
-In the 50 ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm dbh size class [@Condit.2017].
+\     In the 50 ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm dbh size class [@Condit.2017].
 This inventory represents 321 woody species across the eight censuses.
@@
 ### Model description
-\
-To estimate dynamics of water availability by depth over the forest’s rooting zone, we used the Energy Exascale Earth System Land Model (ELM), coupled with the Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES).
+\     To estimate dynamics of water availability by depth over the forest’s rooting zone, we used the Energy Exascale Earth System Land Model (ELM), coupled with the Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES).
 ELM is a land-model that, among many features, simulates the physics and conservative dynamics of water, energy and carbon fluxes.
@@
 
-\
-ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\Delta W_{can}$, surface water $\Delta W_{sfc}$, soil water $\Delta W_{liq, i}$ at depth z~i~ and water in an unconfined aquifer $\Delta W_{a}$ (omitting processes relevant to snow, wetlands or lakes), and conservation of these terms is expressed as follows:
+\     ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\Delta W_{can}$, surface water $\Delta W_{sfc}$, soil water $\Delta W_{liq, i}$ at depth z~i~ and water in an unconfined aquifer $\Delta W_{a}$ (omitting processes relevant to snow, wetlands or lakes), and conservation of these terms is expressed as follows:
 
@@
 ### Model parameterization
-\
-To parameterize ELM-FATES, we identified eleven parameters relevant for Eq.
+\     To parameterize ELM-FATES, we identified eleven parameters relevant for Eq.
 \@ref(eq:water-balance), determined their ranges based on literature for the study site or for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling (LHS) of this global parameter space.
@@
 \
-We ran ELM-FATES for the 5000-member parameter ensemble with hourly climate drivers from 1985-2018 [@Faybishenko.2020] initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode.
+|   We ran ELM-FATES for the 5000-member parameter ensemble with hourly climate drivers from 1985-2018 [@Faybishenko.2020] initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode.
 As our key interest here was in deriving soil water availability, we ran ELM-FATES in a lower-complexity configuration: static stand structure, and with a single plant functional type (PFT) of evergreen trees.
@@
 
-\begin{align*}
-\overline{O} = \frac{1}{6}\sum \left(RMSE^{*}(ET,~E_{v} + E_{g}) + RMSE^{*}(q_{run},~q_{over} + q_{drai}) + RMSE^{*}(\theta_{0-0.15m}, {W_{liq,~\overline{z}}})\\
-+ RMSE^{*}(\theta_{0.15m}, W_{liq,~`r soil.depths[4]`m}) + RMSE^{*}(\theta_{0.4m}, W_{liq,~`r soil.depths[6]`m}) + RMSE^{*}(\theta_{1 m}, W_{liq,~`r soil.depths[8]`m})\right)
+<!-- # For html -->
+<!-- \begin{align*} -->
+<!-- \overline{O} = \frac{1}{6}\sum RMSE^{*}(ET,~E_{v} + E_{g}) + RMSE^{*}(q_{run},~q_{over} + q_{drai}) + RMSE^{*}(\theta_{0-0.15m}, {W_{liq,~\overline{z}}})\\ -->
+<!-- + RMSE^{*}(\theta_{0.15m}, W_{liq,~`r soil.depths[4]`m}) + RMSE^{*}(\theta_{0.4~m}, W_{liq,~`r soil.depths[6]`~m}) + RMSE^{*}(\theta_{1~m}, W_{liq,~`r soil.depths[8]`~m}) -->
+<!-- (\#eq:obj-func) -->
+<!-- \end{align*} -->
+
+\begin{equation}
+\begin{split}
+\overline{O} = \frac{1}{6}\sum \left(RMSE^{*}(ET,~E_{v} + E_{g}) + RMSE^{*}(q_{run},~q_{over} + q_{drai}) + RMSE^{*}(\theta_{0-0.15m}, {W_{liq,~\overline{z}}}) \\
+& + RMSE^{*}(\theta_{0.15m}, W_{liq,~`r soil.depths[4]`m}) + RMSE^{*}(\theta_{0.4~m}, W_{liq,~`r soil.depths[6]`~m}) + RMSE^{*}(\theta_{1~m}, W_{liq,~`r soil.depths[8]`~m})\right)
 (\#eq:obj-func)
-\end{align*}
+\end{split}
+\end{equation}
 
@@
 We identified scaling relationships between fitted parameters $A$ and $B$ in Eq.
-\@ref(eq:vc-exp) and two traits, namely, WSG, the wood specific gravity after drying at 100^$\\circ$^C and LMA, the leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules for leaves receiving direct sunlight (@Wright:2010; see SI for details).
+\@ref(eq:vc-exp) and two traits, namely, WSG, the wood specific gravity after drying at 100^$\circ$^C and LMA, the leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules for leaves receiving direct sunlight (@Wright:2010; see SI for details).
 We fitted polynomial equations described as,
@@
 Figure \@ref(fig:ab-fit): Parameters $A$ (panel A) & $B$ (panel B) for the leaf vulnerability curves that were fitted to observed data on K~leaf~ vs.
-$\\Psi$~leaf~ for `r length(sp.exp.param$sp)` species (Eq.
+$\Psi$~leaf~ for `r length(sp.exp.param$sp)` species (Eq.
 \@ref(eq:vc-exp)) vs.
@@
 Each circle is color coded by the Spearman's *r* with red shades for positive *r* values, while blue shades for negative values, respectively.
-*p*-values are given only when they are $\\ge$ 0.05.
-Units are as follows: ERD in m, K~max~ in mmol m^-1^ s^-1^ MPa^-1^), all variables with $\\Psi$ in MPa.
+*p*-values are given only when they are $\ge$ 0.05.
+Units are as follows: ERD in m, K~max~ in mmol m^-1^ s^-1^ MPa^-1^), all variables with $\Psi$ in MPa.
 
@@
 w: vertical wind speed; t: air temperature; q: water vapor density; c: CO~2~ density; u: horizontal wind speed.
-The bar indicates 30-min averaging and the apex indicates fluctuation ($\\overline{x` = x - \\bar{x}}$)', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(eddy.table))))
+The bar indicates 30-min averaging and the carat indicates fluctuation ($\\overline{\\hat{x} = x - \\bar{x}}$)', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(eddy.table))))
 ```

### 2020-10-22 manuscript.rmd
1. Updated description of LAI seasonality

@@
 output:
-  bookdown::pdf_document2: default
   bookdown::html_document2:
@@
     fig_caption: yes
+  bookdown::pdf_document2: default
   bookdown::word_document2:
@@
 \
-To explore the effect of species seasonality in leaf phenology on growth, we assumed species-specific seasonal curves for leaf area index ($LAI$, range 0-1, unitless).
-For evergreen species we assumed $LAI$ = 1 throughout the year, whereas for non-evergreen species (various deciduous classes), we assumed $LAI$ = 1 during the wet season (DOY = 150 to 300) given the high forest-level leaf area index for this period (not shown), and obtained an estimate of $LAI$ for the rest of the year as follows.
-We interpolated species-specific weekly leaf-fall data collected at two similar sites in BCI (1985-2020) to a daily estimate, scaled to express daily leaf fall as a fraction of the total leaf fall in a given year and obtained an average leaf fall fraction by DOY estimate across years ($LeafFall$, range 0-1, unitless).
-Thus for non-evergreen species, we cumulatively deducted $LeafFall$ from $LAI$ = 1 beginning DOY = 300 and obtained a $LAI$ by DOY curve (Fig.
-S\@ref(fig:leaf-cover)).
+To explore the effect of species seasonality in leaf phenology on growth, we assumed species-specific seasonal curves for leaf area index normalised between 0 and 1 (LAI, unitless), using a combination of observations for long-term records of leaf-fall and leaf-lifespan (see SI).
+
 We obtained species-specific parameters $A$ & $B$ for Eq.
@@
 
+### Leaf Area Index
+
+We generated an estimate of the mean seasonality curve for normalised leaf area index (LAI$*$, range 0-1) for individual species as follows.
+We interpolated species-specific weekly leaf-fall data collected at two similar sites in BCI (1985-2020) to a daily estimate.
+We lead this daily leaf-fall time series by the species-specific average leaf-lifespan (in days) and used it as an estimate of the daily leaf-gain time-series.
+The difference between leaf-fall and leaf-gain gave an estimate of daily leaf-change time series, which was then scaled to express as a fraction of the total leaf fall in a given year.
+Beginning with the LAI$*$ of 1 at the end of the wet season in the first year of the record, we acccumulated leaf-change over it to obtain a running estimate of the leaf-cover fraction, avaraged it by DOY across years and normalised it to obtain the species-specific seasonality curve in LAI$*$ (range 0-1, unitless).
+[@Osnas:2018].
+(Fig.
+S\@ref(fig:leaf-cover)).
+
+For five species, an estimate of leaf lifespan was available from direct observations [@Osnan:2018].
+For the rest erd.sp.wo.ll species, leaf lifespan was predicted from LMA ($\overline{lifespan} = `r lma.ll.cf[1]` + `r lma.ll.cf[2]`LMA$; *R*^2^ = `r lma.ll.r2`; *p* `r lma.ll.p`, *n* = `r lma.ll.n`).
+
 ### Leaf hydraulic conductivity and vulnerability to cavitation {#SIData.joseph}

### 2020-10-22 code/15.0_Prepare_data_for_correlative analyses.R

@@ -2277,7 +2277,7 @@ gap.models.ll$LMA.lifetime <- lm(lifetime ~ LMA, data =
 save(gap.models.ll, file = file.path(results.folder, "gap.models.ll.Rdata"))
 
 bci.lifetime <- bci.ll %>%
-  left_join(ll.lma %>%
+  full_join(ll.lma %>%
             subset(strata == "CANOPY" & site == "FTS" & lifeform6 != "LIANA") %>%
               select(sp, lifetime), by = "sp") %>%
   mutate(lifetime.filled = lifetime,
@@ -2289,6 +2289,7 @@ bci.lifetime <- bci.ll %>%
 
 ## there shouldn't be any duplicate records for species
 # bci.lifetime[duplicated(bci.lifetime$sp),]
+save(bci.lifetime, file = file.path(results.folder, "bci.lifetime.Rdata"))
 
 formula <- y ~ x
 ll.lma.comm <- ggplot(bci.lifetime, aes(x = LMA, y = lifetime.filled)) +

### 2020-10-22 manuscript.rmd
1. model fit for leaf lifetime from LMA

@@
 \end{equation} 
-where $H$ $\in$ {1, .., 100}, refer to the 100 hydrological realisations and $Z$ $\in$ 0.5, {`r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m, covering the range of soil depths in ELM-FATES after above parametarision [@Chonggang, Depths 0.21, 0.37 and 0.62 were all labelled as 0.5 m for model evaluation.
+where $H$ $\in$ {1, .., 100}, refer to the 100 hydrological realisations and $Z$ $\in$ 0.5, {`r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m, covering the range of soil depths in ELM-FATES after above parametarision [AT:Chonggang, So that we can perform model evaluation, depths 0.21, 0.37 and 0.62 were all taken to be as 0.5 m (not averaged).
 How do I explain that?
-Do they need to be retained as their original depths?].
+So hereafter 0.21, 0.37 and 0.62 were not retained, but only 0.5 m.
+How do I explain that?
+Or should I retain them?].
 
@@
 We generated an estimate of the mean seasonality curve for normalised leaf area index (LAI$*$, range 0-1) for individual species as follows.
-We interpolated species-specific weekly leaf-fall data collected at two similar sites in BCI (1985-2020) to a daily estimate.
+We interpolated species-specific weekly leaf-fall data collected at two similar sites in BCI (1985-2020) to a daily estimate [@Wright.1990aeo].
 We lead this daily leaf-fall time series by the species-specific average leaf-lifespan (in days) and used it as an estimate of the daily leaf-gain time-series.
@@
 Beginning with the LAI$*$ of 1 at the end of the wet season in the first year of the record, we acccumulated leaf-change over it to obtain a running estimate of the leaf-cover fraction, avaraged it by DOY across years and normalised it to obtain the species-specific seasonality curve in LAI$*$ (range 0-1, unitless).
-[@Osnas:2018].
 (Fig.
@@
 
-For five species, an estimate of leaf lifespan was available from direct observations [@Osnan:2018].
-For the rest erd.sp.wo.ll species, leaf lifespan was predicted from LMA ($\overline{lifespan} = `r lma.ll.cf[1]` + `r lma.ll.cf[2]`LMA$; *R*^2^ = `r lma.ll.r2`; *p* `r lma.ll.p`, *n* = `r lma.ll.n`).
+For five species, an estimate of leaf lifespan was available from direct observations [@Osnas.2018].
+For the rest erd.sp.wo.ll species, leaf lifespan was predicted from LMA ($\overline{lifespan} = `r ll.lma.cf[1]` + `r ll.lma.cf[2]`LMA$; *R*^2^ = `r ll.lma.r2`; *p* `r ll.lma.p`, *n* = `r ll.lma.n`).
 

### 2020-10-22 code/manuscript_data_figs.R

@@ -16,6 +16,8 @@ load(file = file.path(results.folder, "obs.data.model.AB.Rdata"))
 load(file = file.path(results.folder, "data.model.AB.sub.Rdata"))
 load(file = file.path(results.folder, "obs.sp.vcurves.1.Rdata"))
 load(file = file.path(results.folder, "comm.sp.vcurves.1.Rdata"))
+load(file = file.path(results.folder, "bci.lifetime.Rdata"))
+
 
 #****************************
 ###   Custom Functions   ####
@@ -885,3 +887,8 @@ matched.rows.2 <- match(ab.table$Family, correct.family.2$misspelt)
 ab.table$Family[rows.to.replace.2] <-
   correct.family.2$correct[matched.rows.2[!is.na(matched.rows.2)]]
 
+#****************************
+## species with lifespan----
+#****************************
+erd.sp.with.ll <- length(erd.sp[erd.sp %in% unique(bci.lifetime$sp[!is.na(bci.lifetime$lifetime)])])
+erd.sp.wo.ll <- length(erd.sp[!erd.sp %in% unique(bci.lifetime$sp[!is.na(bci.lifetime$lifetime)])])

### 2020-10-22 code/load.R
1. model fit for leaf lifetime from LMA

@@ -127,14 +127,11 @@ load(file = file.path(results.folder, "data.model.AB.Rdata"))
 load(file = file.path(results.folder, "sp.soft.filled.Rdata"))
 load(file = file.path(results.folder, "sp.exp.param.Rdata"))
 #******************************************************
-### Load Leaf Cohort tracking data from the crane sites------
+### Load Leaf Life-span and Cover estiamtes------
 #******************************************************
 load(file = file.path(results.folder, "sp.leaf_cover.Rdata"))
 # load(file = file.path(results.folder, "sp.leaf_cover.mean.Rdata")) # without leaf gain
 load(file = file.path(results.folder, "sp.leaf_cover.for.model.Rdata"))
-
-load(file = file.path(results.folder, "cohort.Rdata"))
-load(file = file.path(results.folder, "coh.sp.summ.Rdata"))
 
 soil.depths <- unique(psi$depth)
 
@@ -243,8 +240,18 @@ a.wsg.cf <- as.numeric(round(k_by_psi.models$A.WSG100$coefficients, 2))
 
 vpd.cf <- as.numeric(round(gpp.models$eq.gpp.vpd$coefficients, 2))
 
+#************************
+### Leaf lifespan from LMA model
+#************************
+
+ll.lma.cf <- as.numeric(round(gap.models.ll$LMA.lifetime$coefficients, 2))
+ll.lma.r2 <- round(summary(gap.models.ll$LMA.lifetime)$r.squared, 2)
+ll.lma.n <- length(gap.models.ll$LMA.lifetime$residuals)
+ll.lma.p <- ifelse(broom::glance(gap.models.ll$LMA.lifetime)$p.value < 0.001,
+                  paste0("< 0.001"), paste0("= ", signif(broom::glance(gap.models.ll$LMA.lifetime)$p.value, 2)))
+
 #******************************************************
-### Tables
+### Tables-----
 #******************************************************
 hypo.table <-
   data.frame(

### 2020-10-23 manuscript.rmd
1. added second para indentation 2. Removed spaces between set symbols such as i~|~z

@@
 We validated ERD against existing tree and soil stable water isotope data.
-- Across species, deeper ERD was associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates over 35 years.
+- Across species, deeper ERD was associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates during extreme droughts.
 Species with deeper ERD spent less time below critical hydraulic thresholds indicating that trees compensate for water-stress related mortality risk through deeper water access.
@@
 # Introduction
-
 \
 Tropical forests play a disproportionately large role in the global carbon and energy cycles [@Bonan:2008], and support half of global biodiversity [@Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019].
-Tropical forests are logically considered to be an especially drought-vulnerable among biomes given the combination of climate risk and vegetation sensitivity [@Meir:2015trw], yet mortality events in the tropics are rarely as large as those in the temperate and boreal zones (McDowell et al.
-2018), leading to questions regarding the role of large trait-diversity and hydraulic strategies in mitigating mortality events.
+Tropical forests are logically considered to be an especially drought-vulnerable among biomes given the combination of climate risk and vegetation sensitivity [@Meir:2015trw], yet mortality events in the tropics are rarely as large as those in the temperate and boreal zones (@McDowell.2018), leading to questions regarding the role of large trait-diversity and hydraulic strategies in mitigating mortality events.
 Despite this, widespread drought-induced tree mortality has occurred across the tropics for specific functional groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010] and tropical mortality rates are increasing in some regions [@Brienen.2015; @Hubau.2020].
@@
 
-\     Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation to stem water storage, deep water access, increased root production, hydraulic redistribution and embolism resistance.
+\     Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation to stem water storage, deep water access, increased root production, hydraulic redistribution and embolism resistance[@McDowell.2008].
 Deep water access is arguably the most long-standing, yet under-studied mechanism.
@@
 \@ref(fig:schematic)).
-This model incorporated the impact of atmospheric drought on growth and was constrained by above-ground hydraulic traits, namely, species-specific leaf vulnerability curves, by leveraging existing data for 8 species and developing trait-based proxies for the rest.
-We evaluated ERD against existing data on plant and soil water isotopes and tested coordination of ERD with hydraulic traits and mortality rates.
+This model incorporated the impact of atmospheric drought on growth and was constrained by above-ground hydraulic traits, namely, species-specific leaf vulnerability curves, by leveraging existing data for eight species and developing trait-based proxies for the rest.
+We evaluated ERD against existing dueterium to hydrogen content data for a subset of six tree species as an independent observation of ERD, given the corresponding gradient in isotopic composition of water in the soil profile.
+We tested coordination of ERD with hydraulic traits for a subset of eight species with both datasets.
+We finally analysed species' mortality rates across the censuses spanning 35 years to test whether ERD explained risk of drought-induced mortality in periods of extreme hydrological drought.
 Specifically, we tested an overarching hypothesis that inter-specific variation in trees' effective rooting depths (ERD) enables increasingly reliable water availability with depth and thus mitigation of hydraulic vulnerability.
-Our hypotheses are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallow rooted species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deep rooted species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallow rooted species may have higher mortality rates as a result of higher variability water availability in shallow soil layers.
+Our hypotheses are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallow rooted species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deep rooted species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallow rooted species may have higher mortality rates as a result of higher variability in water availability in shallow soil layers.
 To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.
@@
 ## Study site description
-\     We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and data on stable water isotopes on tree species and soil from an independent study of tree water-sourcing depths [@Meinzer:1999].
+\
+We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and data on stable water isotopes on tree species and soil from an independent study of tree water-sourcing depths [@Meinzer:1999].
 The 1,560ha island was formed after the river Chagres was dammed in 1914 to create Gatun Lake and complete the Panama Canal.
@@
 ### Model description
-\     To estimate dynamics of water availability by depth over the forest’s rooting zone, we used the Energy Exascale Earth System Land Model (ELM), coupled with the Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES).
+\
+To estimate dynamics of water availability by depth over the forest’s rooting zone, we used the Energy Exascale Earth System Land Model (ELM), coupled with the Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES).
 ELM is a land-model that, among many features, simulates the physics and conservative dynamics of water, energy and carbon fluxes.
@@
 
-\     ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\Delta W_{can}$, surface water $\Delta W_{sfc}$, soil water $\Delta W_{liq, i}$ at depth z~i~ and water in an unconfined aquifer $\Delta W_{a}$ (omitting processes relevant to snow, wetlands or lakes), and conservation of these terms is expressed as follows:
+\     ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\Delta W$~can~, surface water $\Delta W$~sfc~, soil water $\Delta W$~liq,z~ at depth z and water in an unconfined aquifer $\Delta W_{a}$ (omitting processes relevant to snow, wetlands or lakes).
+Conservation of these terms is expressed as follows:
 
@@
 
-where, $q_{rain}$ is rainfall, $E_{v}$ is evapotranspiration (ET) of the forest, $E_{g}$ is ground evaporation, $q_{over}$ is surface runoff, $q_{h2osfc}$ is runoff from surface water storage, $q_{drai}$ is sub-surface drainage (all in mm H~2~O) and $\Delta t$ is the time-step (typically, 1 hour).
-The soil profile is discretized into 15 exponentially distributed soil layers with depth z $\simeq$ 0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13, 21 and 35 (m).
+where, $q$~rain~ is rainfall, $E_{v}$ is transpiration of the forest, $E_{g}$ is ground evaporation, $q$~over~ is surface runoff, $q$~h2osfc~ is runoff from surface water storage, $q$~drai~ is sub-surface drainage (all in mm H~2~O) and $\Delta t$ is the time-step (typically, 1 hour).
+The soil profile is discretized into 15 exponentially distributed soil layers with depth z, where z $\in$ Z, and Z = {0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13, 21, 35} m.
 
 ### Model parameterization
-\     To parameterize ELM-FATES, we identified eleven parameters relevant for Eq.
-\@ref(eq:water-balance), determined their ranges based on literature for the study site or for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling (LHS) of this global parameter space.
+\
+To parameterize ELM-FATES, we identified eleven parameters relevant for Eq.
+\@ref(eq:water-balance), determined their ranges based on literature for the study site, when available, else for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling (LHS) of this global parameter space.
 In the Conrad catchment, existing data on soil hydraulic conductivity for the depths of 12.5 and 60 cm showed a decline of conductivity by depth (@Kinner:2004; Table \@ref(tab:param)).
@@
 where $\Psi_e$ is the air-entry matric potential, $\theta_{s}$ is the saturated volumetric water content, and $b$, an index for soil pore-size distribution, respectively.
-$\Psi_e$, $\theta_{s}$ and $b$ taht best-fit the data were 200 (mm H~2~O), 0.51 (cm^3^ cm^-3^) and 10 (unitless), respectively.
+$\Psi_e$, $\theta_{s}$ and $b$ that best-fit the data were 200 (mm H~2~O), 0.51 (cm^3^ cm^-3^) and 10 (unitless), respectively.
 
-\
-|   We ran ELM-FATES for the 5000-member parameter ensemble with hourly climate drivers from 1985-2018 [@Faybishenko.2020] initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode.
+\   We ran ELM-FATES for the 5000-member parameter ensemble with hourly climate drivers from 1985-2018 [@Faybishenko.2020] initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode.
 As our key interest here was in deriving soil water availability, we ran ELM-FATES in a lower-complexity configuration: static stand structure, and with a single plant functional type (PFT) of evergreen trees.
@@
 We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq.
-\@ref(eq:water-balance), namely, (i) evapotranspiration ($ET \simeq E_{v} + E_{g}$) from the eddy flux tower (2012-2017; see [SI](#SIData.matteo)), (ii) stream discharge from the Conrad catchment ($q_{run} \simeq q_{over} + q_{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC, $\theta_{z} \simeq W_{liq, z}$) from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we obtained with a vertical TDR probe over 0-15 cm depth ($\theta_{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m (1299 samples in total covering all soil types and habitats; see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\theta_{0.15 m}$, $\theta_{0.4m}$ and $\theta_{1 m}$ refer to plot-wide averages).
+\@ref(eq:water-balance), namely, (i) evapotranspiration (ET) from the eddy flux tower (2012-2017; see [SI](#SIData.matteo)), comparable to $E_{v} + E_{g}$ in Eq.
+\@ref(eq:water-balance)), (ii) stream discharge $q_{run} from the Conrad catchment, comprabel to $q_{over} + q_{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC or $\theta_{z}) comparable to W_{liq, z}$) from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we obtained with a vertical TDR probe over 0-15 cm depth ($\theta_{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m with 1299 samples in total covering all soil types and habitats (see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\theta_{0.15 m}$, $\theta_{0.4 m}$ and $\theta_{1 m}$ refer to plot-wide averages).
 
@@
 \begin{split}
-\overline{O} = \frac{1}{6}\sum \left(RMSE^{*}(ET,~E_{v} + E_{g}) + RMSE^{*}(q_{run},~q_{over} + q_{drai}) + RMSE^{*}(\theta_{0-0.15m}, {W_{liq,~\overline{z}}}) \\
-& + RMSE^{*}(\theta_{0.15m}, W_{liq,~`r soil.depths[4]`m}) + RMSE^{*}(\theta_{0.4~m}, W_{liq,~`r soil.depths[6]`~m}) + RMSE^{*}(\theta_{1~m}, W_{liq,~`r soil.depths[8]`~m})\right)
+\overline{O} = \frac{1}{6}\sum \left(RMSE^{*}(ET,E_{v} + E_{g}) + RMSE^{*}(q_{run},q_{over} + q_{drai}) + RMSE^{*}(\theta_{0-0.15m}, {W_{liq,\overline{z}}}) \\
+& + RMSE^{*}(\theta_{0.15m}, W_{liq,`r soil.depths[4]`m}) + RMSE^{*}(\theta_{0.4~m}, W_{liq,`r soil.depths[6]`~m}) + RMSE^{*}(\theta_{1~m}, W_{liq,`r soil.depths[8]`~m})\right)
 (\#eq:obj-func)
@@
 
-based on monthly sums for both $ET$ and $q_{run}$ and daily averages for $\theta$s.
-$W_{liq,~\overline{z}}$ is the average of daily $W_{liq,~z}$ for $z$ $\in$ {`r paste(soil.depths[1:4], collapse=", ")`} m.
+based on monthly sums for both ET and $q$~run~ and daily averages for $\theta$s.
+$W$~liq,~$_{\overline{z}}$ is the average of daily $W$~liq,z~, where z $\in$ Z, and Z = {`r paste(soil.depths[1:4], collapse=", ")`} m.
 Superscript * indicates that RMSE is standardised between 0 and 1 across the 5000 simulations.
 
-Preliminary analyses showed that minimising the objective function $\overline{O}$, that is, simultaneously fitting the model to all states and fluxes did not capture the soil moisture dynamics at 1 m depth well, our key priority for model calibration.
-To ensure model fit with soil moisture dynamics at 1 m depth, we first chose 300 parameter ensemble members that minimised RMSE between observations and simulations for VWC at the depth of 1m ($RMSE^{*}(\theta_{1 m}, W_{liq,~`r soil.depths[8]`m})$ < 0.2), and among those 300 ensemble members we chose 100 members with the lowest values for the objective function $\overline{O}$.
+\   Preliminary analyses showed that minimising the objective function $\overline{O}$, that is, simultaneously fitting the model to all states and fluxes did not capture the soil moisture dynamics at 1 m depth well, our key priority for model calibration.
+To ensure model fit at 1 m depth, we first chose 300 parameter ensemble members that minimised RMSE between observations and simulations for VWC at the depth of 1 m ($RMSE^{*}(\theta_{1~m}, W_{liq,`r soil.depths[8]`~m}) < 0.2$), and among those 300 ensemble members we chose 100 members with the lowest values for the objective function $\overline{O}$.
 See Table \@ref(tab:param) for the ranges of best-fit values for different parameters.
@@
 
-\
-We ran ELM-FATES with the chosen 100 ensemble members from 1985-2018, with the first 5 years used for model spin-up.
+## Hydrological droughts
+\   We ran ELM-FATES with the chosen 100 ensemble members from 1985-2018, with the first 5 years used for model spin-up.
+Extreme soil droughts were identified by depth *z* as days that fell below the 5 percentile of mean $\Psi$~soil,*z*~ across all hydrological realisations values at that depth.
 
@@
 \
-To estimate effective rooting depth, we tested alternate models for growth (Eqs.
+To estimate effective rooting depth, we tested alternate models for growth (Eq.
 \@ref(eq:M-chosen) and those in SI), which use effective rooting depth as an input.
@@
 In our models K~leaf~ therefore incorporates the effects of rooting depth and rhizosphere water potential.
-For each species' growth dynamics across census intervals was regressed against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, K~leaf~ and the leaf area index (LAI), all standardized between 0 and 1.
+For each species', growth dynamics across census intervals was regressed against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, K~leaf~ and the leaf area index (LAI), all standardized between 0 and 1.
 Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting.
@@
 
-For the chosen model (see Model validation and selection below), daily average growth for species $s$, $G_{s,~t~|~z}$, in the census interval $t$ for depth $z$ and for a given hydrological realisation $h$ is described as follows:
+\    For the chosen model (see Model validation and selection below), daily average growth for species $s$, $G_{s,t|z}$, in the census interval $t$ for depth $z$ and for a given hydrological realisation $h$ is described as follows:
 
 \begin{equation}
-G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,~i~|~z}^{*} VPD_{i}^{*}) + \epsilon_{t~|~z}
+G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,i|z}^{*} VPD_{i}^{*}) + \epsilon_{t|z}
 (\#eq:M-chosen)
@@
 
-where n~t~ be the total number of days in census interval *t*, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is a normally distributed model error term.
+where n~t~ be the total number of days in census interval *t*, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is the model error term.
 See SI for equations for all the alternate models tested.
 
-As VPD may have a non-linear relationship with growth, with growth increasing for the lower range of VPD, and decreasing for the higher range of VPD due to stomatal closure (reference), we used a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see Model parametarisation) to represent the impact of VPD on growth and used GPP standardised between 0 and 1 (hereafter, VPD*) in Eq.
+\    As VPD may have a non-linear relationship with growth, with growth increasing with VPD intially, but decreasing with VPD beyond an optimal value due to stomatal closure [AT:Chonggang, add a reference], we used a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see Model parametarisation) to represent the impact of VPD on growth and used GPP standardised between 0 and 1 (hereafter, VPD*) in Eq.
 \@ref(eq:M-chosen).
 
-For each species, we estimated diurnal maximum value of $K~{leaf,~s,~i~|~z,~h}$ for a species $s$ by substituting $\Psi$~soil,~i~|~z,~h~ in the leaf vulnerability curves defined below in Eq.
+\    For each species, we estimated diurnal maximum value of K~leaf~$_{,s,i|z,h}$ for a species $s$ by substituting $\Psi$~soil~$_{,s,i|z,h}$ in the leaf vulnerability curves defined below in Eq.
 \@ref(eq:vc-exp) using species-specific parameters $A_{s}$ and $B_{s}$ as:
@@
 \begin{equation}
-K_{leaf,~s,~i~|~z,~h} = A_{s}e^{-B_{s}\Psi_{soil,~i~|~z,~h}}
+K_{leaf,s,i|z,h} = A_{s}e^{-B_{s}\Psi_{soil,i|z,h}}
 (\#eq:k-soil)
@@
 
-By varying ‘z’, we estimated species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi$~soil,z~) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015).
+\    By varying *z*, we estimated species effective rooting depths (ERD), defined as the depth *z* at which soil water dynamics ($\Psi$~soil,*z*~) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015).
 Using Eq.
-\@ref(eq:M-chosen), for a species $s$, we obtained likelihood based on Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $r$ for each soil depth $z$ for each of the 100 best-fit hydrological realizations.
-For each hydrological realization we retained only one depth that corresponded to the least AIC score (the best-fit), and within those depths retained the ones that had Pearson's $r$ >= 0.71 (or, $r^2$ >= 0.5) as a valid set.
-For a species $s$, we defined effective rooting depth (ERD) as the median of these set of depths, described as :
+\@ref(eq:M-chosen), for a species *s*, we obtained likelihood based on Akaike's Information Criterion (AIC) and Pearson's correlation coefficient *r* for each soil depth *z* for each of the 100 best-fit hydrological realizations.
+For each hydrological realization *h* we retained only one depth that had the maximum likelihood (the minimum AIC score), and within those depths retained the ones that had Pearson's *r* >= 0.71 (or, *R*^2 >= 0.5) as a valid set.
+For a species *s*, we defined effective rooting depth (ERD) as the median of these set of the retained depths, which can be described as :
 
 \begin{equation}
-ERD_{s} = \underset{\forall \, h \, \in \, H}{median} \Big\{z,h\,\Big|\, \Big(Likelihood_{s~|~z,~h} = \underset{\forall \, z \,\in\,Z}{maximum}(Likelihood_{s~|~z,~h})\Big) \cap \Big(r_{s~|~z,~h} \ge 0.71\Big)\Big\}
+ERD_{s} = \underset{\forall \, h \, \in \, H}{median} \Big\{z,h\,\Big|\, \Big(Likelihood_{s|z,h} = \underset{\forall \, z \,\in\,Z}{maximum}(Likelihood_{s|~z,h})\Big) \cap \Big(r_{s|z,h} \ge 0.71\Big)\Big\}
 (\#eq:erd)
 \end{equation} 
-where $H$ $\in$ {1, .., 100}, refer to the 100 hydrological realisations and $Z$ $\in$ 0.5, {`r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m, covering the range of soil depths in ELM-FATES after above parametarision [AT:Chonggang, So that we can perform model evaluation, depths 0.21, 0.37 and 0.62 were all taken to be as 0.5 m (not averaged).
+where $H$ = {1, .., 100}, refer to the 100 hydrological realisations and $Z$ = {0.5, `r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m, covering the range of soil depths in ELM-FATES after above parametarision [AT:Chonggang, so that we can perform model evaluation, depths 0.21, 0.37 and 0.62 were all taken as if they were 0.5 m (not averaged).
 How do I explain that?
-So hereafter 0.21, 0.37 and 0.62 were not retained, but only 0.5 m.
-How do I explain that?
-Or should I retain them?].
+So hereafter 0.21, 0.37 and 0.62 were not retained, but only 0.5 m.].
 
@@
 \
-For growth estimates, to minimize the effect of light variation among trees, we selected only large trees ($\ge$ 30 cm dbh, the diameter at breast height) and also species whose maximum height was $\ge$ 30 m (hereafter, canopy species) and thus are likely to be fully exposed to the sun.
+For growth estimates, to minimize the effect of light variation among trees, we selected only large trees ($\ge$30 cm dbh, the diameter at breast height) and also species whose maximum height was $\ge$30 m (hereafter, canopy species) and thus are likely to be fully exposed to the sun.
 We selected trees for which the height of the diameter measurements were unchanged across all censuses (1990, 1995, 2000, 2005, 2010, 2015).
-For each individual tree (*i*), we calculated absolute growth rates ($G_{i,t}$; cm yr^-1^) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses, divided by the inter-census interval based on the exact date of measurements.
-We eliminated positive outliers as any growth rates amounting to > 75 mm yr^-1^, as this was the highest rate observed with confidence for fastest growing species [@Condit.2017].
+For each individual tree (*i*), we calculated absolute growth rates ($G_{i,t}$, cm yr^-1^) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses prorated by the inter-census interval based on the exact date of measurements.
+We eliminated positive outliers as any growth rates amounting to >75 mm yr^-1^, as this was the highest rate observed with confidence for fastest growing species [@Condit.2017].
 Negative outliers were removed based on known measurement error, $\sigma_{e} = 0.006214d + 0.903$, where *d* is dbh.
-Any stem whose later dbh measurement fell < 4$\sigma_{e}$ the earlier measurement was removed [@Condit.2017].
-As tree growth rates vary with size [@Muller-Landau:2006], to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($Resid_{i,t} = G_{i,t} - f(d_{i, t})$).
-The diameter model of growth ($f(d_{i, t})$), a B-spline based polynomial equation with 5 degrees of freedom, was fitted to growth rates from all trees pooled together as there was insufficient species-specific data for large trees (Fig.
+Any stem whose later dbh measurement fell <4$\sigma_{e}$ the earlier measurement was removed [@Condit.2017].
+As tree growth rates vary with size [@Muller-Landau:2006], to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($Resid_{i,t} = G_{i,t} - f(d_{i,t})$).
+The diameter model of growth, $f(d_{i, t})$, a B-spline based polynomial equation with five degrees of freedom, was fitted to growth rates from all trees pooled together as there was insufficient species-specific data for large trees (Fig.
 S\@ref(fig:dbh-model)).
@@
 \begin{equation}
-K_{leaf,~s} = A_{s}e^{-B_{s}\Psi_{leaf_{s}}} + error
+K_{leaf,s} = A_{s}e^{-B_{s}\Psi_{leaf_{s}}} + error
 (\#eq:vc-exp)
@@
 
-As these fits explained a large proportion of variation in parameters $A$ and $B$ (see results), we sequentially used Eq.
+\   As these fits explained a large proportion of variation in parameters $A$ and $B$ (see results), we sequentially used Eq.
 \@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r sp.n.model` ERD species without direct data and estimated leaf hydraulic vulnerability curves using Eq.
@@
 For all the 31 ERD species we estimated species $\Psi$~50,leaf~ based on their vulnerability curves (see Table \@ref(tab:symbols) for definition).
+
+\   We obtained species-specific parameters $A$ & $B$ for Eq.
+\@ref(eq:k-soil) either from data when available, or a model based on LMA or WSG as described in the earlier section, and used K standardized between 0 and 1 within species (K*) in Eq.
+\@ref(eq:M-chosen).
 
@@
 \
-To explore the effect of species seasonality in leaf phenology on growth, we assumed species-specific seasonal curves for leaf area index normalised between 0 and 1 (LAI, unitless), using a combination of observations for long-term records of leaf-fall and leaf-lifespan (see SI).
-
-We obtained species-specific parameters $A$ & $B$ for Eq.
-\@ref(eq:k-soil) either from data when available, or a model based on LMA or WSG as described in the earlier section, and used K standardized between 0 and 1 within species (K*) in Eq.
-\@ref(eq:M-chosen).
+In the exploratory model for effective rooting depth, we explored the effect of seasonality in leaf area index (LAI, unitless) on growth.
+We assumed species-specific mean seasonal curves for LAI normalised between 0 and 1 (LAI, unitless), informed by a combination of long-term records for leaf-fall [@Wright.1990aeo] and leaf-lifespan [@Osnas.2018] (see SI).
 
@@
 \
-We fitted a polynomial curve between mean midday values (11:00 to 15:00 hrs) of gross primary productivity (gC m^2^ day^-1^) and VPD (kPa) observations we collected from the eddy flux tower located 75 m from the 50-ha plot (for any day *i*, $GPP_i = `r vpd.cf[1]` + `r vpd.cf[2]`VPD_i + `r vpd.cf[3]`VPD_i^2 + error$; 2012-2017; Fig.
-S\@ref(fig:gpp-vpd)) and used $GPP$ standardized between 0 and 1 as $VPD^*$ in Eq.
+We fitted a polynomial curve between monthly average of mean midday values (11:00 to 15:00 hrs) of gross primary productivity (gC m^2^ day^-1^) we collected from the eddy flux tower located 75 m from the 50-ha plot (2012-2017) and VPD (kPa) calculated using the climate forcing data (for any day *i*, $GPP_i = `r vpd.cf[1]` + `r vpd.cf[2]`VPD_i + `r vpd.cf[3]`VPD_i^2 + error$; Fig.
+S\@ref(fig:gpp-vpd)); used VPD to predict GPP from 1990-2015, standardized it between 0 and 1, and used it as $VPD^*$ in Eq.
 \@ref(eq:M-chosen)).
@@
 \
-To validate the ERD models we used data for xylem sap $\delta$^2H (permil) as an independent observation of ERD.
+To validate the ERD models we used data for xylem sap $\delta$^2^H (‰) as an independent observation of ERD.
+We used xylem sap $\delta$^2^H values given in @Meinzer:2001 for the dry-season of March 1997 as this period showed largest seasonal divergence among species, with a corresponding vertical gradient of $\delta$^2^H in the soil (Fig.
+S\@ref(fig:iso-data)).
+Briefly, this method makes use of a naturally formed gradient in deuterium to hydrogen ratio of soil water vertically belowground.
+Generally due to evaporative enrichment (here, during the dry season), shallow soil water has a greater deuterium to hydrogen content than deep water (here, $\delta$^2^H ranging from -16.5 to -50 ‰ over 0.05 to 1 m depth), with ground-water having the lowest deuterium to hydrogen content (here, -60 ‰), typically reflecting un-fractionated rain water.
+As root water-uptake is generally a non-fractionating process, $\delta$^2^H of water in the xylem sap of a tree reflects signature of the source water, and the tree's water-sourcing depth can therefore be identified as the soil depth with the same value of $\delta$^2^H as of the Xylem sap, or a mixture of shallower or deeper depths [@Dawson.1991].
+Analyses of xylem sap $\delta$^2^H of the 12 species studied by @Meinzer:2001 showed that four species sourced water between 1 m and ground water.
+Two other had water access at least 0.8 m deep, one was restricted to water access between 0 to 0.2 m and five fell in the range over which the relationship between soil depth and soil water $\delta$^2^H was variable in the 0.2 to 0.7 m depth range.
+
+\   We compared six species with data for $\delta$^2^H as well as modelled estimates of ERD, after removing those species in @Meinzer:2001 that are likely to have no leaves in March 1997 because xylem sap $\delta$^2^H of these species may not be linked with water sourced at the time of measurement and thus may not be comparable with other species that had leaves.
+We also removed *Guapira standleyana*, whose xylem sap isotope value was not well constrained by the soil water isotope depth profile.
 For each model of ERD (Eq.
 \@ref(eq:M-chosen) and Eqs.
-in SI), estimated species ERD was regressed against species xylem sap $\delta$^2H (permil).
-We used xylem sap $\delta$^2H values given in @Meinzer:2001 for the dry-season of March 1997 as this period showed largest seasonal divergence among species, with a corresponding gradient of $\delta$^2H in the soil (Fig.
-S\@ref(fig:iso-data)).
-We removed species that are likely to have no leaves in March as xylem sap $\delta$^2H of these species may not be linked with water sourced at the time of measurement and thus may not be comparable with other species that had leaves.
-We also removed *Guapira standleyana*, whose xylem sap isotope value was not well constrained by the soil water isotope depth profile.
-We were thus left with 7 species with data on $\delta$^2H as well as modelled estimates of ERD.
+in SI), estimated species ERD was regressed against species xylem sap $\delta$^2^H (‰).
 
@@
 \
-We calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$ 30 m) and average abundance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; *n* = `r length(erd.mrate.sp)`).
-For each of the seven census intervals ($t$) in the 35-yr record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (% yr^-1^) for two successive censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{time}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees of a species present in cen~1~ and dead in cen~2~, respectively, in the 50 ha plot and $time$ is the duration based on mean dates of cen~1~ and cen~2~.
+We calculated mortality rates for large trees ($\ge$10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$30 m) and average abundance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; *n* = `r length(erd.mrate.sp)`).
+For each of the seven census intervals ($t$) in the 35-yr record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (% yr^-1^) for two successive censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{time}$, where $N$~cen~$_{_{1}}$ and $D$~cen~$_{_{1}}$ are the total number of trees of a species present in cen~1~ and dead in cen~2~, respectively, in the 50 ha plot and $time$ is the duration based on mean dates of cen~1~ and cen~2~.
 For each species we calculated mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{t=1}^{6} {Mort_{t}}$.
 
-We regressed species ERD against species ($Mort_{mean}$).
-To assess whether deeper ERD species escape the experience of significant soil drought, we calculated for each speceis the proportion of days in the simulation period (1990-2015) during which soil water potential in the soil zone matching species ERD ($\Psi$~soil,z~=~ERD~) were below species $\Psi$~crit~.
+\   We regressed species ERD against species (Mort~mean~).
+To assess whether deeper ERD species escape the experience of significant soil drought, we calculated for each speceis the proportion of days in the simulation period (1990-2015) during which soil water potential in the soil zone matching species ERD ($\Psi$~soil,z=ERD~) were below species $\Psi$~crit~.
 $\Psi$~crit~ was defined as $\Psi$~50,leaf~.
@@
 
-  Soil Water Potential dynamics ($\Psi$~soil,z~) by depth (z = 0.1 through 13) obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of such extreme years, drought intensity at depth, drought seasonality and duration (Fig.
-S\@ref(fig:swp-droughts)).
-Nevertheless, across the simulation period (1990-2015) $\Psi$~soil,z~ remained above -0.5 MPa from depths 1.7 to 13 m (Fig.
+\   Soil Water Potential dynamics ($\Psi$~soil,z~) by depth z (from 0.1 through 13 m) obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of such extreme years, drought intensity at depth, drought seasonality and duration (Fig.
+\@ref(fig:swp-droughts)).
+Soil droughts in census intervals 2000-2005 and 2005-2010 distinctly occurred in the wet season, with one to two extreme soil drought years, depending on depth (Fig.
+S\@ref(fig:swp-droughts); Fig.
+\@ref(fig:erd-mort)).
+Across the simulation period (1990-2015) $\Psi$~soil,z~ remained above -0.5 MPa from depths 1.7 to 13 m (Fig.
 S\@ref(fig:swp-all-depths)).
@@
 *R*^2^ = `r bfit.r2`; *p* `r  bfit.p`; also see Fig.
-S\@ref(fig:obs-plc-curves).
-Vulnerability curve parameters were predicted for `r sp.n.model` out of `r length(unique(erd.iso$sp))` ERD species without direct data using Eqs.
+S\@ref(fig:obs-plc-curves)).
+Vulnerability curve parameters were predicted for `r sp.n.model` out of `r length(erd.sp)` ERD species that lacked direct observations using Eqs.
 \@ref(eq:a-mod)-\@ref(eq:b-mod) (Table S\@ref(tab:ab-param), Fig.
@@
 \
-ERD estimated from the model with K~leaf~, LAI and an additive effect of VPD as explanatory variables (Eq.
-\@ref(eq:M2)) was strongly related to the measured isotopic signal ($\delta$^2H), that is, with highest goodness-of-fit (*R*^2^ = 0.65; *p* = 0.027; Fig.
+ERD estimated from the model with K~leaf~ and a multiplicative effect of VPD as explanatory variables (Eq.
+\@ref(eq:M2)) was strongly related to the measured isotopic signal ($\delta$^2^H), that is, with a high goodness-of-fit (*R*^2^ = 0.65; *p* = 0.027, *n* = `r length(unique(erd.iso$sp))`; Fig.
 \@ref(fig:erd-iso)).
-The model with K~leaf~, but without LAI and an additive effect of VPD showed no species with sufficient species-level goodness-of-fit.
-ERD from all the other models showed a positive, but non-significant relationship with $\delta$^2H at $\alpha$ = 0.05 (Fig.
+Two of the alternative models tested--with K~leaf~ and LAI, with or without an effect of VPD--had a strong relationship with $\delta$^2^H as well, but one that was based on a smaller set of species (*n* = 4) as fewer species (<20 out of the `r length(erd.sp)` tested) passed the criteria of sufficient species-level goodness-of-fit (Fig.
 S\@ref(fig:other-erd-mods)).
@@
 seemannii* and *T.
-tuberculata*, that @Meinzer:1999 found to have xylem sap $\delta$^2H values between soil water and groundwater, suggesting that these species sourced most of the water from depths > 1 m, were also identified by our model with ERD > 1 m (2.9 m for both species; Fig.
+tuberculata*, that @Meinzer:1999 found to have xylem sap $\delta$^2^H values between soil water and groundwater, suggesting that these species sourced most of the water from depths > 1 m, were also identified by our model with ERD > 1 m (2.9 m for both species; Fig.
 \@ref(fig:erd-iso)).
@@
 
-\
-  In this forest, modeled effective rooting depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from `r min(df.erd.to.plot$depth)` m to `r max(df.erd.to.plot$depth)` m (Fig.
+\   In this forest, modeled effective rooting depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from `r min(df.erd.to.plot$depth)` m to `r max(df.erd.to.plot$depth)` m (Fig.
 \@ref(fig:sp-erd)).
@@
 ## Coordination of ERD with other hydraulic traits
-
-  For the subset of 8 species with ERD for which hydraulic traits data were available, species with deeper ERD showed greater maximum stem hydraulic conductance (K~max,stem~, Spearmans' *r* = `r cor.k`, *p* = `r p.k` at $\alpha$ = 0.05; Fig.
+\
+For the subset of eight species with ERD for which hydraulic traits data were available, species with deeper ERD showed greater maximum stem hydraulic conductance (K~max,stem~, Spearmans' *r* = `r cor.k`, *p* = `r p.k` at $\alpha$ = 0.05; Fig.
 \@ref(fig:erd-hyd)A), less negative leaf $\Psi$~tlp~ (*r* = `r cor.tlp`, *p* = `r p.tlp`; Fig.
@@
 Species with deeper ERD showed lower mean mortality rates across the 34-yr record (1982-2015; *R*^2^ = 0.11, *p* = 0.075, *n* = `r length(erd.mrate.sp)`; Fig.
-\@ref(fig:erd-mort)).
-For five out of seven census intervals, species with shallower ERD had significantly higher mortality anomalies than deeper ERD species (Fig.
+\@ref(fig:erd-mort)), and this relationship was stronger among evergreen species.
+Species with shallower ERD had significantly higher mortality rates than deeper ERD species in the census intervals when soil droughts occurrend in the wet season (Fig.
 S\@ref(fig:erd-mort-census)).
-Analysis of time spent under significant soil drought showed that the proportion of days species $\Psi$~ERD~ was more negative than species $\Psi$~crit~ decreased with species ERD, that is the shallower the ERD, the greater proportion of days $\Psi$~ERD~ < $\Psi$~crit~ (Fig.
+Analysis of time spent under significant soil drought showed that species with shallower ERD spent greater proportion of days such that $\Psi$~soil,z=ERD~ < $\Psi$~crit~ (Fig.
 \@ref(fig:erd-time)).
-Species with ERD deeper than 1 m did not cross species $\Psi$~crit~ for the entire 25-yr period for which soil water potential $\Psi$~soil,z~ was modeled (1990-2015), as soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`;  median, 95% CI) through all the dry seasons and drought periods, which is above ($\Psi$~crit~) of the most sensitive tree species (-0.28 MPa).
+Species with ERD deeper than 1 m did not cross species $\Psi$~crit~ for the entire 25-yr period for which soil water potential $\Psi$~soil,z~ was modeled (1990-2015).
+Soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`;  median, 95% CI) through all the dry seasons and drought periods, which is above ($\Psi$~crit~) of the most sensitive tree species (-0.28 MPa).
 
@@
 
-\
-Our result of lower mortality in deep rooted trees contrasts with the finding of @Chitra-Tarak.2018 in which species identified by their inverse ERD model as effectively deep rooted died *more* in a prolonged, extreme drought.
+\   Our result of lower mortality in deep rooted trees contrasts with the finding of @Chitra-Tarak.2018 in which species identified by their inverse ERD model as effectively deep rooted died *more* in a prolonged, extreme drought.
 Their hydrological model revealed that the multi-year drought exhausted the deep soil and even bedrock water availablity.
@@
 By estimating ERD for large trees of `r length(erd.sp)` canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species-rich tropical forests.
-Our chosen ERD model is an advance over the only other inverse model of ERD (@Chitra-Tarak.2018; *n* = 12 species) as we utilize physiologically meaningful representation, fewer parameters and corroboration with tree and soil data on $\delta$^2H.
+Our chosen ERD model is an advance over the only other inverse model of ERD (@Chitra-Tarak.2018; *n* = 12 species) as we utilize physiologically meaningful representation, fewer parameters and corroboration with tree and soil data on $\delta$^2^H.
 
-\
-We benefited from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA and WSG) and parameters of leaf vulnerability (or percent loss of conductance, (PLC)) curves for a few species (*n* = `r length(sp.exp.param$sp)`), a 'hard' trait.
+\   We benefited from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA and WSG) and parameters of leaf vulnerability (or percent loss of conductance, (PLC)) curves for a few species (*n* = `r length(sp.exp.param$sp)`), a 'hard' trait.
 This allowed us to scale-up the leaf vulnerability curves at the community scale (*n* = `r length(ab.table$Species)` species, see Table S\@ref(tab:ab-param) for model predicted parameters).
@@
 
-\ 
-Another important piece of information for understanding effective rooting depth is stable water isotopes of xylem water and below-ground sources, in the vadose zone and water table.
+\   Another important piece of information for understanding effective rooting depth is stable water isotopes of xylem water and below-ground sources, in the vadose zone and water table.
 Although the interpretation of these data is still under research [@Deurwaerder.2020], they are critical for model calibration or validation.
@@
 
-\
-Our model estimated ERD as a single value, unlike the dynamic ERD modeled by @Chitra-Tarak.2018, but future model developments should investigate seasonal dynamics of ERD, using corresponding calibration data on isotopically determined water-sourcing depths, as species are expected to shift to deeper soil water sources as shallow soils dry out [@Meinzer:1999].
+\   Our model estimated ERD as a single value, unlike the dynamic ERD modeled by @Chitra-Tarak.2018, but future model developments should investigate seasonal dynamics of ERD, using corresponding calibration data on isotopically determined water-sourcing depths, as species are expected to shift to deeper soil water sources as shallow soils dry out [@Meinzer:1999].
 Our inverse ERD model was parameterized on 5-year growth data, with five data points over a 25-yr period, which decouples climate events and demographic outcomes.
@@
 Use of dendrometer data would thus entail accounting for stem hydrostatic flexing [@Zweifel:2018].
-Three of our exploratory ERD models included seasonality in leaf area fraction, but we did not select them as they worsened the fit with growth data and evaluation data ($\delta$^2H, Fig.
+Three of our exploratory ERD models included seasonality in leaf area fraction, but we did not select them as they worsened the fit with growth data and evaluation data ($\delta$^2^H, Fig.
 S\@ref(fig:other-erd-mods)).
@@
 
+Figure \@ref(fig:ab-fit): Occurrence of extreme soil droughts by census intervals (horizontal panels) at three representative depths (z): 0.1, 0.6 and 1 m (vertical panels).
+Mean (black lines) and lower half of 95% distribution (grey areas) of soil water potential ($\Psi$~soil,z~) for a given DOY and depth z are shown (same across census intervals).
+$\Psi$~soil,z~ by DOY for a year (colored line) is only shown if for some part of that year $\Psi$~soil,z~ declined below the 5 percentile distribution.
+
 Figure \@ref(fig:ab-fit): Parameters $A$ (panel A) & $B$ (panel B) for the leaf vulnerability curves that were fitted to observed data on K~leaf~ vs.
@@
 
-Figure \@ref(fig:erd-iso): Modelled Effective Rooting Depth (ERD; mean $\pm$ 1SE; m) versus dry season stable isotopic concentration $\delta$^2H (mean $\pm$ 1SE; ‰) for `r length(unique(erd.iso$sp))` canopy species from Barro Colorado Island.
+Figure \@ref(fig:erd-iso): Modelled Effective Rooting Depth (ERD; mean $\pm$ 1SE; m) versus dry season stable isotopic concentration $\delta$^2^H (mean $\pm$ 1SE; ‰) for `r length(unique(erd.iso$sp))` canopy species from Barro Colorado Island.
 Data from [@Meinzer:1999].
@@
 Each circle is color coded by the Spearman's *r* with red shades for positive *r* values, while blue shades for negative values, respectively.
-*p*-values are given only when they are $\ge$ 0.05.
+*p*-values are given only when they are $\ge$0.05.
 Units are as follows: ERD in m, K~max~ in mmol m^-1^ s^-1^ MPa^-1^), all variables with $\Psi$ in MPa.
@@
 # Tables
-
-Hypotheses for association between Effective Rooting Depth (ERD) and hydraulic traits.
-The symbol, definition and units of key traits used.
 
@@
 
+```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts}"}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q5_by_depth.jpeg"))
+```
+
 ```{r ab-fit, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ab-fit}")}
@@
 ```{r erd-mort, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-mort}")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/erd_mrate_interval_droughts.jpeg"))
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_by rdi.gr_evergreen.jpeg"))
 ```
@@
 
-We also leveraged exisiting GWC measurements by @Kupers:2019 at the depths of 0.15, 0.4 and 1 m in summer 2015 (February, March and April) and 0.15, 0.4 in summer 2016 (March)--1299 samples that covered all soil types and habitats in the plot.
+We also leveraged exisiting GWC measurements by @Kupers:2019 at the depths of 0.15, 0.4 and 1 m in summer 2015 (February, March and April) and 0.15, 0.4 in summer 2016 (March), with 1299 samples in total that covered all soil types and habitats in the plot.
 The sampling periods were 6, 5, 10 and 8 days long, respectively.
@@
 
-We generated an estimate of the mean seasonality curve for normalised leaf area index (LAI$*$, range 0-1) for individual species as follows.
+We generated an estimate of the mean seasonality curve for normalised leaf area index (LAI$^*$, range 0-1, unitless) for individual species as follows.
 We interpolated species-specific weekly leaf-fall data collected at two similar sites in BCI (1985-2020) to a daily estimate [@Wright.1990aeo].
@@
 The difference between leaf-fall and leaf-gain gave an estimate of daily leaf-change time series, which was then scaled to express as a fraction of the total leaf fall in a given year.
-Beginning with the LAI$*$ of 1 at the end of the wet season in the first year of the record, we acccumulated leaf-change over it to obtain a running estimate of the leaf-cover fraction, avaraged it by DOY across years and normalised it to obtain the species-specific seasonality curve in LAI$*$ (range 0-1, unitless).
+Beginning with the LAI$^*$ of 1 at the end of the wet season in the first year of the record, we acccumulated leaf-change over it to obtain a running estimate of the leaf-cover fraction, avaraged it by DOY across years and normalised it to obtain the species-specific seasonality curve in LAI$^*$ (range 0-1, unitless).
 (Fig.
@@
 ### Leaf hydraulic conductivity and vulnerability to cavitation {#SIData.joseph}
-
+\
 The K~leaf~ was measured under steady state and high irradiance following the Evaporative Flux Method (FEM) [@Sack.2012] using a flowmeter.
@@
 Brett to edit.
-
+\
 Methods for Stem Maximum hydraulic conductivity and vulnerability to cavitation closely followed [Christman.2012] and [Wolfe.2016].
@@
 The entire dataset and R script for the analyses are archived in the NGEE-Tropics archive (Wolfe 2020, To be done).
-\
-Canopy branches were cut from 4–10 trees of 26 species across three sites (Table SXX).
+
+\   Canopy branches were cut from 4–10 trees of 26 species across three sites (Table SXX).
 The branch segments were enclosed in opaque plastic bags, and brought to the laboratory for measurements.
@@
 Among species, maximum vessel length ranged XXX and length of stem between the cut that was made in the air to collect the branch and the section of stem used for Kh measurement was XXX (Table SXX).
-\
-Three leaves that were distal to the Kh segment were sealed in plastic bags and the entire branch was enclosed in black plastic bags for at least 30 minutes to stop transpiration.
+
+\   Three leaves that were distal to the Kh segment were sealed in plastic bags and the entire branch was enclosed in black plastic bags for at least 30 minutes to stop transpiration.
 Just prior to measurement of Kh, leaf water potential of the three bagged leaves was measured with a pressure chamber.
 Since transpiration was stopped by bagging the leaves, leaf water potential was assumed to equal stem water potential (REFERENCE).
-\
-For each Kh segment, three measurements of stem water potential were made on stem sections that were located adjacent to the Kh segment.
+
+\   For each Kh segment, three measurements of stem water potential were made on stem sections that were located adjacent to the Kh segment.
 Stem sections that were approximately 40 mm in length were placed in nickel-plated aluminum chambers that were connected to Peltier-type psychrometers (Merrill Instruments, Logan, UT) and placed in a water bath at 30°C for 3 h to equilibrate the humidity within the cambers.
@@
 Each psychrometer probe was calibrated against salt solutions beforehand and the plateau of the psychromatic response curve was used to assess water potential.
-\
-The Kh segments were removed from the branches by repeatedly cutting back the branch on both ends while it was submerged in water to relax the hydraulic tension within the xylem ((REFERENCE)).
+
+\   The Kh segments were removed from the branches by repeatedly cutting back the branch on both ends while it was submerged in water to relax the hydraulic tension within the xylem ((REFERENCE)).
 Then the ends were shaved clean with a fresh razor.
@@
 To calculate cross-sectional stem area, stem diameter was measured twice perpendicularly on both ends of the segment and averaged, then area was calculated assuming a circular cross section.
-\
-To assess vulnerability to cavitation, the branches were allowed to dry in the lab for 0 – 180 hours before they were measured for stem water potential and Ks.
+
+\   To assess vulnerability to cavitation, the branches were allowed to dry in the lab for 0 – 180 hours before they were measured for stem water potential and Ks.
 We used the water potential measured by the psychrometers because they are a more direct measurement.
@@
 Below we present all the altervative models for Effective Rooting Depth (ERD).
-Daily average species-specific growth, $G_{t~|~z}$, in the census interval *t* given an ERD of depth *z*, is described as follows:
+Daily average species-specific growth, $G_{t|z}$, in the census interval *t* given an ERD of depth *z*, is described as follows:
 
 \begin{equation}
-G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,~i~|~z}^{*}) + \epsilon_{t~|~z}
+G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,i|z}^{*}) + \epsilon_{t|z}
 (\#eq:M1)
@@
 \begin{equation}
-G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,~i~|~z}^{*} + VPD_{i}^{*}) + \epsilon_{t~|~z}
+G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,i|z}^{*} + VPD_{i}^{*}) + \epsilon_{t|z}
 (\#eq:M2)
@@
 \begin{equation}
-G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,~i~|~z}^{*} VPD_{i}^{*}) + \epsilon_{t~|~z}
+G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,i|z}^{*} VPD_{i}^{*}) + \epsilon_{t|z}
 (\#eq:M3)
@@
 \begin{equation}
-G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,~i~|~z}^{*} LAI_{doy~|~i}^{*}) + \epsilon_{t~|~z}
+G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,i|z}^{*} LAI_{doy|i}^{*}) + \epsilon_{t|z}
 (\#eq:M4)
@@
 \begin{equation}
-G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} (K^{*}_{leaf,~i~|~z} + VPD^{*}_{i}) LAI^{*}_{doy~|~i}) + \epsilon_{t~|~z}
+G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} (K^{*}_{leaf,i|z} + VPD^{*}_{i}) LAI^{*}_{doy|i}) + \epsilon_{t|z}
 (\#eq:M5)
@@
 \begin{equation}
-G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,~i~|~z}^{*}VPD_{i}^{*}LAI_{doy~|~i}^{*}) + \epsilon_{t~|~z}
+G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,i|z}^{*}VPD_{i}^{*}LAI_{doy|i}^{*}) + \epsilon_{t|z}
 (\#eq:M6)
@@
 
-```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts}Occurrence of extreme soil droughts by census intervals (horizontal panels) at three representative depths (z): 0.1, 0.6 and 1 m (vertical panels).
-Mean (black lines) and lower half of 95% distribution (grey areas) of soil water potential ($\\Psi$~soil,z~) for a given DOY and depth z are shown (same across census intervals).
-$\\Psi$~soil,z~ by DOY for a year (colored line) is only shown if for some part of that year $\\Psi$~soil,z~ declined below the 2.5 percentile distribution."}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q5_by_depth.jpeg"))
-```
-
 ```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}Vulnerability curves fitted to observed K~leaf~ vs.
@@
 ```{r ab-param, results = 'asis', echo = FALSE}
-knitr::kable(ab.table, caption = 'Leaf vulnerability curves parameters A & B, K~max,leaf~ and $\\Psi$ for the " length(erd.sp)" species with ERD estimates.
+knitr::kable(ab.table, caption = 'Leaf vulnerability curves parameters A & B, K~max,leaf~ and $\\Psi$ for the ", length(erd.sp), " species with ERD estimates.
 Species for which curves were fitted to observed data for K~leaf~ vs.

### 2020-10-23 manuscript.rmd
lifespan replaced with lifetime

@@
 In the exploratory model for effective rooting depth, we explored the effect of seasonality in leaf area index (LAI, unitless) on growth.
-We assumed species-specific mean seasonal curves for LAI normalised between 0 and 1 (LAI, unitless), informed by a combination of long-term records for leaf-fall [@Wright.1990aeo] and leaf-lifespan [@Osnas.2018] (see SI).
+We assumed species-specific mean seasonal curves for LAI normalised between 0 and 1 (LAI, unitless), informed by a combination of long-term records for leaf-fall [@Wright.1990aeo] and leaf-lifetime [@Osnas.2018] (see SI).
 
@@
 We interpolated species-specific weekly leaf-fall data collected at two similar sites in BCI (1985-2020) to a daily estimate [@Wright.1990aeo].
-We lead this daily leaf-fall time series by the species-specific average leaf-lifespan (in days) and used it as an estimate of the daily leaf-gain time-series.
+We lead this daily leaf-fall time series by the species-specific average leaf-lifetime (in days) and used it as an estimate of the daily leaf-gain time-series.
 The difference between leaf-fall and leaf-gain gave an estimate of daily leaf-change time series, which was then scaled to express as a fraction of the total leaf fall in a given year.
@@
 
-For five species, an estimate of leaf lifespan was available from direct observations [@Osnas.2018].
-For the rest erd.sp.wo.ll species, leaf lifespan was predicted from LMA ($\overline{lifespan} = `r ll.lma.cf[1]` + `r ll.lma.cf[2]`LMA$; *R*^2^ = `r ll.lma.r2`; *p* `r ll.lma.p`, *n* = `r ll.lma.n`).
+For five species, an estimate of leaf lifetime was available from direct observations [@Osnas.2018].
+For the rest erd.sp.wo.ll species, leaf lifetime was predicted from LMA ($\overline{lifetime} = `r ll.lma.cf[1]` + `r ll.lma.cf[2]`LMA$; *R*^2^ = `r ll.lma.r2`; *p* `r ll.lma.p`, *n* = `r ll.lma.n`).
 

### 2020-10-23 code/15.0_PhenoDemoTraitsPsi.R

@@ -636,7 +636,7 @@ psi.study <- as.data.table(psi.m)[!is.na(interval),][,
 # psi.stat.4 <- psi.stat.4 %>%
 #   left_join(psi.stat.4.extreme.yr, by = "depth_year")
 #
-# psi.stat.4 <- psi.stat.4 %>% mutate(plot.depth = paste0(round(depth, 1), "m"))
+# psi.stat.4 <- psi.stat.4 %>% mutate(plot.depth = paste0(round(depth, 1), " m"))
 #
 # save(psi.stat.4, file = file.path(results.folder, "psi.stat.4.Rdata"))
 # save(psi.stat.4.select, file = file.path(results.folder, "psi.stat.4.select.Rdata"))

### 2020-10-23 manuscript.rmd
Added high-level M&M setup para

@@
 # Materials and Methods
+\
+In the following sections we describe the study site, the models and data used to estimate and evaluate (1) soil water potential dynamics by depth in the whole-soil column and (2) species-specific effective rooting depths.
+These are followed by analyses to address the proposed hypotheses.
+Details of all datasets used are described in the Supplementary Information.
+
 ## Study site description
@@
 We calculated mortality rates for large trees ($\ge$10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$30 m) and average abundance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; *n* = `r length(erd.mrate.sp)`).
-For each of the seven census intervals ($t$) in the 35-yr record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (% yr^-1^) for two successive censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{time}$, where $N$~cen~$_{_{1}}$ and $D$~cen~$_{_{1}}$ are the total number of trees of a species present in cen~1~ and dead in cen~2~, respectively, in the 50 ha plot and $time$ is the duration based on mean dates of cen~1~ and cen~2~.
+For each of the seven census intervals ($t$) in the 35-yr record (1981-82, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (% yr^-1^) for two successive censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{time}$, where $N$~cen~$_{_{1}}$ and $D$~cen~$_{_{1}}$ are the total number of trees of a species present in cen~1~ and dead in cen~2~, respectively, in the 50 ha plot and $time$ is the duration based on mean dates of cen~1~ and cen~2~.
 For each species we calculated mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{t=1}^{6} {Mort_{t}}$.
@@
 \@ref(fig:swp-droughts)).
-Soil droughts in census intervals 2000-2005 and 2005-2010 distinctly occurred in the wet season, with one to two extreme soil drought years, depending on depth (Fig.
+Notably, soil droughts in census intervals 2000-2005 and 2005-2010 distinctly occurred in the wet season, with one to two extreme soil drought years each, depending on depth location (Fig.
 S\@ref(fig:swp-droughts); Fig.
@@
 \
-ERD estimated from the model with K~leaf~ and a multiplicative effect of VPD as explanatory variables (Eq.
+ERD estimated from the chosen model with K~leaf~ and a multiplicative effect of VPD as explanatory variables (Eq.
 \@ref(eq:M2)) was strongly related to the measured isotopic signal ($\delta$^2^H), that is, with a high goodness-of-fit (*R*^2^ = 0.65; *p* = 0.027, *n* = `r length(unique(erd.iso$sp))`; Fig.
 \@ref(fig:erd-iso)).
-Two of the alternative models tested--with K~leaf~ and LAI, with or without an effect of VPD--had a strong relationship with $\delta$^2^H as well, but one that was based on a smaller set of species (*n* = 4) as fewer species (<20 out of the `r length(erd.sp)` tested) passed the criteria of sufficient species-level goodness-of-fit (Fig.
+Two of the alternative models tested--with K~leaf~ and LAI, with or without an effect of VPD--had a strong relationship with $\delta$^2^H as well, but one that was based on a smaller set of species (*n* = 4) as fewer species passed the criteria of sufficient species-level goodness-of-fit; <20 vs.
+`r length(erd.sp)` species for the chosen model (Fig.
 S\@ref(fig:other-erd-mods)).
-For the rest of the analyses we therefore used ERD estimates from the best-fit model (Eq.
+For the rest of the analyses we therefore used ERD estimates from the less parimonious, chosen model (Eq.
 \@ref(eq:M-chosen)).
-Notably, two species, *L.
-seemannii* and *T.
-tuberculata*, that @Meinzer:1999 found to have xylem sap $\delta$^2^H values between soil water and groundwater, suggesting that these species sourced most of the water from depths > 1 m, were also identified by our model with ERD > 1 m (2.9 m for both species; Fig.
-\@ref(fig:erd-iso)).
 
@@
 \@ref(fig:sp-erd)).
+Notably, two species, *L.
+seemannii* and *T.
+tuberculata*, that @Meinzer:1999 found to have xylem sap $\delta$^2^H values between soil water and groundwater, suggesting that these species sourced most of the water from depths >1 m, were also identified by our model with ERD >1 m (2.9 m for both species; Fig.
+\@ref(fig:erd-iso)).
 
@@
 \
-For the subset of eight species with ERD for which hydraulic traits data were available, species with deeper ERD showed greater maximum stem hydraulic conductance (K~max,stem~, Spearmans' *r* = `r cor.k`, *p* = `r p.k` at $\alpha$ = 0.05; Fig.
-\@ref(fig:erd-hyd)A), less negative leaf $\Psi$~tlp~ (*r* = `r cor.tlp`, *p* = `r p.tlp`; Fig.
-\@ref(fig:erd-hyd)B), less negative $\Psi$~88,stem~ and thus greater vulnerability to embolism from cavitation (*r* = `r cor.stem.88`, *p* = `r p.stem.88`; Fig.
-\@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi$~min~ - $\Psi$~88,stem~; *r* = `r cor.hsm`, *p* = `r p.hsm`; Fig.
+For the subset of eight species with ERD for which hydraulic traits data were available, species with deeper ERD showed greater maximum stem hydraulic conductance (K~max,stem~, Spearmans' *r* = `r cor.k`, *p* = `r p.k`; Fig.
+\@ref(fig:erd-hyd)A), less negative leaf $\Psi$~tlp~ (Spearmans' *r* = `r cor.tlp`, *p* = `r p.tlp`; Fig.
+\@ref(fig:erd-hyd)B), less negative $\Psi$~88,stem~ and thus greater vulnerability to embolism from cavitation (Spearmans' *r* = `r cor.stem.88`, *p* = `r p.stem.88`; Fig.
+\@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi$~min~ - $\Psi$~88,stem~; Spearmans' *r* = `r cor.hsm`, *p* = `r p.hsm`; Fig.
 \@ref(fig:erd-hyd)D).
@@
 \
-Species with deeper ERD showed lower mean mortality rates across the 34-yr record (1982-2015; *R*^2^ = 0.11, *p* = 0.075, *n* = `r length(erd.mrate.sp)`; Fig.
-\@ref(fig:erd-mort)), and this relationship was stronger among evergreen species.
-Species with shallower ERD had significantly higher mortality rates than deeper ERD species in the census intervals when soil droughts occurrend in the wet season (Fig.
+Species with deeper ERD had significantly lower mortality rates than deeper ERD species in the census intervals 1985-1990, 2000-2005 and 2005-2010.
+(*n* = `r nrow(mrate.depth.mean)`, Fig.
 S\@ref(fig:erd-mort-census)).
-Analysis of time spent under significant soil drought showed that species with shallower ERD spent greater proportion of days such that $\Psi$~soil,z=ERD~ < $\Psi$~crit~ (Fig.
-\@ref(fig:erd-time)).
+These relationships were stronger among evergreen species ( *n* = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`; Fig.
+\@ref(fig:erd-mort-census-evg)).
+Census intervals 2000-2005 and 2005-2010 were also the periods during which extreme soil droughts distinctly occurred in the wet season, effectively extending the dry season (Fig.
+\@ref(fig:swp-droughts)).
+Soil water availability was not modelled for 1985-1990.
+Analysis of a species' time spent under significant soil drought (proportion of days $\Psi$~soil,z=ERD~ < $\Psi$~crit~) showed that species with shallower ERD spent greater proportion of days below critical hydraulic thresholds (Fig.
+\@ref(fig:erd-time); S\@ref(fig:erd-time-evg)).
 Species with ERD deeper than 1 m did not cross species $\Psi$~crit~ for the entire 25-yr period for which soil water potential $\Psi$~soil,z~ was modeled (1990-2015).
@@
 
-Fig. \@ref(fig:erd-mort) here.
+Fig. \@ref(fig:erd-mort-census-evg) here.
 Fig. \@ref(fig:erd-time) here.
@@
 \
-Our analyses revealed that co-occurring large canopy tree species with deeper effective rooting depths (ERD) were associated with higher hydraulic efficiency, but not safety.
+Our analyses revealed that co-occurring large canopy tree species with deeper effective rooting depths (ERD) were associated with higher hydraulic efficiency, but not safety (Fig.
+\@ref(fig:erd-hyd)).
 Species with deeper ERD showed higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species.
-Nevertheless, deeper ERD species showed lower mortality rates than shallower ERD species over the 35-yr record (1982-2015).
-Analysis of time spent below ($\Psi$~crit~) in the soil zone matching ERD $\Psi$~soil,z~=~ERD~ showed that this was likely due to decreasing realized risk of hydraulic failure with increasing access to water for species with deeper ERD.
+Nevertheless, deeper ERD species showed lower mortality rates than shallower ERD species over the 35-yr record (1981-2015) (Fig.
+S\@ref(fig:erd-mort-census), especially the evergreen species (Fig.
+\@ref(fig:erd-mort-census-evg)).
+Analysis of time spent below ($\Psi$~crit~) in the soil zone matching ERD $\Psi$~soil,z=ERD~ showed that this was likely due to decreasing realized risk of hydraulic failure with increasing access to water for species with deeper ERD (Fig.
+\@ref(fig:erd-time); Fig.
+S\@ref(fig:erd-time-evg)).
 We thus demonstrate, for the first time, a link between realized drought experience, hydraulic architecture and long-term mortality outcomes for co-occurring adult tree species.
@@
 \
-Our study brings attention to the need for assessing drought-sensitivity in terms of species realized drought risk.
-We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins, that is, traits that are commonly identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through a study period of greater recurrence of drought.
-These results are consistent with the two studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016].
+Our study brings attention to the need for assessing drought-sensitivity in terms of species realized drought risk by accounting for the actual hydrological drought and trees' deep water access, beyond the universal drought indices and trees' above-ground hydraulic vulnerability.
+We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins, that is, traits that are commonly identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through periods of significant droughts (Figs.
+\@ref(fig:erd-mort-census-evg),  S\@ref(fig:erd-mort-census), \@ref(fig:swp-droughts)).
+These results are consistent with a few studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016].
 @Johnson:2018fad found that mortality decreased with deep rooting even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi$~88,stem~) and had narrower safety-margins ($\Psi$~min~ - $\Psi$~88,stem~).
@@
 S\@ref(fig:swp-all-depths); @Chitra-Tarak.2018).
-The variation in mortality explained by ERD in our study is low (*R*^2^ = 0.11, *p* = 0.075), and other causes may explain observed mortality [@McDowell.2018].
+And the kind of droughts that buffered the deep rooted species in our study were recurrent sub-annual droughts that extended the dry season into the wet season.
+The variation in mortality explained by ERD in our study is upto 40%; other causes may explain observed mortality [@McDowell.2018].
 
@@
 By estimating ERD for large trees of `r length(erd.sp)` canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species-rich tropical forests.
-Our chosen ERD model is an advance over the only other inverse model of ERD (@Chitra-Tarak.2018; *n* = 12 species) as we utilize physiologically meaningful representation, fewer parameters and corroboration with tree and soil data on $\delta$^2^H.
+Our chosen ERD model is an advance over the only other inverse model of ERD (@Chitra-Tarak.2018; *n* = 12 species) as we employ a physiologically meaningful representation, fewer parameters and corroboration with tree and soil data for $\delta$^2^H.
 
@@
 
-Figure \@ref(fig:erd-mort): Mortality rate (mean $\pm$ 1SE) versus Modelled Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) for `r length(erd.mrate.sp)` canopy species found on Barro Colorado Island.
-*R*^2^ and *p*-values are given in panel insets.
+Figure \@ref(fig:erd-mort-census-evg): Mortality rate (mean $\pm$ 1SE) versus Modelled Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) for `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))` evergreen, canopy species found on Barro Colorado Island over seven census intervals (1981-2015).
+*R*^2^ and *p*-values for linear model fits are given in panel insets.
+Model fits with SE (gray area) are shown for significant fits at $\alpha = 0.1$.
 
@@
 
-```{r erd-mort, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-mort}")}
+```{r erd-mort-census-evg, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-mort}")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_by rdi.gr_evergreen.jpeg"))
@@
 
-```{r erd-mort-census, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-census}Species mortality rate in each census interval during the study period plotted against ERD."}
+```{r erd-mort-census, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-census}Species mortality rate in each census interval during the study period plotted against ERD.
+*R*^2^ and *p*-values for linear model fits are given in panel insets.
+Model fits with SE (gray area) are shown for significant fits at $\alpha = 0.1$."}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_by rdi.gr.jpeg"))
+```
+
+```{r erd-time-evg, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time-evg}Modelled species Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\\Psi$~soil,z~ was below $\\Psi$~crit~, defined as $\\Psi$~80,leaf~, over 1990-2015, where z is the soil depth matching species ERD--for ", nrow(mrate.depth.mean %>% subset(deciduous == "E")), " canopy species found on Barro Colorado Island.
+Each circle represents a species.
+Circles are colored and jittered vertically around the named depths (ticks) for clarity.")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr_evergreen.jpeg"))
 ```

### 2020-10-24 manuscript.rmd
1. corrected italics and romal letters in equations and elsewhere 2. added space before paragraphs

@@
     theme: flatly
-    fig_caption: yes
-  bookdown::pdf_document2: default
-  bookdown::word_document2:
-        reference_docx: manuscriptstyle.docx
     fig_caption: yes
@@
 # Introduction
-\
+
 Tropical forests play a disproportionately large role in the global carbon and energy cycles [@Bonan:2008], and support half of global biodiversity [@Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019].
@@
 
+\
 \     Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation to stem water storage, deep water access, increased root production, hydraulic redistribution and embolism resistance[@McDowell.2008].
@@
 
+\
 \     Trade-offs in plant 'traits' shape how plants make a living in their environment leading to alternative ecological strategies.
@@
 
+\
 \     We used a locally parameterized and calibrated 1-D hydrological water balance model to obtain daily dynamics of soil water potential by depth in the whole-soil column and then applied a novel inverse model to estimate species effective rooting depths (ERD), defined as the depth at which soil water potential best explained observed dynamics of species-specific growth for large trees of `r length(erd.sp)` dominant species (Fig.
@@
 
+\
 Fig. \@ref(fig:schematic) here.
@@
 # Materials and Methods
-\
+
 In the following sections we describe the study site, the models and data used to estimate and evaluate (1) soil water potential dynamics by depth in the whole-soil column and (2) species-specific effective rooting depths.
@@
 ## Study site description
-\
+
 We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and data on stable water isotopes on tree species and soil from an independent study of tree water-sourcing depths [@Meinzer:1999].
@@
 
+\
 \     In the 50 ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm dbh size class [@Condit.2017].
@@
 ### Model description
-\
+
 To estimate dynamics of water availability by depth over the forest’s rooting zone, we used the Energy Exascale Earth System Land Model (ELM), coupled with the Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES).
@@
 
-\     ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\Delta W$~can~, surface water $\Delta W$~sfc~, soil water $\Delta W$~liq,z~ at depth z and water in an unconfined aquifer $\Delta W_{a}$ (omitting processes relevant to snow, wetlands or lakes).
+\
+\     ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\mathrm{\Delta}W_{\textrm{can}}$, surface water $\mathrm{\Delta}W_{\textrm{sfc}}$, soil water $\mathrm{\Delta}W_{\textrm{liq},z}$ at depth $z$ and water in an unconfined aquifer $\mathrm{\Delta} W_{a}$ (omitting processes relevant to snow, wetlands or lakes).
 Conservation of these terms is expressed as follows:
@@
 \begin{split}
-\Delta W_{can} + \Delta W_{sfc} + \sum_{i=1}^{15} \Delta W_{liq,i} + \Delta W_{a} = \left(q_{rain} - E_{v} - E_{g} - q_{over} - q_{h2osfc} - q_{drai}\right)\Delta t
+\mathrm{\Delta}W_\textrm{can} + \mathrm{\Delta}W_\textrm{sfc} + \sum_{i=1}^{15} \mathrm{\Delta}W_{\textrm{liq},i} + \mathrm{\Delta}W_{a} = \left(q_\textrm{rain} - E_{v} - E_{g} - q_\textrm{over} - q_\textrm{h2osfc} - q_\textrm{drai}\right)\mathrm{\Delta}t
 \end{split}
@@
 \end{equation}
-
-where, $q$~rain~ is rainfall, $E_{v}$ is transpiration of the forest, $E_{g}$ is ground evaporation, $q$~over~ is surface runoff, $q$~h2osfc~ is runoff from surface water storage, $q$~drai~ is sub-surface drainage (all in mm H~2~O) and $\Delta t$ is the time-step (typically, 1 hour).
-The soil profile is discretized into 15 exponentially distributed soil layers with depth z, where z $\in$ Z, and Z = {0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13, 21, 35} m.
+where, $q_{\textrm{rain}}$ is rainfall, $E_{v}$ is transpiration of the forest, $E_{g}$ is ground evaporation, $q_{\textrm{over}}$ is surface runoff, $q_{\textrm{h2osfc}}$ is runoff from surface water storage, $q$~drai~ is sub-surface drainage (all in mm H~2~O) and $\mathrm{\Delta}t$ is the time-step (typically, 1 hour).
+The soil profile is discretized into 15 exponentially distributed soil layers with depth $z$, where $z$ $\in$ $Z$, and $Z$ = {0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13, 21, 35} m.
 
 ### Model parameterization
-\
+
 To parameterize ELM-FATES, we identified eleven parameters relevant for Eq.
@@
 \begin{equation}
-\Psi_{soil} = \Psi_e\left(\frac{\theta}{\theta_{s}}\right)^{-b}
+\Psi_\textrm{soil} = \Psi_e\left(\frac{\theta}{\theta_{s}}\right)^{-b}
 (\#eq:soil-ret-curve)
@@
 
-\   We ran ELM-FATES for the 5000-member parameter ensemble with hourly climate drivers from 1985-2018 [@Faybishenko.2020] initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode.
+\
+\   We ran ELM-FATES for the 5000-member parameter ensemble with hourly climate drivers from 1985-2018 (Faybishenko et al.
+2020) initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode.
 As our key interest here was in deriving soil water availability, we ran ELM-FATES in a lower-complexity configuration: static stand structure, and with a single plant functional type (PFT) of evergreen trees.
@@
 ### Model calibration
-\
+
 We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq.
 \@ref(eq:water-balance), namely, (i) evapotranspiration (ET) from the eddy flux tower (2012-2017; see [SI](#SIData.matteo)), comparable to $E_{v} + E_{g}$ in Eq.
-\@ref(eq:water-balance)), (ii) stream discharge $q_{run} from the Conrad catchment, comprabel to $q_{over} + q_{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC or $\theta_{z}) comparable to W_{liq, z}$) from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we obtained with a vertical TDR probe over 0-15 cm depth ($\theta_{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m with 1299 samples in total covering all soil types and habitats (see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\theta_{0.15 m}$, $\theta_{0.4 m}$ and $\theta_{1 m}$ refer to plot-wide averages).
+\@ref(eq:water-balance)), (ii) stream discharge $q_\textrm{run}$ from the Conrad catchment, comparabel to $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC or $\theta_{z}$) comparable to $W_{\textrm{liq,} z}$ from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we obtained with a vertical TDR probe over 0-15 cm depth ($\theta_\textrm{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m with 1299 samples in total covering all soil types and habitats (see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter,  $\theta_{\textrm{0-0.15 m}}$, $\theta_{\textrm{0.4 m}}$ and $\theta_{\textrm{1 m}}$ refer to plot-wide averages).
 
-For ELM-FATES calibration we calculated an objective function for each of the 5000-member ensemble by equally weighting standardized RMSE between observations and simulations across all fluxes and states.
+\
+\    For ELM-FATES calibration we calculated an objective function for each of the 5000-member ensemble by equally weighting standardized RMSE between observations and simulations across all fluxes and states.
 We describe the objective function $O$ as :
@@
 \begin{split}
-\overline{O} = \frac{1}{6}\sum \left(RMSE^{*}(ET,E_{v} + E_{g}) + RMSE^{*}(q_{run},q_{over} + q_{drai}) + RMSE^{*}(\theta_{0-0.15m}, {W_{liq,\overline{z}}}) \\
-& + RMSE^{*}(\theta_{0.15m}, W_{liq,`r soil.depths[4]`m}) + RMSE^{*}(\theta_{0.4~m}, W_{liq,`r soil.depths[6]`~m}) + RMSE^{*}(\theta_{1~m}, W_{liq,`r soil.depths[8]`~m})\right)
+\overline{O} = \frac{1}{6}\sum \textrm{RMSE}^{*}\textrm{(ET,}E_{v} + E_{g}) + \textrm{RMSE}^{*}(q_{\textrm{run}},q_{\textrm{over}} + q_{\textrm{drai}}) + \textrm{RMSE}^{*}(\theta_{\textrm{0-0.15~m}}, {W_{\textrm{liq,}\overline{z}}}) \\
+& + \textrm{RMSE}^{*}(\theta_{\textrm{0.15~m}}, W_{\textrm{liq,`r soil.depths[4]`m}}) + \textrm{RMSE}^{*}(\theta_{\textrm{0.4~m}}, W_{\textrm{liq,`r soil.depths[6]`~m}}) + \textrm{RMSE}^{*}(\theta_{1~m}, W_{\textrm{liq,`r soil.depths[8]`~m}})
 (\#eq:obj-func)
@@
 based on monthly sums for both ET and $q$~run~ and daily averages for $\theta$s.
-$W$~liq,~$_{\overline{z}}$ is the average of daily $W$~liq,z~, where z $\in$ Z, and Z = {`r paste(soil.depths[1:4], collapse=", ")`} m.
+$W_{\textrm{liq,}\overline{z}}$ is the average of daily $W_{\textrm{liq,} z}$, where $z$ $\in$ $Z$, and $Z$ = {`r paste(soil.depths[1:4], collapse=", ")`} m.
 Superscript * indicates that RMSE is standardised between 0 and 1 across the 5000 simulations.
@@
 \   Preliminary analyses showed that minimising the objective function $\overline{O}$, that is, simultaneously fitting the model to all states and fluxes did not capture the soil moisture dynamics at 1 m depth well, our key priority for model calibration.
-To ensure model fit at 1 m depth, we first chose 300 parameter ensemble members that minimised RMSE between observations and simulations for VWC at the depth of 1 m ($RMSE^{*}(\theta_{1~m}, W_{liq,`r soil.depths[8]`~m}) < 0.2$), and among those 300 ensemble members we chose 100 members with the lowest values for the objective function $\overline{O}$.
+To ensure model fit at 1 m depth, we first chose 300 parameter ensemble members that minimised RMSE between observations and simulations for VWC at the depth of 1 m ($\textrm{RMSE}^{*}(\theta_{\textrm{1 m}}, W_{\textrm{liq,`r soil.depths[8]` m}})$ < 0.2), and among those 300 ensemble members we chose 100 members with the lowest values for the objective function $\overline{O}$.
 See Table \@ref(tab:param) for the ranges of best-fit values for different parameters.
@@
 ### Model evaluation
-\
+
 We evaluated ELM-FATES by calculating RMSE between simulated and observed long-term daily VWC for the depths of 0.1, 0.4 and 1 m (2016-2018), baesd on a dataset we had left-out during calibration.
@@
 ## Hydrological droughts
-\   We ran ELM-FATES with the chosen 100 ensemble members from 1985-2018, with the first 5 years used for model spin-up.
-Extreme soil droughts were identified by depth *z* as days that fell below the 5 percentile of mean $\Psi$~soil,*z*~ across all hydrological realisations values at that depth.
+
+We ran ELM-FATES with the chosen 100 ensemble members from 1985-2018, with the first 5 years used for model spin-up.
+Extreme soil droughts were identified by depth $z$ as days that fell below the 5th percentile of $\Psi_{\textrm{soil},z}$ for a given day of the year.
 
 ## Model for effective rooting depth (ERD)
-\
+
 To estimate effective rooting depth, we tested alternate models for growth (Eq.
@@
 There are many factors that could limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019].
-In this study, we used maximum diurnal leaf conductance (K~leaf~) approximated using rhizosphere soil water potential ($\Psi$~soil~) as a proxy for the soil water limiting factor on growth.
-In our models K~leaf~ therefore incorporates the effects of rooting depth and rhizosphere water potential.
-For each species', growth dynamics across census intervals was regressed against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, K~leaf~ and the leaf area index (LAI), all standardized between 0 and 1.
+In this study, we used maximum diurnal leaf conductance ($K_\textrm{leaf}$) approximated using rhizosphere soil water potential ($\Psi$~soil~) as a proxy for the soil water limiting factor on growth.
+In our models $K_\textrm{leaf}$ therefore incorporates the effects of rooting depth and rhizosphere water potential.
+For each species', growth dynamics across census intervals was regressed against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1.
 Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting.
@@
 
+\
 \    For the chosen model (see Model validation and selection below), daily average growth for species $s$, $G_{s,t|z}$, in the census interval $t$ for depth $z$ and for a given hydrological realisation $h$ is described as follows:
@@
 \begin{equation}
-G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,i|z}^{*} VPD_{i}^{*}) + \epsilon_{t|z}
+G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*} \textrm{VPD}_{i}^{*}) + \epsilon_{t|z}
 (\#eq:M-chosen)
 \end{equation}
-
-where n~t~ be the total number of days in census interval *t*, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is the model error term.
+where n~t~ be the total number of days in census interval $t$, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is the model error term.
 See SI for equations for all the alternate models tested.
 
+\
 \    As VPD may have a non-linear relationship with growth, with growth increasing with VPD intially, but decreasing with VPD beyond an optimal value due to stomatal closure [AT:Chonggang, add a reference], we used a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see Model parametarisation) to represent the impact of VPD on growth and used GPP standardised between 0 and 1 (hereafter, VPD*) in Eq.
@@
 
-\    For each species, we estimated diurnal maximum value of K~leaf~$_{,s,i|z,h}$ for a species $s$ by substituting $\Psi$~soil~$_{,s,i|z,h}$ in the leaf vulnerability curves defined below in Eq.
+\
+\    For each species, we estimated diurnal maximum value of $K_{\textrm{leaf},s,i|z,h}$ for a species $s$ by substituting $\Psi_{\textrm{soil},s,i|z,h}$in the leaf vulnerability curves defined below in Eq.
 \@ref(eq:vc-exp) using species-specific parameters $A_{s}$ and $B_{s}$ as:
@@
 \begin{equation}
-K_{leaf,s,i|z,h} = A_{s}e^{-B_{s}\Psi_{soil,i|z,h}}
+K_{\textrm{leaf},s,i|z,h} = A_{s}e^{-B_{s}\Psi_{\textrm{soil},i|z,h}}
 (\#eq:k-soil)
 \end{equation}
-
-\    By varying *z*, we estimated species effective rooting depths (ERD), defined as the depth *z* at which soil water dynamics ($\Psi$~soil,*z*~) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015).
+\    By varying $z$, we estimated species effective rooting depths (ERD), defined as the depth $z$ at which soil water dynamics ($\Psi_{\textrm{soil,} z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015).
 Using Eq.
-\@ref(eq:M-chosen), for a species *s*, we obtained likelihood based on Akaike's Information Criterion (AIC) and Pearson's correlation coefficient *r* for each soil depth *z* for each of the 100 best-fit hydrological realizations.
-For each hydrological realization *h* we retained only one depth that had the maximum likelihood (the minimum AIC score), and within those depths retained the ones that had Pearson's *r* >= 0.71 (or, *R*^2 >= 0.5) as a valid set.
-For a species *s*, we defined effective rooting depth (ERD) as the median of these set of the retained depths, which can be described as :
+\@ref(eq:M-chosen), for a species $s$, we obtained likelihood based on Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $r$ for each soil depth $z$ for each of the 100 best-fit hydrological realizations.
+For each hydrological realization *h* we retained only one depth that had the maximum likelihood (or, the minimum AIC score), and within those depths retained the ones that had Pearson's $r$ >= 0.71 (or, $R$^2 >= 0.5) as a valid set.
+For a species $s$, we defined effective rooting depth (ERD) as the median of these set of the retained depths, which can be described as :
 
 \begin{equation}
-ERD_{s} = \underset{\forall \, h \, \in \, H}{median} \Big\{z,h\,\Big|\, \Big(Likelihood_{s|z,h} = \underset{\forall \, z \,\in\,Z}{maximum}(Likelihood_{s|~z,h})\Big) \cap \Big(r_{s|z,h} \ge 0.71\Big)\Big\}
+\textrm{ERD}_{s} = \underset{\forall \, h \, \in \, H}{\textrm{median}} \Big\{z,h\,\Big|\, \Big(\textrm{L}_{s|z,h} = \underset{\forall \, z \,\in\,Z}{\textrm{maximum}}(\textrm{L}_{s|~z,h})\Big) \cap \Big(r_{s|z,h} \ge 0.71\Big)\Big\}
 (\#eq:erd)
 \end{equation}
-where $H$ = {1, .., 100}, refer to the 100 hydrological realisations and $Z$ = {0.5, `r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m, covering the range of soil depths in ELM-FATES after above parametarision [AT:Chonggang, so that we can perform model evaluation, depths 0.21, 0.37 and 0.62 were all taken as if they were 0.5 m (not averaged).
-How do I explain that?
-So hereafter 0.21, 0.37 and 0.62 were not retained, but only 0.5 m.].
+where $L$ is the likelihood, $H$ = {1, .., 100}, refer to the 100 hydrological realisations and $Z$ = {0.4, `r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m.
+We removed finely resolved soil layers below 0.21 m, averaged $\Psi_{\textrm{soil},z}$ for depths 0.21, 0.37 and 0.62 at 0.4 m.
 
@@
 #### Growth data
-\
-For growth estimates, to minimize the effect of light variation among trees, we selected only large trees ($\ge$30 cm dbh, the diameter at breast height) and also species whose maximum height was $\ge$30 m (hereafter, canopy species) and thus are likely to be fully exposed to the sun.
+
+For growth estimates, to minimize the effect of light variation among trees, we selected only large trees ($\ge$ 30 cm dbh, the diameter at breast height) and also species whose maximum height was $\ge$ 30 m (hereafter, canopy species) and thus are likely to be fully exposed to the sun.
 We selected trees for which the height of the diameter measurements were unchanged across all censuses (1990, 1995, 2000, 2005, 2010, 2015).
-For each individual tree (*i*), we calculated absolute growth rates ($G_{i,t}$, cm yr^-1^) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses prorated by the inter-census interval based on the exact date of measurements.
+For each individual tree ($i$), we calculated absolute growth rates ($G_{i,t}$, cm yr^-1^) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses prorated by the inter-census interval based on the exact date of measurements.
 We eliminated positive outliers as any growth rates amounting to >75 mm yr^-1^, as this was the highest rate observed with confidence for fastest growing species [@Condit.2017].
@@
 Any stem whose later dbh measurement fell <4$\sigma_{e}$ the earlier measurement was removed [@Condit.2017].
-As tree growth rates vary with size [@Muller-Landau:2006], to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($Resid_{i,t} = G_{i,t} - f(d_{i,t})$).
+As tree growth rates vary with size [@Muller-Landau:2006], to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($\textrm{Resid}_{i,t} = G_{i,t} - f(d_{i,t})$).
 The diameter model of growth, $f(d_{i, t})$, a B-spline based polynomial equation with five degrees of freedom, was fitted to growth rates from all trees pooled together as there was insufficient species-specific data for large trees (Fig.
 S\@ref(fig:dbh-model)).
-We retained only those trees that had $Resid_{i,t}$ estimates for all census intervals (total trees = `r sum(g.n$n)`) and standardized (centered and scaled) individual tree time series of $Resid_{i,t}$.
-We then obtained `r length(erd.sp)` species-level time-series of medians of $Resid_{i,t}$ ($G_{t}$).
-These were obtained for only those species with complete records on at least three trees (median *n* = `r median(g.n$n)`, max *n* = `r max(g.n$n)` trees per species).
+We retained only those trees that had $\textrm{Resid}_{i,t}$ estimates for all census intervals (total trees = `r sum(g.n$n)`) and standardized (centered and scaled) individual tree time series of $\textrm{Resid}_{i,t}$.
+We then obtained `r length(erd.sp)` species-level time-series of medians of $\textrm{Resid}_{i,t}$ ($G_{t}$).
+These were obtained for only those species with complete records on at least three trees (median $n$ = `r median(g.n$n)`, max $n$ = `r max(g.n$n)` trees per species).
 
 #### Leaf hydraulic vulnerability curves
-\
-We obtained leaf hydraulic vulnerability curves (K~leaf~ vs.
+
+We obtained leaf hydraulic vulnerability curves ($K_\textrm{leaf}$ vs.
 $\Psi$~leaf~; see Table \@ref(tab:symbols) for trait symbols and their definitions) for adult trees of 21 common species from Zailaa et al.
@@
 \begin{equation}
-K_{leaf,s} = A_{s}e^{-B_{s}\Psi_{leaf_{s}}} + error
+K_{leaf,s} = A_{s}e^{-B_{s}\Psi_{leaf_{s}}} + \epsilon_{s}
 (\#eq:vc-exp)
 \end{equation}
-where $A$ and $B$ are fitted species-specific parameters.
+where $A$ and $B$ are fitted species-specific parameters, $\epsilon_{s}$ an error term.
 These 21 species included `r sp.n.data` of the `r length(erd.sp)` species selected for ERD estimation.
@@
 \begin{equation}
-A_{s} = `r acf[1]` `r acf[2]`B_{s} `r acf[3]`B_{s}^2 + `r acf[4]`B_{s}LMA_{s} + `r acf[5]`LMA_{s} `r acf.6`LMA_{s}^2 + error
+A_{s} = `r acf[1]` `r acf[2]`B_{s} `r acf[3]`B_{s}^2 + `r acf[4]`B_{s}\textrm{LMA}_{s} + `r acf[5]`\textrm{LMA}_{s} `r acf.6`\textrm{LMA}_{s}^2 + \epsilon_{s,a}
 (\#eq:a-mod)
@@
 \begin{equation}
-B_{s} = `r bcf[1]` `r bcf[2]`WSG_{s} + `r bcf[3]`WSG_{s}^2 + `r bcf[4]`WSG_{s}LMA_{s} + `r bcf[5]`LMA_{s} `r bcf[6]`LMA_{s}^2 + error
+\begin{split}
+B_{s} = `r bcf[1]` `r bcf[2]`\textrm{WSG}_{s} + `r bcf[3]`\textrm{WSG}_{s}^2 
+& `r bcf[4]`\textrm{WSG}_{s}\textrm{LMA}_{s} + `r bcf[5]`\textrm{LMA}_{s} `r bcf[6]`\textrm{LMA}_{s}^2 + \epsilon_{s,b}
 (\#eq:b-mod)
+\end{split}
 \end{equation}
+where $\epsilon_{s,a}$ and $\epsilon_{s,b}$ are error terms.
 
+\
 \   As these fits explained a large proportion of variation in parameters $A$ and $B$ (see results), we sequentially used Eq.
@@
 \#ref(eq:vc-exp).
-For all the 31 ERD species we estimated species $\Psi$~50,leaf~ based on their vulnerability curves (see Table \@ref(tab:symbols) for definition).
+For all the `r length(erd.sp)` ERD species we estimated species $\Psi$~50,leaf~ based on their vulnerability curves (see Table \@ref(tab:symbols) for definition).
 
+\
 \   We obtained species-specific parameters $A$ & $B$ for Eq.
@@
 #### Leaf Area Index (LAI)
-\
+
 In the exploratory model for effective rooting depth, we explored the effect of seasonality in leaf area index (LAI, unitless) on growth.
@@
 #### Vapor Pressure Deficit (VPD)
-\
-We fitted a polynomial curve between monthly average of mean midday values (11:00 to 15:00 hrs) of gross primary productivity (gC m^2^ day^-1^) we collected from the eddy flux tower located 75 m from the 50-ha plot (2012-2017) and VPD (kPa) calculated using the climate forcing data (for any day *i*, $GPP_i = `r vpd.cf[1]` + `r vpd.cf[2]`VPD_i + `r vpd.cf[3]`VPD_i^2 + error$; Fig.
-S\@ref(fig:gpp-vpd)); used VPD to predict GPP from 1990-2015, standardized it between 0 and 1, and used it as $VPD^*$ in Eq.
+
+We fitted a polynomial curve between monthly average of mean midday values (11:00 to 15:00 hrs) of gross primary productivity (gC m^2^ day^-1^) we collected from the eddy flux tower located 75 m from the 50-ha plot (2012-2017) and VPD (kPa) calculated using the climate forcing data (for any day $i$, $\textrm{GPP}_i = `r vpd.cf[1]` + `r vpd.cf[2]`\textrm{VPD}_i + `r vpd.cf[3]`\textrm{VPD}_i^2 + error_{i}$; Fig.
+S\@ref(fig:gpp-vpd)); used VPD to predict GPP from 1990-2015, standardized it between 0 and 1, and used it as $\textrm{VPD}^*$ in Eq.
 \@ref(eq:M-chosen)).
@@
 ### Model validation and selection
-\
-To validate the ERD models we used data for xylem sap $\delta$^2^H (‰) as an independent observation of ERD.
-We used xylem sap $\delta$^2^H values given in @Meinzer:2001 for the dry-season of March 1997 as this period showed largest seasonal divergence among species, with a corresponding vertical gradient of $\delta$^2^H in the soil (Fig.
+
+To validate the ERD models we used data for xylem sap $\mathrm{\delta}$^2^H (‰) as an independent observation of ERD.
+We used xylem sap $\mathrm{\delta}$^2^H values given in @Meinzer:2001 for the dry-season of March 1997 as this period showed largest seasonal divergence among species, with a corresponding vertical gradient of $\mathrm{\delta}$^2^H in the soil (Fig.
 S\@ref(fig:iso-data)).
 Briefly, this method makes use of a naturally formed gradient in deuterium to hydrogen ratio of soil water vertically belowground.
-Generally due to evaporative enrichment (here, during the dry season), shallow soil water has a greater deuterium to hydrogen content than deep water (here, $\delta$^2^H ranging from -16.5 to -50 ‰ over 0.05 to 1 m depth), with ground-water having the lowest deuterium to hydrogen content (here, -60 ‰), typically reflecting un-fractionated rain water.
-As root water-uptake is generally a non-fractionating process, $\delta$^2^H of water in the xylem sap of a tree reflects signature of the source water, and the tree's water-sourcing depth can therefore be identified as the soil depth with the same value of $\delta$^2^H as of the Xylem sap, or a mixture of shallower or deeper depths [@Dawson.1991].
-Analyses of xylem sap $\delta$^2^H of the 12 species studied by @Meinzer:2001 showed that four species sourced water between 1 m and ground water.
-Two other had water access at least 0.8 m deep, one was restricted to water access between 0 to 0.2 m and five fell in the range over which the relationship between soil depth and soil water $\delta$^2^H was variable in the 0.2 to 0.7 m depth range.
+Generally due to evaporative enrichment (here, during the dry season), shallow soil water has a greater deuterium to hydrogen content than deep water (here, $\mathrm{\delta}$^2^H ranging from -16.5 to -50 ‰ over 0.05 to 1 m depth), with ground-water having the lowest deuterium to hydrogen content (here, -60 ‰), typically reflecting un-fractionated rain water.
+As root water-uptake is generally a non-fractionating process, $\mathrm{\delta}$^2^H of water in the xylem sap of a tree reflects signature of the source water, and the tree's water-sourcing depth can therefore be identified as the soil depth with the same value of $\mathrm{\delta}$^2^H as of the Xylem sap, or a mixture of shallower or deeper depths [@Dawson.1991].
+Analyses of xylem sap $\mathrm{\delta}$^2^H of the 12 species studied by @Meinzer:2001 showed that four species sourced water between 1 m and ground water.
+Two other had water access at least 0.8 m deep, one was restricted to water access between 0 to 0.2 m and five fell in the range over which the relationship between soil depth and soil water $\mathrm{\delta}$^2^H was variable in the 0.2 to 0.7 m depth range.
 
-\   We compared six species with data for $\delta$^2^H as well as modelled estimates of ERD, after removing those species in @Meinzer:2001 that are likely to have no leaves in March 1997 because xylem sap $\delta$^2^H of these species may not be linked with water sourced at the time of measurement and thus may not be comparable with other species that had leaves.
+\
+\   We compared six species with data for $\mathrm{\delta}$^2^H as well as modelled estimates of ERD, after removing those species in @Meinzer:2001 that are likely to have no leaves in March 1997 because xylem sap $\mathrm{\delta}$^2^H of these species may not be linked with water sourced at the time of measurement and thus may not be comparable with other species that had leaves.
 We also removed *Guapira standleyana*, whose xylem sap isotope value was not well constrained by the soil water isotope depth profile.
@@
 \@ref(eq:M-chosen) and Eqs.
-in SI), estimated species ERD was regressed against species xylem sap $\delta$^2^H (‰).
+in SI), estimated species ERD was regressed against species xylem sap $\mathrm{\delta}$^2^H (‰).
 
 ## ERD coordination with above-ground hydraulic traits
-\
-To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD against maximum stem hydraulic conductance (K~max,stem~), leaf turgor loss point ($\Psi$~tlp~), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi$~88,stem~), and hydraulic safety margins ($\Psi$~min~ - $\Psi$~88,stem~).
+
+To evaluate whether ERD was associated with above-ground hydraulic traits ($n$ = 8 species with both datasets), we regressed species ERD against maximum stem hydraulic conductance ($K_\textrm{max,stem}$), leaf turgor loss point ($\Psi$~tlp~), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi$~88,stem~), and hydraulic safety margins ($\Psi$~min~ - $\Psi$~88,stem~).
 See [SI](#SIData.brett) for data collection and estimation of these variables.
@@
 ## Mortality analyses and species-specific drought experience
+
+We calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$30 m) and average abundance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; $n$ = `r length(erd.mrate.sp)`).
+For each of the seven census intervals $t$ in the 35-yr record (1981-82, 1985, 1990, 2000, 2005, 2010, 2015, 2015), $\textrm{Mort}_{t}$ (% yr^-1^) for species $s$ for two successive censuses, $c_{1}$ and $c_{2}$, was calculated as $\textrm{Mort}_{s,t} = \frac{D_{s,c_{1}}}{N_{s,c_{2}}}\times\frac{100}{\textrm{d}}$, where $N_{s,c_{1}}$ and $D_{s,c_{2}}$ are the total number of trees of a species $s$ present in $c_{1}$ and dead in $c_{2}$, respectively, in the 50 ha plot and $d$ is the duration based on mean dates of $c_{1}$ and $c_{2}$.
+
 \
-We calculated mortality rates for large trees ($\ge$10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$30 m) and average abundance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; *n* = `r length(erd.mrate.sp)`).
-For each of the seven census intervals ($t$) in the 35-yr record (1981-82, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (% yr^-1^) for two successive censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{time}$, where $N$~cen~$_{_{1}}$ and $D$~cen~$_{_{1}}$ are the total number of trees of a species present in cen~1~ and dead in cen~2~, respectively, in the 50 ha plot and $time$ is the duration based on mean dates of cen~1~ and cen~2~.
-For each species we calculated mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{t=1}^{6} {Mort_{t}}$.
-
-\   We regressed species ERD against species (Mort~mean~).
-To assess whether deeper ERD species escape the experience of significant soil drought, we calculated for each speceis the proportion of days in the simulation period (1990-2015) during which soil water potential in the soil zone matching species ERD ($\Psi$~soil,z=ERD~) were below species $\Psi$~crit~.
+\   We regressed species ERD against species ($\textrm{Mort}_{s,t}$ for each of the census interval $t$).
+To assess whether deeper ERD species escaped the experience of significant soil drought, we calculated for each species the proportion of days in the simulation period (1990-2015) during which soil water potential in the soil zone matching species ERD ($\Psi_{\textrm{soil},z=\textrm{ERD}}$) were below species $\Psi$~crit~.
 $\Psi$~crit~ was defined as $\Psi$~50,leaf~.
@@
 ## Soil water dynamics by depth
-\
+
 The ELM-FATES simulations with the chosen best-fit parameter ensembles (Table \@ref(tab:param)) successfully captured the dynamics of soil moisture by depth (Fig.
@@
 
+\
 Fig. \@ref(fig:hydro-calib) here.
 
-\   Soil Water Potential dynamics ($\Psi$~soil,z~) by depth z (from 0.1 through 13 m) obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of such extreme years, drought intensity at depth, drought seasonality and duration (Fig.
+\   Soil Water Potential dynamics ($\Psi_{\textrm{soil},z}$) by depth $z$ (from 0.1 through 13 m) obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of such extreme years, drought intensity at depth, drought seasonality and duration (Fig.
 \@ref(fig:swp-droughts)).
 Notably, soil droughts in census intervals 2000-2005 and 2005-2010 distinctly occurred in the wet season, with one to two extreme soil drought years each, depending on depth location (Fig.
-S\@ref(fig:swp-droughts); Fig.
-\@ref(fig:erd-mort)).
-Across the simulation period (1990-2015) $\Psi$~soil,z~ remained above -0.5 MPa from depths 1.7 to 13 m (Fig.
+S\@ref(fig:swp-droughts)).
+Across the simulation period (1990-2015) $\Psi_{\textrm{soil},z}$ remained above -0.5 MPa from depths 1.7 to 13 m (Fig.
 S\@ref(fig:swp-all-depths)).
 
+\
 Fig. \@ref(fig:swp-droughts) here.
@@
 ## Leaf hydraulic vulnerability curves
-\
-For the leaf vulnerability curve parameters $A$ and $B$ fitted to the observed data for K~leaf~ vs.
+
+For the leaf vulnerability curve parameters $A$ and $B$ fitted to the observed data for $K_\textrm{leaf}$ vs.
 $\Psi$~leaf~, WSG and parameter $B$ explained a large proportion of variation in parameter $A$ (Eq.
 \@ref(eq:a-mod); *Adj.
-R*^2^ = `r afit.r2`; *p* `r  afit.p`; Fig.
+R*^2^ = `r afit.r2`; $p$ `r  afit.p`; Fig.
 \@ref(fig:ab-fit)).
@@
 \@ref(eq:b-mod); Adj.
-*R*^2^ = `r bfit.r2`; *p* `r  bfit.p`; also see Fig.
+$R$^2^ = `r bfit.r2`; $p$ `r  bfit.p`; also see Fig.
 S\@ref(fig:obs-plc-curves)).
@@
 ## Effective rooting depths (ERD)
-\
-ERD estimated from the chosen model with K~leaf~ and a multiplicative effect of VPD as explanatory variables (Eq.
-\@ref(eq:M2)) was strongly related to the measured isotopic signal ($\delta$^2^H), that is, with a high goodness-of-fit (*R*^2^ = 0.65; *p* = 0.027, *n* = `r length(unique(erd.iso$sp))`; Fig.
+
+ERD estimated from the chosen model with $K_\textrm{leaf}$ and a multiplicative effect of VPD as explanatory variables (Eq.
+\@ref(eq:M2)) was strongly related to the measured isotopic signal ($\mathrm{\delta}$^2^H), that is, with a high goodness-of-fit ($R$^2^ = 0.65; $p$ = 0.027, $n$ = `r length(unique(erd.iso$sp))`; Fig.
 \@ref(fig:erd-iso)).
-Two of the alternative models tested--with K~leaf~ and LAI, with or without an effect of VPD--had a strong relationship with $\delta$^2^H as well, but one that was based on a smaller set of species (*n* = 4) as fewer species passed the criteria of sufficient species-level goodness-of-fit; <20 vs.
+Two of the alternative models tested--with $K_\textrm{leaf}$ and LAI, with or without an effect of VPD--had a strong relationship with $\mathrm{\delta}$^2^H as well, but one that was based on a smaller set of species ($n$ = 4) as fewer species passed the criteria of sufficient species-level goodness-of-fit; <20 vs.
 `r length(erd.sp)` species for the chosen model (Fig.
@@
 
+\
 Fig. \@ref(fig:erd-iso) here.
@@
 seemannii* and *T.
-tuberculata*, that @Meinzer:1999 found to have xylem sap $\delta$^2^H values between soil water and groundwater, suggesting that these species sourced most of the water from depths >1 m, were also identified by our model with ERD >1 m (2.9 m for both species; Fig.
+tuberculata*, that @Meinzer:1999 found to have xylem sap $\mathrm{\delta}$^2^H values between soil water and groundwater, suggesting that these species sourced most of the water from depths >1 m, were also identified by our model with ERD >1 m (2.9 m for both species; Fig.
 \@ref(fig:erd-iso)).
 
+\
 Fig. \@ref(fig:sp-erd) here.
@@
 ## Coordination of ERD with other hydraulic traits
-\
-For the subset of eight species with ERD for which hydraulic traits data were available, species with deeper ERD showed greater maximum stem hydraulic conductance (K~max,stem~, Spearmans' *r* = `r cor.k`, *p* = `r p.k`; Fig.
-\@ref(fig:erd-hyd)A), less negative leaf $\Psi$~tlp~ (Spearmans' *r* = `r cor.tlp`, *p* = `r p.tlp`; Fig.
-\@ref(fig:erd-hyd)B), less negative $\Psi$~88,stem~ and thus greater vulnerability to embolism from cavitation (Spearmans' *r* = `r cor.stem.88`, *p* = `r p.stem.88`; Fig.
-\@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi$~min~ - $\Psi$~88,stem~; Spearmans' *r* = `r cor.hsm`, *p* = `r p.hsm`; Fig.
+
+For the subset of eight species with ERD for which hydraulic traits data were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_\textrm{max,stem}$, Spearmans' $r$ = `r cor.k`, $p$ = `r p.k`; Fig.
+\@ref(fig:erd-hyd)A), less negative leaf $\Psi$~tlp~ (Spearmans' $r$ = `r cor.tlp`, $p$ = `r p.tlp`; Fig.
+\@ref(fig:erd-hyd)B), less negative $\Psi$~88,stem~ and thus greater vulnerability to embolism from cavitation (Spearmans' $r$ = `r cor.stem.88`, $p$ = `r p.stem.88`; Fig.
+\@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi$~min~ - $\Psi$~88,stem~; Spearmans' $r$ = `r cor.hsm`, $p$ = `r p.hsm`; Fig.
 \@ref(fig:erd-hyd)D).
 
+\
 Fig. \@ref(fig:erd-hyd) here.
@@
 ## Effective rooting depths, experienced drought and mortality
-\
+
 Species with deeper ERD had significantly lower mortality rates than deeper ERD species in the census intervals 1985-1990, 2000-2005 and 2005-2010.
-(*n* = `r nrow(mrate.depth.mean)`, Fig.
+($n$ = `r nrow(mrate.depth.mean)`, Fig.
 S\@ref(fig:erd-mort-census)).
-These relationships were stronger among evergreen species ( *n* = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`; Fig.
+These relationships were stronger among evergreen species ( $n$ = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`; Fig.
 \@ref(fig:erd-mort-census-evg)).
@@
 Soil water availability was not modelled for 1985-1990.
-Analysis of a species' time spent under significant soil drought (proportion of days $\Psi$~soil,z=ERD~ < $\Psi$~crit~) showed that species with shallower ERD spent greater proportion of days below critical hydraulic thresholds (Fig.
+Analysis of a species' time spent under significant soil drought (proportion of days $\Psi_{\textrm{soil},z=\textrm{ERD}}$ < $\Psi$~crit~) showed that species with shallower ERD spent greater proportion of days below critical hydraulic thresholds (Fig.
 \@ref(fig:erd-time); S\@ref(fig:erd-time-evg)).
-Species with ERD deeper than 1 m did not cross species $\Psi$~crit~ for the entire 25-yr period for which soil water potential $\Psi$~soil,z~ was modeled (1990-2015).
+Species with ERD deeper than 1 m did not cross species $\Psi$~crit~ for the entire 25-yr period for which soil water potential $\Psi_{\textrm{soil},z}$ was modeled (1990-2015).
 Soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`;  median, 95% CI) through all the dry seasons and drought periods, which is above ($\Psi$~crit~) of the most sensitive tree species (-0.28 MPa).
 
+\
 Fig. \@ref(fig:erd-mort-census-evg) here.
@@
 # Discussion
-\
+
 Our analyses revealed that co-occurring large canopy tree species with deeper effective rooting depths (ERD) were associated with higher hydraulic efficiency, but not safety (Fig.
@@
 \@ref(fig:erd-mort-census-evg)).
-Analysis of time spent below ($\Psi$~crit~) in the soil zone matching ERD $\Psi$~soil,z=ERD~ showed that this was likely due to decreasing realized risk of hydraulic failure with increasing access to water for species with deeper ERD (Fig.
+Analysis of time spent below ($\Psi$~crit~) in the soil zone matching ERD $\Psi_{\textrm{soil},z=\textrm{ERD}}$ showed that this was likely due to decreasing realized risk of hydraulic failure with increasing access to water for species with deeper ERD (Fig.
 \@ref(fig:erd-time); Fig.
@@
 ### Realised drought risk integral to assessing drought-vulnerability
-\
+
 Our study brings attention to the need for assessing drought-sensitivity in terms of species realized drought risk by accounting for the actual hydrological drought and trees' deep water access, beyond the universal drought indices and trees' above-ground hydraulic vulnerability.
@@
 
+\
 \   Our result of lower mortality in deep rooted trees contrasts with the finding of @Chitra-Tarak.2018 in which species identified by their inverse ERD model as effectively deep rooted died *more* in a prolonged, extreme drought.
@@
 ### Effective rooting depths at the tree community level
+
+By estimating ERD for large trees of `r length(erd.sp)` canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species-rich tropical forests.
+Our chosen ERD model is an advance over the only other inverse model of ERD (@Chitra-Tarak.2018; $n$ = 12 species) as we employ a physiologically meaningful representation, fewer parameters and corroboration with tree and soil data for $\mathrm{\delta}$^2^H.
+
 \
-By estimating ERD for large trees of `r length(erd.sp)` canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species-rich tropical forests.
-Our chosen ERD model is an advance over the only other inverse model of ERD (@Chitra-Tarak.2018; *n* = 12 species) as we employ a physiologically meaningful representation, fewer parameters and corroboration with tree and soil data for $\delta$^2^H.
-
-\   We benefited from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA and WSG) and parameters of leaf vulnerability (or percent loss of conductance, (PLC)) curves for a few species (*n* = `r length(sp.exp.param$sp)`), a 'hard' trait.
-This allowed us to scale-up the leaf vulnerability curves at the community scale (*n* = `r length(ab.table$Species)` species, see Table S\@ref(tab:ab-param) for model predicted parameters).
+\   We benefited from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA and WSG) and parameters of leaf vulnerability (or percent loss of conductance, (PLC)) curves for a few species ($n$ = `r length(sp.exp.param$sp)`), a 'hard' trait.
+This allowed us to scale-up the leaf vulnerability curves at the community scale ($n$ = `r length(ab.table$Species)` species, see Table S\@ref(tab:ab-param) for model predicted parameters).
 We find that leaf conductance declined more rapidly with declining $\Psi$~leaf~ for species with higher values of WSG and LMA (Fig.
@@
 
+\
 \   Another important piece of information for understanding effective rooting depth is stable water isotopes of xylem water and below-ground sources, in the vadose zone and water table.
@@
 ### Future directions for model development
-\
+
 The trade-offs that we identify between above- and below-ground traits, the water environment vertically below-ground and mortality rates are important for representing diversity in Dynamic Global Vegetation Models especially in species-rich forests.
-The hydraulic efficiency-safety trade-off between high values of maximum stem conductivity (high K~max,stem~) and cavitation-resistance (low $\Psi$~50,stem~ or $\Psi$~88,stem~) is hypothesized although community-wide examples are rare [@Hacke:2006; @Gleason.2016].
-We report such a trade-off (n = 26 species) at our study site stem level ().
+The hydraulic efficiency-safety trade-off between high values of maximum stem conductivity (high $K_\textrm{max,stem}$) and cavitation-resistance (low $\Psi$~50,stem~ or $\Psi$~88,stem~) is hypothesized although community-wide examples are rare [@Hacke:2006; @Gleason.2016].
+We report such a trade-off ($n$ = 26 species) at our study site stem level ().
 We also found that LMA and WSG were strong predictors of PLC curves for leaves, but not for stems (not shown).
-WSG alone explained `r round(a.wsg.r2*100,0)` % of variation in parameter $A$ ($A = `r a.wsg.cf[1]` `r a.wsg.cf[2]`B + `r a.wsg.cf[3]`B^2 + error$; *p* `r  a.wsg.p`) and `r round(b.wsg.r2*100,0)` % of variation in $B$ ($B = `r b.wsg.cf[1]` `r b.wsg.cf[2]`B + `r b.wsg.cf[3]`B^2 + error$; *p* `r  b.wsg.p`).
-Parameter $B$ alone explained `r round(a.b.r2*100,0)` % of variation in $A$ (*p* `r  a.b.p`), while LMA alone was weakly related to $A$ and was not related to $B$ (not shown).
+WSG alone explained `r round(a.wsg.r2*100,0)` % of variation in parameter $A$ ($A_{s} = `r a.wsg.cf[1]` `r a.wsg.cf[2]`B_{s} + `r a.wsg.cf[3]`B^2_{s} + \epsilon_{s}$; $p$ `r  a.wsg.p`) and `r round(b.wsg.r2*100,0)` % of variation in $B$ ($B_{s} = `r b.wsg.cf[1]` `r b.wsg.cf[2]`B_{s} + `r b.wsg.cf[3]`B^2_{s} + \epsilon_{s}$; $p$ `r  b.wsg.p`).
+Parameter $B$ alone explained `r round(a.b.r2*100,0)` % of variation in $A$ ($p$ `r  a.b.p`), while LMA alone was weakly related to $A$ and was not related to $B$ (not shown).
 The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits, including hydraulic vulnerability, in order to better represent the functional diversity of tropical forests, and their trajectories into the future.
 
+\
 \   Our model estimated ERD as a single value, unlike the dynamic ERD modeled by @Chitra-Tarak.2018, but future model developments should investigate seasonal dynamics of ERD, using corresponding calibration data on isotopically determined water-sourcing depths, as species are expected to shift to deeper soil water sources as shallow soils dry out [@Meinzer:1999].
@@
 Use of dendrometer data would thus entail accounting for stem hydrostatic flexing [@Zweifel:2018].
-Three of our exploratory ERD models included seasonality in leaf area fraction, but we did not select them as they worsened the fit with growth data and evaluation data ($\delta$^2^H, Fig.
+Three of our exploratory ERD models included seasonality in leaf area fraction, but we did not select them as they worsened the fit with growth data and evaluation data ($\mathrm{\delta}$^2^H, Fig.
 S\@ref(fig:other-erd-mods)).
@@
 # Captions for Tables and Figures
-
+\
 Table \@ref(tab:hypo): Hypotheses for association between Effective Rooting Depth (ERD) and hydraulic traits.
 
+\
 Table \@ref(tab:symbols): The symbol, definition and units of key traits used in the simulations and analyses.
 
+\
 Figure \@ref(fig:schematic): A schematic for the methods workflow.
 
+\
 Figure \@ref(fig:hydro-calib): ELM-FATES calibration and evaluations.
@@
 
+\
 Figure \@ref(fig:ab-fit): Occurrence of extreme soil droughts by census intervals (horizontal panels) at three representative depths (z): 0.1, 0.6 and 1 m (vertical panels).
-Mean (black lines) and lower half of 95% distribution (grey areas) of soil water potential ($\Psi$~soil,z~) for a given DOY and depth z are shown (same across census intervals).
-$\Psi$~soil,z~ by DOY for a year (colored line) is only shown if for some part of that year $\Psi$~soil,z~ declined below the 5 percentile distribution.
+Mean (black lines) and lower half of 95% distribution (grey areas) of soil water potential ($\Psi_{\textrm{soil},z}$) for a given DOY and depth $z$ are shown (same across census intervals).
+$\Psi_{\textrm{soil},z}$ by DOY for a year (colored line) is only shown if for some part of that year $\Psi_{\textrm{soil},z}$ declined below the 5 percentile distribution.
 
-Figure \@ref(fig:ab-fit): Parameters $A$ (panel A) & $B$ (panel B) for the leaf vulnerability curves that were fitted to observed data on K~leaf~ vs.
+\
+Figure \@ref(fig:ab-fit): Parameters $A$ (panel A) & $B$ (panel B) for the leaf vulnerability curves that were fitted to observed data on $K_\textrm{leaf}$ vs.
 $\Psi$~leaf~ for `r length(sp.exp.param$sp)` species (Eq.
@@
 \@ref(eq:a-mod)-\@ref(eq:b-mod)).
-Goodness-of-fit (*R*^2^) and *p*-values are given in inset.
+Goodness-of-fit ($R$^2^) and $p$-values are given in inset.
 
-Figure \@ref(fig:erd-iso): Modelled Effective Rooting Depth (ERD; mean $\pm$ 1SE; m) versus dry season stable isotopic concentration $\delta$^2^H (mean $\pm$ 1SE; ‰) for `r length(unique(erd.iso$sp))` canopy species from Barro Colorado Island.
+\
+Figure \@ref(fig:erd-iso): Modelled Effective Rooting Depth (ERD; mean $\pm$ 1SE; m) versus dry season stable isotopic concentration $\mathrm{\delta}$^2^H (mean $\pm$ 1SE; ‰) for `r length(unique(erd.iso$sp))` canopy species from Barro Colorado Island.
 Data from [@Meinzer:1999].
@@
 
+\
 Figure \@ref(fig:sp-erd): Modelled Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) for `r  length(erd.sp)` large, canopy species of Barro Colorado Island.
 
+\
 Figure \@ref(fig:erd-iso): Matrix for pairwise correlations (circles) among Effective Rooting Depth (ERD) and hydraulic properties for `r nrow(cor.data)` canopy species found on Barro Colorado Island (Panama).
-Each circle is color coded by the Spearman's *r* with red shades for positive *r* values, while blue shades for negative values, respectively.
-*p*-values are given only when they are $\ge$0.05.
+Each circle is color coded by the Spearman's $r$ with red shades for positive $r$ values, while blue shades for negative values, respectively.
+$p$-values are given only when they are $\ge$0.05.
 Units are as follows: ERD in m, K~max~ in mmol m^-1^ s^-1^ MPa^-1^), all variables with $\Psi$ in MPa.
 
+\
 Figure \@ref(fig:erd-mort-census-evg): Mortality rate (mean $\pm$ 1SE) versus Modelled Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) for `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))` evergreen, canopy species found on Barro Colorado Island over seven census intervals (1981-2015).
-*R*^2^ and *p*-values for linear model fits are given in panel insets.
-Model fits with SE (gray area) are shown for significant fits at $\alpha = 0.1$.
+$R$^2^ and $p$-values for linear model fits are given in panel insets.
+Model fits with SE (gray area) are shown for significant fits at $\alpha$ = 0.1.
 
-Figure \@ref(fig:erd-time): Modelled species Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\Psi$~soil,z~ was below $\Psi$~crit~, defined as $\Psi$~80,leaf~, over 1990-2015, where z is the soil depth matching species ERD--for `r length(erd.mrate.sp)` canopy species found on Barro Colorado Island.
+\
+Figure \@ref(fig:erd-time): Modelled species Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\Psi_{\textrm{soil},z}$ was below $\Psi$~crit~, defined as $\Psi$~80,leaf~, over 1990-2015, where $z$ is the soil depth matching species ERD--for `r length(erd.mrate.sp)` canopy species found on Barro Colorado Island.
 Each circle represents a species.
@@
 S\@ref(fig:TDR-calib)).
-GWC was converted to VWC based on the observed mean bulk dry density of 0.8 g cm^-3^ near the TDR probe (*n* = 5, cylinder volume = 1250 cm^3^).
+GWC was converted to VWC based on the observed mean bulk dry density of 0.8 g cm^-3^ near the TDR probe ($n$ = 5, cylinder volume = 1250 cm^3^).
 Beginning in 2016 we additionally installed three horizontal probes at depths of 0.15, 0.4 and 1 m and we applied the same calibration equation to these data (2016-2018) as that for the vertical probe.
@@
 For five species, an estimate of leaf lifetime was available from direct observations [@Osnas.2018].
-For the rest erd.sp.wo.ll species, leaf lifetime was predicted from LMA ($\overline{lifetime} = `r ll.lma.cf[1]` + `r ll.lma.cf[2]`LMA$; *R*^2^ = `r ll.lma.r2`; *p* `r ll.lma.p`, *n* = `r ll.lma.n`).
+For the rest erd.sp.wo.ll species, leaf lifetime was predicted from LMA ($\overline{lifetime} = `r ll.lma.cf[1]` + `r ll.lma.cf[2]`LMA$; $R$^2^ = `r ll.lma.r2`; $p$ `r ll.lma.p`, $n$ = `r ll.lma.n`).
 
@@
 \
-The K~leaf~ was measured under steady state and high irradiance following the Evaporative Flux Method (FEM) [@Sack.2012] using a flowmeter.
+The $K_\textrm{leaf}$ was measured under steady state and high irradiance following the Evaporative Flux Method (FEM) [@Sack.2012] using a flowmeter.
 Briefly, the leaf was excised under filtered deionised water and attached to a water source (graduated cylinder) using clear plastic tubing.
 The system includes tubing of a known resistance, separated upstream and downstream by two sensors measuring the transpiration-driven water flow pressure in the EFM system [@Sack.2011].
-The K~leaf~ is calculated as the ratio of the transpiration rate (E~leaf~; mmol s^-1^) to the difference in flow pressure (MPa) upstream and downstream from the resistance tubing, and further standardized to 25°C and by leaf area.
+The $K_\textrm{leaf}$ is calculated as the ratio of the transpiration rate (*E*~leaf~; mmol s^-1^) to the difference in flow pressure (MPa) upstream and downstream from the resistance tubing, and further standardized to 25°C and by leaf area.
 At stability (CV < 5%), the leaf is left to equilibrate in plastic bags with moist paper towels and measured for $\Psi$~leaf~ using a Scholander pressure chamber.
-To construct vulnerability curves for each species, four different functions were fitted to species-specific K~leaf~ vs.
+To construct vulnerability curves for each species, four different functions were fitted to species-specific $K_\textrm{leaf}$ vs.
 $\Psi$~leaf~ data using the *anneal* optimization function in the R package likelihood: exponential without an intercept ($K_{leaf} = Ae^{-B\Psi_{leaf}}$), exponential with an intercept ($K_{leaf} = C + Ae^{-B\Psi_{leaf}}$), logistic ($K_{leaf} = \frac{A}{1 + \left(\frac{\Psi_{leaf}}{x_{0}}\right)^B}$) and sigmoidal ($K_{leaf} = \frac{A}{1 + e^{-\left(\frac{\Psi_{leaf} - x_{0}}{B}\right)}}$) [@Sack.2012].
@@
 
-Among the `r length(sp.exp.param$sp)` species with observed data on (K~leaf~ vs.
+Among the `r length(sp.exp.param$sp)` species with observed data on ($K_\textrm{leaf}$ vs.
 $\Psi$~leaf~) used to build the community-level models of leaf vulnerability curves (Eqs.
@@
 We filled three gaps in LMA using predictions from a strong linear relationship between LMA and LMA~disc~, the latter defined as the mean leaf mass per unit area measured for a 1.483 cm^2^ leaf disc taken to avoid veins (g m^-2^) for leaves receiving direct sunlight ($\overline{LMA} = `r disc.cf[1]` + `r disc.cf[2]`LMA_{disc}$; Adj.
-*R*^2^ = `r disc.r2`; *p* `r disc.p`, *n* = `r disc.n`).
-We filled the two remaining gaps in LMA using predictions from a linear relationship between LMA and LMA~lamina~, as measured during the smaller, K~leaf~ campaign ($\overline{LMA} = `r lma.cf[1]` + `r lma.cf[2]`LMA_{lamina}$; Adj.
-*R*^2^ = `r lma.r2`; *p* `r lma.p`, *n* = `r lma.n`).
+$R$^2^ = `r disc.r2`; $p$ `r disc.p`, $n$ = `r disc.n`).
+We filled the two remaining gaps in LMA using predictions from a linear relationship between LMA and LMA~lamina~, as measured during the smaller, $K_\textrm{leaf}$ campaign ($\overline{LMA} = `r lma.cf[1]` + `r lma.cf[2]`LMA_{lamina}$; Adj.
+$R$^2^ = `r lma.r2`; $p$ `r lma.p`, $n$ = `r lma.n`).
 We chose not to fill the two data gaps in WSG as available data were poor predictors of WSG.
@@
 Below we present all the altervative models for Effective Rooting Depth (ERD).
-Daily average species-specific growth, $G_{t|z}$, in the census interval *t* given an ERD of depth *z*, is described as follows:
+Daily average species-specific growth, $G_{t|z}$, in the census interval $t$ given an ERD of depth $z$, is described as follows:
 
@@
 \begin{equation}
-G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,i|z}^{*} + VPD_{i}^{*}) + \epsilon_{t|z}
+G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*} + \textrm{VPD}_{i}^{*}) + \epsilon_{t|z}
 (\#eq:M2)
@@
 \begin{equation}
-G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,i|z}^{*} VPD_{i}^{*}) + \epsilon_{t|z}
+G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*} \textrm{VPD}_{i}^{*}) + \epsilon_{t|z}
 (\#eq:M3)
@@
 \begin{equation}
-G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,i|z}^{*} LAI_{doy|i}^{*}) + \epsilon_{t|z}
+G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*} \textrm{LAI}_{doy|i}^{*}) + \epsilon_{t|z}
 (\#eq:M4)
@@
 \begin{equation}
-G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} (K^{*}_{leaf,i|z} + VPD^{*}_{i}) LAI^{*}_{doy|i}) + \epsilon_{t|z}
+G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} (K^{*}_{\textrm{leaf},i|z} + \textrm{VPD}^{*}_{i}) \textrm{LAI}^{*}_{doy|i}) + \epsilon_{t|z}
 (\#eq:M5)
@@
 \begin{equation}
-G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,i|z}^{*}VPD_{i}^{*}LAI_{doy|i}^{*}) + \epsilon_{t|z}
+G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*}\textrm{VPD}_{i}^{*}\textrm{LAI}_{doy|i}^{*}) + \epsilon_{t|z}
 (\#eq:M6)
 \end{equation}
-where n~t~ be the total number of days in census interval *t*, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is a normally distributed model error term.
+where n~t~ be the total number of days in census interval $t$, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is a normally distributed model error term.
 
@@
 The model equation is shown in inset along with goodness-of-fit (Adj.
-*R*^2^).")}
+$R$^2^).")}
 knitr::include_graphics(file.path("figures/dbh/growthrate_dbh_predicted_growth_with_spline_degree_5_dbh_cutoff_100_model_no_color.jpeg"))
@@
 
-```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($Psi$~soil,z~) by depth z (m; panels) over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines).
+```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($Psi$~soil,z~) by depth $z$ (m; panels) over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines).
 Vertical gray lines demarcate median dates of the demographic censuses at the BCI 50 ha plot."}
@@
 
-```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}Vulnerability curves fitted to observed K~leaf~ vs.
+```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}Vulnerability curves fitted to observed $K_\textrm{leaf}$ vs.
 $\\Psi$~leaf~ data on ", length(sp.exp.param$sp)," tree species from BCI (A & B) with equation \\@ref(eq:vc-exp)--in absolute terms (A, B), as well as, in terms of loss of conductivity (%) (C, D).
@@
 ```{r obs-comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-comm-plc-curves}Percent loss of vulnerability curves for species with ERD estimates.
-Each line represents a species, color coded by source of the curve: fitted to observed data for K~leaf~ vs.
+Each line represents a species, color coded by source of the curve: fitted to observed data for $K_\textrm{leaf}$ vs.
 $\\Psi$~leaf~ are in yellow and those based on scaling relationship with WSG and LMA are in gray.")}
@@
 ```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-other-erd-mods}Effective Rooting Depth versus $\\delta^2H$ for each of the ERD models tested (panels).
-For each ERD model only those species have an estimate of ERD that have passed the test of Pearson's *r* > 0 and *R*^2^ > 0.5 during ERD parameterisation.
+For each ERD model only those species have an estimate of ERD that have passed the test of Pearson's $r$ > 0 and $R$^2^ > 0.5 during ERD parameterisation.
 The number of species with an ERD estimate as well as $\\delta^2H$.
-or each model are shown in inset, along with the goodness-of-fit (*R*^2^) and *p*-values for a linear relationship between ERD and $\\delta^2H$.
+or each model are shown in inset, along with the goodness-of-fit ($R$^2^) and $p$-values for a linear relationship between ERD and $\\delta^2H$.
 ERD models in panels A-E correspond to Eq.
@@
 ```{r erd-mort-census, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-census}Species mortality rate in each census interval during the study period plotted against ERD.
-*R*^2^ and *p*-values for linear model fits are given in panel insets.
-Model fits with SE (gray area) are shown for significant fits at $\alpha = 0.1$."}
+$R$^2^ and $p$-values for linear model fits are given in panel insets.
+Model fits with SE (gray area) are shown for significant fits at $\\alpha$ = 0.1."}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_by rdi.gr.jpeg"))
@@
 
-```{r erd-time-evg, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time-evg}Modelled species Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\\Psi$~soil,z~ was below $\\Psi$~crit~, defined as $\\Psi$~80,leaf~, over 1990-2015, where z is the soil depth matching species ERD--for ", nrow(mrate.depth.mean %>% subset(deciduous == "E")), " canopy species found on Barro Colorado Island.
+```{r erd-time-evg, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time-evg}Modelled species Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\\Psi_{\textrm{soil},z}$ was below $\\Psi$~crit~, defined as $\\Psi$~80,leaf~, over 1990-2015, where $z$ is the soil depth matching species ERD--for ", nrow(mrate.depth.mean %>% subset(deciduous == "E")), " canopy species found on Barro Colorado Island.
 Each circle represents a species.
@@
 ```{r ab-param, results = 'asis', echo = FALSE}
-knitr::kable(ab.table, caption = 'Leaf vulnerability curves parameters A & B, K~max,leaf~ and $\\Psi$ for the ", length(erd.sp), " species with ERD estimates.
-Species for which curves were fitted to observed data for K~leaf~ vs.
+knitr::kable(ab.table, caption = 'Leaf vulnerability curves parameters A & B, $K_\\textrm{max,leaf}$ and $\\Psi$ for the ", length(erd.sp), " species with ERD estimates.
+Species for which curves were fitted to observed data for $K_\textrm{leaf}$ vs.
 $\\Psi$~leaf~ are identified with Source as Data, while species for which curves were obtained based on trait-proxies are identified with Source as Model', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(ab.table))))

### 2020-10-24 manuscript.rmd
1. corrected italic

@@
 
-```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}Vulnerability curves fitted to observed $K_\textrm{leaf}$ vs.
+```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}Vulnerability curves fitted to observed $K_\\textrm{leaf}$ vs.
 $\\Psi$~leaf~ data on ", length(sp.exp.param$sp)," tree species from BCI (A & B) with equation \\@ref(eq:vc-exp)--in absolute terms (A, B), as well as, in terms of loss of conductivity (%) (C, D).
@@
 ```{r obs-comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-comm-plc-curves}Percent loss of vulnerability curves for species with ERD estimates.
-Each line represents a species, color coded by source of the curve: fitted to observed data for $K_\textrm{leaf}$ vs.
+Each line represents a species, color coded by source of the curve: fitted to observed data for $K_\\textrm{leaf}$ vs.
 $\\Psi$~leaf~ are in yellow and those based on scaling relationship with WSG and LMA are in gray.")}
@@
 knitr::kable(ab.table, caption = 'Leaf vulnerability curves parameters A & B, $K_\\textrm{max,leaf}$ and $\\Psi$ for the ", length(erd.sp), " species with ERD estimates.
-Species for which curves were fitted to observed data for $K_\textrm{leaf}$ vs.
+Species for which curves were fitted to observed data for $K_\\textrm{leaf}$ vs.
 $\\Psi$~leaf~ are identified with Source as Data, while species for which curves were obtained based on trait-proxies are identified with Source as Model', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(ab.table))))

### 2020-10-24 manuscript.rmd
1. formatted equations in mortality analyses 2. presented stats in mortality results 2. as well as ERD vs. iso results

@@
 B_{s} = `r bcf[1]` `r bcf[2]`\textrm{WSG}_{s} + `r bcf[3]`\textrm{WSG}_{s}^2 
-& `r bcf[4]`\textrm{WSG}_{s}\textrm{LMA}_{s} + `r bcf[5]`\textrm{LMA}_{s} `r bcf[6]`\textrm{LMA}_{s}^2 + \epsilon_{s,b}
+&`r bcf[4]`\textrm{WSG}_{s}\textrm{LMA}_{s} + `r bcf[5]`\textrm{LMA}_{s} `r bcf[6]`\textrm{LMA}_{s}^2 + \epsilon_{s,b}
 (\#eq:b-mod)
@@
 
-We calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$30 m) and average abundance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; $n$ = `r length(erd.mrate.sp)`).
+To test whether ERD plays a role in alleviating mortality risk, ee calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$30 m) and average abundance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; $n$ = `r length(erd.mrate.sp)`).
 For each of the seven census intervals $t$ in the 35-yr record (1981-82, 1985, 1990, 2000, 2005, 2010, 2015, 2015), $\textrm{Mort}_{t}$ (% yr^-1^) for species $s$ for two successive censuses, $c_{1}$ and $c_{2}$, was calculated as $\textrm{Mort}_{s,t} = \frac{D_{s,c_{1}}}{N_{s,c_{2}}}\times\frac{100}{\textrm{d}}$, where $N_{s,c_{1}}$ and $D_{s,c_{2}}$ are the total number of trees of a species $s$ present in $c_{1}$ and dead in $c_{2}$, respectively, in the 50 ha plot and $d$ is the duration based on mean dates of $c_{1}$ and $c_{2}$.
@@
 
-\
 Fig. \@ref(fig:hydro-calib) here.
@@
 
-\
 Fig. \@ref(fig:swp-droughts) here.
@@
 
-ERD estimated from the chosen model with $K_\textrm{leaf}$ and a multiplicative effect of VPD as explanatory variables (Eq.
-\@ref(eq:M2)) was strongly related to the measured isotopic signal ($\mathrm{\delta}$^2^H), that is, with a high goodness-of-fit ($R$^2^ = 0.65; $p$ = 0.027, $n$ = `r length(unique(erd.iso$sp))`; Fig.
+The best model (with a mulicative effect of VPD and not LAI; Eq.
+\@ref(eq:M-chosen)) explained a large fraction of the variance in the xylem sap isotopic signature ($\mathrm{\delta}$^2^H); $R$^2^ = `r lm.erd.iso.r2`; $p$ = `r lm.erd.iso.p`, $n$ = `r length(unique(erd.iso$sp))`; Fig.
 \@ref(fig:erd-iso)).
-Two of the alternative models tested--with $K_\textrm{leaf}$ and LAI, with or without an effect of VPD--had a strong relationship with $\mathrm{\delta}$^2^H as well, but one that was based on a smaller set of species ($n$ = 4) as fewer species passed the criteria of sufficient species-level goodness-of-fit; <20 vs.
-`r length(erd.sp)` species for the chosen model (Fig.
+Two of the alternative models tested with LAI had a strong relationship with $\mathrm{\delta}$^2^H as well, but one that was based on a smaller set of species ($n$ = 4) as fewer species passed the criteria of sufficient species-level goodness-of-fit with these models; <20 speceis vs.
+`r length(erd.sp)` species for the best model (Fig.
 S\@ref(fig:other-erd-mods)).
-For the rest of the analyses we therefore used ERD estimates from the less parimonious, chosen model (Eq.
+For the rest of the analyses we therefore used ERD estimates from the less parimonious, best model (with VPD and not LAI, Eq.
 \@ref(eq:M-chosen)).
 
-\
 Fig. \@ref(fig:erd-iso) here.
 
+\
 \   In this forest, modeled effective rooting depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from `r min(df.erd.to.plot$depth)` m to `r max(df.erd.to.plot$depth)` m (Fig.
@@
 
-\
 Fig. \@ref(fig:erd-hyd) here.
@@
 
-Species with deeper ERD had significantly lower mortality rates than deeper ERD species in the census intervals 1985-1990, 2000-2005 and 2005-2010.
-($n$ = `r nrow(mrate.depth.mean)`, Fig.
-S\@ref(fig:erd-mort-census)).
-These relationships were stronger among evergreen species ( $n$ = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`; Fig.
+Amongst the evergreen species ERD explained `r m.evg.r2.90`, `r m.evg.r2.05` and `r m.evg.r2.10`% variation in mortality rates during census intervals 1985-1990, 2000-2005 and 2005-2010, respectively, such that species with deeper ERD had significantly lower mortality rates than deeper ERD species ($p$ = `r m.evg.p.90`, `r m.evg.p.05` and `r m.evg.p.10`, respectively; $n$ = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`; Fig.
 \@ref(fig:erd-mort-census-evg)).
-Census intervals 2000-2005 and 2005-2010 were also the periods during which extreme soil droughts distinctly occurred in the wet season, effectively extending the dry season (Fig.
+When deciduous species were included, ERD explained lower but significant proportion of variation in mortality in those intervals ($R$^2^ = `r m.r2[1]`, `r m.r2[2]` and `r m.r2[3]`% and $p$ = `r m.p.90`, `r m.p.05` and `r m.p.10`, respectively; $n$ = `r nrow(mrate.depth.mean)`; Fig.
+\@ref(fig:erd-mort-census)).
+Census intervals 2000-2005 and 2005-2010 were also the periods during which extreme soil droughts distinctly occurred in the wet season, effectively extending the dry season into the wet season (Fig.
 \@ref(fig:swp-droughts)).
@@
 
-\
 Fig. \@ref(fig:erd-mort-census-evg) here.
@@
 The hydraulic efficiency-safety trade-off between high values of maximum stem conductivity (high $K_\textrm{max,stem}$) and cavitation-resistance (low $\Psi$~50,stem~ or $\Psi$~88,stem~) is hypothesized although community-wide examples are rare [@Hacke:2006; @Gleason.2016].
-We report such a trade-off ($n$ = 26 species) at our study site stem level ().
+We report such a trade-off ($n$ = 26 species) at our study site stem level.
 We also found that LMA and WSG were strong predictors of PLC curves for leaves, but not for stems (not shown).
-WSG alone explained `r round(a.wsg.r2*100,0)` % of variation in parameter $A$ ($A_{s} = `r a.wsg.cf[1]` `r a.wsg.cf[2]`B_{s} + `r a.wsg.cf[3]`B^2_{s} + \epsilon_{s}$; $p$ `r  a.wsg.p`) and `r round(b.wsg.r2*100,0)` % of variation in $B$ ($B_{s} = `r b.wsg.cf[1]` `r b.wsg.cf[2]`B_{s} + `r b.wsg.cf[3]`B^2_{s} + \epsilon_{s}$; $p$ `r  b.wsg.p`).
-Parameter $B$ alone explained `r round(a.b.r2*100,0)` % of variation in $A$ ($p$ `r  a.b.p`), while LMA alone was weakly related to $A$ and was not related to $B$ (not shown).
+WSG alone explained `r round(b.wsg.r2*100,0)`% of variation in $B$ ($p$ `r  b.wsg.p`).
 The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits, including hydraulic vulnerability, in order to better represent the functional diversity of tropical forests, and their trajectories into the future.
@@
 \end{equation}
-where n~t~ be the total number of days in census interval $t$, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is a normally distributed model error term.
+where n~t~ are the total number of days in census interval $t$, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is the model error term.
 

### 2020-10-24 manuscript.rmd
1. edited vulnerability curve results

@@
 We describe the objective function $O$ as :
-
-<!-- # For html -->
-<!-- \begin{align*} -->
-<!-- \overline{O} = \frac{1}{6}\sum RMSE^{*}(ET,~E_{v} + E_{g}) + RMSE^{*}(q_{run},~q_{over} + q_{drai}) + RMSE^{*}(\theta_{0-0.15m}, {W_{liq,~\overline{z}}})\\ -->
-<!-- + RMSE^{*}(\theta_{0.15m}, W_{liq,~`r soil.depths[4]`m}) + RMSE^{*}(\theta_{0.4~m}, W_{liq,~`r soil.depths[6]`~m}) + RMSE^{*}(\theta_{1~m}, W_{liq,~`r soil.depths[8]`~m}) -->
-<!-- (\#eq:obj-func) -->
-<!-- \end{align*} -->
 
@@
 
-For the leaf vulnerability curve parameters $A$ and $B$ fitted to the observed data for $K_\textrm{leaf}$ vs.
-$\Psi$~leaf~, WSG and parameter $B$ explained a large proportion of variation in parameter $A$ (Eq.
+Parameters of leaf vulnerability curves were predictable from WSG and LMA.
+WSG and parameter $B$ explained a large proportion of variance in parameter $A$ (Eq.
 \@ref(eq:a-mod); *Adj.
@@
 \@ref(fig:ab-fit)).
-WSG and LMA explained a large proportion of variation in parameter $B$ (Eq.
+WSG and LMA explained a large proportion of variance in parameter $B$ (Eq.
 \@ref(eq:b-mod); Adj.
@@
 S\@ref(fig:obs-plc-curves)).
-Vulnerability curve parameters were predicted for `r sp.n.model` out of `r length(erd.sp)` ERD species that lacked direct observations using Eqs.
-\@ref(eq:a-mod)-\@ref(eq:b-mod) (Table S\@ref(tab:ab-param), Fig.
+This predictive power allowed us to estimate leaf vulnerability curve parameters for `r sp.n.model` (out of `r length(erd.sp)` ERD species) that lacked direct observations; Table S\@ref(tab:ab-param), Fig.
 S\@ref(fig:obs-comm-plc-curves)).
+These vulenrability curves were used in the models for effective rooting depth.
+
 Fig. \@ref(fig:ab-fit) here.
@@
 
-The best model (with a mulicative effect of VPD and not LAI; Eq.
+The chosen model (with a mulicative effect of VPD and not LAI; Eq.
 \@ref(eq:M-chosen)) explained a large fraction of the variance in the xylem sap isotopic signature ($\mathrm{\delta}$^2^H); $R$^2^ = `r lm.erd.iso.r2`; $p$ = `r lm.erd.iso.p`, $n$ = `r length(unique(erd.iso$sp))`; Fig.
@@
 Two of the alternative models tested with LAI had a strong relationship with $\mathrm{\delta}$^2^H as well, but one that was based on a smaller set of species ($n$ = 4) as fewer species passed the criteria of sufficient species-level goodness-of-fit with these models; <20 speceis vs.
-`r length(erd.sp)` species for the best model (Fig.
+`r length(erd.sp)` species for the chosen model (Fig.
 S\@ref(fig:other-erd-mods)).
-For the rest of the analyses we therefore used ERD estimates from the less parimonious, best model (with VPD and not LAI, Eq.
+For the rest of the analyses we therefore used ERD estimates from the less parimonious, chosen model (with VPD and not LAI, Eq.
 \@ref(eq:M-chosen)).
@@
 This allowed us to scale-up the leaf vulnerability curves at the community scale ($n$ = `r length(ab.table$Species)` species, see Table S\@ref(tab:ab-param) for model predicted parameters).
+WSG alone explained `r round(b.wsg.r2*100,0)`% of the variance in $B$ ($p$ `r  b.wsg.p`).
 We find that leaf conductance declined more rapidly with declining $\Psi$~leaf~ for species with higher values of WSG and LMA (Fig.
@@
 We also found that LMA and WSG were strong predictors of PLC curves for leaves, but not for stems (not shown).
-WSG alone explained `r round(b.wsg.r2*100,0)`% of variation in $B$ ($p$ `r  b.wsg.p`).
 The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits, including hydraulic vulnerability, in order to better represent the functional diversity of tropical forests, and their trajectories into the future.

### 2020-10-24 manuscript.rmd
Replaced coordination with relationship

@@
 We evaluated ERD against existing dueterium to hydrogen content data for a subset of six tree species as an independent observation of ERD, given the corresponding gradient in isotopic composition of water in the soil profile.
-We tested coordination of ERD with hydraulic traits for a subset of eight species with both datasets.
+We tested relationship of ERD with above-ground hydraulic traits for a subset of eight species with both datasets.
 We finally analysed species' mortality rates across the censuses spanning 35 years to test whether ERD explained risk of drought-induced mortality in periods of extreme hydrological drought.
@@
 
-## ERD coordination with above-ground hydraulic traits
+## Relationship of ERD with above-ground hydraulic traits
 
@@
 
-## Coordination of ERD with other hydraulic traits
+## Relationship of ERD with above-ground hydraulic traits
 

### 2020-10-24 manuscript.rmd
Included some of Laura and Rosies suggestions in the results

@@
 We evaluated ERD against existing dueterium to hydrogen content data for a subset of six tree species as an independent observation of ERD, given the corresponding gradient in isotopic composition of water in the soil profile.
-We tested relationship of ERD with above-ground hydraulic traits for a subset of eight species with both datasets.
-We finally analysed species' mortality rates across the censuses spanning 35 years to test whether ERD explained risk of drought-induced mortality in periods of extreme hydrological drought.
+We tested relationship of ERD with above-ground hydraulic traits,namely, xylem vulnerability to embolism, branch hydraulic safety margins, turgor loss points and maximum stem-specific hydraulic conductivity, for a subset of eight species with both datasets.
+Finally, ee analysed species' mortality rates across the censuses spanning 35 years to test whether ERD explained risk of drought-induced mortality in periods of extreme hydrological drought.
 Specifically, we tested an overarching hypothesis that inter-specific variation in trees' effective rooting depths (ERD) enables increasingly reliable water availability with depth and thus mitigation of hydraulic vulnerability.
@@
 
-The ELM-FATES simulations with the chosen best-fit parameter ensembles (Table \@ref(tab:param)) successfully captured the dynamics of soil moisture by depth (Fig.
+The 100 ELM-FATES simulations with the chosen best-fit parameter sets (Table \@ref(tab:param)) successfully captured the dynamics of soil moisture by depth (Fig.
 \@ref(fig:hydro-calib) A & Fig.
@@
 These simulations also captured the dynamics in stream discharge (Fig.
-\@ref(fig:hydro-calib)C) and to a moderate extent the dynamics in evapotranspiration (Fig.
-\@ref(fig:hydro-calib)D).
-The reduction in parameter range in the best-fit ensembles compared to the global ranges tested showed that the above calibration was sensitive to the following parameters: the Ball-Berry stomatal slope parameter, the ELM rooting distribution parameter that regulates the depth of the rooting profile, soil hydraulic conductivity profile especially at depth, and our adjustment factor to soil hydraulic conductivity to account for macroporosity and direct flow paths (Table \@ref(tab:param)).
+\@ref(fig:hydro-calib)C) and in evapotranspiration (Fig.
+\@ref(fig:hydro-calib)D), slightly underestimating peak discharge in wet years and slightly overestimating peak ET.
+The reduction in parameter range in the best-fit ensembles compared to the global ranges tested showed that the model calibration was sensitive to the Ball-Berry stomatal slope parameter, the ELM root distribution parameter (that regulates the depth of the rooting profile), soil hydraulic conductivity profile, especially at depth, and the adjustment factor that varies soil hydraulic conductivity to account for macroporosity and direct flow paths (Table \@ref(tab:param)).
 
@@
 S\@ref(fig:swp-droughts)).
-Across the simulation period (1990-2015) $\Psi_{\textrm{soil},z}$ remained above -0.5 MPa from depths 1.7 to 13 m (Fig.
+Across the simulation period (1990-2015) $\Psi_{\textrm{soil},z}$ remained above -0.5 MPa from depths 1.7 to 13 m in the simulations (Fig.
 S\@ref(fig:swp-all-depths)).
@@
 
+Our novel approach for estimating ERD using 25 years of growth, species-specific leaf vulnerability curves, modeled soil moisture through the profile, and observed VPD was consistent with estimates using xylem water isotopes.
 Our analyses revealed that co-occurring large canopy tree species with deeper effective rooting depths (ERD) were associated with higher hydraulic efficiency, but not safety (Fig.
 \@ref(fig:erd-hyd)).
-Species with deeper ERD showed higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species.
+Species with deeper ERD showed higher maximum stem hydraulic conductance, and greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species.
 Nevertheless, deeper ERD species showed lower mortality rates than shallower ERD species over the 35-yr record (1981-2015) (Fig.
-S\@ref(fig:erd-mort-census), especially the evergreen species (Fig.
+S\@ref(fig:erd-mort-census), especially amongst the evergreen species (Fig.
 \@ref(fig:erd-mort-census-evg)).
@@
 S\@ref(fig:erd-time-evg)).
-We thus demonstrate, for the first time, a link between realized drought experience, hydraulic architecture and long-term mortality outcomes for co-occurring adult tree species.
+We thus demonstrate, for the first time, a link between realized drought experience, hydraulic architecture and long-term mortality outcomes for co-occurring large tree species.
 
@@
 
-Our study brings attention to the need for assessing drought-sensitivity in terms of species realized drought risk by accounting for the actual hydrological drought and trees' deep water access, beyond the universal drought indices and trees' above-ground hydraulic vulnerability.
-We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins, that is, traits that are commonly identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through periods of significant droughts (Figs.
+Our study brings attention to the need for assessing drought sensitivity in terms of species realized drought exposure by accounting for the actual hydrological drought and trees' deep water access, beyond the universal drought indices and trees' above-ground hydraulic vulnerability.
+We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins, that is, traits that are commonly identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through periods of extreme droughts (Figs.
 \@ref(fig:erd-mort-census-evg), S\@ref(fig:erd-mort-census), \@ref(fig:swp-droughts)).
-These results are consistent with a few studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016].
+Deeper ERD species on average had more hydraulically efficient, but risky, suite of traits optimized for an environment of more reliable and greater water source at depth that resulted in a lower mortality risk.
+By contrast shallow ERD species had inefficient, but safer, xylem likely optimized for high variability in $\Psi_{soil}$ in the shallow soil layers, and resulting in higher variance in mortality rates across censuses as $\Psi_{soil,~z = ERD}$ crossed ERD $\Psi_{crit}$.
+Our results are consistent with a few studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016].
 @Johnson:2018fad found that mortality decreased with deep rooting even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi$~88,stem~) and had narrower safety-margins ($\Psi$~min~ - $\Psi$~88,stem~).
@@
 \
-\   Our result of lower mortality in deep rooted trees contrasts with the finding of @Chitra-Tarak.2018 in which species identified by their inverse ERD model as effectively deep rooted died *more* in a prolonged, extreme drought.
-Their hydrological model revealed that the multi-year drought exhausted the deep soil and even bedrock water availablity.
+\   Our result--of lower mortality in deep rooted trees--contrasts with the finding of @Chitra-Tarak.2018 (CT2018) in which species identified by their inverse ERD model as effectively deep rooted died *more* in a rare, prolonged drought.
+The hydrological model in that study revealed that the multi-year drought exhausted the deep soil and even bedrock water availablity.
 In contrast, our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period, $\Psi$~soil~ below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and droughts, that is above ($\Psi$~crit~) of the most sensitive tree species (-0.28 MPa).
-The difference in our results are likely attributable to site differences in the precipitation regime.
-In their dry forest site (Mudumalai National Park, India) precipitation is low: 1095 $\pm$ 296 mm (mean $\pm$ SD) compared to `r round(rain.man.stats$rain)` $\pm$ `r round(rain.man.stats$rain.sd)` mm (`r min(clim$Year)`-`r max(clim$Year)`) at BCI, and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (Penman-Monteith PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (Penman-Monteith PET = `r round(mean(clim$pet.PM))` $\pm$ `r round(sd(clim$pet.PM))` mm;`r min(clim$Year)`-`r max(clim$Year)`; this study, @Guo.2016), and thus at BCI deep soil layers are effectively recharged annually (Fig.
+The difference in the results shown here are likely attributable to site differences in the precipitation regime.
+In the dry forest site in CT2018 (Mudumalai National Park, India) precipitation is low: 1095 $\pm$ 296 mm (mean $\pm$ SD) compared to `r round(rain.man.stats$rain)` $\pm$ `r round(rain.man.stats$rain.sd)` mm (`r min(clim$Year)`-`r max(clim$Year)`) at BCI, and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (Penman-Monteith PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (Penman-Monteith PET = `r round(mean(clim$pet.PM))` $\pm$ `r round(sd(clim$pet.PM))` mm;`r min(clim$Year)`-`r max(clim$Year)`; this study, @Guo.2016), and thus at BCI deep soil layers are effectively recharged annually (Fig.
 S\@ref(fig:swp-all-depths); @Chitra-Tarak.2018).
-And the kind of droughts that buffered the deep rooted species in our study were recurrent sub-annual droughts that extended the dry season into the wet season.
-The variation in mortality explained by ERD in our study is upto 40%; other causes may explain observed mortality [@McDowell.2018].
+The kind of droughts that buffered the deep rooted species in our study were recurrent sub-annual droughts that extended the dry season into the wet season.
+The variation in mortality explained by ERD in our study is upto 40%, and other causes may explain observed mortality [@McDowell.2018].
 
@@
 \
-\   We benefited from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA and WSG) and parameters of leaf vulnerability (or percent loss of conductance, (PLC)) curves for a few species ($n$ = `r length(sp.exp.param$sp)`), a 'hard' trait.
+\   We found a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA and WSG) and parameters of leaf vulnerability (or percent loss of conductance, (PLC)) curves for a few species ($n$ = `r length(sp.exp.param$sp)`), a 'hard' trait.
 This allowed us to scale-up the leaf vulnerability curves at the community scale ($n$ = `r length(ab.table$Species)` species, see Table S\@ref(tab:ab-param) for model predicted parameters).
@@
 comm.).
-The model of below-ground water-availability in this study (ELM-FATES) is far more advanced compared to that used by @Chitra-Tarak.2018 and was calibrated with data on long-term soil moisture dynamics at a point location (0-0.15m) and plot-wide measurements by depth (0.15, 0.4 and 1 m), as well as, evapotranspiration and stream discharge.
-@Chitra-Tarak.2018 was calibrated only on stream discharge and water table levels.
+The model of below-ground water-availability in this study (ELM-FATES) differs from the conceptual water balance model used by CT2018 and was calibrated with data on long-term soil moisture dynamics at a point location (0-0.15m) and plot-wide measurements by depth (0.15, 0.4 and 1 m), as well as, evapotranspiration and stream discharge.
+The model in CT2018 was calibrated only on stream discharge and water table levels.
 We lacked data on water table levels which would have been useful to constrain water availability dynamics in the unobserved soil layers and the spatial variability within the plot.
@@
 
-### Future directions for model development
+### Future directions
 
-The trade-offs that we identify between above- and below-ground traits, the water environment vertically below-ground and mortality rates are important for representing diversity in Dynamic Global Vegetation Models especially in species-rich forests.
+The relationships that we identify between above- and below-ground traits, the water environment vertically below-ground and mortality rates are important for representing diversity in Dynamic Global Vegetation Models especially in species-rich forests, which intrinsically rely on the parameterization of contrasting life history strategies [@Schieter:2012] and the simulation of competition between those strategies.
 The hydraulic efficiency-safety trade-off between high values of maximum stem conductivity (high $K_\textrm{max,stem}$) and cavitation-resistance (low $\Psi$~50,stem~ or $\Psi$~88,stem~) is hypothesized although community-wide examples are rare [@Hacke:2006; @Gleason.2016].
@@
 We also found that LMA and WSG were strong predictors of PLC curves for leaves, but not for stems (not shown).
+This offers the promise of a greater ability to parameterie the 'hard' hydraulic traits with the abundant soft trait data, thus allowing much better representation of forest hydrodynamics.
 The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits, including hydraulic vulnerability, in order to better represent the functional diversity of tropical forests, and their trajectories into the future.
@@
 \
-\   Our model estimated ERD as a single value, unlike the dynamic ERD modeled by @Chitra-Tarak.2018, but future model developments should investigate seasonal dynamics of ERD, using corresponding calibration data on isotopically determined water-sourcing depths, as species are expected to shift to deeper soil water sources as shallow soils dry out [@Meinzer:1999].
+\   Our model estimated ERD as a single value, unlike the temporally dynamic ERD modeled by @Chitra-Tarak.2018, but future model developments should investigate seasonal dynamics of ERD, using corresponding calibration data on isotopically determined water-sourcing depths, as species are expected to shift to deeper soil water sources as shallow soils dry out [@Meinzer:1999].
 Our inverse ERD model was parameterized on 5-year growth data, with five data points over a 25-yr period, which decouples climate events and demographic outcomes.
-Future studies could better constrain the model with higher resolution growth data such as those from dendrometer bands.
+Future studies could better constrain the ERD model with higher resolution growth data such as those from dendrometer bands.
 This is especially important for modeling seasonally dynamic ERD.
@@
 VWC), hydraulic conductivity (K~soil~) by depth, soil moisture by depth, stream discharge and evapotranspiration were important for effective calibration of our 1-D hydrological model.
-We recommend wide-spread data collection of these variables as well as water-table levels in forest-inventory sites to allow for estimation of trees' water environments.
-Finally, developing an inverse ERD model that accounts for movement of water within the soil-plant-atmosphere-continuum (SPAC) would be a step change from the current statistical models of ERD (this study and @Chitra-Tarak.2018).
+We recommend wide-spread and coordinated collection of these variables as well as water-table levels in forest-inventory sites to allow for estimation of trees' water environments.
+Finally, developing an inverse ERD model that accounts for movement of water within the soil-plant-atmosphere-continuum (SPAC) [@Christoffersen:2016kd2d; @Sperry.1998] would be a step change from the current statistical models of ERD (this study and @Chitra-Tarak.2018).
 Although data needs for parameterizing such models [@Christoffersen:2016kd2d; @Sperry.1998] could be greater (e.g.
@@
 
-Establishing relationships between plant's environment, traits and demographic outcomes is imperative for developing a predictive plant ecology, but trees realized water environments and water-access are rarely studied.
+Establishing relationships between environment, traits and demographic outcomes of plants is imperative for developing a predictive plant ecology.
+Trees realized water environments and water-access are nonetheless rarely studied.
 To our knowledge this is the first study to demonstrate a mechanistic link between trees' realized drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.
-The role of deep-water access that we find in alleviating mortality of drought-sensitive trees has important implications for our predictive understanding of tropical forest dynamics under current and future climate.
-Our community-scale framework for modeling effective rooting depths or leaf vulnerability curves indicate the possibilities in expanding the use of critical, hard-earned observations in species-rich forests towards community-scale generalizations.
+We report here that deep-water access plays a role in alleviating mortality of hydrualically vulnerable trees.
+This has important implications for our predictive understanding of tropical forest dynamics under current and future climate, and suggests that greater emphasis should be given to predictive modeling of tree root allocation and distribution (e.g.
+@Wijk.2001) as well as trees' hydrological environments.
+Our community-scale framework for modeling effective rooting depths and leaf vulnerability curves indicate the possibilities in expanding the use of these critical, hard-earned observations in species-rich forests towards community-scale generalizations.
 

### 2020-10-24 old_ms_sections.Rmd
1. corrected italic
minor

@@
 
+For each species $s$ we calculated mean mortality rate across all intervals as $\textrm{Mort}_{s,\textrm{mean}} = \frac{1}{6}\sum_{t=1}^{6} {\textrm{Mort}_{s,t}}$.
+
+Species with deeper ERD showed lower mean mortality rates across the 35-yrs record (1982-2015; *R*^2^ = 0.11, *p* = 0.075, *n* = `r length(erd.mrate.sp)`; Fig.
+\@ref(fig:erd-mort)), and this relationship was stronger among evergreen species.
+
 Specifically for the water environment (or, resource), observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified an acquisitive-conservative resource economic spectrum similar to those for light and nutrients, although, for water, examples of all trade-offs observed simultaneously are rare, and below-ground traits or environment linkages are weak [@Reich:2014; @Sterck:2011].

### 2020-10-24 manuscript.rmd
1. Edited abstract 2. Moved methods summary text to ME&M 3. Edited intro to ERD model

@@
 - Among the mechanisms that plants employ to tolerate drought, deep water access is arguably the most commonly invoked, but under-studied.
-Vulnerability to embolism and hydraulic safety margins are shown to be predictors of drought-induced mortality risk, but deep-water access may delay plants from reaching critical safety thresholds.
+Vulnerability to embolism and hydraulic safety margins can predict mortality risk at given levels of dehydration, but deep-water access may delay plants from dehydration.
 Here, we tested the role of deep-water access in enabling survival within a diverse tropical forest community (Panama) using a novel data-model approach.
@@
 These efforts have resulted in identification of correlations between aboveground hydraulic traits such as vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013, @Venturas.2016].
-Deep roots may mitigate hydraulic vulnerability and mortality-risk, yet plants' realized water environments, specifically their ability to acquire water via differing rooting depths, are rarely studied, especially in the tropical region.
-We tested how rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tree species of a tropical forest over 35 years that witnessed droughts of various intensity and duration.
+Deep roots may mitigate hydraulic vulnerability and mortality-risk, yet plants' realized water environments, specifically plant's ability to acquire water via differing rooting depths in terms of hydrological rather than meteorological droughts, are rarely studied, especially in the tropical region.
 
 \
-\     We used a locally parameterized and calibrated 1-D hydrological water balance model to obtain daily dynamics of soil water potential by depth in the whole-soil column and then applied a novel inverse model to estimate species effective rooting depths (ERD), defined as the depth at which soil water potential best explained observed dynamics of species-specific growth for large trees of `r length(erd.sp)` dominant species (Fig.
-\@ref(fig:schematic)).
-This model incorporated the impact of atmospheric drought on growth and was constrained by above-ground hydraulic traits, namely, species-specific leaf vulnerability curves, by leveraging existing data for eight species and developing trait-based proxies for the rest.
-We evaluated ERD against existing dueterium to hydrogen content data for a subset of six tree species as an independent observation of ERD, given the corresponding gradient in isotopic composition of water in the soil profile.
-We tested relationship of ERD with above-ground hydraulic traits,namely, xylem vulnerability to embolism, branch hydraulic safety margins, turgor loss points and maximum stem-specific hydraulic conductivity, for a subset of eight species with both datasets.
-Finally, ee analysed species' mortality rates across the censuses spanning 35 years to test whether ERD explained risk of drought-induced mortality in periods of extreme hydrological drought.
+\   We tested how rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tree species of a tropical forest over 35 years that witnessed droughts of various intensity and duration.
 Specifically, we tested an overarching hypothesis that inter-specific variation in trees' effective rooting depths (ERD) enables increasingly reliable water availability with depth and thus mitigation of hydraulic vulnerability.
@@
 
-\
-Fig. \@ref(fig:schematic) here.
 Table \@ref(tab:hypo) here.
@@
 # Materials and Methods
+\
+\      We developed a novel, empirical inverse model to estimate species effective rooting depths (ERD) for large trees of `r length(erd.sp)` species (Fig.
+\@ref(fig:schematic)).
+We defined species ERD as the depth at which soil water potential best explained species-specific 5-yrly mean growth rates over 25 years.
+To obtain the daily dynamics of soil water potential in the whole-soil column we locally parameterized and calibrated a 1-D hydrological water balance model.
+The ERD model incorporated the impact of atmospheric drought on growth and was constrained by above-ground hydraulic traits, namely, species-specific leaf vulnerability curves, by using existing data for eight species and developing trait-based proxies for the rest.
+We evaluated ERD against existing dueterium to hydrogen content data for a subset of six tree species as an independent observation of ERD.
+We tested relationship of ERD with above-ground hydraulic traits, namely, xylem vulnerability to embolism, branch hydraulic safety margins, leaf turgor loss point and maximum stem-specific hydraulic conductivity, for a subset of eight species with both datasets.
+Finally, we analysed species' mortality rates over 35 years to test whether ERD explained risk of drought-induced mortality in periods of extreme hydrological drought.
+
+\
+Fig. \@ref(fig:schematic) here.
 
@@
 ELM is a land-model that, among many features, simulates the physics and conservative dynamics of water, energy and carbon fluxes.
-In particular, soil hydrologic fluxes are vertically resolved among discrete soil layers (1-D) [@Bonan:2013].
+In particular, soil hydrologic fluxes are vertically resolved among discrete soil layers (1-D) similar to that in CLM4.5 [@Bonan:2013].
 FATES is a community-based, open-source model used for studying climate-vegetation interactions.
@@
 \
-\   We ran ELM-FATES for the 5000-member parameter ensemble with hourly climate drivers from 1985-2018 (Faybishenko et al.
+\   We ran ELM-FATES for the 5000-member parameter ensemble with hourly climate drivers measured at the BCI meteorological station for the period from 1985-2018 (Faybishenko et al.
 2020) initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode.
-As our key interest here was in deriving soil water availability, we ran ELM-FATES in a lower-complexity configuration: static stand structure, and with a single plant functional type (PFT) of evergreen trees.
+As our key interest here was on deriving soil water availability, we ran ELM-FATES in a lower-complexity configuration: static stand structure, and with a single plant functional type (PFT) of evergreen trees.
 The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of a dry-deciduous PFT did not significantly alter results (not shown).
@@
 \@ref(eq:M-chosen) and those in SI), which use effective rooting depth as an input.
-There are many factors that could limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019].
-In this study, we used maximum diurnal leaf conductance ($K_\textrm{leaf}$) approximated using rhizosphere soil water potential ($\Psi$~soil~) as a proxy for the soil water limiting factor on growth.
-In our models $K_\textrm{leaf}$ therefore incorporates the effects of rooting depth and rhizosphere water potential.
-For each species', growth dynamics across census intervals was regressed against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1.
+There are many intrinsic and extrinsic factors that limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019].
+Rhizosphere soil water potential ($\Psi$~soil~) down-regulates plant tissue water potentials [@Christoffersen.2018].
+In moist forests $\Psi$~leaf~ equilibriates with $\Psi$~soil~ at night and declines through the day as plants transpire.
+$\Psi$~leaf~ in turn controls leaf conductance ($K_\textrm{leaf}$), which in turn controls stomatal conductance and photosynthetic rates [@Sack.2012].
+We therefore used maximum diurnal leaf conductance ($K_\textrm{leaf}$) as determined by $\Psi$~soil~ as a proxy for the soil water limitation on growth.
+In our ERD models $K_\textrm{leaf}$ therefore incorporates the effects of rooting depth and rhizosphere water potential.
+This approach ignores the buffering effect of stem water storage on $K_\textrm{leaf}$, which may be a subject of future work (see Discussion).
+VPD has a positive impact on transpiration and thus photosynthesis and growth upto a VPD threshold, after which VPD has a negative impact [@Grossiord.2020].
+The fraction of a tree's leaf cover present at a given time further limits it's growth potential.
+We therefore regressed species-specific growth in a census interval against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1.
 Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting.
@@
 \
-\    As VPD may have a non-linear relationship with growth, with growth increasing with VPD intially, but decreasing with VPD beyond an optimal value due to stomatal closure [AT:Chonggang, add a reference], we used a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see Model parametarisation) to represent the impact of VPD on growth and used GPP standardised between 0 and 1 (hereafter, VPD*) in Eq.
+\    To account for non-linear impact of VPD on growth, we used a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see Model parametarisation) to represent the impact of VPD on growth and used GPP standardised between 0 and 1 (hereafter, VPD*) in Eq.
 \@ref(eq:M-chosen).
@@
 \end{equation}
-\    By varying $z$, we estimated species effective rooting depths (ERD), defined as the depth $z$ at which soil water dynamics ($\Psi_{\textrm{soil,} z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015).
+\    By varying $z$, we estimated species effective rooting depths (ERD), defined as the depth $z$ at which soil water dynamics ($\Psi_{\textrm{soil,} z}$) best explained observed dynamics of species growth (six 5-yrly censuses, 1990-2015).
 Using Eq.
@@
 \
-\   Our result--of lower mortality in deep rooted trees--contrasts with the finding of @Chitra-Tarak.2018 (CT2018) in which species identified by their inverse ERD model as effectively deep rooted died *more* in a rare, prolonged drought.
+\   Our result--of lower mortality in deep rooted trees--contrasts with the finding of @Chitra-Tarak.2018 (hereafter CT2018) in which species identified by their inverse ERD model as effectively deep rooted died *more* in a rare, prolonged drought.
 The hydrological model in that study revealed that the multi-year drought exhausted the deep soil and even bedrock water availablity.
@@
 The relationships that we identify between above- and below-ground traits, the water environment vertically below-ground and mortality rates are important for representing diversity in Dynamic Global Vegetation Models especially in species-rich forests, which intrinsically rely on the parameterization of contrasting life history strategies [@Schieter:2012] and the simulation of competition between those strategies.
-The hydraulic efficiency-safety trade-off between high values of maximum stem conductivity (high $K_\textrm{max,stem}$) and cavitation-resistance (low $\Psi$~50,stem~ or $\Psi$~88,stem~) is hypothesized although community-wide examples are rare [@Hacke:2006; @Gleason.2016].
-We report such a trade-off ($n$ = 26 species) at our study site stem level.
-We also found that LMA and WSG were strong predictors of PLC curves for leaves, but not for stems (not shown).
+We found that LMA and WSG were strong predictors of PLC curves for leaves, but not for stems (not shown).
 This offers the promise of a greater ability to parameterie the 'hard' hydraulic traits with the abundant soft trait data, thus allowing much better representation of forest hydrodynamics.

### 2020-10-24 manuscript.rmd
1. added a comment on impact of stem water storage to ERD in M&M

@@
 In our ERD models $K_\textrm{leaf}$ therefore incorporates the effects of rooting depth and rhizosphere water potential.
-This approach ignores the buffering effect of stem water storage on $K_\textrm{leaf}$, which may be a subject of future work (see Discussion).
+This approach ignores the buffering effect of stem water storage on $K_\textrm{leaf}$, which may cause an assignment of a slightly deeper ERD to a species.
+Future work should therefore explore the influence of stem water storage on estimation of ERD (see Discussion).
 VPD has a positive impact on transpiration and thus photosynthesis and growth upto a VPD threshold, after which VPD has a negative impact [@Grossiord.2020].

### 2020-10-24 old_ms_sections.Rmd
1. added a comment on impact of stem water storage to ERD in M&M

@@
 
+## Intro
+
+Historic climatic variation is relatively small (Mora et al., 2013), and, though data have been scant, trees' hydraulic safety margins are reportedly relatively narrow (Choat et al., 2012; but see Oliveira et al., 2019).
+
+## M&M
+
 For each species $s$ we calculated mean mortality rate across all intervals as $\textrm{Mort}_{s,\textrm{mean}} = \frac{1}{6}\sum_{t=1}^{6} {\textrm{Mort}_{s,t}}$.
+
+## Results
+
+The hydraulic efficiency-safety trade-off between high values of maximum stem conductivity (high $K_\textrm{max,stem}$) and cavitation-resistance (low $\Psi$~50,stem~ or $\Psi$~88,stem~) is hypothesized although community-wide examples are rare [@Hacke:2006; @Gleason.2016].
+We report such a trade-off ($n$ = 26 species) at our study site stem level.
 
@@
 
-Deeper ERD species had on average more hydraulically efficient, but risky, suite of traits optimized for an environment of more reliable and greater water source at depth that resulted in a lower mortality risk.
-By contrast shallow ERD species had inefficient, but safer, xylem likely optimized for high variability in $\Psi_{soil}$ in the shallow soil layers, and resulting in higher variance in mortality rates across censuses as $\Psi_{soil,~z = ERD}$ crossed ERD $\Psi_{crit}$.
 Mortality rates of shallow ERD species may have been exacerbated by increased drought events during the study period.

### 2020-10-24 manuscript.rmd
M & M intro moved back to Intro

@@
 Deep roots may mitigate hydraulic vulnerability and mortality-risk, yet plants' realized water environments, specifically plant's ability to acquire water via differing rooting depths in terms of hydrological rather than meteorological droughts, are rarely studied, especially in the tropical region.
+We tested how rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tree species of a tropical forest over 35 years that witnessed droughts of various intensity and duration.
 
-\
-\   We tested how rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tree species of a tropical forest over 35 years that witnessed droughts of various intensity and duration.
-Specifically, we tested an overarching hypothesis that inter-specific variation in trees' effective rooting depths (ERD) enables increasingly reliable water availability with depth and thus mitigation of hydraulic vulnerability.
-Our hypotheses are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallow rooted species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deep rooted species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallow rooted species may have higher mortality rates as a result of higher variability in water availability in shallow soil layers.
-To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.
-
-Table \@ref(tab:hypo) here.
-
-# Materials and Methods
 \
@@
 Finally, we analysed species' mortality rates over 35 years to test whether ERD explained risk of drought-induced mortality in periods of extreme hydrological drought.
+Specifically, we tested an overarching hypothesis that inter-specific variation in trees' effective rooting depths (ERD) enables increasingly reliable water availability with depth and thus mitigation of hydraulic vulnerability.
+Our hypotheses are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallow rooted species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deep rooted species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallow rooted species may have higher mortality rates as a result of higher variability in water availability in shallow soil layers.
+To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.
 
@@
 Fig. \@ref(fig:schematic) here.
+Table \@ref(tab:hypo) here.
 
+# Materials and Methods
+\
 In the following sections we describe the study site, the models and data used to estimate and evaluate (1) soil water potential dynamics by depth in the whole-soil column and (2) species-specific effective rooting depths.

### 2020-10-24 manuscript.rmd
# edited isotopic data methods

@@
 \
-\    To account for non-linear impact of VPD on growth, we used a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see Model parametarisation) to represent the impact of VPD on growth and used GPP standardised between 0 and 1 (hereafter, VPD*) in Eq.
+\    To account for the non-linear impact of VPD on growth, we used a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see Model parametarisation) to represent the impact of VPD on growth and used GPP standardised between 0 and 1 (hereafter, VPD*) in Eq.
 \@ref(eq:M-chosen).
@@
 
-To validate the ERD models we used data for xylem sap $\mathrm{\delta}$^2^H (‰) as an independent observation of ERD.
-We used xylem sap $\mathrm{\delta}$^2^H values given in @Meinzer:2001 for the dry-season of March 1997 as this period showed largest seasonal divergence among species, with a corresponding vertical gradient of $\mathrm{\delta}$^2^H in the soil (Fig.
+Xylem sap $\mathrm{\delta}$^2^H values can provide an index of rooting depth, which we used to validate the ERD models as an independent observation of ERD.
+We used xylem sap $\mathrm{\delta}$^2^H values from BCI given in @Meinzer:2001 for the dry-season of March 1997 as this period showed largest seasonal divergence among species, with a corresponding vertical gradient of $\mathrm{\delta}$^2^H in the soil.
+(Fig.
 S\@ref(fig:iso-data)).
-Briefly, this method makes use of a naturally formed gradient in deuterium to hydrogen ratio of soil water vertically belowground.
-Generally due to evaporative enrichment (here, during the dry season), shallow soil water has a greater deuterium to hydrogen content than deep water (here, $\mathrm{\delta}$^2^H ranging from -16.5 to -50 ‰ over 0.05 to 1 m depth), with ground-water having the lowest deuterium to hydrogen content (here, -60 ‰), typically reflecting un-fractionated rain water.
-As root water-uptake is generally a non-fractionating process, $\mathrm{\delta}$^2^H of water in the xylem sap of a tree reflects signature of the source water, and the tree's water-sourcing depth can therefore be identified as the soil depth with the same value of $\mathrm{\delta}$^2^H as of the Xylem sap, or a mixture of shallower or deeper depths [@Dawson.1991].
+Briefly, this isotopic approach makes use of a naturally formed gradient in deuterium to hydrogen ratio of soil water vertically belowground, which is generally formed due to evaporative enrichment (here, during the dry season).
+Shallow soil water has a greater deuterium to hydrogen content than deep soil water (here, $\mathrm{\delta}$^2^H ranged from -16.5 to -50 ‰ over 0.05 to 1 m depth), with ground-water having the lowest deuterium to hydrogen content (here, -60 ‰), typically reflecting un-fractionated rain water.
+As root water-uptake is generally a non-fractionating process, $\mathrm{\delta}$^2^H of water in the xylem sap of a tree reflects the signature of the source water.
+Thus water-sourcing depth of the tree can be identified as the depth with the same value of $\mathrm{\delta}$^2^H as that of the xylem sap, with the uncertainty that the value of $\mathrm{\delta}$^2^H may also result from a mixing of water from shallower or deeper depths [@Dawson.1991].
 Analyses of xylem sap $\mathrm{\delta}$^2^H of the 12 species studied by @Meinzer:2001 showed that four species sourced water between 1 m and ground water.
@@
 
-\
 \   We compared six species with data for $\mathrm{\delta}$^2^H as well as modelled estimates of ERD, after removing those species in @Meinzer:2001 that are likely to have no leaves in March 1997 because xylem sap $\mathrm{\delta}$^2^H of these species may not be linked with water sourced at the time of measurement and thus may not be comparable with other species that had leaves.

### 2020-10-24 manuscript.rmd
1. Explained why ERD is non-coninuous in the community-wide ERD chart

@@
 
-To test whether ERD plays a role in alleviating mortality risk, ee calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$30 m) and average abundance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; $n$ = `r length(erd.mrate.sp)`).
+To test whether ERD plays a role in alleviating mortality risk, we calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$30 m) and average abundance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; $n$ = `r length(erd.mrate.sp)`).
 For each of the seven census intervals $t$ in the 35-yr record (1981-82, 1985, 1990, 2000, 2005, 2010, 2015, 2015), $\textrm{Mort}_{t}$ (% yr^-1^) for species $s$ for two successive censuses, $c_{1}$ and $c_{2}$, was calculated as $\textrm{Mort}_{s,t} = \frac{D_{s,c_{1}}}{N_{s,c_{2}}}\times\frac{100}{\textrm{d}}$, where $N_{s,c_{1}}$ and $D_{s,c_{2}}$ are the total number of trees of a species $s$ present in $c_{1}$ and dead in $c_{2}$, respectively, in the 50 ha plot and $d$ is the duration based on mean dates of $c_{1}$ and $c_{2}$.
@@
 Figure \@ref(fig:sp-erd): Modelled Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) for `r  length(erd.sp)` large, canopy species of Barro Colorado Island.
+Each species' ERD is a median of the species' ERDs (depths) obtained from all hydrological realisations, in which soil layers are vertically resolved into exponentially increasing depths.
+Hence the non-continuous nature of the ERD axis.
 

### 2020-10-24 manuscript.rmd
Moving M&M intro right beck in M&M

@@
 \
+\      We tested an overarching hypothesis that inter-specific variation in trees' effective rooting depths (ERD) enables increasingly reliable water availability with depth and thus mitigation of hydraulic vulnerability.
+Our hypotheses are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallow rooted species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deep rooted species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallow rooted species may have higher mortality rates as a result of higher variability in water availability in shallow soil layers.
+To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.
+
+\
+Table \@ref(tab:hypo) here.
+
+# Materials and Methods
+\
 \   We developed a novel, empirical inverse model to estimate species effective rooting depths (ERD) for large trees of `r length(erd.sp)` species (Fig.
@@
 Finally, we analysed species' mortality rates over 35 years to test whether ERD explained risk of drought-induced mortality in periods of extreme hydrological drought.
-Specifically, we tested an overarching hypothesis that inter-specific variation in trees' effective rooting depths (ERD) enables increasingly reliable water availability with depth and thus mitigation of hydraulic vulnerability.
-Our hypotheses are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallow rooted species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deep rooted species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallow rooted species may have higher mortality rates as a result of higher variability in water availability in shallow soil layers.
-To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.
-
 \
 Fig. \@ref(fig:schematic) here.
-Table \@ref(tab:hypo) here.
 
-# Materials and Methods
 \

### 2020-10-24 manuscript.rmd
1. Edited study site rainfall description 2. Edited discussion

@@
 Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014].
-Rainfall at BCI is seasonal with a mean annual rainfall of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a four-month long dry season from mid-December through mid-April ($\le$ 100 mm rain month^-1^).
+Rainfall at BCI is seasonal with a mean annual total of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a pronounced dry season from mid-December through April with a mean total of 292 mm [@Paton.2020].
 Among the tree *species* found in the forest canopy, 28% are identified to be at least partly deciduous during the dry season, but this amounts to a maximum of 6.3% *trees* in the canopy that are deciduous, and the percentage of total forest crown area that is deciduous is estimated at 9.7% [@Condit:2000].
@@
 There are many intrinsic and extrinsic factors that limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019].
-Rhizosphere soil water potential ($\Psi$~soil~) down-regulates plant tissue water potentials [@Christoffersen.2018].
-In moist forests $\Psi$~leaf~ equilibriates with $\Psi$~soil~ at night and declines through the day as plants transpire.
+Rhizosphere soil water potential ($\Psi$~soil~) down-regulates plant tissue water potentials [@Christoffersen:2016kd2d].
+In moist forests $\Psi$~leaf~ equilibrates with $\Psi$~soil~ at night and declines through the day as plants transpire.
 $\Psi$~leaf~ in turn controls leaf conductance ($K_\textrm{leaf}$), which in turn controls stomatal conductance and photosynthetic rates [@Sack.2012].
@@
 In our ERD models $K_\textrm{leaf}$ therefore incorporates the effects of rooting depth and rhizosphere water potential.
-This approach ignores the buffering effect of stem water storage on $K_\textrm{leaf}$, which may cause an assignment of a slightly deeper ERD to a species.
-Future work should therefore explore the influence of stem water storage on estimation of ERD (see Discussion).
+This approach does not specifically incorporate the buffering effect of stem water storage on $K_\textrm{leaf}$, which may cause the model to slightly overestimate ERD (see Discussion).
 VPD has a positive impact on transpiration and thus photosynthesis and growth upto a VPD threshold, after which VPD has a negative impact [@Grossiord.2020].
@@
 \
-\    To account for the non-linear impact of VPD on growth, we used a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see Model parametarisation) to represent the impact of VPD on growth and used GPP standardised between 0 and 1 (hereafter, VPD*) in Eq.
+\    To account for the non-linear impact of VPD on growth, we used a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see Model parametarisation) and used GPP standardised between 0 and 1 (hereafter, VPD*) in Eq.
 \@ref(eq:M-chosen).
@@
 \
-\   We found a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA and WSG) and parameters of leaf vulnerability (or percent loss of conductance, (PLC)) curves for a few species ($n$ = `r length(sp.exp.param$sp)`), a 'hard' trait.
+\   We found a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA and WSG) and parameters of leaf vulnerability (or percent loss of conductance, PLC) curves for a few species ($n$ = `r length(sp.exp.param$sp)`), a 'hard' trait.
 This allowed us to scale-up the leaf vulnerability curves at the community scale ($n$ = `r length(ab.table$Species)` species, see Table S\@ref(tab:ab-param) for model predicted parameters).
@@
 Although the interpretation of these data is still under research [@Deurwaerder.2020], they are critical for model calibration or validation.
-The model in @Chitra-Tarak.2018 were not corroborated with such data as existing methods were inadequate to extract water bound to the saprolite found in their dry forest study site (Jean Riotte, pers.
-comm.).
-The model of below-ground water-availability in this study (ELM-FATES) differs from the conceptual water balance model used by CT2018 and was calibrated with data on long-term soil moisture dynamics at a point location (0-0.15m) and plot-wide measurements by depth (0.15, 0.4 and 1 m), as well as, evapotranspiration and stream discharge.
-The model in CT2018 was calibrated only on stream discharge and water table levels.
 We lacked data on water table levels which would have been useful to constrain water availability dynamics in the unobserved soil layers and the spatial variability within the plot.
@@
 
-The relationships that we identify between above- and below-ground traits, the water environment vertically below-ground and mortality rates are important for representing diversity in Dynamic Global Vegetation Models especially in species-rich forests, which intrinsically rely on the parameterization of contrasting life history strategies [@Schieter:2012] and the simulation of competition between those strategies.
+The relationships that we identify between above- and below-ground traits, the water environment vertically below-ground and mortality rates are important for representing diversity in Dynamic Global Vegetation Models, which intrinsically rely on the parameterization of contrasting life history strategies [@Scheiter:2013] and the simulation of competition between those strategies.
 We found that LMA and WSG were strong predictors of PLC curves for leaves, but not for stems (not shown).
-This offers the promise of a greater ability to parameterie the 'hard' hydraulic traits with the abundant soft trait data, thus allowing much better representation of forest hydrodynamics.
+This offers the promise of a greater ability to parameterise the 'hard' hydraulic traits with the abundant soft trait data, thus allowing much better representation of forest hydrodynamics.
 The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits, including hydraulic vulnerability, in order to better represent the functional diversity of tropical forests, and their trajectories into the future.
@@
 \
-\   Our model estimated ERD as a single value, unlike the temporally dynamic ERD modeled by @Chitra-Tarak.2018, but future model developments should investigate seasonal dynamics of ERD, using corresponding calibration data on isotopically determined water-sourcing depths, as species are expected to shift to deeper soil water sources as shallow soils dry out [@Meinzer:1999].
-Our inverse ERD model was parameterized on 5-year growth data, with five data points over a 25-yr period, which decouples climate events and demographic outcomes.
+\   Our inverse ERD model was parameterized on 5-year growth data, with five data points over a 25-yr period, which decouples climate events and demographic outcomes.
 Future studies could better constrain the ERD model with higher resolution growth data such as those from dendrometer bands.
-This is especially important for modeling seasonally dynamic ERD.
-However, trees' stem diameter change is a sum total of not just growth, the addition of wood, but also changes due to hydrostatic stem flexing (xylem, phloem and bark hydration), the importance of which increases with temporal resolution [@Chitra-Tarak.2016; @Chitra-Tarak.2015].
-Use of dendrometer data would thus entail accounting for stem hydrostatic flexing [@Zweifel:2018].
-Three of our exploratory ERD models included seasonality in leaf area fraction, but we did not select them as they worsened the fit with growth data and evaluation data ($\mathrm{\delta}$^2^H, Fig.
+However, trees' stem diameter change is a sum total of not just growth, the addition of wood, but also changes due to reversible stem dehydration (xylem, phloem and bark), the importance of which increases with temporal resolution [@Chitra-Tarak.2016; @Chitra-Tarak.2015].
+Use of dendrometer band data would thus entail accounting for stem hydrostatic flexing [@Zweifel:2018], but may also be an avenue to account for stem water storage and temporal dynamics of ERD, a limitation in our current approach.
+Three of our exploratory ERD models included seasonality in leaf area fraction, but we did not select them as they worsened the fit with growth data for most species (Fig.
 S\@ref(fig:other-erd-mods)).
-We interpret this result as that VPD and PLC curves may be adequate to explain growth, but also acknowledge that our estimates of seasonality of LAI is a tentative estimate combining leaf fall data and assumption of leaf presence during the wet season, and omits inter-annual variation.
-Future studies could make use of data from new technologies such as drone based monitoring of species-specific leaf area fraction in species-rich tropical forests [@Park.2019].
+We interpret this result as that VPD and PLC curves may be adequate to explain growth, but also acknowledge that our estimates of seasonality of LAI is a tentative estimate combining leaf fall data and the backtracked timinng of leaf-gain from leaf life-time, and omits inter-annual variation.
+Future studies could further develop models of LAI seasonality and make use of data from new technologies such as drone based monitoring of LAI in species-rich tropical forests [@Park.2019].
 In forests such as BCI where understory tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes, especially from a few growth data points, may need accounting for variation in trees' light environment over time and among trees of the same size class, an issue we minimized by limiting ERD estimates to canopy trees.
-To estimate topographic variation in ERD within or across species future studies may use soil moisture dynamics from a distributed hydrological model [e.g.
+In sites where water does not limit growth, it may not be possible to parametarize ERD models from growth dynamics, and other direct avenues such as sap-flux dynamics could be explored.
+To estimate topographic variation in ERD within or across species, future studies may use soil moisture dynamics from a distributed hydrological model [e.g.
 @Schwantes:2018].
@@
 We recommend wide-spread and coordinated collection of these variables as well as water-table levels in forest-inventory sites to allow for estimation of trees' water environments.
-Finally, developing an inverse ERD model that accounts for movement of water within the soil-plant-atmosphere-continuum (SPAC) [@Christoffersen:2016kd2d; @Sperry.1998] would be a step change from the current statistical models of ERD (this study and @Chitra-Tarak.2018).
+While our model empirically predicts growth via $\Psi$~leaf~ from $\Psi$~soil~ at a specific depth, models that mechanistically account for $\Psi$~soil~ and $\Psi$~root~ over the entire rooting profile and then $\Psi$~stem~ and $\Psi$~leaf~, accounting for water content and embolised tissues, and downstream photosynthetic pathways and non-water limiting processes are likely to be more accurate in predicting growth [@Christoffersen:2016kd2d; @Sperry.1998].
 Although data needs for parameterizing such models [@Christoffersen:2016kd2d; @Sperry.1998] could be greater (e.g.
-retention curves for roots), which limits immediate community-wide application, such models hold greater promise to improve our understanding of plant physiology [e.g.
+retention curves, capacitance for roots, stems and leaves), which increases parameter uncertainty in community-wide application, such models hold greater promise in improving our understanding of plant physiology [e.g.
 see @Johnson:2018fad].

### 2020-10-24 old_ms_sections.Rmd
# moved in discussion on Chitra-Tarak 2018

@@
 Our model was constrained on VPD and species leaf vulnerability curves, variables that are shown to be physiologically meaningful.
+
+The model in CT2018 were not corroborated with such data as existing methods were inadequate to extract water bound to the saprolite found in their dry forest study site (Jean Riotte, pers.
+comm.).
+The model of below-ground water-availability in this study (ELM-FATES) differs from the conceptual water balance model used by CT2018 and was calibrated with data on long-term soil moisture dynamics at a point location (0-0.15m) and plot-wide measurements by depth (0.15, 0.4 and 1 m), as well as, evapotranspiration and stream discharge.
+The model in CT2018 was calibrated only on stream discharge and water table levels.

### 2020-10-25 manuscript.rmd
1. Changed isotope data label 2. Removed italics for Psi and delta

@@
 These efforts have resulted in identification of correlations between aboveground hydraulic traits such as vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013, @Venturas.2016].
-Deep roots may mitigate hydraulic vulnerability and mortality-risk, yet plants' realized water environments, specifically plant's ability to acquire water via differing rooting depths in terms of hydrological rather than meteorological droughts, are rarely studied, especially in the tropical region.
+Deep roots may mitigate hydraulic vulnerability and mortality-risk, yet plants' realized water environments, specifically plant's ability to acquire water via differing rooting depths in hydrological rather than meteorological droughts, are rarely studied, especially in the tropical region.
 We tested how rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tree species of a tropical forest over 35 years that witnessed droughts of various intensity and duration.
@@
 \
-\     In the 50 ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm dbh size class [@Condit.2017].
+\     In the 50 ha plot, all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm dbh size class [@Condit.2017].
 This inventory represents 321 woody species across the eight censuses.
@@
 For greater accuracy, instead of the ELM default soil texture-based pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content vs.
-$\Psi$~soil~ for depths of 0.15, 0.4 and 1 m [@Kupers:2019].
+$\mathrm{\Psi}$~soil~ for depths of 0.15, 0.4 and 1 m [@Kupers:2019].
 We converted gravimetric water content to volumetric (VWC, $\theta$) using a bulk density value of of 0.8 g cm^-3^ measured at a point location (see below).
@@
 \end{equation}
-where $\Psi_e$ is the air-entry matric potential, $\theta_{s}$ is the saturated volumetric water content, and $b$, an index for soil pore-size distribution, respectively.
-$\Psi_e$, $\theta_{s}$ and $b$ that best-fit the data were 200 (mm H~2~O), 0.51 (cm^3^ cm^-3^) and 10 (unitless), respectively.
+where $\mathrm{\Psi}_e$ is the air-entry matric potential, $\theta_{s}$ is the saturated volumetric water content, and $b$, an index for soil pore-size distribution, respectively.
+$\mathrm{\Psi}_e$, $\theta_{s}$ and $b$ that best-fit the data were 200 (mm H~2~O), 0.51 (cm^3^ cm^-3^) and 10 (unitless), respectively.
 
@@
 We ran ELM-FATES with the chosen 100 ensemble members from 1985-2018, with the first 5 years used for model spin-up.
-Extreme soil droughts were identified by depth $z$ as days that fell below the 5th percentile of $\Psi_{\textrm{soil},z}$ for a given day of the year.
+Extreme soil droughts were identified by depth $z$ as days that fell below the 5th percentile of $\mathrm{\Psi}_{\textrm{soil},z}$ for a given day of the year.
 
@@
 There are many intrinsic and extrinsic factors that limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019].
-Rhizosphere soil water potential ($\Psi$~soil~) down-regulates plant tissue water potentials [@Christoffersen:2016kd2d].
-In moist forests $\Psi$~leaf~ equilibrates with $\Psi$~soil~ at night and declines through the day as plants transpire.
-$\Psi$~leaf~ in turn controls leaf conductance ($K_\textrm{leaf}$), which in turn controls stomatal conductance and photosynthetic rates [@Sack.2012].
-We therefore used maximum diurnal leaf conductance ($K_\textrm{leaf}$) as determined by $\Psi$~soil~ as a proxy for the soil water limitation on growth.
+Rhizosphere soil water potential ($\mathrm{\Psi}$~soil~) down-regulates plant tissue water potentials [@Christoffersen:2016kd2d].
+In moist forests $\mathrm{\Psi}$~leaf~ equilibrates with $\mathrm{\Psi}$~soil~ at night and declines through the day as plants transpire.
+$\mathrm{\Psi}$~leaf~ in turn controls leaf conductance ($K_\textrm{leaf}$), which in turn controls stomatal conductance and photosynthetic rates [@Sack.2012].
+We therefore used maximum diurnal leaf conductance ($K_\textrm{leaf}$) as determined by $\mathrm{\Psi}$~soil~ as a proxy for the soil water limitation on growth.
 In our ERD models $K_\textrm{leaf}$ therefore incorporates the effects of rooting depth and rhizosphere water potential.
@@
 \
-\    For each species, we estimated diurnal maximum value of $K_{\textrm{leaf},s,i|z,h}$ for a species $s$ by substituting $\Psi_{\textrm{soil},s,i|z,h}$in the leaf vulnerability curves defined below in Eq.
+\    For each species, we estimated diurnal maximum value of $K_{\textrm{leaf},s,i|z,h}$ for a species $s$ by substituting $\mathrm{\Psi}_{\textrm{soil},s,i|z,h}$in the leaf vulnerability curves defined below in Eq.
 \@ref(eq:vc-exp) using species-specific parameters $A_{s}$ and $B_{s}$ as:
@@
 \end{equation}
-\    By varying $z$, we estimated species effective rooting depths (ERD), defined as the depth $z$ at which soil water dynamics ($\Psi_{\textrm{soil,} z}$) best explained observed dynamics of species growth (six 5-yrly censuses, 1990-2015).
+\    By varying $z$, we estimated species effective rooting depths (ERD), defined as the depth $z$ at which soil water dynamics ($\mathrm{\Psi}_{\textrm{soil,} z}$) best explained observed dynamics of species growth (six 5-yrly censuses, 1990-2015).
 Using Eq.
@@
 where $L$ is the likelihood, $H$ = {1, .., 100}, refer to the 100 hydrological realisations and $Z$ = {0.4, `r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m.
-We removed finely resolved soil layers below 0.21 m, averaged $\Psi_{\textrm{soil},z}$ for depths 0.21, 0.37 and 0.62 at 0.4 m.
+We removed finely resolved soil layers below 0.21 m, averaged $\mathrm{\Psi}_{\textrm{soil},z}$ for depths 0.21, 0.37 and 0.62 at 0.4 m.
 
@@
 We obtained leaf hydraulic vulnerability curves ($K_\textrm{leaf}$ vs.
-$\Psi$~leaf~; see Table \@ref(tab:symbols) for trait symbols and their definitions) for adult trees of 21 common species from Zailaa et al.
+$\mathrm{\Psi}$~leaf~; see Table \@ref(tab:symbols) for trait symbols and their definitions) for adult trees of 21 common species from Zailaa et al.
 (to be submitted; see SI for brief description of methods) described as:
@@
 \#ref(eq:vc-exp).
-For all the `r length(erd.sp)` ERD species we estimated species $\Psi$~50,leaf~ based on their vulnerability curves (see Table \@ref(tab:symbols) for definition).
+For all the `r length(erd.sp)` ERD species we estimated species $\mathrm{\Psi}$~50,leaf~ based on their vulnerability curves (see Table \@ref(tab:symbols) for definition).
 
@@
 
-Xylem sap $\mathrm{\delta}$^2^H values can provide an index of rooting depth, which we used to validate the ERD models as an independent observation of ERD.
-We used xylem sap $\mathrm{\delta}$^2^H values from BCI given in @Meinzer:2001 for the dry-season of March 1997 as this period showed largest seasonal divergence among species, with a corresponding vertical gradient of $\mathrm{\delta}$^2^H in the soil.
+Xylem sap $\mathrm{\delta}$^2^H values (hereafter, $\mathrm{\delta}$^2^H~xylem~) can provide an index of rooting depth, which we used to validate the ERD models as an independent observation of ERD.
+We used species $\mathrm{\delta}$^2^H~xylem~ values from BCI given in @Meinzer:2001 for the dry-season of March 1997 as this period showed largest seasonal divergence among species, with a corresponding vertical gradient of $\mathrm{\delta}$^2^H in the soil (hereafter, $\mathrm{\delta}$^2^H~soil~ ).
 (Fig.
@@
 Briefly, this isotopic approach makes use of a naturally formed gradient in deuterium to hydrogen ratio of soil water vertically belowground, which is generally formed due to evaporative enrichment (here, during the dry season).
-Shallow soil water has a greater deuterium to hydrogen content than deep soil water (here, $\mathrm{\delta}$^2^H ranged from -16.5 to -50 ‰ over 0.05 to 1 m depth), with ground-water having the lowest deuterium to hydrogen content (here, -60 ‰), typically reflecting un-fractionated rain water.
-As root water-uptake is generally a non-fractionating process, $\mathrm{\delta}$^2^H of water in the xylem sap of a tree reflects the signature of the source water.
-Thus water-sourcing depth of the tree can be identified as the depth with the same value of $\mathrm{\delta}$^2^H as that of the xylem sap, with the uncertainty that the value of $\mathrm{\delta}$^2^H may also result from a mixing of water from shallower or deeper depths [@Dawson.1991].
-Analyses of xylem sap $\mathrm{\delta}$^2^H of the 12 species studied by @Meinzer:2001 showed that four species sourced water between 1 m and ground water.
-Two other had water access at least 0.8 m deep, one was restricted to water access between 0 to 0.2 m and five fell in the range over which the relationship between soil depth and soil water $\mathrm{\delta}$^2^H was variable in the 0.2 to 0.7 m depth range.
+Shallow soil water has a greater deuterium to hydrogen content than deep soil water (here, $\mathrm{\delta}$^2^H~soil~ ranged from -16.5 to -50 ‰ over the depth of 0.5-1 m), with ground-water having the lowest deuterium to hydrogen content (here, -60 ‰), typically reflecting un-fractionated rain water.
+As root water-uptake is generally a non-fractionating process, $\mathrm{\delta}$^2^H~xylem~ of a tree reflects the signature of the source water.
+Thus water-sourcing depth of the tree can be identified as the depth with the same value of $\mathrm{\delta}$^2^H~soil~, or as that of $\mathrm{\delta}$^2^H~xylem~, with the uncertainty that the value of $\mathrm{\delta}$^2^H~xylem~ may also result from a mixing of water from shallower or deeper depths [@Dawson.1991].
+Analyses of $\mathrm{\delta}$^2^H~xylem~ of the 12 species studied by @Meinzer:2001 showed that four species sourced water between 1 m and ground water.
+Two other had water access at least 0.8 m deep, one was restricted to water access between 0 to 0.2 m and five fell in the range over which the relationship between soil depth and $\mathrm{\delta}$^2^H~soil~ was variable in the 0.2 to 0.7 m depth range.
 
-\   We compared six species with data for $\mathrm{\delta}$^2^H as well as modelled estimates of ERD, after removing those species in @Meinzer:2001 that are likely to have no leaves in March 1997 because xylem sap $\mathrm{\delta}$^2^H of these species may not be linked with water sourced at the time of measurement and thus may not be comparable with other species that had leaves.
+\   We compared six species with data for $\mathrm{\delta}$^2^H~xylem~ as well as modelled estimates of ERD, after removing those species in @Meinzer:2001 that are likely to have no leaves in March 1997 because xylem sap $\mathrm{\delta}$^2^H of these species may not be linked with water sourced at the time of measurement and thus may not be comparable with other species that had leaves.
 We also removed *Guapira standleyana*, whose xylem sap isotope value was not well constrained by the soil water isotope depth profile.
@@
 \@ref(eq:M-chosen) and Eqs.
-in SI), estimated species ERD was regressed against species xylem sap $\mathrm{\delta}$^2^H (‰).
+in SI), estimated species ERD was regressed against species $\mathrm{\delta}$^2^H~xylem~ (‰).
 
@@
 
-To evaluate whether ERD was associated with above-ground hydraulic traits ($n$ = 8 species with both datasets), we regressed species ERD against maximum stem hydraulic conductance ($K_\textrm{max,stem}$), leaf turgor loss point ($\Psi$~tlp~), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi$~88,stem~), and hydraulic safety margins ($\Psi$~min~ - $\Psi$~88,stem~).
+To evaluate whether ERD was associated with above-ground hydraulic traits ($n$ = 8 species with both datasets), we regressed species ERD against maximum stem hydraulic conductance ($K_\textrm{max,stem}$), leaf turgor loss point ($\mathrm{\Psi}$~tlp~), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\mathrm{\Psi}$~88,stem~), and hydraulic safety margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~).
 See [SI](#SIData.brett) for data collection and estimation of these variables.
@@
 \   We regressed species ERD against species ($\textrm{Mort}_{s,t}$ for each of the census interval $t$).
-To assess whether deeper ERD species escaped the experience of significant soil drought, we calculated for each species the proportion of days in the simulation period (1990-2015) during which soil water potential in the soil zone matching species ERD ($\Psi_{\textrm{soil},z=\textrm{ERD}}$) were below species $\Psi$~crit~.
-$\Psi$~crit~ was defined as $\Psi$~50,leaf~.
+To assess whether deeper ERD species escaped the experience of significant soil drought, we calculated for each species the proportion of days in the simulation period (1990-2015) during which soil water potential in the soil zone matching species ERD ($\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$) were below species $\mathrm{\Psi}$~crit~.
+$\mathrm{\Psi}$~crit~ was defined as $\mathrm{\Psi}$~50,leaf~.
 
@@
 
-\   Soil Water Potential dynamics ($\Psi_{\textrm{soil},z}$) by depth $z$ (from 0.1 through 13 m) obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of such extreme years, drought intensity at depth, drought seasonality and duration (Fig.
+\   Soil Water Potential dynamics ($\mathrm{\Psi}_{\textrm{soil},z}$) by depth $z$ (from 0.1 through 13 m) obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of such extreme years, drought intensity at depth, drought seasonality and duration (Fig.
 \@ref(fig:swp-droughts)).
@@
 S\@ref(fig:swp-droughts)).
-Across the simulation period (1990-2015) $\Psi_{\textrm{soil},z}$ remained above -0.5 MPa from depths 1.7 to 13 m in the simulations (Fig.
+Across the simulation period (1990-2015) $\mathrm{\Psi}_{\textrm{soil},z}$ remained above -0.5 MPa from depths 1.7 to 13 m in the simulations (Fig.
 S\@ref(fig:swp-all-depths)).
@@
 The chosen model (with a mulicative effect of VPD and not LAI; Eq.
-\@ref(eq:M-chosen)) explained a large fraction of the variance in the xylem sap isotopic signature ($\mathrm{\delta}$^2^H); $R$^2^ = `r lm.erd.iso.r2`; $p$ = `r lm.erd.iso.p`, $n$ = `r length(unique(erd.iso$sp))`; Fig.
+\@ref(eq:M-chosen)) explained a large fraction of the variance in $\mathrm{\delta}$^2^H~xylem~ ($R$^2^ = `r lm.erd.iso.r2`; $p$ = `r lm.erd.iso.p`, $n$ = `r length(unique(erd.iso$sp))`; Fig.
 \@ref(fig:erd-iso)).
-Two of the alternative models tested with LAI had a strong relationship with $\mathrm{\delta}$^2^H as well, but one that was based on a smaller set of species ($n$ = 4) as fewer species passed the criteria of sufficient species-level goodness-of-fit with these models; <20 speceis vs.
+Two of the alternative models tested with LAI had a strong relationship with $\mathrm{\delta}$^2^H~xylem~ as well, but one that was based on a smaller set of species ($n$ = 4) as fewer species passed the criteria of sufficient species-level goodness-of-fit with these models; <20 speceis vs.
 `r length(erd.sp)` species for the chosen model (Fig.
@@
 seemannii* and *T.
-tuberculata*, that @Meinzer:1999 found to have xylem sap $\mathrm{\delta}$^2^H values between soil water and groundwater, suggesting that these species sourced most of the water from depths >1 m, were also identified by our model with ERD >1 m (2.9 m for both species; Fig.
+tuberculata*, that @Meinzer:1999 found to have $\mathrm{\delta}$^2^H~xylem~ values between soil water and groundwater, suggesting that these species sourced most of the water from depths >1 m, were also identified by our model with ERD >1 m (2.9 m for both species; Fig.
 \@ref(fig:erd-iso)).
@@
 For the subset of eight species with ERD for which hydraulic traits data were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_\textrm{max,stem}$, Spearmans' $r$ = `r cor.k`, $p$ = `r p.k`; Fig.
-\@ref(fig:erd-hyd)A), less negative leaf $\Psi$~tlp~ (Spearmans' $r$ = `r cor.tlp`, $p$ = `r p.tlp`; Fig.
-\@ref(fig:erd-hyd)B), less negative $\Psi$~88,stem~ and thus greater vulnerability to embolism from cavitation (Spearmans' $r$ = `r cor.stem.88`, $p$ = `r p.stem.88`; Fig.
-\@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi$~min~ - $\Psi$~88,stem~; Spearmans' $r$ = `r cor.hsm`, $p$ = `r p.hsm`; Fig.
+\@ref(fig:erd-hyd)A), less negative leaf $\mathrm{\Psi}$~tlp~ (Spearmans' $r$ = `r cor.tlp`, $p$ = `r p.tlp`; Fig.
+\@ref(fig:erd-hyd)B), less negative $\mathrm{\Psi}$~88,stem~ and thus greater vulnerability to embolism from cavitation (Spearmans' $r$ = `r cor.stem.88`, $p$ = `r p.stem.88`; Fig.
+\@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~; Spearmans' $r$ = `r cor.hsm`, $p$ = `r p.hsm`; Fig.
 \@ref(fig:erd-hyd)D).
@@
 Soil water availability was not modelled for 1985-1990.
-Analysis of a species' time spent under significant soil drought (proportion of days $\Psi_{\textrm{soil},z=\textrm{ERD}}$ < $\Psi$~crit~) showed that species with shallower ERD spent greater proportion of days below critical hydraulic thresholds (Fig.
+Analysis of a species' time spent under significant soil drought (proportion of days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) showed that species with shallower ERD spent greater proportion of days below critical hydraulic thresholds (Fig.
 \@ref(fig:erd-time); S\@ref(fig:erd-time-evg)).
-Species with ERD deeper than 1 m did not cross species $\Psi$~crit~ for the entire 25-yr period for which soil water potential $\Psi_{\textrm{soil},z}$ was modeled (1990-2015).
-Soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`;  median, 95% CI) through all the dry seasons and drought periods, which is above ($\Psi$~crit~) of the most sensitive tree species (-0.28 MPa).
+Species with ERD deeper than 1 m did not cross species $\mathrm{\Psi}$~crit~ for the entire 25-yr period for which soil water potential $\mathrm{\Psi}_{\textrm{soil},z}$ was modeled (1990-2015).
+Soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`;  median, 95% CI) through all the dry seasons and drought periods, which is above ($\mathrm{\Psi}$~crit~) of the most sensitive tree species (-0.28 MPa).
 
@@
 \@ref(fig:erd-mort-census-evg)).
-Analysis of time spent below ($\Psi$~crit~) in the soil zone matching ERD $\Psi_{\textrm{soil},z=\textrm{ERD}}$ showed that this was likely due to decreasing realized risk of hydraulic failure with increasing access to water for species with deeper ERD (Fig.
+Analysis of time spent below ($\mathrm{\Psi}$~crit~) in the soil zone matching ERD $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ showed that this was likely due to decreasing realized risk of hydraulic failure with increasing access to water for species with deeper ERD (Fig.
 \@ref(fig:erd-time); Fig.
@@
 Deeper ERD species on average had more hydraulically efficient, but risky, suite of traits optimized for an environment of more reliable and greater water source at depth that resulted in a lower mortality risk.
-By contrast shallow ERD species had inefficient, but safer, xylem likely optimized for high variability in $\Psi_{soil}$ in the shallow soil layers, and resulting in higher variance in mortality rates across censuses as $\Psi_{soil,~z = ERD}$ crossed ERD $\Psi_{crit}$.
-Our results are consistent with a few studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016].
-@Johnson:2018fad found that mortality decreased with deep rooting even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi$~88,stem~) and had narrower safety-margins ($\Psi$~min~ - $\Psi$~88,stem~).
-@Nardini:2015 found that a species sourcing deep water in a Karst system showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi$~88,stem~ and narrower safety margin compared to two species sourcing shallow-water.
+By contrast shallow ERD species had inefficient, but safer, xylem likely optimized for high variability in $\mathrm{\Psi}_{soil}$ in the shallow soil layers, and resulting in higher variance in mortality rates across censuses as $\mathrm{\Psi}_{soil,~z = ERD}$ crossed ERD $\mathrm{\Psi}_{crit}$.
+Our result of deep rooted species being more hydraulically vulnerble, is consistent with a seasonal forest in the Amazon, where among 12 co-occurring species trees sourcing deeper water were 50 and 70% of variance in $\mathrm{\Psi}_\textrm{88,stem}$ ($\mathrm{\delta}$^18^O~xylem~) Our results are also consistent with a few studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016].
+@Johnson:2018fad found that mortality decreased with deep rooting even though deep rooted species were more vulnerable to xylem cavitation (high $\mathrm{\Psi}$~88,stem~) and had narrower safety-margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~).
+@Nardini:2015 found that a species sourcing deep water in a Karst system showed lower hydraulic failure and mortality in an extreme drought despite having high $\mathrm{\Psi}$~88,stem~ and narrower safety margin compared to two species sourcing shallow-water.
 
@@
 \   Our result--of lower mortality in deep rooted trees--contrasts with the finding of @Chitra-Tarak.2018 (hereafter CT2018) in which species identified by their inverse ERD model as effectively deep rooted died *more* in a rare, prolonged drought.
-The hydrological model in that study revealed that the multi-year drought exhausted the deep soil and even bedrock water availablity.
-In contrast, our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period, $\Psi$~soil~ below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and droughts, that is above ($\Psi$~crit~) of the most sensitive tree species (-0.28 MPa).
+The hydrological model in that study revealed that the multi-year drought exhausted the deep soil and even bedrock water availablity (also see, @Goulden.2019).
+In contrast, our hydrological modeling at BCI finds that during the study period, $\mathrm{\Psi}$~soil~ below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and droughts, that is above ($\mathrm{\Psi}$~crit~) of the most sensitive tree species (-0.28 MPa).
 The difference in the results shown here are likely attributable to site differences in the precipitation regime.
@@
 S\@ref(fig:swp-all-depths); @Chitra-Tarak.2018).
-The kind of droughts that buffered the deep rooted species in our study were recurrent sub-annual droughts that extended the dry season into the wet season.
+The kind of droughts that buffered the deep rooted species in our study were repeated annual droughts that extended the dry season into the wet season.
 The variation in mortality explained by ERD in our study is upto 40%, and other causes may explain observed mortality [@McDowell.2018].
@@
 WSG alone explained `r round(b.wsg.r2*100,0)`% of the variance in $B$ ($p$ `r  b.wsg.p`).
-We find that leaf conductance declined more rapidly with declining $\Psi$~leaf~ for species with higher values of WSG and LMA (Fig.
+We find that leaf conductance declined more rapidly with declining $\mathrm{\Psi}$~leaf~ for species with higher values of WSG and LMA (Fig.
 S\@ref(fig:obs-plc-curves)).
@@
 However, trees' stem diameter change is a sum total of not just growth, the addition of wood, but also changes due to reversible stem dehydration (xylem, phloem and bark), the importance of which increases with temporal resolution [@Chitra-Tarak.2016; @Chitra-Tarak.2015].
-Use of dendrometer band data would thus entail accounting for stem hydrostatic flexing [@Zweifel:2018], but may also be an avenue to account for stem water storage and temporal dynamics of ERD, a limitation in our current approach.
-Three of our exploratory ERD models included seasonality in leaf area fraction, but we did not select them as they worsened the fit with growth data for most species (Fig.
+Use of dendrometer band data would thus entail accounting for reversible stem dehydration [@Zweifel:2018], but may also be an avenue to account for stem water storage and temporal dynamics of ERD, a limitation in our current approach.
+Three of our exploratory ERD models included seasonality in leaf area fraction, but we did not select them as they worsened the fit with growth data for many species (Fig.
 S\@ref(fig:other-erd-mods)).
-We interpret this result as that VPD and PLC curves may be adequate to explain growth, but also acknowledge that our estimates of seasonality of LAI is a tentative estimate combining leaf fall data and the backtracked timinng of leaf-gain from leaf life-time, and omits inter-annual variation.
+We interpret this result as that VPD and PLC curves may be adequate to explain growth, but also acknowledge that our estimates of seasonality of LAI is a tentative estimate combining leaf fall data and the timing of  leaf-gain backtracked from leaf lifetime, and omits inter-annual variation.
 Future studies could further develop models of LAI seasonality and make use of data from new technologies such as drone based monitoring of LAI in species-rich tropical forests [@Park.2019].
-In forests such as BCI where understory tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes, especially from a few growth data points, may need accounting for variation in trees' light environment over time and among trees of the same size class, an issue we minimized by limiting ERD estimates to canopy trees.
-In sites where water does not limit growth, it may not be possible to parametarize ERD models from growth dynamics, and other direct avenues such as sap-flux dynamics could be explored.
-To estimate topographic variation in ERD within or across species, future studies may use soil moisture dynamics from a distributed hydrological model [e.g.
+
+\
+\   In forests such as BCI where understory tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes, especially from a few growth data points, may need accounting for variation in trees' light environment over time and among trees of the same size class, an issue we minimized by limiting ERD estimates to canopy trees.
+In sites where water does not limit growth, it may not be possible to parametarize ERD models from growth dynamics, and other avenues such as sap-flux or leaf water-potential dynamics could be explored [@Brum.2017].
+While our ERD model empirically predicts growth via $\mathrm{\Psi}$~leaf~ from $\mathrm{\Psi}$~soil~ at a specific depth, models that mechanistically account for $\mathrm{\Psi}$~soil~ and $\mathrm{\Psi}$~root~ over the entire rooting profile and then $\mathrm{\Psi}$~stem~ and $\mathrm{\Psi}$~leaf~, accounting for water content and embolised tissues, and downstream photosynthetic pathways and non-water limiting processes are likely to be more accurate in predicting growth [@Christoffersen:2016kd2d; @Sperry.1998].
+Although data needs for parameterizing such models [@Christoffersen:2016kd2d; @Sperry.1998] could be greater (e.g.
+retention curves, capacitance for roots, stems and leaves), adding a degree of parameter uncertainty in community-wide application, such models hold greater promise in improving our understanding of plant physiology [e.g.
+see @Johnson:2018fad].
+
+\
+\    To estimate topographic variation in ERD within or across species, future studies may use soil moisture dynamics from a distributed hydrological model [e.g.
 @Schwantes:2018].
-We note that data availability on soil water retention curve ($\Psi$~soil~ vs.
+We note that data availability on soil water retention curves ($\mathrm{\Psi}$~soil~ vs.
 VWC), hydraulic conductivity (K~soil~) by depth, soil moisture by depth, stream discharge and evapotranspiration were important for effective calibration of our 1-D hydrological model.
 We recommend wide-spread and coordinated collection of these variables as well as water-table levels in forest-inventory sites to allow for estimation of trees' water environments.
-While our model empirically predicts growth via $\Psi$~leaf~ from $\Psi$~soil~ at a specific depth, models that mechanistically account for $\Psi$~soil~ and $\Psi$~root~ over the entire rooting profile and then $\Psi$~stem~ and $\Psi$~leaf~, accounting for water content and embolised tissues, and downstream photosynthetic pathways and non-water limiting processes are likely to be more accurate in predicting growth [@Christoffersen:2016kd2d; @Sperry.1998].
-Although data needs for parameterizing such models [@Christoffersen:2016kd2d; @Sperry.1998] could be greater (e.g.
-retention curves, capacitance for roots, stems and leaves), which increases parameter uncertainty in community-wide application, such models hold greater promise in improving our understanding of plant physiology [e.g.
-see @Johnson:2018fad].
 
@@
 Trees realized water environments and water-access are nonetheless rarely studied.
-To our knowledge this is the first study to demonstrate a mechanistic link between trees' realized drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.
+To our knowledge, this is the first study to demonstrate a mechanistic link between realized drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a tree community.
 We report here that deep-water access plays a role in alleviating mortality of hydrualically vulnerable trees.
@@
 @Wijk.2001) as well as trees' hydrological environments.
-Our community-scale framework for modeling effective rooting depths and leaf vulnerability curves indicate the possibilities in expanding the use of these critical, hard-earned observations in species-rich forests towards community-scale generalizations.
+Our community-scale framework for modeling effective rooting depths and leaf vulnerability curves indicates the possibilities in expanding the use of these critical, hard-earned observations in species-rich forests towards community-scale generalizations.
 
@@
 Figure \@ref(fig:ab-fit): Occurrence of extreme soil droughts by census intervals (horizontal panels) at three representative depths (z): 0.1, 0.6 and 1 m (vertical panels).
-Mean (black lines) and lower half of 95% distribution (grey areas) of soil water potential ($\Psi_{\textrm{soil},z}$) for a given DOY and depth $z$ are shown (same across census intervals).
-$\Psi_{\textrm{soil},z}$ by DOY for a year (colored line) is only shown if for some part of that year $\Psi_{\textrm{soil},z}$ declined below the 5 percentile distribution.
+Mean (black lines) and lower half of 95% distribution (grey areas) of soil water potential ($\mathrm{\Psi}_{\textrm{soil},z}$) for a given DOY and depth $z$ are shown (same across census intervals).
+$\mathrm{\Psi}_{\textrm{soil},z}$ by DOY for a year (colored line) is only shown if for some part of that year $\mathrm{\Psi}_{\textrm{soil},z}$ declined below the 5 percentile distribution.
 
@@
 Figure \@ref(fig:ab-fit): Parameters $A$ (panel A) & $B$ (panel B) for the leaf vulnerability curves that were fitted to observed data on $K_\textrm{leaf}$ vs.
-$\Psi$~leaf~ for `r length(sp.exp.param$sp)` species (Eq.
+$\mathrm{\Psi}$~leaf~ for `r length(sp.exp.param$sp)` species (Eq.
 \@ref(eq:vc-exp)) vs.
@@
 \
-Figure \@ref(fig:erd-iso): Modelled Effective Rooting Depth (ERD; mean $\pm$ 1SE; m) versus dry season stable isotopic concentration $\mathrm{\delta}$^2^H (mean $\pm$ 1SE; ‰) for `r length(unique(erd.iso$sp))` canopy species from Barro Colorado Island.
+Figure \@ref(fig:erd-iso): Modelled Effective Rooting Depth (ERD; mean $\pm$ 1SE; m) versus dry season stable isotopic concentration $\mathrm{\delta}$^2^H~xylem~ (mean $\pm$ 1SE; ‰) for `r length(unique(erd.iso$sp))` canopy species from Barro Colorado Island.
 Data from [@Meinzer:1999].
@@
 $p$-values are given only when they are $\ge$0.05.
-Units are as follows: ERD in m, K~max~ in mmol m^-1^ s^-1^ MPa^-1^), all variables with $\Psi$ in MPa.
+Units are as follows: ERD in m, K~max~ in mmol m^-1^ s^-1^ MPa^-1^), all variables with $\mathrm{\Psi}$ in MPa.
 
@@
 \
-Figure \@ref(fig:erd-time): Modelled species Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\Psi_{\textrm{soil},z}$ was below $\Psi$~crit~, defined as $\Psi$~80,leaf~, over 1990-2015, where $z$ is the soil depth matching species ERD--for `r length(erd.mrate.sp)` canopy species found on Barro Colorado Island.
+Figure \@ref(fig:erd-time): Modelled species Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\mathrm{\Psi}_{\textrm{soil},z}$ was below $\mathrm{\Psi}$~crit~, defined as $\mathrm{\Psi}$~80,leaf~, over 1990-2015, where $z$ is the soil depth matching species ERD--for `r length(erd.mrate.sp)` canopy species found on Barro Colorado Island.
 Each circle represents a species.
-Circles are colored and jittered vertically around the named depths (ticks) for clarity.
+Circles are colored and slightly jittered horizontally for clarity.
 
@@
 The $K_\textrm{leaf}$ is calculated as the ratio of the transpiration rate (*E*~leaf~; mmol s^-1^) to the difference in flow pressure (MPa) upstream and downstream from the resistance tubing, and further standardized to 25°C and by leaf area.
-At stability (CV < 5%), the leaf is left to equilibrate in plastic bags with moist paper towels and measured for $\Psi$~leaf~ using a Scholander pressure chamber.
+At stability (CV < 5%), the leaf is left to equilibrate in plastic bags with moist paper towels and measured for $\mathrm{\Psi}$~leaf~ using a Scholander pressure chamber.
 To construct vulnerability curves for each species, four different functions were fitted to species-specific $K_\textrm{leaf}$ vs.
-$\Psi$~leaf~ data using the *anneal* optimization function in the R package likelihood: exponential without an intercept ($K_{leaf} = Ae^{-B\Psi_{leaf}}$), exponential with an intercept ($K_{leaf} = C + Ae^{-B\Psi_{leaf}}$), logistic ($K_{leaf} = \frac{A}{1 + \left(\frac{\Psi_{leaf}}{x_{0}}\right)^B}$) and sigmoidal ($K_{leaf} = \frac{A}{1 + e^{-\left(\frac{\Psi_{leaf} - x_{0}}{B}\right)}}$) [@Sack.2012].
+$\mathrm{\Psi}$~leaf~ data using the *anneal* optimization function in the R package likelihood: exponential without an intercept ($K_{leaf} = Ae^{-B\Psi_{leaf}}$), exponential with an intercept ($K_{leaf} = C + Ae^{-B\Psi_{leaf}}$), logistic ($K_{leaf} = \frac{A}{1 + \left(\frac{\Psi_{leaf}}{x_{0}}\right)^B}$) and sigmoidal ($K_{leaf} = \frac{A}{1 + e^{-\left(\frac{\Psi_{leaf} - x_{0}}{B}\right)}}$) [@Sack.2012].
 The differences among AICc values for these functions were not large within species (see Zailaa et al.
@@
 Among the `r length(sp.exp.param$sp)` species with observed data on ($K_\textrm{leaf}$ vs.
-$\Psi$~leaf~) used to build the community-level models of leaf vulnerability curves (Eqs.
+$\mathrm{\Psi}$~leaf~) used to build the community-level models of leaf vulnerability curves (Eqs.
 \@ref(eq:b-mod)-\@ref(eq:a-mod)), two lacked data on WSG and five on LMA~lamina~.
@@
 
-```{r iso-data, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-iso-data}Stable isotopic ratio ($\\delta^2H$) of soil water by depth measured at Barro Colorado Island in March 1997.
+```{r iso-data, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-iso-data}Stable isotopic ratio ($\\mathrm{\\delta}^2H$) of soil water by depth measured at Barro Colorado Island in March 1997.
 Data from @Meinzer:1999."}
@@
 ```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}Vulnerability curves fitted to observed $K_\\textrm{leaf}$ vs.
-$\\Psi$~leaf~ data on ", length(sp.exp.param$sp)," tree species from BCI (A & B) with equation \\@ref(eq:vc-exp)--in absolute terms (A, B), as well as, in terms of loss of conductivity (%) (C, D).
+$\\mathrm{\\Psi}$~leaf~ data on ", length(sp.exp.param$sp)," tree species from BCI (A & B) with equation \\@ref(eq:vc-exp)--in absolute terms (A, B), as well as, in terms of loss of conductivity (%) (C, D).
 Each line represents a species, color coded by WSG (A, C), or LMA (B, D), with blue tones indicating smaller values of WSG or LMA and red tones indicating larger values.")}
@@
 Each line represents a species, color coded by source of the curve: fitted to observed data for $K_\\textrm{leaf}$ vs.
-$\\Psi$~leaf~ are in yellow and those based on scaling relationship with WSG and LMA are in gray.")}
+$\\mathrm{\\Psi}$~leaf~ are in yellow and those based on scaling relationship with WSG and LMA are in gray.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/obs.mod_plccurves_fits_ggplot_no_col.jpeg"))
@@
 
-```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-other-erd-mods}Effective Rooting Depth versus $\\delta^2H$ for each of the ERD models tested (panels).
+```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-other-erd-mods}Effective Rooting Depth versus $\\mathrm{\\delta}^2H$ for each of the ERD models tested (panels).
 For each ERD model only those species have an estimate of ERD that have passed the test of Pearson's $r$ > 0 and $R$^2^ > 0.5 during ERD parameterisation.
-The number of species with an ERD estimate as well as $\\delta^2H$.
-or each model are shown in inset, along with the goodness-of-fit ($R$^2^) and $p$-values for a linear relationship between ERD and $\\delta^2H$.
+The number of species with an ERD estimate as well as $\\mathrm{\\delta}^2H$.
+or each model are shown in inset, along with the goodness-of-fit ($R$^2^) and $p$-values for a linear relationship between ERD and $\\mathrm{\\delta}^2H$.
 ERD models in panels A-E correspond to Eq.
@@
 
-```{r erd-time-evg, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time-evg}Modelled species Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\\Psi_{\textrm{soil},z}$ was below $\\Psi$~crit~, defined as $\\Psi$~80,leaf~, over 1990-2015, where $z$ is the soil depth matching species ERD--for ", nrow(mrate.depth.mean %>% subset(deciduous == "E")), " canopy species found on Barro Colorado Island.
+```{r erd-time-evg, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time-evg}Modelled species Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\\mathrm{\\Psi}_{\textrm{soil},z}$ was below $\\mathrm{\\Psi}$~crit~, defined as $\\mathrm{\\Psi}$~80,leaf~, over 1990-2015, where $z$ is the soil depth matching species ERD--for ", nrow(mrate.depth.mean %>% subset(deciduous == "E")), " canopy species found on Barro Colorado Island.
 Each circle represents a species.
@@
 ```{r ab-param, results = 'asis', echo = FALSE}
-knitr::kable(ab.table, caption = 'Leaf vulnerability curves parameters A & B, $K_\\textrm{max,leaf}$ and $\\Psi$ for the ", length(erd.sp), " species with ERD estimates.
+knitr::kable(ab.table, caption = 'Leaf vulnerability curves parameters A & B, $K_\\textrm{max,leaf}$ and $\\mathrm{\\Psi}$ for the ", length(erd.sp), " species with ERD estimates.
 Species for which curves were fitted to observed data for $K_\\textrm{leaf}$ vs.
-$\\Psi$~leaf~ are identified with Source as Data, while species for which curves were obtained based on trait-proxies are identified with Source as Model', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(ab.table))))
+$\\mathrm{\\Psi}$~leaf~ are identified with Source as Data, while species for which curves were obtained based on trait-proxies are identified with Source as Model', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(ab.table))))
 ```

### 2020-10-25 manuscript.rmd
1. added references to Brum et al.

@@
 By contrast shallow ERD species had inefficient, but safer, xylem likely optimized for high variability in $\mathrm{\Psi}_{soil}$ in the shallow soil layers, and resulting in higher variance in mortality rates across censuses as $\mathrm{\Psi}_{soil,~z = ERD}$ crossed ERD $\mathrm{\Psi}_{crit}$.
-Our result of deep rooted species being more hydraulically vulnerble, is consistent with a seasonal forest in the Amazon, where among 12 co-occurring species trees sourcing deeper water were 50 and 70% of variance in $\mathrm{\Psi}_\textrm{88,stem}$ ($\mathrm{\delta}$^18^O~xylem~) Our results are also consistent with a few studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016].
+Our result of deep rooted species being more hydraulically vulnerable is consistent with observations for 12 co-occurring species in a seasonal forest in the Amazon, amongst which species $\mathrm{\delta}$^18^O~xylem~ explained 70% variance in species $\mathrm{\Psi}_\textrm{88,stem}$ [@Brum.2019].
+Our results are also consistent with a few studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016].
 @Johnson:2018fad found that mortality decreased with deep rooting even though deep rooted species were more vulnerable to xylem cavitation (high $\mathrm{\Psi}$~88,stem~) and had narrower safety-margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~).

### 2020-10-25 manuscript.rmd
1. Edited discussion

@@
 ## Soil water dynamics by depth
-
-The 100 ELM-FATES simulations with the chosen best-fit parameter sets (Table \@ref(tab:param)) successfully captured the dynamics of soil moisture by depth (Fig.
-\@ref(fig:hydro-calib) A & Fig.
-\@ref(fig:hydro-calib)B), including the observations from the horizontal TDR probes that were not included in the calibration.
+\
+The 100 ELM-FATES ensemble members with the best fits to observed soil moisture dynamics captured soil moisture seasonality at multiple depths (Fig.
+\@ref(fig:hydro-calib)A, B), including the out of sample observations.
 These simulations also captured the dynamics in stream discharge (Fig.
@@
 \   Our result--of lower mortality in deep rooted trees--contrasts with the finding of @Chitra-Tarak.2018 (hereafter CT2018) in which species identified by their inverse ERD model as effectively deep rooted died *more* in a rare, prolonged drought.
-The hydrological model in that study revealed that the multi-year drought exhausted the deep soil and even bedrock water availablity (also see, @Goulden.2019).
+The hydrological model in that study revealed that the multi-year drought exhausted the deep soil and even bedrock water availablity (also see, @Goulden.2019 and @Ivanov.2012).
 In contrast, our hydrological modeling at BCI finds that during the study period, $\mathrm{\Psi}$~soil~ below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and droughts, that is above ($\mathrm{\Psi}$~crit~) of the most sensitive tree species (-0.28 MPa).
@@
 
+\
+\   As our ERD model assumes diurnal maximum $\mathrm{\Psi}$~leaf~ as equivalent to $\mathrm{\Psi}$~soil~, ignoring the buffering effect of stem water storage capacitance, this approach may overestimate ERD in species where stem water storage capacitance is important as a drought-avoidance strategy, that is, the group of deciduous species [Borchert.2005jma].
+
 ### Future directions
@@
 Future studies could further develop models of LAI seasonality and make use of data from new technologies such as drone based monitoring of LAI in species-rich tropical forests [@Park.2019].
+This would also open the possibility of testing, whether deciduous species are more shallow rooted than evergreen species as generally observed [@Fan:2017].
 
@@
 In sites where water does not limit growth, it may not be possible to parametarize ERD models from growth dynamics, and other avenues such as sap-flux or leaf water-potential dynamics could be explored [@Brum.2017].
-While our ERD model empirically predicts growth via $\mathrm{\Psi}$~leaf~ from $\mathrm{\Psi}$~soil~ at a specific depth, models that mechanistically account for $\mathrm{\Psi}$~soil~ and $\mathrm{\Psi}$~root~ over the entire rooting profile and then $\mathrm{\Psi}$~stem~ and $\mathrm{\Psi}$~leaf~, accounting for water content and embolised tissues, and downstream photosynthetic pathways and non-water limiting processes are likely to be more accurate in predicting growth [@Christoffersen:2016kd2d; @Sperry.1998].
+
+\
+\   While our ERD model empirically predicts growth via $\mathrm{\Psi}$~leaf~ from $\mathrm{\Psi}$~soil~ at a specific depth, models that mechanistically account for $\mathrm{\Psi}$ through soil, the entire rooting profile, as well as, $\mathrm{\Psi}$~stem~ and $\mathrm{\Psi}$~leaf~, accounting for water content and accumulation of embolised tissues, and then downstream photosynthetic pathways and non-water limiting processes influencing growth are likely to be more accurate [@Christoffersen:2016kd2d; @Sperry.1998].
 Although data needs for parameterizing such models [@Christoffersen:2016kd2d; @Sperry.1998] could be greater (e.g.
-retention curves, capacitance for roots, stems and leaves), adding a degree of parameter uncertainty in community-wide application, such models hold greater promise in improving our understanding of plant physiology [e.g.
+retention curves and capacitance for roots, stems and leaves), adding a degree of parameter uncertainty in community-wide application, such models hold greater promise in improving our understanding of plant physiology [e.g.
 see @Johnson:2018fad].
@@
 We recommend wide-spread and coordinated collection of these variables as well as water-table levels in forest-inventory sites to allow for estimation of trees' water environments.
+
+\
+\ Finally, rooting depths may be parametarized within DGVMs such as ELM-FATES using plant hydrodynamic schemes [@Christoffersen:2016kd2d; @Koven.2020], simultaneously modeling plant water-availability with plant water-extraction.
+In general, including differences in tree rooting depths in ELM-FATES would improve understanding of plant competition for water (and other resources) and more accurately represent plant's drought strategies, leading to better predictions of existing and future forest function and composition.
 

### 2020-10-25 manuscript.rmd
1. edited symbols in SI to match with the manuscript

@@
 The tower used for these measurements is 41 m above ground, on a plateau on BCI [@Pau.2018].
-The eddy covariance system includes a sonic anemometer (CSAT3, Campbell Scientific, Logan, UT) and an open-path infrared CO2/H2O gas analyzer (LI7500, LiCOR.
+The eddy covariance system includes a sonic anemometer (CSAT3, Campbell Scientific, Logan, UT) and an open-path infrared CO~2~/H~2~O gas analyzer (LI7500, LiCOR.
 Lincoln, NE).
@@
 We converted these GWC measurements to VWC assuming a bulk density of 0.8 g cm^-3^.
-$\theta_{0.15 m}$, $\theta_{0.4 m}$ and $\theta_{1 m}$ refer to the plot-wide VWC averages by depth over each sampling period.
+$\theta_{\textrm{0.15 m}}$, $\theta_{\textrm{0.4 m}}$ and $\theta_{\textrm{1 m}}$ refer to the plot-wide VWC averages by depth over each sampling period.
 Middle date of a sampling period is used for comparison with model simulations.
@@
 Values for some of the species reported here were previously reported by @Dickman.2019 and @Wu.2020.
-The entire dataset and R script for the analyses are archived in the NGEE-Tropics archive (Wolfe 2020, To be done).
+The entire dataset and *R* script for the analyses are archived in the NGEE-Tropics archive (Wolfe 2020, To be done).
 
@@
 The branch segments were enclosed in opaque plastic bags, and brought to the laboratory for measurements.
-Stem segments used for hydraulic conductivity (Kh) measurements were located at least 1 m from the cut end of the branch to minimize the number of open vessels that extended into the segment.
+Stem segments used for hydraulic conductivity ($K_h$) measurements were located at least 1 m from the cut end of the branch to minimize the number of open vessels that extended into the segment.
 Maximum vessel length was measured on a subset of samples from each species.
-Among species, maximum vessel length ranged XXX and length of stem between the cut that was made in the air to collect the branch and the section of stem used for Kh measurement was XXX (Table SXX).
+Among species, maximum vessel length ranged XXX and length of stem between the cut that was made in the air to collect the branch and the section of stem used for $K_h$ measurement was XXX (Table SXX).
 
-\   Three leaves that were distal to the Kh segment were sealed in plastic bags and the entire branch was enclosed in black plastic bags for at least 30 minutes to stop transpiration.
-Just prior to measurement of Kh, leaf water potential of the three bagged leaves was measured with a pressure chamber.
+\   Three leaves that were distal to the $K_h$ segment were sealed in plastic bags and the entire branch was enclosed in black plastic bags for at least 30 minutes to stop transpiration.
+Just prior to measurement of $K_h$, leaf water potential of the three bagged leaves was measured with a pressure chamber.
 Since transpiration was stopped by bagging the leaves, leaf water potential was assumed to equal stem water potential (REFERENCE).
 
-\   For each Kh segment, three measurements of stem water potential were made on stem sections that were located adjacent to the Kh segment.
+\   For each $K_h$ segment, three measurements of stem water potential were made on stem sections that were located adjacent to the $K_h$ segment.
 Stem sections that were approximately 40 mm in length were placed in nickel-plated aluminum chambers that were connected to Peltier-type psychrometers (Merrill Instruments, Logan, UT) and placed in a water bath at 30°C for 3 h to equilibrate the humidity within the cambers.
@@
 
-\   The Kh segments were removed from the branches by repeatedly cutting back the branch on both ends while it was submerged in water to relax the hydraulic tension within the xylem ((REFERENCE)).
+\   The $K_h$ segments were removed from the branches by repeatedly cutting back the branch on both ends while it was submerged in water to relax the hydraulic tension within the xylem ((REFERENCE)).
 Then the ends were shaved clean with a fresh razor.
@@
 Following Torres-Ruiz et al.
-(2012), Kh was calculated as the slope of the flow rate versus pressure gradient (i.e., pressure head divided by segment length) among the three pressure heads.
-Stem area specific hydraulic conductivity was (Ks) was calculated by dividing Kh by the cross-sectional area of stem.
+(2012), $K_h$ was calculated as the slope of the flow rate versus pressure gradient (i.e., pressure head divided by segment length) among the three pressure heads.
+Stem area specific hydraulic conductivity was ($K_stem$) was calculated by dividing $K_h$ by the cross-sectional area of stem.
 To calculate cross-sectional stem area, stem diameter was measured twice perpendicularly on both ends of the segment and averaged, then area was calculated assuming a circular cross section.
 
-\   To assess vulnerability to cavitation, the branches were allowed to dry in the lab for 0 – 180 hours before they were measured for stem water potential and Ks.
+\   To assess vulnerability to cavitation, the branches were allowed to dry in the lab for 0 – 180 hours before they were measured for stem water potential and $K_stem$.
 We used the water potential measured by the psychrometers because they are a more direct measurement.
@@
 For some species, we noted this pattern while for others it was not apparent (Fig SXX).
-For each species, Ks was plotted against stem water potential and a Weibull curve was fit through the 90th  percentile of the points, following Wolfe et al.
+For each species, $K_stem$ was plotted against stem water potential and a Weibull curve was fit through the 90th  percentile of the points, following Wolfe et al.
 (2016)).
-The intercept of the curve was interpreted as maximum Ks and the water potential where the Weibull curve predicted 12, 50, 80, and 88 loss of maximum Ks were interpreted as P12, P50, P80, and P88, respectively.
+The intercept of the curve was interpreted as maximum $K_stem$ and the water potential where the Weibull curve predicted 12, 50, 80, and 88 loss of maximum $K_stem$ were interpreted as $\\mathrm{\\Psi}$~12,stem~, $\\mathrm{\\Psi}$~50,stem~, $\\mathrm{\\Psi}$~80,stem~, and $\\mathrm{\\Psi}$~88,stem~, respectively.
 
@@
 \@ref(eq:b-mod)-\@ref(eq:a-mod)), two lacked data on WSG and five on LMA~lamina~.
-We filled three gaps in LMA using predictions from a strong linear relationship between LMA and LMA~disc~, the latter defined as the mean leaf mass per unit area measured for a 1.483 cm^2^ leaf disc taken to avoid veins (g m^-2^) for leaves receiving direct sunlight ($\overline{LMA} = `r disc.cf[1]` + `r disc.cf[2]`LMA_{disc}$; Adj.
+We filled three gaps in LMA using predictions from a strong linear relationship between LMA and LMA~disc~, the latter defined as the mean leaf mass per unit area measured for a 1.483 cm^2^ leaf disc taken to avoid veins (g m^-2^) for leaves receiving direct sunlight ($\overline{\textrm{LMA}} = `r disc.cf[1]` + `r disc.cf[2]`\textrm{LMA}_{disc}$; Adj.
 $R$^2^ = `r disc.r2`; $p$ `r disc.p`, $n$ = `r disc.n`).
-We filled the two remaining gaps in LMA using predictions from a linear relationship between LMA and LMA~lamina~, as measured during the smaller, $K_\textrm{leaf}$ campaign ($\overline{LMA} = `r lma.cf[1]` + `r lma.cf[2]`LMA_{lamina}$; Adj.
+We filled the two remaining gaps in LMA using predictions from a linear relationship between LMA and LMA~lamina~, as measured during the smaller, $K_\textrm{leaf}$ campaign ($\overline{\textrm{LMA}} = `r lma.cf[1]` + `r lma.cf[2]`\textrm{LMA}_{lamina}$; Adj.
 $R$^2^ = `r lma.r2`; $p$ `r lma.p`, $n$ = `r lma.n`).

### 2020-10-25 manuscript.rmd
To size in main-text 1. Reduced description of growth data in main text, moved to SI 2. moved VPD model description moved to SI 3. cut down on isope data description.

@@
 T: +1(505) 665 1647
+\
+\
+
+Section  | Word Count
+------------- | -------------
+Summary | 197
+Main text  | 7412
+Introduction | 892
+Methods | 3567
+Results | 871
+Discussion | 2082
+Key-words | 6 out of 8
+
+\
+\
+\
+
+Item | Count
+------------- | -------------
+Figures (all in color)  | 8
+Tables | 2
+Supplementary Material |  1 file
 
@@
 \
-Keywords: below-ground traits, deep-water access, drought-induced mortality, drought tolerance, rooting depths, tropical forest, life-history strategies
+Keywords: hydraulic vulnerability, deep-water access, drought-induced mortality, drought tolerance, rooting depths, tropical forest, life-history strategies, safety margins
 
@@
 Rhizosphere soil water potential ($\mathrm{\Psi}$~soil~) down-regulates plant tissue water potentials [@Christoffersen:2016kd2d].
-In moist forests $\mathrm{\Psi}$~leaf~ equilibrates with $\mathrm{\Psi}$~soil~ at night and declines through the day as plants transpire.
-$\mathrm{\Psi}$~leaf~ in turn controls leaf conductance ($K_\textrm{leaf}$), which in turn controls stomatal conductance and photosynthetic rates [@Sack.2012].
-We therefore used maximum diurnal leaf conductance ($K_\textrm{leaf}$) as determined by $\mathrm{\Psi}$~soil~ as a proxy for the soil water limitation on growth.
-In our ERD models $K_\textrm{leaf}$ therefore incorporates the effects of rooting depth and rhizosphere water potential.
-This approach does not specifically incorporate the buffering effect of stem water storage on $K_\textrm{leaf}$, which may cause the model to slightly overestimate ERD (see Discussion).
-VPD has a positive impact on transpiration and thus photosynthesis and growth upto a VPD threshold, after which VPD has a negative impact [@Grossiord.2020].
-The fraction of a tree's leaf cover present at a given time further limits it's growth potential.
-We therefore regressed species-specific growth in a census interval against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1.
+In moist forests $\mathrm{\Psi}$~leaf~ equilibrates with $\mathrm{\Psi}$~soil~ at night declining through the day as plants transpire.
+$\mathrm{\Psi}$~leaf~ in turn controls leaf conductance ($K_\textrm{leaf}$), which can in turn control transpiration and photosynthetic rates [@Sack.2012].
+We therefore used maximum diurnal ($K_\textrm{leaf}$) as determined by $\mathrm{\Psi}$~soil~ as a proxy for the soil water limitation on growth, and thus $K_\textrm{leaf}$ in our models incorporates the effects of rooting depth and rhizosphere water potential.
+This does not specifically account for stem water storage capacitance, which may cause the model to slightly overestimate ERD (see Discussion).
+Transpiration increases with VPD upto a VPD threshold, after which VPD has a negative impact as leaf pores (stomata) close, reducing photosynthesis and growth  [@Grossiord.2020].
+Apart from stomatal control leaf-drop may further limit water-uptake and transpiration.
+We therefore regressed species-specific growth rates from 5-yrly censuses against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1.
 Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting.
@@
 \
-\    To account for the non-linear impact of VPD on growth, we used a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see Model parametarisation) and used GPP standardised between 0 and 1 (hereafter, VPD*) in Eq.
+\    To account for the non-linear impact of VPD on growth, we used a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see SI) and used GPP standardised between 0 and 1 (hereafter, VPD*) in Eq.
 \@ref(eq:M-chosen).
@@
 For growth estimates, to minimize the effect of light variation among trees, we selected only large trees ($\ge$ 30 cm dbh, the diameter at breast height) and also species whose maximum height was $\ge$ 30 m (hereafter, canopy species) and thus are likely to be fully exposed to the sun.
-We selected trees for which the height of the diameter measurements were unchanged across all censuses (1990, 1995, 2000, 2005, 2010, 2015).
-For each individual tree ($i$), we calculated absolute growth rates ($G_{i,t}$, cm yr^-1^) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses prorated by the inter-census interval based on the exact date of measurements.
-We eliminated positive outliers as any growth rates amounting to >75 mm yr^-1^, as this was the highest rate observed with confidence for fastest growing species [@Condit.2017].
-Negative outliers were removed based on known measurement error, $\sigma_{e} = 0.006214d + 0.903$, where *d* is dbh.
-Any stem whose later dbh measurement fell <4$\sigma_{e}$ the earlier measurement was removed [@Condit.2017].
-As tree growth rates vary with size [@Muller-Landau:2006], to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($\textrm{Resid}_{i,t} = G_{i,t} - f(d_{i,t})$).
-The diameter model of growth, $f(d_{i, t})$, a B-spline based polynomial equation with five degrees of freedom, was fitted to growth rates from all trees pooled together as there was insufficient species-specific data for large trees (Fig.
-S\@ref(fig:dbh-model)).
-We retained only those trees that had $\textrm{Resid}_{i,t}$ estimates for all census intervals (total trees = `r sum(g.n$n)`) and standardized (centered and scaled) individual tree time series of $\textrm{Resid}_{i,t}$.
-We then obtained `r length(erd.sp)` species-level time-series of medians of $\textrm{Resid}_{i,t}$ ($G_{t}$).
-These were obtained for only those species with complete records on at least three trees (median $n$ = `r median(g.n$n)`, max $n$ = `r max(g.n$n)` trees per species).
+We calculated individual tree diameter growth rates (cm yr^-1^) across six censuses (1990, 1995, 2000, 2005, 2010, 2015), as the difference between tree dbh measured prorated by exact date of measurements.
+After removing growth outliers due to measurement error, to account for temporal changes in growth rates due to increasing tree size, we obtained residuals from a dbh model of tree growth.
+(see SI for details).
+Retaining trees with a complete record in all censuses (total trees = `r sum(g.n$n)`) and centering and scaling residuals of each tree, we obtained species-level median time series of residuals ($G_{t}$) used in the ERD model, for only those species ($n$ = `r length(erd.sp)`) with complete records on at least three trees (median $n$ = `r median(g.n$n)`, max $n$ = `r max(g.n$n)` trees per species).
 
@@
 
-In the exploratory model for effective rooting depth, we explored the effect of seasonality in leaf area index (LAI, unitless) on growth.
+In some of the alternative models for effective rooting depth, we explored the effect of seasonality in leaf area index (LAI, unitless) on growth.
 We assumed species-specific mean seasonal curves for LAI normalised between 0 and 1 (LAI, unitless), informed by a combination of long-term records for leaf-fall [@Wright.1990aeo] and leaf-lifetime [@Osnas.2018] (see SI).
 
-#### Vapor Pressure Deficit (VPD)
-
-We fitted a polynomial curve between monthly average of mean midday values (11:00 to 15:00 hrs) of gross primary productivity (gC m^2^ day^-1^) we collected from the eddy flux tower located 75 m from the 50-ha plot (2012-2017) and VPD (kPa) calculated using the climate forcing data (for any day $i$, $\textrm{GPP}_i = `r vpd.cf[1]` + `r vpd.cf[2]`\textrm{VPD}_i + `r vpd.cf[3]`\textrm{VPD}_i^2 + error_{i}$; Fig.
-S\@ref(fig:gpp-vpd)); used VPD to predict GPP from 1990-2015, standardized it between 0 and 1, and used it as $\textrm{VPD}^*$ in Eq.
-\@ref(eq:M-chosen)).
-
 ### Model validation and selection
-
-Xylem sap $\mathrm{\delta}$^2^H values (hereafter, $\mathrm{\delta}$^2^H~xylem~) can provide an index of rooting depth, which we used to validate the ERD models as an independent observation of ERD.
-We used species $\mathrm{\delta}$^2^H~xylem~ values from BCI given in @Meinzer:2001 for the dry-season of March 1997 as this period showed largest seasonal divergence among species, with a corresponding vertical gradient of $\mathrm{\delta}$^2^H in the soil (hereafter, $\mathrm{\delta}$^2^H~soil~ ).
-(Fig.
+\
+As root water-uptake is generally a non-fractionating process, 2^H/1^H content of the tree xylem sap (hereafter, $\mathrm{\delta}$^2^H~xylem~) reflects the signature of the source water.
+Given a vertical gradient in $\mathrm{\delta}$^2^H water by depth, $\mathrm{\delta}$^2^H~xylem~ can be used as an index of rooting depth [@Dawson.1991].
+We thus used $\mathrm{\delta}$^2^H~xylem~ as an independent observation to validate the ERD models.
+We leveraged $\mathrm{\delta}$^2^H~xylem~ values from BCI given in @Meinzer:2001 for the dry-season of March 1997 as this period showed largest seasonal divergence in $\mathrm{\delta}$^2^H~xylem~ among species, in soil and ground water $\mathrm{\delta}$^2^H at natural abundance level (Fig.
 S\@ref(fig:iso-data)).
-Briefly, this isotopic approach makes use of a naturally formed gradient in deuterium to hydrogen ratio of soil water vertically belowground, which is generally formed due to evaporative enrichment (here, during the dry season).
-Shallow soil water has a greater deuterium to hydrogen content than deep soil water (here, $\mathrm{\delta}$^2^H~soil~ ranged from -16.5 to -50 ‰ over the depth of 0.5-1 m), with ground-water having the lowest deuterium to hydrogen content (here, -60 ‰), typically reflecting un-fractionated rain water.
-As root water-uptake is generally a non-fractionating process, $\mathrm{\delta}$^2^H~xylem~ of a tree reflects the signature of the source water.
-Thus water-sourcing depth of the tree can be identified as the depth with the same value of $\mathrm{\delta}$^2^H~soil~, or as that of $\mathrm{\delta}$^2^H~xylem~, with the uncertainty that the value of $\mathrm{\delta}$^2^H~xylem~ may also result from a mixing of water from shallower or deeper depths [@Dawson.1991].
-Analyses of $\mathrm{\delta}$^2^H~xylem~ of the 12 species studied by @Meinzer:2001 showed that four species sourced water between 1 m and ground water.
-Two other had water access at least 0.8 m deep, one was restricted to water access between 0 to 0.2 m and five fell in the range over which the relationship between soil depth and $\mathrm{\delta}$^2^H~soil~ was variable in the 0.2 to 0.7 m depth range.
-
-\   We compared six species with data for $\mathrm{\delta}$^2^H~xylem~ as well as modelled estimates of ERD, after removing those species in @Meinzer:2001 that are likely to have no leaves in March 1997 because xylem sap $\mathrm{\delta}$^2^H of these species may not be linked with water sourced at the time of measurement and thus may not be comparable with other species that had leaves.
-We also removed *Guapira standleyana*, whose xylem sap isotope value was not well constrained by the soil water isotope depth profile.
+Analyses of $\mathrm{\delta}$^2^H~xylem~ of the 12 species studied by @Meinzer:2001 showed that four species sourced water between 1 m and ground water, two other at least 0.8 m deep, one between 0 to 0.2 m, and five in the 0.2 to 0.7 m depth range over which the relationship between soil depth and $\mathrm{\delta}$^2^H~soil~ was variable.
+\
+\   We removed those species that would be leafless in March 1997 because xylem sap $\mathrm{\delta}$^2^H for these species may not be linked with water sourced at the time of measurement and thus may not be comparable with other species that had leaves.
+We also removed one species whose $\mathrm{\delta}$^2^H~xylem~ was in the uncertain depth range.
 For each model of ERD (Eq.
 \@ref(eq:M-chosen) and Eqs.
-in SI), estimated species ERD was regressed against species $\mathrm{\delta}$^2^H~xylem~ (‰).
+in SI), estimated species ERD was regressed against species $\mathrm{\delta}$^2^H~xylem~, for a maximum of six species.
 
@@
 
-To test whether ERD plays a role in alleviating mortality risk, we calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$30 m) and average abundance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; $n$ = `r length(erd.mrate.sp)`).
+To test whether ERD plays a role in alleviating mortality risk, we calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$ 30 m) and average abundance of at least 20 trees ($n$ = `r length(erd.mrate.sp)`).
 For each of the seven census intervals $t$ in the 35-yr record (1981-82, 1985, 1990, 2000, 2005, 2010, 2015, 2015), $\textrm{Mort}_{t}$ (% yr^-1^) for species $s$ for two successive censuses, $c_{1}$ and $c_{2}$, was calculated as $\textrm{Mort}_{s,t} = \frac{D_{s,c_{1}}}{N_{s,c_{2}}}\times\frac{100}{\textrm{d}}$, where $N_{s,c_{1}}$ and $D_{s,c_{2}}$ are the total number of trees of a species $s$ present in $c_{1}$ and dead in $c_{2}$, respectively, in the 50 ha plot and $d$ is the duration based on mean dates of $c_{1}$ and $c_{2}$.
@@
 \
-\   We regressed species ERD against species ($\textrm{Mort}_{s,t}$ for each of the census interval $t$).
+\   We regressed species ERD against species ($\textrm{Mort}_{s,t}$ for each of the census interval $t$) (with and without deciduous species).
 To assess whether deeper ERD species escaped the experience of significant soil drought, we calculated for each species the proportion of days in the simulation period (1990-2015) during which soil water potential in the soil zone matching species ERD ($\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$) were below species $\mathrm{\Psi}$~crit~.
@@
 S\@ref(fig:other-erd-mods)).
-We interpret this result as that VPD and PLC curves may be adequate to explain growth, but also acknowledge that our estimates of seasonality of LAI is a tentative estimate combining leaf fall data and the timing of  leaf-gain backtracked from leaf lifetime, and omits inter-annual variation.
+We interpret this result as that VPD and PLC curves may be adequate to explain inter-census differences in growth, but also acknowledge that our estimates of seasonality of LAI is a tentative estimate combining leaf fall data and the timing of leaf-gain backtracked from leaf lifetime, and omits inter-annual variation.
 Future studies could further develop models of LAI seasonality and make use of data from new technologies such as drone based monitoring of LAI in species-rich tropical forests [@Park.2019].
@@
 $R$^2^ and $p$-values for linear model fits are given in panel insets.
-Model fits with SE (gray area) are shown for significant fits at $\alpha$ = 0.1.
+Model fits (blue lines) with SE (gray area) are shown for significant fits at $\alpha$ = 0.1.
 
@@
 
+\newpage
 ```{r hypo, results = 'asis', echo = FALSE}
@@
 
+\newpage
 ```{r symbols, results = 'asis', echo = FALSE}
@@
 ## Data Descriptions
+
+### Growth Data
+\
+We selected trees for which the height of the diameter measurements were unchanged across all censuses (1990, 1995, 2000, 2005, 2010, 2015).
+For each individual tree ($i$), we calculated absolute growth rates ($G_{i,t}$, cm yr^-1^) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses prorated by the inter-census interval based on the exact date of measurements.
+We eliminated positive outliers as any growth rates amounting to >75 mm yr^-1^, as this was the highest rate observed with confidence for fastest growing species [@Condit.2017].
+Negative outliers were removed based on known measurement error, $\sigma_{e} = 0.006214d + 0.903$, where *d* is dbh.
+Any stem whose later dbh measurement fell <4$\sigma_{e}$ the earlier measurement was removed [@Condit.2017].
+As tree growth rates vary with size [@Muller-Landau:2006], to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($\textrm{Resid}_{i,t} = G_{i,t} - f(d_{i,t})$).
+The diameter model of growth, $f(d_{i, t})$, a B-spline based polynomial equation with five degrees of freedom, was fitted to growth rates from all trees pooled together as there was insufficient species-specific data for large trees (Fig.
+S\@ref(fig:dbh-model)).
+We retained only those trees that had $\textrm{Resid}_{i,t}$ estimates for all census intervals (total trees = `r sum(g.n$n)`) and standardized (centered and scaled) individual tree time series of $\textrm{Resid}_{i,t}$.
+We then obtained `r length(erd.sp)` species-level time-series of medians of $\textrm{Resid}_{i,t}$ ($G_{t}$).
+These were obtained for only those species with complete records on at least three trees (median $n$ = `r median(g.n$n)`, max $n$ = `r max(g.n$n)` trees per species).
 
@@
 $R$^2^ and $p$-values for linear model fits are given in panel insets.
-Model fits with SE (gray area) are shown for significant fits at $\\alpha$ = 0.1."}
+Model fits (blue lines) with SE (gray area) are shown for significant fits at $\\alpha$ = 0.1."}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_by rdi.gr.jpeg"))

### 2020-10-25 old_ms_sections.Rmd
1. moved extra isotopic data description

@@
 
+Briefly, this isotopic approach makes use of a naturally formed gradient in deuterium to hydrogen ratio of soil water vertically belowground, which is generally formed due to evaporative enrichment (here, during the dry season).
+Shallow soil water has a greater deuterium to hydrogen content than deep soil water (here, $\mathrm{\delta}$^2^H~soil~ ranged from -16.5 to -50 ‰ over the depth of 0.5-1 m), with ground-water having the lowest deuterium to hydrogen content (here, -60 ‰), typically reflecting un-fractionated rain water.
+As root water-uptake is generally a non-fractionating process, $\mathrm{\delta}$^2^H~xylem~ of a tree reflects the signature of the source water.
+Thus water-sourcing depth of the tree can be identified as the depth with the same value of $\mathrm{\delta}$^2^H~soil~, or as that of $\mathrm{\delta}$^2^H~xylem~, with the uncertainty that the value of $\mathrm{\delta}$^2^H~xylem~ may also result from a mixing of water from shallower or deeper depths .
+
 ## Results

### 2020-10-25 manuscript.rmd
Corrected spellings

@@
 Summary | 197
-Main text  | 7412
+Main Text  | 6640
 Introduction | 892
-Methods | 3567
+Methods | 3455
 Results | 871
-Discussion | 2082
-Key-words | 6 out of 8
+Discussion | 1422
+Keywords | 6
 
-\
-\
 \
@@
 ------------- | -------------
-Figures (all in color)  | 8
+Figures in color | 8
 Tables | 2
@@
 - Among the mechanisms that plants employ to tolerate drought, deep water access is arguably the most commonly invoked, but under-studied.
-Vulnerability to embolism and hydraulic safety margins can predict mortality risk at given levels of dehydration, but deep-water access may delay plants from dehydration.
+Vulnerability to embolism and hydraulic safety margins can predict mortality risk at given levels of dehydration, but deep-water access may delay plant dehydration.
 Here, we tested the role of deep-water access in enabling survival within a diverse tropical forest community (Panama) using a novel data-model approach.
@@
 Tropical forests play a disproportionately large role in the global carbon and energy cycles [@Bonan:2008], and support half of global biodiversity [@Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019].
-Tropical forests are logically considered to be an especially drought-vulnerable among biomes given the combination of climate risk and vegetation sensitivity [@Meir:2015trw], yet mortality events in the tropics are rarely as large as those in the temperate and boreal zones (@McDowell.2018), leading to questions regarding the role of large trait-diversity and hydraulic strategies in mitigating mortality events.
+Tropical forests are logically considered to be an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw], yet mortality events in the tropics are rarely as large as those in the temperate and boreal zones (@McDowell.2018), leading to questions regarding the role of large trait-diversity and hydraulic strategies in mitigating mortality events.
 Despite this, widespread drought-induced tree mortality has occurred across the tropics for specific functional groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010] and tropical mortality rates are increasing in some regions [@Brienen.2015; @Hubau.2020].
@@
 Significant efforts have been invested into identifying and linking universal drought indices, above-ground traits, and demographic rates.
-These efforts have resulted in identification of correlations between aboveground hydraulic traits such as vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013, @Venturas.2016].
+These efforts have resulted in identification of correlations between above-ground hydraulic traits such as vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013, @Venturas.2016].
 Deep roots may mitigate hydraulic vulnerability and mortality-risk, yet plants' realized water environments, specifically plant's ability to acquire water via differing rooting depths in hydrological rather than meteorological droughts, are rarely studied, especially in the tropical region.
@@
 \@ref(fig:schematic)).
-We defined species ERD as the depth at which soil water potential best explained species-specific 5-yrly mean growth rates over 25 years.
+We defined species ERD as the depth at which soil water potential best explained species-specific 5-yearly mean growth rates over 25 years.
 To obtain the daily dynamics of soil water potential in the whole-soil column we locally parameterized and calibrated a 1-D hydrological water balance model.
-The ERD model incorporated the impact of atmospheric drought on growth and was constrained by above-ground hydraulic traits, namely, species-specific leaf vulnerability curves, by using existing data for eight species and developing trait-based proxies for the rest.
-We evaluated ERD against existing dueterium to hydrogen content data for a subset of six tree species as an independent observation of ERD.
+The ERD model incorporated the impact of atmospheric drought on growth and was constrained with species-specific leaf vulnerability curves, by using existing data for eight species and developing trait-based proxies for the rest.
+We evaluated ERD against existing deuterium to hydrogen content data for a subset of six tree species as an independent observation of ERD.
 We tested relationship of ERD with above-ground hydraulic traits, namely, xylem vulnerability to embolism, branch hydraulic safety margins, leaf turgor loss point and maximum stem-specific hydraulic conductivity, for a subset of eight species with both datasets.
@@
 Rainfall at BCI is seasonal with a mean annual total of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a pronounced dry season from mid-December through April with a mean total of 292 mm [@Paton.2020].
-Among the tree *species* found in the forest canopy, 28% are identified to be at least partly deciduous during the dry season, but this amounts to a maximum of 6.3% *trees* in the canopy that are deciduous, and the percentage of total forest crown area that is deciduous is estimated at 9.7% [@Condit:2000].
+Among the tree *species* found in the forest canopy, 28% are identified to be at least partly deciduous during the dry season, but at a maximum only 6.3% *trees* in the canopy, and only 9.7% of the total forest crown area, is deciduous [@Condit:2000].
 
@@
 \@ref(eq:water-balance), namely, (i) evapotranspiration (ET) from the eddy flux tower (2012-2017; see [SI](#SIData.matteo)), comparable to $E_{v} + E_{g}$ in Eq.
-\@ref(eq:water-balance)), (ii) stream discharge $q_\textrm{run}$ from the Conrad catchment, comparabel to $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC or $\theta_{z}$) comparable to $W_{\textrm{liq,} z}$ from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we obtained with a vertical TDR probe over 0-15 cm depth ($\theta_\textrm{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m with 1299 samples in total covering all soil types and habitats (see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter,  $\theta_{\textrm{0-0.15 m}}$, $\theta_{\textrm{0.4 m}}$ and $\theta_{\textrm{1 m}}$ refer to plot-wide averages).
+\@ref(eq:water-balance)), (ii) stream discharge $q_\textrm{run}$ from the Conrad catchment, comparable to $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC or $\theta_{z}$) comparable to $W_{\textrm{liq,} z}$ from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we obtained with a vertical TDR probe over 0-15 cm depth ($\theta_\textrm{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m with 1299 samples in total covering all soil types and habitats (see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter,  $\theta_{\textrm{0-0.15 m}}$, $\theta_{\textrm{0.4 m}}$ and $\theta_{\textrm{1 m}}$ refer to plot-wide averages).
 
@@
 $W_{\textrm{liq,}\overline{z}}$ is the average of daily $W_{\textrm{liq,} z}$, where $z$ $\in$ $Z$, and $Z$ = {`r paste(soil.depths[1:4], collapse=", ")`} m.
-Superscript * indicates that RMSE is standardised between 0 and 1 across the 5000 simulations.
+Superscript * indicates that RMSE is standardized between 0 and 1 across the 5000 simulations.
 
-\   Preliminary analyses showed that minimising the objective function $\overline{O}$, that is, simultaneously fitting the model to all states and fluxes did not capture the soil moisture dynamics at 1 m depth well, our key priority for model calibration.
-To ensure model fit at 1 m depth, we first chose 300 parameter ensemble members that minimised RMSE between observations and simulations for VWC at the depth of 1 m ($\textrm{RMSE}^{*}(\theta_{\textrm{1 m}}, W_{\textrm{liq,`r soil.depths[8]` m}})$ < 0.2), and among those 300 ensemble members we chose 100 members with the lowest values for the objective function $\overline{O}$.
+\   Preliminary analyses showed that minimizing the objective function $\overline{O}$, that is, simultaneously fitting the model to all states and fluxes did not capture the soil moisture dynamics at 1 m depth well, our key priority for model calibration.
+To ensure model fit at 1 m depth, we first chose 300 parameter ensemble members that minimized RMSE between observations and simulations for VWC at the depth of 1 m ($\textrm{RMSE}^{*}(\theta_{\textrm{1 m}}, W_{\textrm{liq,`r soil.depths[8]` m}})$ < 0.2), and among those 300 ensemble members we chose 100 members with the lowest values for the objective function $\overline{O}$.
 See Table \@ref(tab:param) for the ranges of best-fit values for different parameters.
@@
 
-We evaluated ELM-FATES by calculating RMSE between simulated and observed long-term daily VWC for the depths of 0.1, 0.4 and 1 m (2016-2018), baesd on a dataset we had left-out during calibration.
+We evaluated ELM-FATES by calculating RMSE between simulated and observed long-term daily VWC for the depths of 0.1, 0.4 and 1 m (2016-2018), based on a dataset we had left-out during calibration.
 We obtained these observations from three horizontal time domain reflectometry probes at the depths of 0.1, 0.4 and 1 m at a point location near the eddy flux tower (see [SI](#SIData.VWC)).
@@
 
-To estimate effective rooting depth, we tested alternate models for growth (Eq.
+To estimate effective rooting depth, we tested alternate empirical models for species 5-yearly census growth (Eq.
 \@ref(eq:M-chosen) and those in SI), which use effective rooting depth as an input.
-There are many intrinsic and extrinsic factors that limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019].
-Rhizosphere soil water potential ($\mathrm{\Psi}$~soil~) down-regulates plant tissue water potentials [@Christoffersen:2016kd2d].
-In moist forests $\mathrm{\Psi}$~leaf~ equilibrates with $\mathrm{\Psi}$~soil~ at night declining through the day as plants transpire.
-$\mathrm{\Psi}$~leaf~ in turn controls leaf conductance ($K_\textrm{leaf}$), which can in turn control transpiration and photosynthetic rates [@Sack.2012].
-We therefore used maximum diurnal ($K_\textrm{leaf}$) as determined by $\mathrm{\Psi}$~soil~ as a proxy for the soil water limitation on growth, and thus $K_\textrm{leaf}$ in our models incorporates the effects of rooting depth and rhizosphere water potential.
-This does not specifically account for stem water storage capacitance, which may cause the model to slightly overestimate ERD (see Discussion).
-Transpiration increases with VPD upto a VPD threshold, after which VPD has a negative impact as leaf pores (stomata) close, reducing photosynthesis and growth  [@Grossiord.2020].
-Apart from stomatal control leaf-drop may further limit water-uptake and transpiration.
-We therefore regressed species-specific growth rates from 5-yrly censuses against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1.
+Many factors limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019].
+Because pre-dawn $\mathrm{\Psi}$~leaf~ approaches $\mathrm{\Psi}$~soil~, we used the latter to predict maximum diurnal ($K_\textrm{leaf}$), which in turn controls growth.
+$K_\textrm{leaf}$ in our models thus incorporates the effects of rooting depth and rhizosphere water potential.
+This approach, however, does not account for the buffering effect of stem water storage capacitance on $\mathrm{\Psi}$~leaf~, and may thus overestimation ERD (see Discussion).
+VPD affects growth non-linearly, with growth increasing with VPD up-to a VPD threshold, then decreasing as leaf pores (stomata) close [@Grossiord.2020].
+Apart from stomatal control, leaf-drop may further limit water-uptake and growth.
+We therefore regressed species-specific growth against multiplicative or additive effects of one or more factors calculated daily and averaged over 5-yearly censuses: VPD, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1.
 Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting.
@@
 \
-\    For the chosen model (see Model validation and selection below), daily average growth for species $s$, $G_{s,t|z}$, in the census interval $t$ for depth $z$ and for a given hydrological realisation $h$ is described as follows:
+\    For the chosen model (see Model validation and selection below), daily average growth for species $s$, $G_{s,t|z}$, in the census interval $t$ for depth $z$ and for a given hydrological realization $h$ is described as follows:
 
@@
 \
-\    To account for the non-linear impact of VPD on growth, we used a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see SI) and used GPP standardised between 0 and 1 (hereafter, VPD*) in Eq.
+\    To account for the non-linear impact of VPD on growth, we used a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see SI) and used GPP standardized between 0 and 1 (hereafter, VPD*) in Eq.
 \@ref(eq:M-chosen).
@@
 \end{equation}
-\    By varying $z$, we estimated species effective rooting depths (ERD), defined as the depth $z$ at which soil water dynamics ($\mathrm{\Psi}_{\textrm{soil,} z}$) best explained observed dynamics of species growth (six 5-yrly censuses, 1990-2015).
+\    By varying $z$, we estimated species effective rooting depths (ERD), defined as the depth $z$ at which soil water dynamics ($\mathrm{\Psi}_{\textrm{soil,} z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015).
 Using Eq.
@@
 \end{equation}
-where $L$ is the likelihood, $H$ = {1, .., 100}, refer to the 100 hydrological realisations and $Z$ = {0.4, `r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m.
+where $L$ is the likelihood, $H$ = {1, .., 100}, refer to the 100 hydrological realizations and $Z$ = {0.4, `r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m.
 We removed finely resolved soil layers below 0.21 m, averaged $\mathrm{\Psi}_{\textrm{soil},z}$ for depths 0.21, 0.37 and 0.62 at 0.4 m.
@@
 For growth estimates, to minimize the effect of light variation among trees, we selected only large trees ($\ge$ 30 cm dbh, the diameter at breast height) and also species whose maximum height was $\ge$ 30 m (hereafter, canopy species) and thus are likely to be fully exposed to the sun.
-We calculated individual tree diameter growth rates (cm yr^-1^) across six censuses (1990, 1995, 2000, 2005, 2010, 2015), as the difference between tree dbh measured prorated by exact date of measurements.
-After removing growth outliers due to measurement error, to account for temporal changes in growth rates due to increasing tree size, we obtained residuals from a dbh model of tree growth.
-(see SI for details).
-Retaining trees with a complete record in all censuses (total trees = `r sum(g.n$n)`) and centering and scaling residuals of each tree, we obtained species-level median time series of residuals ($G_{t}$) used in the ERD model, for only those species ($n$ = `r length(erd.sp)`) with complete records on at least three trees (median $n$ = `r median(g.n$n)`, max $n$ = `r max(g.n$n)` trees per species).
+Calculating individual tree growth rates across six censuses (1990-2015), removing outliers, accounting for temporal changes in growth rates due to increasing tree size [see SE], we obtained species-level median time series of growth rates ($G_{t}$, cm yr^-1^) used in the ERD model, for only those species ($n$ = `r length(erd.sp)`) with complete records on at least three trees (median $n$ = `r median(g.n$n)`, max $n$ = `r max(g.n$n)` trees per species).
 
@@
 In some of the alternative models for effective rooting depth, we explored the effect of seasonality in leaf area index (LAI, unitless) on growth.
-We assumed species-specific mean seasonal curves for LAI normalised between 0 and 1 (LAI, unitless), informed by a combination of long-term records for leaf-fall [@Wright.1990aeo] and leaf-lifetime [@Osnas.2018] (see SI).
+We assumed species-specific mean seasonal curves for LAI normalized between 0 and 1 (LAI, unitless), informed by a combination of long-term records for leaf-fall [@Wright.1990aeo] and leaf-lifetime [@Osnas.2018] (see SI).
 
@@
 \   We regressed species ERD against species ($\textrm{Mort}_{s,t}$ for each of the census interval $t$) (with and without deciduous species).
-To assess whether deeper ERD species escaped the experience of significant soil drought, we calculated for each species the proportion of days in the simulation period (1990-2015) during which soil water potential in the soil zone matching species ERD ($\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$) were below species $\mathrm{\Psi}$~crit~.
+To assess whether deeper ERD species escaped the experience of significant soil drought, we calculated for each species the proportion of days in the simulation period (1990-2015) during which soil water potential in the soil layer matching species ERD ($\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$) were below species $\mathrm{\Psi}$~crit~.
 $\mathrm{\Psi}$~crit~ was defined as $\mathrm{\Psi}$~50,leaf~.
@@
 S\@ref(fig:obs-comm-plc-curves)).
-These vulenrability curves were used in the models for effective rooting depth.
+These vulnerability curves were used in the models for effective rooting depth.
 
@@
 
-The chosen model (with a mulicative effect of VPD and not LAI; Eq.
+The chosen model (with a multiplicative effect of VPD and not LAI; Eq.
 \@ref(eq:M-chosen)) explained a large fraction of the variance in $\mathrm{\delta}$^2^H~xylem~ ($R$^2^ = `r lm.erd.iso.r2`; $p$ = `r lm.erd.iso.p`, $n$ = `r length(unique(erd.iso$sp))`; Fig.
 \@ref(fig:erd-iso)).
-Two of the alternative models tested with LAI had a strong relationship with $\mathrm{\delta}$^2^H~xylem~ as well, but one that was based on a smaller set of species ($n$ = 4) as fewer species passed the criteria of sufficient species-level goodness-of-fit with these models; <20 speceis vs.
+Two of the alternative models tested with LAI had a strong relationship with $\mathrm{\delta}$^2^H~xylem~ as well, but one that was based on a smaller set of species ($n$ = 4) as fewer species passed the criteria of sufficient species-level goodness-of-fit with these models; <20 species vs.
 `r length(erd.sp)` species for the chosen model (Fig.
 S\@ref(fig:other-erd-mods)).
-For the rest of the analyses we therefore used ERD estimates from the less parimonious, chosen model (with VPD and not LAI, Eq.
+For the rest of the analyses we therefore used ERD estimates from the less parsimonious, chosen model (with VPD and not LAI, Eq.
 \@ref(eq:M-chosen)).
@@
 
-For the subset of eight species with ERD for which hydraulic traits data were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_\textrm{max,stem}$, Spearmans' $r$ = `r cor.k`, $p$ = `r p.k`; Fig.
-\@ref(fig:erd-hyd)A), less negative leaf $\mathrm{\Psi}$~tlp~ (Spearmans' $r$ = `r cor.tlp`, $p$ = `r p.tlp`; Fig.
-\@ref(fig:erd-hyd)B), less negative $\mathrm{\Psi}$~88,stem~ and thus greater vulnerability to embolism from cavitation (Spearmans' $r$ = `r cor.stem.88`, $p$ = `r p.stem.88`; Fig.
-\@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~; Spearmans' $r$ = `r cor.hsm`, $p$ = `r p.hsm`; Fig.
+For the subset of eight species with ERD for which hydraulic traits data were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_\textrm{max,stem}$, Spearman's' $r$ = `r cor.k`, $p$ = `r p.k`; Fig.
+\@ref(fig:erd-hyd)A), less negative leaf $\mathrm{\Psi}$~tlp~ (Spearman's' $r$ = `r cor.tlp`, $p$ = `r p.tlp`; Fig.
+\@ref(fig:erd-hyd)B), less negative $\mathrm{\Psi}$~88,stem~ and thus greater vulnerability to embolism from cavitation (Spearman's' $r$ = `r cor.stem.88`, $p$ = `r p.stem.88`; Fig.
+\@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~; Spearman's' $r$ = `r cor.hsm`, $p$ = `r p.hsm`; Fig.
 \@ref(fig:erd-hyd)D).
@@
 
-Amongst the evergreen species ERD explained `r m.evg.r2.90`, `r m.evg.r2.05` and `r m.evg.r2.10`% variation in mortality rates during census intervals 1985-1990, 2000-2005 and 2005-2010, respectively, such that species with deeper ERD had significantly lower mortality rates than deeper ERD species ($p$ = `r m.evg.p.90`, `r m.evg.p.05` and `r m.evg.p.10`, respectively; $n$ = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`; Fig.
+Amongst evergreen species ERD explained `r m.evg.r2.90`, `r m.evg.r2.05` and `r m.evg.r2.10`% variation in mortality rates during census intervals 1985-1990, 2000-2005 and 2005-2010, respectively, such that species with deeper ERD had significantly lower mortality rates than deeper ERD species ($p$ = `r m.evg.p.90`, `r m.evg.p.05` and `r m.evg.p.10`, respectively; $n$ = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`; Fig.
 \@ref(fig:erd-mort-census-evg)).
@@
 \@ref(fig:swp-droughts)).
-Soil water availability was not modelled for 1985-1990.
+Soil water availability was not modeled for 1985-1990.
 Analysis of a species' time spent under significant soil drought (proportion of days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) showed that species with shallower ERD spent greater proportion of days below critical hydraulic thresholds (Fig.
@@
 By contrast shallow ERD species had inefficient, but safer, xylem likely optimized for high variability in $\mathrm{\Psi}_{soil}$ in the shallow soil layers, and resulting in higher variance in mortality rates across censuses as $\mathrm{\Psi}_{soil,~z = ERD}$ crossed ERD $\mathrm{\Psi}_{crit}$.
-Our result of deep rooted species being more hydraulically vulnerable is consistent with observations for 12 co-occurring species in a seasonal forest in the Amazon, amongst which species $\mathrm{\delta}$^18^O~xylem~ explained 70% variance in species $\mathrm{\Psi}_\textrm{88,stem}$ [@Brum.2019].
-Our results are also consistent with a few studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016].
-@Johnson:2018fad found that mortality decreased with deep rooting even though deep rooted species were more vulnerable to xylem cavitation (high $\mathrm{\Psi}$~88,stem~) and had narrower safety-margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~).
-@Nardini:2015 found that a species sourcing deep water in a Karst system showed lower hydraulic failure and mortality in an extreme drought despite having high $\mathrm{\Psi}$~88,stem~ and narrower safety margin compared to two species sourcing shallow-water.
+Our results are also consistent with a few studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016; also see @Brum.2019].
 
@@
 \   Our result--of lower mortality in deep rooted trees--contrasts with the finding of @Chitra-Tarak.2018 (hereafter CT2018) in which species identified by their inverse ERD model as effectively deep rooted died *more* in a rare, prolonged drought.
-The hydrological model in that study revealed that the multi-year drought exhausted the deep soil and even bedrock water availablity (also see, @Goulden.2019 and @Ivanov.2012).
+The hydrological model in that study revealed that the multi-year drought exhausted the deep soil and even bedrock water availability (also see, @Goulden.2019 and @Ivanov:2012).
 In contrast, our hydrological modeling at BCI finds that during the study period, $\mathrm{\Psi}$~soil~ below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and droughts, that is above ($\mathrm{\Psi}$~crit~) of the most sensitive tree species (-0.28 MPa).
-The difference in the results shown here are likely attributable to site differences in the precipitation regime.
-In the dry forest site in CT2018 (Mudumalai National Park, India) precipitation is low: 1095 $\pm$ 296 mm (mean $\pm$ SD) compared to `r round(rain.man.stats$rain)` $\pm$ `r round(rain.man.stats$rain.sd)` mm (`r min(clim$Year)`-`r max(clim$Year)`) at BCI, and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (Penman-Monteith PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (Penman-Monteith PET = `r round(mean(clim$pet.PM))` $\pm$ `r round(sd(clim$pet.PM))` mm;`r min(clim$Year)`-`r max(clim$Year)`; this study, @Guo.2016), and thus at BCI deep soil layers are effectively recharged annually (Fig.
+The difference in the results shown here are likely attributable to site differences in the precipitation regime: with mean annual rainfall of 1095 $\pm$ 296 mm (mean $\pm$ SD) in CT2018 compared to `r round(rain.man.stats$rain)` $\pm$ `r round(rain.man.stats$rain.sd)` mm (`r min(clim$Year)`-`r max(clim$Year)`) at BCI, and precipitation to potential evapotranspiration ratio nearly one in CT2019 compared to nearly two in BCI (Penman-Monteith PET = `r round(mean(clim$pet.PM))` $\pm$ `r round(sd(clim$pet.PM))` mm;`r min(clim$Year)`-`r max(clim$Year)`; this study, @Guo.2016).
+Thus at BCI deep soil layers are effectively recharged annually (Fig.
 S\@ref(fig:swp-all-depths); @Chitra-Tarak.2018).
 The kind of droughts that buffered the deep rooted species in our study were repeated annual droughts that extended the dry season into the wet season.
-The variation in mortality explained by ERD in our study is upto 40%, and other causes may explain observed mortality [@McDowell.2018].
+The variation in mortality explained by ERD in our study is up-to 40%, and other causes may explain observed mortality [@McDowell.2018].
 
@@
 Our chosen ERD model is an advance over the only other inverse model of ERD (@Chitra-Tarak.2018; $n$ = 12 species) as we employ a physiologically meaningful representation, fewer parameters and corroboration with tree and soil data for $\mathrm{\delta}$^2^H.
-
-\
-\   We found a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA and WSG) and parameters of leaf vulnerability (or percent loss of conductance, PLC) curves for a few species ($n$ = `r length(sp.exp.param$sp)`), a 'hard' trait.
-This allowed us to scale-up the leaf vulnerability curves at the community scale ($n$ = `r length(ab.table$Species)` species, see Table S\@ref(tab:ab-param) for model predicted parameters).
-WSG alone explained `r round(b.wsg.r2*100,0)`% of the variance in $B$ ($p$ `r  b.wsg.p`).
-We find that leaf conductance declined more rapidly with declining $\mathrm{\Psi}$~leaf~ for species with higher values of WSG and LMA (Fig.
-S\@ref(fig:obs-plc-curves)).
-Whether such scaling relationships between hard and soft traits hold true for other forests, and their wider applicability, remains to be seen as data on PLC curves are limited in tropical forests, but supportive and counter examples both exist [@McCulloh.2011; @Hoffmann.2011].
-
-\
-\   Another important piece of information for understanding effective rooting depth is stable water isotopes of xylem water and below-ground sources, in the vadose zone and water table.
-Although the interpretation of these data is still under research [@Deurwaerder.2020], they are critical for model calibration or validation.
-We lacked data on water table levels which would have been useful to constrain water availability dynamics in the unobserved soil layers and the spatial variability within the plot.
+Stable water isotopes for plant and below-ground sources are critical for calibration or validation of ERD model at community scale, although the interpretation of these data is still under research [@Deurwaerder.2020].
 As in @Chitra-Tarak.2018, we also interpret our ERD estimates as revealing relative differences among species' effective rooting depths rather than in absolute terms.
-
-\
-\   As our ERD model assumes diurnal maximum $\mathrm{\Psi}$~leaf~ as equivalent to $\mathrm{\Psi}$~soil~, ignoring the buffering effect of stem water storage capacitance, this approach may overestimate ERD in species where stem water storage capacitance is important as a drought-avoidance strategy, that is, the group of deciduous species [Borchert.2005jma].
+As our ERD model ignores stem water storage capacitance it may have overestimated ERD especially in species for which stem water storage capacitance is important as a drought-avoidance strategy; for example, the deciduous species [Borchert.2005jma].
 
@@
 We found that LMA and WSG were strong predictors of PLC curves for leaves, but not for stems (not shown).
-This offers the promise of a greater ability to parameterise the 'hard' hydraulic traits with the abundant soft trait data, thus allowing much better representation of forest hydrodynamics.
+This offers the promise of a greater ability to parameterize the 'hard' hydraulic traits with the abundant soft trait data, thus allowing much better representation of forest hydrodynamics.
 The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits, including hydraulic vulnerability, in order to better represent the functional diversity of tropical forests, and their trajectories into the future.
@@
 \   Our inverse ERD model was parameterized on 5-year growth data, with five data points over a 25-yr period, which decouples climate events and demographic outcomes.
-Future studies could better constrain the ERD model with higher resolution growth data such as those from dendrometer bands.
-However, trees' stem diameter change is a sum total of not just growth, the addition of wood, but also changes due to reversible stem dehydration (xylem, phloem and bark), the importance of which increases with temporal resolution [@Chitra-Tarak.2016; @Chitra-Tarak.2015].
-Use of dendrometer band data would thus entail accounting for reversible stem dehydration [@Zweifel:2018], but may also be an avenue to account for stem water storage and temporal dynamics of ERD, a limitation in our current approach.
-Three of our exploratory ERD models included seasonality in leaf area fraction, but we did not select them as they worsened the fit with growth data for many species (Fig.
+Future studies could better constrain the ERD model with high frequency growth data such as those from dendrometer bands.
+At high temporal resolution, however, the role of reversible dehydration in tree diameter change increases [@Chitra-Tarak.2016; @Chitra-Tarak.2015], but it may also be an avenue to include stem water storage  and dynamic rooting depths in ERD models.
+Three of our exploratory ERD models included LAI seasonality, but we did not select them as they worsened the fit with growth data for many species (Fig.
 S\@ref(fig:other-erd-mods)).
-We interpret this result as that VPD and PLC curves may be adequate to explain inter-census differences in growth, but also acknowledge that our estimates of seasonality of LAI is a tentative estimate combining leaf fall data and the timing of leaf-gain backtracked from leaf lifetime, and omits inter-annual variation.
-Future studies could further develop models of LAI seasonality and make use of data from new technologies such as drone based monitoring of LAI in species-rich tropical forests [@Park.2019].
-This would also open the possibility of testing, whether deciduous species are more shallow rooted than evergreen species as generally observed [@Fan:2017].
+We interpret this result as that VPD and leaf hydraulic vulnerability curves may be adequate to explain inter-census differences in growth (via stomatal control), but also acknowledge that our estimates of seasonality of LAI is a tentative estimate combining leaf-fall data and the timing of leaf-gain backtracked from leaf lifetime, and omits inter-annual variation.
+Future studies may further develop models of LAI seasonality and make use of data from new technologies such as drone based monitoring of LAI in species-rich tropical forests [@Park.2019].
 
 \
-\   In forests such as BCI where understory tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes, especially from a few growth data points, may need accounting for variation in trees' light environment over time and among trees of the same size class, an issue we minimized by limiting ERD estimates to canopy trees.
-In sites where water does not limit growth, it may not be possible to parametarize ERD models from growth dynamics, and other avenues such as sap-flux or leaf water-potential dynamics could be explored [@Brum.2017].
+\   In forests such as BCI where understory tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes may need accounting for temporal variation in trees' light environment, an issue we minimized by limiting ERD estimates to canopy trees.
+In sites where growth is not water-limited, ERD models may explore parameterization on sap-flux or leaf water-potential dynamics [@Brum.2017].
 
 \
-\   While our ERD model empirically predicts growth via $\mathrm{\Psi}$~leaf~ from $\mathrm{\Psi}$~soil~ at a specific depth, models that mechanistically account for $\mathrm{\Psi}$ through soil, the entire rooting profile, as well as, $\mathrm{\Psi}$~stem~ and $\mathrm{\Psi}$~leaf~, accounting for water content and accumulation of embolised tissues, and then downstream photosynthetic pathways and non-water limiting processes influencing growth are likely to be more accurate [@Christoffersen:2016kd2d; @Sperry.1998].
-Although data needs for parameterizing such models [@Christoffersen:2016kd2d; @Sperry.1998] could be greater (e.g.
+\   While our ERD model empirically predicts growth via $\mathrm{\Psi}$~leaf~ from $\mathrm{\Psi}$~soil~ at a specific depth, models that mechanistic models that account for plant hydrodynamics and other processes influencing growth are likely to be more accurate [@Christoffersen:2016kd2d; @Sperry.1998].
+Although data needs for parameterizing such models could be greater (e.g.
 retention curves and capacitance for roots, stems and leaves), adding a degree of parameter uncertainty in community-wide application, such models hold greater promise in improving our understanding of plant physiology [e.g.
@@
 
-\
-\ Finally, rooting depths may be parametarized within DGVMs such as ELM-FATES using plant hydrodynamic schemes [@Christoffersen:2016kd2d; @Koven.2020], simultaneously modeling plant water-availability with plant water-extraction.
-In general, including differences in tree rooting depths in ELM-FATES would improve understanding of plant competition for water (and other resources) and more accurately represent plant's drought strategies, leading to better predictions of existing and future forest function and composition.
-
 # Conclusions
@@
 To our knowledge, this is the first study to demonstrate a mechanistic link between realized drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a tree community.
-We report here that deep-water access plays a role in alleviating mortality of hydrualically vulnerable trees.
-This has important implications for our predictive understanding of tropical forest dynamics under current and future climate, and suggests that greater emphasis should be given to predictive modeling of tree root allocation and distribution (e.g.
-@Wijk.2001) as well as trees' hydrological environments.
+We report here that deep-water access plays a role in alleviating mortality of hydraulically vulnerable trees.
+This has important implications for our predictive understanding of tropical forest dynamics under current and future climate.
 Our community-scale framework for modeling effective rooting depths and leaf vulnerability curves indicates the possibilities in expanding the use of these critical, hard-earned observations in species-rich forests towards community-scale generalizations.
@@
 \
-Figure \@ref(fig:erd-iso): Modelled Effective Rooting Depth (ERD; mean $\pm$ 1SE; m) versus dry season stable isotopic concentration $\mathrm{\delta}$^2^H~xylem~ (mean $\pm$ 1SE; ‰) for `r length(unique(erd.iso$sp))` canopy species from Barro Colorado Island.
+Figure \@ref(fig:erd-iso): Modeled Effective Rooting Depth (ERD; mean $\pm$ 1SE; m) versus dry season stable isotopic concentration $\mathrm{\delta}$^2^H~xylem~ (mean $\pm$ 1SE; ‰) for `r length(unique(erd.iso$sp))` canopy species from Barro Colorado Island.
 Data from [@Meinzer:1999].
@@
 \
-Figure \@ref(fig:sp-erd): Modelled Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) for `r  length(erd.sp)` large, canopy species of Barro Colorado Island.
-Each species' ERD is a median of the species' ERDs (depths) obtained from all hydrological realisations, in which soil layers are vertically resolved into exponentially increasing depths.
+Figure \@ref(fig:sp-erd): Modeled Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) for `r  length(erd.sp)` large, canopy species of Barro Colorado Island.
+Each species' ERD is a median of the species' ERDs (depths) obtained from all hydrological realizations, in which soil layers are vertically resolved into exponentially increasing depths.
 Hence the non-continuous nature of the ERD axis.
@@
 \
-Figure \@ref(fig:erd-mort-census-evg): Mortality rate (mean $\pm$ 1SE) versus Modelled Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) for `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))` evergreen, canopy species found on Barro Colorado Island over seven census intervals (1981-2015).
+Figure \@ref(fig:erd-mort-census-evg): Mortality rate (mean $\pm$ 1SE) versus Modeled Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) for `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))` evergreen, canopy species found on Barro Colorado Island over seven census intervals (1981-2015).
 $R$^2^ and $p$-values for linear model fits are given in panel insets.
@@
 \
-Figure \@ref(fig:erd-time): Modelled species Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\mathrm{\Psi}_{\textrm{soil},z}$ was below $\mathrm{\Psi}$~crit~, defined as $\mathrm{\Psi}$~80,leaf~, over 1990-2015, where $z$ is the soil depth matching species ERD--for `r length(erd.mrate.sp)` canopy species found on Barro Colorado Island.
+Figure \@ref(fig:erd-time): Modeled species Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\mathrm{\Psi}_{\textrm{soil},z}$ was below $\mathrm{\Psi}$~crit~, defined as $\mathrm{\Psi}$~80,leaf~, over 1990-2015, where $z$ is the soil depth matching species ERD--for `r length(erd.mrate.sp)` canopy species found on Barro Colorado Island.
 Each circle represents a species.
@@
 
-We also leveraged exisiting GWC measurements by @Kupers:2019 at the depths of 0.15, 0.4 and 1 m in summer 2015 (February, March and April) and 0.15, 0.4 in summer 2016 (March), with 1299 samples in total that covered all soil types and habitats in the plot.
+We also leveraged existing GWC measurements by @Kupers:2019 at the depths of 0.15, 0.4 and 1 m in summer 2015 (February, March and April) and 0.15, 0.4 in summer 2016 (March), with 1299 samples in total that covered all soil types and habitats in the plot.
 The sampling periods were 6, 5, 10 and 8 days long, respectively.
@@
 
-We generated an estimate of the mean seasonality curve for normalised leaf area index (LAI$^*$, range 0-1, unitless) for individual species as follows.
+We generated an estimate of the mean seasonality curve for normalized leaf area index (LAI$^*$, range 0-1, unitless) for individual species as follows.
 We interpolated species-specific weekly leaf-fall data collected at two similar sites in BCI (1985-2020) to a daily estimate [@Wright.1990aeo].
@@
 The difference between leaf-fall and leaf-gain gave an estimate of daily leaf-change time series, which was then scaled to express as a fraction of the total leaf fall in a given year.
-Beginning with the LAI$^*$ of 1 at the end of the wet season in the first year of the record, we acccumulated leaf-change over it to obtain a running estimate of the leaf-cover fraction, avaraged it by DOY across years and normalised it to obtain the species-specific seasonality curve in LAI$^*$ (range 0-1, unitless).
+Beginning with the LAI$^*$ of 1 at the end of the wet season in the first year of the record, we accumulated leaf-change over it to obtain a running estimate of the leaf-cover fraction, averaged it by DOY across years and normalized it to obtain the species-specific seasonality curve in LAI$^*$ (range 0-1, unitless).
 (Fig.
@@
 (2016)).
-The intercept of the curve was interpreted as maximum $K_stem$ and the water potential where the Weibull curve predicted 12, 50, 80, and 88 loss of maximum $K_stem$ were interpreted as $\\mathrm{\\Psi}$~12,stem~, $\\mathrm{\\Psi}$~50,stem~, $\\mathrm{\\Psi}$~80,stem~, and $\\mathrm{\\Psi}$~88,stem~, respectively.
+The intercept of the curve was interpreted as maximum $K_stem$ and the water potential where the Weibull curve predicted 12, 50, 80, and 88 loss of maximum $K_stem$ were interpreted as $\mathrm{\Psi}$~12,stem~, $\mathrm{\Psi}$~50,stem~, $\mathrm{\Psi}$~80,stem~, and $\mathrm{\Psi}$~88,stem~, respectively.
 
@@
 
-Below we present all the altervative models for Effective Rooting Depth (ERD).
+Below we present all the alternative models for Effective Rooting Depth (ERD).
 Daily average species-specific growth, $G_{t|z}$, in the census interval $t$ given an ERD of depth $z$, is described as follows:

### 2020-10-26 manuscript.rmd
Made concise to fit into word count limit

@@
 Summary | 197
-Main Text  | 6640
-Introduction | 892
-Methods | 3455
-Results | 871
-Discussion | 1422
+Main Text  | 6500
+Introduction | 890
+Methods | 3294
+Results | 804
+Discussion | 1512
 Keywords | 6
@@
 \
-
+\
 Item | Count
@@
 Tropical forests play a disproportionately large role in the global carbon and energy cycles [@Bonan:2008], and support half of global biodiversity [@Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019].
-Tropical forests are logically considered to be an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw], yet mortality events in the tropics are rarely as large as those in the temperate and boreal zones (@McDowell.2018), leading to questions regarding the role of large trait-diversity and hydraulic strategies in mitigating mortality events.
+Tropical forests are logically considered to be an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw], yet mortality events in the tropics are rarely as large as those in the temperate and boreal zones [@McDowell.2018], leading to questions regarding the role of large trait-diversity and hydraulic strategies in mitigating mortality events.
 Despite this, widespread drought-induced tree mortality has occurred across the tropics for specific functional groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010] and tropical mortality rates are increasing in some regions [@Brienen.2015; @Hubau.2020].
@@
 \
-\     Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation to stem water storage, deep water access, increased root production, hydraulic redistribution and embolism resistance[@McDowell.2008].
+\     Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation to stem water storage, deep water access, increased root production, hydraulic redistribution and embolism resistance [@McDowell.2008].
 Deep water access is arguably the most long-standing, yet under-studied mechanism.
@@
 However, such data are rare [@Evaristo:2016] and their interpretation is an active area of research given unresolved issues pertaining to soil heterogeneity and the correspondence of soil water extract in the laboratory to actual plant available water [@Adams.2020; @Bowers.2020; @Deurwaerder.2020].
-DNA barcoding of roots [@Jones.2011] may be used to to estimate species-specific rooting depths or rooting profiles, although extent of sampling required may be cost-prohibitive and DNA barcode libraries for tropical forests are still under development.
+DNA barcoding of roots [@Jones.2011] may be used to estimate species-specific rooting depths or profiles, but DNA barcode libraries for tropical forests are still under development, and the method may be cost-prohibitive for the extent of sampling required.
 In general, community-scale data collection for rooting or water-sourcing depths is challenging for species-rich tropical forests.
@@
 \
-\      We tested an overarching hypothesis that inter-specific variation in trees' effective rooting depths (ERD) enables increasingly reliable water availability with depth and thus mitigation of hydraulic vulnerability.
-Our hypotheses are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallow rooted species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deep rooted species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallow rooted species may have higher mortality rates as a result of higher variability in water availability in shallow soil layers.
+\      We tested an overarching hypothesis that inter-specific variation in rooting depths enables increasingly reliable water availability with depth and thus mitigation of hydraulic vulnerability.
+Our hypotheses are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallow rooted species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deep rooted species have lower mortality rates in droughts due to low realized risk of hydraulic failure, whereas shallow rooted species may have higher mortality rates as a result of higher variability in water availability in shallow soil layers.
 To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.
-
-\
-Table \@ref(tab:hypo) here.
 
@@
 \   We developed a novel, empirical inverse model to estimate species effective rooting depths (ERD) for large trees of `r length(erd.sp)` species (Fig.
-\@ref(fig:schematic)).
-We defined species ERD as the depth at which soil water potential best explained species-specific 5-yearly mean growth rates over 25 years.
+\@ref(fig:schematic)), defined as the depth at which soil water potential best explained species-specific growth dynamics.
+The ERD model incorporated the impact of atmospheric drought on growth and was constrained with species-specific leaf vulnerability curves, by using existing data for eight species and developing trait-based proxies for the rest.
 To obtain the daily dynamics of soil water potential in the whole-soil column we locally parameterized and calibrated a 1-D hydrological water balance model.
-The ERD model incorporated the impact of atmospheric drought on growth and was constrained with species-specific leaf vulnerability curves, by using existing data for eight species and developing trait-based proxies for the rest.
 We evaluated ERD against existing deuterium to hydrogen content data for a subset of six tree species as an independent observation of ERD.
 We tested relationship of ERD with above-ground hydraulic traits, namely, xylem vulnerability to embolism, branch hydraulic safety margins, leaf turgor loss point and maximum stem-specific hydraulic conductivity, for a subset of eight species with both datasets.
-Finally, we analysed species' mortality rates over 35 years to test whether ERD explained risk of drought-induced mortality in periods of extreme hydrological drought.
-\
-Fig. \@ref(fig:schematic) here.
+Finally, we analysed species' mortality rates over 35 years to test whether ERD explained risk of drought-induced mortality.
 
@@
 
-We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and data on stable water isotopes on tree species and soil from an independent study of tree water-sourcing depths [@Meinzer:1999].
+We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, hydrology and plant functional ecology, including data for stable water isotopes on trees and soils [@Meinzer:1999].
 The 1,560ha island was formed after the river Chagres was dammed in 1914 to create Gatun Lake and complete the Panama Canal.
@@
 Rainfall at BCI is seasonal with a mean annual total of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a pronounced dry season from mid-December through April with a mean total of 292 mm [@Paton.2020].
-Among the tree *species* found in the forest canopy, 28% are identified to be at least partly deciduous during the dry season, but at a maximum only 6.3% *trees* in the canopy, and only 9.7% of the total forest crown area, is deciduous [@Condit:2000].
+Among the tree *species* found in the forest canopy, 28% are identified to be at least partly deciduous during the dry season, but at a maximum only 6.3% *trees* in the canopy are deciduous, amounting to only 9.7% of the total forest crown area [@Condit:2000].
 
 \
-\     In the 50 ha plot, all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm dbh size class [@Condit.2017].
+\     In the 50 ha plot, all stems above ten mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every five year through 1985-2015 for growth, mortality as well as recruitment of new stems into the ten mm dbh size class [@Condit.2017].
 This inventory represents 321 woody species across the eight censuses.
-Aside from a 2ha patch of ~100-year-old secondary forest at the north-eastern corner of the plot the remaining 48ha represent old growth forest that was never clear cut for agriculture.
+Aside from a two ha patch of ~100-year-old secondary forest at the north-eastern corner of the plot the remaining 48ha represent old growth forest that was never clear cut for agriculture.
 The plot (1000 m $\times$ 500 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001].
@@
 \end{equation}
-where, $q_{\textrm{rain}}$ is rainfall, $E_{v}$ is transpiration of the forest, $E_{g}$ is ground evaporation, $q_{\textrm{over}}$ is surface runoff, $q_{\textrm{h2osfc}}$ is runoff from surface water storage, $q$~drai~ is sub-surface drainage (all in mm H~2~O) and $\mathrm{\Delta}t$ is the time-step (typically, 1 hour).
-The soil profile is discretized into 15 exponentially distributed soil layers with depth $z$, where $z$ $\in$ $Z$, and $Z$ = {0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13, 21, 35} m.
+where, $q_{\textrm{rain}}$ is rainfall, $E_{v}$ is transpiration of the forest, $E_{g}$ is ground evaporation, $q_{\textrm{over}}$ is surface runoff, $q_{\textrm{h2osfc}}$ is runoff from surface water storage, $q$~drai~ is sub-surface drainage (all in mm H~2~O) and $\mathrm{\Delta}t$ is the time-step (here, 1 hr).
+The soil profile is discretized into 15 exponentially distributed soil layers by depth $z$, where $z$ $\in$ $Z$, and $Z$ = {0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13, 21, 35} m.
 
@@
 In the Conrad catchment, existing data on soil hydraulic conductivity for the depths of 12.5 and 60 cm showed a decline of conductivity by depth (@Kinner:2004; Table \@ref(tab:param)).
-We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths < 12.5 cm and > 60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig.
-S\@ref(fig:ksat-profiles) [Need to fix SI Figs.
-numbering to begin at 1]).
-For greater accuracy, instead of the ELM default soil texture-based pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content vs.
+We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths <12.5 cm and >60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig.
+S\@ref(fig:ksat-profiles)).
+For greater accuracy, instead of the ELM default soil texture-based pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content (GWC) vs.
 $\mathrm{\Psi}$~soil~ for depths of 0.15, 0.4 and 1 m [@Kupers:2019].
-We converted gravimetric water content to volumetric (VWC, $\theta$) using a bulk density value of of 0.8 g cm^-3^ measured at a point location (see below).
+We converted GWC to volumetric (VWC, $\theta$) using a bulk density value of of 0.8 g cm^-3^ measured at a point location (see below).
 A @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences in water retention among depths were nominal,
@@
 We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq.
-\@ref(eq:water-balance), namely, (i) evapotranspiration (ET) from the eddy flux tower (2012-2017; see [SI](#SIData.matteo)), comparable to $E_{v} + E_{g}$ in Eq.
-\@ref(eq:water-balance)), (ii) stream discharge $q_\textrm{run}$ from the Conrad catchment, comparable to $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC or $\theta_{z}$) comparable to $W_{\textrm{liq,} z}$ from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we obtained with a vertical TDR probe over 0-15 cm depth ($\theta_\textrm{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m with 1299 samples in total covering all soil types and habitats (see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter,  $\theta_{\textrm{0-0.15 m}}$, $\theta_{\textrm{0.4 m}}$ and $\theta_{\textrm{1 m}}$ refer to plot-wide averages).
+\@ref(eq:water-balance), namely, (i) evapotranspiration (ET) from the eddy flux tower ($\cong$ $E_{v} + E_{g}$; 2012-2017; see [SI](#SIData.matteo))), (ii) stream discharge ($q_\textrm{run}$) from the Conrad catchment ($\cong$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC or $\theta_{z}$) $\cong$ $W_{\textrm{liq,} z}$ from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we obtained with a vertical time domain reflectometry (TDR) probe over 0-15 cm depth ($\theta_\textrm{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m with 1299 samples in total covering all soil types and habitats (see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter,  $\theta_{\textrm{0-0.15 m}}$, $\theta_{\textrm{0.4 m}}$ and $\theta_{\textrm{1 m}}$ refer to plot-wide averages).
 
@@
 \   Preliminary analyses showed that minimizing the objective function $\overline{O}$, that is, simultaneously fitting the model to all states and fluxes did not capture the soil moisture dynamics at 1 m depth well, our key priority for model calibration.
-To ensure model fit at 1 m depth, we first chose 300 parameter ensemble members that minimized RMSE between observations and simulations for VWC at the depth of 1 m ($\textrm{RMSE}^{*}(\theta_{\textrm{1 m}}, W_{\textrm{liq,`r soil.depths[8]` m}})$ < 0.2), and among those 300 ensemble members we chose 100 members with the lowest values for the objective function $\overline{O}$.
+To ensure model fit at 1 m depth, we first chose 300 parameter ensemble members that minimized $\textrm{RMSE}^{*}(\theta_{\textrm{1 m}}, W_{\textrm{liq,`r soil.depths[8]` m}})$ (< 0.2), and among those 300 ensemble members we chose 100 with the lowest values for the objective function $\overline{O}$.
 See Table \@ref(tab:param) for the ranges of best-fit values for different parameters.
@@
 We evaluated ELM-FATES by calculating RMSE between simulated and observed long-term daily VWC for the depths of 0.1, 0.4 and 1 m (2016-2018), based on a dataset we had left-out during calibration.
-We obtained these observations from three horizontal time domain reflectometry probes at the depths of 0.1, 0.4 and 1 m at a point location near the eddy flux tower (see [SI](#SIData.VWC)).
+We obtained these observations from three horizontal TDR probes at the depths of 0.1, 0.4 and 1 m at a point location near vertical probe (see [SI](#SIData.VWC)).
 
@@
 We ran ELM-FATES with the chosen 100 ensemble members from 1985-2018, with the first 5 years used for model spin-up.
-Extreme soil droughts were identified by depth $z$ as days that fell below the 5th percentile of $\mathrm{\Psi}_{\textrm{soil},z}$ for a given day of the year.
+Extreme soil droughts were identified by depth $z$ as days that fell below the 5^th^ percentile of $\mathrm{\Psi}_{\textrm{soil},z}$ for a given day of the year.
 
@@
 Many factors limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019].
-Because pre-dawn $\mathrm{\Psi}$~leaf~ approaches $\mathrm{\Psi}$~soil~, we used the latter to predict maximum diurnal ($K_\textrm{leaf}$), which in turn controls growth.
+Because pre-dawn $\mathrm{\Psi}$~leaf~ approaches $\mathrm{\Psi}$~soil~, we used the latter in leaf hydraulic vulnerability curves to predict maximum diurnal $K_\textrm{leaf}$, which in turn controls growth (see Table \@ref(tab:symbols) for trait symbols and their definitions).
 $K_\textrm{leaf}$ in our models thus incorporates the effects of rooting depth and rhizosphere water potential.
-This approach, however, does not account for the buffering effect of stem water storage capacitance on $\mathrm{\Psi}$~leaf~, and may thus overestimation ERD (see Discussion).
-VPD affects growth non-linearly, with growth increasing with VPD up-to a VPD threshold, then decreasing as leaf pores (stomata) close [@Grossiord.2020].
+This approach, however, does not account for the buffering effect of stem water storage capacitance on $\mathrm{\Psi}$~leaf~, and may thus overestimate ERD (see Discussion).
+VPD affects growth non-linearly, with growth increasing with VPD up-to a VPD threshold, then decreasing as leaf pores (stomata) close reducing water-uptake [@Grossiord.2020].
 Apart from stomatal control, leaf-drop may further limit water-uptake and growth.
@@
 Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting.
-At our study site, due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season, the impact of VPD may as well be a combined impact of not only temperature and humidity, but also radiation.
+Due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season at BCI, the impact of VPD may as well be a combined impact of not only temperature and humidity, but also radiation.
 
@@
 \end{equation}
-where n~t~ be the total number of days in census interval $t$, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is the model error term.
+where n~t~ is the total number of days in census interval $t$, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is the model error term.
 See SI for equations for all the alternate models tested.
@@
 \end{equation}
-\    By varying $z$, we estimated species effective rooting depths (ERD), defined as the depth $z$ at which soil water dynamics ($\mathrm{\Psi}_{\textrm{soil,} z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015).
+\    By varying $z$, we estimated species effective rooting depths (ERD), defined as the depth $z$ at which soil water dynamics ($\mathrm{\Psi}_{\textrm{soil,} z}$) best explained observed dynamics of species growth.
 Using Eq.
 \@ref(eq:M-chosen), for a species $s$, we obtained likelihood based on Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $r$ for each soil depth $z$ for each of the 100 best-fit hydrological realizations.
-For each hydrological realization *h* we retained only one depth that had the maximum likelihood (or, the minimum AIC score), and within those depths retained the ones that had Pearson's $r$ >= 0.71 (or, $R$^2 >= 0.5) as a valid set.
+For each hydrological realization *h* we retained only one depth that had the maximum likelihood, and within those depths retained the ones that had Pearson's $r$ >= 0.71 as a valid set.
 For a species $s$, we defined effective rooting depth (ERD) as the median of these set of the retained depths, which can be described as :
@@
 where $L$ is the likelihood, $H$ = {1, .., 100}, refer to the 100 hydrological realizations and $Z$ = {0.4, `r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m.
-We removed finely resolved soil layers below 0.21 m, averaged $\mathrm{\Psi}_{\textrm{soil},z}$ for depths 0.21, 0.37 and 0.62 at 0.4 m.
+We removed finely resolved soil layers below 0.21 m and averaged $\mathrm{\Psi}_{\textrm{soil},z}$ for depths 0.21, 0.37 and 0.62 at 0.4 m as sufficient resolution for Model validation.
 
@@
 For growth estimates, to minimize the effect of light variation among trees, we selected only large trees ($\ge$ 30 cm dbh, the diameter at breast height) and also species whose maximum height was $\ge$ 30 m (hereafter, canopy species) and thus are likely to be fully exposed to the sun.
-Calculating individual tree growth rates across six censuses (1990-2015), removing outliers, accounting for temporal changes in growth rates due to increasing tree size [see SE], we obtained species-level median time series of growth rates ($G_{t}$, cm yr^-1^) used in the ERD model, for only those species ($n$ = `r length(erd.sp)`) with complete records on at least three trees (median $n$ = `r median(g.n$n)`, max $n$ = `r max(g.n$n)` trees per species).
+Calculating individual tree growth rates across six 5-yearly censuses (1990-2015), removing outliers, accounting for temporal changes in growth rates due to increasing tree size (see SI), we obtained species-level median time series of growth rates ($G_{t}$, cm yr^-1^) used in the ERD model for only those species ($n$ = `r length(erd.sp)`) with complete records on at least three trees (median $n$ = `r median(g.n$n)`, max $n$ = `r max(g.n$n)` trees per species).
 
@@
 We obtained leaf hydraulic vulnerability curves ($K_\textrm{leaf}$ vs.
-$\mathrm{\Psi}$~leaf~; see Table \@ref(tab:symbols) for trait symbols and their definitions) for adult trees of 21 common species from Zailaa et al.
+$\mathrm{\Psi}$~leaf~) for adult trees of 21 common species from Zailaa et al.
 (to be submitted; see SI for brief description of methods) described as:
@@
 We identified scaling relationships between fitted parameters $A$ and $B$ in Eq.
-\@ref(eq:vc-exp) and two traits, namely, WSG, the wood specific gravity after drying at 100^$\circ$^C and LMA, the leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules for leaves receiving direct sunlight (@Wright:2010; see SI for details).
+\@ref(eq:vc-exp) and two traits; namely, WSG, the wood specific gravity, and LMA, the leaf mass per unit area (@Wright:2010; see SI for details).
 We fitted polynomial equations described as,
@@
 \#ref(eq:vc-exp).
-For all the `r length(erd.sp)` ERD species we estimated species $\mathrm{\Psi}$~50,leaf~ based on their vulnerability curves (see Table \@ref(tab:symbols) for definition).
+For all the `r length(erd.sp)` ERD species we estimated species $\mathrm{\Psi}$~50,leaf~ based on their vulnerability curves for definition).
 
@@
 
-Table \@ref(tab:symbols) here.
-
 #### Leaf Area Index (LAI)
 
-In some of the alternative models for effective rooting depth, we explored the effect of seasonality in leaf area index (LAI, unitless) on growth.
-We assumed species-specific mean seasonal curves for LAI normalized between 0 and 1 (LAI, unitless), informed by a combination of long-term records for leaf-fall [@Wright.1990aeo] and leaf-lifetime [@Osnas.2018] (see SI).
+In some of the alternative models for effective rooting depth, we explored the effect of seasonality in leaf area index (LAI) on growth.
+We assumed species-specific mean seasonal curves for LAI (normalized between 0 and 1; unitless), informed by a combination of long-term records for leaf-fall [@Wright.1990aeo] and leaf-lifetime (@Osnas.2018; see SI).
 
@@
 \
-As root water-uptake is generally a non-fractionating process, 2^H/1^H content of the tree xylem sap (hereafter, $\mathrm{\delta}$^2^H~xylem~) reflects the signature of the source water.
-Given a vertical gradient in $\mathrm{\delta}$^2^H water by depth, $\mathrm{\delta}$^2^H~xylem~ can be used as an index of rooting depth [@Dawson.1991].
-We thus used $\mathrm{\delta}$^2^H~xylem~ as an independent observation to validate the ERD models.
-We leveraged $\mathrm{\delta}$^2^H~xylem~ values from BCI given in @Meinzer:2001 for the dry-season of March 1997 as this period showed largest seasonal divergence in $\mathrm{\delta}$^2^H~xylem~ among species, in soil and ground water $\mathrm{\delta}$^2^H at natural abundance level (Fig.
+We used $\mathrm{\delta}$^2^H~xylem~ as an independent observation to validate the ERD models.
+As root water-uptake is generally a non-fractionating process, tree $\mathrm{\delta}$^2^H~xylem~ reflects signature of source water.
+Given a vertical gradient of $\mathrm{\delta}$^2^H in soil and ground water $\mathrm{\delta}$^2^H~xylem~ provides an index of rooting depth [@Dawson.1991].
+We leveraged $\mathrm{\delta}$^2^H~xylem~ from BCI [@Meinzer:2001] for the dry-season of March 1997 as this period showed largest seasonal divergence in $\mathrm{\delta}$^2^H~xylem~ among species and in soil and ground water $\mathrm{\delta}$^2^H at natural abundance level (Fig.
 S\@ref(fig:iso-data)).
-Analyses of $\mathrm{\delta}$^2^H~xylem~ of the 12 species studied by @Meinzer:2001 showed that four species sourced water between 1 m and ground water, two other at least 0.8 m deep, one between 0 to 0.2 m, and five in the 0.2 to 0.7 m depth range over which the relationship between soil depth and $\mathrm{\delta}$^2^H~soil~ was variable.
+Analysing $\mathrm{\delta}$^2^H~xylem~ for 12 tree species @Meinzer:2001 found that four species sourced water between 1 m and ground water, two other at least 0.8 m deep, one between 0 to 0.2 m, and five in the 0.2 to 0.7 m depth range over which the relationship between soil depth and $\mathrm{\delta}$^2^H~soil~ was variable.
 \
-\   We removed those species that would be leafless in March 1997 because xylem sap $\mathrm{\delta}$^2^H for these species may not be linked with water sourced at the time of measurement and thus may not be comparable with other species that had leaves.
+\   We removed those species that would be leafless in March 1997 because xylem sap $\mathrm{\delta}$^2^H for these species may not be linked with water sourced at the time of measurement and thus may be compared to species that had leaves.
 We also removed one species whose $\mathrm{\delta}$^2^H~xylem~ was in the uncertain depth range.
@@
 \@ref(eq:M-chosen) and Eqs.
-in SI), estimated species ERD was regressed against species $\mathrm{\delta}$^2^H~xylem~, for a maximum of six species.
+in SI), modelled species ERD was regressed against species $\mathrm{\delta}$^2^H~xylem~ (six species).
 
@@
 
-To evaluate whether ERD was associated with above-ground hydraulic traits ($n$ = 8 species with both datasets), we regressed species ERD against maximum stem hydraulic conductance ($K_\textrm{max,stem}$), leaf turgor loss point ($\mathrm{\Psi}$~tlp~), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\mathrm{\Psi}$~88,stem~), and hydraulic safety margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~).
+To evaluate whether ERD was associated with above-ground hydraulic traits (eight species with both datasets), we regressed species ERD against maximum stem hydraulic conductance ($K_\textrm{max,stem}$), leaf turgor loss point ($\mathrm{\Psi}$~tlp~), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\mathrm{\Psi}$~88,stem~), and hydraulic safety margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~).
 See [SI](#SIData.brett) for data collection and estimation of these variables.
@@
 To test whether ERD plays a role in alleviating mortality risk, we calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$ 30 m) and average abundance of at least 20 trees ($n$ = `r length(erd.mrate.sp)`).
-For each of the seven census intervals $t$ in the 35-yr record (1981-82, 1985, 1990, 2000, 2005, 2010, 2015, 2015), $\textrm{Mort}_{t}$ (% yr^-1^) for species $s$ for two successive censuses, $c_{1}$ and $c_{2}$, was calculated as $\textrm{Mort}_{s,t} = \frac{D_{s,c_{1}}}{N_{s,c_{2}}}\times\frac{100}{\textrm{d}}$, where $N_{s,c_{1}}$ and $D_{s,c_{2}}$ are the total number of trees of a species $s$ present in $c_{1}$ and dead in $c_{2}$, respectively, in the 50 ha plot and $d$ is the duration based on mean dates of $c_{1}$ and $c_{2}$.
+For each of the seven census intervals $t$ in the 35-yr record (1981, 1985, 1990, 2000, 2005, 2010, 2015, 2015), mortality rate, $\textrm{M}_{t}$ (% yr^-1^) for species $s$ for two successive censuses, $c_{1}$ and $c_{2}$, was calculated as $\textrm{M}_{s,t} = \frac{D_{s,c_{1}}}{N_{s,c_{2}}}\times\frac{100}{\textrm{d}}$, where $N_{s,c_{1}}$ and $D_{s,c_{2}}$ are the total number of trees of species $s$ present in $c_{1}$ and dead in $c_{2}$, respectively, in the 50 ha plot and $d$ is the duration based on mean dates of $c_{1}$ and $c_{2}$.
 
 \
-\   We regressed species ERD against species ($\textrm{Mort}_{s,t}$ for each of the census interval $t$) (with and without deciduous species).
+\   We regressed species ERD against species $\textrm{M}_{s,t}$ for each census interval $t$ with ($n$ = `r nrow(mrate.depth.mean)`) and without deciduous species ($n$ = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`).
 To assess whether deeper ERD species escaped the experience of significant soil drought, we calculated for each species the proportion of days in the simulation period (1990-2015) during which soil water potential in the soil layer matching species ERD ($\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$) were below species $\mathrm{\Psi}$~crit~.
@@
 \@ref(fig:hydro-calib)D), slightly underestimating peak discharge in wet years and slightly overestimating peak ET.
-The reduction in parameter range in the best-fit ensembles compared to the global ranges tested showed that the model calibration was sensitive to the Ball-Berry stomatal slope parameter, the ELM root distribution parameter (that regulates the depth of the rooting profile), soil hydraulic conductivity profile, especially at depth, and the adjustment factor that varies soil hydraulic conductivity to account for macroporosity and direct flow paths (Table \@ref(tab:param)).
+The reduction in parameter range in the best-fit ensembles compared to the tested global ranges showed that the model calibration was sensitive to the Ball-Berry stomatal slope parameter, the ELM root distribution parameter (that regulates the depth of the rooting profile), soil hydraulic conductivity profile, especially at depth, and the adjustment factor that varies soil hydraulic conductivity to account for macroporosity and direct flow paths (Table S\@ref(tab:param)).
 
-Fig. \@ref(fig:hydro-calib) here.
-
-\   Soil Water Potential dynamics ($\mathrm{\Psi}_{\textrm{soil},z}$) by depth $z$ (from 0.1 through 13 m) obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of such extreme years, drought intensity at depth, drought seasonality and duration (Fig.
+\   Soil Water Potential dynamics ($\mathrm{\Psi}_{\textrm{soil},z}$) by depth $z$ (from 0.1 through 13 m) obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of extreme years, drought intensity at depth, drought seasonality and duration (Fig.
 \@ref(fig:swp-droughts)).
-Notably, soil droughts in census intervals 2000-2005 and 2005-2010 distinctly occurred in the wet season, with one to two extreme soil drought years each, depending on depth location (Fig.
+Soil droughts in census intervals 2000-2005 and 2005-2010 distinctly occurred in the wet season, with one to two extreme soil drought years each, depending on the depth location (Fig.
 S\@ref(fig:swp-droughts)).
@@
 S\@ref(fig:swp-all-depths)).
-
-Fig. \@ref(fig:swp-droughts) here.
 
@@
 S\@ref(fig:obs-plc-curves)).
-This predictive power allowed us to estimate leaf vulnerability curve parameters for `r sp.n.model` (out of `r length(erd.sp)` ERD species) that lacked direct observations; Table S\@ref(tab:ab-param), Fig.
+This predictive power allowed us to estimate leaf vulnerability curve parameters for `r sp.n.model` that lacked direct observations (out of `r length(erd.sp)` ERD species; Table S\@ref(tab:ab-param), Fig.
 S\@ref(fig:obs-comm-plc-curves)).
-These vulnerability curves were used in the models for effective rooting depth.
-
-Fig. \@ref(fig:ab-fit) here.
+These vulnerability curves were used in the ERD models.
 
@@
 
-The chosen model (with a multiplicative effect of VPD and not LAI; Eq.
+The chosen ERD model (with a multiplicative effect of VPD and not LAI; Eq.
 \@ref(eq:M-chosen)) explained a large fraction of the variance in $\mathrm{\delta}$^2^H~xylem~ ($R$^2^ = `r lm.erd.iso.r2`; $p$ = `r lm.erd.iso.p`, $n$ = `r length(unique(erd.iso$sp))`; Fig.
 \@ref(fig:erd-iso)).
-Two of the alternative models tested with LAI had a strong relationship with $\mathrm{\delta}$^2^H~xylem~ as well, but one that was based on a smaller set of species ($n$ = 4) as fewer species passed the criteria of sufficient species-level goodness-of-fit with these models; <20 species vs.
-`r length(erd.sp)` species for the chosen model (Fig.
+Two of the alternative models tested with LAI had a strong relationship with $\mathrm{\delta}$^2^H~xylem~ as well, but one that was based on a smaller set of species ($n$ = 4) as fewer species passed the criteria of sufficient species-level goodness-of-fit with these models (<20 species vs.
+`r length(erd.sp)` species for the chosen model; Fig.
 S\@ref(fig:other-erd-mods)).
-For the rest of the analyses we therefore used ERD estimates from the less parsimonious, chosen model (with VPD and not LAI, Eq.
-\@ref(eq:M-chosen)).
-
-Fig. \@ref(fig:erd-iso) here.
+For the rest of the analyses we therefore used ERD estimates from the less parsimonious, chosen model.
 
 \
-\   In this forest, modeled effective rooting depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from `r min(df.erd.to.plot$depth)` m to `r max(df.erd.to.plot$depth)` m (Fig.
+\   Modeled effective rooting depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from `r min(df.erd.to.plot$depth)` m to `r max(df.erd.to.plot$depth)` m (Fig.
 \@ref(fig:sp-erd)).
-Notably, two species, *L.
-seemannii* and *T.
-tuberculata*, that @Meinzer:1999 found to have $\mathrm{\delta}$^2^H~xylem~ values between soil water and groundwater, suggesting that these species sourced most of the water from depths >1 m, were also identified by our model with ERD >1 m (2.9 m for both species; Fig.
+Notably, two species, *Luehea seemannii* and *Trichilia tuberculata*, that @Meinzer:1999 found to have $\mathrm{\delta}$^2^H~xylem~ values between soil water and groundwater, suggesting that these species sourced most of the water from depths >1 m, were also identified by our model with ERD >1 m (2.9 m for both species; Fig.
 \@ref(fig:erd-iso)).
-
-\
-Fig. \@ref(fig:sp-erd) here.
 
@@
 
-For the subset of eight species with ERD for which hydraulic traits data were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_\textrm{max,stem}$, Spearman's' $r$ = `r cor.k`, $p$ = `r p.k`; Fig.
-\@ref(fig:erd-hyd)A), less negative leaf $\mathrm{\Psi}$~tlp~ (Spearman's' $r$ = `r cor.tlp`, $p$ = `r p.tlp`; Fig.
-\@ref(fig:erd-hyd)B), less negative $\mathrm{\Psi}$~88,stem~ and thus greater vulnerability to embolism from cavitation (Spearman's' $r$ = `r cor.stem.88`, $p$ = `r p.stem.88`; Fig.
-\@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~; Spearman's' $r$ = `r cor.hsm`, $p$ = `r p.hsm`; Fig.
+Species with deeper ERD showed greater maximum stem hydraulic conductance ($K_\textrm{max,stem}$, Spearman's' $r$ = `r cor.k`, $p$ = `r p.k`; Fig.
+\@ref(fig:erd-hyd)A), less negative leaf $\mathrm{\Psi}$~tlp~ ($r$ = `r cor.tlp`, $p$ = `r p.tlp`; Fig.
+\@ref(fig:erd-hyd)B), less negative $\mathrm{\Psi}$~88,stem~ and thus greater vulnerability to embolism from cavitation ($r$ = `r cor.stem.88`, $p$ = `r p.stem.88`; Fig.
+\@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~; $r$ = `r cor.hsm`, $p$ = `r p.hsm`; Fig.
 \@ref(fig:erd-hyd)D).
-
-Fig. \@ref(fig:erd-hyd) here.
 
@@
 
-Amongst evergreen species ERD explained `r m.evg.r2.90`, `r m.evg.r2.05` and `r m.evg.r2.10`% variation in mortality rates during census intervals 1985-1990, 2000-2005 and 2005-2010, respectively, such that species with deeper ERD had significantly lower mortality rates than deeper ERD species ($p$ = `r m.evg.p.90`, `r m.evg.p.05` and `r m.evg.p.10`, respectively; $n$ = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`; Fig.
+Among evergreen species, ERD explained `r m.evg.r2.90`, `r m.evg.r2.05` and `r m.evg.r2.10`% variation in mortality rates during census intervals 1985-1990, 2000-2005 and 2005-2010, respectively, such that species with mortality rates significantly decreased with ERD ($p$ = `r m.evg.p.90`, `r m.evg.p.05` and `r m.evg.p.10`, respectively; Fig.
 \@ref(fig:erd-mort-census-evg)).
-When deciduous species were included, ERD explained lower but significant proportion of variation in mortality in those intervals ($R$^2^ = `r m.r2[1]`, `r m.r2[2]` and `r m.r2[3]`% and $p$ = `r m.p.90`, `r m.p.05` and `r m.p.10`, respectively; $n$ = `r nrow(mrate.depth.mean)`; Fig.
+When deciduous species were included, ERD explained lower but significant proportion of variation in mortality in those intervals ($R$^2^ = `r m.r2[1]`, `r m.r2[2]` and `r m.r2[3]`% and $p$ = `r m.p.90`, `r m.p.05` and `r m.p.10`, respectively; Fig.
 \@ref(fig:erd-mort-census)).
@@
 \@ref(fig:erd-time); S\@ref(fig:erd-time-evg)).
-Species with ERD deeper than 1 m did not cross species $\mathrm{\Psi}$~crit~ for the entire 25-yr period for which soil water potential $\mathrm{\Psi}_{\textrm{soil},z}$ was modeled (1990-2015).
+Species with ERD deeper than 2 m did not cross species $\mathrm{\Psi}$~crit~ for the entire 25-yr period for which soil water potential $\mathrm{\Psi}_{\textrm{soil},z}$ was modeled (1990-2015).
 Soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and drought periods, which is above ($\mathrm{\Psi}$~crit~) of the most sensitive tree species (-0.28 MPa).
 
-Fig. \@ref(fig:erd-mort-census-evg) here.
-Fig. \@ref(fig:erd-time) here.
-
 # Discussion
-
-Our novel approach for estimating ERD using 25 years of growth, species-specific leaf vulnerability curves, modeled soil moisture through the profile, and observed VPD was consistent with estimates using xylem water isotopes.
-Our analyses revealed that co-occurring large canopy tree species with deeper effective rooting depths (ERD) were associated with higher hydraulic efficiency, but not safety (Fig.
+\
+Our novel approach for estimating ERD using 25 years of growth, species-specific leaf vulnerability curves, modeled soil moisture through the profile, and observed VPD was consistent with estimates using $\mathrm{\delta}$^2^H~xylem~.
+Our analyses revealed that co-occurring large canopy tree species with deeper ERD were associated with higher hydraulic efficiency, but not safety (Fig.
 \@ref(fig:erd-hyd)).
 Species with deeper ERD showed higher maximum stem hydraulic conductance, and greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species.
-Nevertheless, deeper ERD species showed lower mortality rates than shallower ERD species over the 35-yr record (1981-2015) (Fig.
+Nevertheless, deeper ERD species showed lower mortality rates than shallower ERD species in significant wet season droughts over the 35-yr record (1981-2015) (Fig.
 S\@ref(fig:erd-mort-census), especially amongst the evergreen species (Fig.
@@
 
-### Realised drought risk integral to assessing drought-vulnerability
+### Realized drought risk integral to assessing drought-vulnerability
 
 Our study brings attention to the need for assessing drought sensitivity in terms of species realized drought exposure by accounting for the actual hydrological drought and trees' deep water access, beyond the universal drought indices and trees' above-ground hydraulic vulnerability.
-We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins, that is, traits that are commonly identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through periods of extreme droughts (Figs.
+We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins (Fig.
+\@ref(fig:erd-hyd)), that is, traits that are commonly identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through periods of extreme droughts (Fig.
 \@ref(fig:erd-mort-census-evg), S\@ref(fig:erd-mort-census), \@ref(fig:swp-droughts)).
 Deeper ERD species on average had more hydraulically efficient, but risky, suite of traits optimized for an environment of more reliable and greater water source at depth that resulted in a lower mortality risk.
-By contrast shallow ERD species had inefficient, but safer, xylem likely optimized for high variability in $\mathrm{\Psi}_{soil}$ in the shallow soil layers, and resulting in higher variance in mortality rates across censuses as $\mathrm{\Psi}_{soil,~z = ERD}$ crossed ERD $\mathrm{\Psi}_{crit}$.
-Our results are also consistent with a few studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016; also see @Brum.2019].
+By contrast shallow ERD species had inefficient, but safer, xylem likely optimized for high variability in $\mathrm{\Psi}_{soil}$ in the shallow soil layers, and resulting in higher variance in mortality rates across censuses as $\mathrm{\Psi}_{soil,~z = ERD}$ crossed ERD $\mathrm{\Psi}_{crit}$ (Fig.
+\@ref(fig:erd-mort-census-evg); Fig.
+\@ref(fig:erd-time)).
+Our results are consistent with a few studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016; also see @Brum.2019].
 
 \
-\   Our result--of lower mortality in deep rooted trees--contrasts with the finding of @Chitra-Tarak.2018 (hereafter CT2018) in which species identified by their inverse ERD model as effectively deep rooted died *more* in a rare, prolonged drought.
-The hydrological model in that study revealed that the multi-year drought exhausted the deep soil and even bedrock water availability (also see, @Goulden.2019 and @Ivanov:2012).
-In contrast, our hydrological modeling at BCI finds that during the study period, $\mathrm{\Psi}$~soil~ below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and droughts, that is above ($\mathrm{\Psi}$~crit~) of the most sensitive tree species (-0.28 MPa).
-The difference in the results shown here are likely attributable to site differences in the precipitation regime: with mean annual rainfall of 1095 $\pm$ 296 mm (mean $\pm$ SD) in CT2018 compared to `r round(rain.man.stats$rain)` $\pm$ `r round(rain.man.stats$rain.sd)` mm (`r min(clim$Year)`-`r max(clim$Year)`) at BCI, and precipitation to potential evapotranspiration ratio nearly one in CT2019 compared to nearly two in BCI (Penman-Monteith PET = `r round(mean(clim$pet.PM))` $\pm$ `r round(sd(clim$pet.PM))` mm;`r min(clim$Year)`-`r max(clim$Year)`; this study, @Guo.2016).
+\   Our result--of lower mortality in deep rooted trees--contrasts with the finding of @Chitra-Tarak.2018 (hereafter, CT2018) in which inversely modelled, deeper ERD species died *more* in a rare, prolonged drought.
+The hydrological model in CT2018 revealed that the multi-year drought exhausted the deep soil and even bedrock water availability [also see, @Goulden.2019 and @Ivanov:2012].
+In contrast, our hydrological modeling at BCI finds that during the study period, $\mathrm{\Psi}$~soil~ below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and droughts, that is above $\mathrm{\Psi}$~crit~ of the most sensitive tree species (-0.28 MPa).
+The difference in the results shown here are likely attributable to site differences in the precipitation regime: with mean annual rainfall of 1095 in CT2018 compared to `r round(rain.man.stats$rain)`mm at BCI, and precipitation to potential evapotranspiration ratio nearly one in CT2018 compared to nearly two in BCI.
 Thus at BCI deep soil layers are effectively recharged annually (Fig.
-S\@ref(fig:swp-all-depths); @Chitra-Tarak.2018).
-The kind of droughts that buffered the deep rooted species in our study were repeated annual droughts that extended the dry season into the wet season.
-The variation in mortality explained by ERD in our study is up-to 40%, and other causes may explain observed mortality [@McDowell.2018].
+S\@ref(fig:swp-all-depths)).
+The kind of droughts that buffered the deep rooted species from mortality in our study were repeated annual droughts that extended the dry season into the wet season (Fig.
+\@ref(fig:swp-droughts)).
+This highlights the role of site-specific hydrology in how ERD modulates mortality-risk in different kinds of droughts.
 
 ### Effective rooting depths at the tree community level
-
+\
 By estimating ERD for large trees of `r length(erd.sp)` canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species-rich tropical forests.
 Our chosen ERD model is an advance over the only other inverse model of ERD (@Chitra-Tarak.2018; $n$ = 12 species) as we employ a physiologically meaningful representation, fewer parameters and corroboration with tree and soil data for $\mathrm{\delta}$^2^H.
-Stable water isotopes for plant and below-ground sources are critical for calibration or validation of ERD model at community scale, although the interpretation of these data is still under research [@Deurwaerder.2020].
+We highlight that stable water isotopes for plant and below-ground sources are critical for calibration or validation of ERD model at community scale, although the interpretation of these data is still under research [@Deurwaerder.2020].
 As in @Chitra-Tarak.2018, we also interpret our ERD estimates as revealing relative differences among species' effective rooting depths rather than in absolute terms.
-As our ERD model ignores stem water storage capacitance it may have overestimated ERD especially in species for which stem water storage capacitance is important as a drought-avoidance strategy; for example, the deciduous species [Borchert.2005jma].
+As our ERD model ignores stem water storage capacitance it may have overestimated ERD especially in species for which stem water storage capacitance is important as a drought-avoidance strategy, for example, the deciduous species [@Borchert.2005jma].
 
 ### Future directions
-
-The relationships that we identify between above- and below-ground traits, the water environment vertically below-ground and mortality rates are important for representing diversity in Dynamic Global Vegetation Models, which intrinsically rely on the parameterization of contrasting life history strategies [@Scheiter:2013] and the simulation of competition between those strategies.
+\
+The relationships that we identify between above- and below-ground traits, the water environment vertically below-ground and mortality rates are important for representing diversity in DGVMs, which intrinsically rely on the parameterization of contrasting life history strategies [@Scheiter:2013] and the simulation of competition between those strategies.
 We found that LMA and WSG were strong predictors of PLC curves for leaves, but not for stems (not shown).
-This offers the promise of a greater ability to parameterize the 'hard' hydraulic traits with the abundant soft trait data, thus allowing much better representation of forest hydrodynamics.
+This offers the promise of a greater ability to parameterize the 'hard' hydraulic traits with the abundant 'soft' trait data, thus allowing for better representation of forest hydrodynamics.
 The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits, including hydraulic vulnerability, in order to better represent the functional diversity of tropical forests, and their trajectories into the future.
@@
 \
-\   Our inverse ERD model was parameterized on 5-year growth data, with five data points over a 25-yr period, which decouples climate events and demographic outcomes.
+\   Our inverse ERD model was parameterized on 5-year growth data, with five data points over a 25-yr period, which decoupled climate events and demographic outcomes.
 Future studies could better constrain the ERD model with high frequency growth data such as those from dendrometer bands.
-At high temporal resolution, however, the role of reversible dehydration in tree diameter change increases [@Chitra-Tarak.2016; @Chitra-Tarak.2015], but it may also be an avenue to include stem water storage  and dynamic rooting depths in ERD models.
+At high temporal resolution, however, the role of reversible dehydration in tree diameter change increases [@Chitra-Tarak.2016; @Chitra-Tarak.2015], but that may provide an avenue to include stem water storage  and dynamic rooting depths in ERD models.
 Three of our exploratory ERD models included LAI seasonality, but we did not select them as they worsened the fit with growth data for many species (Fig.
@@
 \
-\   While our ERD model empirically predicts growth via $\mathrm{\Psi}$~leaf~ from $\mathrm{\Psi}$~soil~ at a specific depth, models that mechanistic models that account for plant hydrodynamics and other processes influencing growth are likely to be more accurate [@Christoffersen:2016kd2d; @Sperry.1998].
+\   While our ERD model empirically predicts growth via estimating $\mathrm{\Psi}$~leaf~ from $\mathrm{\Psi}$~soil~ of a specific depth, mechanistic models that account for plant hydrodynamics and other processes influencing growth are likely to predict growth more accurately, and thus ERD [@Christoffersen:2016kd2d; @Sperry.1998].
 Although data needs for parameterizing such models could be greater (e.g.
@@
 
-**RCT: Upload on NGET-Archive and site DOIs:
+To be uploaded on NGET-Archive and site DOIs:
 (1) Boris to upload climate forcing data
@@
 \
-Barro Colorado Forest Census Plot Data can be obtained online[@Condit.2019a; @Condit.2019b].
+Barro Colorado Forest Census Plot Data can be obtained online [@Condit.2019a; @Condit.2019b].
 
@@
 \
-Figure \@ref(fig:hydro-calib): ELM-FATES calibration and evaluations.
-Observations (red lines) versus simulations (blue lines) from 100 best-fit parameter ensemble runs of ELM-FATES are shown for daily Volumetric Water Content (VWC) by depth for horizontal time domain reflectometry (TDR) probes at three depths (0.1, 0.4 and 1 m; A) and a vertical probe (0-0.15 m; B); monthly stream discharge (Discharge, C) and monthly evapotranspiration from eddy flux tower (ET, D).
+Figure \@ref(fig:hydro-calib): ELM-FATES calibration and evaluation.
+Observations (red lines) versus simulations (gray lines) from 100 best-fit ensemble-member runs of ELM-FATES are shown for daily Volumetric Water Content (VWC) by depth for horizontal time domain reflectometry (TDR) probes at three depths (0.1, 0.4 and 1 m; A) and a vertical probe (0-0.15 m; B); monthly stream discharge (Discharge, C) and monthly evapotranspiration from eddy flux tower (ET, D).
 Points (red) in (A) show average plot-wide observations of VWC with 95% CI (error bars).
-VWC data from the horizontal probes (lines in A) were not used for ELM-FATES calibration but evaluation.
+VWC data from the horizontal probes (lines in A) were not used for ELM-FATES calibration, but evaluation.
 Values in inset are average RMSE across the 100 best-fit simulations.
-For panel A, values at the top are for plot-wide VWC and those at the bottom are for TDR probes.
+For panel A, values at the top are for manual, plot-wide VWC and those at the bottom are for VWC from TDR probes.
 VWC is in units of cm^3^ cm^-3^, while ET and Discharge in mm month^-1^.
@@
 \
-Figure \@ref(fig:ab-fit): Occurrence of extreme soil droughts by census intervals (horizontal panels) at three representative depths (z): 0.1, 0.6 and 1 m (vertical panels).
+Figure \@ref(fig:ab-fit): Occurrence of extreme soil droughts by census intervals (horizontal panels) at three representative depths ($z$): 0.1, 0.6 and 1 m (vertical panels).
 Mean (black lines) and lower half of 95% distribution (grey areas) of soil water potential ($\mathrm{\Psi}_{\textrm{soil},z}$) for a given DOY and depth $z$ are shown (same across census intervals).
-$\mathrm{\Psi}_{\textrm{soil},z}$ by DOY for a year (colored line) is only shown if for some part of that year $\mathrm{\Psi}_{\textrm{soil},z}$ declined below the 5 percentile distribution.
+$\mathrm{\Psi}_{\textrm{soil},z}$ by DOY for a year (colored line) is only shown if, at least one day of that year $\mathrm{\Psi}_{\textrm{soil},z}$ declined below the 5^th^ percentile for that day of year (DOY).
 
@@
 \
-Figure \@ref(fig:erd-iso): Modeled Effective Rooting Depth (ERD; mean $\pm$ 1SE; m) versus dry season stable isotopic concentration $\mathrm{\delta}$^2^H~xylem~ (mean $\pm$ 1SE; ‰) for `r length(unique(erd.iso$sp))` canopy species from Barro Colorado Island.
+Figure \@ref(fig:erd-iso): Modeled Effective Rooting Depth (ERD; mean $\pm$ 1SE; m) versus dry-season stable isotopic concentration $\mathrm{\delta}$^2^H~xylem~ (mean $\pm$ 1SE; ‰) for `r length(unique(erd.iso$sp))` canopy species from Barro Colorado Island.
 Data from [@Meinzer:1999].
-See Table  \@ref(tab:ab-param) for the full form of scientific names.
 
@@
 Figure \@ref(fig:sp-erd): Modeled Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) for `r  length(erd.sp)` large, canopy species of Barro Colorado Island.
-Each species' ERD is a median of the species' ERDs (depths) obtained from all hydrological realizations, in which soil layers are vertically resolved into exponentially increasing depths.
-Hence the non-continuous nature of the ERD axis.
+Species ERD is a median of the species' ERDs (depths) obtained from the 100 best-fit hydrological realizations, which are vertically resolved into exponentially increasing depths.
+Hence, the non-continuous nature of the ERD axis.
 
@@
 Figure \@ref(fig:erd-iso): Matrix for pairwise correlations (circles) among Effective Rooting Depth (ERD) and hydraulic properties for `r nrow(cor.data)` canopy species found on Barro Colorado Island (Panama).
-Each circle is color coded by the Spearman's $r$ with red shades for positive $r$ values, while blue shades for negative values, respectively.
+Each circle is color coded by the Spearman's $r$ with red shades for positive $r$ values and blue shades for negative values.
 $p$-values are given only when they are $\ge$0.05.
-Units are as follows: ERD in m, K~max~ in mmol m^-1^ s^-1^ MPa^-1^), all variables with $\mathrm{\Psi}$ in MPa.
+Units are as follows: ERD in m, K~max,stem~ in mmol m^-1^ s^-1^ MPa^-1^) and variables with $\mathrm{\Psi}$ in MPa.
 
 \
-Figure \@ref(fig:erd-mort-census-evg): Mortality rate (mean $\pm$ 1SE) versus Modeled Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) for `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))` evergreen, canopy species found on Barro Colorado Island over seven census intervals (1981-2015).
+Figure \@ref(fig:erd-mort-census-evg): Mortality rate (mean $\pm$ 1SE in % yr^-1^) versus Modeled Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) for `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))` evergreen, canopy species found on Barro Colorado Island over seven census intervals (1981-2015).
 $R$^2^ and $p$-values for linear model fits are given in panel insets.
@@
 \
-Figure \@ref(fig:erd-time): Modeled species Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\mathrm{\Psi}_{\textrm{soil},z}$ was below $\mathrm{\Psi}$~crit~, defined as $\mathrm{\Psi}$~80,leaf~, over 1990-2015, where $z$ is the soil depth matching species ERD--for `r length(erd.mrate.sp)` canopy species found on Barro Colorado Island.
+Figure \@ref(fig:erd-time): Modeled species Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) versus time spent below critical hydraulic threshold--% days for which $\mathrm{\Psi}_{\textrm{soil},z}$ was below $\mathrm{\Psi}$~crit~, defined as $\mathrm{\Psi}$~50,leaf~, over 1990-2015, where $z$ is the soil depth matching species ERD--for `r length(erd.mrate.sp)` canopy tree species on Barro Colorado Island.
 Each circle represents a species.
@@
 
-\newpage
 ```{r hypo, results = 'asis', echo = FALSE}
@@
 
-\newpage
+#
 ```{r symbols, results = 'asis', echo = FALSE}
@@
 
-### Data gaps in WSG and LMA {#SIData.rutuja}
-
-Among the `r length(sp.exp.param$sp)` species with observed data on ($K_\textrm{leaf}$ vs.
+### WSG and LMA {#SIData.rutuja}
+\
+We used WSG, the wood specific gravity after drying at 100^$\circ$^C and LMA, the leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules for leaves receiving direct sunlight as described in @Wright:2010.
+\
+\   Among the `r length(sp.exp.param$sp)` species with observed data on ($K_\textrm{leaf}$ vs.
 $\mathrm{\Psi}$~leaf~) used to build the community-level models of leaf vulnerability curves (Eqs.
@@
 ```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-other-erd-mods}Effective Rooting Depth versus $\\mathrm{\\delta}^2H$ for each of the ERD models tested (panels).
-For each ERD model only those species have an estimate of ERD that have passed the test of Pearson's $r$ > 0 and $R$^2^ > 0.5 during ERD parameterisation.
-The number of species with an ERD estimate as well as $\\mathrm{\\delta}^2H$.
-or each model are shown in inset, along with the goodness-of-fit ($R$^2^) and $p$-values for a linear relationship between ERD and $\\mathrm{\\delta}^2H$.
+For each ERD model only those species have an estimate of ERD that have passed the test of Pearson's $r$ > 0.71 during ERD parameterisation.
+The number of species with an ERD estimate as well as $\\mathrm{\\delta}^2H$ for each model are shown in inset, along with the goodness-of-fit ($R$^2^) and $p$-values for a linear relationship between ERD and $\\mathrm{\\delta}^2H$.
 ERD models in panels A-E correspond to Eq.
@@
 
-```{r erd-time-evg, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time-evg}Modelled species Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\\mathrm{\\Psi}_{\textrm{soil},z}$ was below $\\mathrm{\\Psi}$~crit~, defined as $\\mathrm{\\Psi}$~80,leaf~, over 1990-2015, where $z$ is the soil depth matching species ERD--for ", nrow(mrate.depth.mean %>% subset(deciduous == "E")), " canopy species found on Barro Colorado Island.
+```{r erd-time-evg, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time-evg}Modelled species Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\\mathrm{\\Psi}_{\\textrm{soil},z}$ was below $\\mathrm{\\Psi}$~crit~, defined as $\\mathrm{\\Psi}$~80,leaf~, over 1990-2015, where $z$ is the soil depth matching species ERD--for ", nrow(mrate.depth.mean %>% subset(deciduous == "E")), " canopy species found on Barro Colorado Island.
 Each circle represents a species.
-Circles are colored and jittered vertically around the named depths (ticks) for clarity.")}
+Circles are colored and slightly jittered horizontally for clarity.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr_evergreen.jpeg"))

### 2020-10-26 old_ms_sections.Rmd
moved in lots of text to fit in word count limit

@@
 
+To estimate effective rooting depth, we tested alternate empirical models for species 5-yrly census growth (Eq.
+\@ref(eq:M-chosen) and those in SI), which use effective rooting depth as an input.
+Many intrinsic and extrinsic factors limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019].
+Rhizosphere soil water potential ($\mathrm{\Psi}$~soil~) down-regulates $\mathrm{\Psi}$~leaf~ [@Christoffersen:2016kd2d] such that in moist forests $\mathrm{\Psi}$~leaf~ equilibrates with $\mathrm{\Psi}$~soil~ at night, declining through the day.
+Using species-specific leaf hydraulic vulnerability curves, $\mathrm{\Psi}$~leaf~ can be converted to $K_\textrm{leaf}$, which in turn controls growth [@Sack.2012].
+We therefore used maximum diurnal ($K_\textrm{leaf}$) determined from $\mathrm{\Psi}$~soil~ as a proxy for the soil water limitation on growth.
+$K_\textrm{leaf}$ in our models thus incorporates the effects of rooting depth and rhizosphere water potential.
+This approach, however, does not account for the buffering effect of stem water storage capacitance on $\mathrm{\Psi}$~leaf~, which may cause the model to slightly overestimate ERD (see Discussion).
+VPD affects growth non-linearly, with growth increasing with VPD upto a VPD threshold, then declining as leaf pores (stomata) close [@Grossiord.2020].
+Apart from stomatal control, leaf-drop may further limit water-uptake and growth.
+We therefore regressed species-specific growth against multiplicative or additive effects of one or more factors calculated daily and averaged over 5-yrly censuses: VPD, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1.
+Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting.
+At our study site, due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season, the impact of VPD may as well be a combined impact of not only temperature and humidity, but also radiation.
+
 For each species $s$ we calculated mean mortality rate across all intervals as $\textrm{Mort}_{s,\textrm{mean}} = \frac{1}{6}\sum_{t=1}^{6} {\textrm{Mort}_{s,t}}$.
@@
 
-From Discussion first para
+#Discussion
 
@@
 @Condit.2017 showed that half of the extreme dry seasons (< 1 SD annual water deficit) during 1931-2015 (85-yr) occurred during 1982-2005 (24-yr) of the 34-yr long study period (1982-2015).
+
+Our result of deep rooted species being more hydraulically vulnerable is consistent with observations for a seasonal forest in the Amazon .
+@Johnson:2018fad found that mortality decreased with deep rooting even though deep rooted species were more vulnerable to xylem cavitation (high $\mathrm{\Psi}$~88,stem~) and had narrower safety-margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~).
+@Nardini:2015 found that a species sourcing deep water in a Karst system showed lower hydraulic failure and mortality in an extreme drought despite having high $\mathrm{\Psi}$~88,stem~ and narrower safety margin compared to two species sourcing shallow-water.
+
+at Mudumalai is nearly one (Penman-Monteith PET = 980 $\pm$ 64 mm)
+and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (Penman-Monteith PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (Penman-Monteith PET = `r round(mean(clim$pet.PM))` $\pm$ `r round(sd(clim$pet.PM))` mm;`r min(clim$Year)`-`r max(clim$Year)`; this study, @Guo.2016),
+
+\
+\   We found a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA and WSG) and parameters of leaf vulnerability (or percent loss of conductance, PLC) curves, a 'hard' trait, based on one of the largest dataset for the tropics ($n$ = `r length(sp.exp.param$sp)`).
+This allowed us to scale-up the leaf vulnerability curves at the community scale.
+We find that leaf conductance declined more rapidly with declining $\mathrm{\Psi}$~leaf~ for species with higher values of WSG and LMA (Fig.
+S\@ref(fig:obs-plc-curves)).
+Whether such scaling relationships between hard and soft traits hold true for other forests, and their wider applicability, remains to be seen as data on PLC curves are limited in tropical forests, but supportive and counter examples both exist [@McCulloh.2011; @Hoffmann.2011].
+WSG alone explained `r round(b.wsg.r2*100,0)`% of the variance in $B$ ($p$ `r  b.wsg.p`).
+
+However, trees' stem diameter change is a sum total of not just growth, the addition of wood, but also changes due to reversible stem dehydration (xylem, phloem and bark), the importance of which increases with temporal resolution [@Chitra-Tarak.2016; @Chitra-Tarak.2015].
+
+Use of dendrometer band data would thus entail accounting for reversible stem dehydration
+
+This would also open the possibility of testing, whether deciduous species are more shallow rooted than evergreen species as generally observed [@Fan:2017]
+
+, and suggests that greater emphasis should be given to predictive modeling of tree root allocation and distribution (e.g. @Wijk.2001) as well as trees' hydrological environments
+
+\
+\ Finally, rooting depths may be parametarized within DGVMs such as ELM-FATES using plant hydrodynamic schemes [@Christoffersen:2016kd2d; @Koven.2020], simultaneously modeling plant water-availability with plant water-extraction.
+In general, including differences in tree rooting depths in ELM-FATES would improve understanding of plant competition for water (and other resources) and more accurately represent plant's drought strategies, leading to better predictions of existing and future forest function and composition.
 

### 2020-10-26 manuscript.rmd
1. edited symbols. 1. mfac figures now shown by censuses

@@
 \begin{equation}
-\Psi_\textrm{soil} = \Psi_e\left(\frac{\theta}{\theta_{s}}\right)^{-b}
+\mathrm{\theta}_\textrm{soil} = \Psi_e\left(\frac{\mathrm{\theta}}{\mathrm{\theta}_{s}}\right)^{-b}
 (\#eq:soil-ret-curve)
@@
 We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq.
-\@ref(eq:water-balance), namely, (i) evapotranspiration (ET) from the eddy flux tower ($\cong$ $E_{v} + E_{g}$; 2012-2017; see [SI](#SIData.matteo))), (ii) stream discharge ($q_\textrm{run}$) from the Conrad catchment ($\cong$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC or $\theta_{z}$) $\cong$ $W_{\textrm{liq,} z}$ from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we obtained with a vertical time domain reflectometry (TDR) probe over 0-15 cm depth ($\theta_\textrm{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m with 1299 samples in total covering all soil types and habitats (see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter,  $\theta_{\textrm{0-0.15 m}}$, $\theta_{\textrm{0.4 m}}$ and $\theta_{\textrm{1 m}}$ refer to plot-wide averages).
+\@ref(eq:water-balance), namely, (i) evapotranspiration (ET) from the eddy flux tower ($\cong$ $E_{v} + E_{g}$; 2012-2017; see [SI](#SIData.matteo))), (ii) stream discharge ($q_\textrm{run}$) from the Conrad catchment ($\cong$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC or $\theta_{z}$) $\cong$ $W_{\textrm{liq,} z}$ from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we obtained with a vertical time domain reflectometry (TDR) probe over 0-15 cm depth ($\mathrm{\theta}_\textrm{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m with 1299 samples in total covering all soil types and habitats (see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter,  $\mathrm{\theta}_{\textrm{0-0.15 m}}$, $\mathrm{\theta}_{\textrm{0.4 m}}$ and $\mathrm{\theta}_{\textrm{1 m}}$ refer to plot-wide averages).
 
@@
 \begin{split}
-\overline{O} = \frac{1}{6}\sum \textrm{RMSE}^{*}\textrm{(ET,}E_{v} + E_{g}) + \textrm{RMSE}^{*}(q_{\textrm{run}},q_{\textrm{over}} + q_{\textrm{drai}}) + \textrm{RMSE}^{*}(\theta_{\textrm{0-0.15~m}}, {W_{\textrm{liq,}\overline{z}}}) \\
-& + \textrm{RMSE}^{*}(\theta_{\textrm{0.15~m}}, W_{\textrm{liq,`r soil.depths[4]`m}}) + \textrm{RMSE}^{*}(\theta_{\textrm{0.4~m}}, W_{\textrm{liq,`r soil.depths[6]`~m}}) + \textrm{RMSE}^{*}(\theta_{1~m}, W_{\textrm{liq,`r soil.depths[8]`~m}})
+\overline{O} = \frac{1}{6}\sum \textrm{RMSE}^{*}\textrm{(ET,}E_{v} + E_{g}) + \textrm{RMSE}^{*}(q_{\textrm{run}},q_{\textrm{over}} + q_{\textrm{drai}}) + \textrm{RMSE}^{*}(\mathrm{\theta}_{\textrm{0-0.15~m}}, {W_{\textrm{liq,}\overline{z}}}) \\
+& + \textrm{RMSE}^{*}(\mathrm{\theta}_{\textrm{0.15~m}}, W_{\textrm{liq,`r soil.depths[4]`m}}) + \textrm{RMSE}^{*}(\mathrm{\theta}_{\textrm{0.4~m}}, W_{\textrm{liq,`r soil.depths[6]`~m}}) + \textrm{RMSE}^{*}(\mathrm{\theta}_{1~m}, W_{\textrm{liq,`r soil.depths[8]`~m}})
 (\#eq:obj-func)
@@
 
-based on monthly sums for both ET and $q$~run~ and daily averages for $\theta$s.
+based on monthly sums for both ET and $q$~run~ and daily averages for $\mathrm{\theta}$s.
 $W_{\textrm{liq,}\overline{z}}$ is the average of daily $W_{\textrm{liq,} z}$, where $z$ $\in$ $Z$, and $Z$ = {`r paste(soil.depths[1:4], collapse=", ")`} m.
@@
 \   Preliminary analyses showed that minimizing the objective function $\overline{O}$, that is, simultaneously fitting the model to all states and fluxes did not capture the soil moisture dynamics at 1 m depth well, our key priority for model calibration.
-To ensure model fit at 1 m depth, we first chose 300 parameter ensemble members that minimized $\textrm{RMSE}^{*}(\theta_{\textrm{1 m}}, W_{\textrm{liq,`r soil.depths[8]` m}})$ (< 0.2), and among those 300 ensemble members we chose 100 with the lowest values for the objective function $\overline{O}$.
+To ensure model fit at 1 m depth, we first chose 300 parameter ensemble members that minimized $\textrm{RMSE}^{*}({\mathrm{\theta}_{\textrm{1 m}}, W_{\textrm{liq,`r soil.depths[8]` m}})$ (< 0.2), and among those 300 ensemble members we chose 100 with the lowest values for the objective function $\overline{O}$.
 See Table \@ref(tab:param) for the ranges of best-fit values for different parameters.
@@
 Many factors limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019].
-Because pre-dawn $\mathrm{\Psi}$~leaf~ approaches $\mathrm{\Psi}$~soil~, we used the latter in leaf hydraulic vulnerability curves to predict maximum diurnal $K_\textrm{leaf}$, which in turn controls growth (see Table \@ref(tab:symbols) for trait symbols and their definitions).
+Because pre-dawn $\mathrm{\Psi}$~leaf~ approaches $\mathrm{\Psi}$~soil~, we used the latter in leaf hydraulic vulnerability curves to predict maximum diurnal $K$~leaf~, which in turn controls growth (see Table \@ref(tab:symbols) for trait symbols and their definitions).
 $K_\textrm{leaf}$ in our models thus incorporates the effects of rooting depth and rhizosphere water potential.
@@
 \begin{equation}
-K_{\textrm{leaf},s,i|z,h} = A_{s}e^{-B_{s}\Psi_{\textrm{soil},i|z,h}}
+K_{\textrm{leaf},s,i|z,h} = A_{s}e^{-B_{s}\mathrm{\Psi}_{\textrm{soil},i|z,h}}
 (\#eq:k-soil)
@@
 \begin{equation}
-K_{leaf,s} = A_{s}e^{-B_{s}\Psi_{leaf_{s}}} + \epsilon_{s}
+K_{\textrm{leaf},s}} = A_{s}e^{-B_{s}\mathrm{\Psi}_{\textrm{leaf}_{s}}} + \epsilon_{s}
 (\#eq:vc-exp)
@@
 Soil water availability was not modeled for 1985-1990.
-Analysis of a species' time spent under significant soil drought (proportion of days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) showed that species with shallower ERD spent greater proportion of days below critical hydraulic thresholds (Fig.
+Analysis of a species' time spent under significant soil drought (proportion of days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) showed that species with shallower ERD spent greater proportion of days below critical hydraulic thresholds and increasingly so over 1995-2000, 2000-2005 and 2005-2010 (Fig.
 \@ref(fig:erd-time); S\@ref(fig:erd-time-evg)).
@@
 Deeper ERD species on average had more hydraulically efficient, but risky, suite of traits optimized for an environment of more reliable and greater water source at depth that resulted in a lower mortality risk.
-By contrast shallow ERD species had inefficient, but safer, xylem likely optimized for high variability in $\mathrm{\Psi}_{soil}$ in the shallow soil layers, and resulting in higher variance in mortality rates across censuses as $\mathrm{\Psi}_{soil,~z = ERD}$ crossed ERD $\mathrm{\Psi}_{crit}$ (Fig.
+By contrast shallow ERD species had inefficient, but safer, xylem likely optimized for high variability in $\mathrm{\Psi}_\textrm{soil}$ in the shallow soil layers, and resulting in higher variance in mortality rates across censuses as $\mathrm{\Psi}_{\textrm{soil},z\textrm{ = ERD}}$ crossed ERD $\mathrm{\Psi}_\textrm{crit}$ (Fig.
 \@ref(fig:erd-mort-census-evg); Fig.
@@
 We interpret this result as that VPD and leaf hydraulic vulnerability curves may be adequate to explain inter-census differences in growth (via stomatal control), but also acknowledge that our estimates of seasonality of LAI is a tentative estimate combining leaf-fall data and the timing of leaf-gain backtracked from leaf lifetime, and omits inter-annual variation.
-Future studies may further develop models of LAI seasonality and make use of data from new technologies such as drone based monitoring of LAI in species-rich tropical forests [@Park.2019].
+Future studies may improve models of LAI seasonality and make use of data from new technologies such as drone based monitoring of LAI in species-rich tropical forests [@Park.2019].
 
@@
 \
-Figure \@ref(fig:erd-time): Modeled species Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) versus time spent below critical hydraulic threshold--% days for which $\mathrm{\Psi}_{\textrm{soil},z}$ was below $\mathrm{\Psi}$~crit~, defined as $\mathrm{\Psi}$~50,leaf~, over 1990-2015, where $z$ is the soil depth matching species ERD--for `r length(erd.mrate.sp)` canopy tree species on Barro Colorado Island.
+Figure \@ref(fig:erd-time): Modeled species Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) versus time spent below critical hydraulic threshold--% days for which $\mathrm{\Psi}_{\textrm{soil},z}$ was below $\mathrm{\Psi}$~crit~, defined as $\mathrm{\Psi}$~50,leaf~, over each of the five census intervals (1990-2015), where $z$ is the soil depth matching species ERD--for `r length(erd.mrate.sp)` canopy tree species on Barro Colorado Island.
 Each circle represents a species.
@@
 ```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time}")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr.jpeg"))
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mfac vs. rdi.gr.jpeg"))
 ```
@@
 
-```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($Psi$~soil,z~) by depth $z$ (m; panels) over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines).
+```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($\\Psi_{\\textrm{soil},z}$) by depth $z$ (m; panels) over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines).
 Vertical gray lines demarcate median dates of the demographic censuses at the BCI 50 ha plot."}
@@
 
-```{r erd-time-evg, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time-evg}Modelled species Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\\mathrm{\\Psi}_{\\textrm{soil},z}$ was below $\\mathrm{\\Psi}$~crit~, defined as $\\mathrm{\\Psi}$~80,leaf~, over 1990-2015, where $z$ is the soil depth matching species ERD--for ", nrow(mrate.depth.mean %>% subset(deciduous == "E")), " canopy species found on Barro Colorado Island.
+```{r erd-time-evg, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time-evg}Modelled species Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\\mathrm{\\Psi}_{\\textrm{soil},z}$ was below $\\mathrm{\\Psi}$~crit~, defined as $\\mathrm{\\Psi}$~80,leaf~, over five census intervals (1990-2015), where $z$ is the soil depth matching species ERD--for ", nrow(mrate.depth.mean %>% subset(deciduous == "E")), " canopy species found on Barro Colorado Island.
 Each circle represents a species.
 Circles are colored and slightly jittered horizontally for clarity.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr_evergreen.jpeg"))
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mfac vs. rdi.gr_evergreen.jpeg"))
 ```

### 2020-10-26 manuscript.rmd
Corrected spellings, symbols, para spacing

@@
 
-\
-\
 Item | Count
@@
 # Introduction
-
+\
 Tropical forests play a disproportionately large role in the global carbon and energy cycles [@Bonan:2008], and support half of global biodiversity [@Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019].
@@
 \
-\     Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation to stem water storage, deep water access, increased root production, hydraulic redistribution and embolism resistance [@McDowell.2008].
+Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation to stem water storage, deep water access, increased root production, hydraulic redistribution and embolism resistance [@McDowell.2008].
 Deep water access is arguably the most long-standing, yet under-studied mechanism.
@@
 \
-\     Trade-offs in plant 'traits' shape how plants make a living in their environment leading to alternative ecological strategies.
+  Trade-offs in plant 'traits' shape how plants make a living in their environment leading to alternative ecological strategies.
 Characterizing the essential constraints necessary to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010; @Bartlett.2019] entails identifying the topography of trait trade-offs in different environments, and how these relate to demographic rates (growth, recruitment and mortality).
@@
 \
-\      We tested an overarching hypothesis that inter-specific variation in rooting depths enables increasingly reliable water availability with depth and thus mitigation of hydraulic vulnerability.
+   We tested an overarching hypothesis that inter-specific variation in rooting depths enables increasingly reliable water availability with depth and thus mitigation of hydraulic vulnerability.
 Our hypotheses are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallow rooted species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deep rooted species have lower mortality rates in droughts due to low realized risk of hydraulic failure, whereas shallow rooted species may have higher mortality rates as a result of higher variability in water availability in shallow soil layers.
@@
 \
-\   We developed a novel, empirical inverse model to estimate species effective rooting depths (ERD) for large trees of `r length(erd.sp)` species (Fig.
+We developed a novel, empirical inverse model to estimate species effective rooting depths (ERD) for large trees of `r length(erd.sp)` species (Fig.
 \@ref(fig:schematic)), defined as the depth at which soil water potential best explained species-specific growth dynamics.
@@
 We evaluated ERD against existing deuterium to hydrogen content data for a subset of six tree species as an independent observation of ERD.
-We tested relationship of ERD with above-ground hydraulic traits, namely, xylem vulnerability to embolism, branch hydraulic safety margins, leaf turgor loss point and maximum stem-specific hydraulic conductivity, for a subset of eight species with both datasets.
+We tested the relationship of ERD with above-ground hydraulic traits, namely, xylem vulnerability to embolism, branch hydraulic safety margins, leaf turgor loss point and maximum stem-specific hydraulic conductivity, for a subset of eight species with both datasets.
 Finally, we analysed species' mortality rates over 35 years to test whether ERD explained risk of drought-induced mortality.
@@
 ## Study site description
-
+\
 We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, hydrology and plant functional ecology, including data for stable water isotopes on trees and soils [@Meinzer:1999].
@@
 \
-\     In the 50 ha plot, all stems above ten mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every five year through 1985-2015 for growth, mortality as well as recruitment of new stems into the ten mm dbh size class [@Condit.2017].
+  In the 50 ha plot, all stems above ten mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every five year through 1985-2015 for growth, mortality as well as recruitment of new stems into the ten mm dbh size class [@Condit.2017].
 This inventory represents 321 woody species across the eight censuses.
-Aside from a two ha patch of ~100-year-old secondary forest at the north-eastern corner of the plot the remaining 48ha represent old growth forest that was never clear cut for agriculture.
+Aside from a two ha patch of ~100-year-old secondary forest at the north-eastern corner of the plot the remaining 48ha represent old growth forest that was never cleared for agriculture.
 The plot (1000 m $\times$ 500 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001].
@@
 ## Model ELM-FATES for soil matric potentials
-
+\
 ### Model description 
-
+\
 To estimate dynamics of water availability by depth over the forest’s rooting zone, we used the Energy Exascale Earth System Land Model (ELM), coupled with the Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES).
@@
 \
-\     ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\mathrm{\Delta}W_{\textrm{can}}$, surface water $\mathrm{\Delta}W_{\textrm{sfc}}$, soil water $\mathrm{\Delta}W_{\textrm{liq},z}$ at depth $z$ and water in an unconfined aquifer $\mathrm{\Delta} W_{a}$ (omitting processes relevant to snow, wetlands or lakes).
+  ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\mathrm{\Delta}W_{\textrm{can}}$, surface water $\mathrm{\Delta}W_{\textrm{sfc}}$, soil water $\mathrm{\Delta}W_{\textrm{liq},z}$ at depth $z$ and water in an unconfined aquifer $\mathrm{\Delta} W_{a}$ (omitting processes relevant to snow, wetlands or lakes).
 Conservation of these terms is expressed as follows:
@@
 \end{equation}
-where, $q_{\textrm{rain}}$ is rainfall, $E_{v}$ is transpiration of the forest, $E_{g}$ is ground evaporation, $q_{\textrm{over}}$ is surface runoff, $q_{\textrm{h2osfc}}$ is runoff from surface water storage, $q$~drai~ is sub-surface drainage (all in mm H~2~O) and $\mathrm{\Delta}t$ is the time-step (here, 1 hr).
+where, $q_{\textrm{rain}}$ is rainfall, $E_{v}$ is transpiration of the forest, $E_{g}$ is ground evaporation, $q_{\textrm{over}}$ is surface runoff, $q_{\textrm{h2osfc}}$ is runoff from surface water storage, $q$~drai~ is subsurface drainage (all in mm H~2~O) and $\mathrm{\Delta}t$ is the time-step (here, 1 hr).
 The soil profile is discretized into 15 exponentially distributed soil layers by depth $z$, where $z$ $\in$ $Z$, and $Z$ = {0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13, 21, 35} m.
@@
 ### Model parameterization
-
+\
 To parameterize ELM-FATES, we identified eleven parameters relevant for Eq.
@@
 \begin{equation}
-\mathrm{\theta}_\textrm{soil} = \Psi_e\left(\frac{\mathrm{\theta}}{\mathrm{\theta}_{s}}\right)^{-b}
+\mathrm{\theta}_{\textrm{soil}} = \Psi_e\left(\frac{\mathrm{\theta}}{\mathrm{\theta}_{s}}\right)^{-b}
 (\#eq:soil-ret-curve)
@@
 \
-\   We ran ELM-FATES for the 5000-member parameter ensemble with hourly climate drivers measured at the BCI meteorological station for the period from 1985-2018 (Faybishenko et al.
+We ran ELM-FATES for the 5000-member parameter ensemble with hourly climate drivers measured at the BCI meteorological station for the period from 1985-2018 (Faybishenko et al.
 2020) initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode.
@@
 \
-\    For ELM-FATES calibration we calculated an objective function for each of the 5000-member ensemble by equally weighting standardized RMSE between observations and simulations across all fluxes and states.
+ For ELM-FATES calibration we calculated an objective function for each of the 5000-member ensemble by equally weighting standardized RMSE between observations and simulations across all fluxes and states.
 We describe the objective function $O$ as :
@@
 
-\   Preliminary analyses showed that minimizing the objective function $\overline{O}$, that is, simultaneously fitting the model to all states and fluxes did not capture the soil moisture dynamics at 1 m depth well, our key priority for model calibration.
-To ensure model fit at 1 m depth, we first chose 300 parameter ensemble members that minimized $\textrm{RMSE}^{*}({\mathrm{\theta}_{\textrm{1 m}}, W_{\textrm{liq,`r soil.depths[8]` m}})$ (< 0.2), and among those 300 ensemble members we chose 100 with the lowest values for the objective function $\overline{O}$.
+Preliminary analyses showed that minimizing the objective function $\overline{O}$, that is, simultaneously fitting the model to all states and fluxes did not capture the soil moisture dynamics at 1 m depth well, our key priority for model calibration.
+To ensure model fit at 1 m depth, we first chose 300 parameter ensemble members that minimized $\textrm{RMSE}^{*}({\mathrm{\theta}_{\textrm{1 m}}}, W_{\textrm{liq,`r soil.depths[8]` m}})$ (< 0.2), and among those 300 ensemble members we chose 100 with the lowest values for the objective function $\overline{O}$.
 See Table \@ref(tab:param) for the ranges of best-fit values for different parameters.
@@
 We evaluated ELM-FATES by calculating RMSE between simulated and observed long-term daily VWC for the depths of 0.1, 0.4 and 1 m (2016-2018), based on a dataset we had left-out during calibration.
-We obtained these observations from three horizontal TDR probes at the depths of 0.1, 0.4 and 1 m at a point location near vertical probe (see [SI](#SIData.VWC)).
+We obtained these observations from three horizontal TDR probes at the depths of 0.1, 0.4 and 1 m at a point location near the vertical probe (see [SI](#SIData.VWC)).
 
-## Hydrological droughts
+### Hydrological droughts
 
@@
 ## Model for effective rooting depth (ERD)
-
+\
+### Model description
+\
 To estimate effective rooting depth, we tested alternate empirical models for species 5-yearly census growth (Eq.
@@
 \
-\    For the chosen model (see Model validation and selection below), daily average growth for species $s$, $G_{s,t|z}$, in the census interval $t$ for depth $z$ and for a given hydrological realization $h$ is described as follows:
+ For the chosen model (see Model validation and selection below), daily average growth for species $s$, $G_{s,t|z}$, in the census interval $t$ for depth $z$ and for a given hydrological realization $h$ is described as follows:
 
@@
 \
-\    To account for the non-linear impact of VPD on growth, we used a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see SI) and used GPP standardized between 0 and 1 (hereafter, VPD*) in Eq.
+ To account for the non-linear impact of VPD on growth, we used a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see SI) and used GPP standardized between 0 and 1 (hereafter, VPD*) in Eq.
 \@ref(eq:M-chosen).
@@
 \
-\    For each species, we estimated diurnal maximum value of $K_{\textrm{leaf},s,i|z,h}$ for a species $s$ by substituting $\mathrm{\Psi}_{\textrm{soil},s,i|z,h}$in the leaf vulnerability curves defined below in Eq.
+ For each species, we estimated diurnal maximum value of $K_{\textrm{leaf},s,i|z,h}$ for a species $s$ by substituting $\mathrm{\Psi}_{\textrm{soil},s,i|z,h}$in the leaf vulnerability curves defined below in Eq.
 \@ref(eq:vc-exp) using species-specific parameters $A_{s}$ and $B_{s}$ as:
@@
 \begin{equation}
-K_{\textrm{leaf},s,i|z,h} = A_{s}e^{-B_{s}\mathrm{\Psi}_{\textrm{soil},i|z,h}}
+K_{\textrm{leaf},s,i|z,h} = A_{s}e^{-B_{s}{\mathrm{\Psi}}_{\textrm{soil},i|z,h}}
 (\#eq:k-soil)
 \end{equation}
-\    By varying $z$, we estimated species effective rooting depths (ERD), defined as the depth $z$ at which soil water dynamics ($\mathrm{\Psi}_{\textrm{soil,} z}$) best explained observed dynamics of species growth.
+ By varying $z$, we estimated species effective rooting depths (ERD), defined as the depth $z$ at which soil water dynamics ($\mathrm{\Psi}_{\textrm{soil,} z}$) best explained observed dynamics of species growth.
 Using Eq.
@@
 ### Model parameterization
-
+\
 #### Growth data
@@
 \begin{equation}
-K_{\textrm{leaf},s}} = A_{s}e^{-B_{s}\mathrm{\Psi}_{\textrm{leaf}_{s}}} + \epsilon_{s}
+K_{\textrm{leaf},s} = A_{s}e^{-B_{s}{\mathrm{\Psi}}_{\textrm{leaf}_{s}}} + \epsilon_{s}
 (\#eq:vc-exp)
@@
 \
-\   As these fits explained a large proportion of variation in parameters $A$ and $B$ (see results), we sequentially used Eq.
+As these fits explained a large proportion of variation in parameters $A$ and $B$ (see results), we sequentially used Eq.
 \@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r sp.n.model` ERD species without direct data and estimated leaf hydraulic vulnerability curves using Eq.
@@
 \
-\   We obtained species-specific parameters $A$ & $B$ for Eq.
+We obtained species-specific parameters $A$ & $B$ for Eq.
 \@ref(eq:k-soil) either from data when available, or a model based on LMA or WSG as described in the earlier section, and used K standardized between 0 and 1 within species (K*) in Eq.
@@
 We used $\mathrm{\delta}$^2^H~xylem~ as an independent observation to validate the ERD models.
-As root water-uptake is generally a non-fractionating process, tree $\mathrm{\delta}$^2^H~xylem~ reflects signature of source water.
-Given a vertical gradient of $\mathrm{\delta}$^2^H in soil and ground water $\mathrm{\delta}$^2^H~xylem~ provides an index of rooting depth [@Dawson.1991].
-We leveraged $\mathrm{\delta}$^2^H~xylem~ from BCI [@Meinzer:2001] for the dry-season of March 1997 as this period showed largest seasonal divergence in $\mathrm{\delta}$^2^H~xylem~ among species and in soil and ground water $\mathrm{\delta}$^2^H at natural abundance level (Fig.
+As root water-uptake is generally a non-fractionating process, tree $\mathrm{\delta}$^2^H~xylem~ reflects the signature of source water.
+Given a vertical gradient of $\mathrm{\delta}$^2^H in soil and groundwater $\mathrm{\delta}$^2^H~xylem~ provides an index of rooting depth [@Dawson.1991].
+We leveraged $\mathrm{\delta}$^2^H~xylem~ from BCI [@Meinzer:2001] for the dry-season of March 1997 as this period showed largest seasonal divergence in $\mathrm{\delta}$^2^H~xylem~ among species and in soil and groundwater $\mathrm{\delta}$^2^H at natural abundance level (Fig.
 S\@ref(fig:iso-data)).
 Analysing $\mathrm{\delta}$^2^H~xylem~ for 12 tree species @Meinzer:2001 found that four species sourced water between 1 m and ground water, two other at least 0.8 m deep, one between 0 to 0.2 m, and five in the 0.2 to 0.7 m depth range over which the relationship between soil depth and $\mathrm{\delta}$^2^H~soil~ was variable.
+
 \
-\   We removed those species that would be leafless in March 1997 because xylem sap $\mathrm{\delta}$^2^H for these species may not be linked with water sourced at the time of measurement and thus may be compared to species that had leaves.
+We removed those species that would be leafless in March 1997 because xylem sap $\mathrm{\delta}$^2^H for these species may not be linked with water sourced at the time of measurement and thus may be compared to species that had leaves.
 We also removed one species whose $\mathrm{\delta}$^2^H~xylem~ was in the uncertain depth range.
@@
 ## Relationship of ERD with above-ground hydraulic traits
-
+\
 To evaluate whether ERD was associated with above-ground hydraulic traits (eight species with both datasets), we regressed species ERD against maximum stem hydraulic conductance ($K_\textrm{max,stem}$), leaf turgor loss point ($\mathrm{\Psi}$~tlp~), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\mathrm{\Psi}$~88,stem~), and hydraulic safety margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~).
@@
 ## Mortality analyses and species-specific drought experience
-
+\
 To test whether ERD plays a role in alleviating mortality risk, we calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$ 30 m) and average abundance of at least 20 trees ($n$ = `r length(erd.mrate.sp)`).
@@
 \
-\   We regressed species ERD against species $\textrm{M}_{s,t}$ for each census interval $t$ with ($n$ = `r nrow(mrate.depth.mean)`) and without deciduous species ($n$ = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`).
+We regressed species ERD against species $\textrm{M}_{s,t}$ for each census interval $t$ with ($n$ = `r nrow(mrate.depth.mean)`) and without deciduous species ($n$ = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`).
 To assess whether deeper ERD species escaped the experience of significant soil drought, we calculated for each species the proportion of days in the simulation period (1990-2015) during which soil water potential in the soil layer matching species ERD ($\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$) were below species $\mathrm{\Psi}$~crit~.
@@
 \
-All statistical analyses were conducted in R statistical environment (version 3.5.3, @R:2019).
+All statistical analyses were conducted in the *R* statistical environment (version 3.5.3, @R:2019).
 
@@
 
-\   Soil Water Potential dynamics ($\mathrm{\Psi}_{\textrm{soil},z}$) by depth $z$ (from 0.1 through 13 m) obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of extreme years, drought intensity at depth, drought seasonality and duration (Fig.
+\
+Soil Water Potential dynamics ($\mathrm{\Psi}_{\textrm{soil},z}$) by depth (from 0.01 through 13 m) obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of extreme years, drought intensity at depth, drought seasonality and duration (Fig.
 \@ref(fig:swp-droughts)).
@@
 ## Leaf hydraulic vulnerability curves
-
+\
 Parameters of leaf vulnerability curves were predictable from WSG and LMA.
@@
 ## Effective rooting depths (ERD)
-
+\
 The chosen ERD model (with a multiplicative effect of VPD and not LAI; Eq.
@@
 \
-\   Modeled effective rooting depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from `r min(df.erd.to.plot$depth)` m to `r max(df.erd.to.plot$depth)` m (Fig.
+Modeled effective rooting depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from `r min(df.erd.to.plot$depth)` m to `r max(df.erd.to.plot$depth)` m (Fig.
 \@ref(fig:sp-erd)).
@@
 ## Relationship of ERD with above-ground hydraulic traits
-
+\
 Species with deeper ERD showed greater maximum stem hydraulic conductance ($K_\textrm{max,stem}$, Spearman's' $r$ = `r cor.k`, $p$ = `r p.k`; Fig.
@@
 ## Effective rooting depths, experienced drought and mortality
-
+\
 Among evergreen species, ERD explained `r m.evg.r2.90`, `r m.evg.r2.05` and `r m.evg.r2.10`% variation in mortality rates during census intervals 1985-1990, 2000-2005 and 2005-2010, respectively, such that species with mortality rates significantly decreased with ERD ($p$ = `r m.evg.p.90`, `r m.evg.p.05` and `r m.evg.p.10`, respectively; Fig.
@@
 Soil water availability was not modeled for 1985-1990.
-Analysis of a species' time spent under significant soil drought (proportion of days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) showed that species with shallower ERD spent greater proportion of days below critical hydraulic thresholds and increasingly so over 1995-2000, 2000-2005 and 2005-2010 (Fig.
+Analysis of a species' time spent under significant soil drought (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) showed that species with shallower ERD spent a greater proportion of days below critical hydraulic thresholds and increasingly so over 1995-2000, 2000-2005 and 2005-2010 (Fig.
 \@ref(fig:erd-time); S\@ref(fig:erd-time-evg)).
@@
 ### Realized drought risk integral to assessing drought-vulnerability
-
+\
 Our study brings attention to the need for assessing drought sensitivity in terms of species realized drought exposure by accounting for the actual hydrological drought and trees' deep water access, beyond the universal drought indices and trees' above-ground hydraulic vulnerability.
@@
 \
-\   Our result--of lower mortality in deep rooted trees--contrasts with the finding of @Chitra-Tarak.2018 (hereafter, CT2018) in which inversely modelled, deeper ERD species died *more* in a rare, prolonged drought.
+Our result--of lower mortality in deep rooted trees--contrasts with the finding of @Chitra-Tarak.2018 (hereafter, CT2018) in which inversely modelled, deeper ERD species died *more* in a rare, prolonged drought.
 The hydrological model in CT2018 revealed that the multi-year drought exhausted the deep soil and even bedrock water availability [also see, @Goulden.2019 and @Ivanov:2012].
@@
 Our chosen ERD model is an advance over the only other inverse model of ERD (@Chitra-Tarak.2018; $n$ = 12 species) as we employ a physiologically meaningful representation, fewer parameters and corroboration with tree and soil data for $\mathrm{\delta}$^2^H.
-We highlight that stable water isotopes for plant and below-ground sources are critical for calibration or validation of ERD model at community scale, although the interpretation of these data is still under research [@Deurwaerder.2020].
+We highlight that stable water isotopes for plant and below-ground sources are critical for calibration or validation of ERD models at community scale, although the interpretation of these data is still under research [@Deurwaerder.2020].
 As in @Chitra-Tarak.2018, we also interpret our ERD estimates as revealing relative differences among species' effective rooting depths rather than in absolute terms.
@@
 \
-\   Our inverse ERD model was parameterized on 5-year growth data, with five data points over a 25-yr period, which decoupled climate events and demographic outcomes.
+Our inverse ERD model was parameterized on 5-year growth data, with five data points over a 25-yr period, which decoupled climate events and demographic outcomes.
 Future studies could better constrain the ERD model with high frequency growth data such as those from dendrometer bands.
@@
 \
-\   In forests such as BCI where understory tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes may need accounting for temporal variation in trees' light environment, an issue we minimized by limiting ERD estimates to canopy trees.
+In forests such as BCI where understory tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes may need accounting for temporal variation in trees' light environment, an issue we minimized by limiting ERD estimates to canopy trees.
 In sites where growth is not water-limited, ERD models may explore parameterization on sap-flux or leaf water-potential dynamics [@Brum.2017].
@@
 \
-\   While our ERD model empirically predicts growth via estimating $\mathrm{\Psi}$~leaf~ from $\mathrm{\Psi}$~soil~ of a specific depth, mechanistic models that account for plant hydrodynamics and other processes influencing growth are likely to predict growth more accurately, and thus ERD [@Christoffersen:2016kd2d; @Sperry.1998].
+While our ERD model empirically predicts growth via estimating $\mathrm{\Psi}$~leaf~ from $\mathrm{\Psi}$~soil~ of a specific depth, mechanistic models that account for plant hydrodynamics and other processes influencing growth are likely to predict growth more accurately, and thus ERD [@Christoffersen:2016kd2d; @Sperry.1998].
 Although data needs for parameterizing such models could be greater (e.g.
@@
 \
-\    To estimate topographic variation in ERD within or across species, future studies may use soil moisture dynamics from a distributed hydrological model [e.g.
+ To estimate topographic variation in ERD within or across species, future studies may use soil moisture dynamics from a distributed hydrological model [e.g.
 @Schwantes:2018].
@@
 # Conclusions
-
+\
 Establishing relationships between environment, traits and demographic outcomes of plants is imperative for developing a predictive plant ecology.
@@
 # Acknowledgements
-
+\
 This research was supported as part of the Next Generation Ecosystem Experiments-Tropics, funded by the U.S.
@@
 RF was supported by the National Center for Atmospheric Research, which is funded by the National Science Foundation.
-**Add acknowledgements for other co-authors.
+**Add acknowledgements for other co-authors**.
 The BCI forest dynamics research project, founded by Stephen P.
@@
 \@ref(eq:vc-exp)) vs.
-those that were predicted from model based on trait-proxies.
+those that were predicted from a set of models based on trait-proxies.
 (Eqs.
@@
 
-# Supplementary material {-}
+# Supplementary Information {-}
 
 ## Data Descriptions 
-
+\
 ### Growth Data 
@@
 For each individual tree ($i$), we calculated absolute growth rates ($G_{i,t}$, cm yr^-1^) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses prorated by the inter-census interval based on the exact date of measurements.
-We eliminated positive outliers as any growth rates amounting to >75 mm yr^-1^, as this was the highest rate observed with confidence for fastest growing species [@Condit.2017].
+We eliminated positive outliers as any growth rates amounting to >75 mm yr^-1^, as this was the highest rate observed with confidence for the fastest growing species [@Condit.2017].
 Negative outliers were removed based on known measurement error, $\sigma_{e} = 0.006214d + 0.903$, where *d* is dbh.
@@
 ### Microclimatic tower {#SIData.matteo} 
-
+\
 The tower used for these measurements is 41 m above ground, on a plateau on BCI [@Pau.2018].
@@
 To train the network, the dataset was randomly divided in a training set (70%), a validation set (15%) and a test set (15%).
-A two-layer feed-forward network with 10 sigmoid hidden neurons and linear output neurons, was trained using the Levenberg-Marquardt algorithm until the mean square error (MSE) of the validation set stop improving [@Hagan.1994].
-Performance, in term of MSE, was evaluated using the test set at the end of the training.
+A two-layer feed-forward network with 10 sigmoid hidden neurons and linear output neurons, was trained using the Levenberg-Marquardt algorithm until the mean square error (MSE) of the validation set stopped improving [@Hagan.1994].
+Performance, in terms of MSE, was evaluated using the test set at the end of the training.
 This procedure was repeated 100 times to produce 100 estimates of GPP.
@@
 $f_{i} = \frac{\sum_{k}f_{ik}^{ANN}/{MSE_{ik}}}{\sum_{k}1/{MSE_{ik}}}$
-
+\
 The ANN was implemented using the Neural Network Toolbox in Matlab 2014a.
@@
 ### Stream discharge from the Conrad catchment {#SIData.steveP} 
-
+\
 The Conrad weir is located on the gently sloping western side of Barro Colorado Island.
@@
 Stream stage is recorded at 5-minute intervals.
-
+\
 The weir drains a catchment of approximately 40 ha of which at least 90% consists of the central plateau area of the island.
@@
 Large storms are sometimes preceded by small, initial peaks generated by the rain falling on the steeper slopes close to the weir.
-
+\
 Streamflow data are analyzed by software custom-built by author SP.
@@
 
+\
 We collected volumetric water content (VWC) data (2012-2018) from a point location at the eddy flux tower with a vertical time domain reflectometry (TDR) probe.
@@
 
+\
 We also leveraged existing GWC measurements by @Kupers:2019 at the depths of 0.15, 0.4 and 1 m in summer 2015 (February, March and April) and 0.15, 0.4 in summer 2016 (March), with 1299 samples in total that covered all soil types and habitats in the plot.
@@
 ### Leaf Area Index 
-
+\
 We generated an estimate of the mean seasonality curve for normalized leaf area index (LAI$^*$, range 0-1, unitless) for individual species as follows.
@@
 S\@ref(fig:leaf-cover)).
-
+\
 For five species, an estimate of leaf lifetime was available from direct observations [@Osnas.2018].
@@
 Brett to edit.
+
 \
@@
 
-\   Canopy branches were cut from 4–10 trees of 26 species across three sites (Table SXX).
+\
+Canopy branches were cut from 4–10 trees of 26 species across three sites (Table SXX).
 The branch segments were enclosed in opaque plastic bags, and brought to the laboratory for measurements.
@@
 
-\   Three leaves that were distal to the $K_h$ segment were sealed in plastic bags and the entire branch was enclosed in black plastic bags for at least 30 minutes to stop transpiration.
+\
+Three leaves that were distal to the $K_h$ segment were sealed in plastic bags and the entire branch was enclosed in black plastic bags for at least 30 minutes to stop transpiration.
 Just prior to measurement of $K_h$, leaf water potential of the three bagged leaves was measured with a pressure chamber.
@@
 
-\   For each $K_h$ segment, three measurements of stem water potential were made on stem sections that were located adjacent to the $K_h$ segment.
+\
+For each $K_h$ segment, three measurements of stem water potential were made on stem sections that were located adjacent to the $K_h$ segment.
 Stem sections that were approximately 40 mm in length were placed in nickel-plated aluminum chambers that were connected to Peltier-type psychrometers (Merrill Instruments, Logan, UT) and placed in a water bath at 30°C for 3 h to equilibrate the humidity within the cambers.
@@
 
-\   The $K_h$ segments were removed from the branches by repeatedly cutting back the branch on both ends while it was submerged in water to relax the hydraulic tension within the xylem ((REFERENCE)).
+\
+The $K_h$ segments were removed from the branches by repeatedly cutting back the branch on both ends while it was submerged in water to relax the hydraulic tension within the xylem ((REFERENCE)).
 Then the ends were shaved clean with a fresh razor.
@@
 (2012), $K_h$ was calculated as the slope of the flow rate versus pressure gradient (i.e., pressure head divided by segment length) among the three pressure heads.
-Stem area specific hydraulic conductivity was ($K_stem$) was calculated by dividing $K_h$ by the cross-sectional area of stem.
+Stem area specific hydraulic conductivity was ($K_\textrm{stem}$) was calculated by dividing $K_h$ by the cross-sectional area of stem.
 To calculate cross-sectional stem area, stem diameter was measured twice perpendicularly on both ends of the segment and averaged, then area was calculated assuming a circular cross section.
 
-\   To assess vulnerability to cavitation, the branches were allowed to dry in the lab for 0 – 180 hours before they were measured for stem water potential and $K_stem$.
+\
+To assess vulnerability to cavitation, the branches were allowed to dry in the lab for 0 – 180 hours before they were measured for stem water potential and $K_stem$.
 We used the water potential measured by the psychrometers because they are a more direct measurement.
@@
 We used WSG, the wood specific gravity after drying at 100^$\circ$^C and LMA, the leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules for leaves receiving direct sunlight as described in @Wright:2010.
+
 \
-\   Among the `r length(sp.exp.param$sp)` species with observed data on ($K_\textrm{leaf}$ vs.
+Among the `r length(sp.exp.param$sp)` species with observed data on ($K_\textrm{leaf}$ vs.
 $\mathrm{\Psi}$~leaf~) used to build the community-level models of leaf vulnerability curves (Eqs.
@@
 ## Alternative models for Effective Rooting Depth (ERD)
-
+\
 Below we present all the alternative models for Effective Rooting Depth (ERD).
@@
 # Supplementary Figures
-
+\
 ```{r ksat-profiles, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ksat-profiles}Generated 5000 curves of soil hydraulic conductivity variation by depth based on Latin Hypercube Sampling within 95% CI of observed data at depths of 12.5 cm and 60 cm [@Kinner:2004], assuming a linear decline between those depths, and values before 12.5 and after 60 cm as that of 12.5 cm and 60 cm, respectively.
@@
 ```
-
+\
 ```{r TDR-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-TDR-calib}Volumetric Water Content vs.
@@
 ```
-
+\
 ```{r dbh-model, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-dbh-model}A scatterplot of tree growth rates (circles) from all census intervals versus dbh at the beginning of the census interval is overlaid with a dbh model of growth (spline).
@@
 ```
-
+\
 ```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "35%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd}Scatterplot of mean midday (11:00 am to 15:00 pm) gross primary productivity (GPP) and mean midday vapor pressure deficit (VPD) measured at the eddy flux tower in Barro Colorado Island over 2012-2017.
@@
 ```
-
+\
 ```{r leaf-cover, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-cover}Seasonal variation in standardised LAI (range 0-1) for ", length(erd.sp)," species at Barro Colorado Island (panels).
@@
 ```
-
+\
 ```{r iso-data, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-iso-data}Stable isotopic ratio ($\\mathrm{\\delta}^2H$) of soil water by depth measured at Barro Colorado Island in March 1997.
@@
 ```
-
+\
 ```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($\\Psi_{\\textrm{soil},z}$) by depth $z$ (m; panels) over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines).
@@
 ```
-
+\
 ```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}Vulnerability curves fitted to observed $K_\\textrm{leaf}$ vs.
@@
 ```
-
+\
 ```{r obs-comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-comm-plc-curves}Percent loss of vulnerability curves for species with ERD estimates.
@@
 ```
-
+\
 ```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-other-erd-mods}Effective Rooting Depth versus $\\mathrm{\\delta}^2H$ for each of the ERD models tested (panels).
@@
 ```
-
+\
 ```{r erd-mort-census, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-census}Species mortality rate in each census interval during the study period plotted against ERD.
@@
 
+\
 ```{r erd-time-evg, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time-evg}Modelled species Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\\mathrm{\\Psi}_{\\textrm{soil},z}$ was below $\\mathrm{\\Psi}$~crit~, defined as $\\mathrm{\\Psi}$~80,leaf~, over five census intervals (1990-2015), where $z$ is the soil depth matching species ERD--for ", nrow(mrate.depth.mean %>% subset(deciduous == "E")), " canopy species found on Barro Colorado Island.
@@
 #Supplementary Tables
-
+\
 ```{r param, results = 'asis', echo = FALSE}
@@
 ```
-
+#
 ```{r eddy, results = 'asis', echo = FALSE}
@@
 ```
-
+#
 ```{r ab-param, results = 'asis', echo = FALSE}
-knitr::kable(ab.table, caption = 'Leaf vulnerability curves parameters A & B, $K_\\textrm{max,leaf}$ and $\\mathrm{\\Psi}$ for the ", length(erd.sp), " species with ERD estimates.
+knitr::kable(ab.table, caption = 'Leaf vulnerability curves parameters A & B, $K_\\textrm{max,leaf}$ and $\\mathrm{\\Psi}$ for the ", `r length(erd.sp)`, " species with ERD estimates.
 Species for which curves were fitted to observed data for $K_\\textrm{leaf}$ vs.

### 2020-10-26 manuscript.rmd
Manuscript draft 2 sent to co-authors

@@
 Summary | 197
-Main Text  | 6500
+Main Text  | 6516
 Introduction | 890
-Methods | 3294
-Results | 804
+Methods | 3300
+Results | 814
 Discussion | 1512
@@
 
+\
 Preliminary analyses showed that minimizing the objective function $\overline{O}$, that is, simultaneously fitting the model to all states and fluxes did not capture the soil moisture dynamics at 1 m depth well, our key priority for model calibration.
@@
 \@ref(fig:hydro-calib)D), slightly underestimating peak discharge in wet years and slightly overestimating peak ET.
-The reduction in parameter range in the best-fit ensembles compared to the tested global ranges showed that the model calibration was sensitive to the Ball-Berry stomatal slope parameter, the ELM root distribution parameter (that regulates the depth of the rooting profile), soil hydraulic conductivity profile, especially at depth, and the adjustment factor that varies soil hydraulic conductivity to account for macroporosity and direct flow paths (Table S\@ref(tab:param)).
+The reduction in parameter range (Table S\@ref(tab:param))  in the best-fit ensembles compared to the tested global ranges showed that the model calibration was sensitive to the Ball-Berry stomatal slope parameter, the ELM root distribution parameter (that regulates the depth of the rooting profile), soil hydraulic conductivity profile, especially at depth, and the adjustment factor that varies soil hydraulic conductivity to account for macroporosity and direct flow paths (Fig.
+S\@ref(tab:sens-sm-et), S\@ref(tab:sens-run)).
 
@@
 
-# Conclusions
+## Conclusions
 \
@@
 # Supplementary Figures
-\
+
 ```{r ksat-profiles, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ksat-profiles}Generated 5000 curves of soil hydraulic conductivity variation by depth based on Latin Hypercube Sampling within 95% CI of observed data at depths of 12.5 cm and 60 cm [@Kinner:2004], assuming a linear decline between those depths, and values before 12.5 and after 60 cm as that of 12.5 cm and 60 cm, respectively.
@@
 ```
-\
+
 ```{r TDR-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-TDR-calib}Volumetric Water Content vs.
@@
 ```
-\
+
 ```{r dbh-model, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-dbh-model}A scatterplot of tree growth rates (circles) from all census intervals versus dbh at the beginning of the census interval is overlaid with a dbh model of growth (spline).
@@
 ```
-\
+
 ```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "35%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd}Scatterplot of mean midday (11:00 am to 15:00 pm) gross primary productivity (GPP) and mean midday vapor pressure deficit (VPD) measured at the eddy flux tower in Barro Colorado Island over 2012-2017.
@@
 ```
-\
+
 ```{r leaf-cover, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-cover}Seasonal variation in standardised LAI (range 0-1) for ", length(erd.sp)," species at Barro Colorado Island (panels).
@@
 ```
-\
+
 ```{r iso-data, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-iso-data}Stable isotopic ratio ($\\mathrm{\\delta}^2H$) of soil water by depth measured at Barro Colorado Island in March 1997.
@@
 ```
-\
+
 ```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($\\Psi_{\\textrm{soil},z}$) by depth $z$ (m; panels) over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines).
@@
 ```
-\
+
+```{r sens-sm-et, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Parameter sensitivity of ELM-FATES soil water content (VWC) is shown for two sample months.
+(A, B)) as well as for ET for another two sample months (C, D), as sensitivity for a particular parameter varies depending on the chosen month.
+Parameters are color coded and are consistent across panels.
+Each point represents a simulation among the 5000-ensemble runs.
+See S\\@ref(tab:param) for description, global ranges and chosen ranges for parameters."}
+knitr::include_graphics(file.path("figures/ELM_FATES_parameter_sensitivity/sens.sm.et.jpeg"))
+```
+
+```{r sens-run, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Parameter sensitivity of ELM-FATES stream discharge (QRUNOFF) is shown for four sample months (A, B, C, D), as sensitivity for a particular parameter varies depending on the chosen month.
+Parameters are color coded and are consistent across panels.
+Each point represents a simulation among the 5000-ensemble runs.
+See S\\@ref(tab:param) for description, global ranges and chosen ranges for parameters."}
+knitr::include_graphics(file.path("figures/ELM_FATES_parameter_sensitivity/sens.run.jpeg"))
+```
+
 ```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}Vulnerability curves fitted to observed $K_\\textrm{leaf}$ vs.
@@
 ```
-\
+
 ```{r obs-comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-comm-plc-curves}Percent loss of vulnerability curves for species with ERD estimates.
@@
 ```
-\
+
 ```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-other-erd-mods}Effective Rooting Depth versus $\\mathrm{\\delta}^2H$ for each of the ERD models tested (panels).
@@
 ```
-\
+
 ```{r erd-mort-census, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-census}Species mortality rate in each census interval during the study period plotted against ERD.
@@
 
-\
 ```{r erd-time-evg, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time-evg}Modelled species Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\\mathrm{\\Psi}_{\\textrm{soil},z}$ was below $\\mathrm{\\Psi}$~crit~, defined as $\\mathrm{\\Psi}$~80,leaf~, over five census intervals (1990-2015), where $z$ is the soil depth matching species ERD--for ", nrow(mrate.depth.mean %>% subset(deciduous == "E")), " canopy species found on Barro Colorado Island.

### 2020-10-26 manuscript.rmd
Fig. panel letters made lower case

@@
 # Materials and Methods
-\
+
 We developed a novel, empirical inverse model to estimate species effective rooting depths (ERD) for large trees of `r length(erd.sp)` species (Fig.
@@
 ## Model ELM-FATES for soil matric potentials 
-\
 ### Model description 
@@
 ## Model for effective rooting depth (ERD)
-\
+
 ### Model description 
@@
 ### Model parameterization
-\
+
 #### Growth data
@@
 \begin{split}
-B_{s} = `r bcf[1]` `r bcf[2]`\textrm{WSG}_{s} + `r bcf[3]`\textrm{WSG}_{s}^2 
+B_{s} = `r bcf[1]` `r bcf[2]`\textrm{WSG}_{s} + `r bcf[3]`\textrm{WSG}_{s}^2 \\
 &`r bcf[4]`\textrm{WSG}_{s}\textrm{LMA}_{s} + `r bcf[5]`\textrm{LMA}_{s} `r bcf[6]`\textrm{LMA}_{s}^2 + \epsilon_{s,b}
@@
 The 100 ELM-FATES ensemble members with the best fits to observed soil moisture dynamics captured soil moisture seasonality at multiple depths (Fig.
-\@ref(fig:hydro-calib)A, B), including the out of sample observations.
+\@ref(fig:hydro-calib)a, b), including the out of sample observations.
 These simulations also captured the dynamics in stream discharge (Fig.
-\@ref(fig:hydro-calib)C) and in evapotranspiration (Fig.
-\@ref(fig:hydro-calib)D), slightly underestimating peak discharge in wet years and slightly overestimating peak ET.
+\@ref(fig:hydro-calib)c) and in evapotranspiration (Fig.
+\@ref(fig:hydro-calib)d), slightly underestimating peak discharge in wet years and slightly overestimating peak ET.
 The reduction in parameter range (Table S\@ref(tab:param))  in the best-fit ensembles compared to the tested global ranges showed that the model calibration was sensitive to the Ball-Berry stomatal slope parameter, the ELM root distribution parameter (that regulates the depth of the rooting profile), soil hydraulic conductivity profile, especially at depth, and the adjustment factor that varies soil hydraulic conductivity to account for macroporosity and direct flow paths (Fig.
-S\@ref(tab:sens-sm-et), S\@ref(tab:sens-run)).
+S\@ref(fig:sens-sm-et), S\@ref(fig:sens-run)).
 
@@
 Species with deeper ERD showed greater maximum stem hydraulic conductance ($K_\textrm{max,stem}$, Spearman's' $r$ = `r cor.k`, $p$ = `r p.k`; Fig.
-\@ref(fig:erd-hyd)A), less negative leaf $\mathrm{\Psi}$~tlp~ ($r$ = `r cor.tlp`, $p$ = `r p.tlp`; Fig.
-\@ref(fig:erd-hyd)B), less negative $\mathrm{\Psi}$~88,stem~ and thus greater vulnerability to embolism from cavitation ($r$ = `r cor.stem.88`, $p$ = `r p.stem.88`; Fig.
-\@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~; $r$ = `r cor.hsm`, $p$ = `r p.hsm`; Fig.
-\@ref(fig:erd-hyd)D).
+\@ref(fig:erd-hyd)a), less negative leaf $\mathrm{\Psi}$~tlp~ ($r$ = `r cor.tlp`, $p$ = `r p.tlp`; Fig.
+\@ref(fig:erd-hyd)b), less negative $\mathrm{\Psi}$~88,stem~ and thus greater vulnerability to embolism from cavitation ($r$ = `r cor.stem.88`, $p$ = `r p.stem.88`; Fig.
+\@ref(fig:erd-hyd)c), and narrower hydraulic safety margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~; $r$ = `r cor.hsm`, $p$ = `r p.hsm`; Fig.
+\@ref(fig:erd-hyd)d).
 
@@
 Figure \@ref(fig:hydro-calib): ELM-FATES calibration and evaluation.
-Observations (red lines) versus simulations (gray lines) from 100 best-fit ensemble-member runs of ELM-FATES are shown for daily Volumetric Water Content (VWC) by depth for horizontal time domain reflectometry (TDR) probes at three depths (0.1, 0.4 and 1 m; A) and a vertical probe (0-0.15 m; B); monthly stream discharge (Discharge, C) and monthly evapotranspiration from eddy flux tower (ET, D).
-Points (red) in (A) show average plot-wide observations of VWC with 95% CI (error bars).
-VWC data from the horizontal probes (lines in A) were not used for ELM-FATES calibration, but evaluation.
+Observations (red lines) versus simulations (gray lines) from 100 best-fit ensemble-member runs of ELM-FATES are shown for daily Volumetric Water Content (VWC) by depth for horizontal time domain reflectometry (TDR) probes at three depths (0.1, 0.4 and 1 m; a) and a vertical probe (0-0.15 m; b); monthly stream discharge (Discharge, c) and monthly evapotranspiration from eddy flux tower (ET, d).
+Points (red) in (a) show average plot-wide observations of VWC with 95% CI (error bars).
+VWC data from the horizontal probes (lines in a) were not used for ELM-FATES calibration, but evaluation.
 Values in inset are average RMSE across the 100 best-fit simulations.
-For panel A, values at the top are for manual, plot-wide VWC and those at the bottom are for VWC from TDR probes.
+For panel a, values at the top are for manual, plot-wide VWC and those at the bottom are for VWC from TDR probes.
 VWC is in units of cm^3^ cm^-3^, while ET and Discharge in mm month^-1^.
@@
 \
-Figure \@ref(fig:ab-fit): Parameters $A$ (panel A) & $B$ (panel B) for the leaf vulnerability curves that were fitted to observed data on $K_\textrm{leaf}$ vs.
+Figure \@ref(fig:ab-fit): Parameters $A$ (panel a) & $B$ (panel b) for the leaf vulnerability curves that were fitted to observed data on $K_\textrm{leaf}$ vs.
 $\mathrm{\Psi}$~leaf~ for `r length(sp.exp.param$sp)` species (Eq.
@@
 (1998) except that flow rate was measured with graduated pipettes instead of a balance.
-The perfusion solution was 10 mM KCl filtered to 0.2 m.
+The perfusion solution was 10 mM KCl filtered to 0.2 mM.
 For each segment, we measured the flow rate at three pressure heads: two that ranged 0.5–2 kPa and at zero pressure.
@@
 
-```{r sens-sm-et, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Parameter sensitivity of ELM-FATES soil water content (VWC) is shown for two sample months.
-(A, B)) as well as for ET for another two sample months (C, D), as sensitivity for a particular parameter varies depending on the chosen month.
+```{r sens-sm-et, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-sm-et}Parameter sensitivity of ELM-FATES soil water content (VWC) is shown for two sample months (a, b) as well as for ET for two other sample months (c, d), as sensitivity for a particular parameter varies depending on the chosen month.
 Parameters are color coded and are consistent across panels.
@@
 
-```{r sens-run, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Parameter sensitivity of ELM-FATES stream discharge (QRUNOFF) is shown for four sample months (A, B, C, D), as sensitivity for a particular parameter varies depending on the chosen month.
+```{r sens-run, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-run}Parameter sensitivity of ELM-FATES stream discharge (QRUNOFF) is shown for four sample months (a-d), as sensitivity for a particular parameter varies depending on the chosen month.
 Parameters are color coded and are consistent across panels.
@@
 ```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}Vulnerability curves fitted to observed $K_\\textrm{leaf}$ vs.
-$\\mathrm{\\Psi}$~leaf~ data on ", length(sp.exp.param$sp)," tree species from BCI (A & B) with equation \\@ref(eq:vc-exp)--in absolute terms (A, B), as well as, in terms of loss of conductivity (%) (C, D).
-Each line represents a species, color coded by WSG (A, C), or LMA (B, D), with blue tones indicating smaller values of WSG or LMA and red tones indicating larger values.")}
+$\\mathrm{\\Psi}$~leaf~ data on ", length(sp.exp.param$sp)," tree species from BCI (a, b) with equation \\@ref(eq:vc-exp)--in absolute terms (a, b), as well as, in terms of loss of conductivity (%) (c, d).
+Each line represents a species, color coded by WSG (a, c), or LMA (b, d), with blue tones indicating smaller values of WSG or LMA and red tones indicating larger values.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/plot.obs.plc.jpeg"))
@@
 ```{r ab-param, results = 'asis', echo = FALSE}
-knitr::kable(ab.table, caption = 'Leaf vulnerability curves parameters A & B, $K_\\textrm{max,leaf}$ and $\\mathrm{\\Psi}$ for the ", `r length(erd.sp)`, " species with ERD estimates.
+knitr::kable(ab.table, caption = paste0("\\label{lab-ab-param}Leaf vulnerability curves parameters *A* & *B*, $K_\\textrm{max,leaf}$ and $\\mathrm{\\Psi}$ for the ", length(erd.sp), " species with ERD estimates.
 Species for which curves were fitted to observed data for $K_\\textrm{leaf}$ vs.
-$\\mathrm{\\Psi}$~leaf~ are identified with Source as Data, while species for which curves were obtained based on trait-proxies are identified with Source as Model', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(ab.table))))
+$\\mathrm{\\Psi}$~leaf~ are identified with Source as Data, while species for which curves were obtained based on trait-proxies are identified with Source as Model."), booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(ab.table))))
 ```

### 2020-10-27 manuscript.rmd
Corrected typos etc.

@@
 
-### Model parameterization
-
 #### Growth data
@@
 \@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r sp.n.model` ERD species without direct data and estimated leaf hydraulic vulnerability curves using Eq.
-\#ref(eq:vc-exp).
-For all the `r length(erd.sp)` ERD species we estimated species $\mathrm{\Psi}$~50,leaf~ based on their vulnerability curves for definition).
+\ref(eq:vc-exp).
+For all the `r length(erd.sp)` ERD species, we estimated species $\mathrm{\Psi}$~50,leaf~ based on their vulnerability curves for definition).
 
@@
 S\@ref(fig:iso-data)).
-Analysing $\mathrm{\delta}$^2^H~xylem~ for 12 tree species @Meinzer:2001 found that four species sourced water between 1 m and ground water, two other at least 0.8 m deep, one between 0 to 0.2 m, and five in the 0.2 to 0.7 m depth range over which the relationship between soil depth and $\mathrm{\delta}$^2^H~soil~ was variable.
+Analysing $\mathrm{\delta}$^2^H~xylem~ for 12 tree species @Meinzer:1999 found that four species sourced water between 1 m and ground water, two other at least 0.8 m deep, one between 0 to 0.2 m, and five in the 0.2 to 0.7 m depth range over which the relationship between soil depth and $\mathrm{\delta}$^2^H~soil~ was variable.
 
@@
 To test whether ERD plays a role in alleviating mortality risk, we calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$ 30 m) and average abundance of at least 20 trees ($n$ = `r length(erd.mrate.sp)`).
-For each of the seven census intervals $t$ in the 35-yr record (1981, 1985, 1990, 2000, 2005, 2010, 2015, 2015), mortality rate, $\textrm{M}_{t}$ (% yr^-1^) for species $s$ for two successive censuses, $c_{1}$ and $c_{2}$, was calculated as $\textrm{M}_{s,t} = \frac{D_{s,c_{1}}}{N_{s,c_{2}}}\times\frac{100}{\textrm{d}}$, where $N_{s,c_{1}}$ and $D_{s,c_{2}}$ are the total number of trees of species $s$ present in $c_{1}$ and dead in $c_{2}$, respectively, in the 50 ha plot and $d$ is the duration based on mean dates of $c_{1}$ and $c_{2}$.
+For each of the seven census intervals $t$ in the 35-yr record (1981, 1985, 1990, 1995, 2000, 2005, 2010, 2015), mortality rate, $\textrm{M}_{t}$ (% yr^-1^) for species $s$ for two successive censuses, $c_{1}$ and $c_{2}$, was calculated as $\textrm{M}_{s,t} = \frac{D_{s,c_{1}}}{N_{s,c_{2}}}\times\frac{100}{\textrm{d}}$, where $N_{s,c_{1}}$ and $D_{s,c_{2}}$ are the total number of trees of species $s$ present in $c_{1}$ and dead in $c_{2}$, respectively, in the 50 ha plot and $d$ is the duration based on mean dates of $c_{1}$ and $c_{2}$.
 
@@
 \@ref(fig:swp-droughts)).
-Soil water availability was not modeled for 1985-1990.
+$\mathrm{\Psi}_{\textrm{soil},z}$ over 1985-1990 is not used for analyses, as that was the ELM-FATES spin-up period, while reliable climate forcing data was unavailable for the prior years.
+However, the difference between precipitation and ET shows that 1985-1990 had three extreme deficit water years, apart from the 1983 El-Niño leading to it [@Condit.2017].
 Analysis of a species' time spent under significant soil drought (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) showed that species with shallower ERD spent a greater proportion of days below critical hydraulic thresholds and increasingly so over 1995-2000, 2000-2005 and 2005-2010 (Fig.
@@
 Our result--of lower mortality in deep rooted trees--contrasts with the finding of @Chitra-Tarak.2018 (hereafter, CT2018) in which inversely modelled, deeper ERD species died *more* in a rare, prolonged drought.
-The hydrological model in CT2018 revealed that the multi-year drought exhausted the deep soil and even bedrock water availability [also see, @Goulden.2019 and @Ivanov:2012].
+The hydrological model in CT2018 revealed that the multi-year drought exhausted the deep soil and even bedrock water availability (also see, @Goulden.2019 and @Ivanov:2012).
 In contrast, our hydrological modeling at BCI finds that during the study period, $\mathrm{\Psi}$~soil~ below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and droughts, that is above $\mathrm{\Psi}$~crit~ of the most sensitive tree species (-0.28 MPa).
@@
 \
-Figure \@ref(fig:ab-fit): Occurrence of extreme soil droughts by census intervals (horizontal panels) at three representative depths ($z$): 0.1, 0.6 and 1 m (vertical panels).
+Figure \@ref(fig:swp-droughts): Occurrence of extreme soil droughts by census intervals (horizontal panels) at three representative depths ($z$): 0.1, 0.6 and 1 m (vertical panels).
 Mean (black lines) and lower half of 95% distribution (grey areas) of soil water potential ($\mathrm{\Psi}_{\textrm{soil},z}$) for a given DOY and depth $z$ are shown (same across census intervals).
@@
 \
-Figure \@ref(fig:erd-iso): Matrix for pairwise correlations (circles) among Effective Rooting Depth (ERD) and hydraulic properties for `r nrow(cor.data)` canopy species found on Barro Colorado Island (Panama).
+Figure \@ref(fig:erd-hyd): Matrix for pairwise correlations (circles) among Effective Rooting Depth (ERD) and hydraulic properties for seven canopy species found on Barro Colorado Island (Panama).
 Each circle is color coded by the Spearman's $r$ with red shades for positive $r$ values and blue shades for negative values.
-$p$-values are given only when they are $\ge$0.05.
+$p$-values are given only when they are $\ge$ 0.05.
 Units are as follows: ERD in m, K~max,stem~ in mmol m^-1^ s^-1^ MPa^-1^) and variables with $\mathrm{\Psi}$ in MPa.
@@
 ## Data Descriptions 
-\
 ### Growth Data 
@@
 We also leveraged existing GWC measurements by @Kupers:2019 at the depths of 0.15, 0.4 and 1 m in summer 2015 (February, March and April) and 0.15, 0.4 in summer 2016 (March), with 1299 samples in total that covered all soil types and habitats in the plot.
-The sampling periods were 6, 5, 10 and 8 days long, respectively.
+The sampling periods were six, five, ten and eight days long, respectively.
 The 2016 dry season was associated with the 2015–2016 El Niño (see @Kupers:2019 for further details).
@@
 \
-To assess vulnerability to cavitation, the branches were allowed to dry in the lab for 0 – 180 hours before they were measured for stem water potential and $K_stem$.
+To assess vulnerability to cavitation, the branches were allowed to dry in the lab for 0 – 180 hours before they were measured for stem water potential and $K_\textrm{stem}$.
 We used the water potential measured by the psychrometers because they are a more direct measurement.
@@
 For some species, we noted this pattern while for others it was not apparent (Fig SXX).
-For each species, $K_stem$ was plotted against stem water potential and a Weibull curve was fit through the 90th  percentile of the points, following Wolfe et al.
+For each species, $K_\textrm{stem}$ was plotted against stem water potential and a Weibull curve was fit through the 90th  percentile of the points, following Wolfe et al.
 (2016)).
-The intercept of the curve was interpreted as maximum $K_stem$ and the water potential where the Weibull curve predicted 12, 50, 80, and 88 loss of maximum $K_stem$ were interpreted as $\mathrm{\Psi}$~12,stem~, $\mathrm{\Psi}$~50,stem~, $\mathrm{\Psi}$~80,stem~, and $\mathrm{\Psi}$~88,stem~, respectively.
+The intercept of the curve was interpreted as maximum $K_\textrm{stem}$ and the water potential where the Weibull curve predicted 12, 50, 80, and 88 loss of maximum $K_\textrm{stem}$ were interpreted as $\mathrm{\Psi}$~12,stem~, $\mathrm{\Psi}$~50,stem~, $\mathrm{\Psi}$~80,stem~, and $\mathrm{\Psi}$~88,stem~, respectively.
 
@@
 
-```{r sens-sm-et, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-sm-et}Parameter sensitivity of ELM-FATES soil water content (VWC) is shown for two sample months (a, b) as well as for ET for two other sample months (c, d), as sensitivity for a particular parameter varies depending on the chosen month.
+```{r sens-sm-et, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-sm-et}Parameter sensitivity of ELM-FATES average monthly soil water content (that is, VWC, a-b) and ET (c-d) is shown for two representative sample months, as sensitivity for a particular parameter varies depending on the chosen month.
+Horizontal axes show global ranges for a given parameter.
 Parameters are color coded and are consistent across panels.
-Each point represents a simulation among the 5000-ensemble runs.
+Each point represents a simulation among the 5000 ensemble-member runs.
 See S\\@ref(tab:param) for description, global ranges and chosen ranges for parameters."}
@@
 ```{r sens-run, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-run}Parameter sensitivity of ELM-FATES stream discharge (QRUNOFF) is shown for four sample months (a-d), as sensitivity for a particular parameter varies depending on the chosen month.
+Horizontal axes show global ranges for a given parameter.
 Parameters are color coded and are consistent across panels.
-Each point represents a simulation among the 5000-ensemble runs.
+Each point represents a simulation among the 5000 ensemble-member runs.
 See S\\@ref(tab:param) for description, global ranges and chosen ranges for parameters."}
@@
 ```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}Vulnerability curves fitted to observed $K_\\textrm{leaf}$ vs.
-$\\mathrm{\\Psi}$~leaf~ data on ", length(sp.exp.param$sp)," tree species from BCI (a, b) with equation \\@ref(eq:vc-exp)--in absolute terms (a, b), as well as, in terms of loss of conductivity (%) (c, d).
+$\\mathrm{\\Psi}$~leaf~ data for ", length(sp.exp.param$sp)," tree species from BCI with Eq.
+\\@ref(eq:vc-exp)--in absolute terms (a, b), as well as, in terms of loss of conductivity (%) (c, d).
 Each line represents a species, color coded by WSG (a, c), or LMA (b, d), with blue tones indicating smaller values of WSG or LMA and red tones indicating larger values.")}
@@
 Each line represents a species, color coded by source of the curve: fitted to observed data for $K_\\textrm{leaf}$ vs.
-$\\mathrm{\\Psi}$~leaf~ are in yellow and those based on scaling relationship with WSG and LMA are in gray.")}
+$\\mathrm{\\Psi}$~leaf~ are in yellow and those based on the scaling relationship with WSG and LMA are in gray.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/obs.mod_plccurves_fits_ggplot_no_col.jpeg"))
@@
 #Supplementary Tables
-\
+
 ```{r param, results = 'asis', echo = FALSE}

### 2020-10-29 manuscript.rmd
1. Corrected erd equation 2. Other minor changes

@@
 These efforts have resulted in identification of correlations between above-ground hydraulic traits such as vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013, @Venturas.2016].
-Deep roots may mitigate hydraulic vulnerability and mortality-risk, yet plants' realized water environments, specifically plant's ability to acquire water via differing rooting depths in hydrological rather than meteorological droughts, are rarely studied, especially in the tropical region.
+Deep roots may mitigate hydraulic vulnerability and mortality-risk, yet plant's ability to acquire water via differing rooting depths and the actual water-availability by depth during droughts are rarely studied, especially in the tropical region.
 We tested how rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tree species of a tropical forest over 35 years that witnessed droughts of various intensity and duration.
@@
 We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq.
-\@ref(eq:water-balance), namely, (i) evapotranspiration (ET) from the eddy flux tower ($\cong$ $E_{v} + E_{g}$; 2012-2017; see [SI](#SIData.matteo))), (ii) stream discharge ($q_\textrm{run}$) from the Conrad catchment ($\cong$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC or $\theta_{z}$) $\cong$ $W_{\textrm{liq,} z}$ from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we obtained with a vertical time domain reflectometry (TDR) probe over 0-15 cm depth ($\mathrm{\theta}_\textrm{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m with 1299 samples in total covering all soil types and habitats (see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter,  $\mathrm{\theta}_{\textrm{0-0.15 m}}$, $\mathrm{\theta}_{\textrm{0.4 m}}$ and $\mathrm{\theta}_{\textrm{1 m}}$ refer to plot-wide averages).
+\@ref(eq:water-balance), namely, (i) evapotranspiration (ET) from the eddy flux tower ($\cong$ $E_{v} + E_{g}$; 2012-2017; see [SI](#SIData.matteo))), (ii) stream discharge ($q_\textrm{run}$) from the Conrad catchment ($\cong$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC or $\theta_{z}$ $\cong$ $W_{\textrm{liq,} z}$) from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we obtained with a vertical time domain reflectometry (TDR) probe over 0-15 cm depth ($\mathrm{\theta}_\textrm{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m with 1299 samples in total covering all soil types and habitats (see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter,  $\mathrm{\theta}_{\textrm{0-0.15 m}}$, $\mathrm{\theta}_{\textrm{0.4 m}}$ and $\mathrm{\theta}_{\textrm{1 m}}$ refer to plot-wide averages).
 
@@
 \begin{equation}
-\textrm{ERD}_{s} = \underset{\forall \, h \, \in \, H}{\textrm{median}} \Big\{z,h\,\Big|\, \Big(\textrm{L}_{s|z,h} = \underset{\forall \, z \,\in\,Z}{\textrm{maximum}}(\textrm{L}_{s|~z,h})\Big) \cap \Big(r_{s|z,h} \ge 0.71\Big)\Big\}
+\textrm{ERD}_{s} = \underset{\forall \, h \, \in \, H}{\textrm{median}} \Big(z\,\Big|\, \textrm{L}_{s|z,h} = \underset{\forall \, z \,\in\,Z}{\textrm{maximum}}(\textrm{L}_{s|z,h}| r_{s|z,h} \ge 0.71\Big)\Big)
 (\#eq:erd)
 \end{equation}
-where $L$ is the likelihood, $H$ = {1, .., 100}, refer to the 100 hydrological realizations and $Z$ = {0.4, `r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m.
+where $L$ is the likelihood of $z$ given the observed growth time series, $H$ = {1, .., 100}, refer to the 100 hydrological realizations and $Z$ = {0.4, `r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m.
 We removed finely resolved soil layers below 0.21 m and averaged $\mathrm{\Psi}_{\textrm{soil},z}$ for depths 0.21, 0.37 and 0.62 at 0.4 m as sufficient resolution for Model validation.
@@
 This has important implications for our predictive understanding of tropical forest dynamics under current and future climate.
-Our community-scale framework for modeling effective rooting depths and leaf vulnerability curves indicates the possibilities in expanding the use of these critical, hard-earned observations in species-rich forests towards community-scale generalizations.
+Our community-scale framework for modeling effective rooting depths and leaf vulnerability curves indicates the possibilities in expanding the use of these critical, rare observations in species-rich forests towards community-scale generalizations.
 
@@
 (2012), $K_h$ was calculated as the slope of the flow rate versus pressure gradient (i.e., pressure head divided by segment length) among the three pressure heads.
-Stem area specific hydraulic conductivity was ($K_\textrm{stem}$) was calculated by dividing $K_h$ by the cross-sectional area of stem.
+Stem area specific hydraulic conductivity was ($K_\textrm{stem}$) was calculated by dividing $K_h$ by the cross-sectional area of the stem.
 To calculate cross-sectional stem area, stem diameter was measured twice perpendicularly on both ends of the segment and averaged, then area was calculated assuming a circular cross section.
@@
 
-```{r sens-sm-et, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-sm-et}Parameter sensitivity of ELM-FATES average monthly soil water content (that is, VWC, a-b) and ET (c-d) is shown for two representative sample months, as sensitivity for a particular parameter varies depending on the chosen month.
+```{r sens-sm-et, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-sm-et}Parameter sensitivity of ELM-FATES soil water content (that is, VWC, a-b) and ET (c-d) is shown for two representative sample months, as sensitivity for a particular parameter varies depending on the chosen month.
 Horizontal axes show global ranges for a given parameter.

### 2020-10-31 manuscript.rmd
1. Correctly added two references
1. more editing and formatting

@@
 
+Chitra-Tarak, Rutuja*, Chonggang Xu, Kristina Anderson-Teixeira, Jeff Chambers, Matteo Detto, Boris Faybishenko, Rosie Fisher, Ryan Knox, Charlie Koven, Lara Kueppers, Nobert Kunert, Stefan J.
+Kupers, Nathan G.
+McDowell, Brent Newman, Steve Paton, Laurent Ruiz, Lawren Sack, Jeffrey M.
+Warren, Brett Wolfe, Cynthia Wright, S.
+Joseph Wright, Joseph Zailaa, Sean M.
+McMahon
+
 \
-**Corresponding author**: 
+*__Corresponding author__: 
 Rutuja Chitra-Tarak
@@
 ------------- | -------------
-Summary | 197
-Main Text  | 6516
-Introduction | 890
-Methods | 3300
-Results | 814
-Discussion | 1512
+Summary | 195
+Main Text  | 6637
+Introduction | 873
+Methods | 3313
+Results | 859
+Discussion | 1578
 Keywords | 6
@@
 Here, we tested the role of deep-water access in enabling survival within a diverse tropical forest community (Panama) using a novel data-model approach.
-- We inversely modeled effective rooting depth (ERD, as the average depth of water extraction), for large trees of `r length(erd.sp)` canopy species by linking growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and to leaf hydraulic vulnerability curves.
-We validated ERD against existing tree and soil stable water isotope data.
+- We inversely modeled the effective rooting depth (ERD, as the average depth of water extraction), for `r length(erd.sp)` canopy species by linking growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and leaf hydraulic vulnerability curves.
+We validated ERD against existing data for stable hydrogen isotopes of trees and soil.
 - Across species, deeper ERD was associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates during extreme droughts.
 Species with deeper ERD spent less time below critical hydraulic thresholds indicating that trees compensate for water-stress related mortality risk through deeper water access.
-- The role of deep-water access in alleviating mortality of hydraulically-vulnerable trees has important implications for our predictive understanding of forest dynamics under current and future climates.
+- The role of deep-water access in mitigating mortality of hydraulically-vulnerable trees has important implications for our predictive understanding of forest dynamics under current and future climates.
 
@@
 # Introduction
-\
+
+Drought-induced mortality in tropical forests may have significant global implications.
 Tropical forests play a disproportionately large role in the global carbon and energy cycles [@Bonan:2008], and support half of global biodiversity [@Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019].
-Tropical forests are logically considered to be an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw], yet mortality events in the tropics are rarely as large as those in the temperate and boreal zones [@McDowell.2018], leading to questions regarding the role of large trait-diversity and hydraulic strategies in mitigating mortality events.
-Despite this, widespread drought-induced tree mortality has occurred across the tropics for specific functional groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010] and tropical mortality rates are increasing in some regions [@Brienen.2015; @Hubau.2020].
-However, state-of-the-art dynamic global vegetation models (DGVMs) struggle to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013; @Powell.2018], as underlying mechanisms of drought-tolerance are not fully understood or quantified.
+Tropical forest has been expected to be an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw].
+Still, mortality events in the tropics are rarely as large as those in temperate and boreal zones [@McDowell.2018], leading to questions regarding the role of large trait-diversity and hydraulic strategies in mitigating mortality events.
+Nevertheless, mortality rates are increasing in some tropical regions [@Brienen.2015; @Hubau.2020] and widespread drought-induced tree mortality has occurred across the tropics for specific functional groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010].
+State-of-the-art dynamic global vegetation models (DGVMs) struggle to capture these drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013; @Powell.2018], because underlying mechanisms of drought-tolerance are not fully understood nor quantified.
 
@@
 Deep water access is arguably the most long-standing, yet under-studied mechanism.
-This mechanism assumes that plant available water varies with depth, and that the rooting depths of co-located trees also vary, by species and size [@Meinzer:1999; @Chitra-Tarak.2018; @Brum.2019].
-Consequently, trees in the same forest may differ in their experience during a drought, and thus in their growth and mortality responses [@Chitra-Tarak.2018].
+As plant available water varies with depth, trees in the same forest with different rooting depths, depending on species and size [@Meinzer:1999; @Chitra-Tarak.2018; @Brum.2019], differ in their experience during a drought, and thus in their growth and mortality responses [@Chitra-Tarak.2018].
 A key bottleneck in community-wide testing of this mechanism has been a lack of data in both trees' rooting or water-sourcing depths and plant available soil water at those depths.
@@
 Further, only a small fraction of those are tropical, even though more than 90% of the world’s tree diversity resides in the tropics [@Slik:2015].
-Use of stable isotopes of water as a tracer provides an indirect measure of water sourcing depths by matching the isotopic value in xylem water to variations in soil pore water with depth.
+The use of stable isotopes of water as a tracer provides an indirect measure of water sourcing depths by matching the isotopic value in xylem water to vertical variations in soil pore water.
 However, such data are rare [@Evaristo:2016] and their interpretation is an active area of research given unresolved issues pertaining to soil heterogeneity and the correspondence of soil water extract in the laboratory to actual plant available water [@Adams.2020; @Bowers.2020; @Deurwaerder.2020].
@@
 \
-  Trade-offs in plant 'traits' shape how plants make a living in their environment leading to alternative ecological strategies.
-Characterizing the essential constraints necessary to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010; @Bartlett.2019] entails identifying the topography of trait trade-offs in different environments, and how these relate to demographic rates (growth, recruitment and mortality).
+  Characterising the essential constraints to model species-rich communities [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010; @Bartlett.2019] entails identifying the topography of plant trait trade-offs in different environments, and how these relate to demographic rates (growth, recruitment and mortality).
 Significant efforts have been invested into identifying and linking universal drought indices, above-ground traits, and demographic rates.
-These efforts have resulted in identification of correlations between above-ground hydraulic traits such as vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013, @Venturas.2016].
-Deep roots may mitigate hydraulic vulnerability and mortality-risk, yet plant's ability to acquire water via differing rooting depths and the actual water-availability by depth during droughts are rarely studied, especially in the tropical region.
-We tested how rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tree species of a tropical forest over 35 years that witnessed droughts of various intensity and duration.
-
+Such efforts have found correlations between vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013, @Venturas.2016].
+Even though deep roots may mitigate hydraulic vulnerability and mortality-risk, few drought studies compare a tree's access to water, i.e.
+rooting depth, vs.
+the actual vertical distribution of plant-available water, especially in tropical regions.
 \
-   We tested an overarching hypothesis that inter-specific variation in rooting depths enables increasingly reliable water availability with depth and thus mitigation of hydraulic vulnerability.
-Our hypotheses are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallow rooted species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deep rooted species have lower mortality rates in droughts due to low realized risk of hydraulic failure, whereas shallow rooted species may have higher mortality rates as a result of higher variability in water availability in shallow soil layers.
-To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.
+  Here, we estimate plant-available water in the whole-soil column in a tropical forest, and inversely estimate rooting depths of co-occurring tree species from their growth responses, with a series of model calibrations and validation.
+We evaluate how rooting depth is linked to above-ground hydraulic traits and mortality rates over a 35 year period which experienced droughts of a variety of intensity, frequency and duration.
+Our hypotheses (see Table \@ref(tab:hypo)) are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem due to greater and more reliable water availability, whereas shallow rooted species are associated with less efficient but safer hydraulic traits due to more variable water availability; and that, (ii) deep rooted species have lower mortality rates during droughts due to lower realized risk of hydraulic failure compared to shallow rooted species.
+To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality rates for a species-rich forest.
 
@@
 
-We developed a novel, empirical inverse model to estimate species effective rooting depths (ERD) for large trees of `r length(erd.sp)` species (Fig.
-\@ref(fig:schematic)), defined as the depth at which soil water potential best explained species-specific growth dynamics.
-The ERD model incorporated the impact of atmospheric drought on growth and was constrained with species-specific leaf vulnerability curves, by using existing data for eight species and developing trait-based proxies for the rest.
 To obtain the daily dynamics of soil water potential in the whole-soil column we locally parameterized and calibrated a 1-D hydrological water balance model.
-We evaluated ERD against existing deuterium to hydrogen content data for a subset of six tree species as an independent observation of ERD.
-We tested the relationship of ERD with above-ground hydraulic traits, namely, xylem vulnerability to embolism, branch hydraulic safety margins, leaf turgor loss point and maximum stem-specific hydraulic conductivity, for a subset of eight species with both datasets.
-Finally, we analysed species' mortality rates over 35 years to test whether ERD explained risk of drought-induced mortality.
+We then developed a novel, empirical inverse model to estimate species effective rooting depths (ERD) for large trees of `r length(erd.sp)` species (Fig.
+\@ref(fig:schematic)).
+We define ERD as the depth at which soil water potential best explained species-specific growth dynamics.
+The ERD model incorporated the impact of atmospheric and soil drought on growth, and was constrained with species-specific leaf vulnerability curves by using existing data for eight species, while developing trait-based proxies for the rest (based on data for a total of 21 species).
+We evaluated ERD against existing stable hydrogen isotope ratios ($\mathrm{\delta}$^2^H~xylem~) for a subset of six tree species as an independent observation of ERD.
+For a subset of eight species with existing data for above-ground hydraulic traits, we tested the relationship of ERD with xylem vulnerability to embolism, branch hydraulic safety margins, leaf turgor loss point and maximum stem-specific hydraulic conductivity.
+Finally, we analysed species-specific large tree mortality dynamics over 35 years and estimated time a species’ spent below its critical hydraulic threshold to test whether ERD explained risk of drought-induced mortality.
 
@@
 ## Study site description
-\
-We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, hydrology and plant functional ecology, including data for stable water isotopes on trees and soils [@Meinzer:1999].
+
+We conducted this study at Barro Colorado Island (BCI), Panama.
+The site is extensively characterized in terms of long-term forest demography, hydrology and plant functional ecology, including data for stable water isotopes on trees and soils [@Meinzer:1999].
 The 1,560ha island was formed after the river Chagres was dammed in 1914 to create Gatun Lake and complete the Panama Canal.
@@
 \
-  In the 50 ha plot, all stems above ten mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every five year through 1985-2015 for growth, mortality as well as recruitment of new stems into the ten mm dbh size class [@Condit.2017].
+  In the 50 ha plot, all stems $ge$ 1 cm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every five year through 1985-2015 for growth, mortality as well as recruitment of new stems into the ten mm dbh size class [@Condit.2017].
 This inventory represents 321 woody species across the eight censuses.
@@
 ## Model ELM-FATES for soil matric potentials
+
 ### Model description
-\
+
 To estimate dynamics of water availability by depth over the forest’s rooting zone, we used the Energy Exascale Earth System Land Model (ELM), coupled with the Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES).
@@
 ### Model parameterization
-\
+
 To parameterize ELM-FATES, we identified eleven parameters relevant for Eq.
@@
 We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq.
-\@ref(eq:water-balance), namely, (i) evapotranspiration (ET) from the eddy flux tower ($\cong$ $E_{v} + E_{g}$; 2012-2017; see [SI](#SIData.matteo))), (ii) stream discharge ($q_\textrm{run}$) from the Conrad catchment ($\cong$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC or $\theta_{z}$ $\cong$ $W_{\textrm{liq,} z}$) from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we obtained with a vertical time domain reflectometry (TDR) probe over 0-15 cm depth ($\mathrm{\theta}_\textrm{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m with 1299 samples in total covering all soil types and habitats (see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter,  $\mathrm{\theta}_{\textrm{0-0.15 m}}$, $\mathrm{\theta}_{\textrm{0.4 m}}$ and $\mathrm{\theta}_{\textrm{1 m}}$ refer to plot-wide averages).
+\@ref(eq:water-balance), namely, (i) evapotranspiration (ET) from the eddy flux tower ($\cong$ $E_{v} + E_{g}$; 2012-2017; see [SI](#SIData.matteo))), (ii) stream discharge ($q_\textrm{run}$) from the Conrad catchment ($\cong$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC or $\theta_{z}$ $\cong$ $W_{\textrm{liq,} z}$) from two sources: (1) a long-term (2012-2018) record of average VWC we obtained with vertical time domain reflectometry (TDR) probes from three locations near the eddy flux tower over 0-15 cm depth   ($\mathrm{\theta}_\textrm{0-0.15 m}$), and (2) existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m with 1299 samples in total covering all soil types and habitats (see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter,  $\mathrm{\theta}_{\textrm{0-0.15 m}}$, $\mathrm{\theta}_{\textrm{0.4 m}}$ and $\mathrm{\theta}_{\textrm{1 m}}$ refer to plot-wide averages).
 
@@
 ### Model description
-\
+
 To estimate effective rooting depth, we tested alternate empirical models for species 5-yearly census growth (Eq.
@@
 ### Model validation and selection
-\
+
 We used $\mathrm{\delta}$^2^H~xylem~ as an independent observation to validate the ERD models.
@@
 ## Relationship of ERD with above-ground hydraulic traits
-\
-To evaluate whether ERD was associated with above-ground hydraulic traits (eight species with both datasets), we regressed species ERD against maximum stem hydraulic conductance ($K_\textrm{max,stem}$), leaf turgor loss point ($\mathrm{\Psi}$~tlp~), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\mathrm{\Psi}$~88,stem~), and hydraulic safety margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~).
-See [SI](#SIData.brett) for data collection and estimation of these variables.
+
+To evaluate whether ERD was associated with above-ground hydraulic traits, for a subset of eight species with both ERD and traits data, we regressed species ERD against maximum stem hydraulic conductance ($K_\textrm{max,stem}$), leaf turgor loss point ($\mathrm{\Psi}$~tlp~), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\mathrm{\Psi}$~88,stem~), and hydraulic safety margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~).
+See @Wolfe.2019 and [SI](#SIData.brett) for data collection and estimation of these variables.
 
 ## Mortality analyses and species-specific drought experience
-\
+
 To test whether ERD plays a role in alleviating mortality risk, we calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$ 30 m) and average abundance of at least 20 trees ($n$ = `r length(erd.mrate.sp)`).
@@
 # Results
+
 ## Soil water dynamics by depth
-\
+
 The 100 ELM-FATES ensemble members with the best fits to observed soil moisture dynamics captured soil moisture seasonality at multiple depths (Fig.
@@
 ## Leaf hydraulic vulnerability curves
-\
+
 Parameters of leaf vulnerability curves were predictable from WSG and LMA.
@@
 S\@ref(fig:obs-plc-curves)).
-This predictive power allowed us to estimate leaf vulnerability curve parameters for `r sp.n.model` that lacked direct observations (out of `r length(erd.sp)` ERD species; Table S\@ref(tab:ab-param), Fig.
+This predictive power allowed us to estimate leaf vulnerability curve parameters for `r sp.n.model` species that lacked direct observations (out of `r length(erd.sp)` ERD species; Table S\@ref(tab:ab-param), Fig.
 S\@ref(fig:obs-comm-plc-curves)).
@@
 ## Effective rooting depths (ERD)
-\
+
 The chosen ERD model (with a multiplicative effect of VPD and not LAI; Eq.
@@
 ## Relationship of ERD with above-ground hydraulic traits
-\
+
 Species with deeper ERD showed greater maximum stem hydraulic conductance ($K_\textrm{max,stem}$, Spearman's' $r$ = `r cor.k`, $p$ = `r p.k`; Fig.
@@
 ## Effective rooting depths, experienced drought and mortality
-\
+
 Among evergreen species, ERD explained `r m.evg.r2.90`, `r m.evg.r2.05` and `r m.evg.r2.10`% variation in mortality rates during census intervals 1985-1990, 2000-2005 and 2005-2010, respectively, such that species with mortality rates significantly decreased with ERD ($p$ = `r m.evg.p.90`, `r m.evg.p.05` and `r m.evg.p.10`, respectively; Fig.
@@
 # Discussion
-\
+
 Our novel approach for estimating ERD using 25 years of growth, species-specific leaf vulnerability curves, modeled soil moisture through the profile, and observed VPD was consistent with estimates using $\mathrm{\delta}$^2^H~xylem~.
@@
 
-### Realized drought risk integral to assessing drought-vulnerability
-\
+## Realized drought risk integral to assessing drought-vulnerability
+
 Our study brings attention to the need for assessing drought sensitivity in terms of species realized drought exposure by accounting for the actual hydrological drought and trees' deep water access, beyond the universal drought indices and trees' above-ground hydraulic vulnerability.
@@
 Our results are consistent with a few studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016; also see @Brum.2019].
+Globally, large trees tend to exhibit greater growth reductions, lower post-drought resilience, and greater increases in mortality relative to their understory counterparts [@Bennett:2015; @Phillips:2010].
+Our finding that deep-water access buffers drought-induced mortality in large trees is relevant for understanding drought resistance, resilience and recovery of this important functional group [@Bennett:2015; @McGregor.2020]
 
@@
 
-### Effective rooting depths at the tree community level
-\
+## Effective rooting depths at the tree community level
+
 By estimating ERD for large trees of `r length(erd.sp)` canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species-rich tropical forests.
@@
 
-### Future directions
-\
+## Future directions
+
 The relationships that we identify between above- and below-ground traits, the water environment vertically below-ground and mortality rates are important for representing diversity in DGVMs, which intrinsically rely on the parameterization of contrasting life history strategies [@Scheiter:2013] and the simulation of competition between those strategies.
@@
 ## Conclusions
-\
+
 Establishing relationships between environment, traits and demographic outcomes of plants is imperative for developing a predictive plant ecology.
@@
 # Acknowledgements
-\
+
 This research was supported as part of the Next Generation Ecosystem Experiments-Tropics, funded by the U.S.
@@
 LR was deputed to Indo-French Cell for Water Sciences on an IRD Fellowship.
-An NSF grant to Smithsonian Institution’s ForestGEO (Dimensions 1046113) also partially supported this work via RCT and SMM.
+An NSF grant to Smithsonian Institution’s ForestGEO (Dimensions 1046113) partially supported this work via RCT and SMM.
 RF was supported by the National Center for Atmospheric Research, which is funded by the National Science Foundation.
-**Add acknowledgements for other co-authors**.
 The BCI forest dynamics research project, founded by Stephen P.
@@
 Numerous organizations have provided funding, principally the US National Science Foundation, and hundreds of field workers have contributed.
+
+# Author contribution
+
+Conceptualization: RCT.
+Data curation: BF, SP for climate drivers, SP for stream discharge, MTT for GPP, ET, TDR, bulk density, SK for manual GWC and GWC vs $\Psi_\textrm{soil}$.
+NK, JZ, KAT, LS for leaf hydraulics, BW for stem hydraulics, JW for maximum tree height, WSG, LMA, deciduousness, leaf fall and leaf lifetime.
+Formal Analysis: RCT, CX.
+MTT for upgrading GPP-VPD relationship and LAI seasonality.
+Funding acquisition: SMM, JC, CX, NM, LK.
+Investigation: RCT, CX (SMM and LR for an exploratory version).
+Methodology: RCT, CX (SMM and LR for an exploratory version).
+Project administration: RCT, Resources: SMM, BF, SP, MTT, SK, NK, JZ, KAT, LS, BW, JW.
+Software: RCT; RF, RK, CF, CX for ELM-FATES.
+Supervision: CX, SMM, BN, Validation: RCT.
+Visualization: RCT, Writing – original draft: RCT.
+Writing – review & editing: All authors.
+See https://casrai.org/credit/ for the taxonomy of credits.
 
@@
 Figure \@ref(fig:hydro-calib): ELM-FATES calibration and evaluation.
-Observations (red lines) versus simulations (gray lines) from 100 best-fit ensemble-member runs of ELM-FATES are shown for daily Volumetric Water Content (VWC) by depth for horizontal time domain reflectometry (TDR) probes at three depths (0.1, 0.4 and 1 m; a) and a vertical probe (0-0.15 m; b); monthly stream discharge (Discharge, c) and monthly evapotranspiration from eddy flux tower (ET, d).
-Points (red) in (a) show average plot-wide observations of VWC with 95% CI (error bars).
-VWC data from the horizontal probes (lines in a) were not used for ELM-FATES calibration, but evaluation.
+Observations (red lines and points) versus simulations (gray lines) from 100 best-fit ensemble-member runs of ELM-FATES are shown for daily Volumetric Water Content (VWC) by depth for horizontal time domain reflectometry (TDR) probes at three depths (0.1, 0.4 and 1 m; A) and a vertical TDR probe (0-0.15 m; B); monthly stream discharge (Discharge, C) and monthly evapotranspiration from eddy flux tower (ET, D).
+Red points in (A) show average, manual plot-wide observations of VWC with 95% CI (error bars).
+All observations (red lines and points) were in-sample, except for VWC data from the horizontal probes (red lines in A) which were out-of-sample.
 Values in inset are average RMSE across the 100 best-fit simulations.
@@
 Each circle is color coded by the Spearman's $r$ with red shades for positive $r$ values and blue shades for negative values.
-$p$-values are given only when they are $\ge$ 0.05.
+$p$-values are shown inside the circles.
 Units are as follows: ERD in m, K~max,stem~ in mmol m^-1^ s^-1^ MPa^-1^) and variables with $\mathrm{\Psi}$ in MPa.
@@
 \
-We collected volumetric water content (VWC) data (2012-2018) from a point location at the eddy flux tower with a vertical time domain reflectometry (TDR) probe.
-The relationship between the output from the TDR probe and VWC was calibrated based on measurements of gravimetric water content (GWC) on several occasions from representative soils near the vertical probe (0-0.15 m) (Fig.
+We collected volumetric water content (VWC) data (2012-2018) from three locations near the eddy flux tower with time domain reflectometry (TDR, CS616, Campbell Scientific) probes inserted with an angle to average the first 15 cm of soil.
+The relationship between the output from the TDR probe and VWC was calibrated based on measurements of gravimetric water content (GWC) on several occasions from representative soils near the vertical probes (0-0.15 m) (Fig.
 S\@ref(fig:TDR-calib)).
@@
 
+### Minimum Leaf Water Potentials
+
+Pre-dawn and diurnal leaf water potential were measured on a monthly basis from Feb to May 2016 at SLZ and PNM.
+Data from BCI only available for March.
+This data was collected as part of the 2016 ENSO campaign.
+See [@Wolfe.2019].
+
 ### Stem maximum hydraulic conductivity and vulnerability to cavitation {#SIData.brett}
@@
 \
-Methods for Stem Maximum hydraulic conductivity and vulnerability to cavitation closely followed [Christman.2012] and [Wolfe.2016].
+Methods for Stem Maximum hydraulic conductivity and vulnerability to cavitation closely followed [@Christman.2012] and [@Wolfe.2016].
 Values for some of the species reported here were previously reported by @Dickman.2019 and @Wu.2020.

### 2020-11-01 manuscript.rmd
1. removed HSM from hyd data table

@@
 
-## Relationship of ERD with above-ground hydraulic traits
+## Relationships between ERD and above-ground hydraulic traits
 
-To evaluate whether ERD was associated with above-ground hydraulic traits, for a subset of eight species with both ERD and traits data, we regressed species ERD against maximum stem hydraulic conductance ($K_\textrm{max,stem}$), leaf turgor loss point ($\mathrm{\Psi}$~tlp~), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\mathrm{\Psi}$~88,stem~), and hydraulic safety margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~).
-See @Wolfe.2019 and [SI](#SIData.brett) for data collection and estimation of these variables.
+To evaluate whether ERD was associated with above-ground hydraulic traits (eight species with both datasets), we regressed species ERD against maximum stem hydraulic conductivity ($K_\textrm{max,stem}$), leaf turgor loss point ($\mathrm{\Psi}$~tlp~), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of $K_\textrm{max,stem}$ is lost ($\mathrm{\Psi}$~88,stem~), and hydraulic safety margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~).
+See @Wolfe.2019, @Wolfe.2020 and [SI](#SIData.brett) for data collection and estimation of these variables.
 
@@
 
-## Relationship of ERD with above-ground hydraulic traits
+## Relationships between ERD and above-ground hydraulic traits
 
-Species with deeper ERD showed greater maximum stem hydraulic conductance ($K_\textrm{max,stem}$, Spearman's' $r$ = `r cor.k`, $p$ = `r p.k`; Fig.
+Species with deeper ERD showed greater $K_\textrm{max,stem}$ (Spearman's' $r$ = `r cor.k`, $p$ = `r p.k`; Fig.
 \@ref(fig:erd-hyd)a), less negative leaf $\mathrm{\Psi}$~tlp~ ($r$ = `r cor.tlp`, $p$ = `r p.tlp`; Fig.
@@
 
-Our novel approach for estimating ERD using 25 years of growth, species-specific leaf vulnerability curves, modeled soil moisture through the profile, and observed VPD was consistent with estimates using $\mathrm{\delta}$^2^H~xylem~.
+Our novel approach for estimating ERD using 25 years of growth, species-specific leaf vulnerability curves, modeled soil moisture profiles, and observed VPD was consistent with estimates using $\mathrm{\delta}$^2^H~xylem~.
 Our analyses revealed that co-occurring large canopy tree species with deeper ERD were associated with higher hydraulic efficiency, but not safety (Fig.
@@
 Nevertheless, deeper ERD species showed lower mortality rates than shallower ERD species in significant wet season droughts over the 35-yr record (1981-2015) (Fig.
-S\@ref(fig:erd-mort-census), especially amongst the evergreen species (Fig.
+S\@ref(fig:erd-mort-census)), especially amongst the evergreen species (Fig.
 \@ref(fig:erd-mort-census-evg)).
@@
 \
-Our result--of lower mortality in deep rooted trees--contrasts with the finding of @Chitra-Tarak.2018 (hereafter, CT2018) in which inversely modelled, deeper ERD species died *more* in a rare, prolonged drought.
+Our result--of lower mortality in deep-rooted trees--contrasts with the finding of @Chitra-Tarak.2018 (hereafter, CT2018) in which inversely modelled, deeper ERD species died *more* in a rare, prolonged drought.
 The hydrological model in CT2018 revealed that the multi-year drought exhausted the deep soil and even bedrock water availability (also see, @Goulden.2019 and @Ivanov:2012).
-In contrast, our hydrological modeling at BCI finds that during the study period, $\mathrm{\Psi}$~soil~ below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and droughts, that is above $\mathrm{\Psi}$~crit~ of the most sensitive tree species (-0.28 MPa).
+In contrast, our hydrological modeling at BCI finds that during the study period, $\mathrm{\Psi}$~soil~ below 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and droughts, that is above $\mathrm{\Psi}$~crit~ of the most sensitive tree species (-0.28 MPa).
 The difference in the results shown here are likely attributable to site differences in the precipitation regime: with mean annual rainfall of 1095 in CT2018 compared to `r round(rain.man.stats$rain)`mm at BCI, and precipitation to potential evapotranspiration ratio nearly one in CT2018 compared to nearly two in BCI.
@@
 By estimating ERD for large trees of `r length(erd.sp)` canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species-rich tropical forests.
-Our chosen ERD model is an advance over the only other inverse model of ERD (@Chitra-Tarak.2018; $n$ = 12 species) as we employ a physiologically meaningful representation, fewer parameters and corroboration with tree and soil data for $\mathrm{\delta}$^2^H.
+Our chosen ERD model is an advance over the only other inverse model of ERD (CT2018; $n$ = 12 species) as we employ a physiologically meaningful representation, fewer parameters and corroboration with tree and soil data for $\mathrm{\delta}$^2^H.
 We highlight that stable water isotopes for plant and below-ground sources are critical for calibration or validation of ERD models at community scale, although the interpretation of these data is still under research [@Deurwaerder.2020].
-As in @Chitra-Tarak.2018, we also interpret our ERD estimates as revealing relative differences among species' effective rooting depths rather than in absolute terms.
-As our ERD model ignores stem water storage capacitance it may have overestimated ERD especially in species for which stem water storage capacitance is important as a drought-avoidance strategy, for example, the deciduous species [@Borchert.2005jma].
+As in CT2018, we also interpret our ERD estimates as revealing relative differences among species' effective rooting depths rather than in absolute terms.
+As our ERD model ignores stem water storage capacitance [@Wolfe.2017] it may have overestimated ERD especially in species for which stem water storage capacitance is important as a drought-avoidance strategy, for example, the deciduous species [@Borchert.2005jma].
 
@@
 Conceptualization: RCT.
-Data curation: BF, SP for climate drivers, SP for stream discharge, MTT for GPP, ET, TDR, bulk density, SK for manual GWC and GWC vs $\Psi_\textrm{soil}$.
-NK, JZ, KAT, LS for leaf hydraulics, BW for stem hydraulics, JW for maximum tree height, WSG, LMA, deciduousness, leaf fall and leaf lifetime.
+Data curation: BF, SP for climate drivers.
+SP for stream discharge.
+MTT for GPP, ET, TDR, bulk density.
+SK for manual GWC and GWC vs $\Psi_\textrm{soil}$.
+NK, JZ, KAT, LS for leaf hydraulics, BW for stem hydraulics.
+JW for maximum tree height, WSG, LMA, deciduousness, leaf fall and leaf lifetime.
 Formal Analysis: RCT, CX.
@@
 Project administration: RCT, Resources: SMM, BF, SP, MTT, SK, NK, JZ, KAT, LS, BW, JW.
-Software: RCT; RF, RK, CF, CX for ELM-FATES.
+Software: RCT; RF, RK, CF, CX, RCT for ELM-FATES.
 Supervision: CX, SMM, BN, Validation: RCT.
@@
 
-To be uploaded on NGET-Archive and site DOIs:
-(1) Boris to upload climate forcing data
-(2) Rutuja to upload R scripts for ELM-FATES parameter generations, ELM-FATES modified code, and all R scripts to reproduce manuscript.
+To be uploaded on NGET-Archive and site DOIs along with manuscript acceptance:
+(1) BF to upload climate forcing data
+(2) RCT to upload R scripts for ELM-FATES parameter generations, ELM-FATES modified code, and all R scripts to reproduce manuscript.
 \
@@
 
-### Minimum Leaf Water Potentials
-
-Pre-dawn and diurnal leaf water potential were measured on a monthly basis from Feb to May 2016 at SLZ and PNM.
-Data from BCI only available for March.
-This data was collected as part of the 2016 ENSO campaign.
-See [@Wolfe.2019].
-
 ### Stem maximum hydraulic conductivity and vulnerability to cavitation {#SIData.brett}
 
-Brett to edit.
+\
+Values of stem maximum hydraulic conductivity and vulnerability to cavitation were obtained from the dataset of @Wolfe.2020.
+The entire dataset includes measurements on 26 tree species collected at three sites in Panama.
+The raw data, *R* script for the analyses, and data summaries are archived in the NGEE-Tropics archive (@Wolfe.2020).
+The data were previously reported by @Dickman.2019 and @Wu.2020.
+Here, the seven species that overlapped with our assessment of ERD were included in analyses (Table S\@ref(tab:hyd-table)).
 
 \
-Methods for Stem Maximum hydraulic conductivity and vulnerability to cavitation closely followed [@Christman.2012] and [@Wolfe.2016].
-Values for some of the species reported here were previously reported by @Dickman.2019 and @Wu.2020.
-The entire dataset and *R* script for the analyses are archived in the NGEE-Tropics archive (Wolfe 2020, To be done).
-
-\
-Canopy branches were cut from 4–10 trees of 26 species across three sites (Table SXX).
-The branch segments were enclosed in opaque plastic bags, and brought to the laboratory for measurements.
-Stem segments used for hydraulic conductivity ($K_h$) measurements were located at least 1 m from the cut end of the branch to minimize the number of open vessels that extended into the segment.
-Maximum vessel length was measured on a subset of samples from each species.
-Among species, maximum vessel length ranged XXX and length of stem between the cut that was made in the air to collect the branch and the section of stem used for $K_h$ measurement was XXX (Table SXX).
-
-\
-Three leaves that were distal to the $K_h$ segment were sealed in plastic bags and the entire branch was enclosed in black plastic bags for at least 30 minutes to stop transpiration.
-Just prior to measurement of $K_h$, leaf water potential of the three bagged leaves was measured with a pressure chamber.
-Since transpiration was stopped by bagging the leaves, leaf water potential was assumed to equal stem water potential (REFERENCE).
-
-\
-For each $K_h$ segment, three measurements of stem water potential were made on stem sections that were located adjacent to the $K_h$ segment.
-Stem sections that were approximately 40 mm in length were placed in nickel-plated aluminum chambers that were connected to Peltier-type psychrometers (Merrill Instruments, Logan, UT) and placed in a water bath at 30°C for 3 h to equilibrate the humidity within the cambers.
-Then water potential was measured using an automated multi-channel micro-voltmeter (CR7, Campbell Scientific, Logan, UT) with a 10 s thermocouple cooling current.
-For samples with water potential < -4 MPa, a 45 s cooling current was used to improve accuracy.
-Each psychrometer probe was calibrated against salt solutions beforehand and the plateau of the psychromatic response curve was used to assess water potential.
-
-\
-The $K_h$ segments were removed from the branches by repeatedly cutting back the branch on both ends while it was submerged in water to relax the hydraulic tension within the xylem ((REFERENCE)).
-Then the ends were shaved clean with a fresh razor.
-Segments were 11–34 cm in length (mean ± SD = 22 ± 4 cm) and 0.2–1.6 cm in diameter (0.7 ± 0.2 cm).
-The segments were placed in a hydraulic conductivity apparatus similar to that described by Sperry et al.
-(1998) except that flow rate was measured with graduated pipettes instead of a balance.
-The perfusion solution was 10 mM KCl filtered to 0.2 mM.
-For each segment, we measured the flow rate at three pressure heads: two that ranged 0.5–2 kPa and at zero pressure.
-Following Torres-Ruiz et al.
-(2012), $K_h$ was calculated as the slope of the flow rate versus pressure gradient (i.e., pressure head divided by segment length) among the three pressure heads.
-Stem area specific hydraulic conductivity was ($K_\textrm{stem}$) was calculated by dividing $K_h$ by the cross-sectional area of the stem.
-To calculate cross-sectional stem area, stem diameter was measured twice perpendicularly on both ends of the segment and averaged, then area was calculated assuming a circular cross section.
+Methods for stem maximum hydraulic conductivity and vulnerability to cavitation closely followed [@Christman.2012] and [@Wolfe.2016].
+Briefly, canopy branches were collected from 4–10 trees of each species.
+Segments measured for hydraulic conductivity were removed from the branches and placed in an apparatus similar to that described by @Sperry.1998 except that flow rate was measured with graduated pipettes instead of a balance.
+Each segment was perfused with a 10 mM KCl solution filtered to 0.2 mM at three pressure heads: two that ranged 0.5–2 kPa and at zero pressure.
+Following @TorresRuiz.2012, hydraulic conductivity was calculated as the slope of the flow rate versus pressure gradient (i.e., pressure head divided by segment length) among the three pressure heads.
+Stem area specific hydraulic conductivity ($K_\textrm{stem}$) was calculated by dividing hydraulic conductivity by the cross-sectional area of the stem.
+For each $K_\textrm{stem}$ segment, three measurements of stem water potential were made on stem sections that were located adjacent to the $K_\textrm{stem}$ segment using psychrometers as described by @Wolfe.2017.
 
@@
 To assess vulnerability to cavitation, the branches were allowed to dry in the lab for 0 – 180 hours before they were measured for stem water potential and $K_\textrm{stem}$.
-We used the water potential measured by the psychrometers because they are a more direct measurement.
-It has been noted that bagged-leaf water potential can deviate from stem water potential, especially when branch segments are relatively dry (REFERENCE).
-For some species, we noted this pattern while for others it was not apparent (Fig SXX).
-For each species, $K_\textrm{stem}$ was plotted against stem water potential and a Weibull curve was fit through the 90th  percentile of the points, following Wolfe et al.
-(2016)).
-The intercept of the curve was interpreted as maximum $K_\textrm{stem}$ and the water potential where the Weibull curve predicted 12, 50, 80, and 88 loss of maximum $K_\textrm{stem}$ were interpreted as $\mathrm{\Psi}$~12,stem~, $\mathrm{\Psi}$~50,stem~, $\mathrm{\Psi}$~80,stem~, and $\mathrm{\Psi}$~88,stem~, respectively.
+For each species, $K_\textrm{stem}$ was plotted against stem water potential and a Weibull curve was fit through the 90th percentile of the points, following @Wolfe.2016.
+The intercept of the curve was interpreted as maximum $K_\textrm{stem}$ and the water potential where the Weibull curve predicted 88% loss of maximum $K_\textrm{stem}$ was interpreted as $\mathrm{\Psi}$~88,stem~.
+
+### Stem hydraulic safety margins
+
+Stem hydraulic safety margin was calculated as the difference between minimum leaf water potentials and $\mathrm{\Psi}$~88,stem~.
+Values for minimum leaf water potentials were obtained from the dataset of @Wolfe.2019, which was collected as part of the 2016 ENSO campaign.
+Pre-dawn and diurnal leaf water potentials were measured on a monthly basis from Feb to May 2016 at San Lorenzo and Parque Natural Metropolitano, Panama, for which leaves were sampled using a canopy crane.
+Measurements in BCI were available only for March 2016, for which leaves were sampled with a pole pruner, as there is no canopy crane on BCI.
+In total 34 species were sampled, with no species re-sampled on another site.
+The raw data and data summaries are archived in the NGEE-Tropics archive [@Wolfe.2019].
+Here, the six species that overlapped with our assessment of ERD and $\mathrm{\Psi}$~88,stem~ were included in analyses.
+We defined species minimum leaf water potential as the minimum of all the diurnal measurements for a species, after taking an average value of measurements on two leaf samples for a given day (Table S\@ref(tab:hyd-table)).
 
@@
 \@ref(eq:b-mod)-\@ref(eq:a-mod)), two lacked data on WSG and five on LMA~lamina~.
-We filled three gaps in LMA using predictions from a strong linear relationship between LMA and LMA~disc~, the latter defined as the mean leaf mass per unit area measured for a 1.483 cm^2^ leaf disc taken to avoid veins (g m^-2^) for leaves receiving direct sunlight ($\overline{\textrm{LMA}} = `r disc.cf[1]` + `r disc.cf[2]`\textrm{LMA}_{disc}$; Adj.
+We filled three gaps in LMA using a linear relationship between LMA and LMA~disc~, the latter defined as the mean leaf mass per unit area measured for a 1.483 cm^2^ leaf disc taken to avoid veins (g m^-2^) for leaves receiving direct sunlight ($\overline{\textrm{LMA}} = `r disc.cf[1]` + `r disc.cf[2]`\textrm{LMA}_{disc}$; Adj.
 $R$^2^ = `r disc.r2`; $p$ `r disc.p`, $n$ = `r disc.n`).
-We filled the two remaining gaps in LMA using predictions from a linear relationship between LMA and LMA~lamina~, as measured during the smaller, $K_\textrm{leaf}$ campaign ($\overline{\textrm{LMA}} = `r lma.cf[1]` + `r lma.cf[2]`\textrm{LMA}_{lamina}$; Adj.
+We filled the two remaining gaps in LMA using a linear relationship between LMA and LMA~lamina~, as measured during the smaller, $K_\textrm{leaf}$ campaign ($\overline{\textrm{LMA}} = `r lma.cf[1]` + `r lma.cf[2]`\textrm{LMA}_{lamina}$; Adj.
 $R$^2^ = `r lma.r2`; $p$ `r lma.p`, $n$ = `r lma.n`).
@@
 ```
+#
+```{r hyd-table, results = 'asis', echo = FALSE}
+knitr::kable(hyd.table, caption = paste0("\\label{lab-hyd-table} Leaf turgor loss point $\\mathrm{\\Psi}$~tlp~, leaf minimum water potential $\\mathrm{\\Psi}$~min~, stem maximum hydraulic conductivity $K_\\textrm{max,stem}$ and stem vulnerability to cavitation $\\mathrm{\\Psi}$~88,stem~ for ", nrow(hyd.table), " species with ERD estimates.
+$\\mathrm{\\Psi}$~tlp~ is sourced from Zailaa et al (to be submitted), $\\mathrm{\\Psi}$~min~ from @Wolfe.2019, $K_\\textrm{max,stem}$ and $\\mathrm{\\Psi}$~88,stem~ from @Wolfe.2020."), booktabs = TRUE, escape = FALSE)
+```
 

### 2020-11-01 manuscript.rmd
Corrected hydraulic trait data sources

@@
 Summary | 195
-Main Text  | 6637
-Introduction | 873
-Methods | 3313
-Results | 859
-Discussion | 1578
+Main Text  | 6659
+Introduction | 860
+Methods | 3370
+Results | 855
+Discussion | 1574
 Keywords | 6
@@
 We evaluated ERD against existing stable hydrogen isotope ratios ($\mathrm{\delta}$^2^H~xylem~) for a subset of six tree species as an independent observation of ERD.
-For a subset of eight species with existing data for above-ground hydraulic traits, we tested the relationship of ERD with xylem vulnerability to embolism, branch hydraulic safety margins, leaf turgor loss point and maximum stem-specific hydraulic conductivity.
-Finally, we analysed species-specific large tree mortality dynamics over 35 years and estimated time a species’ spent below its critical hydraulic threshold to test whether ERD explained risk of drought-induced mortality.
+We evaluated whether ERD was associated with above-ground hydraulic traits sourcing the latter from a set of rare datasets for six to eight species that overlapped with our assessment of ERD.
+We tested the relationship of ERD with xylem vulnerability to embolism, branch hydraulic safety margins, leaf turgor loss point and maximum stem-specific hydraulic conductivity.
+Finally, we analysed species-specific large tree mortality dynamics over 35 years and estimated the time a species’ spent below its critical hydraulic threshold to test whether ERD explained risk of drought-induced mortality.
 
 \
-In the following sections we describe the study site, the models and data used to estimate and evaluate (1) soil water potential dynamics by depth in the whole-soil column and (2) species-specific effective rooting depths.
-These are followed by analyses to address the proposed hypotheses.
+In the following sections, we describe the study site, and the models used to estimate and evaluate soil matric potential by depth in the whole soil column and the species-specific ERD.
+Then we describe the analyses of ERD as related to above-ground hydraulic traits and mortality.
 Details of all datasets used are described in the Supplementary Information.
@@
 We obtained leaf hydraulic vulnerability curves ($K_\textrm{leaf}$ vs.
-$\mathrm{\Psi}$~leaf~) for adult trees of 21 common species from Zailaa et al.
-(to be submitted; see SI for brief description of methods) described as:
+$\mathrm{\Psi}$~leaf~) for adult trees of 21 common species from Zailaa *et* *al.* (to be submitted; see SI for brief description of methods) described as:
 
@@
 
-To evaluate whether ERD was associated with above-ground hydraulic traits (eight species with both datasets), we regressed species ERD against maximum stem hydraulic conductivity ($K_\textrm{max,stem}$), leaf turgor loss point ($\mathrm{\Psi}$~tlp~), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of $K_\textrm{max,stem}$ is lost ($\mathrm{\Psi}$~88,stem~), and hydraulic safety margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~).
-See @Wolfe.2019, @Wolfe.2020 and [SI](#SIData.brett) for data collection and estimation of these variables.
+We evaluated whether ERD was associated with above-ground hydraulic traits sourcing the latter from a set of rare datasets [@Wolfe.2019, @Wolfe.2020] for six to seven species that overlapped with our assessment of ERD.
+We regressed species ERD against maximum stem hydraulic conductivity $K_\textrm{max,stem}$ (*n*=7), leaf turgor loss point $\mathrm{\Psi}$~tlp~ (*n*=7), vulnerability to embolism from cavitation $\mathrm{\Psi}$~88,stem~ measured in terms of pressure at which 88% of $K_\textrm{max,stem}$ is lost (*n*=7), and hydraulic safety margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~ (*n*=6).
+See [SI](#SIData.brett) and Table S\@ref(tab:hyd-table) for data collection and estimation of these variables.
 
@@
 \
-Figure \@ref(fig:erd-hyd): Matrix for pairwise correlations (circles) among Effective Rooting Depth (ERD) and hydraulic properties for seven canopy species found on Barro Colorado Island (Panama).
+Figure \@ref(fig:erd-hyd): Matrix for pairwise correlations (circles) among Effective Rooting Depth (ERD) and hydraulic properties for six to seven canopy species found on Barro Colorado Island (Panama) (see Table S\@ref(tab:hyd-table) for data).
 Each circle is color coded by the Spearman's $r$ with red shades for positive $r$ values and blue shades for negative values.
@@
 
-Stem hydraulic safety margin was calculated as the difference between minimum leaf water potentials and $\mathrm{\Psi}$~88,stem~.
-Values for minimum leaf water potentials were obtained from the dataset of @Wolfe.2019, which was collected as part of the 2016 ENSO campaign.
+Stem hydraulic safety margin was calculated as the difference between minimum leaf water potentials, $\mathrm{\Psi}$~min~, and $\mathrm{\Psi}$~88,stem~.
+Values for $\mathrm{\Psi}$~min~ were obtained from the dataset of @Wolfe.2019, which was collected as part of the 2016 ENSO campaign.
 Pre-dawn and diurnal leaf water potentials were measured on a monthly basis from Feb to May 2016 at San Lorenzo and Parque Natural Metropolitano, Panama, for which leaves were sampled using a canopy crane.
@@
 Here, the six species that overlapped with our assessment of ERD and $\mathrm{\Psi}$~88,stem~ were included in analyses.
-We defined species minimum leaf water potential as the minimum of all the diurnal measurements for a species, after taking an average value of measurements on two leaf samples for a given day (Table S\@ref(tab:hyd-table)).
+We defined species $\mathrm{\Psi}$~min~ as the minimum of all the diurnal measurements for a species, after taking an average value of measurements on two leaf samples for a given day (Table S\@ref(tab:hyd-table)).
 
@@
 ```{r hyd-table, results = 'asis', echo = FALSE}
-knitr::kable(hyd.table, caption = paste0("\\label{lab-hyd-table} Leaf turgor loss point $\\mathrm{\\Psi}$~tlp~, leaf minimum water potential $\\mathrm{\\Psi}$~min~, stem maximum hydraulic conductivity $K_\\textrm{max,stem}$ and stem vulnerability to cavitation $\\mathrm{\\Psi}$~88,stem~ for ", nrow(hyd.table), " species with ERD estimates.
-$\\mathrm{\\Psi}$~tlp~ is sourced from Zailaa et al (to be submitted), $\\mathrm{\\Psi}$~min~ from @Wolfe.2019, $K_\\textrm{max,stem}$ and $\\mathrm{\\Psi}$~88,stem~ from @Wolfe.2020."), booktabs = TRUE, escape = FALSE)
+knitr::kable(hyd.table, caption = paste0("\\label{lab-hyd-table} Minimum leaf water potential $\\mathrm{\\Psi}$~min~, leaf turgor loss point $\\mathrm{\\Psi}$~tlp~, stem maximum hydraulic conductivity $K_\\textrm{max,stem}$ and stem vulnerability to cavitation $\\mathrm{\\Psi}$~88,stem~ for seven species with ERD estimates.
+$\\mathrm{\\Psi}$~min~ is sourced from @Wolfe.2019, while $\\mathrm{\\Psi}$~tlp~, $K_\\textrm{max,stem}$ and $\\mathrm{\\Psi}$~88,stem~ are sourced from from @Wolfe.2020."), booktabs = TRUE, escape = FALSE)
 ```

### 2020-11-01 code/manuscript_data_figs.R

@@ -369,7 +369,7 @@ hyd.table <-  erd.stem.traits %>%
   left_join(erd.sp.names %>% dplyr::rename(sp = Code), by = c("sp")) %>%
   subset(sp %in% erd.sp) %>%
   select(-sp) %>%
-  relocate(Genus, Species, Family, TLP, lwp.min_Predawn) %>%
+  relocate(Genus, Species, Family, lwp.min_Predawn, TLP) %>%
   mutate(TLP = round(TLP, 2))
 
 erd.stem.traits.only <- erd.stem.traits %>%

### 2020-11-02 manuscript.rmd
Changed title and summary with deep roots instead of deep water access

@@
 ---
-title: "Hydraulically-vulnerable trees survive on deep-water access in a tropical forest community"
+title: "Deep-rooted trees have more vulnerable hydraulics, yet lower drought-induced mortality in a tropical forest community"
 author: ''
@@
 ------------- | -------------
-Summary | 195
+Summary | 199
 Main Text  | 6659
@@
 
-- Among the mechanisms that plants employ to tolerate drought, deep water access is arguably the most commonly invoked, but under-studied.
-Vulnerability to embolism and hydraulic safety margins can predict mortality risk at given levels of dehydration, but deep-water access may delay plant dehydration.
-Here, we tested the role of deep-water access in enabling survival within a diverse tropical forest community (Panama) using a novel data-model approach.
+- Among the mechanisms that plants employ to tolerate drought, deep-rooting is arguably the most long-standing, but under-studied.
+Vulnerability to embolism and hydraulic safety margins can predict mortality risk at given levels of dehydration, but deep-rooting may delay plant dehydration.
+Here, we tested the role of deep roots in enabling survival within a diverse tropical forest community (Panama) using a novel data-model approach.
 - We inversely modeled the effective rooting depth (ERD, as the average depth of water extraction), for `r length(erd.sp)` canopy species by linking growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and leaf hydraulic vulnerability curves.
 We validated ERD against existing data for stable hydrogen isotopes of trees and soil.
-- Across species, deeper ERD was associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates during extreme droughts.
-Species with deeper ERD spent less time below critical hydraulic thresholds indicating that trees compensate for water-stress related mortality risk through deeper water access.
-- The role of deep-water access in mitigating mortality of hydraulically-vulnerable trees has important implications for our predictive understanding of forest dynamics under current and future climates.
+- Across species, deeper ERD was associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates during extreme hydrological droughts over 35 years (1981-2015).
+Species with deeper ERD spent less time below critical hydraulic thresholds indicating that trees compensate for water-stress related mortality risk through deep-water access.
+- The role of deep roots in mitigating mortality of hydraulically-vulnerable trees has important implications for our predictive understanding of forest dynamics under current and future climates.
 
 \
-Keywords: hydraulic vulnerability, deep-water access, drought-induced mortality, drought tolerance, rooting depths, tropical forest, life-history strategies, safety margins
+Keywords: rooting depths, deep-water access, hydraulic vulnerability, drought-induced mortality, drought tolerance, tropical forest, safety-efficiency trade-off, hydrological droughts
 
@@
 \
-Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation to stem water storage, deep water access, increased root production, hydraulic redistribution and embolism resistance [@McDowell.2008].
-Deep water access is arguably the most long-standing, yet under-studied mechanism.
-As plant available water varies with depth, trees in the same forest with different rooting depths, depending on species and size [@Meinzer:1999; @Chitra-Tarak.2018; @Brum.2019], differ in their experience during a drought, and thus in their growth and mortality responses [@Chitra-Tarak.2018].
+Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation to stem water storage, deep-rooting, increased root production, hydraulic redistribution and embolism resistance [@McDowell.2008].
+Deep-rootedness is arguably the most long-standing, yet under-studied mechanism.
+As plant-available water varies with depth, trees in the same forest with different rooting depths, depending on species and size [@Meinzer:1999; @Chitra-Tarak.2018; @Brum.2019], differ in their experience during a drought, and thus in their growth and mortality responses [@Chitra-Tarak.2018].
 A key bottleneck in community-wide testing of this mechanism has been a lack of data in both trees' rooting or water-sourcing depths and plant available soil water at those depths.
@@
 We define ERD as the depth at which soil water potential best explained species-specific growth dynamics.
-The ERD model incorporated the impact of atmospheric and soil drought on growth, and was constrained with species-specific leaf vulnerability curves by using existing data for eight species, while developing trait-based proxies for the rest (based on data for a total of 21 species).
+The ERD model incorporated the impact of atmospheric and hydrological drought on growth, and was constrained with species-specific leaf vulnerability curves by using existing data for eight species, while developing trait-based proxies for the rest (based on data for a total of 21 species).
 We evaluated ERD against existing stable hydrogen isotope ratios ($\mathrm{\delta}$^2^H~xylem~) for a subset of six tree species as an independent observation of ERD.
@@
 \
-  In the 50 ha plot, all stems $ge$ 1 cm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every five year through 1985-2015 for growth, mortality as well as recruitment of new stems into the ten mm dbh size class [@Condit.2017].
+  In the 50 ha plot, all stems $\ge$ 1 cm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every five year through 1985-2015 for growth, mortality as well as recruitment of new stems into the ten mm dbh size class [@Condit.2017].
 This inventory represents 321 woody species across the eight censuses.
@@
 We regressed species ERD against species $\textrm{M}_{s,t}$ for each census interval $t$ with ($n$ = `r nrow(mrate.depth.mean)`) and without deciduous species ($n$ = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`).
-To assess whether deeper ERD species escaped the experience of significant soil drought, we calculated for each species the proportion of days in the simulation period (1990-2015) during which soil water potential in the soil layer matching species ERD ($\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$) were below species $\mathrm{\Psi}$~crit~.
+To assess whether deeper ERD species escaped the experience of significant hydrological drought, we calculated for each species the proportion of days in the simulation period (1990-2015) during which soil water potential in the soil layer matching species ERD ($\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$) were below species $\mathrm{\Psi}$~crit~.
 $\mathrm{\Psi}$~crit~ was defined as $\mathrm{\Psi}$~50,leaf~.
@@
 \@ref(fig:swp-droughts)).
-Soil droughts in census intervals 2000-2005 and 2005-2010 distinctly occurred in the wet season, with one to two extreme soil drought years each, depending on the depth location (Fig.
+Soil droughts in census intervals 2000-2005 and 2005-2010 distinctly occurred in the wet season, with one to two extreme hydrological drought years each, depending on the depth location (Fig.
 S\@ref(fig:swp-droughts)).
@@
 However, the difference between precipitation and ET shows that 1985-1990 had three extreme deficit water years, apart from the 1983 El-Niño leading to it [@Condit.2017].
-Analysis of a species' time spent under significant soil drought (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) showed that species with shallower ERD spent a greater proportion of days below critical hydraulic thresholds and increasingly so over 1995-2000, 2000-2005 and 2005-2010 (Fig.
+Analysis of a species' time spent under significant hydrological drought (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) showed that species with shallower ERD spent a greater proportion of days below critical hydraulic thresholds and increasingly so over 1995-2000, 2000-2005 and 2005-2010 (Fig.
 \@ref(fig:erd-time); S\@ref(fig:erd-time-evg)).

### 2020-11-03 manuscript.rmd
1. Title achanged back 2. "rooting depths" changed to "deep-water access" in summary 3. Cut down on site description to get within word limit 3. Added a few missing data sources 4. Added SI note on Leaf turgor loss point

@@
 ---
-title: "Deep-rooted trees have more vulnerable hydraulics, yet lower drought-induced mortality in a tropical forest community"
+title: "Hydraulically-vulnerable trees survive on deep-water access during droughts in a tropical forest"
 author: ''
@@
 
-Chitra-Tarak, Rutuja*, Chonggang Xu, Kristina Anderson-Teixeira, Jeff Chambers, Matteo Detto, Boris Faybishenko, Rosie Fisher, Ryan Knox, Charlie Koven, Lara Kueppers, Nobert Kunert, Stefan J.
+Chitra-Tarak, Rutuja*, Chonggang Xu, Salomón Aguilar, Kristina Anderson-Teixeira, Jeff Chambers, Matteo Detto, Boris Faybishenko, Rosie Fisher, Ryan Knox, Charlie Koven, Lara Kueppers, Nobert Kunert, Stefan J.
 Kupers, Nathan G.
-McDowell, Brent Newman, Steve Paton, Laurent Ruiz, Lawren Sack, Jeffrey M.
+McDowell, Brent Newman, Steve Paton, Rolando Pérez, Laurent Ruiz, Lawren Sack, Jeffrey M.
 Warren, Brett Wolfe, Cynthia Wright, S.
@@
 ------------- | -------------
-Summary | 199
-Main Text  | 6659
-Introduction | 860
-Methods | 3370
+Summary | 197
+Main Text  | 6492
+Introduction | 857
+Methods | 3192
 Results | 855
-Discussion | 1574
+Discussion | 1588
 Keywords | 6
@@
 
-- Among the mechanisms that plants employ to tolerate drought, deep-rooting is arguably the most long-standing, but under-studied.
-Vulnerability to embolism and hydraulic safety margins can predict mortality risk at given levels of dehydration, but deep-rooting may delay plant dehydration.
-Here, we tested the role of deep roots in enabling survival within a diverse tropical forest community (Panama) using a novel data-model approach.
+- Among the mechanisms that plants employ to tolerate drought, deep-water access is arguably the most long-standing, but under-studied.
+Vulnerability to embolism and hydraulic safety margins can predict mortality risk at given levels of dehydration, but deep-water access may delay plant dehydration.
+Here, we tested the role of deep-water access in enabling survival within a diverse tropical forest community (Panama) using a novel data-model approach.
 - We inversely modeled the effective rooting depth (ERD, as the average depth of water extraction), for `r length(erd.sp)` canopy species by linking growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and leaf hydraulic vulnerability curves.
 We validated ERD against existing data for stable hydrogen isotopes of trees and soil.
-- Across species, deeper ERD was associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates during extreme hydrological droughts over 35 years (1981-2015).
+- Across species, deeper ERD was associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates during extreme hydrological droughts over 1981-2015.
 Species with deeper ERD spent less time below critical hydraulic thresholds indicating that trees compensate for water-stress related mortality risk through deep-water access.
-- The role of deep roots in mitigating mortality of hydraulically-vulnerable trees has important implications for our predictive understanding of forest dynamics under current and future climates.
+- The role of deep-water access in mitigating mortality of hydraulically-vulnerable trees has important implications for our predictive understanding of forest dynamics under current and future climates.
 
@@
 \
-Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation to stem water storage, deep-rooting, increased root production, hydraulic redistribution and embolism resistance [@McDowell.2008].
-Deep-rootedness is arguably the most long-standing, yet under-studied mechanism.
+Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation to stem water storage, deep-water access, increased root production, hydraulic redistribution and embolism resistance [@McDowell.2008].
+Deep-water access is arguably the most long-standing, yet under-studied mechanism.
 As plant-available water varies with depth, trees in the same forest with different rooting depths, depending on species and size [@Meinzer:1999; @Chitra-Tarak.2018; @Brum.2019], differ in their experience during a drought, and thus in their growth and mortality responses [@Chitra-Tarak.2018].
@@
 We evaluate how rooting depth is linked to above-ground hydraulic traits and mortality rates over a 35 year period which experienced droughts of a variety of intensity, frequency and duration.
-Our hypotheses (see Table \@ref(tab:hypo)) are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem due to greater and more reliable water availability, whereas shallow rooted species are associated with less efficient but safer hydraulic traits due to more variable water availability; and that, (ii) deep rooted species have lower mortality rates during droughts due to lower realized risk of hydraulic failure compared to shallow rooted species.
-To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality rates for a species-rich forest.
+Our hypotheses (see Table \@ref(tab:hypo)) are that (i) deep-rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem due to greater and more reliable water availability at depth, whereas shallow-rooted species are associated with less efficient but safer hydraulic traits due to more variable availability of shallow soil-water; and that, (ii) deep-rooted species have lower mortality rates during droughts due to lower realized risk of hydraulic failure compared to shallow-rooted species.
+To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.
 
@@
 
-\
-In the following sections, we describe the study site, and the models used to estimate and evaluate soil matric potential by depth in the whole soil column and the species-specific ERD.
-Then we describe the analyses of ERD as related to above-ground hydraulic traits and mortality.
-Details of all datasets used are described in the Supplementary Information.
-
 ## Study site description
@@
 We conducted this study at Barro Colorado Island (BCI), Panama.
-The site is extensively characterized in terms of long-term forest demography, hydrology and plant functional ecology, including data for stable water isotopes on trees and soils [@Meinzer:1999].
-The 1,560ha island was formed after the river Chagres was dammed in 1914 to create Gatun Lake and complete the Panama Canal.
 The entire island is forested and classified as a tropical moist forest in the Holdridge Life Zone system.
@@
 Rainfall at BCI is seasonal with a mean annual total of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a pronounced dry season from mid-December through April with a mean total of 292 mm [@Paton.2020].
-Among the tree *species* found in the forest canopy, 28% are identified to be at least partly deciduous during the dry season, but at a maximum only 6.3% *trees* in the canopy are deciduous, amounting to only 9.7% of the total forest crown area [@Condit:2000].
 
 \
-  In the 50 ha plot, all stems $\ge$ 1 cm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every five year through 1985-2015 for growth, mortality as well as recruitment of new stems into the ten mm dbh size class [@Condit.2017].
-This inventory represents 321 woody species across the eight censuses.
-Aside from a two ha patch of ~100-year-old secondary forest at the north-eastern corner of the plot the remaining 48ha represent old growth forest that was never cleared for agriculture.
-The plot (1000 m $\times$ 500 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001].
-The plot is situated in the Conrad catchment.
-Soils of the plot are homogeneous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot [@Baillie.2017].
+  In the 50 ha plot, all stems $\ge$ 1 cm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every five year through 1985-2015 for growth, mortality as well as recruitment of new stems into the 1 cm dbh size class [@Condit.2017].
+This inventory represents 321 woody species across the eight censuses, 28% of which are at least partly dry-season deciduous [@Condit:2000].
+The elevation range in the plot is only 40 m (120-160 m) [@Harms:2001] and soils are homogeneous with red light clays accounting for 72.4% of the plot [@Baillie.2017].
 The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil.
@@
 \
-  ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\mathrm{\Delta}W_{\textrm{can}}$, surface water $\mathrm{\Delta}W_{\textrm{sfc}}$, soil water $\mathrm{\Delta}W_{\textrm{liq},z}$ at depth $z$ and water in an unconfined aquifer $\mathrm{\Delta} W_{a}$ (omitting processes relevant to snow, wetlands or lakes).
+  ELM-FATES simulates interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\mathrm{\Delta}W_{\textrm{can}}$, surface water $\mathrm{\Delta}W_{\textrm{sfc}}$, soil water $\mathrm{\Delta}W_{\textrm{liq},z}$ at depth $z$ and water in an unconfined aquifer $\mathrm{\Delta} W_{a}$ (omitting processes relevant to snow, wetlands or lakes).
 Conservation of these terms is expressed as follows:
@@
 \@ref(eq:water-balance), determined their ranges based on literature for the study site, when available, else for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling (LHS) of this global parameter space.
-In the Conrad catchment, existing data on soil hydraulic conductivity for the depths of 12.5 and 60 cm showed a decline of conductivity by depth (@Kinner:2004; Table \@ref(tab:param)).
+In the Conrad catchment, existing data on soil hydraulic conductivity for the depths of 12.5 and 60 cm showed a decline of conductivity by depth (@Godsey:2004; Table \@ref(tab:param)).
 We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths <12.5 cm and >60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig.
@@
 \
-We ran ELM-FATES for the 5000-member parameter ensemble with hourly climate drivers measured at the BCI meteorological station for the period from 1985-2018 (Faybishenko et al.
-2020) initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode.
+We ran ELM-FATES for the 5000-member parameter ensemble with hourly climate drivers measured at the BCI meteorological station over 1985-2018 initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode (see https://biogeodb.stri.si.edu/physical_monitoring/research/barrocolorado for meteorological data and @Faybishenko.2018 for QA/QC procedures).
 As our key interest here was on deriving soil water availability, we ran ELM-FATES in a lower-complexity configuration: static stand structure, and with a single plant functional type (PFT) of evergreen trees.
@@
 We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq.
-\@ref(eq:water-balance), namely, (i) evapotranspiration (ET) from the eddy flux tower ($\cong$ $E_{v} + E_{g}$; 2012-2017; see [SI](#SIData.matteo))), (ii) stream discharge ($q_\textrm{run}$) from the Conrad catchment ($\cong$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC or $\theta_{z}$ $\cong$ $W_{\textrm{liq,} z}$) from two sources: (1) a long-term (2012-2018) record of average VWC we obtained with vertical time domain reflectometry (TDR) probes from three locations near the eddy flux tower over 0-15 cm depth   ($\mathrm{\theta}_\textrm{0-0.15 m}$), and (2) existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m with 1299 samples in total covering all soil types and habitats (see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter,  $\mathrm{\theta}_{\textrm{0-0.15 m}}$, $\mathrm{\theta}_{\textrm{0.4 m}}$ and $\mathrm{\theta}_{\textrm{1 m}}$ refer to plot-wide averages).
+\@ref(eq:water-balance), namely, (i) evapotranspiration (ET) from the eddy flux tower ($\cong$ $E_{v} + E_{g}$; 2012-2017; see [SI](#SIData.matteo))), (ii) stream discharge ($q_\textrm{run}$) from the Conrad catchment ($\cong$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see @Paton.2019 and [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC or $\theta_{z}$ $\cong$ $W_{\textrm{liq,} z}$) from two sources: (1) a long-term (2012-2018) record of average VWC we obtained with vertical time domain reflectometry (TDR) probes from three locations near the eddy flux tower over 0-15 cm depth ($\mathrm{\theta}_\textrm{0-0.15 m}$), and (2) existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m with 1299 samples in total covering all soil types and habitats (see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\mathrm{\theta}_{\textrm{0-0.15 m}}$, $\mathrm{\theta}_{\textrm{0.4 m}}$ and $\mathrm{\theta}_{\textrm{1 m}}$ refer to plot-wide averages).
 
@@
 We ran ELM-FATES with the chosen 100 ensemble members from 1985-2018, with the first 5 years used for model spin-up.
-Extreme soil droughts were identified by depth $z$ as days that fell below the 5^th^ percentile of $\mathrm{\Psi}_{\textrm{soil},z}$ for a given day of the year.
+Extreme hydrological droughts were identified by depth $z$ as days that fell below the 5^th^ percentile of $\mathrm{\Psi}_{\textrm{soil},z}$ for a given day of the year.
 
@@
 
-For growth estimates, to minimize the effect of light variation among trees, we selected only large trees ($\ge$ 30 cm dbh, the diameter at breast height) and also species whose maximum height was $\ge$ 30 m (hereafter, canopy species) and thus are likely to be fully exposed to the sun.
-Calculating individual tree growth rates across six 5-yearly censuses (1990-2015), removing outliers, accounting for temporal changes in growth rates due to increasing tree size (see SI), we obtained species-level median time series of growth rates ($G_{t}$, cm yr^-1^) used in the ERD model for only those species ($n$ = `r length(erd.sp)`) with complete records on at least three trees (median $n$ = `r median(g.n$n)`, max $n$ = `r max(g.n$n)` trees per species).
+For growth estimates, to minimize the effect of light variation among trees, we selected only large trees ($\ge$ 30 cm dbh, the diameter at breast height) in the 50 ha plot and also species whose maximum height was $\ge$ 30 m (hereafter, canopy species) and thus are likely to be fully exposed to the sun.
+Calculating individual tree growth rates across six 5-yearly censuses (1990-2015), removing outliers, accounting for temporal changes in growth rates due to increasing tree size (see @Condit.2019a; @Condit.2019b and SI), we obtained species-level median time series of growth rates ($G_{t}$, cm yr^-1^) used in the ERD model for only those species ($n$ = `r length(erd.sp)`) with complete records on at least three trees (median $n$ = `r median(g.n$n)`, max $n$ = `r max(g.n$n)` trees per species).
 
@@
 
-To test whether ERD plays a role in alleviating mortality risk, we calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$ 30 m) and average abundance of at least 20 trees ($n$ = `r length(erd.mrate.sp)`).
+To test whether ERD plays a role in mitigating mortality risk, we calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$ 30 m) and average abundance of at least 20 trees ($n$ = `r length(erd.mrate.sp)`) in the 50ha plot [@Condit.2019a].
 For each of the seven census intervals $t$ in the 35-yr record (1981, 1985, 1990, 1995, 2000, 2005, 2010, 2015), mortality rate, $\textrm{M}_{t}$ (% yr^-1^) for species $s$ for two successive censuses, $c_{1}$ and $c_{2}$, was calculated as $\textrm{M}_{s,t} = \frac{D_{s,c_{1}}}{N_{s,c_{2}}}\times\frac{100}{\textrm{d}}$, where $N_{s,c_{1}}$ and $D_{s,c_{2}}$ are the total number of trees of species $s$ present in $c_{1}$ and dead in $c_{2}$, respectively, in the 50 ha plot and $d$ is the duration based on mean dates of $c_{1}$ and $c_{2}$.
@@
 \@ref(fig:swp-droughts)).
-Soil droughts in census intervals 2000-2005 and 2005-2010 distinctly occurred in the wet season, with one to two extreme hydrological drought years each, depending on the depth location (Fig.
+Hydrological droughts in census intervals 2000-2005 and 2005-2010 distinctly occurred in the wet season, with one to two extreme hydrological drought years each, depending on the depth location (Fig.
 S\@ref(fig:swp-droughts)).
@@
 \@ref(fig:erd-mort-census)).
-Census intervals 2000-2005 and 2005-2010 were also the periods during which extreme soil droughts distinctly occurred in the wet season, effectively extending the dry season into the wet season (Fig.
+Census intervals 2000-2005 and 2005-2010 were also the periods during which extreme hydrological droughts distinctly occurred in the wet season, effectively extending the dry season into the wet season (Fig.
 \@ref(fig:swp-droughts)).
@@
 S\@ref(fig:swp-all-depths)).
-The kind of droughts that buffered the deep rooted species from mortality in our study were repeated annual droughts that extended the dry season into the wet season (Fig.
+The kind of droughts that buffered the deep-rooted species from mortality in our study were repeated annual droughts that extended the dry season into the wet season (Fig.
 \@ref(fig:swp-droughts)).
@@
 Establishing relationships between environment, traits and demographic outcomes of plants is imperative for developing a predictive plant ecology.
-Trees realized water environments and water-access are nonetheless rarely studied.
-To our knowledge, this is the first study to demonstrate a mechanistic link between realized drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a tree community.
+Trees rooting depths and actual water environments through hydrological rather than meteorological droughts are nonetheless rarely studied.
+To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.
 We report here that deep-water access plays a role in alleviating mortality of hydraulically vulnerable trees.
@@
 Hubbell and Robin B.
-Foster and sustained for many years by Richard Condit, is now managed by Stuart Davies, Suzanne Lao, and Rolando Perez under the ForestGEO program of the Smithsonian Tropical Research in Panama.
+Foster and sustained for many years by Richard Condit, is now managed by Rolando Perez, Suzanne Lao and Stuart Davies and under the ForestGEO program of the Smithsonian Tropical Research in Panama.
 Numerous organizations have provided funding, principally the US National Science Foundation, and hundreds of field workers have contributed.
@@
 NK, JZ, KAT, LS for leaf hydraulics, BW for stem hydraulics.
+RP and SA for the BCI 50ha plot censuses.
 JW for maximum tree height, WSG, LMA, deciduousness, leaf fall and leaf lifetime.
@@
 
-To be uploaded on NGET-Archive and site DOIs along with manuscript acceptance:
-(1) BF to upload climate forcing data
-(2) RCT to upload R scripts for ELM-FATES parameter generations, ELM-FATES modified code, and all R scripts to reproduce manuscript.
-\
-Barro Colorado Forest Census Plot Data can be obtained online [@Condit.2019a; @Condit.2019b].
+ELM-FATES source code, simulation outputs for best-fit parameter ensemble members and all the R scripts to reproduce the manuscript are archived (DOI:).
+Data sources for all other datasets used are provided throughout the manuscript.
 
@@
 \
-Figure \@ref(fig:swp-droughts): Occurrence of extreme soil droughts by census intervals (horizontal panels) at three representative depths ($z$): 0.1, 0.6 and 1 m (vertical panels).
+Figure \@ref(fig:swp-droughts): Occurrence of extreme hydrological droughts by census intervals (horizontal panels) at three representative depths ($z$): 0.1, 0.6 and 1 m (vertical panels).
 Mean (black lines) and lower half of 95% distribution (grey areas) of soil water potential ($\mathrm{\Psi}_{\textrm{soil},z}$) for a given DOY and depth $z$ are shown (same across census intervals).
@@
 The ANN was implemented using the Neural Network Toolbox in Matlab 2014a. 
+\
+The associated data can be requested from https://ameriflux.lbl.gov.
 
@@
 Beginning in 2016 we additionally installed three horizontal probes at depths of 0.15, 0.4 and 1 m and we applied the same calibration equation to these data (2016-2018) as that for the vertical probe.
+These data are archived at https://ngt-data.lbl.gov/.
 
@@
 
+### Leaf turgor loss point
+
+For each tree species measured for $K_\textrm{stem}$ and $\mathrm{\Psi}$~88,stem~, two to six leaves were measured for $\mathrm{\Psi}$~tlp~ with pressure-volume analysis following [@Koide.1991].
+Leaf water potential was measured on leaf discs with psychrometers (J.R.D.
+Merrill Specialty Equipment, Logan, Utah, USA).
+This method has been shown to give results similar to those of the pressure chamber method [@Nardini.2008].
+To construct the pressure-volume curves, leaf discs were repeatedly measured for water potential and mass.
+Between measurements the discs were bench dried for 5-30 minutes.
+Dry mass was determined after oven drying the discs at 70°C.
+See [@Wolfe.2020] for archived data.
+
 ### Stem hydraulic safety margins
@@
 Measurements in BCI were available only for March 2016, for which leaves were sampled with a pole pruner, as there is no canopy crane on BCI.
-In total 34 species were sampled, with no species re-sampled on another site.
+In total 25 species were sampled, with no species re-sampled on another site.
 The raw data and data summaries are archived in the NGEE-Tropics archive [@Wolfe.2019].
@@
 knitr::kable(hyd.table, caption = paste0("\\label{lab-hyd-table} Minimum leaf water potential $\\mathrm{\\Psi}$~min~, leaf turgor loss point $\\mathrm{\\Psi}$~tlp~, stem maximum hydraulic conductivity $K_\\textrm{max,stem}$ and stem vulnerability to cavitation $\\mathrm{\\Psi}$~88,stem~ for seven species with ERD estimates.
-$\\mathrm{\\Psi}$~min~ is sourced from @Wolfe.2019, while $\\mathrm{\\Psi}$~tlp~, $K_\\textrm{max,stem}$ and $\\mathrm{\\Psi}$~88,stem~ are sourced from from @Wolfe.2020."), booktabs = TRUE, escape = FALSE)
+$\\mathrm{\\Psi}$~min~ is sourced from @Wolfe.2019, while $\\mathrm{\\Psi}$~tlp~, $K_\\textrm{max,stem}$ and $\\mathrm{\\Psi}$~88,stem~ are sourced from @Wolfe.2020.
+$\\mathrm{\\Psi}$~min~, $\\mathrm{\\Psi}$~tlp~ and $\\mathrm{\\Psi}$~88,stem~ are in MPa, while $K_\\textrm{max,stem}$ in kg m^-1^ s^-1^ MPa^-1^."), booktabs = TRUE, escape = FALSE)
 ```

### 2020-11-04 manuscript.rmd
Incorporated Brent's edits

@@
 
-Chitra-Tarak, Rutuja*, Chonggang Xu, Salomón Aguilar, Kristina Anderson-Teixeira, Jeff Chambers, Matteo Detto, Boris Faybishenko, Rosie Fisher, Ryan Knox, Charlie Koven, Lara Kueppers, Nobert Kunert, Stefan J.
-Kupers, Nathan G.
-McDowell, Brent Newman, Steve Paton, Rolando Pérez, Laurent Ruiz, Lawren Sack, Jeffrey M.
-Warren, Brett Wolfe, Cynthia Wright, S.
-Joseph Wright, Joseph Zailaa, Sean M.
-McMahon
+Chitra-Tarak, Rutuja*^1,2^, Chonggang Xu^1^, Salomón Aguilar^3^, Kristina Anderson-Teixeira^3,4^, Jeff Chambers^5^, Matteo Detto^3,6^, Boris Faybishenko^5^, Rosie A.
+Fisher^7^, Ryan Knox^5^, Charlie Koven^5^, Lara Kueppers^5,8^, Nobert Kunert^3,4,9^, Stefan J.
+Kupers^10^, Nathan G.
+McDowell^11,12^, Brent D.
+Newman^1^, Steve Paton^3^, Rolando Pérez^3^, Laurent Ruiz^13,14,15^, Lawren Sack^16^, Jeffrey M.
+Warren^17^, Brett T.
+Wolfe^3,18^, Cynthia Wright^17^, S.
+Joseph Wright^3^, Joseph Zailaa^4,16,19^, Sean M.
+McMahon^2,3^
 
@@
 \
+
+1. Los Alamos National Laboratory, Earth and Environmental Sciences Division (EES-14)
+MS J495, Los Alamos, NM, USA 87545-1663
+2. Smithsonian Environmental Research Center, 647 Contees Wharf Road, Edgewater, MD, USA 21037-0028
+3. Smithsonian Tropical Research Institute, Panama, PA
+4. Smithsonian Conservation Biology Institute, Conservation Ecology Center, Front Royal, VA, USA
+5. Lawrence Berkeley National Laboratory, Climate and Ecosystem Sciences Division, Berkeley, CA, USA
+6. Princeton University, Ecology and Evolutionary Biology, Princeton, NJ, USA
+7. National Center for Atmospheric Research, Climate and Global Dynamics Division
+Boulder, CO, USA
+8. University of California Berkeley, Energy and Resources Group, 310 Barrows Hall #3050
+Berkeley, CA, USA 94720
+9. University of Natural Resources and Life Sciences Vienna, Institute of Botany, Department of Integrative Biology and Biodiversity Research, Gregor‐Mendel‐Str 33, A‐1190 Wien, Wien, AT
+10. German Centre for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig, Computational Forest Ecology, Leipzig, Saxony, DE
+11. Atmospheric Sciences and Global Change Division, Pacific Northwest National Lab, PO Box 999, Richland, WA, 99352
+12. School of Biological Sciences, Washington State University, PO Box 644236, Pullman, WA 99164-4236
+13. Indian Institute of Science, Indo-French Cell for Water Sciences, Bangalore, Karnataka, IN
+14. IRD, CNRS, UPS, UMR GET, Toulouse, FR
+15. INRAE, Institut Agro, UMR SAS, Rennes, FR
+16. University of California Los Angeles, Ecology and Evolutionary Biology, 612 Charles E. Young Drive South, Los Angeles, CA, USA 90095
+17. Oak Ridge National Laboratory, Environmental Sciences Division, Oak Ridge, TN, USA
+18. School of Renewable Natural Resources, Louisiana State University Agricultural Center, Baton Rouge, LA 70803, USA
+19. California State University Los Angeles, Biological Sciences Department, Los Angeles, CA, USA
 
@@
 Rainfall at BCI is seasonal with a mean annual total of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a pronounced dry season from mid-December through April with a mean total of 292 mm [@Paton.2020].
-
-\
-  In the 50 ha plot, all stems $\ge$ 1 cm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every five year through 1985-2015 for growth, mortality as well as recruitment of new stems into the 1 cm dbh size class [@Condit.2017].
-This inventory represents 321 woody species across the eight censuses, 28% of which are at least partly dry-season deciduous [@Condit:2000].
-The elevation range in the plot is only 40 m (120-160 m) [@Harms:2001] and soils are homogeneous with red light clays accounting for 72.4% of the plot [@Baillie.2017].
+In the 50ha (1000 m × 500 m) old growth forest plot, all stems $\ge$ 1 cm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every five year through 1985-2015 for growth, mortality as well as recruitment of new stems into the 1 cm dbh size class [@Condit.2017].
+This inventory represents 321 woody species, 28% of which are at least partly dry-season deciduous [@Condit:2000].
+The elevation range in the plot is only 40 m (120-160 m) [@Harms:2001] and soils are homogeneous with red light clays accounting for 72% of the plot [@Baillie.2017].
 The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil.
@@
 To parameterize ELM-FATES, we identified eleven parameters relevant for Eq.
-\@ref(eq:water-balance), determined their ranges based on literature for the study site, when available, else for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling (LHS) of this global parameter space.
-In the Conrad catchment, existing data on soil hydraulic conductivity for the depths of 12.5 and 60 cm showed a decline of conductivity by depth (@Godsey:2004; Table \@ref(tab:param)).
+\@ref(eq:water-balance), determined their ranges based on literature for the study site, else for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling (LHS) of this global parameter space.
+In the Conrad catchment soil hydraulic conductivity decreases with depth from 12.5 and 60 cm (@Godsey:2004; Table \@ref(tab:param)).
 We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths <12.5 cm and >60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig.
 S\@ref(fig:ksat-profiles)).
-For greater accuracy, instead of the ELM default soil texture-based pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content (GWC) vs.
+For greater accuracy, instead of the ELM default pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content (GWC) vs.
 $\mathrm{\Psi}$~soil~ for depths of 0.15, 0.4 and 1 m [@Kupers:2019].
@@
 The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of a dry-deciduous PFT did not significantly alter results (not shown).
-Holding the stand structure static rather than allowing it to emerge from physiologically-mediated competition between trees removes a set of internal model feedbacks and thus enables a cleaner experimental design where the only differences between model simulations are the direct physiological effects of perturbed model parameters.
+Holding the stand structure static rather than allowing it to emerge from physiologically-mediated competition between trees removes a set of internal model feedbacks and thus the only differences between model simulations are driven by direct physiological effects of perturbed model parameters.
 
@@
 We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq.
-\@ref(eq:water-balance), namely, (i) evapotranspiration (ET) from the eddy flux tower ($\cong$ $E_{v} + E_{g}$; 2012-2017; see [SI](#SIData.matteo))), (ii) stream discharge ($q_\textrm{run}$) from the Conrad catchment ($\cong$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see @Paton.2019 and [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC or $\theta_{z}$ $\cong$ $W_{\textrm{liq,} z}$) from two sources: (1) a long-term (2012-2018) record of average VWC we obtained with vertical time domain reflectometry (TDR) probes from three locations near the eddy flux tower over 0-15 cm depth ($\mathrm{\theta}_\textrm{0-0.15 m}$), and (2) existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m with 1299 samples in total covering all soil types and habitats (see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\mathrm{\theta}_{\textrm{0-0.15 m}}$, $\mathrm{\theta}_{\textrm{0.4 m}}$ and $\mathrm{\theta}_{\textrm{1 m}}$ refer to plot-wide averages).
+\@ref(eq:water-balance), namely, (i) evapotranspiration (ET) from the eddy flux tower ($\cong$ $E_{v} + E_{g}$; 2012-2017; see [SI](#SIData.matteo))), (ii) local stream discharge ($q_\textrm{run}$) ($\cong$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see @Paton.2019 and [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC or $\theta_{z}$ $\cong$ $W_{\textrm{liq,} z}$) from two sources: (1) a long-term (2012-2018) record of average VWC (vertical time domain reflectometry (TDR) probes from three locations near the eddy flux tower over 0-15 cm depth ($\mathrm{\theta}_\textrm{0-0.15 m}$)), and (2) plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 (depths of 0.15, 0.4 and 1 m, 1299 samples covering all soil types and habitats (see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\mathrm{\theta}_{\textrm{0-0.15 m}}$, $\mathrm{\theta}_{\textrm{0.4 m}}$ and $\mathrm{\theta}_{\textrm{1 m}}$ refer to plot-wide averages).
 
@@
 \
-Preliminary analyses showed that minimizing the objective function $\overline{O}$, that is, simultaneously fitting the model to all states and fluxes did not capture the soil moisture dynamics at 1 m depth well, our key priority for model calibration.
+Preliminary analyses showed that simultaneously fitting the model to all states and fluxes did not capture the soil moisture dynamics at 1 m depth well, our key priority for model calibration.
 To ensure model fit at 1 m depth, we first chose 300 parameter ensemble members that minimized $\textrm{RMSE}^{*}({\mathrm{\theta}_{\textrm{1 m}}}, W_{\textrm{liq,`r soil.depths[8]` m}})$ (< 0.2), and among those 300 ensemble members we chose 100 with the lowest values for the objective function $\overline{O}$.
@@
 We evaluated ELM-FATES by calculating RMSE between simulated and observed long-term daily VWC for the depths of 0.1, 0.4 and 1 m (2016-2018), based on a dataset we had left-out during calibration.
-We obtained these observations from three horizontal TDR probes at the depths of 0.1, 0.4 and 1 m at a point location near the vertical probe (see [SI](#SIData.VWC)).
+We obtained these observations from three horizontal TDR probes at the depths of 0.1, 0.4 and 1 m at a point location near the vertical probes (see [SI](#SIData.VWC)).
 
@@
 
-To estimate effective rooting depth, we tested alternate empirical models for species 5-yearly census growth (Eq.
-\@ref(eq:M-chosen) and those in SI), which use effective rooting depth as an input.
+We tested alternate empirical models for species 5-yearly census growth, which use effective rooting depth as an input (Eq.
+\@ref(eq:M-chosen) and those in SI).
 Many factors limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019].
@@
 $K_\textrm{leaf}$ in our models thus incorporates the effects of rooting depth and rhizosphere water potential.
-This approach, however, does not account for the buffering effect of stem water storage capacitance on $\mathrm{\Psi}$~leaf~, and may thus overestimate ERD (see Discussion).
-VPD affects growth non-linearly, with growth increasing with VPD up-to a VPD threshold, then decreasing as leaf pores (stomata) close reducing water-uptake [@Grossiord.2020].
-Apart from stomatal control, leaf-drop may further limit water-uptake and growth.
+This approach does not account for the buffering effect of stem water storage capacitance on $\mathrm{\Psi}$~leaf~, and may overestimate ERD (see Discussion).
+VPD affects growth non-linearly, with growth increasing with VPD up-to a VPD threshold, then decreasing as leaf pores (stomata) close reducing water-uptake [@Grossiord.2020], and leaf-drop may further limit water-uptake and growth.
 We therefore regressed species-specific growth against multiplicative or additive effects of one or more factors calculated daily and averaged over 5-yearly censuses: VPD, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1.
 Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting.
-Due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season at BCI, the impact of VPD may as well be a combined impact of not only temperature and humidity, but also radiation.
+Incorporation of radiation did not improve model fitting, possibly due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season at BCI.
 
@@
 \
- To account for the non-linear impact of VPD on growth, we used a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see SI) and used GPP standardized between 0 and 1 (hereafter, VPD*) in Eq.
-\@ref(eq:M-chosen).
 
@@
 \begin{equation}
-A_{s} = `r acf[1]` `r acf[2]`B_{s} `r acf[3]`B_{s}^2 + `r acf[4]`B_{s}\textrm{LMA}_{s} + `r acf[5]`\textrm{LMA}_{s} `r acf.6`\textrm{LMA}_{s}^2 + \epsilon_{s,a}
-(\#eq:a-mod)
-\end{equation}
-
-\begin{equation}
 \begin{split}
@@
 \end{split}
+\end{equation}
+
+\begin{equation}
+A_{s} = `r acf[1]` `r acf[2]`B_{s} `r acf[3]`B_{s}^2 + `r acf[4]`B_{s}\textrm{LMA}_{s} + `r acf[5]`\textrm{LMA}_{s} `r acf.6`\textrm{LMA}_{s}^2 + \epsilon_{s,a}
+(\#eq:a-mod)
 \end{equation}
@@
 
-## Effective rooting depths, experienced drought and mortality
+## Effective rooting depths, mortality and hydrological droughts
 
@@
 \@ref(fig:erd-mort-census-evg)).
-When deciduous species were included, ERD explained lower but significant proportion of variation in mortality in those intervals ($R$^2^ = `r m.r2[1]`, `r m.r2[2]` and `r m.r2[3]`% and $p$ = `r m.p.90`, `r m.p.05` and `r m.p.10`, respectively; Fig.
+When deciduous species were included, ERD explained lower but still significant proportion of variation in mortality in those intervals ($R$^2^ = `r m.r2[1]`, `r m.r2[2]` and `r m.r2[3]`% and $p$ = `r m.p.90`, `r m.p.05` and `r m.p.10`, respectively; Fig.
 \@ref(fig:erd-mort-census)).
@@
 However, the difference between precipitation and ET shows that 1985-1990 had three extreme deficit water years, apart from the 1983 El-Niño leading to it [@Condit.2017].
-Analysis of a species' time spent under significant hydrological drought (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) showed that species with shallower ERD spent a greater proportion of days below critical hydraulic thresholds and increasingly so over 1995-2000, 2000-2005 and 2005-2010 (Fig.
+The analysis of a species' time spent under significant hydrological drought (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) showed that species with shallower ERD spent a greater proportion of days below critical hydraulic thresholds and increasingly so over 1995-2000, 2000-2005 and 2005-2010 (Fig.
 \@ref(fig:erd-time); S\@ref(fig:erd-time-evg)).
@@
 \
-Our result--of lower mortality in deep-rooted trees--contrasts with the finding of @Chitra-Tarak.2018 (hereafter, CT2018) in which inversely modelled, deeper ERD species died *more* in a rare, prolonged drought.
+Our result--of lower mortality in deep-rooted trees--contrasts with the inverse model finding of @Chitra-Tarak.2018 (hereafter, CT2018) in which deeper ERD species had higher mortality in a rare, prolonged drought.
 The hydrological model in CT2018 revealed that the multi-year drought exhausted the deep soil and even bedrock water availability (also see, @Goulden.2019 and @Ivanov:2012).
-In contrast, our hydrological modeling at BCI finds that during the study period, $\mathrm{\Psi}$~soil~ below 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and droughts, that is above $\mathrm{\Psi}$~crit~ of the most sensitive tree species (-0.28 MPa).
-The difference in the results shown here are likely attributable to site differences in the precipitation regime: with mean annual rainfall of 1095 in CT2018 compared to `r round(rain.man.stats$rain)`mm at BCI, and precipitation to potential evapotranspiration ratio nearly one in CT2018 compared to nearly two in BCI.
-Thus at BCI deep soil layers are effectively recharged annually (Fig.
-S\@ref(fig:swp-all-depths)).
-The kind of droughts that buffered the deep-rooted species from mortality in our study were repeated annual droughts that extended the dry season into the wet season (Fig.
-\@ref(fig:swp-droughts)).
-This highlights the role of site-specific hydrology in how ERD modulates mortality-risk in different kinds of droughts.
+In contrast, our hydrological modeling at BCI finds that during the study period, $\mathrm{\Psi}$~soil~ below 1 m never went below $\mathrm{\Psi}$~crit~ of the most sensitive tree species (-0.28 MPa) for all the dry seasons and droughts.
+Mean annual rainfall of 1095 mm in CT2018 compared to `r round(rain.man.stats$rain)`mm at BCI, and precipitation to potential evapotranspiration ratio of nearly one in CT2018 compared to nearly two in BCI are major factors in the different mortality responses.
+At BCI deep soil layers are effectively recharged annually (Fig.
+S\@ref(fig:swp-all-depths), \@ref(fig:swp-droughts)), whereas in CT2018 they were not.
+The contrast between the two studies highlights the combined role of seasonal precipitation input and site-specific hydrology in modulating deep ERD mortality-risk.
 
@@
 This research was supported as part of the Next Generation Ecosystem Experiments-Tropics, funded by the U.S.
-Department of Energy, Office of Science, Office of Biological and Environmental Research (LANL DOE Contract #7537545).
+Department of Energy, Office of Science, Office of Biological and Environmental Research.
 Funding was also provided to RCT and SMM through an NSF grant (1137366).
@@
 
-ELM-FATES source code, simulation outputs for best-fit parameter ensemble members and all the R scripts to reproduce the manuscript are archived (DOI:).
+ELM-FATES source code, simulation outputs for best-fit parameter ensemble membersm and all the R scripts to reproduce the manuscript are available [@Chitra.tarak.2020].
 Data sources for all other datasets used are provided throughout the manuscript.
 
-# Captions for Tables and Figures
-\
-Table \@ref(tab:hypo): Hypotheses for association between Effective Rooting Depth (ERD) and hydraulic traits.
-
-\
-Table \@ref(tab:symbols): The symbol, definition and units of key traits used in the simulations and analyses.
-
+# Figure Captions 
 \
@@
 ```{r hypo, results = 'asis', echo = FALSE}
-knitr::kable(hypo.table, caption = '', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(hypo.table))))
+knitr::kable(hypo.table, caption = 'Hypotheses for association between Effective Rooting Depth (ERD) and hydraulic traits.', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(hypo.table))))
 ```
@@
 ```{r symbols, results = 'asis', echo = FALSE}
-knitr::kable(symbols.table, caption = '', booktabs = TRUE, escape = FALSE)
+knitr::kable(symbols.table, caption = 'The symbol, definition and units of key traits used in the simulations and analyses.', booktabs = TRUE, escape = FALSE)
 ```
@@
 Briefly, canopy branches were collected from 4–10 trees of each species.
-Segments measured for hydraulic conductivity were removed from the branches and placed in an apparatus similar to that described by @Sperry.1998 except that flow rate was measured with graduated pipettes instead of a balance.
+Segments measured for hydraulic conductivity were removed from the branches and placed in an apparatus similar to that described by @Sperry.1988 except that flow rate was measured with graduated pipettes instead of a balance.
 Each segment was perfused with a 10 mM KCl solution filtered to 0.2 mM at three pressure heads: two that ranged 0.5–2 kPa and at zero pressure.
@@
 
-For each tree species measured for $K_\textrm{stem}$ and $\mathrm{\Psi}$~88,stem~, two to six leaves were measured for $\mathrm{\Psi}$~tlp~ with pressure-volume analysis following [@Koide.1991].
+For each tree species measured for $K_\textrm{stem}$ and $\mathrm{\Psi}$~88,stem~, two to six leaves were measured for $\mathrm{\Psi}$~tlp~ with pressure-volume analysis following [@Koide.1989].
 Leaf water potential was measured on leaf discs with psychrometers (J.R.D.
@@
 Dry mass was determined after oven drying the discs at 70°C.
-See [@Wolfe.2020] for archived data.
 

### 2020-11-04 manuscript.rmd
Chonggang's edits

@@
 McDowell^11,12^, Brent D.
-Newman^1^, Steve Paton^3^, Rolando Pérez^3^, Laurent Ruiz^13,14,15^, Lawren Sack^16^, Jeffrey M.
+Newman^1^, Steven R.
+Paton^3^, Rolando Pérez^3^, Laurent Ruiz^13,14,15^, Lawren Sack^16^, Jeffrey M.
 Warren^17^, Brett T.
@@
 3. Smithsonian Tropical Research Institute, Panama, PA
-4. Smithsonian Conservation Biology Institute, Conservation Ecology Center, Front Royal, VA, USA
+4. Conservation Ecology Center, Smithsonian Conservation Biology Institute, Front Royal, VA, USA
 5. Lawrence Berkeley National Laboratory, Climate and Ecosystem Sciences Division, Berkeley, CA, USA
@@
 Berkeley, CA, USA 94720
-9. University of Natural Resources and Life Sciences Vienna, Institute of Botany, Department of Integrative Biology and Biodiversity Research, Gregor‐Mendel‐Str 33, A‐1190 Wien, Wien, AT
+9. Institute of Botany, University of Natural Resources and Life Sciences Vienna, Department of Integrative Biology and Biodiversity Research, Gregor‐Mendel‐Str 33, A‐1190 Wien, Wien, AT
 10. German Centre for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig, Computational Forest Ecology, Leipzig, Saxony, DE
@@
 12.  School of Biological Sciences, Washington State University, PO Box 644236, Pullman, WA 99164-4236
-13. Indian Institute of Science, Indo-French Cell for Water Sciences, Bangalore, Karnataka, IN
+13.  Indo-French Cell for Water Sciences, Indian Institute of Science, Bangalore, Karnataka, IN
 14. IRD, CNRS, UPS, UMR GET, Toulouse, FR
@@
 We validated ERD against existing data for stable hydrogen isotopes of trees and soil.
-- Across species, deeper ERD was associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates during extreme hydrological droughts over 1981-2015.
+- Across species, deeper ERD was associated with higher maximum stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates during extreme hydrological droughts over 1981-2015.
 Species with deeper ERD spent less time below critical hydraulic thresholds indicating that trees compensate for water-stress related mortality risk through deep-water access.
@@
 \
-Keywords: rooting depths, deep-water access, hydraulic vulnerability, drought-induced mortality, drought tolerance, tropical forest, safety-efficiency trade-off, hydrological droughts
+Keywords: rooting depths, deep-water access, hydraulic vulnerability and safety margins, drought-induced mortality, drought tolerance, tropical forest, safety-efficiency trade-off, hydrological droughts
 
@@
 Tropical forests play a disproportionately large role in the global carbon and energy cycles [@Bonan:2008], and support half of global biodiversity [@Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019].
-Tropical forest has been expected to be an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw].
+Tropical forests have been expected to be an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw].
 Still, mortality events in the tropics are rarely as large as those in temperate and boreal zones [@McDowell.2018], leading to questions regarding the role of large trait-diversity and hydraulic strategies in mitigating mortality events.
@@
 \
-Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation to stem water storage, deep-water access, increased root production, hydraulic redistribution and embolism resistance [@McDowell.2008].
+Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation, stem water storage, deep-water access, increased root production, hydraulic redistribution and embolism resistance [@McDowell.2008].
 Deep-water access is arguably the most long-standing, yet under-studied mechanism.
@@
 Further, only a small fraction of those are tropical, even though more than 90% of the world’s tree diversity resides in the tropics [@Slik:2015].
-The use of stable isotopes of water as a tracer provides an indirect measure of water sourcing depths by matching the isotopic value in xylem water to vertical variations in soil pore water.
-However, such data are rare [@Evaristo:2016] and their interpretation is an active area of research given unresolved issues pertaining to soil heterogeneity and the correspondence of soil water extract in the laboratory to actual plant available water [@Adams.2020; @Bowers.2020; @Deurwaerder.2020].
+The use of stable isotopes of water as a tracer provides an indirect measure of water sourcing depths by matching the isotopic value in xylem water to those in soil pore water at different depths.
+However, such data are rare [@Evaristo:2016].
 DNA barcoding of roots [@Jones.2011] may be used to estimate species-specific rooting depths or profiles, but DNA barcode libraries for tropical forests are still under development, and the method may be cost-prohibitive for the extent of sampling required.
@@
 Such efforts have found correlations between vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013, @Venturas.2016].
-Even though deep roots may mitigate hydraulic vulnerability and mortality-risk, few drought studies compare a tree's access to water, i.e.
-rooting depth, vs.
-the actual vertical distribution of plant-available water, especially in tropical regions.
+Even though deep roots may mitigate hydraulic vulnerability and mortality-risk, hydrological droughts (rather than meteorological droughts) realized through trees’ water access--the interaction between rooting depths and the actual vertical distribution of plant-available water--are not studied, especially in tropical regions.
 \
   Here, we estimate plant-available water in the whole-soil column in a tropical forest, and inversely estimate rooting depths of co-occurring tree species from their growth responses, with a series of model calibrations and validation.
-We evaluate how rooting depth is linked to above-ground hydraulic traits and mortality rates over a 35 year period which experienced droughts of a variety of intensity, frequency and duration.
-Our hypotheses (see Table \@ref(tab:hypo)) are that (i) deep-rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem due to greater and more reliable water availability at depth, whereas shallow-rooted species are associated with less efficient but safer hydraulic traits due to more variable availability of shallow soil-water; and that, (ii) deep-rooted species have lower mortality rates during droughts due to lower realized risk of hydraulic failure compared to shallow-rooted species.
-To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.
+We evaluate how rooting depth is linked to above-ground hydraulic traits and mortality rates over a 35 year period that experienced droughts of a variety of intensity, frequency and duration.
+Our hypotheses (see Table \@ref(tab:hypo)) are that (i) deep-rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem due to greater and more reliable water availability at depth, whereas shallow-rooted species are associated with less efficient but safer hydraulic traits due to more variable availability of shallow soil-water; and that, (ii) deep-rooted species have lower mortality rates during droughts due to a lower realized risk of hydraulic failure compared to shallow-rooted species.
+To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich tropical forest.
 
@@
 
-To obtain the daily dynamics of soil water potential in the whole-soil column we locally parameterized and calibrated a 1-D hydrological water balance model.
+To obtain the daily dynamics of soil water potential in the whole-soil column, we locally parameterized and calibrated a 1-D hydrological water balance model.
 We then developed a novel, empirical inverse model to estimate species effective rooting depths (ERD) for large trees of `r length(erd.sp)` species (Fig.
@@
 The ERD model incorporated the impact of atmospheric and hydrological drought on growth, and was constrained with species-specific leaf vulnerability curves by using existing data for eight species, while developing trait-based proxies for the rest (based on data for a total of 21 species).
-We evaluated ERD against existing stable hydrogen isotope ratios ($\mathrm{\delta}$^2^H~xylem~) for a subset of six tree species as an independent observation of ERD.
+We validated ERD against existing stable hydrogen isotope ratios ($\mathrm{\delta}$^2^H~xylem~) for a subset of six tree species as independent observations.
 We evaluated whether ERD was associated with above-ground hydraulic traits sourcing the latter from a set of rare datasets for six to eight species that overlapped with our assessment of ERD.
@@
 
-To estimate dynamics of water availability by depth over the forest’s rooting zone, we used the Energy Exascale Earth System Land Model (ELM), coupled with the Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES).
+To estimate dynamics of water availability by depth over the forest’s rooting zone, we used the Energy Exascale Earth System Land Model (ELM, @Caldwell.2019), coupled with the Functionally Assembled Terrestrial Ecosystem Simulator (FATES, @Koven:2020) (hereafter, ELM-FATES).
 ELM is a land-model that, among many features, simulates the physics and conservative dynamics of water, energy and carbon fluxes.
@@
 FATES couples to ELM by a common interface of water and carbon fluxes.
-Detailed descriptions for ELM and FATES can be found elsewhere [@Fisher.2010; @Fisher.2015; @Koven:2020].
+Detailed descriptions for ELM and FATES can be found elsewhere [@Bisht.2018; @Fisher.2010; @Fisher.2015; @Koven:2020].
 
@@
 where, $q_{\textrm{rain}}$ is rainfall, $E_{v}$ is transpiration of the forest, $E_{g}$ is ground evaporation, $q_{\textrm{over}}$ is surface runoff, $q_{\textrm{h2osfc}}$ is runoff from surface water storage, $q$~drai~ is subsurface drainage (all in mm H~2~O) and $\mathrm{\Delta}t$ is the time-step (here, 1 hr).
-The soil profile is discretized into 15 exponentially distributed soil layers by depth $z$, where $z$ $\in$ $Z$, and $Z$ = {0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13, 21, 35} m.
+The soil profile is discretized into 15 exponentially distributed soil layers by depth $z$ (user-specified range), where $z$ $\in$ $Z$, and $Z$ = {0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13, 21} m.
 
@@
 To parameterize ELM-FATES, we identified eleven parameters relevant for Eq.
-\@ref(eq:water-balance), determined their ranges based on literature for the study site, else for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling (LHS) of this global parameter space.
+\@ref(eq:water-balance), determined their ranges based on literature for the study site, else for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling (@Stein.1987) of this global parameter space.
 In the Conrad catchment soil hydraulic conductivity decreases with depth from 12.5 and 60 cm (@Godsey:2004; Table \@ref(tab:param)).
@@
 
-We tested alternate empirical models for species 5-yearly census growth, which use effective rooting depth as an input (Eq.
+We tested alternate empirical models for species-specific 5-yearly census growth, which use effective rooting depth as an input (Eq.
 \@ref(eq:M-chosen) and those in SI).
 Many factors limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019].
-Because pre-dawn $\mathrm{\Psi}$~leaf~ approaches $\mathrm{\Psi}$~soil~, we used the latter in leaf hydraulic vulnerability curves to predict maximum diurnal $K$~leaf~, which in turn controls growth (see Table \@ref(tab:symbols) for trait symbols and their definitions).
+Because pre-dawn $\mathrm{\Psi}$~leaf~ generally approaches $\mathrm{\Psi}$~soil~, we used the latter in leaf hydraulic vulnerability curves to predict maximum diurnal $K$~leaf~, which in turn controls growth (see Table \@ref(tab:symbols) for trait symbols and their definitions).
 $K_\textrm{leaf}$ in our models thus incorporates the effects of rooting depth and rhizosphere water potential.
@@
 \end{equation}
- By varying $z$, we estimated species effective rooting depths (ERD), defined as the depth $z$ at which soil water dynamics ($\mathrm{\Psi}_{\textrm{soil,} z}$) best explained observed dynamics of species growth.
+ Checking through each $z$, we estimated species effective rooting depths (ERD), defined as the depth $z$ at which soil water dynamics ($\mathrm{\Psi}_{\textrm{soil,} z}$) best explained observed dynamics of species growth.
 Using Eq.
@@
 
-For growth estimates, to minimize the effect of light variation among trees, we selected only large trees ($\ge$ 30 cm dbh, the diameter at breast height) in the 50 ha plot and also species whose maximum height was $\ge$ 30 m (hereafter, canopy species) and thus are likely to be fully exposed to the sun.
+For diameter growth estimates, to minimize the effect of light variation among trees, we selected only large trees ($\ge$ 30 cm dbh, the diameter at breast height) in the 50 ha plot and also species whose maximum height was $\ge$ 30 m (hereafter, canopy species) and thus are likely to be fully exposed to the sun.
 Calculating individual tree growth rates across six 5-yearly censuses (1990-2015), removing outliers, accounting for temporal changes in growth rates due to increasing tree size (see @Condit.2019a; @Condit.2019b and SI), we obtained species-level median time series of growth rates ($G_{t}$, cm yr^-1^) used in the ERD model for only those species ($n$ = `r length(erd.sp)`) with complete records on at least three trees (median $n$ = `r median(g.n$n)`, max $n$ = `r max(g.n$n)` trees per species).
@@
 
-Among evergreen species, ERD explained `r m.evg.r2.90`, `r m.evg.r2.05` and `r m.evg.r2.10`% variation in mortality rates during census intervals 1985-1990, 2000-2005 and 2005-2010, respectively, such that species with mortality rates significantly decreased with ERD ($p$ = `r m.evg.p.90`, `r m.evg.p.05` and `r m.evg.p.10`, respectively; Fig.
+Among evergreen species, ERD explained `r m.evg.r2.90`, `r m.evg.r2.05` and `r m.evg.r2.10`% variation in mortality rates during census intervals 1985-1990, 2000-2005 and 2005-2010, respectively, such that species mortality rates significantly decreased with ERD ($p$ = `r m.evg.p.90`, `r m.evg.p.05` and `r m.evg.p.10`, respectively; Fig.
 \@ref(fig:erd-mort-census-evg)).
@@
 However, the difference between precipitation and ET shows that 1985-1990 had three extreme deficit water years, apart from the 1983 El-Niño leading to it [@Condit.2017].
-The analysis of a species' time spent under significant hydrological drought (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) showed that species with shallower ERD spent a greater proportion of days below critical hydraulic thresholds and increasingly so over 1995-2000, 2000-2005 and 2005-2010 (Fig.
+Analysis of a species' time spent under significant hydrological drought (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) showed that species with shallower ERD spent a greater proportion of days below critical hydraulic thresholds and increasingly so over 1995-2000, 2000-2005 and 2005-2010 (Fig.
 \@ref(fig:erd-time); S\@ref(fig:erd-time-evg)).
@@
 By estimating ERD for large trees of `r length(erd.sp)` canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species-rich tropical forests.
-Our chosen ERD model is an advance over the only other inverse model of ERD (CT2018; $n$ = 12 species) as we employ a physiologically meaningful representation, fewer parameters and corroboration with tree and soil data for $\mathrm{\delta}$^2^H.
-We highlight that stable water isotopes for plant and below-ground sources are critical for calibration or validation of ERD models at community scale, although the interpretation of these data is still under research [@Deurwaerder.2020].
+Our chosen ERD model is a key improvement over the only other inverse model of ERD (CT2018; $n$ = 12 species) as we employ a physiologically meaningful representation, fewer parameters and corroboration with tree and soil data for $\mathrm{\delta}$^2^H.
+We highlight that stable water isotopes for plant and below-ground sources are critical for calibration or validation of ERD models at community scale, although the interpretation of these data is still under research  [@Deurwaerder.2020; @Adams.2020; @Bowers.2020].
 As in CT2018, we also interpret our ERD estimates as revealing relative differences among species' effective rooting depths rather than in absolute terms.
@@
 We found that LMA and WSG were strong predictors of PLC curves for leaves, but not for stems (not shown).
-This offers the promise of a greater ability to parameterize the 'hard' hydraulic traits with the abundant 'soft' trait data, thus allowing for better representation of forest hydrodynamics.
+This offers the promise of a greater ability to parameterize the 'hard' hydraulic traits with the abundant 'soft' trait data, thus allowing for a better representation of forest hydrodynamics.
 The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits, including hydraulic vulnerability, in order to better represent the functional diversity of tropical forests, and their trajectories into the future.
@@
 Our inverse ERD model was parameterized on 5-year growth data, with five data points over a 25-yr period, which decoupled climate events and demographic outcomes.
-Future studies could better constrain the ERD model with high frequency growth data such as those from dendrometer bands.
+Future studies could better constrain the ERD model with higher frequency growth data such as those from dendrometer bands.
 At high temporal resolution, however, the role of reversible dehydration in tree diameter change increases [@Chitra-Tarak.2016; @Chitra-Tarak.2015], but that may provide an avenue to include stem water storage  and dynamic rooting depths in ERD models.
@@
 Trees rooting depths and actual water environments through hydrological rather than meteorological droughts are nonetheless rarely studied.
-To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.
+To our knowledge this is the first study to test for a mechanistic link between plant-available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.
 We report here that deep-water access plays a role in alleviating mortality of hydraulically vulnerable trees.
@@
 
-ELM-FATES source code, simulation outputs for best-fit parameter ensemble membersm and all the R scripts to reproduce the manuscript are available [@Chitra.tarak.2020].
+ELM-FATES source code, simulation outputs for best-fit parameter ensemble membersm and all the R scripts to reproduce the manuscript are available [@Chitra-Tarak.2020].
 Data sources for all other datasets used are provided throughout the manuscript.

### 2020-11-04 manuscript.rmd
1. edits from Nate and Chonggang 2. addresses

@@
 Summary | 197
-Main Text  | 6492
-Introduction | 857
-Methods | 3192
+Main Text  | 6349
+Introduction | 823
+Methods | 3109
 Results | 855
-Discussion | 1588
+Discussion | 1562
 Keywords | 6
@@
 
-- Among the mechanisms that plants employ to tolerate drought, deep-water access is arguably the most long-standing, but under-studied.
+- Deep-water access is arguably the most long-standing, but under-studied mechanisms that plants employ to survive during drought.
 Vulnerability to embolism and hydraulic safety margins can predict mortality risk at given levels of dehydration, but deep-water access may delay plant dehydration.
-Here, we tested the role of deep-water access in enabling survival within a diverse tropical forest community (Panama) using a novel data-model approach.
-- We inversely modeled the effective rooting depth (ERD, as the average depth of water extraction), for `r length(erd.sp)` canopy species by linking growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and leaf hydraulic vulnerability curves.
-We validated ERD against existing data for stable hydrogen isotopes of trees and soil.
-- Across species, deeper ERD was associated with higher maximum stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates during extreme hydrological droughts over 1981-2015.
-Species with deeper ERD spent less time below critical hydraulic thresholds indicating that trees compensate for water-stress related mortality risk through deep-water access.
+Here, we tested the role of deep-water access in enabling survival within a diverse tropical forest community in Panama using a novel data-model approach.
+- We inversely estimated the effective rooting depth (ERD, as the average depth of water extraction), for 32 canopy species by linking diameter growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and leaf hydraulic vulnerability curves.
+We validated ERD estimates against existing empirical data of potential water access depths.
+- Across species, deeper ERD was associated with higher maximum stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates during extreme droughts over 1981-2015.
+Species with deeper ERD spent did not cross critical hydraulic thresholds indicating that trees compensate for water-stress related mortality risk through deeper water access.
 - The role of deep-water access in mitigating mortality of hydraulically-vulnerable trees has important implications for our predictive understanding of forest dynamics under current and future climates.
-
 \
@@
 \
-  Characterising the essential constraints to model species-rich communities [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010; @Bartlett.2019] entails identifying the topography of plant trait trade-offs in different environments, and how these relate to demographic rates (growth, recruitment and mortality).
+  Characterising the essential constraints to model species-rich communities [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010; @Bartlett.2019; @Lu.2020] entails identifying the topography of plant trait trade-offs in different environments, and how these relate to demographic rates (growth, recruitment and mortality).
 Significant efforts have been invested into identifying and linking universal drought indices, above-ground traits, and demographic rates.
@@
 This approach does not account for the buffering effect of stem water storage capacitance on $\mathrm{\Psi}$~leaf~, and may overestimate ERD (see Discussion).
-VPD affects growth non-linearly, with growth increasing with VPD up-to a VPD threshold, then decreasing as leaf pores (stomata) close reducing water-uptake [@Grossiord.2020], and leaf-drop may further limit water-uptake and growth.
+VPD affects growth non-linearly, with growth increasing with VPD up-to a VPD threshold, then decreasing as leaf pores (stomata) close reducing water-uptake [@Grossiord.2020; @Yang.2019], and leaf-drop may further limit water-uptake and growth.
 We therefore regressed species-specific growth against multiplicative or additive effects of one or more factors calculated daily and averaged over 5-yearly censuses: VPD, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1.
@@
 
-Our study brings attention to the need for assessing drought sensitivity in terms of species realized drought exposure by accounting for the actual hydrological drought and trees' deep water access, beyond the universal drought indices and trees' above-ground hydraulic vulnerability.
+Our study brings attention to the need for assessing drought sensitivity in terms of species realized drought exposure by accounting for the actual hydrological drought and trees' deep water access.
 We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins (Fig.
-\@ref(fig:erd-hyd)), that is, traits that are commonly identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through periods of extreme droughts (Fig.
+\@ref(fig:erd-hyd)), that is, traits that are commonly identified as proxies for mortality risk [@Anderegg:2016], but they were less droughted and had lower mortality compared to species relying on shallower water through periods of extreme droughts (Fig.
 \@ref(fig:erd-mort-census-evg), S\@ref(fig:erd-mort-census), \@ref(fig:swp-droughts)).
@@
 \@ref(fig:erd-time)).
-Our results are consistent with a few studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016; also see @Brum.2019].
+Our results are consistent with studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016; also see @Brum.2019].
 Globally, large trees tend to exhibit greater growth reductions, lower post-drought resilience, and greater increases in mortality relative to their understory counterparts [@Bennett:2015; @Phillips:2010].
@@
 \
-Our result--of lower mortality in deep-rooted trees--contrasts with the inverse model finding of @Chitra-Tarak.2018 (hereafter, CT2018) in which deeper ERD species had higher mortality in a rare, prolonged drought.
+Our result--of lower mortality in deep-rooted trees--contrasts with the inverse model finding of @Chitra-Tarak.2018 (hereafter, CT2018) in which deeper ERD species in a south-asian seasonally tropical dry forest had higher mortality in a rare, prolonged drought.
 The hydrological model in CT2018 revealed that the multi-year drought exhausted the deep soil and even bedrock water availability (also see, @Goulden.2019 and @Ivanov:2012).
@@
 The relationships that we identify between above- and below-ground traits, the water environment vertically below-ground and mortality rates are important for representing diversity in DGVMs, which intrinsically rely on the parameterization of contrasting life history strategies [@Scheiter:2013] and the simulation of competition between those strategies.
-We found that LMA and WSG were strong predictors of PLC curves for leaves, but not for stems (not shown).
+We found that LMA and WSG were strong predictors of PLC curves for leaves.
 This offers the promise of a greater ability to parameterize the 'hard' hydraulic traits with the abundant 'soft' trait data, thus allowing for a better representation of forest hydrodynamics.
@@
 Future studies could better constrain the ERD model with higher frequency growth data such as those from dendrometer bands.
-At high temporal resolution, however, the role of reversible dehydration in tree diameter change increases [@Chitra-Tarak.2016; @Chitra-Tarak.2015], but that may provide an avenue to include stem water storage  and dynamic rooting depths in ERD models.
+At high temporal resolution, however, the role of reversible dehydration in tree diameter change increases [@Chitra-Tarak.2016; @Chitra-Tarak.2015; @Mencuccini.2017], but that may provide an avenue to include stem water storage  and dynamic rooting depths in ERD models.
 Three of our exploratory ERD models included LAI seasonality, but we did not select them as they worsened the fit with growth data for many species (Fig.
@@
 \
-While our ERD model empirically predicts growth via estimating $\mathrm{\Psi}$~leaf~ from $\mathrm{\Psi}$~soil~ of a specific depth, mechanistic models that account for plant hydrodynamics and other processes influencing growth are likely to predict growth more accurately, and thus ERD [@Christoffersen:2016kd2d; @Sperry.1998].
+While our ERD model empirically predicts growth via estimating $\mathrm{\Psi}$~leaf~ from $\mathrm{\Psi}$~soil~ of a specific depth, mechanistic models that account for plant hydrodynamics and other processes influencing growth are likely to predict growth more accurately, and thus ERD and hydraulic risk [@Christoffersen:2016kd2d; @Sperry.1998; @Yang.2019; @Duursma:2018].
 Although data needs for parameterizing such models could be greater (e.g.

### 2020-11-05 manuscript.rmd
1. Added Lawrens's edits for ERD model description

@@
 
-We tested alternate empirical models for species-specific 5-yearly census growth, which use effective rooting depth as an input (Eq.
+To estimate effective rooting depth, we tested alternate empirical models for species-specific 5-yearly census growth, which use effective rooting depth as an input (Eq.
 \@ref(eq:M-chosen) and those in SI).
 Many factors limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019].
-Because pre-dawn $\mathrm{\Psi}$~leaf~ generally approaches $\mathrm{\Psi}$~soil~, we used the latter in leaf hydraulic vulnerability curves to predict maximum diurnal $K$~leaf~, which in turn controls growth (see Table \@ref(tab:symbols) for trait symbols and their definitions).
-$K_\textrm{leaf}$ in our models thus incorporates the effects of rooting depth and rhizosphere water potential.
-This approach does not account for the buffering effect of stem water storage capacitance on $\mathrm{\Psi}$~leaf~, and may overestimate ERD (see Discussion).
-VPD affects growth non-linearly, with growth increasing with VPD up-to a VPD threshold, then decreasing as leaf pores (stomata) close reducing water-uptake [@Grossiord.2020; @Yang.2019], and leaf-drop may further limit water-uptake and growth.
-We therefore regressed species-specific growth against multiplicative or additive effects of one or more factors calculated daily and averaged over 5-yearly censuses: VPD, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1.
+Because pre-dawn $\mathrm{\Psi}$~leaf~ generally approaches $\mathrm{\Psi}$~soil~, we used the latter in leaf hydraulic vulnerability curves to predict maximum diurnal $K$~leaf~, which we assumed controls photosynthetic rate and growth (@Brodribb.2002; @Brodribb.2017; see Table \@ref(tab:symbols) for trait symbols and their definitions).
+We assumed $K_\textrm{leaf}$ to be a hydraulic bottleneck in the plant, and thus to determine or co-determine whole plant hydraulic response to soil drying; notably, root hydraulic vulnerability may also be a strong driver, and recent studies suggest that its vulnerability would be similar to, or correlated with that of leaves (see @Cochard.2020 and references therein).
+$K_\textrm{leaf}$ in our models, being based on vulnerability curves, is dynamic and thus depends on rooting depth and rhizosphere water potential.
+The use of maximum $\mathrm{\Psi}$~leaf~ based on pre-dawn $\mathrm{\Psi}$~soil~ as a proxy for diurnal $\mathrm{\Psi}$~leaf~ and its influence on growth at diurnal time-scales (to relate to 5-year average growth values) thus does not consider diurnal dynamics of $\mathrm{\Psi}$~leaf~ and other factors that would play a role in such diurnal dynamics, such as the buffering effect of stem water storage capacitance on $\mathrm{\Psi}$~leaf~, and may thus overestimate ERD (see Discussion).
+VPD affects growth non-linearly, with growth increasing with VPD up-to a VPD threshold, then decreasing as leaf pores (stomata) close reducing water-uptake [@Grossiord.2020; @Yang.2019].
+To account for the non-linear impact of VPD on growth, we used predicted GPP (hereafter, $\hat{\textrm{VPD}}$) from a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see SI).
+Apart from stomatal control, leaf-drop may further limit water-uptake and growth.
+We therefore regressed species-specific growth against multiplicative or additive effects of one or more factors calculated daily and averaged over 5-yearly censuses: $\hat{\textrm{VPD}}$, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1.
 Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting.
@@
 \begin{equation}
-G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*} \textrm{VPD}_{i}^{*}) + \epsilon_{t|z}
+G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*} \hat{\textrm{VPD}}_{i}^{*}) + \epsilon_{t|z}
 (\#eq:M-chosen)
@@
 See SI for equations for all the alternate models tested.
-
-\
 
@@
 \
-Figure \@ref(fig:erd-hyd): Matrix for pairwise correlations (circles) among Effective Rooting Depth (ERD) and hydraulic properties for six to seven canopy species found on Barro Colorado Island (Panama) (see Table S\@ref(tab:hyd-table) for data).
-Each circle is color coded by the Spearman's $r$ with red shades for positive $r$ values and blue shades for negative values.
-$p$-values are shown inside the circles.
-Units are as follows: ERD in m, K~max,stem~ in mmol m^-1^ s^-1^ MPa^-1^) and variables with $\mathrm{\Psi}$ in MPa.
+Figure \@ref(fig:erd-hyd): Pairwise correlations among modelled Effective Rooting Depth (ERD) and hydraulic properties for seven canopy species found on Barro Colorado Island (Panama).
+Spearman's $r$ with red shades for positive $r$ values and blue shades for negative values.
+$R$^2^ and $p$-values for linear model fits are given in panel insets.
+See Fig.
+S\@ref(fig:erd-hyd-full) for the full correlation matrix.
 
@@
 ```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/erd.stem.traits_cor.matrix_all_p-val.jpeg"))
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/erd.stem.traits.jpeg"))
 ```
@@
 As tree growth rates vary with size [@Muller-Landau:2006], to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($\textrm{Resid}_{i,t} = G_{i,t} - f(d_{i,t})$).
-The diameter model of growth, $f(d_{i, t})$, a B-spline based polynomial equation with five degrees of freedom, was fitted to growth rates from all trees pooled together as there was insufficient species-specific data for large trees (Fig.
-S\@ref(fig:dbh-model)).
+The diameter model of growth, $f(d_{i, t})$, a B-spline based polynomial equation with five degrees of freedom, was fitted to growth rates from all trees pooled together as there was insufficient species-specific data for large trees ($R$^2 = 0.27; $p$-value < 0.01).
 We retained only those trees that had $\textrm{Resid}_{i,t}$ estimates for all census intervals (total trees = `r sum(g.n$n)`) and standardized (centered and scaled) individual tree time series of $\textrm{Resid}_{i,t}$.
@@
 
-```{r dbh-model, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-dbh-model}A scatterplot of tree growth rates (circles) from all census intervals versus dbh at the beginning of the census interval is overlaid with a dbh model of growth (spline).
-Growth rates are color coded by census interval.
-The model equation is shown in inset along with goodness-of-fit (Adj.
-$R$^2^).")}
-knitr::include_graphics(file.path("figures/dbh/growthrate_dbh_predicted_growth_with_spline_degree_5_dbh_cutoff_100_model_no_color.jpeg"))
-```
-
 ```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "35%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd}Scatterplot of mean midday (11:00 am to 15:00 pm) gross primary productivity (GPP) and mean midday vapor pressure deficit (VPD) measured at the eddy flux tower in Barro Colorado Island over 2012-2017.
@@
 The number of species with an ERD estimate as well as $\\mathrm{\\delta}^2H$ for each model are shown in inset, along with the goodness-of-fit ($R$^2^) and $p$-values for a linear relationship between ERD and $\\mathrm{\\delta}^2H$.
-ERD models in panels A-E correspond to Eq.
+ERD models in panels a-e correspond to Eq.
 \\@ref(eq:M1)-\\@ref(eq:M6), respectively.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_sp_color_Meinzer.jpeg"))
+```
+
+```{r erd-hyd-full, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd-full}.
+Pairwise correlations (upper triangle) and corresponding linear regressions (lower triangle) among modelled Effective Rooting Depth (ERD) and hydraulic properties for seven canopy species found on Barro Colorado Island (Panama).
+Spearman's correlation coefficients and significance levels are shown (upper-half; *$p$ < 0.05; .
+$p$ < 0.1)")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/erd.stem.traits_cor.chart.jpeg"))
 ```

### 2020-11-05 manuscript.rmd
1. Moved out a disucssion section due to word limit

@@
 Rutuja Chitra-Tarak  
-\
 E: arutuj@gmail.com  
-\
-T: +1(505) 665 1647
-\
-\
+T: +1 (443) 758 1257
 
@@
 10. German Centre for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig, Computational Forest Ecology, Leipzig, Saxony, DE 
-11. Atmospheric Sciences and Global Change Division, Pacific Northwest National Lab, PO Box 999, Richland, WA, 99352
-12. School of Biological Sciences, Washington State University, PO Box 644236, Pullman, WA 99164-4236
+11. Atmospheric Sciences and Global Change Division, Pacific Northwest National Lab, PO Box 999, Richland, WA, USA 99352 
+12. School of Biological Sciences, Washington State University, PO Box 644236, Pullman, WA, USA 99164-4236 
 13. Indo-French Cell for Water Sciences, Indian Institute of Science, Bangalore, Karnataka, IN 
@@
 Summary | 197  
-Main Text  | 6349
-Introduction | 823
-Methods | 3109
+Main Text  | 6532  
+Introduction | 828  
+Methods | 3278  
 Results | 855  
-Discussion | 1562
+Discussion | 1571  
 Keywords | 6
@@
 
-## Soil water dynamics by depth
+## Soil water dynamics by depth and hydrological droughts
 
@@
 \@ref(fig:erd-mort-census-evg)).
-Analysis of time spent below ($\mathrm{\Psi}$~crit~) in the soil zone matching ERD $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ showed that this was likely due to decreasing realized risk of hydraulic failure with increasing access to water for species with deeper ERD (Fig.
+Analysis of time spent beyond $\mathrm{\Psi}$~crit~ in the soil zone matching ERD $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ showed that this was likely due to decreasing realized risk of hydraulic failure with increasing access to water for species with deeper ERD (Fig.
 \@ref(fig:erd-time); Fig.
@@
 \@ref(fig:erd-time)).
-Our results are consistent with studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016; also see @Brum.2019].
+Our results are consistent with studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016; also see @Brum.2019 and @Brum.2017].
 Globally, large trees tend to exhibit greater growth reductions, lower post-drought resilience, and greater increases in mortality relative to their understory counterparts [@Bennett:2015; @Phillips:2010].
@@
 Future studies may improve models of LAI seasonality and make use of data from new technologies such as drone based monitoring of LAI in species-rich tropical forests [@Park.2019].
-
-\
-In forests such as BCI where understory tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes may need accounting for temporal variation in trees' light environment, an issue we minimized by limiting ERD estimates to canopy trees.
-In sites where growth is not water-limited, ERD models may explore parameterization on sap-flux or leaf water-potential dynamics [@Brum.2017].
 

### 2020-11-05 old_ms_sections.Rmd
1. more secrtions moved out from manuscript

@@
 \
+In forests such as BCI where understory tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes may need accounting for temporal variation in trees' light environment, an issue we minimized by limiting ERD estimates to canopy trees.
+In sites where growth is not water-limited, ERD models may explore parameterization on sap-flux or leaf water-potential dynamics [@Brum.2017].
+
+\
 \ Finally, rooting depths may be parametarized within DGVMs such as ELM-FATES using plant hydrodynamic schemes [@Christoffersen:2016kd2d; @Koven.2020], simultaneously modeling plant water-availability with plant water-extraction.
 In general, including differences in tree rooting depths in ELM-FATES would improve understanding of plant competition for water (and other resources) and more accurately represent plant's drought strategies, leading to better predictions of existing and future forest function and composition.
-
+\
 The model in @Chitra-Tarak.2018 estimated six species-specific phenomenological parameters: three as root fraction for corresponding lumped vadose zone layers, and three as shape parameters for water uptake vs availability curves.
@@
 Our model was constrained on VPD and species leaf vulnerability curves, variables that are shown to be physiologically meaningful.
-
+\
 The model in CT2018 were not corroborated with such data as existing methods were inadequate to extract water bound to the saprolite found in their dry forest study site (Jean Riotte, pers.
@@
 The model in CT2018 was calibrated only on stream discharge and water table levels.
+
+\
+Figure \@ref(fig:erd-hyd): Matrix for pairwise correlations (circles) among Effective Rooting Depth (ERD) and hydraulic properties for six to seven canopy species found on Barro Colorado Island (Panama) (see Table S\@ref(tab:hyd-table) for data).
+Each circle is color coded by the Spearman's $r$ with red shades for positive $r$ values and blue shades for negative values.
+$p$-values are shown inside the circles.
+Units are as follows: ERD in m, K~max,stem~ in mmol m^-1^ s^-1^ MPa^-1^) and variables with $\mathrm{\Psi}$ in MPa.
+
+```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/erd.stem.traits_cor.matrix_all_p-val.jpeg"))
+```
+
+```{r dbh-model, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-dbh-model}A scatterplot of tree growth rates (circles) from all census intervals versus dbh at the beginning of the census interval is overlaid with a dbh model of growth (spline).
+Growth rates are color coded by census interval.
+The model equation is shown in inset along with goodness-of-fit (Adj.
+$R$^2^).")}
+knitr::include_graphics(file.path("figures/dbh/growthrate_dbh_predicted_growth_with_spline_degree_5_dbh_cutoff_100_model_no_color.jpeg"))
+```

### 2020-11-05 manuscript.rmd
1 Front pages formatting

@@
 \
-*__Corresponding author__:  
-Rutuja Chitra-Tarak  
-E: arutuj@gmail.com  
-T: +1 (443) 758 1257
+*__Corresponding author__: Rutuja Chitra-Tarak  
+E: arutuj@gmail.com ;  T: +1 (443) 758 1257
 
+\  
 1. Los Alamos National Laboratory, Earth and Environmental Sciences Division (EES-14) 
@@
 
+\
+
 Section  | Word Count  
@@
 
+\  
 Item | Count  
 ------------- | -------------  
-Figures in color | 8  
+Figures in color | 9  
 Tables | 2  

### 2020-11-05 manuscript.rmd
First page formatted
Moved SI to a separate .rmd

@@
 
-\  
 1. Los Alamos National Laboratory, Earth and Environmental Sciences Division (EES-14) 
@@
 \@ref(fig:erd-time)).
-Our results are consistent with studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016; also see @Brum.2019 and @Brum.2017].
+Our results are consistent with studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016; also see @Brum.2019; @Brum.2017].
 Globally, large trees tend to exhibit greater growth reductions, lower post-drought resilience, and greater increases in mortality relative to their understory counterparts [@Bennett:2015; @Phillips:2010].
@@
 
-# Supplementary Information {-}
-
-## Data Descriptions 
-### Growth Data 
-\
-We selected trees for which the height of the diameter measurements were unchanged across all censuses (1990, 1995, 2000, 2005, 2010, 2015).
-For each individual tree ($i$), we calculated absolute growth rates ($G_{i,t}$, cm yr^-1^) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses prorated by the inter-census interval based on the exact date of measurements.
-We eliminated positive outliers as any growth rates amounting to >75 mm yr^-1^, as this was the highest rate observed with confidence for the fastest growing species [@Condit.2017].
-Negative outliers were removed based on known measurement error, $\sigma_{e} = 0.006214d + 0.903$, where *d* is dbh.
-Any stem whose later dbh measurement fell <4$\sigma_{e}$ the earlier measurement was removed [@Condit.2017].
-As tree growth rates vary with size [@Muller-Landau:2006], to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($\textrm{Resid}_{i,t} = G_{i,t} - f(d_{i,t})$).
-The diameter model of growth, $f(d_{i, t})$, a B-spline based polynomial equation with five degrees of freedom, was fitted to growth rates from all trees pooled together as there was insufficient species-specific data for large trees ($R$^2 = 0.27; $p$-value < 0.01).
-We retained only those trees that had $\textrm{Resid}_{i,t}$ estimates for all census intervals (total trees = `r sum(g.n$n)`) and standardized (centered and scaled) individual tree time series of $\textrm{Resid}_{i,t}$.
-We then obtained `r length(erd.sp)` species-level time-series of medians of $\textrm{Resid}_{i,t}$ ($G_{t}$).
-These were obtained for only those species with complete records on at least three trees (median $n$ = `r median(g.n$n)`, max $n$ = `r max(g.n$n)` trees per species).
-
-### Microclimatic tower {#SIData.matteo} 
-\
-The tower used for these measurements is 41 m above ground, on a plateau on BCI [@Pau.2018].
-The eddy covariance system includes a sonic anemometer (CSAT3, Campbell Scientific, Logan, UT) and an open-path infrared CO~2~/H~2~O gas analyzer (LI7500, LiCOR.
-Lincoln, NE).
-Hi-frequency (10Hz) measurements were acquired by a datalogger (CR1000, Campbell Scientific) and stored on a local PC.
-Data were processed with a custom program using a standard routine described in [@Detto.2010].
-QA/QC criteria for removing erroneous values are listed in table 1 (see below).
-GPP was derived from daytime values of NEE by adding the corresponding mean daily ecosystem respiration obtained as the intercept of the light response curve [@Lasslop.2010].
-The light curve was fitted on a 15-days moving window using a rectangular hyperbolic function (runs with friction velocity less than 0.4 m s^-1^ were excluded).
-In order to compute daily time integrated budgets, gaps were filled using Artificial Neural Network [@Papale.2003] with hydro-meteorological inputs as predictors (soil moisture, solar radiation, temperature, VPD and air pressure, all measured on the same tower).
-To train the network, the dataset was randomly divided in a training set (70%), a validation set (15%) and a test set (15%).
-A two-layer feed-forward network with 10 sigmoid hidden neurons and linear output neurons, was trained using the Levenberg-Marquardt algorithm until the mean square error (MSE) of the validation set stopped improving [@Hagan.1994].
-Performance, in terms of MSE, was evaluated using the test set at the end of the training.
-This procedure was repeated 100 times to produce 100 estimates of GPP.
-Training multiple times generates different results due to different initial conditions and random sampling of the three sets.
-Ensemble was obtained as weighted average from the 100 ANN predictions using the MSE of the test set as weights according to:
-
-$f_{i} = \frac{\sum_{k}f_{ik}^{ANN}/{MSE_{ik}}}{\sum_{k}1/{MSE_{ik}}}$
-\
-The ANN was implemented using the Neural Network Toolbox in Matlab 2014a. 
-\
-The associated data can be requested from https://ameriflux.lbl.gov.
-
-### Stream discharge from the Conrad catchment {#SIData.steveP} 
-\
-The Conrad weir is located on the gently sloping western side of Barro Colorado Island.
-The weir consists of a 90 degree ‘V’ notch with a rectangular upper section.
-The weir was constructed in several stages between 1993 and 1996.
-Stream stage is recorded at 5-minute intervals.
-\
-The weir drains a catchment of approximately 40 ha of which at least 90% consists of the central plateau area of the island.
-Peak flow usually follows peak rain by approximately 2-3 hours.
-Large storms are sometimes preceded by small, initial peaks generated by the rain falling on the steeper slopes close to the weir.
-\
-Streamflow data are analyzed by software custom-built by author SP.
-The program is used to correct for blockages of the weir ‘V’ by fallen vegetation (a common occurrence), sensor drift, small gaps and other problems.
-Discharge is then calculated using an empirically derived rating curve calculated for each 1mm stream stage.
-
-### Volumetric water content
-
-\
-We collected volumetric water content (VWC) data (2012-2018) from three locations near the eddy flux tower with time domain reflectometry (TDR, CS616, Campbell Scientific) probes inserted with an angle to average the first 15 cm of soil.
-The relationship between the output from the TDR probe and VWC was calibrated based on measurements of gravimetric water content (GWC) on several occasions from representative soils near the vertical probes (0-0.15 m) (Fig.
-S\@ref(fig:TDR-calib)).
-GWC was converted to VWC based on the observed mean bulk dry density of 0.8 g cm^-3^ near the TDR probe ($n$ = 5, cylinder volume = 1250 cm^3^).
-Beginning in 2016 we additionally installed three horizontal probes at depths of 0.15, 0.4 and 1 m and we applied the same calibration equation to these data (2016-2018) as that for the vertical probe.
-These data are archived at https://ngt-data.lbl.gov/.
-
-\
-We also leveraged existing GWC measurements by @Kupers:2019 at the depths of 0.15, 0.4 and 1 m in summer 2015 (February, March and April) and 0.15, 0.4 in summer 2016 (March), with 1299 samples in total that covered all soil types and habitats in the plot.
-The sampling periods were six, five, ten and eight days long, respectively.
-The 2016 dry season was associated with the 2015–2016 El Niño (see @Kupers:2019 for further details).
-We converted these GWC measurements to VWC assuming a bulk density of 0.8 g cm^-3^.
-$\theta_{\textrm{0.15 m}}$, $\theta_{\textrm{0.4 m}}$ and $\theta_{\textrm{1 m}}$ refer to the plot-wide VWC averages by depth over each sampling period.
-Middle date of a sampling period is used for comparison with model simulations.
-
-### Leaf Area Index 
-\
-We generated an estimate of the mean seasonality curve for normalized leaf area index (LAI$^*$, range 0-1, unitless) for individual species as follows.
-We interpolated species-specific weekly leaf-fall data collected at two similar sites in BCI (1985-2020) to a daily estimate [@Wright.1990aeo].
-We lead this daily leaf-fall time series by the species-specific average leaf-lifetime (in days) and used it as an estimate of the daily leaf-gain time-series.
-The difference between leaf-fall and leaf-gain gave an estimate of daily leaf-change time series, which was then scaled to express as a fraction of the total leaf fall in a given year.
-Beginning with the LAI$^*$ of 1 at the end of the wet season in the first year of the record, we accumulated leaf-change over it to obtain a running estimate of the leaf-cover fraction, averaged it by DOY across years and normalized it to obtain the species-specific seasonality curve in LAI$^*$ (range 0-1, unitless).
-(Fig.
-S\@ref(fig:leaf-cover)).
-\
-For five species, an estimate of leaf lifetime was available from direct observations [@Osnas.2018].
-For the rest erd.sp.wo.ll species, leaf lifetime was predicted from LMA ($\overline{lifetime} = `r ll.lma.cf[1]` + `r ll.lma.cf[2]`LMA$; $R$^2^ = `r ll.lma.r2`; $p$ `r ll.lma.p`, $n$ = `r ll.lma.n`).
-
-### Leaf hydraulic conductivity and vulnerability to cavitation {#SIData.joseph}
-\
-The $K_\textrm{leaf}$ was measured under steady state and high irradiance following the Evaporative Flux Method (FEM) [@Sack.2012] using a flowmeter.
-Briefly, the leaf was excised under filtered deionised water and attached to a water source (graduated cylinder) using clear plastic tubing.
-The system includes tubing of a known resistance, separated upstream and downstream by two sensors measuring the transpiration-driven water flow pressure in the EFM system [@Sack.2011].
-The $K_\textrm{leaf}$ is calculated as the ratio of the transpiration rate (*E*~leaf~; mmol s^-1^) to the difference in flow pressure (MPa) upstream and downstream from the resistance tubing, and further standardized to 25°C and by leaf area.
-At stability (CV < 5%), the leaf is left to equilibrate in plastic bags with moist paper towels and measured for $\mathrm{\Psi}$~leaf~ using a Scholander pressure chamber.
-To construct vulnerability curves for each species, four different functions were fitted to species-specific $K_\textrm{leaf}$ vs.
-$\mathrm{\Psi}$~leaf~ data using the *anneal* optimization function in the R package likelihood: exponential without an intercept ($K_{leaf} = Ae^{-B\Psi_{leaf}}$), exponential with an intercept ($K_{leaf} = C + Ae^{-B\Psi_{leaf}}$), logistic ($K_{leaf} = \frac{A}{1 + \left(\frac{\Psi_{leaf}}{x_{0}}\right)^B}$) and sigmoidal ($K_{leaf} = \frac{A}{1 + e^{-\left(\frac{\Psi_{leaf} - x_{0}}{B}\right)}}$) [@Sack.2012].
-The differences among AICc values for these functions were not large within species (see Zailaa et al.
-for more detailed methods description).
-We chose the exponential function without the intercept to fit observed data across species, and as a general function to scale-up to the community, because it had the lowest AICc for most species, captured the form of the observed data across all species well, and had the lowest number of parameters, that is, two.
-
-### Stem maximum hydraulic conductivity and vulnerability to cavitation {#SIData.brett}
-
-\
-Values of stem maximum hydraulic conductivity and vulnerability to cavitation were obtained from the dataset of @Wolfe.2020.
-The entire dataset includes measurements on 26 tree species collected at three sites in Panama.
-The raw data, *R* script for the analyses, and data summaries are archived in the NGEE-Tropics archive (@Wolfe.2020).
-The data were previously reported by @Dickman.2019 and @Wu.2020.
-Here, the seven species that overlapped with our assessment of ERD were included in analyses (Table S\@ref(tab:hyd-table)).
-
-\
-Methods for stem maximum hydraulic conductivity and vulnerability to cavitation closely followed [@Christman.2012] and [@Wolfe.2016].
-Briefly, canopy branches were collected from 4–10 trees of each species.
-Segments measured for hydraulic conductivity were removed from the branches and placed in an apparatus similar to that described by @Sperry.1988 except that flow rate was measured with graduated pipettes instead of a balance.
-Each segment was perfused with a 10 mM KCl solution filtered to 0.2 mM at three pressure heads: two that ranged 0.5–2 kPa and at zero pressure.
-Following @TorresRuiz.2012, hydraulic conductivity was calculated as the slope of the flow rate versus pressure gradient (i.e., pressure head divided by segment length) among the three pressure heads.
-Stem area specific hydraulic conductivity ($K_\textrm{stem}$) was calculated by dividing hydraulic conductivity by the cross-sectional area of the stem.
-For each $K_\textrm{stem}$ segment, three measurements of stem water potential were made on stem sections that were located adjacent to the $K_\textrm{stem}$ segment using psychrometers as described by @Wolfe.2017.
-
-\
-To assess vulnerability to cavitation, the branches were allowed to dry in the lab for 0 – 180 hours before they were measured for stem water potential and $K_\textrm{stem}$.
-For each species, $K_\textrm{stem}$ was plotted against stem water potential and a Weibull curve was fit through the 90th percentile of the points, following @Wolfe.2016.
-The intercept of the curve was interpreted as maximum $K_\textrm{stem}$ and the water potential where the Weibull curve predicted 88% loss of maximum $K_\textrm{stem}$ was interpreted as $\mathrm{\Psi}$~88,stem~.
-
-### Leaf turgor loss point
-
-For each tree species measured for $K_\textrm{stem}$ and $\mathrm{\Psi}$~88,stem~, two to six leaves were measured for $\mathrm{\Psi}$~tlp~ with pressure-volume analysis following [@Koide.1989].
-Leaf water potential was measured on leaf discs with psychrometers (J.R.D.
-Merrill Specialty Equipment, Logan, Utah, USA).
-This method has been shown to give results similar to those of the pressure chamber method [@Nardini.2008].
-To construct the pressure-volume curves, leaf discs were repeatedly measured for water potential and mass.
-Between measurements the discs were bench dried for 5-30 minutes.
-Dry mass was determined after oven drying the discs at 70°C.
-
-### Stem hydraulic safety margins
-
-Stem hydraulic safety margin was calculated as the difference between minimum leaf water potentials, $\mathrm{\Psi}$~min~, and $\mathrm{\Psi}$~88,stem~.
-Values for $\mathrm{\Psi}$~min~ were obtained from the dataset of @Wolfe.2019, which was collected as part of the 2016 ENSO campaign.
-Pre-dawn and diurnal leaf water potentials were measured on a monthly basis from Feb to May 2016 at San Lorenzo and Parque Natural Metropolitano, Panama, for which leaves were sampled using a canopy crane.
-Measurements in BCI were available only for March 2016, for which leaves were sampled with a pole pruner, as there is no canopy crane on BCI.
-In total 25 species were sampled, with no species re-sampled on another site.
-The raw data and data summaries are archived in the NGEE-Tropics archive [@Wolfe.2019].
-Here, the six species that overlapped with our assessment of ERD and $\mathrm{\Psi}$~88,stem~ were included in analyses.
-We defined species $\mathrm{\Psi}$~min~ as the minimum of all the diurnal measurements for a species, after taking an average value of measurements on two leaf samples for a given day (Table S\@ref(tab:hyd-table)).
-
-### WSG and LMA {#SIData.rutuja}
-\
-We used WSG, the wood specific gravity after drying at 100^$\circ$^C and LMA, the leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules for leaves receiving direct sunlight as described in @Wright:2010.
-
-\
-Among the `r length(sp.exp.param$sp)` species with observed data on ($K_\textrm{leaf}$ vs.
-$\mathrm{\Psi}$~leaf~) used to build the community-level models of leaf vulnerability curves (Eqs.
-\@ref(eq:b-mod)-\@ref(eq:a-mod)), two lacked data on WSG and five on LMA~lamina~.
-We filled three gaps in LMA using a linear relationship between LMA and LMA~disc~, the latter defined as the mean leaf mass per unit area measured for a 1.483 cm^2^ leaf disc taken to avoid veins (g m^-2^) for leaves receiving direct sunlight ($\overline{\textrm{LMA}} = `r disc.cf[1]` + `r disc.cf[2]`\textrm{LMA}_{disc}$; Adj.
-$R$^2^ = `r disc.r2`; $p$ `r disc.p`, $n$ = `r disc.n`).
-We filled the two remaining gaps in LMA using a linear relationship between LMA and LMA~lamina~, as measured during the smaller, $K_\textrm{leaf}$ campaign ($\overline{\textrm{LMA}} = `r lma.cf[1]` + `r lma.cf[2]`\textrm{LMA}_{lamina}$; Adj.
-$R$^2^ = `r lma.r2`; $p$ `r lma.p`, $n$ = `r lma.n`).
-We chose not to fill the two data gaps in WSG as available data were poor predictors of WSG.
-
-## Alternative models for Effective Rooting Depth (ERD)
-\
-Below we present all the alternative models for Effective Rooting Depth (ERD).
-Daily average species-specific growth, $G_{t|z}$, in the census interval $t$ given an ERD of depth $z$, is described as follows:
-
-\begin{equation}
-G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,i|z}^{*}) + \epsilon_{t|z}
-(\#eq:M1)
-\end{equation}
-
-\begin{equation}
-G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*} + \textrm{VPD}_{i}^{*}) + \epsilon_{t|z}
-(\#eq:M2)
-\end{equation}
-
-\begin{equation}
-G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*} \textrm{VPD}_{i}^{*}) + \epsilon_{t|z}
-(\#eq:M3)
-\end{equation}
-
-\begin{equation}
-G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*} \textrm{LAI}_{doy|i}^{*}) + \epsilon_{t|z}
-(\#eq:M4)
-\end{equation}
-
-\begin{equation}
-G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} (K^{*}_{\textrm{leaf},i|z} + \textrm{VPD}^{*}_{i}) \textrm{LAI}^{*}_{doy|i}) + \epsilon_{t|z}
-(\#eq:M5)
-\end{equation}
-
-\begin{equation}
-G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*}\textrm{VPD}_{i}^{*}\textrm{LAI}_{doy|i}^{*}) + \epsilon_{t|z}
-(\#eq:M6)
-\end{equation}
-where n~t~ are the total number of days in census interval $t$, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is the model error term.
-
-# Supplementary Figures
-
-```{r ksat-profiles, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ksat-profiles}Generated 5000 curves of soil hydraulic conductivity variation by depth based on Latin Hypercube Sampling within 95% CI of observed data at depths of 12.5 cm and 60 cm [@Kinner:2004], assuming a linear decline between those depths, and values before 12.5 and after 60 cm as that of 12.5 cm and 60 cm, respectively.
-Only 100 curves are shown for clarity.
-Note the log scale on horizontal axis.")}
-knitr::include_graphics(file.path("figures/Bootstrapped Ksat profiles within observed 95CI_random_two_hundred.jpeg"))
-```
-
-```{r TDR-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-TDR-calib}Volumetric Water Content vs.
-TDR data for a vertical TDR probe (0-15 cm depth) at the base of the eddy flux tower near the 50 ha plot at BCI."}
-knitr::include_graphics(file.path("figures/Matteos_TDR_SWC_data_linear_model.jpeg"))
-```
-
-```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "35%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd}Scatterplot of mean midday (11:00 am to 15:00 pm) gross primary productivity (GPP) and mean midday vapor pressure deficit (VPD) measured at the eddy flux tower in Barro Colorado Island over 2012-2017.
-Red line shows a fit from a polynomial model described in inset, and used to standardize VPD employed in equations \\@ref(eq:M2) and \\@ref(eq:M4)."}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/gpp_by_env_variables/monthly_gpp.vpd.jpeg"))
-```
-
-```{r leaf-cover, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-cover}Seasonal variation in standardised LAI (range 0-1) for ", length(erd.sp)," species at Barro Colorado Island (panels).
-Species are color-coded by deciduousness.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf.cover_BCI_multi_panel.jpeg"))
-```
-
-```{r iso-data, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-iso-data}Stable isotopic ratio ($\\mathrm{\\delta}^2H$) of soil water by depth measured at Barro Colorado Island in March 1997.
-Data from @Meinzer:1999."}
-knitr::include_graphics(file.path("figures/UDI_confidence/Meinzer_etal_1999_Depth_vs_soil_deltaD.jpeg"))
-```
-
-```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($\\Psi_{\\textrm{soil},z}$) by depth $z$ (m; panels) over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines).
-Vertical gray lines demarcate median dates of the demographic censuses at the BCI 50 ha plot."}
-knitr::include_graphics(file.path("figures/2019-10-14_5000/best-fits/psi_model_daily_all_depths_params.top.few_full.jpeg"))
-```
-
-```{r sens-sm-et, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-sm-et}Parameter sensitivity of ELM-FATES soil water content (that is, VWC, a-b) and ET (c-d) is shown for two representative sample months, as sensitivity for a particular parameter varies depending on the chosen month.
-Horizontal axes show global ranges for a given parameter.
-Parameters are color coded and are consistent across panels.
-Each point represents a simulation among the 5000 ensemble-member runs.
-See S\\@ref(tab:param) for description, global ranges and chosen ranges for parameters."}
-knitr::include_graphics(file.path("figures/ELM_FATES_parameter_sensitivity/sens.sm.et.jpeg"))
-```
-
-```{r sens-run, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-run}Parameter sensitivity of ELM-FATES stream discharge (QRUNOFF) is shown for four sample months (a-d), as sensitivity for a particular parameter varies depending on the chosen month.
-Horizontal axes show global ranges for a given parameter.
-Parameters are color coded and are consistent across panels.
-Each point represents a simulation among the 5000 ensemble-member runs.
-See S\\@ref(tab:param) for description, global ranges and chosen ranges for parameters."}
-knitr::include_graphics(file.path("figures/ELM_FATES_parameter_sensitivity/sens.run.jpeg"))
-```
-
-```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}Vulnerability curves fitted to observed $K_\\textrm{leaf}$ vs.
-$\\mathrm{\\Psi}$~leaf~ data for ", length(sp.exp.param$sp)," tree species from BCI with Eq.
-\\@ref(eq:vc-exp)--in absolute terms (a, b), as well as, in terms of loss of conductivity (%) (c, d).
-Each line represents a species, color coded by WSG (a, c), or LMA (b, d), with blue tones indicating smaller values of WSG or LMA and red tones indicating larger values.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/plot.obs.plc.jpeg"))
-```
-
-```{r obs-comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-comm-plc-curves}Percent loss of vulnerability curves for species with ERD estimates.
-Each line represents a species, color coded by source of the curve: fitted to observed data for $K_\\textrm{leaf}$ vs.
-$\\mathrm{\\Psi}$~leaf~ are in yellow and those based on the scaling relationship with WSG and LMA are in gray.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/obs.mod_plccurves_fits_ggplot_no_col.jpeg"))
-```
-
-```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-other-erd-mods}Effective Rooting Depth versus $\\mathrm{\\delta}^2H$ for each of the ERD models tested (panels).
-For each ERD model only those species have an estimate of ERD that have passed the test of Pearson's $r$ > 0.71 during ERD parameterisation.
-The number of species with an ERD estimate as well as $\\mathrm{\\delta}^2H$ for each model are shown in inset, along with the goodness-of-fit ($R$^2^) and $p$-values for a linear relationship between ERD and $\\mathrm{\\delta}^2H$.
-ERD models in panels a-e correspond to Eq.
-\\@ref(eq:M1)-\\@ref(eq:M6), respectively.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_sp_color_Meinzer.jpeg"))
-```
-
-```{r erd-hyd-full, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd-full}.
-Pairwise correlations (upper triangle) and corresponding linear regressions (lower triangle) among modelled Effective Rooting Depth (ERD) and hydraulic properties for seven canopy species found on Barro Colorado Island (Panama).
-Spearman's correlation coefficients and significance levels are shown (upper-half; *$p$ < 0.05; .
-$p$ < 0.1)")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/erd.stem.traits_cor.chart.jpeg"))
-```
-
-```{r erd-mort-census, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-census}Species mortality rate in each census interval during the study period plotted against ERD.
-$R$^2^ and $p$-values for linear model fits are given in panel insets.
-Model fits (blue lines) with SE (gray area) are shown for significant fits at $\\alpha$ = 0.1."}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_by rdi.gr.jpeg"))
-```
-
-```{r erd-time-evg, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time-evg}Modelled species Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\\mathrm{\\Psi}_{\\textrm{soil},z}$ was below $\\mathrm{\\Psi}$~crit~, defined as $\\mathrm{\\Psi}$~80,leaf~, over five census intervals (1990-2015), where $z$ is the soil depth matching species ERD--for ", nrow(mrate.depth.mean %>% subset(deciduous == "E")), " canopy species found on Barro Colorado Island.
-Each circle represents a species.
-Circles are colored and slightly jittered horizontally for clarity.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mfac vs. rdi.gr_evergreen.jpeg"))
-```
-
-#Supplementary Tables
-
-```{r param, results = 'asis', echo = FALSE}
-knitr::kable(param.table, caption = 'ELM-FATES parameters used to generate ensembles, with description, prescribed global ranges, rationale for the choice of ranges and references.
-Ranges for best-fit chosen ensembles are also given.
-Note that no.
-1-6 are parameters specific to FATES.', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(param.table))))
-```
-#
-```{r eddy, results = 'asis', echo = FALSE}
-knitr::kable(eddy.table, caption = 'QA/QC procedure applied to eddy covariance fluxes.
-w: vertical wind speed; t: air temperature; q: water vapor density; c: CO~2~ density; u: horizontal wind speed.
-The bar indicates 30-min averaging and the carat indicates fluctuation ($\\overline{\\hat{x} = x - \\bar{x}}$)', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(eddy.table))))
-```
-#
-```{r ab-param, results = 'asis', echo = FALSE}
-knitr::kable(ab.table, caption = paste0("\\label{lab-ab-param}Leaf vulnerability curves parameters *A* & *B*, $K_\\textrm{max,leaf}$ and $\\mathrm{\\Psi}$ for the ", length(erd.sp), " species with ERD estimates.
-Species for which curves were fitted to observed data for $K_\\textrm{leaf}$ vs.
-$\\mathrm{\\Psi}$~leaf~ are identified with Source as Data, while species for which curves were obtained based on trait-proxies are identified with Source as Model."), booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(ab.table))))
-```
-#
-```{r hyd-table, results = 'asis', echo = FALSE}
-knitr::kable(hyd.table, caption = paste0("\\label{lab-hyd-table} Minimum leaf water potential $\\mathrm{\\Psi}$~min~, leaf turgor loss point $\\mathrm{\\Psi}$~tlp~, stem maximum hydraulic conductivity $K_\\textrm{max,stem}$ and stem vulnerability to cavitation $\\mathrm{\\Psi}$~88,stem~ for seven species with ERD estimates.
-$\\mathrm{\\Psi}$~min~ is sourced from @Wolfe.2019, while $\\mathrm{\\Psi}$~tlp~, $K_\\textrm{max,stem}$ and $\\mathrm{\\Psi}$~88,stem~ are sourced from @Wolfe.2020.
-$\\mathrm{\\Psi}$~min~, $\\mathrm{\\Psi}$~tlp~ and $\\mathrm{\\Psi}$~88,stem~ are in MPa, while $K_\\textrm{max,stem}$ in kg m^-1^ s^-1^ MPa^-1^."), booktabs = TRUE, escape = FALSE)
-```
-
 # References

### 2020-11-05 Supplementary_Materials.Rmd
1. added SI from manuscript.rmd

---
title: "Hydraulically-vulnerable trees survive on deep-water access during droughts in a tropical forest"
author: ''
date: ''
output:
  bookdown::word_document2:
        reference_docx: manuscriptstyle.docx
        fig_caption: yes
  bookdown::pdf_document2: default
  bookdown::html_document2:
    number_sections: no
    theme: flatly
    fig_caption: yes
bibliography: references.bib
csl: new-phytologist.csl
---

```{r, include=FALSE, echo = FALSE}
if(!require("pacman", quietly=TRUE)) install.packages("pacman")
suppressPackageStartupMessages(
  pacman::p_load(bookdown, knitr, pander, citr, broom))
options(tinytex.verbose = TRUE)
source("code/load.R")
source("code/manuscript_data_figs.R")
# load("results/manuWorkSpace.RData")
# install.packages(c("vctrs", "rlang"))
```

```{r setup, echo = FALSE}
library(knitr)
opts_knit$set(eval.after = "fig.cap")
```

#Supplementary Information

## Data Descriptions 
### Growth Data 
\
We selected trees for which the height of the diameter measurements were unchanged across all censuses (1990, 1995, 2000, 2005, 2010, 2015). For each individual tree ($i$), we calculated absolute growth rates ($G_{i,t}$, cm yr^-1^) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses prorated by the inter-census interval based on the exact date of measurements. We eliminated positive outliers as any growth rates amounting to >75 mm yr^-1^, as this was the highest rate observed with confidence for the fastest growing species [@Condit.2017]. Negative outliers were removed based on known measurement error, $\sigma_{e} = 0.006214d + 0.903$, where *d* is dbh. Any stem whose later dbh measurement fell <4$\sigma_{e}$ the earlier measurement was removed [@Condit.2017]. As tree growth rates vary with size [@Muller-Landau:2006], to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($\textrm{Resid}_{i,t} = G_{i,t} - f(d_{i,t})$). The diameter model of growth, $f(d_{i, t})$, a B-spline based polynomial equation with five degrees of freedom, was fitted to growth rates from all trees pooled together as there was insufficient species-specific data for large trees ($R$^2 = 0.27; $p$-value < 0.01). We retained only those trees that had $\textrm{Resid}_{i,t}$ estimates for all census intervals (total trees = `r sum(g.n$n)`) and standardized (centered and scaled) individual tree time series of $\textrm{Resid}_{i,t}$. We then obtained `r length(erd.sp)` species-level time-series of medians of $\textrm{Resid}_{i,t}$ ($G_{t}$). These were obtained for only those species with complete records on at least three trees (median $n$ = `r median(g.n$n)`, max $n$ = `r max(g.n$n)` trees per species).

### Microclimatic tower {#SIData.matteo} 
\
The tower used for these measurements is 41 m above ground, on a plateau on BCI [@Pau.2018]. The eddy covariance system includes a sonic anemometer (CSAT3, Campbell Scientific, Logan, UT) and an open-path infrared CO~2~/H~2~O gas analyzer (LI7500, LiCOR. Lincoln, NE). Hi-frequency (10Hz) measurements were acquired by a datalogger (CR1000, Campbell Scientific) and stored on a local PC. Data were processed with a custom program using a standard routine described in [@Detto.2010]. QA/QC criteria for removing erroneous values are listed in table 1 (see below). GPP was derived from daytime values of NEE by adding the corresponding mean daily ecosystem respiration obtained as the intercept of the light response curve [@Lasslop.2010]. The light curve was fitted on a 15-days moving window using a rectangular hyperbolic function (runs with friction velocity less than 0.4 m s^-1^ were excluded). In order to compute daily time integrated budgets, gaps were filled using Artificial Neural Network [@Papale.2003] with hydro-meteorological inputs as predictors (soil moisture, solar radiation, temperature, VPD and air pressure, all measured on the same tower). To train the network, the dataset was randomly divided in a training set (70%), a validation set (15%) and a test set (15%). A two-layer feed-forward network with 10 sigmoid hidden neurons and linear output neurons, was trained using the Levenberg-Marquardt algorithm until the mean square error (MSE) of the validation set stopped improving [@Hagan.1994]. Performance, in terms of MSE, was evaluated using the test set at the end of the training. This procedure was repeated 100 times to produce 100 estimates of GPP. Training multiple times generates different results due to different initial conditions and random sampling of the three sets. Ensemble was obtained as weighted average from the 100 ANN predictions using the MSE of the test set as weights according to:

$f_{i} = \frac{\sum_{k}f_{ik}^{ANN}/{MSE_{ik}}}{\sum_{k}1/{MSE_{ik}}}$
\
The ANN was implemented using the Neural Network Toolbox in Matlab 2014a. 
\
The associated data can be requested from https://ameriflux.lbl.gov.

### Stream discharge from the Conrad catchment {#SIData.steveP} 
\
The Conrad weir is located on the gently sloping western side of Barro Colorado Island. The weir consists of a 90 degree ‘V’ notch with a rectangular upper section. The weir was constructed in several stages between 1993 and 1996. Stream stage is recorded at 5-minute intervals.
\
The weir drains a catchment of approximately 40 ha of which at least 90% consists of the central plateau area of the island. Peak flow usually follows peak rain by approximately 2-3 hours. Large storms are sometimes preceded by small, initial peaks generated by the rain falling on the steeper slopes close to the weir.
\
Streamflow data are analyzed by software custom-built by author SP. The program is used to correct for blockages of the weir ‘V’ by fallen vegetation (a common occurrence), sensor drift, small gaps and other problems. Discharge is then calculated using an empirically derived rating curve calculated for each 1mm stream stage.

### Volumetric water content

\
We collected volumetric water content (VWC) data (2012-2018) from three locations near the eddy flux tower with time domain reflectometry (TDR, CS616, Campbell Scientific) probes inserted with an angle to average the first 15 cm of soil. The relationship between the output from the TDR probe and VWC was calibrated based on measurements of gravimetric water content (GWC) on several occasions from representative soils near the vertical probes (0-0.15 m) (Fig. S\@ref(fig:TDR-calib)). GWC was converted to VWC based on the observed mean bulk dry density of 0.8 g cm^-3^ near the TDR probe ($n$ = 5, cylinder volume = 1250 cm^3^). Beginning in 2016 we additionally installed three horizontal probes at depths of 0.15, 0.4 and 1 m and we applied the same calibration equation to these data (2016-2018) as that for the vertical probe. These data are archived at https://ngt-data.lbl.gov/.

\
We also leveraged existing GWC measurements by @Kupers:2019 at the depths of 0.15, 0.4 and 1 m in summer 2015 (February, March and April) and 0.15, 0.4 in summer 2016 (March), with 1299 samples in total that covered all soil types and habitats in the plot. The sampling periods were six, five, ten and eight days long, respectively. The 2016 dry season was associated with the 2015–2016 El Niño (see @Kupers:2019 for further details). We converted these GWC measurements to VWC assuming a bulk density of 0.8 g cm^-3^. $\theta_{\textrm{0.15 m}}$, $\theta_{\textrm{0.4 m}}$ and $\theta_{\textrm{1 m}}$ refer to the plot-wide VWC averages by depth over each sampling period. Middle date of a sampling period is used for comparison with model simulations.

### Leaf Area Index 
\
We generated an estimate of the mean seasonality curve for normalized leaf area index (LAI$^*$, range 0-1, unitless) for individual species as follows. We interpolated species-specific weekly leaf-fall data collected at two similar sites in BCI (1985-2020) to a daily estimate [@Wright.1990aeo]. We lead this daily leaf-fall time series by the species-specific average leaf-lifetime (in days) and used it as an estimate of the daily leaf-gain time-series. The difference between leaf-fall and leaf-gain gave an estimate of daily leaf-change time series, which was then scaled to express as a fraction of the total leaf fall in a given year. Beginning with the LAI$^*$ of 1 at the end of the wet season in the first year of the record, we accumulated leaf-change over it to obtain a running estimate of the leaf-cover fraction, averaged it by DOY across years and normalized it to obtain the species-specific seasonality curve in LAI$^*$ (range 0-1, unitless). (Fig. S\@ref(fig:leaf-cover)).
\
For five species, an estimate of leaf lifetime was available from direct observations [@Osnas.2018]. For the rest erd.sp.wo.ll species, leaf lifetime was predicted from LMA ($\overline{lifetime} = `r ll.lma.cf[1]` + `r ll.lma.cf[2]`LMA$; $R$^2^ = `r ll.lma.r2`; $p$ `r ll.lma.p`, $n$ = `r ll.lma.n`).

### Leaf hydraulic conductivity and vulnerability to cavitation {#SIData.joseph}
\
The $K_\textrm{leaf}$ was measured under steady state and high irradiance following the Evaporative Flux Method (FEM) [@Sack.2012] using a flowmeter. Briefly, the leaf was excised under filtered deionised water and attached to a water source (graduated cylinder) using clear plastic tubing. The system includes tubing of a known resistance, separated upstream and downstream by two sensors measuring the transpiration-driven water flow pressure in the EFM system [@Sack.2011]. The $K_\textrm{leaf}$ is calculated as the ratio of the transpiration rate (*E*~leaf~; mmol s^-1^) to the difference in flow pressure (MPa) upstream and downstream from the resistance tubing, and further standardized to 25°C and by leaf area. At stability (CV < 5%), the leaf is left to equilibrate in plastic bags with moist paper towels and measured for $\mathrm{\Psi}$~leaf~ using a Scholander pressure chamber. To construct vulnerability curves for each species, four different functions were fitted to species-specific $K_\textrm{leaf}$ vs. $\mathrm{\Psi}$~leaf~ data using the *anneal* optimization function in the R package likelihood: exponential without an intercept ($K_{leaf} = Ae^{-B\Psi_{leaf}}$), exponential with an intercept ($K_{leaf} = C + Ae^{-B\Psi_{leaf}}$), logistic ($K_{leaf} = \frac{A}{1 + \left(\frac{\Psi_{leaf}}{x_{0}}\right)^B}$) and sigmoidal ($K_{leaf} = \frac{A}{1 + e^{-\left(\frac{\Psi_{leaf} - x_{0}}{B}\right)}}$) [@Sack.2012]. The differences among AICc values for these functions were not large within species (see Zailaa et al. for more detailed methods description). We chose the exponential function without the intercept to fit observed data across species, and as a general function to scale-up to the community, because it had the lowest AICc for most species, captured the form of the observed data across all species well, and had the lowest number of parameters, that is, two.

### Stem maximum hydraulic conductivity and vulnerability to cavitation {#SIData.brett}

\
Values of stem maximum hydraulic conductivity and vulnerability to cavitation were obtained from the dataset of @Wolfe.2020. The entire dataset includes measurements on 26 tree species collected at three sites in Panama. The raw data, *R* script for the analyses, and data summaries are archived in the NGEE-Tropics archive (@Wolfe.2020). The data were previously reported by @Dickman.2019 and @Wu.2020. Here, the seven species that overlapped with our assessment of ERD were included in analyses (Table S\@ref(tab:hyd-table)).

\
Methods for stem maximum hydraulic conductivity and vulnerability to cavitation closely followed [@Christman.2012] and [@Wolfe.2016]. Briefly, canopy branches were collected from 4–10 trees of each species. Segments measured for hydraulic conductivity were removed from the branches and placed in an apparatus similar to that described by @Sperry.1988 except that flow rate was measured with graduated pipettes instead of a balance. Each segment was perfused with a 10 mM KCl solution filtered to 0.2 mM at three pressure heads: two that ranged 0.5–2 kPa and at zero pressure. Following @TorresRuiz.2012, hydraulic conductivity was calculated as the slope of the flow rate versus pressure gradient (i.e., pressure head divided by segment length) among the three pressure heads. Stem area specific hydraulic conductivity ($K_\textrm{stem}$) was calculated by dividing hydraulic conductivity by the cross-sectional area of the stem. For each $K_\textrm{stem}$ segment, three measurements of stem water potential were made on stem sections that were located adjacent to the $K_\textrm{stem}$ segment using psychrometers as described by @Wolfe.2017.

\
To assess vulnerability to cavitation, the branches were allowed to dry in the lab for 0 – 180 hours before they were measured for stem water potential and $K_\textrm{stem}$. For each species, $K_\textrm{stem}$ was plotted against stem water potential and a Weibull curve was fit through the 90th percentile of the points, following @Wolfe.2016. The intercept of the curve was interpreted as maximum $K_\textrm{stem}$ and the water potential where the Weibull curve predicted 88% loss of maximum $K_\textrm{stem}$ was interpreted as $\mathrm{\Psi}$~88,stem~.

### Leaf turgor loss point

For each tree species measured for $K_\textrm{stem}$ and $\mathrm{\Psi}$~88,stem~, two to six leaves were measured for $\mathrm{\Psi}$~tlp~ with pressure-volume analysis following [@Koide.1989]. Leaf water potential was measured on leaf discs with psychrometers (J.R.D. Merrill Specialty Equipment, Logan, Utah, USA). This method has been shown to give results similar to those of the pressure chamber method [@Nardini.2008]. To construct the pressure-volume curves, leaf discs were repeatedly measured for water potential and mass. Between measurements the discs were bench dried for 5-30 minutes. Dry mass was determined after oven drying the discs at 70°C.

### Stem hydraulic safety margins

Stem hydraulic safety margin was calculated as the difference between minimum leaf water potentials, $\mathrm{\Psi}$~min~, and $\mathrm{\Psi}$~88,stem~. Values for $\mathrm{\Psi}$~min~ were obtained from the dataset of @Wolfe.2019, which was collected as part of the 2016 ENSO campaign. Pre-dawn and diurnal leaf water potentials were measured on a monthly basis from Feb to May 2016 at San Lorenzo and Parque Natural Metropolitano, Panama, for which leaves were sampled using a canopy crane. Measurements in BCI were available only for March 2016, for which leaves were sampled with a pole pruner, as there is no canopy crane on BCI. In total 25 species were sampled, with no species re-sampled on another site. The raw data and data summaries are archived in the NGEE-Tropics archive [@Wolfe.2019]. Here, the six species that overlapped with our assessment of ERD and $\mathrm{\Psi}$~88,stem~ were included in analyses. We defined species $\mathrm{\Psi}$~min~ as the minimum of all the diurnal measurements for a species, after taking an average value of measurements on two leaf samples for a given day (Table S\@ref(tab:hyd-table)).

### WSG and LMA {#SIData.rutuja}
\
We used WSG, the wood specific gravity after drying at 100^$\circ$^C and LMA, the leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules for leaves receiving direct sunlight as described in @Wright:2010.

\
Among the `r length(sp.exp.param$sp)` species with observed data on ($K_\textrm{leaf}$ vs. $\mathrm{\Psi}$~leaf~) used to build the community-level models of leaf vulnerability curves (Eqs. \@ref(eq:b-mod)-\@ref(eq:a-mod)), two lacked data on WSG and five on LMA~lamina~. We filled three gaps in LMA using a linear relationship between LMA and LMA~disc~, the latter defined as the mean leaf mass per unit area measured for a 1.483 cm^2^ leaf disc taken to avoid veins (g m^-2^) for leaves receiving direct sunlight ($\overline{\textrm{LMA}} = `r disc.cf[1]` + `r disc.cf[2]`\textrm{LMA}_{disc}$; Adj. $R$^2^ = `r disc.r2`; $p$ `r disc.p`, $n$ = `r disc.n`). We filled the two remaining gaps in LMA using a linear relationship between LMA and LMA~lamina~, as measured during the smaller, $K_\textrm{leaf}$ campaign ($\overline{\textrm{LMA}} = `r lma.cf[1]` + `r lma.cf[2]`\textrm{LMA}_{lamina}$; Adj. $R$^2^ = `r lma.r2`; $p$ `r lma.p`, $n$ = `r lma.n`). We chose not to fill the two data gaps in WSG as available data were poor predictors of WSG.

## Alternative models for Effective Rooting Depth (ERD)
\
Below we present all the alternative models for Effective Rooting Depth (ERD). Daily average species-specific growth, $G_{t|z}$, in the census interval $t$ given an ERD of depth $z$, is described as follows:

\begin{equation}
G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,i|z}^{*}) + \epsilon_{t|z}
(\#eq:M1)
\end{equation}

\begin{equation}
G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*} + \textrm{VPD}_{i}^{*}) + \epsilon_{t|z}
(\#eq:M2)
\end{equation}

\begin{equation}
G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*} \textrm{VPD}_{i}^{*}) + \epsilon_{t|z}
(\#eq:M3)
\end{equation}

\begin{equation}
G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*} \textrm{LAI}_{doy|i}^{*}) + \epsilon_{t|z}
(\#eq:M4)
\end{equation}

\begin{equation}
G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} (K^{*}_{\textrm{leaf},i|z} + \textrm{VPD}^{*}_{i}) \textrm{LAI}^{*}_{doy|i}) + \epsilon_{t|z}
(\#eq:M5)
\end{equation}

\begin{equation}
G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*}\textrm{VPD}_{i}^{*}\textrm{LAI}_{doy|i}^{*}) + \epsilon_{t|z}
(\#eq:M6)
\end{equation}
where n~t~ are the total number of days in census interval $t$, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is the model error term.

# Supplementary Figures

```{r ksat-profiles, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ksat-profiles}Generated 5000 curves of soil hydraulic conductivity variation by depth based on Latin Hypercube Sampling within 95% CI of observed data at depths of 12.5 cm and 60 cm [@Kinner:2004], assuming a linear decline between those depths, and values before 12.5 and after 60 cm as that of 12.5 cm and 60 cm, respectively. Only 100 curves are shown for clarity. Note the log scale on horizontal axis.")}
knitr::include_graphics(file.path("figures/Bootstrapped Ksat profiles within observed 95CI_random_two_hundred.jpeg"))
```

```{r TDR-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-TDR-calib}Volumetric Water Content vs. TDR data for a vertical TDR probe (0-15 cm depth) at the base of the eddy flux tower near the 50 ha plot at BCI."}
knitr::include_graphics(file.path("figures/Matteos_TDR_SWC_data_linear_model.jpeg"))
```

```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "35%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd}Scatterplot of mean midday (11:00 am to 15:00 pm) gross primary productivity (GPP) and mean midday vapor pressure deficit (VPD) measured at the eddy flux tower in Barro Colorado Island over 2012-2017. Red line shows a fit from a polynomial model described in inset, and used to standardize VPD employed in equations \\@ref(eq:M2) and \\@ref(eq:M4)."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/gpp_by_env_variables/monthly_gpp.vpd.jpeg"))
```

```{r leaf-cover, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-cover}Seasonal variation in standardised LAI (range 0-1) for ", length(erd.sp)," species at Barro Colorado Island (panels). Species are color-coded by deciduousness.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf.cover_BCI_multi_panel.jpeg"))
```

```{r iso-data, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-iso-data}Stable isotopic ratio ($\\mathrm{\\delta}^2H$) of soil water by depth measured at Barro Colorado Island in March 1997. Data from @Meinzer:1999."}
knitr::include_graphics(file.path("figures/UDI_confidence/Meinzer_etal_1999_Depth_vs_soil_deltaD.jpeg"))
```

```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($\\Psi_{\\textrm{soil},z}$) by depth $z$ (m; panels) over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines). Vertical gray lines demarcate median dates of the demographic censuses at the BCI 50 ha plot."}
knitr::include_graphics(file.path("figures/2019-10-14_5000/best-fits/psi_model_daily_all_depths_params.top.few_full.jpeg"))
```

```{r sens-sm-et, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-sm-et}Parameter sensitivity of ELM-FATES soil water content (that is, VWC, a-b) and ET (c-d) is shown for two representative sample months, as sensitivity for a particular parameter varies depending on the chosen month. Horizontal axes show global ranges for a given parameter. Parameters are color coded and are consistent across panels. Each point represents a simulation among the 5000 ensemble-member runs. See S\\@ref(tab:param) for description, global ranges and chosen ranges for parameters."}
knitr::include_graphics(file.path("figures/ELM_FATES_parameter_sensitivity/sens.sm.et.jpeg"))
```

```{r sens-run, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-run}Parameter sensitivity of ELM-FATES stream discharge (QRUNOFF) is shown for four sample months (a-d), as sensitivity for a particular parameter varies depending on the chosen month. Horizontal axes show global ranges for a given parameter. Parameters are color coded and are consistent across panels. Each point represents a simulation among the 5000 ensemble-member runs. See S\\@ref(tab:param) for description, global ranges and chosen ranges for parameters."}
knitr::include_graphics(file.path("figures/ELM_FATES_parameter_sensitivity/sens.run.jpeg"))
```

```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}Vulnerability curves fitted to observed $K_\\textrm{leaf}$ vs. $\\mathrm{\\Psi}$~leaf~ data for ", length(sp.exp.param$sp)," tree species from BCI with Eq. \\@ref(eq:vc-exp)--in absolute terms (a, b), as well as, in terms of loss of conductivity (%) (c, d). Each line represents a species, color coded by WSG (a, c), or LMA (b, d), with blue tones indicating smaller values of WSG or LMA and red tones indicating larger values.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/plot.obs.plc.jpeg"))
```

```{r obs-comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-comm-plc-curves}Percent loss of vulnerability curves for species with ERD estimates. Each line represents a species, color coded by source of the curve: fitted to observed data for $K_\\textrm{leaf}$ vs. $\\mathrm{\\Psi}$~leaf~ are in yellow and those based on the scaling relationship with WSG and LMA are in gray.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/obs.mod_plccurves_fits_ggplot_no_col.jpeg"))
```

```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-other-erd-mods}Effective Rooting Depth versus $\\mathrm{\\delta}^2H$ for each of the ERD models tested (panels). For each ERD model only those species have an estimate of ERD that have passed the test of Pearson's $r$ > 0.71 during ERD parameterisation. The number of species with an ERD estimate as well as $\\mathrm{\\delta}^2H$ for each model are shown in inset, along with the goodness-of-fit ($R$^2^) and $p$-values for a linear relationship between ERD and $\\mathrm{\\delta}^2H$. ERD models in panels a-e correspond to Eq. \\@ref(eq:M1)-\\@ref(eq:M6), respectively.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_sp_color_Meinzer.jpeg"))
```

```{r erd-hyd-full, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd-full}. Pairwise correlations (upper triangle) and corresponding linear regressions (lower triangle) among modelled Effective Rooting Depth (ERD) and hydraulic properties for seven canopy species found on Barro Colorado Island (Panama). Spearman's correlation coefficients and significance levels are shown (upper-half; *$p$ < 0.05; . $p$ < 0.1)")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/erd.stem.traits_cor.chart.jpeg"))
```

```{r erd-mort-census, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-census}Species mortality rate in each census interval during the study period plotted against ERD. $R$^2^ and $p$-values for linear model fits are given in panel insets. Model fits (blue lines) with SE (gray area) are shown for significant fits at $\\alpha$ = 0.1."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_by rdi.gr.jpeg"))
```

```{r erd-time-evg, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time-evg}Modelled species Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\\mathrm{\\Psi}_{\\textrm{soil},z}$ was below $\\mathrm{\\Psi}$~crit~, defined as $\\mathrm{\\Psi}$~80,leaf~, over five census intervals (1990-2015), where $z$ is the soil depth matching species ERD--for ", nrow(mrate.depth.mean %>% subset(deciduous == "E")), " canopy species found on Barro Colorado Island. Each circle represents a species. Circles are colored and slightly jittered horizontally for clarity.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mfac vs. rdi.gr_evergreen.jpeg"))
```

#Supplementary Tables

```{r param, results = 'asis', echo = FALSE}
knitr::kable(param.table, caption = 'ELM-FATES parameters used to generate ensembles, with description, prescribed global ranges, rationale for the choice of ranges and references. Ranges for best-fit chosen ensembles are also given. Note that no. 1-6 are parameters specific to FATES.', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(param.table))))
```
#
```{r eddy, results = 'asis', echo = FALSE}
knitr::kable(eddy.table, caption = 'QA/QC procedure applied to eddy covariance fluxes. w: vertical wind speed; t: air temperature; q: water vapor density; c: CO~2~ density; u: horizontal wind speed. The bar indicates 30-min averaging and the carat indicates fluctuation ($\\overline{\\hat{x} = x - \\bar{x}}$)', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(eddy.table))))
```
#
```{r ab-param, results = 'asis', echo = FALSE}
knitr::kable(ab.table, caption = paste0("\\label{lab-ab-param}Leaf vulnerability curves parameters *A* & *B*, $K_\\textrm{max,leaf}$ and $\\mathrm{\\Psi}$ for the ", length(erd.sp), " species with ERD estimates. Species for which curves were fitted to observed data for $K_\\textrm{leaf}$ vs. $\\mathrm{\\Psi}$~leaf~ are identified with Source as Data, while species for which curves were obtained based on trait-proxies are identified with Source as Model."), booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(ab.table))))
```
#
```{r hyd-table, results = 'asis', echo = FALSE}
knitr::kable(hyd.table, caption = paste0("\\label{lab-hyd-table} Minimum leaf water potential $\\mathrm{\\Psi}$~min~, leaf turgor loss point $\\mathrm{\\Psi}$~tlp~, stem maximum hydraulic conductivity $K_\\textrm{max,stem}$ and stem vulnerability to cavitation $\\mathrm{\\Psi}$~88,stem~ for seven species with ERD estimates. $\\mathrm{\\Psi}$~min~ is sourced from @Wolfe.2019, while $\\mathrm{\\Psi}$~tlp~, $K_\\textrm{max,stem}$ and $\\mathrm{\\Psi}$~88,stem~ are sourced from @Wolfe.2020. $\\mathrm{\\Psi}$~min~, $\\mathrm{\\Psi}$~tlp~ and $\\mathrm{\\Psi}$~88,stem~ are in MPa, while $K_\\textrm{max,stem}$ in kg m^-1^ s^-1^ MPa^-1^."), booktabs = TRUE, escape = FALSE)
```

# References

### 2020-11-06 manuscript.rmd
1. Re-arranged Tables and Figures sequence
1. In reference to Psi using more/less negative rather than above/below 2. mortality results include explanation for 1990-1995 census 3. figure captions edited

@@
 Summary | 197  
-Main Text  | 6532  
+Main Text  | 6498  
 Introduction | 828  
-Methods | 3278  
-Results | 855  
-Discussion | 1571  
+Methods | 3283  
+Results | 883  
+Discussion | 1504  
 Keywords | 6
@@
 \
+
 Item | Count  
@@
 We tested the relationship of ERD with xylem vulnerability to embolism, branch hydraulic safety margins, leaf turgor loss point and maximum stem-specific hydraulic conductivity.
-Finally, we analysed species-specific large tree mortality dynamics over 35 years and estimated the time a species’ spent below its critical hydraulic threshold to test whether ERD explained risk of drought-induced mortality.
+Finally, we analysed species-specific large tree mortality dynamics over 35 years and estimated the time a species’ spent beyond its critical hydraulic threshold to test whether ERD explained risk of drought-induced mortality.
 
@@
 We ran ELM-FATES with the chosen 100 ensemble members from 1985-2018, with the first 5 years used for model spin-up.
-Extreme hydrological droughts were identified by depth $z$ as days that fell below the 5^th^ percentile of $\mathrm{\Psi}_{\textrm{soil},z}$ for a given day of the year.
+Extreme hydrological droughts were identified by depth $z$ as days for which $\mathrm{\Psi}_{\textrm{soil},z}$ was more negative than the 5^th^ percentile of $\mathrm{\Psi}_{\textrm{soil},z}$ for a given day of the year.
 
@@
 where $L$ is the likelihood of $z$ given the observed growth time series, $H$ = {1, .., 100}, refer to the 100 hydrological realizations and $Z$ = {0.4, `r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m.
-We removed finely resolved soil layers below 0.21 m and averaged $\mathrm{\Psi}_{\textrm{soil},z}$ for depths 0.21, 0.37 and 0.62 at 0.4 m as sufficient resolution for Model validation.
+We removed finely resolved soil layers <0.21 m and averaged $\mathrm{\Psi}_{\textrm{soil},z}$ for depths 0.21, 0.37 and 0.62 at 0.4 m as sufficient resolution for Model validation.
 
@@
 We regressed species ERD against species $\textrm{M}_{s,t}$ for each census interval $t$ with ($n$ = `r nrow(mrate.depth.mean)`) and without deciduous species ($n$ = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`).
-To assess whether deeper ERD species escaped the experience of significant hydrological drought, we calculated for each species the proportion of days in the simulation period (1990-2015) during which soil water potential in the soil layer matching species ERD ($\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$) were below species $\mathrm{\Psi}$~crit~.
+To assess whether deeper ERD species escaped the experience of significant hydrological drought, we calculated for each species the proportion of days in the simulation period (1990-2015) during which soil water potential in the soil layer matching species ERD ($\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$) were more negative than species $\mathrm{\Psi}$~crit~.
 $\mathrm{\Psi}$~crit~ was defined as $\mathrm{\Psi}$~50,leaf~.
@@
 
-Among evergreen species, ERD explained `r m.evg.r2.90`, `r m.evg.r2.05` and `r m.evg.r2.10`% variation in mortality rates during census intervals 1985-1990, 2000-2005 and 2005-2010, respectively, such that species mortality rates significantly decreased with ERD ($p$ = `r m.evg.p.90`, `r m.evg.p.05` and `r m.evg.p.10`, respectively; Fig.
+Among evergreen species, ERD explained `r m.evg.r2.90`, `r m.evg.r2.05` and `r m.evg.r2.10`% variation in mortality rates during census intervals 1985-1990, 2000-2005 and 2005-2010, respectively, such that species mortality rates significantly decreased with deeper ERD ($p$ = `r m.evg.p.90`, `r m.evg.p.05` and `r m.evg.p.10`, respectively; Fig.
 \@ref(fig:erd-mort-census-evg)).
@@
 \@ref(fig:erd-mort-census)).
-Census intervals 2000-2005 and 2005-2010 were also the periods during which extreme hydrological droughts distinctly occurred in the wet season, effectively extending the dry season into the wet season (Fig.
-\@ref(fig:swp-droughts)).
-$\mathrm{\Psi}_{\textrm{soil},z}$ over 1985-1990 is not used for analyses, as that was the ELM-FATES spin-up period, while reliable climate forcing data was unavailable for the prior years.
-However, the difference between precipitation and ET shows that 1985-1990 had three extreme deficit water years, apart from the 1983 El-Niño leading to it [@Condit.2017].
-Analysis of a species' time spent under significant hydrological drought (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) showed that species with shallower ERD spent a greater proportion of days below critical hydraulic thresholds and increasingly so over 1995-2000, 2000-2005 and 2005-2010 (Fig.
+Census intervals 2000-2005 and 2005-2010 were also the periods during which extreme soil droughts distinctly occurred in the wet season, effectively extending the dry season into the wet season (Fig.
+\@ref(fig:swp-droughts).
+While our drought assessment did not extend to the model spin-up period of 1985-1990, for which ERD explained significant mortality, a prior study found that during this period BCI experienced three extreme deficit water years, following the 1983 El-Niño year [@Condit.2017].
+Analysis of a species' time spent under significant hydrological drought (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) showed that species with shallower ERD spent a greater proportion of days beyond critical hydraulic thresholds and increasingly so over 2000-2005 and 2005-2010 (Fig.
 \@ref(fig:erd-time); S\@ref(fig:erd-time-evg)).
-Species with ERD deeper than 2 m did not cross species $\mathrm{\Psi}$~crit~ for the entire 25-yr period for which soil water potential $\mathrm{\Psi}_{\textrm{soil},z}$ was modeled (1990-2015).
-Soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and drought periods, which is above ($\mathrm{\Psi}$~crit~) of the most sensitive tree species (-0.28 MPa).
+For the census interval 1990-1995, species with the shallowest ERD spent a high propotion of time beyond $\mathrm{\Psi}$~crit~ (Fig.
+\@ref(fig:erd-time)), but during the dry season (; Fig.
+\@ref(fig:swp-droughts), and did not have increased mortality rates (Fig.
+\@ref(fig:erd-mort-census).
+Overall, Species with ERD deeper than 1.7 m did not cross species $\mathrm{\Psi}$~crit~ for the entire 25-yr period for which soil water potential $\mathrm{\Psi}_{\textrm{soil},z}$ was modeled (1990-2015).
+Soil depths deeper than 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and drought periods, which is above $\mathrm{\Psi}$~crit~ of the most sensitive tree species (-0.28 MPa).
 
@@
 The hydrological model in CT2018 revealed that the multi-year drought exhausted the deep soil and even bedrock water availability (also see, @Goulden.2019 and @Ivanov:2012).
-In contrast, our hydrological modeling at BCI finds that during the study period, $\mathrm{\Psi}$~soil~ below 1 m never went below $\mathrm{\Psi}$~crit~ of the most sensitive tree species (-0.28 MPa) for all the dry seasons and droughts.
+In contrast, our hydrological modeling at BCI finds that during the study period, $\mathrm{\Psi}$~soil~ for depths deeper than 1 m did not cross $\mathrm{\Psi}$~crit~ of the most sensitive tree species (-0.28 MPa) for all the dry seasons and droughts.
 Mean annual rainfall of 1095 mm in CT2018 compared to `r round(rain.man.stats$rain)`mm at BCI, and precipitation to potential evapotranspiration ratio of nearly one in CT2018 compared to nearly two in BCI are major factors in the different mortality responses.
@@
 
+# References
+
 # Figure Captions 
@@
 Figure \@ref(fig:hydro-calib): ELM-FATES calibration and evaluation.
-Observations (red lines and points) versus simulations (gray lines) from 100 best-fit ensemble-member runs of ELM-FATES are shown for daily Volumetric Water Content (VWC) by depth for horizontal time domain reflectometry (TDR) probes at three depths (0.1, 0.4 and 1 m; A) and a vertical TDR probe (0-0.15 m; B); monthly stream discharge (Discharge, C) and monthly evapotranspiration from eddy flux tower (ET, D).
-Red points in (A) show average, manual plot-wide observations of VWC with 95% CI (error bars).
-All observations (red lines and points) were in-sample, except for VWC data from the horizontal probes (red lines in A) which were out-of-sample.
+Observations (red lines and points) versus simulations (gray lines) from 100 best-fit ensemble-member runs of ELM-FATES are shown for daily Volumetric Water Content (VWC) by depth for horizontal time domain reflectometry (TDR) probes at three depths (0.1, 0.4 and 1 m; **a**) and for an average of three vertical TDR probes (0-0.15 m; **b**); monthly stream discharge (Discharge, **c**) and monthly evapotranspiration from eddy flux tower (ET, **d**).
+Red points in (**a**) show average, manual plot-wide observations of VWC with 95% CI (error bars).
+All observations (red lines and points) were in-sample, except for VWC data from the horizontal probes (red lines in **a**) which were out-of-sample.
 Values in inset are average RMSE across the 100 best-fit simulations.
-For panel a, values at the top are for manual, plot-wide VWC and those at the bottom are for VWC from TDR probes.
-VWC is in units of cm^3^ cm^-3^, while ET and Discharge in mm month^-1^.
+For panel **a**, values at the top are for manual, plot-wide VWC and those at the bottom are for VWC from TDR probes.
+VWC is in units of cm^3^ cm^-3^, while ET and Discharge are in mm month^-1^.
 
@@
 Mean (black lines) and lower half of 95% distribution (grey areas) of soil water potential ($\mathrm{\Psi}_{\textrm{soil},z}$) for a given DOY and depth $z$ are shown (same across census intervals).
-$\mathrm{\Psi}_{\textrm{soil},z}$ by DOY for a year (colored line) is only shown if, at least one day of that year $\mathrm{\Psi}_{\textrm{soil},z}$ declined below the 5^th^ percentile for that day of year (DOY).
+$\mathrm{\Psi}_{\textrm{soil},z}$ by DOY for a year (colored line) is only shown if at least one day of that year $\mathrm{\Psi}_{\textrm{soil},z}$ was more negative than the 5^th^ percentile for that day of year (DOY).
 
 \
-Figure \@ref(fig:ab-fit): Parameters $A$ (panel a) & $B$ (panel b) for the leaf vulnerability curves that were fitted to observed data on $K_\textrm{leaf}$ vs.
+Figure \@ref(fig:ab-fit): Parameters $A$ (panel **a**) & $B$ (panel **b**) for the leaf vulnerability curves that were fitted to observed data on $K_\textrm{leaf}$ vs.
 $\mathrm{\Psi}$~leaf~ for `r length(sp.exp.param$sp)` species (Eq.
@@
 \@ref(eq:a-mod)-\@ref(eq:b-mod)).
-Goodness-of-fit ($R$^2^) and $p$-values are given in inset.
+Goodness-of-fit ($R$^2^) and significance levels are given in inset.
 
 \
-Figure \@ref(fig:erd-iso): Modeled Effective Rooting Depth (ERD; mean $\pm$ 1SE; m) versus dry-season stable isotopic concentration $\mathrm{\delta}$^2^H~xylem~ (mean $\pm$ 1SE; ‰) for `r length(unique(erd.iso$sp))` canopy species from Barro Colorado Island.
-Data from [@Meinzer:1999].
+Figure \@ref(fig:erd-iso): Modeled effective rooting depth (ERD; mean $\pm$ 1SE; m) versus dry-season stable isotopic concentration $\mathrm{\delta}$^2^H~xylem~ (mean $\pm$ 1SE; ‰) for `r length(unique(erd.iso$sp))` canopy species from Barro Colorado Island.
+$\mathrm{\delta}$^2^H~xylem~ data are from @Meinzer:1999.
 
 \
-Figure \@ref(fig:sp-erd): Modeled Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) for `r  length(erd.sp)` large, canopy species of Barro Colorado Island.
-Species ERD is a median of the species' ERDs (depths) obtained from the 100 best-fit hydrological realizations, which are vertically resolved into exponentially increasing depths.
-Hence, the non-continuous nature of the ERD axis.
+Figure \@ref(fig:sp-erd): Modeled effective rooting depth (ERD; mean $\pm$ 1SE in m) for `r  length(erd.sp)` large, canopy species of Barro Colorado Island.
+A species ERD is defined as the median, across 100 hydrological realisations, of soil layer depth $z$ at which soil water dynamics ($\mathrm{\Psi}_{\textrm{soil,} z}$) best explained observed dynamics of species growth for each realisation.
+As soil layers in the hydrological model are discretely resolved into exponentially increasing depths, so is the ERD axis.
+Deeper depths are shown with darker blue colors.
 
 \
-Figure \@ref(fig:erd-hyd): Pairwise correlations among modelled Effective Rooting Depth (ERD) and hydraulic properties for seven canopy species found on Barro Colorado Island (Panama).
+Figure \@ref(fig:erd-hyd): Pairwise correlations among modeled effective rooting depth (ERD) and hydraulic properties for seven canopy species found on Barro Colorado Island (Panama).
 Spearman's $r$ with red shades for positive $r$ values and blue shades for negative values.
-$R$^2^ and $p$-values for linear model fits are given in panel insets.
+$R$^2^ and significance levels for linear model fits are given in panel insets.
 See Fig.
@@
 \
-Figure \@ref(fig:erd-mort-census-evg): Mortality rate (mean $\pm$ 1SE in % yr^-1^) versus Modeled Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) for `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))` evergreen, canopy species found on Barro Colorado Island over seven census intervals (1981-2015).
-$R$^2^ and $p$-values for linear model fits are given in panel insets.
-Model fits (blue lines) with SE (gray area) are shown for significant fits at $\alpha$ = 0.1.
+Figure \@ref(fig:erd-mort-census-evg): Mortality rate (mean $\pm$ 1SE in % yr^-1^) versus modeled effective rooting depth (ERD; mean $\pm$ 1SE in m) for `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))` evergreen, canopy species found on Barro Colorado Island over seven census intervals (1981-2015).
+$R$^2^ and significance levels for linear model fits are given in panel insets.
+Model fits (blue lines) with SE (gray area) are only shown for periods with significant fits at $\alpha$ = 0.1, identified by census-intervals with a superscript *.
 
 \
-Figure \@ref(fig:erd-time): Modeled species Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) versus time spent below critical hydraulic threshold--% days for which $\mathrm{\Psi}_{\textrm{soil},z}$ was below $\mathrm{\Psi}$~crit~, defined as $\mathrm{\Psi}$~50,leaf~, over each of the five census intervals (1990-2015), where $z$ is the soil depth matching species ERD--for `r length(erd.mrate.sp)` canopy tree species on Barro Colorado Island.
-Each circle represents a species.
-Circles are colored and slightly jittered horizontally for clarity.
-
-# Tables
-
-```{r hypo, results = 'asis', echo = FALSE}
-knitr::kable(hypo.table, caption = 'Hypotheses for association between Effective Rooting Depth (ERD) and hydraulic traits.', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(hypo.table))))
-```
-
-#
-```{r symbols, results = 'asis', echo = FALSE}
-knitr::kable(symbols.table, caption = 'The symbol, definition and units of key traits used in the simulations and analyses.', booktabs = TRUE, escape = FALSE)
-```
+Figure \@ref(fig:erd-time): Modeled effective rooting depth (ERD; horizontal-axis) versus time spent beyond critical hydraulic threshold (vertical axis) by census interval (colored bars).
+Each bar represents the average time species of the same ERD spent beyond species-specific critical hydraulic thresholds in a given interval, that is, the proportion of days for which $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ was more negative than $\mathrm{\Psi}$~crit~, defined as $\mathrm{\Psi}$~50,leaf~, and where $z$ is the soil depth matching species ERD.
+Standard error of the mean are shown over each bar.
+Data includes all the `r length(erd.mrate.sp)` canopy tree species on Barro Colorado Island that were included in the mortality analyses.
+Trends were similar for evergreen speceis alone.
 
@@
 
-\newpage
+# Tables
 
-# References
+```{r hypo, results = 'asis', echo = FALSE}
+knitr::kable(hypo.table, caption = 'Hypotheses for association between Effective Rooting Depth (ERD) and hydraulic traits.', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(hypo.table))))
+```
+
+#
+```{r symbols, results = 'asis', echo = FALSE}
+knitr::kable(symbols.table, caption = 'The symbol, definition and units of key traits used in the simulations and analyses.', booktabs = TRUE, escape = FALSE)
+```

### 2020-11-06 Supporting_Information.Rmd
1. File name changed
1. Corrected figure captions, lower panel letters 2. Removed evergreen ERD vs. time figure

--- 2020-11-05 Supplementary_Materials.Rmd
@@
 
-#Supplementary Information
+#Supporting Information
 
@@
 
-```{r sens-sm-et, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-sm-et}Parameter sensitivity of ELM-FATES soil water content (that is, VWC, a-b) and ET (c-d) is shown for two representative sample months, as sensitivity for a particular parameter varies depending on the chosen month.
+```{r sens-sm-et, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-sm-et}Parameter sensitivity of ELM-FATES soil water content (that is, VWC, **a**-**b**) and ET (**c**-**d**) is shown for two representative sample months, as sensitivity for a particular parameter varies depending on the chosen month.
 Horizontal axes show global ranges for a given parameter.
@@
 
-```{r sens-run, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-run}Parameter sensitivity of ELM-FATES stream discharge (QRUNOFF) is shown for four sample months (a-d), as sensitivity for a particular parameter varies depending on the chosen month.
+```{r sens-run, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-run}Parameter sensitivity of ELM-FATES stream discharge (QRUNOFF) is shown for four sample months (**a**-**d**), as sensitivity for a particular parameter varies depending on the chosen month.
 Horizontal axes show global ranges for a given parameter.
@@
 $\\mathrm{\\Psi}$~leaf~ data for ", length(sp.exp.param$sp)," tree species from BCI with Eq.
-\\@ref(eq:vc-exp)--in absolute terms (a, b), as well as, in terms of loss of conductivity (%) (c, d).
-Each line represents a species, color coded by WSG (a, c), or LMA (b, d), with blue tones indicating smaller values of WSG or LMA and red tones indicating larger values.")}
+\\@ref(eq:vc-exp)--in absolute terms (**a**, **b**), as well as, in terms of loss of conductivity (%) (**c**, **d**).
+Each line represents a species, color coded by WSG (**a**, **c**), or LMA (**b**, **d**), with blue tones indicating smaller values of WSG or LMA and red tones indicating larger values.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/plot.obs.plc.jpeg"))
@@
 The number of species with an ERD estimate as well as $\\mathrm{\\delta}^2H$ for each model are shown in inset, along with the goodness-of-fit ($R$^2^) and $p$-values for a linear relationship between ERD and $\\mathrm{\\delta}^2H$.
-ERD models in panels a-e correspond to Eq.
+ERD models in panels **a**-**e** correspond to Eq.
 \\@ref(eq:M1)-\\@ref(eq:M6), respectively.")}
@@
 Pairwise correlations (upper triangle) and corresponding linear regressions (lower triangle) among modelled Effective Rooting Depth (ERD) and hydraulic properties for seven canopy species found on Barro Colorado Island (Panama).
-Spearman's correlation coefficients and significance levels are shown (upper-half; *$p$ < 0.05; .
-$p$ < 0.1)")}
+Spearman's correlation coefficients and significance levels are shown (upper-half; '*' $p$ < 0.05; '.' $p$ < 0.1; ' ' $p$ < 1)")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/erd.stem.traits_cor.chart.jpeg"))
@@
 ```{r erd-mort-census, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-census}Species mortality rate in each census interval during the study period plotted against ERD.
-$R$^2^ and $p$-values for linear model fits are given in panel insets.
-Model fits (blue lines) with SE (gray area) are shown for significant fits at $\\alpha$ = 0.1."}
+$R$^2^ and significance levels for linear model fits are given in panel insets.
+Linear model fits (blue lines) with confidance bands (gray area) are shown for significant fits at $\\alpha$ = 0.1."}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_by rdi.gr.jpeg"))

### 2020-11-06 manuscript.rmd
1. Corrected figure captions, lower panel letters 2. Removed evergreen ERD vs. time figure

@@
 Analysis of a species' time spent under significant hydrological drought (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) showed that species with shallower ERD spent a greater proportion of days beyond critical hydraulic thresholds and increasingly so over 2000-2005 and 2005-2010 (Fig.
-\@ref(fig:erd-time); S\@ref(fig:erd-time-evg)).
+\@ref(fig:erd-time)).
 For the census interval 1990-1995, species with the shallowest ERD spent a high propotion of time beyond $\mathrm{\Psi}$~crit~ (Fig.
@@
 Analysis of time spent beyond $\mathrm{\Psi}$~crit~ in the soil zone matching ERD $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ showed that this was likely due to decreasing realized risk of hydraulic failure with increasing access to water for species with deeper ERD (Fig.
-\@ref(fig:erd-time); Fig.
-S\@ref(fig:erd-time-evg)).
+\@ref(fig:erd-time)).
 We thus demonstrate, for the first time, a link between realized drought experience, hydraulic architecture and long-term mortality outcomes for co-occurring large tree species.
@@
 \
-Figure \@ref(fig:erd-hyd): Pairwise correlations among modeled effective rooting depth (ERD) and hydraulic properties for seven canopy species found on Barro Colorado Island (Panama).
-Spearman's $r$ with red shades for positive $r$ values and blue shades for negative values.
-$R$^2^ and significance levels for linear model fits are given in panel insets.
+Figure \@ref(fig:erd-hyd): Modeled effective rooting depth (ERD) versus hydraulic properties for seven canopy species found on Barro Colorado Island (Panama).
+Spearman's $r$ and significance levels.
+Linear model fits (blue lines) with confidance bands (gray area) are shown for significant fits at $\alpha$ = 0.05.
 See Fig.
@@
 $R$^2^ and significance levels for linear model fits are given in panel insets.
-Model fits (blue lines) with SE (gray area) are only shown for periods with significant fits at $\alpha$ = 0.1, identified by census-intervals with a superscript *.
+Model fits (blue lines) with confidance bands (gray area) are only shown for periods with significant fits at $\alpha$ = 0.1, identified by census-intervals with a superscript *.
 

### 2020-11-06 Supporting_Information.Rmd
Removed reference to equations in the main text

@@
 Among the `r length(sp.exp.param$sp)` species with observed data on ($K_\textrm{leaf}$ vs.
-$\mathrm{\Psi}$~leaf~) used to build the community-level models of leaf vulnerability curves (Eqs.
-\@ref(eq:b-mod)-\@ref(eq:a-mod)), two lacked data on WSG and five on LMA~lamina~.
+$\mathrm{\Psi}$~leaf~) used to build the community-level models of leaf vulnerability curves, two lacked data on WSG and five on LMA~lamina~.
 We filled three gaps in LMA using a linear relationship between LMA and LMA~disc~, the latter defined as the mean leaf mass per unit area measured for a 1.483 cm^2^ leaf disc taken to avoid veins (g m^-2^) for leaves receiving direct sunlight ($\overline{\textrm{LMA}} = `r disc.cf[1]` + `r disc.cf[2]`\textrm{LMA}_{disc}$; Adj.
@@
 ```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}Vulnerability curves fitted to observed $K_\\textrm{leaf}$ vs.
-$\\mathrm{\\Psi}$~leaf~ data for ", length(sp.exp.param$sp)," tree species from BCI with Eq.
-\\@ref(eq:vc-exp)--in absolute terms (**a**, **b**), as well as, in terms of loss of conductivity (%) (**c**, **d**).
+$\\mathrm{\\Psi}$~leaf~ data for ", length(sp.exp.param$sp)," tree species from BCI--in absolute terms (**a**, **b**), as well as, in terms of loss of conductivity (%) (**c**, **d**).
 Each line represents a species, color coded by WSG (**a**, **c**), or LMA (**b**, **d**), with blue tones indicating smaller values of WSG or LMA and red tones indicating larger values.")}

### 2020-11-06 Supporting_Information.Rmd
!. hat added to VPD

@@
 \begin{equation}
-G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*} + \textrm{VPD}_{i}^{*}) + \epsilon_{t|z}
+G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*} + \hat{\textrm{VPD}}_{i}^{*}) + \epsilon_{t|z}
 (\#eq:M2)
@@
 \begin{equation}
-G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*} \textrm{VPD}_{i}^{*}) + \epsilon_{t|z}
+G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*} \hat{\textrm{VPD}}_{i}^{*}) + \epsilon_{t|z}
 (\#eq:M3)
@@
 \begin{equation}
-G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} (K^{*}_{\textrm{leaf},i|z} + \textrm{VPD}^{*}_{i}) \textrm{LAI}^{*}_{doy|i}) + \epsilon_{t|z}
+G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} (K^{*}_{\textrm{leaf},i|z} + \hat{\textrm{VPD}}^{*}_{i}) \textrm{LAI}^{*}_{doy|i}) + \epsilon_{t|z}
 (\#eq:M5)
@@
 \begin{equation}
-G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*}\textrm{VPD}_{i}^{*}\textrm{LAI}_{doy|i}^{*}) + \epsilon_{t|z}
+G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*}\hat{\textrm{VPD}}_{i}^{*}\textrm{LAI}_{doy|i}^{*}) + \epsilon_{t|z}
 (\#eq:M6)
@@
 where n~t~ are the total number of days in census interval $t$, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is the model error term.
+$\hat{\textrm{VPD}}$ is the predicted GPP from a locally derived polynomial relationship between gross primary productivity (GPP) and VPD so as to account for the non-linear impact of VPD on growth.
+See the ERD model description in the main text for more details.
 
@@
 ```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "35%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd}Scatterplot of mean midday (11:00 am to 15:00 pm) gross primary productivity (GPP) and mean midday vapor pressure deficit (VPD) measured at the eddy flux tower in Barro Colorado Island over 2012-2017.
-Red line shows a fit from a polynomial model described in inset, and used to standardize VPD employed in equations \\@ref(eq:M2) and \\@ref(eq:M4)."}
+Red line shows the fit from a polynomial model described in inset.
+GPP predicted from this equation is used as \\hat{\\textrm{VPD}} in equations \\@ref(eq:M2), \\@ref(eq:M3), \\@ref(eq:M5) and \\@ref(eq:M6)."}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/gpp_by_env_variables/monthly_gpp.vpd.jpeg"))

### 2020-11-06 manuscript.rmd
1. Corrections while reading

@@
 We validated ERD estimates against existing empirical data of potential water access depths.
-- Across species, deeper ERD was associated with higher maximum stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates during extreme droughts over 1981-2015.
-Species with deeper ERD spent did not cross critical hydraulic thresholds indicating that trees compensate for water-stress related mortality risk through deeper water access.
+- Across species, deeper ERD was associated with higher maximum stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates during extreme droughts over 35 years (1981-2015).
+Species with deeper ERD did not cross critical hydraulic thresholds indicating that trees compensate for water-stress related mortality risk through deep-water access.
 - The role of deep-water access in mitigating mortality of hydraulically-vulnerable trees has important implications for our predictive understanding of forest dynamics under current and future climates.
@@
 As plant-available water varies with depth, trees in the same forest with different rooting depths, depending on species and size [@Meinzer:1999; @Chitra-Tarak.2018; @Brum.2019], differ in their experience during a drought, and thus in their growth and mortality responses [@Chitra-Tarak.2018].
-A key bottleneck in community-wide testing of this mechanism has been a lack of data in both trees' rooting or water-sourcing depths and plant available soil water at those depths.
+A key bottleneck in community-wide testing of this mechanism has been a lack of data in both trees' rooting or water-sourcing depths and plant-available soil water at those depths.
 A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the world’s more than 60 thousand tree species [@Fan:2017; @Beech:2017].
@@
 DNA barcoding of roots [@Jones.2011] may be used to estimate species-specific rooting depths or profiles, but DNA barcode libraries for tropical forests are still under development, and the method may be cost-prohibitive for the extent of sampling required.
-In general, community-scale data collection for rooting or water-sourcing depths is challenging for species-rich tropical forests.
+In general, community-scale data collection for rooting or water-sourcing depths is a formidable challenge in species-rich tropical forests.
 
 \
-  Characterising the essential constraints to model species-rich communities [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010; @Bartlett.2019; @Lu.2020] entails identifying the topography of plant trait trade-offs in different environments, and how these relate to demographic rates (growth, recruitment and mortality).
+  Characterizing the essential constraints to model species-rich communities [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010; @Bartlett.2019; @Lu.2020] entails identifying the topography of plant trait trade-offs in different environments, and how these relate to demographic rates (growth, recruitment and mortality).
 Significant efforts have been invested into identifying and linking universal drought indices, above-ground traits, and demographic rates.
 Such efforts have found correlations between vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013, @Venturas.2016].
-Even though deep roots may mitigate hydraulic vulnerability and mortality-risk, hydrological droughts (rather than meteorological droughts) realized through trees’ water access--the interaction between rooting depths and the actual vertical distribution of plant-available water--are not studied, especially in tropical regions.
+Even though deep roots may mitigate hydraulic vulnerability and mortality-risk, hydrological (rather than meteorological) droughts realized through trees’ water access--the interaction between rooting depths and the actual vertical distribution of plant-available water--are not studied, especially in tropical regions.
+
 \
@@
 We evaluate how rooting depth is linked to above-ground hydraulic traits and mortality rates over a 35 year period that experienced droughts of a variety of intensity, frequency and duration.
-Our hypotheses (see Table \@ref(tab:hypo)) are that (i) deep-rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem due to greater and more reliable water availability at depth, whereas shallow-rooted species are associated with less efficient but safer hydraulic traits due to more variable availability of shallow soil-water; and that, (ii) deep-rooted species have lower mortality rates during droughts due to a lower realized risk of hydraulic failure compared to shallow-rooted species.
+Our hypotheses are that (i) deep-rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem due to greater and more reliable water availability at depth, whereas shallow-rooted species are associated with less efficient but safer hydraulic traits due to more variable availability of shallow soil-water (see Table \@ref(tab:hypo)); and that, (ii) deep-rooted species have lower mortality rates during droughts due to a lower realized risk of hydraulic failure compared to shallow-rooted species.
 To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich tropical forest.
@@
 We validated ERD against existing stable hydrogen isotope ratios ($\mathrm{\delta}$^2^H~xylem~) for a subset of six tree species as independent observations.
-We evaluated whether ERD was associated with above-ground hydraulic traits sourcing the latter from a set of rare datasets for six to eight species that overlapped with our assessment of ERD.
+We evaluated whether ERD was associated with above-ground hydraulic traits sourcing the latter from a set of rare datasets for six to seven species that overlapped with our assessment of ERD.
 We tested the relationship of ERD with xylem vulnerability to embolism, branch hydraulic safety margins, leaf turgor loss point and maximum stem-specific hydraulic conductivity.
@@
 \
-  ELM-FATES simulates interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\mathrm{\Delta}W_{\textrm{can}}$, surface water $\mathrm{\Delta}W_{\textrm{sfc}}$, soil water $\mathrm{\Delta}W_{\textrm{liq},z}$ at depth $z$ and water in an unconfined aquifer $\mathrm{\Delta} W_{a}$ (omitting processes relevant to snow, wetlands or lakes).
+  ELM-FATES simulates interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\mathrm{\Delta}W_{\textrm{can}}$, surface water $\mathrm{\Delta}W_{\textrm{sfc}}$, soil water $\mathrm{\Delta}W_{\textrm{liq},z_{i}}$ at depth $z_{i}$ and water in an unconfined aquifer $\mathrm{\Delta} W_{a}$ (omitting processes relevant to snow, wetlands or lakes).
 Conservation of these terms is expressed as follows:
@@
 \begin{split}
-\mathrm{\Delta}W_\textrm{can} + \mathrm{\Delta}W_\textrm{sfc} + \sum_{i=1}^{15} \mathrm{\Delta}W_{\textrm{liq},i} + \mathrm{\Delta}W_{a} = \left(q_\textrm{rain} - E_{v} - E_{g} - q_\textrm{over} - q_\textrm{h2osfc} - q_\textrm{drai}\right)\mathrm{\Delta}t
+\mathrm{\Delta}W_\textrm{can} + \mathrm{\Delta}W_\textrm{sfc} + \sum_{i=1}^{15} \mathrm{\Delta}W_{\textrm{liq},z_{i}} + \mathrm{\Delta}W_{a} = \left(q_\textrm{rain} - E_{v} - E_{g} - q_\textrm{over} - q_\textrm{h2osfc} - q_\textrm{drai}\right)\mathrm{\Delta}t
 \end{split}
@@
 where, $q_{\textrm{rain}}$ is rainfall, $E_{v}$ is transpiration of the forest, $E_{g}$ is ground evaporation, $q_{\textrm{over}}$ is surface runoff, $q_{\textrm{h2osfc}}$ is runoff from surface water storage, $q$~drai~ is subsurface drainage (all in mm H~2~O) and $\mathrm{\Delta}t$ is the time-step (here, 1 hr).
-The soil profile is discretized into 15 exponentially distributed soil layers by depth $z$ (user-specified range), where $z$ $\in$ $Z$, and $Z$ = {0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13, 21} m.
+The soil profile is discretized into maximum 15 exponentially distributed soil layers with layer node depth $z$.
+Here, $z$ $\in$ $Z$; $Z$ = {0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13} m.
 
@@
 To parameterize ELM-FATES, we identified eleven parameters relevant for Eq.
-\@ref(eq:water-balance), determined their ranges based on literature for the study site, else for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling (@Stein.1987) of this global parameter space.
+\@ref(eq:water-balance), determined their ranges based on literature for the study site, else for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling [LHS; @Stein.1987] from this global parameter space.
 In the Conrad catchment soil hydraulic conductivity decreases with depth from 12.5 and 60 cm (@Godsey:2004; Table \@ref(tab:param)).
@@
 $\mathrm{\Psi}$~soil~ for depths of 0.15, 0.4 and 1 m [@Kupers:2019].
-We converted GWC to volumetric (VWC, $\theta$) using a bulk density value of of 0.8 g cm^-3^ measured at a point location (see below).
+We converted GWC to volumetric (VWC, $\theta$) using a measured bulk density value of 0.8 g cm^-3^ (see below).
 A @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences in water retention among depths were nominal,
@@
 \begin{equation}
-\mathrm{\theta}_{\textrm{soil}} = \Psi_e\left(\frac{\mathrm{\theta}}{\mathrm{\theta}_{s}}\right)^{-b}
+\mathrm{\theta}_{\textrm{soil}} = \mathrm{\Psi}_e\left(\frac{\mathrm{\theta}}{\mathrm{\theta}_{s}}\right)^{-b}
 (\#eq:soil-ret-curve)
 \end{equation}
-where $\mathrm{\Psi}_e$ is the air-entry matric potential, $\theta_{s}$ is the saturated volumetric water content, and $b$, an index for soil pore-size distribution, respectively.
-$\mathrm{\Psi}_e$, $\theta_{s}$ and $b$ that best-fit the data were 200 (mm H~2~O), 0.51 (cm^3^ cm^-3^) and 10 (unitless), respectively.
+where $\mathrm{\Psi}_e$ is the air-entry matric potential, $\mathrm{\theta}_{s}$ is the saturated volumetric water content, and $b$, an index for soil pore-size distribution, respectively.
+$\mathrm{\Psi}_e$, $\mathrm{\theta}_{s}$ and $b$ that best-fit the data were 200 (mm H~2~O), 0.51 (cm^3^ cm^-3^) and 10 (unitless), respectively.
 
@@
 As our key interest here was on deriving soil water availability, we ran ELM-FATES in a lower-complexity configuration: static stand structure, and with a single plant functional type (PFT) of evergreen trees.
-The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of a dry-deciduous PFT did not significantly alter results (not shown).
+The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous [@Condit:2000] and addition of a dry-deciduous PFT did not significantly alter results (not shown).
 Holding the stand structure static rather than allowing it to emerge from physiologically-mediated competition between trees removes a set of internal model feedbacks and thus the only differences between model simulations are driven by direct physiological effects of perturbed model parameters.
@@
 We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq.
-\@ref(eq:water-balance), namely, (i) evapotranspiration (ET) from the eddy flux tower ($\cong$ $E_{v} + E_{g}$; 2012-2017; see [SI](#SIData.matteo))), (ii) local stream discharge ($q_\textrm{run}$) ($\cong$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see @Paton.2019 and [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC or $\theta_{z}$ $\cong$ $W_{\textrm{liq,} z}$) from two sources: (1) a long-term (2012-2018) record of average VWC (vertical time domain reflectometry (TDR) probes from three locations near the eddy flux tower over 0-15 cm depth ($\mathrm{\theta}_\textrm{0-0.15 m}$)), and (2) plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 (depths of 0.15, 0.4 and 1 m, 1299 samples covering all soil types and habitats (see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\mathrm{\theta}_{\textrm{0-0.15 m}}$, $\mathrm{\theta}_{\textrm{0.4 m}}$ and $\mathrm{\theta}_{\textrm{1 m}}$ refer to plot-wide averages).
+\@ref(eq:water-balance), namely, (1) evapotranspiration ET from the eddy flux tower by the 50 ha plot ($\cong$ $E_{v} + E_{g}$; 2012-2017; see [SI](#SIData.matteo))), (2) local stream discharge $q_\textrm{run}$ ($\cong$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see @Paton.2019 and [SI](#SIData.steveP)) and (3) soil volumetric water content (VWC) $\mathrm{\theta}_{z}$ ($\cong$ $W_{\textrm{liq,} z}$) from two sources: (i) a long-term (2012-2018) record of VWC $\mathrm{\theta}_\textrm{0-0.15 m}$ averaged across vertical (0-15 cm depth) time domain reflectometry (TDR) probes from three locations near the eddy flux tower, and (ii) plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 at depths of 0.15, 0.4 and 1 m (1299 samples covering all soil types and habitats; see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\mathrm{\theta}_{\textrm{0-0.15 m}}$, $\mathrm{\theta}_{\textrm{0.4 m}}$ and $\mathrm{\theta}_{\textrm{1 m}}$ refer to plot-wide averages).
 
@@
 based on monthly sums for both ET and $q$~run~ and daily averages for $\mathrm{\theta}$s.
-$W_{\textrm{liq,}\overline{z}}$ is the average of daily $W_{\textrm{liq,} z}$, where $z$ $\in$ $Z$, and $Z$ = {`r paste(soil.depths[1:4], collapse=", ")`} m.
+$W_{\textrm{liq,}\overline{z}}$ is the average of daily $W_{\textrm{liq,} z}$, where $z$ $\in$ $Z$; $Z$ = {`r paste(soil.depths[1:4], collapse=", ")`} m.
 Superscript * indicates that RMSE is standardized between 0 and 1 across the 5000 simulations.
@@
 We evaluated ELM-FATES by calculating RMSE between simulated and observed long-term daily VWC for the depths of 0.1, 0.4 and 1 m (2016-2018), based on a dataset we had left-out during calibration.
-We obtained these observations from three horizontal TDR probes at the depths of 0.1, 0.4 and 1 m at a point location near the vertical probes (see [SI](#SIData.VWC)).
+We obtained these observations from three horizontal TDR probes at the depths of 0.1, 0.4 and 1 m at a location near the vertical probes (see [SI](#SIData.VWC)).
 
@@
 
-To estimate effective rooting depth, we tested alternate empirical models for species-specific 5-yearly census growth, which use effective rooting depth as an input (Eq.
+To inversely estimate effective rooting depth, we tested alternate empirical models for species-specific 5-yearly census growth, which use effective rooting depth as an input (Eq.
 \@ref(eq:M-chosen) and those in SI).
 Many factors limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019].
-Because pre-dawn $\mathrm{\Psi}$~leaf~ generally approaches $\mathrm{\Psi}$~soil~, we used the latter in leaf hydraulic vulnerability curves to predict maximum diurnal $K$~leaf~, which we assumed controls photosynthetic rate and growth (@Brodribb.2002; @Brodribb.2017; see Table \@ref(tab:symbols) for trait symbols and their definitions).
+Because pre-dawn $\mathrm{\Psi}$~leaf~ generally approaches $\mathrm{\Psi}$~soil~, we used $\mathrm{\Psi}$~soil~ in leaf hydraulic vulnerability curves to predict maximum diurnal $K$~leaf~, which we assumed controls photosynthetic rate and growth (@Brodribb.2002; @Brodribb.2017; see Table \@ref(tab:symbols) for trait symbols and their definitions).
 We assumed $K_\textrm{leaf}$ to be a hydraulic bottleneck in the plant, and thus to determine or co-determine whole plant hydraulic response to soil drying; notably, root hydraulic vulnerability may also be a strong driver, and recent studies suggest that its vulnerability would be similar to, or correlated with that of leaves (see @Cochard.2020 and references therein).
 $K_\textrm{leaf}$ in our models, being based on vulnerability curves, is dynamic and thus depends on rooting depth and rhizosphere water potential.
-The use of maximum $\mathrm{\Psi}$~leaf~ based on pre-dawn $\mathrm{\Psi}$~soil~ as a proxy for diurnal $\mathrm{\Psi}$~leaf~ and its influence on growth at diurnal time-scales (to relate to 5-year average growth values) thus does not consider diurnal dynamics of $\mathrm{\Psi}$~leaf~ and other factors that would play a role in such diurnal dynamics, such as the buffering effect of stem water storage capacitance on $\mathrm{\Psi}$~leaf~, and may thus overestimate ERD (see Discussion).
+The use of maximum diurnal $\mathrm{\Psi}$~leaf~ based on pre-dawn $\mathrm{\Psi}$~soil~ as a proxy for diurnal $\mathrm{\Psi}$~leaf~ and its influence on growth at diurnal time-scales (to relate to 5-year average growth values) thus does not consider diurnal dynamics of $\mathrm{\Psi}$~leaf~ and other factors that would play a role in such diurnal dynamics, such as the buffering effect of stem water storage capacitance on $\mathrm{\Psi}$~leaf~, and may thus overestimate ERD (see Discussion).
 VPD affects growth non-linearly, with growth increasing with VPD up-to a VPD threshold, then decreasing as leaf pores (stomata) close reducing water-uptake [@Grossiord.2020; @Yang.2019].
-To account for the non-linear impact of VPD on growth, we used predicted GPP (hereafter, $\hat{\textrm{VPD}}$) from a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see SI).
+To account for the non-linear impact of VPD on growth, we used predicted gross primary productivity (GPP; hereafter, $\hat{\textrm{VPD}}$) from a locally derived polynomial relationship between GPP and VPD (see SI).
 Apart from stomatal control, leaf-drop may further limit water-uptake and growth.
 We therefore regressed species-specific growth against multiplicative or additive effects of one or more factors calculated daily and averaged over 5-yearly censuses: $\hat{\textrm{VPD}}$, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1.
-Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting.
 Incorporation of radiation did not improve model fitting, possibly due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season at BCI.
@@
 \
- For the chosen model (see Model validation and selection below), daily average growth for species $s$, $G_{s,t|z}$, in the census interval $t$ for depth $z$ and for a given hydrological realization $h$ is described as follows:
+ For the best model (see model validation and selection below), daily average growth for species $s$, $G_{s,t|z}$, in the census interval $t$ for depth $z$ and for a given hydrological realization $h$ is described as follows:
 
 \begin{equation}
-G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*} \hat{\textrm{VPD}}_{i}^{*}) + \epsilon_{t|z}
+G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z,h}^{*} \hat{\textrm{VPD}}_{i}^{*}) + \epsilon_{t|z,h}
 (\#eq:M-chosen)
 \end{equation}
-where n~t~ is the total number of days in census interval $t$, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is the model error term.
+where n~t~ is the total number of days in census interval $t$, the superscript * indicates that the variable has been standardized between 0 and 1 ( $K_{\textrm{leaf},s,i|z,h}$ standardised within species), and $\epsilon$ is the model error term.
 See SI for equations for all the alternate models tested.
@@
 For each hydrological realization *h* we retained only one depth that had the maximum likelihood, and within those depths retained the ones that had Pearson's $r$ >= 0.71 as a valid set.
-For a species $s$, we defined effective rooting depth (ERD) as the median of these set of the retained depths, which can be described as :
+For a species $s$, we defined effective rooting depth (ERD) as the median of this valid set of depths, which can be described as :
 
@@
 \end{equation}
-where $L$ is the likelihood of $z$ given the observed growth time series, $H$ = {1, .., 100}, refer to the 100 hydrological realizations and $Z$ = {0.4, `r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m.
-We removed finely resolved soil layers <0.21 m and averaged $\mathrm{\Psi}_{\textrm{soil},z}$ for depths 0.21, 0.37 and 0.62 at 0.4 m as sufficient resolution for Model validation.
+where $L$ is the likelihood of $z$ given the observed growth time series $G_{t}$, $H$ = {1, .., 100}, refer to the 100 hydrological realizations and $Z$ = {0.4, `r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m.
+We removed finely resolved soil layers <0.21 m and averaged $\mathrm{\Psi}_{\textrm{soil},z}$ for depths 0.21, 0.37 and 0.62 (=0.4) m as sufficient resolution for model validation.
 
@@
 
-For diameter growth estimates, to minimize the effect of light variation among trees, we selected only large trees ($\ge$ 30 cm dbh, the diameter at breast height) in the 50 ha plot and also species whose maximum height was $\ge$ 30 m (hereafter, canopy species) and thus are likely to be fully exposed to the sun.
-Calculating individual tree growth rates across six 5-yearly censuses (1990-2015), removing outliers, accounting for temporal changes in growth rates due to increasing tree size (see @Condit.2019a; @Condit.2019b and SI), we obtained species-level median time series of growth rates ($G_{t}$, cm yr^-1^) used in the ERD model for only those species ($n$ = `r length(erd.sp)`) with complete records on at least three trees (median $n$ = `r median(g.n$n)`, max $n$ = `r max(g.n$n)` trees per species).
+For diameter growth estimates, to minimize the effect of light variation among trees, we selected only large trees ($\ge$ 30 cm dbh) in the 50 ha plot and also species whose maximum height was $\ge$ 30 m (hereafter, canopy species) and thus are likely to be fully exposed to the sun.
+Calculating individual tree growth rates across six 5-yearly censuses (1990-2015), removing outliers, accounting for temporal changes in growth rates due to increasing tree size (see @Condit.2019a; @Condit.2019b and SI), we obtained species-level median time series of growth rates ($G_{t}$, cm yr^-1^) used in the ERD model, for only those species ($n$ = `r length(erd.sp)`) with complete records for at least three trees (median $n$ = `r median(g.n$n)`, max $n$ = `r max(g.n$n)` trees per species).
 
@@
 where $A$ and $B$ are fitted species-specific parameters, $\epsilon_{s}$ an error term.
-These 21 species included `r sp.n.data` of the `r length(erd.sp)` species selected for ERD estimation.
+These 21 species included eight of the `r length(erd.sp)` species selected for ERD estimation.
 For the rest `r sp.n.model` species, we obtained vulnerability curves using trait-based proxies.
@@
 \begin{split}
-B_{s} = `r bcf[1]` `r bcf[2]`\textrm{WSG}_{s} + `r bcf[3]`\textrm{WSG}_{s}^2 \\
-&`r bcf[4]`\textrm{WSG}_{s}\textrm{LMA}_{s} + `r bcf[5]`\textrm{LMA}_{s} `r bcf[6]`\textrm{LMA}_{s}^2 + \epsilon_{s,b}
+B_{s} = `r bcf[1]` `r bcf[2]`\textrm{WSG}_{s} + `r bcf[3]`\textrm{WSG}_{s}^2 `r bcf[4]`\textrm{WSG}_{s}\textrm{LMA}_{s} \\
+& + `r bcf[5]`\textrm{LMA}_{s} `r bcf[6]`\textrm{LMA}_{s}^2 + \epsilon_{s,b}
 (\#eq:b-mod)
@@
 \@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r sp.n.model` ERD species without direct data and estimated leaf hydraulic vulnerability curves using Eq.
-\ref(eq:vc-exp).
-For all the `r length(erd.sp)` ERD species, we estimated species $\mathrm{\Psi}$~50,leaf~ based on their vulnerability curves for definition).
-
-\
-We obtained species-specific parameters $A$ & $B$ for Eq.
-\@ref(eq:k-soil) either from data when available, or a model based on LMA or WSG as described in the earlier section, and used K standardized between 0 and 1 within species (K*) in Eq.
-\@ref(eq:M-chosen).
+\@ref(eq:vc-exp).
+We thus obtained parameters $A$ and $B$ for all the `r length(erd.sp)` ERD species for use in Eq.
+\@ref(eq:b-mod).
+We also estimated species $\mathrm{\Psi}$~50,leaf~ from their vulnerability curves.
 
@@
 In some of the alternative models for effective rooting depth, we explored the effect of seasonality in leaf area index (LAI) on growth.
-We assumed species-specific mean seasonal curves for LAI (normalized between 0 and 1; unitless), informed by a combination of long-term records for leaf-fall [@Wright.1990aeo] and leaf-lifetime (@Osnas.2018; see SI).
+We assumed species-specific mean seasonal curves for LAI (standardised between 0 and 1; unitless), informed by a combination of long-term records for leaf-fall [@Wright.1990aeo] and leaf-lifetime (@Osnas.2018; see SI).
 
@@
 S\@ref(fig:iso-data)).
-Analysing $\mathrm{\delta}$^2^H~xylem~ for 12 tree species @Meinzer:1999 found that four species sourced water between 1 m and ground water, two other at least 0.8 m deep, one between 0 to 0.2 m, and five in the 0.2 to 0.7 m depth range over which the relationship between soil depth and $\mathrm{\delta}$^2^H~soil~ was variable.
+Analyzing $\mathrm{\delta}$^2^H~xylem~ for 12 tree species @Meinzer:1999 found that four species sourced water between 1 m and ground water, two other at least 0.8 m deep, one between 0 to 0.2 m, and five in the 0.2 to 0.7 m depth range over which the relationship between soil depth and $\mathrm{\delta}$^2^H~soil~ was variable.
 
@@
 \@ref(eq:M-chosen) and Eqs.
-in SI), modelled species ERD was regressed against species $\mathrm{\delta}$^2^H~xylem~ (six species).
+in SI), modeled species ERD was regressed against species $\mathrm{\delta}$^2^H~xylem~ (six species).
 
@@
 \@ref(fig:erd-time)).
-For the census interval 1990-1995, species with the shallowest ERD spent a high propotion of time beyond $\mathrm{\Psi}$~crit~ (Fig.
+For the census interval 1990-1995, species with the shallowest ERD spent a high proportion of time beyond $\mathrm{\Psi}$~crit~ (Fig.
 \@ref(fig:erd-time)), but during the dry season (; Fig.
@@
 \
-Our result--of lower mortality in deep-rooted trees--contrasts with the inverse model finding of @Chitra-Tarak.2018 (hereafter, CT2018) in which deeper ERD species in a south-asian seasonally tropical dry forest had higher mortality in a rare, prolonged drought.
+Our result--of lower mortality in deep-rooted trees--contrasts with the inverse model finding of @Chitra-Tarak.2018 (hereafter, CT2018) in which deeper ERD species in a South-Asian seasonally tropical dry forest had higher mortality in a rare, prolonged drought.
 The hydrological model in CT2018 revealed that the multi-year drought exhausted the deep soil and even bedrock water availability (also see, @Goulden.2019 and @Ivanov:2012).
@@
 
-ELM-FATES source code, simulation outputs for best-fit parameter ensemble membersm and all the R scripts to reproduce the manuscript are available [@Chitra-Tarak.2020].
+ELM-FATES source code, simulation outputs for best-fit parameter ensemble members and all the R scripts to reproduce the manuscript are available [@Chitra-Tarak.2020].
 Data sources for all other datasets used are provided throughout the manuscript.
@@
 Figure \@ref(fig:sp-erd): Modeled effective rooting depth (ERD; mean $\pm$ 1SE in m) for `r  length(erd.sp)` large, canopy species of Barro Colorado Island.
-A species ERD is defined as the median, across 100 hydrological realisations, of soil layer depth $z$ at which soil water dynamics ($\mathrm{\Psi}_{\textrm{soil,} z}$) best explained observed dynamics of species growth for each realisation.
+A species ERD is defined as the median, across 100 hydrological realizations, of soil layer depth $z$ at which soil water dynamics ($\mathrm{\Psi}_{\textrm{soil,} z}$) best explained observed dynamics of species growth for each realization.
 As soil layers in the hydrological model are discretely resolved into exponentially increasing depths, so is the ERD axis.
@@
 Spearman's $r$ and significance levels.
-Linear model fits (blue lines) with confidance bands (gray area) are shown for significant fits at $\alpha$ = 0.05.
+Linear model fits (blue lines) with confidence bands (gray area) are shown for significant fits at $\alpha$ = 0.05.
 See Fig.
@@
 $R$^2^ and significance levels for linear model fits are given in panel insets.
-Model fits (blue lines) with confidance bands (gray area) are only shown for periods with significant fits at $\alpha$ = 0.1, identified by census-intervals with a superscript *.
+Model fits (blue lines) with confidence bands (gray area) are only shown for periods with significant fits at $\alpha$ = 0.1, identified by census-intervals with a superscript *.
 
@@
 Data includes all the `r length(erd.mrate.sp)` canopy tree species on Barro Colorado Island that were included in the mortality analyses.
-Trends were similar for evergreen speceis alone.
+Trends were similar for evergreen species alone.
 

### 2020-11-06 Supporting_Information.Rmd
1. Added space between para
1. minor edit

@@
 The ANN was implemented using the Neural Network Toolbox in Matlab 2014a.
+
 \
-The associated data can be requested from https://ameriflux.lbl.gov.
+Data can be requested from https://ameriflux.lbl.gov.
 
@@
 Stream stage is recorded at 5-minute intervals.
+
 \
@@
 Large storms are sometimes preceded by small, initial peaks generated by the rain falling on the steeper slopes close to the weir.
+
 \
@@
 S\@ref(fig:leaf-cover)).
+
 \
 For five species, an estimate of leaf lifetime was available from direct observations [@Osnas.2018].
-For the rest erd.sp.wo.ll species, leaf lifetime was predicted from LMA ($\overline{lifetime} = `r ll.lma.cf[1]` + `r ll.lma.cf[2]`LMA$; $R$^2^ = `r ll.lma.r2`; $p$ `r ll.lma.p`, $n$ = `r ll.lma.n`).
+For the rest `r erd.sp.wo.ll` species, leaf lifetime was predicted from LMA ($\overline{lifetime} = `r ll.lma.cf[1]` + `r ll.lma.cf[2]`LMA$; $R$^2^ = `r ll.lma.r2`; $p$ `r ll.lma.p`, $n$ = `r ll.lma.n`).
 

### 2020-11-06 manuscript.rmd
1. Updated ERD model description and eq with correct z selection process 2. Tightened text elsewhere, formatted

@@
 Summary | 197  
-Main Text  | 6498  
-Introduction | 828  
-Methods | 3283  
-Results | 883  
-Discussion | 1504  
+Main Text  | 6460  
+Introduction | 827  
+Methods | 3246  
+Results | 885  
+Discussion | 1502  
 Keywords | 6
@@
 The soil profile is discretized into maximum 15 exponentially distributed soil layers with layer node depth $z$.
-Here, $z$ $\in$ $Z$; $Z$ = {0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13} m.
+Here, $z$ $\in$ $Z$; $Z$ = (0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13) m.
 
@@
 We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq.
-\@ref(eq:water-balance), namely, (1) evapotranspiration ET from the eddy flux tower by the 50 ha plot ($\cong$ $E_{v} + E_{g}$; 2012-2017; see [SI](#SIData.matteo))), (2) local stream discharge $q_\textrm{run}$ ($\cong$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see @Paton.2019 and [SI](#SIData.steveP)) and (3) soil volumetric water content (VWC) $\mathrm{\theta}_{z}$ ($\cong$ $W_{\textrm{liq,} z}$) from two sources: (i) a long-term (2012-2018) record of VWC $\mathrm{\theta}_\textrm{0-0.15 m}$ averaged across vertical (0-15 cm depth) time domain reflectometry (TDR) probes from three locations near the eddy flux tower, and (ii) plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 at depths of 0.15, 0.4 and 1 m (1299 samples covering all soil types and habitats; see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\mathrm{\theta}_{\textrm{0-0.15 m}}$, $\mathrm{\theta}_{\textrm{0.4 m}}$ and $\mathrm{\theta}_{\textrm{1 m}}$ refer to plot-wide averages).
+\@ref(eq:water-balance), namely, (1) evapotranspiration ET from the eddy flux tower by the 50 ha plot ($\backsimeq$ $E_{v} + E_{g}$ in Eq.
+\@ref(eq:water-balance); 2012-2017; see [SI](#SIData.matteo))), (2) local stream discharge $q_\textrm{run}$ ($\backsimeq$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see @Paton.2019 and [SI](#SIData.steveP)) and (3) soil volumetric water content (VWC) $\mathrm{\theta}_{z}$ ($\backsimeq$ $W_{\textrm{liq,} z}$) from two sources: (i) a long-term (2012-2018) record of VWC $\mathrm{\theta}_\textrm{0-0.15 m}$ averaged across three vertical time domain reflectometry (TDR) probes over depth 0-15 cm from three locations near the eddy flux tower, and (ii) plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 at depths of 0.15, 0.4 and 1 m (1299 samples covering all soil types and habitats; see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\mathrm{\theta}_{\textrm{0-0.15 m}}$, $\mathrm{\theta}_{\textrm{0.4 m}}$ and $\mathrm{\theta}_{\textrm{1 m}}$ refer to plot-wide averages).
 
@@
 based on monthly sums for both ET and $q$~run~ and daily averages for $\mathrm{\theta}$s.
-$W_{\textrm{liq,}\overline{z}}$ is the average of daily $W_{\textrm{liq,} z}$, where $z$ $\in$ $Z$; $Z$ = {`r paste(soil.depths[1:4], collapse=", ")`} m.
+$W_{\textrm{liq,}\overline{z}}$ is the average of daily $W_{\textrm{liq,} z}$, where $z$ $\in$ $Z$; $Z$ = (`r paste(soil.depths[1:4], collapse=", ")`) m.
 Superscript * indicates that RMSE is standardized between 0 and 1 across the 5000 simulations.
@@
 \begin{equation}
-G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z,h}^{*} \hat{\textrm{VPD}}_{i}^{*}) + \epsilon_{t|z,h}
+G_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},s,i|z,h}^{*} \hat{\textrm{VPD}}_{i}^{*}) + \epsilon_{s,t|z,h}
 (\#eq:M-chosen)
 \end{equation}
-where n~t~ is the total number of days in census interval $t$, the superscript * indicates that the variable has been standardized between 0 and 1 ( $K_{\textrm{leaf},s,i|z,h}$ standardised within species), and $\epsilon$ is the model error term.
-See SI for equations for all the alternate models tested.
+where n~t~ is the total number of days in census interval $t$, the superscript * indicates that the variable has been standardized between 0 and 1 (within species for $K_{\textrm{leaf}}$), $\beta_{0}$ and $\beta_{1}$ are model coefficients and $\epsilon$ is the model error term.
+See SI for all the alternate models tested.
 
 \
- For each species, we estimated diurnal maximum value of $K_{\textrm{leaf},s,i|z,h}$ for a species $s$ by substituting $\mathrm{\Psi}_{\textrm{soil},s,i|z,h}$in the leaf vulnerability curves defined below in Eq.
+ For each species, we estimated diurnal maximum value of $K_{\textrm{leaf},s,i|z,h}$ for a species $s$ by substituting $\mathrm{\Psi}_{\textrm{soil},s,i|z,h}$ in the leaf hydraulic vulnerability curves defined below in Eq.
 \@ref(eq:vc-exp) using species-specific parameters $A_{s}$ and $B_{s}$ as:
@@
 Using Eq.
-\@ref(eq:M-chosen), for a species $s$, we obtained likelihood based on Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $r$ for each soil depth $z$ for each of the 100 best-fit hydrological realizations.
-For each hydrological realization *h* we retained only one depth that had the maximum likelihood, and within those depths retained the ones that had Pearson's $r$ >= 0.71 as a valid set.
-For a species $s$, we defined effective rooting depth (ERD) as the median of this valid set of depths, which can be described as :
+\@ref(eq:M-chosen), for a species $s$ and hydrological realization *h*, we obtained Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $r$ for each soil depth $z$.
+For each *h*, we retained depths with the maximum likelihood (similar AIC based support, see below), and amongst them selected the best-fit depth as the one with the maximum Pearson's $r$, if $r$ >= 0.71.
+For a species $s$, we defined effective rooting depth (ERD) as the median of the best-fit depth over all *h*, which can be described as :
 
 \begin{equation}
-\textrm{ERD}_{s} = \underset{\forall \, h \, \in \, H}{\textrm{median}} \Big(z\,\Big|\, \textrm{L}_{s|z,h} = \underset{\forall \, z \,\in\,Z}{\textrm{maximum}}(\textrm{L}_{s|z,h}| r_{s|z,h} \ge 0.71\Big)\Big)
+\textrm{ERD}_{s} = \underset{\forall \, h \, \in \, H}{\textrm{med}} \Big(z\,\Big|\, \Big(r_{s|z,h} = \underset{\forall \, z \,\in\,Z}{\textrm{max}}(r_{s|z,h}| \mathrm{\Delta}_{s|z,h} \le 2\Big)\Big) \, \bigcap \, r_{s|z,h} \ge 0.71\Big)
 (\#eq:erd)
 \end{equation}
-where $L$ is the likelihood of $z$ given the observed growth time series $G_{t}$, $H$ = {1, .., 100}, refer to the 100 hydrological realizations and $Z$ = {0.4, `r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m.
+
+where $\mathrm{\Delta}_{s|z,h} = \textrm{AIC}_{s|z,h} - \underset{\forall \, z \,\in\,Z}{\textrm{min}}(\textrm{AIC}_{s|z,h})$,  $\textrm{AIC}_{s|z,h} = -2\textrm{log}(\mathcal{L}(\hat{\beta}_{s|z,h}|{G_{s}})) - 2k$, $\mathcal{L}$ is the likelihood function, $G$ the observed growth time series with $k$ observations (or, census intervals $t$).
+$H$ = (1, .., 100), the 100 best-fit hydrological realizations and $Z$ = (0.4, `r paste(soil.depths[8:length(soil.depths)], collapse=", ")`) m.
 We removed finely resolved soil layers <0.21 m and averaged $\mathrm{\Psi}_{\textrm{soil},z}$ for depths 0.21, 0.37 and 0.62 (=0.4) m as sufficient resolution for model validation.
@@
 For diameter growth estimates, to minimize the effect of light variation among trees, we selected only large trees ($\ge$ 30 cm dbh) in the 50 ha plot and also species whose maximum height was $\ge$ 30 m (hereafter, canopy species) and thus are likely to be fully exposed to the sun.
-Calculating individual tree growth rates across six 5-yearly censuses (1990-2015), removing outliers, accounting for temporal changes in growth rates due to increasing tree size (see @Condit.2019a; @Condit.2019b and SI), we obtained species-level median time series of growth rates ($G_{t}$, cm yr^-1^) used in the ERD model, for only those species ($n$ = `r length(erd.sp)`) with complete records for at least three trees (median $n$ = `r median(g.n$n)`, max $n$ = `r max(g.n$n)` trees per species).
+Calculating individual tree growth rates across six 5-yearly censuses (1990-2015; @Condit.2019a; @Condit.2019b), removing outliers, accounting for the effect of increasing size on growth by obtaining residuals from a dbh model of growth (see SI), we estimated the growth time series $G_{s}$ (cm yr^-1^) for species $s$ as the median of the standardized (centered and scaled) dbh model residuals to use in the ERD model.
+$G_{s}$ was determined for only those species ($n$ = `r length(erd.sp)`) with complete records for at least three trees (median $n$ = `r median(g.n$n)`, max $n$ = `r max(g.n$n)` trees per species).
 

### 2020-11-06 Supporting_Information.Rmd
Updated growth data description to match with the main text

@@
 We selected trees for which the height of the diameter measurements were unchanged across all censuses (1990, 1995, 2000, 2005, 2010, 2015).
-For each individual tree ($i$), we calculated absolute growth rates ($G_{i,t}$, cm yr^-1^) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses prorated by the inter-census interval based on the exact date of measurements.
+For each individual tree ($i$), we calculated absolute growth rates ($\grave{G}_{i,t}$, cm yr^-1^) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses prorated by the inter-census interval based on the exact date of measurements.
 We eliminated positive outliers as any growth rates amounting to >75 mm yr^-1^, as this was the highest rate observed with confidence for the fastest growing species [@Condit.2017].
@@
 Any stem whose later dbh measurement fell <4$\sigma_{e}$ the earlier measurement was removed [@Condit.2017].
-As tree growth rates vary with size [@Muller-Landau:2006], to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($\textrm{Resid}_{i,t} = G_{i,t} - f(d_{i,t})$).
+As tree growth rates vary with size [@Muller-Landau:2006], to keep growth rates comparable across census intervals, we obtained residuals $\textrm{R}_{i,t}$ from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($\textrm{R}_{i,t} = \grave{G}_{i,t} - f(d_{i,t})$).
 The diameter model of growth, $f(d_{i, t})$, a B-spline based polynomial equation with five degrees of freedom, was fitted to growth rates from all trees pooled together as there was insufficient species-specific data for large trees ($R$^2 = 0.27; $p$-value < 0.01).
-We retained only those trees that had $\textrm{Resid}_{i,t}$ estimates for all census intervals (total trees = `r sum(g.n$n)`) and standardized (centered and scaled) individual tree time series of $\textrm{Resid}_{i,t}$.
-We then obtained `r length(erd.sp)` species-level time-series of medians of $\textrm{Resid}_{i,t}$ ($G_{t}$).
-These were obtained for only those species with complete records on at least three trees (median $n$ = `r median(g.n$n)`, max $n$ = `r max(g.n$n)` trees per species).
+We retained only those trees that had $\textrm{R}_{i,t}$ estimates for all census intervals (total trees = `r sum(g.n$n)`) and standardized (centered and scaled) individual tree time series of $\textrm{Resid}_{i,t}$.
+We then obtained growth time series $G_{s}$ for each species $s$ as the time-series of medians of $\textrm{R}_{i,t}$ for use in the ERD model.
+These were obtained for only those species ($n$ = `r length(erd.sp)`) with complete records on at least three trees (median $n$ = `r median(g.n$n)`, max $n$ = `r max(g.n$n)` trees per species).
 

### 2020-11-06 manuscript.rmd
1. edited discussion

@@
 Summary | 197  
-Main Text  | 6460  
-Introduction | 827  
-Methods | 3246  
-Results | 885  
+Main Text  | 6487  
+Introduction | 832  
+Methods | 3277  
+Results | 876  
 Discussion | 1502  
@@
 Such efforts have found correlations between vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013, @Venturas.2016].
-Even though deep roots may mitigate hydraulic vulnerability and mortality-risk, hydrological (rather than meteorological) droughts realized through trees’ water access--the interaction between rooting depths and the actual vertical distribution of plant-available water--are not studied, especially in tropical regions.
+Even though deep roots may mitigate hydraulic vulnerability and mortality-risk, hydraulic risk realized through hydrological (rather than meteorological) droughts--the outcome of the interaction between rooting depths, critical hydraulic thresholds and the actual vertical distribution of plant-available water--are not studied, especially in tropical regions.
 
@@
 We tested the relationship of ERD with xylem vulnerability to embolism, branch hydraulic safety margins, leaf turgor loss point and maximum stem-specific hydraulic conductivity.
-Finally, we analysed species-specific large tree mortality dynamics over 35 years and estimated the time a species’ spent beyond its critical hydraulic threshold to test whether ERD explained risk of drought-induced mortality.
+Finally, to test whether ERD explained risk of drought-induced mortality, we analysed species-specific large tree mortality dynamics over 35 years and estimated species-level realized hydraulic risk as the time a species’ spent beyond its critical hydraulic threshold.
 
@@
 For diameter growth estimates, to minimize the effect of light variation among trees, we selected only large trees ($\ge$ 30 cm dbh) in the 50 ha plot and also species whose maximum height was $\ge$ 30 m (hereafter, canopy species) and thus are likely to be fully exposed to the sun.
-Calculating individual tree growth rates across six 5-yearly censuses (1990-2015; @Condit.2019a; @Condit.2019b), removing outliers, accounting for the effect of increasing size on growth by obtaining residuals from a dbh model of growth (see SI), we estimated the growth time series $G_{s}$ (cm yr^-1^) for species $s$ as the median of the standardized (centered and scaled) dbh model residuals to use in the ERD model.
-$G_{s}$ was determined for only those species ($n$ = `r length(erd.sp)`) with complete records for at least three trees (median $n$ = `r median(g.n$n)`, max $n$ = `r max(g.n$n)` trees per species).
+Calculating individual tree growth rates across six 5-yearly censuses (1990-2015; @Condit.2019a; @Condit.2019b), removing outliers, obtaining residuals from a dbh model of growth to account for the size effect on growth (see SI), we estimated species $s$ growth time series $G_{s}$ (cm yr^-1^) as the median of standardized dbh model residuals, for only those species ($n$ = `r length(erd.sp)`) with complete records for at least three trees (median `r median(g.n$n)`, max `r max(g.n$n)` trees per species).
 
@@
 Given a vertical gradient of $\mathrm{\delta}$^2^H in soil and groundwater $\mathrm{\delta}$^2^H~xylem~ provides an index of rooting depth [@Dawson.1991].
-We leveraged $\mathrm{\delta}$^2^H~xylem~ from BCI [@Meinzer:2001] for the dry-season of March 1997 as this period showed largest seasonal divergence in $\mathrm{\delta}$^2^H~xylem~ among species and in soil and groundwater $\mathrm{\delta}$^2^H at natural abundance level (Fig.
+We leveraged $\mathrm{\delta}$^2^H~xylem~ from BCI [@Meinzer:2001] for the dry-season of March 1997 as this period showed largest seasonal divergence in $\mathrm{\delta}$^2^H~xylem~ among species and vertically in soil and groundwater $\mathrm{\delta}$^2^H at natural abundance level (Fig.
 S\@ref(fig:iso-data)).
-Analyzing $\mathrm{\delta}$^2^H~xylem~ for 12 tree species @Meinzer:1999 found that four species sourced water between 1 m and ground water, two other at least 0.8 m deep, one between 0 to 0.2 m, and five in the 0.2 to 0.7 m depth range over which the relationship between soil depth and $\mathrm{\delta}$^2^H~soil~ was variable.
+Analyzing $\mathrm{\delta}$^2^H~xylem~ for 12 tree species @Meinzer:1999 found that four species sourced water deeper than 1 m, two other at least 0.8 m deep, one between 0 to 0.2 m, and five in the 0.2 to 0.7 m depth range over which the relationship between soil depth and $\mathrm{\delta}$^2^H~soil~ was variable.
 
 \
-We removed those species that would be leafless in March 1997 because xylem sap $\mathrm{\delta}$^2^H for these species may not be linked with water sourced at the time of measurement and thus may be compared to species that had leaves.
+We removed those species that would be leafless in March 1997 because $\mathrm{\delta}$^2^H~xylem~ for these species may not be linked with water sourced at the time of measurement and thus not comparable to species that had leaves.
 We also removed one species whose $\mathrm{\delta}$^2^H~xylem~ was in the uncertain depth range.
 For each model of ERD (Eq.
-\@ref(eq:M-chosen) and Eqs.
-in SI), modeled species ERD was regressed against species $\mathrm{\delta}$^2^H~xylem~ (six species).
+\@ref(eq:M-chosen) and alternatives in SI), modeled species ERD was regressed against species $\mathrm{\delta}$^2^H~xylem~ for a maximum of six species.
 
@@
 
-We evaluated whether ERD was associated with above-ground hydraulic traits sourcing the latter from a set of rare datasets [@Wolfe.2019, @Wolfe.2020] for six to seven species that overlapped with our assessment of ERD.
-We regressed species ERD against maximum stem hydraulic conductivity $K_\textrm{max,stem}$ (*n*=7), leaf turgor loss point $\mathrm{\Psi}$~tlp~ (*n*=7), vulnerability to embolism from cavitation $\mathrm{\Psi}$~88,stem~ measured in terms of pressure at which 88% of $K_\textrm{max,stem}$ is lost (*n*=7), and hydraulic safety margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~ (*n*=6).
+We evaluated whether ERD was associated with above-ground hydraulic traits sourcing the latter from a set of rare datasets [@Wolfe.2019; @Wolfe.2020] for six to seven species that overlapped with our assessment of ERD.
+We regressed species ERD against maximum stem hydraulic conductivity $K_\textrm{max,stem}$ (*n*=7), leaf turgor loss point $\mathrm{\Psi}$~tlp~ (*n*=7), vulnerability to embolism from cavitation $\mathrm{\Psi}$~88,stem~ measured in terms of pressure at which 88% of $K_\textrm{max,stem}$ is lost (*n*=7), and hydraulic safety margins $\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~ (*n*=6).
 See [SI](#SIData.brett) and Table S\@ref(tab:hyd-table) for data collection and estimation of these variables.
@@
 
-To test whether ERD plays a role in mitigating mortality risk, we calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$ 30 m) and average abundance of at least 20 trees ($n$ = `r length(erd.mrate.sp)`) in the 50ha plot [@Condit.2019a].
-For each of the seven census intervals $t$ in the 35-yr record (1981, 1985, 1990, 1995, 2000, 2005, 2010, 2015), mortality rate, $\textrm{M}_{t}$ (% yr^-1^) for species $s$ for two successive censuses, $c_{1}$ and $c_{2}$, was calculated as $\textrm{M}_{s,t} = \frac{D_{s,c_{1}}}{N_{s,c_{2}}}\times\frac{100}{\textrm{d}}$, where $N_{s,c_{1}}$ and $D_{s,c_{2}}$ are the total number of trees of species $s$ present in $c_{1}$ and dead in $c_{2}$, respectively, in the 50 ha plot and $d$ is the duration based on mean dates of $c_{1}$ and $c_{2}$.
+To test whether ERD plays a role in mitigating mortality risk, we calculated mortality rates for large trees (here, $\ge$ 10 cm dbh) in the 50ha plot for species with ERD estimates (all canopy species with maximum height $\ge$ 30 m) as well as average abundance of at least 20 trees in the plot ($n$ = `r length(erd.mrate.sp)`) [@Condit.2019a].
+For each of the seven census intervals $t$ in the 35-yr record (1981, 1985, 1990, 1995, 2000, 2005, 2010, 2015), mortality rate, $\textrm{M}_{t}$ (% yr^-1^) for species $s$ for two successive censuses, $c_{1}$ and $c_{2}$, was calculated as $\textrm{M}_{s,t} = \frac{D_{s,c_{2}}}{N_{s,c_{1}}}\times\frac{100}{\textrm{d}}$, where $N_{s,c_{1}}$ and $D_{s,c_{2}}$ are the total number of large trees of species $s$ present in the 50 ha plot in $c_{1}$ and dead in $c_{2}$, respectively, and $d$ is the duration based on mean dates of $c_{1}$ and $c_{2}$.
 
 \
-We regressed species ERD against species $\textrm{M}_{s,t}$ for each census interval $t$ with ($n$ = `r nrow(mrate.depth.mean)`) and without deciduous species ($n$ = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`).
-To assess whether deeper ERD species escaped the experience of significant hydrological drought, we calculated for each species the proportion of days in the simulation period (1990-2015) during which soil water potential in the soil layer matching species ERD ($\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$) were more negative than species $\mathrm{\Psi}$~crit~.
+We regressed species ERD against species $\textrm{M}_{s,t}$ for each census interval $t$ for the whole pool of ($n$ = `r nrow(mrate.depth.mean)`) species with $\textrm{M}_{s,t}$ estimates and for evergreen species alone ($n$ = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`).
+To assess whether deeper ERD species escaped the experience of significant hydrological drought, we calculated realized hydrulic risk for each species as the proportion of days in each census interval $t$ of the $\mathrm{\Psi}_\textrm{soil}$ simulation period (1990-2015) during which soil water potential in the soil layer matching species ERD, $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$, were more negative than species $\mathrm{\Psi}$~crit~.
 $\mathrm{\Psi}$~crit~ was defined as $\mathrm{\Psi}$~50,leaf~.
@@
 The 100 ELM-FATES ensemble members with the best fits to observed soil moisture dynamics captured soil moisture seasonality at multiple depths (Fig.
-\@ref(fig:hydro-calib)a, b), including the out of sample observations.
+\@ref(fig:hydro-calib)a, b), including the out-of-sample observations.
 These simulations also captured the dynamics in stream discharge (Fig.
@@
 \@ref(fig:hydro-calib)d), slightly underestimating peak discharge in wet years and slightly overestimating peak ET.
-The reduction in parameter range (Table S\@ref(tab:param)) in the best-fit ensembles compared to the tested global ranges showed that the model calibration was sensitive to the Ball-Berry stomatal slope parameter, the ELM root distribution parameter (that regulates the depth of the rooting profile), soil hydraulic conductivity profile, especially at depth, and the adjustment factor that varies soil hydraulic conductivity to account for macroporosity and direct flow paths (Fig.
+The reduction in parameter range (Table S\@ref(tab:param)) in the best-fit ensembles compared to the tested global ranges showed that the model calibration was sensitive to the Ball-Berry stomatal slope parameter, the ELM root distribution parameter that regulates the depth of the rooting profile, soil hydraulic conductivity profile especially at depth, and the adjustment factor that modifies soil hydraulic conductivity to account for macroporosity and direct flow paths (Fig.
 S\@ref(fig:sens-sm-et), S\@ref(fig:sens-run)).
@@
 \
-Soil Water Potential dynamics ($\mathrm{\Psi}_{\textrm{soil},z}$) by depth (from 0.01 through 13 m) obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of extreme years, drought intensity at depth, drought seasonality and duration (Fig.
+Soil water potential dynamics $\mathrm{\Psi}_{\textrm{soil},z}$ by depth z from 0.01 through 13 m obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of extreme years, drought intensity at depth, drought seasonality and duration (Fig.
 \@ref(fig:swp-droughts)).
@@
 S\@ref(fig:swp-droughts)).
-Across the simulation period (1990-2015) $\mathrm{\Psi}_{\textrm{soil},z}$ remained above -0.5 MPa from depths 1.7 to 13 m in the simulations (Fig.
+Across the simulation period 1990-2015, simulated $\mathrm{\Psi}_{\textrm{soil},z}$ remained above -0.5 MPa from depths 1.7 to 13 m (Fig.
 S\@ref(fig:swp-all-depths)).
@@
 
-Parameters of leaf vulnerability curves were predictable from WSG and LMA.
-WSG and parameter $B$ explained a large proportion of variance in parameter $A$ (Eq.
-\@ref(eq:a-mod); *Adj.
-R*^2^ = `r afit.r2`; $p$ `r  afit.p`; Fig.
+Parameters of leaf vulnerability curves were predictable from WSG and LMA (Fig.
 \@ref(fig:ab-fit)).
@@
 \@ref(eq:b-mod); Adj.
-$R$^2^ = `r bfit.r2`; $p$ `r  bfit.p`; also see Fig.
+$R$^2^ = `r bfit.r2`; $p$ `r  bfit.p`).
+WSG and parameter $B$ explained a large proportion of variance in parameter $A$ (Eq.
+\@ref(eq:a-mod); *Adj.
+R*^2^ = `r afit.r2`; $p$ `r  afit.p`; also see Fig.
 S\@ref(fig:obs-plc-curves)).
-This predictive power allowed us to estimate leaf vulnerability curve parameters for `r sp.n.model` species that lacked direct observations (out of `r length(erd.sp)` ERD species; Table S\@ref(tab:ab-param), Fig.
+This predictive power allowed us to estimate leaf vulnerability curves used in the ERD models for `r sp.n.model` out of `r length(erd.sp)` ERD species that lacked direct observations (Table S\@ref(tab:ab-param), Fig.
 S\@ref(fig:obs-comm-plc-curves)).
-These vulnerability curves were used in the ERD models.
 
@@
 
-The chosen ERD model (with a multiplicative effect of VPD and not LAI; Eq.
+The best ERD model (with a multiplicative effect of VPD and not LAI, Eq.
 \@ref(eq:M-chosen)) explained a large fraction of the variance in $\mathrm{\delta}$^2^H~xylem~ ($R$^2^ = `r lm.erd.iso.r2`; $p$ = `r lm.erd.iso.p`, $n$ = `r length(unique(erd.iso$sp))`; Fig.
 \@ref(fig:erd-iso)).
-Two of the alternative models tested with LAI had a strong relationship with $\mathrm{\delta}$^2^H~xylem~ as well, but one that was based on a smaller set of species ($n$ = 4) as fewer species passed the criteria of sufficient species-level goodness-of-fit with these models (<20 species vs.
-`r length(erd.sp)` species for the chosen model; Fig.
+Two of the alternative models with LAI had a strong relationship with $\mathrm{\delta}$^2^H~xylem~ as well, but one that was based on a smaller set of species ($n$ = 4), because fewer species passed the criteria of sufficient species-level goodness-of-fit for these models (<20 species vs.
+`r length(erd.sp)` species for the best model; Fig.
 S\@ref(fig:other-erd-mods)).
-For the rest of the analyses we therefore used ERD estimates from the less parsimonious, chosen model.
+For the rest of the analyses we therefore used ERD estimates from the less parsimonious, best model.
 
@@
 Census intervals 2000-2005 and 2005-2010 were also the periods during which extreme soil droughts distinctly occurred in the wet season, effectively extending the dry season into the wet season (Fig.
-\@ref(fig:swp-droughts).
-While our drought assessment did not extend to the model spin-up period of 1985-1990, for which ERD explained significant mortality, a prior study found that during this period BCI experienced three extreme deficit water years, following the 1983 El-Niño year [@Condit.2017].
-Analysis of a species' time spent under significant hydrological drought (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) showed that species with shallower ERD spent a greater proportion of days beyond critical hydraulic thresholds and increasingly so over 2000-2005 and 2005-2010 (Fig.
+\@ref(fig:swp-droughts)).
+While our $\mathrm{\Psi}_{\textrm{soil}}$ estimates did not extend prior to 1990, due to a lack of reliable climate forcing, a prior study found that during the period 1985-1990, for which ERD explained significant mortality, BCI experienced three extreme deficit water years, following the 1983 El-Niño year [@Condit.2017].
+Analysis of a species' realized hydraulic risk as the time spent under significant hydrological drought (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) showed that species with shallower ERD experienced greater realized hydraulic risk and increasingly so over 2000-2005 and 2005-2010 (Fig.
 \@ref(fig:erd-time)).
-For the census interval 1990-1995, species with the shallowest ERD spent a high proportion of time beyond $\mathrm{\Psi}$~crit~ (Fig.
+For the census interval 1990-1995, species with the shallowest ERD also showed greater realized hydraulic risk (Fig.
 \@ref(fig:erd-time)), but during the dry season (; Fig.
-\@ref(fig:swp-droughts), and did not have increased mortality rates (Fig.
+\@ref(fig:swp-droughts), and without elevated mortality rates (Fig.
 \@ref(fig:erd-mort-census).
-Overall, Species with ERD deeper than 1.7 m did not cross species $\mathrm{\Psi}$~crit~ for the entire 25-yr period for which soil water potential $\mathrm{\Psi}_{\textrm{soil},z}$ was modeled (1990-2015).
+Overall, species with ERD deeper than 1.7 m did not cross species $\mathrm{\Psi}$~crit~ for the entire 25-yr period for which $\mathrm{\Psi}_{\textrm{soil},z}$ was modeled (1990-2015).
 Soil depths deeper than 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and drought periods, which is above $\mathrm{\Psi}$~crit~ of the most sensitive tree species (-0.28 MPa).
@@
 
-Our novel approach for estimating ERD using 25 years of growth, species-specific leaf vulnerability curves, modeled soil moisture profiles, and observed VPD was consistent with estimates using $\mathrm{\delta}$^2^H~xylem~.
-Our analyses revealed that co-occurring large canopy tree species with deeper ERD were associated with higher hydraulic efficiency, but not safety (Fig.
+Our novel approach for estimating ERD using 25 years of tree growth, species-specific leaf vulnerability curves, modeled soil moisture profiles, and observed VPD was consistent with estimates using $\mathrm{\delta}$^2^H~xylem~.
+Our analyses revealed that co-occurring large canopy tree species with deeper ERD were associated with higher stem hydraulic efficiency, but not safety (Fig.
 \@ref(fig:erd-hyd)).
@@
 \@ref(fig:erd-time)).
-We thus demonstrate, for the first time, a link between realized drought experience, hydraulic architecture and long-term mortality outcomes for co-occurring large tree species.
+We thus demonstrate, for the first time, a link between realized drought risk, hydraulic architecture and long-term mortality outcomes for large tree species of a tropical forest.
 
@@
 
-Our study brings attention to the need for assessing drought sensitivity in terms of species realized drought exposure by accounting for the actual hydrological drought and trees' deep water access.
+Our study brings attention to the need for assessing drought sensitivity in terms of species realized drought exposure by accounting for hydrological drought and trees' rooting depths.
 We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins (Fig.
@@
 \@ref(fig:erd-mort-census-evg), S\@ref(fig:erd-mort-census), \@ref(fig:swp-droughts)).
-Deeper ERD species on average had more hydraulically efficient, but risky, suite of traits optimized for an environment of more reliable and greater water source at depth that resulted in a lower mortality risk.
+Deeper ERD species on average had more hydraulically efficient, but risky, suite of above-ground traits optimized for an environment of more reliable and greater water source at depth that resulted in a lower mortality risk.
 By contrast shallow ERD species had inefficient, but safer, xylem likely optimized for high variability in $\mathrm{\Psi}_\textrm{soil}$ in the shallow soil layers, and resulting in higher variance in mortality rates across censuses as $\mathrm{\Psi}_{\textrm{soil},z\textrm{ = ERD}}$ crossed ERD $\mathrm{\Psi}_\textrm{crit}$ (Fig.
@@
 \@ref(fig:erd-time)).
-Our results are consistent with studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016; also see @Brum.2019; @Brum.2017].
+Our results are consistent with studies that included three or four species each with known rooting or water-sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016; also see @Brum.2019; @Brum.2017].
 Globally, large trees tend to exhibit greater growth reductions, lower post-drought resilience, and greater increases in mortality relative to their understory counterparts [@Bennett:2015; @Phillips:2010].
@@
 Our result--of lower mortality in deep-rooted trees--contrasts with the inverse model finding of @Chitra-Tarak.2018 (hereafter, CT2018) in which deeper ERD species in a South-Asian seasonally tropical dry forest had higher mortality in a rare, prolonged drought.
-The hydrological model in CT2018 revealed that the multi-year drought exhausted the deep soil and even bedrock water availability (also see, @Goulden.2019 and @Ivanov:2012).
-In contrast, our hydrological modeling at BCI finds that during the study period, $\mathrm{\Psi}$~soil~ for depths deeper than 1 m did not cross $\mathrm{\Psi}$~crit~ of the most sensitive tree species (-0.28 MPa) for all the dry seasons and droughts.
-Mean annual rainfall of 1095 mm in CT2018 compared to `r round(rain.man.stats$rain)`mm at BCI, and precipitation to potential evapotranspiration ratio of nearly one in CT2018 compared to nearly two in BCI are major factors in the different mortality responses.
-At BCI deep soil layers are effectively recharged annually (Fig.
+The hydrological model in CT2018 revealed that the multi-year drought exhausted the deep soil and even bedrock water availability (also see [@Goulden.2019; @Ivanov:2012]).
+In contrast, our hydrological modeling at BCI finds that $\mathrm{\Psi}$~soil~ for depths deeper than 1 m did not cross $\mathrm{\Psi}$~crit~ of the most sensitive tree species (-0.28 MPa) for all the dry seasons and droughts during 1990-2015.
+Mean annual rainfall of 1095 mm in CT2018 compared to `r round(rain.man.stats$rain)` mm at BCI, and precipitation to potential evapotranspiration ratio of nearly one in CT2018 compared to nearly two in BCI are major factors in the different mortality responses.
+At BCI deep soil layers were recharged annually (Fig.
 S\@ref(fig:swp-all-depths), \@ref(fig:swp-droughts)), whereas in CT2018 they were not.
-The contrast between the two studies highlights the combined role of seasonal precipitation input and site-specific hydrology in modulating deep ERD mortality-risk.
+The contrast between the two studies highlights the combined role of seasonal precipitation input and site-specific hydrology in modulating mortality-risk of deep-rooted species.
 
@@
 By estimating ERD for large trees of `r length(erd.sp)` canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species-rich tropical forests.
-Our chosen ERD model is a key improvement over the only other inverse model of ERD (CT2018; $n$ = 12 species) as we employ a physiologically meaningful representation, fewer parameters and corroboration with tree and soil data for $\mathrm{\delta}$^2^H.
-We highlight that stable water isotopes for plant and below-ground sources are critical for calibration or validation of ERD models at community scale, although the interpretation of these data is still under research  [@Deurwaerder.2020; @Adams.2020; @Bowers.2020].
-As in CT2018, we also interpret our ERD estimates as revealing relative differences among species' effective rooting depths rather than in absolute terms.
-As our ERD model ignores stem water storage capacitance [@Wolfe.2017] it may have overestimated ERD especially in species for which stem water storage capacitance is important as a drought-avoidance strategy, for example, the deciduous species [@Borchert.2005jma].
+Our chosen ERD model is a key improvement over the model of CT2018 ($n$ = 12 species) as we employ a physiologically meaningful representation, fewer parameters and corroboration with tree and soil data for $\mathrm{\delta}$^2^H.
+We note that stable water isotopes for plant and below-ground sources are critical for calibration or validation of ERD models for scaling upto community, although the interpretation of isotopic data is still under research [@Deurwaerder.2020; @Adams.2020; @Bowers.2020].
+As in CT2018, we also interpret our ERD estimates as revealing relative differences among species' effective rooting depths rather than absolute depths.
+As our ERD model ignores stem water storage capacitance [@Wolfe.2017] it may have overestimated ERD especially in those species for which capacitance is important as a drought-avoidance strategy, for example, the deciduous species [@Borchert.2005jma].
 
@@
 The relationships that we identify between above- and below-ground traits, the water environment vertically below-ground and mortality rates are important for representing diversity in DGVMs, which intrinsically rely on the parameterization of contrasting life history strategies [@Scheiter:2013] and the simulation of competition between those strategies.
-We found that LMA and WSG were strong predictors of PLC curves for leaves.
+We found that LMA and WSG were strong predictors of leaf hydraulic vulnerability curves.
 This offers the promise of a greater ability to parameterize the 'hard' hydraulic traits with the abundant 'soft' trait data, thus allowing for a better representation of forest hydrodynamics.
-The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits, including hydraulic vulnerability, in order to better represent the functional diversity of tropical forests, and their trajectories into the future.
+The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits in order to better represent the functional diversity of tropical forests, and their trajectories into the future.
 
@@
 S\@ref(fig:other-erd-mods)).
-We interpret this result as that VPD and leaf hydraulic vulnerability curves may be adequate to explain inter-census differences in growth (via stomatal control), but also acknowledge that our estimates of seasonality of LAI is a tentative estimate combining leaf-fall data and the timing of leaf-gain backtracked from leaf lifetime, and omits inter-annual variation.
+We interpret this result as that VPD and leaf hydraulic vulnerability curves may be adequate to explain inter-census differences in growth (via stomatal control), but also acknowledge that our estimates of seasonality of LAI is a tentative estimate that combines leaf-fall data and the timing of leaf-gain backtracked from leaf lifetime, and omits inter-annual variation.
 Future studies may improve models of LAI seasonality and make use of data from new technologies such as drone based monitoring of LAI in species-rich tropical forests [@Park.2019].
@@
 Although data needs for parameterizing such models could be greater (e.g.
-retention curves and capacitance for roots, stems and leaves), adding a degree of parameter uncertainty in community-wide application, such models hold greater promise in improving our understanding of plant physiology [e.g.
+vulnerability curves and capacitance for roots, stems and leaves), adding a degree of parameter uncertainty in community-wide application, such models hold greater promise in improving our understanding of plant physiology [e.g.
 see @Johnson:2018fad].
@@
 To our knowledge this is the first study to test for a mechanistic link between plant-available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.
-We report here that deep-water access plays a role in alleviating mortality of hydraulically vulnerable trees.
+We report here that deep-water access plays a role in mitigating mortality of otherwise vulnerable stem hydraulics.
 This has important implications for our predictive understanding of tropical forest dynamics under current and future climate.
@@
 Hubbell and Robin B.
-Foster and sustained for many years by Richard Condit, is now managed by Rolando Perez, Suzanne Lao and Stuart Davies and under the ForestGEO program of the Smithsonian Tropical Research in Panama.
+Foster and sustained for many years by Richard Condit, is now managed by Rolando Perez, Suzanne Lao and Stuart Davies under the ForestGEO program of the Smithsonian Tropical Research in Panama.
 Numerous organizations have provided funding, principally the US National Science Foundation, and hundreds of field workers have contributed.
@@
 MTT for GPP, ET, TDR, bulk density.
-SK for manual GWC and GWC vs $\Psi_\textrm{soil}$.
+SK for manual GWC and GWC vs $\mathrm{\Psi}_\textrm{soil}$.
 NK, JZ, KAT, LS for leaf hydraulics, BW for stem hydraulics.
@@
 Observations (red lines and points) versus simulations (gray lines) from 100 best-fit ensemble-member runs of ELM-FATES are shown for daily Volumetric Water Content (VWC) by depth for horizontal time domain reflectometry (TDR) probes at three depths (0.1, 0.4 and 1 m; **a**) and for an average of three vertical TDR probes (0-0.15 m; **b**); monthly stream discharge (Discharge, **c**) and monthly evapotranspiration from eddy flux tower (ET, **d**).
-Red points in (**a**) show average, manual plot-wide observations of VWC with 95% CI (error bars).
+Red points in panel **a** show average, manual plot-wide observations of VWC with 95% CI (error bars).
 All observations (red lines and points) were in-sample, except for VWC data from the horizontal probes (red lines in **a**) which were out-of-sample.
@@
 \
-Figure \@ref(fig:swp-droughts): Occurrence of extreme hydrological droughts by census intervals (horizontal panels) at three representative depths ($z$): 0.1, 0.6 and 1 m (vertical panels).
-Mean (black lines) and lower half of 95% distribution (grey areas) of soil water potential ($\mathrm{\Psi}_{\textrm{soil},z}$) for a given DOY and depth $z$ are shown (same across census intervals).
-$\mathrm{\Psi}_{\textrm{soil},z}$ by DOY for a year (colored line) is only shown if at least one day of that year $\mathrm{\Psi}_{\textrm{soil},z}$ was more negative than the 5^th^ percentile for that day of year (DOY).
+Figure \@ref(fig:swp-droughts): Occurrence of extreme hydrological droughts by census intervals (horizontal panels) at three representative depths: 0.1, 0.6 and 1 m (vertical panels).
+Mean (black lines) and lower half of 95% distribution (grey areas) of soil water potential $\mathrm{\Psi}_{\textrm{soil},z}$ for a given DOY and depth $z$ are shown (same across all census intervals).
+$\mathrm{\Psi}_{\textrm{soil},z}$ by DOY for a year (colored line) is only shown if at least one day of the year $\mathrm{\Psi}_{\textrm{soil},z}$ was more negative than the 5^th^ percentile of $\mathrm{\Psi}_{\textrm{soil},z}$ for that day of year (DOY).
 
@@
 ```{r schematic, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-schematic}")}
-knitr::include_graphics(file.path("figures/schematic.png"))
+knitr::include_graphics(file.path("schematic4/schematic4.001.jpeg"))
 ```

### 2020-11-06 Supporting_Information.Rmd
1. Edited equations to match with the main text 2. Other corrections

@@
 We selected trees for which the height of the diameter measurements were unchanged across all censuses (1990, 1995, 2000, 2005, 2010, 2015).
-For each individual tree ($i$), we calculated absolute growth rates ($\grave{G}_{i,t}$, cm yr^-1^) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses prorated by the inter-census interval based on the exact date of measurements.
+For each individual tree ($i$), we calculated absolute growth rates $\grave{G}_{i,t}$ (cm yr^-1^) for each of the five census intervals $t$ as the difference between tree dbh measured in successive censuses prorated by the inter-census interval based on the exact date of measurements.
 We eliminated positive outliers as any growth rates amounting to >75 mm yr^-1^, as this was the highest rate observed with confidence for the fastest growing species [@Condit.2017].
-Negative outliers were removed based on known measurement error, $\sigma_{e} = 0.006214d + 0.903$, where *d* is dbh.
+Negative outliers were removed based on known measurement error, $\sigma_{e}$ = 0.006214$d$ + 0.903, where *d* is dbh.
 Any stem whose later dbh measurement fell <4$\sigma_{e}$ the earlier measurement was removed [@Condit.2017].
 As tree growth rates vary with size [@Muller-Landau:2006], to keep growth rates comparable across census intervals, we obtained residuals $\textrm{R}_{i,t}$ from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($\textrm{R}_{i,t} = \grave{G}_{i,t} - f(d_{i,t})$).
-The diameter model of growth, $f(d_{i, t})$, a B-spline based polynomial equation with five degrees of freedom, was fitted to growth rates from all trees pooled together as there was insufficient species-specific data for large trees ($R$^2 = 0.27; $p$-value < 0.01).
-We retained only those trees that had $\textrm{R}_{i,t}$ estimates for all census intervals (total trees = `r sum(g.n$n)`) and standardized (centered and scaled) individual tree time series of $\textrm{Resid}_{i,t}$.
-We then obtained growth time series $G_{s}$ for each species $s$ as the time-series of medians of $\textrm{R}_{i,t}$ for use in the ERD model.
-These were obtained for only those species ($n$ = `r length(erd.sp)`) with complete records on at least three trees (median $n$ = `r median(g.n$n)`, max $n$ = `r max(g.n$n)` trees per species).
+The diameter model of growth, $f(d_{i, t})$, a B-spline based polynomial equation with five degrees of freedom, was fitted to growth rates from all trees pooled together as there was insufficient species-specific data for large trees ($R$^2^ = 0.27; $p$-value < 0.01).
+We retained only those trees that had $\textrm{R}_{i,t}$ estimates for all census intervals (total trees = `r sum(g.n$n)`) and standardized (centered and scaled) individual tree time series of $\textrm{R}_{i,t}$.
+For each species $s$ we then obtained growth time series $G_{s}$ as the time-series of medians of $\textrm{R}_{i,t}$ for the five census intervals $t$ to use in the ERD model.
+$G_{s}$ were obtained for only those species ($n$ = `r length(erd.sp)`) with complete records on at least three trees (median `r median(g.n$n)`, max `r max(g.n$n)` trees per species).
 
-### Microclimatic tower {#SIData.matteo} 
+### Microclimatic and flux tower {#SIData.matteo} 
 \
-The tower used for these measurements is 41 m above ground, on a plateau on BCI [@Pau.2018].
+The tower used for these measurements is 41 m above ground, on a plateau on BCI at the north west corner of the 50ha plot [@Pau.2018].
 The eddy covariance system includes a sonic anemometer (CSAT3, Campbell Scientific, Logan, UT) and an open-path infrared CO~2~/H~2~O gas analyzer (LI7500, LiCOR.
@@
 Stream stage is recorded at 5-minute intervals.
-
-\
 The weir drains a catchment of approximately 40 ha of which at least 90% consists of the central plateau area of the island.
@@
 Large storms are sometimes preceded by small, initial peaks generated by the rain falling on the steeper slopes close to the weir.
-
-\
 Streamflow data are analyzed by software custom-built by author SP.
@@
 Beginning in 2016 we additionally installed three horizontal probes at depths of 0.15, 0.4 and 1 m and we applied the same calibration equation to these data (2016-2018) as that for the vertical probe.
-These data are archived at https://ngt-data.lbl.gov/.
+These data are archived at https://ngt-data.lbl.gov/ (DOI pending).
 
 \
-We also leveraged existing GWC measurements by @Kupers:2019 at the depths of 0.15, 0.4 and 1 m in summer 2015 (February, March and April) and 0.15, 0.4 in summer 2016 (March), with 1299 samples in total that covered all soil types and habitats in the plot.
+We also leveraged existing GWC measurements by @Kupers:2019 at the depths of 0.15, 0.4 and 1 m in summer 2015 (February, March and April) and 0.15, 0.4 in summer 2016 (March), with 1299 samples in total that covered all soil types and habitats in the 50ha plot.
 The sampling periods were six, five, ten and eight days long, respectively.
@@
 \
-We generated an estimate of the mean seasonality curve for normalized leaf area index (LAI$^*$, range 0-1, unitless) for individual species as follows.
+We generated an estimate of the mean seasonality curve for normalized leaf area index ($\textrm{LAI}^*$, range 0-1, unitless) for individual species as follows.
 We interpolated species-specific weekly leaf-fall data collected at two similar sites in BCI (1985-2020) to a daily estimate [@Wright.1990aeo].
@@
 The difference between leaf-fall and leaf-gain gave an estimate of daily leaf-change time series, which was then scaled to express as a fraction of the total leaf fall in a given year.
-Beginning with the LAI$^*$ of 1 at the end of the wet season in the first year of the record, we accumulated leaf-change over it to obtain a running estimate of the leaf-cover fraction, averaged it by DOY across years and normalized it to obtain the species-specific seasonality curve in LAI$^*$ (range 0-1, unitless).
+Beginning with the $\textrm{LAI}^*$ of 1 at the end of the wet season in the first year of the record, we accumulated leaf-change over it to obtain a running estimate of the leaf-cover fraction, averaged it by DOY across years and normalized it to obtain the species-specific seasonality curve in $\textrm{LAI}^*$ (range 0-1, unitless).
 (Fig.
@@
 \
-The $K_\textrm{leaf}$ was measured under steady state and high irradiance following the Evaporative Flux Method (FEM) [@Sack.2012] using a flowmeter.
+$K_\textrm{leaf}$ was measured under steady state and high irradiance following the Evaporative Flux Method (FEM) [@Sack.2012] using a flowmeter.
 Briefly, the leaf was excised under filtered deionised water and attached to a water source (graduated cylinder) using clear plastic tubing.
 The system includes tubing of a known resistance, separated upstream and downstream by two sensors measuring the transpiration-driven water flow pressure in the EFM system [@Sack.2011].
-The $K_\textrm{leaf}$ is calculated as the ratio of the transpiration rate (*E*~leaf~; mmol s^-1^) to the difference in flow pressure (MPa) upstream and downstream from the resistance tubing, and further standardized to 25°C and by leaf area.
+$K_\textrm{leaf}$ is calculated as the ratio of the transpiration rate (*E*~leaf~; mmol s^-1^) to the difference in flow pressure (MPa) upstream and downstream from the resistance tubing, and further standardized to 25°C and by leaf area.
 At stability (CV < 5%), the leaf is left to equilibrate in plastic bags with moist paper towels and measured for $\mathrm{\Psi}$~leaf~ using a Scholander pressure chamber.
@@
 The differences among AICc values for these functions were not large within species (see Zailaa et al.
-for more detailed methods description).
+for more detailed methods description.
+To be submitted.).
 We chose the exponential function without the intercept to fit observed data across species, and as a general function to scale-up to the community, because it had the lowest AICc for most species, captured the form of the observed data across all species well, and had the lowest number of parameters, that is, two.
@@
 The entire dataset includes measurements on 26 tree species collected at three sites in Panama.
-The raw data, *R* script for the analyses, and data summaries are archived in the NGEE-Tropics archive (@Wolfe.2020).
+The raw data, *R* script for the analyses, and data summaries are archived in the NGEE-Tropics archive [@Wolfe.2020].
 The data were previously reported by @Dickman.2019 and @Wu.2020.
@@
 \
-Methods for stem maximum hydraulic conductivity and vulnerability to cavitation closely followed [@Christman.2012] and [@Wolfe.2016].
+Methods for stem maximum hydraulic conductivity and vulnerability to cavitation closely followed @Christman.2012 and @Wolfe.2016.
 Briefly, canopy branches were collected from 4–10 trees of each species.
@@
 
-For each tree species measured for $K_\textrm{stem}$ and $\mathrm{\Psi}$~88,stem~, two to six leaves were measured for $\mathrm{\Psi}$~tlp~ with pressure-volume analysis following [@Koide.1989].
+For each tree species measured for $K_\textrm{stem}$ and $\mathrm{\Psi}$~88,stem~, two to six leaves were measured for $\mathrm{\Psi}$~tlp~ with pressure-volume analysis following @Koide.1989.
 Leaf water potential was measured on leaf discs with psychrometers (J.R.D.
@@
 
-Stem hydraulic safety margin was calculated as the difference between minimum leaf water potentials, $\mathrm{\Psi}$~min~, and $\mathrm{\Psi}$~88,stem~.
+Stem hydraulic safety margin was calculated as the difference between minimum leaf water potential $\mathrm{\Psi}$~min~ and $\mathrm{\Psi}$~88,stem~.
 Values for $\mathrm{\Psi}$~min~ were obtained from the dataset of @Wolfe.2019, which was collected as part of the 2016 ENSO campaign.
@@
 Here, the six species that overlapped with our assessment of ERD and $\mathrm{\Psi}$~88,stem~ were included in analyses.
-We defined species $\mathrm{\Psi}$~min~ as the minimum of all the diurnal measurements for a species, after taking an average value of measurements on two leaf samples for a given day (Table S\@ref(tab:hyd-table)).
+We defined species $\mathrm{\Psi}$~min~ as the minimum of all the diurnal measurements for a species after taking an average value of measurements on two leaf samples for a given diurnal measurement (Table S\@ref(tab:hyd-table)).
 
@@
 \begin{equation}
-G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,i|z}^{*}) + \epsilon_{t|z}
+G_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},s,i|z,h}^{*}) + \epsilon_{s,t|z,h}
 (\#eq:M1)
@@
 \begin{equation}
-G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*} + \hat{\textrm{VPD}}_{i}^{*}) + \epsilon_{t|z}
+G_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},s,i|z,h}^{*} + \hat{\textrm{VPD}}_{i}^{*}) + \epsilon_{s,t|z,h}
 (\#eq:M2)
@@
 \begin{equation}
-G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*} \hat{\textrm{VPD}}_{i}^{*}) + \epsilon_{t|z}
-(\#eq:M3)
+G_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},s,i|z,h}^{*} \hat{\textrm{VPD}}_{i}^{*}) + \epsilon_{s,t|z,h}
+(\#eq:M-3)
 \end{equation}
@@
 \begin{equation}
-G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*} \textrm{LAI}_{doy|i}^{*}) + \epsilon_{t|z}
+G_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},s,i|z,h}^{*} \textrm{LAI}_{s,doy|i}^{*}) + \epsilon_{s,t|z,h}
 (\#eq:M4)
@@
 \begin{equation}
-G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} (K^{*}_{\textrm{leaf},i|z} + \hat{\textrm{VPD}}^{*}_{i}) \textrm{LAI}^{*}_{doy|i}) + \epsilon_{t|z}
+G_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},s,i|z,h}^{*} + \hat{\textrm{VPD}}_{i}^{*}) \textrm{LAI}^{*}_{s,doy|i}) + \epsilon_{s,t|z,h}
 (\#eq:M5)
@@
 \begin{equation}
-G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*}\hat{\textrm{VPD}}_{i}^{*}\textrm{LAI}_{doy|i}^{*}) + \epsilon_{t|z}
+G_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},s,i|z,h}^{*}\hat{\textrm{VPD}}_{i}^{*}\textrm{LAI}_{s,doy|i}^{*}) + \epsilon_{s,t|z,h}
 (\#eq:M6)
@@
 where n~t~ are the total number of days in census interval $t$, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is the model error term.
-$\hat{\textrm{VPD}}$ is the predicted GPP from a locally derived polynomial relationship between gross primary productivity (GPP) and VPD so as to account for the non-linear impact of VPD on growth.
-See the ERD model description in the main text for more details.
+$\hat{\textrm{VPD}}$ is the predicted GPP from a locally derived polynomial relationship between gross primary productivity (GPP) and VPD, so as to account for the non-linear impact of VPD on growth.
+See the ERD model description in the main text for other details.
 

### 2020-11-06 manuscript.rmd
1. Hypotheses table now use symbols 2. Symbols table refered with it 3. Other edits

@@
 We evaluate how rooting depth is linked to above-ground hydraulic traits and mortality rates over a 35 year period that experienced droughts of a variety of intensity, frequency and duration.
-Our hypotheses are that (i) deep-rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem due to greater and more reliable water availability at depth, whereas shallow-rooted species are associated with less efficient but safer hydraulic traits due to more variable availability of shallow soil-water (see Table \@ref(tab:hypo)); and that, (ii) deep-rooted species have lower mortality rates during droughts due to a lower realized risk of hydraulic failure compared to shallow-rooted species.
+Our hypotheses are that (i) deep-rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem due to greater and more reliable water availability at depth, whereas shallow-rooted species are associated with less efficient but safer hydraulic traits due to more variable availability of shallow soil-water (see Tables \@ref(tab:symbols) and \@ref(tab:hypo)); and that, (ii) deep-rooted species have lower mortality rates during droughts due to a lower realized risk of hydraulic failure compared to shallow-rooted species.
 To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich tropical forest.
@@
 \@ref(eq:water-balance), determined their ranges based on literature for the study site, else for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling [LHS; @Stein.1987] from this global parameter space.
-In the Conrad catchment soil hydraulic conductivity decreases with depth from 12.5 and 60 cm (@Godsey:2004; Table \@ref(tab:param)).
+In the Conrad catchment, in which the 50 ha plot is situated, soil hydraulic conductivity decreases with depth from 12.5 and 60 cm (@Godsey:2004; Table \@ref(tab:param)).
 We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths <12.5 cm and >60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig.
@@
 We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq.
-\@ref(eq:water-balance), namely, (1) evapotranspiration ET from the eddy flux tower by the 50 ha plot ($\backsimeq$ $E_{v} + E_{g}$ in Eq.
-\@ref(eq:water-balance); 2012-2017; see [SI](#SIData.matteo))), (2) local stream discharge $q_\textrm{run}$ ($\backsimeq$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see @Paton.2019 and [SI](#SIData.steveP)) and (3) soil volumetric water content (VWC) $\mathrm{\theta}_{z}$ ($\backsimeq$ $W_{\textrm{liq,} z}$) from two sources: (i) a long-term (2012-2018) record of VWC $\mathrm{\theta}_\textrm{0-0.15 m}$ averaged across three vertical time domain reflectometry (TDR) probes over depth 0-15 cm from three locations near the eddy flux tower, and (ii) plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 at depths of 0.15, 0.4 and 1 m (1299 samples covering all soil types and habitats; see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\mathrm{\theta}_{\textrm{0-0.15 m}}$, $\mathrm{\theta}_{\textrm{0.4 m}}$ and $\mathrm{\theta}_{\textrm{1 m}}$ refer to plot-wide averages).
+\@ref(eq:water-balance), namely, (1) evapotranspiration ET from the flux tower by the 50 ha plot ($\backsimeq$ $E_{v} + E_{g}$ in Eq.
+\@ref(eq:water-balance); 2012-2017; see [SI](#SIData.matteo))), (2) local stream discharge $q_\textrm{run}$ ($\backsimeq$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see @Paton.2019 and [SI](#SIData.steveP)) and (3) soil volumetric water content (VWC) $\mathrm{\theta}_{z}$ ($\backsimeq$ $W_{\textrm{liq,} z}$) from two sources: (i) a long-term (2012-2018) record of VWC $\mathrm{\theta}_\textrm{0-0.15 m}$ averaged across three vertical time domain reflectometry (TDR) probes over depth 0-15 cm from three locations near the flux tower, and (ii) plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 at depths of 0.15, 0.4 and 1 m (1299 samples covering all soil types and habitats; see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\mathrm{\theta}_{\textrm{0-0.15 m}}$, $\mathrm{\theta}_{\textrm{0.4 m}}$ and $\mathrm{\theta}_{\textrm{1 m}}$ refer to plot-wide averages).
 
@@
 Figure \@ref(fig:hydro-calib): ELM-FATES calibration and evaluation.
-Observations (red lines and points) versus simulations (gray lines) from 100 best-fit ensemble-member runs of ELM-FATES are shown for daily Volumetric Water Content (VWC) by depth for horizontal time domain reflectometry (TDR) probes at three depths (0.1, 0.4 and 1 m; **a**) and for an average of three vertical TDR probes (0-0.15 m; **b**); monthly stream discharge (Discharge, **c**) and monthly evapotranspiration from eddy flux tower (ET, **d**).
+Observations (red lines and points) versus simulations (gray lines) from 100 best-fit ensemble-member runs of ELM-FATES are shown for daily Volumetric Water Content (VWC) by depth for horizontal time domain reflectometry (TDR) probes at three depths (0.1, 0.4 and 1 m; **a**) and for an average of three vertical TDR probes (0-0.15 m; **b**); monthly stream discharge (Discharge, **c**) and monthly evapotranspiration from the flux tower (ET, **d**).
 Red points in panel **a** show average, manual plot-wide observations of VWC with 95% CI (error bars).
@@
 # Tables
+```{r symbols, results = 'asis', echo = FALSE}
+knitr::kable(symbols.table, caption = 'The symbol, definition and units of key traits used in the simulations and analyses.', booktabs = TRUE, escape = FALSE)
+```
 
+#
 ```{r hypo, results = 'asis', echo = FALSE}
@@
 ```
-
-#
-```{r symbols, results = 'asis', echo = FALSE}
-knitr::kable(symbols.table, caption = 'The symbol, definition and units of key traits used in the simulations and analyses.', booktabs = TRUE, escape = FALSE)
-```

### 2020-11-06 manuscript.rmd
1. Fig references corrected, added

@@
 The 100 ELM-FATES ensemble members with the best fits to observed soil moisture dynamics captured soil moisture seasonality at multiple depths (Fig.
-\@ref(fig:hydro-calib)a, b), including the out-of-sample observations.
+\@ref(fig:hydro-calib)**a**, **b**), including the out-of-sample observations.
 These simulations also captured the dynamics in stream discharge (Fig.
-\@ref(fig:hydro-calib)c) and in evapotranspiration (Fig.
-\@ref(fig:hydro-calib)d), slightly underestimating peak discharge in wet years and slightly overestimating peak ET.
-The reduction in parameter range (Table S\@ref(tab:param)) in the best-fit ensembles compared to the tested global ranges showed that the model calibration was sensitive to the Ball-Berry stomatal slope parameter, the ELM root distribution parameter that regulates the depth of the rooting profile, soil hydraulic conductivity profile especially at depth, and the adjustment factor that modifies soil hydraulic conductivity to account for macroporosity and direct flow paths (Fig.
+\@ref(fig:hydro-calib)**c**) and in evapotranspiration (Fig.
+\@ref(fig:hydro-calib)**d**), slightly underestimating peak discharge in wet years and slightly overestimating peak ET.
+The reduction in parameter range (Table S\@ref(tab:param)) in the best-fit ensembles compared to the tested global ranges showed that the model calibration was sensitive to the Ball-Berry stomatal slope parameter, the ELM root distribution parameter that regulates the depth of the rooting profile, soil hydraulic conductivity profile especially at depth, and the adjustment factor that modifies soil hydraulic conductivity to account for macroporosity and direct flow paths (Figs.
 S\@ref(fig:sens-sm-et), S\@ref(fig:sens-run)).
@@
 WSG and LMA explained a large proportion of variance in parameter $B$ (Eq.
-\@ref(eq:b-mod); Adj.
-$R$^2^ = `r bfit.r2`; $p$ `r  bfit.p`).
+\@ref(eq:b-mod), Adj.
+$R$^2^ = `r bfit.r2`, $p$ `r  bfit.p`).
 WSG and parameter $B$ explained a large proportion of variance in parameter $A$ (Eq.
-\@ref(eq:a-mod); *Adj.
-R*^2^ = `r afit.r2`; $p$ `r  afit.p`; also see Fig.
+\@ref(eq:a-mod), *Adj.
+R*^2^ = `r afit.r2`, $p$ `r  afit.p`; also see Fig.
 S\@ref(fig:obs-plc-curves)).
@@
 The best ERD model (with a multiplicative effect of VPD and not LAI, Eq.
-\@ref(eq:M-chosen)) explained a large fraction of the variance in $\mathrm{\delta}$^2^H~xylem~ ($R$^2^ = `r lm.erd.iso.r2`; $p$ = `r lm.erd.iso.p`, $n$ = `r length(unique(erd.iso$sp))`; Fig.
+\@ref(eq:M-chosen)) explained a large fraction of the variance in $\mathrm{\delta}$^2^H~xylem~ ($R$^2^ = `r lm.erd.iso.r2`, $p$ = `r lm.erd.iso.p`, $n$ = `r length(unique(erd.iso$sp))`, Fig.
 \@ref(fig:erd-iso)).
 Two of the alternative models with LAI had a strong relationship with $\mathrm{\delta}$^2^H~xylem~ as well, but one that was based on a smaller set of species ($n$ = 4), because fewer species passed the criteria of sufficient species-level goodness-of-fit for these models (<20 species vs.
-`r length(erd.sp)` species for the best model; Fig.
+`r length(erd.sp)` species for the best model, Fig.
 S\@ref(fig:other-erd-mods)).
@@
 \@ref(fig:sp-erd)).
-Notably, two species, *Luehea seemannii* and *Trichilia tuberculata*, that @Meinzer:1999 found to have $\mathrm{\delta}$^2^H~xylem~ values between soil water and groundwater, suggesting that these species sourced most of the water from depths >1 m, were also identified by our model with ERD >1 m (2.9 m for both species; Fig.
+Notably, two species, *Luehea seemannii* and *Trichilia tuberculata*, that @Meinzer:1999 found to have $\mathrm{\delta}$^2^H~xylem~ values between soil water and groundwater, suggesting that these species sourced most of the water from depths >1 m, were also identified by our model with ERD >1 m (2.9 m for both species, Fig.
 \@ref(fig:erd-iso)).
@@
 
-Species with deeper ERD showed greater $K_\textrm{max,stem}$ (Spearman's' $r$ = `r cor.k`, $p$ = `r p.k`; Fig.
-\@ref(fig:erd-hyd)a), less negative leaf $\mathrm{\Psi}$~tlp~ ($r$ = `r cor.tlp`, $p$ = `r p.tlp`; Fig.
-\@ref(fig:erd-hyd)b), less negative $\mathrm{\Psi}$~88,stem~ and thus greater vulnerability to embolism from cavitation ($r$ = `r cor.stem.88`, $p$ = `r p.stem.88`; Fig.
-\@ref(fig:erd-hyd)c), and narrower hydraulic safety margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~; $r$ = `r cor.hsm`, $p$ = `r p.hsm`; Fig.
-\@ref(fig:erd-hyd)d).
+Species with deeper ERD showed greater $K_\textrm{max,stem}$ (Spearman's' $r$ = `r cor.k`, $p$ = `r p.k`, Fig.
+\@ref(fig:erd-hyd)**a**), less negative leaf $\mathrm{\Psi}$~tlp~ ($r$ = `r cor.tlp`, $p$ = `r p.tlp`, Fig.
+\@ref(fig:erd-hyd)**b**), less negative $\mathrm{\Psi}$~88,stem~ and thus greater vulnerability to xylem embolism from cavitation ($r$ = `r cor.stem.88`, $p$ = `r p.stem.88`, Fig.
+\@ref(fig:erd-hyd)**c**), and more negative above-ground hydraulic safety margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~, $r$ = `r cor.hsm`, $p$ = `r p.hsm`, Fig.
+\@ref(fig:erd-hyd)**d**).
 
@@
 When deciduous species were included, ERD explained lower but still significant proportion of variation in mortality in those intervals ($R$^2^ = `r m.r2[1]`, `r m.r2[2]` and `r m.r2[3]`% and $p$ = `r m.p.90`, `r m.p.05` and `r m.p.10`, respectively; Fig.
-\@ref(fig:erd-mort-census)).
+S\@ref(fig:erd-mort-census)).
 Census intervals 2000-2005 and 2005-2010 were also the periods during which extreme soil droughts distinctly occurred in the wet season, effectively extending the dry season into the wet season (Fig.
@@
 For the census interval 1990-1995, species with the shallowest ERD also showed greater realized hydraulic risk (Fig.
-\@ref(fig:erd-time)), but during the dry season (; Fig.
-\@ref(fig:swp-droughts), and without elevated mortality rates (Fig.
-\@ref(fig:erd-mort-census).
+\@ref(fig:erd-time)), but during the dry season ( Fig.
+\@ref(fig:swp-droughts)), and without elevated mortality rates (Fig.
+\@ref(fig:erd-mort-census-evg)).
 Overall, species with ERD deeper than 1.7 m did not cross species $\mathrm{\Psi}$~crit~ for the entire 25-yr period for which $\mathrm{\Psi}_{\textrm{soil},z}$ was modeled (1990-2015).
@@
 
-Our novel approach for estimating ERD using 25 years of tree growth, species-specific leaf vulnerability curves, modeled soil moisture profiles, and observed VPD was consistent with estimates using $\mathrm{\delta}$^2^H~xylem~.
-Our analyses revealed that co-occurring large canopy tree species with deeper ERD were associated with higher stem hydraulic efficiency, but not safety (Fig.
+Our novel approach for estimating ERD using 25 years of tree growth, species-specific leaf vulnerability curves, modeled soil moisture profiles, and observed VPD was consistent with estimates using $\mathrm{\delta}$^2^H~xylem~ (Figs.
+\@ref(fig:sp-erd), \@ref(fig:erd-iso)).
+Our analyses revealed that co-occurring large canopy tree species with deeper ERD were associated with higher above-ground hydraulic efficiency, but not safety (Fig.
 \@ref(fig:erd-hyd)).
-Species with deeper ERD showed higher maximum stem hydraulic conductance, and greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species.
+Species with deeper ERD showed higher maximum stem hydraulic conductance, and greater xylem vulnerability to embolism and more negative above-ground hydraulic safety margins compared to shallow ERD species.
 Nevertheless, deeper ERD species showed lower mortality rates than shallower ERD species in significant wet season droughts over the 35-yr record (1981-2015) (Fig.
@@
 We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins (Fig.
-\@ref(fig:erd-hyd)), that is, traits that are commonly identified as proxies for mortality risk [@Anderegg:2016], but they were less droughted and had lower mortality compared to species relying on shallower water through periods of extreme droughts (Fig.
-\@ref(fig:erd-mort-census-evg), S\@ref(fig:erd-mort-census), \@ref(fig:swp-droughts)).
+\@ref(fig:erd-hyd)), that is, traits that are commonly identified as proxies for mortality risk [@Anderegg:2016], but they were less droughted and had lower mortality compared to species relying on shallower water through periods of extreme droughts (Figs.
+\@ref(fig:swp-droughts), \@ref(fig:erd-mort-census-evg), S\@ref(fig:erd-mort-census)).
 Deeper ERD species on average had more hydraulically efficient, but risky, suite of above-ground traits optimized for an environment of more reliable and greater water source at depth that resulted in a lower mortality risk.
-By contrast shallow ERD species had inefficient, but safer, xylem likely optimized for high variability in $\mathrm{\Psi}_\textrm{soil}$ in the shallow soil layers, and resulting in higher variance in mortality rates across censuses as $\mathrm{\Psi}_{\textrm{soil},z\textrm{ = ERD}}$ crossed ERD $\mathrm{\Psi}_\textrm{crit}$ (Fig.
-\@ref(fig:erd-mort-census-evg); Fig.
-\@ref(fig:erd-time)).
+By contrast shallow ERD species had inefficient, but safer, xylem likely optimized for high variability in $\mathrm{\Psi}_\textrm{soil}$ in the shallow soil layers, and resulting in higher variance in mortality rates across censuses as $\mathrm{\Psi}_{\textrm{soil},z\textrm{ = ERD}}$ crossed ERD $\mathrm{\Psi}_\textrm{crit}$ (Figs.
+\@ref(fig:erd-mort-census-evg), \@ref(fig:erd-time)).
 Our results are consistent with studies that included three or four species each with known rooting or water-sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016; also see @Brum.2019; @Brum.2017].
@@
 Mean annual rainfall of 1095 mm in CT2018 compared to `r round(rain.man.stats$rain)` mm at BCI, and precipitation to potential evapotranspiration ratio of nearly one in CT2018 compared to nearly two in BCI are major factors in the different mortality responses.
-At BCI deep soil layers were recharged annually (Fig.
+At BCI deep soil layers were recharged annually (Figs.
 S\@ref(fig:swp-all-depths), \@ref(fig:swp-droughts)), whereas in CT2018 they were not.
@@
 
-By estimating ERD for large trees of `r length(erd.sp)` canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species-rich tropical forests.
-Our chosen ERD model is a key improvement over the model of CT2018 ($n$ = 12 species) as we employ a physiologically meaningful representation, fewer parameters and corroboration with tree and soil data for $\mathrm{\delta}$^2^H.
+By estimating ERD for large trees of `r length(erd.sp)` canopy tree species (Fig.
+\@ref(fig:sp-erd)), we make an important advance in modeling effective rooting depths at the community level in species-rich tropical forests.
+Our best ERD model is a key improvement over the model of CT2018 ($n$ = 12 species) as we employ a physiologically meaningful representation, fewer parameters and corroboration with tree and soil data for $\mathrm{\delta}$^2^H.
 We note that stable water isotopes for plant and below-ground sources are critical for calibration or validation of ERD models for scaling upto community, although the interpretation of isotopic data is still under research [@Deurwaerder.2020; @Adams.2020; @Bowers.2020].
@@
 ```{r hypo, results = 'asis', echo = FALSE}
-knitr::kable(hypo.table, caption = 'Hypotheses for association between Effective Rooting Depth (ERD) and hydraulic traits.', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(hypo.table))))
+knitr::kable(hypo.table, caption = 'Hypotheses for association between Effective Rooting Depth (ERD) and above-ground hydraulic traits.', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(hypo.table))))
 ```

### 2020-11-06 manuscript.rmd
1. SI figure number fixed, not dynamic anymore

@@
 To parameterize ELM-FATES, we identified eleven parameters relevant for Eq.
-\@ref(eq:water-balance), determined their ranges based on literature for the study site, else for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling [LHS; @Stein.1987] from this global parameter space.
-In the Conrad catchment, in which the 50 ha plot is situated, soil hydraulic conductivity decreases with depth from 12.5 and 60 cm (@Godsey:2004; Table \@ref(tab:param)).
+\@ref(eq:water-balance), determined their ranges based on literature for the study site, else for the tropics (Table S1), and ran 5000 simulations using Latin Hypercube Sampling [LHS; @Stein.1987] from this global parameter space.
+In the Conrad catchment, in which the 50 ha plot is situated, soil hydraulic conductivity decreases with depth from 12.5 and 60 cm (@Godsey:2004; Table S1).
 We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths <12.5 cm and >60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig.
-S\@ref(fig:ksat-profiles)).
+S1).
 For greater accuracy, instead of the ELM default pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content (GWC) vs.
@@
 VPD affects growth non-linearly, with growth increasing with VPD up-to a VPD threshold, then decreasing as leaf pores (stomata) close reducing water-uptake [@Grossiord.2020; @Yang.2019].
-To account for the non-linear impact of VPD on growth, we used predicted gross primary productivity (GPP; hereafter, $\hat{\textrm{VPD}}$) from a locally derived polynomial relationship between GPP and VPD (see SI).
+To account for the non-linear impact of VPD on growth, we used predicted gross primary productivity (GPP; hereafter, $\hat{\textrm{VPD}}$) from a locally derived polynomial relationship between GPP and VPD (see Fig.
+S2 and SI).
 Apart from stomatal control, leaf-drop may further limit water-uptake and growth.
@@
 We leveraged $\mathrm{\delta}$^2^H~xylem~ from BCI [@Meinzer:2001] for the dry-season of March 1997 as this period showed largest seasonal divergence in $\mathrm{\delta}$^2^H~xylem~ among species and vertically in soil and groundwater $\mathrm{\delta}$^2^H at natural abundance level (Fig.
-S\@ref(fig:iso-data)).
+S3).
 Analyzing $\mathrm{\delta}$^2^H~xylem~ for 12 tree species @Meinzer:1999 found that four species sourced water deeper than 1 m, two other at least 0.8 m deep, one between 0 to 0.2 m, and five in the 0.2 to 0.7 m depth range over which the relationship between soil depth and $\mathrm{\delta}$^2^H~soil~ was variable.
@@
 We regressed species ERD against maximum stem hydraulic conductivity $K_\textrm{max,stem}$ (*n*=7), leaf turgor loss point $\mathrm{\Psi}$~tlp~ (*n*=7), vulnerability to embolism from cavitation $\mathrm{\Psi}$~88,stem~ measured in terms of pressure at which 88% of $K_\textrm{max,stem}$ is lost (*n*=7), and hydraulic safety margins $\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~ (*n*=6).
-See [SI](#SIData.brett) and Table S\@ref(tab:hyd-table) for data collection and estimation of these variables.
+See [SI](#SIData.brett) and Table S2 for data collection and estimation of these variables.
 
@@
 \@ref(fig:hydro-calib)**d**), slightly underestimating peak discharge in wet years and slightly overestimating peak ET.
-The reduction in parameter range (Table S\@ref(tab:param)) in the best-fit ensembles compared to the tested global ranges showed that the model calibration was sensitive to the Ball-Berry stomatal slope parameter, the ELM root distribution parameter that regulates the depth of the rooting profile, soil hydraulic conductivity profile especially at depth, and the adjustment factor that modifies soil hydraulic conductivity to account for macroporosity and direct flow paths (Figs.
-S\@ref(fig:sens-sm-et), S\@ref(fig:sens-run)).
+The reduction in parameter range (Table S1) in the best-fit ensembles compared to the tested global ranges showed that the model calibration was sensitive to the Ball-Berry stomatal slope parameter, the ELM root distribution parameter that regulates the depth of the rooting profile, soil hydraulic conductivity profile especially at depth, and the adjustment factor that modifies soil hydraulic conductivity to account for macroporosity and direct flow paths (Figs.
+S4, S5).
 
@@
 Hydrological droughts in census intervals 2000-2005 and 2005-2010 distinctly occurred in the wet season, with one to two extreme hydrological drought years each, depending on the depth location (Fig.
-S\@ref(fig:swp-droughts)).
+\@ref(fig:swp-droughts)).
 Across the simulation period 1990-2015, simulated $\mathrm{\Psi}_{\textrm{soil},z}$ remained above -0.5 MPa from depths 1.7 to 13 m (Fig.
-S\@ref(fig:swp-all-depths)).
+S6).
 
@@
 R*^2^ = `r afit.r2`, $p$ `r  afit.p`; also see Fig.
-S\@ref(fig:obs-plc-curves)).
-This predictive power allowed us to estimate leaf vulnerability curves used in the ERD models for `r sp.n.model` out of `r length(erd.sp)` ERD species that lacked direct observations (Table S\@ref(tab:ab-param), Fig.
-S\@ref(fig:obs-comm-plc-curves)).
+S7).
+This predictive power allowed us to estimate leaf vulnerability curves used in the ERD models for `r sp.n.model` out of `r length(erd.sp)` ERD species that lacked direct observations (Table S3, Fig.
+S8).
 
@@
 `r length(erd.sp)` species for the best model, Fig.
-S\@ref(fig:other-erd-mods)).
+S9).
 For the rest of the analyses we therefore used ERD estimates from the less parsimonious, best model.
@@
 \@ref(fig:erd-hyd)**d**).
+See Fig.
+S10 for the full correlation matrix.
 
@@
 When deciduous species were included, ERD explained lower but still significant proportion of variation in mortality in those intervals ($R$^2^ = `r m.r2[1]`, `r m.r2[2]` and `r m.r2[3]`% and $p$ = `r m.p.90`, `r m.p.05` and `r m.p.10`, respectively; Fig.
-S\@ref(fig:erd-mort-census)).
+S11).
 Census intervals 2000-2005 and 2005-2010 were also the periods during which extreme soil droughts distinctly occurred in the wet season, effectively extending the dry season into the wet season (Fig.
@@
 Nevertheless, deeper ERD species showed lower mortality rates than shallower ERD species in significant wet season droughts over the 35-yr record (1981-2015) (Fig.
-S\@ref(fig:erd-mort-census)), especially amongst the evergreen species (Fig.
+S11), especially amongst the evergreen species (Fig.
 \@ref(fig:erd-mort-census-evg)).
@@
 \@ref(fig:erd-hyd)), that is, traits that are commonly identified as proxies for mortality risk [@Anderegg:2016], but they were less droughted and had lower mortality compared to species relying on shallower water through periods of extreme droughts (Figs.
-\@ref(fig:swp-droughts), \@ref(fig:erd-mort-census-evg), S\@ref(fig:erd-mort-census)).
+\@ref(fig:swp-droughts), \@ref(fig:erd-mort-census-evg), S11).
 Deeper ERD species on average had more hydraulically efficient, but risky, suite of above-ground traits optimized for an environment of more reliable and greater water source at depth that resulted in a lower mortality risk.
@@
 At BCI deep soil layers were recharged annually (Figs.
-S\@ref(fig:swp-all-depths), \@ref(fig:swp-droughts)), whereas in CT2018 they were not.
+S6, \@ref(fig:swp-droughts)), whereas in CT2018 they were not.
 The contrast between the two studies highlights the combined role of seasonal precipitation input and site-specific hydrology in modulating mortality-risk of deep-rooted species.
@@
 Three of our exploratory ERD models included LAI seasonality, but we did not select them as they worsened the fit with growth data for many species (Fig.
-S\@ref(fig:other-erd-mods)).
+S9).
 We interpret this result as that VPD and leaf hydraulic vulnerability curves may be adequate to explain inter-census differences in growth (via stomatal control), but also acknowledge that our estimates of seasonality of LAI is a tentative estimate that combines leaf-fall data and the timing of leaf-gain backtracked from leaf lifetime, and omits inter-annual variation.

### 2020-11-06 Supporting_Information.Rmd
1. Figures refer to in the main text are presented first, then tables then SI text

@@
 
-#Supporting Information
+# Supporting Information
+
+## Supplementary Figures
+
+```{r ksat-profiles, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ksat-profiles}Generated 5000 curves of soil hydraulic conductivity variation by depth based on Latin Hypercube Sampling within 95% CI of observed data at depths of 12.5 cm and 60 cm [@Kinner:2004], assuming a linear decline between those depths, and values before 12.5 and after 60 cm as that of 12.5 cm and 60 cm, respectively.
+Only 100 curves are shown for clarity.
+Note the log scale on horizontal axis.")}
+knitr::include_graphics(file.path("figures/Bootstrapped Ksat profiles within observed 95CI_random_two_hundred.jpeg"))
+```
+
+```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "35%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd}Scatterplot of mean midday (11:00 am to 15:00 pm) gross primary productivity (GPP) and mean midday vapor pressure deficit (VPD) measured at the eddy flux tower in Barro Colorado Island over 2012-2017.
+Red line shows the fit from a polynomial model described in inset."}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/gpp_by_env_variables/monthly_gpp.vpd.jpeg"))
+```
+
+```{r iso-data, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-iso-data}Stable isotopic ratio ($\\mathrm{\\delta}^2H$) of soil water by depth measured at Barro Colorado Island in March 1997.
+Data from @Meinzer:1999."}
+knitr::include_graphics(file.path("figures/UDI_confidence/Meinzer_etal_1999_Depth_vs_soil_deltaD.jpeg"))
+```
+
+```{r sens-sm-et, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-sm-et}Parameter sensitivity of ELM-FATES soil water content (that is, VWC, **a**-**b**) and ET (**c**-**d**) is shown for two representative sample months, as sensitivity for a particular parameter varies depending on the chosen month.
+Horizontal axes show global ranges for a given parameter.
+Parameters are color coded and are consistent across panels.
+Each point represents a simulation among the 5000 ensemble-member runs.
+See S\\@ref(tab:param) for description, global ranges and chosen ranges for parameters."}
+knitr::include_graphics(file.path("figures/ELM_FATES_parameter_sensitivity/sens.sm.et.jpeg"))
+```
+
+```{r sens-run, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-run}Parameter sensitivity of ELM-FATES stream discharge (QRUNOFF) is shown for four sample months (**a**-**d**), as sensitivity for a particular parameter varies depending on the chosen month.
+Horizontal axes show global ranges for a given parameter.
+Parameters are color coded and are consistent across panels.
+Each point represents a simulation among the 5000 ensemble-member runs.
+See S\\@ref(tab:param) for description, global ranges and chosen ranges for parameters."}
+knitr::include_graphics(file.path("figures/ELM_FATES_parameter_sensitivity/sens.run.jpeg"))
+```
+
+```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($\\Psi_{\\textrm{soil},z}$) by depth $z$ (m; panels) over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines).
+Vertical gray lines demarcate median dates of the demographic censuses at the BCI 50 ha plot."}
+knitr::include_graphics(file.path("figures/2019-10-14_5000/best-fits/psi_model_daily_all_depths_params.top.few_full.jpeg"))
+```
+
+```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}Vulnerability curves fitted to observed $K_\\textrm{leaf}$ vs.
+$\\mathrm{\\Psi}$~leaf~ data for ", length(sp.exp.param$sp)," tree species from BCI--in absolute terms (**a**, **b**), as well as, in terms of loss of conductivity (%) (**c**, **d**).
+Each line represents a species, color coded by WSG (**a**, **c**), or LMA (**b**, **d**), with blue tones indicating smaller values of WSG or LMA and red tones indicating larger values.")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/plot.obs.plc.jpeg"))
+```
+
+```{r obs-comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-comm-plc-curves}Percent loss of vulnerability curves for species with ERD estimates.
+Each line represents a species, color coded by source of the curve: fitted to observed data for $K_\\textrm{leaf}$ vs.
+$\\mathrm{\\Psi}$~leaf~ are in yellow and those based on the scaling relationship with WSG and LMA are in gray.")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/obs.mod_plccurves_fits_ggplot_no_col.jpeg"))
+```
+
+```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-other-erd-mods}Effective Rooting Depth versus $\\mathrm{\\delta}^2H$ for each of the ERD models tested (panels).
+For each ERD model only those species have an estimate of ERD that have passed the test of Pearson's $r$ > 0.71 during ERD parameterisation.
+The number of species with an ERD estimate as well as $\\mathrm{\\delta}^2H$ for each model are shown in inset, along with the goodness-of-fit ($R$^2^) and $p$-values for a linear relationship between ERD and $\\mathrm{\\delta}^2H$.
+ERD models in panels **a**-**e** correspond to Eq.
+\\@ref(eq:M1)-\\@ref(eq:M6), respectively.")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_sp_color_Meinzer.jpeg"))
+```
+
+```{r erd-hyd-full, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd-full}.
+Pairwise correlations (upper triangle) and corresponding linear regressions (lower triangle) among modelled Effective Rooting Depth (ERD) and hydraulic properties for seven canopy species found on Barro Colorado Island (Panama).
+Spearman's correlation coefficients and significance levels are shown (upper-half; '*' $p$ < 0.05; '.' $p$ < 0.1; ' ' $p$ < 1)")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/erd.stem.traits_cor.chart.jpeg"))
+```
+
+```{r erd-mort-census, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-census}Species mortality rate in each census interval during the study period plotted against ERD.
+$R$^2^ and significance levels for linear model fits are given in panel insets.
+Linear model fits (blue lines) with confidance bands (gray area) are shown for significant fits at $\\alpha$ = 0.1."}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_by rdi.gr.jpeg"))
+```
+
+## Supplementary Tables
+
+```{r param, results = 'asis', echo = FALSE}
+knitr::kable(param.table, caption = 'ELM-FATES parameters used to generate ensembles, with description, prescribed global ranges, rationale for the choice of ranges and references.
+Ranges for best-fit chosen ensembles are also given.
+Note that no.
+1-6 are parameters specific to FATES.', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(param.table))))
+```
+#
+```{r hyd-table, results = 'asis', echo = FALSE}
+knitr::kable(hyd.table, caption = paste0("\\label{lab-hyd-table} Minimum leaf water potential $\\mathrm{\\Psi}$~min~, leaf turgor loss point $\\mathrm{\\Psi}$~tlp~, stem maximum hydraulic conductivity $K_\\textrm{max,stem}$ and stem vulnerability to cavitation $\\mathrm{\\Psi}$~88,stem~ for seven species with ERD estimates.
+$\\mathrm{\\Psi}$~min~ is sourced from @Wolfe.2019, while $\\mathrm{\\Psi}$~tlp~, $K_\\textrm{max,stem}$ and $\\mathrm{\\Psi}$~88,stem~ are sourced from @Wolfe.2020.
+$\\mathrm{\\Psi}$~min~, $\\mathrm{\\Psi}$~tlp~ and $\\mathrm{\\Psi}$~88,stem~ are in MPa, while $K_\\textrm{max,stem}$ in kg m^-1^ s^-1^ MPa^-1^."), booktabs = TRUE, escape = FALSE)
+```
+
+#
+```{r ab-param, results = 'asis', echo = FALSE}
+knitr::kable(ab.table, caption = paste0("\\label{lab-ab-param}Leaf vulnerability curves parameters *A* & *B*, $K_\\textrm{max,leaf}$ and $\\mathrm{\\Psi}$ for the ", length(erd.sp), " species with ERD estimates.
+Species for which curves were fitted to observed data for $K_\\textrm{leaf}$ vs.
+$\\mathrm{\\Psi}$~leaf~ are identified with Source as Data, while species for which curves were obtained based on trait-proxies are identified with Source as Model."), booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(ab.table))))
+```
 
@@
 
+#
+```{r eddy, results = 'asis', echo = FALSE}
+knitr::kable(eddy.table, caption = 'QA/QC procedure applied to eddy covariance fluxes.
+w: vertical wind speed; t: air temperature; q: water vapor density; c: CO~2~ density; u: horizontal wind speed.
+The bar indicates 30-min averaging and the carat indicates fluctuation ($\\overline{\\hat{x} = x - \\bar{x}}$)', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(eddy.table))))
+```
+
 ### Stream discharge from the Conrad catchment {#SIData.steveP} 
@@
 
+```{r TDR-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-TDR-calib}Volumetric Water Content vs.
+TDR data for a vertical TDR probe (0-15 cm depth) at the base of the eddy flux tower near the 50 ha plot at BCI."}
+knitr::include_graphics(file.path("figures/Matteos_TDR_SWC_data_linear_model.jpeg"))
+```
+
 ### Leaf Area Index 
@@
 For the rest `r erd.sp.wo.ll` species, leaf lifetime was predicted from LMA ($\overline{lifetime} = `r ll.lma.cf[1]` + `r ll.lma.cf[2]`LMA$; $R$^2^ = `r ll.lma.r2`; $p$ `r ll.lma.p`, $n$ = `r ll.lma.n`).
+
+```{r leaf-cover, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-cover}Seasonal variation in standardised LAI (range 0-1) for ", length(erd.sp)," species at Barro Colorado Island (panels).
+Species are color-coded by deciduousness.")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf.cover_BCI_multi_panel.jpeg"))
+```
 
@@
 
-# Supplementary Figures
-
-```{r ksat-profiles, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ksat-profiles}Generated 5000 curves of soil hydraulic conductivity variation by depth based on Latin Hypercube Sampling within 95% CI of observed data at depths of 12.5 cm and 60 cm [@Kinner:2004], assuming a linear decline between those depths, and values before 12.5 and after 60 cm as that of 12.5 cm and 60 cm, respectively.
-Only 100 curves are shown for clarity.
-Note the log scale on horizontal axis.")}
-knitr::include_graphics(file.path("figures/Bootstrapped Ksat profiles within observed 95CI_random_two_hundred.jpeg"))
-```
-
-```{r TDR-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-TDR-calib}Volumetric Water Content vs.
-TDR data for a vertical TDR probe (0-15 cm depth) at the base of the eddy flux tower near the 50 ha plot at BCI."}
-knitr::include_graphics(file.path("figures/Matteos_TDR_SWC_data_linear_model.jpeg"))
-```
-
-```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "35%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd}Scatterplot of mean midday (11:00 am to 15:00 pm) gross primary productivity (GPP) and mean midday vapor pressure deficit (VPD) measured at the eddy flux tower in Barro Colorado Island over 2012-2017.
-Red line shows the fit from a polynomial model described in inset.
-GPP predicted from this equation is used as \\hat{\\textrm{VPD}} in equations \\@ref(eq:M2), \\@ref(eq:M3), \\@ref(eq:M5) and \\@ref(eq:M6)."}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/gpp_by_env_variables/monthly_gpp.vpd.jpeg"))
-```
-
-```{r leaf-cover, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-cover}Seasonal variation in standardised LAI (range 0-1) for ", length(erd.sp)," species at Barro Colorado Island (panels).
-Species are color-coded by deciduousness.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf.cover_BCI_multi_panel.jpeg"))
-```
-
-```{r iso-data, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-iso-data}Stable isotopic ratio ($\\mathrm{\\delta}^2H$) of soil water by depth measured at Barro Colorado Island in March 1997.
-Data from @Meinzer:1999."}
-knitr::include_graphics(file.path("figures/UDI_confidence/Meinzer_etal_1999_Depth_vs_soil_deltaD.jpeg"))
-```
-
-```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($\\Psi_{\\textrm{soil},z}$) by depth $z$ (m; panels) over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines).
-Vertical gray lines demarcate median dates of the demographic censuses at the BCI 50 ha plot."}
-knitr::include_graphics(file.path("figures/2019-10-14_5000/best-fits/psi_model_daily_all_depths_params.top.few_full.jpeg"))
-```
-
-```{r sens-sm-et, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-sm-et}Parameter sensitivity of ELM-FATES soil water content (that is, VWC, **a**-**b**) and ET (**c**-**d**) is shown for two representative sample months, as sensitivity for a particular parameter varies depending on the chosen month.
-Horizontal axes show global ranges for a given parameter.
-Parameters are color coded and are consistent across panels.
-Each point represents a simulation among the 5000 ensemble-member runs.
-See S\\@ref(tab:param) for description, global ranges and chosen ranges for parameters."}
-knitr::include_graphics(file.path("figures/ELM_FATES_parameter_sensitivity/sens.sm.et.jpeg"))
-```
-
-```{r sens-run, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-run}Parameter sensitivity of ELM-FATES stream discharge (QRUNOFF) is shown for four sample months (**a**-**d**), as sensitivity for a particular parameter varies depending on the chosen month.
-Horizontal axes show global ranges for a given parameter.
-Parameters are color coded and are consistent across panels.
-Each point represents a simulation among the 5000 ensemble-member runs.
-See S\\@ref(tab:param) for description, global ranges and chosen ranges for parameters."}
-knitr::include_graphics(file.path("figures/ELM_FATES_parameter_sensitivity/sens.run.jpeg"))
-```
-
-```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}Vulnerability curves fitted to observed $K_\\textrm{leaf}$ vs.
-$\\mathrm{\\Psi}$~leaf~ data for ", length(sp.exp.param$sp)," tree species from BCI--in absolute terms (**a**, **b**), as well as, in terms of loss of conductivity (%) (**c**, **d**).
-Each line represents a species, color coded by WSG (**a**, **c**), or LMA (**b**, **d**), with blue tones indicating smaller values of WSG or LMA and red tones indicating larger values.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/plot.obs.plc.jpeg"))
-```
-
-```{r obs-comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-comm-plc-curves}Percent loss of vulnerability curves for species with ERD estimates.
-Each line represents a species, color coded by source of the curve: fitted to observed data for $K_\\textrm{leaf}$ vs.
-$\\mathrm{\\Psi}$~leaf~ are in yellow and those based on the scaling relationship with WSG and LMA are in gray.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/obs.mod_plccurves_fits_ggplot_no_col.jpeg"))
-```
-
-```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-other-erd-mods}Effective Rooting Depth versus $\\mathrm{\\delta}^2H$ for each of the ERD models tested (panels).
-For each ERD model only those species have an estimate of ERD that have passed the test of Pearson's $r$ > 0.71 during ERD parameterisation.
-The number of species with an ERD estimate as well as $\\mathrm{\\delta}^2H$ for each model are shown in inset, along with the goodness-of-fit ($R$^2^) and $p$-values for a linear relationship between ERD and $\\mathrm{\\delta}^2H$.
-ERD models in panels **a**-**e** correspond to Eq.
-\\@ref(eq:M1)-\\@ref(eq:M6), respectively.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_sp_color_Meinzer.jpeg"))
-```
-
-```{r erd-hyd-full, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd-full}.
-Pairwise correlations (upper triangle) and corresponding linear regressions (lower triangle) among modelled Effective Rooting Depth (ERD) and hydraulic properties for seven canopy species found on Barro Colorado Island (Panama).
-Spearman's correlation coefficients and significance levels are shown (upper-half; '*' $p$ < 0.05; '.' $p$ < 0.1; ' ' $p$ < 1)")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/erd.stem.traits_cor.chart.jpeg"))
-```
-
-```{r erd-mort-census, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-census}Species mortality rate in each census interval during the study period plotted against ERD.
-$R$^2^ and significance levels for linear model fits are given in panel insets.
-Linear model fits (blue lines) with confidance bands (gray area) are shown for significant fits at $\\alpha$ = 0.1."}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_by rdi.gr.jpeg"))
-```
-
-```{r erd-time-evg, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time-evg}Modelled species Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\\mathrm{\\Psi}_{\\textrm{soil},z}$ was below $\\mathrm{\\Psi}$~crit~, defined as $\\mathrm{\\Psi}$~80,leaf~, over five census intervals (1990-2015), where $z$ is the soil depth matching species ERD--for ", nrow(mrate.depth.mean %>% subset(deciduous == "E")), " canopy species found on Barro Colorado Island.
-Each circle represents a species.
-Circles are colored and slightly jittered horizontally for clarity.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mfac vs. rdi.gr_evergreen.jpeg"))
-```
-
-#Supplementary Tables
-
-```{r param, results = 'asis', echo = FALSE}
-knitr::kable(param.table, caption = 'ELM-FATES parameters used to generate ensembles, with description, prescribed global ranges, rationale for the choice of ranges and references.
-Ranges for best-fit chosen ensembles are also given.
-Note that no.
-1-6 are parameters specific to FATES.', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(param.table))))
-```
-#
-```{r eddy, results = 'asis', echo = FALSE}
-knitr::kable(eddy.table, caption = 'QA/QC procedure applied to eddy covariance fluxes.
-w: vertical wind speed; t: air temperature; q: water vapor density; c: CO~2~ density; u: horizontal wind speed.
-The bar indicates 30-min averaging and the carat indicates fluctuation ($\\overline{\\hat{x} = x - \\bar{x}}$)', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(eddy.table))))
-```
-#
-```{r ab-param, results = 'asis', echo = FALSE}
-knitr::kable(ab.table, caption = paste0("\\label{lab-ab-param}Leaf vulnerability curves parameters *A* & *B*, $K_\\textrm{max,leaf}$ and $\\mathrm{\\Psi}$ for the ", length(erd.sp), " species with ERD estimates.
-Species for which curves were fitted to observed data for $K_\\textrm{leaf}$ vs.
-$\\mathrm{\\Psi}$~leaf~ are identified with Source as Data, while species for which curves were obtained based on trait-proxies are identified with Source as Model."), booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(ab.table))))
-```
-#
-```{r hyd-table, results = 'asis', echo = FALSE}
-knitr::kable(hyd.table, caption = paste0("\\label{lab-hyd-table} Minimum leaf water potential $\\mathrm{\\Psi}$~min~, leaf turgor loss point $\\mathrm{\\Psi}$~tlp~, stem maximum hydraulic conductivity $K_\\textrm{max,stem}$ and stem vulnerability to cavitation $\\mathrm{\\Psi}$~88,stem~ for seven species with ERD estimates.
-$\\mathrm{\\Psi}$~min~ is sourced from @Wolfe.2019, while $\\mathrm{\\Psi}$~tlp~, $K_\\textrm{max,stem}$ and $\\mathrm{\\Psi}$~88,stem~ are sourced from @Wolfe.2020.
-$\\mathrm{\\Psi}$~min~, $\\mathrm{\\Psi}$~tlp~ and $\\mathrm{\\Psi}$~88,stem~ are in MPa, while $K_\\textrm{max,stem}$ in kg m^-1^ s^-1^ MPa^-1^."), booktabs = TRUE, escape = FALSE)
-```
-
 # References

### 2020-11-06 manuscript.rmd
1. Figures refer to in the main text are presented first, then tables then SI text

@@
 To ensure model fit at 1 m depth, we first chose 300 parameter ensemble members that minimized $\textrm{RMSE}^{*}({\mathrm{\theta}_{\textrm{1 m}}}, W_{\textrm{liq,`r soil.depths[8]` m}})$ (< 0.2), and among those 300 ensemble members we chose 100 with the lowest values for the objective function $\overline{O}$.
-See Table \@ref(tab:param) for the ranges of best-fit values for different parameters.
+See Table S1 for the ranges of best-fit values for different parameters.
 
@@
 See Fig.
-S\@ref(fig:erd-hyd-full) for the full correlation matrix.
+S10 for the full correlation matrix.
 

### 2020-11-09 Supporting_Information.Rmd
1. Minor edit

@@
 
-### Stem hydraulic safety margins
+### Above-ground hydraulic safety margins
 
-Stem hydraulic safety margin was calculated as the difference between minimum leaf water potential $\mathrm{\Psi}$~min~ and $\mathrm{\Psi}$~88,stem~.
+Above-ground hydraulic safety margin was calculated as the difference between minimum leaf water potential $\mathrm{\Psi}$~min~ and $\mathrm{\Psi}$~88,stem~.
 Values for $\mathrm{\Psi}$~min~ were obtained from the dataset of @Wolfe.2019, which was collected as part of the 2016 ENSO campaign.

### 2020-11-09 manuscript.rmd
1. Minor edit

--- 2020-11-06 manuscript.rmd
@@
 Summary | 197  
-Main Text  | 6487  
-Introduction | 832  
-Methods | 3277  
-Results | 876  
-Discussion | 1502  
-Keywords | 6
+Main Text  | 6491  
+Introduction | 833  
+Methods | 3285  
+Results | 873  
+Discussion | 1500  
+Keywords | 8
 
@@
 Such efforts have found correlations between vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013, @Venturas.2016].
-Even though deep roots may mitigate hydraulic vulnerability and mortality-risk, hydraulic risk realized through hydrological (rather than meteorological) droughts--the outcome of the interaction between rooting depths, critical hydraulic thresholds and the actual vertical distribution of plant-available water--are not studied, especially in tropical regions.
+Even though deep roots may mitigate hydraulic vulnerability and mortality-risk, hydraulic-risk realized through hydrological (rather than meteorological) droughts--the outcome of the interaction between rooting depths, critical hydraulic thresholds and the actual vertical distribution of plant-available water--are not studied, especially in tropical regions.
 
@@
 
-To obtain the daily dynamics of soil water potential in the whole-soil column, we locally parameterized and calibrated a 1-D hydrological water balance model.
-We then developed a novel, empirical inverse model to estimate species effective rooting depths (ERD) for large trees of `r length(erd.sp)` species (Fig.
+To obtain the daily dynamics of soil water potential in the whole-soil column, we locally parameterized and calibrated a 1-D hydrological water balance model (Fig.
 \@ref(fig:schematic)).
+We then developed a novel, empirical inverse model to estimate species effective rooting depths (ERD) for large trees of `r length(erd.sp)` species.
 We define ERD as the depth at which soil water potential best explained species-specific growth dynamics.
@@
 We tested the relationship of ERD with xylem vulnerability to embolism, branch hydraulic safety margins, leaf turgor loss point and maximum stem-specific hydraulic conductivity.
-Finally, to test whether ERD explained risk of drought-induced mortality, we analysed species-specific large tree mortality dynamics over 35 years and estimated species-level realized hydraulic risk as the time a species’ spent beyond its critical hydraulic threshold.
+Finally, to test whether ERD explained risk of drought-induced mortality, we analysed species-specific large tree mortality dynamics over 35 years and estimated species-level realized hydraulic-risk as the time a species’ spent beyond its critical hydraulic threshold.
 
@@
 VPD affects growth non-linearly, with growth increasing with VPD up-to a VPD threshold, then decreasing as leaf pores (stomata) close reducing water-uptake [@Grossiord.2020; @Yang.2019].
-To account for the non-linear impact of VPD on growth, we used predicted gross primary productivity (GPP; hereafter, $\hat{\textrm{VPD}}$) from a locally derived polynomial relationship between GPP and VPD (see Fig.
-S2 and SI).
+To account for the non-linear impact of VPD on growth, we used predicted gross primary productivity (GPP; hereafter, $\hat{\textrm{VPD}}$) from a locally derived polynomial relationship between GPP and VPD (see SI and Fig.
+S2).
 Apart from stomatal control, leaf-drop may further limit water-uptake and growth.
@@
 \@ref(fig:erd-hyd)**d**).
-See Fig.
-S10 for the full correlation matrix.
 
@@
 While our $\mathrm{\Psi}_{\textrm{soil}}$ estimates did not extend prior to 1990, due to a lack of reliable climate forcing, a prior study found that during the period 1985-1990, for which ERD explained significant mortality, BCI experienced three extreme deficit water years, following the 1983 El-Niño year [@Condit.2017].
-Analysis of a species' realized hydraulic risk as the time spent under significant hydrological drought (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) showed that species with shallower ERD experienced greater realized hydraulic risk and increasingly so over 2000-2005 and 2005-2010 (Fig.
+Analysis of a species' realized hydraulic-risk as the time spent under significant hydrological drought (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) showed that species with shallower ERD experienced greater realized hydraulic-risk and increasingly so over 2000-2005 and 2005-2010 (Fig.
 \@ref(fig:erd-time)).
-For the census interval 1990-1995, species with the shallowest ERD also showed greater realized hydraulic risk (Fig.
+For the census interval 1990-1995, species with the shallowest ERD also showed greater realized hydraulic-risk (Fig.
 \@ref(fig:erd-time)), but during the dry season ( Fig.
@@
 Our novel approach for estimating ERD using 25 years of tree growth, species-specific leaf vulnerability curves, modeled soil moisture profiles, and observed VPD was consistent with estimates using $\mathrm{\delta}$^2^H~xylem~ (Figs.
-\@ref(fig:sp-erd), \@ref(fig:erd-iso)).
+\@ref(fig:erd-iso), \@ref(fig:sp-erd)).
 Our analyses revealed that co-occurring large canopy tree species with deeper ERD were associated with higher above-ground hydraulic efficiency, but not safety (Fig.
@@
 \
-While our ERD model empirically predicts growth via estimating $\mathrm{\Psi}$~leaf~ from $\mathrm{\Psi}$~soil~ of a specific depth, mechanistic models that account for plant hydrodynamics and other processes influencing growth are likely to predict growth more accurately, and thus ERD and hydraulic risk [@Christoffersen:2016kd2d; @Sperry.1998; @Yang.2019; @Duursma:2018].
+While our ERD model empirically predicts growth via estimating $\mathrm{\Psi}$~leaf~ from $\mathrm{\Psi}$~soil~ of a specific depth, mechanistic models that account for plant hydrodynamics and other processes influencing growth are likely to predict growth more accurately, and thus ERD and hydraulic-risk [@Christoffersen:2016kd2d; @Sperry.1998; @Yang.2019; @Duursma:2018].
 Although data needs for parameterizing such models could be greater (e.g.
@@
 \@ref(eq:vc-exp)) vs.
-those that were predicted from a set of models based on trait-proxies.
-(Eqs.
+those that were predicted from a set of models based on trait-proxies (Eqs.
 \@ref(eq:a-mod)-\@ref(eq:b-mod)).

### 2020-11-09 code/load.R

@@ -288,7 +288,7 @@ symbols.table <-
       "$\\mathrm{\\Psi}$~leaf~ at 50% loss of leaf conductance",
       "$\\mathrm{\\Psi}$~stem~ at 88% loss of stem conductivity",
       "Seasonal minimum water potential, the most negative $\\mathrm{\\Psi}$~leaf~ measured at midday in the dry season",
-      "Above-ground Hydraulic Safety Margin",
+      "Above-ground hydraulic safety margin",
       "Leaf-area specific hydraulic conductance of leaf",
       "Maximum leaf area-specific hydraulic conductance of leaf",
       "Maximum stem area-specific hydraulic conductivity of stem",

### 2020-11-09 code/manuscript_data_figs.R

@@ -232,7 +232,7 @@ erd.sp.plot <- ggplot(df.erd.to.plot,
         axis.text.y = element_text(face = "plain")) +
   scale_y_continuous(trans = "reverse", breaks = unique(ml.rsq.combine$depth))
 ggsave("ERD_by_sp_large_canopy.tiff",
-       plot = erd.sp.plot, file.path(figures.folder), device = "tiff", height = 3.5, width = 5, units='in')
+       plot = erd.sp.plot, file.path(figures.folder), device = "tiff", height = 4.5, width = 5, units='in')
 ggsave("ERD_by_sp_large_canopy.jpeg",
        plot = erd.sp.plot, file.path(figures.folder), device = "jpeg", height = 4.5, width = 5, units='in')
 
@@ -281,6 +281,8 @@ p4 <- ggplot(ml.rsq.combine.chosen,
   theme(legend.text = element_text(face = "italic", size = 8))
 ggsave("psi.corr_best.depth_xylem_sap_deltaD_phenology_Meinzer_gr.Psi.VPD.jpeg",
        plot = p4, file.path(figures.folder), device = "jpeg", height = 3, width = 4.3, units = 'in')
+ggsave("psi.corr_best.depth_xylem_sap_deltaD_phenology_Meinzer_gr.Psi.VPD.tiff",
+       plot = p4, file.path(figures.folder), device = "tiff", height = 3, width = 4.3, units = 'in')
 
 
 ## as "gr.Psi.VPD.add" is not present

### 2020-11-10 manuscript.rmd
Printe references where needed

@@
 # References
+<div id="refs"></div>
 
@@
 ```{r hydro-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-hydro-calib}")}
-knitr::include_graphics(file.path("figures/hydro-calib2.jpeg"))
+knitr::include_graphics(file.path("figures/2019-10-14_5000/best-fits/hydro-calib2.jpeg"))
 ```

### 2020-11-10 code/02.0_Rooting_profiles.R
1. Removed manuscripts non-essential files 2. Renumbered file sequentially

@@ -1,431 +1,0 @@
-
-#-------------------------------
-# Title: Generate Rooting profiles
-# Author : Rutuja Chitra-Tarak
-# Original date: 2019-12-20
-#-------------------------------
-
-rm(list=ls())
-
-if (!require("pacman")) install.packages("pacman"); library(pacman)
-pacman::p_load(ncdf4, easyNCDF, tidyverse)
-
-require(scales);
-rev_sqrt_trans <- function() {
-  scales::trans_new(
-    name = "rev_sqrt",
-    transform = function(x) -sqrt(abs(x)),
-    inverse = function(x) x^2);
-}
-
-# graphics info
-theme_set(theme_bw())
-theme_update(text = element_text(size=14),
-             panel.grid.major = element_blank(),
-             panel.grid.minor = element_blank(),
-             strip.background = element_blank()
-)
-
-file.path.figures <- file.path("figures", "rooting_profiles")
-if(!dir.exists(file.path.figures)) {dir.create(file.path.figures)}
-####--------------------------------------------
-####      Rooting profiles for inversion
-####--------------------------------------------
-
-####--------------------------------------------
-## --------------in ELM-FATES-------------------
-####--------------------------------------------
-# FATES uses equation (2) in Zeng et al.
-## y = 1 - 0.5*(exp(-a*d) + exp(-b*d)) ....(eq 1)
-# where d = depth and Y = root fraction, a = fates_roota_par, b = fates_rootb_par
-
-# fates_roota_par: Parameter a in Eq. 2 in (Zeng 2001) that regulates the shape of the rooting profile.
-# Range 5.9 - 7.4 per m
-# Range corresponds to this parameter specified for BATS (or IGBP) land cover classified
-# Deciduous Broadleaf Trees (5.9 m-1) and Evergreen Broadleaf Trees  (7.4 m-1) as given in
-# Table 1 (or 2) of Zeng 2001.
-
-# fates_rootb_par : Parameter b in Eq. 2 in (Zeng 2001) that regulates the depth of the rooting profile.
-# Chosen range of b is derived using this equation so as to fit the observed range of rooting depth (dr)
[…]

### 2020-11-10 code/03.0_GLUEsetup_part1_BCI.R

@@ -1,373 +1,0 @@
-
-#------------------------------------------------------------------------------
-# Title: Collate swp data, rooting profiles, generated btran and census dates
-# Author : Rutuja Chitra-Tarak
-# Original date: December 18, 2019
-#------------------------------------------------------------------------------
-
-rm(list=ls())
-for (i in 1:10) {gc(reset = TRUE)}
-graphics.off()
-if (!require("pacman")) install.packages("pacman"); library(pacman)
-pacman::p_load(tidyverse, doParallel, foreach, data.table)
-
-GLUEsetup_part1 <- function(current.folder = current.folder, intervals = intervals, nsam = nsam) {
-  # current.folder = "2019-10-14_5000"; intervals = 5; nsam = 200
-  if(!dir.exists(file.path("results"))) {dir.create(file.path("results"))}
-  if(!dir.exists(file.path("results", current.folder))) {dir.create(file.path("results", current.folder))}
-
-  ####-------------------------------------
-  #### 1. To enable hydro time series into intervals
-  ####-------------------------------------
-  ##
-  # # Storing census med and interval start and end dates--------------
-  # load("data-raw/CTFScensuses/bci.tree1.Rdata")
-  # load("data-raw/CTFScensuses/bci.tree2.Rdata")
-  # load("data-raw/CTFScensuses/bci.tree3.Rdata")
-  # load("data-raw/CTFScensuses/bci.tree4.Rdata")
-  # load("data-raw/CTFScensuses/bci.tree5.Rdata")
-  # load("data-raw/CTFScensuses/bci.tree6.Rdata")
-  # load("data-raw/CTFScensuses/bci.tree7.Rdata")
-  # load("data-raw/CTFScensuses/bci.tree8.Rdata")
-  # head(bci.tree1)
-  #
-  # census.years <- c(1982, 1985, 1990, 1995, 2000, 2005, 2010, 2015)
-  # bci.tree1$census <- census.years[1]
-  # bci.tree2$census <- census.years[2]
-  # bci.tree3$census <- census.years[3]
-  # bci.tree4$census <- census.years[4]
-  # bci.tree5$census <- census.years[5]
-  # bci.tree6$census <- census.years[6]
-  # bci.tree7$census <- census.years[7]
-  # bci.tree8$census <- census.years[8]
-  #
-  # bci.tree <- rbind.data.frame(bci.tree1, bci.tree2, bci.tree3, bci.tree4, bci.tree5, bci.tree6, bci.tree7, bci.tree8)
[…]

### 2020-11-10 code/04.0_GLUEsetup_part2_BCI.R

@@ -1,75 +1,0 @@
-#----------------------------------------------
-# Title: Collate growth and btran data and metadata
-# Author : Rutuja Chitra-Tarak
-# Original date: Feb 18, 2019
-#----------------------------------------------
-
-rm(list = ls())
-gc()
-graphics.off()
-if (!require("pacman")) install.packages("pacman"); library(pacman)
-pacman::p_load(tidyverse)
-
-GLUEsetup_part2 <- function(growth.type, dbh.residuals, solar.residuals, growth.selection) { #n.ensembles = 5000
-  # n.ensembles = 100000,  #which ensemble set to use
-    #growth.selection  <- "size_class_predefined_cc_scaled"
-    load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
-    intervals <- info$intervals
-    ## load Result of 3.GLUErun_part1_BCI.R:
-    load("results/GLUEsetup_part1.3_BCI.RData")
-    n.ensembles <- length(info.3$btran.int.sam[[1]])
-    ## Observed growth
-    ## individual, species or sp by size level
-
-    growth.name <- load(file =  paste0("results/sp_size.", growth.type, "_growth_dbh.residuals_", dbh.residuals, "_ci_", intervals, "_", growth.selection, ".Rdata"))
-    growth <- get(growth.name); rm(growth.name)
-    sp_size.name <- load(file = paste0("results/sp_size.", growth.type, ".names_", intervals, "_", growth.selection, ".Rdata"))
-    growth.meta <- data.frame(sp_size = get(sp_size.name)) %>%
-      separate(sp_size, into = c("sp", "size"), sep = "_", remove = FALSE)
-    rm(sp_size.name)
-
-    # growth.type <- "med_growth_sp_by_size"
-    # load(paste0("results/", growth.type, "_intervals_", intervals ,".Rdata"))
-    # growth.long <- med_growth_sp_by_size %>% unite("sp_size", sp, size, remove = TRUE) %>% select(-interval.2) %>%
-    #   pivot_longer(cols = starts_with("interval."), names_to = "interval", names_prefix = "interval.", values_to = "growth") %>% mutate(interval = as.numeric(interval))
-    # growth.prep <- split(growth.long, growth.long$sp_size)
-    # growth <- lapply(growth.prep, function(x) {x %>% select(-sp_size)})
-    #
-    # sp_size.median.names <- names(growth)
-    # growth.meta <- data.frame(sp_size = sp_size.median.names) %>%
[…]

### 2020-11-10 code/05.0_GLUErun_part2_func_BCI.R

@@ -1,165 +1,0 @@
-
-#---------------------------------
-# Title: Function to calculate various goodness of fit
-# stats between observed growth and simulated btran
-# This is sourced in GLUErun_part2_BCI.R
-# Author : Rutuja Chitra-Tarak
-# Original date: December 21, 2019
-#---------------------------------
-
-growth_by_si.func <- function(fun.version = fun.version,
-                              growth_by_si.info = growth_by_si.info,
-                              splevel = splevel,
-                              goodness.fit = goodness.fit,
-                              drop.months = drop.months,
-                              statistic = statistic) {
-
-  if (splevel == "on") {
-    sp.with.tlp <- names(info.2$sp.si[[drop.months]])
-    growth.meta <- growth_by_si.info$growth.meta %>%
-      subset(sp %in% sp.with.tlp)
-    sp_size.on  <- growth.meta$sp_size #.full %>% subset(size == "large" || size == "small") %>% select(-size, -sp, -interval.2) %>% as.matrix()
-    growth <- growth_by_si.info$growth[sp_size.on]
-    sp_size <- names(growth)
-    ### to create si for each sp_size, sp.si elements should be repeated as sps in growth
-    sp.in.sp_size <- stringr::str_split_fixed(names(growth), "_", 2)[,1]
-    ## Taking the sp.si list element for the prescribed drop.months
-    sp.si.match <- info.2$sp.si[[drop.months]][sp.in.sp_size] ## this repeats sp.si elements (with sp names) in the order of sp.in.sp_size
-    ## need to repeat elements of si according to growth[[j]]$interval
-    ## since intervals are from 3:7 (8-intervals), and so are corresponding si vec, this would mean removing 3-1 = 2
-    location.adj <- as.numeric(strsplit(colnames(sp.si.match[[1]]),".", fixed = TRUE)[[1]][2]) - 1
-    sp.si.match <- lapply(sp.si.match, unname)
-  } else {
-    growth <- growth_by_si.info$growth
-    if (drop.months != "None"){
-      # only species that are deciduous need to be retained from the growth matrix
-      deci <- read.csv("data-raw/traits/HydraulicTraits_Kunert/deciduous_species_Meakem.csv")
[…]

### 2020-11-10 code/06.1_udi.calculator.R

@@ -1,271 +1,0 @@
-
-#----------------------------------------------
-# Title: Calculate UDI & SDI for all ensembles
-# Author : Rutuja Chitra-Tarak
-# Original date: January 22, 2020
-#----------------------------------------------
-
-# .rs.restartR()
-# for 50 ha obs species groups
-rm(list = ls())
-for (i in 1:5) {gc()}
-
-if (!require("pacman")) install.packages("pacman"); library(pacman)
-pacman::p_load(tidyverse, doParallel, foreach, data.table)
-
-udi.calculator <- function(splevel = splevel, dryseason = dryseason, rsq.thresh = rsq.thresh,
-                           n.ci.thresh = n.ci.thresh, iso.subset = iso.subset, drop.months = drop.months) {
-  # load interval and n.ensembles
-  load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
-  load(file.path("results/GLUEsetup_part2_BCI.RData")) # has n.ensembles and growth and si matrix
-
-  intervals <- info$intervals
-  n.ensembles <- growth_by_si.info$n.ensembles
-  growth.type <- growth_by_si.info$growth.type ## this is not data but type such as "individual"
-  growth.selection <- growth_by_si.info$growth.selection
-  dbh.residuals <- growth_by_si.info$dbh.residuals
-  si.type <- growth_by_si.info$si.type
-  si.param.rel <- growth_by_si.info$si.param.rel[[drop.months]]
-  ncor <- detectCores() - 1
-
-  if (splevel == "on") {
-    level.folder <- "splevel"
-    } else {
-    level.folder <- "commlevel"
-    }
-
-  rf.sam.vec <- unique(si.param.rel$rf.sam)
-  n.par.sam <- length(unique(si.param.rel$par.sam)) ## this is the sequence repeated in growth_by_si.info$si.param.rel
-  par.sam.vec <- si.param.rel$par.sam[1:n.par.sam] ## this is the sequence used when GLUE.rsq was constructed
-  n.rf.sam <- length(rf.sam.vec)
-  rm(info); rm(growth_by_si.info)
-  ## sdt is chosen based on dryseason on/off
-  file.extension.base1 <- paste0("drop.months", drop.months, "_cor", rsq.thresh, "_", si.type, "_", n.ensembles, "_",
[…]

### 2020-11-10 code/07.0_summarising_udi_by_sp_size.R

@@ -1,133 +1,0 @@
-
-#---------------------------------
-# Title: Calculates summary UDI & SDI and rooting profiles by sp_size
-# Author : Rutuja Chitra-Tarak
-# Original date: January 22, 2020
-#--------------------------------
-
-rm(list = ls())
-gc()
-if (!require("pacman")) install.packages("pacman"); library(pacman)
-pacman::p_load(tidyverse, data.table)
-
-summarise.udi <- function(splevel = splevel, dryseason = dryseason, rsq.thresh = rsq.thresh,
-                          root.selection = root.selection, iso.subset = iso.subset,
-                          n.rank = n.rank, drop.months = drop.months) {
-  load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
-  load(file.path("results/GLUEsetup_part2_BCI.RData")) # has n.ensembles and growth and si matrix
-
-  intervals <- info$intervals
-  rm(info)
-  n.ensembles <- growth_by_si.info$n.ensembles
-  growth.type <- growth_by_si.info$growth.type ## this is not data but type such as "individual"
-  si.type <- growth_by_si.info$si.type
-  growth.selection <- growth_by_si.info$growth.selection
-  dbh.residuals <- growth_by_si.info$dbh.residuals
-  rsq.thresh <-  rsq.thresh
-  if (splevel == "on") {
-    level.folder <- "splevel"
-  } else {
-
-    level.folder <- "commlevel"
-  }
-  file.extension.base3 <- paste0("drop.months", drop.months, "_cor", rsq.thresh, "_",
-                                 si.type, "_", n.ensembles, "_", growth.type, "_",
-                                 growth.selection, "_", dbh.residuals, "_", intervals,
-                                 "_dryseason_", dryseason, "_iso.subset_", iso.subset)
-  file.extension.base4 <- paste0(file.extension.base3, "_root.selection_", root.selection)
-
-  load(file = paste0("results/", level.folder, "/ds.bestfit.all_", file.extension.base3, ".Rdata"))
-
-
-  if (root.selection == "on") {
[…]

### 2020-11-10 code/08.0_Sensitivity analysis for UDI of deciduous species copy.R

@@ -1,395 +1,0 @@
-
-#-----------------------------------------------------
-# Title: Sensitivity analysis for deciduous species:
-#         Effect of dry season month exclusion on UDI
-# Author : Rutuja Chitra-Tarak
-# Original date: April 11, 2020
-#-----------------------------------------------------
-
-rm(list=ls())
-gc()
-
-pacman::p_load(tidyverse, readxl, gridExtra, magrittr, multipanelfigure)
-# graphics info
-theme_set(theme_bw())
-theme_update(text = element_text(size = 14),
-             panel.grid.major = element_blank(),
-             panel.grid.minor = element_blank(),
-             strip.background = element_blank()
-)
-
-rev_sqrt_trans <- function() {
-  scales::trans_new(
-    name = "rev_sqrt",
-    transform = function(x) -sqrt(abs(x)),
-    inverse = function(x) x^2)
-}
-reverselog_trans <- function(base = exp(1)) {
-  scales::trans_new(name = paste0("reverselog-", format(base)),
-                    log_breaks(base = base),
-                    domain = c(1e-100, Inf),
-                    transform = function(x) -log(x, base),
-                    inverse = function(x) base^(-x))
-}
-sensitivity.analysis <- function(goodness.fit = goodness.fit,
-                        dryseason.vec = dryseason.vec,
-                        root.selection = root.selection,
-                        iso.subset = iso.subset) {
-
-deci <- read_excel(file.path("data-raw/traits/nomenclature_R_20190524_Rready_Osvaldo Calderon & JoeWright_expert_opinion.xlsx")) %>%
-  mutate(sp = tolower(sp6)) %>% select(sp, deciduous, tree)
-tlp <- read.csv(file.path("data-raw/traits/HydraulicTraits_Kunert/tlp_sp_mean.csv")) %>%
-  subset(!is.na(tlp))
-
-# load interval and working.iter
-load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
-load(file.path("results/GLUEsetup_part2_BCI.RData")) # has working.iter and growth and si matrix
-
-soil.depths <- unique(info$root.param.long$depth)
-intervals <- info$intervals
[…]

### 2020-11-10 code/08.0_Sensitivity analysis for UDI of deciduous species.R

@@ -1,215 +1,0 @@
-
-#-----------------------------------------------------
-# Title: Sensitivity analysis for deciduous species:
-#         Effect of dry season month exclusion on UDI
-# Author : Rutuja Chitra-Tarak
-# Original date: April 11, 2020
-#-----------------------------------------------------
-
-rm(list=ls())
-gc()
-
-pacman::p_load(tidyverse, readxl)
-sensitivity.analysis <- function(goodness.fit = goodness.fit,
-                        dryseason = dryseason,
-                        root.selection =  root.selection,
-                        iso.subset = iso.subset, drop.months = drop.months){
-
-deci <- read_excel(file.path("data-raw/traits/nomenclature_R_20190524_Rready_Osvaldo Calderon & JoeWright_expert_opinion.xlsx")) %>%
-  mutate(sp = tolower(sp6)) %>% select(sp, deciduous, tree)
-tlp <- read.csv(file.path("data-raw/traits/HydraulicTraits_Kunert/tlp_sp_mean.csv")) %>%
-  subset(!is.na(tlp))
-
-# load interval and working.iter
-load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
-load(file.path("results/GLUEsetup_part2_BCI.RData")) # has working.iter and growth and si matrix
-
-soil.depths <- unique(info$root.param.long$depth)
-intervals <- info$intervals
-n.ensembles <- growth_by_si.info$n.ensembles
-growth.type <- growth_by_si.info$growth.type
-growth.selection <- growth_by_si.info$growth.selection
-dbh.residuals <- growth_by_si.info$dbh.residuals
-si.type <- growth_by_si.info$si.type
-##
-level.folder <- c("splevel", "commlevel")
-
-max.ll.list <- vector(mode = "list", length = length(level.folder))
-names(max.ll.list) <- level.folder
-load(file.path("results/GLUEsetup_part1.2_BCI.RData"))
-drop.months.vec <- names(info.2$sp.si)
-
-for (j in 1: length(level.folder)){
-  for (i in 1: length(drop.months.vec)) {
-    file.extension.base1 <- paste0("drop.months", drop.months.vec[i], "_cor", goodness.fit, "_", si.type, "_", n.ensembles, "_",
[…]

### 2020-11-10 code/09.1_Comparison_with_isotopic_data.R

@@ -1,776 +1,0 @@
-## Comparison with Meinzer et al. 1999 results.
-
-rm(list=ls())
-gc()
-
-iso.compare <- function(goodness.fit = goodness.fit,
-                         dryseason = dryseason,
-                         root.selection =  root.selection,
-                         iso.subset = iso.subset, drop.months = drop.months){
-  pacman::p_load(tidyverse, scales, ggpmisc)
-  # graphics info
-  theme_set(theme_bw())
-  theme_update(text = element_text(size = 14),
-               panel.grid.major = element_blank(),
-               panel.grid.minor = element_blank(),
-               strip.background = element_blank()
-  )
-
-  rev_sqrt_trans <- function() {
-    scales::trans_new(
-      name = "rev_sqrt",
-      transform = function(x) -sqrt(abs(x)),
-      inverse = function(x) x^2);
-  }
-
-  # load interval and working.iter
-  load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
-  load(file.path("results/GLUEsetup_part2_BCI.RData")) # has working.iter and growth and si matrix
-
-  soil.depths <- unique(info$root.param.long$depth)
-  intervals <- info$intervals
-  n.ensembles <- growth_by_si.info$n.ensembles
-  growth.type <- growth_by_si.info$growth.type
-  growth.selection <- growth_by_si.info$growth.selection
-  dbh.residuals <- growth_by_si.info$dbh.residuals
-  si.type <- growth_by_si.info$si.type
-  ##
-
-  file.extension.base4 <- paste0("drop.months", drop.months, "_cor", goodness.fit, "_", si.type, "_", n.ensembles, "_", growth.type,
-                                 "_", growth.selection, "_", dbh.residuals, "_", intervals,
-                                 "_dryseason_", dryseason, "_iso.subset_", iso.subset, "_root.selection_", root.selection)
-
-  load(file = paste0("results/splevel/ds.bestfit_", file.extension.base4, ".Rdata"))
-  ds <- ds.bestfit
[…]

### 2020-11-10 code/11.1_mortality_BCI.R

@@ -1,514 +1,0 @@
-#-----------------------------------------------------
-# Title: Analysing mortality by Uptake Depth Index
-# Author : Rutuja Chitra-Tarak
-# Original date: Mar 20, 2020
-#-----------------------------------------------------
-
-rm(list=ls())
-gc()
-# load("/Library/Frameworks/R.framework/Versions/3.4/Resources/library/CTFSRPackage/CTFSRPackage.Rdata")
-if (!require("pacman")) install.packages("pacman"); library(pacman)
-pacman::p_load(tidyverse, hms, ggpmisc)
-
-# graphics info
-theme_set(theme_bw())
-theme_update(text = element_text(size=14),
-             panel.grid.major = element_blank(),
-             panel.grid.minor = element_blank(),
-             strip.background = element_blank()
-)
-
-census.years <- c(1982, 1985, 1990, 1995, 2000, 2005, 2010, 2015)
-
-require(scales)
-rev_sqrt_trans <- function() {
-  scales::trans_new(
-    name = "rev_sqrt",
-    transform = function(x) -sqrt(abs(x)),
-    inverse = function(x) x^2);
-}
-
-####********************************************************************
-##### Load demographic data ----------------------------------------
-##### If you dont need to update the above, start directly here:
-####********************************************************************
-
-demo.graphs <- function(level.folder = level.folder, n.threshold = n.threshold,
-                        dbh.residuals = dbh.residuals,
-                        dryseason = dryseason,
-                        root.selection = root.selection,
-                        iso.subset = iso.subset) {
-
-  load("results/demo.sp.RData")
-  load("results/demo.sp_size.RData")
-  load("results/mrate.long.RData")
-  load("results/sp.mrate.long.RData")
-  load("results/adult.mrate.long.RData")
-
-  ##plotting mortality rates by sp
-  census.years.short <- format(strptime(census.years, "%Y"), "%y")
-  # mrate.long$censusint <- mapvalues(mrate.long$census, from = census.years[-1], to =  paste(census.years[-length(census.years)],census.years[-1], sep = "-"))
[…]

### 2020-11-10 code/12.0_hydraulic_traits_BCI.R

@@ -1,560 +1,0 @@
-##-------------------------
-## Date : 22 Jul 2018
-## Author : Rutuja
-## Title : Uptake depth and hydraulic traits
-##-------------------------
-## Do species with shallow uptake depth have greater drought tolerance? (more -ve TLP)
-
-rm(list = ls())
-gc()
-if (!require("pacman")) install.packages("pacman"); library(pacman)
-pacman::p_load(tidyverse, ggcorrplot, Hmisc, PerformanceAnalytics,
-               corrplot, devtools, multcompView, agricolae, readxl, ggpmisc)
-# install_github("vqv/ggbiplot")
-  # graphics info
-theme_set(theme_bw())
-theme_update(text = element_text(size=14),
-             panel.grid.major = element_blank(),
-             panel.grid.minor = element_blank(),
-             strip.background = element_blank()
-)
-require(scales)
-rev_sqrt_trans <- function() {
-  scales::trans_new(
-    name = "rev_sqrt",
-    transform = function(x) -sqrt(abs(x)),
-    inverse = function(x) x^2);
-}
-## adding uptake depth index
-load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
-load("results/GLUEsetup_part2_BCI.RData") # has n.ensembles and growth and si matrix
-load("results/demo.sp_size.RData")
-load("results/demo.sp.RData") ## Mortality rate only for those cases for which avg.abundance greater than 10
-intervals <- info$intervals
-n.ensembles <- growth_by_si.info$n.ensembles
-growth.type <- growth_by_si.info$growth.type
-growth.selection <- growth_by_si.info$growth.selection
-si.type <- growth_by_si.info$si.type
-goodness.fit <- 0.3
-soil.depths <- unique(info$root.param.long$depth)
-##
-dbh.residuals <- "on"#growth_by_si.info$dbh.residuals
-dryseason <- "on"
-root.selection <- "on"
-iso.subset <- "off"
-##
-file.extension.base4 <- paste0(goodness.fit, "_", si.type, "_", n.ensembles, "_", growth.type, "_", growth.selection,
[…]

### 2020-11-10 code/14.0_spatial_hydrological_niche.R

@@ -1,222 +1,0 @@
-##-------------------------
-## Date : 22 Jul 2018
-## Edited: Jan 23 2020
-## Author : Rutuja
-## Title : Spatial hydrological niche
-##-------------------------
-rm(list = ls())
-
-if (!require("pacman")) install.packages("pacman"); library(pacman)
-pacman::p_load(tidyverse, devtools, data.table, reshape2, fgeo, remotes)
-# remotes::install_github("forestgeo/bciex")
-# had to load XQuartz for Mac first
-# install.packages("installr")
-# available.packages("utils")
-# packageUrl<- "https://cran.r-project.org/src/contrib/Archive/R.utils/R.utils_2.6.0.tar.gz"
-# install.packages(packageUrl, repos=NULL, type='source')
-# graphics info
-theme_set(theme_bw())
-theme_update(text = element_text(size = 14),
-             panel.grid.major = element_blank(),
-             panel.grid.minor = element_blank(),
-             strip.background = element_blank()
-)
-require(scales)
-rev_sqrt_trans <- function() {
-  scales::trans_new(
-    name = "rev_sqrt",
-    transform = function(x) -sqrt(abs(x)),
-    inverse = function(x) x^2);
-}
-file.path.spatial <- file.path("figures", "spatial")
-if(!dir.exists(file.path.spatial)) {dir.create(file.path.spatial)}
-#------------------
-## map trees at BCI color coded by UDImean
-# load tree locations data
-load("data-raw/CTFScensuses/BCI.tree8.Rdata")
-# load interval and working.iter
-load("results/GLUEsetup_part1_BCI.RData") # has model info and data on obs
-load(file.path("results/GLUEsetup_part2_BCI.RData")) # has working.iter and growth and si matrix
-
-intervals <- info$intervals
-n.ensembles <- growth_by_si.info$n.ensembles
-growth.type <- growth_by_si.info$growth.type
-growth.selection <- growth_by_si.info$growth.selection
-dbh.residuals <- growth_by_si.info$dbh.residuals
-si.type <- growth_by_si.info$si.type
-goodness.fit <- 0.3
-soil.depths <- unique(info$root.param.long$depth)
-dryseason <- "on"
-root.selection <- "on"
-rm(info); rm(growth_by_si.info)
-##
[…]

### 2020-11-10 code/17.0_Extra_parts_of_PhenoDemoTraits.R

@@ -1,1793 +1,0 @@
-###---------------------------------------------------------------------------------------------------------
-### Parts of PhenoDemoTraits.R after running RED model, but not required for the manucript
-### Could be run after running PhenoDemoTraits.R
-###---------------------------------------------------------------------------------------------------------
-
-load(file = file.path(results.folder, "chosen.model.Rdata"))
-load(file = file.path(results.folder, "psi.stat.4.Rdata"))
-load(file = file.path(results.folder, "psi.stat.4.select.Rdata"))
-
-## tlp was used in an earlier version
-# tlp.sp <-  traits.long %>% subset(trait == "TLP") %>% select(sp, value) %>% rename(tlp = value)
-# n.tlp.sp <- length(tlp.sp$sp)
-# tlp.sp <- tlp.sp %>%
-#   mutate(psi_threshold = tlp*0.8)
-# tlp.sp.ls <- split(tlp.sp, f = list(tlp.sp$sp), drop = TRUE)
-
-grate.gfac <- gfac.interval.long <- vector(mode = "list", length = length(names.gfac))
-names(gfac.interval.long) <- names(grate.gfac) <- names.gfac
-for (i in 1:length(names.gfac)) {
-  gfac.interval.long[[i]] <- gfac.interval[[i]] %>%
-    pivot_longer(cols = c(-sp, -interval),
-                 names_to = "depth", values_to = "gfac") %>%
-    rename(interval.num =  interval) %>%
-    mutate(depth = as.numeric(depth),
-           interval.num =  as.numeric(as.character(interval.num)),
-           size = "large")
-  grate.gfac[[i]] <- growth.sub %>% rename(interval.num = interval) %>%
-    left_join(gfac.interval.long[[i]], by = c("interval.num", "sp", "size")) %>%
-    subset(!is.na(gfac))
-}
-
-grate.gfac.best <- dplyr::bind_rows(grate.gfac, .id = "corr.func") %>%
-  mutate(censusint.m = recode(interval.num, `1` = "1982-85", `2` = "1985-90",
-                              `3` = "1990-95", `4` = "1995-00", `5` = "2000-05", `6` = "2005-10", `7` = "2010-15")) %>%
-  transform(corr.func = factor(corr.func, levels = names.gfac)) %>%
[…]

### 2020-11-10 code/Miscellaneous/01.0_Preparing_stem_growth_time_series_for_Chonggang_FATES_HYDRO.R

@@ -1,264 +1,0 @@
-#---------------------------------
-# Title: Preparing growth data for inverse modeling
-# Author : Rutuja Chitra-Tarak
-# Original date: December 11, 2019
-#---------------------------------
-
-# for 50 ha obs species groups
-rm(list = ls())
-gc()
-if (!require("pacman")) install.packages("pacman"); library(pacman)
-pacman::p_load(tidyverse, purrrlyr, scales, grid, gridExtra, lme4)
-
-# graphics info
-theme_set(theme_bw())
-theme_update(text = element_text(size = 14),
-             panel.grid.major = element_blank(),
-             panel.grid.minor = element_blank(),
-             strip.background = element_blank()
-)
-# time series info
-years <- seq(1990, 2015, by = 5)
-ncensus  <- length(years)
-nint <- length(years) - 1
-intervals <- nint
-# sp4 <- c("ANOL", "TECG", "LAGL", "TERT")
-##------------------------------------------------
-## Collating tree.full data for all censuses
-##------------------------------------------------
-# https://repository.si.edu/handle/10088/20925
-# load("data-raw/CTFScensuses/BCI.stem1.Rdata") # 1982-1985
-load("data-raw/CTFScensuses/BCI.stem2.Rdata") # 1985-1990
-load("data-raw/CTFScensuses/BCI.stem3.Rdata")
-load("data-raw/CTFScensuses/BCI.stem4.Rdata")
-load("data-raw/CTFScensuses/BCI.stem5.Rdata")
-load("data-raw/CTFScensuses/BCI.stem6.Rdata")
-load("data-raw/CTFScensuses/BCI.stem7.Rdata")
-load("data-raw/CTFScensuses/BCI.stem8.Rdata")
-
-length(unique(bci.stem3$sp))
-# all censuses in one list
-stem.full <- bind_rows(bci.stem2, bci.stem3, bci.stem4, bci.stem5, bci.stem6, bci.stem7, bci.stem8)
-# > nrow(stem.full)
-# [1] 4740360
-# > nrow(stem.full)*6
-# [1] 28442160
-###------------selection----------
-census <- 2:8
-###-------------------------------
-stem.full$census <- rep(census, each = nrow(bci.stem3))
-## remove duplicated tags if any
-nrow(stem.full) # 4740360
-stem.full <- stem.full %>% mutate(ExactDate = as.Date(ExactDate))
[…]

### 2020-11-10 code/Miscellaneous/01_data_availability_by_species.R

@@ -1,100 +1,0 @@
-##--------------------
-## Data by species
-## Author: Rutuja
-## Date April 14, 2020
-##-------------------
-rm(list = ls())
-gc()
-if (!require("pacman")) install.packages("pacman"); library(pacman)
-pacman::p_load(tidyverse, naniar)
-
-### Demographic data -----
-load("results/demo.sp.RData")
-## keeping data only for adult
-adult.demo.sp <- demo.sp %>% select(sp, grate.adult, mrate.adult, avg.abund) %>%
-  subset(!(is.na(grate.adult) & is.na(mrate.adult))) %>% droplevels()
-View(adult.demo.sp)
-adult.demo.sp <- adult.demo.sp %>%
-  mutate(mortrate.adult = ifelse(!is.na(mrate.adult), "1", NA))
-
-### Growth data available for inversion- for adult trees----
-load(file.path("results/4.1GLUEsetup_part2_BCI.RData"))
-above10.sp.for.inversion <- growth_by_si.info$growth.meta %>% subset(size %in% c("medium"))
-above30.sp.for.inversion <- growth_by_si.info$growth.meta %>% subset(size %in% c("large"))
-growth.ts.for.inversion <- data.frame(sp = above10.sp.for.inversion$sp, growth.ts.above10cmDBH = "1") %>%
-  full_join(data.frame(sp = above30.sp.for.inversion$sp, growth.ts.above30cmDBH = "1"), by= "sp")
-growth.ts.for.inversion <- replace_with_na_all(growth.ts.for.inversion, condition = ~.x == "<NA>")
-
-data.by.sp <- adult.demo.sp %>% select(-grate.adult, -mrate.adult, -grate.adult) %>%
-  left_join(growth.ts.for.inversion, by = "sp")
-### Canopy Species-----
-## canopy or not?
-# Metadata Panama Traits Definitions.doc:
-# GRWFRM2$ - Values are as in GRWFRM1 except free-standing species can have multiple values if maximum size varies widely within Panama
-# GRWFRM1$ - Values are Climber, HERB, S, U, M and T. S, U, M and T are free-standing species with maximum heights of 5, 10, 20 and > 30 m, respectively.
-
-bci.traits <- read.csv("data-raw/traits/BCITRAITS_20101220.csv") %>%
[…]

### 2020-11-10 code/Utilities/plot.ticks.R

@@ -1,62 +1,0 @@
-###
-if (!require("pacman")) install.packages("pacman"); library(pacman)
-p_load(ggplot2, gtable, grid) # v0.2.0 for gtable
-
-# Get a plot
-plot.ticks <- function(plot.gg) {
-  p = plot.gg +
-    theme_bw() +
-    theme(panel.grid = element_blank(),
-          plot.margin = margin(1, 1, 1, 1, "cm"),
-          axis.ticks = element_line(size = 0.25),
-          axis.ticks.length = unit(-0.25, "cm"),
-          axis.text.x = element_text(vjust = 2, margin = margin(t = 0.5, unit = "cm"), size = 14),
-          axis.text.y = element_text(hjust = 1.5, margin = margin(r = 0.5, unit = "cm"), size = 14))
-
-  # Convert the plot to a grob
-  gt <- ggplotGrob(p)
-
-  # Get the position of the panel in the layout
-  panel <-c(subset(gt$layout, name=="panel", se=t:r))
-
-  ## For the bottom axis
-  # Get the row number of the bottom axis in the layout
-  rn <- which(gt$layout$name == "axis-b")
-
-  # Extract the axis (tick marks only)
-  axis.grob <- gt$grobs[[rn]]
-  axisb <- axis.grob$children[[2]]  # Two children - get the second
-  axisb  # Note: two grobs - tick marks and text
-
-  # Get the tick marks
-  xaxis = axisb$grobs[[1]]  # NOTE: tick marks first
-  xaxis$y = xaxis$y - unit(0.25, "cm")  # Position them inside the panel
-
-  # Add a new row to gt, and insert the revised xaxis grob into the new row.
-  gt <- gtable_add_rows(gt, unit(0, "lines"), panel$t-1)
-  gt <- gtable_add_grob(gt, xaxis, l = panel$l, t = panel$t, r = panel$r, name = "ticks")
-
-  ## Repeat for the left axis
-  # Get the row number of the left axis in the layout
-  panel <-c(subset(gt$layout, name=="panel", se=t:r))
-  rn <- which(gt$layout$name == "axis-l")
-
-  # Extract the axis (tick marks and axis text)
-  axis.grob <- gt$grobs[[rn]]
-  axisl <- axis.grob$children[[2]]  # Two children - get the second
-  axisl  # Note: two grobs -  text and tick marks
-
-  # Get the tick marks
[…]

### 2020-11-10 code/howmanyrefs.R

@@ -1,7 +1,0 @@
-
-
-rl <- readLines("manuscript.rmd")
-
-library(stringr)
-ref <- str_extract_all(rl, "@[a-zA-Z0-9]++")
-length(unique(do.call(c, ref)))

### 2020-11-10 code/old_code/01.0.1_Species_growth_BCI_Condit_data.R

@@ -1,871 +1,0 @@
-#------------------------------------------------------------------------------
-# Title: Old script to prepare data for inversion
-# Tries to match data preparation with Condit et al 2017, including modulus transformation of growth rates
-# Author : Rutuja Chitra-Tarak
-# Original date: February 2, 2019
-#------------------------------------------------------------------------------
-
-# for 50 ha obs species groups
-rm(list = ls())
-source(file = "R/utilities.R")
-if (!require("pacman")) install.packages("pacman"); library(pacman)
-pacman::p_load(tidyverse, purrrlyr, scales, grid, gridExtra)
-# graphics info
-tex <- element_text(size = 16, face = "plain") # , family = "gara"
-tex <- element_text(size = 5, face = "plain")
-my.theme <-  theme(axis.text = tex, axis.title = tex,
-                   title = tex, legend.title = tex, legend.text = tex, strip.text.y = tex, strip.text.x = tex)
-my.bg <- theme_bw() + theme(panel.grid.major.y = element_blank(), panel.grid.major.x = element_line(colour = "black", size = 0.25),
-                            panel.grid.minor = element_blank())
-my.adjust <- theme(axis.title.y = element_text(vjust = 1), axis.title.x = element_text(vjust = -0.6), title = element_text(vjust = 2))
-
-# time series info
-years <- c(1982, seq(1985, 2015, by = 5))
-ncensus  <- length(years)
-nint <- length(years) - 1
-# sp4 <- c("ANOL", "TECG", "LAGL", "TERT")
-##------------------------------------------------
-## Collating tree.full data for all censuses
-##------------------------------------------------
-## All data frames have same number of rows. Each row represent a tree,
-## and a given row number across censuses represent the same tree
-load("data-raw/CTFScensuses/BCI.tree1.Rdata")
-load("data-raw/CTFScensuses/BCI.tree2.Rdata")
-load("data-raw/CTFScensuses/BCI.tree3.Rdata")
-load("data-raw/CTFScensuses/BCI.tree4.Rdata")
-load("data-raw/CTFScensuses/BCI.tree5.Rdata")
-load("data-raw/CTFScensuses/BCI.tree6.Rdata")
-load("data-raw/CTFScensuses/BCI.tree7.Rdata")
-load("data-raw/CTFScensuses/BCI.tree8.Rdata")
-
-length(unique(bci.tree1$sp))
-# all censuses in one list
[…]

### 2020-11-10 code/09.0_manuscript.rmd
updated sourced filenames

--- 2020-11-10 manuscript.rmd
@@
 options(tinytex.verbose = TRUE)
-source("code/load.R")
-source("code/manuscript_data_figs.R")
+source("code/07.0_load.R")
+source("code/08.0_manuscript_data_figs.R")
 # load("results/manuWorkSpace.RData")

### 2020-11-10 code/10.0_Supporting_Information.Rmd
1. Removed manuscripts non-essential files 2. Renumbered file sequentially
updated sourced filenames

--- 2020-11-09 Supporting_Information.Rmd
@@
 
-```{r setup, echo = FALSE}
+```{r setup2, echo = FALSE}
 library(knitr)

### 2020-11-10 code/06.0_PhenoDemoTraitsPsi.R

@@ -16,7 +16,7 @@ rm(list=ls())
 #******************************************************
 ### Load data -------
 #******************************************************
-source("code/load.R")
+source("code/07.0_load.R")
 
 #*******************************************
 ####   Load Libraries, Prep for graphics, folders  ##

### 2020-11-10 code/11.0_Readme.Rmd
readme moved to the code

---
title: "Workflow of Inverse Model for Rooting Profiles"
author: "Rutuja"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, include = TRUE, eval = FALSE)
```

# Model Setup

#### Pre-requisite data

**Growth data**

Calculate species-size wise growth time series from forest inventory data. Calculate growth residuals after removing effect of size.

```{r Chunk 01.0, eval = FALSE}
source("code/01.0_Preparing_stem_growth_time_series.R")
```

**Water Availability**

Soil Water Potential data -- best-fit 100 ensembles from ELM-FATES during the study period.
This is sourced from data-package `bci.elm.fates.hydro` with library(bci.elm.fates.hydro)

#### Potential rooting profile trait-space

```{r Chunk 02.0, eval = FALSE}
source("code/02.0_Rooting_profiles.R")
```

#### Calculate btran and organise

Gather all pre-requisites in a place or two

1. `GLUEsetup_part1() in 03.0_GLUEsetup_part1_BCI.R`

 * Loads best.fit ensembles of Soil Water Potential (SWP) from bci.elm.fates.hydro::psi of a specific version (given by the "current folder")
 * Loads hypothetical rooting profiles generated by R/3.090_root_parameters_b.R
 * Loads Turgor Loss Point (TLP) data
 * Loads tree growth census dates for specific no. of intervals
 * Converts swp into a growth factor (range 0-1) using a step-fuction, such that growth factor is 0 at sp-level or community mean level Turgor Loss Point. Species wise growth factor is stores separately (sp.hydro.output.chosen)

Then using a combination of a range of hypothetical root profiles obtained from 2.rooting_profiles.R and SWP based growth factor ensemble simulations generate a btran time-series using equation : Btran= i=1 to i=z RootFrac_z * f(swp_z),
where f(swp_z) is a growth.factor_z based on a linear step relationship with swp_z ap species specific or community mean TLP level.

 * Calculates btran time-series by census-interval for each rooting profile for all the best.fit SWP ensemble
  + by individual species with species-specific TLP (N of btran time series == No. of species x No. of SWP ensembles x No. of root profiles)
  + at community-mean TLP (No. of btran time series == No. of SWP ensembles x No. of root profiles)
  + both community and species level btran is calculated by dropping none or one or more number of dry season months as sublists
  * Saves ELM-FATES hydro output, rooting profiles, census dates, number of intervals in a list called info, community level btran in a list named info.3, and species level btran is saved in info.2, along with sub-list names of dropped months

```{r Chunk 03.0, eval = FALSE}
source("code/03.0_GLUEsetup_part1_BCI.R")
GLUEsetup_part1(current.folder = "2019-10-14_5000", intervals = 5, nsam = 200) # no.of root profiles
## On macbook paralle processing to generate sp-level btran by months crashes Rstudio. Need to serialise and split job on two computers, then stitch info.2 together. 
```

2. `GLUEsetup_part2() in 04.0_GLUEsetup_part2_BCI.R`

This function stores together growth data, growth meta-data, and btran data (at community mean TLP level by dropped months) in a list named growth_by_si.info.

```{r Chunk 04.0, eval = FALSE}
source("code/04.0_GLUEsetup_part2_BCI.R")
GLUEsetup_part2(growth.type = "med", dbh.residuals = "on", solar.residuals = "off", 
                growth.selection = "size_class_predefined_cc_scaled") #"size_class_varying_non_cc")
#n.ensembles = 100000
# growth.type = "med"; dbh.residuals = "on"; solar.residuals = "off"; growth.selection = "size_class_predefined_cc_scaled"
```

# Model Parameterisation

#### Run model for all ensembles

`GLUE_run() in 05.1_GLUErun_part2_BCI.R`

This function generates a few statistics for the fit between growth and btran:

Rsq, residuals, sum of negative log likelihood and the number of data points of growth factor that pass through 95% confidence interval of growth. TLP level and any months to drop while calculating growth factor need be specified.  Threshold (goodness.fit) as to when to calculate Rs value also needs to be specified.

```{r Chunk 05.1, eval = FALSE}
# This sources "code/4.2GLUErun_part2_func_BCI.R"
source("code/05.1_GLUErun_part2_BCI.R")
# splevel = "on"; goodness.fit = 0.3; drop.months = "None"
GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "None")
GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "Jan")
GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "Feb")
GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "Mar")
GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "JanFeb")
GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "FebMar")
GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "JanFebMar")
GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "None")
GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "Jan")
GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "Feb")
GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "Mar")
GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "JanFeb")
GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "FebMar")
GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "JanFebMar")
```

#### Get water availability and btran during a specific period

`prep.udi.calculator() in 06.0_prep.udi.calculator.R`

This subsets SWP-based growth factor data (species-specific or community level) to various specified time period or months: late dryseason (Mar Apr) 1997, Jan 1997, dryseason 1992 etc.
For example, restricted to early and late 1997 dry season to later calculate UDI only for this period to compare with isotopic tree Xylem sap data that is collected in early and later 1997.

```{r Chunk 06.0, eval = FALSE}
## These need not be updated if rooting profile space changes, but only if hydrology changes or dropped months; or the period one is interested in finding UDI for.
source("code/06.0_prep.udi.calculator.R")
# splevel = "on"; drop.months = "JanFebMar"
prep.udi.calculator(splevel = "on", drop.months = "None")
prep.udi.calculator(splevel = "on", drop.months = "Jan")
prep.udi.calculator(splevel = "on", drop.months = "Feb")
prep.udi.calculator(splevel = "on", drop.months = "Mar")
prep.udi.calculator(splevel = "on", drop.months = "JanFeb")
prep.udi.calculator(splevel = "on", drop.months = "FebMar")
prep.udi.calculator(splevel = "on", drop.months = "JanFebMar")

prep.udi.calculator(splevel = "off", drop.months = "None")
prep.udi.calculator(splevel = "off", drop.months = "Jan")
prep.udi.calculator(splevel = "off", drop.months = "Feb")
prep.udi.calculator(splevel = "off", drop.months = "Mar")
prep.udi.calculator(splevel = "off", drop.months = "JanFeb")
prep.udi.calculator(splevel = "off", drop.months = "FebMar")
prep.udi.calculator(splevel = "off", drop.months = "JanFebMar")
```

#### Calculate water Uptake Depth Index for a specific period

`udi.calculator() in 06.1_udi.calculator.R`

This calculates Uptake Depth Index and SDI (water Stress Depth Index, without taking into account total water availability) for each simulation and stores corresponding values of model Rsq, negative loLikelihood, number of data points that pass through confidence interval of growth, species and size in one data-frame named "ds.bestfit.all.Rdata"

UDI or SDI is only calculated for a simulation for which R-sq is greater than the given therehold (e.g. rsq.thresh = 0.3) and if at least 3 data-points have passed through 95% CI of growth data calculated based on (e.g. n.ci.thresh = 3)

```{r Chunk 06.1, eval = FALSE}
source("code/06.1_udi.calculator.R")
##---needs to be run manually on macbook for splevel = "off" and iso.subset = "off"---
# udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.5, iso.subset = "on") ## udi is not calculated below this rsq.thresh
# splevel = "off"; dryseason = "on"; rsq.thresh = 0.3;  n.ci.thresh = 3; iso.subset = "on"; drop.months = "None"
udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.3,
               n.ci.thresh = 3, iso.subset = "off", drop.months = "None")
udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.3,
               n.ci.thresh = 3, iso.subset = "off", drop.months = "Jan")
udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.3,
               n.ci.thresh = 3, iso.subset = "off", drop.months = "Feb")
udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.3,
               n.ci.thresh = 3, iso.subset = "off", drop.months = "Mar")
udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.3,
               n.ci.thresh = 3, iso.subset = "off", drop.months = "JanFeb")
udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.3,
               n.ci.thresh = 3, iso.subset = "off", drop.months = "FebMar")
udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.3,
               n.ci.thresh = 3, iso.subset = "off", drop.months = "JanFebMar")

udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.3,
               n.ci.thresh = 3, iso.subset = "off", drop.months = "None")
udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.3,
               n.ci.thresh = 3, iso.subset = "off", drop.months = "Jan")
udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.3,
               n.ci.thresh = 3, iso.subset = "off", drop.months = "Feb")
udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.3,
               n.ci.thresh = 3, iso.subset = "off", drop.months = "Mar")
udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.3,
               n.ci.thresh = 3, iso.subset = "off", drop.months = "JanFeb")
udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.3,
               n.ci.thresh = 3, iso.subset = "off", drop.months = "FebMar")
udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.3,
               n.ci.thresh = 3, iso.subset = "off", drop.months = "JanFebMar")

udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.3,
               n.ci.thresh = 3, iso.subset = "off", drop.months = "None")
udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.3,
               n.ci.thresh = 3, iso.subset = "off", drop.months = "Jan")
udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.3,
               n.ci.thresh = 3, iso.subset = "off", drop.months = "Feb")
udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.3,
               n.ci.thresh = 3, iso.subset = "off", drop.months = "Mar")
udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.3,
               n.ci.thresh = 3, iso.subset = "off", drop.months = "JanFeb")
udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.3,
               n.ci.thresh = 3, iso.subset = "off", drop.months = "FebMar")
udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.3,
               n.ci.thresh = 3, iso.subset = "off", drop.months = "JanFebMar")

udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, 
               n.ci.thresh = 3, iso.subset = "off", drop.months = "None")
udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, 
               n.ci.thresh = 3, iso.subset = "off", drop.months = "Jan")
udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.3,
               n.ci.thresh = 3, iso.subset = "off", drop.months = "Feb")
udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.3,
               n.ci.thresh = 3, iso.subset = "off", drop.months = "Mar")
udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, 
               n.ci.thresh = 3, iso.subset = "off", drop.months = "JanFeb")
udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.3,
               n.ci.thresh = 3, iso.subset = "off", drop.months = "FebMar")
udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, 
               n.ci.thresh = 3, iso.subset = "off", drop.months = "JanFebMar")
```

#### Summarise Uptake Depth Index by species for best-fit models

UDI corresponding to model with maximum Rsq, likelihood 
median, upr & lwr confidence interval for UDI for top ranking models

```{r Chunk 07.0, eval = FALSE}
source("code/07.0_summarising_udi_by_sp_size.R")
# splevel = "off"; dryseason = "on"; rsq.thresh = 0.3; root.selection = "exp"; iso.subset = "on"; drop.months = "None"; n.rank = 10

summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "None")
summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "Jan")
summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "Feb")
summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "Mar")
summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "JanFeb")
summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "FebMar")
summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "JanFebMar")

summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "None")
summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "Jan")
summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "Feb")
summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "Mar")
summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "JanFeb")
summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "FebMar")
summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "JanFebMar")

summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "None")
summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "Jan")
summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "Feb")
summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "Mar")
summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "JanFeb")
summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "FebMar")
summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "JanFebMar")

summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "None")
summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "Jan")
summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "Feb")
summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "Mar")
summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "JanFeb")
summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "FebMar")
summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "JanFebMar")
```

# Model Evaluation

#### Sensitivity of UDI of deciduous species to inclusion of deciduous months

```{r Chunk 08.0, eval = FALSE}
source("code/08.0_Sensitivity analysis for UDI of deciduous species.R")
# goodness.fit = 0.3; dryseason = "on"; root.selection = "zeng"; iso.subset = "on"
sensitivity.analysis(goodness.fit = 0.3, dryseason = "on", 
            root.selection = "zeng", iso.subset = "on")
sensitivity.analysis(goodness.fit = 0.3, dryseason = "jan", 
            root.selection = "zeng", iso.subset = "on")
# goodness.fit = 0.3; dryseason.vec = c("on", "jan"); root.selection = "zeng"; iso.subset = "on"
```

#### Comparison with deuterium isotopic signatures of water sourcing depths

<!-- For example: -->
<!-- `source("code/09.0_Comparison_with_Meinzer1999.R") -->
<!-- iso.compare(goodness.fit = 0.3, dryseason = "on",  -->
<!--             root.selection = "exp", iso.subset = "on", drop.months = "None")` -->

```{r Chunk 09.0, eval = FALSE}
source("code/09.1_Comparison_with_Meinzer1999.R")
iso.compare(goodness.fit = 0.3, dryseason = "on", 
            root.selection = "zeng", iso.subset = "on", drop.months = "None")
iso.compare(goodness.fit = 0.3, dryseason = "on", 
            root.selection = "zeng", iso.subset = "on", drop.months = "Jan")
iso.compare(goodness.fit = 0.3, dryseason = "on", 
            root.selection = "zeng", iso.subset = "on", drop.months = "Feb")
iso.compare(goodness.fit = 0.3, dryseason = "on", 
            root.selection = "zeng", iso.subset = "on", drop.months = "Mar")
iso.compare(goodness.fit = 0.3, dryseason = "on", 
            root.selection = "zeng", iso.subset = "on", drop.months = "JanFeb")
iso.compare(goodness.fit = 0.3, dryseason = "on", 
            root.selection = "zeng", iso.subset = "on", drop.months = "FebMar")
iso.compare(goodness.fit = 0.3, dryseason = "on", 
            root.selection = "zeng", iso.subset = "on", drop.months = "JanFebMar")
# goodness.fit = 0.3; dryseason = "on"; root.selection = "exp"; iso.subset = "on"; drop.months = "None"
```

#### Plot species and size-wise Uptake Depth Index
```{r Chunk 10.0, eval = FALSE}
source("code/10.0_species_graphs_BCI.R")
```

# Evaluating whole plant drought strategies

#### Calculate mortality rates from forest inventory data

```{r Chunk 11.0, eval = FALSE}
source("code/11.0_demographic_data_BCI.R")
```

#### Analyse mortality against UDI
```{r Chunk 11.1, eval = FALSE}
source("code/11.1_mortality_BCI.R")
demo.graphs(level.folder = "commlevel", n.threshold = 30, dbh.residuals = "on", 
           dryseason = "on", root.selection = "on", iso.subset = "off")
# level.folder = "commlevel"; n.threshold = 30; dbh.residuals = "on";
#            dryseason = "on"; root.selection = "on"; iso.subset = "off"
```

#### Analyse functional traits against UDI
```{r Chunk 12.0, eval = FALSE}
source("code/12.0_hydraulic_traits_BCI.R")
```

### 2020-11-10 code/11.0_Readme.Rmd
Minor edits

@@
 ---
-title: "Workflow of Inverse Model for Rooting Profiles"
-author: "Rutuja"
+title: "Workflow"
+author: "Rutuja Chitra-Tarak"
 date: "`r format(Sys.time(), '%d %B, %Y')`"
 output:
-  github_document
+  pdf_document: default
+  html_document:
+    df_print: paged
+  word_document: default
+subtitle: Hydraulically-vulnerable trees survive on deep-water access during droughts
+  in a tropical forest
 ---
@@
 
-# Model Setup
-
-#### Pre-requisite data
+# Pre-requisite data
 
@@
 
-Calculate species-size wise growth time series from forest inventory data. Calculate growth residuals after removing effect of size.
+Calculate species-specific growth time series from forest inventory data.
 
@@
 
-**Water Availability**
-
-Soil Water Potential data -- best-fit 100 ensembles from ELM-FATES during the study period.
-This is sourced from data-package `bci.elm.fates.hydro` with library(bci.elm.fates.hydro)
-
-#### Potential rooting profile trait-space
+**Mortality data**
 
 ```{r Chunk 02.0, eval = FALSE}
-source("code/02.0_Rooting_profiles.R")
+source("code/02.0_demographic_data_BCI.R")
 ```
 
-#### Calculate btran and organise
+#### All pre-requisites
 
-Gather all pre-requisites in a place or two
+Gather all pre-requisites for rooting depth inverse model (effective rooting depth ) in a place and traits data for post-processing
 
-1. `GLUEsetup_part1() in 03.0_GLUEsetup_part1_BCI.R`
+- Soil Water Potentials
+- VPD-GPP relationship
+- Leaf vulnerability curves
+- LAI seasonality
+- Hydraulic traits
 
- * Loads best.fit ensembles of Soil Water Potential (SWP) from bci.elm.fates.hydro::psi of a specific version (given by the "current folder")
- * Loads hypothetical rooting profiles generated by R/3.090_root_parameters_b.R
- * Loads Turgor Loss Point (TLP) data
- * Loads tree growth census dates for specific no. of intervals
- * Converts swp into a growth factor (range 0-1) using a step-fuction, such that growth factor is 0 at sp-level or community mean level Turgor Loss Point.
-Species wise growth factor is stores separately (sp.hydro.output.chosen)
-
-Then using a combination of a range of hypothetical root profiles obtained from 2.rooting_profiles.R and SWP based growth factor ensemble simulations generate a btran time-series using equation : Btran= i=1 to i=z RootFrac_z * f(swp_z),
-where f(swp_z) is a growth.factor_z based on a linear step relationship with swp_z ap species specific or community mean TLP level.
-
- * Calculates btran time-series by census-interval for each rooting profile for all the best.fit SWP ensemble
-  + by individual species with species-specific TLP (N of btran time series == No. of species x No. of SWP ensembles x No. of root profiles)
-  + at community-mean TLP (No. of btran time series == No. of SWP ensembles x No. of root profiles)
-  + both community and species level btran is calculated by dropping none or one or more number of dry season months as sublists
-  * Saves ELM-FATES hydro output, rooting profiles, census dates, number of intervals in a list called info, community level btran in a list named info.3, and species level btran is saved in info.2, along with sub-list names of dropped months
-
-```{r Chunk 03.0, eval = FALSE}
-source("code/03.0_GLUEsetup_part1_BCI.R")
-GLUEsetup_part1(current.folder = "2019-10-14_5000", intervals = 5, nsam = 200) # no.of root profiles
-## On macbook paralle processing to generate sp-level btran by months crashes Rstudio. Need to serialise and split job on two computers, then stitch info.2 together. 
+```{r Chunk 05.0, eval = FALSE}
+source("code/05.0_Prepare_data_for_correlative analyses.R")
 ```
 
-2. `GLUEsetup_part2() in 04.0_GLUEsetup_part2_BCI.R`
+Script to load data generated from the data preparation script above for use in inverse model as well as manuscript
 
-This function stores together growth data, growth meta-data, and btran data (at community mean TLP level by dropped months) in a list named growth_by_si.info.
-
-```{r Chunk 04.0, eval = FALSE}
-source("code/04.0_GLUEsetup_part2_BCI.R")
-GLUEsetup_part2(growth.type = "med", dbh.residuals = "on", solar.residuals = "off", 
-                growth.selection = "size_class_predefined_cc_scaled") #"size_class_varying_non_cc")
-#n.ensembles = 100000
-# growth.type = "med"; dbh.residuals = "on"; solar.residuals = "off"; growth.selection = "size_class_predefined_cc_scaled"
+```{r Chunk 06.0, eval = FALSE}
+source("code/06.0_load.R")
 ```
 
-# Model Parameterisation
+# Model Setup
 
-#### Run model for all ensembles
+Set-up and run effective rooting depth models, evaluate model, and join ERD from the predicted model with hydraulic traits and mortality data, and calcualted realized hydraulic risk
 
-`GLUE_run() in 05.1_GLUErun_part2_BCI.R`
-
-This function generates a few statistics for the fit between growth and btran:
-
-Rsq, residuals, sum of negative log likelihood and the number of data points of growth factor that pass through 95% confidence interval of growth.
-TLP level and any months to drop while calculating growth factor need be specified.
-Threshold (goodness.fit) as to when to calculate Rs value also needs to be specified.
-
-```{r Chunk 05.1, eval = FALSE}
-# This sources "code/4.2GLUErun_part2_func_BCI.R"
-source("code/05.1_GLUErun_part2_BCI.R")
-# splevel = "on"; goodness.fit = 0.3; drop.months = "None"
-GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "None")
-GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "Jan")
-GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "Feb")
-GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "Mar")
-GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "JanFeb")
-GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "FebMar")
-GLUE_run(splevel = "off", goodness.fit = 0.3, drop.months = "JanFebMar")
-GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "None")
-GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "Jan")
-GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "Feb")
-GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "Mar")
-GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "JanFeb")
-GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "FebMar")
-GLUE_run(splevel = "on", goodness.fit = 0.3, drop.months = "JanFebMar")
+```{r Chunk 07.0, eval = FALSE}
+source("code/07.0_PhenoDemoTraitsPsi.R")
 ```
 
-#### Get water availability and btran during a specific period
+# Manuscript
 
-`prep.udi.calculator() in 06.0_prep.udi.calculator.R`
+**Figures**
 
-This subsets SWP-based growth factor data (species-specific or community level) to various specified time period or months: late dryseason (Mar Apr) 1997, Jan 1997, dryseason 1992 etc.
-For example, restricted to early and late 1997 dry season to later calculate UDI only for this period to compare with isotopic tree Xylem sap data that is collected in early and later 1997.
-
-```{r Chunk 06.0, eval = FALSE}
-## These need not be updated if rooting profile space changes, but only if hydrology changes or dropped months; or the period one is interested in finding UDI for.
-source("code/06.0_prep.udi.calculator.R")
-# splevel = "on"; drop.months = "JanFebMar"
-prep.udi.calculator(splevel = "on", drop.months = "None")
-prep.udi.calculator(splevel = "on", drop.months = "Jan")
-prep.udi.calculator(splevel = "on", drop.months = "Feb")
-prep.udi.calculator(splevel = "on", drop.months = "Mar")
-prep.udi.calculator(splevel = "on", drop.months = "JanFeb")
-prep.udi.calculator(splevel = "on", drop.months = "FebMar")
-prep.udi.calculator(splevel = "on", drop.months = "JanFebMar")
-
-prep.udi.calculator(splevel = "off", drop.months = "None")
-prep.udi.calculator(splevel = "off", drop.months = "Jan")
-prep.udi.calculator(splevel = "off", drop.months = "Feb")
-prep.udi.calculator(splevel = "off", drop.months = "Mar")
-prep.udi.calculator(splevel = "off", drop.months = "JanFeb")
-prep.udi.calculator(splevel = "off", drop.months = "FebMar")
-prep.udi.calculator(splevel = "off", drop.months = "JanFebMar")
+```{r Chunk 08.0, eval = FALSE}
+source("code/08.0_manuscript_data_figs.R")
 ```
 
-#### Calculate water Uptake Depth Index for a specific period
+**Main text**
 
-`udi.calculator() in 06.1_udi.calculator.R`
-
-This calculates Uptake Depth Index and SDI (water Stress Depth Index, without taking into account total water availability) for each simulation and stores corresponding values of model Rsq, negative loLikelihood, number of data points that pass through confidence interval of growth, species and size in one data-frame named "ds.bestfit.all.Rdata"
-
-UDI or SDI is only calculated for a simulation for which R-sq is greater than the given therehold (e.g.
-rsq.thresh = 0.3) and if at least 3 data-points have passed through 95% CI of growth data calculated based on (e.g.
-n.ci.thresh = 3)
-
-```{r Chunk 06.1, eval = FALSE}
-source("code/06.1_udi.calculator.R")
-##---needs to be run manually on macbook for splevel = "off" and iso.subset = "off"---
-# udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.5, iso.subset = "on") ## udi is not calculated below this rsq.thresh
-# splevel = "off"; dryseason = "on"; rsq.thresh = 0.3;  n.ci.thresh = 3; iso.subset = "on"; drop.months = "None"
-udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.3,
-               n.ci.thresh = 3, iso.subset = "off", drop.months = "None")
-udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.3,
-               n.ci.thresh = 3, iso.subset = "off", drop.months = "Jan")
-udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.3,
-               n.ci.thresh = 3, iso.subset = "off", drop.months = "Feb")
-udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.3,
-               n.ci.thresh = 3, iso.subset = "off", drop.months = "Mar")
-udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.3,
-               n.ci.thresh = 3, iso.subset = "off", drop.months = "JanFeb")
-udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.3,
-               n.ci.thresh = 3, iso.subset = "off", drop.months = "FebMar")
-udi.calculator(splevel = "on", dryseason = "on", rsq.thresh = 0.3,
-               n.ci.thresh = 3, iso.subset = "off", drop.months = "JanFebMar")
-
-udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.3,
-               n.ci.thresh = 3, iso.subset = "off", drop.months = "None")
-udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.3,
-               n.ci.thresh = 3, iso.subset = "off", drop.months = "Jan")
-udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.3,
-               n.ci.thresh = 3, iso.subset = "off", drop.months = "Feb")
-udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.3,
-               n.ci.thresh = 3, iso.subset = "off", drop.months = "Mar")
-udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.3,
-               n.ci.thresh = 3, iso.subset = "off", drop.months = "JanFeb")
-udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.3,
-               n.ci.thresh = 3, iso.subset = "off", drop.months = "FebMar")
-udi.calculator(splevel = "off", dryseason = "on", rsq.thresh = 0.3,
-               n.ci.thresh = 3, iso.subset = "off", drop.months = "JanFebMar")
-
-udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.3,
-               n.ci.thresh = 3, iso.subset = "off", drop.months = "None")
-udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.3,
-               n.ci.thresh = 3, iso.subset = "off", drop.months = "Jan")
-udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.3,
-               n.ci.thresh = 3, iso.subset = "off", drop.months = "Feb")
-udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.3,
-               n.ci.thresh = 3, iso.subset = "off", drop.months = "Mar")
-udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.3,
-               n.ci.thresh = 3, iso.subset = "off", drop.months = "JanFeb")
-udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.3,
-               n.ci.thresh = 3, iso.subset = "off", drop.months = "FebMar")
-udi.calculator(splevel = "on", dryseason = "jan", rsq.thresh = 0.3,
-               n.ci.thresh = 3, iso.subset = "off", drop.months = "JanFebMar")
-
-udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, 
-               n.ci.thresh = 3, iso.subset = "off", drop.months = "None")
-udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, 
-               n.ci.thresh = 3, iso.subset = "off", drop.months = "Jan")
-udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.3,
-               n.ci.thresh = 3, iso.subset = "off", drop.months = "Feb")
-udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.3,
-               n.ci.thresh = 3, iso.subset = "off", drop.months = "Mar")
-udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, 
-               n.ci.thresh = 3, iso.subset = "off", drop.months = "JanFeb")
-udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.3,
-               n.ci.thresh = 3, iso.subset = "off", drop.months = "FebMar")
-udi.calculator(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, 
-               n.ci.thresh = 3, iso.subset = "off", drop.months = "JanFebMar")
+```{r Chunk 09.0, eval = FALSE}
+source("code/09.0_manuscript.rmd")
 ```
 
-#### Summarise Uptake Depth Index by species for best-fit models
+**Supporting Information**
 
-UDI corresponding to model with maximum Rsq, likelihood 
-median, upr & lwr confidence interval for UDI for top ranking models
-
-```{r Chunk 07.0, eval = FALSE}
-source("code/07.0_summarising_udi_by_sp_size.R")
-# splevel = "off"; dryseason = "on"; rsq.thresh = 0.3; root.selection = "exp"; iso.subset = "on"; drop.months = "None"; n.rank = 10
-
-summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "None")
-summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "Jan")
-summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "Feb")
-summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "Mar")
-summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "JanFeb")
-summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "FebMar")
-summarise.udi(splevel = "on", dryseason = "on", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "JanFebMar")
-
-summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "None")
-summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "Jan")
-summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "Feb")
-summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "Mar")
-summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "JanFeb")
-summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "FebMar")
-summarise.udi(splevel = "off", dryseason = "on", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "JanFebMar")
-
-summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "None")
-summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "Jan")
-summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "Feb")
-summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "Mar")
-summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "JanFeb")
-summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "FebMar")
-summarise.udi(splevel = "on", dryseason = "jan", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "JanFebMar")
-
-summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "None")
-summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "Jan")
-summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "Feb")
-summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "Mar")
-summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "JanFeb")
-summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "FebMar")
-summarise.udi(splevel = "off", dryseason = "jan", rsq.thresh = 0.3, root.selection = "zeng", iso.subset = "off", n.rank = 10, drop.months = "JanFebMar")
+```{r Chunk 10.0, eval = FALSE}
+source("code/10.0_Supporting_Information.Rmd")
 ```
-
-# Model Evaluation
-
-#### Sensitivity of UDI of deciduous species to inclusion of deciduous months
-
-```{r Chunk 08.0, eval = FALSE}
-source("code/08.0_Sensitivity analysis for UDI of deciduous species.R")
-# goodness.fit = 0.3; dryseason = "on"; root.selection = "zeng"; iso.subset = "on"
-sensitivity.analysis(goodness.fit = 0.3, dryseason = "on", 
-            root.selection = "zeng", iso.subset = "on")
-sensitivity.analysis(goodness.fit = 0.3, dryseason = "jan", 
-            root.selection = "zeng", iso.subset = "on")
-# goodness.fit = 0.3; dryseason.vec = c("on", "jan"); root.selection = "zeng"; iso.subset = "on"
-```
-
-#### Comparison with deuterium isotopic signatures of water sourcing depths
-
-<!-- For example: -->
-<!-- `source("code/09.0_Comparison_with_Meinzer1999.R") -->
-<!-- iso.compare(goodness.fit = 0.3, dryseason = "on",  -->
-<!--             root.selection = "exp", iso.subset = "on", drop.months = "None")` -->
-
-```{r Chunk 09.0, eval = FALSE}
-source("code/09.1_Comparison_with_Meinzer1999.R")
-iso.compare(goodness.fit = 0.3, dryseason = "on", 
-            root.selection = "zeng", iso.subset = "on", drop.months = "None")
-iso.compare(goodness.fit = 0.3, dryseason = "on", 
-            root.selection = "zeng", iso.subset = "on", drop.months = "Jan")
-iso.compare(goodness.fit = 0.3, dryseason = "on", 
-            root.selection = "zeng", iso.subset = "on", drop.months = "Feb")
-iso.compare(goodness.fit = 0.3, dryseason = "on", 
-            root.selection = "zeng", iso.subset = "on", drop.months = "Mar")
-iso.compare(goodness.fit = 0.3, dryseason = "on", 
-            root.selection = "zeng", iso.subset = "on", drop.months = "JanFeb")
-iso.compare(goodness.fit = 0.3, dryseason = "on", 
-            root.selection = "zeng", iso.subset = "on", drop.months = "FebMar")
-iso.compare(goodness.fit = 0.3, dryseason = "on", 
-            root.selection = "zeng", iso.subset = "on", drop.months = "JanFebMar")
-# goodness.fit = 0.3; dryseason = "on"; root.selection = "exp"; iso.subset = "on"; drop.months = "None"
-```
-
-#### Plot species and size-wise Uptake Depth Index
-```{r Chunk 10.0, eval = FALSE}
-source("code/10.0_species_graphs_BCI.R")
-```
-
-# Evaluating whole plant drought strategies
-
-#### Calculate mortality rates from forest inventory data
-
-```{r Chunk 11.0, eval = FALSE}
-source("code/11.0_demographic_data_BCI.R")
-```
-
-#### Analyse mortality against UDI
-```{r Chunk 11.1, eval = FALSE}
-source("code/11.1_mortality_BCI.R")
-demo.graphs(level.folder = "commlevel", n.threshold = 30, dbh.residuals = "on", 
-           dryseason = "on", root.selection = "on", iso.subset = "off")
-# level.folder = "commlevel"; n.threshold = 30; dbh.residuals = "on";
-#            dryseason = "on"; root.selection = "on"; iso.subset = "off"
-```
-
-#### Analyse functional traits against UDI
-```{r Chunk 12.0, eval = FALSE}
-source("code/12.0_hydraulic_traits_BCI.R")
-```

### 2020-11-12 code/09.0_manuscript.rmd
1. corrected sourced file name

--- 2020-11-10 manuscript.rmd
@@
 options(tinytex.verbose = TRUE)
-source("code/load.R")
-source("code/manuscript_data_figs.R")
+source("code/06.0_load.R")
+source("code/08.0_manuscript_data_figs.R")
 # load("results/manuWorkSpace.RData")

### 2020-11-12 Readme.Rmd
updated title and description

--- 2020-11-10 code/11.0_Readme.Rmd
@@
 ---
-title: "Workflow"
+title: "Workflow Part II: for inverse root modeling and post-processing"
 author: "Rutuja Chitra-Tarak"
@@
 
-- Soil Water Potentials
+- Soil Water Potentials (outputs from ELM-FATES)
 - VPD-GPP relationship

### 2020-11-12 Readme.Rmd
updated title and description

@@
 ---
-title: "Workflow Part II: for inverse root modeling and post-processing"
+title: "Workflow Part II: Inverse root modeling and post-processing"
 author: "Rutuja Chitra-Tarak"

### 2021-01-20 LAI.Rmd
First commit

---
title: "LAI workflow"
author: "Rutuja Chitra-Tarak"
date: "1/15/2021"
output:
      bookdown::html_document2:
        number_sections: no
        theme: flatly
        fig_caption: yes
      bookdown::pdf_document2: default
---
```{r setup, echo = FALSE, include=FALSE}
library(knitr)
opts_knit$set(eval.after = "fig.cap")
knitr::opts_chunk$set(echo = TRUE)
if(!require("pacman", quietly=TRUE)) install.packages("pacman")
suppressPackageStartupMessages(
  pacman::p_load(bookdown, knitr, pander, citr, broom))
options(tinytex.verbose = TRUE)
```

1. Built leaf lifespan (LL) to LMA relationship based on both, PNM and San-Lorenzo (FTS) based on canopy leaves of trees (Figure \@ref(fig:ll-lma)), and used that to gap-fill leaf lifespan data.

```{r ll-lma, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", results = "asis", fig.cap = paste0("\\label{lab-ll-lma}Species-specific Leaf lifespan vs. LMA observed for canopy leaves at two sites at Panama. Each observation (circle) is color coded by site. A linear model fit (line) and equation with R2 and p-value are shown.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_cohort/lifespan_by_lma_canopy_sitecolored.jpeg"))
```

```{r ll-lma-by-site, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", results = "asis", fig.cap = paste0("\\label{lab-ll-lma-by-site} Species-specific Leaf lifespan vs. LMA observed for canopy leaves at two sites, shown separately. A linear model fit (line) and equation with coefficient of determination and p-value are shown separately for each site.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_cohort/lifespan_by_lma_canopy_two sites.jpeg"))
```

2. Restricted leaf-litter data to the last 7 years for BCI-Poachers and BCI-50ha, converted observed leaf fall data column in mass units (leaf_gm) to mass per unit ground area as follows:  
    Leaf in gm per m2 ground area at a site = Leaf biomass in gm / (Site-specific trap area in m2  x Number of leaves were collected from) --- (Eq 1)  
    In Joe's the dataframe, column "leaf_gm"" represents leaf~biomass~in~gm and column "n_traps" represents the number of traps the leaf_gm observation came from.

3. Removed outliers in leaf_gm_per_m2 as those greater than 5 standard deviation by species

4. Remove a small set of species-site combinations with too few records for that species at that site by visually inspecting data

4. Distributed weekly leaf_gm_per_m2 over the days of the week with interpolation

5. When a species' leaf_gm_per_m2 was recorded at both the sites, obtained an average.

6. Converted leaf_gm_per_m2 to LAI units as   
    leaf_fall_LAI (in m2/m2) = leaf_gm_per_m2 / LMA (in g/m2) --- (Eq 2)  
    LMA units are from Metadata_Lftraits_VariableDefinitionsOnly.doc for data in BCITRAITS_20101220.csv)

7. For each species, prepended its leaf_gm_per_m2 data-frame by adding days equal to the species leaf lifespan, sp.lifespan  
    Then LAI time series is calculated: For t in 1: total.no.of.days in the data frame,   
    1. Assuming zero LAI on previous date, leaf fall for a future day sp.lifespan + 1 is considered as leaf gain on day 1. That is,    
    If t == 1,  
    LAI[t] = leaf_fall_LAI[t + sp.lifespan]  
    1. New leaf gain is added to LAI of the previous day, but leaf fall does not begin yet until the day of sp.lifespan + 1:  
    For t > 1, if t < sp.lifespan + 1,   
    LAI[t] <- LAI[t-1] + leaf_fall_LAI[t + sp.lifespan]  
    1. LEaf fall begins on day sp.lifespan + 1, and is removed  
    For t > 1, if t >= sp.lifespan + 1,  
    LAI[t] <- LAI[t-1] + leaf_fall_LAI[t + sp.lifespan] - leaf_fall_LAI[t]  
    1. I then removed the prepended dates, that is, for t in 1: sp.lifespan, and thus obtained the LAI time series (Figure \@ref(fig:LAI-ts)).

```{r LAI-ts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ts}LAI time series, beginning with zero prior LAI. Absolute values of LAI are way too high and need correction by a factor, and I haven't figure out what that is (let me know if you do), but look at the time series trend for now. For some species, time series ranges are quite stable over the years, and for others they increase over the years time. LL = species-specific leaf lifespan in days.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/LAI_ts.jpeg"))
```

8. I then built a species-specific LAI seasonality curve by taking a mean of LAI by DOY. Then normalized this curve by dividing by the max LAI value in the year (Figure \@ref(fig:LAI-ssn)).

```{r LAI-ssn, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ssn}LAI seasonality curves, beginning with zero prior LAI. LL = species-specific leaf lifespan in days. Species are color coded by deciduousness.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/LAI.seasonality_mean.jpeg"))
```

9. As step 6.1 starts from zero prior LAI, I thought to modify that and use the expected LAI from step 7 for the DOY of day 1, then recalculated LAI, but that that did not change the pattern observed in some soecies of increasing LAI over the years (Figure \@ref(fig:LAI-ts-mod)) and resulted in a marginally different LAI seasonality curves (Figure \@ref(fig:LAI-ssn-mod)).

```{r LAI-ts-mod, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ts-mod}Modified LAI time series, beginning with expected LAI for the DOY of day 1. The absolute values of LAI are way too high and need correction by a factor, and I haven't figure out what that is (let me know if you do), but look at the time series trend for now. LL = species-specific leaf lifespan in days.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/LAI_ts_mod.jpeg"))
```

```{r LAI-ssn-mod, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ssn-mod}Modified LAI seasonality curves, beginning with expected LAI for the DOY of day 1. LL = species-specific leaf lifespan in days. Species are color coded by deciduousness.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/LAI.seasonality_mean_mod.jpeg"))
```

### 2021-01-20 LAI.Rmd
First commit

@@
 
-```{r ll-lma-by-site, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", results = "asis", fig.cap = paste0("\\label{lab-ll-lma-by-site} Species-specific Leaf lifespan vs.
+```{r ll-lma, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", results = "asis", fig.cap = paste0("\\label{lab-ll-lma} Species-specific Leaf lifespan vs.
 LMA observed for canopy leaves at two sites, shown separately.
@@
 
-3. Removed outliers in leaf_gm_per_m2 as those greater than 5 standard deviation by species
+3. Removed outliers by species as those leaf_gm_per_m2 > 5 standard deviation
 
@@
 6. Converted leaf_gm_per_m2 to LAI units as   
-    leaf_fall_LAI (in m2/m2) = leaf_gm_per_m2 / LMA (in g/m2) --- (Eq 2)  
+    leaf_fall_LAI (in m2/m2) = leaf_gm_per_m2 / LMA (in g/m2)
     LMA units are from Metadata_Lftraits_VariableDefinitionsOnly.doc for data in BCITRAITS_20101220.csv)

### 2021-01-20 LAI.Rmd
Removed modified graphs built by starting with expected LAI for the beginning DOY of the time series

--- 2021-01-20 LAI.Rmd
@@
 ```
-
-9.
-As step 6.1 starts from zero prior LAI, I thought to modify that and use the expected LAI from step 7 for the DOY of day 1, then recalculated LAI, but that that did not change the pattern observed in some soecies of increasing LAI over the years (Figure \@ref(fig:LAI-ts-mod)) and resulted in a marginally different LAI seasonality curves (Figure \@ref(fig:LAI-ssn-mod)).
-
-```{r LAI-ts-mod, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ts-mod}Modified LAI time series, beginning with expected LAI for the DOY of day 1.
-The absolute values of LAI are way too high and need correction by a factor, and I haven't figure out what that is (let me know if you do), but look at the time series trend for now.
-LL = species-specific leaf lifespan in days.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/LAI_ts_mod.jpeg"))
-```
-
-```{r LAI-ssn-mod, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ssn-mod}Modified LAI seasonality curves, beginning with expected LAI for the DOY of day 1.
-LL = species-specific leaf lifespan in days.
-Species are color coded by deciduousness.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/LAI.seasonality_mean_mod.jpeg"))
-```

### 2021-01-20 LAI.Rmd
1. Tightened language 2. Removed end period of LAI time series for which leaf gain data not present 3. Added detrended time series

@@
 7. For each species, prepended its leaf_gm_per_m2 data-frame by adding days equal to the species leaf lifespan, sp.lifespan  
-    Then LAI time series is calculated: For t in 1: total.no.of.days in the data frame,   
-    1. Assuming zero LAI on previous date, leaf fall for a future day sp.lifespan + 1 is considered as leaf gain on day 1. That is,    
+    Then LAI time series is calculated: For t in 1 to c(total no of days in the data frame - leaf.lifespan) ,   
+    1. Assuming zero LAI to begin with, leaf gain on day 1 is leaf fall for the future day sp.lifespan. That is,    
     If t == 1,  
-    LAI[t] = leaf_fall_LAI[t + sp.lifespan]  
-    1. New leaf gain is added to LAI of the previous day, but leaf fall does not begin yet until the day of sp.lifespan + 1:  
-    For t > 1, if t < sp.lifespan + 1,   
-    LAI[t] <- LAI[t-1] + leaf_fall_LAI[t + sp.lifespan]  
-    1. LEaf fall begins on day sp.lifespan + 1, and is removed  
-    For t > 1, if t >= sp.lifespan + 1,  
-    LAI[t] <- LAI[t-1] + leaf_fall_LAI[t + sp.lifespan] - leaf_fall_LAI[t]  
+    LAI[t] = leaf_fall_LAI[sp.lifespan]  
+    1. New leaf gain is added to LAI of the previous day, but leaf fall does not begin yet until the day sp.lifespan:  
+    For t > 1, if t < sp.lifespan,   
+    LAI[t] <- LAI[t-1] + leaf_fall_LAI[sp.lifespan]  
+    1. LEaf fall begins on day sp.lifespan, and is removed  
+    For t > 1, if t >= sp.lifespan,  
+    LAI[t] <- LAI[t-1] + leaf_fall_LAI[sp.lifespan] - leaf_fall_LAI[t]  
     1. I then removed the prepended dates, that is, for t in 1: sp.lifespan, and thus obtained the LAI time series (Figure \@ref(fig:LAI-ts)).
 
-```{r LAI-ts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ts}LAI time series, beginning with zero prior LAI.
-Absolute values of LAI are way too high and need correction by a factor, and I haven't figure out what that is (let me know if you do), but look at the time series trend for now.
-For some species, time series ranges are quite stable over the years, and for others they increase over the years time.
+```{r LAI-ts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ts}LAI time series.
 LL = species-specific leaf lifespan in days.")}
@@
 
-8. I then built a species-specific LAI seasonality curve by taking a mean of LAI by DOY. Then normalized this curve by dividing by the max LAI value in the year (Figure \@ref(fig:LAI-ssn)).
+8. For some species LAI increases/decreases over time likely due to change in population levels. Therefore detrended the time series by removing the slope (Figure \@ref(fig:LAI-ts)).
 
-```{r LAI-ssn, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ssn}LAI seasonality curves, beginning with zero prior LAI.
+```{r LAI-ts-det, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ts-det}Detrended LAI time series.
+LL = species-specific leaf lifespan in days.")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/LAI_ts_detrended.jpeg"))
+```
+
+9. I then built a species-specific LAI seasonality curve by taking a mean of LAI by DOY. Then normalized this curve by dividing by the max LAI value in the year (Figure \@ref(fig:LAI-ts-det)).
+
+```{r LAI-ssn, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ssn}Mean LAI seasonality curves.
 LL = species-specific leaf lifespan in days.

### 2021-01-20 LAI.Rmd
Language correction

@@
 2. Restricted leaf-litter data to the last 7 years for BCI-Poachers and BCI-50ha, converted observed leaf fall data column in mass units (leaf_gm) to mass per unit ground area as follows:  
-    Leaf in gm per m2 ground area at a site = Leaf biomass in gm / (Site-specific trap area in m2  x Number of leaves were collected from) --- (Eq 1)  
+    Leaf in gm per m2 ground area at a site = Leaf biomass in gm / (Site-specific trap area in m2  x Number of traps leaves were collected from) --- (Eq 1)  
     In Joe's the dataframe, column "leaf_gm"" represents leaf~biomass~in~gm and column "n_traps" represents the number of traps the leaf_gm observation came from.

### 2021-01-20 09.0_manuscript.rmd
Moved .Rmd to the root

--- 2020-11-12 code/09.0_manuscript.rmd
@@
 An NSF grant to Smithsonian Institution’s ForestGEO (Dimensions 1046113) partially supported this work via RCT and SMM.
+BTW was supported by the National Institute of Food and Agriculture, U.S.
+Department of Agriculture, McIntire Stennis project under LAB94493.
 RF was supported by the National Center for Atmospheric Research, which is funded by the National Science Foundation.

### 2021-01-28 LAI.Rmd
Made correction in workflow matching rcode

--- 2021-01-20 LAI.Rmd
@@
 ---
-title: "LAI workflow"
+title: "LAI workflow documentation"
 author: "Rutuja Chitra-Tarak"
@@
 
-1. Built leaf lifespan (LL) to LMA relationship based on both, PNM and San-Lorenzo (FTS) based on canopy leaves of trees (Figure \@ref(fig:ll-lma)), and used that to gap-fill leaf lifespan data.
+1.
+Built leaf lifespan (LL) to LMA relationship for San-Lorenzo (FTS) based on canopy leaves and those with sufficient sample size to determine (n_lifespan > 100) (Figure \@ref(fig:ll-lma)).
+Per Joe, BCI climate is more similar to San Lorenzo, while PNM site is much drier.
+More importantly, the BCI 50-ha plot and San Lorenzo support old-growth forests.
+PNM supports much younger, secondary forest (age unknown, perhaps 100 years).
+So, used the leaf lifespan-LMA relationship for San Lorenzo.
+This was used to predict leaf lifespan data for species without leaf lifespan data based on their LMA, after gap-filling species LMA data.
+For the latter, I used LMA_lamina, which was in turn  gap-filled based on a linear relationship with LMALEAF_AVD or LMADISC_AVD.
 
 ```{r ll-lma, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", results = "asis", fig.cap = paste0("\\label{lab-ll-lma}Species-specific Leaf lifespan vs.
-LMA observed for canopy leaves at two sites at Panama.
-Each observation (circle) is color coded by site.
+LMA observed for canopy leaves at San Lorenzo in Panama.
+Each observation (circle) is color coded by deciduous class.
 A linear model fit (line) and equation with R2 and p-value are shown.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_cohort/lifespan_by_lma_canopy_sitecolored.jpeg"))
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_cohort/lifespan_by_lma_canopy.jpeg"))
 ```
 
-```{r ll-lma-by-site, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", results = "asis", fig.cap = paste0("\\label{lab-ll-lma-by-site} Species-specific Leaf lifespan vs.
-LMA observed for canopy leaves at two sites, shown separately.
-A linear model fit (line) and equation with coefficient of determination and p-value are shown separately for each site.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_cohort/lifespan_by_lma_canopy_two sites.jpeg"))
-```
+2.
+BCI-Poachers data goes back to 1985, while BCI-50ha data begins in 2013.
+I used this entire record.
+BCI-Poachers has 59 traps of 0.25 m2 each, while BCI-50ha has 62 traps of 0.5 m2 each.
+Before sending me the leaf-fall dataset, Joe had removed extra traps at BCI-50ha, which were discontinued after a few weeks of observation.
+I converted observed leaf fall data column in mass units (leaf_gm) to mass per unit ground area as follows:
+    Leaf in gm per m2 ground area at a site = Leaf biomass in gm / (Site-specific trap area in m2 x Total number of traps used ) --- (Eq 1)  
+    In Joe's dataframe, column "leaf_gm"" represents leaf biomass in gm.
 
-2. Restricted leaf-litter data to the last 7 years for BCI-Poachers and BCI-50ha, converted observed leaf fall data column in mass units (leaf_gm) to mass per unit ground area as follows:  
-    Leaf in gm per m2 ground area at a site = Leaf biomass in gm / (Site-specific trap area in m2  x Number of leaves were collected from) --- (Eq 1)  
-    In Joe's the dataframe, column "leaf_gm"" represents leaf~biomass~in~gm and column "n_traps" represents the number of traps the leaf_gm observation came from.
+3. Distributed weekly leaf_gm_per_m2 over the days of the week with interpolation using approx function.
 
-3. Removed outliers in leaf_gm_per_m2 as those greater than 5 standard deviation by species
+4. When a species' leaf_gm_per_m2 was recorded at both the sites, obtained an average.
 
-4. Remove a small set of species-site combinations with too few records for that species at that site by visually inspecting data
-
-4. Distributed weekly leaf_gm_per_m2 over the days of the week with interpolation
-
-5. When a species' leaf_gm_per_m2 was recorded at both the sites, obtained an average.
-
-6. Converted leaf_gm_per_m2 to LAI units as   
+5. Converted leaf_gm_per_m2 to LAI units as   
     leaf_fall_LAI (in m2/m2) = leaf_gm_per_m2 / LMA (in g/m2) --- (Eq 2)  
@@
 
-7. For each species, prepended its leaf_gm_per_m2 data-frame by adding days equal to the species leaf lifespan, sp.lifespan  
-    Then LAI time series is calculated: For t in 1 to c(total no of days in the data frame - leaf.lifespan) ,   
-    1. Assuming zero LAI to begin with, leaf gain on day 1 is leaf fall for the future day sp.lifespan. That is,    
+6. For each species, prepended its leaf_gm_per_m2 data-frame by adding the number of prior dates equivalent to days in the species leaf lifespan, named 'sp.lifespan'  
+    Then LAI time series is calculated with the following steps: For t in 1 to c(total no of days in the data frame - leaf.lifespan),   
+    1. Assuming zero LAI to begin with, leaf gain on date 1 is leaf fall for the future date shifted by sp.lifespan. That is,    
     If t == 1,  
     LAI[t] = leaf_fall_LAI[sp.lifespan]  
-    1. New leaf gain is added to LAI of the previous day, but leaf fall does not begin yet until the day sp.lifespan:  
+    1. New leaf gain is added to LAI of the previous date, but leaf fall does not begin until the future date equivalent to sp.lifespan. So,  
     For t > 1, if t < sp.lifespan,   
     LAI[t] <- LAI[t-1] + leaf_fall_LAI[sp.lifespan]  
-    1. LEaf fall begins on day sp.lifespan, and is removed  
+    1. Leaf fall begins on date equivalent to sp.lifespan. So,  
     For t > 1, if t >= sp.lifespan,  
     LAI[t] <- LAI[t-1] + leaf_fall_LAI[sp.lifespan] - leaf_fall_LAI[t]  
-    1. I then removed the prepended dates, that is, for t in 1: sp.lifespan, and thus obtained the LAI time series (Figure \@ref(fig:LAI-ts)).
+    1.
+Because dates before the date when leaf-fall begins, that is, the prepended dates from t = 1: sp.lifespan, would underestimate LAI because of lack of a prior, I removed the time series for the prepended dates.
+The remaining LAI time series for each species is shown in (Figure \@ref(fig:LAI-ts)).
 
@@
 
-8. For some species LAI increases/decreases over time likely due to change in population levels. Therefore detrended the time series by removing the slope (Figure \@ref(fig:LAI-ts)).
+7. Per Joe, for some species LAI increases/decreases over time likely due to change in population levels. Therefore detrended the LAI time series by removing the slope (Figure \@ref(fig:LAI-ts-det)).
 
@@
 
-9. I then built a species-specific LAI seasonality curve by taking a mean of LAI by DOY. Then normalized this curve by dividing by the max LAI value in the year (Figure \@ref(fig:LAI-ts-det)).
+8.
+A species-specific LAI seasonality curve is then built by taking a mean of LAI by DOY for the detrended LAI time series.
+This curve is then normalized by dividing by the max LAI value across DOYs (Figure \@ref(fig:LAI-ssn)).
 
@@
 ```
+
+9.
+The prior LAI on date 1 (Step 8.1 above) for calculating the LAI time series in Figure \@ref(fig:LAI-ts) could be estimated from the mean LAI estimate for the DOY of day 1 Figure \@ref(fig:LAI-ssn)).
+So Step 8.1 becomes: If t == 1,  LAI[t] =  Expected_LAI[t] + leaf_fall_LAI[sp.lifespan].
+Using this prior, and following rest of the steps 2-4, the resultant LAI time series is (Figure \@ref(fig:LAI-ts-mod)), and after detrending, mean LAI by DOY time series is calculated.
+These are shown in Figure \@ref(fig:LAI-ssn-mod) as thick lines overlaid on LAI by DOY curves calculated without a prior (thin lines).
+Using a prior significantly reduces the dip in LAI for many a species.
+This may be more realistic for evergreen species (Joe, your thoughts?).
+
+```{r LAI-ts-mod, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ts-mod}LAI time series based on a prior LAI estimate.
+LL = species-specific leaf lifespan in days.")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/LAI_ts_mod.jpeg"))
+```
+
+```{r LAI-ssn-mod, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ssn-mod}Mean LAI seasonality curves based on a prior LAI estimate (thick lines) overlaid on those without a prior (thin lines).
+LL = species-specific leaf lifespan in days.
+Species are color coded by deciduousness.")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/LAI.seasonality_mean_overlaid_mod.jpeg"))
+```
+
+10.
+To make non-evergreen species fully deciduous two options are tried: (1) By rescaling the *whole* mean LAI seasonality curves to range from 0-1 (Figure \@ref(fig:LAI-ssn-mod-deci-all); rescale1 <- function (x) {x/max(x)}.
+This hardly allows trees to have full LAI through the wet season.
+Especially, see SPONMO and TRATAS.
+Therefore (2), I rescaled *only the lowest periods* of mean LAI seasonality curves in Figure \@ref(fig:LAI-ssn-mod-deci-all)) such that the minimum is zero (rescale0 <- function (x) {c(x - min(x))/(max(x) - min(x))}.
+Brevideciduous species are deciduous for a short while, while obligate deciduous species for longer; so I varied the length of the deciduousness period based on deciduous type: For brevideciduous I chose the part of the curve for which LAI was less than 0.15 quantile of the LAI distribution in the mean seasonality curve; for facultative deciduous species < 0.2 quantile; and for obligate deciduous species < 0.3 quantile.
+These modified curves are shown in figure (Figure \@ref(fig:LAI-ssn-mod-deci)).
+
+```{r LAI-ssn-mod-deci-all, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ssn-mod-deci-all}Mean LAI seasonality curves based on a prior LAI estimate are used: Curves are then rescaled so that for evergreen species the maximum is 1, while for non-evergreen species, the range is 0-1.
+Species are color coded by deciduousness.")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/LAI.seasonality_mean_mod_deci_unrestricted_range01.jpeg"))
+```
+
+```{r LAI-ssn-mod-deci, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ssn-mod-deci}Mean LAI seasonality curves based on a prior LAI estimate are used.
+Then all species' curves are first rescaled so that the maximum is 1.
+For non-evergreen species, lowest periods based on quantiles are rescaled further such the minimum is 0.
+Species are color coded by deciduousness.")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/LAI.seasonality_mean_mod_deci_restricted.jpeg"))
+```

### 2021-01-28 LAI.Rmd
1. Added Joe's calculation 2. Code to show interpolation

@@
     In Joe's dataframe, column "leaf_gm"" represents leaf biomass in gm.
+    Retained only canopy tree species.
 
-3. Distributed weekly leaf_gm_per_m2 over the days of the week with interpolation using approx function.
+3. Distributed weekly leaf_gm_per_m2 over the days of the week with interpolation using base R's approx function.
+```r
+For a single species, the data-frame looks like this: 
+head(df.sp_site[[1]])
+       sp4 census leaf_gm n_traps       date         site site.trap.area leaf_gm_per_m2     sp
+30024 ALCC      1     0.2       1 1985-11-25 BCI-Poachers          14.75     0.01355932 alchco
+30025 ALCC      3    12.0       1 1985-12-09 BCI-Poachers          14.75     0.81355932 alchco
+30026 ALCC      4     0.4       1 1985-12-16 BCI-Poachers          14.75     0.02711864 alchco
+30027 ALCC      5     0.6       3 1985-12-23 BCI-Poachers          14.75     0.04067797 alchco
+30028 ALCC      6     0.2       1 1985-12-30 BCI-Poachers          14.75     0.01355932 alchco
+30029 ALCC      7     0.4       2 1986-01-06 BCI-Poachers          14.75     0.02711864 alchco
+                  sp_site   form
+30024 alchco_BCI-Poachers Canopy
+30025 alchco_BCI-Poachers Canopy
+30026 alchco_BCI-Poachers Canopy
+30027 alchco_BCI-Poachers Canopy
+30028 alchco_BCI-Poachers Canopy
+30029 alchco_BCI-Poachers Canopy
 
-4. When a species' leaf_gm_per_m2 was recorded at both the sites, obtained an average.
+## First I gap-fill dates in the weekly record data-frame with the following function:
+
+fill.day.gaps <- function(df) {
+  df.1 <- df %>% arrange(date) %>%
+    # for the first census assume leaf fall is collected over the past week
+    mutate(n.days = c(7, as.numeric(date - lag(date))[-1])) %>%
+    #  leaf_fall rate per day for the census interval is stored on the day of the census:
+    mutate(leaf_gm_rate = leaf_gm_per_m2/n.days,
+           # approx() will interpolate this forward until the next census,
+           # we want it to interpolate this value backward over the past census interval,
+           # so just for computation purposes, leading this value
+           leaf_gm_rate_lead = lead(leaf_gm_rate, n = 1))
+  # Gap-filling dates
+  full.date.df <- data.frame(date = seq(from = min(df$date, na.rm = TRUE),
+                                        to = max(df$date, na.rm = TRUE), by = 1)) %>%
+    mutate(day_number = as.numeric(difftime(date, min(df$date, na.rm = TRUE) - 1)))
+  df.2 <- df.1 %>%
+    full_join(full.date.df, by = "date") %>% arrange(date, site)
+  return(df.2)
+}
+# To get leaf fall on each day of the interval
+leaf.fall.daygaps <- lapply(df.sp_site, fill.day.gaps)
+
+head(leaf.fall.daygaps[[1]])
+   sp4 census leaf_gm n_traps       date         site site.trap.area leaf_gm_per_m2     sp             sp_site
+1 ALCC      1     0.2       1 1985-11-25 BCI-Poachers          14.75     0.01355932 alchco alchco_BCI-Poachers
+2 <NA>     NA      NA      NA 1985-11-26         <NA>             NA             NA   <NA>                <NA>
+3 <NA>     NA      NA      NA 1985-11-27         <NA>             NA             NA   <NA>                <NA>
+4 <NA>     NA      NA      NA 1985-11-28         <NA>             NA             NA   <NA>                <NA>
+5 <NA>     NA      NA      NA 1985-11-29         <NA>             NA             NA   <NA>                <NA>
+6 <NA>     NA      NA      NA 1985-11-30         <NA>             NA             NA   <NA>                <NA>
+    form n.days leaf_gm_rate leaf_gm_rate_lead day_number
+1 Canopy      7  0.001937046        0.05811138          1
+2   <NA>     NA           NA                NA          2
+3   <NA>     NA           NA                NA          3
+4   <NA>     NA           NA                NA          4
+5   <NA>     NA           NA                NA          5
+6   <NA>     NA           NA                NA          6s
+
+## Interpolating from weekly sums to daily leaf_gm
+leaf.interp.approx <- function(df) {
+  x <- df$day_number
+  y <- df$leaf_gm_rate_lead
+  xout <- df$day_number[is.na(df$leaf_gm_rate_lead)]
+  # yout <- approx(x, y, xout, method = "linear")
+  ## method = "constant" would be more parsimonious
+  yout <- approx(x, y, xout, method = "constant")
+  df.1 <- df %>%
+    left_join(data.frame(day_number = yout$x, leaf_gm.int.raw = yout$y), by = "day_number") %>%
+    ## filling interpolation gap on the day of the census
+    mutate(leaf_gm.int.raw = ifelse(is.na(leaf_gm_rate), leaf_gm.int.raw, leaf_gm_rate),
+           sp = df$sp[1],
+           site = df$site[1],
+           site.trap.area = df$site.trap.area[1])
+  return(df.1)
+}
+
+leaf.fall.int <- lapply(lapply(leaf.fall.daygaps, leaf.interp.approx),
+                        as.data.frame) %>%
+  bind_rows(.id = "sp_site") %>%
+  select(-leaf_gm_rate_lead)
+
+head(leaf.fall.int)
+sp4 census leaf_gm n_traps       date         site site.trap.area leaf_gm_per_m2     sp             sp_site
+1 ALCC      1     0.2       1 1985-11-25 BCI-Poachers          14.75     0.01355932 alchco alchco_BCI-Poachers
+2 <NA>     NA      NA      NA 1985-11-26 BCI-Poachers          14.75             NA alchco alchco_BCI-Poachers
+3 <NA>     NA      NA      NA 1985-11-27 BCI-Poachers          14.75             NA alchco alchco_BCI-Poachers
+4 <NA>     NA      NA      NA 1985-11-28 BCI-Poachers          14.75             NA alchco alchco_BCI-Poachers
+5 <NA>     NA      NA      NA 1985-11-29 BCI-Poachers          14.75             NA alchco alchco_BCI-Poachers
+6 <NA>     NA      NA      NA 1985-11-30 BCI-Poachers          14.75             NA alchco alchco_BCI-Poachers
+    form n.days leaf_gm_rate day_number leaf_gm.int.raw
+1 Canopy      7  0.001937046          1     0.001937046
+2   <NA>     NA           NA          2     0.058111380
+3   <NA>     NA           NA          3     0.058111380
+4   <NA>     NA           NA          4     0.058111380
+5   <NA>     NA           NA          5     0.058111380
+6   <NA>     NA           NA          6     0.058111380
+  ```
+
+4. When a species' leaf_gm_per_m2 was recorded at both the sites, obtained an average wieghted by the total trap area to accountfor differential sample sizes
 
@@
 
-6. For each species, prepended its leaf_gm_per_m2 data-frame by adding the number of prior dates equivalent to days in the species leaf lifespan, named 'sp.lifespan'  
+6.
+New: Using Joe's suggestion: Could you estimate living canopy leaf mass per unit ground area present at time t (L[t]) from mean leaf longevity and leaf mass fall per unit ground area at time t (F[t]) as follows:  L[t] = F[t+1] + F[t+2] + … + F[t + mean leaf longevity]
+
+    1.
+For each species, calculated a day of the year average of leaf_fall_LAI (m2/m2), weighted by total trap area of sites observed, to account for differential sampling intensity over time.
+This gives more weight to dates after 2013 when BCI-50ha site observations began.
+    2. LAI, for the changing fraction on the tree is obtained by 
+   looping over days t in 1:365 
+    LAI[t] <- sum(leaf_fall_LAI[c(t: (t+sp.lifespan)) %% 365])
+    where sp.lifespan are the number of days in species leaf lifespan, named 'sp.lifespan'
+
+```{r LAI-ssn-loop, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ssn-loop}Mean LAI seasonality curves based on Joe's calculation.
+LL = species-specific leaf lifespan in days.
+Species are color coded by deciduousness.If LL is estimated from LMA it is shown in solid line, if estimated from observations it is in a broken line.")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/LAI.seasonality_mean_loop.jpeg"))
+```
+
+## Comparison between LAI seasonality based on Rutuja's calculation (step #9, with Prior) vs. Joe's (Figure \@ref(fig:LAI-ssn-loop-mod))
+
+```{r LAI-ssn-loop-mod, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ssn-loop-mod}Mean LAI seasonality curves based on Rutuja's calculation (thick lines) vs.
+Joe's (thin lines).
+LL = species-specific leaf lifespan in days.
+Species are color coded by deciduousness.")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/LAI.seasonality_mean_loop_overlaid_mod.jpeg"))
+```
+
+# IGNORE THE REST
+6. Old : For each species, prepended its leaf_gm_per_m2 data-frame by adding the number of prior dates equivalent to days in the species leaf lifespan, named 'sp.lifespan'  
     Then LAI time series is calculated with the following steps: For t in 1 to c(total no of days in the data frame - leaf.lifespan),   

### 2021-01-29 LAI.Rmd
1. Added new gap-filling approach, prioritising genus and family level over site-level and LMA based rel 2. addeda graph with color-coding by the source of leaf-lifespan

@@
 
-1.
-Built leaf lifespan (LL) to LMA relationship for San-Lorenzo (FTS) based on canopy leaves and those with sufficient sample size to determine (n_lifespan > 100) (Figure \@ref(fig:ll-lma)).
+2.
+Gap-filled species-level mean leaf lifespan, with genus-level mean.
+If genus-level not available, then family-level mean.
+When neither were available: (1) for deciduous species, used mean lifespan at PNM site, (2) for evergreen species, used a relationship of leaf lifespan (LL) with LMA.
+This LMA-based relationship was based on data at San-Lorenzo (FTS) for canopy leaves and for species with sufficient sample sizes (n_lifespan > 100) (Figure \@ref(fig:ll-lma)).
 Per Joe, BCI climate is more similar to San Lorenzo, while PNM site is much drier.
@@
 So, used the leaf lifespan-LMA relationship for San Lorenzo.
-This was used to predict leaf lifespan data for species without leaf lifespan data based on their LMA, after gap-filling species LMA data.
-For the latter, I used LMA_lamina, which was in turn  gap-filled based on a linear relationship with LMALEAF_AVD or LMADISC_AVD.
+This was used to predict leaf lifespan data for species without leaf lifespan data based on their LMA, after gap-filling species LMA data:
+    Gap-filled LMALAM_AVD with the following sequence of priority: 
+    LMALAM_AVD relationship with LMADISC_AVD > LMALAM_AVD rel.
+with LMALEAF_AVD > Genus-level mean LMALAM_AVD (gap-filled with LMADISC_AVD/LMALEAF_AVD) > Family-level mean LMALAM_AVD (gap-filled with LMADISC_AVD/LMALEAF_AVD)
 
@@
 This gives more weight to dates after 2013 when BCI-50ha site observations began.
-    2. LAI, for the changing fraction on the tree is obtained by 
+(Figure \@ref(fig:leaf_fall_LAI)))
+    2. LAI, for the changing fraction on the tree is obtained (Figure \@ref(fig:LAI-ssn-loop)) by   
     looping over days t in 1:365   
@@
     where sp.lifespan are the number of days in species leaf lifespan, named 'sp.lifespan'  
+    To normalise LAI, it is the divided by the maximum value.
 
-```{r LAI-ssn-loop, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ssn-loop}Mean LAI seasonality curves based on Joe's calculation.
+```{r leaf_fall_LAI, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{leaf_fall_LAI}Seasonality curves for Leaf Fall in LAI units based on Joe's calculation.
 LL = species-specific leaf lifespan in days.
-Species are color coded by deciduousness.If LL is estimated from LMA it is shown in solid line, if estimated from observations it is in a broken line.")}
+Species are color coded by deciduousness.")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/Leaf_fall_LAI.seasonality_mean.jpeg"))
+```
+
+```{r LAI-ssn-loop, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ssn-loop}Mean LAI seasonality curves based on Joe's approach to calculation.
+LL = species-specific leaf lifespan in days.
+Species are color coded by deciduousness.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/LAI.seasonality_mean_loop.jpeg"))
+```
+
+```{r LAI-ssn-loop-subcol, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ssn-loop-subcol}Mean LAI seasonality curves based on Joe's approach to calculation.
+LL = species-specific leaf lifespan in days.
+Species are color coded by the method/source of leaf-lifespan substitution.")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/LAI.seasonality_mean_loop_color_substitution.jpeg"))
 ```
@@
 # IGNORE THE REST
-6. Old : For each species, prepended its leaf_gm_per_m2 data-frame by adding the number of prior dates equivalent to days in the species leaf lifespan, named 'sp.lifespan'  
+6. Old : For each species, prepended its leaf_gm_per_m2 data-frame by adding the number of prior dates equivalent to days in the species leaf lifespan
     Then LAI time series is calculated with the following steps: For t in 1 to c(total no of days in the data frame - leaf.lifespan),   

### 2021-02-03 LAI.Rmd
Removed Rutuja's LAI calculation

@@
 ---
-title: "LAI workflow documentation"
+title: "LAI Workflow Documentation"
 author: "Rutuja Chitra-Tarak"
@@
 This gives more weight to dates after 2013 when BCI-50ha site observations began.
-(Figure \@ref(fig:leaf_fall_LAI)))
+(Figure \@ref(fig:leaf-fall-LAI)))
     2. LAI, for the changing fraction on the tree is obtained (Figure \@ref(fig:LAI-ssn-loop)) by   
@@
     LAI[t] <- sum(leaf_fall_LAI[c(t: (t+sp.lifespan)) %% 365])  
-    where sp.lifespan are the number of days in species leaf lifespan, named 'sp.lifespan'  
+    where sp.lifespan are the number of days in species leaf lifespan  
     To normalise LAI, it is the divided by the maximum value.
 
-```{r leaf_fall_LAI, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{leaf_fall_LAI}Seasonality curves for Leaf Fall in LAI units based on Joe's calculation.
+```{r leaf-fall-LAI, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{leaf-fall-LAI}Seasonality curves for Leaf Fall in LAI units based on Joe's calculation.
 LL = species-specific leaf lifespan in days.
@@
 ```
-
-## Comparison between LAI seasonality based on Rutuja's calculation (step #9, with Prior) vs. Joe's (Figure \@ref(fig:LAI-ssn-loop-mod))
-
-```{r LAI-ssn-loop-mod, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ssn-loop-mod}Mean LAI seasonality curves based on Rutuja's calculation (thick lines) vs.
-Joe's (thin lines).
-LL = species-specific leaf lifespan in days.
-Species are color coded by deciduousness.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/LAI.seasonality_mean_loop_overlaid_mod.jpeg"))
-```
-
-# IGNORE THE REST
-6. Old : For each species, prepended its leaf_gm_per_m2 data-frame by adding the number of prior dates equivalent to days in the species leaf lifespan
-    Then LAI time series is calculated with the following steps: For t in 1 to c(total no of days in the data frame - leaf.lifespan),   
-    1. Assuming zero LAI to begin with, leaf gain on date 1 is leaf fall for the future date shifted by sp.lifespan. That is,    
-    If t == 1,  
-    LAI[t] = leaf_fall_LAI[sp.lifespan]  
-    1. New leaf gain is added to LAI of the previous date, but leaf fall does not begin until the future date equivalent to sp.lifespan. So,  
-    For t > 1, if t < sp.lifespan,   
-    LAI[t] <- LAI[t-1] + leaf_fall_LAI[sp.lifespan]  
-    1. Leaf fall begins on date equivalent to sp.lifespan. So,  
-    For t > 1, if t >= sp.lifespan,  
-    LAI[t] <- LAI[t-1] + leaf_fall_LAI[sp.lifespan] - leaf_fall_LAI[t]  
-    1.
-Because dates before the date when leaf-fall begins, that is, the prepended dates from t = 1: sp.lifespan, would underestimate LAI because of lack of a prior, I removed the time series for the prepended dates.
-The remaining LAI time series for each species is shown in (Figure \@ref(fig:LAI-ts)).
-
-```{r LAI-ts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ts}LAI time series.
-LL = species-specific leaf lifespan in days.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/LAI_ts.jpeg"))
-```
-
-7. Per Joe, for some species LAI increases/decreases over time likely due to change in population levels. Therefore detrended the LAI time series by removing the slope (Figure \@ref(fig:LAI-ts-det)).
-
-```{r LAI-ts-det, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ts-det}Detrended LAI time series.
-LL = species-specific leaf lifespan in days.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/LAI_ts_detrended.jpeg"))
-```
-
-8.
-A species-specific LAI seasonality curve is then built by taking a mean of LAI by DOY for the detrended LAI time series.
-This curve is then normalized by dividing by the max LAI value across DOYs (Figure \@ref(fig:LAI-ssn)).
-
-```{r LAI-ssn, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ssn}Mean LAI seasonality curves.
-LL = species-specific leaf lifespan in days.
-Species are color coded by deciduousness.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/LAI.seasonality_mean.jpeg"))
-```
-
-9.
-The prior LAI on date 1 (Step 8.1 above) for calculating the LAI time series in Figure \@ref(fig:LAI-ts) could be estimated from the mean LAI estimate for the DOY of day 1 Figure \@ref(fig:LAI-ssn)).
-So Step 8.1 becomes: If t == 1,  LAI[t] =  Expected_LAI[t] + leaf_fall_LAI[sp.lifespan].
-Using this prior, and following rest of the steps 2-4, the resultant LAI time series is (Figure \@ref(fig:LAI-ts-mod)), and after detrending, mean LAI by DOY time series is calculated.
-These are shown in Figure \@ref(fig:LAI-ssn-mod) as thick lines overlaid on LAI by DOY curves calculated without a prior (thin lines).
-Using a prior significantly reduces the dip in LAI for many a species.
-This may be more realistic for evergreen species (Joe, your thoughts?).
-
-```{r LAI-ts-mod, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ts-mod}LAI time series based on a prior LAI estimate.
-LL = species-specific leaf lifespan in days.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/LAI_ts_mod.jpeg"))
-```
-
-```{r LAI-ssn-mod, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ssn-mod}Mean LAI seasonality curves based on a prior LAI estimate (thick lines) overlaid on those without a prior (thin lines).
-LL = species-specific leaf lifespan in days.
-Species are color coded by deciduousness.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/LAI.seasonality_mean_overlaid_mod.jpeg"))
-```
-
-10.
-To make non-evergreen species fully deciduous two options are tried: (1) By rescaling the *whole* mean LAI seasonality curves to range from 0-1 (Figure \@ref(fig:LAI-ssn-mod-deci-all); rescale1 <- function (x) {x/max(x)}.
-This hardly allows trees to have full LAI through the wet season.
-Especially, see SPONMO and TRATAS.
-Therefore (2), I rescaled *only the lowest periods* of mean LAI seasonality curves in Figure \@ref(fig:LAI-ssn-mod-deci-all)) such that the minimum is zero (rescale0 <- function (x) {c(x - min(x))/(max(x) - min(x))}.
-Brevideciduous species are deciduous for a short while, while obligate deciduous species for longer; so I varied the length of the deciduousness period based on deciduous type: For brevideciduous I chose the part of the curve for which LAI was less than 0.15 quantile of the LAI distribution in the mean seasonality curve; for facultative deciduous species < 0.2 quantile; and for obligate deciduous species < 0.3 quantile.
-These modified curves are shown in figure (Figure \@ref(fig:LAI-ssn-mod-deci)).
-
-```{r LAI-ssn-mod-deci-all, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ssn-mod-deci-all}Mean LAI seasonality curves based on a prior LAI estimate are used: Curves are then rescaled so that for evergreen species the maximum is 1, while for non-evergreen species, the range is 0-1.
-Species are color coded by deciduousness.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/LAI.seasonality_mean_mod_deci_unrestricted_range01.jpeg"))
-```
-
-```{r LAI-ssn-mod-deci, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ssn-mod-deci}Mean LAI seasonality curves based on a prior LAI estimate are used.
-Then all species' curves are first rescaled so that the maximum is 1.
-For non-evergreen species, lowest periods based on quantiles are rescaled further such the minimum is 0.
-Species are color coded by deciduousness.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/LAI.seasonality_mean_mod_deci_restricted.jpeg"))
-```

### 2021-02-06 LAI.Rmd
Reformatted adn updated

@@
 
+1. Gap-filled species-level mean leaf lifespan with observations from crane sites (data referred to as "cohort""); if not with genus-level mean; else family-level mean. When neither were available: 
+    1. For deciduous species, used mean lifespan at PNM site
     2.
-Gap-filled species-level mean leaf lifespan, with genus-level mean.
-If genus-level not available, then family-level mean.
-When neither were available: (1) for deciduous species, used mean lifespan at PNM site, (2) for evergreen species, used a relationship of leaf lifespan (LL) with LMA.
+For evergreen species, used a relationship of leaf lifespan (LL) with LMA.
 This LMA-based relationship was based on data at San-Lorenzo (FTS) for canopy leaves and for species with sufficient sample sizes (n_lifespan > 100) (Figure \@ref(fig:ll-lma)).
+
     Per Joe, BCI climate is more similar to San Lorenzo, while PNM site is much drier.
@@
     mutate(leaf_gm_rate = leaf_gm_per_m2/n.days,
-           # approx() will interpolate this forward until the next census,
+           # approx() in the next function will interpolate this forward until the next census,
            # we want it to interpolate this value backward over the past census interval,
@@
 6.
-New: Using Joe's suggestion: Could you estimate living canopy leaf mass per unit ground area present at time t (L[t]) from mean leaf longevity and leaf mass fall per unit ground area at time t (F[t]) as follows:  L[t] = F[t+1] + F[t+2] + … + F[t + mean leaf longevity]
+Joe's suggestion: Estimate living canopy leaf mass per unit ground area present at time t (L[t]) from mean leaf longevity and leaf mass fall per unit ground area at time t (F[t]) as follows:  L[t] = F[t+1] + F[t+2] + … + F[t + mean leaf longevity]
 
@@
 
-```{r leaf-fall-LAI, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{leaf-fall-LAI}Seasonality curves for Leaf Fall in LAI units based on Joe's calculation.
+```{r leaf-fall-LAI, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{leaf-fall-LAI}Seasonality curves for Leaf Fall in LAI units.
 LL = species-specific leaf lifespan in days.
@@
 
-```{r LAI-ssn-loop, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ssn-loop}Mean LAI seasonality curves based on Joe's approach to calculation.
+```{r LAI-ssn-loop, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ssn-loop}Mean LAI seasonality curves.
 LL = species-specific leaf lifespan in days.
@@
 
-```{r LAI-ssn-loop-subcol, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ssn-loop-subcol}Mean LAI seasonality curves based on Joe's approach to calculation.
-LL = species-specific leaf lifespan in days.
-Species are color coded by the method/source of leaf-lifespan substitution.")}
+```{r LAI-ssn-loop-subcol, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "90%", results = "asis", fig.cap = paste0("\\label{LAI-ssn-loop-subcol}Mean LAI seasonality curves with species color coded by the source of leaf-lifespan observation/substitution.
+LL = species-specific leaf lifespan in days.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/LAI.seasonality_mean_loop_color_substitution.jpeg"))

### 2021-02-18 09.0_manuscript.rmd
changed to groundhog

@@
 ```{r, include=FALSE, echo = FALSE}
-if(!require("pacman", quietly=TRUE)) install.packages("pacman")
-suppressPackageStartupMessages(
-  pacman::p_load(bookdown, knitr, pander, citr, broom))
+if (!require("groundhog", quietly=TRUE)) install.packages("groundhog"); library(groundhog)
+groundhog.day = "2020-04-01"
+pkgs=c('knitr', 'pander', 'citr', 'broom')
+suppressPackageStartupMessages(groundhog.library(pkgs, groundhog.day))
 options(tinytex.verbose = TRUE)
@@
 
-\
-We removed those species that would be leafless in March 1997 because $\mathrm{\delta}$^2^H~xylem~ for these species may not be linked with water sourced at the time of measurement and thus not comparable to species that had leaves.
-We also removed one species whose $\mathrm{\delta}$^2^H~xylem~ was in the uncertain depth range.
+\We removed one species with a $\mathrm{\delta}$^2^H value of -33.5 (SE$\pm$2.6) which is particularly difficult to assign a source-water depth given the uncertainty in soil water $\mathrm{\delta}$^2^H data by depth (see Fig S3).
+$\mathrm{\delta}$^2^H~xylem~ from leafless trees may not be linked with water sourced at the time of measurement and thus may not be comparable to species that had leaves.
+Leaflessness status of sampled trees is not recorded in [@Meinzer:2001].
+We therefore removed five species that are typically leafless in March (Joseph Wright, personal observation), that is, the month of $\mathrm{\delta}$^2^H~xylem~ sampling.
 For each model of ERD (Eq.

### 2021-02-18 code/06.0_load.R

@@ -6,7 +6,7 @@
 # Original date: Summer, 2020
 #-----------------------------------------------------
 #
-# rm(list=ls())
+rm(list=ls())
 #
 # #*******************************************
 # ####   Load Libraries, Prep for graphics, folders  ####
@@ -14,8 +14,12 @@
 # #### Written  with R version 3.6.3 ###
 # #*******************************************
 if (!require("groundhog")) install.packages("groundhog"); library(groundhog)
+set.groundhog.folder('/Users/ftuser/Library/R/3.6/library')
 groundhog.day = "2020-04-01"
-pkgs=c('lattice', 'magick', 'cowplot', 'corrplot', 'ggcorrplot', 'tidyverse', 'readxl', 'forcats', 'scales', 'data.table', 'ggpmisc', 'GGally')
+groundhog.library('devtools', groundhog.day)
+devtools::unload('lattice') # this is to remove a newer version that conflicts with what groundhog needs
+pkgs=c('lattice', 'magick', 'cowplot', 'corrplot', 'ggcorrplot', 'tidyverse', 'readxl',
+       'forcats', 'scales', 'data.table', 'ggpmisc', 'GGally', 'bookdown', 'rmarkdown')
 groundhog.library(pkgs, groundhog.day)
 
 # graphics info

### 2021-02-18 09.0_manuscript.rmd
reverted groundhog

@@
 ```{r, include=FALSE, echo = FALSE}
-if (!require("groundhog", quietly=TRUE)) install.packages("groundhog"); library(groundhog)
-groundhog.day = "2020-04-01"
-pkgs=c('knitr', 'pander', 'citr', 'broom')
-suppressPackageStartupMessages(groundhog.library(pkgs, groundhog.day))
+if (!require("pacman")) install.packages("pacman"); library(pacman)
+pacman::p_load(knitr, pander, citr, broom)
 options(tinytex.verbose = TRUE)

### 2021-02-18 code/05.0_Prepare_data_for_correlative analyses.R

@@ -8,20 +8,10 @@
 
 rm(list=ls())
 
-#*******************************************
-####   Load Libraries, Prep for graphics, folders  ####
-#*******************************************
-#### Written  with R version 3.6.3 ###
-#*******************************************
-if (!require("groundhog")) install.packages("groundhog"); library(groundhog)
-groundhog.day = "2020-04-01"
-pkgs=c('tidyverse','readxl', 'forcats', 'scales', 'data.table', 'ggpmisc', 'GGally',
-       'agricolae', 'gridExtra', 'zoo', 'Evapotranspiration',
-       'data.table', 'mgcv', 'lubridate', 'smooth', 'viridis')
-groundhog.library(pkgs, groundhog.day)
-library('bci.hydromet') # This is processed BCI climate data with Eddy Flux tower data
-library('bci.elm.fates.hydrology') # This has calibrated ELM-FATES Psi
-
+if (!require("pacman")) install.packages("pacman"); library(pacman)
+pacman::p_load(tidyverse, readxl, forcats, agricolae, gridExtra, zoo,
+               scales, GGally, ggpmisc, Evapotranspiration,
+               data.table, bci.elm.fates.hydro, mgcv, lubridate, smooth, bci.hydromet, viridis)
 # graphics info
 theme_set(theme_bw())
 theme_update(text = element_text(size = 14),
@@ -133,7 +123,7 @@ deci <- read_excel(file.path("data-raw/traits/nomenclature_R_20190524_Rready_Osv
 deci.level_key <- c("Evg" = "1", "DF" = "2", "DB" = "3", "DO" = "4") #c(a = "apple", b = "banana", c = "carrot")
 
 deci <- deci %>% mutate(sp = tolower(sp6)) %>%
-  dplyr::select(sp4, sp, deciduous) %>%
+  select(sp4, sp, deciduous) %>%
   subset(deciduous %in% c("E", "DF", "DO", "DB")) %>% droplevels() %>%
   mutate(deciduousness = recode_factor(as.factor(deciduous), `E` = "Evergreen", `DB` = "Brevideciduous",
                                        `DF` = "Facultative Deciduous", `DO` = "Obligate Deciduous"),
@@ -148,14 +138,14 @@ deci <- deci %>% mutate(sp = tolower(sp6)) %>%
[…]

### 2021-02-18 code/07.0_PhenoDemoTraitsPsi.R

@@ -22,14 +22,10 @@ source("code/06.0_load.R")
 ####   Load Libraries, Prep for graphics, folders  ##
 #*******************************************
 
-if (!require("groundhog")) install.packages("groundhog"); library(groundhog)
-groundhog.day = "2020-04-01"
-pkgs=c('tidyverse', 'readxl', 'forcats', 'agricolae', 'gridExtra',
-       'scales', 'GGally', 'ggpmisc', 'Evapotranspiration',
-       'data.table', 'mgcv')
-
-groundhog.library(pkgs, groundhog.day)
-
+if (!require("pacman")) install.packages("pacman"); library(pacman)
+pacman::p_load(tidyverse, readxl, forcats, agricolae, gridExtra,
+               scales, GGally, ggpmisc, Evapotranspiration,
+               data.table, bci.elm.fates.hydro, mgcv)
 # graphics info
 theme_set(theme_bw())
 theme_update(text = element_text(size = 14),

### 2021-02-18 code/08.0_manuscript_data_figs.R

@@ -384,14 +384,14 @@ hyd.table <-  erd.stem.traits %>%
   left_join(erd.sp.names, by = c("sp")) %>%
   subset(sp %in% erd.sp) %>%
   select(-sp) %>%
-  select(Genus, Species, Family, lwp.min_Predawn, TLP, everything()) %>%
+  relocate(Genus, Species, Family, lwp.min_Predawn, TLP) %>%
   mutate(lwp.min_Predawn = round(lwp.min_Predawn, 2),
          TLP = round(TLP, 2))
 
 erd.stem.traits.only <- erd.stem.traits %>%
   subset(!trait %in% c("lwp.min_Diurnal", "lwp.min_Predawn")) %>%
   left_join(df.erd.to.plot %>%
-              select(sp, depth, depth.se), by = "sp") %>%
+              dplyr::select(sp, depth, depth.se), by = "sp") %>%
   subset(!is.na(depth)) %>%
   droplevels()
 erd.stem.traits.sp <- unique(erd.stem.traits.only$sp)

### 2021-02-18 code/01.0_Preparing_stem_growth_time_series.R
reverted groundhog

@@ -7,17 +7,8 @@
 # for 50 ha obs species groups
 rm(list = ls())
 gc()
-
-#*******************************************
-####   Load Libraries, Prep for graphics, folders  ####
-#*******************************************
-#### Written with R version 3.6.3 ###
-#*******************************************
-if (!require("groundhog")) install.packages("groundhog"); library(groundhog)
-groundhog.day = "2020-04-01"
-pkgs=c('tidyverse', 'purrrlyr', 'scales', 'grid', 'gridExtra', 'lme4')
-groundhog.library(pkgs, groundhog.day)
-
+if (!require("pacman")) install.packages("pacman"); library(pacman)
+pacman::p_load(tidyverse, purrrlyr, scales, grid, gridExtra, lme4)
 
 # graphics info
 theme_set(theme_bw())

### 2021-02-18 code/02.0_demographic_data_BCI.R

@@ -6,17 +6,9 @@
 
 rm(list=ls())
 gc()
-
-#*******************************************
-####   Load Libraries, Prep for graphics, folders  ####
-#*******************************************
-#### Written with R version 3.6.3 ###
-#*******************************************
-if (!require("groundhog")) install.packages("groundhog"); library(groundhog)
-groundhog.day = "2020-04-01"
-pkgs=c('tidyverse', 'hms', 'ggpmisc')
-groundhog.library(pkgs, groundhog.day)
-
+# load("/Library/Frameworks/R.framework/Versions/3.4/Resources/library/CTFSRPackage/CTFSRPackage.Rdata")
+if (!require("pacman")) install.packages("pacman"); library(pacman)
+pacman::p_load(tidyverse, hms, ggpmisc)
 
 # graphics info
 theme_set(theme_bw())

### 2021-02-18 code/04.0_distribution_qqplots_growth_rate_inverse_model_residuals_UDI.R

@@ -6,18 +6,10 @@
 ###-------------------------------
 
 rm(list = ls())
+
 gc()
-
-#*******************************************
-####   Load Libraries, Prep for graphics, folders  ####
-#*******************************************
-#### Written  with R version 3.6.3 ###
-#*******************************************
-if (!require("groundhog")) install.packages("groundhog"); library(groundhog)
-groundhog.day = "2020-04-01"
-pkgs=c('tidyverse', 'gridExtra', 'spatstat', 'MASS')
-groundhog.library(pkgs, groundhog.day)
-
+if (!require("pacman")) install.packages("pacman")
+p_load(tidyverse, gridExtra, bci.elm.fates.hydro, spatstat, MASS)
 figures.folder <- paste0("figures/Distributions")
 if(!dir.exists(file.path(figures.folder))) {dir.create(file.path(figures.folder))}
 

### 2021-02-18 10.0_Supporting_Information.Rmd

--- 2020-11-10 code/10.0_Supporting_Information.Rmd
@@
 options(tinytex.verbose = TRUE)
-source("code/07.0_load.R")
+source("code/06.0_load.R")
 source("code/08.0_manuscript_data_figs.R")

### 2021-02-18 10.0_Supporting_Information.Rmd
Revert "reverted groundhog"
This reverts commit 2855abdeeba5c4cba9cdcfae1ad149e044029a51.

Really trying to get groundhog back

@@
 \
-We collected volumetric water content (VWC) data (2012-2018) from three locations near the eddy flux tower with time domain reflectometry (TDR, CS616, Campbell Scientific) probes inserted with an angle to average the first 15 cm of soil.
+We collected volumetric water content (VWC) data (2012-2018) from three locations near the eddy flux tower with time domain reflectometry (TDR, CS616, Campbell Scientific) probes inserted vertically ranging over the first 15 cm of soil.
 The relationship between the output from the TDR probe and VWC was calibrated based on measurements of gravimetric water content (GWC) on several occasions from representative soils near the vertical probes (0-0.15 m) (Fig.
@@
 We interpolated species-specific weekly leaf-fall data collected at two similar sites in BCI (1985-2020) to a daily estimate [@Wright.1990aeo].
-We lead this daily leaf-fall time series by the species-specific average leaf-lifetime (in days) and used it as an estimate of the daily leaf-gain time-series.
+We lead this daily leaf-fall time series by the species-specific average leaf lifespan (in days) and used it as an estimate of the daily leaf-gain time-series.
 The difference between leaf-fall and leaf-gain gave an estimate of daily leaf-change time series, which was then scaled to express as a fraction of the total leaf fall in a given year.
@@
 \
-For five species, an estimate of leaf lifetime was available from direct observations [@Osnas.2018].
-For the rest `r erd.sp.wo.ll` species, leaf lifetime was predicted from LMA ($\overline{lifetime} = `r ll.lma.cf[1]` + `r ll.lma.cf[2]`LMA$; $R$^2^ = `r ll.lma.r2`; $p$ `r ll.lma.p`, $n$ = `r ll.lma.n`).
+For five species, an estimate of leaf lifespan was available from direct observations [@Osnas.2018].
+For the rest `r erd.sp.wo.ll` species, leaf lifespan was predicted from LMA ($\overline{lifespan} = `r ll.lma.cf[1]` + `r ll.lma.cf[2]`LMA$; $R$^2^ = `r ll.lma.r2`; $p$ `r ll.lma.p`, $n$ = `r ll.lma.n`).
 

### 2021-02-18 code/05.0_Prepare_data_for_correlative analyses.R

@@ -8,10 +8,20 @@
 
 rm(list=ls())
 
-if (!require("pacman")) install.packages("pacman"); library(pacman)
-pacman::p_load(tidyverse, readxl, forcats, agricolae, gridExtra, zoo,
-               scales, GGally, ggpmisc, Evapotranspiration,
-               data.table, bci.elm.fates.hydro, mgcv, lubridate, smooth, bci.hydromet, viridis)
+#*******************************************
+####   Load Libraries, Prep for graphics, folders  ####
+#*******************************************
+#### Written  with R version 3.6.3 ###
+#*******************************************
+if (!require("groundhog")) install.packages("groundhog"); library(groundhog)
+groundhog.day = "2020-04-01"
+pkgs=c('tidyverse','readxl', 'forcats', 'scales', 'data.table', 'ggpmisc', 'GGally',
+       'agricolae', 'gridExtra', 'zoo', 'Evapotranspiration',
+       'data.table', 'mgcv', 'lubridate', 'smooth', 'viridis')
+groundhog.library(pkgs, groundhog.day)
+library('bci.hydromet') # This is processed BCI climate data with Eddy Flux tower data
+library('bci.elm.fates.hydrology') # This has calibrated ELM-FATES Psi
+
 # graphics info
 theme_set(theme_bw())
 theme_update(text = element_text(size = 14),
@@ -123,7 +133,7 @@ deci <- read_excel(file.path("data-raw/traits/nomenclature_R_20190524_Rready_Osv
 deci.level_key <- c("Evg" = "1", "DF" = "2", "DB" = "3", "DO" = "4") #c(a = "apple", b = "banana", c = "carrot")
 
 deci <- deci %>% mutate(sp = tolower(sp6)) %>%
-  select(sp4, sp, deciduous) %>%
+  dplyr::select(sp4, sp, deciduous) %>%
   subset(deciduous %in% c("E", "DF", "DO", "DB")) %>% droplevels() %>%
   mutate(deciduousness = recode_factor(as.factor(deciduous), `E` = "Evergreen", `DB` = "Brevideciduous",
                                        `DF` = "Facultative Deciduous", `DO` = "Obligate Deciduous"),
@@ -138,14 +148,14 @@ deci <- deci %>% mutate(sp = tolower(sp6)) %>%
[…]

### 2021-02-20 09.0_manuscript.rmd
updated groundhog.day and updated to make sure package-functions work

@@
 
-```{r, include=FALSE, echo = FALSE}
-if (!require("pacman")) install.packages("pacman"); library(pacman)
-pacman::p_load(knitr, pander, citr, broom)
+```{r setup1, include=FALSE, echo = FALSE}
+if (!require("groundhog", quietly=TRUE)) install.packages("groundhog"); library(groundhog)
+groundhog.day = "2020-05-01"
+pkgs=c('knitr', 'pander', 'citr', 'broom')
+suppressPackageStartupMessages(groundhog.library(pkgs, groundhog.day))
 options(tinytex.verbose = TRUE)
@@
 
-```{r setup, echo = FALSE}
-library(knitr)
+```{r setup2, echo = FALSE}
 opts_knit$set(eval.after = "fig.cap")

### 2021-02-20 10.0_Supporting_Information.Rmd
updated groundhog.day and updated to make sure package-functions work

@@
 
-```{r, include=FALSE, echo = FALSE}
+```{r setup, include=FALSE, echo = FALSE}
 if(!require("pacman", quietly=TRUE)) install.packages("pacman")
@@
 ```{r setup2, echo = FALSE}
-library(knitr)
 opts_knit$set(eval.after = "fig.cap")

### 2021-02-20 code/05.0_Prepare_data_for_correlative analyses.R

@@ -11,11 +11,11 @@ rm(list=ls())
 #*******************************************
 ####   Load Libraries, Prep for graphics, folders  ####
 #*******************************************
-#### Written  with R version 3.6.3 ###
+#### Written  with R version 4.0.3 ###
 #*******************************************
 if (!require("groundhog")) install.packages("groundhog"); library(groundhog)
-groundhog.day = "2020-04-01"
-pkgs=c('tidyverse','readxl', 'forcats', 'scales', 'data.table', 'ggpmisc', 'GGally',
+groundhog.day = "2020-05-01"
+pkgs=c('magick', 'cowplot', 'tidyverse','readxl', 'forcats', 'scales', 'data.table', 'ggpmisc', 'GGally',
        'agricolae', 'gridExtra', 'zoo', 'Evapotranspiration',
        'data.table', 'mgcv', 'lubridate', 'smooth', 'viridis')
 groundhog.library(pkgs, groundhog.day)
@@ -614,7 +614,8 @@ gpp.rel.daily <- obs.gpp.d %>%
   subset(date < max(obs.gpp.d$date, na.rm = TRUE)) %>%
   droplevels()
 
-gpp.rel.daily.long <- gpp.rel.daily %>% pivot_longer(-date, values_to = "value", names_to = "variable")
+gpp.rel.daily.long <- gpp.rel.daily %>% dplyr::select(-.groups) %>%
+  pivot_longer(-date, values_to = "value", names_to = "variable")
 
 gpp.rel.daily.plot <- ggplot(gpp.rel.daily.long) +
   # geom_rect(data=rectangles, aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax),
@@ -1070,7 +1071,10 @@ traits.long <- traits.long %>%
          deci_sp.plot = factor(deci_sp, levels=unique(deci_sp[order(deciduousness)]), ordered=TRUE))
 # just for sp with hyd.traits, but traits.long does not have all those sp, and hab preference and WSG traits will be missed
 ## so beginning with those other traits
-traits.wide <- traits.long %>% dplyr::select(-trait.plot, -trait.plot.chart) %>% pivot_wider(names_from = trait, values_from = value)
+traits.wide <- traits.long %>% dplyr::select(-trait.plot, -trait.plot.chart) %>%
+  mutate(row = row_number()) %>%
[…]

### 2021-02-20 code/06.0_load.R

@@ -11,16 +11,15 @@ rm(list=ls())
 # #*******************************************
 # ####   Load Libraries, Prep for graphics, folders  ####
 # #*******************************************
-# #### Written  with R version 3.6.3 ###
+# #### Written  with R version 4.0.3 ###
 # #*******************************************
-if (!require("groundhog")) install.packages("groundhog"); library(groundhog)
-set.groundhog.folder('/Users/ftuser/Library/R/3.6/library')
-groundhog.day = "2020-04-01"
-groundhog.library('devtools', groundhog.day)
-devtools::unload('lattice') # this is to remove a newer version that conflicts with what groundhog needs
-pkgs=c('lattice', 'magick', 'cowplot', 'corrplot', 'ggcorrplot', 'tidyverse', 'readxl',
-       'forcats', 'scales', 'data.table', 'ggpmisc', 'GGally', 'bookdown', 'rmarkdown')
-groundhog.library(pkgs, groundhog.day)
+# if (!require("groundhog")) install.packages("groundhog"); library(groundhog)
+# # set.groundhog.folder('/Users/ftuser/Library/R/3.6/library')
+# groundhog.day = "2020-05-01"
+# groundhog.library('devtools', groundhog.day)
+# pkgs=c('lattice', 'magick', 'cowplot', 'corrplot', 'ggcorrplot', 'tidyverse', 'readxl',
+#        'forcats', 'scales', 'data.table', 'ggpmisc', 'GGally', 'bookdown', 'rmarkdown', 'broom')
+# groundhog.library(pkgs, groundhog.day)
 
 # graphics info
 theme_set(theme_bw())
@@ -41,7 +40,7 @@ if(!dir.exists(file.path(results.folder))) {dir.create(file.path(results.folder)
 ## Load ELM-FATES best-fit parameters -----
 #******************************************************
 params.obj.top.few <- read.csv(file = file.path("results/2019-10-14_5000/params.obj.top.few_100.csv"), header = TRUE)
-p.best <- params.obj.top.few[,1:13]
+p.best <- params.obj.top.few[, 1:13]
 #******************************************************
 ## Load Deciduousness-----
 #******************************************************
@@ -110,8 +109,8 @@ growth <- get(growth.name); rm(growth.name)
 ## No. of trees by sp for grpwth data
 g.n <- lapply(growth[grep("large", names(growth))], as.data.frame) %>%
[…]

### 2021-02-20 code/07.0_PhenoDemoTraitsPsi.R

@@ -23,7 +23,7 @@ source("code/06.0_load.R")
 #*******************************************
 
 if (!require("groundhog")) install.packages("groundhog"); library(groundhog)
-groundhog.day = "2020-04-01"
+groundhog.day = "2020-05-01"
 pkgs=c('tidyverse', 'readxl', 'forcats', 'agricolae', 'gridExtra',
        'scales', 'GGally', 'ggpmisc', 'Evapotranspiration',
        'data.table', 'mgcv')

### 2021-02-20 code/08.0_manuscript_data_figs.R

@@ -374,7 +374,7 @@ mrate.mfac.depth.gr.mean.mfac <- mrate.mfac.depth.select %>%
 # save.image("results/manuWorkSpace.RData")
 
 #****************************
-### Hydraulic traits vs. ERD---
+### Hydraulic traits vs. ERD----
 #****************************
 
 hyd.table <-  erd.stem.traits %>%
@@ -383,15 +383,15 @@ hyd.table <-  erd.stem.traits %>%
   pivot_wider(names_from = trait, values_from = value) %>%
   left_join(erd.sp.names, by = c("sp")) %>%
   subset(sp %in% erd.sp) %>%
-  select(-sp) %>%
-  select(Genus, Species, Family, lwp.min_Predawn, TLP, everything()) %>%
+  dplyr::select(-sp) %>%
+  dplyr::select(Genus, Species, Family, lwp.min_Predawn, TLP, everything()) %>%
   mutate(lwp.min_Predawn = round(lwp.min_Predawn, 2),
          TLP = round(TLP, 2))
 
 erd.stem.traits.only <- erd.stem.traits %>%
   subset(!trait %in% c("lwp.min_Diurnal", "lwp.min_Predawn")) %>%
   left_join(df.erd.to.plot %>%
-              select(sp, depth, depth.se), by = "sp") %>%
+              dplyr::select(sp, depth, depth.se), by = "sp") %>%
   subset(!is.na(depth)) %>%
   droplevels()
 erd.stem.traits.sp <- unique(erd.stem.traits.only$sp)
@@ -425,9 +425,9 @@ erd.pairs <- erd.stem.traits.only.lab %>%
   pivot_wider(names_from = trait.plot.chart, values_from = value) %>%
   rename(ERD = depth)
 
-p_load(corrplot, ggcorrplot)
+
 cor.data <- erd.pairs %>% dplyr::select(-sp) %>%
-  relocate(ERD, `italic(\"K\")[\"max,stem\"]`)
+  dplyr::select(ERD, `italic(\"K\")[\"max,stem\"]`, everything())
 M <- cor(cor.data, use = "pairwise.complete.obs", method = "spearman")
 
 res1 <- cor.mtest(cor.data, conf.level = 0.95,
@@ -471,9 +471,9 @@ corrplot(M, type = "upper", order = "original",
          insig = "p-val", sig.level = -1, pch.col = "grey90",
          diag = FALSE, cl.ratio = .2, cl.align = "l", win.asp= 1)
 graphics.off()
[…]

### 2021-02-20 Readme.Rmd
Cleaned

@@
 
-Script to load data generated from the data preparation script above for use in inverse model as well as manuscript
+Script to load data generated from the data preparation script above for use in inverse model as well as manuscript. It's needed for rest of the scripts.
 
-```{r Chunk 06.0, eval = FALSE}
-source("code/06.0_load.R")
-```
+<!-- ```{r Chunk 06.0, eval = FALSE} -->
+<!-- source("code/06.0_load.R") -->
+<!-- ``` -->
 
@@
 ```{r Chunk 08.0, eval = FALSE}
+source("code/06.0_load.R")
 source("code/08.0_manuscript_data_figs.R")
@@
 ```{r Chunk 09.0, eval = FALSE}
-source("code/09.0_manuscript.rmd")
+render("09.0_manuscript.rmd", output_format = "bookdown::html_document2")
 ```
@@
 ```{r Chunk 10.0, eval = FALSE}
-source("code/10.0_Supporting_Information.Rmd")
+render("10.0_Supporting_Information.Rmd", output_format = "bookdown::html_document2")
 ```

### 2021-02-22 09.0_manuscript.rmd
ground hog works with one package at a time

--- 2021-02-18 09.0_manuscript.rmd
@@
 
-```{r, include=FALSE, echo = FALSE}
-if (!require("pacman")) install.packages("pacman"); library(pacman)
-pacman::p_load(knitr, pander, citr, broom)
-options(tinytex.verbose = TRUE)
+```{r setup1, include=FALSE, echo = FALSE}
+if (!require("groundhog", quietly=TRUE)) install.packages("groundhog"); library(groundhog)
+groundhog.day = "2020-05-01"
+# pkgs=c('knitr', 'pander', 'citr', 'broom')
+groundhog.library('knitr', groundhog.day)
+groundhog.library('pander', groundhog.day)
+# groundhog.library('citr', groundhog.day)
+groundhog.library('broom', groundhog.day)
+
+# options(tinytex.verbose = TRUE)
 source("code/06.0_load.R")

### 2021-02-22 10.0_Supporting_Information.Rmd
ground hog works with one package at a time

@@
 
-```{r setup, include=FALSE, echo = FALSE}
-if(!require("pacman", quietly=TRUE)) install.packages("pacman")
-suppressPackageStartupMessages(
-  pacman::p_load(bookdown, knitr, pander, citr, broom))
+```{r setup1, include=FALSE, echo = FALSE}
+if (!require("groundhog", quietly=TRUE)) install.packages("groundhog"); library(groundhog)
+groundhog.day = "2020-05-01"
+# pkgs=c('knitr', 'pander', 'citr', 'broom')
+groundhog.library('knitr', groundhog.day)
+groundhog.library('pander', groundhog.day)
+# groundhog.library('citr', groundhog.day)
+groundhog.library('broom', groundhog.day)
 options(tinytex.verbose = TRUE)
@@
 source("code/08.0_manuscript_data_figs.R")
-# load("results/manuWorkSpace.RData")
-# install.packages(c("vctrs", "rlang"))
 ```

### 2021-02-22 code/06.0_load.R

@@ -13,13 +13,28 @@ rm(list=ls())
 # #*******************************************
 # #### Written  with R version 4.0.3 ###
 # #*******************************************
-# if (!require("groundhog")) install.packages("groundhog"); library(groundhog)
-# # set.groundhog.folder('/Users/ftuser/Library/R/3.6/library')
-# groundhog.day = "2020-05-01"
-# groundhog.library('devtools', groundhog.day)
+if (!require("groundhog")) install.packages("groundhog"); library(groundhog)
+# set.groundhog.folder('/Users/ftuser/Library/R/3.6/library')
+groundhog.day = "2020-05-01"
+groundhog.library('devtools', groundhog.day)
 # pkgs=c('lattice', 'magick', 'cowplot', 'corrplot', 'ggcorrplot', 'tidyverse', 'readxl',
 #        'forcats', 'scales', 'data.table', 'ggpmisc', 'GGally', 'bookdown', 'rmarkdown', 'broom')
-# groundhog.library(pkgs, groundhog.day)
+
+groundhog.library('lattice', groundhog.day)
+groundhog.library('magick', groundhog.day)
+groundhog.library('cowplot', groundhog.day)
+groundhog.library('corrplot', groundhog.day)
+groundhog.library('ggcorrplot', groundhog.day)
+groundhog.library('tidyverse', groundhog.day)
+groundhog.library('readxl', groundhog.day)
+groundhog.library('forcats', groundhog.day)
+groundhog.library('scales', groundhog.day)
+groundhog.library('data.table', groundhog.day)
+groundhog.library('ggpmisc', groundhog.day)
+groundhog.library('GGally', groundhog.day)
+groundhog.library('bookdown', groundhog.day)
+groundhog.library('rmarkdown', groundhog.day)
+groundhog.library('broom', groundhog.day)
 
 # graphics info
 theme_set(theme_bw())

### 2021-02-22 Readme.Rmd
reorganised

@@
 
-Script to load data generated from the data preparation script above for use in inverse model as well as manuscript. It's needed for rest of the scripts.
+Script to load data generated from the data preparation script above for use in inverse model as well as manuscript. It's used within the rest of the scripts.
 
@@
 
-**Figures**
-
-```{r Chunk 08.0, eval = FALSE}
-source("code/06.0_load.R")
-source("code/08.0_manuscript_data_figs.R")
-```
-
 **Main text**
-
 ```{r Chunk 09.0, eval = FALSE}
-render("09.0_manuscript.rmd", output_format = "bookdown::html_document2")
+# Two files are used within the report
+# For data, 06.0_load.R and 
+# For figures and data, code/08.0_manuscript_data_figs.R
+groundhog.day = "2020-05-01"
+groundhog.library('rmarkdown', groundhog.day)
+groundhog.library('bookdown', groundhog.day)
+rmarkdown::render("09.0_manuscript.rmd", output_format = "bookdown::html_document2")
 ```

### 2021-03-16 09.0_manuscript.rmd
1. resistance curve instead of vulnerability 2. updated hypothesis 2 3. drought exposure rather than hydraulic risk 4. updated species removal criteria 5. updated Methods narrative 6. updated ERD model description 7. Discussion: ERD model validation, note on capacitance, updated why the first study work

@@
 if (!require("groundhog", quietly=TRUE)) install.packages("groundhog"); library(groundhog)
-groundhog.day = "2020-05-01"
-# pkgs=c('knitr', 'pander', 'citr', 'broom')
-groundhog.library('knitr', groundhog.day)
-groundhog.library('pander', groundhog.day)
-# groundhog.library('citr', groundhog.day)
-groundhog.library('broom', groundhog.day)
+groundhog.day = "2021-01-01"
+pkgs=c('knitr', 'pander', 'broom')
+groundhog.library(pkgs, groundhog.day)
 
@@
 Here, we tested the role of deep-water access in enabling survival within a diverse tropical forest community in Panama using a novel data-model approach.
-- We inversely estimated the effective rooting depth (ERD, as the average depth of water extraction), for 32 canopy species by linking diameter growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and leaf hydraulic vulnerability curves.
+- We inversely estimated the effective rooting depth (ERD, as the average depth of water extraction), for 32 canopy species by linking diameter growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and leaf hydraulic resistance curves.
 We validated ERD estimates against existing empirical data of potential water access depths.
@@
 A key bottleneck in community-wide testing of this mechanism has been a lack of data in both trees' rooting or water-sourcing depths and plant-available soil water at those depths.
-A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the world’s more than 60 thousand tree species [@Fan:2017; @Beech:2017].
-Further, only a small fraction of those are tropical, even though more than 90% of the world’s tree diversity resides in the tropics [@Slik:2015].
+A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of more than 60 thousand tree species in the World [@Fan:2017; @Beech:2017].
+Further, only a small fraction of those are tropical, even though more than 90% of the World’s tree diversity resides in the tropics [@Slik:2015].
 The use of stable isotopes of water as a tracer provides an indirect measure of water sourcing depths by matching the isotopic value in xylem water to those in soil pore water at different depths.
@@
 Such efforts have found correlations between vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013, @Venturas.2016].
-Even though deep roots may mitigate hydraulic vulnerability and mortality-risk, hydraulic-risk realized through hydrological (rather than meteorological) droughts--the outcome of the interaction between rooting depths, critical hydraulic thresholds and the actual vertical distribution of plant-available water--are not studied, especially in tropical regions.
+Even though deep roots may mitigate hydraulic vulnerability [@Brum.2019] and mortality-risk, hydraulic-risk realized through hydrological (rather than meteorological) droughts--the outcome of the interaction between rooting depths, critical hydraulic thresholds and the actual vertical distribution of plant-available water--are not studied, especially in tropical regions.
 
@@
 We evaluate how rooting depth is linked to above-ground hydraulic traits and mortality rates over a 35 year period that experienced droughts of a variety of intensity, frequency and duration.
-Our hypotheses are that (i) deep-rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem due to greater and more reliable water availability at depth, whereas shallow-rooted species are associated with less efficient but safer hydraulic traits due to more variable availability of shallow soil-water (see Tables \@ref(tab:symbols) and \@ref(tab:hypo)); and that, (ii) deep-rooted species have lower mortality rates during droughts due to a lower realized risk of hydraulic failure compared to shallow-rooted species.
-To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich tropical forest.
+Our hypotheses are that (i) deep-rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem due to greater and more reliable water availability at depth, whereas shallow-rooted species are associated with less efficient but safer hydraulic traits due to more variable availability of shallow soil-water (see Tables \@ref(tab:symbols) and \@ref(tab:hypo)); and that, (ii) deep-rooted species have lower mortality rates during droughts due to a lower exposure to water-stress compared to shallow-rooted species.
+To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic traits and long-term mortality outcomes for a species-rich tropical forest.
 
@@
 
-To obtain the daily dynamics of soil water potential in the whole-soil column, we locally parameterized and calibrated a 1-D hydrological water balance model (Fig.
-\@ref(fig:schematic)).
-We then developed a novel, empirical inverse model to estimate species effective rooting depths (ERD) for large trees of `r length(erd.sp)` species.
-We define ERD as the depth at which soil water potential best explained species-specific growth dynamics.
-The ERD model incorporated the impact of atmospheric and hydrological drought on growth, and was constrained with species-specific leaf vulnerability curves by using existing data for eight species, while developing trait-based proxies for the rest (based on data for a total of 21 species).
+This work combines hydrology, physiology and demography of a tropical forest at Barro Colorado Island, Panama, to inform the inverse model for rooting depths; validates the model, and tests hypotheses pertaining to relationship of rooting depths with above-ground hydraulic traits, drought exposure, and mortality (see Fig.
+\@ref(fig:schematic) for a workflow).
+We defined species-specific effective rooting depths (ERD) as the depth at which soil water potential best explained species' growth dynamics.
+To estimate ERDs we used two models in succession: first, we obtained the daily dynamics of soil water potential in the whole-soil column.
+For this, we locally parameterized and calibrated 1-D hydrological water balance of the forest via an Earth System Model.
+Second, we developed a novel, empirical inverse model of ERD (run offline) and estimated ERDs for large trees of `r length(erd.sp)` species.
+The ERD model incorporated the impact of atmospheric and hydrological drought on growth, and was constrained with species-specific leaf resistance curves.
+For the latter, we used existing data for eight species and developed trait-based proxies for the rest (based on data for a total of 21 species).
 We validated ERD against existing stable hydrogen isotope ratios ($\mathrm{\delta}$^2^H~xylem~) for a subset of six tree species as independent observations.
@@
 We tested the relationship of ERD with xylem vulnerability to embolism, branch hydraulic safety margins, leaf turgor loss point and maximum stem-specific hydraulic conductivity.
-Finally, to test whether ERD explained risk of drought-induced mortality, we analysed species-specific large tree mortality dynamics over 35 years and estimated species-level realized hydraulic-risk as the time a species’ spent beyond its critical hydraulic threshold.
+We analysed whether ERD correlated with mortality dynamics of large trees over 35 years that experienced several El-Nino droughts.
+Finally, to test whether ERD explained risk of drought-induced mortality, we analysed species-level exposure to water-stress as the time a species’ spent beyond its critical hydraulic threshold, given its ERD.
 
@@
 \begin{equation}
-\mathrm{\theta}_{\textrm{soil}} = \mathrm{\Psi}_e\left(\frac{\mathrm{\theta}}{\mathrm{\theta}_{s}}\right)^{-b}
+\mathrm{\Psi}_{\textrm{soil}} = \mathrm{\Psi}_e\left(\frac{\mathrm{\theta}}{\mathrm{\theta}_{s}}\right)^{-b}
 (\#eq:soil-ret-curve)
@@
 
-To inversely estimate effective rooting depth, we tested alternate empirical models for species-specific 5-yearly census growth, which use effective rooting depth as an input (Eq.
+We defined species-specific effective rooting depth (ERD) as the depth at which soil water potential best explained species' growth dynamics.
+We inversely estimated ERD, by testing alternate empirical models of species-specific 5-yearly census growth, which use ERD as an input (Eq.
 \@ref(eq:M-chosen) and those in SI).
-Many factors limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019].
-Because pre-dawn $\mathrm{\Psi}$~leaf~ generally approaches $\mathrm{\Psi}$~soil~, we used $\mathrm{\Psi}$~soil~ in leaf hydraulic vulnerability curves to predict maximum diurnal $K$~leaf~, which we assumed controls photosynthetic rate and growth (@Brodribb.2002; @Brodribb.2017; see Table \@ref(tab:symbols) for trait symbols and their definitions).
-We assumed $K_\textrm{leaf}$ to be a hydraulic bottleneck in the plant, and thus to determine or co-determine whole plant hydraulic response to soil drying; notably, root hydraulic vulnerability may also be a strong driver, and recent studies suggest that its vulnerability would be similar to, or correlated with that of leaves (see @Cochard.2020 and references therein).
-$K_\textrm{leaf}$ in our models, being based on vulnerability curves, is dynamic and thus depends on rooting depth and rhizosphere water potential.
-The use of maximum diurnal $\mathrm{\Psi}$~leaf~ based on pre-dawn $\mathrm{\Psi}$~soil~ as a proxy for diurnal $\mathrm{\Psi}$~leaf~ and its influence on growth at diurnal time-scales (to relate to 5-year average growth values) thus does not consider diurnal dynamics of $\mathrm{\Psi}$~leaf~ and other factors that would play a role in such diurnal dynamics, such as the buffering effect of stem water storage capacitance on $\mathrm{\Psi}$~leaf~, and may thus overestimate ERD (see Discussion).
+
+Many intrinsic and extrinsic factors limit plant growth, including soil moisture, VPD, radiation, leaf area seasonality, rooting depth and the rate at which plant hydraulic conductance declines with water-stress [@Lawrence.2019; @Brodribb.2002].
+The latter could be studied at the organ level, for example, $K$~leaf~ vs.
+$\mathrm{\Psi}$~leaf~ at leaf-level, referred to as the leaf hydraulic resistance curve (@Sack.2012; see Table \@ref(tab:symbols) for trait symbols and their definitions).
+We assumed $K_\textrm{leaf}$ to be a hydraulic bottleneck in the plant and that it determines or co-determines whole plant hydraulic response to soil drying; notably, root hydraulic resistance may also be a strong driver, but recent studies suggest that its resistance would be similar to, or correlated with that of leaves (see @Cochard.2020 and references therein).
+Because pre-dawn $\mathrm{\Psi}$~leaf~ generally approaches $\mathrm{\Psi}$~soil~, this allowed us to use $\mathrm{\Psi}$~soil~ in leaf hydraulic resistance curves to predict maximum diurnal $K$~leaf~, which we assumed controls photosynthetic rate and growth (to relate to 5-year average growth values), ignoring diurnal dynamics of $K$~leaf~ [@Brodribb.2002; @Brodribb.2017].
+$K_\textrm{leaf}$ in our models, being based on resistance curves, is thus dynamic and depends on rooting depth and rhizosphere water potential.
+
 VPD affects growth non-linearly, with growth increasing with VPD up-to a VPD threshold, then decreasing as leaf pores (stomata) close reducing water-uptake [@Grossiord.2020; @Yang.2019].
@@
 Apart from stomatal control, leaf-drop may further limit water-uptake and growth.
-We therefore regressed species-specific growth against multiplicative or additive effects of one or more factors calculated daily and averaged over 5-yearly censuses: $\hat{\textrm{VPD}}$, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1.
+
+We therefore regressed species-specific growth against multiplicative or additive effects of one or more growth factors calculated daily and averaged over 5-yearly censuses: $\hat{\textrm{VPD}}$, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1.
 Incorporation of radiation did not improve model fitting, possibly due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season at BCI.
@@
 \
- For the best model (see model validation and selection below), daily average growth for species $s$, $G_{s,t|z}$, in the census interval $t$ for depth $z$ and for a given hydrological realization $h$ is described as follows:
+ For the best model (see model validation and selection below), daily average growth for species $s$, $G_{s,t|z,h}$, in the census interval $t$ for depth $z$ and for a given hydrological realization $h$ is described as follows:
 
 \begin{equation}
-G_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},s,i|z,h}^{*} \hat{\textrm{VPD}}_{i}^{*}) + \epsilon_{s,t|z,h}
+G_{s,t|z,h} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},s,i|z,h}^{*} \hat{\textrm{VPD}}_{i}^{*}) + \epsilon_{s,t|z,h}
 (\#eq:M-chosen)
 \end{equation}
-where n~t~ is the total number of days in census interval $t$, the superscript * indicates that the variable has been standardized between 0 and 1 (within species for $K_{\textrm{leaf}}$), $\beta_{0}$ and $\beta_{1}$ are model coefficients and $\epsilon$ is the model error term.
+where | refers to conditionals, n~t~ is the total number of days in census interval $t$, the superscript * indicates that the variable has been standardized between 0 and 1 (within species for $K_{\textrm{leaf}}$), $\beta_{0}$ and $\beta_{1}$ are model coefficients and $\epsilon$ is the model error term.
 See SI for all the alternate models tested.
@@
 \
- For each species, we estimated diurnal maximum value of $K_{\textrm{leaf},s,i|z,h}$ for a species $s$ by substituting $\mathrm{\Psi}_{\textrm{soil},s,i|z,h}$ in the leaf hydraulic vulnerability curves defined below in Eq.
+ For each species, we estimated diurnal maximum value of $K_{\textrm{leaf},s,i|z,h}$ for a species $s$ by substituting $\mathrm{\Psi}_{\textrm{soil},s,i|z,h}$ in the leaf hydraulic resistance curves defined below in Eq.
 \@ref(eq:vc-exp) using species-specific parameters $A_{s}$ and $B_{s}$ as:
@@
 Using Eq.
-\@ref(eq:M-chosen), for a species $s$ and hydrological realization *h*, we obtained Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $r$ for each soil depth $z$.
-For each *h*, we retained depths with the maximum likelihood (similar AIC based support, see below), and amongst them selected the best-fit depth as the one with the maximum Pearson's $r$, if $r$ >= 0.71.
-For a species $s$, we defined effective rooting depth (ERD) as the median of the best-fit depth over all *h*, which can be described as :
+\@ref(eq:M-chosen), for a species $s$ and hydrological realization *h*, we obtained Akaike's Information Criterion (AIC) and Goodness-of-fit ($R$^2) for each soil depth $z$.
+We used AIC as the first selection criterion to ensure that the fitted values are within the confidence interval of the error distribution (as the same model Eq.
+\@ref(eq:M-chosen) was being compared for each *h*, number of model parameters were the same).
+For each *h*, we retained those depths with the maximum likelihood (similar AIC based support, see below) filtering out the rest.
+However, AIC does not ensure that fitted values will have a positive trend with growth.
+We therefore used $R$^2 as a goodness-of-fit measure and retained depths associated with a relatively high threshold $R$^2 value of 0.5; after filtering out depths associated with negative correlations if any.
+Finally, among these we chose the depth with the highest $R$^2 per *h*.
+Results were not sensitive to inclusion of this final step of retaining only a single depth as the best-fit (not shown).
+For a species $s$, we defined effective rooting depth (ERD) as the median of the best-fit depth over all *h*, which can be described as:
 
@@
 
-where $\mathrm{\Delta}_{s|z,h} = \textrm{AIC}_{s|z,h} - \underset{\forall \, z \,\in\,Z}{\textrm{min}}(\textrm{AIC}_{s|z,h})$,  $\textrm{AIC}_{s|z,h} = -2\textrm{log}(\mathcal{L}(\hat{\beta}_{s|z,h}|{G_{s}})) - 2k$, $\mathcal{L}$ is the likelihood function, $G$ the observed growth time series with $k$ observations (or, census intervals $t$).
+where $\mathrm{\Delta}_{s|z,h} = \textrm{AIC}_{s|z,h} - \underset{\forall \, z \,\in\,Z}{\textrm{min}}(\textrm{AIC}_{s|z,h})$,  $\textrm{AIC}_{s|z,h} = -2\textrm{log}(\mathcal{L}(\hat{\beta}_{s|z,h}|{G_{s}})) - 2k$, $\mathcal{L}$ is the likelihood function, $\hat{\beta}$ is the maximum-likelihood estimate, $G$ the observed growth time series with $k$ observations (or, census intervals $t$).
 $H$ = (1, .., 100), the 100 best-fit hydrological realizations and $Z$ = (0.4, `r paste(soil.depths[8:length(soil.depths)], collapse=", ")`) m.
@@
 
-#### Leaf hydraulic vulnerability curves
+#### Leaf hydraulic resistance curves
 
-We obtained leaf hydraulic vulnerability curves ($K_\textrm{leaf}$ vs.
+We obtained leaf hydraulic resistance curves ($K_\textrm{leaf}$ vs.
 $\mathrm{\Psi}$~leaf~) for adult trees of 21 common species from Zailaa *et* *al.* (to be submitted; see SI for brief description of methods) described as:
@@
 These 21 species included eight of the `r length(erd.sp)` species selected for ERD estimation.
-For the rest `r sp.n.model` species, we obtained vulnerability curves using trait-based proxies.
+For the rest `r sp.n.model` species, we obtained resistance curves using trait-based proxies.
 We identified scaling relationships between fitted parameters $A$ and $B$ in Eq.
@@
 As these fits explained a large proportion of variation in parameters $A$ and $B$ (see results), we sequentially used Eq.
-\@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r sp.n.model` ERD species without direct data and estimated leaf hydraulic vulnerability curves using Eq.
+\@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r sp.n.model` ERD species without direct data and estimated leaf hydraulic resistance curves using Eq.
 \@ref(eq:vc-exp).
@@
 \@ref(eq:b-mod).
-We also estimated species $\mathrm{\Psi}$~50,leaf~ from their vulnerability curves.
+We also estimated species $\mathrm{\Psi}$~50,leaf~ using their resistance curves.
 
@@
 We regressed species ERD against species $\textrm{M}_{s,t}$ for each census interval $t$ for the whole pool of ($n$ = `r nrow(mrate.depth.mean)`) species with $\textrm{M}_{s,t}$ estimates and for evergreen species alone ($n$ = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`).
-To assess whether deeper ERD species escaped the experience of significant hydrological drought, we calculated realized hydrulic risk for each species as the proportion of days in each census interval $t$ of the $\mathrm{\Psi}_\textrm{soil}$ simulation period (1990-2015) during which soil water potential in the soil layer matching species ERD, $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$, were more negative than species $\mathrm{\Psi}$~crit~.
-$\mathrm{\Psi}$~crit~ was defined as $\mathrm{\Psi}$~50,leaf~.
+We assessed whether exposure to hydrological drought decreased with ERD.
+For this, we assumed crossing of  species-specific critical hydraulic threshold, $\mathrm{\Psi}$~crit~, here defined as $\mathrm{\Psi}$~50,leaf~, as an indicator of exposure to water-stress.
+Extended water-stress could lead to mortality via hydraulic failure or carbon starvation.
+The duration of exposure to water-stress, and thus realized hydraulic-risk, was defined for each species as the proportion of days in each census interval $t$ of the $\mathrm{\Psi}_\textrm{soil}$ simulation period (1990-2015) during which soil water potential in the soil layer matching species ERD, $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$, were more negative than species $\mathrm{\Psi}$~crit~.
 
@@
 
-## Leaf hydraulic vulnerability curves
+## Predictors of leaf hydraulic resistance curves
 
-Parameters of leaf vulnerability curves were predictable from WSG and LMA (Fig.
+Parameters of leaf resistance curves were predictable from WSG and LMA (Fig.
 \@ref(fig:ab-fit)).
@@
 S7).
-This predictive power allowed us to estimate leaf vulnerability curves used in the ERD models for `r sp.n.model` out of `r length(erd.sp)` ERD species that lacked direct observations (Table S3, Fig.
+This predictive power allowed us to estimate leaf resistance curves used in the ERD models for `r sp.n.model` ERD species that lacked direct observations (among `r length(erd.sp)` ERD species) (Table S3, Fig.
 S8).
@@
 
-Among evergreen species, ERD explained `r m.evg.r2.90`, `r m.evg.r2.05` and `r m.evg.r2.10`% variation in mortality rates during census intervals 1985-1990, 2000-2005 and 2005-2010, respectively, such that species mortality rates significantly decreased with deeper ERD ($p$ = `r m.evg.p.90`, `r m.evg.p.05` and `r m.evg.p.10`, respectively; Fig.
+Among evergreen species, ERD explained `r m.evg.r2.90`, `r m.evg.r2.05` and `r m.evg.r2.10`% variation in mortality rates during census intervals 1985-1990, 1985-1990, 2000-2005 and 2005-2010, respectively, such that species mortality rates significantly decreased with deeper ERD ($p$ = `r m.evg.p.90`, `r m.evg.p.05` and `r m.evg.p.10`, respectively; Fig.
 \@ref(fig:erd-mort-census-evg)).
-When deciduous species were included, ERD explained lower but still significant proportion of variation in mortality in those intervals ($R$^2^ = `r m.r2[1]`, `r m.r2[2]` and `r m.r2[3]`% and $p$ = `r m.p.90`, `r m.p.05` and `r m.p.10`, respectively; Fig.
+When deciduous species were included, ERD explained lower but still significant proportion of variation in mortality in those intervals ($R$^2^ = `r m.r2[1]`, `r m.r2[2]` and `r m.r2[3]`% and $p$ = `r m.p.90`, `r m.p.05` and , respectively; Fig.
 S11).
@@
 While our $\mathrm{\Psi}_{\textrm{soil}}$ estimates did not extend prior to 1990, due to a lack of reliable climate forcing, a prior study found that during the period 1985-1990, for which ERD explained significant mortality, BCI experienced three extreme deficit water years, following the 1983 El-Niño year [@Condit.2017].
-Analysis of a species' realized hydraulic-risk as the time spent under significant hydrological drought (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) showed that species with shallower ERD experienced greater realized hydraulic-risk and increasingly so over 2000-2005 and 2005-2010 (Fig.
+Species exposure to water-stress (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) non-linearly declined with ERD, indicating that species with shallower ERD spent greater time under significant hydrological drought, and experienced greater hydraulic-risk, which increased over 2000-2005 and 2005-2010 (Fig.
 \@ref(fig:erd-time)).
 For the census interval 1990-1995, species with the shallowest ERD also showed greater realized hydraulic-risk (Fig.
-\@ref(fig:erd-time)), but during the dry season ( Fig.
+\@ref(fig:erd-time)), but during the dry season (Fig.
 \@ref(fig:swp-droughts)), and without elevated mortality rates (Fig.
@@
 
-Our novel approach for estimating ERD using 25 years of tree growth, species-specific leaf vulnerability curves, modeled soil moisture profiles, and observed VPD was consistent with estimates using $\mathrm{\delta}$^2^H~xylem~ (Figs.
+Our novel approach for estimating ERD using 25 years of tree growth, species-specific leaf resistance curves, modeled soil moisture profiles, and observed VPD was consistent with estimates using $\mathrm{\delta}$^2^H~xylem~ (Figs.
 \@ref(fig:erd-iso), \@ref(fig:sp-erd)).
@@
 \@ref(fig:erd-mort-census-evg)).
-Analysis of time spent beyond $\mathrm{\Psi}$~crit~ in the soil zone matching ERD $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ showed that this was likely due to decreasing realized risk of hydraulic failure with increasing access to water for species with deeper ERD (Fig.
+Analysis of time spent beyond $\mathrm{\Psi}$~crit~ in the soil zone matching ERD $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ showed that this was likely due to decreasing realized risk of water-stress with increasing access to water for species with deeper ERD (Fig.
 \@ref(fig:erd-time)).
-We thus demonstrate, for the first time, a link between realized drought risk, hydraulic architecture and long-term mortality outcomes for large tree species of a tropical forest.
+We thus demonstrate, for the first time, a link between drought exposure based on water-availability in the whole soil-column, above- and below-ground hydraulic traits, and long-term mortality outcomes for trees in a tropical forest canopy.
 
-## Realized drought risk integral to assessing drought-vulnerability
+## Drought exposure integral to assessing drought-induced mortality
 
@@
 Globally, large trees tend to exhibit greater growth reductions, lower post-drought resilience, and greater increases in mortality relative to their understory counterparts [@Bennett:2015; @Phillips:2010].
-Our finding that deep-water access buffers drought-induced mortality in large trees is relevant for understanding drought resistance, resilience and recovery of this important functional group [@Bennett:2015; @McGregor.2020]
+Our finding that deep-water access buffers drought-induced mortality in large trees is relevant for understanding drought resistance, resilience and recovery of this important functional group [@Bennett:2015; @McGregor.2020].
+Future studies should test the ERD-mortality relationship on a greater number of species in the community.
 
@@
 Our best ERD model is a key improvement over the model of CT2018 ($n$ = 12 species) as we employ a physiologically meaningful representation, fewer parameters and corroboration with tree and soil data for $\mathrm{\delta}$^2^H.
-We note that stable water isotopes for plant and below-ground sources are critical for calibration or validation of ERD models for scaling upto community, although the interpretation of isotopic data is still under research [@Deurwaerder.2020; @Adams.2020; @Bowers.2020].
+
 As in CT2018, we also interpret our ERD estimates as revealing relative differences among species' effective rooting depths rather than absolute depths.
-As our ERD model ignores stem water storage capacitance [@Wolfe.2017] it may have overestimated ERD especially in those species for which capacitance is important as a drought-avoidance strategy, for example, the deciduous species [@Borchert.2005jma].
+Besides, our ERD model predicts only daily maximum diurnal $\mathrm{\Psi}$~leaf~ assuming it is equivalent to $\mathrm{\Psi}$~soil~.
+While this scale may be adequate to relate to 5-year average growth values, this approach ignores diurnal dynamics of $\mathrm{\Psi}$~leaf~ and other factors that would play a role in such diurnal dynamics, such as the buffering effect of stem water storage capacitance on $\mathrm{\Psi}$~leaf~.
+By ignoring stem water storage capacitance [@Wolfe.2017] our model may thus have overestimated ERD especially in those species for which capacitance is important as a drought-avoidance strategy, for example, the deciduous species [@Borchert.2005jma].
+Future studies should investigate the role of capacitance on estimating ERD.
+
+While ERD models have the potential to estimate ERDs for entire tree communities, during the their development phase, ERD models may need to be validated against data for a subset of representative species of the community, as this study does.
+ERD models should be tested across varied climates and forest types, covering contrasting plant strategies and possibly seasonality in ERDs.
+Direct observations of rooting depths and stable water isotopes based water-sourcing depths will be important datasets for such validation; although the interpretation of isotopic data is still under research [@Deurwaerder.2020; @Adams.2020; @Bowers.2020].
 
@@
 The relationships that we identify between above- and below-ground traits, the water environment vertically below-ground and mortality rates are important for representing diversity in DGVMs, which intrinsically rely on the parameterization of contrasting life history strategies [@Scheiter:2013] and the simulation of competition between those strategies.
-We found that LMA and WSG were strong predictors of leaf hydraulic vulnerability curves.
-This offers the promise of a greater ability to parameterize the 'hard' hydraulic traits with the abundant 'soft' trait data, thus allowing for a better representation of forest hydrodynamics.
+ERDs could be assimilated into DGVMs by setting rooting parameters that result into the centroid of root water-uptake profiles equivalent to ERDs.
+Current DGVMS could thus provide an informative evaluation of ERD model as well.
+We found that LMA and WSG were strong predictors of leaf hydraulic resistance curves.
+Albeit future studies should undertake sensitivity analyses for the errors involved, our finding offers the promise of a greater ability to parameterize the 'hard' hydraulic traits with the abundant 'soft' trait data, thus allowing for a better representation of forest hydrodynamics.
 The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits in order to better represent the functional diversity of tropical forests, and their trajectories into the future.
@@
 S9).
-We interpret this result as that VPD and leaf hydraulic vulnerability curves may be adequate to explain inter-census differences in growth (via stomatal control), but also acknowledge that our estimates of seasonality of LAI is a tentative estimate that combines leaf-fall data and the timing of leaf-gain backtracked from leaf lifetime, and omits inter-annual variation.
+We interpret this result as that VPD and leaf hydraulic resistance curves may be adequate to explain inter-census differences in growth (via stomatal control), but also acknowledge that our estimates of seasonality of LAI is a tentative estimate that combines leaf-fall data and the timing of leaf-gain backtracked from a leaf lifetime, and omits inter-annual variation.
 Future studies may improve models of LAI seasonality and make use of data from new technologies such as drone based monitoring of LAI in species-rich tropical forests [@Park.2019].
@@
 Although data needs for parameterizing such models could be greater (e.g.
-vulnerability curves and capacitance for roots, stems and leaves), adding a degree of parameter uncertainty in community-wide application, such models hold greater promise in improving our understanding of plant physiology [e.g.
+hydraulic vulnerability curves and capacitance for roots, stems and leaves), adding a degree of parameter uncertainty in community-wide application, such models hold greater promise in improving our understanding of plant physiology [e.g.
 see @Johnson:2018fad].
@@
 This has important implications for our predictive understanding of tropical forest dynamics under current and future climate.
-Our community-scale framework for modeling effective rooting depths and leaf vulnerability curves indicates the possibilities in expanding the use of these critical, rare observations in species-rich forests towards community-scale generalizations.
+Our community-scale framework for modeling effective rooting depths and leaf resistance curves indicates the possibilities in expanding the use of these critical, rare observations in species-rich forests towards community-scale generalizations.
 
@@
 \
-Figure \@ref(fig:schematic): A schematic for the methods workflow.
+Figure \@ref(fig:schematic): A schematic outlining the methods workflow, which combines hydrology, physiology and demography of the tropical forest at Barro Colorado Island to inform the inverse model for rooting depths; validates it, and tests hypotheses pertaining to relationship of rooting depths with above-ground hydraulic traits, drought exposure, and drought-induced mortality.
 
@@
 \
-Figure \@ref(fig:ab-fit): Parameters $A$ (panel **a**) & $B$ (panel **b**) for the leaf vulnerability curves that were fitted to observed data on $K_\textrm{leaf}$ vs.
+Figure \@ref(fig:ab-fit): Parameters $A$ (panel **a**) & $B$ (panel **b**) for the leaf resistance curves that were fitted to observed data on $K_\textrm{leaf}$ vs.
 $\mathrm{\Psi}$~leaf~ for `r length(sp.exp.param$sp)` species (Eq.
@@
 $R$^2^ and significance levels for linear model fits are given in panel insets.
-Model fits (blue lines) with confidence bands (gray area) are only shown for periods with significant fits at $\alpha$ = 0.1, identified by census-intervals with a superscript *.
+Model fits (blue lines) with confidence bands (gray area) are only shown for periods with significant fits at $\alpha$ = 0.05, identified by census-intervals with a superscript *.
 
 \
-Figure \@ref(fig:erd-time): Modeled effective rooting depth (ERD; horizontal-axis) versus time spent beyond critical hydraulic threshold (vertical axis) by census interval (colored bars).
+Figure \@ref(fig:erd-time): Modeled effective rooting depth (ERD; horizontal-axis) versus time spent beyond critical hydraulic threshold (vertical axis) by census interval (colored bars; note the squared y-axis scale).
 Each bar represents the average time species of the same ERD spent beyond species-specific critical hydraulic thresholds in a given interval, that is, the proportion of days for which $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ was more negative than $\mathrm{\Psi}$~crit~, defined as $\mathrm{\Psi}$~50,leaf~, and where $z$ is the soil depth matching species ERD.
 Standard error of the mean are shown over each bar.
-Data includes all the `r length(erd.mrate.sp)` canopy tree species on Barro Colorado Island that were included in the mortality analyses.
+Data includes all the `r length(erd.mrate.sp)` canopy tree species that were included in the mortality analyses.
 Trends were similar for evergreen species alone.
@@
 ```{r schematic, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-schematic}")}
-knitr::include_graphics(file.path("schematic4/schematic4.001.jpeg"))
+knitr::include_graphics(file.path("schematic_easy/schematic_easy.001.jpeg"))
 ```

### 2021-03-16 09.0_manuscript.rmd
created groundhog folder and changed grounhog day

@@
 ```{r setup1, include=FALSE, echo = FALSE}
-if (!require("groundhog", quietly=TRUE)) install.packages("groundhog"); library(groundhog)
+if (!require("groundhog")) install.packages("groundhog");
+groundhog.folder <- paste0("Users/ftuser/Library/R/4.0/library/bci.roots.inversion")
+if(!dir.exists(file.path(groundhog.folder))) {dir.create(file.path(groundhog.folder))}
+set.groundhog.folder(groundhog.folder)
+library(groundhog)
+
 groundhog.day = "2021-01-01"

### 2021-03-16 10.0_Supporting_Information.Rmd
created groundhog folder and changed grounhog day

--- 2021-02-20 10.0_Supporting_Information.Rmd
@@
 
-```{r setup, include=FALSE, echo = FALSE}
-if(!require("pacman", quietly=TRUE)) install.packages("pacman")
-suppressPackageStartupMessages(
-  pacman::p_load(bookdown, knitr, pander, citr, broom))
+```{r setup1, include=FALSE, echo = FALSE}
+if (!require("groundhog")) install.packages("groundhog");
+groundhog.folder <- paste0("Users/ftuser/Library/R/4.0/library/bci.roots.inversion")
+if(!dir.exists(file.path(groundhog.folder))) {dir.create(file.path(groundhog.folder))}
+set.groundhog.folder(groundhog.folder)
+library(groundhog)
+
+groundhog.day = "2021-01-01"
+pkgs=c('knitr', 'pander', 'broom')
+groundhog.library(pkgs, groundhog.day)
 options(tinytex.verbose = TRUE)
@@
 ```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($\\Psi_{\\textrm{soil},z}$) by depth $z$ (m; panels) over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines).
-Vertical gray lines demarcate median dates of the demographic censuses at the BCI 50 ha plot."}
-knitr::include_graphics(file.path("figures/2019-10-14_5000/best-fits/psi_model_daily_all_depths_params.top.few_full.jpeg"))
+Vertical gray lines demarcate median dates of the demographic censuses at the BCI 50 ha plot.
+Years 1991 & 1992 for depths > 2.9 m showed marks of model initialisation phase for ELM-FATES, which was run from ($\\Psi_{\\textrm{soil},z}$) = 0 for all depths, starting from 1985-01-01.
+We therefore removed $\\Psi_{\\textrm{soil}}$ for years 1991 & 1992 for depths greater than 2.9 m, and conservatively removed $\\Psi_{\\textrm{soil}}$ for year 1991 for the depth of 2.9 m."}
+knitr::include_graphics(file.path("figures/2019-10-14_5000/best-fits/psi_model_daily_all_depths_params.top.few_full_beg_1991.jpeg"))
 ```
@@
 
-```{r erd-hyd-full, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd-full}.
-Pairwise correlations (upper triangle) and corresponding linear regressions (lower triangle) among modelled Effective Rooting Depth (ERD) and hydraulic properties for seven canopy species found on Barro Colorado Island (Panama).
-Spearman's correlation coefficients and significance levels are shown (upper-half; '*' $p$ < 0.05; '.' $p$ < 0.1; ' ' $p$ < 1)")}
+```{r erd-hyd-full, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd-full}Pairwise Spearman's correlations (upper triangle) and corresponding linear regressions (lower triangle) among modelled Effective Rooting Depth (ERD) and hydraulic properties for seven canopy species found on Barro Colorado Island (Panama).Spearman's correlation coefficients and significance levels are shown (upper-half; '*' $p$ < 0.05; '.' $p$ < 0.1; ' ' $p$ < 1).")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/erd.stem.traits_cor.chart.jpeg"))
@@
 $R$^2^ and significance levels for linear model fits are given in panel insets.
-Linear model fits (blue lines) with confidance bands (gray area) are shown for significant fits at $\\alpha$ = 0.1."}
+Linear model fits (blue lines) with confidance bands (gray area) are shown for significant fits at $\\alpha$ = 0.05, identified by census-intervals with a superscript *."}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_by rdi.gr.jpeg"))
@@
 ```{r leaf-cover, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-cover}Seasonal variation in standardised LAI (range 0-1) for ", length(erd.sp)," species at Barro Colorado Island (panels).
-Species are color-coded by deciduousness.")}
+Species are color-coded by deciduous category as defined by expert botanists.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf.cover_BCI_multi_panel.jpeg"))

### 2021-03-16 code/06.0_load.R

@@ -13,28 +13,18 @@ rm(list=ls())
 # #*******************************************
 # #### Written  with R version 4.0.3 ###
 # #*******************************************
-if (!require("groundhog")) install.packages("groundhog"); library(groundhog)
+if (!require("groundhog")) install.packages("groundhog");
+groundhog.folder <- paste0("Users/ftuser/Library/R/4.0/library/bci.roots.inversion")
+if(!dir.exists(file.path(groundhog.folder))) {dir.create(file.path(groundhog.folder))}
+set.groundhog.folder(groundhog.folder)
+library(groundhog)
 # set.groundhog.folder('/Users/ftuser/Library/R/3.6/library')
-groundhog.day = "2020-05-01"
+groundhog.day = "2021-01-01"
 groundhog.library('devtools', groundhog.day)
-# pkgs=c('lattice', 'magick', 'cowplot', 'corrplot', 'ggcorrplot', 'tidyverse', 'readxl',
-#        'forcats', 'scales', 'data.table', 'ggpmisc', 'GGally', 'bookdown', 'rmarkdown', 'broom')
+pkgs=c('lattice', 'magick', 'cowplot', 'corrplot', 'ggcorrplot', 'tidyverse', 'readxl',
+       'forcats', 'scales', 'data.table', 'ggpmisc', 'GGally', 'bookdown', 'rmarkdown', 'broom')
+groundhog.library(pkgs, groundhog.day)
 
-groundhog.library('lattice', groundhog.day)
-groundhog.library('magick', groundhog.day)
-groundhog.library('cowplot', groundhog.day)
-groundhog.library('corrplot', groundhog.day)
-groundhog.library('ggcorrplot', groundhog.day)
-groundhog.library('tidyverse', groundhog.day)
-groundhog.library('readxl', groundhog.day)
-groundhog.library('forcats', groundhog.day)
-groundhog.library('scales', groundhog.day)
-groundhog.library('data.table', groundhog.day)
-groundhog.library('ggpmisc', groundhog.day)
-groundhog.library('GGally', groundhog.day)
-groundhog.library('bookdown', groundhog.day)
-groundhog.library('rmarkdown', groundhog.day)
-groundhog.library('broom', groundhog.day)
 
 # graphics info
 theme_set(theme_bw())

### 2021-03-16 code/07.0_PhenoDemoTraitsPsi.R

@@ -27,7 +27,7 @@ groundhog.folder <- paste0("Users/ftuser/Library/R/4.0/library/bci.roots.inversi
 if(!dir.exists(file.path(groundhog.folder))) {dir.create(file.path(groundhog.folder))}
 set.groundhog.folder(groundhog.folder)
 library(groundhog)
-groundhog.day = "2020-05-01"
+groundhog.day = "2021-01-01"
 pkgs=c('tidyverse', 'readxl', 'forcats', 'agricolae', 'gridExtra',
        'scales', 'GGally', 'ggpmisc', 'Evapotranspiration',
        'data.table', 'mgcv')
@@ -491,7 +491,7 @@ data.model.AB.sub <- data.model.AB %>%
   mutate(A = if_else(is.na(data.A), model.A, data.A),
          B = if_else(is.na(data.B), model.B, data.B)) %>%
   subset(sp %in% growth.sub$sp) %>%
-  # ## And LMA from LMALAM from LEAF and DISC
+  ## And LMA from LMALAM from LEAF and DISC
   subset(sp.LMA.sub == "original") %>%
   # removing A & B predicted from gap-filled WSG
   subset(sp.WSG.sub == "original")
@@ -518,7 +518,7 @@ save(sp.ab.growth, file = file.path(results.folder, "sp.ab.growth_for_LAI.Rdata"
 sp.growth.not.LAI <- "guapst"
 sp.LAI.for.model.mod <- sp.LAI.for.model %>%
   mutate(sp = as.character(sp)) %>%
-  select(sp, doy, L.norm) %>%
+  dplyr::select(sp, doy, L.norm) %>%
   bind_rows(data.frame(sp = rep(sp.growth.not.LAI, each = max(sp.LAI.for.model$doy)),
                        doy = rep(unique(sp.LAI.for.model$doy), times = length(sp.growth.not.LAI)),
                        L.norm = NA)) %>%
@@ -657,9 +657,8 @@ ml.rsq.combine.best <- ml.rsq.combine.best %>%
   left_join(iso.1.3.join %>%
               subset(!sp %in% as.character(leafless_mar.apr$sp[leafless_mar.apr$leafless_in_mar_apr_from_notes == "Yes"])) %>%
               group_by(sp) %>%
-              summarise(Xylem_sap_deltaD_permil.mean = mean(Xylem_sap_deltaD_permil, na.rm = TRUE),
[…]

### 2021-03-16 Readme.Rmd

--- 2021-02-20 Readme.Rmd
@@
 
-Set-up and run effective rooting depth models, evaluate model, and join ERD from the predicted model with hydraulic traits and mortality data, and calcualted realized hydraulic risk
+Set-up and run effective rooting depth models, evaluate model, and join ERD from the predicted model with hydraulic traits and mortality data, and calculated realized hydraulic risk
 
@@
 
-**Figures**
-
-```{r Chunk 08.0, eval = FALSE}
-source("code/06.0_load.R")
-source("code/08.0_manuscript_data_figs.R")
-```
-
 **Main text**
-
 ```{r Chunk 09.0, eval = FALSE}
-render("09.0_manuscript.rmd", output_format = "bookdown::html_document2")
+# Two files are used within the report
+# For data, 06.0_load.R and 
+# For figures and data, code/08.0_manuscript_data_figs.R
+if (!require("groundhog")) install.packages("groundhog"); 
+groundhog.folder <- paste0("Users/ftuser/Library/R/4.0/library/bci.roots.inversion")
+if(!dir.exists(file.path(groundhog.folder))) {dir.create(file.path(groundhog.folder))}
+set.groundhog.folder(groundhog.folder)
+library(groundhog)
+groundhog.day = "2021-01-01"
+groundhog.library('rmarkdown', groundhog.day)
+groundhog.library('bookdown', groundhog.day)
+rmarkdown::render("09.0_manuscript.rmd", output_format = "bookdown::html_document2")
 ```
@@
 ```
+
+```{r Chunk 11.0, eval = FALSE}
+remotes::install_github("ropenscilabs/reviewer")
+modified_file <- system.file("extdata/CheatSheet-modified.Rmd", package = "reviewer")
+reference_file <- system.file("extdata/CheatSheet.Rmd", package = "reviewer")
+library(reviewer)
+result <- diff_rmd(modified_file, reference_file)
+groundhog.library('diffobj', groundhog.day)
+browseVignettes("diffobj")
+remotes::install_github("ropenscilabs/trackmd")
+trackmd::trackChanges("09.0_manuscript.rmd")
+```
+<!-- groundhog.library('diffr', groundhog.day) -->
+<!-- diffr("09.0_manuscript_NP1.rmd", "09.0_manuscript.rmd") -->

### 2021-03-16 09.0_manuscript.rmd
Updated mortality results

@@
 
-Among evergreen species, ERD explained `r m.evg.r2.90`, `r m.evg.r2.05` and `r m.evg.r2.10`% variation in mortality rates during census intervals 1985-1990, 1985-1990, 2000-2005 and 2005-2010, respectively, such that species mortality rates significantly decreased with deeper ERD ($p$ = `r m.evg.p.90`, `r m.evg.p.05` and `r m.evg.p.10`, respectively; Fig.
+Among the seven census-intervals over 1982-2015 for which we analysed relationship of ERD with mortality rates, six intervals were associated with an El Niño event identiﬁed based on Oceanic Niño Index values > 0.5 °C for at least five consecutive months [@Detto.2018; @Condit.2017].
+Out of those six intervals, among evergreen species, ERD explained 30-40% variation in mortality rates during five intervals--1982-1985, 1985-1990, 1995-2000, 2000-2005 and 2005-2010--such that species mortality rates decreased with deeper ERD (p-values significant at alpha = 0.05 for four intervals and at alpha = 0.1 for one interval; Fig.
 \@ref(fig:erd-mort-census-evg)).
-When deciduous species were included, ERD explained lower but still significant proportion of variation in mortality in those intervals ($R$^2^ = `r m.r2[1]`, `r m.r2[2]` and `r m.r2[3]`% and $p$ = `r m.p.90`, `r m.p.05` and , respectively; Fig.
+When deciduous species were included, ERD explained 10-16% variation in mortality in four intervals with El-Nino events (p-values marginally significant, at alpha = 0.1; Fig.
 S11).
-Census intervals 2000-2005 and 2005-2010 were also the periods during which extreme soil droughts distinctly occurred in the wet season, effectively extending the dry season into the wet season (Fig.
+
+Our analysis of hydrological droughts ranged from 1990 to 2015 and revealed distinctive extreme hydrological droughts for the census-intervals 1995-2000, 2000-2005 and 2005-2010 for which ERD explained significant mortality (Fig.
 \@ref(fig:swp-droughts)).
-While our $\mathrm{\Psi}_{\textrm{soil}}$ estimates did not extend prior to 1990, due to a lack of reliable climate forcing, a prior study found that during the period 1985-1990, for which ERD explained significant mortality, BCI experienced three extreme deficit water years, following the 1983 El-Niño year [@Condit.2017].
-Species exposure to water-stress (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) non-linearly declined with ERD, indicating that species with shallower ERD spent greater time under significant hydrological drought, and experienced greater hydraulic-risk, which increased over 2000-2005 and 2005-2010 (Fig.
+Census interval 1995-2000 was marked with a prolonged and extreme hydrological drought during the 1997-1998 El-Nino event.
+During 2000-2005 and 2005-2010 extreme hydrological drought occurred in the wet season, effectively extending the dry season into the wet season (Fig.
+3).
+While our $\mathrm{\Psi}_{\textrm{soil}}$ estimates did not extend prior to 1990, due to a lack of reliable climate forcing, a prior study found that during census-intervals 1982-1985 and 1985-1990, for which ERD explained significant mortality, BCI experienced one and three extreme deficit water years, respectively; and with the El-Niño of 1983 marked by the harshest dry season since 1931 [@Condit.2017].
+
+On an average, species exposure to water-stress (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) declined with ERD, indicating that species with shallower ERD spent greater time under significant hydrological drought, and thus likely experienced greater hydraulic-risk, which steadily increased over 1995-2000, 2000-2005 and 2005-2010 (Fig.
 \@ref(fig:erd-time)).
 For the census interval 1990-1995, species with the shallowest ERD also showed greater realized hydraulic-risk (Fig.
-\@ref(fig:erd-time)), but during the dry season (Fig.
-\@ref(fig:swp-droughts)), and without elevated mortality rates (Fig.
+\@ref(fig:erd-time)), but without elevated mortality rates (Fig.
 \@ref(fig:erd-mort-census-evg)).

### 2021-03-19 09.0_manuscript.rmd
Changed Psi_crit = KL20 istead of KL50

--- 2021-03-16 09.0_manuscript.rmd
@@
   Here, we estimate plant-available water in the whole-soil column in a tropical forest, and inversely estimate rooting depths of co-occurring tree species from their growth responses, with a series of model calibrations and validation.
-We evaluate how rooting depth is linked to above-ground hydraulic traits and mortality rates over a 35 year period that experienced droughts of a variety of intensity, frequency and duration.
+We evaluate how rooting depth is linked to above-ground hydraulic traits and mortality rates through seve census-intervals over a 35 year period that experienced El Niño droughts of a variety of intensity, frequency and duration [@Detto.2018; @Condit.2017].
 Our hypotheses are that (i) deep-rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem due to greater and more reliable water availability at depth, whereas shallow-rooted species are associated with less efficient but safer hydraulic traits due to more variable availability of shallow soil-water (see Tables \@ref(tab:symbols) and \@ref(tab:hypo)); and that, (ii) deep-rooted species have lower mortality rates during droughts due to a lower exposure to water-stress compared to shallow-rooted species.
@@
 For this, we locally parameterized and calibrated 1-D hydrological water balance of the forest via an Earth System Model.
-Second, we developed a novel, empirical inverse model of ERD (run offline) and estimated ERDs for large trees of `r length(erd.sp)` species.
+Second, Model for effective rooting depth (ERD) we developed a novel, empirical inverse model of ERD (run offline) and estimated ERDs for large trees of `r length(erd.sp)` species.
 The ERD model incorporated the impact of atmospheric and hydrological drought on growth, and was constrained with species-specific leaf resistance curves.
@@
 We tested the relationship of ERD with xylem vulnerability to embolism, branch hydraulic safety margins, leaf turgor loss point and maximum stem-specific hydraulic conductivity.
-We analysed whether ERD correlated with mortality dynamics of large trees over 35 years that experienced several El-Nino droughts.
-Finally, to test whether ERD explained risk of drought-induced mortality, we analysed species-level exposure to water-stress as the time a species’ spent beyond its critical hydraulic threshold, given its ERD.
+We analysed whether ERD correlated with mortality dynamics of large trees over 35 years marked by several El Niño events.
+Finally, to test whether ERD explained risk of drought-induced mortality, we analysed species-level exposure to water-stress as the time a species’ spent beyond its critical hydraulic threshold given its ERD.
 
@@
 \@ref(eq:b-mod).
-We also estimated species $\mathrm{\Psi}$~50,leaf~ using their resistance curves.
+We also estimated species $\mathrm{\Psi}$~20,leaf~ using their resistance curves.
 
@@
 
-\We removed one species with a $\mathrm{\delta}$^2^H value of -33.5 (SE$\pm$2.6) which is particularly difficult to assign a source-water depth given the uncertainty in soil water $\mathrm{\delta}$^2^H data by depth (see Fig S3).
+\
+For comparison with modelled ERD, we removed six species from the @Meinzer:1999 dataset to account for the uncertainty in their water-sourcing depths.
+One species with a $\mathrm{\delta}$^2^H value of -33.5 (SE $\pm$ 2.6) which is particularly difficult to assign a source-water depth given the uncertainty in soil water $\mathrm{\delta}$^2^H data by depth (see Fig S3).
 $\mathrm{\delta}$^2^H~xylem~ from leafless trees may not be linked with water sourced at the time of measurement and thus may not be comparable to species that had leaves.
@@
 We assessed whether exposure to hydrological drought decreased with ERD.
-For this, we assumed crossing of  species-specific critical hydraulic threshold, $\mathrm{\Psi}$~crit~, here defined as $\mathrm{\Psi}$~50,leaf~, as an indicator of exposure to water-stress.
-Extended water-stress could lead to mortality via hydraulic failure or carbon starvation.
+For this, we assumed crossing of  species-specific critical hydraulic threshold, $\mathrm{\Psi}$~crit~, here defined as $\mathrm{\Psi}$~20,leaf~, as an indicator of exposure to water-stress.
+Extreme and prolonged water-stress could lead to mortality via hydraulic failure and/or carbon starvation [@Rowland:2015; @@McDowell:2018].
 The duration of exposure to water-stress, and thus realized hydraulic-risk, was defined for each species as the proportion of days in each census interval $t$ of the $\mathrm{\Psi}_\textrm{soil}$ simulation period (1990-2015) during which soil water potential in the soil layer matching species ERD, $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$, were more negative than species $\mathrm{\Psi}$~crit~.
@@
 \
-All statistical analyses were conducted in the *R* statistical environment (version 3.5.3, @R:2019).
+All statistical analyses were conducted in the *R* statistical environment (version 4.0.3, @R:2020).
 
@@
 \@ref(fig:swp-droughts)).
-Hydrological droughts in census intervals 2000-2005 and 2005-2010 distinctly occurred in the wet season, with one to two extreme hydrological drought years each, depending on the depth location (Fig.
+Hydrological droughts in census interval 1990-1995 was markedly prolonged over the dry season, while those during 2000-2005 and 2005-2010 distinctly occurred in the wet season, effectively extending the dry season into the wet season, with one to two extreme hydrological drought years each, depending on the depth location (Fig.
 \@ref(fig:swp-droughts)).
@@
 \@ref(fig:erd-iso)).
-Two of the alternative models with LAI had a strong relationship with $\mathrm{\delta}$^2^H~xylem~ as well, but one that was based on a smaller set of species ($n$ = 4), because fewer species passed the criteria of sufficient species-level goodness-of-fit for these models (<20 species vs.
+Two of the alternative models with LAI had a strong relationship with $\mathrm{\delta}$^2^H~xylem~ as well, but one that was based on a smaller set of species ($n$ = 4 or 5), because fewer species passed the criteria of sufficient species-level goodness-of-fit for these models (<17 species vs.
 `r length(erd.sp)` species for the best model, Fig.
@@
 \
-Modeled effective rooting depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from `r min(df.erd.to.plot$depth)` m to `r max(df.erd.to.plot$depth)` m (Fig.
+Modeled effective rooting depths (ERD) for the `r length(erd.sp)` large ($\ge$ 30 cm DBH) trees of canopy species varied from `r min(df.erd.to.plot$depth)` m to `r max(df.erd.to.plot$depth)` m (Fig.
 \@ref(fig:sp-erd)).
-Notably, two species, *Luehea seemannii* and *Trichilia tuberculata*, that @Meinzer:1999 found to have $\mathrm{\delta}$^2^H~xylem~ values between soil water and groundwater, suggesting that these species sourced most of the water from depths >1 m, were also identified by our model with ERD >1 m (2.9 m for both species, Fig.
+Evergreen species constituted a greater proportion of species with deeper ERDs compared with deciduous species, a group composed of a variety of categories of deciduousness.
+The range of ERD largely overlapped for both phenology groups (Fig.
+\@ref(fig:sp-erd)).
+Notably, two species, *Luehea seemannii* and *Trichilia tuberculata*, that @Meinzer:1999 found to have $\mathrm{\delta}$^2^H~xylem~ values between soil water and groundwater, suggesting that these species sourced most of the water from depths >1 m and likely to have sourced some portion of ground water, were also identified by our model with ERD >1 m (2.9 m for both species, Fig.
 \@ref(fig:erd-iso)).
@@
 
-Among evergreen species, ERD explained `r m.evg.r2.90`, `r m.evg.r2.05` and `r m.evg.r2.10`% variation in mortality rates during census intervals 1985-1990, 1985-1990, 2000-2005 and 2005-2010, respectively, such that species mortality rates significantly decreased with deeper ERD ($p$ = `r m.evg.p.90`, `r m.evg.p.05` and `r m.evg.p.10`, respectively; Fig.
+Among the seven census-intervals over 1982-2015 for which we analysed relationship of ERD with mortality rates, six intervals were associated with occurrence of one or more El Niño event, identiﬁed based on Oceanic Niño Index values > 0.5°C for at least five consecutive months [@Detto.2018; @Condit.2017].
+Out of the six intervals with El Niño events, among evergreen species, ERD explained 30-40% variation in mortality rates during five intervals--1982-1985, 1985-1990, 1995-2000, 2000-2005 and 2005-2010--such that species mortality rates decreased with deeper ERD (p-values significant at alpha = 0.05 for four intervals and at alpha = 0.1 for one interval; Fig.
 \@ref(fig:erd-mort-census-evg)).
-When deciduous species were included, ERD explained lower but still significant proportion of variation in mortality in those intervals ($R$^2^ = `r m.r2[1]`, `r m.r2[2]` and `r m.r2[3]`% and $p$ = `r m.p.90`, `r m.p.05` and , respectively; Fig.
+When deciduous species were included, ERD explained 10-16% variation in mortality in four intervals with El Niño events (p-values marginally significant at alpha = 0.1; Fig.
 S11).
-Census intervals 2000-2005 and 2005-2010 were also the periods during which extreme soil droughts distinctly occurred in the wet season, effectively extending the dry season into the wet season (Fig.
+
+Our analysis of hydrological droughts ranged from 1990 to 2015 and revealed distinctive extreme hydrological droughts for the census-intervals 1995-2000, 2000-2005 and 2005-2010, for which ERD explained significant mortality (Fig.
 \@ref(fig:swp-droughts)).
-While our $\mathrm{\Psi}_{\textrm{soil}}$ estimates did not extend prior to 1990, due to a lack of reliable climate forcing, a prior study found that during the period 1985-1990, for which ERD explained significant mortality, BCI experienced three extreme deficit water years, following the 1983 El-Niño year [@Condit.2017].
-Species exposure to water-stress (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) non-linearly declined with ERD, indicating that species with shallower ERD spent greater time under significant hydrological drought, and experienced greater hydraulic-risk, which increased over 2000-2005 and 2005-2010 (Fig.
+Census interval 1995-2000 experienced a prolonged and extreme hydrological drought during the 1997-1998 El Niño event.
+During 2000-2005 and 2005-2010 extreme hydrological drought occurred in the wet season, effectively extending the dry season water-stress into the wet season (Fig.
+3).
+While our $\mathrm{\Psi}_{\textrm{soil}}$ estimates did not extend prior to 1990, due to a lack of reliable climate forcing, a prior study found that during census-intervals 1982-1985 and 1985-1990, for which ERD explained significant mortality, BCI experienced one and three extreme deficit water years, respectively; the El-Niño of 1983 was marked by the harshest dry season since 1931 [@Condit.2017].
+
+On an average, species exposure to water-stress (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) steeply declined with ERD, indicating that species with shallower ERD spent greater time under significant hydrological drought, and thus likely experienced greater hydraulic-risk, which steadily increased over 1995-2000, 2000-2005 and 2005-2010 (Fig.
 \@ref(fig:erd-time)).
 For the census interval 1990-1995, species with the shallowest ERD also showed greater realized hydraulic-risk (Fig.
-\@ref(fig:erd-time)), but during the dry season (Fig.
-\@ref(fig:swp-droughts)), and without elevated mortality rates (Fig.
+\@ref(fig:erd-time)), but without elevated mortality rates (Fig.
 \@ref(fig:erd-mort-census-evg)).
@@
 Species with deeper ERD showed higher maximum stem hydraulic conductance, and greater xylem vulnerability to embolism and more negative above-ground hydraulic safety margins compared to shallow ERD species.
-Nevertheless, deeper ERD species showed lower mortality rates than shallower ERD species in significant wet season droughts over the 35-yr record (1981-2015) (Fig.
-S11), especially amongst the evergreen species (Fig.
+Nevertheless, species with deeper ERD, among the evergreen but not deciduous phenology group, showed lower mortality rates than shallower ERD species in five out of six census-intervals that had experienced one or more El-Nino events (over seven census-intervals studied in total; 1982-2015).
+Quantifying extreme droughts in the whole soil column over 1990-2015 revealed that ERD explained mortality in periods when extreme soil droughts prolonged over the dry season or occurred over the wet season effectively extending the dry season water-stress (Fig.
 \@ref(fig:erd-mort-census-evg)).
-Analysis of time spent beyond $\mathrm{\Psi}$~crit~ in the soil zone matching ERD $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ showed that this was likely due to decreasing realized risk of water-stress with increasing access to water for species with deeper ERD (Fig.
+Analysis of species time spent beyond $\mathrm{\Psi}$~crit~ in the soil zone matching ERD $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ showed that exposure to water-stress steeply declined with deeper ERD (Fig.
 \@ref(fig:erd-time)).
-We thus demonstrate, for the first time, a link between drought exposure based on water-availability in the whole soil-column, above- and below-ground hydraulic traits, and long-term mortality outcomes for trees in a tropical forest canopy.
+Because extreme, sustained water-stress increases risk of mortality via hydraulic-failure and/or carbon starvation, deeper ERD may mitigate drought-induced mortality by limiting exposure to water-stress [@Rowland:2015; @McDowell:2018].
+We thus demonstrate, for the first time, a link between species trade-offs in above- and below-ground hydraulic traits, drought exposure based on hydrological droughts quantified over the whole soil-column, and large ($\ge$ 10 cm DBH) tree mortality through El-Nino events occurred over 35 years.
 
@@
 
-Our study brings attention to the need for assessing drought sensitivity in terms of species realized drought exposure by accounting for hydrological drought and trees' rooting depths.
+Our study brings attention to the need for assessing drought sensitivity in terms of species drought exposure and realized hydraulic-risk by accounting for hydrological drought and trees' rooting depths.
 We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins (Fig.
-\@ref(fig:erd-hyd)), that is, traits that are commonly identified as proxies for mortality risk [@Anderegg:2016], but they were less droughted and had lower mortality compared to species relying on shallower water through periods of extreme droughts (Figs.
-\@ref(fig:swp-droughts), \@ref(fig:erd-mort-census-evg), S11).
+\@ref(fig:erd-hyd)), that is, traits that are commonly identified as proxies for mortality risk [@Anderegg:2016], but they were less droughted and, among the evergreen but not necessarily deciduous species, they had lower mortality compared to species relying on shallower water through periods of extreme droughts (Figs.
+\@ref(fig:swp-droughts), \@ref(fig:erd-mort-census-evg)).
 Deeper ERD species on average had more hydraulically efficient, but risky, suite of above-ground traits optimized for an environment of more reliable and greater water source at depth that resulted in a lower mortality risk.
@@
 \@ref(fig:erd-mort-census-evg), \@ref(fig:erd-time)).
-Our results are consistent with studies that included three or four species each with known rooting or water-sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016; also see @Brum.2019; @Brum.2017].
+We found that a greater proportion of deeper ERD species had evergreen phenology increased with deeper ERD.
+The trend of mortality under extreme soil drought in evergreen but not deciduous species is consistent with the expectation that deciduous species avoid exposure to drought-stress via leaf-drop.
+
+Our results are consistent with recent studies that analyses rooting or water-sourcing depths versus hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016; also see @Brum.2019; @Brum.2017; @Rowland:2015].
 Globally, large trees tend to exhibit greater growth reductions, lower post-drought resilience, and greater increases in mortality relative to their understory counterparts [@Bennett:2015; @Phillips:2010].
@@
 $\mathrm{\Psi}_{\textrm{soil},z}$ by DOY for a year (colored line) is only shown if at least one day of the year $\mathrm{\Psi}_{\textrm{soil},z}$ was more negative than the 5^th^ percentile of $\mathrm{\Psi}_{\textrm{soil},z}$ for that day of year (DOY).
+Note the distinctive features of extreme droughts that occurred during the three periods indicated by a $*$: all three periods featured extreme hydrological droughts that either distinctly prolonged over the dry season (1995-2000), or distinctly occurred in the wet season, effectively extending the dry season (2000-2005, 2005-2010).
 
@@
 Figure \@ref(fig:erd-time): Modeled effective rooting depth (ERD; horizontal-axis) versus time spent beyond critical hydraulic threshold (vertical axis) by census interval (colored bars; note the squared y-axis scale).
-Each bar represents the average time species of the same ERD spent beyond species-specific critical hydraulic thresholds in a given interval, that is, the proportion of days for which $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ was more negative than $\mathrm{\Psi}$~crit~, defined as $\mathrm{\Psi}$~50,leaf~, and where $z$ is the soil depth matching species ERD.
+Each bar represents the average time species of the same ERD spent beyond species-specific critical hydraulic thresholds in a given interval, that is, the proportion of days for which $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ was more negative than species $\mathrm{\Psi}$~crit~, defined as $\mathrm{\Psi}$~20,leaf~, and where $z$ is the soil depth matching species ERD.
 Standard error of the mean are shown over each bar.

### 2021-03-19 code/07.0_PhenoDemoTraitsPsi.R

@@ -159,17 +159,17 @@ psi.corr.fun.ls.2 <- list(
 )
 
 get.mfac.ls <- list(
-  "mr.kl50.I" =
+  "mr.kl20.I" =
     function(df) {
       result.df <-
-        as.data.table(psi.study)[, psi.mod := indicator(psi, df$psi_kl50, greater.than = FALSE)][
+        as.data.table(psi.study)[, psi.mod := indicator(psi, df$psi_kl20, greater.than = FALSE)][
           , keyby = .(depth, interval, par.sam), .(mfac = sum(psi.mod, na.rm = TRUE))][
             , keyby = .(depth, interval), .(mfac = mean(mfac, na.rm = TRUE))]
       result.df <- data.frame(result.df) %>% pivot_wider(names_from = "depth", values_from = "mfac")
       return(result.df)
     }
 )
-# Originally written by Sean M McMahon, then modified
+# This likelihood function Originally written by Sean M McMahon, then modified
 
 get.ts.lk <- function(df) {
   demo.ts <- df$demo.rate
@@ -867,7 +867,7 @@ for (i in 1:length(names.mfac)) {
 save(mfac.interval.long, file = file.path(results.folder, "mfac.interval.long.Rdata"))
 ## Ordered along Rooting Depth Index
 # load(file = file.path(results.folder, "mfac.interval.long.Rdata"))
-mfac.on <- "mr.kl50.I"
+mfac.on <- "mr.kl20.I"
 mrate.depth <-
   adult.mrate.long %>% mutate(size = "large") %>%
   # mrate.long %>%

### 2021-03-19 10.0_Supporting_Information.Rmd
Edited figure legend

@@
 
-```{r erd-mort-census, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-census}Species mortality rate in each census interval during the study period plotted against ERD.
+```{r erd-mort-census, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-census}Species (circles) Mortality rates in species of various deciduous leaf-habit categories (colors) across census intervals from 1985-2015 plotted against ERD.
 $R$^2^ and significance levels for linear model fits are given in panel insets.
-Linear model fits (blue lines) with confidance bands (gray area) are shown for significant fits at $\\alpha$ = 0.05, identified by census-intervals with a superscript *."}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_by rdi.gr.jpeg"))
+Species are color colded by deciduous category.
+Census-intervals with a superscript * are shown for comparison and indicate intervals in which significant mortality was explained by ERD among evergreen species (Fig.
+8 in the Main text)."}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_by rdi.gr_deciduous.jpeg"))
 ```

### 2021-04-05 09.0_manuscript.rmd
included co-authors' feedback

@@
 output:
+  bookdown::word_document2:
+        reference_docx: manuscriptstyle.docx
+        fig_caption: yes
+        number_sections: false
   bookdown::html_document2:
@@
 groundhog.day = "2021-01-01"
-pkgs=c('knitr', 'pander', 'broom')
+pkgs=c('knitr','pander', 'broom')
 groundhog.library(pkgs, groundhog.day)
 
-# options(tinytex.verbose = TRUE)
 source("code/06.0_load.R")
@@
 Chitra-Tarak, Rutuja*^1,2^, Chonggang Xu^1^, Salomón Aguilar^3^, Kristina Anderson-Teixeira^3,4^, Jeff Chambers^5^, Matteo Detto^3,6^, Boris Faybishenko^5^, Rosie A.
-Fisher^7^, Ryan Knox^5^, Charlie Koven^5^, Lara Kueppers^5,8^, Nobert Kunert^3,4,9^, Stefan J.
-Kupers^10^, Nathan G.
-McDowell^11,12^, Brent D.
+Fisher^7,8^, Ryan Knox^5^, Charlie Koven^5^, Lara Kueppers^5,9^, Nobert Kunert^3,4,10^, Stefan J.
+Kupers^11^, Nathan G.
+McDowell^12,13^, Brent D.
 Newman^1^, Steven R.
-Paton^3^, Rolando Pérez^3^, Laurent Ruiz^13,14,15^, Lawren Sack^16^, Jeffrey M.
-Warren^17^, Brett T.
-Wolfe^3,18^, Cynthia Wright^17^, S.
-Joseph Wright^3^, Joseph Zailaa^4,16,19^, Sean M.
+Paton^3^, Rolando Pérez^3^, Laurent Ruiz^14,15,16^, Lawren Sack^17^, Jeffrey M.
+Warren^18^, Brett T.
+Wolfe^3,19^, Cynthia Wright^18^, S.
+Joseph Wright^3^, Joseph Zailaa^4,17,20^, Sean M.
 McMahon^2,3^
@@
 Boulder, CO, USA 
-8. University of California Berkeley, Energy and Resources Group, 310 Barrows Hall #3050
+8. Laboratoire Évolution & Diversité Biologique, CNRS:UMR 5174, Université Paul Sabatier, Toulouse, France.
+9. University of California Berkeley, Energy and Resources Group, 310 Barrows Hall #3050
 Berkeley, CA, USA 94720 
-9. Institute of Botany, University of Natural Resources and Life Sciences Vienna, Department of Integrative Biology and Biodiversity Research, Gregor‐Mendel‐Str 33, A‐1190 Wien, Wien, AT 
-10. German Centre for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig, Computational Forest Ecology, Leipzig, Saxony, DE 
-11. Atmospheric Sciences and Global Change Division, Pacific Northwest National Lab, PO Box 999, Richland, WA, USA 99352 
-12. School of Biological Sciences, Washington State University, PO Box 644236, Pullman, WA, USA 99164-4236 
-13. Indo-French Cell for Water Sciences, Indian Institute of Science, Bangalore, Karnataka, IN 
-14. IRD, CNRS, UPS, UMR GET, Toulouse, FR 
-15. INRAE, Institut Agro, UMR SAS, Rennes, FR 
-16. University of California Los Angeles, Ecology and Evolutionary Biology, 612 Charles E. Young Drive South, Los Angeles, CA, USA 90095 
-17. Oak Ridge National Laboratory, Environmental Sciences Division, Oak Ridge, TN, USA 
-18. School of Renewable Natural Resources, Louisiana State University Agricultural Center, Baton Rouge, LA 70803, USA 
-19. California State University Los Angeles, Biological Sciences Department, Los Angeles, CA, USA
+10. Institute of Botany, University of Natural Resources and Life Sciences Vienna, Department of Integrative Biology and Biodiversity Research, Gregor‐Mendel‐Str 33, A‐1190 Wien, Wien, AT 
+11. German Centre for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig, Computational Forest Ecology, Leipzig, Saxony, DE 
+12. Atmospheric Sciences and Global Change Division, Pacific Northwest National Lab, PO Box 999, Richland, WA, USA 99352 
+13. School of Biological Sciences, Washington State University, PO Box 644236, Pullman, WA, USA 99164-4236 
+14. Indo-French Cell for Water Sciences, Indian Institute of Science, Bangalore, Karnataka, IN 
+15. IRD, CNRS, UPS, UMR GET, Toulouse, FR 
+16. INRAE, Institut Agro, UMR SAS, Rennes, FR 
+17. University of California Los Angeles, Ecology and Evolutionary Biology, 612 Charles E. Young Drive South, Los Angeles, CA, USA 90095 
+18. Oak Ridge National Laboratory, Environmental Sciences Division, Oak Ridge, TN, USA 
+19. School of Renewable Natural Resources, Louisiana State University Agricultural Center, Baton Rouge, LA 70803, USA 
+20. California State University Los Angeles, Biological Sciences Department, Los Angeles, CA, USA
 
@@
 ------------- | -------------  
-Summary | 197  
-Main Text  | 6491  
+Summary | 200  
+Main Text  | 8024  
 Introduction | 833  
-Methods | 3285  
-Results | 873  
-Discussion | 1500  
+Methods | 4132  
+Results | 924  
+Discussion | 2135  
 Keywords | 8
@@
 - We inversely estimated the effective rooting depth (ERD, as the average depth of water extraction), for 32 canopy species by linking diameter growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and leaf hydraulic resistance curves.
-We validated ERD estimates against existing empirical data of potential water access depths.
-- Across species, deeper ERD was associated with higher maximum stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates during extreme droughts over 35 years (1981-2015).
+We validated ERD estimates against existing isotopic data of potential water access depths.
+- Across species, deeper ERD was associated with higher maximum stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates especially in evergreen leaf-habit during extreme droughts over 35 years (1981-2015).
 Species with deeper ERD did not cross critical hydraulic thresholds indicating that trees compensate for water-stress related mortality risk through deep-water access.
@@
 Drought-induced mortality in tropical forests may have significant global implications.
-Tropical forests play a disproportionately large role in the global carbon and energy cycles [@Bonan:2008], and support half of global biodiversity [@Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019].
+Tropical forests play a disproportionately large role in the global carbon and energy cycles [@Bonan:2008], and support half of global biodiversity [@Wright:20050th], but face a threat from intensifying droughts [@Malhi.2008; @Doughty:2015; @Xu.2019].
 Tropical forests have been expected to be an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw].
 Still, mortality events in the tropics are rarely as large as those in temperate and boreal zones [@McDowell.2018], leading to questions regarding the role of large trait-diversity and hydraulic strategies in mitigating mortality events.
-Nevertheless, mortality rates are increasing in some tropical regions [@Brienen.2015; @Hubau.2020] and widespread drought-induced tree mortality has occurred across the tropics for specific functional groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010].
+Further, mortality rates are increasing in some tropical regions [@Brienen.2015; @Hubau.2020] and widespread drought-induced tree mortality has occurred across the tropics for specific functional groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010].
 State-of-the-art dynamic global vegetation models (DGVMs) struggle to capture these drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013; @Powell.2018], because underlying mechanisms of drought-tolerance are not fully understood nor quantified.
@@
 Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation, stem water storage, deep-water access, increased root production, hydraulic redistribution and embolism resistance [@McDowell.2008].
-Deep-water access is arguably the most long-standing, yet under-studied mechanism.
+Deep-water access is arguably the most long-standing [@Cannon.1911], yet under-studied mechanism.
 As plant-available water varies with depth, trees in the same forest with different rooting depths, depending on species and size [@Meinzer:1999; @Chitra-Tarak.2018; @Brum.2019], differ in their experience during a drought, and thus in their growth and mortality responses [@Chitra-Tarak.2018].
@@
 \
-  Characterizing the essential constraints to model species-rich communities [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010; @Bartlett.2019; @Lu.2020] entails identifying the topography of plant trait trade-offs in different environments, and how these relate to demographic rates (growth, recruitment and mortality).
+  Characterizing the essential constraints to model species-rich communities [@Koven:2020; @Christoffersen:2016kd2d; @Maréchaux.2017; @Wright:2010; @Bartlett.2019; @Lu.2020] entails identifying the topography of plant trait trade-offs in different environments, and how these relate to demographic rates (growth, recruitment and mortality).
 Significant efforts have been invested into identifying and linking universal drought indices, above-ground traits, and demographic rates.
-Such efforts have found correlations between vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013, @Venturas.2016].
-Even though deep roots may mitigate hydraulic vulnerability [@Brum.2019] and mortality-risk, hydraulic-risk realized through hydrological (rather than meteorological) droughts--the outcome of the interaction between rooting depths, critical hydraulic thresholds and the actual vertical distribution of plant-available water--are not studied, especially in tropical regions.
+Such efforts have found correlations between vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013; @Venturas.2016].
+Deep roots may mitigate hydraulic vulnerability [@Brum.2019] and mortality-risk, in particular during hydrological (rather than meteorological) droughts.
+Nonetheless, the interaction between rooting depths, above-ground hydraulic traits, hydrological droughts quantified over the whole soil-column and mortality outcomes are hardly studied.
 
 \
-  Here, we estimate plant-available water in the whole-soil column in a tropical forest, and inversely estimate rooting depths of co-occurring tree species from their growth responses, with a series of model calibrations and validation.
-We evaluate how rooting depth is linked to above-ground hydraulic traits and mortality rates through seve census-intervals over a 35 year period that experienced El Niño droughts of a variety of intensity, frequency and duration [@Detto.2018; @Condit.2017].
-Our hypotheses are that (i) deep-rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem due to greater and more reliable water availability at depth, whereas shallow-rooted species are associated with less efficient but safer hydraulic traits due to more variable availability of shallow soil-water (see Tables \@ref(tab:symbols) and \@ref(tab:hypo)); and that, (ii) deep-rooted species have lower mortality rates during droughts due to a lower exposure to water-stress compared to shallow-rooted species.
-To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic traits and long-term mortality outcomes for a species-rich tropical forest.
+  In this paper, we estimate plant-available water in the whole-soil column in a tropical forest, and inversely estimate rooting depths of co-occurring tree species from their growth responses, with a series of model calibrations and validation.
+We evaluate how rooting depth is linked to above-ground hydraulic traits and mortality rates through several census-intervals over a 35 year period that experienced El Niño droughts of a variety of intensity, frequency and duration [@Detto.2018; @Condit.2017].
+Our hypotheses are that (i) deep-rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem due to greater and more reliable water availability at depth (see Tables \@ref(tab:symbols) and \@ref(tab:hypo)); and that, (ii) deep-rooted species have lower mortality rates during droughts due to a lower exposure to water-stress compared to shallow-rooted species.
+To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic traits and multi-decadal mortality outcomes for a species-rich tropical forest.
 
@@
 \@ref(fig:schematic) for a workflow).
-We defined species-specific effective rooting depths (ERD) as the depth at which soil water potential best explained species' growth dynamics.
-To estimate ERDs we used two models in succession: first, we obtained the daily dynamics of soil water potential in the whole-soil column.
-For this, we locally parameterized and calibrated 1-D hydrological water balance of the forest via an Earth System Model.
-Second, Model for effective rooting depth (ERD) we developed a novel, empirical inverse model of ERD (run offline) and estimated ERDs for large trees of `r length(erd.sp)` species.
+We defined species-specific effective rooting depths (ERD) as the depth at which the growth factor determined by soil water potential and leaf hydraulic traits best explained species' growth dynamics over 25 years.
+We developed a novel, empirical inverse model of ERD and estimated ERDs for large trees of `r length(erd.sp)` species.
 The ERD model incorporated the impact of atmospheric and hydrological drought on growth, and was constrained with species-specific leaf resistance curves.
 For the latter, we used existing data for eight species and developed trait-based proxies for the rest (based on data for a total of 21 species).
-We validated ERD against existing stable hydrogen isotope ratios ($\mathrm{\delta}$^2^H~xylem~) for a subset of six tree species as independent observations.
+We obtained the daily dynamics of soil water potential in the whole-soil column (up to 13 m) over the 25 years by locally parameterizing a 1-D hydrological water balance of the forest within a land surface model for an average vegetation type.
+The water balance was calibrated on available measurements, in particular continuous soil moisture data in three surface layers (over the first meter of soil), stream discharge and ET fluxes.
+
+We validated our ERD estimates against existing stable hydrogen isotope ratios ($\mathrm{\delta}$^2^H~xylem~) for a subset of six tree species as independent observations.
 We evaluated whether ERD was associated with above-ground hydraulic traits sourcing the latter from a set of rare datasets for six to seven species that overlapped with our assessment of ERD.
 We tested the relationship of ERD with xylem vulnerability to embolism, branch hydraulic safety margins, leaf turgor loss point and maximum stem-specific hydraulic conductivity.
-We analysed whether ERD correlated with mortality dynamics of large trees over 35 years marked by several El Niño events.
+We analysed whether ERD is correlated with mortality dynamics of large trees over 35 years marked by several El Niño events for evergreen and deciduous species.
 Finally, to test whether ERD explained risk of drought-induced mortality, we analysed species-level exposure to water-stress as the time a species’ spent beyond its critical hydraulic threshold given its ERD.
@@
 Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014].
-Rainfall at BCI is seasonal with a mean annual total of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a pronounced dry season from mid-December through April with a mean total of 292 mm [@Paton.2020].
+Rainfall at BCI is seasonal with a mean annual total of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a pronounced dry season from mid-December through April with less than 100 mm of rainfall per month [@Paton.2020].
 In the 50ha (1000 m × 500 m) old growth forest plot, all stems $\ge$ 1 cm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every five year through 1985-2015 for growth, mortality as well as recruitment of new stems into the 1 cm dbh size class [@Condit.2017].
@@
 
-## Model ELM-FATES for soil matric potentials
+## Model for effective rooting depth (ERD)
 
-### Model description
+### ERD model description
 
-To estimate dynamics of water availability by depth over the forest’s rooting zone, we used the Energy Exascale Earth System Land Model (ELM, @Caldwell.2019), coupled with the Functionally Assembled Terrestrial Ecosystem Simulator (FATES, @Koven:2020) (hereafter, ELM-FATES).
-ELM is a land-model that, among many features, simulates the physics and conservative dynamics of water, energy and carbon fluxes.
-In particular, soil hydrologic fluxes are vertically resolved among discrete soil layers (1-D) similar to that in CLM4.5 [@Bonan:2013].
-FATES is a community-based, open-source model used for studying climate-vegetation interactions.
-FATES is a vegetation demography model, with a size-structured group of plants (cohorts) and successional trajectory-based patches based on the ecosystem demography [@Moorcroft.2001] approach.
-FATES couples to ELM by a common interface of water and carbon fluxes.
-Detailed descriptions for ELM and FATES can be found elsewhere [@Bisht.2018; @Fisher.2010; @Fisher.2015; @Koven:2020].
-
-\
-  ELM-FATES simulates interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\mathrm{\Delta}W_{\textrm{can}}$, surface water $\mathrm{\Delta}W_{\textrm{sfc}}$, soil water $\mathrm{\Delta}W_{\textrm{liq},z_{i}}$ at depth $z_{i}$ and water in an unconfined aquifer $\mathrm{\Delta} W_{a}$ (omitting processes relevant to snow, wetlands or lakes).
-Conservation of these terms is expressed as follows:
+Roots could impact tree growth through two factors: 1) water uptake; and 2) nutrient uptake.
+Because our study is situated in an old-growth tropical forest, we assume that nutrients are mainly concentrated in the shallow soil layers derived from litter decomposition, and that rooting depths do not substantially affect nutrient uptake.
+Because our study site has a dry season, we assume that rooting depth is a key factor affecting tree growth.
+We defined species-specific effective rooting depth (ERD) as the depth at which a soil moisture growth limitation factor, $B$, determined by the soil water potential and hydraulic traits best explained species’ diameter growth dynamics.
+$B$ is used to approximate the amount of stomatal closure due to water-stress in the soil.
+In this study, we used the loss of conductivity of leaf ($FLC$~leaf~) to estimate $B$.
+$FLC$~leaf~ is calculated as a fraction of maximum leaf conductivity $K_\textrm{max, leaf}$:
 
 \begin{equation}
-\begin{split}
-\mathrm{\Delta}W_\textrm{can} + \mathrm{\Delta}W_\textrm{sfc} + \sum_{i=1}^{15} \mathrm{\Delta}W_{\textrm{liq},z_{i}} + \mathrm{\Delta}W_{a} = \left(q_\textrm{rain} - E_{v} - E_{g} - q_\textrm{over} - q_\textrm{h2osfc} - q_\textrm{drai}\right)\mathrm{\Delta}t
-\end{split}
-(\#eq:water-balance)
-\end{equation}
-where, $q_{\textrm{rain}}$ is rainfall, $E_{v}$ is transpiration of the forest, $E_{g}$ is ground evaporation, $q_{\textrm{over}}$ is surface runoff, $q_{\textrm{h2osfc}}$ is runoff from surface water storage, $q$~drai~ is subsurface drainage (all in mm H~2~O) and $\mathrm{\Delta}t$ is the time-step (here, 1 hr).
-The soil profile is discretized into maximum 15 exponentially distributed soil layers with layer node depth $z$.
-Here, $z$ $\in$ $Z$; $Z$ = (0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13) m.
-
-### Model parameterization
-
-To parameterize ELM-FATES, we identified eleven parameters relevant for Eq.
-\@ref(eq:water-balance), determined their ranges based on literature for the study site, else for the tropics (Table S1), and ran 5000 simulations using Latin Hypercube Sampling [LHS; @Stein.1987] from this global parameter space.
-In the Conrad catchment, in which the 50 ha plot is situated, soil hydraulic conductivity decreases with depth from 12.5 and 60 cm (@Godsey:2004; Table S1).
-We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths <12.5 cm and >60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig.
-S1).
-For greater accuracy, instead of the ELM default pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content (GWC) vs.
-$\mathrm{\Psi}$~soil~ for depths of 0.15, 0.4 and 1 m [@Kupers:2019].
-We converted GWC to volumetric (VWC, $\theta$) using a measured bulk density value of 0.8 g cm^-3^ (see below).
-A @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences in water retention among depths were nominal,
-
-\begin{equation}
-\mathrm{\Psi}_{\textrm{soil}} = \mathrm{\Psi}_e\left(\frac{\mathrm{\theta}}{\mathrm{\theta}_{s}}\right)^{-b}
-(\#eq:soil-ret-curve)
-\end{equation}
-where $\mathrm{\Psi}_e$ is the air-entry matric potential, $\mathrm{\theta}_{s}$ is the saturated volumetric water content, and $b$, an index for soil pore-size distribution, respectively.
-$\mathrm{\Psi}_e$, $\mathrm{\theta}_{s}$ and $b$ that best-fit the data were 200 (mm H~2~O), 0.51 (cm^3^ cm^-3^) and 10 (unitless), respectively.
-
-\
-We ran ELM-FATES for the 5000-member parameter ensemble with hourly climate drivers measured at the BCI meteorological station over 1985-2018 initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode (see https://biogeodb.stri.si.edu/physical_monitoring/research/barrocolorado for meteorological data and @Faybishenko.2018 for QA/QC procedures).
-As our key interest here was on deriving soil water availability, we ran ELM-FATES in a lower-complexity configuration: static stand structure, and with a single plant functional type (PFT) of evergreen trees.
-The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous [@Condit:2000] and addition of a dry-deciduous PFT did not significantly alter results (not shown).
-Holding the stand structure static rather than allowing it to emerge from physiologically-mediated competition between trees removes a set of internal model feedbacks and thus the only differences between model simulations are driven by direct physiological effects of perturbed model parameters.
-
-### Model calibration
-
-We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq.
-\@ref(eq:water-balance), namely, (1) evapotranspiration ET from the flux tower by the 50 ha plot ($\backsimeq$ $E_{v} + E_{g}$ in Eq.
-\@ref(eq:water-balance); 2012-2017; see [SI](#SIData.matteo))), (2) local stream discharge $q_\textrm{run}$ ($\backsimeq$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see @Paton.2019 and [SI](#SIData.steveP)) and (3) soil volumetric water content (VWC) $\mathrm{\theta}_{z}$ ($\backsimeq$ $W_{\textrm{liq,} z}$) from two sources: (i) a long-term (2012-2018) record of VWC $\mathrm{\theta}_\textrm{0-0.15 m}$ averaged across three vertical time domain reflectometry (TDR) probes over depth 0-15 cm from three locations near the flux tower, and (ii) plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 at depths of 0.15, 0.4 and 1 m (1299 samples covering all soil types and habitats; see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\mathrm{\theta}_{\textrm{0-0.15 m}}$, $\mathrm{\theta}_{\textrm{0.4 m}}$ and $\mathrm{\theta}_{\textrm{1 m}}$ refer to plot-wide averages).
-
-\
- For ELM-FATES calibration we calculated an objective function for each of the 5000-member ensemble by equally weighting standardized RMSE between observations and simulations across all fluxes and states.
-We describe the objective function $O$ as :
-
-\begin{equation}
-\begin{split}
-\overline{O} = \frac{1}{6}\sum \textrm{RMSE}^{*}\textrm{(ET,}E_{v} + E_{g}) + \textrm{RMSE}^{*}(q_{\textrm{run}},q_{\textrm{over}} + q_{\textrm{drai}}) + \textrm{RMSE}^{*}(\mathrm{\theta}_{\textrm{0-0.15~m}}, {W_{\textrm{liq,}\overline{z}}}) \\
-& + \textrm{RMSE}^{*}(\mathrm{\theta}_{\textrm{0.15~m}}, W_{\textrm{liq,`r soil.depths[4]`m}}) + \textrm{RMSE}^{*}(\mathrm{\theta}_{\textrm{0.4~m}}, W_{\textrm{liq,`r soil.depths[6]`~m}}) + \textrm{RMSE}^{*}(\mathrm{\theta}_{1~m}, W_{\textrm{liq,`r soil.depths[8]`~m}})
-(\#eq:obj-func)
-\end{split}
+FLC_{\textrm{leaf}} = K_{\textrm{leaf}}/K_{\textrm{max, leaf}}
+(\#eq:k-soil)
 \end{equation}
 
-based on monthly sums for both ET and $q$~run~ and daily averages for $\mathrm{\theta}$s.
-$W_{\textrm{liq,}\overline{z}}$ is the average of daily $W_{\textrm{liq,} z}$, where $z$ $\in$ $Z$; $Z$ = (`r paste(soil.depths[1:4], collapse=", ")`) m.
-Superscript * indicates that RMSE is standardized between 0 and 1 across the 5000 simulations.
+where $K$~leaf~ is the current leaf hydraulic conductance and $K_\textrm{max, leaf}$ is the maximum leaf hydraulic conductance (see Table \@ref(tab:symbols) for trait symbols and their definitions)).
+The dynamics of $K$~leaf~, and thus $FLC$~leaf~, are estimated from the species-specific relationship between $K$~leaf~ vs.
+$\mathrm{\Psi}$~leaf~, referred to as the leaf hydraulic resistance curve [@Sack.2012].
+Because pre-dawn $\mathrm{\Psi}$~leaf~ generally approaches $\mathrm{\Psi}$~soil~, this allowed us to use $\mathrm{\Psi}$~soil~ in leaf hydraulic resistance curves to predict maximum diurnal $K$~leaf~.
+$K$~leaf~ is strongly related to photosynthetic capacity [@Brodribb.2002].
+Because we were interested in relating $K$~leaf~ to 5-year average diameter growth observations, we ignored diurnal dynamics of $K$~leaf~.
+$K_\textrm{leaf}$ in our models, being based on resistance curves, is thus dynamic and depends on rooting depth and rhizosphere water potential.
+We considered $K_\textrm{leaf}$ to be a hydraulic bottleneck in the plant and that it determines or co-determines whole plant hydraulic response to soil drying; notably, root hydraulic resistance may also be a strong driver, but recent studies suggest that its resistance would be similar to, or correlated with that of leaves (see @Cochard.2020 and references therein).
 
-\
-Preliminary analyses showed that simultaneously fitting the model to all states and fluxes did not capture the soil moisture dynamics at 1 m depth well, our key priority for model calibration.
-To ensure model fit at 1 m depth, we first chose 300 parameter ensemble members that minimized $\textrm{RMSE}^{*}({\mathrm{\theta}_{\textrm{1 m}}}, W_{\textrm{liq,`r soil.depths[8]` m}})$ (< 0.2), and among those 300 ensemble members we chose 100 with the lowest values for the objective function $\overline{O}$.
-See Table S1 for the ranges of best-fit values for different parameters.
-
-### Model evaluation
-
-We evaluated ELM-FATES by calculating RMSE between simulated and observed long-term daily VWC for the depths of 0.1, 0.4 and 1 m (2016-2018), based on a dataset we had left-out during calibration.
-We obtained these observations from three horizontal TDR probes at the depths of 0.1, 0.4 and 1 m at a location near the vertical probes (see [SI](#SIData.VWC)).
-
-### Hydrological droughts
-
-We ran ELM-FATES with the chosen 100 ensemble members from 1985-2018, with the first 5 years used for model spin-up.
-Extreme hydrological droughts were identified by depth $z$ as days for which $\mathrm{\Psi}_{\textrm{soil},z}$ was more negative than the 5^th^ percentile of $\mathrm{\Psi}_{\textrm{soil},z}$ for a given day of the year.
-
-## Model for effective rooting depth (ERD)
-
-### Model description
-
-We defined species-specific effective rooting depth (ERD) as the depth at which soil water potential best explained species' growth dynamics.
-We inversely estimated ERD, by testing alternate empirical models of species-specific 5-yearly census growth, which use ERD as an input (Eq.
-\@ref(eq:M-chosen) and those in SI).
-
-Many intrinsic and extrinsic factors limit plant growth, including soil moisture, VPD, radiation, leaf area seasonality, rooting depth and the rate at which plant hydraulic conductance declines with water-stress [@Lawrence.2019; @Brodribb.2002].
-The latter could be studied at the organ level, for example, $K$~leaf~ vs.
-$\mathrm{\Psi}$~leaf~ at leaf-level, referred to as the leaf hydraulic resistance curve (@Sack.2012; see Table \@ref(tab:symbols) for trait symbols and their definitions).
-We assumed $K_\textrm{leaf}$ to be a hydraulic bottleneck in the plant and that it determines or co-determines whole plant hydraulic response to soil drying; notably, root hydraulic resistance may also be a strong driver, but recent studies suggest that its resistance would be similar to, or correlated with that of leaves (see @Cochard.2020 and references therein).
-Because pre-dawn $\mathrm{\Psi}$~leaf~ generally approaches $\mathrm{\Psi}$~soil~, this allowed us to use $\mathrm{\Psi}$~soil~ in leaf hydraulic resistance curves to predict maximum diurnal $K$~leaf~, which we assumed controls photosynthetic rate and growth (to relate to 5-year average growth values), ignoring diurnal dynamics of $K$~leaf~ [@Brodribb.2002; @Brodribb.2017].
-$K_\textrm{leaf}$ in our models, being based on resistance curves, is thus dynamic and depends on rooting depth and rhizosphere water potential.
-
+Many other intrinsic and extrinsic factors limit plant growth, including soil moisture, VPD, radiation, leaf area seasonality [@Lawrence.2019; @Brodribb.2002].
 VPD affects growth non-linearly, with growth increasing with VPD up-to a VPD threshold, then decreasing as leaf pores (stomata) close reducing water-uptake [@Grossiord.2020; @Yang.2019].
@@
 S2).
-Apart from stomatal control, leaf-drop may further limit water-uptake and growth.
+Apart from stomatal control, leaf deciduousness may further limit water-uptake and growth.
 
-We therefore regressed species-specific growth against multiplicative or additive effects of one or more growth factors calculated daily and averaged over 5-yearly censuses: $\hat{\textrm{VPD}}$, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1.
+We therefore tested alternate empirical growth models, or inverse models of ERD, in which we regressed species-specific growth against multiplicative or additive effects of one or more growth factors calculated daily and averaged over 5-yearly censuses: $\hat{\textrm{VPD}}$, $FLC_\textrm{leaf}$ and the leaf area index (LAI), all standardized to range between 0 and 1 (Eq.
+\@ref(eq:M-chosen) and those in SI).
 Incorporation of radiation did not improve model fitting, possibly due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season at BCI.
@@
 \
- For the best model (see model validation and selection below), daily average growth for species $s$, $G_{s,t|z,h}$, in the census interval $t$ for depth $z$ and for a given hydrological realization $h$ is described as follows:
+ The best empirical model we found to describe daily average tree growth (see model validation and selection below), daily average growth for species $s$, $\hat{G}$, in the census interval $t$ for depth $z$ and for a given hydrological realization $h$ is described as follows:
 
 \begin{equation}
-G_{s,t|z,h} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},s,i|z,h}^{*} \hat{\textrm{VPD}}_{i}^{*}) + \epsilon_{s,t|z,h}
+\hat{G}_{s,t|z,h} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} FLC_{\textrm{leaf},s,i|z,h}^{*} \hat{\textrm{VPD}}_{i}^{*}) + \epsilon_{s,t|z,h}
 (\#eq:M-chosen)
 \end{equation}
-where | refers to conditionals, n~t~ is the total number of days in census interval $t$, the superscript * indicates that the variable has been standardized between 0 and 1 (within species for $K_{\textrm{leaf}}$), $\beta_{0}$ and $\beta_{1}$ are model coefficients and $\epsilon$ is the model error term.
+where | refers to conditionals, n~t~ is the total number of days in census interval $t$, the superscript * indicates that the variable has been standardized to range between 0 and 1 (within species for $FLC_{\textrm{leaf}}$), $\beta_{0}$ and $\beta_{1}$ are model coefficients and $\epsilon$ is the model error term.
 See SI for all the alternate models tested.
@@
 \
- For each species, we estimated diurnal maximum value of $K_{\textrm{leaf},s,i|z,h}$ for a species $s$ by substituting $\mathrm{\Psi}_{\textrm{soil},s,i|z,h}$ in the leaf hydraulic resistance curves defined below in Eq.
+ For each species, we estimated diurnal maximum value of $K_{\textrm{leaf},s,i|z,h}$ for a species $s$ by substituting $\mathrm{\Psi}_{\textrm{leaf}}$ with $\mathrm{\Psi}_{\textrm{soil}}$ in the leaf hydraulic resistance curves defined below in Eq.
 \@ref(eq:vc-exp) using species-specific parameters $A_{s}$ and $B_{s}$ as:
@@
 \end{equation}
-Checking through each $z$, we estimated species effective rooting depths (ERD), defined as the depth $z$ at which soil water dynamics ($\mathrm{\Psi}_{\textrm{soil,} z}$) best explained observed dynamics of species growth.
-Using Eq.
-\@ref(eq:M-chosen), for a species $s$ and hydrological realization *h*, we obtained Akaike's Information Criterion (AIC) and Goodness-of-fit ($R$^2) for each soil depth $z$.
-We used AIC as the first selection criterion to ensure that the fitted values are within the confidence interval of the error distribution (as the same model Eq.
+$K_{\textrm{max, leaf},s}$, was obtained as the maximum value of $K_{\textrm{leaf},s}$ using Eq. \@ref(eq:k-soil).
+
+We estimated species effective rooting depths (ERD) as the depth $z$ at which soil water dynamics ($\mathrm{\Psi}_{\textrm{soil,} z}$) best explained observed dynamics of growth via modeled growth (Eq.\@ref(eq:M-chosen)).
+Our ERD model does not explicitly use rooting profiles, but finds soil water dynamics at a single depth $z$ as the central tendency that influences the observed growth dynamics the most.
+We modelled multiple hydrological realisations of soil water-potential dynamics with a water-balance model (see below), and thus defined ERD as the median of best-fit depths across all hydrological realisations.
+
+Formally, species-specific best-fit ERD was selected as follows: using Eq.
+\@ref(eq:M-chosen), for a species $s$ and hydrological realization *h*, we obtained Akaike's Information Criterion (AIC) and Goodness-of-fit ($R$^2^) for each modeled soil layer depth $z$.
+We used AIC as the first selection criterion to ensure that the fitted growth values from Eq.
+\@ref(eq:M-chosen) are within the confidence interval of the error distribution of observed growth (as the same model Eq.
 \@ref(eq:M-chosen) was being compared for each *h*, number of model parameters were the same).
-For each *h*, we retained those depths with the maximum likelihood (similar AIC based support, see below) filtering out the rest.
+For each *h*, we retained those depths with the maximum likelihood of growth (similar AIC based support, see below) filtering out the rest.
 However, AIC does not ensure that fitted values will have a positive trend with growth.
-We therefore used $R$^2 as a goodness-of-fit measure and retained depths associated with a relatively high threshold $R$^2 value of 0.5; after filtering out depths associated with negative correlations if any.
-Finally, among these we chose the depth with the highest $R$^2 per *h*.
+We therefore used $R$^2^ between observed and fitted growth values as a goodness-of-fit measure and retained depths associated with a relatively high threshold $R$^2^ value of 0.5; after filtering out depths associated with negative correlations $r$ if any.
+Finally, among these we chose the depth with the highest $R$^2^ per *h*.
 Results were not sensitive to inclusion of this final step of retaining only a single depth as the best-fit (not shown).
@@
 \begin{equation}
-\textrm{ERD}_{s} = \underset{\forall \, h \, \in \, H}{\textrm{med}} \Big(z\,\Big|\, \Big(r_{s|z,h} = \underset{\forall \, z \,\in\,Z}{\textrm{max}}(r_{s|z,h}| \mathrm{\Delta}_{s|z,h} \le 2\Big)\Big) \, \bigcap \, r_{s|z,h} \ge 0.71\Big)
+\textrm{ERD}_{s} = \underset{\forall \, h \, \in \, H}{\textrm{med}} \Big(z\,\Big|\, \Big(R^2_{s|z,h} = \underset{\forall \, z \,\in\,Z}{\textrm{max}}(R^2_{s|z,h}| \mathrm{\Delta}_{s|z,h} \le 2\Big)\Big) \, \bigcap \, R^2_{s|z,h} \ge 0.5 \, \bigcap \,  r_{s|z,h} \ge 0\Big)
 (\#eq:erd)
@@
 
-where $\mathrm{\Delta}_{s|z,h} = \textrm{AIC}_{s|z,h} - \underset{\forall \, z \,\in\,Z}{\textrm{min}}(\textrm{AIC}_{s|z,h})$,  $\textrm{AIC}_{s|z,h} = -2\textrm{log}(\mathcal{L}(\hat{\beta}_{s|z,h}|{G_{s}})) - 2k$, $\mathcal{L}$ is the likelihood function, $\hat{\beta}$ is the maximum-likelihood estimate, $G$ the observed growth time series with $k$ observations (or, census intervals $t$).
+where
+
+\begin{equation}
+\mathrm{\Delta}_{s|z,h} = \textrm{AIC}_{s|z,h} - \underset{\forall \, z \,\in\,Z}{\textrm{min}}(\textrm{AIC}_{s|z,h})
+(\#eq:delta)
+\end{equation}
+
+and
+
+\begin{equation}
+\textrm{AIC}_{s|z,h} = -2\textrm{log}(\mathcal{L}(\hat{\beta}_{s|z,h}|{G_{s}})) - 2\textrm{k}
+(\#eq:aic)
+\end{equation}
+
+where r is the correlation, and $R^2$ the Goodness-of-fit, between observed growth $G$ and estimated growth $\hat{G}.
+$\mathcal{L}$ is the likelihood function, $\hat{\beta}$ is the maximum-likelihood estimate, $G$ the observed growth with $k$ observations (or, census intervals $t$).
 $H$ = (1, .., 100), the 100 best-fit hydrological realizations and $Z$ = (0.4, `r paste(soil.depths[8:length(soil.depths)], collapse=", ")`) m.
@@
 We obtained leaf hydraulic resistance curves ($K_\textrm{leaf}$ vs.
-$\mathrm{\Psi}$~leaf~) for adult trees of 21 common species from Zailaa *et* *al.* (to be submitted; see SI for brief description of methods) described as:
+$\mathrm{\Psi}$~leaf~) for adult trees of 21 common species at BCI from Zailaa *et* *al.* (in prep; see SI for brief description of methods) described as:
 
@@
 These 21 species included eight of the `r length(erd.sp)` species selected for ERD estimation.
-For the rest `r sp.n.model` species, we obtained resistance curves using trait-based proxies.
+For the remaining `r sp.n.model` species, we obtained resistance curves using trait-based proxies.
 We identified scaling relationships between fitted parameters $A$ and $B$ in Eq.
@@
 
-### Model validation and selection
+#### Using ELM-FATES to model soil matric potentials
+
+##### ELM-FATES model description
+
+To estimate dynamics of water availability by depth over the forest’s rooting zone, we used the Energy Exascale Earth System Land Model (ELM, @Caldwell.2019), coupled with the Functionally Assembled Terrestrial Ecosystem Simulator (FATES, @Koven:2020) (hereafter, ELM-FATES).
+ELM is a land-model that, among many features, simulates the physics and conservative dynamics of water, energy and carbon fluxes.
+In particular, soil hydrologic fluxes are vertically resolved among discrete soil layers (1-D) similarly to the CLM4.5 [@Oleson.2013].
+FATES is a community-based, open-source model used for studying climate-vegetation interactions.
+FATES is a vegetation demography model, with a size-structured group of plants (cohorts) and successional trajectory-based patches based on the ecosystem demography [@Moorcroft.2001] approach.
+FATES couples to ELM by a common interface of water and carbon fluxes.
+Detailed descriptions for ELM and FATES can be found elsewhere [@Bisht.2018; @Fisher.2010; @Fisher.2014; @Koven:2020].
+
+\
+  We ran ELM with FATES vegetation, in which the ELM model simulates interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\mathrm{\Delta}W_{\textrm{can}}$, surface water $\mathrm{\Delta}W_{\textrm{sfc}}$, soil water $\mathrm{\Delta}W_{\textrm{liq},z_{i}}$ at depth $z_{i}$ and water in an unconfined aquifer $\mathrm{\Delta} W_{a}$ (omitting processes relevant to snow, wetlands or lakes).
+Conservation of these terms is expressed as follows:
+
+\begin{equation}
+\begin{split}
+\mathrm{\Delta}W_\textrm{can} + \mathrm{\Delta}W_\textrm{sfc} + \sum_{i=1}^{15} \mathrm{\Delta}W_{\textrm{liq},z_{i}} + \mathrm{\Delta}W_{a} = \left(q_\textrm{rain} - E_{v} - E_{g} - q_\textrm{over} - q_\textrm{h2osfc} - q_\textrm{drai}\right)\mathrm{\Delta}t
+\end{split}
+(\#eq:water-balance)
+\end{equation}
+where, $q_{\textrm{rain}}$ is rainfall, $E_{v}$ is transpiration of the forest, $E_{g}$ is ground evaporation, $q_{\textrm{over}}$ is surface runoff, $q_{\textrm{h2osfc}}$ is runoff from surface water storage, $q$~drai~ is subsurface drainage (all in mm H~2~O) and $\mathrm{\Delta}t$ is the time-step (here, 1 hr).
+The soil profile is discretized into maximum 15 exponentially distributed soil layers with layer node depth $z$.
+Here, $z$ $\in$ $Z$; $Z$ = (0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13) m.
+
+In ELM, soil water is predicted from a multi-layer model, in which the vertical soil moisture transport is governed by infiltration, surface and sub-surface runoff, gradient diffusion, gravity, canopy transpiration through root extraction, and interactions with groundwater [@Oleson.2013].
+FATES simulates size-structured tree community together with individual tree rooting profiles.
+ELM removes total community canopy transpiration from the soil layers in proportion to the total root fraction in each layer in the 1-D column.
+
+##### ELM-FATES model parameterization
+
+To parameterize catchment hydrology in ELM-FATES, we identified eleven parameters relevant for Eq.
+\@ref(eq:water-balance), determined their ranges based on literature for the study site, else for the tropics (Table S1), and ran 5000 simulations using Latin Hypercube Sampling [LHS; @Stein.1987] from this global parameter space.
+In the Conrad catchment, in which the 50 ha plot is situated, soil hydraulic conductivity decreases with depth from 12.5 and 60 cm (@Godsey:2004; Table S1).
+We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths <12.5 cm and >60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig.
+S1).
+For greater accuracy, instead of the ELM default pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content (GWC) vs.
+$\mathrm{\Psi}$~soil~ for depths of 0.15, 0.4 and 1 m [@Kupers:2019].
+We converted GWC to volumetric (VWC, $\theta$) using a measured bulk density value of 0.8 g cm^-3^ (see below).
+The @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences in water retention among depths were nominal,
+
+\begin{equation}
+\mathrm{\Psi}_{\textrm{soil}} = \mathrm{\Psi}_e\left(\frac{\mathrm{\theta}}{\mathrm{\theta}_{s}}\right)^{-b}
+(\#eq:soil-ret-curve)
+\end{equation}
+where $\mathrm{\Psi}_e$ is the air-entry matric potential, $\mathrm{\theta}_{s}$ is the saturated volumetric water content, and $b$, an index for soil pore-size distribution, respectively.
+$\mathrm{\Psi}_e$, $\mathrm{\theta}_{s}$ and $b$ that best-fit the data were 200 (mm H~2~O), 0.51 (cm^3^ cm^-3^) and 10 (unitless), respectively.
+
+\
+We ran ELM-FATES for the 5000-member 11 parameter ensemble with hourly climate drivers measured at the BCI meteorological station over 1985-2018 [@Faybishenko.2021] initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode.
+As our key interest here was on deriving soil water availability, we ran ELM-FATES in a lower-complexity configuration: static stand structure, and with a single plant functional type (PFT) of evergreen trees.
+The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous [@Condit:2000] and addition of a dry-deciduous PFT did not significantly alter results (not shown).
+In static stand structure model, the model is started from an observed forest inventory and recruitment, growth and mortality are deactivated.
+Holding the stand structure static rather than allowing it to emerge from physiologically-mediated competition between trees removes a set of internal model feedbacks and thus the only differences between model simulations are driven by direct physiological effects of perturbed model parameters.
+This allows us to focus on the hydraulic and hydrological questions explored in this paper.
+
+Note that we use the soil water potential dynamics obtained from ELM-FATES as an ecosystem-level outcome to drive our species-level, empirical ERD model, and thus believe that there is no conflict/circularity between tree-level rooting profiles and water uptake processes in ELM-FATES and the ERD model.
+
+##### ELM-FATES model calibration
+
+We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq.
+\@ref(eq:water-balance), namely, (1) evapotranspiration ET from the flux tower by the 50 ha plot ($\backsimeq$ $E_{v} + E_{g}$ in Eq.
+\@ref(eq:water-balance); 2012-2017; see [SI](#SIData.matteo))), (2) local stream discharge $q_\textrm{run}$ ($\backsimeq$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see @Paton.2019 and [SI](#SIData.steveP)) and (3) soil volumetric water content (VWC) $\mathrm{\theta}_{z}$ ($\backsimeq$ $W_{\textrm{liq,} z}$) from two sources: (i) a long-term (2012-2018) record of VWC $\mathrm{\theta}_\textrm{0-0.15 m}$ averaged across three vertical time domain reflectometry (TDR) probes over depth 0-15 cm from three locations near the flux tower, and (ii) plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 at depths of 0.15, 0.4 and 1 m (1299 samples covering all soil types and habitats; see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\mathrm{\theta}_{\textrm{0-0.15 m}}$, $\mathrm{\theta}_{\textrm{0.4 m}}$ and $\mathrm{\theta}_{\textrm{1 m}}$ refer to plot-wide averages).
+
+\
+ For ELM-FATES calibration we calculated an objective function for each of the 5000-member ensemble by equally weighting standardized RMSE between observations and simulations across all fluxes and states.
+We describe the objective function $O$ as :
+
+\begin{equation}
+\begin{split}
+\overline{O} = \frac{1}{6}\sum \textrm{RMSE}^{*}\textrm{(ET,}E_{v} + E_{g}) + \textrm{RMSE}^{*}(q_{\textrm{run}},q_{\textrm{over}} + q_{\textrm{drai}}) + \textrm{RMSE}^{*}(\mathrm{\theta}_{\textrm{0-0.15~m}}, {W_{\textrm{liq,}\overline{z}}}) \\
+& + \textrm{RMSE}^{*}(\mathrm{\theta}_{\textrm{0.15~m}}, W_{\textrm{liq,`r soil.depths[4]`m}}) + \textrm{RMSE}^{*}(\mathrm{\theta}_{\textrm{0.4~m}}, W_{\textrm{liq,`r soil.depths[6]`~m}}) + \textrm{RMSE}^{*}(\mathrm{\theta}_{1~m}, W_{\textrm{liq,`r soil.depths[8]`~m}})
+(\#eq:obj-func)
+\end{split}
+\end{equation}
+
+based on monthly sums for both ET and $q$~run~ and daily averages for $\mathrm{\theta}$s.
+$W_{\textrm{liq,}\overline{z}}$ is the average of daily $W_{\textrm{liq,} z}$, where $z$ $\in$ $Z$; $Z$ = (`r paste(soil.depths[1:4], collapse=", ")`) m.
+Superscript * indicates that RMSE is standardized between 0 and 1 across the 5000 simulations.
+
+\
+Model fit to soil moisture dynamics by depth was one of our key priority.
+Preliminary analyses showed that fitting the model to all states and fluxes simultaneously captured soil moisture dynamics for shallower depths but not at 1 m depth.
+To ensure model fit at 1 m depth, we thus imposed a two-step procedure, wherein we first chose the 300 parameter ensemble members that minimized $\textrm{RMSE}^{*}({\mathrm{\theta}_{\textrm{1 m}}}, W_{\textrm{liq,`r soil.depths[8]` m}})$ (< 0.2), and among those 300 ensemble members we chose 100 with the lowest values for the objective function $\overline{O}$.
+See Table S1 for the ranges of best-fit values for different parameters.
+
+##### Hydrological droughts
+
+We ran ELM-FATES with the chosen 100 ensemble members from 1985-2018, with the first 5 years used for model spin-up.
+Extreme hydrological droughts were identified by depth $z$ as days for which $\mathrm{\Psi}_{\textrm{soil},z}$ was more negative than the 5^th^ percentile of $\mathrm{\Psi}_{\textrm{soil},z}$ for a given day of the year.
+
+##### ELM-FATES model evaluation
+
+We evaluated ELM-FATES by calculating RMSE between simulated and observed long-term daily VWC for the depths of 0.1, 0.4 and 1 m (2016-2018), based on a dataset we had left-out during calibration.
+We obtained these observations from three horizontal TDR probes at the depths of 0.1, 0.4 and 1 m at a location near the vertical probes (see [SI](#SIData.VWC)).
+
+### ERD model validation and selection
 
@@
 
-We evaluated whether ERD was associated with above-ground hydraulic traits sourcing the latter from a set of rare datasets [@Wolfe.2019; @Wolfe.2020] for six to seven species that overlapped with our assessment of ERD.
+We evaluated whether ERD was associated with above-ground hydraulic traits sourcing the latter from existing datasets [@Wolfe.2019; @Wolfe.2021] for seven species that overlapped with our assessment of ERD.
 We regressed species ERD against maximum stem hydraulic conductivity $K_\textrm{max,stem}$ (*n*=7), leaf turgor loss point $\mathrm{\Psi}$~tlp~ (*n*=7), vulnerability to embolism from cavitation $\mathrm{\Psi}$~88,stem~ measured in terms of pressure at which 88% of $K_\textrm{max,stem}$ is lost (*n*=7), and hydraulic safety margins $\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~ (*n*=6).
@@
 \
-We regressed species ERD against species $\textrm{M}_{s,t}$ for each census interval $t$ for the whole pool of ($n$ = `r nrow(mrate.depth.mean)`) species with $\textrm{M}_{s,t}$ estimates and for evergreen species alone ($n$ = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`).
-We assessed whether exposure to hydrological drought decreased with ERD.
-For this, we assumed crossing of  species-specific critical hydraulic threshold, $\mathrm{\Psi}$~crit~, here defined as $\mathrm{\Psi}$~20,leaf~, as an indicator of exposure to water-stress.
-Extreme and prolonged water-stress could lead to mortality via hydraulic failure and/or carbon starvation [@Rowland:2015; @@McDowell:2018].
+Dry season deciduous species may escape drought exposure via leaf deciduousness, so we analysed deciduous and evergreen species separately.
+Tree species on BCI are scored by expert botanists as one among four leaf-habits; evergreen, brevideciduous, facultative deciduous and obligate deciduous [@Meakem.2018; see SI].
+We pooled all deciduous leaf habits together (hereafter 'deciduous' for brevity) and regressed species ERD against species $\textrm{M}_{s,t}$ for each census interval $t$ for deciduous ($n$ = `r nrow(mrate.depth.mean %>% subset(deciduous != "E"))`) and evergreen species ($n$ = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`) separately.
+
+Extreme and prolonged water-stress could lead to mortality via hydraulic failure and/or carbon starvation [@Rowland:2015; @McDowell:2018].
+As an indicator of exposure to water-stress, we use a species-specific critical hydraulic threshold, $\mathrm{\Psi}$~crit~, here defined as $\mathrm{\Psi}$~20,leaf~.
 The duration of exposure to water-stress, and thus realized hydraulic-risk, was defined for each species as the proportion of days in each census interval $t$ of the $\mathrm{\Psi}_\textrm{soil}$ simulation period (1990-2015) during which soil water potential in the soil layer matching species ERD, $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$, were more negative than species $\mathrm{\Psi}$~crit~.
@@
 \@ref(fig:hydro-calib)**d**), slightly underestimating peak discharge in wet years and slightly overestimating peak ET.
-The reduction in parameter range (Table S1) in the best-fit ensembles compared to the tested global ranges showed that the model calibration was sensitive to the Ball-Berry stomatal slope parameter, the ELM root distribution parameter that regulates the depth of the rooting profile, soil hydraulic conductivity profile especially at depth, and the adjustment factor that modifies soil hydraulic conductivity to account for macroporosity and direct flow paths (Figs.
+The reduction in parameter range (Table S1) in the best-fit ensembles compared to the tested global ranges showed that the model calibration was primarily sensitive to the Ball-Berry stomatal slope parameter (fates_leaf_BB_slope), the ELM root distribution parameter that regulates the depth of the rooting profile (fates_rootb_par), soil hydraulic conductivity (HKSAT) profile especially at depth, and the adjustment factor (HKSAT_ADJ) that modifies soil hydraulic conductivity to account for macroporosity and direct flow paths (Figs.
 S4, S5).
@@
 \
-Soil water potential dynamics $\mathrm{\Psi}_{\textrm{soil},z}$ by depth z from 0.01 through 13 m obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of extreme years, drought intensity at depth, drought seasonality and duration (Fig.
+Soil water potential dynamics $\mathrm{\Psi}_{\textrm{soil},z}$ by depth z from 0.01 through 13 m obtained from the best-fit simulations revealed that every 5-yr census interval had at least one extreme hydrological drought year, but varied in terms of number of extreme years, drought intensity, drought seasonality and duration (Fig.
 \@ref(fig:swp-droughts)).
-Hydrological droughts in census interval 1990-1995 was markedly prolonged over the dry season, while those during 2000-2005 and 2005-2010 distinctly occurred in the wet season, effectively extending the dry season into the wet season, with one to two extreme hydrological drought years each, depending on the depth location (Fig.
+Hydrological droughts in census interval 1990-1995 was marked by a prolonged dry season, while those during 2000-2005 and 2005-2010 distinctly occurred in the wet season, effectively extending the dry season into the wet season (Fig.
 \@ref(fig:swp-droughts)).
@@
 
-The best ERD model (with a multiplicative effect of VPD and not LAI, Eq.
-\@ref(eq:M-chosen)) explained a large fraction of the variance in $\mathrm{\delta}$^2^H~xylem~ ($R$^2^ = `r lm.erd.iso.r2`, $p$ = `r lm.erd.iso.p`, $n$ = `r length(unique(erd.iso$sp))`, Fig.
-\@ref(fig:erd-iso)).
-Two of the alternative models with LAI had a strong relationship with $\mathrm{\delta}$^2^H~xylem~ as well, but one that was based on a smaller set of species ($n$ = 4 or 5), because fewer species passed the criteria of sufficient species-level goodness-of-fit for these models (<17 species vs.
-`r length(erd.sp)` species for the best model, Fig.
-S9).
-For the rest of the analyses we therefore used ERD estimates from the less parsimonious, best model.
+The best ERD model explained a large fraction of the variance in $\mathrm{\delta}$^2^H~xylem~ ($R$^2^ = `r lm.erd.iso.r2`, $p$ = `r lm.erd.iso.p`, $n$ = `r length(unique(erd.iso$sp))`, Fig.
+\@ref(fig:erd-iso); also see SI).
+This model included an effect of VPD and not LAI (Eq.
+\@ref(eq:M-chosen)).
 
@@
 \@ref(fig:erd-hyd)**b**), less negative $\mathrm{\Psi}$~88,stem~ and thus greater vulnerability to xylem embolism from cavitation ($r$ = `r cor.stem.88`, $p$ = `r p.stem.88`, Fig.
-\@ref(fig:erd-hyd)**c**), and more negative above-ground hydraulic safety margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~, $r$ = `r cor.hsm`, $p$ = `r p.hsm`, Fig.
+\@ref(fig:erd-hyd)**c**), and narrower above-ground hydraulic safety margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~, $r$ = `r cor.hsm`, $p$ = `r p.hsm`, Fig.
 \@ref(fig:erd-hyd)**d**).
@@
 
-Among the seven census-intervals over 1982-2015 for which we analysed relationship of ERD with mortality rates, six intervals were associated with occurrence of one or more El Niño event, identiﬁed based on Oceanic Niño Index values > 0.5°C for at least five consecutive months [@Detto.2018; @Condit.2017].
+Among the seven census-intervals over 1982-2015 for which we analysed relationship of ERD with mortality rates, six intervals were associated with occurrence of one or more El Niño event [@Detto.2018; @Condit.2017].
 Out of the six intervals with El Niño events, among evergreen species, ERD explained 30-40% variation in mortality rates during five intervals--1982-1985, 1985-1990, 1995-2000, 2000-2005 and 2005-2010--such that species mortality rates decreased with deeper ERD (p-values significant at alpha = 0.05 for four intervals and at alpha = 0.1 for one interval; Fig.
@@
 3).
-While our $\mathrm{\Psi}_{\textrm{soil}}$ estimates did not extend prior to 1990, due to a lack of reliable climate forcing, a prior study found that during census-intervals 1982-1985 and 1985-1990, for which ERD explained significant mortality, BCI experienced one and three extreme deficit water years, respectively; the El-Niño of 1983 was marked by the harshest dry season since 1931 [@Condit.2017].
+ERD also explained significant mortality in earlier droughts [1982-1985 and 1985-1990; @Condit.2017], not covered by our $\mathrm{\Psi}_{\textrm{soil}}$ estimates.
 
-On an average, species exposure to water-stress (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) steeply declined with ERD, indicating that species with shallower ERD spent greater time under significant hydrological drought, and thus likely experienced greater hydraulic-risk, which steadily increased over 1995-2000, 2000-2005 and 2005-2010 (Fig.
-\@ref(fig:erd-time)).
+On an average, species exposure to water-stress (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) steeply declined with ERD, indicating that species with shallower ERD spent greater time under significant hydrological drought, and thus likely experienced greater hydraulic-risk, which exponentially increased over 1995-2000, 2000-2005 and 2005-2010 (Fig.
+\@ref(fig:erd-time-evg)).
 For the census interval 1990-1995, species with the shallowest ERD also showed greater realized hydraulic-risk (Fig.
-\@ref(fig:erd-time)), but without elevated mortality rates (Fig.
+\@ref(fig:erd-time-evg)), but without elevated mortality rates (Fig.
 \@ref(fig:erd-mort-census-evg)).
-Overall, species with ERD deeper than 1.7 m did not cross species $\mathrm{\Psi}$~crit~ for the entire 25-yr period for which $\mathrm{\Psi}_{\textrm{soil},z}$ was modeled (1990-2015).
-Soil depths deeper than 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and drought periods, which is above $\mathrm{\Psi}$~crit~ of the most sensitive tree species (-0.28 MPa).
+Overall, species with ERD deeper than 2.9 m did not cross species $\mathrm{\Psi}$~crit~ for the entire 25-yr period for which $\mathrm{\Psi}_{\textrm{soil},z}$ was modeled (1990-2015).
+Soil depths deeper than 1.7 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and drought periods, which is above $\mathrm{\Psi}$~crit~ of the most sensitive tree species (`r min.psi_crit` MPa).
 
@@
 
-Our novel approach for estimating ERD using 25 years of tree growth, species-specific leaf resistance curves, modeled soil moisture profiles, and observed VPD was consistent with estimates using $\mathrm{\delta}$^2^H~xylem~ (Figs.
+We introduce a novel approach for estimating ERD using 25 years of tree growth, species-specific leaf resistance curves, modeled soil water potential profiles, and observed VPD.
+Our predictions of ERD were consistent with estimates using $\mathrm{\delta}$^2^H~xylem~ (Figs.
 \@ref(fig:erd-iso), \@ref(fig:sp-erd)).
-Our analyses revealed that co-occurring large canopy tree species with deeper ERD were associated with higher above-ground hydraulic efficiency, but not safety (Fig.
+Our analyses suggest that co-occurring large canopy tree species with deeper ERD were associated with higher above-ground hydraulic efficiency, but not safety (Fig.
 \@ref(fig:erd-hyd)).
-Species with deeper ERD showed higher maximum stem hydraulic conductance, and greater xylem vulnerability to embolism and more negative above-ground hydraulic safety margins compared to shallow ERD species.
-Nevertheless, species with deeper ERD, among the evergreen but not deciduous phenology group, showed lower mortality rates than shallower ERD species in five out of six census-intervals that had experienced one or more El-Nino events (over seven census-intervals studied in total; 1982-2015).
-Quantifying extreme droughts in the whole soil column over 1990-2015 revealed that ERD explained mortality in periods when extreme soil droughts prolonged over the dry season or occurred over the wet season effectively extending the dry season water-stress (Fig.
+Species with deeper ERD showed higher maximum stem hydraulic conductance, and greater xylem vulnerability to embolism and narrower above-ground hydraulic safety margins compared to shallow ERD species.
+Nevertheless, species with deeper ERD, among the evergreen but not necessarily deciduous leaf habits, showed lower mortality rates than shallower ERD species.
+This ERD-mortality relationship was significant in five out of six census-intervals that had experienced one or more El-Nino events over seven census-intervals studied in total (1982-2015).
+Quantifying extreme droughts in the whole soil column over 1990-2015 revealed that ERD explained significant levels of mortality in periods when extreme soil droughts prolonged the dry season water-stress (Fig.
 \@ref(fig:erd-mort-census-evg)).
-Analysis of species time spent beyond $\mathrm{\Psi}$~crit~ in the soil zone matching ERD $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ showed that exposure to water-stress steeply declined with deeper ERD (Fig.
-\@ref(fig:erd-time)).
+Analysis of species time spent beyond $\mathrm{\Psi}$~crit~ in the soil zone matching ERD $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ showed that exposure to water-stress exponentially declined with deeper ERD (Fig.
+\@ref(fig:erd-time-evg)).
 Because extreme, sustained water-stress increases risk of mortality via hydraulic-failure and/or carbon starvation, deeper ERD may mitigate drought-induced mortality by limiting exposure to water-stress [@Rowland:2015; @McDowell:2018].
-We thus demonstrate, for the first time, a link between species trade-offs in above- and below-ground hydraulic traits, drought exposure based on hydrological droughts quantified over the whole soil-column, and large ($\ge$ 10 cm DBH) tree mortality through El-Nino events occurred over 35 years.
+We thus demonstrate, for the first time, a link between species trade-offs in above- and below-ground hydraulic traits, drought exposure through extreme hydrological droughts quantified over the whole soil-column across several El-Nino events, and large ($\ge$ 10 cm DBH) tree mortality over 35 years.
+
+## Drought strategies designed to mitigate realised hydraulic-risks
+
+Our findings indicate how eco-evolutionary drought strategies may have been shaped by trait coordination and investments adapted for realised risks in the environment.
+If investments in stress-tolerance traits come at a cost, such insurance may not pay-off if the risk of stress is not real.
+Our findings suggest that shallow rooted species pay the cost of hydraulic safety in terms of hydraulic efficiency (Fig.
+\@ref(fig:erd-hyd)), adapted for an environment in which hydraulic-risk is significant, as extreme droughts cause exponentially greater water-stress in shallow soil layers (Fig.
+S6); whereas species with investment in deep roots can afford a risky suite of hydraulic traits, because access to a reliable deep-water resource ensures that for them hydraulic-risk is not realised (Figs.
+\@ref(fig:swp-droughts)).
+
+Our findings suggest that the above- and below-ground trait coordination associated with differential environmental risks has adaptive value in terms of plant performance.
+Shallow rooted species on an average had greater mortality rates, especially for evergreen species, than deep rooted species through prolonged extreme droughts (Fig.
+\@ref(fig:erd-mort-census-evg)) that may have crossed even the greater tolerance limits of shallow-rooted species (Fig.
+\@ref(fig:erd-hyd)).
+Hydraulic-risk for shallow rooted species may have been exacerbated in our 35-yr study period in which extreme water-deficit years were more frequent than those in the last century according to a prior study [@Condit.2017].
+Deep rooted evergreen species on an average had greater survival than shallow rooted species through extreme droughts; although their survival may not be guaranteed if droughts intensify in future.
+
+Our finding of greater mortality under extreme soil water-stress among evergreen but not deciduous species is consistent with the expectation that deciduous species can also avoid drought-stress via leaf deciduousness.
+Studies on simultaneous coordination between leaf phenology, rooting depth and hydraulic traits are almost absent for the tropics and warrant future consideration [@Oliviera.2021].
+Our analysis for ERD-hydraulic traits relationship was limited to only seven species, but covered all species of different leaf phonologies (Fig.
+\@ref(fig:erd-hyd)), so here we assume that the trend in the ERD-hydraulic trait relationships holds true across all species of different leaf phonologies.
+We also found that the range of ERDs overlapped across leaf phenology, but evergreen species tend to have deeper ERDs, consistent with observations by @Meinzer:1999 and elsewhere [@Smithmartin.2020; @Fan:2017; @Oliviera.2021].
+Whether deciduous species also have more efficient and vulnerable hydraulics at BCI as is observed elsewhere remains to be seen [@Markesteijn:2010; @Markesteijn:2011; @Xu:2016; @Gleason.2016].
+Leaf pehnologies at BCI are numerous and complicated and need further research in the future.
 
@@
 We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins (Fig.
-\@ref(fig:erd-hyd)), that is, traits that are commonly identified as proxies for mortality risk [@Anderegg:2016], but they were less droughted and, among the evergreen but not necessarily deciduous species, they had lower mortality compared to species relying on shallower water through periods of extreme droughts (Figs.
+\@ref(fig:erd-hyd)).
+These traits are commonly identified as proxies for mortality risk [@Anderegg:2016], but in fact were associated with less drought exposure and had lower mortality, especially in evergreen species.
+Hydraulic risk was balanced by investment in deep roots (Figs.
 \@ref(fig:swp-droughts), \@ref(fig:erd-mort-census-evg)).
-Deeper ERD species on average had more hydraulically efficient, but risky, suite of above-ground traits optimized for an environment of more reliable and greater water source at depth that resulted in a lower mortality risk.
-By contrast shallow ERD species had inefficient, but safer, xylem likely optimized for high variability in $\mathrm{\Psi}_\textrm{soil}$ in the shallow soil layers, and resulting in higher variance in mortality rates across censuses as $\mathrm{\Psi}_{\textrm{soil},z\textrm{ = ERD}}$ crossed ERD $\mathrm{\Psi}_\textrm{crit}$ (Figs.
-\@ref(fig:erd-mort-census-evg), \@ref(fig:erd-time)).
-We found that a greater proportion of deeper ERD species had evergreen phenology increased with deeper ERD.
-The trend of mortality under extreme soil drought in evergreen but not deciduous species is consistent with the expectation that deciduous species avoid exposure to drought-stress via leaf-drop.
 
-Our results are consistent with recent studies that analyses rooting or water-sourcing depths versus hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016; also see @Brum.2019; @Brum.2017; @Rowland:2015].
+Our results are consistent with recent studies that analysed rooting or water-sourcing depths versus hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016; also see @Brum.2019; @Brum.2017; @Rowland:2015].
 Globally, large trees tend to exhibit greater growth reductions, lower post-drought resilience, and greater increases in mortality relative to their understory counterparts [@Bennett:2015; @Phillips:2010].
-Our finding that deep-water access buffers drought-induced mortality in large trees is relevant for understanding drought resistance, resilience and recovery of this important functional group [@Bennett:2015; @McGregor.2020].
-Future studies should test the ERD-mortality relationship on a greater number of species in the community.
+Our finding that deep-water access buffers drought-induced mortality in large trees is relevant for understanding drought resistance, resilience and recovery [@Bennett:2015; @McGregor.2020].
+Future studies should test the ERD-mortality relationship on a greater number of species.
 
@@
 As in CT2018, we also interpret our ERD estimates as revealing relative differences among species' effective rooting depths rather than absolute depths.
-Besides, our ERD model predicts only daily maximum diurnal $\mathrm{\Psi}$~leaf~ assuming it is equivalent to $\mathrm{\Psi}$~soil~.
+Our ERD model predicts only daily maximum diurnal $\mathrm{\Psi}$~leaf~ assuming it is equivalent to $\mathrm{\Psi}$~soil~.
 While this scale may be adequate to relate to 5-year average growth values, this approach ignores diurnal dynamics of $\mathrm{\Psi}$~leaf~ and other factors that would play a role in such diurnal dynamics, such as the buffering effect of stem water storage capacitance on $\mathrm{\Psi}$~leaf~.
@@
 
-While ERD models have the potential to estimate ERDs for entire tree communities, during the their development phase, ERD models may need to be validated against data for a subset of representative species of the community, as this study does.
+While ERD models have the potential to estimate ERDs for entire tree communities, during their development phase, ERD models may need to be validated against data for a subset of representative species of the community, as this study did.
 ERD models should be tested across varied climates and forest types, covering contrasting plant strategies and possibly seasonality in ERDs.
@@
 
-The relationships that we identify between above- and below-ground traits, the water environment vertically below-ground and mortality rates are important for representing diversity in DGVMs, which intrinsically rely on the parameterization of contrasting life history strategies [@Scheiter:2013] and the simulation of competition between those strategies.
+The relationships that we identify between above- and below-ground traits, vertical profile of soil water status and mortality rates are important for representing diversity in DGVMs, which intrinsically rely on the parameterization of contrasting life history strategies [@Scheiter:2013] and the simulation of competition between those strategies.
 ERDs could be assimilated into DGVMs by setting rooting parameters that result into the centroid of root water-uptake profiles equivalent to ERDs.
@@
 \
-Figure \@ref(fig:erd-time): Modeled effective rooting depth (ERD; horizontal-axis) versus time spent beyond critical hydraulic threshold (vertical axis) by census interval (colored bars; note the squared y-axis scale).
+Figure \@ref(fig:erd-time-evg): Modeled effective rooting depth (ERD; horizontal-axis) versus time spent beyond critical hydraulic threshold (vertical axis) by census interval (colored bars; note the squared y-axis scale).
 Each bar represents the average time species of the same ERD spent beyond species-specific critical hydraulic thresholds in a given interval, that is, the proportion of days for which $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ was more negative than species $\mathrm{\Psi}$~crit~, defined as $\mathrm{\Psi}$~20,leaf~, and where $z$ is the soil depth matching species ERD.
 Standard error of the mean are shown over each bar.
-Data includes all the `r length(erd.mrate.sp)` canopy tree species that were included in the mortality analyses.
+Figure includes all the `r length(erd.mrate.sp)` canopy tree species, evergreen as wells as deciduous, that were included in the mortality analyses.
 Trends were similar for evergreen species alone.
@@
 
-```{r erd-mort-census-evg, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-mort}")}
+```{r erd-mort-census-evg, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-mort-evg}")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_by rdi.gr_evergreen.jpeg"))
@@
 
-```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time}")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mfac vs. rdi.gr.jpeg"))
+```{r erd-time-evg, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time-evg}")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mfac vs. rdi.gr_evergreen.jpeg"))
 ```

### 2021-04-05 10.0_Supporting_Information.Rmd
included co-authors' feedback

@@
 options(tinytex.verbose = TRUE)
+if (!require("groundhog")) install.packages("groundhog");
+groundhog.folder <- paste0("Users/ftuser/Library/R/4.0/library/bci.roots.inversion")
+if(!dir.exists(file.path(groundhog.folder))) {dir.create(file.path(groundhog.folder))}
+set.groundhog.folder(groundhog.folder)
+library(groundhog)
+# set.groundhog.folder('/Users/ftuser/Library/R/3.6/library')
+groundhog.day = "2021-01-01"
+groundhog.library('devtools', groundhog.day)
+pkgs=c('lattice', 'magick', 'cowplot', 'corrplot', 'ggcorrplot', 'tidyverse', 'readxl',
+       'forcats', 'scales', 'data.table', 'ggpmisc', 'GGally', 'bookdown', 'rmarkdown', 'broom')
+groundhog.library(pkgs, groundhog.day)
+
 source("code/06.0_load.R")
@@
 ```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($\\Psi_{\\textrm{soil},z}$) by depth $z$ (m; panels) over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines).
-Vertical gray lines demarcate median dates of the demographic censuses at the BCI 50 ha plot.
-Years 1991 & 1992 for depths > 2.9 m showed marks of model initialisation phase for ELM-FATES, which was run from ($\\Psi_{\\textrm{soil},z}$) = 0 for all depths, starting from 1985-01-01.
-We therefore removed $\\Psi_{\\textrm{soil}}$ for years 1991 & 1992 for depths greater than 2.9 m, and conservatively removed $\\Psi_{\\textrm{soil}}$ for year 1991 for the depth of 2.9 m."}
+Vertical gray lines demarcate median dates of the demographic censuses at the BCI 50 ha plot."}
 knitr::include_graphics(file.path("figures/2019-10-14_5000/best-fits/psi_model_daily_all_depths_params.top.few_full_beg_1991.jpeg"))
@@
 
-```{r erd-mort-census, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-census}Species (circles) Mortality rates in species of various deciduous leaf-habit categories (colors) across census intervals from 1985-2015 plotted against ERD.
+```{r erd-mort-deci, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-cdeci}Species (circles) Mortality rates in species of various deciduous leaf-habit categories (colors) across census intervals from 1985-2015 plotted against ERD.
 $R$^2^ and significance levels for linear model fits are given in panel insets.
 Species are color colded by deciduous category.
-Census-intervals with a superscript * are shown for comparison and indicate intervals in which significant mortality was explained by ERD among evergreen species (Fig.
+Census-intervals with a superscript * are shown for comparison and indicate intervals in which significant mortality was explained by ERD among evergreen species (see Fig.
 8 in the Main text)."}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_by rdi.gr_deciduous.jpeg"))
+```
+
+```{r erd-time-deci, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time-deci}")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mfac vs. rdi.gr_deciduous.jpeg"))
 ```
@@
 knitr::kable(hyd.table, caption = paste0("\\label{lab-hyd-table} Minimum leaf water potential $\\mathrm{\\Psi}$~min~, leaf turgor loss point $\\mathrm{\\Psi}$~tlp~, stem maximum hydraulic conductivity $K_\\textrm{max,stem}$ and stem vulnerability to cavitation $\\mathrm{\\Psi}$~88,stem~ for seven species with ERD estimates.
-$\\mathrm{\\Psi}$~min~ is sourced from @Wolfe.2019, while $\\mathrm{\\Psi}$~tlp~, $K_\\textrm{max,stem}$ and $\\mathrm{\\Psi}$~88,stem~ are sourced from @Wolfe.2020.
+$\\mathrm{\\Psi}$~min~ is sourced from @Wolfe.2019, while $\\mathrm{\\Psi}$~tlp~, $K_\\textrm{max,stem}$ and $\\mathrm{\\Psi}$~88,stem~ are sourced from @Wolfe.2021.
 $\\mathrm{\\Psi}$~min~, $\\mathrm{\\Psi}$~tlp~ and $\\mathrm{\\Psi}$~88,stem~ are in MPa, while $K_\\textrm{max,stem}$ in kg m^-1^ s^-1^ MPa^-1^."), booktabs = TRUE, escape = FALSE)
@@
 
+## Leaf habit
+
+AT BCI expert botanists have identified four leaf habits: evergreen species ($n$ = `r nrow(subset(erd.data, deciduous != "E"))`, for species with ERD) keep most but not all of their canopy in the dry season; obligate deciduous species ($n$ = `r nrow(subset(erd.data, deciduous != "DO"))`) shed all their leaves in the beginning of the dry season; facultative deciduous ($n$ = `r nrow(subset(erd.data, deciduous != "DF"))`) species drop leaves steadily as the dry season develops, and both facultative and obligate deciduous species leaf out in the beginning of the wet season; brevi-deciduous species ($n$ = `r nrow(subset(erd.data, deciduous != "DB"))`) shed all their leaves in the dry season, but leaf out in a few days to six weeks under ongoing dry season.
+
 ### Leaf Area Index 
@@
 We generated an estimate of the mean seasonality curve for normalized leaf area index ($\textrm{LAI}^*$, range 0-1, unitless) for individual species as follows.
-We interpolated species-specific weekly leaf-fall data collected at two similar sites in BCI (1985-2020) to a daily estimate [@Wright.1990aeo].
-We lead this daily leaf-fall time series by the species-specific average leaf lifespan (in days) and used it as an estimate of the daily leaf-gain time-series.
-The difference between leaf-fall and leaf-gain gave an estimate of daily leaf-change time series, which was then scaled to express as a fraction of the total leaf fall in a given year.
-Beginning with the $\textrm{LAI}^*$ of 1 at the end of the wet season in the first year of the record, we accumulated leaf-change over it to obtain a running estimate of the leaf-cover fraction, averaged it by DOY across years and normalized it to obtain the species-specific seasonality curve in $\textrm{LAI}^*$ (range 0-1, unitless).
+We interpolated species-specific weekly leaf-fall data collected at two similar sites in BCI (1985-2020) to a daily estimate of leaf mass fall per unit ground area [@Wright.1990aeo].
+We estimated living canopy leaf mass per unit ground area present at time t (L[t]) from mean leaf longevity in days (LL) and leaf mass fall per unit ground area at time t (F[t]) as follows:  L[t] = F[t+1] + F[t+2] + … + F[t + LL].
+L[t] was averaged by DOY across years and divided by the max value to obtain the species-specific seasonality curve in $\textrm{LAI}^*$ (range 0-1, unitless).
 (Fig.
 S\@ref(fig:leaf-cover)).
+\
+For `r sp.LAI.obs` species, an estimate of leaf lifespan was available from direct observations at a dry and wet site in Panama [@Osnas.2018].
+For the rest it was gap-filled with genus-level (n = `r sp.LAI.gen`) or family-level means (`r sp.LAI.fam`), if available.
+For `r sp.LAI.pnm.site` deciduous species, dry site-level mean was used.
+For one species, leaf lifespan was predicted from LMA ($\overline{lifespan} = `r ll.lma.cf[1]` + `r ll.lma.cf[2]`LMA$; $R$^2^ = `r ll.lma.r2`; $p$ `r ll.lma.p`, $n$ = `r ll.lma.n`).
 
-\
-For five species, an estimate of leaf lifespan was available from direct observations [@Osnas.2018].
-For the rest `r erd.sp.wo.ll` species, leaf lifespan was predicted from LMA ($\overline{lifespan} = `r ll.lma.cf[1]` + `r ll.lma.cf[2]`LMA$; $R$^2^ = `r ll.lma.r2`; $p$ `r ll.lma.p`, $n$ = `r ll.lma.n`).
+Although deciduous leaf-habits are generally expected to be fully deciduous, at BCI there is a great within-species variation in the timing of leaf-drop and leaf-out and neither do all individuals of a deciduous species may sreach full deciduousness every year [@Condit.2000].
+Our reconstruction of $\textrm{LAI}^*$ seasonality is based upon species leaf-fall data, which captures the within-species variation in the timing of leaf-drop, so in our reconstruction the seasonal decline in $\textrm{LAI}^*$ appears to be more spread out and does not reach zero, in contrast to what is expected for an idealised individual of each deciduous leaf habit.
+Greater uncertainty in leaf-lifespan for deciduous species may have compounded this issue further.
+We did not normalise the deciduous species' $\textrm{LAI}^*$ curves to reach zero as it may further distort the shape of the $\textrm{LAI}^*$ curve shifting a larger period under a lower $\textrm{LAI}^*$.
 
 ```{r leaf-cover, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-cover}Seasonal variation in standardised LAI (range 0-1) for ", length(erd.sp)," species at Barro Colorado Island (panels).
-Species are color-coded by deciduous category as defined by expert botanists.")}
+Species are color-coded by leaf habit as defined by expert botanists.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf.cover_BCI_multi_panel.jpeg"))
@@
 \
-Values of stem maximum hydraulic conductivity and vulnerability to cavitation were obtained from the dataset of @Wolfe.2020.
+Values of stem maximum hydraulic conductivity and vulnerability to cavitation were obtained from the dataset of @Wolfe.2021.
 The entire dataset includes measurements on 26 tree species collected at three sites in Panama.
-The raw data, *R* script for the analyses, and data summaries are archived in the NGEE-Tropics archive [@Wolfe.2020].
+The raw data, *R* script for the analyses, and data summaries are archived in the NGEE-Tropics archive [@Wolfe.2021].
 The data were previously reported by @Dickman.2019 and @Wu.2020.
@@
 G_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},s,i|z,h}^{*} \hat{\textrm{VPD}}_{i}^{*}) + \epsilon_{s,t|z,h}
-(\#eq:M-3)
+(\#eq:M3)
 \end{equation}
@@
 
+The best ERD model explained a large fraction of the variance in $\mathrm{\delta}$^2^H~xylem~.
+This model included an effect of VPD and not LAI (Eq.
+\@ref(eq:M3)).
+Two of the alternative models (Eq.
+\@ref(eq:M4) & \@ref(eq:M6)) with LAI had a strong relationship with $\mathrm{\delta}$^2^H~xylem~ as well, but one that was based on a smaller set of species ($n$ = 4 or 5), because fewer species passed the criteria of sufficient species-level goodness-of-fit for these models (<17 species vs.
+`r length(erd.sp)` species for the best model, Fig.
+S9).
+For the rest of the analyses we therefore used ERD estimates from the less parsimonious model (Eq.
+\@ref(eq:M-3)).
+
 # References

### 2021-04-05 LAI.Rmd
typo

@@
     where sp.lifespan are the number of days in species leaf lifespan  
-    To normalise LAI, it is the divided by the maximum value.
+    To normalise LAI, it is then divided by the maximum value.
 

### 2021-04-05 Readme.Rmd
minor

@@
 groundhog.day = "2021-01-01"
-groundhog.library('rmarkdown', groundhog.day)
+# groundhog.library('rmarkdown', groundhog.day)
+# .rs.restartR() to restart R session
 groundhog.library('bookdown', groundhog.day)
-rmarkdown::render("09.0_manuscript.rmd", output_format = "bookdown::html_document2")
+groundhog.library('knitr', groundhog.day)
+# rmarkdown::render("09.0_manuscript.rmd", output_format = "bookdown::html_document2")
+render("09.0_manuscript.rmd", output_format = "bookdown::word_document2")
 ```
@@
 ```{r Chunk 10.0, eval = FALSE}
-render("10.0_Supporting_Information.Rmd", output_format = "bookdown::html_document2")
+# render("10.0_Supporting_Information.Rmd", output_format = "bookdown::html_document2")
+render("10.0_Supporting_Information.Rmd", output_format = "bookdown::word_document2")
 ```

### 2021-04-06 Readme.Rmd
rmarkdown, bookdown via groundhog

@@
 # For figures and data, code/08.0_manuscript_data_figs.R
-if (!require("groundhog")) install.packages("groundhog"); 
+if (!require("groundhog")) install.packages("groundhog")
+library(groundhog)
 groundhog.folder <- paste0("Users/ftuser/Library/R/4.0/library/bci.roots.inversion")
@@
 set.groundhog.folder(groundhog.folder)
-library(groundhog)
+
 groundhog.day = "2021-01-01"
-# groundhog.library('rmarkdown', groundhog.day)
+groundhog.library('rmarkdown', groundhog.day)
 # .rs.restartR() to restart R session
 groundhog.library('bookdown', groundhog.day)
-groundhog.library('knitr', groundhog.day)
+# groundhog.library('knitr', groundhog.day)
 # rmarkdown::render("09.0_manuscript.rmd", output_format = "bookdown::html_document2")
-render("09.0_manuscript.rmd", output_format = "bookdown::word_document2")
+rmarkdown::render("09.0_manuscript.rmd", output_format = "bookdown::word_document2")
 ```

### 2021-04-08 09.0_manuscript.rmd
1. More changes with co-authors feedback 2. resistance curve usage changed back to vulnerability

@@
 Summary | 200  
-Main Text  | 8024  
-Introduction | 833  
-Methods | 4132  
-Results | 924  
-Discussion | 2135  
+Main Text  | 6971  
+Introduction | 832  
+Methods |  3211  
+Results | 855  
+Discussion | 2073  
 Keywords | 8
@@
 
-- Deep-water access is arguably the most long-standing, but under-studied mechanisms that plants employ to survive during drought.
+- Deep-water access is arguably the most effective, but under-studied mechanism that plants employ to survive during drought.
 Vulnerability to embolism and hydraulic safety margins can predict mortality risk at given levels of dehydration, but deep-water access may delay plant dehydration.
 Here, we tested the role of deep-water access in enabling survival within a diverse tropical forest community in Panama using a novel data-model approach.
-- We inversely estimated the effective rooting depth (ERD, as the average depth of water extraction), for 32 canopy species by linking diameter growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and leaf hydraulic resistance curves.
-We validated ERD estimates against existing isotopic data of potential water access depths.
-- Across species, deeper ERD was associated with higher maximum stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates especially in evergreen leaf-habit during extreme droughts over 35 years (1981-2015).
-Species with deeper ERD did not cross critical hydraulic thresholds indicating that trees compensate for water-stress related mortality risk through deep-water access.
-- The role of deep-water access in mitigating mortality of hydraulically-vulnerable trees has important implications for our predictive understanding of forest dynamics under current and future climates.
+- We inversely estimated the effective rooting depth (ERD, as the average depth of water extraction), for 32 canopy species by linking diameter growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and leaf hydraulic vulnerability curves.
+We validated ERD estimates against existing isotopic data of potential water-access depths.
+- Across species, deeper ERD was associated with higher maximum stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates during extreme droughts over 35 years (1981-2015), especially in evergreen species.
+Species exposure to water-stress declined with deeper ERD indicating that trees compensate for water-stress related mortality risk through deep-water access.
+- The role of deep-water access in mitigating mortality of hydraulically vulnerable trees has important implications for our predictive understanding of forest dynamics under current and future climates.
 \
@@
 Tropical forests play a disproportionately large role in the global carbon and energy cycles [@Bonan:2008], and support half of global biodiversity [@Wright:20050th], but face a threat from intensifying droughts [@Malhi.2008; @Doughty:2015; @Xu.2019].
-Tropical forests have been expected to be an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw].
+Tropical forests are considered an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw].
 Still, mortality events in the tropics are rarely as large as those in temperate and boreal zones [@McDowell.2018], leading to questions regarding the role of large trait-diversity and hydraulic strategies in mitigating mortality events.
@@
 \
-Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation, stem water storage, deep-water access, increased root production, hydraulic redistribution and embolism resistance [@McDowell.2008].
-Deep-water access is arguably the most long-standing [@Cannon.1911], yet under-studied mechanism.
-As plant-available water varies with depth, trees in the same forest with different rooting depths, depending on species and size [@Meinzer:1999; @Chitra-Tarak.2018; @Brum.2019], differ in their experience during a drought, and thus in their growth and mortality responses [@Chitra-Tarak.2018].
+Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from deep-water access, increased root production, hydraulic redistribution, embolism resistance, adjustment of leaf area (deciduousness) to change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation and stem water storage capacitance  [@McDowell.2008].
+Deep-water access is arguably the most effective, yet under-studied mechanism.
+As plant-available water varies with depth, trees within the same forest with different rooting depths, depending on species and size [@Meinzer:1999; @Chitra-Tarak.2018; @Brum.2019], differ in their experience during a drought, and thus in their growth and mortality responses [@Chitra-Tarak.2018].
 A key bottleneck in community-wide testing of this mechanism has been a lack of data in both trees' rooting or water-sourcing depths and plant-available soil water at those depths.
@@
 Such efforts have found correlations between vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013; @Venturas.2016].
-Deep roots may mitigate hydraulic vulnerability [@Brum.2019] and mortality-risk, in particular during hydrological (rather than meteorological) droughts.
+Deep roots may mitigate hydraulic vulnerability [@Brum.2019] and mortality risk, in particular during hydrological (rather than meteorological) droughts.
 Nonetheless, the interaction between rooting depths, above-ground hydraulic traits, hydrological droughts quantified over the whole soil-column and mortality outcomes are hardly studied.
@@
 
-This work combines hydrology, physiology and demography of a tropical forest at Barro Colorado Island, Panama, to inform the inverse model for rooting depths; validates the model, and tests hypotheses pertaining to relationship of rooting depths with above-ground hydraulic traits, drought exposure, and mortality (see Fig.
-\@ref(fig:schematic) for a workflow).
+This work combines hydrology, physiology and demography of a tropical forest at Barro Colorado Island, Panama, to inform the inverse model for rooting depths; validates the model, and tests hypotheses pertaining to relationship of rooting depths with above-ground hydraulic traits, drought exposure, and mortality (Fig.
+\@ref(fig:schematic)).
 We defined species-specific effective rooting depths (ERD) as the depth at which the growth factor determined by soil water potential and leaf hydraulic traits best explained species' growth dynamics over 25 years.
-We developed a novel, empirical inverse model of ERD and estimated ERDs for large trees of `r length(erd.sp)` species.
-The ERD model incorporated the impact of atmospheric and hydrological drought on growth, and was constrained with species-specific leaf resistance curves.
+Developing a novel, empirical inverse model, we estimated ERDs for large trees of `r length(erd.sp)` species.
+The ERD model incorporated the impact of atmospheric and hydrological drought on growth, and was constrained with species-specific leaf vulnerability curves.
 For the latter, we used existing data for eight species and developed trait-based proxies for the rest (based on data for a total of 21 species).
 We obtained the daily dynamics of soil water potential in the whole-soil column (up to 13 m) over the 25 years by locally parameterizing a 1-D hydrological water balance of the forest within a land surface model for an average vegetation type.
-The water balance was calibrated on available measurements, in particular continuous soil moisture data in three surface layers (over the first meter of soil), stream discharge and ET fluxes.
+The water balance was calibrated on available measurements, in particular continuous soil moisture data in three surface layers (over the first meter of soil), stream discharge and evapo-transpiration.
 
+\
 We validated our ERD estimates against existing stable hydrogen isotope ratios ($\mathrm{\delta}$^2^H~xylem~) for a subset of six tree species as independent observations.
@@
 We tested the relationship of ERD with xylem vulnerability to embolism, branch hydraulic safety margins, leaf turgor loss point and maximum stem-specific hydraulic conductivity.
-We analysed whether ERD is correlated with mortality dynamics of large trees over 35 years marked by several El Niño events for evergreen and deciduous species.
-Finally, to test whether ERD explained risk of drought-induced mortality, we analysed species-level exposure to water-stress as the time a species’ spent beyond its critical hydraulic threshold given its ERD.
+We analysed whether ERD is correlated with mortality dynamics of large trees for evergreen and deciduous species over 35 years marked by several El Niño events.
+Finally, to test whether ERD explained risk of drought-induced mortality, we analysed species-level exposure to water-stress.
 
@@
 Because our study is situated in an old-growth tropical forest, we assume that nutrients are mainly concentrated in the shallow soil layers derived from litter decomposition, and that rooting depths do not substantially affect nutrient uptake.
-Because our study site has a dry season, we assume that rooting depth is a key factor affecting tree growth.
+Because our study site has a dry season, we expect that rooting depth is a key factor affecting tree growth.
 We defined species-specific effective rooting depth (ERD) as the depth at which a soil moisture growth limitation factor, $B$, determined by the soil water potential and hydraulic traits best explained species’ diameter growth dynamics.
 $B$ is used to approximate the amount of stomatal closure due to water-stress in the soil.
-In this study, we used the loss of conductivity of leaf ($FLC$~leaf~) to estimate $B$.
-$FLC$~leaf~ is calculated as a fraction of maximum leaf conductivity $K_\textrm{max, leaf}$:
+In this study, we used the fractional loss of hydraulic conductivity of the leaf ($\textrm{FLC}$~leaf~) to estimate $B$.
+$\textrm{FLC}$~leaf~ is calculated as a fraction of maximum leaf conductivity $K_\textrm{max, leaf}$:
 
 \begin{equation}
-FLC_{\textrm{leaf}} = K_{\textrm{leaf}}/K_{\textrm{max, leaf}}
+\textrm{FLC}_{\textrm{leaf}} = K_{\textrm{leaf}}/K_{\textrm{max, leaf}}
+(\#eq:flc)
+\end{equation}
+
+where $K$~leaf~ is the current leaf hydraulic conductance and $K_\textrm{max, leaf}$ is the maximum leaf hydraulic conductance (see Table \@ref(tab:symbols) for trait symbols and their definitions).
+The dynamics of $K$~leaf~ are estimated from the species-specific relationship between $K$~leaf~ vs.
+$\mathrm{\Psi}$~leaf~, referred to as the leaf hydraulic vulnerability curve [@Sack.2012].
+The vulnerability to loss of hydraulic conductivity arises not only from embolism but also from extra-xylem processes [@Scoffoni:2017].
+Because pre-dawn $\mathrm{\Psi}$~leaf~ generally approaches $\mathrm{\Psi}$~soil~, this allowed us to substitute $\mathrm{\Psi}$~leaf~ with $\mathrm{\Psi}$~soil~ in leaf hydraulic vulnerability curves for each species $s$---defined below in Eq.
+\@ref(eq:vc-exp) using species-specific parameters $A_{s}$ and $B_{s}$---to predict maximum diurnal $K_{\textrm{leaf},s, i}$ for each day $i$ as:
+
+\begin{equation}
+K_{\textrm{leaf},s, i} = A_{s}e^{-B_{s}{\mathrm{\Psi}}_{\textrm{soil, i}}}
 (\#eq:k-soil)
 \end{equation}
-
-where $K$~leaf~ is the current leaf hydraulic conductance and $K_\textrm{max, leaf}$ is the maximum leaf hydraulic conductance (see Table \@ref(tab:symbols) for trait symbols and their definitions)).
-The dynamics of $K$~leaf~, and thus $FLC$~leaf~, are estimated from the species-specific relationship between $K$~leaf~ vs.
-$\mathrm{\Psi}$~leaf~, referred to as the leaf hydraulic resistance curve [@Sack.2012].
-Because pre-dawn $\mathrm{\Psi}$~leaf~ generally approaches $\mathrm{\Psi}$~soil~, this allowed us to use $\mathrm{\Psi}$~soil~ in leaf hydraulic resistance curves to predict maximum diurnal $K$~leaf~.
+$K_{\textrm{max, leaf}, s}$, was obtained as the maximum value of $K_{\textrm{leaf}, s}$ using Eq.
+\@ref(eq:k-soil).
 $K$~leaf~ is strongly related to photosynthetic capacity [@Brodribb.2002].
 Because we were interested in relating $K$~leaf~ to 5-year average diameter growth observations, we ignored diurnal dynamics of $K$~leaf~.
-$K_\textrm{leaf}$ in our models, being based on resistance curves, is thus dynamic and depends on rooting depth and rhizosphere water potential.
-We considered $K_\textrm{leaf}$ to be a hydraulic bottleneck in the plant and that it determines or co-determines whole plant hydraulic response to soil drying; notably, root hydraulic resistance may also be a strong driver, but recent studies suggest that its resistance would be similar to, or correlated with that of leaves (see @Cochard.2020 and references therein).
 
+\
 Many other intrinsic and extrinsic factors limit plant growth, including soil moisture, VPD, radiation, leaf area seasonality [@Lawrence.2019; @Brodribb.2002].
@@
 To account for the non-linear impact of VPD on growth, we used predicted gross primary productivity (GPP; hereafter, $\hat{\textrm{VPD}}$) from a locally derived polynomial relationship between GPP and VPD (see SI and Fig.
-S2).
+S1).
 Apart from stomatal control, leaf deciduousness may further limit water-uptake and growth.
 
-We therefore tested alternate empirical growth models, or inverse models of ERD, in which we regressed species-specific growth against multiplicative or additive effects of one or more growth factors calculated daily and averaged over 5-yearly censuses: $\hat{\textrm{VPD}}$, $FLC_\textrm{leaf}$ and the leaf area index (LAI), all standardized to range between 0 and 1 (Eq.
+\
+We therefore tested alternate structures of empirical growth models, or inverse models of ERD, in which we regressed species-specific growth against multiplicative or additive effects of one or more growth factors calculated daily and averaged over 5-yearly censuses: $\hat{\textrm{VPD}}$, $textrm{FLC}_\textrm{leaf}$ and the leaf area index (LAI) (Eq.
 \@ref(eq:M-chosen) and those in SI).
@@
 \
- The best empirical model we found to describe daily average tree growth (see model validation and selection below), daily average growth for species $s$, $\hat{G}$, in the census interval $t$ for depth $z$ and for a given hydrological realization $h$ is described as follows:
+The best empirical growth model, or inverse ERD model, structure that we found (see model validation and selection below) to describe daily average tree growth $\hat{G}$ for species $s$ in the census interval $t$ is described as follows, in which $K_\textrm{leaf}$, and thus $FLC_\textrm{leaf}$, is driven by soil water dynamics at $z$:
 
 \begin{equation}
-\hat{G}_{s,t|z,h} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} FLC_{\textrm{leaf},s,i|z,h}^{*} \hat{\textrm{VPD}}_{i}^{*}) + \epsilon_{s,t|z,h}
+\hat{G}_{s,t|z} = \beta_{0,s|z} + \beta_{1,s|z}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} \textrm{FLC}_{\textrm{leaf},s,i|z}^{*} \hat{\textrm{VPD}}_{i}^{*}) + \epsilon_{s,t|z}
 (\#eq:M-chosen)
 \end{equation}
-where | refers to conditionals, n~t~ is the total number of days in census interval $t$, the superscript * indicates that the variable has been standardized to range between 0 and 1 (within species for $FLC_{\textrm{leaf}}$), $\beta_{0}$ and $\beta_{1}$ are model coefficients and $\epsilon$ is the model error term.
+where | refers to conditionals, n~t~ is the total number of days in census interval $t$, the superscript * indicates that the variable has been standardized to range between 0 and 1 (within species for $\textrm{FLC}$~leaf~), $\beta_{0}$ and $\beta_{1}$ are model coefficients and $\epsilon$ is the model error term.
 See SI for all the alternate models tested.
@@
 \
- For each species, we estimated diurnal maximum value of $K_{\textrm{leaf},s,i|z,h}$ for a species $s$ by substituting $\mathrm{\Psi}_{\textrm{leaf}}$ with $\mathrm{\Psi}_{\textrm{soil}}$ in the leaf hydraulic resistance curves defined below in Eq.
-\@ref(eq:vc-exp) using species-specific parameters $A_{s}$ and $B_{s}$ as:
+We evaluated different model structures (see SI) and for each we estimated species effective rooting depths (ERD) as the depth $z$ at which soil water dynamics ($\mathrm{\Psi}_{\textrm{soil,} z}$) best explained observed dynamics of growth via modeled growth (e.g.
+Eq.\@ref(eq:M-chosen)).
+Our growth model, or inverse ERD model, does not explicitly use rooting profiles, but identifies soil water dynamics at a single depth $z$ as the central tendency that influences the observed growth dynamics the most.
+We modelled multiple hydrological realisations of soil water-potential dynamics (see below).
+Incorporating this uncertainty, we defined species ERD as the median ($\pm$ SE) of best-fit depths across all hydrological realisations.
+See SI for statistics for identifying best-fit ERD.
 
-\begin{equation}
-K_{\textrm{leaf},s,i|z,h} = A_{s}e^{-B_{s}{\mathrm{\Psi}}_{\textrm{soil},i|z,h}}
-(\#eq:k-soil)
-\end{equation}
-$K_{\textrm{max, leaf},s}$, was obtained as the maximum value of $K_{\textrm{leaf},s}$ using Eq. \@ref(eq:k-soil).
-
-We estimated species effective rooting depths (ERD) as the depth $z$ at which soil water dynamics ($\mathrm{\Psi}_{\textrm{soil,} z}$) best explained observed dynamics of growth via modeled growth (Eq.\@ref(eq:M-chosen)).
-Our ERD model does not explicitly use rooting profiles, but finds soil water dynamics at a single depth $z$ as the central tendency that influences the observed growth dynamics the most.
-We modelled multiple hydrological realisations of soil water-potential dynamics with a water-balance model (see below), and thus defined ERD as the median of best-fit depths across all hydrological realisations.
-
-Formally, species-specific best-fit ERD was selected as follows: using Eq.
-\@ref(eq:M-chosen), for a species $s$ and hydrological realization *h*, we obtained Akaike's Information Criterion (AIC) and Goodness-of-fit ($R$^2^) for each modeled soil layer depth $z$.
-We used AIC as the first selection criterion to ensure that the fitted growth values from Eq.
-\@ref(eq:M-chosen) are within the confidence interval of the error distribution of observed growth (as the same model Eq.
-\@ref(eq:M-chosen) was being compared for each *h*, number of model parameters were the same).
-For each *h*, we retained those depths with the maximum likelihood of growth (similar AIC based support, see below) filtering out the rest.
-However, AIC does not ensure that fitted values will have a positive trend with growth.
-We therefore used $R$^2^ between observed and fitted growth values as a goodness-of-fit measure and retained depths associated with a relatively high threshold $R$^2^ value of 0.5; after filtering out depths associated with negative correlations $r$ if any.
-Finally, among these we chose the depth with the highest $R$^2^ per *h*.
-Results were not sensitive to inclusion of this final step of retaining only a single depth as the best-fit (not shown).
-For a species $s$, we defined effective rooting depth (ERD) as the median of the best-fit depth over all *h*, which can be described as:
-
-\begin{equation}
-\textrm{ERD}_{s} = \underset{\forall \, h \, \in \, H}{\textrm{med}} \Big(z\,\Big|\, \Big(R^2_{s|z,h} = \underset{\forall \, z \,\in\,Z}{\textrm{max}}(R^2_{s|z,h}| \mathrm{\Delta}_{s|z,h} \le 2\Big)\Big) \, \bigcap \, R^2_{s|z,h} \ge 0.5 \, \bigcap \,  r_{s|z,h} \ge 0\Big)
-(\#eq:erd)
-\end{equation}
-
-where
-
-\begin{equation}
-\mathrm{\Delta}_{s|z,h} = \textrm{AIC}_{s|z,h} - \underset{\forall \, z \,\in\,Z}{\textrm{min}}(\textrm{AIC}_{s|z,h})
-(\#eq:delta)
-\end{equation}
-
-and
-
-\begin{equation}
-\textrm{AIC}_{s|z,h} = -2\textrm{log}(\mathcal{L}(\hat{\beta}_{s|z,h}|{G_{s}})) - 2\textrm{k}
-(\#eq:aic)
-\end{equation}
-
-where r is the correlation, and $R^2$ the Goodness-of-fit, between observed growth $G$ and estimated growth $\hat{G}.
-$\mathcal{L}$ is the likelihood function, $\hat{\beta}$ is the maximum-likelihood estimate, $G$ the observed growth with $k$ observations (or, census intervals $t$).
-$H$ = (1, .., 100), the 100 best-fit hydrological realizations and $Z$ = (0.4, `r paste(soil.depths[8:length(soil.depths)], collapse=", ")`) m.
-We removed finely resolved soil layers <0.21 m and averaged $\mathrm{\Psi}_{\textrm{soil},z}$ for depths 0.21, 0.37 and 0.62 (=0.4) m as sufficient resolution for model validation.
-
-#### Growth data
+### Growth data
 
@@
 
-#### Leaf hydraulic resistance curves
+### Leaf hydraulic vulnerability curves
 
-We obtained leaf hydraulic resistance curves ($K_\textrm{leaf}$ vs.
+We obtained leaf hydraulic vulnerability curves ($K_\textrm{leaf}$ vs.
 $\mathrm{\Psi}$~leaf~) for adult trees of 21 common species at BCI from Zailaa *et* *al.* (in prep; see SI for brief description of methods) described as:
@@
 These 21 species included eight of the `r length(erd.sp)` species selected for ERD estimation.
-For the remaining `r sp.n.model` species, we obtained resistance curves using trait-based proxies.
+For the remaining `r sp.n.model` species, we obtained vulnerability curves using trait-based proxies.
 We identified scaling relationships between fitted parameters $A$ and $B$ in Eq.
-\@ref(eq:vc-exp) and two traits; namely, WSG, the wood specific gravity, and LMA, the leaf mass per unit area (@Wright:2010; see SI for details).
+\@ref(eq:vc-exp) and two traits; namely, WSG, the wood specific gravity, and LMA, the leaf mass per unit area (@Wright:2010; see SI).
 We fitted polynomial equations described as,
@@
 As these fits explained a large proportion of variation in parameters $A$ and $B$ (see results), we sequentially used Eq.
-\@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r sp.n.model` ERD species without direct data and estimated leaf hydraulic resistance curves using Eq.
+\@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r sp.n.model` ERD species without direct data and estimated leaf hydraulic vulnerability curves using Eq.
 \@ref(eq:vc-exp).
@@
 \@ref(eq:b-mod).
-We also estimated species $\mathrm{\Psi}$~20,leaf~ using their resistance curves.
+We also estimated species $\mathrm{\Psi}$~20,leaf~ using their vulnerability curves.
 
-#### Leaf Area Index (LAI)
+### Leaf Area Index (LAI)
 
@@
 
-#### Using ELM-FATES to model soil matric potentials
+### Using ELM-FATES to model soil matric potentials
 
-##### ELM-FATES model description
+#### ELM-FATES model description
 
@@
 \
-  We ran ELM with FATES vegetation, in which the ELM model simulates interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\mathrm{\Delta}W_{\textrm{can}}$, surface water $\mathrm{\Delta}W_{\textrm{sfc}}$, soil water $\mathrm{\Delta}W_{\textrm{liq},z_{i}}$ at depth $z_{i}$ and water in an unconfined aquifer $\mathrm{\Delta} W_{a}$ (omitting processes relevant to snow, wetlands or lakes).
-Conservation of these terms is expressed as follows:
-
-\begin{equation}
-\begin{split}
-\mathrm{\Delta}W_\textrm{can} + \mathrm{\Delta}W_\textrm{sfc} + \sum_{i=1}^{15} \mathrm{\Delta}W_{\textrm{liq},z_{i}} + \mathrm{\Delta}W_{a} = \left(q_\textrm{rain} - E_{v} - E_{g} - q_\textrm{over} - q_\textrm{h2osfc} - q_\textrm{drai}\right)\mathrm{\Delta}t
-\end{split}
-(\#eq:water-balance)
-\end{equation}
-where, $q_{\textrm{rain}}$ is rainfall, $E_{v}$ is transpiration of the forest, $E_{g}$ is ground evaporation, $q_{\textrm{over}}$ is surface runoff, $q_{\textrm{h2osfc}}$ is runoff from surface water storage, $q$~drai~ is subsurface drainage (all in mm H~2~O) and $\mathrm{\Delta}t$ is the time-step (here, 1 hr).
+  We ran ELM with FATES vegetation, in which the ELM model simulates interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water, surface water, soil water by depth and water in an unconfined aquifer (omitting processes relevant to snow, wetlands or lakes).
+See SI for a water balance equation and a note on how soil water dynamics is simulated in ELM-FATES.
 The soil profile is discretized into maximum 15 exponentially distributed soil layers with layer node depth $z$.
@@
 
-In ELM, soil water is predicted from a multi-layer model, in which the vertical soil moisture transport is governed by infiltration, surface and sub-surface runoff, gradient diffusion, gravity, canopy transpiration through root extraction, and interactions with groundwater [@Oleson.2013].
-FATES simulates size-structured tree community together with individual tree rooting profiles.
-ELM removes total community canopy transpiration from the soil layers in proportion to the total root fraction in each layer in the 1-D column.
+#### ELM-FATES model parameterization
 
-##### ELM-FATES model parameterization
-
-To parameterize catchment hydrology in ELM-FATES, we identified eleven parameters relevant for Eq.
-\@ref(eq:water-balance), determined their ranges based on literature for the study site, else for the tropics (Table S1), and ran 5000 simulations using Latin Hypercube Sampling [LHS; @Stein.1987] from this global parameter space.
-In the Conrad catchment, in which the 50 ha plot is situated, soil hydraulic conductivity decreases with depth from 12.5 and 60 cm (@Godsey:2004; Table S1).
-We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths <12.5 cm and >60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig.
-S1).
-For greater accuracy, instead of the ELM default pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content (GWC) vs.
-$\mathrm{\Psi}$~soil~ for depths of 0.15, 0.4 and 1 m [@Kupers:2019].
-We converted GWC to volumetric (VWC, $\theta$) using a measured bulk density value of 0.8 g cm^-3^ (see below).
-The @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences in water retention among depths were nominal,
-
-\begin{equation}
-\mathrm{\Psi}_{\textrm{soil}} = \mathrm{\Psi}_e\left(\frac{\mathrm{\theta}}{\mathrm{\theta}_{s}}\right)^{-b}
-(\#eq:soil-ret-curve)
-\end{equation}
-where $\mathrm{\Psi}_e$ is the air-entry matric potential, $\mathrm{\theta}_{s}$ is the saturated volumetric water content, and $b$, an index for soil pore-size distribution, respectively.
-$\mathrm{\Psi}_e$, $\mathrm{\theta}_{s}$ and $b$ that best-fit the data were 200 (mm H~2~O), 0.51 (cm^3^ cm^-3^) and 10 (unitless), respectively.
+To parameterize catchment hydrology in ELM-FATES, we identified eleven parameters relevant for the water balance, determined their ranges based on literature for the study site, else for the tropics (Table S1), and ran 5000 simulations using Latin Hypercube Sampling [LHS; @Stein.1987] from this global parameter space.
+Notably, we leveraged local data for soil hydraulic conductivity by depth (@Godsey:2004; Fig.
+S2), and instead of the ELM default soil texture-based pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content vs.
+$\mathrm{\Psi}$~soil~ [@Kupers:2019].
+See SI for details.
 
@@
 We ran ELM-FATES for the 5000-member 11 parameter ensemble with hourly climate drivers measured at the BCI meteorological station over 1985-2018 [@Faybishenko.2021] initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode.
-As our key interest here was on deriving soil water availability, we ran ELM-FATES in a lower-complexity configuration: static stand structure, and with a single plant functional type (PFT) of evergreen trees.
+As our key interest here was on deriving soil water availability, we ran ELM-FATES in a lower-complexity configuration: static stand structure (see SI), and with a single plant functional type (PFT) of evergreen trees.
 The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous [@Condit:2000] and addition of a dry-deciduous PFT did not significantly alter results (not shown).
-In static stand structure model, the model is started from an observed forest inventory and recruitment, growth and mortality are deactivated.
-Holding the stand structure static rather than allowing it to emerge from physiologically-mediated competition between trees removes a set of internal model feedbacks and thus the only differences between model simulations are driven by direct physiological effects of perturbed model parameters.
-This allows us to focus on the hydraulic and hydrological questions explored in this paper.
 
-Note that we use the soil water potential dynamics obtained from ELM-FATES as an ecosystem-level outcome to drive our species-level, empirical ERD model, and thus believe that there is no conflict/circularity between tree-level rooting profiles and water uptake processes in ELM-FATES and the ERD model.
+#### ELM-FATES model calibration
 
-##### ELM-FATES model calibration
-
-We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq.
-\@ref(eq:water-balance), namely, (1) evapotranspiration ET from the flux tower by the 50 ha plot ($\backsimeq$ $E_{v} + E_{g}$ in Eq.
-\@ref(eq:water-balance); 2012-2017; see [SI](#SIData.matteo))), (2) local stream discharge $q_\textrm{run}$ ($\backsimeq$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see @Paton.2019 and [SI](#SIData.steveP)) and (3) soil volumetric water content (VWC) $\mathrm{\theta}_{z}$ ($\backsimeq$ $W_{\textrm{liq,} z}$) from two sources: (i) a long-term (2012-2018) record of VWC $\mathrm{\theta}_\textrm{0-0.15 m}$ averaged across three vertical time domain reflectometry (TDR) probes over depth 0-15 cm from three locations near the flux tower, and (ii) plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 at depths of 0.15, 0.4 and 1 m (1299 samples covering all soil types and habitats; see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\mathrm{\theta}_{\textrm{0-0.15 m}}$, $\mathrm{\theta}_{\textrm{0.4 m}}$ and $\mathrm{\theta}_{\textrm{1 m}}$ refer to plot-wide averages).
+We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in the water balance equation, namely, (1) evapotranspiration ET from the flux tower by the 50 ha plot (2012-2017), (2) local stream discharge (2012-2018; see @Paton.2019) and (3) soil volumetric water content (VWC) from two sources: (i) a long-term (2012-2018) record of VWC averaged across three vertical time domain reflectometry (TDR) probes over depth 0-15 cm from three locations near the flux tower, and (ii) plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 at depths of 0.15, 0.4 and 1 m (1299 samples covering all soil types and habitats (@Kupers:2019).
 
 \
- For ELM-FATES calibration we calculated an objective function for each of the 5000-member ensemble by equally weighting standardized RMSE between observations and simulations across all fluxes and states.
-We describe the objective function $O$ as :
+ For ELM-FATES calibration we calculated an objective function for each of the 5000-member ensemble by equally weighting standardized RMSE between observations and simulations across all fluxes and states mentioned above, and then identified 100 parameter ensemble members that minimised the objective function, ensuring that soil moisture dynamics by depth is captured correctly.
+See SI for details and Table S1 for the ranges of best-fit values for different parameters.
 
-\begin{equation}
-\begin{split}
-\overline{O} = \frac{1}{6}\sum \textrm{RMSE}^{*}\textrm{(ET,}E_{v} + E_{g}) + \textrm{RMSE}^{*}(q_{\textrm{run}},q_{\textrm{over}} + q_{\textrm{drai}}) + \textrm{RMSE}^{*}(\mathrm{\theta}_{\textrm{0-0.15~m}}, {W_{\textrm{liq,}\overline{z}}}) \\
-& + \textrm{RMSE}^{*}(\mathrm{\theta}_{\textrm{0.15~m}}, W_{\textrm{liq,`r soil.depths[4]`m}}) + \textrm{RMSE}^{*}(\mathrm{\theta}_{\textrm{0.4~m}}, W_{\textrm{liq,`r soil.depths[6]`~m}}) + \textrm{RMSE}^{*}(\mathrm{\theta}_{1~m}, W_{\textrm{liq,`r soil.depths[8]`~m}})
-(\#eq:obj-func)
-\end{split}
-\end{equation}
+#### Soil Water Potential Dynamics and Hydrological droughts
 
-based on monthly sums for both ET and $q$~run~ and daily averages for $\mathrm{\theta}$s.
-$W_{\textrm{liq,}\overline{z}}$ is the average of daily $W_{\textrm{liq,} z}$, where $z$ $\in$ $Z$; $Z$ = (`r paste(soil.depths[1:4], collapse=", ")`) m.
-Superscript * indicates that RMSE is standardized between 0 and 1 across the 5000 simulations.
-
-\
-Model fit to soil moisture dynamics by depth was one of our key priority.
-Preliminary analyses showed that fitting the model to all states and fluxes simultaneously captured soil moisture dynamics for shallower depths but not at 1 m depth.
-To ensure model fit at 1 m depth, we thus imposed a two-step procedure, wherein we first chose the 300 parameter ensemble members that minimized $\textrm{RMSE}^{*}({\mathrm{\theta}_{\textrm{1 m}}}, W_{\textrm{liq,`r soil.depths[8]` m}})$ (< 0.2), and among those 300 ensemble members we chose 100 with the lowest values for the objective function $\overline{O}$.
-See Table S1 for the ranges of best-fit values for different parameters.
-
-##### Hydrological droughts
-
-We ran ELM-FATES with the chosen 100 ensemble members from 1985-2018, with the first 5 years used for model spin-up.
+We ran ELM-FATES with the best-fit 100 ensemble members from 1985-2018, with the first 5 years used for model spin-up.
 Extreme hydrological droughts were identified by depth $z$ as days for which $\mathrm{\Psi}_{\textrm{soil},z}$ was more negative than the 5^th^ percentile of $\mathrm{\Psi}_{\textrm{soil},z}$ for a given day of the year.
 
-##### ELM-FATES model evaluation
+#### ELM-FATES model evaluation
 
@@
 
-### ERD model validation and selection
+### ERD model structure selection
 
@@
 For comparison with modelled ERD, we removed six species from the @Meinzer:1999 dataset to account for the uncertainty in their water-sourcing depths.
-One species with a $\mathrm{\delta}$^2^H value of -33.5 (SE $\pm$ 2.6) which is particularly difficult to assign a source-water depth given the uncertainty in soil water $\mathrm{\delta}$^2^H data by depth (see Fig S3).
+One species, *Guapira standleyana*, was removed because it's $\mathrm{\delta}$^2^H value of -33.5 (SE $\pm$ 2.6) was particularly difficult to assign a source-water depth given the uncertainty in soil water $\mathrm{\delta}$^2^H data by depth (see Fig S3).
 $\mathrm{\delta}$^2^H~xylem~ from leafless trees may not be linked with water sourced at the time of measurement and thus may not be comparable to species that had leaves.
-Leaflessness status of sampled trees is not recorded in [@Meinzer:2001].
-We therefore removed five species that are typically leafless in March (Joseph Wright, personal observation), that is, the month of $\mathrm{\delta}$^2^H~xylem~ sampling.
+Leaflessness status of sampled trees is not recorded in @Meinzer:2001.
+We therefore removed five species that are typically leafless in March-April (Joseph Wright, personal observation), that is, the months of $\mathrm{\delta}$^2^H~xylem~ sampling.
 For each model of ERD (Eq.
@@
 
-## Mortality analyses and species-specific drought experience
+## Mortality analyses and species-specific drought exposure
 
@@
 Dry season deciduous species may escape drought exposure via leaf deciduousness, so we analysed deciduous and evergreen species separately.
-Tree species on BCI are scored by expert botanists as one among four leaf-habits; evergreen, brevideciduous, facultative deciduous and obligate deciduous [@Meakem.2018; see SI].
+Tree species on BCI are scored by expert botanists as one among four leaf-habits; evergreen, brevideciduous, facultative deciduous and obligate deciduous (@Meakem.2018; see SI).
 We pooled all deciduous leaf habits together (hereafter 'deciduous' for brevity) and regressed species ERD against species $\textrm{M}_{s,t}$ for each census interval $t$ for deciduous ($n$ = `r nrow(mrate.depth.mean %>% subset(deciduous != "E"))`) and evergreen species ($n$ = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`) separately.
 
-Extreme and prolonged water-stress could lead to mortality via hydraulic failure and/or carbon starvation [@Rowland:2015; @McDowell:2018].
+\
 As an indicator of exposure to water-stress, we use a species-specific critical hydraulic threshold, $\mathrm{\Psi}$~crit~, here defined as $\mathrm{\Psi}$~20,leaf~.
-The duration of exposure to water-stress, and thus realized hydraulic-risk, was defined for each species as the proportion of days in each census interval $t$ of the $\mathrm{\Psi}_\textrm{soil}$ simulation period (1990-2015) during which soil water potential in the soil layer matching species ERD, $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$, were more negative than species $\mathrm{\Psi}$~crit~.
+The duration of exposure to water-stress, and thus the potential for realized hydraulic risk, was defined for each species as the proportion of days in each census interval $t$ of the $\mathrm{\Psi}_\textrm{soil}$ simulation period (1990-2015) during which soil water potential in the soil layer matching species ERD, $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$, were more negative than species $\mathrm{\Psi}$~crit~.
 
@@
 S4, S5).
+The distribution of maximum depth of soil water dynamics, and thus ecosystem root zone depth, across the 100 hydrological realisations encompassed 95% CI of `r root.zone.CI1` m - `r root.zone.CI2` m with the median at `r root.zone.med` m.
 
@@
 
-## Predictors of leaf hydraulic resistance curves
+## Predictors of leaf hydraulic vulnerability curves
 
-Parameters of leaf resistance curves were predictable from WSG and LMA (Fig.
+Parameters of leaf vulnerability curves were predictable from WSG and LMA (Fig.
 \@ref(fig:ab-fit)).
@@
 S7).
-This predictive power allowed us to estimate leaf resistance curves used in the ERD models for `r sp.n.model` ERD species that lacked direct observations (among `r length(erd.sp)` ERD species) (Table S3, Fig.
+This predictive power allowed us to estimate leaf vulnerability curves used in the ERD models for `r sp.n.model` ERD species that lacked direct observations (among `r length(erd.sp)` ERD species) (Table S3, Fig.
 S8).
@@
 \@ref(fig:sp-erd)).
-Evergreen species constituted a greater proportion of species with deeper ERDs compared with deciduous species, a group composed of a variety of categories of deciduousness.
-The range of ERD largely overlapped for both phenology groups (Fig.
+Evergreen and deciduous species had similar range of ERDs, but a greater proportion of deeper ERD species tended to be evergreen rather than deciduous---a group composed of a variety of categories of deciduousness (Fig.
 \@ref(fig:sp-erd)).
@@
 
-Among the seven census-intervals over 1982-2015 for which we analysed relationship of ERD with mortality rates, six intervals were associated with occurrence of one or more El Niño event [@Detto.2018; @Condit.2017].
-Out of the six intervals with El Niño events, among evergreen species, ERD explained 30-40% variation in mortality rates during five intervals--1982-1985, 1985-1990, 1995-2000, 2000-2005 and 2005-2010--such that species mortality rates decreased with deeper ERD (p-values significant at alpha = 0.05 for four intervals and at alpha = 0.1 for one interval; Fig.
+Among the seven census-intervals over 1982-2015 for which we analysed relationship of ERD with mortality rates, six intervals were associated with occurrence of one or more El-Niño event [@Detto.2018; @Condit.2017].
+Out of the six intervals with El-Niño events ERD explained 30-40% variation in mortality rates among evergreen species during five intervals--1982-1985, 1985-1990, 1995-2000, 2000-2005 and 2005-2010--such that species mortality rates decreased with deeper ERD (p-values < 0.05 for four intervals and 0.06 for one interval; Fig.
 \@ref(fig:erd-mort-census-evg)).
-When deciduous species were included, ERD explained 10-16% variation in mortality in four intervals with El Niño events (p-values marginally significant at alpha = 0.1; Fig.
+For deciduous species ERD explained 11-16% variation in mortality in four intervals with El-Niño events, but p-values were not significant (Fig.
 S11).
 
-Our analysis of hydrological droughts ranged from 1990 to 2015 and revealed distinctive extreme hydrological droughts for the census-intervals 1995-2000, 2000-2005 and 2005-2010, for which ERD explained significant mortality (Fig.
+\
+Our analysis of hydrological droughts ranged from 1990 to 2015 and revealed distinctive extreme, prolonged hydrological droughts for the census-intervals 1995-2000, 2000-2005 and 2005-2010, for which ERD explained significant mortality (Fig.
 \@ref(fig:swp-droughts)).
-Census interval 1995-2000 experienced a prolonged and extreme hydrological drought during the 1997-1998 El Niño event.
-During 2000-2005 and 2005-2010 extreme hydrological drought occurred in the wet season, effectively extending the dry season water-stress into the wet season (Fig.
-3).
 ERD also explained significant mortality in earlier droughts [1982-1985 and 1985-1990; @Condit.2017], not covered by our $\mathrm{\Psi}_{\textrm{soil}}$ estimates.
 
-On an average, species exposure to water-stress (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) steeply declined with ERD, indicating that species with shallower ERD spent greater time under significant hydrological drought, and thus likely experienced greater hydraulic-risk, which exponentially increased over 1995-2000, 2000-2005 and 2005-2010 (Fig.
-\@ref(fig:erd-time-evg)).
-For the census interval 1990-1995, species with the shallowest ERD also showed greater realized hydraulic-risk (Fig.
+\
+On an average, species exposure to water-stress (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) exponentially declined with ERD, indicating that species with shallower ERD spent greater time under significant hydrological drought, and thus likely experienced greater hydraulic risk (also see SI).
+Exposure to water-stress increased over the three periods for which ERD explained significant mortality (Figs.
+\@ref(fig:erd-time-evg), S11), although it was also high in 1990-1995 for the shallowest ERD (Fig.
 \@ref(fig:erd-time-evg)), but without elevated mortality rates (Fig.
 \@ref(fig:erd-mort-census-evg)).
-Overall, species with ERD deeper than 2.9 m did not cross species $\mathrm{\Psi}$~crit~ for the entire 25-yr period for which $\mathrm{\Psi}_{\textrm{soil},z}$ was modeled (1990-2015).
-Soil depths deeper than 1.7 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and drought periods, which is above $\mathrm{\Psi}$~crit~ of the most sensitive tree species (`r min.psi_crit` MPa).
 
@@
 
-We introduce a novel approach for estimating ERD using 25 years of tree growth, species-specific leaf resistance curves, modeled soil water potential profiles, and observed VPD.
+We introduce a novel approach for estimating ERD using 25 years of tree growth, species-specific leaf vulnerability curves, modeled soil water potential profiles, and observed VPD (\@ref(eq:M-chosen), Figs.
+\@ref(fig:schematic), \@ref(fig:sp-erd)).
 Our predictions of ERD were consistent with estimates using $\mathrm{\delta}$^2^H~xylem~ (Figs.
-\@ref(fig:erd-iso), \@ref(fig:sp-erd)).
-Our analyses suggest that co-occurring large canopy tree species with deeper ERD were associated with higher above-ground hydraulic efficiency, but not safety (Fig.
+\@ref(fig:erd-iso)).
+Our analyses suggest that co-occurring large canopy tree species with deeper ERD were associated with higher above-ground hydraulic efficiency, but lower safety (Fig.
 \@ref(fig:erd-hyd)).
-Species with deeper ERD showed higher maximum stem hydraulic conductance, and greater xylem vulnerability to embolism and narrower above-ground hydraulic safety margins compared to shallow ERD species.
-Nevertheless, species with deeper ERD, among the evergreen but not necessarily deciduous leaf habits, showed lower mortality rates than shallower ERD species.
-This ERD-mortality relationship was significant in five out of six census-intervals that had experienced one or more El-Nino events over seven census-intervals studied in total (1982-2015).
+Nevertheless, evergreen but not deciduous species with deeper ERD showed significantly lower mortality rates than shallower ERD species (Fig.
+\@ref(fig:erd-mort-census-evg), S11).
+This ERD-mortality relationship was significant in five out of six census-intervals that had experienced one or more El-Niño events---over seven census-intervals studied in total (1982-2015).
 Quantifying extreme droughts in the whole soil column over 1990-2015 revealed that ERD explained significant levels of mortality in periods when extreme soil droughts prolonged the dry season water-stress (Fig.
-\@ref(fig:erd-mort-census-evg)).
-Analysis of species time spent beyond $\mathrm{\Psi}$~crit~ in the soil zone matching ERD $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ showed that exposure to water-stress exponentially declined with deeper ERD (Fig.
+\@ref(fig:swp-droughts)).
+Species exposure to water-stress exponentially declined with deeper ERD (Fig.
 \@ref(fig:erd-time-evg)).
 Because extreme, sustained water-stress increases risk of mortality via hydraulic-failure and/or carbon starvation, deeper ERD may mitigate drought-induced mortality by limiting exposure to water-stress [@Rowland:2015; @McDowell:2018].
-We thus demonstrate, for the first time, a link between species trade-offs in above- and below-ground hydraulic traits, drought exposure through extreme hydrological droughts quantified over the whole soil-column across several El-Nino events, and large ($\ge$ 10 cm DBH) tree mortality over 35 years.
+We thus demonstrate, for the first time, a link between species trade-offs in above- and below-ground hydraulic traits, drought exposure through extreme hydrological droughts quantified over the whole soil-column across several El-Niño events, and large ($\ge$ 10 cm DBH) tree mortality over 35 years.
 
-## Drought strategies designed to mitigate realised hydraulic-risks
+## Drought strategies designed to mitigate realised hydraulic risks
 
-Our findings indicate how eco-evolutionary drought strategies may have been shaped by trait coordination and investments adapted for realised risks in the environment.
-If investments in stress-tolerance traits come at a cost, such insurance may not pay-off if the risk of stress is not real.
-Our findings suggest that shallow rooted species pay the cost of hydraulic safety in terms of hydraulic efficiency (Fig.
-\@ref(fig:erd-hyd)), adapted for an environment in which hydraulic-risk is significant, as extreme droughts cause exponentially greater water-stress in shallow soil layers (Fig.
-S6); whereas species with investment in deep roots can afford a risky suite of hydraulic traits, because access to a reliable deep-water resource ensures that for them hydraulic-risk is not realised (Figs.
-\@ref(fig:swp-droughts)).
+If investments in stress-tolerance traits come at a cost, such insurance may not pay-off if the risk of stress is not realised.
+Our findings suggest that species with investment in deep roots can afford the hydraulically efficient, but risky suite of traits (Fig.
+\@ref(fig:erd-hyd)), because access to a reliable deep-water resource ensures that for them hydraulic risk is not realised (Figs.
+\@ref(fig:erd-time-evg)), whereas shallow rooted species pay the cost of hydraulic safety in terms of efficiency, adapted for an environment in which hydraulic risk is significant, as extreme droughts cause exponentially greater water-stress in shallow soil layers (Fig.
+S6).
 
-Our findings suggest that the above- and below-ground trait coordination associated with differential environmental risks has adaptive value in terms of plant performance.
-Shallow rooted species on an average had greater mortality rates, especially for evergreen species, than deep rooted species through prolonged extreme droughts (Fig.
-\@ref(fig:erd-mort-census-evg)) that may have crossed even the greater tolerance limits of shallow-rooted species (Fig.
-\@ref(fig:erd-hyd)).
-Hydraulic-risk for shallow rooted species may have been exacerbated in our 35-yr study period in which extreme water-deficit years were more frequent than those in the last century according to a prior study [@Condit.2017].
-Deep rooted evergreen species on an average had greater survival than shallow rooted species through extreme droughts; although their survival may not be guaranteed if droughts intensify in future.
+\
+Extreme, prolonged El-Niño droughts (Fig.
+\@ref(fig:swp-droughts)) in our study may have crossed even the greater tolerance limits of shallow-rooted species leading to their greater mortality compared to deep-rooted species (Fig.
+\@ref(fig:erd-mort-census-evg)).
+Hydraulic risk for shallow rooted species may have been exacerbated in our 35-yr study period in which extreme water-deficit years were more frequent than those in the last century [@Condit.2017].
+Greater survival of deep-rooted species that we observed may change into the future if droughts intensify.
 
-Our finding of greater mortality under extreme soil water-stress among evergreen but not deciduous species is consistent with the expectation that deciduous species can also avoid drought-stress via leaf deciduousness.
-Studies on simultaneous coordination between leaf phenology, rooting depth and hydraulic traits are almost absent for the tropics and warrant future consideration [@Oliviera.2021].
+\
+We found significant ERD-mortality relationship during extreme water-stress for evergreen but not deciduous species (Fig.
+\@ref(fig:erd-mort-census-evg), S11); consistent with the expectation that deciduous species can also avoid water-stress via leaf-drop.
+This is also consistent with the observation that species distribution along local (BCI) and regional (Panama) moisture gradients is correlated with $\mathrm{\Psi}$~tlp~ for evergreen but not deciduous species [@Kunert.2021].
+Studies assessing coordination between leaf phenology, rooting depth and hydraulic traits simultaneously are almost absent for the tropics and warrant future consideration [@Oliviera.2021].
 Our analysis for ERD-hydraulic traits relationship was limited to only seven species, but covered all species of different leaf phonologies (Fig.
 \@ref(fig:erd-hyd)), so here we assume that the trend in the ERD-hydraulic trait relationships holds true across all species of different leaf phonologies.
-We also found that the range of ERDs overlapped across leaf phenology, but evergreen species tend to have deeper ERDs, consistent with observations by @Meinzer:1999 and elsewhere [@Smithmartin.2020; @Fan:2017; @Oliviera.2021].
-Whether deciduous species also have more efficient and vulnerable hydraulics at BCI as is observed elsewhere remains to be seen [@Markesteijn:2010; @Markesteijn:2011; @Xu:2016; @Gleason.2016].
-Leaf pehnologies at BCI are numerous and complicated and need further research in the future.
+We also found that the range of ERDs overlapped across leaf phenology, but evergreen species tend to have deeper ERDs (Fig.
+\@ref(fig:sp-erd)), consistent with observations by @Meinzer:1999 and elsewhere [@Smithmartin.2020; @Fan:2017; @Oliviera.2021].
+Whether deciduous species also have more efficient and vulnerable hydraulics at BCI as is observed elsewhere remains to be studied [@Markesteijn:2010; @Markesteijn:2011; @Xu:2016; @Gleason.2016].
+Leaf phenologies at BCI are numerous and complicated and warrant research.
 
@@
 
-Our study brings attention to the need for assessing drought sensitivity in terms of species drought exposure and realized hydraulic-risk by accounting for hydrological drought and trees' rooting depths.
+Our study brings attention to the need for assessing drought sensitivity in terms of species drought exposure and realized hydraulic risk by accounting for hydrological drought and trees' rooting depths.
 We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins (Fig.
 \@ref(fig:erd-hyd)).
-These traits are commonly identified as proxies for mortality risk [@Anderegg:2016], but in fact were associated with less drought exposure and had lower mortality, especially in evergreen species.
+These traits are commonly identified as proxies for mortality risk [@Anderegg:2016], but in fact were associated with species with less drought exposure and had lower mortality.
 Hydraulic risk was balanced by investment in deep roots (Figs.
@@
 
+\
 Our results are consistent with recent studies that analysed rooting or water-sourcing depths versus hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016; also see @Brum.2019; @Brum.2017; @Rowland:2015].
@@
 \
-Our result--of lower mortality in deep-rooted trees--contrasts with the inverse model finding of @Chitra-Tarak.2018 (hereafter, CT2018) in which deeper ERD species in a South-Asian seasonally tropical dry forest had higher mortality in a rare, prolonged drought.
+Our result of lower mortality in deep-rooted trees contrasts with the inverse model finding of @Chitra-Tarak.2018 (hereafter, CT2018) in which deeper ERD species in a South-Asian seasonally dry tropical forest had higher mortality in a rare, prolonged drought.
 The hydrological model in CT2018 revealed that the multi-year drought exhausted the deep soil and even bedrock water availability (also see [@Goulden.2019; @Ivanov:2012]).
@@
 
-## Effective rooting depths at the tree community level
+## Modeling effective rooting depths at the tree community level
 
@@
 \@ref(fig:sp-erd)), we make an important advance in modeling effective rooting depths at the community level in species-rich tropical forests.
-Our best ERD model is a key improvement over the model of CT2018 ($n$ = 12 species) as we employ a physiologically meaningful representation, fewer parameters and corroboration with tree and soil data for $\mathrm{\delta}$^2^H.
+Our best ERD model is a key improvement over the model of CT2018 as we employ a physiologically meaningful representation, fewer parameters and corroboration with tree and soil data for $\mathrm{\delta}$^2^H.
 
-As in CT2018, we also interpret our ERD estimates as revealing relative differences among species' effective rooting depths rather than absolute depths.
-Our ERD model predicts only daily maximum diurnal $\mathrm{\Psi}$~leaf~ assuming it is equivalent to $\mathrm{\Psi}$~soil~.
-While this scale may be adequate to relate to 5-year average growth values, this approach ignores diurnal dynamics of $\mathrm{\Psi}$~leaf~ and other factors that would play a role in such diurnal dynamics, such as the buffering effect of stem water storage capacitance on $\mathrm{\Psi}$~leaf~.
-By ignoring stem water storage capacitance [@Wolfe.2017] our model may thus have overestimated ERD especially in those species for which capacitance is important as a drought-avoidance strategy, for example, the deciduous species [@Borchert.2005jma].
+\
+Our ERD model predicts daily maximum diurnal $\mathrm{\Psi}$~leaf~ assuming it is equivalent to $\mathrm{\Psi}$~soil~.
+Our model ignores other factors that buffer soil drying such as stem water storage capacitance [@Wolfe.2017], and may thus have overestimated ERD, especially in those species for which capacitance is important as a drought-avoidance strategy, for example, the deciduous species [@Borchert.2005jma].
 Future studies should investigate the role of capacitance on estimating ERD.
 
+\
 While ERD models have the potential to estimate ERDs for entire tree communities, during their development phase, ERD models may need to be validated against data for a subset of representative species of the community, as this study did.
@@
 The relationships that we identify between above- and below-ground traits, vertical profile of soil water status and mortality rates are important for representing diversity in DGVMs, which intrinsically rely on the parameterization of contrasting life history strategies [@Scheiter:2013] and the simulation of competition between those strategies.
-ERDs could be assimilated into DGVMs by setting rooting parameters that result into the centroid of root water-uptake profiles equivalent to ERDs.
-Current DGVMS could thus provide an informative evaluation of ERD model as well.
-We found that LMA and WSG were strong predictors of leaf hydraulic resistance curves.
-Albeit future studies should undertake sensitivity analyses for the errors involved, our finding offers the promise of a greater ability to parameterize the 'hard' hydraulic traits with the abundant 'soft' trait data, thus allowing for a better representation of forest hydrodynamics.
-The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits in order to better represent the functional diversity of tropical forests, and their trajectories into the future.
+In the context of trait filtering models, if we used the hydraulic trait information without knowledge of their relationship to rooting depth, models would likely kill the 'risky' strategy trees in droughts, which would in fact be the opposite result from that observed in this study.
+A key outcome of this study is thus the relationships between hydraulic traits and ERD that could be plugged into a DGVM of BCI.
+To assimilate ERDs, DGVMs could vary rooting parameters such that the centroid of the species water-uptake profiles match ERDs.
+We found that LMA and WSG were strong predictors of leaf hydraulic vulnerability curves.
+Albeit future studies should undertake sensitivity analyses for uncertainties involved, our finding offers the promise of a greater ability to parameterize the 'hard' hydraulic traits with the abundant 'soft' trait data, thus allowing for a better representation of forest hydrodynamics.
+The relationships between ERD and above-ground hydraulic traits that we find thus provide important insights on how to model rooting depths and their coordination or trade-offs with other traits in order to better represent the functional diversity of tropical forests, and their trajectories into the future.
 
@@
 S9).
-We interpret this result as that VPD and leaf hydraulic resistance curves may be adequate to explain inter-census differences in growth (via stomatal control), but also acknowledge that our estimates of seasonality of LAI is a tentative estimate that combines leaf-fall data and the timing of leaf-gain backtracked from a leaf lifetime, and omits inter-annual variation.
+Our interpretation of this result is that VPD and leaf hydraulic vulnerability curves may be adequate to explain inter-census differences in growth (via stomatal control), but also acknowledge that our estimates of seasonality of LAI is a tentative estimate that combines leaf-fall data and the timing of leaf-gain backtracked from a leaf lifetime, and omits inter-annual variation.
+Species-level leaf-fall data from litter-traps that we use includes large within-species variability in leaf-fall timing, and thus does not capture typical tree-level deciduous seasonality.
 Future studies may improve models of LAI seasonality and make use of data from new technologies such as drone based monitoring of LAI in species-rich tropical forests [@Park.2019].
@@
 \
-While our ERD model empirically predicts growth via estimating $\mathrm{\Psi}$~leaf~ from $\mathrm{\Psi}$~soil~ of a specific depth, mechanistic models that account for plant hydrodynamics and other processes influencing growth are likely to predict growth more accurately, and thus ERD and hydraulic-risk [@Christoffersen:2016kd2d; @Sperry.1998; @Yang.2019; @Duursma:2018].
+While our ERD model empirically predicts growth via estimating $\mathrm{\Psi}$~leaf~ from $\mathrm{\Psi}$~soil~ of a specific depth, mechanistic models that account for plant hydrodynamics and other processes influencing growth are likely to predict growth more accurately, and thus ERD and hydraulic risk [@Christoffersen:2016kd2d; @Sperry.1998; @Yang.2019; @Duursma:2018].
 Although data needs for parameterizing such models could be greater (e.g.
@@
 VWC), hydraulic conductivity (K~soil~) by depth, soil moisture by depth, stream discharge and evapotranspiration were important for effective calibration of our 1-D hydrological model.
-We recommend wide-spread and coordinated collection of these variables as well as water-table levels in forest-inventory sites to allow for estimation of trees' water environments.
+We recommend widespread and coordinated collection of these variables as well as water-table levels in forest-inventory sites to allow for estimation of trees' water environments.
 
@@
 Establishing relationships between environment, traits and demographic outcomes of plants is imperative for developing a predictive plant ecology.
-Trees rooting depths and actual water environments through hydrological rather than meteorological droughts are nonetheless rarely studied.
+Trees' rooting depths and actual water environments through hydrological rather than meteorological droughts are nonetheless rarely studied.
 To our knowledge this is the first study to test for a mechanistic link between plant-available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.
@@
 This has important implications for our predictive understanding of tropical forest dynamics under current and future climate.
-Our community-scale framework for modeling effective rooting depths and leaf resistance curves indicates the possibilities in expanding the use of these critical, rare observations in species-rich forests towards community-scale generalizations.
+Our community-scale framework for modeling effective rooting depths and leaf vulnerability curves indicates the possibilities in expanding the use of these critical, rare observations in species-rich forests towards community-scale generalizations.
 
@@
 Department of Agriculture, McIntire Stennis project under LAB94493.
+LS was supported by National Science Foundation award 2017949.
 RF was supported by the National Center for Atmospheric Research, which is funded by the National Science Foundation.
@@
 $\mathrm{\Psi}_{\textrm{soil},z}$ by DOY for a year (colored line) is only shown if at least one day of the year $\mathrm{\Psi}_{\textrm{soil},z}$ was more negative than the 5^th^ percentile of $\mathrm{\Psi}_{\textrm{soil},z}$ for that day of year (DOY).
-Note the distinctive features of extreme droughts that occurred during the three periods indicated by a $*$: all three periods featured extreme hydrological droughts that either distinctly prolonged over the dry season (1995-2000), or distinctly occurred in the wet season, effectively extending the dry season (2000-2005, 2005-2010).
+Note the distinctive features of extreme droughts that occurred during the three periods indicated by a $*$: all three periods featured extreme hydrological droughts that either prolonged over the dry season (1995-2000), or occurred in the wet season, effectively extending the dry season (2000-2005, 2005-2010).
 
 \
-Figure \@ref(fig:ab-fit): Parameters $A$ (panel **a**) & $B$ (panel **b**) for the leaf resistance curves that were fitted to observed data on $K_\textrm{leaf}$ vs.
+Figure \@ref(fig:ab-fit): Parameters $A$ (panel **a**) & $B$ (panel **b**) for the leaf vulnerability curves that were fitted to observed data on $K_\textrm{leaf}$ vs.
 $\mathrm{\Psi}$~leaf~ for `r length(sp.exp.param$sp)` species (Eq.
@@
 \
-Figure \@ref(fig:erd-time-evg): Modeled effective rooting depth (ERD; horizontal-axis) versus time spent beyond critical hydraulic threshold (vertical axis) by census interval (colored bars; note the squared y-axis scale).
+Figure \@ref(fig:erd-time-evg): Modeled effective rooting depth (ERD; horizontal-axis) versus time spent beyond critical hydraulic threshold (vertical axis) by census interval (colored bars) for `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))` evergreen species included in the mortality analyses (Fig.
+\@ref(fig:erd-mort-evg)).
 Each bar represents the average time species of the same ERD spent beyond species-specific critical hydraulic thresholds in a given interval, that is, the proportion of days for which $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ was more negative than species $\mathrm{\Psi}$~crit~, defined as $\mathrm{\Psi}$~20,leaf~, and where $z$ is the soil depth matching species ERD.
-Standard error of the mean are shown over each bar.
-Figure includes all the `r length(erd.mrate.sp)` canopy tree species, evergreen as wells as deciduous, that were included in the mortality analyses.
-Trends were similar for evergreen species alone.
+Standard error of the mean are shown over each bar when available.
+Note the squared y-axis scale.
 

### 2021-04-08 10.0_Supporting_Information.Rmd
ELM-FATES and ERD model details moved to SI

@@
 
+```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "35%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd}Scatterplot of mean midday (11:00 am to 15:00 pm) gross primary productivity (GPP) and mean midday vapor pressure deficit (VPD) measured at the eddy flux tower in Barro Colorado Island over 2012-2017.
+Red line shows the fit from a polynomial model described in inset."}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/gpp_by_env_variables/monthly_gpp.vpd.jpeg"))
+```
+
+\newpage
 ```{r ksat-profiles, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ksat-profiles}Generated 5000 curves of soil hydraulic conductivity variation by depth based on Latin Hypercube Sampling within 95% CI of observed data at depths of 12.5 cm and 60 cm [@Kinner:2004], assuming a linear decline between those depths, and values before 12.5 and after 60 cm as that of 12.5 cm and 60 cm, respectively.
@@
 
-```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "35%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd}Scatterplot of mean midday (11:00 am to 15:00 pm) gross primary productivity (GPP) and mean midday vapor pressure deficit (VPD) measured at the eddy flux tower in Barro Colorado Island over 2012-2017.
-Red line shows the fit from a polynomial model described in inset."}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/gpp_by_env_variables/monthly_gpp.vpd.jpeg"))
-```
-
+\newpage
 ```{r iso-data, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-iso-data}Stable isotopic ratio ($\\mathrm{\\delta}^2H$) of soil water by depth measured at Barro Colorado Island in March 1997.
@@
 
+\newpage
 ```{r sens-sm-et, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-sm-et}Parameter sensitivity of ELM-FATES soil water content (that is, VWC, **a**-**b**) and ET (**c**-**d**) is shown for two representative sample months, as sensitivity for a particular parameter varies depending on the chosen month.
@@
 
+\newpage
 ```{r sens-run, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-run}Parameter sensitivity of ELM-FATES stream discharge (QRUNOFF) is shown for four sample months (**a**-**d**), as sensitivity for a particular parameter varies depending on the chosen month.
@@
 
+\newpage
 ```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($\\Psi_{\\textrm{soil},z}$) by depth $z$ (m; panels) over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines).
@@
 
+\newpage
 ```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}Vulnerability curves fitted to observed $K_\\textrm{leaf}$ vs.
@@
 
+\newpage
 ```{r obs-comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-comm-plc-curves}Percent loss of vulnerability curves for species with ERD estimates.
@@
 
+\newpage
 ```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-other-erd-mods}Effective Rooting Depth versus $\\mathrm{\\delta}^2H$ for each of the ERD models tested (panels).
@@
 
+\newpage
 ```{r erd-hyd-full, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd-full}Pairwise Spearman's correlations (upper triangle) and corresponding linear regressions (lower triangle) among modelled Effective Rooting Depth (ERD) and hydraulic properties for seven canopy species found on Barro Colorado Island (Panama).Spearman's correlation coefficients and significance levels are shown (upper-half; '*' $p$ < 0.05; '.' $p$ < 0.1; ' ' $p$ < 1).")}
@@
 
-```{r erd-mort-deci, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-cdeci}Species (circles) Mortality rates in species of various deciduous leaf-habit categories (colors) across census intervals from 1985-2015 plotted against ERD.
+\newpage
+```{r erd-mort-deci, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-cdeci}Mortality rates across census intervals from 1985-2015 for species (circles) of various deciduous leaf-habit categories (colors) plotted against ERD.
 $R$^2^ and significance levels for linear model fits are given in panel insets.
-Species are color colded by deciduous category.
 Census-intervals with a superscript * are shown for comparison and indicate intervals in which significant mortality was explained by ERD among evergreen species (see Fig.
-8 in the Main text)."}
+8 in the main text)."}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_by rdi.gr_deciduous.jpeg"))
@@
 
-```{r erd-time-deci, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time-deci}")}
+\newpage
+```{r erd-time-deci, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time-deci}Modeled effective rooting depth (ERD; horizontal-axis) versus time spent beyond critical hydraulic threshold (vertical axis) by census interval (colored bars) for deciduous species in mortality analyses (Fig.
+\\@ref(fig:erd-mort-deci)).
+Each bar represents the average time species of the same ERD spent beyond species-specific critical hydraulic thresholds in a given interval, that is, the proportion of days for which $\\mathrm{\\Psi}_{\\textrm{soil},z=\\textrm{ERD}}$ was more negative than species $\\mathrm{\\Psi}$~crit~, defined as $\\mathrm{\\Psi}$~20,leaf~, and where $z$ is the soil depth matching species ERD.
+Standard error of the mean are shown over each bar.
+Note the squared y-axis scale.")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mfac vs. rdi.gr_deciduous.jpeg"))
@@
 ```
-#
+
+\newpage
 ```{r hyd-table, results = 'asis', echo = FALSE}
 knitr::kable(hyd.table, caption = paste0("\\label{lab-hyd-table} Minimum leaf water potential $\\mathrm{\\Psi}$~min~, leaf turgor loss point $\\mathrm{\\Psi}$~tlp~, stem maximum hydraulic conductivity $K_\\textrm{max,stem}$ and stem vulnerability to cavitation $\\mathrm{\\Psi}$~88,stem~ for seven species with ERD estimates.
-$\\mathrm{\\Psi}$~min~ is sourced from @Wolfe.2019, while $\\mathrm{\\Psi}$~tlp~, $K_\\textrm{max,stem}$ and $\\mathrm{\\Psi}$~88,stem~ are sourced from @Wolfe.2021.
-$\\mathrm{\\Psi}$~min~, $\\mathrm{\\Psi}$~tlp~ and $\\mathrm{\\Psi}$~88,stem~ are in MPa, while $K_\\textrm{max,stem}$ in kg m^-1^ s^-1^ MPa^-1^."), booktabs = TRUE, escape = FALSE)
+$\\mathrm{\\Psi}$~min~ is sourced from @Wolfe.2019, while $\\mathrm{\\Psi}$~tlp~, $K_\\textrm{max,stem}$, $\\mathrm{\\Psi}$~50,stem~ and $\\mathrm{\\Psi}$~88,stem~ are sourced from @Wolfe.2021.
+$\\mathrm{\\Psi}$~min~, $\\mathrm{\\Psi}$~tlp~, $\\mathrm{\\Psi}$~50,stem~ and $\\mathrm{\\Psi}$~88,stem~ are in MPa, while $K_\\textrm{max,stem}$ in kg m^-1^ s^-1^ MPa^-1^."), booktabs = TRUE, escape = FALSE)
 ```
 
-#
+\newpage
 ```{r ab-param, results = 'asis', echo = FALSE}
-knitr::kable(ab.table, caption = paste0("\\label{lab-ab-param}Leaf vulnerability curves parameters *A* & *B*, $K_\\textrm{max,leaf}$ and $\\mathrm{\\Psi}$ for the ", length(erd.sp), " species with ERD estimates.
+knitr::kable(ab.table, caption = paste0("\\label{lab-ab-param}Leaf vulnerability curves parameters *A* & *B*, $K_\\textrm{max,leaf}$ and $\\mathrm{\\Psi}$~20,leaf~ for the ", length(erd.sp), " species with ERD estimates.
 Species for which curves were fitted to observed data for $K_\\textrm{leaf}$ vs.
-$\\mathrm{\\Psi}$~leaf~ are identified with Source as Data, while species for which curves were obtained based on trait-proxies are identified with Source as Model."), booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(ab.table))))
+$\\mathrm{\\Psi}$~leaf~ are identified with column Source as Data, while species for which curves were obtained based on trait-proxies are identified as Model."), booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(ab.table))))
 ```
+
+\newpage
+## Additional details of the ELM-FATES Model
+
+In ELM, soil water is predicted from a multi-layer model, in which the vertical soil moisture transport is governed by infiltration, surface and sub-surface runoff, gradient diffusion, gravity, canopy transpiration through root extraction, and interactions with groundwater (see @Oleson.2013 for process details and equations).
+FATES simulates size-structured tree community complete with individual tree rooting profiles.
+In our use of reduced-complexity configuration of ELM-FATES with a single plant functional type, ELM removes total community canopy transpiration as water-uptake from the soil layers in proportion to the total root fraction in each layer in the 1-D column.
+Note that we use the soil water potential dynamics obtained from ELM-FATES as an ecosystem-level outcome to drive our species-level, empirical ERD model, and thus believe that there is no conflict/circularity between tree-level rooting profiles or water uptake processes in ELM-FATES and the ERD model.
+
+\
+We ran ELM with FATES vegetation, in which the ELM model simulates interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\mathrm{\Delta}W_{\textrm{can}}$, surface water $\mathrm{\Delta}W_{\textrm{sfc}}$, soil water $\mathrm{\Delta}W_{\textrm{liq},z_{i}}$ at depth $z_{i}$ and water in an unconfined aquifer $\mathrm{\Delta} W_{a}$ (omitting processes relevant to snow, wetlands or lakes).
+Conservation of these terms is expressed as follows:
+
+\begin{equation}
+\begin{split}
+\mathrm{\Delta}W_\textrm{can} + \mathrm{\Delta}W_\textrm{sfc} + \sum_{i=1}^{15} \mathrm{\Delta}W_{\textrm{liq},z_{i}} + \mathrm{\Delta}W_{a} = \left(q_\textrm{rain} - E_{v} - E_{g} - q_\textrm{over} - q_\textrm{h2osfc} - q_\textrm{drai}\right)\mathrm{\Delta}t
+\end{split}
+(\#eq:water-balance)
+\end{equation}
+where, $q_{\textrm{rain}}$ is rainfall, $E_{v}$ is transpiration of the forest, $E_{g}$ is ground evaporation, $q_{\textrm{over}}$ is surface runoff, $q_{\textrm{h2osfc}}$ is runoff from surface water storage, $q$~drai~ is subsurface drainage (all in mm H~2~O) and $\mathrm{\Delta}t$ is the time-step (here, 1 hr).
+
+\
+Justification for the use of static stand structure: In static stand structure model, the model is started from an observed forest inventory and recruitment, growth and mortality are deactivated.
+Holding the stand structure static rather than allowing it to emerge from physiologically-mediated competition between trees removes a set of internal model feedbacks and thus the only differences between model simulations are driven by direct physiological effects of perturbed model parameters.
+This allows us to focus on the hydraulic and hydrological questions explored in this paper.
+
+## Additional details for ELM-FATES model parameterization
+
+In the Conrad catchment, in which the 50 ha plot is situated, soil hydraulic conductivity decreases with depth from 12.5 and 60 cm (@Godsey:2004; Table \@ref(tab:param)).
+We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths <12.5 cm and >60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig.
+\@ref(fig:ksat-profiles)).
+\
+For greater accuracy, instead of the ELM default pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content (GWC) vs.
+$\mathrm{\Psi}$~soil~ for depths of 0.15, 0.4 and 1 m [@Kupers:2019].
+We converted GWC to volumetric (VWC, $\theta$) using a measured bulk density value of 0.8 g cm^-3^ (see below).
+The @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences in water retention among depths were nominal,
+
+\begin{equation}
+\mathrm{\Psi}_{\textrm{soil}} = \mathrm{\Psi}_e\left(\frac{\mathrm{\theta}}{\mathrm{\theta}_{s}}\right)^{-b}
+(\#eq:soil-ret-curve)
+\end{equation}
+where $\mathrm{\Psi}_e$ is the air-entry matric potential, $\mathrm{\theta}_{s}$ is the saturated volumetric water content, and $b$, an index for soil pore-size distribution, respectively.
+$\mathrm{\Psi}_e$, $\mathrm{\theta}_{s}$ and $b$ that best-fit the data were 200 (mm H~2~O), 0.51 (cm^3^ cm^-3^) and 10 (unitless), respectively.
+
+## ELM-FATES calibration
+
+We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq.
+\@ref(eq:water-balance), namely, (1) evapotranspiration ET from the flux tower by the 50 ha plot ($\backsimeq$ $E_{v} + E_{g}$ in Eq.
+\@ref(eq:water-balance); 2012-2017; see [SI](#SIData.matteo))), (2) local stream discharge $q_\textrm{run}$ ($\backsimeq$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see @Paton.2019 and [SI](#SIData.steveP)) and (3) soil volumetric water content (VWC) $\mathrm{\theta}_{z}$ ($\backsimeq$ $W_{\textrm{liq,} z}$) from two sources: (i) a long-term (2012-2018) record of VWC $\mathrm{\theta}_\textrm{0-0.15 m}$ averaged across three vertical time domain reflectometry (TDR) probes over depth 0-15 cm from three locations near the flux tower, and (ii) plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 at depths of 0.15, 0.4 and 1 m (1299 samples covering all soil types and habitats; see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\mathrm{\theta}_{\textrm{0-0.15 m}}$, $\mathrm{\theta}_{\textrm{0.4 m}}$ and $\mathrm{\theta}_{\textrm{1 m}}$ refer to plot-wide averages).
+
+\
+For ELM-FATES calibration we calculated an objective function for each of the 5000-member ensemble by equally weighting standardized RMSE between observations and simulations across all fluxes and states mentioned above.
+We describe the objective function $O$ as:
+
+\begin{equation}
+\begin{split}
+\overline{O} = \frac{1}{6}\sum \textrm{RMSE}^{*}\textrm{(ET,}E_{v} + E_{g}) + \textrm{RMSE}^{*}(q_{\textrm{run}},q_{\textrm{over}} + q_{\textrm{drai}}) + \textrm{RMSE}^{*}(\mathrm{\theta}_{\textrm{0-0.15~m}}, {W_{\textrm{liq,}\overline{z}}}) \\
+& + \textrm{RMSE}^{*}(\mathrm{\theta}_{\textrm{0.15~m}}, W_{\textrm{liq,`r soil.depths[4]`m}}) + \textrm{RMSE}^{*}(\mathrm{\theta}_{\textrm{0.4~m}}, W_{\textrm{liq,`r soil.depths[6]`~m}}) + \textrm{RMSE}^{*}(\mathrm{\theta}_{1~m}, W_{\textrm{liq,`r soil.depths[8]`~m}})
+(\#eq:obj-func)
+\end{split}
+\end{equation}
+
+based on monthly sums for both ET and $q$~run~ and daily averages for $\mathrm{\theta}$s.
+$W_{\textrm{liq,}\overline{z}}$ is the average of daily $W_{\textrm{liq,} z}$, where $z$ $\in$ $Z$; $Z$ = (`r paste(soil.depths[1:4], collapse=", ")`) m.
+Superscript * indicates that RMSE is standardized between 0 and 1 across the 5000 simulations.
+
+\
+Model fit to soil moisture dynamics by depth was one of our key priority.
+Preliminary analyses showed that fitting the model to all states and fluxes simultaneously captured soil moisture dynamics for shallower depths but not at 1 m depth.
+To ensure model fit at 1 m depth, we thus imposed a two-step procedure, wherein we first chose the 300 parameter ensemble members that minimized $\textrm{RMSE}^{*}({\mathrm{\theta}_{\textrm{1 m}}}, W_{\textrm{liq,`r soil.depths[8]` m}})$ (< 0.2), and among those 300 ensemble members we chose 100 with the lowest values for the objective function $\overline{O}$.
+
+## Alternative model structures for Effective Rooting Depth (ERD)
+
+Below we present all the alternative models for Effective Rooting Depth (ERD).
+Daily average species-specific growth, $G_{t|z}$, in the census interval $t$ given an ERD of depth $z$, is described as follows:
+
+\begin{equation}
+G_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} \textrm{FLC}_{\textrm{leaf},s,i|z,h}^{*}) + \epsilon_{s,t|z,h}
+(\#eq:M1)
+\end{equation}
+
+\begin{equation}
+G_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} \textrm{FLC}_{\textrm{leaf},s,i|z,h}^{*} + \hat{\textrm{VPD}}_{i}^{*}) + \epsilon_{s,t|z,h}
+(\#eq:M2)
+\end{equation}
+
+\begin{equation}
+G_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} \textrm{FLC}_{\textrm{leaf},s,i|z,h}^{*} \hat{\textrm{VPD}}_{i}^{*}) + \epsilon_{s,t|z,h}
+(\#eq:M3)
+\end{equation}
+
+\begin{equation}
+G_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} \textrm{FLC}_{\textrm{leaf},s,i|z,h}^{*} \textrm{LAI}_{s,doy|i}^{*}) + \epsilon_{s,t|z,h}
+(\#eq:M4)
+\end{equation}
+
+\begin{equation}
+G_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} \textrm{FLC}_{\textrm{leaf},s,i|z,h}^{*} + \hat{\textrm{VPD}}_{i}^{*}) \textrm{LAI}^{*}_{s,doy|i}) + \epsilon_{s,t|z,h}
+(\#eq:M5)
+\end{equation}
+
+\begin{equation}
+G_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} \textrm{FLC}_{\textrm{leaf},s,i|z,h}^{*}\hat{\textrm{VPD}}_{i}^{*}\textrm{LAI}_{s,doy|i}^{*}) + \epsilon_{s,t|z,h}
+(\#eq:M6)
+\end{equation}
+where n~t~ are the total number of days in census interval $t$, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is the model error term.
+$\hat{\textrm{VPD}}$ is the predicted GPP from a locally derived polynomial relationship between gross primary productivity (GPP) and VPD, so as to account for the non-linear impact of VPD on growth.
+See the ERD model description in the main text for other details.
+
+## ERD estimation
+
+For each ERD model structure (Eqs.
+\@ref(eq:M1)-\@ref(eq:M6)), species-specific best-fit ERD was identified as follows: For example, for the chosen model (Eq.
+\@ref(eq:M3)), for a species $s$ and hydrological realization *h*, we obtained Akaike's Information Criterion (AIC) and Goodness-of-fit ($R$^2^) for each modeled soil layer depth $z$.
+We used AIC as the first selection criterion to ensure that the fitted growth values from Eq.
+\@ref(eq:M3) are within the confidence interval of the error distribution of observed growth (as the same model Eq.
+\@ref(eq:M3) was being compared for each *h*, number of model parameters were the same).
+For each *h*, we retained those depths with the maximum likelihood of growth (similar AIC based support, see below) filtering out the rest.
+However, AIC does not ensure that fitted values will have a positive trend with growth.
+We therefore used $R$^2^ between observed and fitted growth values as a goodness-of-fit measure and retained depths associated with a relatively high threshold $R$^2^ value of 0.5; after filtering out depths associated with negative correlations $r$ if any.
+Finally, among these we chose the depth with the highest $R$^2^ per *h*.
+Results were not sensitive to inclusion of this final step of retaining only a single depth as the best-fit (not shown).
+For a species $s$, we defined effective rooting depth (ERD) as the median of the best-fit depth over all *h*, which can be described as:
+
+\begin{equation}
+\textrm{ERD}_{s} = \underset{\forall \, h \, \in \, H}{\textrm{med}} \Big(z\,\Big|\, \Big(R^2_{s|z,h} = \underset{\forall \, z \,\in\,Z}{\textrm{max}}(R^2_{s|z,h}| \mathrm{\Delta}_{s|z,h} \le 2\Big)\Big) \, \bigcap \, R^2_{s|z,h} \ge 0.5 \, \bigcap \,  r_{s|z,h} \ge 0\Big)
+(\#eq:erd)
+\end{equation}
+
+where
+
+\begin{equation}
+\mathrm{\Delta}_{s|z,h} = \textrm{AIC}_{s|z,h} - \underset{\forall \, z \,\in\,Z}{\textrm{min}}(\textrm{AIC}_{s|z,h})
+(\#eq:delta)
+\end{equation}
+
+and
+
+\begin{equation}
+\textrm{AIC}_{s|z,h} = -2\textrm{log}(\mathcal{L}(\hat{\beta}_{s|z,h}|{G_{s}})) - 2\textrm{k}
+(\#eq:aic)
+\end{equation}
+
+where r is the correlation, and $R^2$ the Goodness-of-fit, between observed growth $G$ and estimated growth $\hat{G}.
+$\mathcal{L}$ is the likelihood function, $\hat{\beta}$ is the maximum-likelihood estimate, $G$ the observed growth with $k$ observations (or, census intervals $t$).
+$H$ = (1, .., 100), the 100 best-fit hydrological realizations and $Z$ = (0.4, `r paste(soil.depths[8:length(soil.depths)], collapse=", ")`) m.
+We removed finely resolved soil layers <0.21 m and averaged $\mathrm{\Psi}_{\textrm{soil},z}$ for depths 0.21, 0.37 and 0.62 (=0.4) m as sufficient resolution for model validation.
+
+## ERD model structure selection
+
+The best ERD model structure explained a large fraction of the variance in $\mathrm{\delta}$^2^H~xylem~.
+This model included an effect of VPD and not LAI (Eq.
+\@ref(eq:M3)).
+Two of the alternative models (Eqs.
+\@ref(eq:M4) & \@ref(eq:M6)) with LAI had a strong relationship with $\mathrm{\delta}$^2^H~xylem~ as well, but one that was based on a smaller set of species ($n$ = 4 or 5), because fewer species passed the criteria of sufficient species-level goodness-of-fit for these models (<17 species vs.
+`r length(erd.sp)` species for the best model, Fig.
+S9).
+For the rest of the analyses we therefore used ERD estimates from the less parsimonious model (Eq.
+\@ref(eq:M3)).
+
+# Additional results on exposure to water-stress
+
+Overall, species with ERD deeper than 2.9 m did not cross species $\mathrm{\Psi}$~crit~ for the entire 25-yr period for which $\mathrm{\Psi}_{\textrm{soil},z}$ was modeled (1990-2015).
+Soil depths deeper than 1.7 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and drought periods, which is above $\mathrm{\Psi}$~crit~ of the most sensitive tree species (`r min.psi_crit` MPa).
 
@@
 ### Growth Data
-\
+
 We selected trees for which the height of the diameter measurements were unchanged across all censuses (1990, 1995, 2000, 2005, 2010, 2015).
@@
 ### Microclimatic and flux tower {#SIData.matteo}
-\
+
 The tower used for these measurements is 41 m above ground, on a plateau on BCI at the north west corner of the 50ha plot [@Pau.2018].
@@
 ### Stream discharge from the Conrad catchment {#SIData.steveP}
-\
+
 The Conrad weir is located on the gently sloping western side of Barro Colorado Island.
@@
 
-\
 We collected volumetric water content (VWC) data (2012-2018) from three locations near the eddy flux tower with time domain reflectometry (TDR, CS616, Campbell Scientific) probes inserted vertically ranging over the first 15 cm of soil.
@@
 
-## Leaf habit
+### Leaf habit
 
@@
 
-Although deciduous leaf-habits are generally expected to be fully deciduous, at BCI there is a great within-species variation in the timing of leaf-drop and leaf-out and neither do all individuals of a deciduous species may sreach full deciduousness every year [@Condit.2000].
-Our reconstruction of $\textrm{LAI}^*$ seasonality is based upon species leaf-fall data, which captures the within-species variation in the timing of leaf-drop, so in our reconstruction the seasonal decline in $\textrm{LAI}^*$ appears to be more spread out and does not reach zero, in contrast to what is expected for an idealised individual of each deciduous leaf habit.
+Although deciduous leaf-habits are generally expected to be fully deciduous, at BCI there is a great within-species variation in the timing of leaf-drop and leaf-out, and neither do all individuals of a deciduous species reach full deciduousness every year [@Condit.2000].
+Our reconstruction of $\textrm{LAI}^*$ seasonality is based upon species leaf-fall data, which captures the within-species variation in the timing of leaf-drop, so in our reconstruction the seasonal decline in $\textrm{LAI}^*$ appears to be more spread out and does not reach zero, in contrast to what is ideally expected for an individual of each deciduous leaf habit.
 Greater uncertainty in leaf-lifespan for deciduous species may have compounded this issue further.
@@
 
-\
 Values of stem maximum hydraulic conductivity and vulnerability to cavitation were obtained from the dataset of @Wolfe.2021.
@@
 ### WSG and LMA {#SIData.rutuja}
-\
+
 We used WSG, the wood specific gravity after drying at 100^$\circ$^C and LMA, the leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules for leaves receiving direct sunlight as described in @Wright:2010.
@@
 
-## Alternative models for Effective Rooting Depth (ERD)
-\
-Below we present all the alternative models for Effective Rooting Depth (ERD).
-Daily average species-specific growth, $G_{t|z}$, in the census interval $t$ given an ERD of depth $z$, is described as follows:
-
-\begin{equation}
-G_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},s,i|z,h}^{*}) + \epsilon_{s,t|z,h}
-(\#eq:M1)
-\end{equation}
-
-\begin{equation}
-G_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},s,i|z,h}^{*} + \hat{\textrm{VPD}}_{i}^{*}) + \epsilon_{s,t|z,h}
-(\#eq:M2)
-\end{equation}
-
-\begin{equation}
-G_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},s,i|z,h}^{*} \hat{\textrm{VPD}}_{i}^{*}) + \epsilon_{s,t|z,h}
-(\#eq:M3)
-\end{equation}
-
-\begin{equation}
-G_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},s,i|z,h}^{*} \textrm{LAI}_{s,doy|i}^{*}) + \epsilon_{s,t|z,h}
-(\#eq:M4)
-\end{equation}
-
-\begin{equation}
-G_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},s,i|z,h}^{*} + \hat{\textrm{VPD}}_{i}^{*}) \textrm{LAI}^{*}_{s,doy|i}) + \epsilon_{s,t|z,h}
-(\#eq:M5)
-\end{equation}
-
-\begin{equation}
-G_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},s,i|z,h}^{*}\hat{\textrm{VPD}}_{i}^{*}\textrm{LAI}_{s,doy|i}^{*}) + \epsilon_{s,t|z,h}
-(\#eq:M6)
-\end{equation}
-where n~t~ are the total number of days in census interval $t$, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is the model error term.
-$\hat{\textrm{VPD}}$ is the predicted GPP from a locally derived polynomial relationship between gross primary productivity (GPP) and VPD, so as to account for the non-linear impact of VPD on growth.
-See the ERD model description in the main text for other details.
-
-The best ERD model explained a large fraction of the variance in $\mathrm{\delta}$^2^H~xylem~.
-This model included an effect of VPD and not LAI (Eq.
-\@ref(eq:M3)).
-Two of the alternative models (Eq.
-\@ref(eq:M4) & \@ref(eq:M6)) with LAI had a strong relationship with $\mathrm{\delta}$^2^H~xylem~ as well, but one that was based on a smaller set of species ($n$ = 4 or 5), because fewer species passed the criteria of sufficient species-level goodness-of-fit for these models (<17 species vs.
-`r length(erd.sp)` species for the best model, Fig.
-S9).
-For the rest of the analyses we therefore used ERD estimates from the less parsimonious model (Eq.
-\@ref(eq:M-3)).
-
 # References

### 2021-04-09 09.0_manuscript.rmd
Explained uncertainty in isotopic data

@@
 Summary | 200  
-Main Text  | 6971  
+Main Text  | 6939  
 Introduction | 832  
-Methods |  3211  
+Methods |  3150  
 Results | 855  
-Discussion | 2073  
+Discussion | 2102  
 Keywords | 8
@@
 In this paper, we estimate plant-available water in the whole-soil column in a tropical forest, and inversely estimate rooting depths of co-occurring tree species from their growth responses, with a series of model calibrations and validation.
-We evaluate how rooting depth is linked to above-ground hydraulic traits and mortality rates through several census-intervals over a 35 year period that experienced El Niño droughts of a variety of intensity, frequency and duration [@Detto.2018; @Condit.2017].
+We evaluate how rooting depth is linked to above-ground hydraulic traits and mortality rates through seven census-intervals over a 35-year period that experienced El Niño droughts of a variety of intensity, frequency and duration [@Detto.2018; @Condit.2017].
 Our hypotheses are that (i) deep-rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem due to greater and more reliable water availability at depth (see Tables \@ref(tab:symbols) and \@ref(tab:hypo)); and that, (ii) deep-rooted species have lower mortality rates during droughts due to a lower exposure to water-stress compared to shallow-rooted species.
@@
 In this study, we used the fractional loss of hydraulic conductivity of the leaf ($\textrm{FLC}$~leaf~) to estimate $B$.
-$\textrm{FLC}$~leaf~ is calculated as a fraction of maximum leaf conductivity $K_\textrm{max, leaf}$:
+$\textrm{FLC}$~leaf~ is calculated as a fraction of maximum leaf conductivity $K_\textrm{max, leaf}$,
 
@@
 Because pre-dawn $\mathrm{\Psi}$~leaf~ generally approaches $\mathrm{\Psi}$~soil~, this allowed us to substitute $\mathrm{\Psi}$~leaf~ with $\mathrm{\Psi}$~soil~ in leaf hydraulic vulnerability curves for each species $s$---defined below in Eq.
-\@ref(eq:vc-exp) using species-specific parameters $A_{s}$ and $B_{s}$---to predict maximum diurnal $K_{\textrm{leaf},s, i}$ for each day $i$ as:
+\@ref(eq:vc-exp) using species-specific parameters $A_{s}$ and $B_{s}$---to predict maximum diurnal $K_{\textrm{leaf},s, i}$ for each day $i$,
 
@@
 \end{equation}
-$K_{\textrm{max, leaf}, s}$, was obtained as the maximum value of $K_{\textrm{leaf}, s}$ using Eq.
+$K_{\textrm{max, leaf}, s}$ was obtained as the maximum value of $K_{\textrm{leaf}, s}$ using Eq.
 \@ref(eq:k-soil).
@@
 \
-We therefore tested alternate structures of empirical growth models, or inverse models of ERD, in which we regressed species-specific growth against multiplicative or additive effects of one or more growth factors calculated daily and averaged over 5-yearly censuses: $\hat{\textrm{VPD}}$, $textrm{FLC}_\textrm{leaf}$ and the leaf area index (LAI) (Eq.
-\@ref(eq:M-chosen) and those in SI).
+We therefore tested alternate structures of empirical growth models (Eq.
+\@ref(eq:M-chosen) and those in SI), or inverse models of ERD, in which we regressed species-specific growth against multiplicative or additive effects of one or more growth factors calculated daily and averaged over 5-yearly censuses: $\hat{\textrm{VPD}}$, $\textrm{FLC}_\textrm{leaf}$ and the leaf area index (LAI).
 Incorporation of radiation did not improve model fitting, possibly due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season at BCI.
@@
 \
-We evaluated different model structures (see SI) and for each we estimated species effective rooting depths (ERD) as the depth $z$ at which soil water dynamics ($\mathrm{\Psi}_{\textrm{soil,} z}$) best explained observed dynamics of growth via modeled growth (e.g.
-Eq.\@ref(eq:M-chosen)).
+We evaluated different model structures and for each we estimated species effective rooting depths (ERD) as the depth $z$ at which soil water dynamics ($\mathrm{\Psi}_{\textrm{soil,} z}$) best explained observed dynamics of growth $G$ (see below) via modeled growth $\hat{G}$.
 Our growth model, or inverse ERD model, does not explicitly use rooting profiles, but identifies soil water dynamics at a single depth $z$ as the central tendency that influences the observed growth dynamics the most.
@@
 We thus obtained parameters $A$ and $B$ for all the `r length(erd.sp)` ERD species for use in Eq.
-\@ref(eq:b-mod).
+\@ref(eq:k-soil).
 We also estimated species $\mathrm{\Psi}$~20,leaf~ using their vulnerability curves.
@@
 In some of the alternative models for effective rooting depth, we explored the effect of seasonality in leaf area index (LAI) on growth.
-We assumed species-specific mean seasonal curves for LAI (standardised between 0 and 1; unitless), informed by a combination of long-term records for leaf-fall [@Wright.1990aeo] and leaf-lifetime (@Osnas.2018; see SI).
+We assumed species-specific mean seasonal curves for LAI (standardised between 0 and 1; unitless), informed by a combination of long-term records for leaf-fall [@Wright.1990aeo] and leaf-lifetime [@Osnas.2018].
+See SI.
 
@@
 
-To estimate dynamics of water availability by depth over the forest’s rooting zone, we used the Energy Exascale Earth System Land Model (ELM, @Caldwell.2019), coupled with the Functionally Assembled Terrestrial Ecosystem Simulator (FATES, @Koven:2020) (hereafter, ELM-FATES).
+To estimate dynamics of water availability by depth over the forest’s rooting zone, we used the Energy Exascale Earth System Land Model (ELM; @Caldwell.2019), coupled with the Functionally Assembled Terrestrial Ecosystem Simulator (FATES; @Koven:2020) (hereafter, ELM-FATES).
 ELM is a land-model that, among many features, simulates the physics and conservative dynamics of water, energy and carbon fluxes.
@@
 $\mathrm{\Psi}$~soil~ [@Kupers:2019].
-See SI for details.
+See SI.
 
@@
 
-We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in the water balance equation, namely, (1) evapotranspiration ET from the flux tower by the 50 ha plot (2012-2017), (2) local stream discharge (2012-2018; see @Paton.2019) and (3) soil volumetric water content (VWC) from two sources: (i) a long-term (2012-2018) record of VWC averaged across three vertical time domain reflectometry (TDR) probes over depth 0-15 cm from three locations near the flux tower, and (ii) plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 at depths of 0.15, 0.4 and 1 m (1299 samples covering all soil types and habitats (@Kupers:2019).
+We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in the water balance equation, namely, (1) evapotranspiration ET from the flux tower by the 50 ha plot (2012-2017), (2) local stream discharge (2012-2018; see @Paton.2019) and (3) soil volumetric water content (VWC) from two sources: (i) a long-term (2012-2018) record of VWC averaged across three vertical time domain reflectometry (TDR) probes over depth 0-15 cm from three locations near the flux tower, and (ii) plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 at depths of 0.15, 0.4 and 1 m (1299 samples covering all soil types and habitats; @Kupers:2019).
 
 \
- For ELM-FATES calibration we calculated an objective function for each of the 5000-member ensemble by equally weighting standardized RMSE between observations and simulations across all fluxes and states mentioned above, and then identified 100 parameter ensemble members that minimised the objective function, ensuring that soil moisture dynamics by depth is captured correctly.
-See SI for details and Table S1 for the ranges of best-fit values for different parameters.
+ For ELM-FATES calibration we calculated an objective function for each of the 5000-member ensemble by equally weighting standardized root mean square error (RMSE) between observations and simulations across all fluxes and states mentioned above, and then identified 100 parameter ensemble members that minimised the objective function, ensuring that soil moisture dynamics by depth is captured correctly (see SI).
+See Table S1 for the ranges of best-fit values for different parameters.
 
@@
 
-We ran ELM-FATES with the best-fit 100 ensemble members from 1985-2018, with the first 5 years used for model spin-up.
+We ran ELM-FATES with the best-fit 100 ensemble members from 1985-2018, with the first five years used for model spin-up.
 Extreme hydrological droughts were identified by depth $z$ as days for which $\mathrm{\Psi}_{\textrm{soil},z}$ was more negative than the 5^th^ percentile of $\mathrm{\Psi}_{\textrm{soil},z}$ for a given day of the year.
@@
 As root water-uptake is generally a non-fractionating process, tree $\mathrm{\delta}$^2^H~xylem~ reflects the signature of source water.
-Given a vertical gradient of $\mathrm{\delta}$^2^H in soil and groundwater $\mathrm{\delta}$^2^H~xylem~ provides an index of rooting depth [@Dawson.1991].
+Given a vertical gradient of $\mathrm{\delta}$^2^H in soil and groundwater, $\mathrm{\delta}$^2^H~xylem~ provides an index of rooting depth [@Dawson.1991].
 We leveraged $\mathrm{\delta}$^2^H~xylem~ from BCI [@Meinzer:2001] for the dry-season of March 1997 as this period showed largest seasonal divergence in $\mathrm{\delta}$^2^H~xylem~ among species and vertically in soil and groundwater $\mathrm{\delta}$^2^H at natural abundance level (Fig.
 S3).
-Analyzing $\mathrm{\delta}$^2^H~xylem~ for 12 tree species @Meinzer:1999 found that four species sourced water deeper than 1 m, two other at least 0.8 m deep, one between 0 to 0.2 m, and five in the 0.2 to 0.7 m depth range over which the relationship between soil depth and $\mathrm{\delta}$^2^H~soil~ was variable.
 
@@
 For comparison with modelled ERD, we removed six species from the @Meinzer:1999 dataset to account for the uncertainty in their water-sourcing depths.
-One species, *Guapira standleyana*, was removed because it's $\mathrm{\delta}$^2^H value of -33.5 (SE $\pm$ 2.6) was particularly difficult to assign a source-water depth given the uncertainty in soil water $\mathrm{\delta}$^2^H data by depth (see Fig S3).
 $\mathrm{\delta}$^2^H~xylem~ from leafless trees may not be linked with water sourced at the time of measurement and thus may not be comparable to species that had leaves.
@@
 We therefore removed five species that are typically leafless in March-April (Joseph Wright, personal observation), that is, the months of $\mathrm{\delta}$^2^H~xylem~ sampling.
+$\mathrm{\delta}^2H$ to depth relationship was particularly uncertain for $\mathrm{\delta}^2H$ > -40 ‰, so from the remaining dataset, we removed one species, *Guapira standleyana*, with $\mathrm{\delta}$^2^H~xylem~ of -28.9 ‰ $\\pm$ 3.7SE (see Fig.
+S3).
 For each model of ERD (Eq.
@@
 Because extreme, sustained water-stress increases risk of mortality via hydraulic-failure and/or carbon starvation, deeper ERD may mitigate drought-induced mortality by limiting exposure to water-stress [@Rowland:2015; @McDowell:2018].
-We thus demonstrate, for the first time, a link between species trade-offs in above- and below-ground hydraulic traits, drought exposure through extreme hydrological droughts quantified over the whole soil-column across several El-Niño events, and large ($\ge$ 10 cm DBH) tree mortality over 35 years.
+We thus demonstrate, for the first time, a link between species trade-offs in above- and below-ground hydraulic traits, drought exposure through extreme hydrological droughts quantified over the whole soil-column, and large ($\ge$ 10 cm DBH) tree mortality across several El-Niño events over 35 years.
 
@@
 \
-We found significant ERD-mortality relationship during extreme water-stress for evergreen but not deciduous species (Fig.
+We found significant ERD-mortality relationships during extreme water-stress for evergreen but not deciduous species (Fig.
 \@ref(fig:erd-mort-census-evg), S11); consistent with the expectation that deciduous species can also avoid water-stress via leaf-drop.
@@
 Studies assessing coordination between leaf phenology, rooting depth and hydraulic traits simultaneously are almost absent for the tropics and warrant future consideration [@Oliviera.2021].
-Our analysis for ERD-hydraulic traits relationship was limited to only seven species, but covered all species of different leaf phonologies (Fig.
-\@ref(fig:erd-hyd)), so here we assume that the trend in the ERD-hydraulic trait relationships holds true across all species of different leaf phonologies.
+Our analyses for ERD-hydraulic traits relationships were limited to only seven species, but covered all species of different leaf phenologies (Fig.
+\@ref(fig:erd-hyd)), so here we assume that the trend in the ERD-hydraulic trait relationships holds true across all species of different leaf phenologies.
 We also found that the range of ERDs overlapped across leaf phenology, but evergreen species tend to have deeper ERDs (Fig.
@@
 \
-To estimate topographic variation in ERD within or across species, future studies may use soil moisture dynamics from a distributed hydrological model [e.g.
+As in CT2018, with water-availability resolved for a 1-D column, we interpret our ERD estimates as revealing relative differences among species’ effective rooting depths rather than absolute depths.
+To estimate absolute depths and topographic variation in ERD within or across species, future studies may use soil moisture dynamics from a distributed hydrological model [e.g.
 @Schwantes:2018].
@@
 $R$^2^ and significance levels for linear model fits are given in panel insets.
-Model fits (blue lines) with confidence bands (gray area) are only shown for periods with significant fits at $\alpha$ = 0.05, identified by census-intervals with a superscript *.
+Model fits (blue lines) with confidence bands (gray area) are only shown for periods with significant fits at $\alpha$ = 0.05 or $\alpha$ = 0.1 (1982-85), identified by census-intervals with a superscript *.
 
@@
 Figure \@ref(fig:erd-time-evg): Modeled effective rooting depth (ERD; horizontal-axis) versus time spent beyond critical hydraulic threshold (vertical axis) by census interval (colored bars) for `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))` evergreen species included in the mortality analyses (Fig.
-\@ref(fig:erd-mort-evg)).
+\@ref(fig:erd-mort-census-evg)).
 Each bar represents the average time species of the same ERD spent beyond species-specific critical hydraulic thresholds in a given interval, that is, the proportion of days for which $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ was more negative than species $\mathrm{\Psi}$~crit~, defined as $\mathrm{\Psi}$~20,leaf~, and where $z$ is the soil depth matching species ERD.

### 2021-04-09 10.0_Supporting_Information.Rmd
Explained uncertainty in isotopic data

@@
 ```{r iso-data, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-iso-data}Stable isotopic ratio ($\\mathrm{\\delta}^2H$) of soil water by depth measured at Barro Colorado Island in March 1997.
-Data from @Meinzer:1999."}
+Data from @Meinzer:1999.
+$\\mathrm{\\delta}^2H$ to depth relationship was particularly uncertain for $\\mathrm{\\delta}^2H$ values > -40 ‰.
+Two species, *Guapira standleyana* (-28.9 $\\pm$ 3.7; mean $\\pm$ SE) and *Spondias radlkoferi* (-33.5 $\\pm$ 2.6), in @Meinzer:1999 dataset had $\\mathrm{\\delta}$^2^H~xylem~ values > -40 ‰.
+For example, *Guapira standleyana*, with $\\mathrm{\\delta}$^2^H value of -28.9 $\\pm$ 3.7 could have a source-water depth anywhere between 10 to 50 cm.
+*Spondias radlkoferi* is also typically leafless in March-April.
+These two species were not included in the ERD model-data comparison."}
 knitr::include_graphics(file.path("figures/UDI_confidence/Meinzer_etal_1999_Depth_vs_soil_deltaD.jpeg"))
@@
 However, AIC does not ensure that fitted values will have a positive trend with growth.
-We therefore used $R$^2^ between observed and fitted growth values as a goodness-of-fit measure and retained depths associated with a relatively high threshold $R$^2^ value of 0.5; after filtering out depths associated with negative correlations $r$ if any.
+We therefore used $R$^2^ between observed and fitted growth values as a goodness-of-fit measure and retained depths associated with a relatively high threshold $R$^2^ value of 0.5; after filtering out depths associated with negative correlations $r$, if any.
 Finally, among these we chose the depth with the highest $R$^2^ per *h*.
@@
 
+Deciduous leaf-habits are generally expected to be fully deciduous at individual tree-level.
+At BCI, however, there is a great within-species variation in the timing of leaf-drop and leaf-out, and neither do all individuals of a deciduous species reach full deciduousness every year (Condit et al., 2000).
+Our reconstruction of LAI seasonality is based upon species-level leaf-fall data, which captures the within-species variation in the timing of leaf-drop, so in our reconstruction seasonal decline in $\textrm{LAI}^*$ appears to be more spread out and does not reach zero, in contrast to what may be  expected for an idealised individual of each deciduous leaf habit.
+Greater uncertainty in our estimates of leaf-lifespan for deciduous species may have compounded this issue further.
+We chose not to normalise the deciduous species’ $\textrm{LAI}^*$ curves to reach zero as it would further distort the shape of the curve, shifting a larger period under a lower $\textrm{LAI}^*$.
+
 Although deciduous leaf-habits are generally expected to be fully deciduous, at BCI there is a great within-species variation in the timing of leaf-drop and leaf-out, and neither do all individuals of a deciduous species reach full deciduousness every year [@Condit.2000].

### 2021-04-09 09.0_manuscript.rmd
Added grid lines, added more axis breaks

@@
 We therefore removed five species that are typically leafless in March-April (Joseph Wright, personal observation), that is, the months of $\mathrm{\delta}$^2^H~xylem~ sampling.
-$\mathrm{\delta}^2H$ to depth relationship was particularly uncertain for $\mathrm{\delta}^2H$ > -40 ‰, so from the remaining dataset, we removed one species, *Guapira standleyana*, with $\mathrm{\delta}$^2^H~xylem~ of -28.9 ‰ $\\pm$ 3.7SE (see Fig.
+$\mathrm{\delta}^2H$ to depth relationship was particularly uncertain for $\mathrm{\delta}^2H$ > -40 ‰, so from the remaining dataset, we removed one species, *Guapira standleyana*, with $\mathrm{\delta}$^2^H~xylem~ of -28.9 ‰ $\pm$ 3.7SE (see Fig.
 S3).

### 2021-04-09 code/03.0_Preparing_isotopic_data.R

@@ -17,7 +17,7 @@ gc()
 #### Written with R version 4 ###
 #*******************************************
 if (!require("groundhog")) install.packages("groundhog"); library(groundhog)
-groundhog.day = "2020-05-01"
+groundhog.day = "2021-01-01"
 pkgs=c('tidyverse', 'scales', 'ggpmisc')
 groundhog.library(pkgs, groundhog.day)
 
@@ -97,6 +97,7 @@ iso <- iso %>%
   unite("deci_sp", deciduous, sp, remove = FALSE) %>%
   mutate(deci_sp.plot = factor(deci_sp, levels=unique(deci_sp[order(Xylem_sap_deltaD_permil)]), ordered=TRUE))
 
+xylem.label <- expression(delta^2*H[xylem]~"( \u2030)")
 g1 <- ggplot(iso %>% subset(!is.na(Xylem_sap_deltaD_permil)),
              aes(x = deci_sp.plot, y = Xylem_sap_deltaD_permil)) +
   geom_col(aes(fill = deciduousness)) +
@@ -104,7 +105,9 @@ g1 <- ggplot(iso %>% subset(!is.na(Xylem_sap_deltaD_permil)),
   theme(legend.position = "top") +
   ylab(xylem.label) + xlab("Species") +
   geom_errorbar(aes(ymax = Xylem_sap_deltaD_permil + se, ymin = Xylem_sap_deltaD_permil - se), width = 0.2, size = 0.5) +
-  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
+  scale_y_continuous(breaks = seq(from = -60, to = 0, by = 5)) +
+  theme(panel.grid.major = element_line(colour = "gray", size = 0.1))
 ggsave(file.path(paste0("data-raw/traits/isotopes/Meinzer1999_Table1_Xylem_Sap_deltaD_Fig4.jpeg")),
        plot = g1, height = 5, width = 6, units='in')
 g2 <- g1 + coord_flip() +
@@ -150,8 +153,11 @@ depth.m1.label2 = paste0("R2 = ", round(summ.depth.m1$r.squared, 2),
 formula <- 'y ~ x'
 g1 <- ggplot(iso.soil.1, aes(y = depth, x = soil.deltaD)) +
   xlab(soil.label) + ylab("Depth (cm)") +
-  scale_y_reverse() +
-  xlim(-60,0) +  theme(text = element_text(size = 20)) +
[…]

### 2021-04-09 09.0_manuscript.rmd
Typo corrections and other minor edits

@@
 Summary | 200  
-Main Text  | 6939  
-Introduction | 832  
+Main Text  | 6934  
+Introduction | 831    
 Methods |  3150  
-Results | 855  
-Discussion | 2102  
+Results | 838  
+Discussion | 2115  
 Keywords | 8
@@
 Tree species on BCI are scored by expert botanists as one among four leaf-habits; evergreen, brevideciduous, facultative deciduous and obligate deciduous (@Meakem.2018; see SI).
-We pooled all deciduous leaf habits together (hereafter 'deciduous' for brevity) and regressed species ERD against species $\textrm{M}_{s,t}$ for each census interval $t$ for deciduous ($n$ = `r nrow(mrate.depth.mean %>% subset(deciduous != "E"))`) and evergreen species ($n$ = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`) separately.
+We pooled all deciduous leaf habits together (hereafter, 'deciduous' for brevity) and regressed species ERD against species $\textrm{M}_{s,t}$ for each census interval $t$ for deciduous ($n$ = `r nrow(mrate.depth.mean %>% subset(deciduous != "E"))`) and evergreen species ($n$ = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`) separately.
 
@@
 S4, S5).
-The distribution of maximum depth of soil water dynamics, and thus ecosystem root zone depth, across the 100 hydrological realisations encompassed 95% CI of `r root.zone.CI1` m - `r root.zone.CI2` m with the median at `r root.zone.med` m.
+The distribution of maximum depth of soil water dynamics, and thus ecosystem root zone depth, across the 100 hydrological realisations encompassed 95% CI of `r root.zone.CI1` m - `r root.zone.CI2` m (median at `r root.zone.med` m).
 
 \
-Soil water potential dynamics $\mathrm{\Psi}_{\textrm{soil},z}$ by depth z from 0.01 through 13 m obtained from the best-fit simulations revealed that every 5-yr census interval had at least one extreme hydrological drought year, but varied in terms of number of extreme years, drought intensity, drought seasonality and duration (Fig.
+Soil water potential dynamics $\mathrm{\Psi}_{\textrm{soil},z}$ by depth z from 0.01 through 13 m obtained from the best-fit simulations revealed that every 5-yr census interval had at least one extreme hydrological drought year, but each interval varied in terms of number of extreme years, drought intensity, drought seasonality and duration (Fig.
 \@ref(fig:swp-droughts)).
@@
 
-The best ERD model explained a large fraction of the variance in $\mathrm{\delta}$^2^H~xylem~ ($R$^2^ = `r lm.erd.iso.r2`, $p$ = `r lm.erd.iso.p`, $n$ = `r length(unique(erd.iso$sp))`, Fig.
+The best ERD model (Eq.
+(3)) explained a large fraction of the variance in $\mathrm{\delta}$^2^H~xylem~ ($R$^2^ = `r lm.erd.iso.r2`, $p$ = `r lm.erd.iso.p`, $n$ = `r length(unique(erd.iso$sp))`, Fig.
 \@ref(fig:erd-iso); also see SI).
@@
 Among the seven census-intervals over 1982-2015 for which we analysed relationship of ERD with mortality rates, six intervals were associated with occurrence of one or more El-Niño event [@Detto.2018; @Condit.2017].
-Out of the six intervals with El-Niño events ERD explained 30-40% variation in mortality rates among evergreen species during five intervals--1982-1985, 1985-1990, 1995-2000, 2000-2005 and 2005-2010--such that species mortality rates decreased with deeper ERD (p-values < 0.05 for four intervals and 0.06 for one interval; Fig.
+Out of the six intervals with El-Niño events ERD explained 30-40% variation in mortality rates among evergreen species during five intervals--1982-1985, 1985-1990, 1995-2000, 2000-2005 and 2005-2010--such that species mortality rates decreased with deeper ERD ($p$-values < 0.05 for four intervals and 0.06 for one interval; Fig.
 \@ref(fig:erd-mort-census-evg)).
@@
 
-We introduce a novel approach for estimating ERD using 25 years of tree growth, species-specific leaf vulnerability curves, modeled soil water potential profiles, and observed VPD (\@ref(eq:M-chosen), Figs.
+We introduce a novel approach for estimating ERD using 25 years of tree growth, species-specific leaf vulnerability curves, modeled soil water potential profiles, and observed VPD (Eq.
+\@ref(eq:M-chosen), Figs.
 \@ref(fig:schematic), \@ref(fig:sp-erd)).
-Our predictions of ERD were consistent with estimates using $\mathrm{\delta}$^2^H~xylem~ (Figs.
+Our predictions of ERD were consistent with estimates using $\mathrm{\delta}$^2^H~xylem~ (Fig.
 \@ref(fig:erd-iso)).
@@
 Our findings suggest that species with investment in deep roots can afford the hydraulically efficient, but risky suite of traits (Fig.
-\@ref(fig:erd-hyd)), because access to a reliable deep-water resource ensures that for them hydraulic risk is not realised (Figs.
-\@ref(fig:erd-time-evg)), whereas shallow rooted species pay the cost of hydraulic safety in terms of efficiency, adapted for an environment in which hydraulic risk is significant, as extreme droughts cause exponentially greater water-stress in shallow soil layers (Fig.
-S6).
+\@ref(fig:erd-hyd)), because access to a reliable deep-water resource ensures that for them hydraulic risk is not realised (Fig.
+\@ref(fig:erd-time-evg)), whereas shallow rooted species pay the cost of hydraulic safety in terms of efficiency, adapted for an environment in which hydraulic risk is significant, as extreme droughts cause exponentially greater water-stress in shallow soil layers (Figs.
+Figs.
+\@ref(fig:erd-time-evg), S6).
 
@@
 Hydraulic risk for shallow rooted species may have been exacerbated in our 35-yr study period in which extreme water-deficit years were more frequent than those in the last century [@Condit.2017].
-Greater survival of deep-rooted species that we observed may change into the future if droughts intensify.
+Greater survival of deep-rooted species that we observed may not continue into the future if droughts intensify.
 
@@
 \@ref(fig:erd-mort-census-evg), S11); consistent with the expectation that deciduous species can also avoid water-stress via leaf-drop.
-This is also consistent with the observation that species distribution along local (BCI) and regional (Panama) moisture gradients is correlated with $\mathrm{\Psi}$~tlp~ for evergreen but not deciduous species [@Kunert.2021].
-Studies assessing coordination between leaf phenology, rooting depth and hydraulic traits simultaneously are almost absent for the tropics and warrant future consideration [@Oliviera.2021].
+This is also consistent with the observation that species distributions along local (BCI) and regional (Panama) moisture gradients are correlated with $\mathrm{\Psi}$~tlp~ for evergreen but not deciduous species [@Kunert.2021].
+Studies that simultaneously assess coordination between leaf phenology, rooting depth and hydraulic traits are almost absent for the tropics and warrant future consideration [@Oliviera.2021].
 Our analyses for ERD-hydraulic traits relationships were limited to only seven species, but covered all species of different leaf phenologies (Fig.
@@
 Whether deciduous species also have more efficient and vulnerable hydraulics at BCI as is observed elsewhere remains to be studied [@Markesteijn:2010; @Markesteijn:2011; @Xu:2016; @Gleason.2016].
-Leaf phenologies at BCI are numerous and complicated and warrant research.
+Leaf phenologies at BCI are numerous and complicated and warrant further research.
 
@@
 \
-Our results are consistent with recent studies that analysed rooting or water-sourcing depths versus hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016; also see @Brum.2019; @Brum.2017; @Rowland:2015].
+Our results are consistent with recent studies that analysed rooting or water-sourcing depths versus hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016; also see, @Brum.2019; @Brum.2017; @Rowland:2015].
 Globally, large trees tend to exhibit greater growth reductions, lower post-drought resilience, and greater increases in mortality relative to their understory counterparts [@Bennett:2015; @Phillips:2010].
@@
 Our result of lower mortality in deep-rooted trees contrasts with the inverse model finding of @Chitra-Tarak.2018 (hereafter, CT2018) in which deeper ERD species in a South-Asian seasonally dry tropical forest had higher mortality in a rare, prolonged drought.
-The hydrological model in CT2018 revealed that the multi-year drought exhausted the deep soil and even bedrock water availability (also see [@Goulden.2019; @Ivanov:2012]).
-In contrast, our hydrological modeling at BCI finds that $\mathrm{\Psi}$~soil~ for depths deeper than 1 m did not cross $\mathrm{\Psi}$~crit~ of the most sensitive tree species (-0.28 MPa) for all the dry seasons and droughts during 1990-2015.
-Mean annual rainfall of 1095 mm in CT2018 compared to `r round(rain.man.stats$rain)` mm at BCI, and precipitation to potential evapotranspiration ratio of nearly one in CT2018 compared to nearly two in BCI are major factors in the different mortality responses.
-At BCI deep soil layers were recharged annually (Figs.
-S6, \@ref(fig:swp-droughts)), whereas in CT2018 they were not.
+The hydrological model in CT2018 revealed that the multi-year drought exhausted the deep soil and even bedrock water availability (also see, @Goulden.2019; @Ivanov:2012).
+In contrast, our hydrological modeling at BCI finds that $\mathrm{\Psi}$~soil~ for depths deeper than 2.9 m did not cross $\mathrm{\Psi}$~crit~ of the most sensitive tree species (`r min.psi_crit` MPa) for all the dry seasons and droughts during 1990-2015 (see SI).
+Mean annual rainfall of 1095 mm in CT2018 compared to `r round(rain.man.stats$rain)` mm at BCI, and precipitation to potential evapotranspiration ratio of nearly one in CT2018 compared to nearly two in BCI, are major factors in the different mortality responses.
+At BCI deep soil layers were recharged annually (Fig.
+S6), whereas in CT2018 they were not.
 The contrast between the two studies highlights the combined role of seasonal precipitation input and site-specific hydrology in modulating mortality-risk of deep-rooted species.
@@
 To assimilate ERDs, DGVMs could vary rooting parameters such that the centroid of the species water-uptake profiles match ERDs.
-We found that LMA and WSG were strong predictors of leaf hydraulic vulnerability curves.
+We found that LMA and WSG were strong predictors of leaf hydraulic vulnerability curves (Fig.
+\@ref(fig:ab-fit)).
 Albeit future studies should undertake sensitivity analyses for uncertainties involved, our finding offers the promise of a greater ability to parameterize the 'hard' hydraulic traits with the abundant 'soft' trait data, thus allowing for a better representation of forest hydrodynamics.
@@
 S9).
-Our interpretation of this result is that VPD and leaf hydraulic vulnerability curves may be adequate to explain inter-census differences in growth (via stomatal control), but also acknowledge that our estimates of seasonality of LAI is a tentative estimate that combines leaf-fall data and the timing of leaf-gain backtracked from a leaf lifetime, and omits inter-annual variation.
-Species-level leaf-fall data from litter-traps that we use includes large within-species variability in leaf-fall timing, and thus does not capture typical tree-level deciduous seasonality.
+Our interpretation of this result is that VPD and leaf hydraulic vulnerability curves may be adequate to explain inter-census differences in growth (via stomatal control), but also acknowledge that our estimates of seasonality of LAI is a tentative estimate that combines leaf-fall data and the timing of leaf-gain backtracked from leaf lifetime, and omits inter-annual variation.
+Species-level leaf-fall data from litter-traps that we used includes large within-species variability in leaf-fall timing, and so may not have captured  tree-level seasonality in deciduousness (see SI).
 Future studies may improve models of LAI seasonality and make use of data from new technologies such as drone based monitoring of LAI in species-rich tropical forests [@Park.2019].

### 2021-04-09 10.0_Supporting_Information.Rmd
Typo corrections and other minor edits

@@
 
-# Additional results on exposure to water-stress
+## Additional results on exposure to water-stress
 
 Overall, species with ERD deeper than 2.9 m did not cross species $\mathrm{\Psi}$~crit~ for the entire 25-yr period for which $\mathrm{\Psi}_{\textrm{soil},z}$ was modeled (1990-2015).
-Soil depths deeper than 1.7 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and drought periods, which is above $\mathrm{\Psi}$~crit~ of the most sensitive tree species (`r min.psi_crit` MPa).
+Soil depths deeper than 2.9 m remained above `r round(psi.2.9.min$median, 2)` MPa (`r round(as.numeric(psi.2.9.min$q97.5), 2)`, `r round(as.numeric(psi.2.9.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and drought periods, which is above $\mathrm{\Psi}$~crit~ of the most sensitive tree species (`r min.psi_crit` MPa).
 

### 2021-04-13 09.0_manuscript.rmd
Minor corrections

@@
 Summary | 200  
-Main Text  | 6934  
+Main Text  | 6933  
 Introduction | 831    
@@
 As soil layers in the hydrological model are discretely resolved into exponentially increasing depths, so is the ERD axis.
-Deeper depths are shown with darker blue colors.
+Species are color coded by leaf habit.
 
@@
 Figure \@ref(fig:erd-hyd): Modeled effective rooting depth (ERD) versus hydraulic properties for seven canopy species found on Barro Colorado Island (Panama).
-Spearman's $r$ and significance levels.
+Spearman's $r$ and significance levels are given in panel insets.
 Linear model fits (blue lines) with confidence bands (gray area) are shown for significant fits at $\alpha$ = 0.05.
@@
 S10 for the full correlation matrix.
+Species are color coded by leaf habit.
+Multiple species in Facultative Deciduous leaf-habit are distinguished by shapes.
 
@@
 $R$^2^ and significance levels for linear model fits are given in panel insets.
-Model fits (blue lines) with confidence bands (gray area) are only shown for periods with significant fits at $\alpha$ = 0.05 or $\alpha$ = 0.1 (1982-85), identified by census-intervals with a superscript *.
+Model fits (blue lines) with confidence bands (gray area) are only shown for periods with significant fits.
+Census intervals, for which fits are significant at $\alpha$ = 0.05 are identified by superscript $**$, while those significant at $\alpha$ = 0.1 by superscript $*$.
 

### 2021-04-13 Readme.Rmd
Minor corrections

@@
 
-Set-up and run effective rooting depth models, evaluate model, and join ERD from the predicted model with hydraulic traits and mortality data, and calculated realized hydraulic risk
+Set-up and run effective rooting depth models, evaluate model, and join ERD from the predicted model with hydraulic traits and mortality data, and calculate realized hydraulic risk
 
@@
 ```{r Chunk 10.0, eval = FALSE}
-# render("10.0_Supporting_Information.Rmd", output_format = "bookdown::html_document2")
-render("10.0_Supporting_Information.Rmd", output_format = "bookdown::word_document2")
+# rmarkdown::render("10.0_Supporting_Information.Rmd", output_format = "bookdown::html_document2")
+rmarkdown::render("10.0_Supporting_Information.Rmd", output_format = "bookdown::word_document2")
 ```

### 2021-04-13 code/08.0_manuscript_data_figs.R

@@ -153,6 +153,9 @@ plot.psi.stat.7.interval.q5.depth <- plot.psi.stat.7.interval.q5.depth.base +
 ggsave("psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q5_by_depth.jpeg",
        plot = plot.psi.stat.7.interval.q5.depth, file.path(figures.folder), device = "jpeg", height = 5, width = 8, units='in')
 
+ggsave("psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q5_by_depth.tiff",
+       plot = plot.psi.stat.7.interval.q5.depth, file.path(figures.folder), device = "tiff", height = 5, width = 8, units='in')
+
 plot.psi.stat.7.interval.q5.depth.freq.base <-
   ggplot(subset(psi.stat.4.select.freq, depth %in% c(0.5, 1, 1.7) &
                   interval.yrs != "(Missing)") %>% droplevels()) +
@@ -571,7 +574,8 @@ depth.traits.select.plot <- ggplot(erd.stem.traits.only.lab,
                                             label = sprintf('italic(p)~"="~%.2f', p),
                                             group = trait.plot), parse = TRUE, vjust = "inward", hjust = "inward") +
   geom_errorbarh(aes(xmax = depth + depth.se, xmin = depth - depth.se), size = 0.2) +
-  geom_point(shape = 21, color = "white", aes(fill = deciduousness), alpha = 1, size = 2.5) +
+  geom_point(color = "white", aes(fill = deciduousness, shape = sp), alpha = 1, size = 2.5) +
+  scale_shape_manual(values = c(21, 22, 23, 21, 24, 21, 25)) +
   coord_cartesian(xlim = c(0, max(erd.stem.traits.only.lab$depth) + 0.5)) +
   xlab("Effective Rooting Depth (m)") + ylab("") +
   facet_wrap(trait.plot ~ ., scales = "free_y", labeller = label_parsed,
@@ -579,9 +583,11 @@ depth.traits.select.plot <- ggplot(erd.stem.traits.only.lab,
   theme(strip.placement = "outside", panel.spacing.x = unit(0, "lines"),
         strip.text.y.left = element_text(size = 10, angle = 90, vjust = -1),
         plot.margin = margin(0.2, 0.2, 0.2, 0.2, "cm")) +
[…]

### 2021-04-13 09.0_manuscript.rmd
Added a reference

@@
 Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014].
-Rainfall at BCI is seasonal with a mean annual total of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a pronounced dry season from mid-December through April with less than 100 mm of rainfall per month [@Paton.2020].
+Rainfall at BCI is seasonal with a mean annual total of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a pronounced dry season from mid-December through April with less than 100 mm of rainfall per month [@Paton.2020; Paton@2019].
 In the 50ha (1000 m × 500 m) old growth forest plot, all stems $\ge$ 1 cm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every five year through 1985-2015 for growth, mortality as well as recruitment of new stems into the 1 cm dbh size class [@Condit.2017].

### 2021-05-03 09.0_manuscript.rmd
Added/corrected location of groundhog library

@@
 if (!require("groundhog")) install.packages("groundhog");
-groundhog.folder <- paste0("Users/ftuser/Library/R/4.0/library/bci.roots.inversion")
+groundhog.folder <- paste0("groundhog.library")
 if(!dir.exists(file.path(groundhog.folder))) {dir.create(file.path(groundhog.folder))}

### 2021-05-03 Readme.Rmd
Added/corrected location of groundhog library

@@
 library(groundhog)
-groundhog.folder <- paste0("Users/ftuser/Library/R/4.0/library/bci.roots.inversion")
+groundhog.folder <- paste0("groundhog.library")
 if(!dir.exists(file.path(groundhog.folder))) {dir.create(file.path(groundhog.folder))}

### 2021-05-03 code/01.0_Preparing_stem_growth_time_series.R

@@ -14,7 +14,10 @@ gc()
 #### Written with R version 4 ###
 #*******************************************
 if (!require("groundhog")) install.packages("groundhog"); library(groundhog)
-groundhog.day = "2020-05-01"
+groundhog.folder <- paste0("groundhog.library")
+if(!dir.exists(file.path(groundhog.folder))) {dir.create(file.path(groundhog.folder))}
+set.groundhog.folder(groundhog.folder)
+groundhog.day = "2021-01-01" # "2020-05-01"
 pkgs=c('tidyverse', 'purrrlyr', 'scales', 'grid', 'gridExtra', 'lme4')
 groundhog.library(pkgs, groundhog.day)
 

### 2021-05-03 code/02.0_demographic_data_BCI.R

@@ -13,7 +13,10 @@ gc()
 #### Written with R version 4 ###
 #*******************************************
 if (!require("groundhog")) install.packages("groundhog"); library(groundhog)
-groundhog.day = "2020-05-01"
+groundhog.folder <- paste0("groundhog.library")
+if(!dir.exists(file.path(groundhog.folder))) {dir.create(file.path(groundhog.folder))}
+set.groundhog.folder(groundhog.folder)
+groundhog.day = "2021-01-01"
 pkgs=c('tidyverse', 'hms', 'ggpmisc')
 groundhog.library(pkgs, groundhog.day)
 

### 2021-05-03 code/03.0_Preparing_isotopic_data.R

@@ -17,6 +17,9 @@ gc()
 #### Written with R version 4 ###
 #*******************************************
 if (!require("groundhog")) install.packages("groundhog"); library(groundhog)
+groundhog.folder <- paste0("groundhog.library")
+if(!dir.exists(file.path(groundhog.folder))) {dir.create(file.path(groundhog.folder))}
+set.groundhog.folder(groundhog.folder)
 groundhog.day = "2021-01-01"
 pkgs=c('tidyverse', 'scales', 'ggpmisc')
 groundhog.library(pkgs, groundhog.day)

### 2021-05-03 code/04.0_distribution_qqplots_growth_rate_inverse_model_residuals_UDI.R

@@ -14,7 +14,10 @@ gc()
 #### Written  with R version 4 ###
 #*******************************************
 if (!require("groundhog")) install.packages("groundhog"); library(groundhog)
-groundhog.day = "2020-05-01"
+groundhog.folder <- paste0("groundhog.library")
+if(!dir.exists(file.path(groundhog.folder))) {dir.create(file.path(groundhog.folder))}
+set.groundhog.folder(groundhog.folder)
+groundhog.day = "2021-01-01"
 pkgs=c('tidyverse', 'gridExtra', 'spatstat', 'MASS')
 groundhog.library(pkgs, groundhog.day)
 library(bci.elm.fates.hydrology)

### 2021-05-03 code/06.0_load.R

@@ -13,12 +13,11 @@
 # #*******************************************
 # #### Written  with R version 4.0.3 ###
 # #*******************************************
-if (!require("groundhog")) install.packages("groundhog");
-groundhog.folder <- paste0("Users/ftuser/Library/R/4.0/library/bci.roots.inversion")
+if (!require("groundhog")) install.packages("groundhog"); library(groundhog)
+groundhog.folder <- paste0("groundhog.library")
 if(!dir.exists(file.path(groundhog.folder))) {dir.create(file.path(groundhog.folder))}
 set.groundhog.folder(groundhog.folder)
-library(groundhog)
-# set.groundhog.folder('/Users/ftuser/Library/R/3.6/library')
+
 groundhog.day = "2021-01-01"
 groundhog.library('devtools', groundhog.day)
 pkgs=c('lattice', 'magick', 'cowplot', 'corrplot', 'ggcorrplot', 'tidyverse', 'readxl',

### 2021-05-03 code/07.0_PhenoDemoTraitsPsi.R

@@ -22,11 +22,10 @@ source("code/06.0_load.R")
 ####   Load Libraries, Prep for graphics, folders  ##
 #*******************************************
 
-if (!require("groundhog")) install.packages("groundhog");
-groundhog.folder <- paste0("Users/ftuser/Library/R/4.0/library/bci.roots.inversion")
+if (!require("groundhog")) install.packages("groundhog"); library(groundhog)
+groundhog.folder <- paste0("groundhog.library")
 if(!dir.exists(file.path(groundhog.folder))) {dir.create(file.path(groundhog.folder))}
 set.groundhog.folder(groundhog.folder)
-library(groundhog)
 groundhog.day = "2021-01-01"
 pkgs=c('tidyverse', 'readxl', 'forcats', 'agricolae', 'gridExtra',
        'scales', 'GGally', 'ggpmisc', 'Evapotranspiration',

### 2021-05-06 09.0_manuscript.rmd
Incorporated Production Managers comments

@@
 
-Chitra-Tarak, Rutuja*^1,2^, Chonggang Xu^1^, Salomón Aguilar^3^, Kristina Anderson-Teixeira^3,4^, Jeff Chambers^5^, Matteo Detto^3,6^, Boris Faybishenko^5^, Rosie A.
-Fisher^7,8^, Ryan Knox^5^, Charlie Koven^5^, Lara Kueppers^5,9^, Nobert Kunert^3,4,10^, Stefan J.
-Kupers^11^, Nathan G.
+Chitra-Tarak, Rutuja*^1,2^, Chonggang Xu^1^, Salomón Aguilar^3^, Kristina J.
+Anderson-Teixeira^3,4^, Jeff Chambers^5^, Matteo Detto^3,6^, Boris Faybishenko^5^, Rosie A.
+Fisher^7,8^, Ryan G.
+Knox^5^, Charles D.
+Koven^5^, Lara M.
+Kueppers^5,9^, Nobert Kunert^3,4,10^, Stefan J.
+Kupers^11^, Nate G.
 McDowell^12,13^, Brent D.
@@
 
-1. Los Alamos National Laboratory, Earth and Environmental Sciences Division (EES-14) 
-MS J495, Los Alamos, NM, USA 87545-1663 
-2. Smithsonian Environmental Research Center, 647 Contees Wharf Road, Edgewater, MD, USA 21037-0028 
-3. Smithsonian Tropical Research Institute, Panama, PA 
-4. Conservation Ecology Center, Smithsonian Conservation Biology Institute, Front Royal, VA, USA 
-5. Lawrence Berkeley National Laboratory, Climate and Ecosystem Sciences Division, Berkeley, CA, USA 
-6. Princeton University, Ecology and Evolutionary Biology, Princeton, NJ, USA 
-7. National Center for Atmospheric Research, Climate and Global Dynamics Division
-Boulder, CO, USA 
-8. Laboratoire Évolution & Diversité Biologique, CNRS:UMR 5174, Université Paul Sabatier, Toulouse, France.
-9. University of California Berkeley, Energy and Resources Group, 310 Barrows Hall #3050
-Berkeley, CA, USA 94720 
-10. Institute of Botany, University of Natural Resources and Life Sciences Vienna, Department of Integrative Biology and Biodiversity Research, Gregor‐Mendel‐Str 33, A‐1190 Wien, Wien, AT 
-11. German Centre for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig, Computational Forest Ecology, Leipzig, Saxony, DE 
-12. Atmospheric Sciences and Global Change Division, Pacific Northwest National Lab, PO Box 999, Richland, WA, USA 99352 
-13. School of Biological Sciences, Washington State University, PO Box 644236, Pullman, WA, USA 99164-4236 
-14. Indo-French Cell for Water Sciences, Indian Institute of Science, Bangalore, Karnataka, IN 
-15. IRD, CNRS, UPS, UMR GET, Toulouse, FR 
-16. INRAE, Institut Agro, UMR SAS, Rennes, FR 
-17. University of California Los Angeles, Ecology and Evolutionary Biology, 612 Charles E. Young Drive South, Los Angeles, CA, USA 90095 
-18. Oak Ridge National Laboratory, Environmental Sciences Division, Oak Ridge, TN, USA 
+1. Los Alamos National Laboratory, Earth and Environmental Sciences Division (EES-14) MS J495, Los Alamos, NM 87545-1663, USA 
+2. Smithsonian Environmental Research Center, 647 Contees Wharf Road, Edgewater, MD 21037-0028, USA 
+3. Smithsonian Tropical Research Institute, Apartado 0843–03092, Balboa, Republic of Panama
+4. Conservation Ecology Center, Smithsonian Conservation Biology Institute, Front Royal, VA 22630, USA 
+5. Lawrence Berkeley National Laboratory, Climate and Ecosystem Sciences Division, Berkeley, CA 94720, USA 
+6. Princeton University, Ecology and Evolutionary Biology, Princeton, NJ, USA 08544
+7. National Center for Atmospheric Research, Climate and Global Dynamics Division Boulder, CO 80305, USA 
+8. Laboratoire Évolution & Diversité Biologique, CNRS:UMR 5174, Université Paul Sabatier, 31062 Toulouse, France 
+9. University of California Berkeley, Energy and Resources Group, 310 Barrows Hall #3050 Berkeley, CA 94720, USA 
+10. Institute of Botany, University of Natural Resources and Life Sciences Vienna, Department of Integrative Biology and Biodiversity Research, Gregor‐Mendel‐Str 33, A‐1190 Wien, Wien, Austria
+11. German Centre for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig, Computational Forest Ecology, 04103 Leipzig, Saxony, Germany 
+12. Atmospheric Sciences and Global Change Division, Pacific Northwest National Lab, PO Box 999, Richland, WA  99352, USA
+13. School of Biological Sciences, Washington State University, PO Box 644236, Pullman, WA 99164-4236, USA 
+14. Indo-French Cell for Water Sciences, Indian Institute of Science, Bangalore 560012, India
+15. IRD, CNRS, UPS, UMR GET, Toulouse, 31700, France
+16. INRAE, Institut Agro, UMR SAS, Rennes, 35042, France
+17. University of California Los Angeles, Ecology and Evolutionary Biology, 612 Charles E. Young Drive South, Los Angeles, CA  90095, USA
+18. Oak Ridge National Laboratory, Environmental Sciences Division, Oak Ridge, TN 37831, USA 
 19. School of Renewable Natural Resources, Louisiana State University Agricultural Center, Baton Rouge, LA 70803, USA
-20. California State University Los Angeles, Biological Sciences Department, Los Angeles, CA, USA
+20. California State University Los Angeles, Biological Sciences Department, Los Angeles, CA 90032, USA
+
+# ORCID
+
+Kristina J. Anderson-Teixeira https://orcid.org/0000-00018461-9713   
+Rutuja Chitra-Tarak https://orcid.org/0000-0001-9714-7524  
+Matteo Detto https://orcid.org/0000-0003-0494-188X  
+Boris Faybishenko https://orcid.org/0000-0003-0085-8499  
+Ryan G Knox https://orcid.org/0000-0003-1140-3350  
+Charles D. Koven https://orcid.org/0000-0002-3367-0065  
+Lara M. Kueppers https://orcid.org/0000-0002-8134-3579   
+Norbert Kunert https://orcid.org/0000-0002-5602-6221   
+Stefan J. Kupers https://orcid.org/0000-0001-8094-1895  
+Nate G. McDowell https://orcid.org/0000-0002-2178-2254  
+Sean M. McMahon https://orcid.org/0000-0001-8302-6908  
+Steven R. Paton https://orcid.org/0000-0003-2035-6699  
+Laurent Ruiz https://orcid.org/0000-0001-5043-282X   
+Lawren Sack https://orcid.org/0000-0002-7009-7202  
+Jeffrey M. Warren https://orcid.org/0000-0002-0680-4697  
+Brett T. Wolfe https://orcid.org/0000-0001-7535-045X  
+Cynthia Wright https://orcid.org/0000-0003-2571-7334  
+S. Joseph Wright https://orcid.org/0000-0003-4260-5676   
+Joseph Zailaa https://orcid.org/0000-0001-9103-190X  
+Chonggang Xu http://orcid.org/0000-0002-0937-5744
+
+# Social Media Accounts
+
+Next-Generation Ecosystem Experiments-Tropics (NGEE-Tropics) Twitter $@$NGEETropics  
+Los Alamos National Laboratory: Facebook $@$LosAlamosNationalLab; Twitter $@$LosAlamosNatLab   
+Smithsonian Environmental Research Center: Facebook $@$smithsonian.serc   
+Smithsonian Tropical Research Institute: Facebook $@$SmithsonianPanama; Twitter $@$stri_panama   
+Smithsonian Forest Global Earth Observatory: Facebook $@$ForestGEO; Twitter $@$forestgeo  
+Oak Ridge National Laboratory: Twitter: $@$ORNL and $@$ORNLBio_Env ; Facebook $@$Oak.Ridge.National.Laboratory
+
+Rutuja Chitra-Tarak Facebook $@$rutuju   
+Joseph Zailaa Twitter: $@$ZailaaJ  
+Norbert Kunert Twitter: $@$KunertNorbert; Facebook $@$norbert.kunert.7/  
+K Anderson-Teixeira Twitter: $@$K_A_Teixeira
 
@@
 Here, we tested the role of deep-water access in enabling survival within a diverse tropical forest community in Panama using a novel data-model approach.
-- We inversely estimated the effective rooting depth (ERD, as the average depth of water extraction), for 32 canopy species by linking diameter growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and leaf hydraulic vulnerability curves.
+- We inversely estimated the effective rooting depth (ERD, as the average depth of water extraction), for 29 canopy species by linking diameter growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and leaf hydraulic vulnerability curves.
 We validated ERD estimates against existing isotopic data of potential water-access depths.
@@
 Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014].
-Rainfall at BCI is seasonal with a mean annual total of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a pronounced dry season from mid-December through April with less than 100 mm of rainfall per month [@Paton.2020; Paton@2019].
+Rainfall at BCI is seasonal with a mean annual total of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a pronounced dry season from mid-December through April with less than 100 mm of rainfall per month [@Paton.2020yr; @Paton.2019mon].
 In the 50ha (1000 m × 500 m) old growth forest plot, all stems $\ge$ 1 cm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every five year through 1985-2015 for growth, mortality as well as recruitment of new stems into the 1 cm dbh size class [@Condit.2017].
 This inventory represents 321 woody species, 28% of which are at least partly dry-season deciduous [@Condit:2000].
-The elevation range in the plot is only 40 m (120-160 m) [@Harms:2001] and soils are homogeneous with red light clays accounting for 72% of the plot [@Baillie.2017].
+The elevation range in the plot is only 40 m (120-160 m) [@Harms:2001] and soils are homogeneous with red light clays accounting for 72% of the plot [@Baillie.2007].
 The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil.
@@
 VPD affects growth non-linearly, with growth increasing with VPD up-to a VPD threshold, then decreasing as leaf pores (stomata) close reducing water-uptake [@Grossiord.2020; @Yang.2019].
-To account for the non-linear impact of VPD on growth, we used predicted gross primary productivity (GPP; hereafter, $\hat{\textrm{VPD}}$) from a locally derived polynomial relationship between GPP and VPD (see SI and Fig.
+To account for the non-linear impact of VPD on growth, we used predicted gross primary productivity (GPP; hereafter, $\hat{\textrm{VPD}}$) from a locally derived polynomial relationship between GPP and VPD (Dataset S1, Fig.
 S1).
@@
 \
-We therefore tested alternate structures of empirical growth models (Eq.
-\@ref(eq:M-chosen) and those in SI), or inverse models of ERD, in which we regressed species-specific growth against multiplicative or additive effects of one or more growth factors calculated daily and averaged over 5-yearly censuses: $\hat{\textrm{VPD}}$, $\textrm{FLC}_\textrm{leaf}$ and the leaf area index (LAI).
+We therefore tested alternate structures of empirical growth models (Methods S1, Eqs.
+S1-S6, including Eq.
+\@ref(eq:M-chosen)), or inverse models of ERD, in which we regressed species-specific growth against multiplicative or additive effects of one or more growth factors calculated daily and averaged over 5-yearly censuses: $\hat{\textrm{VPD}}$, $\textrm{FLC}_\textrm{leaf}$ and the leaf area index (LAI).
 Incorporation of radiation did not improve model fitting, possibly due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season at BCI.
@@
 where | refers to conditionals, n~t~ is the total number of days in census interval $t$, the superscript * indicates that the variable has been standardized to range between 0 and 1 (within species for $\textrm{FLC}$~leaf~), $\beta_{0}$ and $\beta_{1}$ are model coefficients and $\epsilon$ is the model error term.
-See SI for all the alternate models tested.
+See Methods S1 for all the alternate model structures tested (Eqs.
+S1-S6).
 
@@
 We modelled multiple hydrological realisations of soil water-potential dynamics (see below).
-Incorporating this uncertainty, we defined species ERD as the median ($\pm$ SE) of best-fit depths across all hydrological realisations.
-See SI for statistics for identifying best-fit ERD.
+Incorporating this uncertainty, we defined species ERD as the median ($\pm$ SE) of best-fit depths across all hydrological realisations for soil water dynamics (\Psi_{\mathrm{soil,z}}) (Eq.
+S7).
+See Methods S2 for statistics for identifying best-fit ERD.
 
@@
 For diameter growth estimates, to minimize the effect of light variation among trees, we selected only large trees ($\ge$ 30 cm dbh) in the 50 ha plot and also species whose maximum height was $\ge$ 30 m (hereafter, canopy species) and thus are likely to be fully exposed to the sun.
-Calculating individual tree growth rates across six 5-yearly censuses (1990-2015; @Condit.2019a; @Condit.2019b), removing outliers, obtaining residuals from a dbh model of growth to account for the size effect on growth (see SI), we estimated species $s$ growth time series $G_{s}$ (cm yr^-1^) as the median of standardized dbh model residuals, for only those species ($n$ = `r length(erd.sp)`) with complete records for at least three trees (median `r median(g.n$n)`, max `r max(g.n$n)` trees per species).
+Calculating individual tree growth rates across six 5-yearly censuses (1990-2015; @Condit.2019a; @Condit.2019b), removing outliers, obtaining residuals from a dbh model of growth to account for the size effect on growth (Methods S3), we estimated species $s$ growth time series $G_{s}$ (cm yr^-1^) as the median of standardized dbh model residuals, for only those species ($n$ = `r length(erd.sp)`) with complete records for at least three trees (median `r median(g.n$n)`, max `r max(g.n$n)` trees per species).
 
@@
 We obtained leaf hydraulic vulnerability curves ($K_\textrm{leaf}$ vs.
-$\mathrm{\Psi}$~leaf~) for adult trees of 21 common species at BCI from Zailaa *et* *al.* (in prep; see SI for brief description of methods) described as:
+$\mathrm{\Psi}$~leaf~) for adult trees of 21 common species at BCI from J.
+Zailaa *et* *al.* (unpublished; see Dataset S2 for brief description of methods) described as:
 
@@
 We identified scaling relationships between fitted parameters $A$ and $B$ in Eq.
-\@ref(eq:vc-exp) and two traits; namely, WSG, the wood specific gravity, and LMA, the leaf mass per unit area (@Wright:2010; see SI).
+\@ref(eq:vc-exp) and two traits; namely, WSG, the wood specific gravity, and LMA, the leaf mass per unit area (@Wright:2010; see Dataset S3).
 We fitted polynomial equations described as,
@@
 We assumed species-specific mean seasonal curves for LAI (standardised between 0 and 1; unitless), informed by a combination of long-term records for leaf-fall [@Wright.1990aeo] and leaf-lifetime [@Osnas.2018].
-See SI.
+See Dataset S4, Methods S4, Fig S2.
 
@@
 FATES couples to ELM by a common interface of water and carbon fluxes.
-Detailed descriptions for ELM and FATES can be found elsewhere [@Bisht.2018; @Fisher.2010; @Fisher.2014; @Koven:2020].
+Detailed descriptions for ELM and FATES can be found elsewhere [@Bisht.2018; @Fisher.2010; @Fisher.20159wi; @Koven:2020].
 
@@
   We ran ELM with FATES vegetation, in which the ELM model simulates interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water, surface water, soil water by depth and water in an unconfined aquifer (omitting processes relevant to snow, wetlands or lakes).
-See SI for a water balance equation and a note on how soil water dynamics is simulated in ELM-FATES.
+See Methods S5 for a water balance equation (Eq.
+S10) and a note on how soil water dynamics is simulated in ELM-FATES.
 The soil profile is discretized into maximum 15 exponentially distributed soil layers with layer node depth $z$.
@@
 Notably, we leveraged local data for soil hydraulic conductivity by depth (@Godsey:2004; Fig.
-S2), and instead of the ELM default soil texture-based pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content vs.
-$\mathrm{\Psi}$~soil~ [@Kupers:2019].
-See SI.
+S3), and instead of the ELM default soil texture-based pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content vs.
+$\mathrm{\Psi}$~soil~ [@Kupers:2019; Eq.
+S11].
+See Methods S6.
 
@@
 We ran ELM-FATES for the 5000-member 11 parameter ensemble with hourly climate drivers measured at the BCI meteorological station over 1985-2018 [@Faybishenko.2021] initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode.
-As our key interest here was on deriving soil water availability, we ran ELM-FATES in a lower-complexity configuration: static stand structure (see SI), and with a single plant functional type (PFT) of evergreen trees.
+As our key interest here was on deriving soil water availability, we ran ELM-FATES in a lower-complexity configuration: static stand structure (see Methods S5), and with a single plant functional type (PFT) of evergreen trees.
 The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous [@Condit:2000] and addition of a dry-deciduous PFT did not significantly alter results (not shown).
@@
 
-We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in the water balance equation, namely, (1) evapotranspiration ET from the flux tower by the 50 ha plot (2012-2017), (2) local stream discharge (2012-2018; see @Paton.2019) and (3) soil volumetric water content (VWC) from two sources: (i) a long-term (2012-2018) record of VWC averaged across three vertical time domain reflectometry (TDR) probes over depth 0-15 cm from three locations near the flux tower, and (ii) plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 at depths of 0.15, 0.4 and 1 m (1299 samples covering all soil types and habitats; @Kupers:2019).
+We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in the water balance equation, namely, (1) evapotranspiration ET from the flux tower by the 50 ha plot (2012-2017, Dataset S2, Table S2), (2) local stream discharge (2012-2018; Dataset S5; @Paton.2019run) and (3) soil volumetric water content (VWC) from two sources: (i) a long-term (2012-2018) record of VWC averaged across three vertical time domain reflectometry (TDR) probes over depth 0-15 cm from three locations near the flux tower (Dataset S6, Fig.
+S4), and (ii) plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 at depths of 0.15, 0.4 and 1 m (1299 samples covering all soil types and habitats; @Kupers:2019).
+See Methods S7.
 
 \
- For ELM-FATES calibration we calculated an objective function for each of the 5000-member ensemble by equally weighting standardized root mean square error (RMSE) between observations and simulations across all fluxes and states mentioned above, and then identified 100 parameter ensemble members that minimised the objective function, ensuring that soil moisture dynamics by depth is captured correctly (see SI).
+ For ELM-FATES calibration we calculated an objective function (Eq.
+S12) for each of the 5000-member ensemble by equally weighting standardized root mean square error (RMSE) between observations and simulations across all fluxes and states mentioned above, and then identified 100 parameter ensemble members that minimised the objective function, ensuring that soil moisture dynamics by depth is captured correctly (Methods S7).
 See Table S1 for the ranges of best-fit values for different parameters.
@@
 We evaluated ELM-FATES by calculating RMSE between simulated and observed long-term daily VWC for the depths of 0.1, 0.4 and 1 m (2016-2018), based on a dataset we had left-out during calibration.
-We obtained these observations from three horizontal TDR probes at the depths of 0.1, 0.4 and 1 m at a location near the vertical probes (see [SI](#SIData.VWC)).
+We obtained these observations from three horizontal TDR probes at the depths of 0.1, 0.4 and 1 m at a location near the vertical probes (Dataset S6).
 
@@
 We leveraged $\mathrm{\delta}$^2^H~xylem~ from BCI [@Meinzer:2001] for the dry-season of March 1997 as this period showed largest seasonal divergence in $\mathrm{\delta}$^2^H~xylem~ among species and vertically in soil and groundwater $\mathrm{\delta}$^2^H at natural abundance level (Fig.
-S3).
+S5).
 
@@
 We therefore removed five species that are typically leafless in March-April (Joseph Wright, personal observation), that is, the months of $\mathrm{\delta}$^2^H~xylem~ sampling.
-$\mathrm{\delta}^2H$ to depth relationship was particularly uncertain for $\mathrm{\delta}^2H$ > -40 ‰, so from the remaining dataset, we removed one species, *Guapira standleyana*, with $\mathrm{\delta}$^2^H~xylem~ of -28.9 ‰ $\pm$ 3.7SE (see Fig.
-S3).
-For each model of ERD (Eq.
-\@ref(eq:M-chosen) and alternatives in SI), modeled species ERD was regressed against species $\mathrm{\delta}$^2^H~xylem~ for a maximum of six species.
+$\mathrm{\delta}$^2^H to soil depth relationship was particularly uncertain for $\mathrm{\delta}$^2^H > -40 ‰, so from the remaining dataset, we removed one species, *Guapira standleyana*, with $\mathrm{\delta}$^2^H~xylem~ of -28.9 ‰ $\pm$ 3.7SE (see Fig.
+S5).
+For each model of ERD (Eqs.
+\@ref(eq:M-chosen) and S1-S6), modeled species ERD was regressed against species $\mathrm{\delta}$^2^H~xylem~ for a maximum of six species.
 
@@
 We regressed species ERD against maximum stem hydraulic conductivity $K_\textrm{max,stem}$ (*n*=7), leaf turgor loss point $\mathrm{\Psi}$~tlp~ (*n*=7), vulnerability to embolism from cavitation $\mathrm{\Psi}$~88,stem~ measured in terms of pressure at which 88% of $K_\textrm{max,stem}$ is lost (*n*=7), and hydraulic safety margins $\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~ (*n*=6).
-See [SI](#SIData.brett) and Table S2 for data collection and estimation of these variables.
+See Datasets S7, S8, S9 and Table S3 for data collection and estimation of these variables.
 
@@
 Dry season deciduous species may escape drought exposure via leaf deciduousness, so we analysed deciduous and evergreen species separately.
-Tree species on BCI are scored by expert botanists as one among four leaf-habits; evergreen, brevideciduous, facultative deciduous and obligate deciduous (@Meakem.2018; see SI).
+Tree species on BCI are scored by expert botanists as one among four leaf habits; evergreen, brevideciduous, facultative deciduous and obligate deciduous (@Meakem.2018; Dataset S4).
 We pooled all deciduous leaf habits together (hereafter, 'deciduous' for brevity) and regressed species ERD against species $\textrm{M}_{s,t}$ for each census interval $t$ for deciduous ($n$ = `r nrow(mrate.depth.mean %>% subset(deciduous != "E"))`) and evergreen species ($n$ = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`) separately.
@@
 The reduction in parameter range (Table S1) in the best-fit ensembles compared to the tested global ranges showed that the model calibration was primarily sensitive to the Ball-Berry stomatal slope parameter (fates_leaf_BB_slope), the ELM root distribution parameter that regulates the depth of the rooting profile (fates_rootb_par), soil hydraulic conductivity (HKSAT) profile especially at depth, and the adjustment factor (HKSAT_ADJ) that modifies soil hydraulic conductivity to account for macroporosity and direct flow paths (Figs.
-S4, S5).
+S6, S7).
 The distribution of maximum depth of soil water dynamics, and thus ecosystem root zone depth, across the 100 hydrological realisations encompassed 95% CI of `r root.zone.CI1` m - `r root.zone.CI2` m (median at `r root.zone.med` m).
@@
 \
-Soil water potential dynamics $\mathrm{\Psi}_{\textrm{soil},z}$ by depth z from 0.01 through 13 m obtained from the best-fit simulations revealed that every 5-yr census interval had at least one extreme hydrological drought year, but each interval varied in terms of number of extreme years, drought intensity, drought seasonality and duration (Fig.
+Analysis of soil water potential dynamics $\mathrm{\Psi}_{\textrm{soil},z}$ by depth z from 0.01 through 13 m obtained from the best-fit simulations (Fig.
+S8) revealed that every 5-yr census interval had at least one extreme hydrological drought year, but each interval varied in terms of number of extreme years, drought intensity, drought seasonality and duration (Fig.
 \@ref(fig:swp-droughts)).
@@
 Across the simulation period 1990-2015, simulated $\mathrm{\Psi}_{\textrm{soil},z}$ remained above -0.5 MPa from depths 1.7 to 13 m (Fig.
-S6).
+S8).
 
@@
 R*^2^ = `r afit.r2`, $p$ `r  afit.p`; also see Fig.
-S7).
-This predictive power allowed us to estimate leaf vulnerability curves used in the ERD models for `r sp.n.model` ERD species that lacked direct observations (among `r length(erd.sp)` ERD species) (Table S3, Fig.
-S8).
+S9).
+This predictive power allowed us to estimate leaf vulnerability curves used in the ERD models for `r sp.n.model` ERD species that lacked direct observations (among `r length(erd.sp)` ERD species) (Table S4, Fig.
+S10).
 
@@
 The best ERD model (Eq.
-(3)) explained a large fraction of the variance in $\mathrm{\delta}$^2^H~xylem~ ($R$^2^ = `r lm.erd.iso.r2`, $p$ = `r lm.erd.iso.p`, $n$ = `r length(unique(erd.iso$sp))`, Fig.
-\@ref(fig:erd-iso); also see SI).
+(3); Notes S1) explained a large fraction of the variance in $\mathrm{\delta}$^2^H~xylem~ ($R$^2^ = `r lm.erd.iso.r2`, $p$ = `r lm.erd.iso.p`, $n$ = `r length(unique(erd.iso$sp))`, Fig.
+\@ref(fig:erd-iso); see Fig.
+S11).
 This model included an effect of VPD and not LAI (Eq.
@@
 \@ref(fig:erd-hyd)**d**).
+See Fig.
+S12 for the full correlation matrix.
 
@@
 For deciduous species ERD explained 11-16% variation in mortality in four intervals with El-Niño events, but p-values were not significant (Fig.
-S11).
+S13).
 
@@
 \
-On an average, species exposure to water-stress (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) exponentially declined with ERD, indicating that species with shallower ERD spent greater time under significant hydrological drought, and thus likely experienced greater hydraulic risk (also see SI).
+On an average, species exposure to water-stress (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) exponentially declined with ERD (Figs.
+\@ref(fig:erd-time-evg), S14), indicating that species with shallower ERD spent greater time under significant hydrological drought, and thus likely experienced greater hydraulic risk (Notes S2).
 Exposure to water-stress increased over the three periods for which ERD explained significant mortality (Figs.
@@
 
-We introduce a novel approach for estimating ERD using 25 years of tree growth, species-specific leaf vulnerability curves, modeled soil water potential profiles, and observed VPD (Eq.
+We introduce a novel approach for estimating effective rooting depths (ERD) using 25 years of tree growth, species-specific leaf vulnerability curves, modeled soil water potential profiles, and observed VPD (Eq.
 \@ref(eq:M-chosen), Figs.
@@
 \@ref(fig:erd-hyd)).
-Nevertheless, evergreen but not deciduous species with deeper ERD showed significantly lower mortality rates than shallower ERD species (Fig.
+Nevertheless, evergreen but not deciduous species with deeper ERD showed significantly lower mortality rates than shallower ERD species (Figs.
 \@ref(fig:erd-mort-census-evg), S11).
@@
 \@ref(fig:erd-time-evg)), whereas shallow rooted species pay the cost of hydraulic safety in terms of efficiency, adapted for an environment in which hydraulic risk is significant, as extreme droughts cause exponentially greater water-stress in shallow soil layers (Figs.
-Figs.
 \@ref(fig:erd-time-evg), S6).
@@
 The hydrological model in CT2018 revealed that the multi-year drought exhausted the deep soil and even bedrock water availability (also see, @Goulden.2019; @Ivanov:2012).
-In contrast, our hydrological modeling at BCI finds that $\mathrm{\Psi}$~soil~ for depths deeper than 2.9 m did not cross $\mathrm{\Psi}$~crit~ of the most sensitive tree species (`r min.psi_crit` MPa) for all the dry seasons and droughts during 1990-2015 (see SI).
+In contrast, our hydrological modeling at BCI finds that $\mathrm{\Psi}$~soil~ for depths deeper than 2.9 m did not cross $\mathrm{\Psi}$~crit~ of the most sensitive tree species (`r min.psi_crit` MPa) for all the dry seasons and droughts during 1990-2015 (Notes S1).
 Mean annual rainfall of 1095 mm in CT2018 compared to `r round(rain.man.stats$rain)` mm at BCI, and precipitation to potential evapotranspiration ratio of nearly one in CT2018 compared to nearly two in BCI, are major factors in the different mortality responses.
@@
 Our interpretation of this result is that VPD and leaf hydraulic vulnerability curves may be adequate to explain inter-census differences in growth (via stomatal control), but also acknowledge that our estimates of seasonality of LAI is a tentative estimate that combines leaf-fall data and the timing of leaf-gain backtracked from leaf lifetime, and omits inter-annual variation.
-Species-level leaf-fall data from litter-traps that we used includes large within-species variability in leaf-fall timing, and so may not have captured  tree-level seasonality in deciduousness (see SI).
+Species-level leaf-fall data from litter-traps that we used includes large within-species variability in leaf-fall timing, and so may not have captured tree-level seasonality in deciduousness (Methods S4).
 Future studies may improve models of LAI seasonality and make use of data from new technologies such as drone based monitoring of LAI in species-rich tropical forests [@Park.2019].
@@
 Department of Agriculture, McIntire Stennis project under LAB94493.
-LS was supported by National Science Foundation award 2017949.
-RF was supported by the National Center for Atmospheric Research, which is funded by the National Science Foundation.
+LS was supported by the National Science Foundation award 2017949.
+RAF was supported by the National Center for Atmospheric Research, which is funded by the National Science Foundation.
+MD was supported by the Carbon Mitigation Initiative at Princeton University.
 The BCI forest dynamics research project, founded by Stephen P.
 Hubbell and Robin B.
-Foster and sustained for many years by Richard Condit, is now managed by Rolando Perez, Suzanne Lao and Stuart Davies under the ForestGEO program of the Smithsonian Tropical Research in Panama.
+Foster and sustained for many years by Richard Condit, is now managed by Rolando Pérez (RP), Suzanne Lao and Stuart Davies under the ForestGEO program of the Smithsonian Tropical Research in Panama.
 Numerous organizations have provided funding, principally the US National Science Foundation, and hundreds of field workers have contributed.
+Comments from editor Jarmila Pitterman and three anonymous reviewers helped improve and clarify this manuscript.
 
-# Author contribution
+# Author contributions
 
 Conceptualization: RCT.
-Data curation: BF, SP for climate drivers.
-SP for stream discharge.
-MTT for GPP, ET, TDR, bulk density.
-SK for manual GWC and GWC vs $\mathrm{\Psi}_\textrm{soil}$.
-NK, JZ, KAT, LS for leaf hydraulics, BW for stem hydraulics.
+Data curation: BF, SRP for climate drivers.
+SRP for stream discharge.
+MD for GPP, ET, TDR, bulk density.
+SJK for manual GWC and GWC vs $\mathrm{\Psi}_\textrm{soil}$.
+NK, JZ, KJA, LS for leaf hydraulics, BTW for stem hydraulics.
 RP and SA for the BCI 50ha plot censuses.
-JW for maximum tree height, WSG, LMA, deciduousness, leaf fall and leaf lifetime.
+SJW for maximum tree height, WSG, LMA, deciduousness, leaf fall and leaf lifetime.
 Formal Analysis: RCT, CX.
-MTT for upgrading GPP-VPD relationship and LAI seasonality.
-Funding acquisition: SMM, JC, CX, NM, LK.
+MD for upgrading GPP-VPD relationship and LAI seasonality.
+Funding acquisition: SMM, JC, CX, NM, LMK.
 Investigation: RCT, CX (SMM and LR for an exploratory version).
 Methodology: RCT, CX (SMM and LR for an exploratory version).
-Project administration: RCT, Resources: SMM, BF, SP, MTT, SK, NK, JZ, KAT, LS, BW, JW.
-Software: RCT; RF, RK, CF, CX, RCT for ELM-FATES.
-Supervision: CX, SMM, BN, Validation: RCT.
+Project administration: RCT, Resources: SMM, BF, SRP, MD, SJK, NK, JZ, KJA, LS, BTW, SJW.
+Software: RCT; RAF, RGK, CDK, CX, RCT for ELM-FATES.
+Supervision: CX, SMM, BDN, Validation: RCT.
 Visualization: RCT, Writing – original draft: RCT.
-Writing – review & editing: All authors.
+Writing – review & editing: RCT, CX, KAT, MD, RAF, RGK, CDK, LMK, NK, SJK, NGM, BDN, SRP, LR, LS, JMW, BTW, CW, SJW, JZ, SMM.
 See https://casrai.org/credit/ for the taxonomy of credits.
@@
 
+# Supporting Information
+
+The following Supporting Information is available for this article:
+
+**Figure S1**  Observed relationship between GPP and VPD at Barro Colorado Island.  
+**Figure S2**  Seasonal variation in estimated standardised LAI for 29 species at Barro Colorado Island.  
+**Figure S3**  Ensembles of soil hydraulic conductivity variation by depth for Conrad Trail Stream Catchment, Barro Colorado Island. Data from @Godsey:2004.  
+**Figure S4**  Volumetric water content vs. TDR data for a vertical TDR probe used for TDR probe calibration.  
+**Figure S5**  $\mathrm{\delta}$^2^H of soil water by depth measured at Barro Colorado Island. Data from @Meinzer:1999.   
+**Figure S6**  Parameter sensitivity of ELM-FATES soil water content and evapo-transpiration.  
+**Figure S7**  Parameter sensitivity of ELM-FATES stream discharge.  
+**Figure S8**  ELM-FATES predicted soil water potential dynamics by depth over 1990-2018.  
+**Figure S9**  Leaf hydraulic vulnerability curves fitted to observed $K_\textrm{leaf}$ vs. $\mathrm{\Psi}$~leaf~ data for 21 tree species from Barro Colorado Island.  
+**Figure S10**  Percent loss of leaf hydraulic conductivity curves for ERD species, including curves fitted to $K_\textrm{leaf}$ vs.
+$\mathrm{\Psi}$~leaf~ as well as those based on scaling relationships with leaf mass area and wood specific gravity.
+**Figure S11**  Effective rooting depths (ERD) versus $\mathrm{\delta}$^2^H~xylem~ for each of the ERD model structures tested.   
+**Figure S12**  Correlation matrix between ERD and hydraulic traits.  
+**Figure S13**  Mortality rates for deciduous species across census intervals from 1985-2015 versus ERD.  
+**Figure S14**  Modeled effective rooting depths of deciduous species versus time spent beyond critical hydraulic threshold.
+
+**Table S1**  ELM-FATES parameters used to generate ensembles, with description, prescribed global ranges, rationale for the choice of ranges and references, as well as ranges for best-fit ensembles.
+**Table S2**  QA/QC procedure applied to eddy covariance fluxes.  
+**Table S3**  Above-ground hydraulic traits data from @Wolfe.2019 and @Wolfe.2021 used for comparison with ERD.   
+**Table S4**  Leaf vulnerability curve parameters A & B, $K_\textrm{max,leaf}$ and $\mathrm{\Psi}_\textrm{20,leaf}$ by source (data or model) for the 29 species with ERD estimates.
+
+**Methods S1** Alternative structures for effective rooting depth model.  
+**Methods S2** Statistics for identifying best-fit ERD.   
+**Methods S3** Processing of forest census data for growth estimates.    
+**Methods S4** Leaf Area Index calculations.
+**Methods S5** Details of the ELM-FATES model.  
+**Methods S6** Details for ELM-FATES model parameterization.  
+**Methods S7** ELM-FATES calibration.
+
+**Dataset S1** Microclimatic and flux tower data.   
+**Dataset S2** Leaf hydraulic conductivity and vulnerability to cavitation.   
+**Dataset S3** Wood specific gravity (WSG) and leaf mass area (LMA). 
+**Dataset S4** Leaf deciduousness categories.  
+**Dataset S5** Stream discharge from the Conrad catchment.  
+**Dataset S6** Volumetric water content.  
+**Dataset S7** Stem maximum hydraulic conductivity and vulnerability to cavitation.  
+**Dataset S8** Leaf turgor loss point.  
+**Dataset S9** Above-ground hydraulic safety margins.
+
+**Notes S1** ERD model structure selection.  
+**Notes S2** Additional results for exposure to water-stress.
+
 # Figure Captions 
@@
 Figure \@ref(fig:schematic): A schematic outlining the methods workflow, which combines hydrology, physiology and demography of the tropical forest at Barro Colorado Island to inform the inverse model for rooting depths; validates it, and tests hypotheses pertaining to relationship of rooting depths with above-ground hydraulic traits, drought exposure, and drought-induced mortality.
+ELM-FATES, Energy Exascale Earth System Land Model coupled with the Functionally Assembled Terrestrial Ecosystem Simulator; ET, evapotranspiration; LMA, leaf mass per unit area.
 
@@
 \
-Figure \@ref(fig:erd-hyd): Modeled effective rooting depth (ERD) versus hydraulic properties for seven canopy species found on Barro Colorado Island (Panama).
+Figure \@ref(fig:erd-hyd): Modeled effective rooting depth (ERD) versus hydraulic properties for seven canopy species found on Barro Colorado Island (Panama); namely, maximum stem area-specific hydraulic conductivity of stem ($K_\textrm{max,stem}$; **a**), bulk leaf turgor loss point, the $\mathrm{\Psi}$~leaf~ where turgor potential = 0 ($\mathrm{\Psi}$~tlp~; **b**), $\mathrm{\Psi}$~stem~ at 88% loss of stem conductivity ($\mathrm{\Psi}$~88,stem~; **c**), and above-ground hydraulic safety margin ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~; **d**).
 Spearman's $r$ and significance levels are given in panel insets.
 Linear model fits (blue lines) with confidence bands (gray area) are shown for significant fits at $\alpha$ = 0.05.
-See Fig.
-S10 for the full correlation matrix.
 Species are color coded by leaf habit.
@@
 Figure \@ref(fig:erd-time-evg): Modeled effective rooting depth (ERD; horizontal-axis) versus time spent beyond critical hydraulic threshold (vertical axis) by census interval (colored bars) for `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))` evergreen species included in the mortality analyses (Fig.
-\@ref(fig:erd-mort-census-evg)).
+8).
 Each bar represents the average time species of the same ERD spent beyond species-specific critical hydraulic thresholds in a given interval, that is, the proportion of days for which $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ was more negative than species $\mathrm{\Psi}$~crit~, defined as $\mathrm{\Psi}$~20,leaf~, and where $z$ is the soil depth matching species ERD.
@@
 
-#
 ```{r hypo, results = 'asis', echo = FALSE}

### 2021-05-06 10.0_Supporting_Information.Rmd
Incorporated Production Managers comments

@@
 ---
-title: "Hydraulically-vulnerable trees survive on deep-water access during droughts in a tropical forest"
+title: ""
 author: ''
@@
   bookdown::word_document2:
-        reference_docx: manuscriptstyle.docx
-        fig_caption: yes
+        reference_docx: manuscriptstyle_without_line_number.docx
+        fig_caption: false
         number_sections: false
@@
 csl: new-phytologist.csl
+indent: true
+header-includes:
+    - \usepackage{caption}
+    - \usepackage{indentfirst}
+    - \setlength{\parindent}{24pt}
+    # - \setlength{\parskip}{0em}
 ---
+
+\captionsetup[table]{labelformat=empty}
+\usepackage{float}
 
@@
 
-# Supporting Information
+_**New Phytologist**_ Supporting Information
 
-## Supplementary Figures
+**Hydraulically-vulnerable trees survive on deep-water access during droughts in a tropical forest**
 
-```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "35%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd}Scatterplot of mean midday (11:00 am to 15:00 pm) gross primary productivity (GPP) and mean midday vapor pressure deficit (VPD) measured at the eddy flux tower in Barro Colorado Island over 2012-2017.
-Red line shows the fit from a polynomial model described in inset."}
+Chitra-Tarak, Rutuja, Chonggang Xu, Salomón Aguilar, Kristina J.
+Anderson-Teixeira, Jeff Chambers, Matteo Detto, Boris Faybishenko, Rosie A.
+Fisher, Ryan G.
+Knox, Charles D.
+Koven, Lara M.
+Kueppers, Nobert Kunert, Stefan J.
+Kupers, Nathan G.
+McDowell, Brent D.
+Newman, Steven R.
+Paton, Rolando Pérez, Laurent Ruiz, Lawren Sack, Jeffrey M.
+Warren, Brett T.
+Wolfe, Cynthia Wright, S.
+Joseph Wright, Joseph Zailaa, Sean M.
+McMahon
+
+\
+Article acceptance date: 29 April 2021
+
+\
+The following Supporting Information is available for this article:
+
+**Figure S1**  Observed relationship between GPP and VPD at Barro Colorado Island.  
+**Figure S2**  Seasonal variation in estimated standardised LAI for 29 species at Barro Colorado Island.  
+**Figure S3**  Ensembles of soil hydraulic conductivity variation by depth for Conrad Trail Stream Catchment, Barro Colorado Island. Data from @Godsey:2004.  
+**Figure S4**  Volumetric water content vs. TDR data for a vertical TDR probe used for TDR probe calibration.  
+**Figure S5**  $\mathrm{\delta}$^2^H of soil water by depth measured at Barro Colorado Island. Data from @Meinzer:1999.   
+**Figure S6**  Parameter sensitivity of ELM-FATES soil water content and evapo-transpiration.  
+**Figure S7**  Parameter sensitivity of ELM-FATES stream discharge.  
+**Figure S8**  ELM-FATES predicted soil water potential dynamics by depth over 1990-2018.  
+**Figure S9**  Leaf hydraulic vulnerability curves fitted to observed $K_\textrm{leaf}$ vs. $\mathrm{\Psi}$~leaf~ data for 21 tree species from Barro Colorado Island.  
+**Figure S10**  Percent loss of leaf hydraulic conductivity curves for ERD species, including curves fitted to $K_\textrm{leaf}$ vs.
+$\mathrm{\Psi}$~leaf~ as well as those based on scaling relationships with leaf mass area and wood specific gravity.
+**Figure S11**  Effective rooting depth (ERD) versus $\mathrm{\delta}$^2^H~xylem~ for each of the ERD model structures tested.   
+**Figure S12**  Correlation matrix between ERD and hydraulic traits.  
+**Figure S13**  Mortality rates for deciduous species across census intervals from 1985-2015 versus ERD.  
+**Figure S14**  Modeled effective rooting depths of deciduous species versus time spent beyond critical hydraulic threshold.
+
+**Table S1**  ELM-FATES parameters used to generate ensembles, with description, prescribed global ranges, rationale for the choice of ranges and references, as well as ranges for best-fit ensembles.
+**Table S2**  QA/QC procedure applied to eddy covariance fluxes.  
+**Table S3**  Above-ground hydraulic traits data from @Wolfe.2019 and @Wolfe.2021 used for comparison with ERD.   
+**Table S4**  Leaf vulnerability curve parameters A & B, $K_\textrm{max,leaf}$ and $\mathrm{\Psi}_\textrm{20,leaf}$ by source (data or model) for the 29 species with ERD estimates.
+
+**Methods S1** Alternative structures for effective rooting depth model.  
+**Methods S2** Statistics for identifying best-fit ERD.   
+**Methods S3** Processing of forest census data for growth estimates.     
+**Methods S4** Leaf Area Index calculations.   
+**Methods S5** Details of the ELM-FATES model.  
+**Methods S6** Details for ELM-FATES model parameterization.  
+**Methods S7** ELM-FATES calibration.
+
+**Dataset S1** Microclimatic and flux tower data.   
+**Dataset S2** Leaf hydraulic conductivity and vulnerability to cavitation.   
+**Dataset S3** Wood specific gravity (WSG) and leaf mass area (LMA).  
+**Dataset S4** Leaf deciduousness categories.  
+**Dataset S5** Stream discharge from the Conrad catchment.  
+**Dataset S6** Volumetric water content.  
+**Dataset S7** Stem maximum hydraulic conductivity and vulnerability to cavitation.  
+**Dataset S8** Leaf turgor loss point.  
+**Dataset S9** Above-ground hydraulic safety margins.
+
+**Notes S1** ERD model structure selection.  
+**Notes S2** Additional results for exposure to water-stress.
+
+\newpage
+**Figure S1** Scatterplot of mean midday (11:00 am to 15:00 pm) gross primary productivity (GPP) and mean midday vapor pressure deficit (VPD) measured at the eddy flux tower in Barro Colorado Island over 2012-2017.
+Red line shows the fit from a polynomial model described in inset.
+
+```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "35%", fig.width = 3, fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd}"}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/gpp_by_env_variables/monthly_gpp.vpd.jpeg"))
@@
 \newpage
-```{r ksat-profiles, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ksat-profiles}Generated 5000 curves of soil hydraulic conductivity variation by depth based on Latin Hypercube Sampling within 95% CI of observed data at depths of 12.5 cm and 60 cm [@Kinner:2004], assuming a linear decline between those depths, and values before 12.5 and after 60 cm as that of 12.5 cm and 60 cm, respectively.
+**Figure S2** Seasonal variation in standardised LAI (range 0-1) for `r length(erd.sp)` species at Barro Colorado Island (panels). Species are color-coded by leaf habit as defined by expert botanists.
+
+```{r leaf-cover, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-cover}")}
+knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf.cover_BCI_multi_panel.jpeg"))
+```
+
+\newpage
+**Figure S3** Generated 5000 curves of soil hydraulic conductivity variation by depth based on Latin Hypercube Sampling within 95% CI of observed data at depths of 12.5 cm and 60 cm for Conrad Trail Stream Catchment [@Godsey:2004], assuming a linear decline between those depths, and values before 12.5 and after 60 cm as that of 12.5 cm and 60 cm, respectively.
 Only 100 curves are shown for clarity.
-Note the log scale on horizontal axis.")}
+Note the log scale on horizontal axis.
+
+```{r ksat-profiles, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ksat-profiles}")}
 knitr::include_graphics(file.path("figures/Bootstrapped Ksat profiles within observed 95CI_random_two_hundred.jpeg"))
@@
 \newpage
-```{r iso-data, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-iso-data}Stable isotopic ratio ($\\mathrm{\\delta}^2H$) of soil water by depth measured at Barro Colorado Island in March 1997.
+**Figure S4** Volumetric water content vs. TDR data for a vertical TDR probe (0-15 cm depth) at the base of the eddy flux tower near the 50 ha plot at BCI.
+
+```{r TDR-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-TDR-calib}"}
+knitr::include_graphics(file.path("figures/Matteos_TDR_SWC_data_linear_model.jpeg"))
+```
+
+\newpage
+**Figure S5** Stable isotopic ratio ($\mathrm{\delta}^2H$) of soil water by depth measured at Barro Colorado Island in March 1997.
+Groundwater sampled at the sources of springs on BCI during this period yielded an average $\mathrm{\delta}^2H$ value of –60 ‰ (not shown).
 Data from @Meinzer:1999.
-$\\mathrm{\\delta}^2H$ to depth relationship was particularly uncertain for $\\mathrm{\\delta}^2H$ values > -40 ‰.
-Two species, *Guapira standleyana* (-28.9 $\\pm$ 3.7; mean $\\pm$ SE) and *Spondias radlkoferi* (-33.5 $\\pm$ 2.6), in @Meinzer:1999 dataset had $\\mathrm{\\delta}$^2^H~xylem~ values > -40 ‰.
-For example, *Guapira standleyana*, with $\\mathrm{\\delta}$^2^H value of -28.9 $\\pm$ 3.7 could have a source-water depth anywhere between 10 to 50 cm.
+$\mathrm{\delta}^2H$ to depth relationship was particularly uncertain for $\mathrm{\delta}^2H$ values > -40 ‰.
+Two species, *Guapira standleyana* (-28.9 $\pm$ 3.7; mean $\pm$ SE) and *Spondias radlkoferi* (-33.5 $\pm$ 2.6), in @Meinzer:1999 dataset had $\mathrm{\delta}$^2^H~xylem~ values > -40 ‰.
+For example, *Guapira standleyana*, with $\mathrm{\delta}$^2^H~xylem~ value of -28.9 $\pm$ 3.7 could have a source-water depth anywhere between 10 to 50 cm.
 *Spondias radlkoferi* is also typically leafless in March-April.
-These two species were not included in the ERD model-data comparison."}
+These two species were not included in the ERD model-data comparison.
+
+```{r iso-data, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-iso-data}"}
 knitr::include_graphics(file.path("figures/UDI_confidence/Meinzer_etal_1999_Depth_vs_soil_deltaD.jpeg"))
@@
 \newpage
-```{r sens-sm-et, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-sm-et}Parameter sensitivity of ELM-FATES soil water content (that is, VWC, **a**-**b**) and ET (**c**-**d**) is shown for two representative sample months, as sensitivity for a particular parameter varies depending on the chosen month.
+**Figure S6** Parameter sensitivity of ELM-FATES soil water content (that is, VWC, **a**-**b**) and evapotranspiration (ET, **c**-**d**) is shown for two representative sample months, as sensitivity for a particular parameter varies depending on the chosen month.
 Horizontal axes show global ranges for a given parameter.
@@
 Each point represents a simulation among the 5000 ensemble-member runs.
-See S\\@ref(tab:param) for description, global ranges and chosen ranges for parameters."}
+See Table S\@ref(tab:param) for description, global ranges and chosen ranges for parameters.
+
+```{r sens-sm-et, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-sm-et}"}
 knitr::include_graphics(file.path("figures/ELM_FATES_parameter_sensitivity/sens.sm.et.jpeg"))
@@
 \newpage
-```{r sens-run, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-run}Parameter sensitivity of ELM-FATES stream discharge (QRUNOFF) is shown for four sample months (**a**-**d**), as sensitivity for a particular parameter varies depending on the chosen month.
+**Figure S7** Parameter sensitivity of ELM-FATES stream discharge (QRUNOFF) is shown for four sample months (**a**-**d**), as sensitivity for a particular parameter varies depending on the chosen month.
 Horizontal axes show global ranges for a given parameter.
@@
 Each point represents a simulation among the 5000 ensemble-member runs.
-See S\\@ref(tab:param) for description, global ranges and chosen ranges for parameters."}
+See Table S\@ref(tab:param) for description, global ranges and chosen ranges for parameters.
+
+```{r sens-run, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-run}"}
 knitr::include_graphics(file.path("figures/ELM_FATES_parameter_sensitivity/sens.run.jpeg"))
@@
 \newpage
-```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($\\Psi_{\\textrm{soil},z}$) by depth $z$ (m; panels) over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines).
-Vertical gray lines demarcate median dates of the demographic censuses at the BCI 50 ha plot."}
+**Figure S8** Daily mean soil water potential ($\Psi_{\textrm{soil},z}$) by depth $z$ (in m; panels) over 1990-2018 predicted from each of the 100 best-fit ensembles (colored lines).
+Vertical gray lines demarcate median dates of the tree demographic censuses at the BCI 50 ha plot.
+
+```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}"}
 knitr::include_graphics(file.path("figures/2019-10-14_5000/best-fits/psi_model_daily_all_depths_params.top.few_full_beg_1991.jpeg"))
@@
 \newpage
-```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}Vulnerability curves fitted to observed $K_\\textrm{leaf}$ vs.
-$\\mathrm{\\Psi}$~leaf~ data for ", length(sp.exp.param$sp)," tree species from BCI--in absolute terms (**a**, **b**), as well as, in terms of loss of conductivity (%) (**c**, **d**).
-Each line represents a species, color coded by WSG (**a**, **c**), or LMA (**b**, **d**), with blue tones indicating smaller values of WSG or LMA and red tones indicating larger values.")}
+**Figure S9** Leaf hydraulic vulnerability curves fitted to observed $K_\textrm{leaf}$ vs.
+$\mathrm{\Psi}$~leaf~ data for `r length(sp.exp.param$sp)` tree species from BCI--in absolute terms (**a**, **b**), as well as, in terms of percent loss of conductivity (**c**, **d**).
+Each line represents a species, color coded by wood specific gravity (WSG) (**a**, **c**), or leaf mass area (LMA) (**b**, **d**), with blue tones indicating smaller values of WSG or LMA and red tones indicating larger values.
+
+```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/plot.obs.plc.jpeg"))
@@
 \newpage
-```{r obs-comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-comm-plc-curves}Percent loss of vulnerability curves for species with ERD estimates.
-Each line represents a species, color coded by source of the curve: fitted to observed data for $K_\\textrm{leaf}$ vs.
-$\\mathrm{\\Psi}$~leaf~ are in yellow and those based on the scaling relationship with WSG and LMA are in gray.")}
+**Figure S10** Percent loss of leaf hydraulic conductivity curves for species with ERD estimates.
+Each line represents a species, color coded by source of the curve: fitted to observed data for $K_\textrm{leaf}$ vs.
+$\mathrm{\Psi}$~leaf~ are in yellow (data) and those based on the scaling relationships with leaf mass area and wood specific gravity are in gray (model).
+
+```{r obs-comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-comm-plc-curves}")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/obs.mod_plccurves_fits_ggplot_no_col.jpeg"))
@@
 \newpage
-```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-other-erd-mods}Effective Rooting Depth versus $\\mathrm{\\delta}^2H$ for each of the ERD models tested (panels).
-For each ERD model only those species have an estimate of ERD that have passed the test of Pearson's $r$ > 0.71 during ERD parameterisation.
-The number of species with an ERD estimate as well as $\\mathrm{\\delta}^2H$ for each model are shown in inset, along with the goodness-of-fit ($R$^2^) and $p$-values for a linear relationship between ERD and $\\mathrm{\\delta}^2H$.
+**Figure S11** Effective rooting depth versus $\mathrm{\delta}^2H$~xylem~ for each of the ERD model structures tested (panels).
+For each ERD model structure only those species have an estimate of ERD that have passed the test of Pearson's $r$ > 0.71 during ERD parameterisation.
+The number of species with an ERD estimate as well as $\mathrm{\delta}^2H$ for each model structure are shown in inset, along with the goodness-of-fit ($R$^2^) and $p$-values for a linear relationship between ERD and $\mathrm{\delta}^2H$~xylem~.
 ERD models in panels **a**-**e** correspond to Eq.
-\\@ref(eq:M1)-\\@ref(eq:M6), respectively.")}
+S\@ref(eq:M1)-S\@ref(eq:M6), respectively.
+
+```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-other-erd-mods}")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_sp_color_Meinzer.jpeg"))
@@
 \newpage
-```{r erd-hyd-full, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd-full}Pairwise Spearman's correlations (upper triangle) and corresponding linear regressions (lower triangle) among modelled Effective Rooting Depth (ERD) and hydraulic properties for seven canopy species found on Barro Colorado Island (Panama).Spearman's correlation coefficients and significance levels are shown (upper-half; '*' $p$ < 0.05; '.' $p$ < 0.1; ' ' $p$ < 1).")}
+**Figure S12** Pairwise Spearman's correlations (upper triangle) and corresponding linear regressions (lower triangle) among modelled effective rooting depth (ERD) and hydraulic properties for seven canopy species found on Barro Colorado Island (Panama).
+Spearman's correlation coefficients and significance levels are shown (upper triangle; '$**$' $p$ < 0.01, '$*$' $p$ < 0.05, '$.$' $p$ < 0.1).
+
+```{r erd-hyd-full, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd-full}")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/erd.stem.traits_cor.chart.jpeg"))
@@
 \newpage
-```{r erd-mort-deci, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-cdeci}Mortality rates across census intervals from 1985-2015 for species (circles) of various deciduous leaf-habit categories (colors) plotted against ERD.
+**Figure S13** Mortality rates across census intervals from 1985-2015 for species (circles) of various deciduous leaf-habit categories (colors) plotted against ERD.
 $R$^2^ and significance levels for linear model fits are given in panel insets.
-Census-intervals with a superscript * are shown for comparison and indicate intervals in which significant mortality was explained by ERD among evergreen species (see Fig.
-8 in the main text)."}
+Census-intervals with a superscript $*$ and $**$ are shown for comparison with a corresponding figure for evergreen species (Fig.
+8 in the main text) and indicate intervals in which significant mortality was explained by ERD among evergreen species at $\alpha$ = 0.1 and $\alpha$ = 0.05, respectively.
+
+```{r erd-mort-deci, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-cdeci}"}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_by rdi.gr_deciduous.jpeg"))
@@
 \newpage
-```{r erd-time-deci, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time-deci}Modeled effective rooting depth (ERD; horizontal-axis) versus time spent beyond critical hydraulic threshold (vertical axis) by census interval (colored bars) for deciduous species in mortality analyses (Fig.
-\\@ref(fig:erd-mort-deci)).
-Each bar represents the average time species of the same ERD spent beyond species-specific critical hydraulic thresholds in a given interval, that is, the proportion of days for which $\\mathrm{\\Psi}_{\\textrm{soil},z=\\textrm{ERD}}$ was more negative than species $\\mathrm{\\Psi}$~crit~, defined as $\\mathrm{\\Psi}$~20,leaf~, and where $z$ is the soil depth matching species ERD.
+**Figure S14** Modeled effective rooting depth (ERD; horizontal-axis) versus time spent beyond critical hydraulic threshold (vertical axis) by census interval (colored bars) for deciduous species in the mortality analyses (Fig.
+S\@ref(fig:erd-mort-deci)).
+Each bar represents the average time species of the same ERD spent beyond species-specific critical hydraulic thresholds in a given interval, that is, the proportion of days for which $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ was more negative than species $\mathrm{\Psi}$~crit~, defined as $\mathrm{\Psi}$~20,leaf~, and where $z$ is the soil depth matching species ERD.
 Standard error of the mean are shown over each bar.
-Note the squared y-axis scale.")}
+Note the squared y-axis scale.
+
+```{r erd-time-deci, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time-deci}")}
 knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mfac vs. rdi.gr_deciduous.jpeg"))
@@
 
-## Supplementary Tables
+\newpage
+**Table S1** ELM-FATES parameters used to generate ensembles, with parameter description, prescribed global ranges, rationale for the choice of ranges and references.
+Ranges for best-fit chosen ensembles are also given.
+Note that the first six parameters are specific to FATES.
 
-```{r param, results = 'asis', echo = FALSE}
-knitr::kable(param.table, caption = 'ELM-FATES parameters used to generate ensembles, with description, prescribed global ranges, rationale for the choice of ranges and references.
-Ranges for best-fit chosen ensembles are also given.
-Note that no.
-1-6 are parameters specific to FATES.', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(param.table))))
+```{r param, results = 'asis', echo = FALSE, tab.cap = NULL}
+knitr::kable(param.table, caption = "", booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(param.table))))
+```
+
+**Table S2** QA/QC procedure applied to eddy covariance fluxes.
+w: vertical wind speed; t: air temperature; q: water vapor density; c: CO~2~ density; u: horizontal wind speed.
+The bar indicates 30-min averaging and the carat indicates fluctuation ($\overline{\hat{x} = x - \bar{x}}$)
+
+```{r eddy, results = 'asis', echo = FALSE}
+knitr::kable(eddy.table, caption = "", booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(eddy.table))))
+```
+
+**Table S3** Minimum leaf water potential $\mathrm{\Psi}$~min~, leaf turgor loss point $\mathrm{\Psi}$~tlp~, stem maximum hydraulic conductivity $K_\textrm{max,stem}$ and stem vulnerability to cavitation $\mathrm{\Psi}$~88,stem~ for seven species with ERD estimates.
+$\mathrm{\Psi}$~min~ is sourced from @Wolfe.2019, while $\mathrm{\Psi}$~tlp~, $K_\textrm{max,stem}$, $\mathrm{\Psi}$~50,stem~ and $\mathrm{\Psi}$~88,stem~ are sourced from @Wolfe.2021.
+$\mathrm{\Psi}$~min~, $\mathrm{\Psi}$~tlp~, $\mathrm{\Psi}$~50,stem~ and $\mathrm{\Psi}$~88,stem~ are in MPa, while $K_\textrm{max,stem}$ in kg m^-1^ s^-1^ MPa^-1^.
+
+```{r hyd-table, results = 'asis', echo = FALSE}
+knitr::kable(hyd.table, caption = "", booktabs = TRUE, escape = FALSE)
 ```
@@
 \newpage
-```{r hyd-table, results = 'asis', echo = FALSE}
-knitr::kable(hyd.table, caption = paste0("\\label{lab-hyd-table} Minimum leaf water potential $\\mathrm{\\Psi}$~min~, leaf turgor loss point $\\mathrm{\\Psi}$~tlp~, stem maximum hydraulic conductivity $K_\\textrm{max,stem}$ and stem vulnerability to cavitation $\\mathrm{\\Psi}$~88,stem~ for seven species with ERD estimates.
-$\\mathrm{\\Psi}$~min~ is sourced from @Wolfe.2019, while $\\mathrm{\\Psi}$~tlp~, $K_\\textrm{max,stem}$, $\\mathrm{\\Psi}$~50,stem~ and $\\mathrm{\\Psi}$~88,stem~ are sourced from @Wolfe.2021.
-$\\mathrm{\\Psi}$~min~, $\\mathrm{\\Psi}$~tlp~, $\\mathrm{\\Psi}$~50,stem~ and $\\mathrm{\\Psi}$~88,stem~ are in MPa, while $K_\\textrm{max,stem}$ in kg m^-1^ s^-1^ MPa^-1^."), booktabs = TRUE, escape = FALSE)
+**Table S4** Leaf vulnerability curve parameters *A* & *B*, $K_\textrm{max,leaf}$ and $\mathrm{\Psi}_\textrm{20,leaf}$ for the `r length(erd.sp)` species with ERD estimates.
+Species for which curves were fitted to observed data for $K_\textrm{leaf}$ vs.
+$\mathrm{\Psi}$~leaf~ are identified with column Source as 'Data', while species for which curves were obtained based on trait-proxies are identified as 'Model'.
+
+```{r ab-param, results = 'asis', echo = FALSE}
+knitr::kable(ab.table, caption = "", booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(ab.table))))
 ```
@@
 \newpage
-```{r ab-param, results = 'asis', echo = FALSE}
-knitr::kable(ab.table, caption = paste0("\\label{lab-ab-param}Leaf vulnerability curves parameters *A* & *B*, $K_\\textrm{max,leaf}$ and $\\mathrm{\\Psi}$~20,leaf~ for the ", length(erd.sp), " species with ERD estimates.
-Species for which curves were fitted to observed data for $K_\\textrm{leaf}$ vs.
-$\\mathrm{\\Psi}$~leaf~ are identified with column Source as Data, while species for which curves were obtained based on trait-proxies are identified as Model."), booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(ab.table))))
-```
+**Methods S1** Alternative structures for effective rooting depth model.
 
-\newpage
-## Additional details of the ELM-FATES Model
-
-In ELM, soil water is predicted from a multi-layer model, in which the vertical soil moisture transport is governed by infiltration, surface and sub-surface runoff, gradient diffusion, gravity, canopy transpiration through root extraction, and interactions with groundwater (see @Oleson.2013 for process details and equations).
-FATES simulates size-structured tree community complete with individual tree rooting profiles.
-In our use of reduced-complexity configuration of ELM-FATES with a single plant functional type, ELM removes total community canopy transpiration as water-uptake from the soil layers in proportion to the total root fraction in each layer in the 1-D column.
-Note that we use the soil water potential dynamics obtained from ELM-FATES as an ecosystem-level outcome to drive our species-level, empirical ERD model, and thus believe that there is no conflict/circularity between tree-level rooting profiles or water uptake processes in ELM-FATES and the ERD model.
-
-\
-We ran ELM with FATES vegetation, in which the ELM model simulates interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\mathrm{\Delta}W_{\textrm{can}}$, surface water $\mathrm{\Delta}W_{\textrm{sfc}}$, soil water $\mathrm{\Delta}W_{\textrm{liq},z_{i}}$ at depth $z_{i}$ and water in an unconfined aquifer $\mathrm{\Delta} W_{a}$ (omitting processes relevant to snow, wetlands or lakes).
-Conservation of these terms is expressed as follows:
+Below we present all the alternative models for effective rooting depth (ERD).
+Daily average growth, $\hat{G}_{s,t|z}$ for species $s$, in the census interval $t$ is described as follows, in which $K_\textrm{leaf}$, and thus $FLC_\textrm{leaf}$, is driven by soil water dynamics at depth $z$ for a given  hydrological realisation $h$:
 
 \begin{equation}
-\begin{split}
-\mathrm{\Delta}W_\textrm{can} + \mathrm{\Delta}W_\textrm{sfc} + \sum_{i=1}^{15} \mathrm{\Delta}W_{\textrm{liq},z_{i}} + \mathrm{\Delta}W_{a} = \left(q_\textrm{rain} - E_{v} - E_{g} - q_\textrm{over} - q_\textrm{h2osfc} - q_\textrm{drai}\right)\mathrm{\Delta}t
-\end{split}
-(\#eq:water-balance)
-\end{equation}
-where, $q_{\textrm{rain}}$ is rainfall, $E_{v}$ is transpiration of the forest, $E_{g}$ is ground evaporation, $q_{\textrm{over}}$ is surface runoff, $q_{\textrm{h2osfc}}$ is runoff from surface water storage, $q$~drai~ is subsurface drainage (all in mm H~2~O) and $\mathrm{\Delta}t$ is the time-step (here, 1 hr).
-
-\
-Justification for the use of static stand structure: In static stand structure model, the model is started from an observed forest inventory and recruitment, growth and mortality are deactivated.
-Holding the stand structure static rather than allowing it to emerge from physiologically-mediated competition between trees removes a set of internal model feedbacks and thus the only differences between model simulations are driven by direct physiological effects of perturbed model parameters.
-This allows us to focus on the hydraulic and hydrological questions explored in this paper.
-
-## Additional details for ELM-FATES model parameterization
-
-In the Conrad catchment, in which the 50 ha plot is situated, soil hydraulic conductivity decreases with depth from 12.5 and 60 cm (@Godsey:2004; Table \@ref(tab:param)).
-We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths <12.5 cm and >60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig.
-\@ref(fig:ksat-profiles)).
-\
-For greater accuracy, instead of the ELM default pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content (GWC) vs.
-$\mathrm{\Psi}$~soil~ for depths of 0.15, 0.4 and 1 m [@Kupers:2019].
-We converted GWC to volumetric (VWC, $\theta$) using a measured bulk density value of 0.8 g cm^-3^ (see below).
-The @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences in water retention among depths were nominal,
-
-\begin{equation}
-\mathrm{\Psi}_{\textrm{soil}} = \mathrm{\Psi}_e\left(\frac{\mathrm{\theta}}{\mathrm{\theta}_{s}}\right)^{-b}
-(\#eq:soil-ret-curve)
-\end{equation}
-where $\mathrm{\Psi}_e$ is the air-entry matric potential, $\mathrm{\theta}_{s}$ is the saturated volumetric water content, and $b$, an index for soil pore-size distribution, respectively.
-$\mathrm{\Psi}_e$, $\mathrm{\theta}_{s}$ and $b$ that best-fit the data were 200 (mm H~2~O), 0.51 (cm^3^ cm^-3^) and 10 (unitless), respectively.
-
-## ELM-FATES calibration
-
-We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq.
-\@ref(eq:water-balance), namely, (1) evapotranspiration ET from the flux tower by the 50 ha plot ($\backsimeq$ $E_{v} + E_{g}$ in Eq.
-\@ref(eq:water-balance); 2012-2017; see [SI](#SIData.matteo))), (2) local stream discharge $q_\textrm{run}$ ($\backsimeq$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see @Paton.2019 and [SI](#SIData.steveP)) and (3) soil volumetric water content (VWC) $\mathrm{\theta}_{z}$ ($\backsimeq$ $W_{\textrm{liq,} z}$) from two sources: (i) a long-term (2012-2018) record of VWC $\mathrm{\theta}_\textrm{0-0.15 m}$ averaged across three vertical time domain reflectometry (TDR) probes over depth 0-15 cm from three locations near the flux tower, and (ii) plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 at depths of 0.15, 0.4 and 1 m (1299 samples covering all soil types and habitats; see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\mathrm{\theta}_{\textrm{0-0.15 m}}$, $\mathrm{\theta}_{\textrm{0.4 m}}$ and $\mathrm{\theta}_{\textrm{1 m}}$ refer to plot-wide averages).
-
-\
-For ELM-FATES calibration we calculated an objective function for each of the 5000-member ensemble by equally weighting standardized RMSE between observations and simulations across all fluxes and states mentioned above.
-We describe the objective function $O$ as:
-
-\begin{equation}
-\begin{split}
-\overline{O} = \frac{1}{6}\sum \textrm{RMSE}^{*}\textrm{(ET,}E_{v} + E_{g}) + \textrm{RMSE}^{*}(q_{\textrm{run}},q_{\textrm{over}} + q_{\textrm{drai}}) + \textrm{RMSE}^{*}(\mathrm{\theta}_{\textrm{0-0.15~m}}, {W_{\textrm{liq,}\overline{z}}}) \\
-& + \textrm{RMSE}^{*}(\mathrm{\theta}_{\textrm{0.15~m}}, W_{\textrm{liq,`r soil.depths[4]`m}}) + \textrm{RMSE}^{*}(\mathrm{\theta}_{\textrm{0.4~m}}, W_{\textrm{liq,`r soil.depths[6]`~m}}) + \textrm{RMSE}^{*}(\mathrm{\theta}_{1~m}, W_{\textrm{liq,`r soil.depths[8]`~m}})
-(\#eq:obj-func)
-\end{split}
-\end{equation}
-
-based on monthly sums for both ET and $q$~run~ and daily averages for $\mathrm{\theta}$s.
-$W_{\textrm{liq,}\overline{z}}$ is the average of daily $W_{\textrm{liq,} z}$, where $z$ $\in$ $Z$; $Z$ = (`r paste(soil.depths[1:4], collapse=", ")`) m.
-Superscript * indicates that RMSE is standardized between 0 and 1 across the 5000 simulations.
-
-\
-Model fit to soil moisture dynamics by depth was one of our key priority.
-Preliminary analyses showed that fitting the model to all states and fluxes simultaneously captured soil moisture dynamics for shallower depths but not at 1 m depth.
-To ensure model fit at 1 m depth, we thus imposed a two-step procedure, wherein we first chose the 300 parameter ensemble members that minimized $\textrm{RMSE}^{*}({\mathrm{\theta}_{\textrm{1 m}}}, W_{\textrm{liq,`r soil.depths[8]` m}})$ (< 0.2), and among those 300 ensemble members we chose 100 with the lowest values for the objective function $\overline{O}$.
-
-## Alternative model structures for Effective Rooting Depth (ERD)
-
-Below we present all the alternative models for Effective Rooting Depth (ERD).
-Daily average species-specific growth, $G_{t|z}$, in the census interval $t$ given an ERD of depth $z$, is described as follows:
-
-\begin{equation}
-G_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} \textrm{FLC}_{\textrm{leaf},s,i|z,h}^{*}) + \epsilon_{s,t|z,h}
+\hat{G}_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} \textrm{FLC}_{\textrm{leaf},s,i|z,h}^{*}) + \epsilon_{s,t|z,h}
 (\#eq:M1)
@@
 \begin{equation}
-G_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} \textrm{FLC}_{\textrm{leaf},s,i|z,h}^{*} + \hat{\textrm{VPD}}_{i}^{*}) + \epsilon_{s,t|z,h}
+\hat{G}_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} \textrm{FLC}_{\textrm{leaf},s,i|z,h}^{*} + \hat{\textrm{VPD}}_{i}^{*}) + \epsilon_{s,t|z,h}
 (\#eq:M2)
@@
 \begin{equation}
-G_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} \textrm{FLC}_{\textrm{leaf},s,i|z,h}^{*} \hat{\textrm{VPD}}_{i}^{*}) + \epsilon_{s,t|z,h}
+\hat{G}_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} \textrm{FLC}_{\textrm{leaf},s,i|z,h}^{*} \hat{\textrm{VPD}}_{i}^{*}) + \epsilon_{s,t|z,h}
 (\#eq:M3)
@@
 \begin{equation}
-G_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} \textrm{FLC}_{\textrm{leaf},s,i|z,h}^{*} \textrm{LAI}_{s,doy|i}^{*}) + \epsilon_{s,t|z,h}
+\hat{G}_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} \textrm{FLC}_{\textrm{leaf},s,i|z,h}^{*} \textrm{LAI}_{s,doy|i}^{*}) + \epsilon_{s,t|z,h}
 (\#eq:M4)
@@
 \begin{equation}
-G_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} \textrm{FLC}_{\textrm{leaf},s,i|z,h}^{*} + \hat{\textrm{VPD}}_{i}^{*}) \textrm{LAI}^{*}_{s,doy|i}) + \epsilon_{s,t|z,h}
+\hat{G}_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} \textrm{FLC}_{\textrm{leaf},s,i|z,h}^{*} + \hat{\textrm{VPD}}_{i}^{*}) \textrm{LAI}^{*}_{s,doy|i}) + \epsilon_{s,t|z,h}
 (\#eq:M5)
@@
 \begin{equation}
-G_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} \textrm{FLC}_{\textrm{leaf},s,i|z,h}^{*}\hat{\textrm{VPD}}_{i}^{*}\textrm{LAI}_{s,doy|i}^{*}) + \epsilon_{s,t|z,h}
+\hat{G}_{s,t|z} = \beta_{0,s|z,h} + \beta_{1,s|z,h}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} \textrm{FLC}_{\textrm{leaf},s,i|z,h}^{*}\hat{\textrm{VPD}}_{i}^{*}\textrm{LAI}_{s,doy|i}^{*}) + \epsilon_{s,t|z,h}
 (\#eq:M6)
@@
 
-## ERD estimation
+\newpage
+**Methods S2** Statistics for identifying best-fit ERD.
 
 For each ERD model structure (Eqs.
-\@ref(eq:M1)-\@ref(eq:M6)), species-specific best-fit ERD was identified as follows: For example, for the chosen model (Eq.
-\@ref(eq:M3)), for a species $s$ and hydrological realization *h*, we obtained Akaike's Information Criterion (AIC) and Goodness-of-fit ($R$^2^) for each modeled soil layer depth $z$.
+S\@ref(eq:M1)-S\@ref(eq:M6)), species-specific best-fit ERD was identified as follows: for example, for the chosen model (Eq.
+S\@ref(eq:M3)), for a species $s$ and hydrological realization *h* for soil water potential $\mathrm{\Psi}_{\textrm{soil,} z}$ dynamics, we obtained Akaike's Information Criterion (AIC) and Goodness-of-fit ($R$^2^) for each modeled soil layer depth $z$.
 We used AIC as the first selection criterion to ensure that the fitted growth values from Eq.
-\@ref(eq:M3) are within the confidence interval of the error distribution of observed growth (as the same model Eq.
-\@ref(eq:M3) was being compared for each *h*, number of model parameters were the same).
+S\@ref(eq:M3) are within the confidence interval of the error distribution of observed growth (as the same model Eq.
+S\@ref(eq:M3) was being compared for each *h*, number of model parameters were the same).
 For each *h*, we retained those depths with the maximum likelihood of growth (similar AIC based support, see below) filtering out the rest.
@@
 
-where r is the correlation, and $R^2$ the Goodness-of-fit, between observed growth $G$ and estimated growth $\hat{G}.
+where r is the correlation, and $R^2$ the Goodness-of-fit, between observed growth $G$ and estimated growth $\hat{G}$.
 $\mathcal{L}$ is the likelihood function, $\hat{\beta}$ is the maximum-likelihood estimate, $G$ the observed growth with $k$ observations (or, census intervals $t$).
@@
 
-## ERD model structure selection
+\
+**Methods S3** Processing of forest census data for growth estimates.
 
-The best ERD model structure explained a large fraction of the variance in $\mathrm{\delta}$^2^H~xylem~.
-This model included an effect of VPD and not LAI (Eq.
-\@ref(eq:M3)).
-Two of the alternative models (Eqs.
-\@ref(eq:M4) & \@ref(eq:M6)) with LAI had a strong relationship with $\mathrm{\delta}$^2^H~xylem~ as well, but one that was based on a smaller set of species ($n$ = 4 or 5), because fewer species passed the criteria of sufficient species-level goodness-of-fit for these models (<17 species vs.
-`r length(erd.sp)` species for the best model, Fig.
-S9).
-For the rest of the analyses we therefore used ERD estimates from the less parsimonious model (Eq.
-\@ref(eq:M3)).
-
-## Additional results on exposure to water-stress
-
-Overall, species with ERD deeper than 2.9 m did not cross species $\mathrm{\Psi}$~crit~ for the entire 25-yr period for which $\mathrm{\Psi}_{\textrm{soil},z}$ was modeled (1990-2015).
-Soil depths deeper than 2.9 m remained above `r round(psi.2.9.min$median, 2)` MPa (`r round(as.numeric(psi.2.9.min$q97.5), 2)`, `r round(as.numeric(psi.2.9.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and drought periods, which is above $\mathrm{\Psi}$~crit~ of the most sensitive tree species (`r min.psi_crit` MPa).
-
-## Data Descriptions 
-### Growth Data
-
+We used tree diameter recensus data from the Barro Colorado 50-ha plot [@Condit.2019a].
 We selected trees for which the height of the diameter measurements were unchanged across all censuses (1990, 1995, 2000, 2005, 2010, 2015).
@@
 
-### Microclimatic and flux tower {#SIData.matteo}
+\
+**Methods S4** Leaf Area Index calculations.
 
+We generated an estimate of the mean seasonality curve for normalized leaf area index ($\textrm{LAI}^*$, range 0-1, unitless) for individual species as follows.
+We interpolated species-specific weekly leaf-fall data collected at two similar sites in BCI (1985-2020) to a daily estimate of leaf mass fall per unit ground area [@Wright.1990aeo].
+We estimated living canopy leaf mass per unit ground area present at time t, L[t], from mean leaf longevity in days, LL, and leaf mass fall per unit ground area at time t, F[t], as follows:  L[t] = F[t+1] + F[t+2] + … + F[t + LL].
+L[t] was averaged by DOY across years and divided by the max value to obtain the species-specific seasonality curve in $\textrm{LAI}^*$ (range 0-1, unitless).
+(Fig.
+S\@ref(fig:leaf-cover)).
+
+For `r sp.LAI.obs` species, an estimate of leaf lifespan was available from direct observations at a dry and wet site in Panama [@Osnas.2018].
+For the remaining species leaf lifespan was gap-filled with the following preference order depending on data availability: genus-level mean (n = `r sp.LAI.gen`), family-level mean (n = `r sp.LAI.fam`), site-level mean (n = `r sp.LAI.pnm.site` deciduous species using data from the site Parque Natural Metropolitano situated on the drier side of the rainfall gradient along the Isthmus of Panama).
+For one species, leaf lifespan was predicted from leaf mass area (LMA) ($\overline{lifespan} = `r ll.lma.cf[1]` + `r ll.lma.cf[2]`LMA$; $R$^2^ = `r ll.lma.r2`; $p$ `r ll.lma.p`, $n$ = `r ll.lma.n`).
+
+Although deciduous leaf-habits are generally expected to be fully deciduous, at BCI there is a large within-species variation in the timing of leaf-drop and leaf-out and not all individuals of a deciduous species reach full deciduousness each year [@Condit.2000].
+Our reconstruction of $\textrm{LAI}^*$ seasonality is based upon species leaf-fall data, which captures the within-species variation in the timing of leaf-drop.
+So, in our reconstruction the seasonal decline in $\textrm{LAI}^*$ appears to be more spread out and does not reach zero than that is one expected for an idealized individual representing each of the deciduous leaf habits.
+Greater uncertainty in leaf-lifespan for deciduous species may have compounded this issue further.
+We did not normalize the deciduous species' $\textrm{LAI}^*$ curves to reach zero as it may further distort the shape of the $\textrm{LAI}^*$ curve shifting a larger period under a lower $\textrm{LAI}^*$.
+
+\
+**Methods S5** Details of the ELM-FATES model.
+
+In ELM, soil water is predicted from a multi-layer model, in which the vertical soil moisture transport is governed by infiltration, surface and sub-surface runoff, gradient diffusion, gravity, canopy transpiration through root extraction, and interactions with groundwater (see @Oleson.2013 for process details and equations).
+FATES simulates size-structured tree community complete with individual tree rooting profiles.
+In our use of reduced-complexity configuration of ELM-FATES with a single plant functional type, ELM removes total community canopy transpiration as water-uptake from the soil layers in proportion to the total root fraction in each layer in the 1-D column.
+Note that we use the soil water potential dynamics obtained from ELM-FATES as an ecosystem-level outcome to drive our species-level, empirical ERD model.
+We thus believe that there is no conflict/circularity between tree-level rooting profiles or water uptake processes in ELM-FATES and the ERD model.
+
+We ran ELM with FATES vegetation, in which the ELM model simulates interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\mathrm{\Delta}W_{\textrm{can}}$, surface water $\mathrm{\Delta}W_{\textrm{sfc}}$, soil water $\mathrm{\Delta}W_{\textrm{liq},z_{i}}$ at depth $z_{i}$ and water in an unconfined aquifer $\mathrm{\Delta} W_{a}$ (omitting processes relevant to snow, wetlands or lakes).
+Conservation of these terms is expressed as follows:
+
+\begin{equation}
+\begin{split}
+\mathrm{\Delta}W_\textrm{can} + \mathrm{\Delta}W_\textrm{sfc} + \sum_{i=1}^{15} \mathrm{\Delta}W_{\textrm{liq},z_{i}} + \mathrm{\Delta}W_{a} = \left(q_\textrm{rain} - E_{v} - E_{g} - q_\textrm{over} - q_\textrm{h2osfc} - q_\textrm{drai}\right)\mathrm{\Delta}t
+\end{split}
+(\#eq:water-balance)
+\end{equation}
+where, $q_{\textrm{rain}}$ is rainfall, $E_{v}$ is transpiration of the forest, $E_{g}$ is ground evaporation, $q_{\textrm{over}}$ is surface runoff, $q_{\textrm{h2osfc}}$ is runoff from surface water storage, $q$~drai~ is subsurface drainage (all in mm H~2~O) and $\mathrm{\Delta}t$ is the time-step (here, 1 hr).
+
+Justification for the use of static stand structure: in static stand structure model, the model is started from an observed forest inventory and recruitment, growth and mortality are deactivated.
+Holding the stand structure static rather than allowing it to emerge from physiologically-mediated competition between trees removes a set of internal model feedbacks and thus the only differences between model simulations are driven by direct physiological effects of perturbed model parameters.
+This allows us to focus on the hydraulic and hydrological questions explored in this paper.
+
+**Methods S6** Details for ELM-FATES model parameterization.
+
+In the Conrad Trail Stream catchment, in which the 50 ha plot is situated, soil hydraulic conductivity decreases with depth from 12.5 and 60 cm (@Godsey:2004; Table \@ref(tab:param)).
+We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths <12.5 cm and >60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig.
+S\@ref(fig:ksat-profiles)).
+
+For greater accuracy, instead of the ELM default pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content (GWC) vs.
+$\mathrm{\Psi}$~soil~ for depths of 0.15, 0.4 and 1 m [@Kupers:2019].
+We converted GWC to volumetric (VWC, $\theta$) using a measured bulk density value of 0.8 g cm^-3^ (see below).
+The @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated that differences in water retention among depths were nominal,
+
+\begin{equation}
+\mathrm{\Psi}_{\textrm{soil}} = \mathrm{\Psi}_e\left(\frac{\mathrm{\theta}}{\mathrm{\theta}_{s}}\right)^{-b}
+(\#eq:soil-ret-curve)
+\end{equation}
+where $\mathrm{\Psi}_e$ is the air-entry matric potential, $\mathrm{\theta}_{s}$ is the saturated volumetric water content, and $b$, an index for soil pore-size distribution, respectively.
+$\mathrm{\Psi}_e$, $\mathrm{\theta}_{s}$ and $b$ that best-fit the data were 200 (mm H~2~O), 0.51 (cm^3^ cm^-3^) and 10 (unitless), respectively.
+
+\
+**Methods S7** ELM-FATES calibration.
+
+We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq.
+\@ref(eq:water-balance), namely, (1) evapotranspiration ET from the flux tower by the 50 ha plot ($\backsimeq$ $E_{v} + E_{g}$ in Eq.
+\@ref(eq:water-balance); 2012-2017; Dataset S2), (2) local stream discharge $q_\textrm{run}$ ($\backsimeq$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see @Paton.2019run and Dataset S5) and (3) soil volumetric water content (VWC) $\mathrm{\theta}_{z}$ ($\backsimeq$ $W_{\textrm{liq,} z}$) from two sources: (i) a long-term (2012-2018) record of VWC $\mathrm{\theta}_\textrm{0-0.15 m}$ averaged across three vertical time domain reflectometry (TDR) probes over depth 0-15 cm from three locations near the flux tower, and (ii) plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 at depths of 0.15, 0.4 and 1 m (1299 samples covering all soil types and habitats; see Dataset S6 and @Kupers:2019 for details; hereafter, $\mathrm{\theta}_{\textrm{0-0.15 m}}$, $\mathrm{\theta}_{\textrm{0.4 m}}$ and $\mathrm{\theta}_{\textrm{1 m}}$ refer to plot-wide averages).
+
+For ELM-FATES calibration we calculated an objective function for each of the 5000-member ensemble by equally weighting standardized RMSE between observations and simulations across all fluxes and states mentioned above.
+We describe the objective function $O$ as:
+
+\begin{equation}
+\begin{split}
+\overline{O} = \frac{1}{6}\sum \textrm{RMSE}^{*}\textrm{(ET,}E_{v} + E_{g}) + \textrm{RMSE}^{*}(q_{\textrm{run}},q_{\textrm{over}} + q_{\textrm{drai}}) + \textrm{RMSE}^{*}(\mathrm{\theta}_{\textrm{0-0.15~m}}, {W_{\textrm{liq,}\overline{z}}}) \\
+& + \textrm{RMSE}^{*}(\mathrm{\theta}_{\textrm{0.15~m}}, W_{\textrm{liq,`r soil.depths[4]`m}}) + \textrm{RMSE}^{*}(\mathrm{\theta}_{\textrm{0.4~m}}, W_{\textrm{liq,`r soil.depths[6]`~m}}) + \textrm{RMSE}^{*}(\mathrm{\theta}_{1~m}, W_{\textrm{liq,`r soil.depths[8]`~m}})
+(\#eq:obj-func)
+\end{split}
+\end{equation}
+
+based on monthly sums for both ET and $q$~run~ and daily averages for $\mathrm{\theta}$s.
+$W_{\textrm{liq,}\overline{z}}$ is the average of daily $W_{\textrm{liq,} z}$, where $z$ $\in$ $Z$; $Z$ = (`r paste(soil.depths[1:4], collapse=", ")`) m.
+Superscript * indicates that RMSE is standardized between 0 and 1 across the 5000 simulations.
+
+Model fit to soil moisture dynamics by depth was one of our key priority.
+Preliminary analyses showed that fitting the model to all states and fluxes simultaneously captured soil moisture dynamics for shallower depths but not at 1 m depth.
+To ensure model fit at 1 m depth, we thus imposed a two-step procedure, wherein we first chose the 300 parameter ensemble members that minimized $\textrm{RMSE}^{*}({\mathrm{\theta}_{\textrm{1 m}}}, W_{\textrm{liq,`r soil.depths[8]` m}})$ (< 0.2), and among those 300 ensemble members we chose 100 with the lowest values for the objective function $\overline{O}$.
+
+\
+**Dataset S1** Microclimatic and flux tower data.
+\
 The tower used for these measurements is 41 m above ground, on a plateau on BCI at the north west corner of the 50ha plot [@Pau.2018].
@@
 Data were processed with a custom program using a standard routine described in [@Detto.2010].
-QA/QC criteria for removing erroneous values are listed in table 1 (see below).
+QA/QC criteria for removing erroneous values are listed in Table S2.
 GPP was derived from daytime values of NEE by adding the corresponding mean daily ecosystem respiration obtained as the intercept of the light response curve [@Lasslop.2010].
@@
 $f_{i} = \frac{\sum_{k}f_{ik}^{ANN}/{MSE_{ik}}}{\sum_{k}1/{MSE_{ik}}}$
-\
+
 The ANN was implemented using the Neural Network Toolbox in Matlab 2014a.
 
-\
 Data can be requested from https://ameriflux.lbl.gov.
 
-#
-```{r eddy, results = 'asis', echo = FALSE}
-knitr::kable(eddy.table, caption = 'QA/QC procedure applied to eddy covariance fluxes.
-w: vertical wind speed; t: air temperature; q: water vapor density; c: CO~2~ density; u: horizontal wind speed.
-The bar indicates 30-min averaging and the carat indicates fluctuation ($\\overline{\\hat{x} = x - \\bar{x}}$)', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(eddy.table))))
-```
+\
+**Dataset S2** Leaf hydraulic conductivity and vulnerability to cavitation.
+\
+$K_\textrm{leaf}$ was measured under steady state and high irradiance following the Evaporative Flux Method (FEM) [@Sack.2012] using a flowmeter (J.
+Zailaa et al., unpublished).
+Briefly, the leaf was excised under filtered deionised water and attached to a water source (graduated cylinder) using clear plastic tubing.
+The system includes tubing of a known resistance, separated upstream and downstream by two sensors measuring the transpiration-driven water flow pressure in the EFM system [@Sack.2011].
+$K_\textrm{leaf}$ is calculated as the ratio of the transpiration rate (*E*~leaf~; mmol s^-1^) to the difference in flow pressure (MPa) upstream and downstream from the resistance tubing, and further standardized to 25°C and by leaf area.
+At stability (CV < 5%), the leaf is left to equilibrate in plastic bags with moist paper towels and measured for $\mathrm{\Psi}$~leaf~ using a Scholander pressure chamber.
+To construct vulnerability curves for each species, four different functions were fitted to species-specific $K_\textrm{leaf}$ vs.
+$\mathrm{\Psi}$~leaf~ data using the *anneal* optimization function in the R package likelihood: exponential without an intercept ($K_{leaf} = Ae^{-B\Psi_{leaf}}$), exponential with an intercept ($K_{leaf} = C + Ae^{-B\Psi_{leaf}}$), logistic ($K_{leaf} = \frac{A}{1 + \left(\frac{\Psi_{leaf}}{x_{0}}\right)^B}$) and sigmoidal ($K_{leaf} = \frac{A}{1 + e^{-\left(\frac{\Psi_{leaf} - x_{0}}{B}\right)}}$) [@Sack.2012].
+The differences among AICc values for these functions were not large within species.
+We chose the exponential function without the intercept to fit observed data across species, and as a general function to scale-up to the community, because it had the lowest AICc for most species, captured the form of the observed data across all species well, and had the lowest number of parameters, that is, two.
 
-### Stream discharge from the Conrad catchment {#SIData.steveP}
+\
+**Dataset S3** Wood specific gravity (WSG) and leaf mass area (LMA).
+\
+We used WSG, the wood specific gravity after drying at 100^$\circ$^C, and LMA, the leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules for leaves receiving direct sunlight as described in @Wright:2010.
 
+Among the `r length(sp.exp.param$sp)` species with observed data for ($K_\textrm{leaf}$ vs.
+$\mathrm{\Psi}$~leaf~) used to build the community-level models of leaf vulnerability curves, two lacked data for WSG and five in LMA.
+We filled three gaps in LMA by substituting species-level means for LMA~disc~, the latter defined as the mean leaf mass per unit area measured for a 1.483 cm^2^ leaf disc taken to avoid veins (g m^-2^) for leaves receiving direct sunlight, in a linear relationship between LMA and LMA~disc~ ($\overline{\textrm{LMA}} = `r disc.cf[1]` + `r disc.cf[2]`\textrm{LMA}_{disc}$; Adj.
+$R$^2^ = `r disc.r2`; $p$ `r disc.p`, $n$ = `r disc.n`).
+We filled the two remaining gaps in LMA by substituting LMA measured during the $K_\textrm{leaf}$ campaign, LMA~campaign~, in a linear relationship between LMA and ($\overline{\textrm{LMA}} = `r lma.cf[1]` + `r lma.cf[2]`\textrm{LMA}_{campaign}$; Adj.
+$R$^2^ = `r lma.r2`; $p$ `r lma.p`, $n$ = `r lma.n`).
+We chose not to fill the two data gaps in WSG as available data were poor predictors of WSG.
+
+\
+**Dataset S4** Leaf deciduousness categories.
+\
+At Barro Colorado Island SJW and Osvaldo Calderón have drawn on 62+ years of experience in Panama to score each species as one among four leaf habits [@Meakem.2018]: evergreen species ($n$ = `r nrow(subset(erd.data, deciduous != "E"))`, for species with ERD) keep most but not all of their canopy in the dry season; obligate deciduous species ($n$ = `r nrow(subset(erd.data, deciduous != "DO"))`) shed all their leaves in the beginning of the dry season; facultative deciduous ($n$ = `r nrow(subset(erd.data, deciduous != "DF"))`) species drop leaves steadily as the dry season develops, and both facultative and obligate deciduous species leaf out in the beginning of the wet season; brevi-deciduous species ($n$ = `r nrow(subset(erd.data, deciduous != "DB"))`) shed all their leaves in the dry season, but leaf out in a few days to six weeks under ongoing dry season.
+
+\
+**Dataset S5** Stream discharge from the Conrad catchment.
+\
 The Conrad weir is located on the gently sloping western side of Barro Colorado Island.
@@
 Large storms are sometimes preceded by small, initial peaks generated by the rain falling on the steeper slopes close to the weir.
-Streamflow data are analyzed by software custom-built by author SP.
+Streamflow data are analyzed by software custom-built by author SRP.
 The program is used to correct for blockages of the weir ‘V’ by fallen vegetation (a common occurrence), sensor drift, small gaps and other problems.
@@
 
-### Volumetric water content
-
+\
+**Dataset S6** Volumetric water content.
+\
 We collected volumetric water content (VWC) data (2012-2018) from three locations near the eddy flux tower with time domain reflectometry (TDR, CS616, Campbell Scientific) probes inserted vertically ranging over the first 15 cm of soil.
@@
 
-\
 We also leveraged existing GWC measurements by @Kupers:2019 at the depths of 0.15, 0.4 and 1 m in summer 2015 (February, March and April) and 0.15, 0.4 in summer 2016 (March), with 1299 samples in total that covered all soil types and habitats in the 50ha plot.
@@
 
-```{r TDR-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-TDR-calib}Volumetric Water Content vs.
-TDR data for a vertical TDR probe (0-15 cm depth) at the base of the eddy flux tower near the 50 ha plot at BCI."}
-knitr::include_graphics(file.path("figures/Matteos_TDR_SWC_data_linear_model.jpeg"))
-```
-
-### Leaf habit
-
-AT BCI expert botanists have identified four leaf habits: evergreen species ($n$ = `r nrow(subset(erd.data, deciduous != "E"))`, for species with ERD) keep most but not all of their canopy in the dry season; obligate deciduous species ($n$ = `r nrow(subset(erd.data, deciduous != "DO"))`) shed all their leaves in the beginning of the dry season; facultative deciduous ($n$ = `r nrow(subset(erd.data, deciduous != "DF"))`) species drop leaves steadily as the dry season develops, and both facultative and obligate deciduous species leaf out in the beginning of the wet season; brevi-deciduous species ($n$ = `r nrow(subset(erd.data, deciduous != "DB"))`) shed all their leaves in the dry season, but leaf out in a few days to six weeks under ongoing dry season.
-
-### Leaf Area Index 
 \
-We generated an estimate of the mean seasonality curve for normalized leaf area index ($\textrm{LAI}^*$, range 0-1, unitless) for individual species as follows.
-We interpolated species-specific weekly leaf-fall data collected at two similar sites in BCI (1985-2020) to a daily estimate of leaf mass fall per unit ground area [@Wright.1990aeo].
-We estimated living canopy leaf mass per unit ground area present at time t (L[t]) from mean leaf longevity in days (LL) and leaf mass fall per unit ground area at time t (F[t]) as follows:  L[t] = F[t+1] + F[t+2] + … + F[t + LL].
-L[t] was averaged by DOY across years and divided by the max value to obtain the species-specific seasonality curve in $\textrm{LAI}^*$ (range 0-1, unitless).
-(Fig.
-S\@ref(fig:leaf-cover)).
+**Dataset S7** Stem maximum hydraulic conductivity and vulnerability to cavitation.
 \
-For `r sp.LAI.obs` species, an estimate of leaf lifespan was available from direct observations at a dry and wet site in Panama [@Osnas.2018].
-For the rest it was gap-filled with genus-level (n = `r sp.LAI.gen`) or family-level means (`r sp.LAI.fam`), if available.
-For `r sp.LAI.pnm.site` deciduous species, dry site-level mean was used.
-For one species, leaf lifespan was predicted from LMA ($\overline{lifespan} = `r ll.lma.cf[1]` + `r ll.lma.cf[2]`LMA$; $R$^2^ = `r ll.lma.r2`; $p$ `r ll.lma.p`, $n$ = `r ll.lma.n`).
-
-Deciduous leaf-habits are generally expected to be fully deciduous at individual tree-level.
-At BCI, however, there is a great within-species variation in the timing of leaf-drop and leaf-out, and neither do all individuals of a deciduous species reach full deciduousness every year (Condit et al., 2000).
-Our reconstruction of LAI seasonality is based upon species-level leaf-fall data, which captures the within-species variation in the timing of leaf-drop, so in our reconstruction seasonal decline in $\textrm{LAI}^*$ appears to be more spread out and does not reach zero, in contrast to what may be  expected for an idealised individual of each deciduous leaf habit.
-Greater uncertainty in our estimates of leaf-lifespan for deciduous species may have compounded this issue further.
-We chose not to normalise the deciduous species’ $\textrm{LAI}^*$ curves to reach zero as it would further distort the shape of the curve, shifting a larger period under a lower $\textrm{LAI}^*$.
-
-Although deciduous leaf-habits are generally expected to be fully deciduous, at BCI there is a great within-species variation in the timing of leaf-drop and leaf-out, and neither do all individuals of a deciduous species reach full deciduousness every year [@Condit.2000].
-Our reconstruction of $\textrm{LAI}^*$ seasonality is based upon species leaf-fall data, which captures the within-species variation in the timing of leaf-drop, so in our reconstruction the seasonal decline in $\textrm{LAI}^*$ appears to be more spread out and does not reach zero, in contrast to what is ideally expected for an individual of each deciduous leaf habit.
-Greater uncertainty in leaf-lifespan for deciduous species may have compounded this issue further.
-We did not normalise the deciduous species' $\textrm{LAI}^*$ curves to reach zero as it may further distort the shape of the $\textrm{LAI}^*$ curve shifting a larger period under a lower $\textrm{LAI}^*$.
-
-```{r leaf-cover, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-cover}Seasonal variation in standardised LAI (range 0-1) for ", length(erd.sp)," species at Barro Colorado Island (panels).
-Species are color-coded by leaf habit as defined by expert botanists.")}
-knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf.cover_BCI_multi_panel.jpeg"))
-```
-
-### Leaf hydraulic conductivity and vulnerability to cavitation {#SIData.joseph}
-\
-$K_\textrm{leaf}$ was measured under steady state and high irradiance following the Evaporative Flux Method (FEM) [@Sack.2012] using a flowmeter.
-Briefly, the leaf was excised under filtered deionised water and attached to a water source (graduated cylinder) using clear plastic tubing.
-The system includes tubing of a known resistance, separated upstream and downstream by two sensors measuring the transpiration-driven water flow pressure in the EFM system [@Sack.2011].
-$K_\textrm{leaf}$ is calculated as the ratio of the transpiration rate (*E*~leaf~; mmol s^-1^) to the difference in flow pressure (MPa) upstream and downstream from the resistance tubing, and further standardized to 25°C and by leaf area.
-At stability (CV < 5%), the leaf is left to equilibrate in plastic bags with moist paper towels and measured for $\mathrm{\Psi}$~leaf~ using a Scholander pressure chamber.
-To construct vulnerability curves for each species, four different functions were fitted to species-specific $K_\textrm{leaf}$ vs.
-$\mathrm{\Psi}$~leaf~ data using the *anneal* optimization function in the R package likelihood: exponential without an intercept ($K_{leaf} = Ae^{-B\Psi_{leaf}}$), exponential with an intercept ($K_{leaf} = C + Ae^{-B\Psi_{leaf}}$), logistic ($K_{leaf} = \frac{A}{1 + \left(\frac{\Psi_{leaf}}{x_{0}}\right)^B}$) and sigmoidal ($K_{leaf} = \frac{A}{1 + e^{-\left(\frac{\Psi_{leaf} - x_{0}}{B}\right)}}$) [@Sack.2012].
-The differences among AICc values for these functions were not large within species (see Zailaa et al.
-for more detailed methods description.
-To be submitted.).
-We chose the exponential function without the intercept to fit observed data across species, and as a general function to scale-up to the community, because it had the lowest AICc for most species, captured the form of the observed data across all species well, and had the lowest number of parameters, that is, two.
-
-### Stem maximum hydraulic conductivity and vulnerability to cavitation {#SIData.brett}
-
 Values of stem maximum hydraulic conductivity and vulnerability to cavitation were obtained from the dataset of @Wolfe.2021.
@@
 
-\
 Methods for stem maximum hydraulic conductivity and vulnerability to cavitation closely followed @Christman.2012 and @Wolfe.2016.
@@
 
-\
 To assess vulnerability to cavitation, the branches were allowed to dry in the lab for 0 – 180 hours before they were measured for stem water potential and $K_\textrm{stem}$.
@@
 
-### Leaf turgor loss point
-
+\
+**Dataset S8** Leaf turgor loss point.
+\
 For each tree species measured for $K_\textrm{stem}$ and $\mathrm{\Psi}$~88,stem~, two to six leaves were measured for $\mathrm{\Psi}$~tlp~ with pressure-volume analysis following @Koide.1989.
@@
 
-### Above-ground hydraulic safety margins
-
+\
+**Dataset S9** Above-ground hydraulic safety margins.
+\
 Above-ground hydraulic safety margin was calculated as the difference between minimum leaf water potential $\mathrm{\Psi}$~min~ and $\mathrm{\Psi}$~88,stem~.
@@
 
-### WSG and LMA {#SIData.rutuja}
-
-We used WSG, the wood specific gravity after drying at 100^$\circ$^C and LMA, the leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules for leaves receiving direct sunlight as described in @Wright:2010.
+\
+**Notes S1** ERD model structure selection.
+\
+The best ERD model structure explained a large fraction of the variance in $\mathrm{\delta}$^2^H~xylem~.
+This model included an effect of VPD and not LAI (Eq.
+S\@ref(eq:M3)).
+Two of the alternative models (Eqs.
+S\@ref(eq:M4) & S\@ref(eq:M6)) with LAI had a strong relationship with $\mathrm{\delta}$^2^H~xylem~ as well, but one that was based on a smaller set of species ($n$ = 4 or 5), because fewer species passed the criteria of sufficient species-level goodness-of-fit for these models (<17 species vs.
+`r length(erd.sp)` species for the best model, Fig.
+S\@ref(fig:other-erd-mods)).
+For the rest of the analyses we therefore used ERD estimates from the less parsimonious model (Eq.
+S\@ref(eq:M3)).
 
 \
-Among the `r length(sp.exp.param$sp)` species with observed data on ($K_\textrm{leaf}$ vs.
-$\mathrm{\Psi}$~leaf~) used to build the community-level models of leaf vulnerability curves, two lacked data on WSG and five on LMA~lamina~.
-We filled three gaps in LMA using a linear relationship between LMA and LMA~disc~, the latter defined as the mean leaf mass per unit area measured for a 1.483 cm^2^ leaf disc taken to avoid veins (g m^-2^) for leaves receiving direct sunlight ($\overline{\textrm{LMA}} = `r disc.cf[1]` + `r disc.cf[2]`\textrm{LMA}_{disc}$; Adj.
-$R$^2^ = `r disc.r2`; $p$ `r disc.p`, $n$ = `r disc.n`).
-We filled the two remaining gaps in LMA using a linear relationship between LMA and LMA~lamina~, as measured during the smaller, $K_\textrm{leaf}$ campaign ($\overline{\textrm{LMA}} = `r lma.cf[1]` + `r lma.cf[2]`\textrm{LMA}_{lamina}$; Adj.
-$R$^2^ = `r lma.r2`; $p$ `r lma.p`, $n$ = `r lma.n`).
-We chose not to fill the two data gaps in WSG as available data were poor predictors of WSG.
+**Notes S2** Additional results for exposure to water-stress.
+\
+Overall, species with ERD deeper than 2.9 m did not cross species $\mathrm{\Psi}$~crit~ for the entire 25-yr period for which $\mathrm{\Psi}_{\textrm{soil},z}$ was modeled (1990-2015).
+Soil depths deeper than 2.9 m remained above `r round(psi.2.9.min$median, 2)` MPa (`r round(as.numeric(psi.2.9.min$q97.5), 2)`, `r round(as.numeric(psi.2.9.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and drought periods, which is above $\mathrm{\Psi}$~crit~ of the most sensitive tree species (`r min.psi_crit` MPa).
 

### 2021-05-07 09.0_manuscript.rmd
rephrased a sentence with evergreen species

@@
 We validated ERD estimates against existing isotopic data of potential water-access depths.
-- Across species, deeper ERD was associated with higher maximum stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates during extreme droughts over 35 years (1981-2015), especially in evergreen species.
+- Across species, deeper ERD was associated with higher maximum stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates, especially in evergreen species, during extreme droughts over 35 years (1981-2015).
 Species exposure to water-stress declined with deeper ERD indicating that trees compensate for water-stress related mortality risk through deep-water access.

### 2021-07-02 Readme.Rmd
Added copyright info and associated publicaiton

@@
 ---
-title: "Workflow Part II: Inverse root modeling and post-processing"
+title: "Workflow Part II: Inverse root depth modeling and post-processing"
 author: "Rutuja Chitra-Tarak"
@@
   word_document: default
+  md_document:
 subtitle: Hydraulically-vulnerable trees survive on deep-water access during droughts
@@
 ```
+
+Scripts in this repo are a part of the workflow associated with the following manuscript:
+
+Chitra-Tarak, R, C Xu, S Aguilar, K Anderson-Teixeira, J Chambers, M Detto, B Faybishenko, RA Fisher, R Knox, C Koven, L Kueppers, N Kunert, SJ Kupers, NG McDowell, BD Newman, SR Paton, R Pérez, L Ruiz, L Sack, JM Warren, BT Wolfe, C Wright, SJ Wright, J Zailaa, SM McMahon (2021) Hydraulically vulnerable trees survive on deep-water access during droughts in a tropical forest.
+New Phytologist.
+https: //doi.org/10.1111/nph.17464
+
+For the full work-flow and output datasets associated with the above manuscript see the following archived dataset:
+
+Chitra-Tarak R, Xu C, Aguilar S, Anderson-Teixeira K, Chambers J, Detto M, Faybishenko B, Fisher R, Knox R, Koven C et al.
+2020.
+Soil water potentials (1990–2018) from a calibrated ELM-FATES, and rooting depth analyses scripts, PA-BCI, Panama.
+2.0.
+NGEE Tropics Data Collection.
+doi: 10.15486/ngt/1696806
+
+# Workflow Part II: Inverse root modeling and post-processing
 
@@
 ```
-<!-- groundhog.library('diffr', groundhog.day) -->
-<!-- diffr("09.0_manuscript_NP1.rmd", "09.0_manuscript.rmd") -->
+
+## Copyrights
+
+© 2021. Triad National Security, LLC. All rights reserved.
+
+This program was produced under U.S. Government contract 89233218CNA000001 for Los Alamos
+National Laboratory (LANL), which is operated by Triad National Security, LLC for the U.S.
+Department of Energy/National Nuclear Security Administration. All rights in the program are
+reserved by Triad National Security, LLC, and the U.S. Department of Energy/National Nuclear
+Security Administration.
+The Government is granted for itself and others acting on its behalf a nonexclusive, paid-up, irrevocable worldwide license in this material to reproduce, prepare derivative works, distribute copies to the public, perform publicly and display publicly, and to permit others to do so.
+
+## Open Source Redistribution License
+
+This program is open source under the BSD-3 License.
+
+Redistribution and use in source and binary forms, with or without modification, are permitted provided that the following conditions are met:
+
+    1. Redistributions of source code must retain the above copyright notice, this list of conditions and the following disclaimer.
+
+    2.
+Redistributions in binary form must reproduce the above copyright notice, this list of conditions and the following disclaimer in the documentation and/or other materials provided with the distribution.
+
+    3. Neither the name of the copyright holder nor the names of its contributors may be used to endorse or promote products derived from this software without specific prior written permission.
+
+THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS "AS
+IS" AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE
+IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR
+PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT HOLDER OR
+CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL,
+EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT LIMITED TO,
+PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE, DATA, OR PROFITS;
+OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY THEORY OF LIABILITY,
+WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT (INCLUDING NEGLIGENCE OR
+OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF THIS SOFTWARE, EVEN IF
+ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.

### 2021-07-02 Readme.Rmd
Added LANL C #

@@
 output:
-  pdf_document: default
+  md_document:
   html_document:
@@
   word_document: default
-  md_document:
+  pdf_document: default
 subtitle: Hydraulically-vulnerable trees survive on deep-water access during droughts
@@
 
-© 2021. Triad National Security, LLC. All rights reserved.
+© 2021. Triad National Security, LLC. All rights reserved. LANL C20130.
 
